### 2013-01-25 Length Bias and Seach Limitations in CGP/main.tex
Adding all of the source files for papers

% This is "sig-alternate.tex" V2.0 May 2012
% This file should be compiled with V2.5 of "sig-alternate.cls" May 2012
%
% This example file demonstrates the use of the 'sig-alternate.cls'
% V2.5 LaTeX2e document class file. It is for those submitting
% articles to ACM Conference Proceedings WHO DO NOT WISH TO
% STRICTLY ADHERE TO THE SIGS (PUBS-BOARD-ENDORSED) STYLE.
% The 'sig-alternate.cls' file will produce a similar-looking,
% albeit, 'tighter' paper resulting in, invariably, fewer pages.
%
% ----------------------------------------------------------------------------------------------------------------
% This .tex file (and associated .cls V2.5) produces:
%       1) The Permission Statement
%       2) The Conference (location) Info information
%       3) The Copyright Line with ACM data
%       4) NO page numbers
%
% as against the acm_proc_article-sp.cls file which
% DOES NOT produce 1) thru' 3) above.
%
% Using 'sig-alternate.cls' you have control, however, from within
% the source .tex file, over both the CopyrightYear
% (defaulted to 200X) and the ACM Copyright Data
% (defaulted to X-XXXXX-XX-X/XX/XX).
% e.g.
% \CopyrightYear{2007} will cause 2007 to appear in the copyright line.
% \crdata{0-12345-67-8/90/12} will cause 0-12345-67-8/90/12 to appear in the copyright line.
%
% ---------------------------------------------------------------------------------------------------------------
% This .tex source is an example which *does* use
% the .bib file (from which the .bbl file % is produced).
% REMEMBER HOWEVER: After having produced the .bbl file,
% and prior to final submission, you *NEED* to 'insert'
% your .bbl file into your source .tex file so as to provide
% ONE 'self-contained' source file.
%
% ================= IF YOU HAVE QUESTIONS =======================
% Questions regarding the SIGS styles, SIGS policies and
% procedures, Conferences etc. should be sent to
% Adrienne Griscti (griscti@acm.org)
%
% Technical questions _only_ to
% Gerald Murray (murray@hq.acm.org)
% ===============================================================
%
% For tracking purposes - this is V2.0 - May 2012

\documentclass{sig-alternate}
%\linespread{2.5}
\usepackage{url}
\usepackage{verbatim}
\newcommand{\includegraphicsfit}[1]
{\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}
\begin{document}
%
% --- Author Metadata here ---
\conferenceinfo{GECCO'13,} {July 6-10, 2013, Philadelphia, Pennsylvania, USA.}
\CopyrightYear{2013}
\crdata{TBA}
\clubpenalty=10000
\widowpenalty = 10000
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

\title{Length Bias and Search Limitations in Cartesian Genetic Programming}
\subtitle{[Genetic Programming Track]}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

\numberofauthors{2} %
\begin{comment}
\author{
% 1st. author
\alignauthor
Brian W. Goldman\\
       \affaddr{BEACON Center for the Study of Evolution in Action}\\
       \affaddr{Michigan State University, U.S.A.}\\
       \email{brianwgoldman@acm.org}
% 2nd. author
\alignauthor
William Punch\\
       \affaddr{BEACON Center for the Study of Evolution in Action}\\
       \affaddr{Michigan State University, U.S.A.}\\
       \email{punch@cse.msu.edu}
}
%\end{comment}
%\begin{comment}
\author{
% 1st. author
\alignauthor
Anonymous\\
       \affaddr{Group}\\
       \affaddr{Organization}\\
       \email{email@site.com}
% 2nd. author
\alignauthor
Anonymous\\
       \affaddr{Group}\\
       \affaddr{Organization}\\
       \email{email@site.com}
}
%\end{comment}

\maketitle
\begin{abstract}
%Understanding how configuration decisions can interact with individual representation and
%evolutionary mechanisms in Cartesian Genetic Programming (CGP) to cause limitations
%on search is vital to proper algorithm application.  
In this paper we examine how Cartesian Genetic Programming's (CGP's) method
for encoding directed acyclic graphs (DAGs) and its mutation operator bias the
effective length of individuals as well as the distribution of inactive nodes
in the genome.  We investigate these biases experimentally using two CGP variants as comparisons: \emph{Reorder},
a method for shuffling node ordering without effecting individual evaluation, and
\emph{DAG}, a method for removing the concept of node position.
Experiments were performed on four problems tailored to highlight potential search
limitations, with further testing on the 3-bit multiplier problem.

Unlike previous work, our experiments show that CGP has an innate parsimony pressure
that makes it very difficult to evolve individuals with a high percentage of active nodes.  This bias is
particularly prevalent as the length of an individual increases.  Furthermore, these problems are compounded by CGP's positional
biases which can make some problems effectively unsolvable.  Both \emph{Reorder}
and \emph{DAG} appear to avoid these problems and outperform \emph{Normal} CGP
on preliminary benchmark testing. Finally, these new techniques require more
reasonable genome sizes than those suggested in current CGP, with some evidence
that solutions are also more terse.

\end{abstract}

% A category with the (minimum) three required fields
\category{Computing Methodologies}{Artificial Intelligence}{Search Methodologies}
%A category including the fourth, optional field follows...
%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

\terms{Algorithms}

\keywords{Cartesian Genetic Programming, Neutrality, Bloat}
\\
\\
\section{Introduction}
In all black box search, search designers must make assumptions about the properties
of the problem in order to outperform random search.  When these assumptions
match the application, search time can be greatly reduced.  As such, understanding how design decisions
effect search effectiveness or even prevent search from finding solutions is very important.
%
In Cartesian Genetic Programming (CGP) the maximum genome size of an individual
is set at initialization.  While this
requires the maximum genome size to be larger than the problem's solution needs,
the question remains; how will the maximum genome size effect evolution's
speed at finding solutions.  As larger genomes inherently represent larger search
spaces, intuition suggests that using the lowest possible maximum size will have the most efficient search.
However, in CGP a mounting body of evidence suggests the optimal
maximum size is much larger than the actual solutions evolved~\cite{miller:2006:redundancy}.
We examine the cause of this discrepancy by applying CGP and a few
modified versions of CGP to problems which test specific edge cases in representation.
The result is a better understanding of the search limitations of CGP.

\section{CGP}
\label{sec-cgp}
In Cartesian Genetic Programming (CGP) individuals use a fixed length string of integers to
encode directed acyclic graphs (DAGs).  
Each node in the DAG is encoded using multiple genes: one specifying a function and
the rest specifying node input connections.
Graph output is determined by additional genes which can connect to any node or
program input.
To prevent cycles,
nodes may only connect to problem inputs or nodes that precede them in the genome.
Early versions of CGP limited valid connections further through user specified graph
topologies.
While specifying topology is still useful
when evolving circuits with known hardware limitations, the less restricted form
allows CGP to represent all possible
DAGs of size \emph{N}, where \emph{N} is the number of nodes in the genome.
%
The most common evolutionary mechanism in CGP is a $1+\lambda$ population
where each generation the best offspring replaces the parent if it is no less
fit.  CGP almost exclusively uses a form of point mutation to create offspring,
with no form of crossover commonly being employed.  As such, these are the methods
considered in our analysis.
For a more detailed description of CGP, see~\cite{miller:2011:chapter2}.

%\subsection{Graph Size, Bloat, and Inactive Genes}
While CGP imposes a maximum graph size \emph{N}, the minimum
effective graph size is variable, as there can be unused portions
of the genome.  These inactive nodes exist because both output locations and
internal connections can be changed by evolution.  A trivial example is when all outputs
are connected directly to problem inputs the entire genome becomes
inactive.  Thus the active nodes are those nodes used when computing output.
%With all
%of these features combined, it should be clear that CGP is capable of evolving
%variable sized DAGs with a bounded maximum size.

A pernicious problem in the Genetic Programming (GP) community is individual bloat, in which solutions
become larger with time without substantive improvement in quality~\cite{luke:2006:bloat}.
Experimental evidence has so far suggests that CGP is able to avoid
bloat, as the evolved number of active
nodes does not grow during evolution~\cite{miller:2001:bloat} nor does it grow significantly
as the maximum number of nodes is raised~\cite{miller:2006:redundancy}.

%\subsection{Neutral Search}
The existence of inactive genes provides CGP with an explicit form of genetic
introns, allowing the evolutionary process to perform neutral search.
CGP's replacement strategy facilitates neutral space
exploration by ensuring offspring replace their parents if they are no less fit.
This allows the population to drift,
which has been shown to be beneficial~\cite{yu:2001:neutrality}.  Furthermore,
there is evidence that as much as 95\% of the genotype should be inactive
to allow for sufficient neutral search~\cite{miller:2006:redundancy}.

However, this approach to neutral search may have disadvantages.  In response to a previous claim
about neutrality's beneficial effects on needle in a haystack style
problems, Collins~\cite{collins:2006:haystacks} showed that on problems with no search
gradient, efficient random sampling can significantly outperform CGP with neutral search.
Collins argued that the benefits CGP claimed to be receiving through neutral search
may be better attributed to higher maximum genome sizes causing an increase in
the total number of active genes.

\section{Length and Positional Biasing}
\label{sec-length-biasing}
Despite CGP's ability to represent all possible DAGs
of size \emph{N}, there is little evidence that it uniformly
searches the space of DAGs. In fact,~\cite{payne:2009:bias} showed that certain types of structures
are more difficult to evolve than others.  In particular, larger structures may be more difficult to
evolve as they are less common in the search space. 
Tree based GP has similar issues,
with strong evidence that specific types of trees are more difficult to
evolve even though they are common in the search space~\cite{daida3:2003:treebias}.
We suggest that
the mechanics used to initialize and evolve connection genes in CGP
impose significant biases on the structures of the DAGs evolved, with more bias against
larger genotypes than should be expected by just representation frequency.

When initializing and mutating a connection gene, its value is randomly set to
either a preceding node in the genome or to an input.  As was discussed in~\cite{collins:2006:haystacks},
this means the expected location of a new connection is $\frac{i}{2}$ where \emph{i}
is the number of nodes and inputs that precede the mutated node.  If all nodes had
a single connection gene, we would expect random individuals to have a length of $\log_2N$.
Precise calculation of the expected length is made difficult
due to node overlap, but the relationship will remain logarithmic to the total genome size.

The probability that a node is active changes greatly depending on where
in the genome it is encoded.  Consider nodes near the end of the genome.
The highest index node expected to be active is given by Equation~\ref{eq-active-end}.
\begin{equation}
  \left ( N+I \right )\frac{O}{O + 1}
  \label{eq-active-end}
\end{equation}
Here, \emph{N} is the number of nodes in the genome, \emph{I} is the number of input locations,
and \emph{O} is the number of output locations.  This equation is derived from the fact that
if numbers are drawn uniformly at random and then ranked, the mean value for each rank
will be evenly spaced in the range.  As a result, the expected highest output location will be
the same as if \emph{O} locations were placed evenly on the range $N + I$.
On problems with a single
output (such as the commonly used parity problem) this means, on average, nodes
in the second half of the genome are inactive.  
%Problems with
%more outputs will be less effected as Equation~\ref{eq-active-end} approaches $N + I$
%as \emph{O} grows.
Some CGP implementations reduce this problem by
fixing the output locations to the last \emph{O} nodes in the genome, which has the effect
of scaling \emph{O} in Equation~\ref{eq-active-end} by the arity of the nodes in use.  However, this does not
remove the probability for large sections of inactive nodes near the end of the genome.

Conversely, nodes near the beginning of the genome are almost always active.
Equation~\ref{eq-incoming} shows how the expected number of incoming connections
increases as the number of nodes following node \emph{i} increases and the
arity (\emph{a}) used increases.  This relationship is non-linear monotonic, with input locations
and nodes near the inputs likely to have more than one incoming connection while nodes
near the output location will likely have less than one incoming connection.
\begin{equation}
  \sum_{j=i+1}^{N} \frac{a}{j}
  \label{eq-incoming}
\end{equation}
Equation~\ref{eq-incoming} is derived from the fact that all nodes following \emph{i}
in the genome have \emph{a} connections which each have a $\frac{1}{j}$ chance of
connecting to \emph{i}.

The probability that a node is active increases as the number of incoming connections
increases, as each could potentially come from an active node.  This creates a self
reinforcing cycle:  The closer a node is to the inputs the more incoming connections
it can receive from other nodes with lots of incoming connections.
As a result, nodes near the inputs are almost
always active while nodes near the outputs are almost never active.  Furthermore,
the output of nodes near the inputs is likely to be used by multiple active nodes.

Having many active nodes near the start and few near the end has important
implications on CGP's ability to perform neutral search.
If there are no inactive nodes preceding a given
node, it cannot benefit from neutral evolution.  Nodes that have a high percentage
of inactive nodes preceding them will conversely have a high probability of attaching
to inactive nodes, but will likely be inactive themselves.

\begin{figure}
  \centering
  \includegraphicsfit{ThoughtExperiment}
  \caption{Thought experiment illustrating limitations in \emph{Normal} CGP.
           Arrows represent the flow of information.} 
  \label{fig-thought}
\end{figure}

How this positional bias of active genes effects search can be examined
using a thought experiment, shown in Figure~\ref{fig-thought}.  
Each of the three Cases given are of the second best phenotype, in which \emph{Z}
takes its input from \emph{X}, while in the optimal phenotype \emph{Y} is applied
on \emph{X}'s output before input to \emph{Z}.  In Case 1 there are inactive nodes
between \emph{X} and \emph{Z}, so CGP is able to use drift to insert \emph{Y}.
Unfortunately, because active nodes cluster heavily near the start of the genome,
this is unlike to happen if even a moderate percentage of the genome is active.
In Case 2, \emph{X} and \emph{Z} are adjacent nodes, or equivalently all nodes between
\emph{X} and \emph{Z} are active.  CGP can only solve this case by recreating \emph{X}
and/or \emph{Z} somewhere else in the genome using drift, converting this back
to Case 1.  That process is highly unlikely as it involves drift of active genes.
In Case 3, \emph{Y} already exists in the genome, but in the wrong order relative to \emph{Z}.
CGP can solve this by either reinventing \emph{Y} as in Case 1 or moving \emph{Z} as in Case 2.
From a phenotypic standpoint, the first two cases should be equally difficult with the third
being easier, as \emph{Y} already exists.  However, due to CGP's positional biases, Case 2 is much harder than Case 1,
as active nodes must be reinvented before \emph{Y} can be evolved and added between \emph{X} and \emph{Z}.
Furthermore, Case 3 is just as hard as Case 1, as even though \emph{Y} already exists, \emph{Z} cannot connect to it.
Case 2 and 3 become even more difficult if \emph{X}, \emph{Y}, and \emph{Z} are thought of as
collections of nodes used to produce a single output, as recreating these complex structures
increases the difficulty.  They also become more common as interwoven structures in the
genome act the same as adjacent nodes in this experiment.

\section{Avoiding Search Limitations}
\label{sec-modifying-cgp}
In order to experimentally examine how the mechanisms of CGP bias search, we propose two CGP variants.
The first
is \emph{Reorder}, which attempts to mitigate positional bias by performing phenotypically
neutral node shuffling.  The second is \emph{DAG}, which allows nodes to connect
to any other node in the genome as long as no cycle is created.  For clarity, we will
refer to standard CGP as \emph{Normal}.  Note that both variants are capable of
representing the exact same set of graphs as \emph{Normal}.  They only differ
in how graphs can be modified by evolution.

\subsection{Reorder}
\label{sec-reorder}
\emph{Reorder} works identically to normal CGP with a single change:
Before producing offspring, the parent's nodes are shuffled without changing the phenotype of the individual.
That is, we perform neutral search on the parent's node ordering.
This process begins by building bidirectional node dependency sets, such that
each node's input and output sets can be easily accessed.  First the
set of \emph{addable} nodes is created, which initially contains all of the
inputs.  The new node ordering is then determined iteratively using the following
steps:
\begin{enumerate}
  \item Select and remove a node $\alpha$ at random from the \emph{addable} set.
  \item If $\alpha$ is not an input, map it sequentially to the next genome location.
  \item For each node $\beta$ which takes output from $\alpha$:
  \begin{enumerate}
    \item Mark $\beta$'s dependency on $\alpha$ as satisfied.
    \item If $\beta$'s dependencies have all been satisfied, add it to \emph{addable}.
  \end{enumerate}
\end{enumerate}
Iteration ends
when all nodes have been given new locations in the sequence.  At this point, nodes can be moved
into their new locations, with their connections being updated using the ordering
map.  Once reordered, nodes maintain the normal CGP requirement that all nodes
must take their inputs from nodes preceding them in the genome, meaning that
the mutation and evaluation process remains the same.  As all connections and functions are
preserved during this process, there is no change in the result obtained by executing
an individual.  \emph{Reorder} is just a neutral permutation of the node ordering.

Repeating the thought experiment in Figure~\ref{fig-thought} with
\emph{Reorder} shows how it can differ from \emph{Normal}.  Because inactive nodes
can be shuffled between \emph{X} and \emph{Z}, Case 1 and 2 become identical.
Case 3 can be
solved by \emph{Reorder} without recreating \emph{Y}, as it can
be shuffled to precede \emph{Z}.

Without breaking connections, a node can be moved forward such that
it transitively depends on all preceding nodes.  A node can also be moved back until all following nodes
transitively depend on it.  Furthermore, the distance between a node and
its input with the highest amount of transitive dependency will be on average smaller
than in \emph{Normal}.  This is illustrated
when considering graphs with a node arity of one.  In \emph{Normal} and at initialization
of \emph{Reorder}, a node is on average $\frac{i}{2}$ indexes away from where it gets its input and $2i$ indexes
from where its output goes.  In \emph{Reorder}, nodes will on average be equidistant to
their their inputs and outputs.  In this case, the expected distance is related to the total number
of nodes in the dependency chain.  This means that an increased number of active nodes can
occupy higher index locations as they are pushed to the center between their inputs and outputs.

This does not directly translate into \emph{Reorder} being able to create more active graphs,
or mean that \emph{Reorder} is without length bias.  When creating new connections,
the expected location is still $\frac{i}{2}$, which means given enough mutations,
the individual can revert back to all of the expectations for \emph{Normal}.
In general however, \emph{Reorder} will allow nodes on which
many nodes transitively depend to increase how many nodes can connect to them, while simultaneously
allowing nodes with only a few transitive dependencies to gain more dependencies.

A final artificial bias in both \emph{Reorder} and \emph{Normal} is
that all connections in a node will tend to have the same transitive
dependencies.  This is because if node $\beta$ is already transitively dependent
on node $\alpha$, then $\alpha$ is always a valid connection when generating new connections in $\beta$.
This is because all transitive dependencies to a node always precede that node, even
when using \emph{Reorder}.  Nodes that $\beta$ is not already transitively dependent
on have a probability of being unreachable, and therefore are connected to less frequently.

\subsection{DAG}
\label{sec-dag}
The concept behind \emph{DAG} is to relax the requirement that nodes must connect
forward in the genome.
To accomplish this, only two portions of CGP have to be
modified: connection gene mutation and evaluation ordering.
In \emph{Normal}, when a connection gene is mutated, its value is randomly changed
to any input or node index less than the mutating node's index.  In
\emph{DAG}, when a connection gene is chosen to be mutated, new connections are
tried randomly and kept if they do not create a cycle.

At the start of each mutation, a mapping is developed to store which nodes are
transitively dependent on the mutating node, and which are not.  Initially this map stores that
the input indexes are not dependent on the mutating node, and
that the node is dependent on itself.  Nodes are then tested in a random
order for transitive dependence.  This is performed by recursively applying
the following rules:
\begin{itemize}
  \item If the node's dependency is already known, return the stored information.
  \item If this node takes input from a dependent node, it is dependent.
  \item If none of its inputs are dependent, it is not dependent.
\end{itemize}
As information is collected, the map is updated to ensure a node is never checked
for dependency twice.  If the randomly chosen node is not dependent on the mutating
node, it is used as the new connection value. While this algorithm can result in
each mutation needing to traverse a large portion of the graph searching for
dependencies, the growth is at worst linear.  Furthermore, our reason for
introducing \emph{DAG} is to analyze limitations in \emph{Normal}, not necessarily as
a computationally efficient alternative.

To determine node ordering during evaluation, \emph{DAG}
performs a recursive walk of the active nodes starting from the output
locations, recording for each node its bidirectional dependencies on other active nodes.
Next, \emph{DAG} builds the ordering of active nodes similar to
the \emph{Reorder} method in Section~\ref{sec-reorder}, ensuring only active nodes are added
and that they are added only after all of their dependencies are in the list.
Using this ordering, which is only built once for each individual, any number of
input sequences can be evaluated with identical run time as \emph{Normal}.

Removing the forward-only requirement found in \emph{Normal} and \emph{Reorder}
allows \emph{DAG} to avoid positional bias.
This is because any node can connect to any structure
that exists in the graph as long as its not already part of that structure.  
In contrast, \emph{Reorder} can only connect to nodes currently shuffled
in front of the mutating node, while in \emph{Normal} structures may have
to be reinvented to achieve the same connection.  \emph{DAG} can also
insert inactive nodes between active nodes, as there is
no concept of adjacency.

As to individual length, \emph{DAG} has a much higher expected percentage of active nodes
than \emph{Normal} or \emph{Reorder}.  When mutating, a connection can be made to
any non-dependent node and input.  Using an arity of one as an example, the
expected number of active nodes is $\frac{N}{I+1}$ where \emph{I} is the number
of inputs.  Higher arity will increase the active proportion, but overlap makes
precise estimations difficult.  Note that unlike \emph{Normal}, the number of
active nodes in \emph{DAG} scales linearly with the total number of nodes.  Also,
since \emph{DAG} still uses \emph{Normal}'s method of initialization, the actual
individuals evolved may be less active than expected.

While \emph{DAG}'s reduced search biases do not necessarily result in better results
on specific problems, it should generally make \emph{DAG} effective on a wider
range of applications.
Also note that while evolution of complete DAGs has been done previously~\cite{niehaus:2007:ggp},
our \emph{DAG} method mirror's CGP more faithfully and allows for more direct comparison.

\section{Tailored Problems}
\label{sec-tailored}
To test the search biases of CGP and our variants,
we propose a few benchmark problems specifically
tailored to highlight certain types of bias.
%When performing
%experiments with any evolutionary system, it is necessary to specify all configuration
%information in order to improve replicability of results.
The following experimental configuration is used across all experiments reported here.
%In order to simplify
%matters, all experiments reported here use the same configuration.
Keeping with existing CGP literature, we opted for a $1+4$ strategy using
mutation as the only variation operator.  To avoid tuning the mutation
rate, we use the \emph{Single} mutation method described in~\cite{goldman:2013:cgpwaste},
which mutates genes at random until exactly one active gene is changed.  
Each run of each experiment used the same initial individual for all algorithms.
All experiments were replicated 50 times, resulting in 50 different starting individuals per experiment.
Having all algorithms start with the same initialization was done to reduce noise.
Each
run was given 10,000,000 evaluations before being terminated, and runs were stopped
immediately on success.

For all problems except \emph{Neutral} (Section~\ref{sec-neutral}), we provide a plot of the median
evaluations until success using different problem sizes.  If the median run
was unsuccessful, no information is plotted.  Note that both axes in these plots
use a log scale.  In order to ensure a good sample of scalability we used 20
evenly spaced problem sizes $\left [ 5 .. 100 \right ]$ at increments of 5.
This information was then used to calculate each algorithms' order of complexity
in regards to problem size, with results given in Table~\ref{table-growth}.
A complete package of our results as well as the source code used in testing
is available from our website.\footnote{\url{https://name.of.site.removed.com/anonymous-CGP}}

\subsection{Neutral Exploration}
\label{sec-neutral}

\begin{figure}
  \centering
  \includegraphicsfit{neutral}
  \caption{Genome sizes required to achieve different mode
           numbers of active nodes on the \emph{Neutral} problem} 
  \label{fig-neutral}
\end{figure}

\begin{figure}
  \centering
  \includegraphicsfit{length_frequencies_100}
  \caption{Frequency of evolving individuals of different active lengths
           on the \emph{Neutral} problem} 
  \label{fig-length}
\end{figure}

To examine how CGP and our variants bias the active length of individuals
based on the maximum genome size, we first experiment using the \emph{Neutral} problem.
Similar to the problems used in~\cite{miller:2001:bloat,collins:2006:haystacks},
\emph{Neutral} has a completely flat landscape, such that all individuals
evaluate to the same fitness.  In CGP this means parents are replaced each
generation with a randomly chosen offspring. Thus there is no evolutionary pressure, only drift.
This problem highlights what kinds of individuals each variant is biased toward
creating, independent of the evolutionary pressure.
In less complex representations, such as those in binary genetic
algorithms, you would expect all representable
search points to have an equal probability of being seen given enough evaluations.  As CGP's representation
involves complex decoding, \emph{Neutral} can help examine the likelihood of searching
different parts of the search space.  
In our experiments, \emph{Neutral} has a single input and a single output,
with a node arity of two.  Furthermore, the function set used is of size one, meaning
\emph{Single} mutation must change an active node's connection gene or the output location
in order to produce new offspring.

\begin{table}
	\centering
	\begin{tabular}{|r|c|c|c|}
	  \hline
& \textbf{\emph{Normal}} & \textbf{\emph{Reorder}} & \textbf{\emph{DAG}} \\ \hline
    \textbf{Mode} & 15      & 20      & 45 \\ \hline
 \textbf{95\% CI} & [5, 31] & [6, 44] & [10, 67] \\ \hline
\textbf{<99.99\%} & 45      & 61      & 81 \\ \hline
 \textbf{Longest} & 59      & 76      & 92 \\ \hline
	\end{tabular}
	\caption{Frequency of different length individuals on \emph{Neutral}. 
	         Includes the most common length (mode), 95\% confidence interval (CI),
	         length that 99.99\% of individuals were below, and the longest individual found.}
	\label{table-length}
\end{table}

We performed two types of experiments on the \emph{Neutral} problem.  The first
tests how each variant's evolved number of active nodes changes
with genome size.  The results provided in Figure~\ref{fig-neutral} can be used
to predict how large of a genome size should be used
given a user's rough estimate of the number of active nodes required by the problem.  
This also assumes that search will perform best when the most commonly searched length
is close to the optimal individual length.  Figure~\ref{fig-neutral} shows that \emph{Normal}
required the largest genome sizes to achieve each mode number of active nodes.  In
terms of scaling, \emph{Normal} needs a genome size of $O(n^{1.79})$ with
$n$ the desired mode number of active nodes.  While \emph{Reorder} only reduced
this to $O(n^{1.53})$, \emph{DAG} scaled much better at $O(n^{1.18})$.  This implies
that as true problem size increases, we should expect \emph{Normal} to require nearly
quadratically increasing genome sizes, while \emph{DAG} scales nearly linearly.

The second experiment looked at the distribution of evolved lengths using a fixed
genome size of 100. This
information is shown in Figure~\ref{fig-length} as a probability density curve,
with numeric results in Table~\ref{table-length}.
The results for \emph{Normal} match our predictions in Section~\ref{sec-reorder};
most individuals were highly inactive, with a mode number of
active nodes of only 15.  The probability of generating an individual drops quickly
as the number of active nodes deviates from the mode, with none of the 0.5 billion
generated individuals having more than 59 active nodes, and only one in 10,000 above 45.  This is in contrast
to the actual number of graphs of each size, which increases exponentially.
This means that \emph{Normal} is biased toward small graphs even more than
Figure~\ref{fig-length} indicates.
\emph{Reorder} and \emph{DAG} show less of a bias, with \emph{DAG}'s mode of 45
close to the prediction made in Section~\ref{sec-dag}.  The implications of these
biases is that each is most likely to find solutions fastest when the number of
active nodes desired by the problem best matches the algorithms search bias, as
otherwise the algorithm will add artificial pressure to search different sized
graphs.  The steepness of the bias reflects how much pressure will be applied.
Combined with the length scaling results in Figure~\ref{fig-neutral}, we would
predict \emph{Normal} to both require a larger genome size than \emph{Reorder} and \emph{DAG},
as well as search a much smaller subset of the search space.

%TODO Consider returning to this problem after having done \emph{Depth} and
%\emph{Flat}, using the optimal versions of those two problems as the
%starting individual.

\subsection{Breadth}

\begin{table}
	\centering
	\begin{tabular}{|r|c|c|c|}
	  \hline
& \textbf{\emph{Normal}} & \textbf{\emph{Reorder}} & \textbf{\emph{DAG}} \\ \hline
\textbf{\emph{Breadth}} & 3.24 & 2.06 & 2.09 \\ \hline
  \textbf{\emph{Depth}} & 7.72 & 3.66 & 4.17 \\ \hline
   \textbf{\emph{Flat}} & 1.86 & 1.87 & 1.94 \\ \hline

	\end{tabular}
	\caption{$O(n^x)$ for each variant on each problem, where $n$ is problem size and $x$ is given in
	         the table.}
	\label{table-growth}
\end{table}

The \emph{Breadth} problem tests CGP's ability to evolve
graphs that connect to all inputs using the minimum possible number of nodes.
The landscape for this test is a problem with \emph{N} nodes which has $N+1$
inputs and a single output.  The optimal structure is a tree with all nodes in the genome
active and all inputs used exactly once as leaves.  To evaluate
an individual, all nodes execute the binary \emph{OR} function,
and possible program inputs are all strings of length $N+1$ with exactly one bit
set.  As a result, the fitness of an individual is equal to the number of inputs
connected to the output.  This problem is similar to the tree graph proposed
in~\cite{payne:2009:bias}, except the problem scales such that all nodes
must always be active, multiple inputs are used,
and the fitness function involves individual execution instead of graph edit distance.

Experimental results for the \emph{Breadth} problem are given in Figure~\ref{fig-breadth}.
This shows that both \emph{Reorder} and \emph{DAG} scale better than \emph{Normal} when solving \emph{Breadth},
but have similar behavior to each other.
From Table~\ref{table-growth} we see that the number of evaluations \emph{Normal} requires
to solve the \emph{Breadth} problem increases cubically while the other variants scale
quadratically.  A likely cause of this growth discrepancy is that the solution to \emph{Breadth} requires
all nodes in the genome to be active, which \emph{Normal} does with difficulty.
The fact that \emph{Reorder} does not share this issue even though its
search bias on \emph{Neutral} was relatively similar implies that \emph{Reorder} is better
able to adapt to problem pressures than \emph{Normal}.  While \emph{DAG} scaled
slightly worse than \emph{Reorder}, the difference is not significant.

\begin{figure}
  \centering
  \includegraphicsfit{breadth}
  \caption{Median evaluations until success for different sizes of the \emph{Breadth} problem} 
  \label{fig-breadth}
\end{figure}

\subsection{Depth}
The \emph{Depth} problem requires CGP to evolve graphs with the maximum possible node
reuse.  In this problem, all nodes use the same function, which takes the minimum of
their inputs, increments it by one, and returns that value.  The fitness function is
equal to the value returned, and each individual is tested on a single input of zero.
As a result, the maximum fitness is achieved when all nodes are connected in a chain
such that all nodes are active and all connections in each node go to the previous node in the graph.
Where \emph{Breadth} required no node reuse, \emph{Depth} requires the maximum amount of node
reuse.  Also in contrast to \emph{Breadth}, \emph{Depth} has a unique solution for
\emph{Normal} and \emph{Reorder}, with \emph{DAG} having multiple ways of representing
the same graph by encoding nodes in different orders.  This problem is similar to the linear graph proposed
in~\cite{payne:2009:bias}, except the problem scales such that all nodes
must always be active and the fitness function involves individual execution instead of graph edit distance.

Figure~\ref{fig-depth} and Table~\ref{table-growth} both show how \emph{Normal}
is effectively unable to solve this problem.  \emph{Normal} was only able to solve
the smallest three problem sizes of the twenty tried, with an experimental growth
rate of $O(n^{7.72})$.  By comparison, \emph{Reorder} and \emph{DAG} solved all
or almost all, respectively, of the problems tried, with far lower growth rates.
As \emph{Breadth} also required all nodes to be active, the reason for \emph{Normal}'s
breakdown must be something else about \emph{Depth}.  A likely explanation is that
\emph{Normal} requires structures to be evolved in a precise location in the genome,
in that a node with all connections set to $i$ is only part of the optimal solution
if that node is at position $i+1$.  If it is not, it must be reinvented at $i+1$ to
move to optimality.  This is especially problematic in the following case.  Consider
a fully optimal individual, except that $i+1$ is inactive and $i+2$ is active and
all of its connections go to $i$.  The only way to improve this individual is to
set all of $i+1$'s connections to $i$ and change all of $i+2$'s connections to $i+1$.
What makes this difficult is the absence of fitness gradient, meaning all changes can
only happen by drift and that changes to $i+2$ can only happen after $i+1$ is relatively
correct.  Otherwise the individual will decrease in fitness.  This problem is further
exacerbated by \emph{Normal}'s mutation method, which on average connects $i$ to $\frac{i}{2}$,
meaning high index nodes are rarely connected to adjacent indexes.
\emph{Reorder} avoids this problem by allowing structures to be moved to where they are
needed, while simultaneously decreasing the length of connections.  \emph{DAG} has
no concept of position, and therefore cannot have this problem.

While \emph{Reorder} and \emph{DAG} have very similar behaviors in Figure~\ref{fig-depth},
Table~\ref{table-growth} shows that \emph{Reorder} scaled slightly better.  A likely
cause of this is that \emph{Reorder} has a higher probability of making nodes
with uniform connections as compared to \emph{DAG}.  This is true
because nodes with no transitive dependency on the mutating node
and can still sometimes be unreachable by mutation in \emph{Reorder}.  Conversely, nodes
that the mutating node is transitively dependent on will always be reachable.  In \emph{DAG}
there is no bias toward connecting to already transitively dependent nodes.
As a result, \emph{Reorder} is able to leverage this
search bias to better make nodes with uniform connection genes, improving search speed.

\begin{figure}
  \centering
  \includegraphicsfit{depth}
  \caption{Median evaluations until success for different sizes of the \emph{Depth} problem} 
  \label{fig-depth}
\end{figure}

\subsection{Flat}
As a counter point to \emph{Breadth} and \emph{Depth}, which require entirely
active genomes, we propose \emph{Flat}.  Unlike the other problems, \emph{Flat}
evaluates an individual's genome directly, with the fitness of
an individual equal to the number of connections made directly to the input.
In this case the optimal individual is one where all nodes and the output location
are connected to the only input location.  This tests how well CGP is able to
create genomes with the least amount of internal structure.
Similar to \emph{Depth}, \emph{Flat} has exactly one solution. 
In~\cite{payne:2009:bias} some evidence was found that solutions
that appear less often in the search space may be harder to find.  If this is the reason
\emph{Depth} was more difficult than \emph{Breadth}, then \emph{Flat} should also
be difficult to solve.  As all nodes now
participate in evaluation regardless of connections, they are all considered active
when performing \emph{Single} mutation.
%Note that a more standard fitness function
%cannot be used as an inverse to \emph{Breadth} and \emph{Depth} as all three methods
%can equally make fully inactive individuals by directly connecting the output
%to the input at a probability of $\frac{1}{N+1}$.

On this problem, all three algorithms have approximately the same overall shape, shown
in Figure~\ref{fig-flat}, and have similar complexities, shown in Table~\ref{table-growth}.
\emph{DAG} does appear to be slower than the other two, although the difference may
just be by a constant scalar.  This makes sense as the probability of a connection
mutating to an input is $\frac{1}{N}$ for \emph{DAG} and $\frac{1}{i}$ for
\emph{Normal} and \emph{Reorder}, with $i\leq N$.  In \emph{DAG}, if there are nodes with
transitive dependencies on the mutating node, then the probability
of connecting to the input increases.  However, it can never surpass that for \emph{Normal}
and \emph{Reorder}.  More generally, these results support the idea that \emph{Reorder}
and \emph{DAG} do not have a harder time making highly unconnected structures
than \emph{Normal}, and that solution rarity in the search space is not why \emph{Normal}
did so poorly on \emph{Depth}.

\begin{figure}
  \centering
  \includegraphicsfit{flat}
  \caption{Median evaluations until success for different sizes of the \emph{Flat} problem} 
  \label{fig-flat}
\end{figure}

\section{Benchmarking}
\label{sec-benchmarking}
Understanding length biases and search limitations allows users to
solve new search problems easier.  In order to examine how our predicted search biases
interact with standard CGP landscapes, we tested all three algorithms on the 3-bit
multiplier problem.  In this problem there are six inputs and six outputs with
a function set of \{AND, OR, NAND, NOR\}.  Instead of scaling problem difficulty
as in Section~\ref{sec-tailored}, each algorithm was tested with scaling genome
sizes.  We used the range $\left [ 10 .. 8000 \right ]$, taking 21 points on an
exponential scale.  Other run parameters were kept the same.
The median number of evaluations to success is shown in
Figure~\ref{fig-multiply} on a log-log scale.  The results from the best genome
size are given in Table~\ref{table-multiply}.

As predicted, both the graphic and tabular results show that \emph{Reorder} and
\emph{DAG} can perform effective search with much smaller genome sizes than \emph{Normal}:
The best tested genome size for both novel
variants had lower median evaluations to success than \emph{Normal}'s best, which
also required a genome twenty times larger.  \emph{Reorder} was statistically
better than \emph{Normal} under the single tailed Mann-Whitney U test. While \emph{DAG} was not
statistically better than \emph{Normal}, the likely reason is \emph{Normal}'s exceptionally large median
absolute deviation.  We believe this deviation is caused by \emph{Normal}'s occasional need to
reinvent useful structures in different genome locations.  In the unlikely event that
active nodes are evolved with the right amount of inactive nodes between them, and in the
right order relative to each other, \emph{Normal} is likely going to receive far better
results.  In cases were these properties do not hold, \emph{Normal} will
have to recreate nodes in new locations, often with no fitness gradient, significantly
increasing run time.  As the number of nodes in the genome increases, the probability
that inactive nodes exist between two active nodes increases.  This may be part of
the reason \emph{Normal} requires such a large genome size to perform well.
Also, more inactive space in the genome may facilitate structure recreation in new locations.
As a final advantage to \emph{Reorder} and \emph{DAG}, we see that both evolved
solutions with half as many median active nodes as \emph{Normal}.  A potential
reason for this is \emph{Normal}'s need to leverage introns in order to perform
neutral search more than the other methods.

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|}
	  \hline
& \textbf{\emph{Normal}} & \textbf{\emph{Reorder}} & \textbf{\emph{DAG}} \\ \hline
\textbf{Genome} & 4000   & 200     & 200 \\ \hline
  \textbf{MES} & 272,324 & 213,660 & 249,031 \\ \hline
  \textbf{MAD} & 125,997 & 79,335  & 86,551 \\ \hline
  \textbf{P value} & N/A & 0.030   & 0.117 \\ \hline
  \textbf{Active} & 255  & 118     & 120 \\ \hline

	\end{tabular}
	\caption{Best configuration of each algorithm on 3-bit Multiplier.
	Includes maximum genome size, Median Evaluations to Success (MES), Median Absolute Deviation (MAD)
	one tailed P value comparison with \emph{Normal}, and median number of active
	nodes in the final solutions found.}
	\label{table-multiply}
\end{table}

\emph{Reorder} and \emph{DAG} are not without drawbacks.  In Figure~\ref{fig-multiply} we see that
\emph{Normal} may underperform at peak performance and at lower genome sizes, but
as the genome size increases it surpasses the other methods.  \emph{DAG} is especially
sensitive to increasing genome sizes, performing worse than \emph{Normal} on sizes above 600.
\emph{Reorder} also falls behind \emph{Normal} once sizes exceed 2000.  This
could indicate that the novel methods are more sensitive to genome size.
While this is true, we believe it to be less of a flaw and more of an unavoidable side
effect of these techniques being able to evolve more complex graphs than \emph{Normal}.
%TODO Consider expanding here.

\begin{figure}
  \centering
  \includegraphicsfit{multiply}
  \caption{Median evaluations until success on the Multiply problem using different genome sizes} 
  \label{fig-multiply}
\end{figure}

\section{Conclusions and Future Work}
\label{sec-conclusion}
Existing literature suggests CGP performs best when maximum genome sizes
far exceed actual solution size~\cite{miller:2006:redundancy}, supported by
evidence that inactive genes improve neutral search~\cite{yu:2001:neutrality},
and that CGP doesn't suffer from bloat~\cite{miller:2001:bloat}.  We, however, find evidence
that the need for highly inactive genomes is likely due to limitations in CGP's
search mechanisms.  When no selective pressure on size exists, CGP highly favors
short individuals, with this search bias continuing even when the problem structure
calls for longer individuals.  Furthermore, when combined with CGP's use of node
positioning, problems requiring significant proportions of the genotype to be
active were found to be difficult or impossible for CGP to solve.  We showed theoretical
and experimental evidence that the blame for this problem lies with the way CGP's mutation
interacts with its representation, as existing mechanisms impose artificial search limitations
not inherent to the search problem.  \emph{Reorder}, which shuffled node positions
without changing node expression, and \emph{DAG}, which removed all concept of node
position, were both able to avoid these limitations and more fully utilize
genome space.  As these methods were able to achieve better results with less
inactive genes, we conclude that while inactive genes are useful, the number required
by \emph{Normal} CGP is a result of the inherent length bias and search limitations in its
representation and evolutionary mechanisms, and not because that many inactive
nodes are needed for neutral search.  Similarly, CGP's resistance to bloat is likely
no more than the work of a previously unexamined form of parsimony pressure, the strength
of which is determined by the maximum number of nodes.  The drawback of \emph{Reorder} and
\emph{DAG} is that in removing this parsimony pressure, they are more able to utilize
all genes in the genome, making them more sensitive to the genome size.  Yet this
may prove to be an advantage, as it may allow CGP to solve more complex problems without
explosively increasing the genome size.

Further analysis of \emph{Reorder} and \emph{DAG} should be performed in light
of the evidence found here.  Both appear to be valid ways of making CGP more flexible
and powerful, but wider testing needs to be performed to solidify that claim.  As
an important aspect of those tests, run time comparisons need to be performed, as
both increase the computational expense of CGP, with \emph{DAG} especially having
the potential for significant overhead.  It is also possible either could benefit
from further modification, such as using different initialization or method for mutation.

More analysis on the tailored problems will likely improve the clarity of results.  Specifically
\emph{Neutral} would likely benefit from further scalability testing to see how search frequencies change
with respect to genome size, especially at larger genome sizes.  Tests with \emph{Breadth} and \emph{Depth} to see how many inactive
nodes are needed before \emph{Normal} surpasses \emph{Reorder} and \emph{DAG}, if ever, may
also help highlight exactly what proportion of inactive nodes \emph{Normal} needs to solve
problems.

%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\bibliography{../main}  % sigproc.bib is the name of the Bibliography in this case
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references
%
% ACM needs 'a single self-contained file'!
%
%APPENDICES are optional
\balancecolumns
%\balancecolumns % GM June 2007
% That's all folks!
\end{document}

### 2013-01-25 Reducing Wasted Evaluations in CGP/main.tex
Adding all of the source files for papers

%%%%%%%%%%%%%%%%%%%%%%% file typeinst.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is the LaTeX source for the instructions to authors using
% the LaTeX document class 'llncs.cls' for contributions to
% the Lecture Notes in Computer Sciences series.
% http://www.springer.com/lncs       Springer Heidelberg 2006/05/04
%
% It may be used as a template for your own input - copy it
% to a new file with a new name and use it as the basis
% for your article.
%
% NB: the document class 'llncs' has its own and detailed documentation, see
% ftp://ftp.springer.de/data/pubftp/pub/tex/latex/llncs/latex2e/llncsdoc.pdf
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[runningheads,a4paper]{llncs}

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{url}
\usepackage{hhline}
\newcommand{\includegraphicsfit}[1]
{\includegraphics[width=.9\textwidth,height=\textheight,keepaspectratio]{#1}}
%\urldef{\mailsa}\path|{alfred.hofmann, ursula.barth, ingrid.haas, frank.holzwarth,|
%\urldef{\mailsb}\path|anna.kramer, leonie.kunz, christine.reiss, nicole.sator,|
%\urldef{\mailsc}\path|erika.siebert-cole, peter.strasser, lncs}@springer.com|    
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

\begin{document}

\mainmatter  % start of an individual contribution

% first the title is needed
\title{Reducing Wasted Evaluations in Cartesian Genetic Programming}

% a short form should be given in case it is too long for the running head
%\titlerunning{Lecture Notes in Computer Science: Authors' Instructions}

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{Brian W.~Goldman\and William Punch}
%\author{Anonymous Author\and Anonymous Author}
%
%\authorrunning{Lecture Notes in Computer Science: Authors' Instructions}
% (feature abused for this document to repeat the title also on left hand pages)

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{BEACON Center for the Study of Evolution in Action,\\
Michigan State University, U.S.A.\\
brianwgoldman@acm.org, punch@cse.msu.edu}
%\institute{Department\\ Organization\\ Email}

%
% NB: a more complex sample for affiliations and the mapping to the
% corresponding authors can be found in the file "llncs.dem"
% (search for the string "\mainmatter" where a contribution starts).
% "llncs.dem" accompanies the document class "llncs.cls".
%

%\toctitle{Lecture Notes in Computer Science}
%\tocauthor{Authors' Instructions}
\maketitle

\begin{abstract}
Cartesian Genetic Programming~(CGP) is a form of Genetic Programming~(GP)
where a large proportion of the genome is identifiably unused
by the phenotype.  This can lead mutation to create offspring that are 
genotypically different but phenotypically identical, and therefore do
not need to be evaluated.  We investigate theoretically and empirically
the effects of avoiding these otherwise wasted evaluations, and provide
evidence that doing so reduces the median number of evaluations to solve
four benchmark problems, as well as reducing CGP's sensitivity to the mutation
rate.  The similarity of results across the problem set in combination with the
theoretical conclusions supports the general need for avoiding these unnecessary
evaluations.

%The abstract should summarize the contents of the paper and should
%contain at least 70 and at most 150 words. It should be written using the
%\emph{abstract} environment.

\keywords{cartesian genetic programming, mutation}
\end{abstract}

\section{Introduction}
%TODO Consider change to "best results"
In Genetic Programming~(GP) the most common metric for measuring algorithm complexity
is the number of fitness evaluations required to solve
a black box problem.  This metric assumes that the cost of evaluating an
individual will dominate search times on real world problems.  Under this
metric, search algorithms that best exploit evaluation information and avoid
unnecessary evaluations will receive the best results.  Cartesian Genetic
Programming~(CGP), a branch of GP, has a particular structure that allows
for the detection of some phenotypically identical genotypes without
requiring fitness evaluation.  No previous work has investigated if this
property can be exploited to meaningfully improve CGP efficiency.  After discussing the
potential theoretical gains of avoiding these evaluations, we propose and
investigate three novel techniques designed to improve CGP's
search.

\section{Cartesian Genetic Programming}
\label{sec:cgp}
Cartesian Genetic Programming is a variant of genetic programming in which
individuals encode directed acyclic graphs (DAG).
An in-depth explanation of CGP is provided in~\cite{miller:2011:chapter2}, but for
our purposes the important features are:
\begin{itemize}
  \item DAGs are represented by a collection of nodes and genes specifying output locations.
  \item Nodes contain genes describing both what function they perform
  and how they connect to other nodes.
  \item Offspring are created using mutation.
  \item Offspring replace parents if they are of the same or greater fitness.
\end{itemize}
CGP's encoding method does not require a node
to connect to an output location, which is required for it to
affect the phenotype of the individual.
These unconnected nodes and their genes are commonly called ``inactive''
and the connected nodes and their genes are called ``active.''
As the topology of the DAG is constantly evolving, the number and location of
active nodes can change from parent to child.
There have been numerous studies on the usefulness of inactive nodes~\cite{vassilev2000neutrality,yu:2001:neutrality}
including an argument that it is desirable to have up
to $95\%$ of the genome inactive~\cite{miller:2006:redundancy}.

CGP's canonical variation operator is point mutation.  Unfortunately the details
of this operator are often ambiguously stated.  For instance, in some papers this
operator chooses a set number of genes at random to be mutated~\cite{miller:2011:chapter2},
while in other papers each gene can be mutated with a certain probability,
allowing any number of genes to be mutated at once~\cite{harding:2012:mtcgp}.  Also, mutation in CGP can either
force a mutated gene's value to change~\cite{harding:2012:mtcgp,miller:2011:chapter2},
or randomly reset it to any valid value including the genes previous value~\cite{miller:2006:redundancy}.
As much of our analysis hinges on
the behavior of mutation, we specify mutation as follows:
when performing mutation, all genes use the mutation rate as the
probability that they are changed to a different value.
%Only if a gene has exactly one valid value\footnote{Genes can have just a single valid
%value in the case of the first node's connection genes in single input problem
%or if there is only a single valid function.}
%will mutation leave the gene unchanged.

The most common form of CGP uses a $1+\lambda$ strategy, where all offspring are
mutants of the parent.  The most fit offspring replaces the parent only if its
fitness is no worse than the parent.  Ties between offspring are broken by random selection
and ties between the offspring and the parent are awarded to the offspring:  in this manner
inactive genes are allowed to drift, which has been shown to significantly improve
performance~\cite{yu:2001:neutrality}.  We will refer to this standard form of
mutation and replacement as \emph{Normal}.

\section{Wasted Evaluations}
Preventing search from repeatedly evaluating the same search point can improve
any search that is limited by the number of evaluations it can perform.
In an evolutionary search employing indirect phenotype encoding, as CGP does,
the same search point can be represented by multiple different genotypes
making duplicate detection more difficult.  While it does not catch all
identical phenotypes, if two CGP individuals are actively identical (contain
the same active genes) then their phenotypes must be identical.
Miller~\cite{miller:2011:chapter2} theorized
that this may lead to wasted evaluations as any number of actively identical
individuals can be evaluated even though they will receive identical fitnesses.  He did not,
however, provide any analysis for how frequently these unnecessary evaluations may occur.

\begin{figure}
  \centering
  \includegraphicsfit{Probability}
  \caption{Probability of no active gene being mutated for different mutation rates} 
  \label{fig:probability}
\end{figure}
\subsection{Detection}
Before evaluating a CGP individual, it is common practice to determine which
nodes are active allowing inactive nodes to be ignored during
evaluation, leading to a reduction in runtime~\cite{vasicek:2012:efficient}.
By storing the set of active genes as part of the individual, it is trivial
to determine when active genes are mutated.  More generally, it takes $O(N)$ additional time to determine
if two randomly chosen individuals contain different active genes, where $N$ is
the number of active nodes, assuming at least one of the individuals has been
evaluated.

\begin{comment}
To exploit this knowledge to its fullest, all newly created individuals could
be compared against all previously evaluated individuals to determine if they
need to be evaluated.  While this can be optimized by storing a hash table of
actively unique individuals, the likelihood that two unrelated individuals are
actively identical may be too low to warrant the memory overhead required by being
this thorough.
\end{comment}

%A reasonable heuristic to reduce this overhead is to only check if
A reasonable heuristic to avoid duplicate evaluations is to determine if
a parent's active genes are identical to its offspring.  This will catch the
majority of cases, as the probability of two individuals being actively identical
decreases as the number of mutation applications
between them increases.  This also has the advantage that
extra individuals need not be stored in memory.  Due
to these advantages, we only compare an offspring against its parent
to detect unnecessary evaluations.

\subsection{Frequency of Offspring Actively Identical to their Parent}
\label{sec:freq}
Using the definition of point mutation given in Section~\ref{sec:cgp},
it is possible to quantify the probability that an offspring is actively
identical to its parent.
\begin{equation}
  \left (1 - mutation\_rate  \right )^{active\_genes}
\label{eq:waste}
\end{equation}
\begin{comment}
First, the number of active genes in each individual
is given by Equation~\ref{eq:active}.  This shows how the number of active genes
can change greatly with the number of active nodes and the arity used for each node.
\begin{equation}
  active\_genes = active\_nodes\cdot\left(arity + 1 \right) + output\_length
  \label{eq:active}
\end{equation}
\end{comment}
Equation~\ref{eq:waste} shows the probability that none of the active genes are
mutated at the given mutation rate.  To help display the effects of this relationship,
Figure~\ref{fig:probability} illustrates the probability that an offspring is actively
identical to its parent for various mutation rates and number of active genes.
Figure~\ref{fig:probability} shows that as the mutation rate and the number of active genes decreases,
the probability of an offspring being actively identical to a parent greatly increases.
While the other versions of mutation discussed in Section~\ref{sec:cgp} will change the details
of this equation, the overall relationship between the variables will be the same.

\begin{comment}
Using the other versions of mutation in Section~\ref{sec:cgp} does not change the
overall relationship of the number of active genes and the mutation rate to the
frequency of wasted evaluations.  If mutated genes are not forced to change to a
new value the probability of actively identical offspring will increase.  While this is intuitively
straightforward, determining the exact amount of increase is complex as the probability
that a randomly reset gene returns to its current value depends on which gene is being mutated.
If mutation is changed to have a set percentage of the genome mutated, the relationship
changes to use the percentage of the genome that is active, but it will still follow the general
shape of Figure~\ref{fig:probability}.
\end{comment}

\section{Methods to Avoid Wasting Evaluations}
\subsection{Skip}
The most straightforward method to use existing evaluation results is to
set the offspring's fitness to the parent's
if they are actively identical. This method will be referred to as \emph{Skip}.
As \emph{Skip} makes no modification to evolutionary mechanisms, any problem
solvable by \emph{Normal} will be solvable by \emph{Skip}, and vice versa. The
primary difference is only that \emph{Normal} will always use at least as many
evaluations as \emph{Skip}, with the potential to use significantly more.  As
a result, \emph{Skip} may solve problems \emph{Normal} cannot if given a
maximum number of evaluations.

A significant theoretic advantage to this technique is that it lowers CGP's
sensitivity to the mutation rate.  In \emph{Normal}, as the mutation rate
decreases the potential for wasted evaluations
increases creating an evaluation penalty.  As the mutation rate increases the expected
number of modified active genes increases making small
changes in phenotype less probable. From this, \emph{Normal} will likely
be most effective when the mutation rate
is high enough to expect at least one active gene to be modified by
mutation, but low enough to allow the small phenotypic moves required
to reach optimal values.
Unlike \emph{Normal}, \emph{Skip} can use mutation rates expected to produce
frequent offspring that are not actively different from their parents without
penalty.  As such \emph{Skip} can use very fine grained mutation allowing it to potentially
hill climb much more effectively.

\subsection{Accumulating Mutation}
As was stated in Section~\ref{sec:cgp}, some of CGP's success stems
from allowing genetic drift of inactive genes~\cite{yu:2001:neutrality}.
%Inactive gene modification can build structures which can potentially be made active
%by later mutations.
This raises the following question: will evolution in CGP improve
if inactive genes are allowed to change at an increased rate?

To test this possibility we propose a method of mutation called \emph{Accumulate}.
The primary concept behind this technique is to allow inactive genes to drift
multiple generations worth of distance in a single actual generation of CGP.
For our example we will refer to one of the offspring produced by the parent as $F_0$.  If $F_0$
is actively identical to its parent, $F_0$ produces a mutant offspring $F_1$ in the same generation.
This chain continues with $F_i$ mutating to produce $F_{i+1}$ until an individual
is produced that has one or more active genes different from the original parent.
This actively different individual is called $F_n$.  $F_n$ is then evaluated.
If $F_n$ is no worse than the original parent, it replaces $F_0$ and CGP continues
evolution and selection as normal.  If $F_n$ is worse than the original parent
$F_{n-1}$ replaces $F_0$.  Thus accumulate's offspring can have zero to multiple
active genes changed.

By chaining mutations each $F_i$ will probabilistically have more changes to
inactive genes than $F_{i-1}$ when compared with the original parent.  By using
$F_{n-1}$ or $F_n$ instead of $F_0$, \emph{Accumulate} can drift faster than
\emph{Normal} and \emph{Skip}.  This can be advantageous on problems that benefit
from high exploration of the neutral space, with the potential drawback that
previously active useful structures are less likely to survive to be reintegrated.

\subsection{Single Active Mutation}
%An alternative method for avoiding unnecessary evaluations is to
%modify mutation to ensure offspring
%are actively different from their parents. 
\emph{Skip}
and \emph{Accumulate} have the potential, similar to \emph{Normal}, to generate large numbers of individuals
with no active differences, which can be computationally expensive even if they
are not evaluated. To avoid
this potential overhead, we propose a method to ensure a \emph{Single} active
gene is mutated every time an offspring is generated.
%
Creating an offspring using \emph{Single} is an iterative process.  Until an
active gene has been mutated, select a gene at random in the individual and
mutate its value.  This has the following properties:
\begin{itemize}
\item Exactly one active gene is mutated for all offspring.
\item Zero or more inactive genes can be mutated.
\item A gene that is active will be mutated more frequently than one that is not.
\item No mutation rate is required.
\end{itemize}
Many of these properties make \emph{Single} distinct from the other described mutation
methods.  By forcing only a single active gene to change, \emph{Single} may have
better results on problems where incremental improvements are possible with single
gene changes and worse results on problems where larger changes are necessary.
This is in contrast to all of the other methods that have the potential to change
multiple active genes.
Limiting mutation to a single active gene change does not, however, prevent \emph{Single}
from making large phenotypic changes, which are still possible by connecting
in inactive nodes.

\emph{Single} has the property that the effective mutation rate is always
$\frac{1}{a}$ for active genes and $\frac{1}{a + 1}$ for
inactive genes, resulting in $\frac{n - a}{a + 1} + 1$ expected mutations, where
$a$ is the number of active genes and $n$ is the total number of genes.  This
means that the number of inactive genes does not affect the probability of an
inactive gene being mutated, just the expected number of inactive genes that are
mutated.  Furthermore, as the number of active genes approaches zero, the expected
number of mutations increases dramatically, even though it is still limited to
changing a single active gene.  As the evolved number of active genes is likely to be
highly problem dependent, the behavior of \emph{Single} will likely change significantly
from problem to problem, with some problems having very little changes to inactive
genes, and others changing a great deal.  It is also possible that this will create
a selective pressure on the number of active genes which may only become apparent
if multiple parents are used.

In comparison with \emph{Skip}, \emph{Single} is even easier to apply to new problems
as it completely removes the need to set the mutation rate, but does so at the
potential expense of lower maximum effectiveness.  In comparison with \emph{Accumulate},
\emph{Single} mutates inactive genes less often with respect to active genes, making it
better at preserving useful inactive structures but worse at exploring the inactive
space.

\section{Experimental Setup}
In order to empirically test the suggested benefits of \emph{Skip},
\emph{Accumulate}, and \emph{Single} in comparison with \emph{Normal}, we evaluated their performance on
four benchmarks: Parity, Multiplier, Binary Decode and Binary Encode.
Both Parity and Multiplier have been used extensively to test
CGP~\cite{walker:2008:cgpmodules,miller:2006:redundancy}.
The 3-bit even parity is included as it is the most commonly used benchmark for CGP and to
help provide some level of comparability with previous work.  In order to try a
more challenging binary problem, we chose to do the 3-bit multiplier, which has
6 inputs and 6 outputs.

Binary Decode is effectively a demultiplexer, where $N$
input lines specify which of the $2^N$ output lines should have a value of one, with
all others set to zero.  For example, if $101$ is the input to a 3-bit Binary Decode
problem, the output is $00100000$.  Binary Encode is the reverse of Binary Decode.
Given $2^N$ input lines, where exactly one line has a bit set to one and all others
are zero, the binary value index of the one bit is output using $N$ lines.
An example of this mapping is that $00100000$
should return the output of $101$.  The reason we use Binary Decode and Binary Encode
as benchmarks is their asymmetric number of inputs to outputs and their relatively
small number of test cases.  For example, on the 16 to 4 Binary Encode and the 4 to 16 Binary
Decode (the instances we use) only 16 test cases are needed.

All problems used the operators \{and, or, nand, nor\} and determine fitness as the average
percentage of correct bits across all possible inputs.  While this is a common function set used
by CGP for these problems,~\cite{walker:2008:cgpmodules} used a different set of operators
for Multiplier.  Preliminary testing shows our operator set makes Multiplier harder than theirs,
but, as always, conclusions drawn by comparing experiments using different operator sets should
be minimal.

While a few of these benchmarks may be overly simple, with parity especially
being considered for obsolescence~\cite{mcdermott:2012:benchmarks}, they are used here
for the specific reason of their simplicity.  The primary
difference between \emph{Normal} and \emph{Skip}
is not that one can solve harder problems than the other, rather it is that \emph{Skip} is thought
to solve problems in less evaluations than \emph{Normal}.  Being able to compare each method's
median number of evaluations until success makes for the most logical metric in this case,
so these problems are useful in that they can be solved often.  While much of the
previous work on CGP uses Koza's \emph{computational effort} to analyze experimental
results,~\cite{christensen:2002:comp_effort} explains how that can introduce
significant underestimates, especially when using small population sizes.

In order to compare with the effective standard CGP, we use the $1+4$
evolutionary strategy.  We use a
genotype size of 3000 nodes, similar to those found optimal in~\cite{miller:2006:redundancy}.
Each run was allowed up to
10 million evaluations before being terminated, and 50 runs of each experiment
were performed.  To help focus the differences between each mutation method,
all starting individuals on the same problem were identical for the same run number,
meaning that the 50 different individuals \emph{Normal} started with were identical
to the 50 starting individuals for \emph{Skip}, \emph{Accumulate}, and \emph{Single}.
For statistical purposes this means all results should be compared as paired tests.
As the number of evaluations to success is likely not normally distributed, we choose
the Wilcoxon signed-rank test to compare each mutation method with \emph{Normal}.
As an attempt to cover a wide range of mutation rates, the 16 mutation rates listed
in Figure~\ref{fig:probability} were used for each method.  These cover from 0.0001
to 0.8 on a log scale, and should encompass all previous levels of mutation used
with CGP.

\begin{comment}
We had intended to include Koza-1 and Pagie-1 from~\cite{mcdermott:2012:benchmarks}
as sample regression problems, but we had difficulty achieving acceptable success
rates to allow for comparisons here.  Future work may be needed to understand why
these problems were significantly harder when using our configuration settings
than the four binary problems. 
\end{comment}

\section{Results}
To summarize the results of testing each mutation rate, Figures~\ref{fig:parity},
\ref{fig:multiply}, \ref{fig:encode}, and \ref{fig:decode} show the median number of evaluations
required over 50 runs to solve the Parity, Multiply, Binary Encode, and Binary Decode problems.  Note
that both axes are presented on a log scale.  In the event that the median run of a configuration
did not find the optimal value before termination, the value is not displayed in the graph.  This can be
clearly observed in Figure~\ref{fig:multiply} for mutation rates greater than 0.01.  As \emph{Single}
does not use a mutation rate, its data is assumed constant for all mutation rates.
%
Table~\ref{table:results} shows fine details
about the best configurations found for each mutation method.  This includes the mutation rate
with the lowest median evaluations until success and the median evaluations it required, the median absolute
deviation from that median, the single tailed P-value received from comparing with the best \emph{Normal}
configuration using the Wilcoxon signed-rank test, and the median number of active genes in the solution.
Note that the number of active genes is a different measure from the number of active nodes, which has
been reported in other work, and is used here for compatibility with Equation~\ref{eq:waste}.
%
A complete package of our results as well as the source code used in testing
is available from our website.\footnote{\url{https://github.com/brianwgoldman/ReducingWastedEvaluationsCGP}}

\begin{figure}
  \centering
  \includegraphicsfit{parity}
  \caption{Median evaluations to success for each method on the Parity problem} 
  \label{fig:parity}
\end{figure}

\begin{figure}
  \centering
  \includegraphicsfit{multiply}
  \caption{Median evaluations to success for each method on the Multiply problem} 
  \label{fig:multiply}
\end{figure}

\begin{comment}
\begin{figure}
  \centering
  \includegraphicsfit{quartic}
  \caption{Mean Evaluations to Success for each method on the Quartic problem} 
  \label{fig:quartic}
\end{figure}
\end{comment}

\begin{figure}
  \centering
  \includegraphicsfit{encode}
  \caption{Median evaluations to success for each method on the Binary Encode problem} 
  \label{fig:encode}
\end{figure}

\begin{figure}
  \centering
  \includegraphicsfit{decode}
  \caption{Median evaluations to success for each method on the Binary Decode problem} 
  \label{fig:decode}
\end{figure}

\section{Discussion}
Even though the Parity, Multiplier, Binary Encode, and Binary Decode problems have
very different numbers of inputs and outputs, different input to output ratios, 
different evolved numbers of active genes (see Table~\ref{table:results}), and
different levels of difficulty, Figures~\ref{fig:parity}, \ref{fig:multiply},
\ref{fig:encode}, and \ref{fig:decode} present surprisingly similar results.  
\begin{comment}
On all four
problems \emph{Normal} has a fairly clear quadratic shape which at its lowest point
approaches or sometimes crosses the line of \emph{Single}.  As the mutation rate increases
beyond the optimal value for \emph{Normal} all techniques except \emph{Single} share
very similar if not identical progressions.  Below \emph{Normal}'s optimal mutation
rate, \emph{Normal} gets significantly worse while \emph{Skip} and \emph{Accumulate}
diverge from \emph{Normal} by maintaining similar qualities as the optimal value.
At exceptionally low mutation rates \emph{Normal} takes over an order of magnitude
more evaluations to find success than all of the other techniques.

The shape of these features is as expected.
\end{comment}
  The steep increases in median evaluations
for \emph{Normal} corresponds with the prediction that overly high mutation rates will
significantly increase the number of evaluations until success, 
while low mutation rates will create a significant portion of
wasted evaluations.  On Multiply, Binary Encode, and Binary Decode, overly high mutation rates
prevent \emph{Normal} from reaching the optimum.
Parity is the only problem CGP could solve using exceptionally high mutation rates, with even
a slight improvement at the very top of the tested range.  This likely happens as the Parity problem
we are using is so simple that a CGP using effectively random search is able to solve it.
\emph{Skip} and \emph{Accumulate}'s tendency to mirror \emph{Normal}'s effectiveness as the
mutation rate increases is because the probability of creating actively identical offspring
approaches zero.  As the divergence point is always near
\emph{Normal}'s optimum mutation rate, it can be inferred that \emph{Normal}'s significant decrease in
quality when using lower mutation rates is likely caused by wasted evaluations.
As \emph{Skip} also decreases
somewhat in quality, some of the impaired fitness can also likely be attributed
to improbability of making large enough genotypic changes to escape local optima
with small mutation rates.

All four figures suggest that \emph{Skip} and \emph{Accumulate}
have a lower sensitivity to the mutation rate than \emph{Normal}.
As the mutation rate goes below the optimal value,
\emph{Skip} and \emph{Accumulate} encounter a relatively minor increase in their median number of evaluations
when compared with \emph{Normal}. As confirmation of Equation~\ref{eq:waste}
we recorded the predicted number of wasted evaluations and found
the aggregate predicted waste was nearly identical to the differences between \emph{Normal} and \emph{Skip}.

\emph{Single}, the only method without a mutation rate, is clearly not as generally effective as
the other methods where mutation rate can be optimized.  However, when the optimal mutation rate is
not known, \emph{Single} does show significant advantages.  When compared against the entire
space of tested mutation rates, \emph{Single} almost always requires
the least number of evaluations to succeed.  Below the optimum mutation rates for each, \emph{Skip}
and \emph{Accumulate} are reasonable approximations of \emph{Single}, but come with the disadvantage
that they are more computationally expensive.
%
As the number of evaluations required to solve a problem increases, \emph{Single}
appears to improve in performance relative to \emph{Normal}.  Parity required the
least number of evaluations to be solved, followed by Binary Encode and Binary Decode,
with Multiply requiring by far the most.
This ordering mirror's \emph{Single}'s relative scaling with \emph{Normal}, 
as it goes from a lower peak performance than \emph{Normal}
on Parity and Binary Encode to surpassing it on Binary Decode and Multiply.

\begin{table}
	\centering
	\caption{Comparisons of Best Mutation Rate for Each Mutation Method on Each Benchmark.
	  All results are from 100\% successful configurations.}
	\begin{tabular}{r|c|c|c|c|c|c|c|c|c|c|c|}
	  \cline{3-6}
		 \multicolumn{2}{r|}{}    & \textbf{\emph{Normal}} & \textbf{\emph{Skip}} & \textbf{\emph{Accumulate}} & \textbf{\emph{Single}} \\ \cline{2-6} 
		                    & Mutation Rate & 0.02 & 0.02 & 0.01 & N/A \\ \cline{2-6}
	       \textbf{3-bit} & Median Evaluations & 1,018 & 959 & 928 & 1358 \\ \cline{2-6}
        \textbf{Parity} & Median Absolute Deviation & 614 & 564 & 416 & 733 \\ \cline{2-6}
	                      & P value  & N/A & 0.0000 & 0.4875 & 0.0790 \\ \cline{2-6}
	                      & Median Final Active Genes & 250 & 250 & 262 & 223 \\ \hhline{~=====}

		                    & Mutation Rate & 0.004 & 0.002 & 0.0008 & N/A \\ \cline{2-6}
	       \textbf{3-bit} & Median Evaluations & 503,424 & 398,354 & 421,859 & 340,165 \\ \cline{2-6}
	  \textbf{Multiplier} & Median Absolute Deviation & 191,037 & 184,979 & 165,218 & 151,063 \\ \cline{2-6}
	                      & P value  & N/A & 0.0212 & 0.3632 & 0.0065 \\ \cline{2-6}
	                      & Median Final Active Genes & 627 & 629 & 684 & 620 \\ \hhline{~=====}

                        & Mutation Rate & 0.004 & 0.004 & 0.008 & N/A \\ \cline{2-6}
	     \textbf{16 to 4} & Median Evaluations & 24,057 & 19,732 & 21,563 & 26,031 \\ \cline{2-6}
	      \textbf{Encode} & Median Absolute Deviation & 8,829 & 7,425 & 7,449 & 10,170 \\ \cline{2-6}
	                      & P value  & N/A & 0.0000 & 0.0864 & 0.3578 \\ \cline{2-6}
	                      & Median Final Active Genes & 472 & 472 & 499 & 510 \\ \hhline{~=====}

	                      & Mutation Rate & 0.004 & 0.002 & 0.004 & N/A \\ \cline{2-6} \cline{2-6}
       \textbf{4 to 16} & Median Evaluations & 73,351 & 70,023 & 65,347 & 69,564 \\ \cline{2-6}
        \textbf{Decode} & Median Absolute Deviation & 17,889 & 17,036 & 18,411 & 21,131 \\ \cline{2-6}
	                      & P value  & N/A & 0.415 & 0.3276 & 0.2398 \\ \cline{2-6}
	                      & Median Final Active Genes & 1,117 & 1,027 & 1,050 & 1,090 \\ \cline{2-6}

  \end{tabular}
  \label{table:results}
\end{table}

Using Table~\ref{table:results}, it is clear that \emph{Normal} is never the best option.
In all cases \emph{Skip} and \emph{Accumulate} can obtain better results,
with even \emph{Single} doing better on Binary Decode and Multiply.  Furthermore, \emph{Skip} always
had a lower median absolute deviation than \emph{Normal}, with \emph{Accumulate} lower
on all problems except Binary Decode.  The statistical analysis is less clear as the deviations
are so large, but \emph{Skip}'s improvements are definitely significant on all problems except
Binary Decode, with \emph{Single} statistically better than \emph{Normal} on Multiply.
As support that our configuration
for \emph{Normal} is valid to compare against, our peak quality is approximately 6 times
better than results published in~\cite{walker:2008:cgpmodules} for their standard CGP and ECGP
on Parity.
%TODO Discuss length stuff

While the true complexity measure for a GP implementation is how many evaluations
it requires to solve a problem, we also performed preliminary timing tests to ensure
our new techniques were less computationally complex than the evaluations they avoid.
When using the best found mutation rate on Multiply, \emph{Normal} required the most time (4597 minutes)
to complete all 50 runs, with \emph{Skip} using 9\% less time, \emph{Accumulate} 16\% less,
and \emph{Single} 29\% less.

\section{Conclusions and Future Work}
CGP's current mutation operator has the potential to create offspring which can
be identified as phenotypically identical to their parents without invoking
the evaluation function.  Evaluating these offspring creates waste, as we can
assign their fitness to be identical to their parent.  Through careful definition
of how the mutation operator works, we have shown how the probability of having
wasted evaluations is dependent on the mutation rate and the number of active genes.
We predicted and then provided empirical evidence that avoiding this waste can
reduce how many evaluations CGP requires to solve problems as well as reduce
CGP's sensitivity to the mutation rate parameter.

We proposed \emph{Skip} and \emph{Accumulate} as two methods for what to do when
actively identical offspring are detected, and \emph{Single} as a method to ensure
they are never created.  On four fairly diverse benchmark problems we showed that
\emph{Skip} and \emph{Accumulate} do at least as well as \emph{Normal} using a
variety of mutation rates and are frequently vastly more efficient than \emph{Normal}.
As there was little discernible difference in their overall quality, for now we suggest \emph{Skip}
be used as a replacement to \emph{Normal}, as it is less complex than \emph{Accumulate},
although future experimentation may suggest otherwise.  In situations where the mutation
rate cannot be optimized, we suggest \emph{Single}, as while it did not achieve
the best peak performance on all problems, it has the least overhead with the widest applicability.
Furthermore, \emph{Single} achieved the best results on the hardest test problem, suggesting more
extensive testing of this method on difficult problems should be performed.

Even though the results appear to be problem independent, \emph{Skip}, \emph{Accumulate},
and \emph{Single} should be tested on a wider range of problem classes to make sure.
More advanced techniques for handling or avoiding wasted evaluations may also improve
results, such as a modified version of \emph{Single} to allow multiple active genes
to be mutated at once.  Nevertheless, we feel that the theoretic and empirical evidence
sufficiently supports a need to avoid wasting evaluations on actively identical
individuals.

\bibliographystyle{splncs}
\bibliography{../main}
\end{document}

### 2013-01-25 Reducing Wasted Evaluations in CGP/sub.sh

@@ -1,0 +1,13 @@
+author=goldman
+mkdir $author
+sed -e "s,bibliography{../main},bibliography{$author},g" main.tex > $author/${author}.tex
+cp ../main.bib $author/${author}.bib
+cp *.eps $author
+cp *.sty *.bst *.cls $author
+sed -e "s,main,$author,g" Makefile > $author/Makefile
+make -C $author/
+make -C $author/ clean
+cp authors.txt $author/
+cp abstract.txt $author/
+cp ${author}cr.pdf $author/
+zip -9 -r ${author}.zip $author

### 2013-02-26 CGP Journal/main.tex
Created initial files for eventual CGP Journal Article, with IEEE Transactions formatting

%% bare_jrnl.tex
%% V1.4
%% 2012/12/27
%% by Michael Shell
%% see http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.8 or later) with an IEEE journal paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
%% and
%% http://www.ieee.org/

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. IEEE's font choices can trigger bugs that do  ***
% *** not appear when using other class files.                            ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE! 
%% User assumes all risk.
%% In no event shall IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%
%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex,
%%                    bare_jrnl_transmag.tex
%%*************************************************************************

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[journal]{IEEEtran}
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[journal]{../sty/IEEEtran}

%\linespread{2.5}

% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

% *** CITATION PACKAGES ***
%
%\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off
% such as if a citation ever needs to be enclosed in parenthesis.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation
% can be obtained at: 
% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found at:
% http://www.ctan.org/tex-archive/info/epslatex/
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

% *** MATH PACKAGES ***
%
%\usepackage[cmex10]{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics. If using
% it, be sure to load this package with the cmex10 option to ensure that
% only type 1 fonts will utilized at all point sizes. Without this option,
% it is possible that some math symbols, particularly those within
% footnotes, will be rendered in bitmap form which will result in a
% document that can not be IEEE Xplore compliant!
%
% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
% There is also a support site at:
% http://algorithms.berlios.de/index.html
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/tools/

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

% *** SUBFIGURE PACKAGES ***
%\ifCLASSOPTIONcompsoc
%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
%\else
%  \usepackage[caption=false,font=footnotesize]{subfig}
%\fi
% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
% for subfigure.sty, the latter of which is no longer maintained and is
% incompatible with some LaTeX packages including fixltx2e. However,
% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
% which will override IEEEtran.cls' handling of captions and this will result
% in non-IEEE style figure/table captions. To prevent this problem, be sure
% and invoke subfig.sty's "caption=false" package option (available since
% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
% handling of captions.
% Note that the Computer Society format requires a larger sans serif font
% than the serif footnote size font used in traditional IEEE formatting
% and thus the need to invoke different subfig.sty package options depending
% on whether compsoc mode has been enabled.
%
% The latest version and documentation of subfig.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/

% *** FLOAT PACKAGES ***
%
%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure. The latest version and documentation can be found at:
% http://www.ctan.org/tex-archive/macros/latex/base/

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
% Do not use the stfloats baselinefloat ability as IEEE does not allow
% \baselineskip to stretch. Authors submitting work to the IEEE should note
% that IEEE rarely uses double column equations and that authors should try
% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
% packages (also by Sigitas Tolusis) as IEEE does not format its papers in
% such ways.
% Do not attempt to use stfloats with fixltx2e as they are incompatible.
% Instead, use Morten Hogholm'a dblfloatfix which combines the features
% of both fixltx2e and stfloats:
%
% \usepackage{dblfloatfix}
% The latest version can be found at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/dblfloatfix/

%\ifCLASSOPTIONcaptionsoff
%  \usepackage[nomarkers]{endfloat}
% \let\MYoriglatexcaption\caption
% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
%\fi
% endfloat.sty was written by James Darrell McCauley, Jeff Goldberg and 
% Axel Sommerfeldt. This package may be useful when used in conjunction with 
% IEEEtran.cls'  captionsoff option. Some IEEE journals/societies require that
% submissions have lists of figures/tables at the end of the paper and that
% figures/tables without any captions are placed on a page by themselves at
% the end of the document. If needed, the draftcls IEEEtran class option or
% \CLASSINPUTbaselinestretch interface can be used to increase the line
% spacing as well. Be sure and use the nomarkers option of endfloat to
% prevent endfloat from "marking" where the figures would have been placed
% in the text. The two hack lines of code above are a slight modification of
% that suggested by in the endfloat docs (section 8.4.1) to ensure that
% the full captions always appear in the list of figures/tables - even if
% the user used the short optional argument of \caption[]{}.
% IEEE papers do not typically make use of \caption[]'s optional argument,
% so this should not be an issue. A similar trick can be used to disable
% captions of packages such as subfig.sty that lack options to turn off
% the subcaptions:
% For subfig.sty:
% \let\MYorigsubfloat\subfloat
% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
% However, the above trick will not work if both optional arguments of
% the \subfloat command are used. Furthermore, there needs to be a
% description of each subfigure *somewhere* and endfloat does not add
% subfigure captions to its list of figures. Thus, the best approach is to
% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
% and instead reference/explain all the subfigures within the main caption.
% The latest version of endfloat.sty and its documentation can obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/endfloat/
%
% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
% later in the document, say, to conditionally put the References on a 
% page by themselves.

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
%\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/url/
% Basically, \url{my_url_here}.

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

% correct bad hyphenation here
%\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
% Do not put math or special symbols in the title.
\title{TODO Title}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

%\author{Michael~Shell,~\IEEEmembership{Member,~IEEE,}
%        John~Doe,~\IEEEmembership{Fellow,~OSA,}
%        and~Jane~Doe,~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
\author{Brian~W.~Goldman, William~F.~Punch}
%\thanks{M. Shell is with the Department
%of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
%GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
%\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
%\thanks{Manuscript received April 19, 2005; revised December 27, 2012.}}

% note the % following the last \IEEEmembership and also \thanks - 
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
% 
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!
%
% a space would be appended to the last name and could cause every name on that
% line to be shifted left slightly. This is one of those "LaTeX things". For
% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
% "AB" then you have to do: "\textbf{A}\textbf{B}"
% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names. For what it is worth,
% this is a minor point as most people would not even notice if the said evil
% space somehow managed to creep in.

% The paper headers
\markboth{IEEE Transactions on Evolutionary Computation,~Vol.~\#\#, No.~\#, Month~Year}%
{IEEE Transactions on Evolutionary Computation,~Vol.~\#\#, No.~\#, Month~Year}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

% If you want to put a publisher's ID mark on the page you can do it like
% this:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2012 IEEE}
% Remember, if you use this you must call \IEEEpubidadjcol in the second
% column for its text to clear the IEEEpubid mark.

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
The abstract goes here.
\end{abstract}

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
TODO Keywords
\end{IEEEkeywords}

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{T}{ODO this} section.
% You must have at least 2 lines in the paragraph with the drop letter
% (should never be an issue)

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation Results.}
%\label{fig_sim}
%\end{figure}

% Note that IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command,
% and the \label for the overall figure must come after \caption.
% \hfil is used as a separator to get equal spacing.
% Watch out that the combined width of all the subfigures on a 
% line do not exceed the text width or a line break will occur.
%
%\begin{figure*}[!t]
%\centering
%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
%\label{fig_second_case}}
%\caption{Simulation results.}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat[]), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals use top floats exclusively.
% Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
% floats. This can be corrected via the \fnbelowfloat command of the
% stfloats package.

\section{Conclusion}
The conclusion goes here.~\cite{goldman:2013:cgpwaste}

% if have a single appendix:
%\appendix[Proof of the Zonklar Equations]
% or
%\appendix  % for no appendix heading
% do not use \section anymore after \appendix, only \section*
% is possibly needed

% use appendices with more than one appendix
% then use \section to start each appendix
% you must declare a \section before using any
% \subsection or using \label (\appendices by itself
% starts a section numbered zero.)
%

\appendices
\section{Example Appendix}
Appendix one text goes here.

% you can choose not to have a title for an appendix
% if you want by leaving the argument blank
\section{}
Appendix two text goes here.

% use section* for acknowledgement
\section*{Acknowledgment}

The authors would like to thank...

% Can use something like this to put references on a page
% by themselves when using endfloat and the captionsoff option.
\ifCLASSOPTIONcaptionsoff
  \newpage
\fi

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
%\begin{thebibliography}{1}

%\bibitem{IEEEhowto:kopka}
%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.

%\end{thebibliography}

\bibliographystyle{IEEEtran}
\bibliography{IEEEabrv,../main}

% biography section
% 
% If you have an EPS/PDF photo (graphicx package needed) extra braces are
% needed around the contents of the optional argument to biography to prevent
% the LaTeX parser from getting confused when it sees the complicated
% \includegraphics command within an optional argument. (You could create
% your own custom macro containing the \includegraphics command to make things
% simpler here.)
%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
% or if you just want to reserve a space for a photo:

\begin{IEEEbiography}{Brian W. Goldman}
TODO
\end{IEEEbiography}

% if you will not have a photo at all:
\begin{IEEEbiographynophoto}{William F. Punch}
TODO
\end{IEEEbiographynophoto}

% insert where needed to balance the two columns on the last page with
% biographies
%\newpage

% You can push biographies down or up by placing
% a \vfill before or after them. The appropriate
% use of \vfill depends on what kind of text is
% on the last page and whether or not the columns
% are being equalized.

%\vfill

% Can be used to pull up biographies so that the bottom of the last one
% is flush with the other column.
%\enlargethispage{-5in}

% that's all folks
\end{document}

### 2013-03-11 CGP Journal/main.tex
Added outline of CGP journal paper, removed windows line endings

@@
 
+\section{Cartesian Genetic Programming}
+
+\section{Duplicate Evaluation Avoidance}
+
+\subsection{Skip}
+
+\subsection{Accumulate}
+
+\subsection{Single}
+
+\section{Genome Ordering}
+
+\subsection{Reorder}
+
+\subsection{DAG}
+
+\section{Qualitative Comparisons}
+
+TODO All possible combinations of Skip, Accumulate, Single, with Normal, Reorder, Dag.
+
+\subsection{Problem Set}
+TODO 3-bit Parity, 6-bit Multiply, 4-16-bit Decode, 16-4-bit Encode
+
+TODO Consider adding 5-bit Parity or Equality
+
+\subsection{Parameter Setting}
+TODO Mutation Rate / Genome Size grid
+
+TODO Method for choosing parameter values.
+
+\subsection{Results}
+TODO Compare using a table the best parameters found for each configuration.
+One table per problem.
+
+\subsection{Generalization}
+TODO Consider dropping this section due to computational limits.
+
+\section{Analysis of Evolutionary Mechanisms}
+TODO Explain footprint concept here.
+\subsection{Inactive Nodes and Node Duplication}
+TODO Look at how much of the genome is duplicated in both active and inactive
+
+TODO Look at how much of the genome is unused.
+
+\subsection{Node Reactivation and Active Node Mutation}
+TODO For each node that was active, then inactive, then active again, see how its changed.
+
+TODO For each node that is active, look at how successful mutations change its behavior.
+
+\subsection{Source of High Variance}
+TODO Look for general indicators of high evaluation runs.
+
+TODO Look for footprint moving.
+
+TODO Look for active node clustering.
+
+TODO Look into how often children replace parents.
+
+\section{Conclusions}
+
 % An example of a floating figure using the graphicx package.

### 2013-03-13 CGP Journal/main.tex
Some initial methodology dump

@@
 
+TODO Explain here why only compared with parent, not global history.
+
 \subsection{Skip}
+The \emph{Skip} method for avoiding duplicate evaluations involves the least
+amount of modification from CGP's normal behavior.  After an offspring is produced
+and its set of active nodes is determined but before it is evaluated, it is compared
+with its parent.  Each gene in each active node is compared for equivalence with
+the corresponding gene in the offspring's parent.  If all genes are found to be equal
+than the individual is not evaluated.  Instead the offspring is given the same fitness
+as its parent, as there is no way for it to be phenotypically different from its parent.
+Note that since inactive genes are not compared for equivalence it is still possible
+for offspring to genetically differ from their parents.
+
+Because \emph{Skip} does not modify any evolutionary mechanisms from normal CGP,
+in all possible cases it can do no worse.  If the mutation rate is high enough relative
+to the number of active genes, \emph{Skip} will act identically to normal CGP as
+the probability of an offspring being actively identical to its parent is effectively zero.
+When the mutation rate is low enough relative to the number of active genes, some number of
+offspring will be actively identical to their parents, resulting in a reduction in evaluations
+but no change in the evolutionary trajectory.
+
+From a tuning perspective we would expect both intuitively and from initial
+experimentation~\cite{goldman:2013:cgpwaste} that \emph{Skip} will be less sensitive
+to the mutation rate and will perform best using a lower mutation rate than normal CGP.
+The former comes from the fact that in normal CGP any mutation rate that has a significant
+probability of creating offspring actively identical to their parents will be penalized
+by wasting evaluations.  In \emph{Skip} there are only two penalties for reducing the
+mutation rate.  First, if search becomes trapped in a local optima, low mutation
+rates may have increased difficulty escaping to find the global optimum.  Second,
+exceptionally low mutation rates may result in large computational overhead as
+more and more time is spent attempting to produce an evaluable useful offspring.
+Note these penalties also exist in normal CGP.
 
 \subsection{Accumulate}
+Similar to \emph{Skip}, \emph{Accumulate} works by adding a step between offspring
+creation and offspring evaluation.  Instead of skipping evaluations with offspring
+are determined to be actively identical to their parents, \emph{Accumulate} enters
+into a cycle of repeated mutation until an individual worth evaluating is created.
+While the working offspring is actively identical to the
+parent, a new individual is produced by applying mutation to the working offspring.
+Note that each mutation step can probabilistically mutate any number of genes, just
+as in the mutation used by normal CGP.
+In this way each iteration the working offspring accumulates mutations to inactive
+genes until an individual is produced that differs by one or more active genes from
+the parent.  This repeatedly mutated offspring is then evaluated.  If its fitness
+is less than the original parent, the final set of mutations are reverted
+and the offspring's fitness is set to the same as the original parent.
+
+Viewed in another way, \emph{Accumulate} performs a micro evolution on each offspring
+before it is fully produced.  The final result of each offspring step is to produce
+an offspring which is either actively identical to its parent, but with mutations
+to its inactive genes, or an offspring which is at least as fit as the parent but
+not actively identical.
+
+Previous experimentation has shown that \emph{Accumulate} acts very similarly to
+\emph{Skip}~\cite{goldman:2013:cgpwaste}, with the exception that \emph{Accumulate}
+favors lower mutation rates. While at first these algorithms may appear
+quite different, further consideration shows how they are similar.  In \emph{Skip}
+an actively identical offspring is likely to be selected as we expect mutations to
+active genes to more often reduce fitness than in prove it.  In the next generation
+this actively identical offspring is then mutated again, with its lineage likely
+continuing until it finally does mutate an active gene.  At this point if the
+mutant is better it replaces the parent, otherwise the parent, which has been accumulating
+mutations to inactive genes, is kept.  In this way \emph{Skip} mirrors \emph{Accumulate},
+except over multiple generations instead of compressed into a single generation.
 
 \subsection{Single}
+While \emph{Skip} and \emph{Accumulate} focus on what to do when offspring are
+produced which are actively identical to their parents, \emph{Skip} changes
+how mutation works to ensure only evaluable offspring are created.  Instead
+of mutating each gene at a set probability, \emph{Single} chooses genes at
+random to mutate until exactly one active gene is mutated.
+
+This modification gives \emph{Single} three properties distinct from the other
+forms of duplicate avoidance.  First, \emph{Single} forces offspring to have
+an active gene which is different than there parent.  This limits drift as the
+mutated active gene must either be to an intron or represent another way to
+code a solution of equal quality.  Second, as a benefit of forced changes
+\emph{Single} avoids the overhead of repeatedly generating individuals without
+creating an evaluable offspring.  Third, \emph{Single} does not require a mutation
+rate parameter, effectively setting the mutation rate to $\frac{1}{a}$ for active
+genes and $\frac{1}{a+1}$ for inactive genes, where $a$ is the number of active genes.
+
+In encodings without inactive genes, limiting mutation to changing exactly one gene
+could prevent an algorithm from escaping some types of local maxima.  Yet because
+CGP allows for inactive genes, \emph{Single} is still able to escape most local
+maxima using sufficient drift of inactive genes.  When a high percentage of
+the genome becomes active or when \emph{Single} is otherwise limited in its ability
+to drift due to lack of introns, it does have an increased potential to become stuck.
+As will be discussed in Section~\ref{sec-ordering} and was shown in~\cite{goldman:2013:ordering}
+this problem of highly active genomes is very unlikely.
 
 \section{Genome Ordering}
+\label{sec-ordering}
 
@@
 TODO All possible combinations of Skip, Accumulate, Single, with Normal, Reorder, Dag.
+
+TODO Explain why you never use normal CGP mutation.
 

### 2013-04-09 Length Bias and Seach Limitations in CGP/main.tex
Added non anonymous version of my paper, fixed Bill's author information
Updated LengthBias to cover reviewer comments, updated bib to cite the paper correctly

@@
 % --- Author Metadata here ---
-\conferenceinfo{GECCO'13,} {July 6-10, 2013, Philadelphia, Pennsylvania, USA.}
+\conferenceinfo{GECCO'13,} {July 6-10, 2013, Amsterdam, The Netherlands.}
 \CopyrightYear{2013}
@@
 \numberofauthors{2} %
-\begin{comment}
+%\begin{comment}
 \author{
@@
 \alignauthor
-William Punch\\
+William F. Punch\\
        \affaddr{BEACON Center for the Study of Evolution in Action}\\
        \affaddr{Michigan State University, U.S.A.}\\
-       \email{punch@cse.msu.edu}
+       \email{punch@msu.edu}
 }
 %\end{comment}
-%\begin{comment}
+\begin{comment}
 \author{
@@
 % A category with the (minimum) three required fields
-\category{Computing Methodologies}{Artificial Intelligence}{Search Methodologies}
+%\category{Computing Methodologies}{Artificial Intelligence}{Search Methodologies}
+\category{I.2.8}{Artificial Intelligence}{Problem Solving, Control Methods, and Search}
+\category{I.2.2}{Artificial Intelligence}{Automatic Programming}
 %A category including the fourth, optional field follows...
@@
 \keywords{Cartesian Genetic Programming, Neutrality, Bloat}
-\\
-\\
+
 \section{Introduction}
@@
 A complete package of our results as well as the source code used in testing
-is available from our website.\footnote{\url{https://name.of.site.removed.com/anonymous-CGP}}
+is available from our website.\footnote{\url{https://github.com/brianwgoldman/LengthBiasCGP}}
 
@@
 The median number of evaluations to success is shown in
-Figure~\ref{fig-multiply} on a log-log scale.  The results from the best genome
-size are given in Table~\ref{table-multiply}.
+Figure~\ref{fig-multiply} on a log-log scale.  The runs using the best genome size for each 
+algorithm were then repeated to allow for statistical comparison free from selection bias.
+The results from the repeated runs are given in Table~\ref{table-multiply}.
 
@@
 variants had lower median evaluations to success than \emph{Normal}'s best, which
-also required a genome twenty times larger.  \emph{Reorder} was statistically
-better than \emph{Normal} under the single tailed Mann-Whitney U test. While \emph{DAG} was not
-statistically better than \emph{Normal}, the likely reason is \emph{Normal}'s exceptionally large median
-absolute deviation.  We believe this deviation is caused by \emph{Normal}'s occasional need to
+also required a genome twenty times larger.  A Kruskal-Wallis Test was first used
+to determine that at least one of the three algorithms had a different median than the others,
+with a resulting p-value of 0.00006.
+Both \emph{Reorder} and \emph{DAG} were statistically better than \emph{Normal} under
+the single tailed Mann-Whitney~U test, even after accounting for multiple comparisons
+using the Bonferroni correction.
+Beyond the difference in median evaluations to success, \emph{Normal} also had
+a much higher median absolute deviation.
+We believe this deviation is caused by \emph{Normal}'s occasional need to
 reinvent useful structures in different genome locations.  In the unlikely event that
@@
 Also, more inactive space in the genome may facilitate structure recreation in new locations.
+The higher number of evaluations, higher deviation, and larger optimal genome size
+also combined to make \emph{Normal} take almost three times more computational time
+than either of the other algorithms to complete all runs.
 As a final advantage to \emph{Reorder} and \emph{DAG}, we see that both evolved
@@
 neutral search more than the other methods.
+
+\begin{figure}
+  \centering
+  \includegraphicsfit{multiply}
+  \caption{Median evaluations until success on the Multiply problem using different genome sizes} 
+  \label{fig-multiply}
+\end{figure}
 
@@
 \textbf{Genome} & 4000   & 200     & 200 \\ \hline
-  \textbf{MES} & 272,324 & 213,660 & 249,031 \\ \hline
-  \textbf{MAD} & 125,997 & 79,335  & 86,551 \\ \hline
-  \textbf{P value} & N/A & 0.030   & 0.117 \\ \hline
-  \textbf{Active} & 255  & 118     & 120 \\ \hline
+  \textbf{MES} & 467,621 & 236,438 & 265,997 \\ \hline
+  \textbf{MAD} & 215,993 & 82,966  & 66,930 \\ \hline
+  \textbf{P-value} & N/A & 0.00012 & 0.00006 \\ \hline
+  \textbf{Active} & 245  & 115     & 116 \\ \hline
 
@@
 
-\begin{figure}
-  \centering
-  \includegraphicsfit{multiply}
-  \caption{Median evaluations until success on the Multiply problem using different genome sizes} 
-  \label{fig-multiply}
-\end{figure}
-
 \section{Conclusions and Future Work}
 \label{sec-conclusion}
-Existing literature suggests CGP performs best when maximum genome sizes
-far exceed actual solution size~\cite{miller:2006:redundancy}, supported by
+Existing literature suggests CGP performs best when the maximum genome size
+far exceeds the actual solution size~\cite{miller:2006:redundancy}, supported by
 evidence that inactive genes improve neutral search~\cite{yu:2001:neutrality},
@@
 and powerful, but wider testing needs to be performed to solidify that claim.  As
-an important aspect of those tests, run time comparisons need to be performed, as
+an important aspect of those tests, run time comparisons using equal genome sizes need to be performed, as
 both increase the computational expense of CGP, with \emph{DAG} especially having

### 2013-04-10 Length Bias and Seach Limitations in CGP/main.tex
Minor fixes

@@
 the question remains; how will the maximum genome size effect evolution's
-speed at finding solutions.  As larger genomes inherently represent larger search
+speed at finding solutions?  As larger genomes inherently represent larger search
 spaces, intuition suggests that using the lowest possible maximum size will have the most efficient search.
@@
 DAGs of size \emph{N}, where \emph{N} is the number of nodes in the genome.
+
 %
@@
 a single connection gene, we would expect random individuals to have a length of $\log_2N$.
-Precise calculation of the expected length is made difficult
+Precise calculation of the expected length when using more connection genes is made difficult
 due to node overlap, but the relationship will remain logarithmic to the total genome size.
@@
 biases is that each is most likely to find solutions fastest when the number of
-active nodes desired by the problem best matches the algorithms search bias, as
+active nodes desired by the problem best matches the algorithm's search bias, as
 otherwise the algorithm will add artificial pressure to search different sized

### 2013-05-03 CGP Journal/main.tex
Added description of how Reorder works, and pseudo code

@@
 %
-%\usepackage{algorithmic}
+\usepackage{algorithmic}
 % algorithmic.sty was written by Peter Williams and Rogerio Brito.
@@
 
+\usepackage{verbatim}
 \begin{document}
@@
 \subsection{Reorder}
+The concept behind \emph{Reorder} is to randomly shuffled node ordering without
+impacting gene expression.  This is possible because for a given node, there can
+be a large number of other nodes which neither have no required ordering,
+as they neither directly or indirectly take input from the node nor provide their
+output to the node.  In general, this process works by assigning nodes new
+locations in the genome at random once all of the nodes they take input from have been assigned
+locations earlier in the genome.
+
+The first step in performing \emph{Reorder} on a genome is to create convenience
+data structures to store direct connection relationships, such that given a node
+we can determine in constant time where it gets its input and which nodes read
+its output.  These structures can be build in order $O(N)$ time, where $N$ is the
+number of nodes.
+
+With this preprocessing complete, the algorithm given in Figure~\ref{fig:reorder}
+is used to assign each node a new location in the modified genome.
+It starts by constructing the \emph{addable} set, which contains all nodes who's direct
+dependencies have already been added.  Initially this contains only the input locations.
+Once a node is assigned a location, all of the nodes that depend on its output
+have that dependency marked as satisfied.  Once all of a nodes dependencies
+are satisfied, it can be put into \emph{addable}.  In this way, nodes are
+randomly removed from \emph{addable}, assigned the next possible location
+in the genome, with new nodes added to \emph{addable} as it becomes semantically
+viable.  Iteration ends when all nodes have been assigned new locations.
+This algorithm requires $O(AN)$, where $A$ is the arity used by the nodes
+and $N$ is the number of nodes.
+
+\begin{figure}
+  \begin{algorithmic}
+
+    \STATE $addable \leftarrow input\_locations$
+    \STATE $index \leftarrow |input\_locations|$
+    \STATE $new\_loc \leftarrow \emptyset$
+    \WHILE{$|addable| > 0$}
+      \STATE $working \leftarrow randpop(addable)$
+      \IF{$working \notin input\_locations$}
+        \STATE $new\_loc[working] \leftarrow index$
+        \STATE $index \leftarrow index + 1$
+      \ENDIF
+      \FORALL{$link \in feeds\_to(working)$}
+        \STATE Satisfy $link$'s dependence on $working$
+        \IF{$|unsatisfied(link)| = 0$}
+          \STATE $addable \leftarrow addable \cup \{link\}$
+        \ENDIF
+      \ENDFOR
+      \RETURN $new\_loc$
+    \ENDWHILE
+  \end{algorithmic}
+  \caption{Algorithm used by \emph{Reorder} to determine new node locations.}
+  \label{fig:reorder}
+\end{figure}
+
+The final step in \emph{Reorder} is to use the list of new locations to convert
+the existing genome into the reordered genome.  This can be done in the straight
+forward manner by converting connection genes and output location genes using
+the $new_loc$ map returned from the algorithm in Figure~\ref{fig:reorder}.
 
@@
 
+TODO Explain reruning.
+
 \subsection{Results}
@@
 TODO Explain footprint concept here.
+
 \subsection{Inactive Nodes and Node Duplication}

### 2013-05-06 CGP Journal/main.tex
More big blocks of basic stuff

@@
 %
-\usepackage{algorithmic}
+%\usepackage{algorithmic}
+\usepackage{algpseudocode}
 % algorithmic.sty was written by Peter Williams and Rogerio Brito.
@@
 \section{Cartesian Genetic Programming}
+TODO Explain basic CGP, use~\cite{miller:2011:chapter2}.
 
@@
 \label{sec-ordering}
+TODO General stuff about ordering, including normal CGP.
 
 \subsection{Reorder}
+\label{sec:reorder}
 The concept behind \emph{Reorder} is to randomly shuffled node ordering without
@@
   \begin{algorithmic}
-
-    \STATE $addable \leftarrow input\_locations$
-    \STATE $index \leftarrow |input\_locations|$
-    \STATE $new\_loc \leftarrow \emptyset$
-    \WHILE{$|addable| > 0$}
-      \STATE $working \leftarrow randpop(addable)$
-      \IF{$working \notin input\_locations$}
-        \STATE $new\_loc[working] \leftarrow index$
-        \STATE $index \leftarrow index + 1$
-      \ENDIF
-      \FORALL{$link \in feeds\_to(working)$}
-        \STATE Satisfy $link$'s dependence on $working$
-        \IF{$|unsatisfied(link)| = 0$}
-          \STATE $addable \leftarrow addable \cup \{link\}$
-        \ENDIF
-      \ENDFOR
-      \RETURN $new\_loc$
-    \ENDWHILE
+    \Procedure{NewOrder}{}
+    \State $addable \leftarrow input\_locations$
+    \State $index \leftarrow |input\_locations|$
+    \State $new\_loc \leftarrow \emptyset$
+    \While{$|addable| > 0$}
+      \State $working \leftarrow randpop(addable)$
+      \If{$working \notin input\_locations$}
+        \State $new\_loc[working] \leftarrow index$
+        \State $index \leftarrow index + 1$
+      \EndIf
+      \ForAll{$link \in feeds\_to(working)$}
+        \State Satisfy $link$'s dependence on $working$
+        \If{$|unsatisfied(link)| = 0$}
+          \State $addable \leftarrow addable \cup \{link\}$
+        \EndIf
+      \EndFor
+    \EndWhile
+    \State\Return $new\_loc$
+    \EndProcedure
   \end{algorithmic}
@@
 the $new_loc$ map returned from the algorithm in Figure~\ref{fig:reorder}.
+This also requires $O(AN)$ time, as all connection genes in all nodes must
+be converted.  As this is the final step unique to \emph{Reorder}, and
+no previous steps were of higher complexity, it is also the complexity of
+the algorithm as a whole.  Furthermore it is of the same complexity class
+of copying the entire genome, meaning it does not change CGP's overall complexity.
+
+\emph{Reorder} is used once each generation to shuffle the nodes of the parent
+organism.  As the shuffling does not semantically change the parent, it does
+not need to be reevaluated.  Furthermore, as shuffling does not change the requirement
+that nodes only depend on those preceding them in the genome, shuffling does not
+require any changes to any other CGP methods, such as evaluation and mutation.
+Yet after shuffling, the potential mutations that can be applied to the parent
+have changed.  Consider two nodes $X$ and $Y$ which are in the genome, but
+have no dependence relationship.  As $X$ does not take input from $Y$ and
+$Y$ does not take input from $X$, either can be placed preceding the other
+in the genome.  Without shuffling, the ordering in which these two nodes originally
+evolved will be preserved indefinitely, such that if $X$ precedes $Y$ at node creation,
+$X$ will always precede $Y$.  No mutation in \emph{Normal} is capable of making $X$
+take input from $Y$.  Yet CGP using \emph{Reorder} is capable of shuffling $Y$ to
+precede $X$, allowing subsequent mutations to make the connection.
+
+This node movement has the potential to reduce node reinvention.  If evolution created
+$X$ and $Y$, and mutating $X$ to depend on $Y$ would improve fitness, mutation's
+ability to make that improvement would depend on happenstance.  If $X$ evolved
+preceding $Y$, mutation would have to recreate either $X$ or $Y$ in new locations
+in order to make the required connection.  \emph{Reorder} could achieve the same
+effect through shuffling, and without costly node reinvention.
+
+The price \emph{Reorder} pays for this potential improvement is increased mutational search
+space and minor runtime costs.  Even though \emph{Reorder} is not capable of representing
+different individuals that \emph{Normal}, their mutational adjacency is significantly different.
+By mutational adjacency, we mean the set of individual's which are most likely to be created
+when mutating a given individual.  For instance in \emph{Normal} if $X$ precedes $Y$, individuals
+containing $Y$ dependent on $X$ could be adjacent to the original individual, but those containing
+$X$ dependent on $Y$ would not.  As \emph{Reorder} can shuffle node order, both sets would be
+adjacent.  In this way ordering limits how \emph{Normal} performs its search.  This limitation
+will benefit \emph{Normal} any time making $X$ dependent on $Y$ is detrimental, as
+\emph{Normal} won't waste evaluations on individuals having that mutation.
 
 \subsection{DAG}
+The principle behind \emph{DAG} is to remove the traditional CGP requirement that
+nodes can only take input from nodes that precede them in the genome.  Instead
+this is replaced with the relaxed requirement than the genome cannot encode for
+any cycles.  To accomplish this goal, modifications must be made to how CGP
+performs mutation and evaluation.
+
+When a connection gene is chosen for mutation, its value is still chosen randomly
+from all possible alternative values.  To achieve this, we incrementally determine
+which nodes transitively depend on the mutating node and which are not, with
+information stored in the $known$ map.  Initially, we know that the node
+is dependent on itself and that input locations and the nodes current inputs
+are not dependent on the mutating node.  Nodes are then
+tested in a random order for dependence using the algorithm given in Figure~\ref{fig:dag}.
+The first node returned that is not dependent on the mutating node is then
+used as the new connection gene value.
+
+\begin{figure}
+  \begin{algorithmic}
+  \Procedure{IsDependent}{$working$, $known$}
+    \If{$working \in known$}
+      \State\Return{$known[working]$}
+    \EndIf
+    \ForAll{$link \in reads\_from(working)$}
+      \If{\Call{IsDependent}{$link$, $known$}}
+        \State $known[working] \leftarrow True$
+        \State\Return{$True$}
+      \EndIf
+    \EndFor
+    \State $known[working] \leftarrow False$
+    \State\Return{$False$}
+  \EndProcedure
+  \end{algorithmic}
+  \caption{Algorithm used by \emph{DAG} to determine valid connection gene values.}
+  \label{fig:dag}
+\end{figure}
+
+The algorithm given in Figure~\ref{fig:dag} is a non-repeated recursive depth
+first search of the individual's DAG.  Search terminates as soon as a dependent
+node is found, and information from previous searches is recorded to prevent
+repeated search.  Each recursion level results in a node having its dependency
+set, meaning early calls will likely result in lots of recursion, but subsequent
+calls will find answers faster.  In the worst case, this algorithm may need
+to examine all nodes in the genome to determine if $working$ is dependent on the
+mutating node, giving it a complexity of $O(N)$.  In the worst case, this algorithm
+may be called on all possible nodes in the genome to check their dependence.
+Yet cumulatively those calls can only take $O(N)$ time due to the non-repeating
+nature of the search.  As a result, the expected worst case run time for \emph{DAG}
+mutation is $O(mAN^2)$, as when $m$ is the mutation rate we expect $mAN$
+connection genes to be mutated, each requiring a worst case of $N$ dependency checks.
+This is in contrast to \emph{Normal} and \emph{Reorder} which require $O(mAN)$ to perform
+mutation, as individual mutations can be performed in constant time.
+
+The method for evaluating \emph{DAG} individuals is very similar to the method for
+efficiently evaluating \emph{Normal} and \emph{Reorder} individuals.  Just as before,
+preprocessing is done to determine which nodes are active, and only those nodes are
+executed during evaluation.  Just as before the process begins from the output locations,
+recursively following connection genes and marking nodes as active until the input locations
+are reached.  Here the process diverges, as in \emph{DAG} we must determine not only the set of
+nodes that are active, but the order in which those nodes should be executed.
+
+The algorithm given in Figure~\ref{fig:reorder} can be reused to determine the order
+in which nodes should be executed during evaluation.  Instead of using $new\_loc$
+to reorder the genome, we can use this map to specify what order nodes should be
+executed in.  For instance, if a node was given the new location of $X$, we know
+that once all nodes at locations preceding $X$ have been evaluated, the node at $X$
+can be evaluated.  Stripping the inactive nodes from $new\_loc$ and inverting the map
+results in an efficient, valid ordering in which to execute the nodes.
+
+The process of determining which nodes are active requires $O(AN)$ time, regardless
+of how the node ordering is handled.  As was discussed in Section~\ref{sec:reorder},
+getting the $new\_loc$ map requires $O(AN)$ time.  Finally, converting the map
+into an ordering of the active nodes takes $O(N)$ time.  Each of these steps are
+sequential, meaning \emph{DAG} does not increase the runtime complexity of evaluation
+preprocessing.  Furthermore, once the preprocessing step is done, \emph{DAG} takes
+identical time to evaluate than the other methods, regardless of the number of
+input combinations the individual is evaluated on.
+
+With these modifications, there is no longer any meaning to the position of a node
+in the genome.  In \emph{Normal} a node near the input locations will always
+have a highly restricted set of values.  More completely, the number of valid
+values for a connection gene in \emph{Normal} is related to the number of nodes
+preceding that gene, and independent of the number of nodes following that gene.
+In \emph{DAG} a connection gene's valid values only depends on the current solution
+represented by the individual, independent of the genes location in the genome.
+As such, there are far less artificial limitations on how \emph{DAG} can modify
+solutions.
+
+While both reduce mutational limitations, \emph{DAG}'s changes are more broad that
+\emph{Reorder}.  While both have the possibility to use mutation to connect any two
+nodes such that no cycle is created, \emph{Reorder} still maintains a bias about
+how often some types of connection can happen.  Consider again two nodes $X$ and $Y$,
+such that neither is dependent on the other.  If $X$ is transitively dependent on very
+few nodes, and $Y$ is transitively dependent on many nodes, in general \emph{Reorder}
+will shuffle the genome such that $X$ is before $Y$.  This is true because the probability
+that all of $X$'s dependencies are added at any point of building the new order is always
+higher than for $Y$'s dependencies.  As a result we expect $X$ to be put into $addable$ sooner,
+and as a result we expect it to be added sooner.  The greater the discrepancy in these
+set sizes the less likely the order between the nodes will be reversed.
+Conversely, \emph{DAG} has no such bias.  It is always possible to mutate $X$ to connect to $Y$.
+
+The probability of connecting $X$ to $Y$ is also influenced by how many nodes transitively depend on each.
+In this example, let us assume $X$ has more nodes transitively dependent on it than $Y$ has.
+In \emph{Reorder}, $X$ would again appear closer to the start of the genome, this time because the number
+of possible nodes preceding it is reduced.  \emph{DAG} is biased in the reverse, such that $X$ will
+have a higher chance of connecting to $Y$ that $Y$ to $X$.  This is because $X$ can connect to fewer
+nodes, so each mutation has a higher chance of choosing $Y$.
 
@@
 \subsection{Parameter Setting}
-TODO Mutation Rate / Genome Size grid
+In order to fairly compare each method, we should ensure proper parameter configuration.
+This is to avoid the potential criticism that the parameter settings chosen benefit
+a specific method more than others.  While tuning can lead to the alternative issue
+of methods only being effective after extensive problem specific tuning, we set out to use
+rough grained values and provide each technique with equal tuning time to alleviate that bias.
 
-TODO Method for choosing parameter values.
+We focus on two parameter values in our configuration: mutation rate and genome size.
+While there are other potential configurable parameters (population size, offspring size, etc),
+we feel the CGP literature has converged on settings for those parameters (1, 4).
+Furthermore, as duplication detection explicitly deals with the relationship between
+mutation rate and number of active genes, and the number of active genes depends
+on the genome size and ordering method, mutation rate and genome size seem to be
+the most likely to impact results.
+
+To set parameter values, we started by defining a grid of parameter values:
+50, 100, 200, 500, 1000, 2000, 5000, 10000 for genome size and 0.05, 0.02, 0.01,
+0.005, 0.002, 0.001, 0.0005, 0.0002 for mutation rate.  Note that here a genome
+size of 50 means there are 50 nodes, each composed of multiple genes, plus the
+required genes to specify output locations.  Also note that mutation rate of 0.05
+is used here to mean each gene has a 5\% probability of being mutated in a single
+application of mutation.  The grid is formed by trying all possible mutation rates
+with all possible genome sizes, resulting in 64 potential configurations for each
+of the algorithmic configurations.  Finally note that as \emph{Single} does not
+use a mutation rate, only the eight configurations with different genome sizes were
+used when \emph{Single} was employed.
+
+To choose which parameter configuration to use with each algorithmic configuration
+for each problem, we used an iterative process of performing runs, comparing means, removing configurations,
+and repeating.  For each algorithm configuration and problem, we first run all
+potential parameter configurations 5 times (320 runs if not using \emph{Single}, 40 otherwise).
+The parameter configurations are then sorted based on their median.  The best half are then
+each run 4 more times, with the best half of that group chosen.  In this way, 32 configurations
+are run 5 times, 16 are run 9 times, 8 are run 13 times, 4 are run 17 times, 2 are run 21 times,
+and 2 are run 25 times.  For configuration with single, all 8 parameter configurations are run 17 times
+before any are removed.  This iterative reduction allows us to focus tuning on
+those parameter configurations most likely to be effective, with more information
+gathered before choosing between high quality configurations.  In total this means
+3924 complete runs are required to set the parameter configuration for all
+algorithm configurations for a single problem.
 

### 2013-05-07 Reducing Wasted Evaluations in CGP/main.tex
Updated gitignore, added "make double" functionality
Added non anonymous version of my paper, fixed Bill's author information
Escaped a bunch of stuff in the bib file, added BGGB journal, fixed citation style

@@
 \documentclass[runningheads,a4paper]{llncs}
-
+%\linespread{2.5}
 \usepackage{amssymb}
@@
 %
-\author{Brian W.~Goldman\and William Punch}
+\author{Brian W.~Goldman\and William F. Punch}
 %\author{Anonymous Author\and Anonymous Author}
@@
 Michigan State University, U.S.A.\\
-brianwgoldman@acm.org, punch@cse.msu.edu}
+brianwgoldman@acm.org, punch@msu.edu}
 %\institute{Department\\ Organization\\ Email}
@@
 active nodes can change from parent to child.
-There have been numerous studies on the usefulness of inactive nodes~\cite{vassilev2000neutrality,yu:2001:neutrality}
+There have been numerous studies on the usefulness of inactive nodes~\cite{vassilev:2000:neutrality,yu:2001:neutrality}
 including an argument that it is desirable to have up

### 2013-05-07 CGP Journal/main.tex
Fill in of the Qualitative comparisons and Problem Set section

@@
 \section{Qualitative Comparisons}
-
-TODO All possible combinations of Skip, Accumulate, Single, with Normal, Reorder, Dag.
-
-TODO Explain why you never use normal CGP mutation.
+While it is possible to theorize the effects of each CGP variant on search, and
+how those effects may interact with each other, without empirical evidence it is
+mostly speculation.  Therefore we propose to start by testing the effectiveness
+of the variants and their combinations in problem solving.  Deeper analysis of
+the causes of their effectiveness are given in Section~\ref{sec:analysis}.
+
+All told we have proposed three new methods for avoiding duplicate evaluations
+(\emph{Skip}, \emph{Accumulate}, and \emph{Single}) and two new methods for
+how to deal with genome ordering (\emph{Reorder} and \emph{DAG}).  As these two
+sets are non overlapping, and there is no intuitive groupings, we chose to try
+all possible combinations.  As a control we included \emph{Normal} ordering,
+as it provides insight to how each of the duplicate prevention techniques work
+if applied alone.  Similar measures are not needed in testing the ordering techniques,
+as \emph{Skip} has no impact on evolutionary search, and is a strict improvement
+to using no duplication detection at all.
+
+In total this creates 9 algorithmic configurations.  Of these previous work
+has been done testing the duplication methods with \emph{Normal} ordering~\cite{goldman:2013:cgpwaste}
+as well as analysis of all three ordering methods with \emph{Single}~\cite{goldman:2013:ordering}.
+This means there are 4 novel combinations, beyond this being the first time a cross
+comparison has been performed.  This is also the first time the ordering methods
+have been rigorously tested, as much of the work in~\cite{goldman:2013:ordering}
+was analysis oriented.
 
 \subsection{Problem Set}
-TODO 3-bit Parity, 6-bit Multiply, 4-16-bit Decode, 16-4-bit Encode
-
-TODO Consider adding 5-bit Parity or Equality
+In order to provide test landscapes for evolution, we chose problems common to
+CGP literature and previous work with the duplication prevention and ordering
+techniques.  We have chosen 4 binary circuit problems, as the binary representation
+allows for some of the precise analysis performed in Section~\ref{sec:analysis}.
+For all problems, we used the function set \{and, or, nand, nor\}.
+To cover a range of different binary problems, each of the four chosen have
+different numbers of inputs, different number of outputs, and different levels
+of difficulty.
+
+The first problem, 3-bit parity, represents probably the most
+common historic application for
+CGP~\cite{yu:2001:neutrality,miller:2006:redundancy,walker:2008:cgpmodules}.
+We include this problem purely for backward comparison, as we agree that in general
+it is too simple of a problem to merit its own conclusions~\cite{white:2013:bggb}.
+Yet as part of the group it may help yield understanding about how each variant
+performs on simple problems.
+
+As representatives of binary problems with varying input and output sizes, we
+reuse the 4-16-bit Decode and 16-4-bit Encode proposed in~\cite{goldman:2013:cgpwaste}.
+In these problems, the circuit evolved must either convert a 4-bit encoded integer
+to a 1 on the corresponding output line (Decode), or take a 1 on one of the 16
+input lines and convert it back into a 4-bit encoded integer (Encode).  These
+problems share properties with the commonly used Multiplexer problem, but include
+multiple output lines, and can be evaluated quicker as each only requires 16 possible
+test points.
+
+Finally we chose the 3-bit Multiply to represent hard binary problems,
+as was suggested in~\cite{white:2013:bggb} and used
+by~\cite{vassilev:2000:neutrality,miller:2006:redundancy,walker:2008:cgpmodules}.
+This problem is very difficult by comparison to the other problems, and has
+the largest number of test points.
+
+TODO Consider adding 5-bit Parity or Equality, 11-bit Multiplexer.
 
@@
 \section{Analysis of Evolutionary Mechanisms}
+\label{sec:analysis}
 TODO Explain footprint concept here.

### 2013-05-07 CGP Journal/main.tex
Fixed duplicated conclusion
Gave the paper a working title, more fill in

@@
 % Do not put math or special symbols in the title.
-\title{TODO Title}
+\title{Analysis of Cartesian Genetic Programming's Evolutionary Mechanisms}
 %
@@
 is used here to mean each gene has a 5\% probability of being mutated in a single
-application of mutation.  The grid is formed by trying all possible mutation rates
+application of mutation.  These values were chosen as they completely cover the
+range of previously used parameter settings, with each value having at least
+one paired value for the other parameter which is feasibly effective.
+The grid is formed by trying all possible mutation rates
 with all possible genome sizes, resulting in 64 potential configurations for each
@@
 
-TODO Explain reruning.
+To ensure termination, all runs were limited to 10,000,000 evaluations, well beyond
+the expected time to completion.  As the median was used to compare configurations,
+and not the mean, selection should be insensitive to this limit unless a large
+portion of runs are failing to optimize.
+
+To prevent undue biasing in the final results, none of these tuning runs were
+used for any other data analysis.  This helps reduce selection's effect on
+result distribution, similar to how it is common to split training and testing
+runs.  In this case, the runs used to set the parameter configuration are considered
+the training runs for that configuration.  As a result, our final results were gather
+from a completely independent set of runs to test the configurations general characteristics.
+Each configuration was run 50 times to ensure statistical power.
 
@@
 \label{sec:analysis}
-TODO Explain footprint concept here.
+Comparing how many evaluations each algorithm takes to solve a set of benchmark
+problems only provides us with a limited amount of understanding about how
+those algorithms actually function.  We propose here to look for further detail,
+both to check on theorized capabilities of each variant and CGP in general, and
+to look for previously hidden aspects of CGP.
+To do so we have devised a number of novel metrics to examine what makes a
+CGP run successful.
+
+As a primary concept to our analysis, we determine the \emph{footprint} of nodes
+in the genome.  What we mean by \emph{footprint} is a measure the semantic use
+of a node.  For each input that is given to an individual, all of its nodes
+will produce a single output.  These output values can be recorded an ordered.
+As the domain in use in binary, the outputs can be viewed as a bit string, with
+each bit the value that a node outputs when an individual is presented with a
+specific input.  Similar recording can be done for other domains, but binary
+allows for by far the simplest encoding.
+
+The \emph{footprint} of a node is a complete description of its functionality.
+Any two nodes with identical \emph{footprint} values have identical behavior,
+even if their method for encoding that behavior is drastically different.  From
+this perspective, we can view CGP evolution as attempting to evolve a node who's
+\emph{footprint} matches the desired output \emph{footprint}.  Doing so allows
+us to examine how the evolutionary operators change the behavior of nodes,
+not just their gene values.
+
+Utilizing \emph{footprint} analysis, we can extract the semantically useful
+portions of an individual, a subset of the active nodes.  Consider two active
+nodes with the same \emph{footprint}.  As CGP is a graph representation, only
+one of these nodes is actually necessary.  Given one in the genome, any node that reads
+from the other can have its connection changed without changing its behavior.
+
+TODO Explain simplify algorithm
 
@@
 TODO Look into how often children replace parents.
-
-\section{Conclusions}
 
@@
 \section{Conclusion}
-The conclusion goes here.~\cite{goldman:2013:cgpwaste}
+The conclusion goes here.
 

### 2013-05-15 CGP Journal/main.tex
Added algorithms for simplifying genomes

@@
 
+\subsection{Scalability}
+TODO Discuss using tuned parameters on different problem sizes and analyze
+results.  Include O(N) table similar to previous paper.
+
 \subsection{Generalization}
@@
 from the other can have its connection changed without changing its behavior.
+
+\begin{figure}
+  \begin{algorithmic}
+  \Procedure{RepeatedFootprint}{}
+    %\State $reachable \leftarrow \{i \mapsto \{ footprint(i) \} | i \in input\_locations\}$
+    \State $reachable \leftarrow \emptyset$
+    \ForAll{$i \in input\_locations$}
+      \State $reachable[i] \leftarrow footprint(i)$
+    \EndFor
+    \State $repeated \leftarrow \emptyset$
+    \ForAll{$node \in active$}
+      \State $working \leftarrow footprint(node)$
+      \State $direct \leftarrow \emptyset$
+      \ForAll{$link \in reads\_from(node)$}
+        \State $direct \leftarrow direct \cup reachable[link]$
+      \EndFor
+      \If{$working \in direct$}
+        \State $repeated \leftarrow repeated \cup \{working\}$
+      \Else
+        \State $direct \leftarrow direct \cup \{working\}$
+      \EndIf
+      \State $reachable[node] \leftarrow direct$
+    \EndFor
+    \State\Return{$repeated$}
+  \EndProcedure
+  \end{algorithmic}
+  \caption{Algorithm to determine which nodes transitively depend on their
+  own footprint.}
+  \label{fig:repeated}
+\end{figure}
+
+\begin{figure}
+  \begin{algorithmic}
+  \Procedure{Simplify}{$p\_r$, $p\_c$, $p\_i$, $best$}
+    \If{$|p\_r| = 0$}
+      \State\Return{$p\_i$}
+    \ElsIf{$|p\_i| + |p\_r| \geq |best|$}
+      \State\Return{$best$}
+    \EndIf
+    \State $working, p\_a \leftarrow randpop(p\_r)$
+    \State $covrd \leftarrow p\_c \cup {working}$
+    \State $ansc \leftarrow p\_a \cup {working}$
+    \ForAll{$node \in has\_footprint(working)$}
+      \State $node\_valid \leftarrow True$
+      \State $incd \leftarrow p\_i \cup {node}$
+      \State $reqd \leftarrow p\_r$
+      \ForAll{$link \in reads\_from(node)$}
+        \State $foot \leftarrow footprint(link)$
+        \If{$foot \in ansc$}
+          \State $node\_valid \leftarrow False$
+          \State \textbf{break}
+        \ElsIf{$foot \notin covrd$}
+          \State $reqd[foot] \leftarrow reqd[foot] \cup ansc$
+        \EndIf
+      \EndFor
+      \If{$node\_valid$}
+        \State $solution \leftarrow \Call{Simplify}{reqd, covrd, incd, best}$
+        \If{$|solution| < |best|$}
+          \State $best \leftarrow solution$
+        \EndIf
+      \EndIf
+    \EndFor
+    \State\Return{$best$}
+  \EndProcedure
+  \end{algorithmic}
+  \caption{Recursive algorithm which determines the minimum set of nodes required
+    to generate a set of required outputs.}
+  \label{fig:simplify}
+\end{figure}
 

### 2013-05-15 CGP Journal/main.tex
Added text explanation of the simplify algorithm chunks

@@
   \Procedure{RepeatedFootprint}{}
-    %\State $reachable \leftarrow \{i \mapsto \{ footprint(i) \} | i \in input\_locations\}$
     \State $reachable \leftarrow \emptyset$
@@
 \end{figure}
+
+An efficient algorithm to remove some of the redundancy in the genome is given
+in Figure~\ref{fig:repeated}.  This algorithm finds the set of nodes which are
+transitively dependent on their own footprint.  In other words, in order for
+each node in the $repeated$ set, some other node in the genome has to produce
+the same output.  As an example, consider node $Y$ which is in the $repeated$
+set.  There must exist some node $X$ which $Y$ transitively depends on that has
+the same footprint as $Y$.  As a result, $Y$ can always be remove, with anything
+previously dependent on $Y$ now dependent on $X$.  All nodes in the $repeated$
+set can therefore be removed from the genome without effecting its ability to
+calculate any value.  This also has the potential to cause cascading removals
+if there are intermediary nodes which are no longer useful once all of the
+$repeated$ nodes have been removed.  In total this algorithm runs in $O(AN^2)$
+time for an arbitrary individual.  Note that this is an offline algorithm,
+so it only needs to be run on individuals used in production or for analysis.
 
@@
 
-TODO Explain simplify algorithm
+While the previous algorithm is guaranteed to only remove redundant nodes, it
+does not remove all redundant nodes.  In order to remove redundant nodes without
+any transitive relationships, we must use the algorithm given in
+Figure~\ref{fig:simplify}.  This is a recursive algorithm that tries all possible
+choices of nodes that produce each footprint to see which combination results
+in the fewest active nodes.  The algorithm takes in a mapping of footprints
+required by previous calls to all ancestor footprints of those requirements ($p\_r$), footprints previously covered ($p\_c$),
+nodes that have been previously included in the solution ($p\_i$), and smallest
+set of nodes found so far which produces all required footprints ($best$).  Initially
+$p\_c$ contains the set of input footprints, with $p\_r$ mapping each of the output
+footprints to empty sets unless that footprint is already in $p\_c$.  $p\_i$
+is initialized to the empty set, and $best$ starts out as the set of all $active$ nodes.
+Recursion ends when the the nodes in $p\_i$ have no unsatisfied requirements or
+all possible solutions containing $p\_i$ that can satisfy all requirements are
+no better than the current best solution.  While this algorithm is exponential
+in nature, in practice it has a reasonable runtime.  Using the algorithm in
+Figure~\ref{fig:repeated} to prevent $repeated$ nodes from being returned by
+$has\_footprint$ significantly reduces the number of combinations that need
+to be tested.  Furthermore, by preventing recursion from exploring solutions
+worse than the current known best, significant time can be saved.  Lastly, as
+with the algorithm in Figure~\ref{fig:repeated}, this is an offline algorithm,
+and has no impact on evolutionary speed.
+
+Using the algorithm given in Figure~\ref{fig:simplify}, we can determine the
+absolute minimum set of nodes in a genome's solution that are necessary to
+reproduce the original output of that genome.  This allows us to discuss
+the usefulness of each node as either being part of the minimum solution, a
+duplicate of a node in the minimum solution, or irrelevant to the solution.
+As such we can then examine how redundancy is being used, how frequently useful
+structures exist in the inactive nodes, and how useful nodes are constructed by evolution.
+This algorithm can also be used to construct simplified genomes when combined
+with the algorithm given in Figure~\ref{fig:reorder}.
 

### 2013-05-16 CGP Journal/main.tex
Added data tables

@@
 \usepackage{verbatim}
+\usepackage{color, colortbl}
+\definecolor{Gray}{gray}{0.9}
 \begin{document}
@@
 \subsection{Results}
+
+\begin{table*}
+	\centering
+	\begin{tabular}{|c|c|c|c|c|c|c|c|}
+	  \hline
+\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{p-value} \\ \hline
+\emph{Accumulate} & \emph{Normal} & 10,000 & 0.02 & 682 & 348 & 122 & 0.5602 \\ \hline
+\emph{Accumulate} & \emph{Reorder}&  5,000 & 0.01 & 796 & 434 & 331 & 0.7174 \\ \hline
+\rowcolor{Gray}
+\emph{Skip} & \emph{Normal}       & 10,000 & 0.02 & 823 & 363 & 119 & NA \\ \hline
+\emph{Skip} & \emph{Reorder}      &  2,000 & 0.01 & 870 & 362 & 198 & 0.3125 \\ \hline
+\emph{Accumulate} & \emph{Dag}    &  5,000 & 0.01 & 964 & 499 & 1,174 & 0.3310 \\ \hline
+\emph{Skip} & \emph{Dag}          & 10,000 & 0.02 & 1,042 & 620 & 2,083 & 0.0248 \\ \hline
+\emph{Single} & \emph{Normal}     & 500 & NA & 1,214 & 497 & 39 & 0.0046 \\ \hline
+\emph{Single} & \emph{Reorder}    & 200 & NA & 1,506 & 823 & 57 & 0 \\ \hline
+\emph{Single} & \emph{Dag}        & 100 & NA & 2,569 & 1,313 & 43 & 0 \\ \hline
+	\end{tabular}
+	\caption{Parity, 5.9956882435042246e-14}
+	\label{tab:parity}
+\end{table*}
+
+\begin{table*}
+	\centering
+	\begin{tabular}{|c|c|c|c|c|c|c|c|}
+	  \hline
+\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{p-value} \\ \hline
+\rowcolor{Gray}
+\emph{Skip} & \emph{Normal}        & 10,000 & 0.005 & 16,226 & 7,455 & 283 & NA \\ \hline
+\emph{Skip} & \emph{Reorder}       & 10,000 & 0.005 & 17,883 & 8,378 & 2,070 & 0.6918 \\ \hline
+\emph{Accumulate} & \emph{Reorder} & 10,000 & 0.002 & 18,132 & 8,926 & 2,027 & 0.8822 \\ \hline
+\emph{Accumulate} & \emph{Normal}  & 10,000 & 0.005 & 21,428 & 6,379 & 279 & 0.2525 \\ \hline
+\emph{Accumulate} & \emph{Dag}     & 10,000 & 0.002 & 24,855 & 12,512 & 4,059 & 0.0736 \\ \hline
+\emph{Skip} & \emph{Dag}           & 10,000 & 0.002 & 25,678 & 12,034 & 4,178 & 0.0211 \\ \hline
+\emph{Single} & \emph{Reorder}     & 100 & NA & 26,558 & 7,858 & 52 & 0.0076 \\ \hline
+\emph{Single} & \emph{Normal}      & 2,000 & NA & 31,057 & 14,475 & 141 & 0.0085 \\ \hline
+\emph{Single} & \emph{Dag}         & 100 & NA & 34,208 & 10,069 & 59 & 0 \\ \hline
+	\end{tabular}
+	\caption{Encode 1.1926088284917749e-06}
+	\label{tab:encode}
+\end{table*}
+
+\begin{table*}
+	\centering
+	\begin{tabular}{|c|c|c|c|c|c|c|c|}
+	  \hline
+\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{p-value} \\ \hline
+\emph{Skip} & \emph{Reorder}       & 2,000 & 0.002 & 62,853 & 2,023 & 757  & 0.5247 \\ \hline
+\emph{Single} & \emph{Reorder}     & 500 & NA & 63,731 & 13,119 & 245 & 0.3610 \\ \hline
+\emph{Accumulate} & \emph{Normal}  & 1,000 & 0.002 & 67,390 & 11,991 & 211 & 0.8767 \\ \hline
+\rowcolor{Gray}
+\emph{Skip} & \emph{Normal}        & 1,000 & 0.002 & 68,231 & 15,115 & 202 & NA \\ \hline
+\emph{Single} & \emph{Normal}      & 1,000 & NA & 68,819 & 14,590 & 203 & 0.7695 \\ \hline
+\emph{Accumulate} & \emph{Reorder} & 2,000 & 0.002 & 72,991 & 18,650 & 760 & 0.1868 \\ \hline
+\emph{Skip} & \emph{Dag}           & 10,000 & 0.001 & 132,821 & 29,432 & 4,983 & 0 \\ \hline
+\emph{Accumulate} & \emph{Dag}     & 5,000 & 0.002 & 133,188 & 29,326 & 2,542 & 0 \\ \hline
+\emph{Single} & \emph{Dag}         & 500 & NA & 160,696 & 392,242 & 299 & 0 \\ \hline
+	\end{tabular}
+	\caption{Decode 1.0182688124966139e-36}
+	\label{tab:decode}
+\end{table*}
+
+\begin{table*}
+	\centering
+	\begin{tabular}{|c|c|c|c|c|c|c|c|}
+	  \hline
+\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{p-value} \\ \hline
+\emph{Skip} & \emph{Reorder}      & 10,000 & 0.0005  & 196,116  & 60,326  & 3,175 & 0  \\ \hline
+\emph{Accumulate} & \emph{Reorder}& 10,000 & 0.001 & 199,193 & 85,970 & 3,117 & 0  \\ \hline
+\emph{Single} & \emph{Dag} & 200  & NA & 239,844 & 78,739 & 114 & 0 \\ \hline
+\emph{Single} & \emph{Reorder}    & 500   & NA & 243,664 & 91,935 & 266 & 0 \\ \hline
+\emph{Skip} & \emph{Dag}          & 2,000 & 0.002 & 369,931 & 174,408 & 952 & 0.0100 \\ \hline
+\emph{Accumulate} & \emph{Dag}    & 2,000 & 0.002  & 379,202 & 147,289 & 973 & 0.0053  \\ \hline
+\emph{Single} & \emph{Normal}     & 5,000  & NA & 399,911 & 194,384 & 273 & 0.0047 \\ \hline
+\emph{Accumulate} & \emph{Normal} & 2,000 & 0.002 & 483,938 & 227,686 & 168 & 0.45 \\ \hline
+\rowcolor{Gray}
+\emph{Skip} & \emph{Normal} & 2,000 & 0.002 & 657,121 & 311,376 & 175 & NA \\ \hline
+
+	\end{tabular}
+	\caption{Multiply 5.1429910779824196e-18}
+	\label{tab:Multiply}
+\end{table*}
+
 TODO Compare using a table the best parameters found for each configuration.

### 2013-05-16 CGP Journal/main.tex
Fixed unescaped underscore

@@
 forward manner by converting connection genes and output location genes using
-the $new_loc$ map returned from the algorithm in Figure~\ref{fig:reorder}.
+the $new\_loc$ map returned from the algorithm in Figure~\ref{fig:reorder}.
 This also requires $O(AN)$ time, as all connection genes in all nodes must

### 2013-05-17 CGP Journal/main.tex
Updated tables to include "reduced" genome sizes

--- 2013-05-15 CGP Journal/main.tex
@@
 \subsection{Results}
+
+\begin{table*}
+	\centering
+	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
+	  \hline
+\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Reduced} & \textbf{p-value} \\ \hline
+\emph{Accumulate} & \emph{Normal} & 10,000 & 0.02 & 682 & 348 & 122 & 22 & 0.5602 \\ \hline
+\emph{Accumulate} & \emph{Reorder}&  5,000 & 0.01 & 796 & 434 & 331 & 27 & 0.7174 \\ \hline
+\rowcolor{Gray}
+\emph{Skip} & \emph{Normal}       & 10,000 & 0.02 & 823 & 363 & 119 & 23 & NA \\ \hline
+\emph{Skip} & \emph{Reorder}      &  2,000 & 0.01 & 870 & 362 & 198 & 23 & 0.3125 \\ \hline
+\emph{Accumulate} & \emph{Dag}    &  5,000 & 0.01 & 964 & 499 & 1,174 & 26 & 0.3310 \\ \hline
+\emph{Skip} & \emph{Dag}          & 10,000 & 0.02 & 1,042 & 620 & 2,083 & 28 & 0.0248 \\ \hline
+\emph{Single} & \emph{Normal}     & 500 & NA & 1,214 & 497 & 39 & 18 & 0.0046 \\ \hline
+\emph{Single} & \emph{Reorder}    & 200 & NA & 1,506 & 823 & 57 & 20 & 0 \\ \hline
+\emph{Single} & \emph{Dag}        & 100 & NA & 2,569 & 1,313 & 43 & 17 & 0 \\ \hline
+	\end{tabular}
+	\caption{Parity, 5.9956882435042246e-14}
+	\label{tab:parity}
+\end{table*}
+
+\begin{table*}
+	\centering
+	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
+	  \hline
+\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Reduced} & \textbf{p-value} \\ \hline
+\rowcolor{Gray}
+\emph{Skip} & \emph{Normal}        & 10,000 & 0.005 & 16,226 & 7,455  &   283 & 62 & NA \\ \hline
+\emph{Skip} & \emph{Reorder}       & 10,000 & 0.005 & 17,883 & 8,378  & 2,070 & 82 & 0.6918 \\ \hline
+\emph{Accumulate} & \emph{Reorder} & 10,000 & 0.002 & 18,132 & 8,926  & 2,027 & 84 & 0.8822 \\ \hline
+\emph{Accumulate} & \emph{Normal}  & 10,000 & 0.005 & 21,428 & 6,379  &   279 & 62 & 0.2525 \\ \hline
+\emph{Accumulate} & \emph{Dag}     & 10,000 & 0.002 & 24,855 & 12,512 & 4,059 & 88 & 0.0736 \\ \hline
+\emph{Skip} & \emph{Dag}           & 10,000 & 0.002 & 25,678 & 12,034 & 4,178 & 81 & 0.0211 \\ \hline
+\emph{Single} & \emph{Reorder}     & 100    & NA    & 26,558 & 7,858  &    52 & 37 & 0.0076 \\ \hline
+\emph{Single} & \emph{Normal}      & 2,000  & NA    & 31,057 & 14,475 &   141 & 50 & 0.0085 \\ \hline
+\emph{Single} & \emph{Dag}         & 100    & NA    & 34,208 & 10,069 &    59 & 40 & 0 \\ \hline
+	\end{tabular}
+	\caption{Encode 1.1926088284917749e-06}
+	\label{tab:encode}
+\end{table*}
+
+\begin{table*}
+	\centering
+	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
+	  \hline
+\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Reduced} & \textbf{p-value} \\ \hline
+\emph{Skip} & \emph{Reorder}       &  2,000 & 0.002 &  62,853 & 20,023 &   757 & 117 & 0.5247 \\ \hline
+\emph{Single} & \emph{Reorder}     &    500 &    NA &  63,731 & 13,119 &   245 & 100 & 0.3610 \\ \hline
+\emph{Accumulate} & \emph{Normal}  &  1,000 & 0.002 &  67,390 & 11,991 &   211 &  90 & 0.8767 \\ \hline
+\rowcolor{Gray}
+\emph{Skip} & \emph{Normal}        &  1,000 & 0.002 &  68,231 & 15,115 &   202 &  88 & NA \\ \hline
+\emph{Single} & \emph{Normal}      &  1,000 &    NA &  68,819 & 14,590 &   203 &  87 & 0.7695 \\ \hline
+\emph{Accumulate} & \emph{Reorder} &  2,000 & 0.002 &  72,991 & 18,650 &   760 & 118 & 0.1868 \\ \hline
+\emph{Skip} & \emph{Dag}           & 10,000 & 0.001 & 132,821 & 29,432 & 4,983 & 127 & 0 \\ \hline
+\emph{Accumulate} & \emph{Dag}     &  5,000 & 0.002 & 133,188 & 29,326 & 2,542 & 129 & 0 \\ \hline
+\emph{Single} & \emph{Dag}         &    500 &    NA & 160,696 & 39,242 &   299 & 100 & 0 \\ \hline
+	\end{tabular}
+	\caption{Decode 1.0182688124966139e-36}
+	\label{tab:decode}
+\end{table*}
+
+\begin{table*}
+	\centering
+	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
+	  \hline
+\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Reduced} & \textbf{p-value} \\ \hline
+\emph{Skip} & \emph{Reorder}      & 10,000 & 0.0005  & 196,116  & 60,326  & 3,175 & 490 & 0  \\ \hline
+\emph{Accumulate} & \emph{Reorder}& 10,000 & 0.001 & 199,193 & 85,970 & 3,117 & 497 & 0  \\ \hline
+\emph{Single} & \emph{Dag} & 200  & NA & 239,844 & 78,739 & 114 & 77 & 0 \\ \hline
+\emph{Single} & \emph{Reorder}    & 500   & NA & 243,664 & 91,935 & 266 & 134 & 0 \\ \hline
+\emph{Skip} & \emph{Dag}          & 2,000 & 0.002 & 369,931 & 174,408 & 952 & 153 & 0.0100 \\ \hline
+\emph{Accumulate} & \emph{Dag}    & 2,000 & 0.002  & 379,202 & 147,289 & 973 & 162 & 0.0053  \\ \hline
+\emph{Single} & \emph{Normal}     & 5,000  & NA & 399,911 & 194,384 & 273 & 114 & 0.0047 \\ \hline
+\emph{Accumulate} & \emph{Normal} & 2,000 & 0.002 & 483,938 & 227,686 & 168 & 87 & 0.45 \\ \hline
+\rowcolor{Gray}
+\emph{Skip} & \emph{Normal} & 2,000 & 0.002 & 657,121 & 311,376 & 175 & 87 & NA \\ \hline
+
+	\end{tabular}
+	\caption{Multiply 5.1429910779824196e-18}
+	\label{tab:Multiply}
+\end{table*}
+
 TODO Compare using a table the best parameters found for each configuration.

### 2013-05-17 CGP Journal/main.tex
Updated "reduce" values caused by broken simplify algorithm
Changed name on the first algorithm to reflect its general usefulness

@@
   \begin{algorithmic}
-    \Procedure{NewOrder}{}
+    \Procedure{RandomSerialization}{}
     \State $addable \leftarrow input\_locations$
@@
   \end{algorithmic}
-  \caption{Algorithm used by \emph{Reorder} to determine new node locations.}
+  \caption{Algorithm which converts any DAG into a random serial ordering such
+           that all nodes take input from nodes that precede them in the genome.
+           Used by \emph{Reorder} to determine new node locations and \emph{DAG} to
+           determine node evaluation ordering.}
   \label{fig:reorder}
@@
 \textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Reduced} & \textbf{p-value} \\ \hline
-\emph{Accumulate} & \emph{Normal} & 10,000 & 0.02 & 682 & 348 & 122 & 22 & 0.5602 \\ \hline
-\emph{Accumulate} & \emph{Reorder}&  5,000 & 0.01 & 796 & 434 & 331 & 27 & 0.7174 \\ \hline
+\emph{Accumulate} & \emph{Normal} & 10,000 & 0.02 & 682 & 348 & 122 & 20 & 0.5602 \\ \hline
+\emph{Accumulate} & \emph{Reorder}&  5,000 & 0.01 & 796 & 434 & 331 & 23 & 0.7174 \\ \hline
 \rowcolor{Gray}
-\emph{Skip} & \emph{Normal}       & 10,000 & 0.02 & 823 & 363 & 119 & 23 & NA \\ \hline
-\emph{Skip} & \emph{Reorder}      &  2,000 & 0.01 & 870 & 362 & 198 & 23 & 0.3125 \\ \hline
-\emph{Accumulate} & \emph{Dag}    &  5,000 & 0.01 & 964 & 499 & 1,174 & 26 & 0.3310 \\ \hline
-\emph{Skip} & \emph{Dag}          & 10,000 & 0.02 & 1,042 & 620 & 2,083 & 28 & 0.0248 \\ \hline
-\emph{Single} & \emph{Normal}     & 500 & NA & 1,214 & 497 & 39 & 18 & 0.0046 \\ \hline
-\emph{Single} & \emph{Reorder}    & 200 & NA & 1,506 & 823 & 57 & 20 & 0 \\ \hline
-\emph{Single} & \emph{Dag}        & 100 & NA & 2,569 & 1,313 & 43 & 17 & 0 \\ \hline
+\emph{Skip} & \emph{Normal}       & 10,000 & 0.02 & 823 & 363 & 119 & 20 & NA \\ \hline
+\emph{Skip} & \emph{Reorder}      &  2,000 & 0.01 & 870 & 362 & 198 & 20 & 0.3125 \\ \hline
+\emph{Accumulate} & \emph{Dag}    &  5,000 & 0.01 & 964 & 499 & 1,174 & 22 & 0.3310 \\ \hline
+\emph{Skip} & \emph{Dag}          & 10,000 & 0.02 & 1,042 & 620 & 2,083 & 24 & 0.0248 \\ \hline
+\emph{Single} & \emph{Normal}     & 500 & NA & 1,214 & 497 & 39 & 17 & 0.0046 \\ \hline
+\emph{Single} & \emph{Reorder}    & 200 & NA & 1,506 & 823 & 57 & 18 & 0 \\ \hline
+\emph{Single} & \emph{Dag}        & 100 & NA & 2,569 & 1,313 & 43 & 15 & 0 \\ \hline
 	\end{tabular}
@@
 \rowcolor{Gray}
-\emph{Skip} & \emph{Normal}        & 10,000 & 0.005 & 16,226 & 7,455  &   283 & 62 & NA \\ \hline
-\emph{Skip} & \emph{Reorder}       & 10,000 & 0.005 & 17,883 & 8,378  & 2,070 & 82 & 0.6918 \\ \hline
-\emph{Accumulate} & \emph{Reorder} & 10,000 & 0.002 & 18,132 & 8,926  & 2,027 & 84 & 0.8822 \\ \hline
-\emph{Accumulate} & \emph{Normal}  & 10,000 & 0.005 & 21,428 & 6,379  &   279 & 62 & 0.2525 \\ \hline
-\emph{Accumulate} & \emph{Dag}     & 10,000 & 0.002 & 24,855 & 12,512 & 4,059 & 88 & 0.0736 \\ \hline
-\emph{Skip} & \emph{Dag}           & 10,000 & 0.002 & 25,678 & 12,034 & 4,178 & 81 & 0.0211 \\ \hline
-\emph{Single} & \emph{Reorder}     & 100    & NA    & 26,558 & 7,858  &    52 & 37 & 0.0076 \\ \hline
-\emph{Single} & \emph{Normal}      & 2,000  & NA    & 31,057 & 14,475 &   141 & 50 & 0.0085 \\ \hline
-\emph{Single} & \emph{Dag}         & 100    & NA    & 34,208 & 10,069 &    59 & 40 & 0 \\ \hline
+\emph{Skip} & \emph{Normal}        & 10,000 & 0.005 & 16,226 & 7,455  &   283 & 49 & NA \\ \hline
+\emph{Skip} & \emph{Reorder}       & 10,000 & 0.005 & 17,883 & 8,378  & 2,070 & 62 & 0.6918 \\ \hline
+\emph{Accumulate} & \emph{Reorder} & 10,000 & 0.002 & 18,132 & 8,926  & 2,027 & 64 & 0.8822 \\ \hline
+\emph{Accumulate} & \emph{Normal}  & 10,000 & 0.005 & 21,428 & 6,379  &   279 & 49 & 0.2525 \\ \hline
+\emph{Accumulate} & \emph{Dag}     & 10,000 & 0.002 & 24,855 & 12,512 & 4,059 & 69 & 0.0736 \\ \hline
+\emph{Skip} & \emph{Dag}           & 10,000 & 0.002 & 25,678 & 12,034 & 4,178 & 62 & 0.0211 \\ \hline
+\emph{Single} & \emph{Reorder}     & 100    & NA    & 26,558 & 7,858  &    52 & 32 & 0.0076 \\ \hline
+\emph{Single} & \emph{Normal}      & 2,000  & NA    & 31,057 & 14,475 &   141 & 43 & 0.0085 \\ \hline
+\emph{Single} & \emph{Dag}         & 100    & NA    & 34,208 & 10,069 &    59 & 34 & 0 \\ \hline
 	\end{tabular}
@@
 \textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Reduced} & \textbf{p-value} \\ \hline
-\emph{Skip} & \emph{Reorder}       &  2,000 & 0.002 &  62,853 & 20,023 &   757 & 117 & 0.5247 \\ \hline
-\emph{Single} & \emph{Reorder}     &    500 &    NA &  63,731 & 13,119 &   245 & 100 & 0.3610 \\ \hline
-\emph{Accumulate} & \emph{Normal}  &  1,000 & 0.002 &  67,390 & 11,991 &   211 &  90 & 0.8767 \\ \hline
+\emph{Skip} & \emph{Reorder}       &  2,000 & 0.002 &  62,853 & 20,023 &   757 & 114 & 0.5247 \\ \hline
+\emph{Single} & \emph{Reorder}     &    500 &    NA &  63,731 & 13,119 &   245 &  97 & 0.3610 \\ \hline
+\emph{Accumulate} & \emph{Normal}  &  1,000 & 0.002 &  67,390 & 11,991 &   211 &  87 & 0.8767 \\ \hline
 \rowcolor{Gray}
-\emph{Skip} & \emph{Normal}        &  1,000 & 0.002 &  68,231 & 15,115 &   202 &  88 & NA \\ \hline
-\emph{Single} & \emph{Normal}      &  1,000 &    NA &  68,819 & 14,590 &   203 &  87 & 0.7695 \\ \hline
-\emph{Accumulate} & \emph{Reorder} &  2,000 & 0.002 &  72,991 & 18,650 &   760 & 118 & 0.1868 \\ \hline
-\emph{Skip} & \emph{Dag}           & 10,000 & 0.001 & 132,821 & 29,432 & 4,983 & 127 & 0 \\ \hline
-\emph{Accumulate} & \emph{Dag}     &  5,000 & 0.002 & 133,188 & 29,326 & 2,542 & 129 & 0 \\ \hline
-\emph{Single} & \emph{Dag}         &    500 &    NA & 160,696 & 39,242 &   299 & 100 & 0 \\ \hline
+\emph{Skip} & \emph{Normal}        &  1,000 & 0.002 &  68,231 & 15,115 &   202 &  85 & NA \\ \hline
+\emph{Single} & \emph{Normal}      &  1,000 &    NA &  68,819 & 14,590 &   203 &  85 & 0.7695 \\ \hline
+\emph{Accumulate} & \emph{Reorder} &  2,000 & 0.002 &  72,991 & 18,650 &   760 & 115 & 0.1868 \\ \hline
+\emph{Skip} & \emph{Dag}           & 10,000 & 0.001 & 132,821 & 29,432 & 4,983 & 125 & 0 \\ \hline
+\emph{Accumulate} & \emph{Dag}     &  5,000 & 0.002 & 133,188 & 29,326 & 2,542 & 125 & 0 \\ \hline
+\emph{Single} & \emph{Dag}         &    500 &    NA & 160,696 & 39,242 &   299 &  97 & 0 \\ \hline
 	\end{tabular}
@@
 \textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Reduced} & \textbf{p-value} \\ \hline
-\emph{Skip} & \emph{Reorder}      & 10,000 & 0.0005  & 196,116  & 60,326  & 3,175 & 490 & 0  \\ \hline
-\emph{Accumulate} & \emph{Reorder}& 10,000 & 0.001 & 199,193 & 85,970 & 3,117 & 497 & 0  \\ \hline
-\emph{Single} & \emph{Dag} & 200  & NA & 239,844 & 78,739 & 114 & 77 & 0 \\ \hline
-\emph{Single} & \emph{Reorder}    & 500   & NA & 243,664 & 91,935 & 266 & 134 & 0 \\ \hline
-\emph{Skip} & \emph{Dag}          & 2,000 & 0.002 & 369,931 & 174,408 & 952 & 153 & 0.0100 \\ \hline
-\emph{Accumulate} & \emph{Dag}    & 2,000 & 0.002  & 379,202 & 147,289 & 973 & 162 & 0.0053  \\ \hline
-\emph{Single} & \emph{Normal}     & 5,000  & NA & 399,911 & 194,384 & 273 & 114 & 0.0047 \\ \hline
+\emph{Skip} & \emph{Reorder}      & 10,000 & 0.0005  & 196,116  & 60,326  & 3,175 & 484 & 0  \\ \hline
+\emph{Accumulate} & \emph{Reorder}& 10,000 & 0.001 & 199,193 & 85,970 & 3,117 & 489 & 0  \\ \hline
+\emph{Single} & \emph{Dag} & 200  & NA & 239,844 & 78,739 & 114 & 74 & 0 \\ \hline
+\emph{Single} & \emph{Reorder}    & 500   & NA & 243,664 & 91,935 & 266 & 131 & 0 \\ \hline
+\emph{Skip} & \emph{Dag}          & 2,000 & 0.002 & 369,931 & 174,408 & 952 & 149 & 0.0100 \\ \hline
+\emph{Accumulate} & \emph{Dag}    & 2,000 & 0.002  & 379,202 & 147,289 & 973 & 158 & 0.0053  \\ \hline
+\emph{Single} & \emph{Normal}     & 5,000  & NA & 399,911 & 194,384 & 273 & 112 & 0.0047 \\ \hline
 \emph{Accumulate} & \emph{Normal} & 2,000 & 0.002 & 483,938 & 227,686 & 168 & 87 & 0.45 \\ \hline
 \rowcolor{Gray}
-\emph{Skip} & \emph{Normal} & 2,000 & 0.002 & 657,121 & 311,376 & 175 & 87 & NA \\ \hline
+\emph{Skip} & \emph{Normal} & 2,000 & 0.002 & 657,121 & 311,376 & 175 & 84 & NA \\ \hline
 
@@
 of a node.  For each input that is given to an individual, all of its nodes
-will produce a single output.  These output values can be recorded an ordered.
+will produce a single output.  These output values can be recorded and ordered.
 As the domain in use in binary, the outputs can be viewed as a bit string, with

### 2013-05-17 CGP Journal/main.tex
Added place holder to LengthBias paper to stop "undefined references" and fixed a few other bib related things

@@
 We include this problem purely for backward comparison, as we agree that in general
-it is too simple of a problem to merit its own conclusions~\cite{white:2013:bggb}.
+it is too simple of a problem to merit its own conclusions~\cite{white:2013:bgpb}.
 Yet as part of the group it may help yield understanding about how each variant
@@
 Finally we chose the 3-bit Multiply to represent hard binary problems,
-as was suggested in~\cite{white:2013:bggb} and used
+as was suggested in~\cite{white:2013:bgpb} and used
 by~\cite{vassilev:2000:neutrality,miller:2006:redundancy,walker:2008:cgpmodules}.

### 2013-05-17 CGP Journal/main.tex
Replaced simplify with simpler algorithm, commented out old section

@@
 
+\begin{figure}
+  \begin{algorithmic}
+  \Procedure{Simplify}{}
+    \State $reconnect \leftarrow \emptyset$
+    \ForAll{$node \in reversed(active)$}
+      \State $reconnect[footprint(node)] \leftarrow node$
+    \EndFor
+    \ForAll{$i \in input\_locations$}
+      \State $reconnect[footprint(i)] \leftarrow i$
+    \EndFor
+    \ForAll{$g \in genes$}
+      \If{$g$ is a connection gene}
+        \If{$g \in reconnect$}
+          \State $g \leftarrow reconnect[footprint(g)]$
+        \EndIf
+      \EndIf
+    \EndFor
+  \EndProcedure
+  \end{algorithmic}
+  \caption{Algorithm to ensure each active \emph{footprint} is only
+           calculated once.}
+  \label{fig:simplify}
+\end{figure}
+
+The \Call{Simplify}{} algorithm given in Figure~\ref{fig:simplify} is designed
+to efficiently reduce the redundancy encoded in a genome.  Generally, this
+algorithm will move all connections to the first node that produces the desired
+\emph{footprint}.  After application, and recalculation of which nodes are still
+active, each active node produces a unique \emph{footprint}.  Any node that is
+transitively dependent on its own \emph{footprint} will always be removed from
+the active set.  In general, if a set of nodes all produce the same \emph{footprint},
+whichever has the least transitive dependencies will be kept.  Note that this
+does guarantee the minimum possible genome, an algorithm capable of that requires
+exponential runtime.  The \Call{Simplify}{} algorithm requires $O(AN)$ time,
+as the final for loop is the most expensive step.  As before, this means
+\Call{Simplify}{} is no more complex than copying a genome.
+
+By examining which nodes remain active after simplification, we can discuss
+the usefulness of each node as either being part of the minimum solution, a
+duplicate of a node in the minimum solution, or irrelevant to the solution.
+As such we can then examine how redundancy is being used, how frequently useful
+structures exist in the inactive nodes, and how useful nodes are constructed by evolution.
+
+\begin{comment}
 \begin{figure}
@@
 with the algorithm given in Figure~\ref{fig:reorder}.
-
+\end{comment}
 \subsection{Inactive Nodes and Node Duplication}

### 2013-05-22 CGP Journal/main.tex
Wrote the CGP introduction section, added some citations

@@
 \section{Cartesian Genetic Programming}
-TODO Explain basic CGP, use~\cite{miller:2011:chapter2}.
+Cartesian Genetic Programming (CGP) was originally proposed as a method for general
+genetic programming in~\cite{miller:2000:CGPorigin}.  While its beginnings and
+its name come from evolving circuits on a two dimensional grid, modern CGP can
+represent any directed acyclic graph (DAG), with applications in evolving binary
+circuits~\cite{walker:2008:cgpmodules},
+robot controllers~\cite{harding:2005:robots},
+neural networks~\cite{khan:2010:cgpann},
+image classifiers~\cite{harding:2012:mtcgp},
+and regression~\cite{harding:2009:smcgp}.
+
+CGP represents DAGs using a linear genome of integer values.  Each node in the
+DAG is encoded using a collection of genes, with one gene specifying the function
+the node applies to its inputs, and the remaining genes expressing where the
+node takes input from.  Nodes can take input from any problem input and from
+any node preceding them in the genome.  To complete the representation, a set
+of extra genes are included at the end of the genome to specify which nodes or
+input locations to use as function outputs.
+
+Unlike tree based GP encodings, CGP is able to perform value reuse.  This means
+that the output of a node can be reused by any number of other nodes as input values.
+CGP's requirement that nodes connect forward in the genome prevents this ability
+from causing cycles.
+
+As output locations and information flow in the DAG are evolvable, there are often
+large portions of the genome which do not participate in creating any of the
+output values.  These nodes are referred to as inactive, with the nodes currently
+being used to create output values referred to as active.  Inactive nodes
+allow CGP individuals to drift genetically, as they can be mutated without effecting
+the fitness of the individual.  This genetic drift can then be incorporated, as
+mutation can change the DAG structure causing previously inactive nodes to become
+active.  Previous work suggests CGP is most efficient when up to 95\% of the genome
+is inactive~\cite{miller:2006:redundancy}, yet other work suggests this may
+be a result of hidden parsimony pressure in CGP~\cite{goldman:2013:ordering}.
+
+CGP uses very simple evolutionary mechanisms.  By far the most common strategy
+is $1+4$, in which each generation a single parent produces four offspring using
+mutation.  The best offspring then competes with the parent, with the offspring
+replacing the parent if it is no less fit.  This replacement strategy encourages
+drift, as completely neutral mutations are always preferred to stagnation.  Mutation
+is commonly performed by mutating each gene at a set probability, where mutation
+involves changing a gene randomly to some different valid value.  For instance,
+if a function gene is chosen for mutation, its new value is randomly chosen from
+all possible functions, with the exception of the genes current value.  Combined,
+this form of CGP can be viewed as a stochastic hill climber with neutrality.
+For a more in depth description of CGP, see~\cite{miller:2011:chapter2}.
 
@@
 results.  Include O(N) table similar to previous paper.
+
+TODO Include scaling comparison with~\cite{walker:2008:cgpmodules}.
 

### 2013-05-23 CGP Journal/main.tex
Updated TODOs, got rid of generalization section because I've already spent too much computation time
Filled in some introductory sections

@@
 % and "HIS" in caps to complete the first word.
-\IEEEPARstart{T}{ODO this} section.
-% You must have at least 2 lines in the paragraph with the drop letter
-% (should never be an issue)
+\IEEEPARstart{W}{hile effective} evolutionary methods are often simple to design
+and implement, it can sometimes be difficult to discern the reason a technique
+or behavior in that system leads to effective search.  This can be exceptionally
+true for Genetic Programming (GP) systems, as they often include nontrivial
+genotype to phenotype maps and operators that act on genotypes without regard to
+phenotypic impact.  Yet understanding the root causes of evolutionary success
+and failure is critical to improving existing techniques, designing new techniques,
+and understanding how and where to apply each optimization system.
+
+TODO Previous work on understanding evolution, especially CGP.
+
+In this work we set out to develop a deeper understanding of what makes Cartesian
+Genetic Programming (CGP) an effective evolutionary optimizer.  Our primary focus is
+on how mutation and genome ordering interact.  In Section~\ref{sec:cgp} we provide
+an introduction to CGP, with sufficient background to put later sections into
+context.  Section~\ref{sec:duplicate} explains previous work about mutation's potential
+to create identifiably wasted evaluations, and proposed methods to prevent that
+waste.  Section~\ref{sec:ordering} discusses the impact of CGP's genome ordering,
+with methods to both examine and bypass the limitations imposed by that ordering.
+Each technique is then rigorously tuned and qualitatively compared on a collection
+of benchmarks in Section~\ref{sec:quality}.  Section~\ref{sec:analysis} then
+provides a deep look into the details of CGP's search, and how each method impacts
+search, with final remarks given in Section~\ref{sec:conclusion}.
 
 \section{Cartesian Genetic Programming}
+\label{sec:cgp}
 Cartesian Genetic Programming (CGP) was originally proposed as a method for general
@@
 \section{Duplicate Evaluation Avoidance}
-
-TODO Explain here why only compared with parent, not global history.
+\label{sec:duplicate}
+CGP's mutation operator is traditionally applied to all genes equally, with no
+consideration if genes are part of an active or inactive node.  This has the
+potential to lead offspring who's only mutated genes are part of inactive nodes.
+These offspring are actively identical (contain identical active genes) to their
+parents, and therefore by definition have identical fitness to their parents.
+As such, evaluating these individuals is wasteful, as their fitness can be
+assigned without the computational complexity of performing an evaluation.
+Previous work~\cite{goldman:2013:cgpwaste} has taken an initial look at the
+effect this waste can have, and proposed \emph{Skip}, \emph{Accumulate}, and
+\emph{Single} as methods for avoiding or preventing these duplicate evaluations.
+
+TODO Explain that inactive / active is knowable before evaluation.
+
+The amount of evaluations wasted on these actively identical individuals is
+directly related to the mutation rate and the number of active genes.  With
+sufficiently high mutation rates and enough active genes, the probability of
+creating an offspring which is actively identical to its parent is very low.
+Yet such configuration make make it difficult for CGP to optimize, as mutation
+is likely to change many active genes in each application.  As was discussed
+in~\cite{goldman:2013:cgpwaste}, this can make normal CGP very sensitive to the
+mutation rate in use.
 
@@
 to drift due to lack of introns, it does have an increased potential to become stuck.
-As will be discussed in Section~\ref{sec-ordering} and was shown in~\cite{goldman:2013:ordering}
+As will be discussed in Section~\ref{sec:ordering} and was shown in~\cite{goldman:2013:ordering}
 this problem of highly active genomes is very unlikely.
@@
 \section{Genome Ordering}
-\label{sec-ordering}
-TODO General stuff about ordering, including normal CGP.
+\label{sec:ordering}
+As was discussed in Section~\ref{sec:cgp}, CGP uses node ordering in the genome
+to prevent cycles, as nodes are only able to receive input from sources that
+precede them in the genome.  This restriction does not limit CGP's ability
+to represent DAGs, as all DAGs can be serialized to fit this requirement.
+Yet this likely has an impact on CGP's ability to evolve specific
+DAGs~\cite{goldman:2013:ordering}.
+
+Primarily, enforcing node ordering adds artificial limitations CGP's ability to
+connect nodes.  Beyond preventing cycles, it also prevents nodes from connecting
+which would not create a cycle based solely on the evolutionary happenstance of
+where those nodes were evolved.  Similarly, it is impossible for adjacent nodes
+to ever be connected through an intermediate node.  As a result, useful structures
+in the genome may need to be evolved repeatedly in order to find the best
+location for that structure in the genome.
+
+Ordering combined with the random reset mutation method may be the cause of CGP's
+immensely inactive genomes.  As was rigorously defined in~\cite{goldman:2013:ordering},
+the number of active genes in a genome is expected to scale logarithmically
+with genome size, independent of problem application.  Furthermore, the less
+active a genome, the lower the likelihood of structure ordering preventing structure
+connection.
+
+To rectify these issues an examine CGP without ordering bias, \cite{goldman:2013:ordering}
+created \emph{Reorder} and \emph{DAG} as methods for genome ordering, and defined
+\emph{Normal} as CGP's historical ordering method.
 
@@
 \section{Qualitative Comparisons}
+\label{sec:quality}
 While it is possible to theorize the effects of each CGP variant on search, and
@@
 TODO Include scaling comparison with~\cite{walker:2008:cgpmodules}.
-
-\subsection{Generalization}
-TODO Consider dropping this section due to computational limits.
 
@@
 \section{Conclusion}
-The conclusion goes here.
+\label{sec:conclusion}
+TODO The conclusion goes here.
 

### 2013-05-23 CGP Journal/main.tex
More fill in, more citations

@@
 
-TODO Previous work on understanding evolution, especially CGP.
-
 In this work we set out to develop a deeper understanding of what makes Cartesian
@@
 
+There have been a number of previous studies into various aspects of GP evolution.
+For instance~\cite{daida3:2003:treebias} looked into how trees of specific shapes,
+which helped inspire the use of new grammar based operations~\cite{xuan:2006:grammar}.
+Analysis of the root cause of bloat has helped to inform methods for controlling
+bloat~\cite{luke:2006:bloat}.  A few studies of evolutionary mechanisms have also
+been done in CGP directly, related to bloat~\cite{miller:2001:bloat},
+neutrality~\cite{vassilev:2000:neutrality}, and structural bias~\cite{payne:2009:bias}.
+
 \section{Cartesian Genetic Programming}
@@
 assigned without the computational complexity of performing an evaluation.
+Furthermore, calculating which nodes are active before evaluating an individual
+improves evaluation speed~\cite{vasicek:2012:efficient}, meaning detecting when duplicated individuals are
+created can be done without significant computational overhead.
 Previous work~\cite{goldman:2013:cgpwaste} has taken an initial look at the
@@
 \emph{Single} as methods for avoiding or preventing these duplicate evaluations.
-
-TODO Explain that inactive / active is knowable before evaluation.
 
@@
 
-TODO Consider adding 5-bit Parity or Equality, 11-bit Multiplexer.
+%TODO Consider adding 5-bit Parity or Equality, 11-bit Multiplexer.
 

### 2013-05-30 CGP Journal/main.tex
Updated some TODOs and subsection layouts

@@
 \end{comment}
-\subsection{Inactive Nodes and Node Duplication}
+\subsection{Active, Inactive, and Never Active Nodes}
+TODO Look at how much of the genome is unused.
+
+\subsection{Node Behaviors}
+TODO High amount of TF.
+
+TODO Breakdown of Active Nodes.
+
+TODO Breakdown of Inactive Nodes.
+
 TODO Look at how much of the genome is duplicated in both active and inactive
-
-TODO Look at how much of the genome is unused.
 

### 2013-05-31 CGP Journal/main.tex
Big block of results

@@
 \subsection{DAG}
+\label{sec:dag}
 The principle behind \emph{DAG} is to remove the traditional CGP requirement that
@@
 \subsection{Parameter Setting}
+\label{sec:parameter}
 In order to fairly compare each method, we should ensure proper parameter configuration.
@@
 	\caption{Multiply 5.1429910779824196e-18}
-	\label{tab:Multiply}
+	\label{tab:multiply}
 \end{table*}
 
-TODO Compare using a table the best parameters found for each configuration.
-One table per problem.
-
+The results from running each tuned parameter configuration for each algorithm
+combination on each problem are summarized by Table~\ref{tab:parity},
+Table~\ref{tab:encode}, Table~\ref{tab:decode}, and Table~\ref{tab:multiply}.
+Rows in each table are ordered by that algorithm combination's median evaluations to
+success (MES), with the best algorithm combination appearing at the top of each table.
+The highlighted row marks \emph{Skip Normal}, the control algorithm combination
+representing traditional CGP's performance.  All configurations for each problem
+were then testing using Kruskal-Wallis to determine if any configuration is
+statistically different from any other.  All problems received highly significant
+values.  As such, each algorithm combination was then compared with the control
+to determine statistical significance using pairwise Mann-Whitney~U tests. The
+resulting p-value for each test is reported in the table.
+
+As discussed in Section~\ref{sec:parameter}, the genome size and mutation rate
+for each algorithm combination was extensively optimized.  In line with previous
+research~\cite{miller:2006:redundancy}, this led to massive genome sizes for
+almost all algorithm combinations on all problems.  With the exception of 
+some combinations using \emph{Single}, all algorithm combinations used over
+1,000 nodes.  Half of those configurations used 10,000 nodes, the maximum allowed
+by our tuning, which may imply these combinations work best with even larger
+genomes.
+
+Of the 12 configurations including \emph{Single}, 11 used a genome size smaller
+than all configurations not using \emph{Single}.  The exception was \emph{Single Normal}
+on Multiply.  No other algorithm had as much impact on genome size tuning as
+the inclusion or exclusion of \emph{Single}.  In previous work~\cite{goldman:2013:ordering}
+it appeared \emph{Reorder} and \emph{DAG} were allowing CGP to use smaller genomes.
+In light of our results, this conclusion should be amended.
+It appears that \emph{Single} is most effective when the genome size is small and
+that \emph{Reorder} and \emph{DAG} are best able to cope with smaller genome sizes.
+Independent of duplication method, the ordering methods do not seem to create any
+pressure on which genome size is chosen.
+
+\begin{comment}
+Genome size explosion seems most prevalent on easier problems, with Parity and
+Encode having much higher genome sizes than Decode and Multiply.  Initially
+this seems counter intuitive, as we would expect harder problems to require
+larger solutions which in turn should require larger genomes.  Yet the genome
+sizes in use are so far beyond what a solution to these problems require that
+solution size is likely not the cause of this behavior.  Instead, consider
+the usefulness of unexplored nodes.
+
+On simple enough problems, exploring lots of randomly generated nodes may be
+sufficient to generate high quality solutions.  At relatively low numbers of
+evaluations, large genotypes likely mean each mutation to a connection gene
+will activate nodes that have never been active, resulting in exploration.
+At higher numbers of evaluations, mutations to connection genes that are not currently
+active have a high chance of connecting to a node that was tested previously
+and found ineffective.  As a result, having a smaller genome size on harder problems
+may reduce the chance of exploratory mutations when optimizing near the global optimum.
+
+Under this theory, we should expect Parity to have the largest genome sizes
+\end{comment}
+
+Unlike genome size, all configurations choose approximately the same mutation rate,
+with different rates for each problem.  With the exception of Parity, there appears
+to be a general consensus that each gene should be mutated with probability 0.002,
+independent of genome size.  Parity used a higher mutation rate, but this is likely
+because 3-bit Parity is by far the easiest problem tested and therefore likely benefits
+from higher exploration.
+
+On many of the problems, many of the configurations have little to no statistical
+difference from the control in their evaluations required to reach the global optimum.
+On Parity and Encode it is probably safe to conclude that \emph{Single} is worse than
+the control, although as these are the easiest problems tested that difference may have
+little real world meaning.  \emph{DAG} is almost certainly worse than the control on
+Decode.  Its results are so much worse that we can also likely conclude that \emph{DAG}
+is worse than all non \emph{DAG} configuration on that problem, even though no statistical
+tests are presented here.  The control on multiply did by far the worst of all configurations,
+with only \emph{Accumulate Normal} likely failing statistical significance.
+
+By ordering the algorithm combinations by MES and examining that ordering across each
+problem, we can look for general trends.  From the tables, it appears on easier problems
+which method of duplication is in use is the strongest predictor of rank, while
+the ordering methods are more important on harder problems.  This comes from the
+observation that in for Parity and Encode (Table~\ref{tab:parity} and Table~\ref{tab:encode})
+the first column is effectively sorted, while for Decode and Multiply
+(Table~\ref{tab:decode} and Table~\ref{tab:multiply}) the second column is sorted.
+Each table includes coincidentally one exception to these rules.  On Parity and
+Encode pairings with \emph{DAG} are ranked lower than other ordering methods, so much
+so as to break the ordering once on both problems.  On Decode \emph{Reorder Duplicate}
+ranks worse than expected.  Finally on Multiply \emph{Single DAG} does far better
+than expected.
+
+On almost every problem, \emph{DAG} required
+more evaluations to success than \emph{Reorder}, and it only outperformed \emph{Normal}
+on Multiply.  Combined with its's algorithmic overhead described in
+Section~\ref{sec:dag}, \emph{DAG} is therefore not likely to be useful for anything
+but comparison purposes.
+
+On all four problems, \emph{Reorder} either finds the solution in comparable
+or significantly less evaluations than \emph{Normal} ordering, independent of
+the duplicate method in use.  Combined with the complexity analysis in
+Section~\ref{sec:reorder} which shows it is no more asymptotically complex,
+we would suggest \emph{Reorder} as an alternative to \emph{Normal} on hard
+problems.
+
+The difference between using \emph{Skip} and \emph{Accumulate} remains
+negligible~\cite{goldman:2013:cgpwaste}, even in combination with different ordering
+methods.  As \emph{Accumulate} is far more complex in terms of algorithm function
+than \emph{Skip}, we would suggest \emph{Skip} be used over \emph{Accumulate} for
+any real world application.
+
+TODO Discuss actual runtime analysis, and how \emph{Single Reorder} is likely
+the fastest method in that regard.
 \subsection{Scalability}

### 2013-06-03 CGP Journal/main.tex
Results section now completely dumped in.

@@
 
-TODO Discuss actual runtime analysis, and how \emph{Single Reorder} is likely
-the fastest method in that regard.
+Our parameter tuning focused on minimizing the number of evaluations each algorithm
+combination required to solve problems, as that is the primary measure of runtime
+for CGP and GP in general.  As such, comparing the true runtime of the tuned parameters
+is likely to give potentially misleading results.  In using profilers and examining
+general trends, one of the largest indicators of runtime per evaluation was genome size.
+This makes sense as each time an offspring is generated all of the parent's genes
+must first be copied.  Also, larger genome sizes correlate with more active nodes,
+each of which must be executed on each input tested during individual evaluation.
+
+With these caveats, the observed runtime for each algorithm combination follows
+what we would expect given the number of evaluations performed and the genome size.
+When \emph{Reorder} and \emph{Normal} use the same genome sizes, they require
+nearly the same amount of time per evaluation, with \emph{Reorder}'s added
+steps and increased number of active nodes constituting a relatively small increase.
+In general this difference in time per evaluation is trumped by differences in
+actual number of evaluations, which remains the primary cause of differential total runtime.
+As predicted \emph{DAG} does take longer than the other algorithms to perform each evaluation,
+reinforcing the conclusion that \emph{DAG} should not be used for applications.
+
+As our implementation of \emph{Skip} and \emph{Accumulate} required each gene
+to generate a random number to determine if it should mutate, these methods took
+significantly longer per evaluation than \emph{Single}.  Utilizing a binomial
+distribution, identical behavior could have been achieved in effectively constant
+instead of linear time.  Further skewing the comparison is that \emph{Single}'s
+tuning resulted in far smaller genomes, giving it a significant edge over the other
+techniques in runtime per evaluation.
+
+If we had selected on runtime as apposed to evaluations during tuning, we would expect whichever
+technique is most effective at small genome sizes to obtain the best results.
+From theoretical work in~\cite{goldman:2013:ordering}, we would expect \emph{Reorder}
+to work better with this limitation than \emph{Normal}.  Assuming the tuned values
+we obtained are any gauge, \emph{Single Reorder} will likely require the least
+amount of runtime when tuned to reduce runtime.
+
 \subsection{Scalability}

### 2013-06-03 CGP Journal/main.tex
Added the "never active" table and its images

@@
 \definecolor{Gray}{gray}{0.9}
+\usepackage{graphicx}
+\usepackage{rotating}
+\usepackage{multirow}
+
+\newcommand{\graphicthird}[1]
+{\includegraphics[width=.3\textwidth,height=.3\textheight,keepaspectratio]{#1}}
+
+\newcommand{\thirdlabel}[1]
+{\multicolumn{1}{|c|}{\raisebox{.15\textwidth}{\rotatebox[origin=c]{90}{\textbf{\em #1}}}}}
+
 \begin{document}
@@
 
+\begin{table*}
+	\centering
+  \begin{tabular}{c|c|c|c|}
+    \cline{2-4}
+    & \textbf{\em Normal} & \textbf{\em Reorder} & \textbf{\em DAG} \\ \hline
+    \thirdlabel{Skip} &\graphicthird{multiply_skip_normal} & \graphicthird{multiply_skip_reorder} & \graphicthird{multiply_skip_dag}\\ \hline
+    \thirdlabel{Accumulate}&\graphicthird{multiply_accumulate_normal} & \graphicthird{multiply_accumulate_reorder} & \graphicthird{multiply_accumulate_dag}\\ \hline
+    \thirdlabel{Single}&\graphicthird{multiply_single_normal} & \graphicthird{multiply_single_reorder} & \graphicthird{multiply_single_dag}\\ \hline
+	\end{tabular}
+	\caption{Location of never active nodes across all 50 runs for all nine algorithm combinations on the Multiply problem.}
+	\label{tab:never_active}
+\end{table*}
+
 \subsection{Node Behaviors}

### 2013-06-04 CGP Journal/main.tex
Big block of never active section

@@
 \end{comment}
-\subsection{Active, Inactive, and Never Active Nodes}
-TODO Look at how much of the genome is unused.
+\subsection{Never Active Nodes}
+In line with the suggestions made in~\cite{miller:2006:redundancy}, the evolved
+number of active nodes across all problems and all algorithm configurations
+in Table~\ref{tab:parity}, Table~\ref{tab:encode}, Table~\ref{tab:decode}, and Table~\ref{tab:multiply}
+was significantly lower than the genome size.  With the exception of a few combinations
+with \emph{DAG} and \emph{Single}, the evolved results were over 50\% inactive.
+\emph{Skip Normal}, our control representing normal CGP, had only 1\%, 3\%, 20\%,
+and 9\% of nodes in the genome active in the median runs of Parity, Encode, Decode,
+and Multiply, respectively.  These final results fall in line with the predictions
+made in~\cite{goldman:2013:ordering}, in that if the problems are ordered based on
+their number of output locations, their evolved percentage of active nodes is also ordered.
+
+With how inactive the final genomes are, we ask ``what is the rest for?''
+Just because a node is not active in the final solution does not mean it was never
+useful.  For a node to be truly of no use to search, we must consider nodes that
+were never active.  Never active nodes are nodes such that for all parents of the
+final solution (the line of descent), no parent had that node active.  Note that
+for \emph{Reorder} this demarcation of node activity follows the node through
+each shuffle, such that if a node was marked as never active before the genome
+was shuffled it is marked as never active after the shuffle, and vice versa.
+This is in contrast to marking locations as never active, which was not done in
+this study.
 
@@
 \end{table*}
+
+Table~\ref{tab:never_active} provides information about never active nodes for
+each algorithm combination on the 3-bit Multiply problem.  The horizontal axis
+in each of the nine plots shows different genome locations, while the vertical
+axis denotes the level of fitness the recorded individual obtained.  The shading
+of each coordinate gives the percentage of 50 runs that had an individual of
+the given fitness level such that the node at the given location had never
+previously been active.  While only Multiply is shown here, all of the
+other problems tested had similar behavior, with allowances for differences in
+number of evaluations performed.
+
+Examining the columns of Table~\ref{tab:never_active} makes it clear that
+each of the ordering methods had significantly different never active behavior.
+Looking at \emph{Normal} we see that only at very high fitness levels does evolution
+begin to activate nodes in the latter half of the genome.  Even by the final solution
+large percentages of the genome were never activated, with \emph{Skip}, \emph{Accumulate},
+and \emph{Single} having median percentages of never active nodes of 43.35\%, 42.05\%, and
+56.17\%, respectively.  This is in stark contrast to \emph{Reorder} and \emph{DAG},
+which never had a median higher than 0.05\%.
+
+Compared to the difference in columns, the difference in the rows of Table~\ref{tab:never_active}
+caused by the different duplication methods were almost non-existent.  Only
+\emph{Single} has a distinguishable behavior, but this is almost certainly the result
+of the genome size and not the duplication method.  \emph{Single Normal} used
+a genome size 2.5 times larger than \emph{Normal} did with the other techniques,
+so the resulting increase in never active nodes makes sense.  When \emph{Single}
+was paired with \emph{Reorder} and \emph{DAG} the genome sizes were far smaller,
+resulting in noisier graphs with less never active nodes.
+
+So now the question is why does \emph{Normal} use so little of its genome?  To
+answer this we must consider what can prevent a node from becoming active.  For
+a node to never be active all tested offspring with mutations that result in that node being activated
+must have been less fit than their parent.  Even if those offspring are no more fit, the selection method
+used by CGP would have allowed that offspring to replace its parent.  This leads
+to two possible causes for a node never being active: bad genes or infrequently tried.
+
+The bad genes cause for a node never being active suggests the behavior encoded
+by a node is simply useless on the problem landscape.  If this was the root cause
+of never active nodes, we would expect \emph{Reorder} and \emph{DAG} to have
+similar problems, as neither technique explicitly changes the likelihood of node
+behaviors.  Furthermore, we would expect the location of never active nodes to be
+uniform in the genome, which is obviously not the case.  Finally, even if bad
+genes were the problem, we would expect the mutational drift allowed by CGP's
+selection method to change their bad genes and eventually allow them to be active.
+
+The infrequently tried cause for a node never being active suggests that few
+individuals are ever produced which have that node active, limiting the likelihood
+of those individuals being selected regardless of potential fitness.  Under this
+cause and combined with the theory given in~\cite{goldman:2013:ordering}, we
+expect a strong positional bias in the location of the never active nodes.  Furthermore,
+we would expect nodes near the end of the genome to be the most likely to be never active.
+Finally we would expect \emph{Reorder} and \emph{DAG} to have less or no positional bias.
+All of these expectations match the results given in Table~\ref{tab:never_active}.
+
+Examining the progression in fitness across the different ordering methods shows
+the effect of initialization on each.  All three methods share the same initialization
+method, and as such all have very similar never active node behavior early in
+evolution.  This initialization similarity explains why \emph{DAG} has any positional
+bias at all, even though its evolutionary mechanisms have none.  Similarly \emph{Reorder}
+diverges from \emph{Normal} given sufficient time, eventually losing its positional
+bias.
+
+This early similarity may be part of the reason \emph{Normal} and \emph{Reorder}
+have such similar behavior on relatively easy problems.  After relatively few
+evaluations both have explored similar sections of their genome.  Yet on hard
+problems that require further exploration \emph{Reorder} is able to do so, while
+\emph{Normal} remains positionally limited.
 

### 2013-06-04 CGP Journal/main.tex
Added TODO

@@
 \subsection{Never Active Nodes}
+%TODO Somewhere in here you should talk about how Normal has the least number of active,
+%which is probably related to how many never become active
 In line with the suggestions made in~\cite{miller:2006:redundancy}, the evolved

### 2013-06-05 CGP Journal/main.tex
Added bar graphs and table

@@
     & \textbf{\em Normal} & \textbf{\em Reorder} & \textbf{\em DAG} \\ \hline
-    \thirdlabel{Skip} &\graphicthird{multiply_skip_normal} & \graphicthird{multiply_skip_reorder} & \graphicthird{multiply_skip_dag}\\ \hline
-    \thirdlabel{Accumulate}&\graphicthird{multiply_accumulate_normal} & \graphicthird{multiply_accumulate_reorder} & \graphicthird{multiply_accumulate_dag}\\ \hline
-    \thirdlabel{Single}&\graphicthird{multiply_single_normal} & \graphicthird{multiply_single_reorder} & \graphicthird{multiply_single_dag}\\ \hline
+    %\begin{comment}
+    \thirdlabel{Skip} & \graphicthird{multiply_skip_normal} &
+                        \graphicthird{multiply_skip_reorder} &
+                        \graphicthird{multiply_skip_dag}\\ \hline
+    \thirdlabel{Accumulate} & \graphicthird{multiply_accumulate_normal} &
+                              \graphicthird{multiply_accumulate_reorder} &
+                              \graphicthird{multiply_accumulate_dag}\\ \hline
+    \thirdlabel{Single} & \graphicthird{multiply_single_normal} &
+                          \graphicthird{multiply_single_reorder} &
+                          \graphicthird{multiply_single_dag}\\ \hline
+    %\end{comment}
 	\end{tabular}
@@
 \subsection{Node Behaviors}
+
+\begin{table*}
+	\centering
+  \begin{tabular}{c|c|c|c|}
+    \cline{2-4}
+    & \textbf{\em Normal} & \textbf{\em Reorder} & \textbf{\em DAG} \\ \hline
+    \thirdlabel{Skip} & \graphicthird{bar_multiply_skip_normal} &
+                        \graphicthird{bar_multiply_skip_reorder} &
+                        \graphicthird{bar_multiply_skip_dag}\\ \hline
+    \thirdlabel{Accumulate} & \graphicthird{bar_multiply_accumulate_normal} &
+                              \graphicthird{bar_multiply_accumulate_reorder} &
+                              \graphicthird{bar_multiply_accumulate_dag}\\ \hline
+    \thirdlabel{Single} & \graphicthird{bar_multiply_single_normal} &
+                          \graphicthird{bar_multiply_single_reorder} &
+                          \graphicthird{bar_multiply_single_dag}\\ \hline
+	\end{tabular}
+	\caption{Average node behavior for all nine algorithm combinations on the Multiply problem.
+	         Excludes never active nodes.}
+	\label{tab:behavior}
+\end{table*}
+
+Normal has the highest "used" bar unless its Single, likely because it also has the least active nodes
+
 TODO High amount of TF.

### 2013-06-05 CGP Journal/main.tex
Added encode bar table
More in the behavior section

@@
 	  \hline
-\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Reduced} & \textbf{p-value} \\ \hline
+\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
 \emph{Accumulate} & \emph{Normal} & 10,000 & 0.02 & 682 & 348 & 122 & 20 & 0.5602 \\ \hline
@@
 	  \hline
-\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Reduced} & \textbf{p-value} \\ \hline
+\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
 \rowcolor{Gray}
@@
 	  \hline
-\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Reduced} & \textbf{p-value} \\ \hline
+\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
 \emph{Skip} & \emph{Reorder}       &  2,000 & 0.002 &  62,853 & 20,023 &   757 & 114 & 0.5247 \\ \hline
@@
 	  \hline
-\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Reduced} & \textbf{p-value} \\ \hline
+\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
 \emph{Skip} & \emph{Reorder}      & 10,000 & 0.0005  & 196,116  & 60,326  & 3,175 & 484 & 0  \\ \hline
@@
     & \textbf{\em Normal} & \textbf{\em Reorder} & \textbf{\em DAG} \\ \hline
+    \thirdlabel{Skip} & \graphicthird{bar_encode_skip_normal} &
+                        \graphicthird{bar_encode_skip_reorder} &
+                        \graphicthird{bar_encode_skip_dag}\\ \hline
+    \thirdlabel{Accumulate} & \graphicthird{bar_encode_accumulate_normal} &
+                              \graphicthird{bar_encode_accumulate_reorder} &
+                              \graphicthird{bar_encode_accumulate_dag}\\ \hline
+    \thirdlabel{Single} & \graphicthird{bar_encode_single_normal} &
+                          \graphicthird{bar_encode_single_reorder} &
+                          \graphicthird{bar_encode_single_dag}\\ \hline
+	\end{tabular}
+	\caption{Average node behavior for all nine algorithm combinations on the Encode problem.
+	         Excludes never active nodes.}
+	\label{tab:encode_behavior}
+\end{table*}
+
+\begin{table*}
+	\centering
+  \begin{tabular}{c|c|c|c|}
+    \cline{2-4}
+    & \textbf{\em Normal} & \textbf{\em Reorder} & \textbf{\em DAG} \\ \hline
     \thirdlabel{Skip} & \graphicthird{bar_multiply_skip_normal} &
@@
 	         Excludes never active nodes.}
-	\label{tab:behavior}
+	\label{tab:multiply_behavior}
 \end{table*}
 
-Normal has the highest "used" bar unless its Single, likely because it also has the least active nodes
-
-TODO High amount of TF.
-
-TODO Breakdown of Active Nodes.
-
-TODO Breakdown of Inactive Nodes.
-
-TODO Look at how much of the genome is duplicated in both active and inactive
+In CGP it is common to report what fraction of the genome is active, as this represents
+the easiest estimate of solution size.  By utilizing \emph{footprint} comparison
+and the \Call{Simplify}{} algorithm, we can get a much better estimate of what
+purpose each node in the genome serves.
+Table~\ref{tab:parity}, Table~\ref{tab:encode}, Table~\ref{tab:decode},
+and Table~\ref{tab:multiply} provide two column relevant to node behavior: Active
+and Used.  The former gives the traditional measure of the median amount of
+active nodes in the solution found.  The latter is the median number of nodes
+still active after applying the \Call{Simplify}{} algorithm.
+
+On all problems and for all algorithm combinations, the number of active nodes is
+far smaller than the genome size, and the number of Used nodes
+is far smaller than the number of Active nodes.  In some extreme cases less
+than 1\% of active nodes in the evolved solution are actually necessary.  Only the
+most extreme configurations have even 50\% of active nodes in use by the
+simplified solution.
+
+Table~\ref{tab:encode_behavior} and Table~\ref{tab:multiply_behavior} help illustrate
+the behavioral breakdown of nodes for each algorithm combination on the Encode
+and Multiply problems, respectively.  These two problems were chosen as they
+represent different ends of the quality spectrum for \emph{Skip Normal} and
+because together the are sufficiently representative of behaviors seen on the other
+problems.
+
+Each diagram describes node behavior as falling into eight distinct categories.
+The first category are the constant nodes, both active and inactive.  These nodes
+return the same value for all tested inputs.  The ``Used'' bar is the average
+number of nodes which are part of the simplified solution.  The ``Useful'' bars
+express how many nodes, either active or inactive, have identical \emph{footprint}
+values to used nodes.  These nodes represent either independent methods of calculating
+the same behavior or are effectively pass through nodes which mimic the behavior
+of their inputs.  Active ``Intron'' nodes are those active nodes that do not
+fall into either the constant or used categories, with inactive ``Intron'' nodes
+being those inactive nodes that have identical \emph{footprint} values to active
+intron nodes.  The final bar is ``Explore'', which marks how many nodes on
+average produce a \emph{footprint} not part of any previous category.  Note that
+all never active nodes are ignored by these plots, as their behavior is meaningless.
+
+The first striking feature of these tables is how many nodes produce constant
+outputs.  As the instruction set used in our genomes never require constant values
+to produce any non constant behavior (TRUE NAND X $\Leftrightarrow$ X NAND X)
+and most operations with constants result in pass through nodes
+(FALSE OR X $\Leftrightarrow$ X), we would not expect evolution to proliferate
+these constant nodes.  If these constants are in fact of no use, future work
+could likely improve CGP's performance by reducing their prevalence.  Yet
+it is possible they serve an evolutionary function by allowing mutation more
+flexibility.  For instance, pass through nodes may be allowing mutation to create
+adjustments not easily done if this behavior was forbidden.
+
+As with previous analysis, \emph{Skip} and \emph{Accumulate} have nearly identical
+plots, regardless of ordering method and problem.  \emph{Single} again has
+a significantly different plot when paired with \emph{Reorder} or \emph{DAG},
+with some hits of a change on Encode with \emph{Normal}.  These combinations
+were also the most likely to have significantly smaller genome sizes.  From the
+results tables we see that the number of used nodes on each problem is relatively
+constant across the combinations even though the number of active nodes and the
+genome sizes changed significantly.  As such, the proportion of nodes in the
+used category will vary as these other features change.
+
+From Table~\ref{tab:encode_behavior} and Table~\ref{tab:multiply_behavior}, having
+the plurality of active nodes in the used category appears to correlate with
+increase evaluations to success.  Notice that in all \emph{Single} plots the
+tallest bar for active nodes is ``Used,'' and the only non \emph{Single} plots
+where this is true are \emph{Normal} on Multiply.  On Encode, \emph{Single}
+runs had the highest median evaluations to success, with \emph{Normal} runs
+doing similarly on Multiply.  This appears to imply that having duplication
+of node behavior in the active nodes, and not just inactive nodes, is beneficial
+to search.  If this is true, it may help explain CGP uses such large genomes,
+as \emph{Normal} is only able to achieve enough active nodes to allow duplication
+when the genome size significantly exceeds the actual solution size.  Also
+the success of \emph{Reorder} and \emph{DAG} on Multiply may also be partially be attributable
+to this ability to increase the number of active nodes.
+
+One significant difference in behavior between the two problems is proportion of
+explore nodes.  A potential explanation is that on 16-4-bit Encode there are only
+$2^{16} = 65,536$ \emph{footprints} while on 3-bit Multiply there are $2^{64} = 1.845\cdot 10^{19}$
+\emph{footprints}.  Combined with the fact that not all \emph{footprints} are equally
+likely to be produced, the difference may just be that on Encode nodes are
+that much more likely to recreate existing behavior.  Another possibility is that
+CGP increases the diversity of behavior of inactive nodes over evolutionary time.
+As such, the fact that Multiply required approximately an order of magnitude more
+evaluations to solve would lead to the discrepancy.
 

### 2013-06-19 CGP Journal/main.tex
More tables and some paragraphs explaining them

@@
 
-\subsection{Node Reactivation and Active Node Mutation}
-TODO For each node that was active, then inactive, then active again, see how its changed.
-
-TODO For each node that is active, look at how successful mutations change its behavior.
+\subsection{Parent Replacement}
+
+\begin{table*}
+	\centering
+	\begin{tabular}{|c|c|c|c|c|}
+	  \hline
+\textbf{Duplicate} & \textbf{Ordering} & \textbf{Active Unchanged} & \textbf{No Reactivation} & \textbf{Reactivated Changed} \\ \hline
+       \emph{Skip} &     \emph{DAG} & 0.00\% & 3.48\% & 55.63\% \\ \hline
+ \emph{Accumulate} &     \emph{DAG} & 0.00\% & 8.48\% & 52.82\% \\ \hline
+ \emph{Accumulate} & \emph{Reorder} & 2.09\% & 33.26\% & 58.46\% \\ \hline
+   \rowcolor{Gray}
+       \emph{Skip} &  \emph{Normal} & 2.25\% & 39.35\% & 67.17\% \\ \hline
+ \emph{Accumulate} &  \emph{Normal} & 2.69\% & 39.99\% & 67.27\% \\ \hline
+       \emph{Skip} & \emph{Reorder} & 11.46\% & 53.99\% & 59.47\% \\ \hline
+     \emph{Single} & \emph{Reorder} & 24.94\% & 80.49\% & 60.63\% \\ \hline
+     \emph{Single} &  \emph{Normal} & 26.44\% & 83.55\% & 59.96\% \\ \hline
+     \emph{Single} &     \emph{DAG} & 27.80\% & 83.73\% & 59.92\% \\ \hline
+	\end{tabular}
+	\caption{Parity}
+	\label{tab:parity_reactivate}
+\end{table*}
+
+\begin{table*}
+	\centering
+	\begin{tabular}{|c|c|c|c|c|}
+	  \hline
+\textbf{Duplicate} & \textbf{Ordering} & \textbf{Active Unchanged} & \textbf{No Reactivation} & \textbf{Reactivated Changed} \\ \hline
+ \emph{Accumulate} &     \emph{DAG} & 0.05\% & 4.06\% & 48.70\% \\ \hline
+       \emph{Skip} &     \emph{DAG} & 0.06\% & 3.85\% & 48.33\% \\ \hline
+ \emph{Accumulate} & \emph{Reorder} & 1.77\% & 19.24\% & 55.47\% \\ \hline
+       \emph{Skip} & \emph{Reorder} & 2.02\% & 19.80\% & 56.15\% \\ \hline
+   \rowcolor{Gray}
+       \emph{Skip} &  \emph{Normal} & 14.16\% & 54.78\% & 64.14\% \\ \hline
+ \emph{Accumulate} &  \emph{Normal} & 14.52\% & 54.79\% & 64.17\% \\ \hline
+     \emph{Single} &  \emph{Normal} & 33.77\% & 79.34\% & 59.86\% \\ \hline
+     \emph{Single} &     \emph{DAG} & 35.98\% & 80.91\% & 58.39\% \\ \hline
+     \emph{Single} & \emph{Reorder} & 39.95\% & 78.91\% & 59.72\% \\ \hline
+	\end{tabular}
+	\caption{Encode}
+	\label{tab:encode_reactivate}
+\end{table*}
+
+\begin{table*}
+	\centering
+	\begin{tabular}{|c|c|c|c|c|}
+	  \hline
+\textbf{Duplicate} & \textbf{Ordering} & \textbf{Active Unchanged} & \textbf{No Reactivation} & \textbf{Reactivated Changed} \\ \hline
+ \emph{Accumulate} &     \emph{DAG} & 0.03\% & 10.10\% & 54.99\% \\ \hline
+       \emph{Skip} &     \emph{DAG} & 0.05\% & 10.21\% & 51.33\% \\ \hline
+ \emph{Accumulate} & \emph{Reorder} & 9.98\% & 49.65\% & 70.57\% \\ \hline
+       \emph{Skip} & \emph{Reorder} & 12.53\% & 51.63\% & 70.24\% \\ \hline
+     \emph{Single} & \emph{Reorder} & 28.38\% & 79.73\% & 66.85\% \\ \hline
+     \emph{Single} &  \emph{Normal} & 29.43\% & 80.86\% & 65.85\% \\ \hline
+     \emph{Single} &     \emph{DAG} & 32.60\% & 81.73\% & 57.70\% \\ \hline
+ \emph{Accumulate} &  \emph{Normal} & 51.11\% & 84.54\% & 72.57\% \\ \hline
+   \rowcolor{Gray}
+       \emph{Skip} &  \emph{Normal} & 65.47\% & 89.17\% & 72.21\% \\ \hline
+
+	\end{tabular}
+	\caption{Decode}
+	\label{tab:decode_reactivate}
+\end{table*}
+
+\begin{table*}
+	\centering
+	\begin{tabular}{|c|c|c|c|c|}
+	  \hline
+\textbf{Duplicate} & \textbf{Ordering} & \textbf{Active Unchanged} & \textbf{No Reactivation} & \textbf{Reactivated Changed} \\ \hline
+ \emph{Accumulate} & \emph{Reorder} & 1.14\% & 22.09\% & 71.75\% \\ \hline
+ \emph{Accumulate} &     \emph{DAG} & 4.52\% & 45.38\% & 63.42\% \\ \hline
+       \emph{Skip} &     \emph{DAG} & 5.05\% & 45.58\% & 63.03\% \\ \hline
+       \emph{Skip} & \emph{Reorder} & 10.40\% & 47.67\% & 71.79\% \\ \hline
+     \emph{Single} &  \emph{Normal} & 18.85\% & 78.19\% & 70.11\% \\ \hline
+     \emph{Single} &     \emph{DAG} & 20.72\% & 80.20\% & 65.10\% \\ \hline
+     \emph{Single} & \emph{Reorder} & 22.14\% & 80.21\% & 69.76\% \\ \hline
+ \emph{Accumulate} &  \emph{Normal} & 57.69\% & 87.27\% & 78.95\% \\ \hline
+   \rowcolor{Gray}
+       \emph{Skip} &  \emph{Normal} & 73.46\% & 92.04\% & 79.07\% \\ \hline
+	\end{tabular}
+	\caption{Multiply}
+	\label{tab:multiply_reactivate}
+\end{table*}
+
+An important aspect of CGP is its replacement strategy that allows for neutral drift.
+Each generation the best offspring replaces its parent if it is no less fit.  Previous
+research has shown the impact removing this feature can have on
+performance~\cite{yu:2001:neutrality}, but how it interacts with the different variants
+of CGP has not been investigated previously.
+
+Table~\ref{tab:parity_reactivate}, Table~\ref{tab:encode_reactivate},
+Table~\ref{tab:decode_reactivate}, and Table~\ref{tab:multiply_reactivate} provide a
+novel look at CGP's replacement behavior.  Each time an offspring replaced a parent
+in the final runs, we recorded how the behavior of each node differed between the
+parent and the offspring.
+
+The \textbf{Active Unchanged} reports how often all active nodes in the offspring
+have identical \emph{footprints} to the corresponding nodes in the offspring's parent.
+This goes beyond the concept of actively identical offspring and checks of the behavior
+of any active node was modified by the mutation.  The rows in each of the four
+tables are sorted using this column.  Smaller values indicate an algorithm
+combination that produced more changes in active node behavior.
+
+The \textbf{No Reactivation} column uses the word reactivation to refer to nodes
+that were active in some ancestor of the offspring, were inactive in the offspring's
+parent, but are active in the offspring.  Reactivation is a measure of how
+often CGP reuses previously useful active nodes.  Reported in the table is how
+frequently an offspring replaces its parent with no reactivated nodes.  Smaller
+values indicate an algorithm combination that reuses nodes, as opposed to
+including randomly generated nodes (never active nodes) or ignoring inactive nodes
+entirely.
+
+The \textbf{Reactivated Changed} column states the percentage of reactivated nodes
+that have a different \emph{footprint} from the last time they were active.  Conversely,
+reactivated nodes without a change mean that CGP has taken a previously useful behavior,
+stored it in the inactive space, and reactivated it in later mutations.  Smaller values
+indicate an algorithm combination that more often preserve node behaviors while
+inactive.
+
+Combined these three measures provide somewhat surprising evidence.  Most notably
+is that \emph{Single}, a technique designed to force offspring to be actively different
+than their parents, produces offspring with no changes in active node behavior
+20\%-40\% of the time.  The further exposes CGP's ability to have genetically different
+yet phenotypically identical individuals.  Even more surprising is that \emph{Single}
+appears to increase the chance of producing offspring without changing active behavior.
+Somewhat unexpectedly \emph{Normal} appears to also reduce changes to active node behavior,
+causing the most effect after \emph{Single}.
+
+The likely cause of both is the number
+of active nodes receiving mutations each generation.  Even though \emph{Single}
+forces a gene in an active node to be mutated, it limits mutation to only one change.
+If this one change does not change that node's behavior, the offspring will be marked
+as active unchanged.  Conversely, \emph{Skip} and \emph{Accumulate} can mutate any
+number of active genes each generation, improving their chance of a change in behavior.
+Similarly, even though \emph{Normal} makes no explicit change in mutation, it is characterized
+by a significantly smaller number of active nodes than \emph{Reorder} or \emph{DAG}.  More
+active nodes means more chances for mutation to change the behavior of at least one.
+
+Somewhat unintuitively, node reactivation and active node \emph{footprint} changes appear
+to be highly correlated.  This is apparent in the fact that the \textbf{No Reactivation}
+column in each table is almost perfectly sorted, even though rows were sorted using the
+\textbf{Active Unchanged} column.  Again the number of active nodes seems the likely
+cause.  Having fewer active nodes reduces the number of introns, making it more
+difficult for inactive nodes to become active.  Fewer active nodes also decreases
+how many inactive nodes could have ever been active.
+
+Potentially the most interesting results come from the control combination of
+\emph{Skip Normal}.  As problem difficulty increases, \emph{Skip Normal} produces
+an increasingly large number of offspring with no change in active node behavior,
+with little to no node reactivation, and a high chance of changing nodes before
+they are reactivated.  This suggests two important features: \emph{Skip} is likely
+saving a significant number of evaluations, and \emph{Normal} CGP is likely using
+inactive nodes primarily to inject random behavior.  The former comes from the
+argument that a large part of the active unchanged individuals are also likely
+actively identical to their parents.  The latter is supported by the fact that
+\emph{Skip Normal} so infrequently reactivates a node with tested behavior.  More
+often than not offspring connecting in inactive nodes are testing that node's behavior
+for the first time.
 

### 2013-06-20 CGP Journal/main.tex
Compressed the tables, fixed references

@@
 \begin{abstract}
-The abstract goes here.
+TODO The abstract goes here.
 \end{abstract}
@@
 \begin{IEEEkeywords}
-TODO Keywords
+Cartesian genetic programming (CGP), Analysis
 \end{IEEEkeywords}
@@
 	  \hline
+	  \multicolumn{9}{|c|}{\textbf{3-bit Parity}, Kruskal-Wallis=$5.99\cdot 10^{-14}$} \\ \hline
 \textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
@@
 \emph{Single} & \emph{Reorder}    & 200 & NA & 1,506 & 823 & 57 & 18 & 0 \\ \hline
-\emph{Single} & \emph{Dag}        & 100 & NA & 2,569 & 1,313 & 43 & 15 & 0 \\ \hline
-	\end{tabular}
-	\caption{Parity, 5.9956882435042246e-14}
-	\label{tab:parity}
-\end{table*}
-
-\begin{table*}
-	\centering
-	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
-	  \hline
+\emph{Single} & \emph{Dag}        & 100 & NA & 2,569 & 1,313 & 43 & 15 & 0 \\ \hline\hline
+
+	  \multicolumn{9}{|c|}{\textbf{16-4-bit Encode}, Kruskal-Wallis=$1.19\cdot 10^{-6}$} \\ \hline
+
 \textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
@@
 \emph{Single} & \emph{Normal}      & 2,000  & NA    & 31,057 & 14,475 &   141 & 43 & 0.0085 \\ \hline
-\emph{Single} & \emph{Dag}         & 100    & NA    & 34,208 & 10,069 &    59 & 34 & 0 \\ \hline
-	\end{tabular}
-	\caption{Encode 1.1926088284917749e-06}
-	\label{tab:encode}
-\end{table*}
-
-\begin{table*}
-	\centering
-	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
-	  \hline
+\emph{Single} & \emph{Dag}         & 100    & NA    & 34,208 & 10,069 &    59 & 34 & 0 \\ \hline\hline
+
+	  \multicolumn{9}{|c|}{\textbf{4-16-bit Decode}, Kruskal-Wallis=$1.02\cdot 10^{-36}$} \\ \hline
+
 \textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
@@
 \emph{Accumulate} & \emph{Dag}     &  5,000 & 0.002 & 133,188 & 29,326 & 2,542 & 125 & 0 \\ \hline
-\emph{Single} & \emph{Dag}         &    500 &    NA & 160,696 & 39,242 &   299 &  97 & 0 \\ \hline
-	\end{tabular}
-	\caption{Decode 1.0182688124966139e-36}
-	\label{tab:decode}
-\end{table*}
-
-\begin{table*}
-	\centering
-	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
-	  \hline
+\emph{Single} & \emph{Dag}         &    500 &    NA & 160,696 & 39,242 &   299 &  97 & 0 \\ \hline\hline
+
+	  \multicolumn{9}{|c|}{\textbf{3-bit Multiply}, Kruskal-Wallis=$5.14\cdot 10^{-18}$} \\ \hline
+
 \textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
@@
 	\end{tabular}
-	\caption{Multiply 5.1429910779824196e-18}
-	\label{tab:multiply}
+	\caption{Results for algorithm configuration on each problem.  Highlighted row
+	is the control configuration \emph{Skip Normal}.  MES=Median Evaluations to Success; MAD=Median
+	Absolute Deviation; p-value=Result from Mann-Whitney~U comparison with control.
+	}
+	\label{tab:results}
 \end{table*}
@@
 The results from running each tuned parameter configuration for each algorithm
-combination on each problem are summarized by Table~\ref{tab:parity},
-Table~\ref{tab:encode}, Table~\ref{tab:decode}, and Table~\ref{tab:multiply}.
+combination on each problem are summarized by Table~\ref{tab:results}.
 Rows in each table are ordered by that algorithm combination's median evaluations to
@@
 the ordering methods are more important on harder problems.  This comes from the
-observation that in for Parity and Encode (Table~\ref{tab:parity} and Table~\ref{tab:encode})
+observation that in for Parity and Encode (Upper half of Table~\ref{tab:results})
 the first column is effectively sorted, while for Decode and Multiply
-(Table~\ref{tab:decode} and Table~\ref{tab:multiply}) the second column is sorted.
+(Lower half of Table~\ref{tab:results})
+the second column is sorted.
 Each table includes coincidentally one exception to these rules.  On Parity and
@@
 number of active nodes across all problems and all algorithm configurations
-in Table~\ref{tab:parity}, Table~\ref{tab:encode}, Table~\ref{tab:decode}, and Table~\ref{tab:multiply}
+in Table~\ref{tab:results}
 was significantly lower than the genome size.  With the exception of a few combinations
@@
 purpose each node in the genome serves.
-Table~\ref{tab:parity}, Table~\ref{tab:encode}, Table~\ref{tab:decode},
-and Table~\ref{tab:multiply} provide two column relevant to node behavior: Active
+Table~\ref{tab:results} provides two columns relevant to node behavior: Active
 and Used.  The former gives the traditional measure of the median amount of
@@
 	  \hline
+	  \multicolumn{5}{|c|}{\textbf{3-bit Parity}} \\ \hline
 \textbf{Duplicate} & \textbf{Ordering} & \textbf{Active Unchanged} & \textbf{No Reactivation} & \textbf{Reactivated Changed} \\ \hline
@@
      \emph{Single} &  \emph{Normal} & 26.44\% & 83.55\% & 59.96\% \\ \hline
-     \emph{Single} &     \emph{DAG} & 27.80\% & 83.73\% & 59.92\% \\ \hline
-	\end{tabular}
-	\caption{Parity}
-	\label{tab:parity_reactivate}
-\end{table*}
-
-\begin{table*}
-	\centering
-	\begin{tabular}{|c|c|c|c|c|}
-	  \hline
+     \emph{Single} &     \emph{DAG} & 27.80\% & 83.73\% & 59.92\% \\ \hline\hline
+
+  \multicolumn{5}{|c|}{\textbf{16-4-bit Encode}} \\ \hline
+
 \textbf{Duplicate} & \textbf{Ordering} & \textbf{Active Unchanged} & \textbf{No Reactivation} & \textbf{Reactivated Changed} \\ \hline
@@
      \emph{Single} &     \emph{DAG} & 35.98\% & 80.91\% & 58.39\% \\ \hline
-     \emph{Single} & \emph{Reorder} & 39.95\% & 78.91\% & 59.72\% \\ \hline
-	\end{tabular}
-	\caption{Encode}
-	\label{tab:encode_reactivate}
-\end{table*}
-
-\begin{table*}
-	\centering
-	\begin{tabular}{|c|c|c|c|c|}
-	  \hline
+     \emph{Single} & \emph{Reorder} & 39.95\% & 78.91\% & 59.72\% \\ \hline\hline
+
+  \multicolumn{5}{|c|}{\textbf{4-16-bit Decode}} \\ \hline
+
 \textbf{Duplicate} & \textbf{Ordering} & \textbf{Active Unchanged} & \textbf{No Reactivation} & \textbf{Reactivated Changed} \\ \hline
@@
    \rowcolor{Gray}
-       \emph{Skip} &  \emph{Normal} & 65.47\% & 89.17\% & 72.21\% \\ \hline
-
-	\end{tabular}
-	\caption{Decode}
-	\label{tab:decode_reactivate}
-\end{table*}
-
-\begin{table*}
-	\centering
-	\begin{tabular}{|c|c|c|c|c|}
-	  \hline
+       \emph{Skip} &  \emph{Normal} & 65.47\% & 89.17\% & 72.21\% \\ \hline\hline
+
+  \multicolumn{5}{|c|}{\textbf{3-bit Multiply}} \\ \hline
+
 \textbf{Duplicate} & \textbf{Ordering} & \textbf{Active Unchanged} & \textbf{No Reactivation} & \textbf{Reactivated Changed} \\ \hline
@@
 	\end{tabular}
-	\caption{Multiply}
-	\label{tab:multiply_reactivate}
+	\caption{Behavior of offspring replacement for each algorithm combination
+	         on each problem.}
+	\label{tab:reactivate}
 \end{table*}
@@
 
-Table~\ref{tab:parity_reactivate}, Table~\ref{tab:encode_reactivate},
-Table~\ref{tab:decode_reactivate}, and Table~\ref{tab:multiply_reactivate} provide a
+Table~\ref{tab:reactivate} provides a
 novel look at CGP's replacement behavior.  Each time an offspring replaced a parent
@@
 
+\begin{comment}
 \appendices
@@
 
+\end{comment}
 % use section* for acknowledgement

### 2013-06-20 CGP Journal/main.tex
Added conclusion

@@
 
-\subsection{Source of High Variance}
-TODO Look for general indicators of high evaluation runs.
-
-TODO Look for footprint moving.
-
-TODO Look for active node clustering.
-
-TODO Look into how often children replace parents.
-
 % An example of a floating figure using the graphicx package.
@@
 \label{sec:conclusion}
-TODO The conclusion goes here.
+Even with CGP's relatively simple evolutionary mechanisms, fully understanding
+the effects those mechanisms have on search takes concerted effort.  By examining
+the effectiveness as well as the actual behavior of traditional CGP and
+a handful of variants we have helped advance this understanding.
+
+Generally, CGP appears to require the least number of evaluations when the genome
+size far exceeds what is necessary for the problem at hand.  When properly tuned,
+almost all variants had only a tiny fraction of the genome active in their final
+solution.  Furthermore, a significant portion of these active nodes were found
+to be redundant, removable using a very simple reduction algorithm.  The only
+variant that preferred more reasonable genome sizes were those using \emph{Single},
+especially when paired with \emph{Reorder} or \emph{DAG}.  Yet we theorize this may
+have more to do with \emph{Single} scaling poorly with the number of active nodes
+than with it overcoming whatever causes traditional CGP to prefer enormous genomes.
+While \emph{Reorder} and \emph{DAG} were able to use smaller genomes than \emph{Normal}
+when paired with \emph{Single}, this decrease in size was not observed in any other
+pairings.  As such, the suggestion in~\cite{goldman:2013:ordering} that these
+techniques generally reduce genome size should be discarded.  Instead it is likely
+more correct to say that \emph{Reorder} and \emph{DAG} are less detrimentally effected
+by reducing the genome size.
+
+Compared on median evaluations to success, \emph{Normal} and \emph{Reorder} are
+likely to be equally good for simple problems, with some suggestion that \emph{Reorder}
+performs better as problems become more difficult.  In general \emph{DAG} was found
+to be unnecessarily time consuming, as it required higher algorithmic complexity
+without any performance gain.  As such we would suggest \emph{Reorder} for general
+future CGP use.
+
+In line with the results from~\cite{goldman:2013:cgpwaste}, \emph{Skip} and
+\emph{Accumulate} continued to show no significant difference in effectiveness
+or behavior.  As \emph{Skip} is simpler to understand and implement, we suggest
+\emph{Accumulate} be ignored in all future CGP applications.  \emph{Single}'s
+behavior continues to be enough different from \emph{Skip} to warrant further
+study, especially with problems requiring higher numbers of active genes.  In
+general though, if evaluation time is significantly impacted by the number of
+active nodes or the total size of the genome, we suggest \emph{Single}.  While
+mutation rate appears relatively variant and problem independent (all of our tuning
+resulted in 0.002 being optimal or close to it), \emph{Single} still has the
+advantage of not requiring this parameter at all.
+
+As to actual behavior, CGP genomes include a surprising amount of redundant and
+unused nodes.  As theoretically predicted in~\cite{goldman:2013:ordering},
+\emph{Normal} ordering leaves large sections of the genome completely unused
+while \emph{Reorder} and \emph{DAG} are more likely to utilize the whole genome.
+Further analysis showed that CGP, at least with our function set, produced a large
+number of constant nodes, such that their output is independent of problem
+inputs.  As the binary domain only has two constants (always true and always false),
+each genome at most should require two constant nodes.  Even so these behaviors
+were found duplicated in both the active and inactive portions of final solutions
+for all algorithm variants.  Also common was duplication of node behavior, even
+though CGP's encoding does not require any duplication to produce any value.  Yet
+attempting to prevent or remove this behavioral duplication may hinder CGP's evolutionary
+success, as it will limit neutral drift.
+
+CGP's use of neutrality has become one of its cornerstone arguments for its success.
+Yet the reason for neutrality's importance appears to be different from previous
+arguments.  In our testing of traditional CGP we found large sections of the
+genome were never used by any ancestor of the final solution.  Furthermore,
+offspring almost never include active nodes that were inactive in their direct
+parent, but active in a previous ancestor.  Even when reactivation does happen,
+the node has almost always changed behavior significantly from when it was active.
+As such we suggest the advantages of neutrality in regards to inactive nodes is
+not that it allows the storage of previously useful structures, but that it allows
+inactive nodes to be constantly changed facilitating mutational inclusion of what
+are effectively random nodes.  Under this theory, CGP would likely benefit from randomly resetting
+all inactive nodes just before they become active.  Note that with \emph{Reorder}
+and \emph{DAG} these conclusions do not hold as a reasonable amount of node
+reactivation occurs.
+
+While this analysis was limited to boolean problems by necessity of some of
+the analysis methods, most of the conclusions have no obvious reason to be
+limited to the boolean domain.  Still, applying \emph{footprint} analysis
+to detect changes in behavior and behavior duplication, along with looking at
+never active nodes and tuned parameter settings, on other domains will help
+solidify our conclusions.
 

### 2013-06-20 CGP Journal/main.tex
Wrote abstract.  Only remaining section is scalability

@@
 \begin{abstract}
-TODO The abstract goes here.
+Understanding how search operators interact with solution representation is
+a critical step to search improvement.  For many search algorithms this may
+be trivial, but in Cartesian Genetic Programming (CGP), and Genetic Programming (GP)
+in general, the complex genotype to phenotype map makes achieving understanding
+challenging.  By examining tuned parameter values, variant success on different
+problem difficulties, node behavior, and offspring replacement properties we
+seek to comprehend observed characteristics of CGP search.  We focus our study
+around methods to prevent wasted CGP evaluations (\emph{Skip}, \emph{Accumulate},
+and \emph{Single}) and methods to overcome CGP's search limitations imposed by
+genome ordering (\emph{Reorder} and \emph{DAG}).
+
+We find strong evidence that currently most CGP techniques evolve genomes that
+are highly inactive, very redundant, and full of seemingly useless constants.
+In some instances we found less than 1\% of the genome was actually required to
+encode the evolved solution.
+Without \emph{Reorder} or \emph{DAG} almost the entire latter portion of the genome
+is never used by any ancestor of the evolved solution.  More generally, our
+results lead us to conclude that \emph{Skip Reorder} and \emph{Single Reorder}
+are most likely to solve hard problems in the least number of evaluations and
+the least amount of time, respectively, with the least amount of degenerate
+behavior.
+
 \end{abstract}

### 2013-06-20 CGP Journal/main.tex
First edit pass - Stopped before DAG

@@
 \IEEEPARstart{W}{hile effective} evolutionary methods are often simple to design
-and implement, it can sometimes be difficult to discern the reason a technique
-or behavior in that system leads to effective search.  This can be exceptionally
-true for Genetic Programming (GP) systems, as they often include nontrivial
+and implement, discerning the reason that system
+leads to effective search can be challenging.
+This is exceptionally
+true for Genetic Programming (GP), which often employs nontrivial
 genotype to phenotype maps and operators that act on genotypes without regard to
@@
 
+There have been a number of previous studies into various aspects of GP evolution.
+For instance~\cite{daida3:2003:treebias} looked into a tree's shape effects is evolvability
+regardless of phenotype,
+which helped inspire the use of new grammar based operations~\cite{xuan:2006:grammar}.
+Analysis of the root cause of bloat has helped to inform methods for controlling
+bloat~\cite{luke:2006:bloat}.
+
 In this work we set out to develop a deeper understanding of what makes Cartesian
-Genetic Programming (CGP) an effective evolutionary optimizer.  Our primary focus is
-on how mutation and genome ordering interact.  In Section~\ref{sec:cgp} we provide
+Genetic Programming (CGP) an effective evolutionary optimizer.
+A few studies of CGP's evolutionary mechanisms have
+been done previously, related to bloat~\cite{miller:2001:bloat},
+neutrality~\cite{vassilev:2000:neutrality}, and structural bias~\cite{payne:2009:bias}.
+Our primary focus is
+on how mutation and genome ordering interact.
+
+In Section~\ref{sec:cgp} we provide
 an introduction to CGP, with sufficient background to put later sections into
@@
 
-There have been a number of previous studies into various aspects of GP evolution.
-For instance~\cite{daida3:2003:treebias} looked into how trees of specific shapes,
-which helped inspire the use of new grammar based operations~\cite{xuan:2006:grammar}.
-Analysis of the root cause of bloat has helped to inform methods for controlling
-bloat~\cite{luke:2006:bloat}.  A few studies of evolutionary mechanisms have also
-been done in CGP directly, related to bloat~\cite{miller:2001:bloat},
-neutrality~\cite{vassilev:2000:neutrality}, and structural bias~\cite{payne:2009:bias}.
-
 \section{Cartesian Genetic Programming}
@@
 Cartesian Genetic Programming (CGP) was originally proposed as a method for general
-genetic programming in~\cite{miller:2000:CGPorigin}.  While its beginnings and
-its name come from evolving circuits on a two dimensional grid, modern CGP can
+genetic programming in~\cite{miller:2000:CGPorigin}.  While its name comes from
+its original application evolving circuits on a two dimensional grid, modern CGP can
 represent any directed acyclic graph (DAG), with applications in evolving binary
@@
 that the output of a node can be reused by any number of other nodes as input values.
-CGP's requirement that nodes connect forward in the genome prevents this ability
-from causing cycles.
+CGP's requirement that nodes connect forward in the genome prevents the formation
+of cycles.
 
@@
 consideration if genes are part of an active or inactive node.  This has the
-potential to lead offspring who's only mutated genes are part of inactive nodes.
+potential to create offspring who's only mutated genes are part of inactive nodes.
 These offspring are actively identical (contain identical active genes) to their
@@
 creating an offspring which is actively identical to its parent is very low.
-Yet such configuration make make it difficult for CGP to optimize, as mutation
+Yet such a configuration may make it difficult for CGP to optimize, as mutation
 is likely to change many active genes in each application.  As was discussed
@@
 the corresponding gene in the offspring's parent.  If all genes are found to be equal
-than the individual is not evaluated.  Instead the offspring is given the same fitness
+than the offspring is not evaluated.  Instead the offspring is given the same fitness
 as its parent, as there is no way for it to be phenotypically different from its parent.
@@
 more and more time is spent attempting to produce an evaluable useful offspring.
-Note these penalties also exist in normal CGP.
+Note that these penalties also exist in normal CGP.
 
@@
 Similar to \emph{Skip}, \emph{Accumulate} works by adding a step between offspring
-creation and offspring evaluation.  Instead of skipping evaluations with offspring
+creation and offspring evaluation.  Instead of skipping evaluations when offspring
 are determined to be actively identical to their parents, \emph{Accumulate} enters
@@
 an actively identical offspring is likely to be selected as we expect mutations to
-active genes to more often reduce fitness than in prove it.  In the next generation
+active genes to more often reduce fitness than improve it.  In the next generation
 this actively identical offspring is then mutated again, with its lineage likely
@@
 While \emph{Skip} and \emph{Accumulate} focus on what to do when offspring are
-produced which are actively identical to their parents, \emph{Skip} changes
+produced which are actively identical to their parents, \emph{Single} changes
 how mutation works to ensure only evaluable offspring are created.  Instead
@@
 forms of duplicate avoidance.  First, \emph{Single} forces offspring to have
-an active gene which is different than there parent.  This limits drift as the
+an active gene which is differs from their parent's gene.  This limits drift as the
 mutated active gene must either be to an intron or represent another way to
@@
 In encodings without inactive genes, limiting mutation to changing exactly one gene
-could prevent an algorithm from escaping some types of local maxima.  Yet because
+could prevent an algorithm from escaping some types of local optima.  Yet because
 CGP allows for inactive genes, \emph{Single} is still able to escape most local
-maxima using sufficient drift of inactive genes.  When a high percentage of
+optima using sufficient drift of inactive genes.  When a high percentage of
 the genome becomes active or when \emph{Single} is otherwise limited in its ability
@@
 As was discussed in Section~\ref{sec:cgp}, CGP uses node ordering in the genome
-to prevent cycles, as nodes are only able to receive input from sources that
+to prevent cycles, by ensuring nodes only receive input from sources that
 precede them in the genome.  This restriction does not limit CGP's ability
@@
 Primarily, enforcing node ordering adds artificial limitations CGP's ability to
-connect nodes.  Beyond preventing cycles, it also prevents nodes from connecting
-which would not create a cycle based solely on the evolutionary happenstance of
-where those nodes were evolved.  Similarly, it is impossible for adjacent nodes
+connect nodes.  Nodes which can be connected without creating a cycle may
+still be prevented from forming that connection because evolutionary happenstance
+has placed one behind the other.
+Similarly, it is impossible for adjacent nodes
 to ever be connected through an intermediate node.  As a result, useful structures
@@
 
-To rectify these issues an examine CGP without ordering bias, \cite{goldman:2013:ordering}
+To rectify these issues and examine CGP without ordering bias, \cite{goldman:2013:ordering}
 created \emph{Reorder} and \emph{DAG} as methods for genome ordering, and defined
@@
 impacting gene expression.  This is possible because for a given node, there can
-be a large number of other nodes which neither have no required ordering,
-as they neither directly or indirectly take input from the node nor provide their
+be a large number of other nodes which have no required ordering,
+as they neither take input (directly or indirectly) from the node nor provide their
 output to the node.  In general, this process works by assigning nodes new
@@
 we can determine in constant time where it gets its input and which nodes read
-its output.  These structures can be build in order $O(N)$ time, where $N$ is the
-number of nodes.
-
-With this preprocessing complete, the algorithm given in Figure~\ref{fig:reorder}
+its output.  These structures can be build in order $O(AN)$ time, where $A$ is the arity used by the nodes
+and $N$ is the number of nodes.
+
+With this preprocessing complete, the \Call{RandomSerialization}{} algorithm given in Figure~\ref{fig:reorder}
 is used to assign each node a new location in the modified genome.
@@
 randomly removed from \emph{addable}, assigned the next possible location
-in the genome, with new nodes added to \emph{addable} as it becomes semantically
+in the genome, with new nodes added to \emph{addable} as they becomes semantically
 viable.  Iteration ends when all nodes have been assigned new locations.
-This algorithm requires $O(AN)$, where $A$ is the arity used by the nodes
-and $N$ is the number of nodes.
+\Call{RandomSerialization}{} requires $O(AN)$, as there can be only $N$ iterations,
+with each iteration updating a distinct subset of the $AN$ dependencies.
 
@@
 forward manner by converting connection genes and output location genes using
-the $new\_loc$ map returned from the algorithm in Figure~\ref{fig:reorder}.
+the $new\_loc$ map returned from \Call{RandomSerialization}{}.
 This also requires $O(AN)$ time, as all connection genes in all nodes must
@@
 no previous steps were of higher complexity, it is also the complexity of
-the algorithm as a whole.  Furthermore it is of the same complexity class
+the \emph{Reorder} algorithm as a whole.  Furthermore, it is of the same complexity class
 of copying the entire genome, meaning it does not change CGP's overall complexity.
 
-\emph{Reorder} is used once each generation to shuffle the nodes of the parent
-organism.  As the shuffling does not semantically change the parent, it does
+\emph{Reorder} is used once each generation to shuffle the nodes of the parent's
+genome.  As the shuffling does not semantically change the parent, it does
 not need to be reevaluated.  Furthermore, as shuffling does not change the requirement
@@
 $X$ and $Y$, and mutating $X$ to depend on $Y$ would improve fitness, mutation's
-ability to make that improvement would depend on happenstance.  If $X$ evolved
+ability to make that improvement in \emph{Normal} would depend on happenstance.  If $X$ evolved
 preceding $Y$, mutation would have to recreate either $X$ or $Y$ in new locations

### 2013-06-20 CGP Journal/main.tex
Edits up to Section VI: Analysis

@@
 are not dependent on the mutating node.  Nodes are then
-tested in a random order for dependence using the algorithm given in Figure~\ref{fig:dag}.
-The first node returned that is not dependent on the mutating node is then
+tested in a random order for dependence using the \Call{IsDependent}{} algorithm given in Figure~\ref{fig:dag}.
+The first node returned at the top level that is not dependent on the mutating node is then
 used as the new connection gene value.
@@
 
-The algorithm given in Figure~\ref{fig:dag} is a non-repeated recursive depth
+\Call{IsDependent}{} is a non-repeated recursive depth
 first search of the individual's DAG.  Search terminates as soon as a dependent
@@
 to examine all nodes in the genome to determine if $working$ is dependent on the
-mutating node, giving it a complexity of $O(N)$.  In the worst case, this algorithm
+mutating node, giving it a complexity of $O(AN)$.  In the worst case, this algorithm
 may be called on all possible nodes in the genome to check their dependence.
-Yet cumulatively those calls can only take $O(N)$ time due to the non-repeating
+Yet cumulatively those calls can only take $O(AN)$ time due to the non-repeating
 nature of the search.  As a result, the expected worst case run time for \emph{DAG}
-mutation is $O(mAN^2)$, as when $m$ is the mutation rate we expect $mAN$
-connection genes to be mutated, each requiring a worst case of $N$ dependency checks.
+mutation is $O(mA^2N^2)$, as when $m$ is the mutation rate we expect $mAN$
+connection genes to be mutated, each requiring a worst case of $N$ dependency checks
+each taking $A$ time.
 This is in contrast to \emph{Normal} and \emph{Reorder} which require $O(mAN)$ to perform
@@
 
-The algorithm given in Figure~\ref{fig:reorder} can be reused to determine the order
+\Call{RandomSerialization}{} can be reused to determine the order
 in which nodes should be executed during evaluation.  Instead of using $new\_loc$
@@
 preprocessing.  Furthermore, once the preprocessing step is done, \emph{DAG} takes
-identical time to evaluate than the other methods, regardless of the number of
+identical time to evaluate as the other methods, regardless of the number of
 input combinations the individual is evaluated on.
@@
 higher than for $Y$'s dependencies.  As a result we expect $X$ to be put into $addable$ sooner,
-and as a result we expect it to be added sooner.  The greater the discrepancy in these
+and therefore we expect it to be added sooner.  The greater the discrepancy in these
 set sizes the less likely the order between the nodes will be reversed.
@@
 \label{sec:quality}
-While it is possible to theorize the effects of each CGP variant on search, and
+While it is possible to theorize about how the effects of each CGP variant on search and
 how those effects may interact with each other, without empirical evidence it is
@@
 
-In total this creates 9 algorithmic configurations.  Of these previous work
+In total this creates 9 algorithm combinations.  Of these previous work
 has been done testing the duplication methods with \emph{Normal} ordering~\cite{goldman:2013:cgpwaste}
@@
 allows for some of the precise analysis performed in Section~\ref{sec:analysis}.
-For all problems, we used the function set \{and, or, nand, nor\}.
+For all problems, we used the function set \{AND, OR, NAND, NOR\}.
 To cover a range of different binary problems, each of the four chosen have
@@
 As representatives of binary problems with varying input and output sizes, we
-reuse the 4-16-bit Decode and 16-4-bit Encode proposed in~\cite{goldman:2013:cgpwaste}.
+reuse the 16-4-bit Encode and 4-16-bit Decode proposed in~\cite{goldman:2013:cgpwaste}.
 In these problems, the circuit evolved must either convert a 4-bit encoded integer
@@
 \label{sec:parameter}
-In order to fairly compare each method, we should ensure proper parameter configuration.
+To perform a fair comparison, we should ensure proper parameter configuration.
 This is to avoid the potential criticism that the parameter settings chosen benefit
@@
 with all possible genome sizes, resulting in 64 potential configurations for each
-of the algorithmic configurations.  Finally note that as \emph{Single} does not
+of the algorithm combinations.  Finally note that as \emph{Single} does not
 use a mutation rate, only the eight configurations with different genome sizes were
@@
 
-To choose which parameter configuration to use with each algorithmic configuration
-for each problem, we used an iterative process of performing runs, comparing means, removing configurations,
-and repeating.  For each algorithm configuration and problem, we first run all
+To choose which parameter configuration to use with each algorithm combination
+for each problem, we used an iterative process of performing runs, comparing medians, removing configurations,
+and repeating.  For each algorithm combination and problem, we first ran all
 potential parameter configurations 5 times (320 runs if not using \emph{Single}, 40 otherwise).
 The parameter configurations are then sorted based on their median.  The best half are then
-each run 4 more times, with the best half of that group chosen.  In this way, 32 configurations
-are run 5 times, 16 are run 9 times, 8 are run 13 times, 4 are run 17 times, 2 are run 21 times,
-and 2 are run 25 times.  For configuration with single, all 8 parameter configurations are run 17 times
+each run 4 more times, with the best half of that group chosen.  In this way, 64 configurations
+are run 5 times, 32 are run 9 times, 16 are run 13 times, 8 are run 17 times, 4 are run 21 times,
+and 2 are run 25 times.  For algorithm combinations using \emph{Single}, all 8 parameter configurations were run 17 times
 before any are removed.  This iterative reduction allows us to focus tuning on
@@
 gathered before choosing between high quality configurations.  In total this means
-3924 complete runs are required to set the parameter configuration for all
-algorithm configurations for a single problem.
+15,696 complete runs are required to set the parameter configuration for all
+algorithm configurations for all problems.
 
@@
 the training runs for that configuration.  As a result, our final results were gather
-from a completely independent set of runs to test the configurations general characteristics.
+from a completely independent set of runs to test the configuration's general characteristics.
 Each configuration was run 50 times to ensure statistical power.
@@
 combination on each problem are summarized by Table~\ref{tab:results}.
-Rows in each table are ordered by that algorithm combination's median evaluations to
-success (MES), with the best algorithm combination appearing at the top of each table.
-The highlighted row marks \emph{Skip Normal}, the control algorithm combination
+Rows for each problem are ordered by that algorithm combination's median evaluations to
+success (MES), with the best algorithm combination appearing at the top of each section.
+The highlighted rows mark \emph{Skip Normal}, the control algorithm combination
 representing traditional CGP's performance.  All configurations for each problem
@@
 some combinations using \emph{Single}, all algorithm combinations used over
-1,000 nodes.  Half of those configurations used 10,000 nodes, the maximum allowed
-by our tuning, which may imply these combinations work best with even larger
+1,000 nodes.  Half of all combinations without \emph{Single} used 10,000 nodes.
+As this was the maximum allowed by our tuning, these combinations may work better with even larger
 genomes.
@@
 On Parity and Encode it is probably safe to conclude that \emph{Single} is worse than
-the control, although as these are the easiest problems tested that difference may have
+the control.  As these are the easiest problems tested that difference may have
 little real world meaning.  \emph{DAG} is almost certainly worse than the control on
@@
 By ordering the algorithm combinations by MES and examining that ordering across each
-problem, we can look for general trends.  From the tables, it appears on easier problems
+problem, we can look for general trends.  From the table, it appears on easier problems
 which method of duplication is in use is the strongest predictor of rank, while

### 2013-06-21 CGP Journal/main.tex
End of edits

@@
 be trivial, but in Cartesian Genetic Programming (CGP), and Genetic Programming (GP)
-in general, the complex genotype to phenotype map makes achieving understanding
-challenging.  By examining tuned parameter values, variant success on different
+in general, complex genotype to phenotype map makes achieving this understanding a
+challenge.  By examining tuned parameter values, variant success on different
 problem difficulties, node behavior, and offspring replacement properties we
@@
 \label{sec:analysis}
-Comparing how many evaluations each algorithm takes to solve a set of benchmark
-problems only provides us with a limited amount of understanding about how
-those algorithms actually function.  We propose here to look for further detail,
-both to check on theorized capabilities of each variant and CGP in general, and
-to look for previously hidden aspects of CGP.
-To do so we have devised a number of novel metrics to examine what makes a
+Comparing how many evaluations each algorithm requires to solve benchmark
+problems provides very little understanding of why one is more efficient than another.
+Thus we must turn to a more detailed view
+to support theorized capabilities of each variant and
+to discover hidden aspects of CGP.
+To accomplish this task we have devised novel metrics to expose what makes a
 CGP run successful.
 
-As a primary concept to our analysis, we determine the \emph{footprint} of nodes
-in the genome.  What we mean by \emph{footprint} is a measure the semantic use
-of a node.  For each input that is given to an individual, all of its nodes
+As a primary concept to our analysis, we determine the semantics of nodes
+in the genome~\cite{mcphee:2008:treesemantics}.
+For each input that is given to an individual, all of its nodes
 will produce a single output.  These output values can be recorded and ordered.
@@
 
-The \emph{footprint} of a node is a complete description of its functionality.
-Any two nodes with identical \emph{footprint} values have identical behavior,
+The semantics of a node is a complete description of its functionality.
+Any two nodes with identical semantic values have identical behavior,
 even if their method for encoding that behavior is drastically different.  From
 this perspective, we can view CGP evolution as attempting to evolve a node who's
-\emph{footprint} matches the desired output \emph{footprint}.  Doing so allows
+semantics match the desired output semantics.  Doing so allows
 us to examine how the evolutionary operators change the behavior of nodes,
@@
 
-Utilizing \emph{footprint} analysis, we can extract the semantically useful
+Utilizing semantic analysis, we can extract the useful
 portions of an individual, a subset of the active nodes.  Consider two active
-nodes with the same \emph{footprint}.  As CGP is a graph representation, only
-one of these nodes is actually necessary.  Given one in the genome, any node that reads
-from the other can have its connection changed without changing its behavior.
+nodes X and Y with the same semantics such that X proceeds Y in the genome.
+As CGP is a graph representation, only
+one of these nodes is actually necessary.  Any node that reads the output of Y
+can instead read the output of X, removing the need to include Y in the solution.
 
@@
   \end{algorithmic}
-  \caption{Algorithm to ensure each active \emph{footprint} is only
-           calculated once.}
+  \caption{Algorithm to ensure each active node calculates
+           a unique semantic.}
   \label{fig:simplify}
@@
 algorithm will move all connections to the first node that produces the desired
-\emph{footprint}.  After application, and recalculation of which nodes are still
-active, each active node produces a unique \emph{footprint}.  Any node that is
-transitively dependent on its own \emph{footprint} will always be removed from
-the active set.  In general, if a set of nodes all produce the same \emph{footprint},
+semantic.  After application, and recalculation of which nodes are still
+active, each active node produces a unique semantic string.  Any node that is
+transitively dependent on its own semantic will always be removed from
+the active set.  In general, if a set of nodes all produce the same semantic,
 whichever has the least transitive dependencies will be kept.  Note that this
-does guarantee the minimum possible genome, an algorithm capable of that requires
+does not guarantee the minimum possible genome, as that would require
 exponential runtime.  The \Call{Simplify}{} algorithm requires $O(AN)$ time,
-as the final for loop is the most expensive step.  As before, this means
+since the final \textbf{for all} loop is the most expensive step.  As before, this means
 \Call{Simplify}{} is no more complex than copying a genome.
-
-By examining which nodes remain active after simplification, we can discuss
-the usefulness of each node as either being part of the minimum solution, a
-duplicate of a node in the minimum solution, or irrelevant to the solution.
-As such we can then examine how redundancy is being used, how frequently useful
-structures exist in the inactive nodes, and how useful nodes are constructed by evolution.
 
@@
 useful.  For a node to be truly of no use to search, we must consider nodes that
-were never active.  Never active nodes are nodes such that for all parents of the
+were never active.  Never active nodes are nodes such that for all ancestors of the
 final solution (the line of descent), no parent had that node active.  Note that
@@
 must have been less fit than their parent.  Even if those offspring are no more fit, the selection method
-used by CGP would have allowed that offspring to replace its parent.  This leads
+used by CGP would have allowed that offspring to replace its parent, at least probabilistically.  This leads
 to two possible causes for a node never being active: bad genes or infrequently tried.
@@
 cause and combined with the theory given in~\cite{goldman:2013:ordering}, we
-expect a strong positional bias in the location of the never active nodes.  Furthermore,
+expect a strong positional bias in the location of the never active nodes;
 we would expect nodes near the end of the genome to be the most likely to be never active.
@@
 In CGP it is common to report what fraction of the genome is active, as this represents
-the easiest estimate of solution size.  By utilizing \emph{footprint} comparison
+the easiest estimate of solution size.  By utilizing semantic comparison
 and the \Call{Simplify}{} algorithm, we can get a much better estimate of what
 purpose each node in the genome serves.
-Table~\ref{tab:results} provides two columns relevant to node behavior: Active
-and Used.  The former gives the traditional measure of the median amount of
+Table~\ref{tab:results} provides two columns relevant to node behavior: \textbf{Active}
+and \textbf{Used}.  The former gives the traditional measure of the median amount of
 active nodes in the solution found.  The latter is the median number of nodes
@@
 On all problems and for all algorithm combinations, the number of active nodes is
-far smaller than the genome size, and the number of Used nodes
-is far smaller than the number of Active nodes.  In some extreme cases less
+far smaller than the genome size, and the number of used nodes
+is far smaller than the number of active nodes.  In some extreme cases less
 than 1\% of active nodes in the evolved solution are actually necessary.  Only the
@@
 
-Each diagram describes node behavior as falling into eight distinct categories.
-The first category are the constant nodes, both active and inactive.  These nodes
-return the same value for all tested inputs.  The ``Used'' bar is the average
-number of nodes which are part of the simplified solution.  The ``Useful'' bars
-express how many nodes, either active or inactive, have identical \emph{footprint}
-values to used nodes.  These nodes represent either independent methods of calculating
-the same behavior or are effectively pass through nodes which mimic the behavior
-of their inputs.  Active ``Intron'' nodes are those active nodes that do not
-fall into either the constant or used categories, with inactive ``Intron'' nodes
-being those inactive nodes that have identical \emph{footprint} values to active
-intron nodes.  The final bar is ``Explore'', which marks how many nodes on
-average produce a \emph{footprint} not part of any previous category.  Note that
+Each diagram describes node behavior as falling into eight distinct categories
+split between active and inactive nodes.  Nodes are put into the first category
+they satisfy.
+\begin{enumerate}
+\item ``Constant'' - Independent of problem input, constant nodes produce a fixed output.
+\item ``Used'' - Nodes still active after applying \Call{Simplify}{}.
+\item ``Useful'' - Nodes with identical semantics to a ``Used'' node.
+\item ``Intron'' - Any node with semantics identical to an active node.
+\item ``Explore'' - All remaining nodes.
+\end{enumerate}
+Used nodes are by definition only in the active portion of the genome.  Similarly
+explore nodes can only be in the inactive portion of the genome.   Note that
 all never active nodes are ignored by these plots, as their behavior is meaningless.
@@
 with some hits of a change on Encode with \emph{Normal}.  These combinations
-were also the most likely to have significantly smaller genome sizes.  From the
-results tables we see that the number of used nodes on each problem is relatively
+were also the most likely to have significantly smaller genome sizes.  From
+Table~\ref{tab:results} we see that the number of used nodes on each problem is relatively
 constant across the combinations even though the number of active nodes and the
@@
 explore nodes.  A potential explanation is that on 16-4-bit Encode there are only
-$2^{16} = 65,536$ \emph{footprints} while on 3-bit Multiply there are $2^{64} = 1.845\cdot 10^{19}$
-\emph{footprints}.  Combined with the fact that not all \emph{footprints} are equally
+$2^{16} = 65,536$ unique semantics while on 3-bit Multiply there are $2^{64} = 1.845\cdot 10^{19}$
+unique semantics.  Combined with the fact that not all semantics are equally
 likely to be produced, the difference may just be that on Encode nodes are
 that much more likely to recreate existing behavior.  Another possibility is that
-CGP increases the diversity of behavior of inactive nodes over evolutionary time.
+CGP increases the semantic diversity of inactive nodes over evolutionary time.
 As such, the fact that Multiply required approximately an order of magnitude more
@@
 Each generation the best offspring replaces its parent if it is no less fit.  Previous
-research has shown the impact removing this feature can have on
+research has shown the negative impact removing this feature can have on
 performance~\cite{yu:2001:neutrality}, but how it interacts with the different variants
@@
 novel look at CGP's replacement behavior.  Each time an offspring replaced a parent
-in the final runs, we recorded how the behavior of each node differed between the
+in 50 runs, we recorded how the behavior of each node differed between the
 parent and the offspring.
 
-The \textbf{Active Unchanged} reports how often all active nodes in the offspring
-have identical \emph{footprints} to the corresponding nodes in the offspring's parent.
-This goes beyond the concept of actively identical offspring and checks of the behavior
+The \textbf{Active Unchanged} column reports how often all active nodes in the offspring
+have identical semantics to the corresponding nodes in the offspring's parent.
+This goes beyond the concept of actively identical offspring and checks if the behavior
 of any active node was modified by the mutation.  The rows in each of the four
@@
 frequently an offspring replaces its parent with no reactivated nodes.  Smaller
-values indicate an algorithm combination that reuses nodes, as opposed to
+values indicate an algorithm combination that frequently reuses nodes, as opposed to
 including randomly generated nodes (never active nodes) or ignoring inactive nodes
@@
 The \textbf{Reactivated Changed} column states the percentage of reactivated nodes
-that have a different \emph{footprint} from the last time they were active.  Conversely,
-reactivated nodes without a change mean that CGP has taken a previously useful behavior,
+that have a different semantic from the last time they were active.  Conversely,
+reactivated nodes without a change mean that CGP has taken a previously active behavior,
 stored it in the inactive space, and reactivated it in later mutations.  Smaller values
@@
 is that \emph{Single}, a technique designed to force offspring to be actively different
-than their parents, produces offspring with no changes in active node behavior
-20\%-40\% of the time.  The further exposes CGP's ability to have genetically different
+from their parents, produces offspring with no changes in active node behavior
+20\%-40\% of the time.  This further exposes CGP's ability to have genetically different
 yet phenotypically identical individuals.  Even more surprising is that \emph{Single}
@@
 
-Somewhat unintuitively, node reactivation and active node \emph{footprint} changes appear
+Somewhat unintuitively, node reactivation and active node semantic changes appear
 to be highly correlated.  This is apparent in the fact that the \textbf{No Reactivation}
@@
 to be redundant, removable using a very simple reduction algorithm.  The only
-variant that preferred more reasonable genome sizes were those using \emph{Single},
+algorithm combinations that preferred more reasonable genome sizes were those using \emph{Single},
 especially when paired with \emph{Reorder} or \emph{DAG}.  Yet we theorize this may
@@
 active nodes or the total size of the genome, we suggest \emph{Single}.  While
-mutation rate appears relatively variant and problem independent (all of our tuning
+mutation rate appears relatively independent of problem and algorithm combination (all of our tuning
 resulted in 0.002 being optimal or close to it), \emph{Single} still has the
@@
 the analysis methods, most of the conclusions have no obvious reason to be
-limited to the boolean domain.  Still, applying \emph{footprint} analysis
+limited to the boolean domain.  Still, applying semantic analysis
 to detect changes in behavior and behavior duplication, along with looking at

### 2013-06-28 CGP Journal/main.tex
First commit of Bill's first edit

@@
 is never used by any ancestor of the evolved solution.  More generally, our
-results lead us to conclude that \emph{Skip Reorder} and \emph{Single Reorder}
+results lead us to conclude that \emph{Skip-Reorder} and \emph{Single-Reorder}
 are most likely to solve hard problems in the least number of evaluations and
@@
 There have been a number of previous studies into various aspects of GP evolution.
-For instance~\cite{daida3:2003:treebias} looked into a tree's shape effects is evolvability
+For instance~\cite{daida3:2003:treebias} showed how a tree's shape effects its evolvability
 regardless of phenotype,
-which helped inspire the use of new grammar based operations~\cite{xuan:2006:grammar}.
+resulting in the use of new grammar based operations~\cite{xuan:2006:grammar}.
 Analysis of the root cause of bloat has helped to inform methods for controlling
@@
 an introduction to CGP, with sufficient background to put later sections into
-context.  Section~\ref{sec:duplicate} explains previous work about mutation's potential
-to create identifiably wasted evaluations, and proposed methods to prevent that
-waste.  Section~\ref{sec:ordering} discusses the impact of CGP's genome ordering,
-with methods to both examine and bypass the limitations imposed by that ordering.
+context.  Section~\ref{sec:duplicate} describes previous work on preventing mutation
+from creating identifiably wasted evaluations.
+Section~\ref{sec:ordering} discusses the impact of CGP's genome ordering,
+along with methods to both examine and bypass the limitations imposed by that ordering.
 Each technique is then rigorously tuned and qualitatively compared on a collection
@@
 its original application evolving circuits on a two dimensional grid, modern CGP can
-represent any directed acyclic graph (DAG), with applications in evolving binary
+represent any directed acyclic graph (DAG), and has been utilized in applications such as binary
 circuits~\cite{walker:2008:cgpmodules},
@@
 CGP represents DAGs using a linear genome of integer values.  Each node in the
-DAG is encoded using a collection of genes, with one gene specifying the function
+DAG is encoded as a tuple of genes, with one gene specifying the function
 the node applies to its inputs, and the remaining genes expressing where the
-node takes input from.  Nodes can take input from any problem input and from
-any node preceding them in the genome.  To complete the representation, a set
+node takes input from.  Nodes can take input from either a problem input or
+any node preceding them in the linear genome.  Restricting connections in this way
+prevents the creation of cycles, while still allowing CGP to reuse values.  This is
+in contrast to tree based GP, which must recalculate a value everywhere it is needed.
+To complete the representation, a set
 of extra genes are included at the end of the genome to specify which nodes or
@@
 
-Unlike tree based GP encodings, CGP is able to perform value reuse.  This means
-that the output of a node can be reused by any number of other nodes as input values.
-CGP's requirement that nodes connect forward in the genome prevents the formation
-of cycles.
-
-As output locations and information flow in the DAG are evolvable, there are often
-large portions of the genome which do not participate in creating any of the
-output values.  These nodes are referred to as inactive, with the nodes currently
-being used to create output values referred to as active.  Inactive nodes
+As both output locations and information flow in the DAG are evolvable, often
+only a tiny fraction of the genome participates in creating the
+output values.  These nodes are referred to as ``active,'' with the nodes not
+being used to create output values referred to as ``inactive.''  Inactive nodes
 allow CGP individuals to drift genetically, as they can be mutated without effecting
@@
 active.  Previous work suggests CGP is most efficient when up to 95\% of the genome
-is inactive~\cite{miller:2006:redundancy}, yet other work suggests this may
+is inactive~\cite{miller:2006:redundancy}, while more recent work suggests this may
 be a result of hidden parsimony pressure in CGP~\cite{goldman:2013:ordering}.
 
-CGP uses very simple evolutionary mechanisms.  By far the most common strategy
-is $1+4$, in which each generation a single parent produces four offspring using
-mutation.  The best offspring then competes with the parent, with the offspring
+CGP uses very simple evolutionary mechanisms.  The most common evolutionary strategy
+is $\mu + \lambda$ where $\mu \leftarrow 1$ and $\lambda \leftarrow 4$.
+This means generation a single parent produces four offspring using
+mutation.  The best offspring then competes with the parent,\footnote{Ties are broken
+randomly.} with the offspring
 replacing the parent if it is no less fit.  This replacement strategy encourages
-drift, as completely neutral mutations are always preferred to stagnation.  Mutation
-is commonly performed by mutating each gene at a set probability, where mutation
-involves changing a gene randomly to some different valid value.  For instance,
+neutral drift.  Mutation
+is commonly performed by mutating each gene of each node at a set probability, where mutation
+involves changing a gene randomly to some different valid value.  For example,
 if a function gene is chosen for mutation, its new value is randomly chosen from
@@
 \label{sec:duplicate}
-CGP's mutation operator is traditionally applied to all genes equally, with no
-consideration if genes are part of an active or inactive node.  This has the
-potential to create offspring who's only mutated genes are part of inactive nodes.
+CGP's mutation operator is traditionally applied to genes in active and inactive
+node's uniformly. This has the
+potential to create offspring who's only mutated genes are in inactive nodes.
 These offspring are actively identical (contain identical active genes) to their
 parents, and therefore by definition have identical fitness to their parents.
-As such, evaluating these individuals is wasteful, as their fitness can be
-assigned without the computational complexity of performing an evaluation.
-Furthermore, calculating which nodes are active before evaluating an individual
-improves evaluation speed~\cite{vasicek:2012:efficient}, meaning detecting when duplicated individuals are
-created can be done without significant computational overhead.
+As such, re-evaluating these individuals is computationally wasteful, as their fitness
+is known.
+Detecting duplicated individuals can be done without significant overhead as
+finding the set of active nodes is already done prior to
+evaluation~\cite{vasicek:2012:efficient}.
 Previous work~\cite{goldman:2013:cgpwaste} has taken an initial look at the
@@
 
-The amount of evaluations wasted on these actively identical individuals is
+The number of wasted evaluations is
 directly related to the mutation rate and the number of active genes.  With
@@
 with its parent.  Each gene in each active node is compared for equivalence with
-the corresponding gene in the offspring's parent.  If all genes are found to be equal
-than the offspring is not evaluated.  Instead the offspring is given the same fitness
-as its parent, as there is no way for it to be phenotypically different from its parent.
+the corresponding gene in the offspring's parent.  If all active genes are found to be equal
+than the offspring is not evaluated and is instead given the same fitness
+as its parent.
 Note that since inactive genes are not compared for equivalence it is still possible
@@
 
-Because \emph{Skip} does not modify any evolutionary mechanisms from normal CGP,
-in all possible cases it can do no worse.  If the mutation rate is high enough relative
-to the number of active genes, \emph{Skip} will act identically to normal CGP as
+Because \emph{Skip} does not modify any evolutionary mechanisms found in traditional CGP,
+it will produce identical individuals, with a potential reduction in evaluations.
+If the mutation rate is high enough relative
+to the number of active genes, \emph{Skip} will require the same number of evaluations as traditional CGP. This is because
 the probability of an offspring being actively identical to its parent is effectively zero.
@@
 
-From a tuning perspective we would expect both intuitively and from initial
+We would expect both intuitively and from initial
 experimentation~\cite{goldman:2013:cgpwaste} that \emph{Skip} will be less sensitive
-to the mutation rate and will perform best using a lower mutation rate than normal CGP.
+to the mutation rate than normal CGP.  We would also expect \emph{Skip} to use a lower
+mutation rate than normal CGP when both are optimized to reduce evaluations.
 The former comes from the fact that in normal CGP any mutation rate that has a significant
-probability of creating offspring actively identical to their parents will be penalized
-by wasting evaluations.  In \emph{Skip} there are only two penalties for reducing the
-mutation rate.  First, if search becomes trapped in a local optima, low mutation
-rates may have increased difficulty escaping to find the global optimum.  Second,
-exceptionally low mutation rates may result in large computational overhead as
-more and more time is spent attempting to produce an evaluable useful offspring.
+probability of creating offspring actively identical to their parents wastes
+evaluations.  In \emph{Skip} there are only two penalties for reducing the
+mutation rate.  From an evolutionary standpoint, if search becomes trapped in a local optima, low mutation
+rates may have increased difficulty escaping to find the global optimum.  From a performance standpoint,
+exceptionally low mutation rates may spend a prohibitive amount of time
+attempting to produce an evaluable offspring.
 Note that these penalties also exist in normal CGP.
@@
 into a cycle of repeated mutation until an individual worth evaluating is created.
-While the working offspring is actively identical to the
-parent, a new individual is produced by applying mutation to the working offspring.
-Note that each mutation step can probabilistically mutate any number of genes, just
-as in the mutation used by normal CGP.
-In this way each iteration the working offspring accumulates mutations to inactive
-genes until an individual is produced that differs by one or more active genes from
-the parent.  This repeatedly mutated offspring is then evaluated.  If its fitness
-is less than the original parent, the final set of mutations are reverted
-and the offspring's fitness is set to the same as the original parent.
-
+To understand this technique, consider an example where the parent $P$ produces
+the offspring $F_0$ using mutation. If $F_0$ is actively identical $P$, $F_0$
+creates its own offspring $F_1$ using mutation.  This process of $F_i$ producing
+$F_{i+1}$ continues until $F_n$ is produced, where $F_n$ is not actively identical
+to $P$.  $F_n$ is then evaluated.  If $F_n$ is no worse than $P$, it replaces
+$F_0$, and evolution continues as though $P$ had produced $F_n$ directly.  If
+$F_n$ is worse than $P$, $F_{n-1}$ replaces $F_0$, as $F_{n-1}$ has accumulated
+the most mutations to inactive genes without reducing fitness.  Note that
+the mutation used each time is identical to normal CGP's mutation, such that
+$F_n$ can have one or more active genes different from $P$.  Furthermore, even
+though this technique produces $n+1$ individuals, only $F_n$ is evaluated.
 Viewed in another way, \emph{Accumulate} performs a micro evolution on each offspring
-before it is fully produced.  The final result of each offspring step is to produce
-an offspring which is either actively identical to its parent, but with mutations
-to its inactive genes, or an offspring which is at least as fit as the parent but
-not actively identical.
+before it is evaluated, with an emphasis on encouraging neutral drift.
 
@@
 \emph{Skip}~\cite{goldman:2013:cgpwaste}, with the exception that \emph{Accumulate}
-favors lower mutation rates. While at first these algorithms may appear
+favors lower mutation rates when tuned to reduce evaluations to success. While at first these algorithms may appear
 quite different, further consideration shows how they are similar.  In \emph{Skip}
@@
 this actively identical offspring is then mutated again, with its lineage likely
-continuing until it finally does mutate an active gene.  At this point if the
+continuing until it finally does mutate one or more active genes.  At this point, if the
 mutant is better it replaces the parent, otherwise the parent, which has been accumulating
 mutations to inactive genes, is kept.  In this way \emph{Skip} mirrors \emph{Accumulate},
-except over multiple generations instead of compressed into a single generation.
+except \emph{Skip} acts over multiple generations instead of compressed into a single generation.
 
 \subsection{Single}
-While \emph{Skip} and \emph{Accumulate} focus on what to do when offspring are
-produced which are actively identical to their parents, \emph{Single} changes
-how mutation works to ensure only evaluable offspring are created.  Instead
-of mutating each gene at a set probability, \emph{Single} chooses genes at
-random to mutate until exactly one active gene is mutated.
+While \emph{Skip} and \emph{Accumulate} save evaluations by not evaluating offspring
+which are actively identical to their parents, \emph{Single} changes
+mutation to ensure that only evaluable offspring are created.  Instead
+of mutating each gene at a set probability, \emph{Single} mutates genes at
+random until exactly one active gene is mutated.
 
@@
 forms of duplicate avoidance.  First, \emph{Single} forces offspring to have
-an active gene which is differs from their parent's gene.  This limits drift as the
+an active gene which differs from their parent.  This limits drift as the
 mutated active gene must either be to an intron or represent another way to
@@
 
-Primarily, enforcing node ordering adds artificial limitations CGP's ability to
+Primarily, enforcing node ordering adds artificial limitations to CGP's ability to
 connect nodes.  Nodes which can be connected without creating a cycle may
-still be prevented from forming that connection because evolutionary happenstance
-has placed one behind the other.
+still be prevented from forming that connection because of random genome ordering.
 Similarly, it is impossible for adjacent nodes
@@
 
-Ordering combined with the random reset mutation method may be the cause of CGP's
+Ordering, when combined with the mutation's uniform resetting of connection genes, may be the cause of CGP's
 immensely inactive genomes.  As was rigorously defined in~\cite{goldman:2013:ordering},
@@
 \label{sec:reorder}
-The concept behind \emph{Reorder} is to randomly shuffled node ordering without
-impacting gene expression.  This is possible because for a given node, there can
+The \emph{Reorder} method shuffles an existing genome's node ordering without
+changing node behavior.  This is possible because for a given node, there can
 be a large number of other nodes which have no required ordering,
@@
 The first step in performing \emph{Reorder} on a genome is to create convenience
-data structures to store direct connection relationships, such that given a node
-we can determine in constant time where it gets its input and which nodes read
-its output.  These structures can be build in order $O(AN)$ time, where $A$ is the arity used by the nodes
+data structures to store direct connection relationships.  These structures are used
+to determine in constant time all connections incident on a node, and 
+can be build in order $O(AN)$ time, where $A$ is the arity of the nodes
 and $N$ is the number of nodes.
@@
 The final step in \emph{Reorder} is to use the list of new locations to convert
-the existing genome into the reordered genome.  This can be done in the straight
-forward manner by converting connection genes and output location genes using
+the existing genome into the reordered genome.  This can be done
+by converting connection genes and output location genes using
 the $new\_loc$ map returned from \Call{RandomSerialization}{}.
@@
 be converted.  As this is the final step unique to \emph{Reorder}, and
-no previous steps were of higher complexity, it is also the complexity of
+no previous steps are of higher complexity, it is also the complexity of
 the \emph{Reorder} algorithm as a whole.  Furthermore, it is of the same complexity class
-of copying the entire genome, meaning it does not change CGP's overall complexity.
+as copying the entire genome, meaning it does not change CGP's overall complexity.
 
@@
 genome.  As the shuffling does not semantically change the parent, it does
-not need to be reevaluated.  Furthermore, as shuffling does not change the requirement
+not require re-evaluated.  Furthermore, as shuffling does not change the requirement
 that nodes only depend on those preceding them in the genome, shuffling does not
 require any changes to any other CGP methods, such as evaluation and mutation.
-Yet after shuffling, the potential mutations that can be applied to the parent
-have changed.  Consider two nodes $X$ and $Y$ which are in the genome, but
+It is important to note, however, that the potential mutations that can be applied to the parent
+have changed.  Consider two nodes $X$ and $Y$ which are in the genome, and
 have no dependence relationship.  As $X$ does not take input from $Y$ and
-$Y$ does not take input from $X$, either can be placed preceding the other
-in the genome.  Without shuffling, the ordering in which these two nodes originally
-evolved will be preserved indefinitely, such that if $X$ precedes $Y$ at node creation,
-$X$ will always precede $Y$.  No mutation in \emph{Normal} is capable of making $X$
-take input from $Y$.  Yet CGP using \emph{Reorder} is capable of shuffling $Y$ to
-precede $X$, allowing subsequent mutations to make the connection.
-
-This node movement has the potential to reduce node reinvention.  If evolution created
-$X$ and $Y$, and mutating $X$ to depend on $Y$ would improve fitness, mutation's
-ability to make that improvement in \emph{Normal} would depend on happenstance.  If $X$ evolved
-preceding $Y$, mutation would have to recreate either $X$ or $Y$ in new locations
-in order to make the required connection.  \emph{Reorder} could achieve the same
-effect through shuffling, and without costly node reinvention.
+$Y$ does not take input from $X$, either can be placed before the other
+in the genome.  Without shuffling, the initial ordering of nodes is fixed.
+No operation in \emph{Normal} is capable of making $X$
+take input from $Y$.  \emph{Reorder}'s shuffling can allow
+subsequent mutations to make the connection.
+
+\emph{Reorder} has the potential to reduce node reinvention.  In \emph{Normal}, if
+$X$ preceded $Y$, mutation would have to recreate either $X$ or $Y$ in new locations
+in order to make the required connection.  \emph{Reorder} provides the opportunity
+to examine different orderings of the genome without such costs.
 
 The price \emph{Reorder} pays for this potential improvement is increased mutational search
-space and minor runtime costs.  Even though \emph{Reorder} is not capable of representing
-different individuals that \emph{Normal}, their mutational adjacency is significantly different.
-By mutational adjacency, we mean the set of individual's which are most likely to be created
-when mutating a given individual.  For instance in \emph{Normal} if $X$ precedes $Y$, individuals
-containing $Y$ dependent on $X$ could be adjacent to the original individual, but those containing
-$X$ dependent on $Y$ would not.  As \emph{Reorder} can shuffle node order, both sets would be
-adjacent.  In this way ordering limits how \emph{Normal} performs its search.  This limitation
-will benefit \emph{Normal} any time making $X$ dependent on $Y$ is detrimental, as
-\emph{Normal} won't waste evaluations on individuals having that mutation.
+space and runtime costs.  Though a shuffled individual is phenotypically unchanged,
+its mutational adjacency is significantly different.  As such the variety of offspring
+an individual can create is increased when using \emph{Reorder}.  While this increase
+makes finding a specific offspring that improves upon the parent less likely, the
+general likelihood of producing an offspring that improves upon the parent may 
+more than offset this drawback.
 
@@
 \label{sec:dag}
-The principle behind \emph{DAG} is to remove the traditional CGP requirement that
-nodes can only take input from nodes that precede them in the genome.  Instead
-this is replaced with the relaxed requirement than the genome cannot encode for
-any cycles.  To accomplish this goal, modifications must be made to how CGP
+\emph{Normal} and \emph{Reorder} use node location in a linear genome to create
+an efficient protection against cycles: nodes must connect forward.  \emph{DAG}
+removes this scheme of location, relying on more complex methodologies to
+prevent cycles.  To accomplish this goal, modifications must be made to how CGP
 performs mutation and evaluation.
 
-When a connection gene is chosen for mutation, its value is still chosen randomly
-from all possible alternative values.  To achieve this, we incrementally determine
+As in \emph{Normal} and \emph{Reorder}, when a connection gene is chosen for mutation, its value is chosen randomly.
+However, the options to choose from are only limited to those connections which do not
+create cycles.
+To achieve this, we incrementally determine
 which nodes transitively depend on the mutating node and which are not, with
-information stored in the $known$ map.  Initially, we know that the node
-is dependent on itself and that input locations and the nodes current inputs
+information stored in the $known$ map.  This data structure is initialized with
+the information that the node is dependent on itself and that all input locations
 are not dependent on the mutating node.  Nodes are then
@@
 The first node returned at the top level that is not dependent on the mutating node is then
-used as the new connection gene value.
+used as the new connection gene value.  As a result, mutating a connection gene
+cannot introduce a cycle.
 
@@
 first search of the individual's DAG.  Search terminates as soon as a dependent
-node is found, and information from previous searches is recorded to prevent
-repeated search.  Each recursion level results in a node having its dependency
-set, meaning early calls will likely result in lots of recursion, but subsequent
-calls will find answers faster.  In the worst case, this algorithm may need
+node is found, and information from previous searches is memoized to prevent
+repeated search.  Each recursion level results in a node being added to $known$.
+As such initial calls will likely result in lots of recursion, but subsequent
+calls will find answers more quickly.  In the worst case, this algorithm may need
 to examine all nodes in the genome to determine if $working$ is dependent on the
-mutating node, giving it a complexity of $O(AN)$.  In the worst case, this algorithm
+mutating node, giving \Call{IsDependent}{} a complexity of $O(AN)$.  In the worst case, this algorithm
 may be called on all possible nodes in the genome to check their dependence.
@@
 nature of the search.  As a result, the expected worst case run time for \emph{DAG}
-mutation is $O(mA^2N^2)$, as when $m$ is the mutation rate we expect $mAN$
+mutation is $O(mA^2N^2)$, where $m$ is the mutation rate.  We expect $mAN$
 connection genes to be mutated, each requiring a worst case of $N$ dependency checks
@@
 The method for evaluating \emph{DAG} individuals is very similar to the method for
-efficiently evaluating \emph{Normal} and \emph{Reorder} individuals.  Just as before,
+efficiently evaluating \emph{Normal} and \emph{Reorder} individuals.  As before,
 preprocessing is done to determine which nodes are active, and only those nodes are
-executed during evaluation.  Just as before the process begins from the output locations,
+executed during evaluation.  Also as before, the process begins from the output locations,
 recursively following connection genes and marking nodes as active until the input locations
-are reached.  Here the process diverges, as in \emph{DAG} we must determine not only the set of
+are reached.  At this point, the process changes.  In \emph{DAG}, we must determine not only the set of
 nodes that are active, but the order in which those nodes should be executed.
@@
 in which nodes should be executed during evaluation.  Instead of using $new\_loc$
-to reorder the genome, we can use this map to specify what order nodes should be
-executed in.  For instance, if a node was given the new location of $X$, we know
+to reorder the genome, we can use this map to specify the order in which nodes should be
+executed.  For instance, if a node was given the new location of $X$, we know
 that once all nodes at locations preceding $X$ have been evaluated, the node at $X$
@@
 The process of determining which nodes are active requires $O(AN)$ time, regardless
-of how the node ordering is handled.  As was discussed in Section~\ref{sec:reorder},
+of node ordering.  As was discussed in Section~\ref{sec:reorder},
 getting the $new\_loc$ map requires $O(AN)$ time.  Finally, converting the map
@@
 identical time to evaluate as the other methods, regardless of the number of
-input combinations the individual is evaluated on.
-
-With these modifications, there is no longer any meaning to the position of a node
+input combinations.
+
+With these modifications, there is no longer any meaning to the location of a node
 in the genome.  In \emph{Normal} a node near the input locations will always
@@
 represented by the individual, independent of the genes location in the genome.
-As such, there are far less artificial limitations on how \emph{DAG} can modify
+As such, there are fewer artificial limitations on how \emph{DAG} can modify
 solutions.
 
-While both reduce mutational limitations, \emph{DAG}'s changes are more broad that
-\emph{Reorder}.  While both have the possibility to use mutation to connect any two
-nodes such that no cycle is created, \emph{Reorder} still maintains a bias about
-how often some types of connection can happen.  Consider again two nodes $X$ and $Y$,
+While \emph{Reorder} and \emph{DAG} both reduce mutational limitations, \emph{DAG}'s changes are more sweeping.
+Though both can create any connect that does not form a cycle,
+\emph{Reorder} maintains a bias toward certain kinds of connections.
+Consider again two nodes $X$ and $Y$
 such that neither is dependent on the other.  If $X$ is transitively dependent on very
-few nodes, and $Y$ is transitively dependent on many nodes, in general \emph{Reorder}
-will shuffle the genome such that $X$ is before $Y$.  This is true because the probability
-that all of $X$'s dependencies are added at any point of building the new order is always
+few nodes, and $Y$ is transitively dependent on many nodes, \emph{Reorder} is more likely
+to shuffle $X$ to precede $Y$.  This is true because at any point during the construction
+of the new ordering, the probability
+that all of $X$'s dependencies have been added is
 higher than for $Y$'s dependencies.  As a result we expect $X$ to be put into $addable$ sooner,
-and therefore we expect it to be added sooner.  The greater the discrepancy in these
+and therefore we expect it to be added sooner.  The greater the discrepancy in the dependency
 set sizes the less likely the order between the nodes will be reversed.
@@
 
-The probability of connecting $X$ to $Y$ is also influenced by how many nodes transitively depend on each.
-In this example, let us assume $X$ has more nodes transitively dependent on it than $Y$ has.
-In \emph{Reorder}, $X$ would again appear closer to the start of the genome, this time because the number
+The probability of connecting $X$ to $Y$ is also influenced by how many nodes transitively depend $X$ and $Y$.
+For example, assume $X$ has more nodes transitively dependent on it than $Y$.
+In \emph{Reorder}, $X$ is again more likely to be placed before $Y$, this time because the number
 of possible nodes preceding it is reduced.  \emph{DAG} is biased in the reverse, such that $X$ will
@@
 \label{sec:quality}
-While it is possible to theorize about how the effects of each CGP variant on search and
-how those effects may interact with each other, without empirical evidence it is
-mostly speculation.  Therefore we propose to start by testing the effectiveness
-of the variants and their combinations in problem solving.  Deeper analysis of
-the causes of their effectiveness are given in Section~\ref{sec:analysis}.
+Empirical results are an important step in confirming our examinations.
+
+We accomplish this by testing the effectiveness
+of the described variants and their combination in problem solving.  Deeper analysis of
+evolutionary side effects are given in Section~\ref{sec:analysis}.
 
@@
 (\emph{Skip}, \emph{Accumulate}, and \emph{Single}) and two new methods for
-how to deal with genome ordering (\emph{Reorder} and \emph{DAG}).  As these two
+genome ordering (\emph{Reorder} and \emph{DAG}).  As these two
 sets are non overlapping, and there is no intuitive groupings, we chose to try
-all possible combinations.  As a control we included \emph{Normal} ordering,
+all possible combinations.  As a control we included \emph{Normal} genome ordering,
 as it provides insight to how each of the duplicate prevention techniques work
-if applied alone.  Similar measures are not needed in testing the ordering techniques,
+if applied alone.  Similar measures are not needed in testing the genome ordering techniques,
 as \emph{Skip} has no impact on evolutionary search, and is a strict improvement
@@
 
-In total this creates 9 algorithm combinations.  Of these previous work
-has been done testing the duplication methods with \emph{Normal} ordering~\cite{goldman:2013:cgpwaste}
+In total this creates 9 algorithm combinations.  Of these, previous work
+has been reported showing the effectiveness of the duplication methods with \emph{Normal} ordering~\cite{goldman:2013:cgpwaste}
 as well as analysis of all three ordering methods with \emph{Single}~\cite{goldman:2013:ordering}.
-This means there are 4 novel combinations, beyond this being the first time a cross
-comparison has been performed.  This is also the first time the ordering methods
-have been rigorously tested, as much of the work in~\cite{goldman:2013:ordering}
+This means there remain 4 novel combinations to be tested.
+This is also the first time the ordering methods
+have been rigorously tested for quality, as much of the work in~\cite{goldman:2013:ordering}
 was analysis oriented.
@@
 In order to provide test landscapes for evolution, we chose problems common to
-CGP literature and previous work with the duplication prevention and ordering
+the CGP literature and previous work with the duplication prevention and ordering
 techniques.  We have chosen 4 binary circuit problems, as the binary representation
@@
 
-The first problem, 3-bit parity, represents probably the most
-common historic application for
+The first problem, 3-bit parity, represents the most
+common application for
 CGP~\cite{yu:2001:neutrality,miller:2006:redundancy,walker:2008:cgpmodules}.
@@
 \label{sec:parameter}
-To perform a fair comparison, we should ensure proper parameter configuration.
-This is to avoid the potential criticism that the parameter settings chosen benefit
-a specific method more than others.  While tuning can lead to the alternative issue
+To perform a fair comparison, we ensure proper parameter configuration
+to avoid potential bias that might benefit
+a specific method.  While tuning can lead to the alternative issue
 of methods only being effective after extensive problem specific tuning, we set out to use
@@
 size of 50 means there are 50 nodes, each composed of multiple genes, plus the
-required genes to specify output locations.  Also note that mutation rate of 0.05
+required genes to specify output locations.  Also note that a mutation rate of 0.05
 is used here to mean each gene has a 5\% probability of being mutated in a single
-application of mutation.  These values were chosen as they completely cover the
-range of previously used parameter settings, with each value having at least
-one paired value for the other parameter which is feasibly effective.
+application of mutation.  These ranges were chosen to completely cover the
+range of previously used parameter settings.
 The grid is formed by trying all possible mutation rates
@@
 and 2 are run 25 times.  For algorithm combinations using \emph{Single}, all 8 parameter configurations were run 17 times
-before any are removed.  This iterative reduction allows us to focus tuning on
+before any were removed.  This iterative reduction allows us to focus tuning on
 those parameter configurations most likely to be effective, with more information
@@
 the expected time to completion.  As the median was used to compare configurations,
-and not the mean, selection should be insensitive to this limit unless a large
-portion of runs are failing to optimize.
-
-To prevent undue biasing in the final results, none of these tuning runs were
-used for any other data analysis.  This helps reduce selection's effect on
-result distribution, similar to how it is common to split training and testing
-runs.  In this case, the runs used to set the parameter configuration are considered
-the training runs for that configuration.  As a result, our final results were gather
+selection should be insensitive to this limit unless a large
+portion of runs failed to optimize.
+
+To prevent undue biasing in the final results, these tuning runs were
+not used in the subsequent data analysis.  Effectively, the runs used to set
+the parameter configuration were considered
+the training runs for that configuration.  As a result, our final results were gathered
 from a completely independent set of runs to test the configuration's general characteristics.
@@
 	\caption{Results for algorithm configuration on each problem.  Highlighted row
-	is the control configuration \emph{Skip Normal}.  MES=Median Evaluations to Success; MAD=Median
+	is the control configuration \emph{Skip-Normal}.  MES=Median Evaluations to Success; MAD=Median
 	Absolute Deviation; p-value=Result from Mann-Whitney~U comparison with control.
@@
 success (MES), with the best algorithm combination appearing at the top of each section.
-The highlighted rows mark \emph{Skip Normal}, the control algorithm combination
-representing traditional CGP's performance.  All configurations for each problem
-were then testing using Kruskal-Wallis to determine if any configuration is
-statistically different from any other.  All problems received highly significant
-values.  As such, each algorithm combination was then compared with the control
+The highlighted rows mark \emph{Skip-Normal}, the control algorithm combination
+representing traditional CGP performance.  All configurations for each problem
+were then tested using Kruskal-Wallis to determine if any configuration was
+statistically different from any other.  All problems were determined to be highly significant.
+As such, each algorithm combination was then compared with the control
 to determine statistical significance using pairwise Mann-Whitney~U tests. The
@@
 Of the 12 configurations including \emph{Single}, 11 used a genome size smaller
-than all configurations not using \emph{Single}.  The exception was \emph{Single Normal}
+than all configurations not using \emph{Single}.  The exception was \emph{Single-Normal}
 on Multiply.  No other algorithm had as much impact on genome size tuning as
 the inclusion or exclusion of \emph{Single}.  In previous work~\cite{goldman:2013:ordering}
-it appeared \emph{Reorder} and \emph{DAG} were allowing CGP to use smaller genomes.
+it appeared that \emph{Reorder} and \emph{DAG} allowed CGP to use smaller genomes.
 In light of our results, this conclusion should be amended.
@@
 By ordering the algorithm combinations by MES and examining that ordering across each
-problem, we can look for general trends.  From the table, it appears on easier problems
-which method of duplication is in use is the strongest predictor of rank, while
-the ordering methods are more important on harder problems.  This comes from the
+problem, we can look for general trends.  From the table, it appears that on easier problems
+the method of duplication used is the strongest predictor of rank, while
+the ordering methods are more predictive on harder problems.  This comes from the
 observation that in for Parity and Encode (Upper half of Table~\ref{tab:results})
@@
 Each table includes coincidentally one exception to these rules.  On Parity and
-Encode pairings with \emph{DAG} are ranked lower than other ordering methods, so much
-so as to break the ordering once on both problems.  On Decode \emph{Reorder Duplicate}
-ranks worse than expected.  Finally on Multiply \emph{Single DAG} does far better
+Encode, pairings with \emph{DAG} are ranked lower than other ordering methods, so much
+so as to break the ordering once on both problems.  On Decode \emph{Accumulate-Reorder}
+ranks worse than expected.  Finally on Multiply \emph{Single-DAG} does far better
 than expected.
@@
 
-On all four problems, \emph{Reorder} either finds the solution in comparable
+On all four problems, \emph{Reorder} finds the solution in either comparable
 or significantly less evaluations than \emph{Normal} ordering, independent of
@@
 actual number of evaluations, which remains the primary cause of differential total runtime.
-As predicted \emph{DAG} does take longer than the other algorithms to perform each evaluation,
+As predicted, \emph{DAG} does take longer than the other algorithms to perform each evaluation,
 reinforcing the conclusion that \emph{DAG} should not be used for applications.
@@
 
-If we had selected on runtime as apposed to evaluations during tuning, we would expect whichever
+If we had tuned using runtime instead of evaluations, we would expect whichever
 technique is most effective at small genome sizes to obtain the best results.
@@
 to work better with this limitation than \emph{Normal}.  Assuming the tuned values
-we obtained are any gauge, \emph{Single Reorder} will likely require the least
+we obtained are any gauge, \emph{Single-Reorder} will likely require the least
 amount of runtime when tuned to reduce runtime.
@@
 will produce a single output.  These output values can be recorded and ordered.
-As the domain in use in binary, the outputs can be viewed as a bit string, with
-each bit the value that a node outputs when an individual is presented with a
+As the domain in use is binary, the outputs can be viewed as a bit string, where
+each bit indicates the value a node outputs when an individual is presented with a
 specific input.  Similar recording can be done for other domains, but binary
@@
 
-The semantics of a node is a complete description of its functionality.
-Any two nodes with identical semantic values have identical behavior,
+This semantics representation of a node is a complete description of its functionality.
+Any two nodes with an identical semantic representation have identical behavior,
 even if their method for encoding that behavior is drastically different.  From
@@
 The \Call{Simplify}{} algorithm given in Figure~\ref{fig:simplify} is designed
-to efficiently reduce the redundancy encoded in a genome.  Generally, this
+to efficiently reduce redundancy encoded in a genome.  Generally, this
 algorithm will move all connections to the first node that produces the desired
@@
 with \emph{DAG} and \emph{Single}, the evolved results were over 50\% inactive.
-\emph{Skip Normal}, our control representing normal CGP, had only 1\%, 3\%, 20\%,
+\emph{Skip-Normal}, our control representing normal CGP, had only 1\%, 3\%, 20\%,
 and 9\% of nodes in the genome active in the median runs of Parity, Encode, Decode,
@@
 \emph{Single} has a distinguishable behavior, but this is almost certainly the result
-of the genome size and not the duplication method.  \emph{Single Normal} used
+of the genome size and not the duplication method.  \emph{Single-Normal} used
 a genome size 2.5 times larger than \emph{Normal} did with the other techniques,
@@
 and Multiply problems, respectively.  These two problems were chosen as they
-represent different ends of the quality spectrum for \emph{Skip Normal} and
+represent different ends of the quality spectrum for \emph{Skip-Normal} and
 because together the are sufficiently representative of behaviors seen on the other
@@
 Potentially the most interesting results come from the control combination of
-\emph{Skip Normal}.  As problem difficulty increases, \emph{Skip Normal} produces
+\emph{Skip-Normal}.  As problem difficulty increases, \emph{Skip-Normal} produces
 an increasingly large number of offspring with no change in active node behavior,
@@
 actively identical to their parents.  The latter is supported by the fact that
-\emph{Skip Normal} so infrequently reactivates a node with tested behavior.  More
+\emph{Skip-Normal} so infrequently reactivates a node with tested behavior.  More
 often than not offspring connecting in inactive nodes are testing that node's behavior

### 2013-07-01 CGP Journal/main.tex
First full pass of Bill's edits

@@
 Understanding how search operators interact with solution representation is
-a critical step to search improvement.  For many search algorithms this may
-be trivial, but in Cartesian Genetic Programming (CGP), and Genetic Programming (GP)
+a critical step to improving search.
+In Cartesian Genetic Programming (CGP), and Genetic Programming (GP)
 in general, complex genotype to phenotype map makes achieving this understanding a
-challenge.  By examining tuned parameter values, variant success on different
+challenge.  By examining tuned parameter values, the search quality of CGP variants at different
 problem difficulties, node behavior, and offspring replacement properties we
-seek to comprehend observed characteristics of CGP search.  We focus our study
-around methods to prevent wasted CGP evaluations (\emph{Skip}, \emph{Accumulate},
+seek to comprehend observed characteristics of CGP search.  Our focus is twofold:
+methods to prevent wasted CGP evaluations (\emph{Skip}, \emph{Accumulate},
 and \emph{Single}) and methods to overcome CGP's search limitations imposed by
@@
 
-We find strong evidence that currently most CGP techniques evolve genomes that
+Our results show that most CGP techniques evolve genomes that
 are highly inactive, very redundant, and full of seemingly useless constants.
-In some instances we found less than 1\% of the genome was actually required to
+On some tested problems we found less than 1\% of the genome was actually required to
 encode the evolved solution.
@@
 is never used by any ancestor of the evolved solution.  More generally, our
-results lead us to conclude that \emph{Skip-Reorder} and \emph{Single-Reorder}
+results suggest that \emph{Skip-Reorder} and \emph{Single-Reorder}
 are most likely to solve hard problems in the least number of evaluations and
-the least amount of time, respectively, with the least amount of degenerate
+the least amount of time with the least amount of degenerate
 behavior.
@@
 was significantly lower than the genome size.  With the exception of a few combinations
-with \emph{DAG} and \emph{Single}, the evolved results were over 50\% inactive.
+with \emph{DAG} and \emph{Single}, the evolved genomes were over 50\% inactive.
 \emph{Skip-Normal}, our control representing normal CGP, had only 1\%, 3\%, 20\%,
@@
 
-With how inactive the final genomes are, we ask ``what is the rest for?''
+With such a small percentage of the final genome active, what is the rest of the
+genome for?
 Just because a node is not active in the final solution does not mean it was never
@@
 answer this we must consider what can prevent a node from becoming active.  For
-a node to never be active all tested offspring with mutations that result in that node being activated
+a node to never be active, all tested offspring with mutations that result in that node being activated
 must have been less fit than their parent.  Even if those offspring are no more fit, the selection method
@@
 Examining the progression in fitness across the different ordering methods shows
-the effect of initialization on each.  All three methods share the same initialization
+the effect of initialization on each.  All three order methods share the same initialization
 method, and as such all have very similar never active node behavior early in
@@
 evaluations both have explored similar sections of their genome.  Yet on hard
-problems that require further exploration \emph{Reorder} is able to do so, while
+problems that require further exploration, \emph{Reorder} is able to do so, while
 \emph{Normal} remains positionally limited.
@@
 Used nodes are by definition only in the active portion of the genome.  Similarly
-explore nodes can only be in the inactive portion of the genome.   Note that
+Explore nodes can only be in the inactive portion of the genome.   Note that
 all never active nodes are ignored by these plots, as their behavior is meaningless.
@@
 From Table~\ref{tab:encode_behavior} and Table~\ref{tab:multiply_behavior}, having
-the plurality of active nodes in the used category appears to correlate with
+the plurality of active nodes in the ``Used'' category appears to correlate with
 increase evaluations to success.  Notice that in all \emph{Single} plots the
@@
 One significant difference in behavior between the two problems is proportion of
-explore nodes.  A potential explanation is that on 16-4-bit Encode there are only
+Explore nodes.  A potential explanation is that on 16-4-bit Encode there are only
 $2^{16} = 65,536$ unique semantics while on 3-bit Multiply there are $2^{64} = 1.845\cdot 10^{19}$
@@
 An important aspect of CGP is its replacement strategy that allows for neutral drift.
-Each generation the best offspring replaces its parent if it is no less fit.  Previous
+In each generation the best offspring replaces its parent if it is no less fit.  Previous
 research has shown the negative impact removing this feature can have on
-performance~\cite{yu:2001:neutrality}, but how it interacts with the different variants
+performance~\cite{yu:2001:neutrality}. However, how it interacts with the different variants
 of CGP has not been investigated previously.
@@
 that were active in some ancestor of the offspring, were inactive in the offspring's
-parent, but are active in the offspring.  Reactivation is a measure of how
+parent, but are now again active in the offspring.  Reactivation is a measure of how
 often CGP reuses previously useful active nodes.  Reported in the table is how
@@
 
-The \textbf{Reactivated Changed} column states the percentage of reactivated nodes
+The \textbf{Reactivated Changed} column shows the percentage of reactivated nodes
 that have a different semantic from the last time they were active.  Conversely,
@@
 
-Combined these three measures provide somewhat surprising evidence.  Most notably
+Combined these three measures provide some surprising insights.  Most notably
 is that \emph{Single}, a technique designed to force offspring to be actively different

### 2013-07-01 CGP Journal/main.tex
TODO statements from today's meeting

@@
 \label{sec:ordering}
+%TODO Add formula for probability node is active based on index
+%p(i) \leftarrow 1 - (\frac{N+I-1}{N+I})^O \cdot \prod_{j=i+1}^{N} 1 - p(j) \cdot (1-(\frac{j+I-1}{j+I})^a)
 As was discussed in Section~\ref{sec:cgp}, CGP uses node ordering in the genome
@@
 \label{sec:quality}
+%TODO Hook here the table of descriptions.
 Empirical results are an important step in confirming our examinations.
-
 We accomplish this by testing the effectiveness
@@
 their number of output locations, their evolved percentage of active nodes is also ordered.
+%TODO Reference here again the probability function, and its ability to predict values
 

### 2013-07-03 CGP Journal/main.tex
Most of the updates suggested in our meeting, still need to move semantic section and reference Equation 1 in discussion section

--- 2013-07-01 CGP Journal/main.tex
@@
 to represent DAGs, as all DAGs can be serialized to fit this requirement.
-Yet this likely has an impact on CGP's ability to evolve specific
+However, this representation likely has an impact on CGP's ability to evolve specific
 DAGs~\cite{goldman:2013:ordering}.
@@
 connection.
+
+As an extension to the theory presented in~\cite{goldman:2013:ordering}, we have
+derived the formula which predicts the probability a node is active based on its
+index, given in Equation~\ref{eq:prob} and shown in Figure~\ref{fig:prob}.
+\begin{equation}
+p(i) \leftarrow 1 - (\frac{N+I-1}{N+I})^O \cdot \prod_{j=i+1}^{N} 1 - p(j) \cdot (1-(\frac{j+I-1}{j+I})^a)
+\label{eq:prob}
+\end{equation}
+In this equation, $i$ is the index of the working node, $N$ is the genome size, $I$ is the number of input locations,
+$O$ is the number of output locations, and $a$ is the arity of each node.
+This equation is constructed as a negation, such that we determine the probability
+the node is not active due to each potential source that could make it active.
+The product term calculates the probability that the node at index $j$ is connected to the
+node at index $i$, times the probability that the node at index $j$ is active.  This is
+negated to determine the probability node $i$ is not active because of node $j$.
+All together, the product term calculates the probability that no active node is
+connected to the node at index $i$.  Nodes can also become active if they are directly
+connected to by an output location.  As such the product is scaled by the probability
+that all output locations connect to inputs or nodes not at index $i$.  This
+formula assumes only operator bias, not selective pressure, is acting on the
+genome.
+
+\begin{figure}
+\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{active_prob}
+\caption{Plot of Equation~\ref{eq:prob} describing the probability that a node
+         at a given index will be active.  $N=2000, a=2, I=6, O=6$.}
+\label{fig:prob}
+\end{figure}
+
+From Figure~\ref{fig:prob} it is clear that nodes near the beginning of the genome
+are almost certainly going to be active, while nodes in the last three quarters
+of the genome are almost never going to be active.
 
@@
 \label{sec:quality}
+
+\begin{table}
+	\centering
+	\begin{tabular}{|c|p{6cm}|}
+	\hline
+  \multicolumn{2}{|c|}{\textbf{Duplication}} \\ \hline
+	\textbf{\emph{Skip}} & Whenever an actively identical offspring is detected, assign it its parent's fitness. \\ & \\
+	\textbf{\emph{Accumulate}} & Repeatedly perform full genome mutation until \textbf{one or more} active genes change. \\ & \\
+	\textbf{\emph{Single}} & Select and mutate individual genes until \textbf{exactly one} active gene changes. \\
+	\hline \multicolumn{2}{|c|}{\textbf{Ordering}} \\ \hline
+	\textbf{\emph{Normal}} & Nodes must take input from other nodes or inputs that precede them in the genome. \\ & \\
+	\textbf{\emph{Reorder}} & Each generation the parent's node ordering is shuffled without effecting node behavior. \\ & \\
+	\textbf{\emph{DAG}} & Connection genes can mutate to make any connection as long as no cycle is formed. \\
+	  \hline
+	\end{tabular}
+	\caption{Short descriptions of each of the algorithm variants.}
+	\label{tab:glossary}
+\end{table}
+
 Empirical results are an important step in confirming our examinations.
-
 We accomplish this by testing the effectiveness
@@
 evolutionary side effects are given in Section~\ref{sec:analysis}.
+Table~\ref{tab:glossary} gives a summary of what each variant means.
 
@@
 from a completely independent set of runs to test the configuration's general characteristics.
-Each configuration was run 50 times to ensure statistical power.
+Each configuration was run 50 times to ensure statistical power.  All told over
+11.5 billion evaluations were performed to gather our results.
 
@@
 
+As a final check that our results do not somehow unfairly handicap traditional CGP,
+we compared our median evaluations to success with previous publications.  The results
+reported for traditional CGP and CGP using modules in~\cite{walker:2008:cgpmodules}
+for 3-bit parity required 6 times as many evaluations as \emph{Skip-Normal}.
+The same source had traditional CGP requiring 4 million evaluations to success,
+as opposed to our \emph{Skip-Normal} requiring 600 thousand, also a factor of 6 improvement.
+CGP with modules required over 800 thousand evaluations, worse than our \emph{Skip-Normal},
+and approximately 4 times more than \emph{Skip-Reorder}.  We must be wary about comparisons
+to their 3-bit multiplier though, as their version used a different function set.
+That said the function set they used appears to have been chosen explicitly to improve
+CGP's performance on that problem, so we suggest that if we had used a the same function
+set our results would still be significantly better.
+
 Our parameter tuning focused on minimizing the number of evaluations each algorithm
@@
 
+\begin{comment}
 \subsection{Scalability}
@@
 TODO Include scaling comparison with~\cite{walker:2008:cgpmodules}.
+\end{comment}
 
@@
 \item ``Useful'' - Nodes with identical semantics to a ``Used'' node.
-\item ``Intron'' - Any node with semantics identical to an active node.
+\item ``Intron'' - Any node with semantics identical to a node that was
+                   active before applying \Call{Simplify}{}, but was inactive afterward.
 \item ``Explore'' - All remaining nodes.

### 2013-07-03 CGP Journal/main.tex
Removed outdated commented out sections

@@
 
-\begin{comment}
-Genome size explosion seems most prevalent on easier problems, with Parity and
-Encode having much higher genome sizes than Decode and Multiply.  Initially
-this seems counter intuitive, as we would expect harder problems to require
-larger solutions which in turn should require larger genomes.  Yet the genome
-sizes in use are so far beyond what a solution to these problems require that
-solution size is likely not the cause of this behavior.  Instead, consider
-the usefulness of unexplored nodes.
-
-On simple enough problems, exploring lots of randomly generated nodes may be
-sufficient to generate high quality solutions.  At relatively low numbers of
-evaluations, large genotypes likely mean each mutation to a connection gene
-will activate nodes that have never been active, resulting in exploration.
-At higher numbers of evaluations, mutations to connection genes that are not currently
-active have a high chance of connecting to a node that was tested previously
-and found ineffective.  As a result, having a smaller genome size on harder problems
-may reduce the chance of exploratory mutations when optimizing near the global optimum.
-
-Under this theory, we should expect Parity to have the largest genome sizes
-\end{comment}
-
 Unlike genome size, all configurations choose approximately the same mutation rate,
@@
 
-\begin{comment}
-\begin{figure}
-  \begin{algorithmic}
-  \Procedure{RepeatedFootprint}{}
-    \State $reachable \leftarrow \emptyset$
-    \ForAll{$i \in input\_locations$}
-      \State $reachable[i] \leftarrow footprint(i)$
-    \EndFor
-    \State $repeated \leftarrow \emptyset$
-    \ForAll{$node \in active$}
-      \State $working \leftarrow footprint(node)$
-      \State $direct \leftarrow \emptyset$
-      \ForAll{$link \in reads\_from(node)$}
-        \State $direct \leftarrow direct \cup reachable[link]$
-      \EndFor
-      \If{$working \in direct$}
-        \State $repeated \leftarrow repeated \cup \{working\}$
-      \Else
-        \State $direct \leftarrow direct \cup \{working\}$
-      \EndIf
-      \State $reachable[node] \leftarrow direct$
-    \EndFor
-    \State\Return{$repeated$}
-  \EndProcedure
-  \end{algorithmic}
-  \caption{Algorithm to determine which nodes transitively depend on their
-  own footprint.}
-  \label{fig:repeated}
-\end{figure}
-
-An efficient algorithm to remove some of the redundancy in the genome is given
-in Figure~\ref{fig:repeated}.  This algorithm finds the set of nodes which are
-transitively dependent on their own footprint.  In other words, in order for
-each node in the $repeated$ set, some other node in the genome has to produce
-the same output.  As an example, consider node $Y$ which is in the $repeated$
-set.  There must exist some node $X$ which $Y$ transitively depends on that has
-the same footprint as $Y$.  As a result, $Y$ can always be remove, with anything
-previously dependent on $Y$ now dependent on $X$.  All nodes in the $repeated$
-set can therefore be removed from the genome without effecting its ability to
-calculate any value.  This also has the potential to cause cascading removals
-if there are intermediary nodes which are no longer useful once all of the
-$repeated$ nodes have been removed.  In total this algorithm runs in $O(AN^2)$
-time for an arbitrary individual.  Note that this is an offline algorithm,
-so it only needs to be run on individuals used in production or for analysis.
-
-\begin{figure}
-  \begin{algorithmic}
-  \Procedure{Simplify}{$p\_r$, $p\_c$, $p\_i$, $best$}
-    \If{$|p\_r| = 0$}
-      \State\Return{$p\_i$}
-    \ElsIf{$|p\_i| + |p\_r| \geq |best|$}
-      \State\Return{$best$}
-    \EndIf
-    \State $working, p\_a \leftarrow randpop(p\_r)$
-    \State $covrd \leftarrow p\_c \cup {working}$
-    \State $ansc \leftarrow p\_a \cup {working}$
-    \ForAll{$node \in has\_footprint(working)$}
-      \State $node\_valid \leftarrow True$
-      \State $incd \leftarrow p\_i \cup {node}$
-      \State $reqd \leftarrow p\_r$
-      \ForAll{$link \in reads\_from(node)$}
-        \State $foot \leftarrow footprint(link)$
-        \If{$foot \in ansc$}
-          \State $node\_valid \leftarrow False$
-          \State \textbf{break}
-        \ElsIf{$foot \notin covrd$}
-          \State $reqd[foot] \leftarrow reqd[foot] \cup ansc$
-        \EndIf
-      \EndFor
-      \If{$node\_valid$}
-        \State $solution \leftarrow \Call{Simplify}{reqd, covrd, incd, best}$
-        \If{$|solution| < |best|$}
-          \State $best \leftarrow solution$
-        \EndIf
-      \EndIf
-    \EndFor
-    \State\Return{$best$}
-  \EndProcedure
-  \end{algorithmic}
-  \caption{Recursive algorithm which determines the minimum set of nodes required
-    to generate a set of required outputs.}
-  \label{fig:simplify}
-\end{figure}
-
-While the previous algorithm is guaranteed to only remove redundant nodes, it
-does not remove all redundant nodes.  In order to remove redundant nodes without
-any transitive relationships, we must use the algorithm given in
-Figure~\ref{fig:simplify}.  This is a recursive algorithm that tries all possible
-choices of nodes that produce each footprint to see which combination results
-in the fewest active nodes.  The algorithm takes in a mapping of footprints
-required by previous calls to all ancestor footprints of those requirements ($p\_r$), footprints previously covered ($p\_c$),
-nodes that have been previously included in the solution ($p\_i$), and smallest
-set of nodes found so far which produces all required footprints ($best$).  Initially
-$p\_c$ contains the set of input footprints, with $p\_r$ mapping each of the output
-footprints to empty sets unless that footprint is already in $p\_c$.  $p\_i$
-is initialized to the empty set, and $best$ starts out as the set of all $active$ nodes.
-Recursion ends when the the nodes in $p\_i$ have no unsatisfied requirements or
-all possible solutions containing $p\_i$ that can satisfy all requirements are
-no better than the current best solution.  While this algorithm is exponential
-in nature, in practice it has a reasonable runtime.  Using the algorithm in
-Figure~\ref{fig:repeated} to prevent $repeated$ nodes from being returned by
-$has\_footprint$ significantly reduces the number of combinations that need
-to be tested.  Furthermore, by preventing recursion from exploring solutions
-worse than the current known best, significant time can be saved.  Lastly, as
-with the algorithm in Figure~\ref{fig:repeated}, this is an offline algorithm,
-and has no impact on evolutionary speed.
-
-Using the algorithm given in Figure~\ref{fig:simplify}, we can determine the
-absolute minimum set of nodes in a genome's solution that are necessary to
-reproduce the original output of that genome.  This allows us to discuss
-the usefulness of each node as either being part of the minimum solution, a
-duplicate of a node in the minimum solution, or irrelevant to the solution.
-As such we can then examine how redundancy is being used, how frequently useful
-structures exist in the inactive nodes, and how useful nodes are constructed by evolution.
-This algorithm can also be used to construct simplified genomes when combined
-with the algorithm given in Figure~\ref{fig:reorder}.
-\end{comment}
 \subsection{Never Active Nodes}
@@
     & \textbf{\em Normal} & \textbf{\em Reorder} & \textbf{\em DAG} \\ \hline
-    %\begin{comment}
     \thirdlabel{Skip} & \graphicthird{multiply_skip_normal} &
@@
                           \graphicthird{multiply_single_dag}\\ \hline
-    %\end{comment}
 	\end{tabular}

### 2013-07-03 CGP Journal/main.tex
Moved semantic stuff, included Equation 1 in discussion

@@
 
-%TODO Consider adding 5-bit Parity or Equality, 11-bit Multiplexer.
-
 \subsection{Parameter Setting}
@@
 
-As a primary concept to our analysis, we determine the semantics of nodes
-in the genome~\cite{mcphee:2008:treesemantics}.
+\subsection{Never Active Nodes}
+In line with the suggestions made in~\cite{miller:2006:redundancy}, the evolved
+number of active nodes across all problems and all algorithm configurations
+in Table~\ref{tab:results}
+was significantly lower than the genome size.  With the exception of a few combinations
+with \emph{DAG} and \emph{Single}, the evolved genomes were over 50\% inactive.
+\emph{Skip-Normal}, our control representing normal CGP, had only 1\%, 3\%, 20\%,
+and 9\% of nodes in the genome active in the median runs of Parity, Encode, Decode,
+and Multiply, respectively.  These final results fall in line with the predictions
+made in~\cite{goldman:2013:ordering}, in that if the problems are ordered based on
+their number of output locations, their evolved percentage of active nodes is also ordered.
+Furthermore, if Equation~\ref{eq:prob} is used to calculate the expected number of
+active nodes for configurations using \emph{Normal} ordering, we find that in all
+cases the evolved number is only marginally higher than the expectation.
+
+With such a small percentage of the final genome active, what is the rest of the
+genome for?
+Just because a node is not active in the final solution does not mean it was never
+useful.  For a node to be truly of no use to search, we must consider nodes that
+were never active.  Never active nodes are nodes such that for all ancestors of the
+final solution (the line of descent), no parent had that node active.  Note that
+for \emph{Reorder} this demarcation of node activity follows the node through
+each shuffle, such that if a node was marked as never active before the genome
+was shuffled it is marked as never active after the shuffle, and vice versa.
+This is in contrast to marking locations as never active, which was not done in
+this study.
+
+\begin{table*}
+	\centering
+  \begin{tabular}{c|c|c|c|}
+    \cline{2-4}
+    & \textbf{\em Normal} & \textbf{\em Reorder} & \textbf{\em DAG} \\ \hline
+    \thirdlabel{Skip} & \graphicthird{multiply_skip_normal} &
+                        \graphicthird{multiply_skip_reorder} &
+                        \graphicthird{multiply_skip_dag}\\ \hline
+    \thirdlabel{Accumulate} & \graphicthird{multiply_accumulate_normal} &
+                              \graphicthird{multiply_accumulate_reorder} &
+                              \graphicthird{multiply_accumulate_dag}\\ \hline
+    \thirdlabel{Single} & \graphicthird{multiply_single_normal} &
+                          \graphicthird{multiply_single_reorder} &
+                          \graphicthird{multiply_single_dag}\\ \hline
+	\end{tabular}
+	\caption{Location of never active nodes across all 50 runs for all nine algorithm combinations on the Multiply problem.}
+	\label{tab:never_active}
+\end{table*}
+
+Table~\ref{tab:never_active} provides information about never active nodes for
+each algorithm combination on the 3-bit Multiply problem.  The horizontal axis
+in each of the nine plots shows different genome locations, while the vertical
+axis denotes the level of fitness the recorded individual obtained.  The shading
+of each coordinate gives the percentage of 50 runs that had an individual of
+the given fitness level such that the node at the given location had never
+previously been active.  While only Multiply is shown here, all of the
+other problems tested had similar behavior, with allowances for differences in
+number of evaluations performed.
+
+Examining the columns of Table~\ref{tab:never_active} makes it clear that
+each of the ordering methods had significantly different never active behavior.
+Looking at \emph{Normal} we see that only at very high fitness levels does evolution
+begin to activate nodes in the latter half of the genome.  Even by the final solution
+large percentages of the genome were never activated, with \emph{Skip}, \emph{Accumulate},
+and \emph{Single} having median percentages of never active nodes of 43.35\%, 42.05\%, and
+56.17\%, respectively.  This is in stark contrast to \emph{Reorder} and \emph{DAG},
+which never had a median higher than 0.05\%.
+
+Compared to the difference in columns, the difference in the rows of Table~\ref{tab:never_active}
+caused by the different duplication methods were almost non-existent.  Only
+\emph{Single} has a distinguishable behavior, but this is almost certainly the result
+of the genome size and not the duplication method.  \emph{Single-Normal} used
+a genome size 2.5 times larger than \emph{Normal} did with the other techniques,
+so the resulting increase in never active nodes makes sense.  When \emph{Single}
+was paired with \emph{Reorder} and \emph{DAG} the genome sizes were far smaller,
+resulting in noisier graphs with less never active nodes.
+
+So now the question is why does \emph{Normal} use so little of its genome?  To
+answer this we must consider what can prevent a node from becoming active.  For
+a node to never be active, all tested offspring with mutations that result in that node being activated
+must have been less fit than their parent.  Even if those offspring are no more fit, the selection method
+used by CGP would have allowed that offspring to replace its parent, at least probabilistically.  This leads
+to two possible causes for a node never being active: bad genes or infrequently tried.
+
+The bad genes cause for a node never being active suggests the behavior encoded
+by a node is simply useless on the problem landscape.  If this was the root cause
+of never active nodes, we would expect \emph{Reorder} and \emph{DAG} to have
+similar problems, as neither technique explicitly changes the likelihood of node
+behaviors.  Furthermore, we would expect the location of never active nodes to be
+uniform in the genome, which is obviously not the case.  Finally, even if bad
+genes were the problem, we would expect the mutational drift allowed by CGP's
+selection method to change their bad genes and eventually allow them to be active.
+
+The infrequently tried cause for a node never being active suggests that few
+individuals are ever produced which have that node active, limiting the likelihood
+of those individuals being selected regardless of potential fitness.  Under this
+cause and combined with Equation~\ref{eq:prob}, we
+expect a strong positional bias in the location of the never active nodes;
+we would expect nodes near the end of the genome to be the most likely to be never active.
+Figure~\ref{fig:prob} shows the expected probability given by Equation~\ref{eq:prob},
+using the same genome size, arity, number of inputs, and number of outputs as our
+3-bit multiply.  This curve bears a striking resemblance to all three \emph{Normal}
+plots in Table~\ref{tab:never_active}.
+Finally, from~\cite{goldman:2013:ordering} we would expect \emph{Reorder} and \emph{DAG}
+to have less or no positional bias, which also matches the results given in Table~\ref{tab:never_active}.
+
+Examining the progression in fitness across the different ordering methods shows
+the effect of initialization on each.  All three order methods share the same initialization
+method, and as such all have very similar never active node behavior early in
+evolution.  This initialization similarity explains why \emph{DAG} has any positional
+bias at all, even though its evolutionary mechanisms have none.  Similarly \emph{Reorder}
+diverges from \emph{Normal} given sufficient time, eventually losing its positional
+bias.
+
+This early similarity may be part of the reason \emph{Normal} and \emph{Reorder}
+have such similar behavior on relatively easy problems.  After relatively few
+evaluations both have explored similar sections of their genome.  Yet on hard
+problems that require further exploration, \emph{Reorder} is able to do so, while
+\emph{Normal} remains positionally limited.
+
+\subsection{Node Behaviors}
+
+In order to understand how each node is being used by CGP,
+we first determine the semantics of each node~\cite{mcphee:2008:treesemantics}.
 For each input that is given to an individual, all of its nodes
@@
 
-\subsection{Never Active Nodes}
-%TODO Somewhere in here you should talk about how Normal has the least number of active,
-%which is probably related to how many never become active
-In line with the suggestions made in~\cite{miller:2006:redundancy}, the evolved
-number of active nodes across all problems and all algorithm configurations
-in Table~\ref{tab:results}
-was significantly lower than the genome size.  With the exception of a few combinations
-with \emph{DAG} and \emph{Single}, the evolved genomes were over 50\% inactive.
-\emph{Skip-Normal}, our control representing normal CGP, had only 1\%, 3\%, 20\%,
-and 9\% of nodes in the genome active in the median runs of Parity, Encode, Decode,
-and Multiply, respectively.  These final results fall in line with the predictions
-made in~\cite{goldman:2013:ordering}, in that if the problems are ordered based on
-their number of output locations, their evolved percentage of active nodes is also ordered.
-%TODO Reference here again the probability function, and its ability to predict values
-
-With such a small percentage of the final genome active, what is the rest of the
-genome for?
-Just because a node is not active in the final solution does not mean it was never
-useful.  For a node to be truly of no use to search, we must consider nodes that
-were never active.  Never active nodes are nodes such that for all ancestors of the
-final solution (the line of descent), no parent had that node active.  Note that
-for \emph{Reorder} this demarcation of node activity follows the node through
-each shuffle, such that if a node was marked as never active before the genome
-was shuffled it is marked as never active after the shuffle, and vice versa.
-This is in contrast to marking locations as never active, which was not done in
-this study.
-
-\begin{table*}
-	\centering
-  \begin{tabular}{c|c|c|c|}
-    \cline{2-4}
-    & \textbf{\em Normal} & \textbf{\em Reorder} & \textbf{\em DAG} \\ \hline
-    \thirdlabel{Skip} & \graphicthird{multiply_skip_normal} &
-                        \graphicthird{multiply_skip_reorder} &
-                        \graphicthird{multiply_skip_dag}\\ \hline
-    \thirdlabel{Accumulate} & \graphicthird{multiply_accumulate_normal} &
-                              \graphicthird{multiply_accumulate_reorder} &
-                              \graphicthird{multiply_accumulate_dag}\\ \hline
-    \thirdlabel{Single} & \graphicthird{multiply_single_normal} &
-                          \graphicthird{multiply_single_reorder} &
-                          \graphicthird{multiply_single_dag}\\ \hline
-	\end{tabular}
-	\caption{Location of never active nodes across all 50 runs for all nine algorithm combinations on the Multiply problem.}
-	\label{tab:never_active}
-\end{table*}
-
-Table~\ref{tab:never_active} provides information about never active nodes for
-each algorithm combination on the 3-bit Multiply problem.  The horizontal axis
-in each of the nine plots shows different genome locations, while the vertical
-axis denotes the level of fitness the recorded individual obtained.  The shading
-of each coordinate gives the percentage of 50 runs that had an individual of
-the given fitness level such that the node at the given location had never
-previously been active.  While only Multiply is shown here, all of the
-other problems tested had similar behavior, with allowances for differences in
-number of evaluations performed.
-
-Examining the columns of Table~\ref{tab:never_active} makes it clear that
-each of the ordering methods had significantly different never active behavior.
-Looking at \emph{Normal} we see that only at very high fitness levels does evolution
-begin to activate nodes in the latter half of the genome.  Even by the final solution
-large percentages of the genome were never activated, with \emph{Skip}, \emph{Accumulate},
-and \emph{Single} having median percentages of never active nodes of 43.35\%, 42.05\%, and
-56.17\%, respectively.  This is in stark contrast to \emph{Reorder} and \emph{DAG},
-which never had a median higher than 0.05\%.
-
-Compared to the difference in columns, the difference in the rows of Table~\ref{tab:never_active}
-caused by the different duplication methods were almost non-existent.  Only
-\emph{Single} has a distinguishable behavior, but this is almost certainly the result
-of the genome size and not the duplication method.  \emph{Single-Normal} used
-a genome size 2.5 times larger than \emph{Normal} did with the other techniques,
-so the resulting increase in never active nodes makes sense.  When \emph{Single}
-was paired with \emph{Reorder} and \emph{DAG} the genome sizes were far smaller,
-resulting in noisier graphs with less never active nodes.
-
-So now the question is why does \emph{Normal} use so little of its genome?  To
-answer this we must consider what can prevent a node from becoming active.  For
-a node to never be active, all tested offspring with mutations that result in that node being activated
-must have been less fit than their parent.  Even if those offspring are no more fit, the selection method
-used by CGP would have allowed that offspring to replace its parent, at least probabilistically.  This leads
-to two possible causes for a node never being active: bad genes or infrequently tried.
-
-The bad genes cause for a node never being active suggests the behavior encoded
-by a node is simply useless on the problem landscape.  If this was the root cause
-of never active nodes, we would expect \emph{Reorder} and \emph{DAG} to have
-similar problems, as neither technique explicitly changes the likelihood of node
-behaviors.  Furthermore, we would expect the location of never active nodes to be
-uniform in the genome, which is obviously not the case.  Finally, even if bad
-genes were the problem, we would expect the mutational drift allowed by CGP's
-selection method to change their bad genes and eventually allow them to be active.
-
-The infrequently tried cause for a node never being active suggests that few
-individuals are ever produced which have that node active, limiting the likelihood
-of those individuals being selected regardless of potential fitness.  Under this
-cause and combined with the theory given in~\cite{goldman:2013:ordering}, we
-expect a strong positional bias in the location of the never active nodes;
-we would expect nodes near the end of the genome to be the most likely to be never active.
-Finally we would expect \emph{Reorder} and \emph{DAG} to have less or no positional bias.
-All of these expectations match the results given in Table~\ref{tab:never_active}.
-
-Examining the progression in fitness across the different ordering methods shows
-the effect of initialization on each.  All three order methods share the same initialization
-method, and as such all have very similar never active node behavior early in
-evolution.  This initialization similarity explains why \emph{DAG} has any positional
-bias at all, even though its evolutionary mechanisms have none.  Similarly \emph{Reorder}
-diverges from \emph{Normal} given sufficient time, eventually losing its positional
-bias.
-
-This early similarity may be part of the reason \emph{Normal} and \emph{Reorder}
-have such similar behavior on relatively easy problems.  After relatively few
-evaluations both have explored similar sections of their genome.  Yet on hard
-problems that require further exploration, \emph{Reorder} is able to do so, while
-\emph{Normal} remains positionally limited.
-
-\subsection{Node Behaviors}
-
 \begin{table*}
@@
 solution.  Furthermore, a significant portion of these active nodes were found
-to be redundant, removable using a very simple reduction algorithm.  The only
+to be redundant; removable using a very simple reduction algorithm.  The only
 algorithm combinations that preferred more reasonable genome sizes were those using \emph{Single},

### 2013-07-16 CGP Journal/main.tex
Added missing file.  Finished edit pass after returning from GECCO

@@
 \usepackage{multirow}
+\usepackage{url}
 
@@
 In Cartesian Genetic Programming (CGP), and Genetic Programming (GP)
-in general, complex genotype to phenotype map makes achieving this understanding a
+in general, the complex genotype to phenotype map makes achieving this understanding a
 challenge.  By examining tuned parameter values, the search quality of CGP variants at different
@@
 neutrality~\cite{vassilev:2000:neutrality}, and structural bias~\cite{payne:2009:bias}.
-Our primary focus is
-on how mutation and genome ordering interact.
+We shall focus on the interaction between mutation and genome ordering.
 
@@
 is $\mu + \lambda$ where $\mu \leftarrow 1$ and $\lambda \leftarrow 4$.
-This means generation a single parent produces four offspring using
+This means each generation a single parent produces four offspring using
 mutation.  The best offspring then competes with the parent,\footnote{Ties are broken
@@
 \label{sec:duplicate}
-CGP's mutation operator is traditionally applied to genes in active and inactive
-node's uniformly. This has the
+CGP's mutation operator is traditionally applied to both active and inactive genes uniformly.
+This has the
 potential to create offspring who's only mutated genes are in inactive nodes.
@@
 in~\cite{goldman:2013:cgpwaste}, this can make normal CGP very sensitive to the
-mutation rate in use.
+mutation rate.
 
@@
 the mutation used each time is identical to normal CGP's mutation, such that
-$F_n$ can have one or more active genes different from $P$.  Furthermore, even
+$F_n$ will have one or more active genes different from $P$.  Furthermore, even
 though this technique produces $n+1$ individuals, only $F_n$ is evaluated.
@@
 quite different, further consideration shows how they are similar.  In \emph{Skip}
-an actively identical offspring is likely to be selected as we expect mutations to
+any actively identical offspring that is produced is likely to be selected as we expect mutations to
 active genes to more often reduce fitness than improve it.  In the next generation
@@
 \begin{equation}
-p(i) \leftarrow 1 - (\frac{N+I-1}{N+I})^O \cdot \prod_{j=i+1}^{N} 1 - p(j) \cdot (1-(\frac{j+I-1}{j+I})^a)
+p(i) \leftarrow 1 - (\frac{N+I-1}{N+I})^O \cdot \prod_{j=i+1}^{N} 1 - p(j) \cdot (1-(\frac{j+I-1}{j+I})^A)
 \label{eq:prob}
@@
 In this equation, $i$ is the index of the working node, $N$ is the genome size, $I$ is the number of input locations,
-$O$ is the number of output locations, and $a$ is the arity of each node.
+$O$ is the number of output locations, and $A$ is the arity of each node.
 This equation is constructed as a negation, such that we determine the probability
@@
 The product term calculates the probability that the node at index $j$ is connected to the
-node at index $i$, times the probability that the node at index $j$ is active.  This is
+node at index $i$, multiplied by the probability that the node at index $j$ is active.  This is
 negated to determine the probability node $i$ is not active because of node $j$.
@@
 \caption{Plot of Equation~\ref{eq:prob} describing the probability that a node
-         at a given index will be active.  $N=2000, a=2, I=6, O=6$.}
+         at a given index will be active.  $N=2000, A=2, I=6, O=6$.
+         Predicts approximately 160 total active nodes, with 80\% of the first
+         100 nodes active and 4\% of the remaining 1900 nodes active.}
 \label{fig:prob}
@@
 as they neither take input (directly or indirectly) from the node nor provide their
-output to the node.  In general, this process works by assigning nodes new
+output (directly or indirectly) to the node.  In general, this process works by assigning nodes new
 locations in the genome at random once all of the nodes they take input from have been assigned
@@
 to determine in constant time all connections incident on a node, and 
-can be build in order $O(AN)$ time, where $A$ is the arity of the nodes
+can be built in order $O(AN)$ time, where $A$ is the arity of the nodes
 and $N$ is the number of nodes.
@@
 The price \emph{Reorder} pays for this potential improvement is increased mutational search
-space and runtime costs.  Though a shuffled individual is phenotypically unchanged,
+space.  Though a shuffled individual is phenotypically unchanged,
 its mutational adjacency is significantly different.  As such the variety of offspring
@@
 To achieve this, we incrementally determine
-which nodes transitively depend on the mutating node and which are not, with
+which nodes transitively depend on the mutating node and which do not, with
 information stored in the $known$ map.  This data structure is initialized with
@@
 With these modifications, there is no longer any meaning to the location of a node
-in the genome.  In \emph{Normal} a node near the input locations will always
-have a highly restricted set of values.  More completely, the number of valid
-values for a connection gene in \emph{Normal} is related to the number of nodes
+in the genome.  In \emph{Normal} the number of valid
+values for a connection gene is related to the number of nodes
 preceding that gene, and independent of the number of nodes following that gene.
@@
 
-Empirical results are an important step in confirming our examinations.
-We accomplish this by testing the effectiveness
+\begin{comment}
+Empirical results are an important step in confirming our theoretical examinations.
+Therefore we have tested the effectiveness
 of the described variants and their combination in problem solving.  Deeper analysis of
@@
 Table~\ref{tab:glossary} gives a summary of what each variant means.
+\end{comment}
 
@@
 (\emph{Skip}, \emph{Accumulate}, and \emph{Single}) and two new methods for
-genome ordering (\emph{Reorder} and \emph{DAG}).  As these two
-sets are non overlapping, and there is no intuitive groupings, we chose to try
-all possible combinations.  As a control we included \emph{Normal} genome ordering,
+genome ordering (\emph{Reorder} and \emph{DAG}).
+See Table~\ref{tab:glossary} for a short description each variant.
+As these two
+sets are non overlapping, and there is no intuitive groupings, we chose to test
+all possible combinations in our empirical analysis.  As a control we included \emph{Normal} genome ordering,
 as it provides insight to how each of the duplicate prevention techniques work
@@
 The parameter configurations are then sorted based on their median.  The best half are then
-each run 4 more times, with the best half of that group chosen.  In this way, 64 configurations
-are run 5 times, 32 are run 9 times, 16 are run 13 times, 8 are run 17 times, 4 are run 21 times,
+each run 4 more times, with the best half of that group chosen.  In this way, 32 configurations
+are run 5 times, 16 are run 9 times, 8 are run 13 times, 4 are run 17 times, 2 are run 21 times,
 and 2 are run 25 times.  For algorithm combinations using \emph{Single}, all 8 parameter configurations were run 17 times
@@
 Each configuration was run 50 times to ensure statistical power.  All told over
-11.5 billion evaluations were performed to gather our results.
+11.5 billion evaluations were performed to gather our results.  The source code
+used to produce our results is available from our website.\footnote{\url{https://website.name.removed.com/AnonymousCGP}}
 
@@
 This makes sense as each time an offspring is generated all of the parent's genes
-must first be copied.  Also, larger genome sizes correlate with more active nodes,
+must first be copied.  Furthermore, all genetic operations in CGP scale with genome size.
+Also, larger genome sizes correlate with more active nodes,
 each of which must be executed on each input tested during individual evaluation.
@@
 axis denotes the level of fitness the recorded individual obtained.  The shading
-of each coordinate gives the percentage of 50 runs that had an individual of
-the given fitness level such that the node at the given location had never
-previously been active.  While only Multiply is shown here, all of the
+of each coordinate gives the percentage of 50 runs in which the first individual of
+the given fitness level had no ancestor in which the node at a given location was active.
+While only Multiply is shown here, all of the
 other problems tested had similar behavior, with allowances for differences in
@@
       \If{$g$ is a connection gene}
-        \If{$g \in reconnect$}
+        \If{$footprint(g) \in reconnect$}
           \State $g \leftarrow reconnect[footprint(g)]$
@@
 the active set.  In general, if a set of nodes all produce the same semantic,
-whichever has the least transitive dependencies will be kept.  Note that this
+whichever is predicted to have the least transitive dependencies will be kept.  Note that this
 does not guarantee the minimum possible genome, as that would require
@@
 inactive nodes to be constantly changed facilitating mutational inclusion of what
-are effectively random nodes.  Under this theory, CGP would likely benefit from randomly resetting
-all inactive nodes just before they become active.  Note that with \emph{Reorder}
+are effectively random nodes.  Under this theory, CGP would likely benefit from periodically resetting
+all inactive nodes.  Note that with \emph{Reorder}
 and \emph{DAG} these conclusions do not hold as a reasonable amount of node

### 2013-07-16 CGP Journal/main.tex
Added a TODO so I would stop forgetting to do the acknowledgements
Manually filled in my citation information, wrote a bibliography

@@
 
+TODO
 The authors would like to thank...
@@
 \begin{IEEEbiography}{Brian W. Goldman}
-TODO
+received the B. S. degree and M.S. degree in computer science from the
+Missouri University of Science and Technology, Rolla, in 2010
+and 2012 respectively.
+
+He is currently a Ph.D. student with the Department of Computer Science
+and Engineering at Michigan State University, East Lansing.  His current
+research interests include evolutionary computation and genetic programming.
+
 \end{IEEEbiography}

### 2013-07-23 CGP Journal/main.tex
Added my image to my biography section
Final content edit, now its time to shrink things

@@
 in general, the complex genotype to phenotype map makes achieving this understanding a
-challenge.  By examining tuned parameter values, the search quality of CGP variants at different
+challenge.  By examining aspects such as tuned parameter values, the search quality of CGP variants at different
 problem difficulties, node behavior, and offspring replacement properties we
-seek to comprehend observed characteristics of CGP search.  Our focus is twofold:
-methods to prevent wasted CGP evaluations (\emph{Skip}, \emph{Accumulate},
-and \emph{Single}) and methods to overcome CGP's search limitations imposed by
+seek to better understand the characteristics of CGP search.  Our focus is twofold:
+creating methods to prevent wasted CGP evaluations (\emph{Skip}, \emph{Accumulate},
+and \emph{Single}) and creating methods to overcome CGP's search limitations imposed by
 genome ordering (\emph{Reorder} and \emph{DAG}).
@@
 encode the evolved solution.
-Without \emph{Reorder} or \emph{DAG} almost the entire latter portion of the genome
-is never used by any ancestor of the evolved solution.  More generally, our
+Furthermore, traditional CGP ordering results in large portions of the genome
+that are never used by any ancestor of the evolved solution.
+\emph{Reorder} and \emph{DAG} allow evolution to utilize the entire genome.
+More generally, our
 results suggest that \emph{Skip-Reorder} and \emph{Single-Reorder}
-are most likely to solve hard problems in the least number of evaluations and
-the least amount of time with the least amount of degenerate
+are most likely to solve hard problems using the least number of evaluations and
+the least amount of time while better avoiding degenerate
 behavior.
@@
 This is exceptionally
-true for Genetic Programming (GP), which often employs nontrivial
+true for Genetic Programming (GP), which often employs both nontrivial
 genotype to phenotype maps and operators that act on genotypes without regard to
@@
 Genetic Programming (CGP) an effective evolutionary optimizer.
-A few studies of CGP's evolutionary mechanisms have
-been done previously, related to bloat~\cite{miller:2001:bloat},
+Previous studies of CGP's evolutionary mechanisms have
+focused on bloat~\cite{miller:2001:bloat},
 neutrality~\cite{vassilev:2000:neutrality}, and structural bias~\cite{payne:2009:bias}.
@@
 prevents the creation of cycles, while still allowing CGP to reuse values.  This is
-in contrast to tree based GP, which must recalculate a value everywhere it is needed.
+in contrast to tree based GP, which must duplicate functionality everywhere the same value is needed.
 To complete the representation, a set
@@
 being used to create output values referred to as ``inactive.''  Inactive nodes
-allow CGP individuals to drift genetically, as they can be mutated without effecting
+allow CGP individuals to drift genetically, as they can be mutated without affecting
 the fitness of the individual.  This genetic drift can then be incorporated, as
@@
 is $\mu + \lambda$ where $\mu \leftarrow 1$ and $\lambda \leftarrow 4$.
-This means each generation a single parent produces four offspring using
+This means during each generation a single parent produces four offspring using
 mutation.  The best offspring then competes with the parent,\footnote{Ties are broken
@@
 if a function gene is chosen for mutation, its new value is randomly chosen from
-all possible functions, with the exception of the genes current value.  Combined,
+all possible functions, excluding the gene's current value.  Combined,
 this form of CGP can be viewed as a stochastic hill climber with neutrality.
@@
 parents, and therefore by definition have identical fitness to their parents.
-As such, re-evaluating these individuals is computationally wasteful, as their fitness
+As such, re-evaluating these individuals is computationally wasteful as their fitness
 is known.
-Detecting duplicated individuals can be done without significant overhead as
+Detecting duplicated individuals can be done without significant overhead since
 finding the set of active nodes is already done prior to
 evaluation~\cite{vasicek:2012:efficient}.
-Previous work~\cite{goldman:2013:cgpwaste} has taken an initial look at the
+Our previous work~\cite{goldman:2013:cgpwaste} has taken an initial look at the
 effect this waste can have, and proposed \emph{Skip}, \emph{Accumulate}, and
@@
 If the mutation rate is high enough relative
-to the number of active genes, \emph{Skip} will require the same number of evaluations as traditional CGP. This is because
+to the number of active genes, \emph{Skip} will require the same number of evaluations as traditional CGP, since
 the probability of an offspring being actively identical to its parent is effectively zero.
@@
 
-We would expect both intuitively and from initial
+We would expect both intuitively and from our initial
 experimentation~\cite{goldman:2013:cgpwaste} that \emph{Skip} will be less sensitive
@@
 mutated active gene must either be to an intron or represent another way to
-code a solution of equal quality.  Second, as a benefit of forced changes
+code a solution of equal quality.  Second, as a benefit of forced changes,
 \emph{Single} avoids the overhead of repeatedly generating individuals without
@@
 Ordering, when combined with the mutation's uniform resetting of connection genes, may be the cause of CGP's
-immensely inactive genomes.  As was rigorously defined in~\cite{goldman:2013:ordering},
+immensely inactive genomes.  As was rigorously defined in our work~\cite{goldman:2013:ordering},
 the number of active genes in a genome is expected to scale logarithmically
@@
 derived the formula which predicts the probability a node is active based on its
-index, given in Equation~\ref{eq:prob} and shown in Figure~\ref{fig:prob}.
+index, given in Equation~\ref{eq:prob} and plotted in Figure~\ref{fig:prob}.
 \begin{equation}
@@
 genome.  As the shuffling does not semantically change the parent, it does
-not require re-evaluated.  Furthermore, as shuffling does not change the requirement
+not require re-evaluation.  Furthermore, as shuffling does not change the requirement
 that nodes only depend on those preceding them in the genome, shuffling does not
@@
 have changed.  Consider two nodes $X$ and $Y$ which are in the genome, and
-have no dependence relationship.  As $X$ does not take input from $Y$ and
+have no dependent relationship.  As $X$ does not take input from $Y$ and
 $Y$ does not take input from $X$, either can be placed before the other
@@
 
-The price \emph{Reorder} pays for this potential improvement is increased mutational search
+The price \emph{Reorder} pays for this potential improvement is a larger mutational search
 space.  Though a shuffled individual is phenotypically unchanged,
-its mutational adjacency is significantly different.  As such the variety of offspring
+its mutational adjacency can be significantly different.  As such the variety of offspring
 an individual can create is increased when using \emph{Reorder}.  While this increase
@@
 an efficient protection against cycles: nodes must connect forward.  \emph{DAG}
-removes this scheme of location, relying on more complex methodologies to
+removes this restriction and allows forward and backward connections, relying on other, more complex, methodologies to
 prevent cycles.  To accomplish this goal, modifications must be made to how CGP
@@
 which nodes transitively depend on the mutating node and which do not, with
-information stored in the $known$ map.  This data structure is initialized with
+information stored in a $known$ map data structure.  This data structure is initialized with
 the information that the node is dependent on itself and that all input locations
@@
 are reached.  At this point, the process changes.  In \emph{DAG}, we must determine not only the set of
-nodes that are active, but the order in which those nodes should be executed.
+nodes that are active, but also the order in which those nodes should be executed.
 
@@
 While \emph{Reorder} and \emph{DAG} both reduce mutational limitations, \emph{DAG}'s changes are more sweeping.
-Though both can create any connect that does not form a cycle,
+Though both can create any connection that does not form a cycle,
 \emph{Reorder} maintains a bias toward certain kinds of connections.
@@
 
-The probability of connecting $X$ to $Y$ is also influenced by how many nodes transitively depend $X$ and $Y$.
+The probability of connecting $X$ to $Y$ is also influenced by how many nodes transitively depend on $X$ and $Y$.
 For example, assume $X$ has more nodes transitively dependent on it than $Y$.
@@
 if applied alone.  Similar measures are not needed in testing the genome ordering techniques,
-as \emph{Skip} has no impact on evolutionary search, and is a strict improvement
+as \emph{Skip} has no impact on evolutionary search, and is strictly an improvement
 to using no duplication detection at all.
 
-In total this creates 9 algorithm combinations.  Of these, previous work
+In total this creates 9 algorithm combinations.  Of these, our previous work
 has been reported showing the effectiveness of the duplication methods with \emph{Normal} ordering~\cite{goldman:2013:cgpwaste}
@@
 of methods only being effective after extensive problem specific tuning, we set out to use
-rough grained values and provide each technique with equal tuning time to alleviate that bias.
+rough-grained values and provide each technique with equal tuning time to alleviate that bias.
 
@@
 	is the control configuration \emph{Skip-Normal}.  MES=Median Evaluations to Success; MAD=Median
-	Absolute Deviation; p-value=Result from Mann-Whitney~U comparison with control.
+	Absolute Deviation; Active=Median number of active nodes in evolved solution;
+	Used=Median number of nodes active in evolved solution after applying SIMPLIFY (See Section~\ref{sec:behavior});
+	p-value=Result from Mann-Whitney~U comparison with control.
 	}
@@
 on Multiply.  No other algorithm had as much impact on genome size tuning as
-the inclusion or exclusion of \emph{Single}.  In previous work~\cite{goldman:2013:ordering}
+the inclusion or exclusion of \emph{Single}.  In our previous work~\cite{goldman:2013:ordering}
 it appeared that \emph{Reorder} and \emph{DAG} allowed CGP to use smaller genomes.
-In light of our results, this conclusion should be amended.
+In light of our current results, this conclusion should be amended.
 It appears that \emph{Single} is most effective when the genome size is small and
@@
 
-Unlike genome size, all configurations choose approximately the same mutation rate,
-with different rates for each problem.  With the exception of Parity, there appears
+Unlike genome size, all configurations choose approximately the same mutation rate
+for each problem.  With the exception of Parity, there appears
 to be a general consensus that each gene should be mutated with probability 0.002,
@@
 On Parity and Encode it is probably safe to conclude that \emph{Single} is worse than
-the control.  As these are the easiest problems tested that difference may have
+the control.  As these are the easiest problems tested, that difference may have
 little real world meaning.  \emph{DAG} is almost certainly worse than the control on
@@
 
-By ordering the algorithm combinations by MES and examining that ordering across each
+By ordering the algorithm combinations by median evaluations to success (MES) and examining that ordering across each
 problem, we can look for general trends.  From the table, it appears that on easier problems
@@
 the ordering methods are more predictive on harder problems.  This comes from the
-observation that in for Parity and Encode (Upper half of Table~\ref{tab:results})
+observation that in Parity and Encode (Upper half of Table~\ref{tab:results})
 the first column is effectively sorted, while for Decode and Multiply
@@
 used by CGP would have allowed that offspring to replace its parent, at least probabilistically.  This leads
-to two possible causes for a node never being active: bad genes or infrequently tried.
-
-The bad genes cause for a node never being active suggests the behavior encoded
+to two possible causes for a node never being active: bad genes or infrequent node sampling.
+
+The bad genes cause suggests the behavior encoded
 by a node is simply useless on the problem landscape.  If this was the root cause
@@
 
-The infrequently tried cause for a node never being active suggests that few
+The infrequent node sampling cause suggests that few
 individuals are ever produced which have that node active, limiting the likelihood
@@
 \subsection{Node Behaviors}
+\label{sec:behavior}
 
@@
 we first determine the semantics of each node~\cite{mcphee:2008:treesemantics}.
-For each input that is given to an individual, all of its nodes
+By semantics we mean that for each input to an individual, each node
 will produce a single output.  These output values can be recorded and ordered.
@@
 The \Call{Simplify}{} algorithm given in Figure~\ref{fig:simplify} is designed
-to efficiently reduce redundancy encoded in a genome.  Generally, this
-algorithm will move all connections to the first node that produces the desired
+to efficiently remove redundancy encoded in a genome.  This
+algorithm moves all connections to the first node that produces the desired
 semantic.  After application, and recalculation of which nodes are still
@@
 the plurality of active nodes in the ``Used'' category appears to correlate with
-increase evaluations to success.  Notice that in all \emph{Single} plots the
+increase evaluations to success in Table~\ref{tab:results}.  Notice that in all \emph{Single} plots the
 tallest bar for active nodes is ``Used,'' and the only non \emph{Single} plots
@@
 often CGP reuses previously useful active nodes.  Reported in the table is how
-frequently an offspring replaces its parent with no reactivated nodes.  Smaller
+frequently an offspring replaces its parent such that is has no reactivated nodes.  Smaller
 values indicate an algorithm combination that frequently reuses nodes, as opposed to
@@
 
-\end{comment}
 % use section* for acknowledgement
@@
 The authors would like to thank...
+\end{comment}
 
@@
 
-\begin{IEEEbiography}{Brian W. Goldman}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{goldman_headshot}}]{Brian W. Goldman}
 received the B. S. degree and M.S. degree in computer science from the
@@
 % if you will not have a photo at all:
-\begin{IEEEbiographynophoto}{William F. Punch}
+\begin{IEEEbiography}{William F. Punch}
 TODO
-\end{IEEEbiographynophoto}
+\end{IEEEbiography}
 

### 2013-07-23 CGP Journal/main.tex
Added stuff for Bill's bio

@@
 % if you will not have a photo at all:
-\begin{IEEEbiography}{William F. Punch}
-TODO
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{headPunch}}]{William F. Punch}
+Bill Punch received his Ph.D. from the Ohio State University and is currently an
+Associate Professor at the Computer Science and Engineering Department of Michigan
+State University. He is also the  director of the MSU High Performance Computing
+Center. His research interests are evolutionary computation, high performance
+computing and computing pedagogy. He co-directs the Genetic Algorithms Research
+and Application Group (GARAGe) and is on the executive committee of the NSF STC,
+BEACON, at MSU. His book with Rich Enbody titled "The Practice of Computing Using
+Python" is now in its second edition.
 \end{IEEEbiography}

### 2013-07-23 CGP Journal/main.tex
Rephased three paragraphs into one in the never active section, made last table single column

@@
 
+\begin{comment}
 So now the question is why does \emph{Normal} use so little of its genome?  To
@@
 to have less or no positional bias, which also matches the results given in Table~\ref{tab:never_active}.
+%\end{comment}
+
+%The above paragraphs were replaced by the below paragraph.
+In Section~\ref{sec:ordering} we used Equation~\ref{eq:prob} to suggest \emph{Normal}
+as a strong bias on the location of active nodes regardless of fitness landscape, and that
+\emph{Reorder} and \emph{DAG} would likely have less or no bias.
+Figure~\ref{fig:prob} shows the expected probability given by Equation~\ref{eq:prob},
+using the same genome size, arity, number of inputs, and number of outputs as our
+3-bit multiply.  This curve bears a striking resemblance to all three \emph{Normal}
+plots in Table~\ref{tab:never_active}.
 
@@
 
-\begin{table*}
+\begin{table}
 	\centering
@@
 	  \multicolumn{5}{|c|}{\textbf{3-bit Parity}} \\ \hline
-\textbf{Duplicate} & \textbf{Ordering} & \textbf{Active Unchanged} & \textbf{No Reactivation} & \textbf{Reactivated Changed} \\ \hline
+%\textbf{Duplicate} & \textbf{Ordering} & \textbf{Active Unchanged} & \textbf{No Reactivation} & \textbf{Reactivated Changed} \\ \hline
+\textbf{Duplicate} & \textbf{Ordering} & \textbf{Act-Un} & \textbf{No React} & \textbf{React Changed} \\ \hline
        \emph{Skip} &     \emph{DAG} & 0.00\% & 3.48\% & 55.63\% \\ \hline
@@
 
-\textbf{Duplicate} & \textbf{Ordering} & \textbf{Active Unchanged} & \textbf{No Reactivation} & \textbf{Reactivated Changed} \\ \hline
+%\textbf{Duplicate} & \textbf{Ordering} & \textbf{Active Unchanged} & \textbf{No Reactivation} & \textbf{Reactivated Changed} \\ \hline
+\textbf{Duplicate} & \textbf{Ordering} & \textbf{Act-Un} & \textbf{No React} & \textbf{React Changed} \\ \hline
  \emph{Accumulate} &     \emph{DAG} & 0.05\% & 4.06\% & 48.70\% \\ \hline
@@
 
-\textbf{Duplicate} & \textbf{Ordering} & \textbf{Active Unchanged} & \textbf{No Reactivation} & \textbf{Reactivated Changed} \\ \hline
+%\textbf{Duplicate} & \textbf{Ordering} & \textbf{Active Unchanged} & \textbf{No Reactivation} & \textbf{Reactivated Changed} \\ \hline
+\textbf{Duplicate} & \textbf{Ordering} & \textbf{Act-Un} & \textbf{No React} & \textbf{React Changed} \\ \hline
  \emph{Accumulate} &     \emph{DAG} & 0.03\% & 10.10\% & 54.99\% \\ \hline
@@
 
-\textbf{Duplicate} & \textbf{Ordering} & \textbf{Active Unchanged} & \textbf{No Reactivation} & \textbf{Reactivated Changed} \\ \hline
+%\textbf{Duplicate} & \textbf{Ordering} & \textbf{Active Unchanged} & \textbf{No Reactivation} & \textbf{Reactivated Changed} \\ \hline
+\textbf{Duplicate} & \textbf{Ordering} & \textbf{Act-Un} & \textbf{No React} & \textbf{React Changed} \\ \hline
  \emph{Accumulate} & \emph{Reorder} & 1.14\% & 22.09\% & 71.75\% \\ \hline
@@
 	\caption{Behavior of offspring replacement for each algorithm combination
-	         on each problem.}
+	         on each problem.
+	         Act-Un=Active Unchanged; No React=No Reactivation; React Changed=Reactivation Changed}
 	\label{tab:reactivate}
-\end{table*}
+\end{table}
 

### 2013-07-23 CGP Journal/main.tex
Shrank figures from .3 column width to .285, removed section from conclusion that restated minor conclusion

@@
 \newcommand{\graphicthird}[1]
-{\includegraphics[width=.3\textwidth,height=.3\textheight,keepaspectratio]{#1}}
+{\includegraphics[width=.285\textwidth,height=.285\textheight,keepaspectratio]{#1}}
 
@@
 than with it overcoming whatever causes traditional CGP to prefer enormous genomes.
+\begin{comment}
 While \emph{Reorder} and \emph{DAG} were able to use smaller genomes than \emph{Normal}
@@
 by reducing the genome size.
+\end{comment}
 

### 2013-07-23 CGP Journal/main.tex
Added NSF stuff to the commented out area, just in case I want it later

@@
 Appendix two text goes here.
-
 % use section* for acknowledgement
@@
 
-TODO
-The authors would like to thank...
+This material is based in part upon work supported by the National Science Foundation
+under Cooperative Agreement No. DBI-0939454. Any opinions, findings, and conclusions
+or recommendations expressed in this material are those of the author(s) and do not
+necessarily reflect the views of the National Science Foundation.
 \end{comment}

### 2013-09-10 Parameterless Population Pyramid/main.tex
Setup for P3 paper

--- 2013-01-25 Length Bias and Seach Limitations in CGP/main.tex
@@
 % --- Author Metadata here ---
-\conferenceinfo{GECCO'13,} {July 6-10, 2013, Philadelphia, Pennsylvania, USA.}
-\CopyrightYear{2013}
-\crdata{TBA}
+\conferenceinfo{GECCO'14,} {July 12--16, 2014, Vancouver, BC, Canada.}
+\CopyrightYear{2014}
+\crdata{978-1-4503-1963-8/13/07}
 \clubpenalty=10000
@@
 
-\title{Length Bias and Search Limitations in Cartesian Genetic Programming}
-\subtitle{[Genetic Programming Track]}
+\title{Parameterless Population Pyramid}
+\subtitle{[Genetic Algorithms Track]}
 %
@@
 \alignauthor
-William Punch\\
+William F. Punch\\
        \affaddr{BEACON Center for the Study of Evolution in Action}\\
        \affaddr{Michigan State University, U.S.A.}\\
-       \email{punch@cse.msu.edu}
+       \email{punch@msu.edu}
 }
@@
 \begin{abstract}
-%Understanding how configuration decisions can interact with individual representation and
-%evolutionary mechanisms in Cartesian Genetic Programming (CGP) to cause limitations
-%on search is vital to proper algorithm application.  
-In this paper we examine how Cartesian Genetic Programming's (CGP's) method
-for encoding directed acyclic graphs (DAGs) and its mutation operator bias the
-effective length of individuals as well as the distribution of inactive nodes
-in the genome.  We investigate these biases experimentally using two CGP variants as comparisons: \emph{Reorder},
-a method for shuffling node ordering without effecting individual evaluation, and
-\emph{DAG}, a method for removing the concept of node position.
-Experiments were performed on four problems tailored to highlight potential search
-limitations, with further testing on the 3-bit multiplier problem.
-
-Unlike previous work, our experiments show that CGP has an innate parsimony pressure
-that makes it very difficult to evolve individuals with a high percentage of active nodes.  This bias is
-particularly prevalent as the length of an individual increases.  Furthermore, these problems are compounded by CGP's positional
-biases which can make some problems effectively unsolvable.  Both \emph{Reorder}
-and \emph{DAG} appear to avoid these problems and outperform \emph{Normal} CGP
-on preliminary benchmark testing. Finally, these new techniques require more
-reasonable genome sizes than those suggested in current CGP, with some evidence
-that solutions are also more terse.
 
@@
 % A category with the (minimum) three required fields
-\category{Computing Methodologies}{Artificial Intelligence}{Search Methodologies}
+%\category{Computing Methodologies}{Artificial Intelligence}{Search Methodologies}
+\category{TODO}{TODO}{TODO}
 %A category including the fourth, optional field follows...
@@
 
-\keywords{Cartesian Genetic Programming, Neutrality, Bloat}
-\\
-\\
+\keywords{TODO}
+
 \section{Introduction}
-In all black box search, search designers must make assumptions about the properties
-of the problem in order to outperform random search.  When these assumptions
-match the application, search time can be greatly reduced.  As such, understanding how design decisions
-effect search effectiveness or even prevent search from finding solutions is very important.
-%
-In Cartesian Genetic Programming (CGP) the maximum genome size of an individual
-is set at initialization.  While this
-requires the maximum genome size to be larger than the problem's solution needs,
-the question remains; how will the maximum genome size effect evolution's
-speed at finding solutions.  As larger genomes inherently represent larger search
-spaces, intuition suggests that using the lowest possible maximum size will have the most efficient search.
-However, in CGP a mounting body of evidence suggests the optimal
-maximum size is much larger than the actual solutions evolved~\cite{miller:2006:redundancy}.
-We examine the cause of this discrepancy by applying CGP and a few
-modified versions of CGP to problems which test specific edge cases in representation.
-The result is a better understanding of the search limitations of CGP.
-
-\section{CGP}
-\label{sec-cgp}
-In Cartesian Genetic Programming (CGP) individuals use a fixed length string of integers to
-encode directed acyclic graphs (DAGs).  
-Each node in the DAG is encoded using multiple genes: one specifying a function and
-the rest specifying node input connections.
-Graph output is determined by additional genes which can connect to any node or
-program input.
-To prevent cycles,
-nodes may only connect to problem inputs or nodes that precede them in the genome.
-Early versions of CGP limited valid connections further through user specified graph
-topologies.
-While specifying topology is still useful
-when evolving circuits with known hardware limitations, the less restricted form
-allows CGP to represent all possible
-DAGs of size \emph{N}, where \emph{N} is the number of nodes in the genome.
-%
-The most common evolutionary mechanism in CGP is a $1+\lambda$ population
-where each generation the best offspring replaces the parent if it is no less
-fit.  CGP almost exclusively uses a form of point mutation to create offspring,
-with no form of crossover commonly being employed.  As such, these are the methods
-considered in our analysis.
-For a more detailed description of CGP, see~\cite{miller:2011:chapter2}.
-
-%\subsection{Graph Size, Bloat, and Inactive Genes}
-While CGP imposes a maximum graph size \emph{N}, the minimum
-effective graph size is variable, as there can be unused portions
-of the genome.  These inactive nodes exist because both output locations and
-internal connections can be changed by evolution.  A trivial example is when all outputs
-are connected directly to problem inputs the entire genome becomes
-inactive.  Thus the active nodes are those nodes used when computing output.
-%With all
-%of these features combined, it should be clear that CGP is capable of evolving
-%variable sized DAGs with a bounded maximum size.
-
-A pernicious problem in the Genetic Programming (GP) community is individual bloat, in which solutions
-become larger with time without substantive improvement in quality~\cite{luke:2006:bloat}.
-Experimental evidence has so far suggests that CGP is able to avoid
-bloat, as the evolved number of active
-nodes does not grow during evolution~\cite{miller:2001:bloat} nor does it grow significantly
-as the maximum number of nodes is raised~\cite{miller:2006:redundancy}.
-
-%\subsection{Neutral Search}
-The existence of inactive genes provides CGP with an explicit form of genetic
-introns, allowing the evolutionary process to perform neutral search.
-CGP's replacement strategy facilitates neutral space
-exploration by ensuring offspring replace their parents if they are no less fit.
-This allows the population to drift,
-which has been shown to be beneficial~\cite{yu:2001:neutrality}.  Furthermore,
-there is evidence that as much as 95\% of the genotype should be inactive
-to allow for sufficient neutral search~\cite{miller:2006:redundancy}.
-
-However, this approach to neutral search may have disadvantages.  In response to a previous claim
-about neutrality's beneficial effects on needle in a haystack style
-problems, Collins~\cite{collins:2006:haystacks} showed that on problems with no search
-gradient, efficient random sampling can significantly outperform CGP with neutral search.
-Collins argued that the benefits CGP claimed to be receiving through neutral search
-may be better attributed to higher maximum genome sizes causing an increase in
-the total number of active genes.
-
-\section{Length and Positional Biasing}
-\label{sec-length-biasing}
-Despite CGP's ability to represent all possible DAGs
-of size \emph{N}, there is little evidence that it uniformly
-searches the space of DAGs. In fact,~\cite{payne:2009:bias} showed that certain types of structures
-are more difficult to evolve than others.  In particular, larger structures may be more difficult to
-evolve as they are less common in the search space. 
-Tree based GP has similar issues,
-with strong evidence that specific types of trees are more difficult to
-evolve even though they are common in the search space~\cite{daida3:2003:treebias}.
-We suggest that
-the mechanics used to initialize and evolve connection genes in CGP
-impose significant biases on the structures of the DAGs evolved, with more bias against
-larger genotypes than should be expected by just representation frequency.
-
-When initializing and mutating a connection gene, its value is randomly set to
-either a preceding node in the genome or to an input.  As was discussed in~\cite{collins:2006:haystacks},
-this means the expected location of a new connection is $\frac{i}{2}$ where \emph{i}
-is the number of nodes and inputs that precede the mutated node.  If all nodes had
-a single connection gene, we would expect random individuals to have a length of $\log_2N$.
-Precise calculation of the expected length is made difficult
-due to node overlap, but the relationship will remain logarithmic to the total genome size.
-
-The probability that a node is active changes greatly depending on where
-in the genome it is encoded.  Consider nodes near the end of the genome.
-The highest index node expected to be active is given by Equation~\ref{eq-active-end}.
-\begin{equation}
-  \left ( N+I \right )\frac{O}{O + 1}
-  \label{eq-active-end}
-\end{equation}
-Here, \emph{N} is the number of nodes in the genome, \emph{I} is the number of input locations,
-and \emph{O} is the number of output locations.  This equation is derived from the fact that
-if numbers are drawn uniformly at random and then ranked, the mean value for each rank
-will be evenly spaced in the range.  As a result, the expected highest output location will be
-the same as if \emph{O} locations were placed evenly on the range $N + I$.
-On problems with a single
-output (such as the commonly used parity problem) this means, on average, nodes
-in the second half of the genome are inactive.  
-%Problems with
-%more outputs will be less effected as Equation~\ref{eq-active-end} approaches $N + I$
-%as \emph{O} grows.
-Some CGP implementations reduce this problem by
-fixing the output locations to the last \emph{O} nodes in the genome, which has the effect
-of scaling \emph{O} in Equation~\ref{eq-active-end} by the arity of the nodes in use.  However, this does not
-remove the probability for large sections of inactive nodes near the end of the genome.
-
-Conversely, nodes near the beginning of the genome are almost always active.
-Equation~\ref{eq-incoming} shows how the expected number of incoming connections
-increases as the number of nodes following node \emph{i} increases and the
-arity (\emph{a}) used increases.  This relationship is non-linear monotonic, with input locations
-and nodes near the inputs likely to have more than one incoming connection while nodes
-near the output location will likely have less than one incoming connection.
-\begin{equation}
-  \sum_{j=i+1}^{N} \frac{a}{j}
-  \label{eq-incoming}
-\end{equation}
-Equation~\ref{eq-incoming} is derived from the fact that all nodes following \emph{i}
-in the genome have \emph{a} connections which each have a $\frac{1}{j}$ chance of
-connecting to \emph{i}.
-
-The probability that a node is active increases as the number of incoming connections
-increases, as each could potentially come from an active node.  This creates a self
-reinforcing cycle:  The closer a node is to the inputs the more incoming connections
-it can receive from other nodes with lots of incoming connections.
-As a result, nodes near the inputs are almost
-always active while nodes near the outputs are almost never active.  Furthermore,
-the output of nodes near the inputs is likely to be used by multiple active nodes.
-
-Having many active nodes near the start and few near the end has important
-implications on CGP's ability to perform neutral search.
-If there are no inactive nodes preceding a given
-node, it cannot benefit from neutral evolution.  Nodes that have a high percentage
-of inactive nodes preceding them will conversely have a high probability of attaching
-to inactive nodes, but will likely be inactive themselves.
-
-\begin{figure}
-  \centering
-  \includegraphicsfit{ThoughtExperiment}
-  \caption{Thought experiment illustrating limitations in \emph{Normal} CGP.
-           Arrows represent the flow of information.} 
-  \label{fig-thought}
-\end{figure}
-
-How this positional bias of active genes effects search can be examined
-using a thought experiment, shown in Figure~\ref{fig-thought}.  
-Each of the three Cases given are of the second best phenotype, in which \emph{Z}
-takes its input from \emph{X}, while in the optimal phenotype \emph{Y} is applied
-on \emph{X}'s output before input to \emph{Z}.  In Case 1 there are inactive nodes
-between \emph{X} and \emph{Z}, so CGP is able to use drift to insert \emph{Y}.
-Unfortunately, because active nodes cluster heavily near the start of the genome,
-this is unlike to happen if even a moderate percentage of the genome is active.
-In Case 2, \emph{X} and \emph{Z} are adjacent nodes, or equivalently all nodes between
-\emph{X} and \emph{Z} are active.  CGP can only solve this case by recreating \emph{X}
-and/or \emph{Z} somewhere else in the genome using drift, converting this back
-to Case 1.  That process is highly unlikely as it involves drift of active genes.
-In Case 3, \emph{Y} already exists in the genome, but in the wrong order relative to \emph{Z}.
-CGP can solve this by either reinventing \emph{Y} as in Case 1 or moving \emph{Z} as in Case 2.
-From a phenotypic standpoint, the first two cases should be equally difficult with the third
-being easier, as \emph{Y} already exists.  However, due to CGP's positional biases, Case 2 is much harder than Case 1,
-as active nodes must be reinvented before \emph{Y} can be evolved and added between \emph{X} and \emph{Z}.
-Furthermore, Case 3 is just as hard as Case 1, as even though \emph{Y} already exists, \emph{Z} cannot connect to it.
-Case 2 and 3 become even more difficult if \emph{X}, \emph{Y}, and \emph{Z} are thought of as
-collections of nodes used to produce a single output, as recreating these complex structures
-increases the difficulty.  They also become more common as interwoven structures in the
-genome act the same as adjacent nodes in this experiment.
-
-\section{Avoiding Search Limitations}
-\label{sec-modifying-cgp}
-In order to experimentally examine how the mechanisms of CGP bias search, we propose two CGP variants.
-The first
-is \emph{Reorder}, which attempts to mitigate positional bias by performing phenotypically
-neutral node shuffling.  The second is \emph{DAG}, which allows nodes to connect
-to any other node in the genome as long as no cycle is created.  For clarity, we will
-refer to standard CGP as \emph{Normal}.  Note that both variants are capable of
-representing the exact same set of graphs as \emph{Normal}.  They only differ
-in how graphs can be modified by evolution.
-
-\subsection{Reorder}
-\label{sec-reorder}
-\emph{Reorder} works identically to normal CGP with a single change:
-Before producing offspring, the parent's nodes are shuffled without changing the phenotype of the individual.
-That is, we perform neutral search on the parent's node ordering.
-This process begins by building bidirectional node dependency sets, such that
-each node's input and output sets can be easily accessed.  First the
-set of \emph{addable} nodes is created, which initially contains all of the
-inputs.  The new node ordering is then determined iteratively using the following
-steps:
-\begin{enumerate}
-  \item Select and remove a node $\alpha$ at random from the \emph{addable} set.
-  \item If $\alpha$ is not an input, map it sequentially to the next genome location.
-  \item For each node $\beta$ which takes output from $\alpha$:
-  \begin{enumerate}
-    \item Mark $\beta$'s dependency on $\alpha$ as satisfied.
-    \item If $\beta$'s dependencies have all been satisfied, add it to \emph{addable}.
-  \end{enumerate}
-\end{enumerate}
-Iteration ends
-when all nodes have been given new locations in the sequence.  At this point, nodes can be moved
-into their new locations, with their connections being updated using the ordering
-map.  Once reordered, nodes maintain the normal CGP requirement that all nodes
-must take their inputs from nodes preceding them in the genome, meaning that
-the mutation and evaluation process remains the same.  As all connections and functions are
-preserved during this process, there is no change in the result obtained by executing
-an individual.  \emph{Reorder} is just a neutral permutation of the node ordering.
-
-Repeating the thought experiment in Figure~\ref{fig-thought} with
-\emph{Reorder} shows how it can differ from \emph{Normal}.  Because inactive nodes
-can be shuffled between \emph{X} and \emph{Z}, Case 1 and 2 become identical.
-Case 3 can be
-solved by \emph{Reorder} without recreating \emph{Y}, as it can
-be shuffled to precede \emph{Z}.
-
-Without breaking connections, a node can be moved forward such that
-it transitively depends on all preceding nodes.  A node can also be moved back until all following nodes
-transitively depend on it.  Furthermore, the distance between a node and
-its input with the highest amount of transitive dependency will be on average smaller
-than in \emph{Normal}.  This is illustrated
-when considering graphs with a node arity of one.  In \emph{Normal} and at initialization
-of \emph{Reorder}, a node is on average $\frac{i}{2}$ indexes away from where it gets its input and $2i$ indexes
-from where its output goes.  In \emph{Reorder}, nodes will on average be equidistant to
-their their inputs and outputs.  In this case, the expected distance is related to the total number
-of nodes in the dependency chain.  This means that an increased number of active nodes can
-occupy higher index locations as they are pushed to the center between their inputs and outputs.
-
-This does not directly translate into \emph{Reorder} being able to create more active graphs,
-or mean that \emph{Reorder} is without length bias.  When creating new connections,
-the expected location is still $\frac{i}{2}$, which means given enough mutations,
-the individual can revert back to all of the expectations for \emph{Normal}.
-In general however, \emph{Reorder} will allow nodes on which
-many nodes transitively depend to increase how many nodes can connect to them, while simultaneously
-allowing nodes with only a few transitive dependencies to gain more dependencies.
-
-A final artificial bias in both \emph{Reorder} and \emph{Normal} is
-that all connections in a node will tend to have the same transitive
-dependencies.  This is because if node $\beta$ is already transitively dependent
-on node $\alpha$, then $\alpha$ is always a valid connection when generating new connections in $\beta$.
-This is because all transitive dependencies to a node always precede that node, even
-when using \emph{Reorder}.  Nodes that $\beta$ is not already transitively dependent
-on have a probability of being unreachable, and therefore are connected to less frequently.
-
-\subsection{DAG}
-\label{sec-dag}
-The concept behind \emph{DAG} is to relax the requirement that nodes must connect
-forward in the genome.
-To accomplish this, only two portions of CGP have to be
-modified: connection gene mutation and evaluation ordering.
-In \emph{Normal}, when a connection gene is mutated, its value is randomly changed
-to any input or node index less than the mutating node's index.  In
-\emph{DAG}, when a connection gene is chosen to be mutated, new connections are
-tried randomly and kept if they do not create a cycle.
-
-At the start of each mutation, a mapping is developed to store which nodes are
-transitively dependent on the mutating node, and which are not.  Initially this map stores that
-the input indexes are not dependent on the mutating node, and
-that the node is dependent on itself.  Nodes are then tested in a random
-order for transitive dependence.  This is performed by recursively applying
-the following rules:
-\begin{itemize}
-  \item If the node's dependency is already known, return the stored information.
-  \item If this node takes input from a dependent node, it is dependent.
-  \item If none of its inputs are dependent, it is not dependent.
-\end{itemize}
-As information is collected, the map is updated to ensure a node is never checked
-for dependency twice.  If the randomly chosen node is not dependent on the mutating
-node, it is used as the new connection value. While this algorithm can result in
-each mutation needing to traverse a large portion of the graph searching for
-dependencies, the growth is at worst linear.  Furthermore, our reason for
-introducing \emph{DAG} is to analyze limitations in \emph{Normal}, not necessarily as
-a computationally efficient alternative.
-
-To determine node ordering during evaluation, \emph{DAG}
-performs a recursive walk of the active nodes starting from the output
-locations, recording for each node its bidirectional dependencies on other active nodes.
-Next, \emph{DAG} builds the ordering of active nodes similar to
-the \emph{Reorder} method in Section~\ref{sec-reorder}, ensuring only active nodes are added
-and that they are added only after all of their dependencies are in the list.
-Using this ordering, which is only built once for each individual, any number of
-input sequences can be evaluated with identical run time as \emph{Normal}.
-
-Removing the forward-only requirement found in \emph{Normal} and \emph{Reorder}
-allows \emph{DAG} to avoid positional bias.
-This is because any node can connect to any structure
-that exists in the graph as long as its not already part of that structure.  
-In contrast, \emph{Reorder} can only connect to nodes currently shuffled
-in front of the mutating node, while in \emph{Normal} structures may have
-to be reinvented to achieve the same connection.  \emph{DAG} can also
-insert inactive nodes between active nodes, as there is
-no concept of adjacency.
-
-As to individual length, \emph{DAG} has a much higher expected percentage of active nodes
-than \emph{Normal} or \emph{Reorder}.  When mutating, a connection can be made to
-any non-dependent node and input.  Using an arity of one as an example, the
-expected number of active nodes is $\frac{N}{I+1}$ where \emph{I} is the number
-of inputs.  Higher arity will increase the active proportion, but overlap makes
-precise estimations difficult.  Note that unlike \emph{Normal}, the number of
-active nodes in \emph{DAG} scales linearly with the total number of nodes.  Also,
-since \emph{DAG} still uses \emph{Normal}'s method of initialization, the actual
-individuals evolved may be less active than expected.
-
-While \emph{DAG}'s reduced search biases do not necessarily result in better results
-on specific problems, it should generally make \emph{DAG} effective on a wider
-range of applications.
-Also note that while evolution of complete DAGs has been done previously~\cite{niehaus:2007:ggp},
-our \emph{DAG} method mirror's CGP more faithfully and allows for more direct comparison.
-
-\section{Tailored Problems}
-\label{sec-tailored}
-To test the search biases of CGP and our variants,
-we propose a few benchmark problems specifically
-tailored to highlight certain types of bias.
-%When performing
-%experiments with any evolutionary system, it is necessary to specify all configuration
-%information in order to improve replicability of results.
-The following experimental configuration is used across all experiments reported here.
-%In order to simplify
-%matters, all experiments reported here use the same configuration.
-Keeping with existing CGP literature, we opted for a $1+4$ strategy using
-mutation as the only variation operator.  To avoid tuning the mutation
-rate, we use the \emph{Single} mutation method described in~\cite{goldman:2013:cgpwaste},
-which mutates genes at random until exactly one active gene is changed.  
-Each run of each experiment used the same initial individual for all algorithms.
-All experiments were replicated 50 times, resulting in 50 different starting individuals per experiment.
-Having all algorithms start with the same initialization was done to reduce noise.
-Each
-run was given 10,000,000 evaluations before being terminated, and runs were stopped
-immediately on success.
-
-For all problems except \emph{Neutral} (Section~\ref{sec-neutral}), we provide a plot of the median
-evaluations until success using different problem sizes.  If the median run
-was unsuccessful, no information is plotted.  Note that both axes in these plots
-use a log scale.  In order to ensure a good sample of scalability we used 20
-evenly spaced problem sizes $\left [ 5 .. 100 \right ]$ at increments of 5.
-This information was then used to calculate each algorithms' order of complexity
-in regards to problem size, with results given in Table~\ref{table-growth}.
-A complete package of our results as well as the source code used in testing
-is available from our website.\footnote{\url{https://name.of.site.removed.com/anonymous-CGP}}
-
-\subsection{Neutral Exploration}
-\label{sec-neutral}
-
-\begin{figure}
-  \centering
-  \includegraphicsfit{neutral}
-  \caption{Genome sizes required to achieve different mode
-           numbers of active nodes on the \emph{Neutral} problem} 
-  \label{fig-neutral}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphicsfit{length_frequencies_100}
-  \caption{Frequency of evolving individuals of different active lengths
-           on the \emph{Neutral} problem} 
-  \label{fig-length}
-\end{figure}
-
-To examine how CGP and our variants bias the active length of individuals
-based on the maximum genome size, we first experiment using the \emph{Neutral} problem.
-Similar to the problems used in~\cite{miller:2001:bloat,collins:2006:haystacks},
-\emph{Neutral} has a completely flat landscape, such that all individuals
-evaluate to the same fitness.  In CGP this means parents are replaced each
-generation with a randomly chosen offspring. Thus there is no evolutionary pressure, only drift.
-This problem highlights what kinds of individuals each variant is biased toward
-creating, independent of the evolutionary pressure.
-In less complex representations, such as those in binary genetic
-algorithms, you would expect all representable
-search points to have an equal probability of being seen given enough evaluations.  As CGP's representation
-involves complex decoding, \emph{Neutral} can help examine the likelihood of searching
-different parts of the search space.  
-In our experiments, \emph{Neutral} has a single input and a single output,
-with a node arity of two.  Furthermore, the function set used is of size one, meaning
-\emph{Single} mutation must change an active node's connection gene or the output location
-in order to produce new offspring.
-
-\begin{table}
-	\centering
-	\begin{tabular}{|r|c|c|c|}
-	  \hline
-& \textbf{\emph{Normal}} & \textbf{\emph{Reorder}} & \textbf{\emph{DAG}} \\ \hline
-    \textbf{Mode} & 15      & 20      & 45 \\ \hline
- \textbf{95\% CI} & [5, 31] & [6, 44] & [10, 67] \\ \hline
-\textbf{<99.99\%} & 45      & 61      & 81 \\ \hline
- \textbf{Longest} & 59      & 76      & 92 \\ \hline
-	\end{tabular}
-	\caption{Frequency of different length individuals on \emph{Neutral}. 
-	         Includes the most common length (mode), 95\% confidence interval (CI),
-	         length that 99.99\% of individuals were below, and the longest individual found.}
-	\label{table-length}
-\end{table}
-
-We performed two types of experiments on the \emph{Neutral} problem.  The first
-tests how each variant's evolved number of active nodes changes
-with genome size.  The results provided in Figure~\ref{fig-neutral} can be used
-to predict how large of a genome size should be used
-given a user's rough estimate of the number of active nodes required by the problem.  
-This also assumes that search will perform best when the most commonly searched length
-is close to the optimal individual length.  Figure~\ref{fig-neutral} shows that \emph{Normal}
-required the largest genome sizes to achieve each mode number of active nodes.  In
-terms of scaling, \emph{Normal} needs a genome size of $O(n^{1.79})$ with
-$n$ the desired mode number of active nodes.  While \emph{Reorder} only reduced
-this to $O(n^{1.53})$, \emph{DAG} scaled much better at $O(n^{1.18})$.  This implies
-that as true problem size increases, we should expect \emph{Normal} to require nearly
-quadratically increasing genome sizes, while \emph{DAG} scales nearly linearly.
-
-The second experiment looked at the distribution of evolved lengths using a fixed
-genome size of 100. This
-information is shown in Figure~\ref{fig-length} as a probability density curve,
-with numeric results in Table~\ref{table-length}.
-The results for \emph{Normal} match our predictions in Section~\ref{sec-reorder};
-most individuals were highly inactive, with a mode number of
-active nodes of only 15.  The probability of generating an individual drops quickly
-as the number of active nodes deviates from the mode, with none of the 0.5 billion
-generated individuals having more than 59 active nodes, and only one in 10,000 above 45.  This is in contrast
-to the actual number of graphs of each size, which increases exponentially.
-This means that \emph{Normal} is biased toward small graphs even more than
-Figure~\ref{fig-length} indicates.
-\emph{Reorder} and \emph{DAG} show less of a bias, with \emph{DAG}'s mode of 45
-close to the prediction made in Section~\ref{sec-dag}.  The implications of these
-biases is that each is most likely to find solutions fastest when the number of
-active nodes desired by the problem best matches the algorithms search bias, as
-otherwise the algorithm will add artificial pressure to search different sized
-graphs.  The steepness of the bias reflects how much pressure will be applied.
-Combined with the length scaling results in Figure~\ref{fig-neutral}, we would
-predict \emph{Normal} to both require a larger genome size than \emph{Reorder} and \emph{DAG},
-as well as search a much smaller subset of the search space.
-
-%TODO Consider returning to this problem after having done \emph{Depth} and
-%\emph{Flat}, using the optimal versions of those two problems as the
-%starting individual.
-
-\subsection{Breadth}
-
-\begin{table}
-	\centering
-	\begin{tabular}{|r|c|c|c|}
-	  \hline
-& \textbf{\emph{Normal}} & \textbf{\emph{Reorder}} & \textbf{\emph{DAG}} \\ \hline
-\textbf{\emph{Breadth}} & 3.24 & 2.06 & 2.09 \\ \hline
-  \textbf{\emph{Depth}} & 7.72 & 3.66 & 4.17 \\ \hline
-   \textbf{\emph{Flat}} & 1.86 & 1.87 & 1.94 \\ \hline
-
-	\end{tabular}
-	\caption{$O(n^x)$ for each variant on each problem, where $n$ is problem size and $x$ is given in
-	         the table.}
-	\label{table-growth}
-\end{table}
-
-The \emph{Breadth} problem tests CGP's ability to evolve
-graphs that connect to all inputs using the minimum possible number of nodes.
-The landscape for this test is a problem with \emph{N} nodes which has $N+1$
-inputs and a single output.  The optimal structure is a tree with all nodes in the genome
-active and all inputs used exactly once as leaves.  To evaluate
-an individual, all nodes execute the binary \emph{OR} function,
-and possible program inputs are all strings of length $N+1$ with exactly one bit
-set.  As a result, the fitness of an individual is equal to the number of inputs
-connected to the output.  This problem is similar to the tree graph proposed
-in~\cite{payne:2009:bias}, except the problem scales such that all nodes
-must always be active, multiple inputs are used,
-and the fitness function involves individual execution instead of graph edit distance.
-
-Experimental results for the \emph{Breadth} problem are given in Figure~\ref{fig-breadth}.
-This shows that both \emph{Reorder} and \emph{DAG} scale better than \emph{Normal} when solving \emph{Breadth},
-but have similar behavior to each other.
-From Table~\ref{table-growth} we see that the number of evaluations \emph{Normal} requires
-to solve the \emph{Breadth} problem increases cubically while the other variants scale
-quadratically.  A likely cause of this growth discrepancy is that the solution to \emph{Breadth} requires
-all nodes in the genome to be active, which \emph{Normal} does with difficulty.
-The fact that \emph{Reorder} does not share this issue even though its
-search bias on \emph{Neutral} was relatively similar implies that \emph{Reorder} is better
-able to adapt to problem pressures than \emph{Normal}.  While \emph{DAG} scaled
-slightly worse than \emph{Reorder}, the difference is not significant.
-
-\begin{figure}
-  \centering
-  \includegraphicsfit{breadth}
-  \caption{Median evaluations until success for different sizes of the \emph{Breadth} problem} 
-  \label{fig-breadth}
-\end{figure}
-
-\subsection{Depth}
-The \emph{Depth} problem requires CGP to evolve graphs with the maximum possible node
-reuse.  In this problem, all nodes use the same function, which takes the minimum of
-their inputs, increments it by one, and returns that value.  The fitness function is
-equal to the value returned, and each individual is tested on a single input of zero.
-As a result, the maximum fitness is achieved when all nodes are connected in a chain
-such that all nodes are active and all connections in each node go to the previous node in the graph.
-Where \emph{Breadth} required no node reuse, \emph{Depth} requires the maximum amount of node
-reuse.  Also in contrast to \emph{Breadth}, \emph{Depth} has a unique solution for
-\emph{Normal} and \emph{Reorder}, with \emph{DAG} having multiple ways of representing
-the same graph by encoding nodes in different orders.  This problem is similar to the linear graph proposed
-in~\cite{payne:2009:bias}, except the problem scales such that all nodes
-must always be active and the fitness function involves individual execution instead of graph edit distance.
-
-Figure~\ref{fig-depth} and Table~\ref{table-growth} both show how \emph{Normal}
-is effectively unable to solve this problem.  \emph{Normal} was only able to solve
-the smallest three problem sizes of the twenty tried, with an experimental growth
-rate of $O(n^{7.72})$.  By comparison, \emph{Reorder} and \emph{DAG} solved all
-or almost all, respectively, of the problems tried, with far lower growth rates.
-As \emph{Breadth} also required all nodes to be active, the reason for \emph{Normal}'s
-breakdown must be something else about \emph{Depth}.  A likely explanation is that
-\emph{Normal} requires structures to be evolved in a precise location in the genome,
-in that a node with all connections set to $i$ is only part of the optimal solution
-if that node is at position $i+1$.  If it is not, it must be reinvented at $i+1$ to
-move to optimality.  This is especially problematic in the following case.  Consider
-a fully optimal individual, except that $i+1$ is inactive and $i+2$ is active and
-all of its connections go to $i$.  The only way to improve this individual is to
-set all of $i+1$'s connections to $i$ and change all of $i+2$'s connections to $i+1$.
-What makes this difficult is the absence of fitness gradient, meaning all changes can
-only happen by drift and that changes to $i+2$ can only happen after $i+1$ is relatively
-correct.  Otherwise the individual will decrease in fitness.  This problem is further
-exacerbated by \emph{Normal}'s mutation method, which on average connects $i$ to $\frac{i}{2}$,
-meaning high index nodes are rarely connected to adjacent indexes.
-\emph{Reorder} avoids this problem by allowing structures to be moved to where they are
-needed, while simultaneously decreasing the length of connections.  \emph{DAG} has
-no concept of position, and therefore cannot have this problem.
-
-While \emph{Reorder} and \emph{DAG} have very similar behaviors in Figure~\ref{fig-depth},
-Table~\ref{table-growth} shows that \emph{Reorder} scaled slightly better.  A likely
-cause of this is that \emph{Reorder} has a higher probability of making nodes
-with uniform connections as compared to \emph{DAG}.  This is true
-because nodes with no transitive dependency on the mutating node
-and can still sometimes be unreachable by mutation in \emph{Reorder}.  Conversely, nodes
-that the mutating node is transitively dependent on will always be reachable.  In \emph{DAG}
-there is no bias toward connecting to already transitively dependent nodes.
-As a result, \emph{Reorder} is able to leverage this
-search bias to better make nodes with uniform connection genes, improving search speed.
-
-\begin{figure}
-  \centering
-  \includegraphicsfit{depth}
-  \caption{Median evaluations until success for different sizes of the \emph{Depth} problem} 
-  \label{fig-depth}
-\end{figure}
-
-\subsection{Flat}
-As a counter point to \emph{Breadth} and \emph{Depth}, which require entirely
-active genomes, we propose \emph{Flat}.  Unlike the other problems, \emph{Flat}
-evaluates an individual's genome directly, with the fitness of
-an individual equal to the number of connections made directly to the input.
-In this case the optimal individual is one where all nodes and the output location
-are connected to the only input location.  This tests how well CGP is able to
-create genomes with the least amount of internal structure.
-Similar to \emph{Depth}, \emph{Flat} has exactly one solution. 
-In~\cite{payne:2009:bias} some evidence was found that solutions
-that appear less often in the search space may be harder to find.  If this is the reason
-\emph{Depth} was more difficult than \emph{Breadth}, then \emph{Flat} should also
-be difficult to solve.  As all nodes now
-participate in evaluation regardless of connections, they are all considered active
-when performing \emph{Single} mutation.
-%Note that a more standard fitness function
-%cannot be used as an inverse to \emph{Breadth} and \emph{Depth} as all three methods
-%can equally make fully inactive individuals by directly connecting the output
-%to the input at a probability of $\frac{1}{N+1}$.
-
-On this problem, all three algorithms have approximately the same overall shape, shown
-in Figure~\ref{fig-flat}, and have similar complexities, shown in Table~\ref{table-growth}.
-\emph{DAG} does appear to be slower than the other two, although the difference may
-just be by a constant scalar.  This makes sense as the probability of a connection
-mutating to an input is $\frac{1}{N}$ for \emph{DAG} and $\frac{1}{i}$ for
-\emph{Normal} and \emph{Reorder}, with $i\leq N$.  In \emph{DAG}, if there are nodes with
-transitive dependencies on the mutating node, then the probability
-of connecting to the input increases.  However, it can never surpass that for \emph{Normal}
-and \emph{Reorder}.  More generally, these results support the idea that \emph{Reorder}
-and \emph{DAG} do not have a harder time making highly unconnected structures
-than \emph{Normal}, and that solution rarity in the search space is not why \emph{Normal}
-did so poorly on \emph{Depth}.
-
-\begin{figure}
-  \centering
-  \includegraphicsfit{flat}
-  \caption{Median evaluations until success for different sizes of the \emph{Flat} problem} 
-  \label{fig-flat}
-\end{figure}
-
-\section{Benchmarking}
-\label{sec-benchmarking}
-Understanding length biases and search limitations allows users to
-solve new search problems easier.  In order to examine how our predicted search biases
-interact with standard CGP landscapes, we tested all three algorithms on the 3-bit
-multiplier problem.  In this problem there are six inputs and six outputs with
-a function set of \{AND, OR, NAND, NOR\}.  Instead of scaling problem difficulty
-as in Section~\ref{sec-tailored}, each algorithm was tested with scaling genome
-sizes.  We used the range $\left [ 10 .. 8000 \right ]$, taking 21 points on an
-exponential scale.  Other run parameters were kept the same.
-The median number of evaluations to success is shown in
-Figure~\ref{fig-multiply} on a log-log scale.  The results from the best genome
-size are given in Table~\ref{table-multiply}.
-
-As predicted, both the graphic and tabular results show that \emph{Reorder} and
-\emph{DAG} can perform effective search with much smaller genome sizes than \emph{Normal}:
-The best tested genome size for both novel
-variants had lower median evaluations to success than \emph{Normal}'s best, which
-also required a genome twenty times larger.  \emph{Reorder} was statistically
-better than \emph{Normal} under the single tailed Mann-Whitney U test. While \emph{DAG} was not
-statistically better than \emph{Normal}, the likely reason is \emph{Normal}'s exceptionally large median
-absolute deviation.  We believe this deviation is caused by \emph{Normal}'s occasional need to
-reinvent useful structures in different genome locations.  In the unlikely event that
-active nodes are evolved with the right amount of inactive nodes between them, and in the
-right order relative to each other, \emph{Normal} is likely going to receive far better
-results.  In cases were these properties do not hold, \emph{Normal} will
-have to recreate nodes in new locations, often with no fitness gradient, significantly
-increasing run time.  As the number of nodes in the genome increases, the probability
-that inactive nodes exist between two active nodes increases.  This may be part of
-the reason \emph{Normal} requires such a large genome size to perform well.
-Also, more inactive space in the genome may facilitate structure recreation in new locations.
-As a final advantage to \emph{Reorder} and \emph{DAG}, we see that both evolved
-solutions with half as many median active nodes as \emph{Normal}.  A potential
-reason for this is \emph{Normal}'s need to leverage introns in order to perform
-neutral search more than the other methods.
-
-\begin{table}
-	\centering
-	\begin{tabular}{|c|c|c|c|}
-	  \hline
-& \textbf{\emph{Normal}} & \textbf{\emph{Reorder}} & \textbf{\emph{DAG}} \\ \hline
-\textbf{Genome} & 4000   & 200     & 200 \\ \hline
-  \textbf{MES} & 272,324 & 213,660 & 249,031 \\ \hline
-  \textbf{MAD} & 125,997 & 79,335  & 86,551 \\ \hline
-  \textbf{P value} & N/A & 0.030   & 0.117 \\ \hline
-  \textbf{Active} & 255  & 118     & 120 \\ \hline
-
-	\end{tabular}
-	\caption{Best configuration of each algorithm on 3-bit Multiplier.
-	Includes maximum genome size, Median Evaluations to Success (MES), Median Absolute Deviation (MAD)
-	one tailed P value comparison with \emph{Normal}, and median number of active
-	nodes in the final solutions found.}
-	\label{table-multiply}
-\end{table}
-
-\emph{Reorder} and \emph{DAG} are not without drawbacks.  In Figure~\ref{fig-multiply} we see that
-\emph{Normal} may underperform at peak performance and at lower genome sizes, but
-as the genome size increases it surpasses the other methods.  \emph{DAG} is especially
-sensitive to increasing genome sizes, performing worse than \emph{Normal} on sizes above 600.
-\emph{Reorder} also falls behind \emph{Normal} once sizes exceed 2000.  This
-could indicate that the novel methods are more sensitive to genome size.
-While this is true, we believe it to be less of a flaw and more of an unavoidable side
-effect of these techniques being able to evolve more complex graphs than \emph{Normal}.
-%TODO Consider expanding here.
-
-\begin{figure}
-  \centering
-  \includegraphicsfit{multiply}
-  \caption{Median evaluations until success on the Multiply problem using different genome sizes} 
-  \label{fig-multiply}
-\end{figure}
-
-\section{Conclusions and Future Work}
-\label{sec-conclusion}
-Existing literature suggests CGP performs best when maximum genome sizes
-far exceed actual solution size~\cite{miller:2006:redundancy}, supported by
-evidence that inactive genes improve neutral search~\cite{yu:2001:neutrality},
-and that CGP doesn't suffer from bloat~\cite{miller:2001:bloat}.  We, however, find evidence
-that the need for highly inactive genomes is likely due to limitations in CGP's
-search mechanisms.  When no selective pressure on size exists, CGP highly favors
-short individuals, with this search bias continuing even when the problem structure
-calls for longer individuals.  Furthermore, when combined with CGP's use of node
-positioning, problems requiring significant proportions of the genotype to be
-active were found to be difficult or impossible for CGP to solve.  We showed theoretical
-and experimental evidence that the blame for this problem lies with the way CGP's mutation
-interacts with its representation, as existing mechanisms impose artificial search limitations
-not inherent to the search problem.  \emph{Reorder}, which shuffled node positions
-without changing node expression, and \emph{DAG}, which removed all concept of node
-position, were both able to avoid these limitations and more fully utilize
-genome space.  As these methods were able to achieve better results with less
-inactive genes, we conclude that while inactive genes are useful, the number required
-by \emph{Normal} CGP is a result of the inherent length bias and search limitations in its
-representation and evolutionary mechanisms, and not because that many inactive
-nodes are needed for neutral search.  Similarly, CGP's resistance to bloat is likely
-no more than the work of a previously unexamined form of parsimony pressure, the strength
-of which is determined by the maximum number of nodes.  The drawback of \emph{Reorder} and
-\emph{DAG} is that in removing this parsimony pressure, they are more able to utilize
-all genes in the genome, making them more sensitive to the genome size.  Yet this
-may prove to be an advantage, as it may allow CGP to solve more complex problems without
-explosively increasing the genome size.
-
-Further analysis of \emph{Reorder} and \emph{DAG} should be performed in light
-of the evidence found here.  Both appear to be valid ways of making CGP more flexible
-and powerful, but wider testing needs to be performed to solidify that claim.  As
-an important aspect of those tests, run time comparisons need to be performed, as
-both increase the computational expense of CGP, with \emph{DAG} especially having
-the potential for significant overhead.  It is also possible either could benefit
-from further modification, such as using different initialization or method for mutation.
-
-More analysis on the tailored problems will likely improve the clarity of results.  Specifically
-\emph{Neutral} would likely benefit from further scalability testing to see how search frequencies change
-with respect to genome size, especially at larger genome sizes.  Tests with \emph{Breadth} and \emph{Depth} to see how many inactive
-nodes are needed before \emph{Normal} surpasses \emph{Reorder} and \emph{DAG}, if ever, may
-also help highlight exactly what proportion of inactive nodes \emph{Normal} needs to solve
-problems.
+TODO~\cite{goldman:2012:ltga}.
 

### 2013-09-11 Parameterless Population Pyramid/main.tex
Outline of P3 paper, some citations

@@
 {\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}
+
+\usepackage{bm}
 \begin{document}
@@
 \begin{abstract}
-
+TODO
 \end{abstract}
@@
 \section{Introduction}
-TODO~\cite{goldman:2012:ltga}.
+TODO Parameterless optimization, model building, etc.
+
+\section{The P3 Algorithm}
+TODO Overview.  Use graphic.
+
+\subsection{Hill Climber}
+TODO Describe First Improvement Hill Climber, including waste prevention, etc.
+
+\subsection{Pyramid}
+TODO Explain population levels, entropy tables $O(n^2)$, uniqueness
+
+\subsection{Crossover}
+TODO Explain how entropy becomes clusters efficiently $O(n^2)$, cluster ordering,
+keep changes that don't reduce fitness, removal of zero entropy sub-clusters,
+keep donating until a change is made.
+
+\subsection{All Together}
+TODO Pieces working together.  HC removes pairwise noise, which allows crossover to detect
+clusters.  Levels allow realization of linkage at high levels, reduction in noise.
+
+\section{Comparison Algorithms}
+
+\subsection{Random Restart First Improvement Hill Climber}
+TODO This is important to have as it shows how the pyramid is able to reach optima
+the hill climber cannot.  Need a citation.
+
+\subsection{Linkage Tree Genetic Algorithm}
+TODO Discuss LTGA~\cite{thierens:2013:ltgahiff}, how it differs from P3, its limitation based on
+initialization, requirement of population size.  Mention how some
+of the ideas in P3 were tried in LTGA and seemed to make improvements~\cite{goldman:2012:ltga}.
+Discuss bisection method and Rule of Three~\cite{jovanovic:1997:ruleofthree}.
+Remember to say that the paper argues for $\frac{3}{N+1}$, so 100 trails is strictly
+less than $3\%$ expected failure rate.
+
+\subsection{\bm{$(1+(\lambda,\lambda))$}}
+TODO Explain briefly how it works, that it is the best theory supported crossover method~\cite{doerr:2013:lambdalambda}.
+Explain changes including: Keep best mutant if no worse than crossover offspring,
+prevent evaluations if crossover offspring identical to either parent, if tied fitness
+in crossover offspring take maximum hamming distance to parent.  If $\lambda \ge N$
+restart with random individual.  Does not use ``mod'' version as restarting is preferable
+when stuck on plateaus.
+
+\section{Test Problems}
+
+\subsection{Single Instance Problems}
+TODO Deceptive Trap, Deceptive Step Trap~\cite{goldman:2012:ltga}, Hierarchical If
+and Only If, Discretized Rastrigin.
+
+\subsection{Randomly Generated Problem Classes}
+TODO Nearest Neighbor NK with wrapping neighborhoods, Ising Spin Glasses on toroidal
+grids with only -1, 1 spins, Maximum Satisfiability constructed with known global
+optimum.
+
+\section{Experimental Results}
+TODO Cover all 7 problems as a group, focusing on the two ``types'' of problems.
+Make it explicit that you outperform by a constant factor on the static problems
+and by an order of complexity on generated problems.
+
+TODO Include graphic for 7 problem results.
+
+TODO Include table comparing results on largest version of each problem.
+
+\section{Promising Theory}
+TODO Easy proof on unimodal (no plateaus) and all functions of unitation due to HC.
+Sketch of how to prove runtime on Deceptive Trap, HIFF, suggestions toward all
+possible non-overlapping functions.
+
+\section{Conclusions and Future Work}
+TODO No parameters required, beats tuned LTGA.  Good at solving hard classes
+of problems, not just single instances of a problem as it scales to problem difficulty.
+Future work in theory.  Also readily applicable to real world problems as no
+problem information is required.
 

### 2013-09-17 Parameterless Population Pyramid/main.tex
Added a diagram, more citation updates

@@
 
+TODO Talk about ALPS~\cite{hornby:2006:alps}.
+
 \section{The P3 Algorithm}
-TODO Overview.  Use graphic.
+TODO Overview.  Use Figure~\ref{fig-p3}.
+
+\begin{figure}
+  \centering
+  \includegraphicsfit{P3_big_fail}
+  \caption{Example}
+  \label{fig-p3}
+\end{figure}
 
@@
 \subsection{Crossover}
-TODO Explain how entropy becomes clusters efficiently $O(n^2)$, cluster ordering,
+TODO Explain how entropy becomes clusters efficiently $O(n^2)$~\cite{gronau:2007:upgma}, cluster ordering,
 keep changes that don't reduce fitness, removal of zero entropy sub-clusters,
@@
 TODO Nearest Neighbor NK with wrapping neighborhoods, Ising Spin Glasses on toroidal
-grids with only -1, 1 spins, Maximum Satisfiability constructed with known global
+grids with only -1, 1 spins~\cite{saul:1994:spinglass}.
+Used\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}, Maximum Satisfiability constructed with known global
 optimum.
@@
 and by an order of complexity on generated problems.
+
+TODO Mention 1.5 trillion evaluations and 150,000 runs (average 10 million evaluations per run).
 

### 2013-09-20 Parameterless Population Pyramid/main.tex
More references, fixed bottom line in example diagram

@@
 
-\keywords{TODO}
+\keywords{Linkage Learning, Local Search, TODO}
 
@@
 TODO Talk about ALPS~\cite{hornby:2006:alps}.
+
+TODO Talk about LO-LIMD~\cite{posik:2011:parameterless} and how it was parameter-less,
+but is incapable of handling overlapping linkages.
 
@@
 of the ideas in P3 were tried in LTGA and seemed to make improvements~\cite{goldman:2012:ltga}.
+Tests of adding hillclimber to LTGA argued no improvement~\cite{bosman:2011:lsbbo}.
 Discuss bisection method and Rule of Three~\cite{jovanovic:1997:ruleofthree}.
@@
 \subsection{Randomly Generated Problem Classes}
-TODO Nearest Neighbor NK with wrapping neighborhoods, Ising Spin Glasses on toroidal
+TODO Nearest Neighbor NK with wrapping neighborhoods~\cite{wright:2000:solvingnk}, Ising Spin Glasses on toroidal
 grids with only -1, 1 spins~\cite{saul:1994:spinglass}.

### 2013-09-24 Parameterless Population Pyramid/main.tex
More fill in on P3

@@
 \section{The P3 Algorithm}
-TODO Overview.  Use Figure~\ref{fig-p3}.
+\label{sec-p3}
+Before going into detail about how each aspect of P3 works, consider the example
+evolution presented in Figure~\ref{fig-p3}.  At the start of optimization there
+is no known problem information, so solution \#1 is randomly generated.  This
+is then brought to a local optima using a hill climber (Section~\ref{sec-hillclimber})
+creating solution \#2, which is added to the pyramid (Section~\ref{sec-pyramid}) level
+$P_0$.  As there is no further exploitable information, \#3 is randomly generated
+and optimized to create \#4, which is added to $P_0$.  \#2 and \#4 are used to
+define a crossover (Section~\ref{sec-crossover}), which improves \#4 to create
+\#5. As \#5 used information from $P_0$ is it added to the next level of the pyramid, $P_1$.
+Again P3 lacks information to continue, so \#6 is randomly generated and improved into
+\#7.  Unlike \#2 and \#4, \#7 is not added in this example because it happens to be identical to
+a solution already in the pyramid.  Even though \#7 is not unique, it still crosses with
+$P_0$, which does make a unique improvement \#8 that is added to $P_1$ and mixed with that
+level to create \#9.  As before, the process creates \#10, \#11, and \#12 using
+random generation, hill climbing, and crossover, respectively.  In this example \#13
+is not added to $P_2$ because it not a strict improvement in fitness over \#12.  There
+is not enough information in $P_2$ to improve \#13 further, so the process would begin
+yet again with a random solution.
 
@@
   \includegraphicsfit{P3_big_fail}
-  \caption{Example}
+  \caption{Example of the first steps of P3's optimization described in Section~\ref{sec-p3}.
+  \#7 is not added because it is a duplicate of a solution already in the pyramid.  \#13 is
+  not added because it is not a fitness improvement over \#12.}
   \label{fig-p3}
@@
 \subsection{Hill Climber}
-TODO Describe First Improvement Hill Climber, including waste prevention, etc.
+\label{sec-hillclimber}
+The terminology used to describe different methods of making minor modifications
+to a solution, evaluating the quality of those changes, and iteratively constructing
+new solutions, is full of duplicated, ambiguously defined terms.  As such we shall
+describe what is likely an existing algorithm and give it a descriptive name that
+may not match all of its previous incarnations.
+
+For our purposes, we use the First Improvement Hill Climber (FIHC), a method of exploiting
+local information through a series of modify-evaluate-accept iterations, to start
+from randomly initialized solutions and end with single change local optimal results.
+FIHC uses the following algorithm.  Consider all gene loci $x_i$ in a random order.
+For a given $x_i$ try all possible alternative values for the gene $x_i$, evaluating
+solution quality with each.  If any alternative value is a strict fitness improvement
+over the current value for $x_i$, replace $x_i$ with the improved value.  If no
+change is made to the solution, stop, otherwise iterate.
+
+To avoid wasting evaluations, FIHC keeps track of all loci tested since the last
+change was made to the solution.  If the algorithm returns to the same loci without
+any changes, it moves on to the next loci without performing any evaluations.
+
+FIHC solves all functions of unitation in $O(N)$ time.  This is in contrast to
+methods like Steepest Ascent Hill Climber~\cite{bosman:2011:lsbbo, goldman:2012:ltga}
+which require $O(N^2)$ as changes are only accepted after all neighbor solutions
+are evaluated.  In general FIHC is designed to find optima using a minimum number
+of evaluations.  Furthermore, because FIHC tests loci in a random order it does
+not cause any positional bias to the search.
 
 \subsection{Pyramid}
-TODO Explain population levels, entropy tables $O(n^2)$, uniqueness
+\label{sec-pyramid}
+Unlike many EC techniques, P3 does not have a single population of solutions.
+Instead to maintains a pyramid like structure, such that each level of the pyramid
+has a population of solutions, with each level representing something akin to different
+generations of evolution.  The exact relationship between the levels is described
+in Section~\ref{sec-alltogether}.
+
+The levels represent independent sets of unique solutions, such that
+$\forall_{i, j \in P, i \neq j} P_i \cap P_j = \emptyset$.  This relationship
+is maintained constructively, such that new solutions can only be added to a level
+if they do not exist in any level of the pyramid.  This uniqueness check can be
+achieved in constant time by maintaining a hash of all solutions in the pyramid.
+
+Along with storing the solutions, each level also maintains a table of pairwise
+gene value frequencies.  This is also built constructively, such that initially
+all frequencies are set to 0.  Each time a solution is added to a level, all
+pairs of loci are queried for their gene values.  The table is then updated,
+incrementing the count for each observed pairing.  This requires $O(N^2)$ operations
+(no evaluations) each time a solution is added to a level of the pyramid, but is
+constant in the number of solutions currently in the level.
 
 \subsection{Crossover}
-TODO Explain how entropy becomes clusters efficiently $O(n^2)$~\cite{gronau:2007:upgma}, cluster ordering,
-keep changes that don't reduce fitness, removal of zero entropy sub-clusters,
-keep donating until a change is made.
+\label{sec-crossover}
+P3 uses a crossover method derived from that defined for LTGA~\cite{thierens:2013:ltgahiff}
+(Section~\ref{sec-ltga}),
+in that it uses gene value entropy to construct clusters of genes that should have their values
+conserved during crossover.  There are two primary steps in
+defining this type of crossover: cluster creation and cluster usage.
+
+During cluster creation, P3 attempts to create a binary tree of clusters such that the leaves
+represent individual gene loci and each internal node represent the subset of loci created
+by joining the subsets encoded in its children.  These clusters represent detected linked
+loci in the genome, and are used when performing crossover.
+Using the table of observed frequencies described in Section~\ref{sec-pyramid}, Equation~\ref{eq-distance}
+efficiently estimates the linkage between clusters of loci using the average pairwise
+entropy, as given in Equation~\ref{eq-entropy}.
+\begin{equation}
+  D(C_i,C_j) = \frac{1}{\left | C_i \right |\cdot \left |C_j \right|}\sum_{c_i \in C_i}\sum_{c_j \in C_j} 
+  2 - \frac{H(C_i) + H(C_j)}{H(C_i \cup C_j)}
+  \label{eq-distance}
+\end{equation}
+\begin{equation}
+  H(C) = -\sum_{s\in S} p_c(s)\log(p_c(s))
+  \label{eq-entropy}
+\end{equation}
+
+Starting from the leaves, such that each loci appears exactly once, the binary tree
+can be constructed using an agglomerative method.  Each iteration determine the
+pairwise linkage between each cluster of loci using Equation~\ref{eq-distance},
+and merge the two closest into a new cluster.  The process ends when there is only
+a single cluster containing all loci.  As our clustering algorithm provides the proper
+criteria this algorithm requires $O(N^2)$ total time to correctly construct the linkage tree
+from the frequency table~\cite{gronau:2007:upgma}.
+
+Not all of the clusters in the linkage tree are worth using during crossover.  The most
+obvious is a cluster containing all the loci.  If we performed a crossover such that all
+of the loci must remain together, no new individual would be created.  We further remove
+clusters that are completely contained within some other cluster, such that if
+$D(C_i,C_j) = 0$, $C_i$ and $C_j$ are not used when performing crossover.  $D(C_i,C_j)$ only
+returns $0$ when the loci in $C_i$ perfectly predict the bits in $C_j$ and vice versa,
+and from this we conclude that the settings for those two loci should never be disrupted.
+When used on particularly small populations this has the potential to greatly reduce the number
+of clusters used during crossover.  For instance, if $|P_i| = 1$ no clusters exist, and if $|P_i| = 2$
+exactly two clusters exist: one where the solutions agree and one where they disagree.
+
+Given the useful set of clusters, we can now define how they are used.  P3 uses
+crossover to recombine known information in a level of the pyramid with a candidate
+solution.  For each cluster, do the following.  Select a random donor from the population.
+Copy the donor's gene values for all loci in the cluster into the candidate solution.
+Evaluate the candidate.  If the fitness of the candidate has been reduced, revert the candidate's
+gene values.  When choosing the donor P3 ensures there is at least one gene value
+different between the donor and the candidate solution.  This can potentially require
+up to $O(|P_i|)$, but is on average more efficient than allowing donations that
+do not change any gene values.
+
+To complete a single crossover event, all of the useful clusters are tested, each
+requiring at most a single evaluation.  Clusters are tested in smallest first order,
+with clusters of equal size ordered randomly.  This ensures smaller building blocks
+are optimized before large changes are made to the genome, which helps preserve
+diversity.  As cluster building and application ordering are not dependent on
+the position of loci in the genome, there is no positional bias in the search points
+evaluated.
 
 \subsection{All Together}
+\label{sec-alltogether}
 TODO Pieces working together.  HC removes pairwise noise, which allows crossover to detect
@@
 \subsection{Random Restart First Improvement Hill Climber}
-TODO This is important to have as it shows how the pyramid is able to reach optima
-the hill climber cannot.  Need a citation.
+TODO Need a citation either here or Section~\ref{sec-hillclimber}.
+
+As P3 utilizes a hill climber to perform optimization, it is necessary to show P3's
+added complexity is able to outperform using the hill climber alone.  Therefore
+as the first comparison we define a complete optimization method based of off the
+First Improvement Hill Climber defined in Section~\ref{sec-hillclimber}.  This definition
+is extended to match P3's style of restarting. Whenever the hill climber
+obtains a solution that cannot be improved using a single bit flip, a new solution
+is randomly generated and optimized.
+
+If P3's quality relies solely on starting at random points and applying hill climber,
+this comparison algorithm should perform even better than P3.  Similar to P3 this
+algorithm is parameterless, makes no assumptions about gene ordering, and will
+find the global optima given a sufficient (potentially exponential) number of evaluations.
 
 \subsection{Linkage Tree Genetic Algorithm}
-TODO Discuss LTGA~\cite{thierens:2013:ltgahiff}, how it differs from P3, its limitation based on
-initialization, requirement of population size.  Mention how some
-of the ideas in P3 were tried in LTGA and seemed to make improvements~\cite{goldman:2012:ltga}.
-Tests of adding hillclimber to LTGA argued no improvement~\cite{bosman:2011:lsbbo}.
-Discuss bisection method and Rule of Three~\cite{jovanovic:1997:ruleofthree}.
-Remember to say that the paper argues for $\frac{3}{N+1}$, so 100 trails is strictly
-less than $3\%$ expected failure rate.
+\label{sec-ltga}
+The closest relative to P3 in existing literature is the Linkage Tree Genetic Algorithm
+(LTGA).  There have been a number of different variations proposed to LTGA since
+its original publication~\cite{thierens:2010:ltga}, so for comparison we have chosen
+to use the variant described in the most recent publication~\cite{thierens:2013:ltgahiff}.
+This algorithm is also the current state of the art, outperforming other algorithms
+in black box optimization across numerous benchmarks.
+
+LTGA works by iteratively improving a population of solutions in a generational manner.
+Each generation the half of the solutions (chosen using standard tournament selection)
+are used to construct a linkage tree just as described in Section~\ref{sec-crossover}.
+Each solution in the population is then crossed with the entire population using that
+linkage tree, with the results put into the next generation.  This process continues
+until the population fails to change between generations, signaling convergence.
+
+Unlike P3, LTGA does not make use of a hill climber, and there is evidence that
+doing so results in decreased efficiency~\cite{bosman:2011:lsbbo}.  LTGA applies
+clusters in least linked first order whereas P3 uses smallest first.  Also, LTGA
+does not search for donors until one containing different genes is found.  Instead
+if the donated genes are identical, it simply skips the evaluation.  LTGA
+requires a population size parameter, maintains a fixed population size across
+generations, and does not prevent duplicate individuals in a population.  LTGA
+does not add new genetic information after initialization, meaning that if
+required diversity is not contained in the original population or lost in future
+generations, LTGA will fail to find the global optimum.
+
+In order to set the population size parameter to solve a class of problems, we use
+the bisection method~\cite{goldman:2012:ltga}, which determines the minimum population
+size required to meet a specified success criteria.  As P3 will run until the global optima
+is found or the evaluation limit is reached, we want to ensure LTGA has a similar requirement
+on its success.  Therefore using the Rule of Three~\cite{jovanovic:1997:ruleofthree} we
+can provide a bound on the probability of failure $\frac{3}{k+1}$ with $95\%$ confidence,
+where $k$ is a number of runs without a failure.  Here we state that a population size
+is successful if it performs 100 runs without failing to find the global optimum, meaning
+the probability of that population failing to solve problems of the same class is bounded above by
+$3\%$.  Note that these 100 runs are considered training, and the results from the bisection
+runs are not included during the comparative testing.  Furthermore, when testing on problem
+classes, bisection is performed on a different set of instances than testing.
+
+Other variants of LTGA have more closely resembled P3~\cite{goldman:2012:ltga},
+but as those variants are not in active use, we did not use them for comparison.
+However, no LTGA variant has been able to remove the population size parameter
+or the problem of premature convergence.
 
 \subsection{\bm{$(1+(\lambda,\lambda))$}}
-TODO Explain briefly how it works, that it is the best theory supported crossover method~\cite{doerr:2013:lambdalambda}.
-Explain changes including: Keep best mutant if no worse than crossover offspring,
-prevent evaluations if crossover offspring identical to either parent, if tied fitness
-in crossover offspring take maximum hamming distance to parent.  If $\lambda \ge N$
-restart with random individual.  Does not use ``mod'' version as restarting is preferable
-when stuck on plateaus.
+As a final comparison we chose the $(1+(\lambda,\lambda))$ algorithm~\cite{doerr:2013:lambdalambda}
+which is currently the best theory supported crossover method.  Unlike P3 and LTGA,
+$(1+(\lambda,\lambda))$ does not use a population of solutions or any model building
+to determine problem variable linkage.  Instead it uses mutation to produce $\lambda$
+offspring, selecting the best to recombine with the original parent to produce $\lambda$
+more offspring using uniform crossover.  In the original work a method for controlling
+the value of $\lambda$ is provided based on offspring success, meaning there are no
+parameters to be set in this algorithm.
+
+In order to make $(1+(\lambda,\lambda))$ a viable method of optimizing problems
+with multiple local optima, we modified the original algorithm.  Primarily, if
+during search $\lambda \ge N$, search is restarted from a random individual with
+$\lambda=1$.  This is done because when $\lambda \ge N$ the mutation rate is greater
+than or equal to $100\%$, nullifying search.  Furthermore, this point is only reached
+when the algorithm has stalled for a significant number of generations.
+
+We also made minor modifications to selection and to prevent wasting evaluations.
+If the best offspring produced in a generation is a mutant offspring, select that
+over any of those produced using crossover.  If there is a fitness tie between the
+best offspring in a generation, select the one with the maximum hamming distance
+to the parent.  This encourages drift over plateaus.  While the original paper
+discusses a ``mod'' version of the $\lambda$ control strategy for how to handle
+when offspring of equal fitness are produced, we found this conflicted with our
+method of restarting.  As such we use the original control strategy, where $\lambda$
+is increased if the best offspring in a generation is not strictly better than its parent.
 

### 2013-10-28 Parameterless Population Pyramid/main.tex
dos2unix to remove misterious windows line endings

@@
 of clusters used during crossover.  For instance, if $|P_i| = 1$ no clusters exist, and if $|P_i| = 2$
-exactly two clusters exist: one where the solutions agree and one where they disagree.
+at most two clusters exist: one where the solutions agree and one where they disagree.
 
@@
 \section{Test Problems}
+TODO Mention that since all of our algorithms are variable order insensitive
+and have no bias toward creating 1s over 0s, all will have identical results if
+problem variables are randomly reordered or XOR'd.
+
+TODO Chose only problems with known global optima to allow comparison on time to
+global optima.
 
 \subsection{Single Instance Problems}
-TODO Deceptive Trap, Deceptive Step Trap~\cite{goldman:2012:ltga}, Hierarchical If
-and Only If, Discretized Rastrigin.
+Well defined test problem landscapes can help when performing exact analysis of
+algorithms.  These problems can be thought of as testing the edge cases of optimizers,
+presenting them with unnatural but interesting problem features.
+
+The classic example for this kind of test problem is the Deceptive Trap problem.
+For this benchmark the genome is broken up into non-overlapping problems of $k$
+genes, with the fitness of a single ``trap'' given by Equation~\ref{eq-deceptivetrap}.
+\begin{equation}
+   trap(t) = \left\{
+     \begin{array}{rl}
+       k-1-t, &  t<k\\
+       k,   &  t = k
+     \end{array}
+   \right.
+  \label{eq-deceptivetrap}
+\end{equation}
+In this equation $t$ represents the number of bits set in the trap.  For each trap
+there is one local optima (all bits set to 0) and one global optima (all bits set to 1).
+This problem requires exponential time to solve unless there is some method to effectively
+move complete traps between solutions, making this an excellent crossover test problem.
+
+However, when using a hill climber and linkage learning, the Deceptive Trap problem
+may be trivially solved~\cite{goldman:2012:ltga}.  As such the Deceptive Step Trap
+problem was defined, given in Equation~\ref{eq-deceptivesteptrap}.
+\begin{equation}
+   step\_trap(t) = \left \lfloor \frac{(k-s)\pmod{s} + trap(t)}{s} \right \rfloor
+  \label{eq-deceptivesteptrap}
+\end{equation}
+This problem leverages the original Deceptive Trap problem definition, with the
+addition of fitness plateaus of configurable size $s$.  This creates an enormous
+number of local optima and makes detection of linkages much more challenging.
+
+While both of these problems have difficult to solve subproblems, there non-overlapping
+nature and single optimal setting may not capture aspects of real world problem solving.
+Therefore the Hierarchical If and only If (HIFF) problem is often used to test an
+algorithms ability to handle these characteristics~\cite{thierens:2013:ltgahiff}.
+In this problem the genome is broken up hierarchically into a complete binary tree,
+with each gene representing a single bit, and each internal node representing the union
+between all genes in that nodes subtree.  A node in the tree contributes to the fitness
+only if all of the genes in its subset have the same value (IE are all 0 or are all 1).
+The amount contributed is equal to the number of bits in the subset.  As a result
+this problem involves creating larger and larger blocks of single value runs which
+only score if the entire block is optimized.
+
+Our final single instance problem is borrowed from the real valued optimization domain.
+The Discretized Rastrigin used the standard Rastrigin evaluation function (Equation~\ref{eq-rast}).
+\begin{equation}
+  An + \sum_{i=1}^{n}\left [ x_i^2-A\cos (2\pi x_i) \right ] \forall x\in [-5.12,5.12]
+  \label{eq-rast}
+\end{equation}
+This function creates a highly multimodal landscape which is completely separable
+and has a single global optima.  In order to apply our algorithms to this complex problem,
+we discretized the representation using 10 bit gray coded numbers evenly distributed
+on the range $[-5.12,5.12)$.
 
 \subsection{Randomly Generated Problem Classes}
-TODO Nearest Neighbor NK with wrapping neighborhoods~\cite{wright:2000:solvingnk}, Ising Spin Glasses on toroidal
-grids with only -1, 1 spins~\cite{saul:1994:spinglass}.
-Used\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}, Maximum Satisfiability constructed with known global
-optimum.
+By enforcing only certain characteristics of a search space while allowing others
+to be generated randomly, we can perform tests on classes of problems instead
+of specific instances.  This can improve the generality of results as experiments
+then predict the expected quality over the entire class even if only a subset
+of possible instances are actually used during experimentation.  Unlike single
+instance problems, the challenge with randomly generated problems is often how
+to know when the global optimum is reached.
+
+NK landscapes define a class of problems such that the fitness of a single gene's
+value is dependent on $k$ other genes in the genome.  Finding the solution which
+maximized fitness on an arbitrary NK landscape requires exponential time, and
+are therefore a good candidate for heuristic techniques.  Nearest Neighbor NK
+landscapes are a subset of NK landscapes such that each gene is dependent on the $k$
+genes following it in the genome.  The advantage of this subset is that they can be
+solved in polynomial time~\cite{wright:2000:solvingnk}.  For our experiments we
+used wrapping neighborhoods, such that genes near the end of the genome depend on
+genes at the start of the genome.
+
+Ising Spin Glasses are another large class of $NP-Hard$ problems.  These problems
+involve minimizing the interaction terms in a graph based on edge weights by changing
+the sign of graph nodes.  Similar to with NK landscapes, there is a subset of Ising problems
+known as $2D\pm J$ Ising Spin Glasses.  This subset restricts graph interactions to a
+toroidal two dimensional grid with edge weights of $\{-1, +1\}$.  This subset can
+be solved in polynomial time~\cite{saul:1994:spinglass}, and there are existing
+tools for solving arbitrary instances.\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}
+
+Our final problem class is a subset of the Maximum Satisfiability (MAXSAT) problem.
+This problem consists of a series of disjunctive clauses, such that each clause contains
+three problem variables, some of which are negated. While
+MAXSAT requires exponential time to solve, we have devised a method of generating
+instances that ensures each has at least one solution that satisfies all clauses.
+First, a random target solution is generated.  For each clause, set the sign of each
+variable in the clause such that that clause will be satisfied by the target solution.
+To ensure there is no bias to variable signs in the generated problem there is a $\frac{1}{6}$ probability
+all clause signs match the target, a $\frac{1}{6}$ probability two of the three signs match,
+and a $\frac{4}{6}$ probability only one sign matches.  By setting the clause signs
+in this way we ensure the target solution is always able to satisfy all clauses,
+meaning all problems have a known best fitness.  Note that its possible there to be
+other solutions that satisfy all clauses.
 
@@
 
+TODO Mention source code is available\footnote{\url{https://name.of.site.removed/anonymous}}.
+
 \section{Promising Theory}
@@
 possible non-overlapping functions.
+
+Due to P3's persistent integration of random solutions, it always has a non zero
+probability of reaching the global optimum.  Because it meets this prerequisite
+(unlike LTGA), it is likely possible to derive expected running times for P3 on
+a number of problems.  While we do not present a rigorous solution to any particular
+problem here, we wish to illuminate potential avenues for future proofs.
+
+The traditional One Max class, as well as all functions of unitation, will be
+trivially solved by the hill climber in at most $N+1$ evaluations.  These come
+from the initial evaluation of the random solution, plus the $N$ evaluations required
+to test each gene for improvement.  While more difficult to bound, any unimodal
+search space without fitness plateaus will be solved by a single application
+of the hill climber.  Notably we expect a bound on Leading Ones of $O(N^2)$ time.
+
+Yet unlike many theory based algorithms, we believe P3 can achieve bounds on more
+complex and interesting landscapes.  Consider a sketch for a proof on the Deceptive
+Trap problem.  Each restart will require at most $2N$ evaluations to evaluate the
+random individual and optimize it to a local optima using the hill climber.  In
+$P_0$ all solutions will either have all 1s or all 0s.  As a result $D(C_i, C_j) = 0$
+if all $C_i$ and $C_j$ come from the same trap.  Therefore all crossover clusters
+will move only complete traps between solutions, and once there is sufficient
+diversity in the population there will be a crossover for each individual trap.
+When performing a crossover of a single trap there can only be a single alternative
+value, meaning if the candidate solution has all 0s the donor must have all 1s, or vice
+versa.  Therefore if $P_i$ contains the optimum solution for a trap in at least one
+individual and has a crossover that moves that trap, all solutions produced by crossing
+$P_i$ with a candidate solution will have the optimized value of that trap.  Extending to
+all traps we can say that once any $P_i$ has at least one solution with the optimized
+version of a trap for all traps the very next crossover with $P_i$ will produce the global optimum.
+The only remaining portion of the proof is to determine how many evaluations are required to reach this point,
+which is relatively trivial.
+
+Using loose requirements, this can likely be extended to any arbitrary fully decomposable
+problem. Assuming that for sufficiently large population sizes the entropy between
+bits in a single subproblem is lower than the entropy of bits between subproblems,
+P3 will construct a cluster capable of moving exactly one subproblem's gene values
+during crossover.  As crossover cannot decrease the fitness of a solution each
+crossover will have a non zero probability of improving subproblem solution quality.
+The probability of low fitness local optima for a subproblem reaching high levels
+is therefore low, which increases the probability of donors only having high quality
+solutions to subproblems.  Combined this is expected to create a runtime potentially
+exponential in subproblem size, but polynomial in problem size.
+
+This proof sketch can likely be extended to the HIFF problem, with minor modifications.
+Each application of the hill climber ensures all subproblems of size 2 will be solved.
+Once $P_0$ has a sufficient number of solutions it will correctly cross all subproblems
+independently, such that all future solutions crossed with $P_0$ will have all subproblems
+of size 4 solved.  While $P_1$ can contain solutions that do not correctly solve some 2 bit
+subproblems (added before $P_0$ reached critical mass), the probability of these solutions
+being selected as donors decreases rapidly (solutions added after $P_0$ reaches critical mass).
+As such each level effectively serves to optimize the next largest subproblem size, leading to
+convergence using a logarithmic number of levels.
 

### 2013-11-12 CGP Journal/main.tex
Fixed "footprint"->"sematic" in the sudocode
First hack of reviewer suggestions

@@
 We shall focus on the interaction between mutation and genome ordering.
+%NEW
+While these features of CGP have trivial definitions, we will argue, using
+experimentation and analysis of alternatives, that they have a great deal
+of impact on search success.
+%/NEW
 
@@
 being used to create output values referred to as ``inactive.''  Inactive nodes
-allow CGP individuals to drift genetically, as they can be mutated without affecting
+allow CGP individuals to drift genetically, as they can be mutated without effecting
 the fitness of the individual.  This genetic drift can then be incorporated, as
@@
 replacing the parent if it is no less fit.  This replacement strategy encourages
-neutral drift.  Mutation
-is commonly performed by mutating each gene of each node at a set probability, where mutation
+neutral drift.
+%NEW
+In CGP mutation is done in a myriad of nearly synonymous ways.  Here we mutate
+%/NEW
+each gene of each node at a set probability, where mutation
 involves changing a gene randomly to some different valid value.  For example,
 if a function gene is chosen for mutation, its new value is randomly chosen from
-all possible functions, excluding the gene's current value.  Combined,
+all possible functions, excluding the gene's current value.
+%NEW
+This version of mutation was chosen to allow precise prediction of changes
+in gene value necessarily in later sections.
+%/NEW
+Combined,
 this form of CGP can be viewed as a stochastic hill climber with neutrality.
@@
 The number of wasted evaluations is
-directly related to the mutation rate and the number of active genes.  With
-sufficiently high mutation rates and enough active genes, the probability of
+directly related to the mutation probability and the number of active genes.  With
+sufficiently high probabilities and enough active genes, the chance of
 creating an offspring which is actively identical to its parent is very low.
@@
 in~\cite{goldman:2013:cgpwaste}, this can make normal CGP very sensitive to the
-mutation rate.
+mutation probability.
 
@@
 the corresponding gene in the offspring's parent.  If all active genes are found to be equal
-than the offspring is not evaluated and is instead given the same fitness
+then the offspring is not evaluated and is instead given the same fitness
 as its parent.
@@
 it will produce identical individuals, with a potential reduction in evaluations.
-If the mutation rate is high enough relative
+If the mutation probability is high enough relative
 to the number of active genes, \emph{Skip} will require the same number of evaluations as traditional CGP, since
 the probability of an offspring being actively identical to its parent is effectively zero.
-When the mutation rate is low enough relative to the number of active genes, some number of
+When the mutation probability is low enough relative to the number of active genes, some number of
 offspring will be actively identical to their parents, resulting in a reduction in evaluations
@@
 
+%NEW
+This technique incurs a minor increase in runtime, as each active gene in the
+offspring must be compared against the parent's gene at the same locus.
+This operation is at most linear in genome size.  Yet
+on problems where evaluation is even somewhat expensive, the evaluations prevented
+by performing this check will result in a net reduction in runtime.
+%/NEW
+
 We would expect both intuitively and from our initial
 experimentation~\cite{goldman:2013:cgpwaste} that \emph{Skip} will be less sensitive
-to the mutation rate than normal CGP.  We would also expect \emph{Skip} to use a lower
-mutation rate than normal CGP when both are optimized to reduce evaluations.
-The former comes from the fact that in normal CGP any mutation rate that has a significant
-probability of creating offspring actively identical to their parents wastes
+to the mutation probability than normal CGP.  We would also expect \emph{Skip} to use a lower
+mutation probability than normal CGP when both are optimized to reduce evaluations.
+The former comes from the fact that in normal CGP any mutation probability that has a significant
+chance of creating offspring actively identical to their parents wastes
 evaluations.  In \emph{Skip} there are only two penalties for reducing the
-mutation rate.  From an evolutionary standpoint, if search becomes trapped in a local optima, low mutation
-rates may have increased difficulty escaping to find the global optimum.  From a performance standpoint,
-exceptionally low mutation rates may spend a prohibitive amount of time
+mutation probability.  From an evolutionary standpoint, if search becomes trapped in a local optima, low mutation
+probabilities may have increased difficulty escaping to find the global optimum.  From a performance standpoint,
+exceptionally low mutation probabilities may spend a prohibitive amount of time
 attempting to produce an evaluable offspring.
@@
 \emph{Skip}~\cite{goldman:2013:cgpwaste}, with the exception that \emph{Accumulate}
-favors lower mutation rates when tuned to reduce evaluations to success. While at first these algorithms may appear
+favors lower mutation probabilities when tuned to reduce evaluations to success. While at first these algorithms may appear
 quite different, further consideration shows how they are similar.  In \emph{Skip}
@@
 
+%NEW
+Similar to \emph{Skip}, \emph{Accumulate} incurs a minor computation cost in
+order to prevent wasting evaluations.  Generating each $F_i$ requires at most
+$O(AN)$, where $A$ is the arity of the nodes and $N$ is the number of nodes.
+While \emph{Accumulate} may produce many $F_i$ in the process of creating
+a single offspring, conceptually \emph{Skip} using the same mutation rate
+will create a similar number of total individuals when performing the same
+number of evaluations.  As such, we suggest \emph{Accumulate}
+will take no more than a constant factor more runtime to produce the same number
+of evaluable individuals as \emph{Skip}.  As before, this increase is likely
+eclipsed when solving problems with sufficiently time consuming evaluation functions.
+%/NEW
+
 \subsection{Single}
@@
 creating an evaluable offspring.  Third, \emph{Single} does not require a mutation
-rate parameter, effectively setting the mutation rate to $\frac{1}{a}$ for active
+parameter, effectively setting the mutation probability to $\frac{1}{a}$ for active
 genes and $\frac{1}{a+1}$ for inactive genes, where $a$ is the number of active genes.
@@
 
+%NEW
+Conceptually this method could be extended to mutate some number of active genes
+before terminating.  Doing so would reintroduce a parameter to govern how many
+active genes to mutate.  Our choice of a single mutation allows for the minimal
+amount of change to ensure the individual should be evaluated, while mutations
+made to inactive genes still give CGP sufficient power in each mutation to escape
+local optimums.
+%/NEW
+
 \section{Genome Ordering}
@@
 to represent DAGs, as all DAGs can be serialized to fit this requirement.
-However, this representation likely has an impact on CGP's ability to evolve specific
+However, this representation is likely having an impact on CGP's ability to evolve specific
 DAGs~\cite{goldman:2013:ordering}.
@@
 To rectify these issues and examine CGP without ordering bias, \cite{goldman:2013:ordering}
-created \emph{Reorder} and \emph{DAG} as methods for genome ordering, and defined
-\emph{Normal} as CGP's historical ordering method.
+created \emph{Reorder} and \emph{DAG} as methods for genome ordering.
+%NEW
+For clarity, we shall refer to CGP's historical ordering method of enforcing
+only forward connections as \emph{Normal}.
+%/NEW
 
@@
 
-The first step in performing \emph{Reorder} on a genome is to create convenience
+The first step in performing \emph{Reorder} on a genome is to create
 data structures to store direct connection relationships.  These structures are used
@@
 Once a node is assigned a location, all of the nodes that depend on its output
-have that dependency marked as satisfied.  Once all of a nodes dependencies
+have that dependency marked as satisfied.  Once all of a node's dependencies
 are satisfied, it can be put into \emph{addable}.  In this way, nodes are
@@
 its mutational adjacency can be significantly different.  As such the variety of offspring
-an individual can create is increased when using \emph{Reorder}.  While this increase
-makes finding a specific offspring that improves upon the parent less likely, the
-general likelihood of producing an offspring that improves upon the parent may 
-more than offset this drawback.
+an individual can create is increased when using \emph{Reorder}.
+%/NEW
+Therefore creating a specific offspring that improves upon the parent is
+less likely, in that the number of alternatives is increased.  Yet many
+of these additional alternatives may also be improvements upon the parent.  This may
+increase the likelihood of producing an offspring that improves
+upon the parent, allowing \emph{Reorder} to overcome this drawback.
+%/NEW
 
@@
 To achieve this, we incrementally determine
+%NEW No change was made to mutating node as that is in fact what we wanted
+%/NEW
 which nodes transitively depend on the mutating node and which do not, with
@@
 nature of the search.  As a result, the expected worst case run time for \emph{DAG}
-mutation is $O(mA^2N^2)$, where $m$ is the mutation rate.  We expect $mAN$
+mutation is $O(mA^2N^2)$, where $m$ is the mutation probability.  We expect $mAN$
 connection genes to be mutated, each requiring a worst case of $N$ dependency checks
@@
 	\centering
+	\caption{Short descriptions of each of the algorithm variants.}
 	\begin{tabular}{|c|p{6cm}|}
@@
 	\end{tabular}
-	\caption{Short descriptions of each of the algorithm variants.}
 	\label{tab:glossary}
@@
 
-We focus on two parameter values in our configuration: mutation rate and genome size.
+We focus on two parameter values in our configuration: mutation probability and genome size.
 While there are other potential configurable parameters (population size, offspring size, etc),
@@
 Furthermore, as duplication detection explicitly deals with the relationship between
-mutation rate and number of active genes, and the number of active genes depends
-on the genome size and ordering method, mutation rate and genome size seem to be
+mutation probability and number of active genes, and the number of active genes depends
+on the genome size and ordering method, mutation probability and genome size seem to be
 the most likely to impact results.
@@
 50, 100, 200, 500, 1000, 2000, 5000, 10000 for genome size and 0.05, 0.02, 0.01,
-0.005, 0.002, 0.001, 0.0005, 0.0002 for mutation rate.  Note that here a genome
+0.005, 0.002, 0.001, 0.0005, 0.0002 for mutation probability.  Note that here a genome
 size of 50 means there are 50 nodes, each composed of multiple genes, plus the
-required genes to specify output locations.  Also note that a mutation rate of 0.05
+required genes to specify output locations.  Also note that a mutation probability of 0.05
 is used here to mean each gene has a 5\% probability of being mutated in a single
@@
 range of previously used parameter settings.
-The grid is formed by trying all possible mutation rates
+The grid is formed by trying all possible mutation probabilities
 with all possible genome sizes, resulting in 64 potential configurations for each
 of the algorithm combinations.  Finally note that as \emph{Single} does not
-use a mutation rate, only the eight configurations with different genome sizes were
+use a mutation probability, only the eight configurations with different genome sizes were
 used when \emph{Single} was employed.
@@
 those parameter configurations most likely to be effective, with more information
-gathered before choosing between high quality configurations.  In total this means
+gathered before choosing between high quality configurations.
+%NEW
+We chose to run each initially 5 times to reduce the chance of
+eliminating a configuration due to outliers.  We incrementally added 4
+to provide enough information to significantly update estimates,
+while keeping the number of runs odd, allowing the median to be a
+representative run and not an interpolation.  As a final consideration
+the number of experiments were set to make the total number of runs feasible.
+%/NEW
+In total this means
 15,696 complete runs are required to set the parameter configuration for all
@@
 	\centering
+	\caption{Results for algorithm configuration on each problem.  Highlighted row
+	is the control configuration \emph{Skip-Normal}.  MES=Median Evaluations to Success; MAD=Median
+	Absolute Deviation; Active=Median number of active nodes in evolved solution;
+	Used=Median number of nodes active in evolved solution after applying SIMPLIFY (See Section~\ref{sec:behavior});
+	p-value=Result from Mann-Whitney~U comparison with control.
+	}
 	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
@@
 	  \multicolumn{9}{|c|}{\textbf{3-bit Parity}, Kruskal-Wallis=$5.99\cdot 10^{-14}$} \\ \hline
-\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
+\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
 \emph{Accumulate} & \emph{Normal} & 10,000 & 0.02 & 682 & 348 & 122 & 20 & 0.5602 \\ \hline
@@
 
-\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
+\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
 \rowcolor{Gray}
@@
 
-\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
+\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
 \emph{Skip} & \emph{Reorder}       &  2,000 & 0.002 &  62,853 & 20,023 &   757 & 114 & 0.5247 \\ \hline
@@
 
-\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
+\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
 \emph{Skip} & \emph{Reorder}      & 10,000 & 0.0005  & 196,116  & 60,326  & 3,175 & 484 & 0  \\ \hline
@@
 	\end{tabular}
-	\caption{Results for algorithm configuration on each problem.  Highlighted row
-	is the control configuration \emph{Skip-Normal}.  MES=Median Evaluations to Success; MAD=Median
-	Absolute Deviation; Active=Median number of active nodes in evolved solution;
-	Used=Median number of nodes active in evolved solution after applying SIMPLIFY (See Section~\ref{sec:behavior});
-	p-value=Result from Mann-Whitney~U comparison with control.
-	}
 	\label{tab:results}
@@
 
-As discussed in Section~\ref{sec:parameter}, the genome size and mutation rate
+As discussed in Section~\ref{sec:parameter}, the genome size and mutation probability
 for each algorithm combination was extensively optimized.  In line with previous
@@
 
-Unlike genome size, all configurations choose approximately the same mutation rate
+Unlike genome size, all configurations choose approximately the same mutation probability
 for each problem.  With the exception of Parity, there appears
 to be a general consensus that each gene should be mutated with probability 0.002,
-independent of genome size.  Parity used a higher mutation rate, but this is likely
+independent of genome size.  Parity used a higher mutation probability, but this is likely
 because 3-bit Parity is by far the easiest problem tested and therefore likely benefits
@@
 negligible~\cite{goldman:2013:cgpwaste}, even in combination with different ordering
-methods.  As \emph{Accumulate} is far more complex in terms of algorithm function
+methods.
+%NEW
+This supports the idea that the two are actually achieving synonymous behavior
+through different means.
+%/NEW
+As \emph{Accumulate} is far more complex in terms of algorithm function
 than \emph{Skip}, we would suggest \emph{Skip} be used over \emph{Accumulate} for
-any real world application.
+%NEW Slight change here, "real world" became "future"
+any future applications.
 
@@
 \begin{table*}
+	\caption{Location of never active nodes across all 50 runs for all nine algorithm combinations on the Multiply problem.}
 	\centering
@@
 	\end{tabular}
-	\caption{Location of never active nodes across all 50 runs for all nine algorithm combinations on the Multiply problem.}
 	\label{tab:never_active}
@@
 to have less or no positional bias, which also matches the results given in Table~\ref{tab:never_active}.
-%\end{comment}
+\end{comment}
 
@@
     \ForAll{$node \in reversed(active)$}
-      \State $reconnect[footprint(node)] \leftarrow node$
+      \State $reconnect[semantics(node)] \leftarrow node$
     \EndFor
     \ForAll{$i \in input\_locations$}
-      \State $reconnect[footprint(i)] \leftarrow i$
+      \State $reconnect[semantics(i)] \leftarrow i$
     \EndFor
@@
       \If{$g$ is a connection gene}
-        \If{$footprint(g) \in reconnect$}
-          \State $g \leftarrow reconnect[footprint(g)]$
+        \If{$semantics(g) \in reconnect$}
+          \State $g \leftarrow reconnect[semantics(g)]$
         \EndIf
@@
 	\centering
+	\caption{Average node behavior for all nine algorithm combinations on the Encode problem.
+	         Excludes never active nodes.}
   \begin{tabular}{c|c|c|c|}
@@
 	\end{tabular}
-	\caption{Average node behavior for all nine algorithm combinations on the Encode problem.
-	         Excludes never active nodes.}
 	\label{tab:encode_behavior}
@@
 	\centering
+	\caption{Average node behavior for all nine algorithm combinations on the Multiply problem.
+	         Excludes never active nodes.}
   \begin{tabular}{c|c|c|c|}
@@
 	\end{tabular}
-	\caption{Average node behavior for all nine algorithm combinations on the Multiply problem.
-	         Excludes never active nodes.}
 	\label{tab:multiply_behavior}
@@
 represent different ends of the quality spectrum for \emph{Skip-Normal} and
-because together the are sufficiently representative of behaviors seen on the other
+because together they are sufficiently representative of behaviors seen on the other
 problems.
@@
 of node behavior in the active nodes, and not just inactive nodes, is beneficial
-to search.  If this is true, it may help explain CGP uses such large genomes,
+to search.  If this is true, it may help explain why CGP uses such large genomes,
 as \emph{Normal} is only able to achieve enough active nodes to allow duplication
@@
 	\centering
+	\caption{Behavior of offspring replacement for each algorithm combination
+	         on each problem.
+	         Act-Un=Active Unchanged; No React=No Reactivation; React Changed=Reactivation Changed}
 	\begin{tabular}{|c|c|c|c|c|}
@@
 	\end{tabular}
-	\caption{Behavior of offspring replacement for each algorithm combination
-	         on each problem.
-	         Act-Un=Active Unchanged; No React=No Reactivation; React Changed=Reactivation Changed}
 	\label{tab:reactivate}
@@
 Somewhat unexpectedly \emph{Normal} appears to also reduce changes to active node behavior,
-causing the most effect after \emph{Single}.
+causing the largest effect after \emph{Single}.
 
@@
 active nodes or the total size of the genome, we suggest \emph{Single}.  While
-mutation rate appears relatively independent of problem and algorithm combination (all of our tuning
+mutation probability appears relatively independent of problem and algorithm combination (all of our tuning
 resulted in 0.002 being optimal or close to it), \emph{Single} still has the

### 2013-11-15 CGP Journal/response.txt
Completed a first pass of addressing reviewer comments

In [8], the authors use AND, OR, NAND, NOR for all of the other boolean problems tested (Even Parity, Digital Adder, and Digital Comparitor),
but use AND, AND with one input inverted, XOR and OR on digital multiplier.
The two papers [8] cites that performed research on this problem used AND, XOR.  They claim this makes the problem easier as
human designed solutions use only AND, XOR.  Therefore eliminating XOR as a choice likely makes this problem even harder.

It is possible to achieve O(1) amortized cost when performing mutations using the probability method.
This can be done leveraging the binomial distribution, which in many languages is implemented to give
the random number of trials k of N total which are true given probability P.  k genes can then
be selected from the genome at random just as when using a mutation rate.  This operator is
advantageous in that allows for a flexible number of mutating genes, up to and including
mutating all genes in a single step.  If I were to extend Single to use a configurable
number of genes, this would be my choice of method.

Over four thousand runs would be required to properly tune and evaluate any additional
problem.
A major focus of our work (ordering effect on the location and number of active genes)
is grounded in theory which makes no assumptions on the type of problem being solved.
As the experimental results matched the theory, we have no reason to suspect our
conclusions on these topics to change when the operators or data type is changed.
Some conclusions that could change depending on data type: The relative effectiveness
of the proposed modifications (IE, skip vs single, reorder vs normal, etc), the
evolved genome sizes, and the charts in Table IV and Table V.

The reason we chose all of our original problems as binary was to allow for the
detailed analysis of behavior in the latter half of Section VI.  If we had done
four problems, but using diverse domains, it would have been much more challenging
to determine why we observed changes in behavior between problems.  By using
four problems with similar domains and the same operators, we are able to determine
how our techniques scale with problem difficulty.

While we agree it is desirable to have more problems, and more diverse problems,
we were limited by how deep we wanted to go in analyzing our set of problems.
As stated elsewhere, each problem required approximately four thousand runs to
determine optimal genome size and mutation rate.  Similarly we wanted to delve
deeply into behavior, which requires time to determine measures and discover patterns,
not all of which are going to be domain independent.  As CGP's foundation is the
binary domain, we felt this was the correct choice for such detail oriented work.

Upon the suggestion of Reviewer 2, we further examined the statistical measures
of evaluations to success which we report in Table II.  As it is not safe to assume the number
of evaluations to success is normally distributed (in fact, they are most likely not normal),
we chose to provide non-parametric (median) instead of parametric (mean) measures,
as was done in [8].  Upon our review, we decided Median Absolute Deviation was not
capturing useful information, and we now show bootstrap confidence intervals, which is a much
more statistically sound and interesting value.

As to the including the wall clock time for our experiments, we attempted to cover our
reasoning for not doing this in the 4rd to last paragraph in the Results section.
In order to provide a fair comparison of runtime, we would need to redo tuning
with a focus on minimizing wall clock time instead of number of evaluations, which
would require another four thousand runs to time just one problem.  An additional
problem in reporting this value is hardware diversity.  To complete our experiments
we utilized a computing cluster with diverse hardware quality.  To avoid this problem,
we would need to recreate all experiments on homogeneous hardware, which is intractable
for experiments of this size.

### 2013-11-15 CGP Journal/main.tex
Completed a first pass of addressing reviewer comments

@@
 
+%NEW
+It is important to note that Equation~\ref{eq:prob} does not apply to all variants of CGP.
+For instance, the early versions of CGP~\cite{miller:2011:chapter2} which utilize a
+levels back parameter $L$ will have
+a flat active probability for all nodes at least $L$ nodes from the start and $L$ nodes from the end
+of the genome.  Yet this does not remove the issue of artificially limiting connections,
+and in fact adds many new limitations.  More recently,~\cite{cai:2005:ircgp} proposed
+a method which reassembled the genome using semantic cues, which circumvents the
+forward only limitation in \emph{Normal}.
+%/NEW
 \subsection{Reorder}
@@
 	\caption{Results for algorithm configuration on each problem.  Highlighted row
-	is the control configuration \emph{Skip-Normal}.  MES=Median Evaluations to Success; MAD=Median
-	Absolute Deviation; Active=Median number of active nodes in evolved solution;
+	is the control configuration \emph{Skip-Normal}.  MES=Median Evaluations to Success;
+	Confidence Interval=95\% bootstrapped confidence interval; Active=Median number of active nodes in evolved solution;
 	Used=Median number of nodes active in evolved solution after applying SIMPLIFY (See Section~\ref{sec:behavior});
@@
 	  \multicolumn{9}{|c|}{\textbf{3-bit Parity}, Kruskal-Wallis=$5.99\cdot 10^{-14}$} \\ \hline
-\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
-\emph{Accumulate} & \emph{Normal} & 10,000 & 0.02 & 682 & 348 & 122 & 20 & 0.5602 \\ \hline
-\emph{Accumulate} & \emph{Reorder}&  5,000 & 0.01 & 796 & 434 & 331 & 23 & 0.7174 \\ \hline
+\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{Confidence Interval} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
+\emph{Accumulate} & \emph{Normal} & 10,000 & 0.02 & 682 & 445 .. 869 & 122 & 20 & 0.5602 \\ \hline
+\emph{Accumulate} & \emph{Reorder}&  5,000 & 0.01 & 796 & 576 .. 1,019 & 331 & 23 & 0.7174 \\ \hline
 \rowcolor{Gray}
-\emph{Skip} & \emph{Normal}       & 10,000 & 0.02 & 823 & 363 & 119 & 20 & NA \\ \hline
-\emph{Skip} & \emph{Reorder}      &  2,000 & 0.01 & 870 & 362 & 198 & 20 & 0.3125 \\ \hline
-\emph{Accumulate} & \emph{Dag}    &  5,000 & 0.01 & 964 & 499 & 1,174 & 22 & 0.3310 \\ \hline
-\emph{Skip} & \emph{Dag}          & 10,000 & 0.02 & 1,042 & 620 & 2,083 & 24 & 0.0248 \\ \hline
-\emph{Single} & \emph{Normal}     & 500 & NA & 1,214 & 497 & 39 & 17 & 0.0046 \\ \hline
-\emph{Single} & \emph{Reorder}    & 200 & NA & 1,506 & 823 & 57 & 18 & 0 \\ \hline
-\emph{Single} & \emph{Dag}        & 100 & NA & 2,569 & 1,313 & 43 & 15 & 0 \\ \hline\hline
+\emph{Skip} & \emph{Normal}       & 10,000 & 0.02 & 823 & 662 .. 1,034 & 119 & 20 & NA \\ \hline
+\emph{Skip} & \emph{Reorder}      &  2,000 & 0.01 & 870 & 708 .. 1,070 & 198 & 20 & 0.3125 \\ \hline
+\emph{Accumulate} & \emph{Dag}    &  5,000 & 0.01 & 964 & 759 .. 1,250 & 1,174 & 22 & 0.3310 \\ \hline
+\emph{Skip} & \emph{Dag}          & 10,000 & 0.02 & 1,042 & 734 .. 1,339 & 2,083 & 24 & 0.0248 \\ \hline
+\emph{Single} & \emph{Normal}     & 500 & NA & 1,214 & 885 .. 1,487 & 39 & 17 & 0.0046 \\ \hline
+\emph{Single} & \emph{Reorder}    & 200 & NA & 1,506 & 552 .. 2,122 & 57 & 18 & 0 \\ \hline
+\emph{Single} & \emph{Dag}        & 100 & NA & 2,569 & 2,109 .. 3,358 & 43 & 15 & 0 \\ \hline\hline
 
@@
 
-\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
+\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{Confidence Interval} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
 \rowcolor{Gray}
-\emph{Skip} & \emph{Normal}        & 10,000 & 0.005 & 16,226 & 7,455  &   283 & 49 & NA \\ \hline
-\emph{Skip} & \emph{Reorder}       & 10,000 & 0.005 & 17,883 & 8,378  & 2,070 & 62 & 0.6918 \\ \hline
-\emph{Accumulate} & \emph{Reorder} & 10,000 & 0.002 & 18,132 & 8,926  & 2,027 & 64 & 0.8822 \\ \hline
-\emph{Accumulate} & \emph{Normal}  & 10,000 & 0.005 & 21,428 & 6,379  &   279 & 49 & 0.2525 \\ \hline
-\emph{Accumulate} & \emph{Dag}     & 10,000 & 0.002 & 24,855 & 12,512 & 4,059 & 69 & 0.0736 \\ \hline
-\emph{Skip} & \emph{Dag}           & 10,000 & 0.002 & 25,678 & 12,034 & 4,178 & 62 & 0.0211 \\ \hline
-\emph{Single} & \emph{Reorder}     & 100    & NA    & 26,558 & 7,858  &    52 & 32 & 0.0076 \\ \hline
-\emph{Single} & \emph{Normal}      & 2,000  & NA    & 31,057 & 14,475 &   141 & 43 & 0.0085 \\ \hline
-\emph{Single} & \emph{Dag}         & 100    & NA    & 34,208 & 10,069 &    59 & 34 & 0 \\ \hline\hline
+\emph{Skip} & \emph{Normal}        & 10,000 & 0.005 & 16,226 & 9,315 .. 22,229  &   283 & 49 & NA \\ \hline
+\emph{Skip} & \emph{Reorder}       & 10,000 & 0.005 & 17,883 & 10,690 .. 23,999 & 2,070 & 62 & 0.6918 \\ \hline
+\emph{Accumulate} & \emph{Reorder} & 10,000 & 0.002 & 18,132 & 13,646 .. 22,983 & 2,027 & 64 & 0.8822 \\ \hline
+\emph{Accumulate} & \emph{Normal}  & 10,000 & 0.005 & 21,428 & 17,536 .. 25,948 &   279 & 49 & 0.2525 \\ \hline
+\emph{Accumulate} & \emph{Dag}     & 10,000 & 0.002 & 24,855 & 18,222 .. 33,069 & 4,059 & 69 & 0.0736 \\ \hline
+\emph{Skip} & \emph{Dag}           & 10,000 & 0.002 & 25,678 & 16,170 .. 33,422 & 4,178 & 62 & 0.0211 \\ \hline
+\emph{Single} & \emph{Reorder}     & 100    & NA    & 26,558 & 20,678 .. 31,613 &    52 & 32 & 0.0076 \\ \hline
+\emph{Single} & \emph{Normal}      & 2,000  & NA    & 31,057 & 24,145 .. 42,278 &   141 & 43 & 0.0085 \\ \hline
+\emph{Single} & \emph{Dag}         & 100    & NA    & 34,208 & 29,840 .. 38,035 &    59 & 34 & 0 \\ \hline\hline
 
@@
 
-\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
-\emph{Skip} & \emph{Reorder}       &  2,000 & 0.002 &  62,853 & 20,023 &   757 & 114 & 0.5247 \\ \hline
-\emph{Single} & \emph{Reorder}     &    500 &    NA &  63,731 & 13,119 &   245 &  97 & 0.3610 \\ \hline
-\emph{Accumulate} & \emph{Normal}  &  1,000 & 0.002 &  67,390 & 11,991 &   211 &  87 & 0.8767 \\ \hline
+\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{Confidence Interval} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
+\emph{Skip} & \emph{Reorder}       &  2,000 & 0.002 &  62,853 & 50,033 .. 72,267 &   757 & 114 & 0.5247 \\ \hline
+\emph{Single} & \emph{Reorder}     &    500 &    NA &  63,731 & 56,070 .. 70,706 &   245 &  97 & 0.3610 \\ \hline
+\emph{Accumulate} & \emph{Normal}  &  1,000 & 0.002 &  67,390 & 60,738 .. 74,624 &   211 &  87 & 0.8767 \\ \hline
 \rowcolor{Gray}
-\emph{Skip} & \emph{Normal}        &  1,000 & 0.002 &  68,231 & 15,115 &   202 &  85 & NA \\ \hline
-\emph{Single} & \emph{Normal}      &  1,000 &    NA &  68,819 & 14,590 &   203 &  85 & 0.7695 \\ \hline
-\emph{Accumulate} & \emph{Reorder} &  2,000 & 0.002 &  72,991 & 18,650 &   760 & 115 & 0.1868 \\ \hline
-\emph{Skip} & \emph{Dag}           & 10,000 & 0.001 & 132,821 & 29,432 & 4,983 & 125 & 0 \\ \hline
-\emph{Accumulate} & \emph{Dag}     &  5,000 & 0.002 & 133,188 & 29,326 & 2,542 & 125 & 0 \\ \hline
-\emph{Single} & \emph{Dag}         &    500 &    NA & 160,696 & 39,242 &   299 &  97 & 0 \\ \hline\hline
+\emph{Skip} & \emph{Normal}        &  1,000 & 0.002 &  68,231 & 63,266 .. 74,463 &   202 &  85 & NA \\ \hline
+\emph{Single} & \emph{Normal}      &  1,000 &    NA &  68,819 & 62,940 .. 74,939 &   203 &  85 & 0.7695 \\ \hline
+\emph{Accumulate} & \emph{Reorder} &  2,000 & 0.002 &  72,991 & 67,163 .. 78,131 &   760 & 115 & 0.1868 \\ \hline
+\emph{Skip} & \emph{Dag}           & 10,000 & 0.001 & 132,821 & 117,034 .. 155,471 & 4,983 & 125 & 0 \\ \hline
+\emph{Accumulate} & \emph{Dag}     &  5,000 & 0.002 & 133,188 & 121,369 .. 147,073 & 2,542 & 125 & 0 \\ \hline
+\emph{Single} & \emph{Dag}         &    500 &    NA & 160,696 & 142,001 .. 160,696 &   299 &  97 & 0 \\ \hline\hline
 
@@
 
-\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
-\emph{Skip} & \emph{Reorder}      & 10,000 & 0.0005  & 196,116  & 60,326  & 3,175 & 484 & 0  \\ \hline
-\emph{Accumulate} & \emph{Reorder}& 10,000 & 0.001 & 199,193 & 85,970 & 3,117 & 489 & 0  \\ \hline
-\emph{Single} & \emph{Dag} & 200  & NA & 239,844 & 78,739 & 114 & 74 & 0 \\ \hline
-\emph{Single} & \emph{Reorder}    & 500   & NA & 243,664 & 91,935 & 266 & 131 & 0 \\ \hline
-\emph{Skip} & \emph{Dag}          & 2,000 & 0.002 & 369,931 & 174,408 & 952 & 149 & 0.0100 \\ \hline
-\emph{Accumulate} & \emph{Dag}    & 2,000 & 0.002  & 379,202 & 147,289 & 973 & 158 & 0.0053  \\ \hline
-\emph{Single} & \emph{Normal}     & 5,000  & NA & 399,911 & 194,384 & 273 & 112 & 0.0047 \\ \hline
-\emph{Accumulate} & \emph{Normal} & 2,000 & 0.002 & 483,938 & 227,686 & 168 & 87 & 0.45 \\ \hline
+\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{Confidence Interval} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
+\emph{Skip} & \emph{Reorder}      & 10,000 & 0.0005& 196,116 & 160,073 .. 223,500  & 3,175 & 484 & 0  \\ \hline
+\emph{Accumulate} & \emph{Reorder}& 10,000 & 0.001 & 199,193 & 123,496 .. 275,013 & 3,117 & 489 & 0  \\ \hline
+\emph{Single} & \emph{Dag}        &   200  & NA    & 239,844 & 204,347 .. 286,418 & 114 & 74 & 0 \\ \hline
+\emph{Single} & \emph{Reorder}    & 500    & NA    & 243,664 & 178,566 .. 301,246 & 266 & 131 & 0 \\ \hline
+\emph{Skip} & \emph{Dag}          & 2,000  & 0.002 & 369,931 & 234,885 .. 483,753 & 952 & 149 & 0.0100 \\ \hline
+\emph{Accumulate} & \emph{Dag}    & 2,000  & 0.002 & 379,202 & 255,165 .. 460,783 & 973 & 158 & 0.0053  \\ \hline
+\emph{Single} & \emph{Normal}     & 5,000  & NA    & 399,911 & 280,070 .. 516,034 & 273 & 112 & 0.0047 \\ \hline
+\emph{Accumulate} & \emph{Normal} & 2,000  & 0.002 & 483,938 & 308,049 .. 611,034 & 168 & 87 & 0.45 \\ \hline
 \rowcolor{Gray}
-\emph{Skip} & \emph{Normal} & 2,000 & 0.002 & 657,121 & 311,376 & 175 & 84 & NA \\ \hline
+\emph{Skip} & \emph{Normal}       & 2,000  & 0.002 & 657,121 & 469,287 .. 864,073 & 175 & 84 & NA \\ \hline
 
@@
 representing traditional CGP performance.  All configurations for each problem
-were then tested using Kruskal-Wallis to determine if any configuration was
+were then tested using
+%NEW
+Kruskal-Wallis\footnote{Kruskal-Wallis, a non-parametric equivalent of ANOVA, determines
+if samples were drawn from the same distribution.  If this test is significant, at
+least one of the samples differs from the others.}
+%/NEW
+to determine if any configuration was
 statistically different from any other.  All problems were determined to be highly significant.
 As such, each algorithm combination was then compared with the control
-to determine statistical significance using pairwise Mann-Whitney~U tests. The
+to determine statistical significance using pairwise
+%NEW
+Mann-Whitney~U\footnote{Mann-Whitney~U, a non-parametric equivalent of Student's t-test,
+determines if two populations were drawn from the same distribution.  If this test is significant,
+the populations are not equal.}
+%/NEW
+tests. The
 resulting p-value for each test is reported in the table.
@@
 we compared our median evaluations to success with previous publications.  The results
-reported for traditional CGP and CGP using modules in~\cite{walker:2008:cgpmodules}
+reported for traditional CGP in~\cite{walker:2008:cgpmodules}
 for 3-bit parity required 6 times as many evaluations as \emph{Skip-Normal}.
-The same source had traditional CGP requiring 4 million evaluations to success,
+On the 3-bit Multiplier problem using a slightly different operator set,
+the same source had traditional CGP requiring 4 million evaluations to success,
 as opposed to our \emph{Skip-Normal} requiring 600 thousand, also a factor of 6 improvement.
-CGP with modules required over 800 thousand evaluations, worse than our \emph{Skip-Normal},
-and approximately 4 times more than \emph{Skip-Reorder}.  We must be wary about comparisons
-to their 3-bit multiplier though, as their version used a different function set.
-That said the function set they used appears to have been chosen explicitly to improve
-CGP's performance on that problem, so we suggest that if we had used a the same function
-set our results would still be significantly better.
+%NEW
+To determine the effect the difference in function set had on results, we ran
+\emph{Skip-Normal} and \emph{Skip-Reorder} using their functions but our parameters.
+The former improved to 453,684 while the latter increased to 303,657, maintaining
+a statistically significant difference.  These
+results make sense as the operator set in~\cite{walker:2008:cgpmodules} is able
+to more compactly represent the solution than our set.  This puts less strain
+on \emph{Normal} to deal with complex solutions, while \emph{Reorder}'s extremely
+large genome size is likely made unnecessary.  Note that as we did not tune
+using their operator set, our results should be considered a lower bound on quality for
+using \emph{Skip-Normal} and \emph{Skip-Reorder} to solve 3-bit Multiplier
+with their operators.
+%/NEW
 

### 2013-12-02 CGP Journal/main.tex
Cut down to length, back on 15 pages!

@@
 \newcommand{\graphicthird}[1]
-{\includegraphics[width=.285\textwidth,height=.285\textheight,keepaspectratio]{#1}}
+{\includegraphics[width=.27\textwidth,height=.27\textheight,keepaspectratio]{#1}}
 
@@
 
+\begin{comment}
 In Section~\ref{sec:cgp} we provide
@@
 search, with final remarks given in Section~\ref{sec:conclusion}.
+\end{comment}
 
@@
 
+\begin{comment}
 The number of wasted evaluations is
@@
 mutation probability.
+\end{comment}
 
@@
 To understand this technique, consider an example where the parent $P$ produces
-the offspring $F_0$ using mutation. If $F_0$ is actively identical $P$, $F_0$
+the offspring $F_0$ using mutation. If $F_0$ is actively identical to $P$, $F_0$
 creates its own offspring $F_1$ using mutation.  This process of $F_i$ producing
@@
 derived the formula which predicts the probability a node is active based on its
-index, given in Equation~\ref{eq:prob} and plotted in Figure~\ref{fig:prob}.
+index, given in Equation~\ref{eq:prob}.
 \begin{equation}
@@
 
+\begin{comment}
 \begin{figure}
@@
 of the genome are almost never going to be active.
+\end{comment}
+%NEW
+Substituting $N=2000$, $A=2$, $I=6$, $O=6$ into Equation~\ref{eq:prob}
+predicts only 160 active nodes, with 80\% of the first 100 nodes expected to be
+active while just 4\% of the remaining 1900 nodes are expected to be active.
+%/NEW
 
@@
 \emph{Reorder} and \emph{DAG} would likely have less or no bias.
-Figure~\ref{fig:prob} shows the expected probability given by Equation~\ref{eq:prob},
-using the same genome size, arity, number of inputs, and number of outputs as our
-3-bit multiply.  This curve bears a striking resemblance to all three \emph{Normal}
-plots in Table~\ref{tab:never_active}.
+%NEW
+The example inputs used in that section match those of \emph{Normal}'s tuned
+configuration on 3-bit multiply.  The predicted number of active nodes and their
+location in the genome match almost exactly
+the behavior displayed in all three \emph{Normal} plots in Table~\ref{tab:never_active}.
+%/NEW
 
@@
 they satisfy.
+\\
+\\
 \begin{enumerate}
@@
 \label{sec:conclusion}
+\begin{comment}
 Even with CGP's relatively simple evolutionary mechanisms, fully understanding
@@
 a handful of variants we have helped advance this understanding.
+\end{comment}
 

### 2013-12-03 CGP Journal/main.tex
Don't need white space things anymore
Minor text fixes found during full readthrough.

@@
 %NEW
-Substituting $N=2000$, $A=2$, $I=6$, $O=6$ into Equation~\ref{eq:prob}
-predicts only 160 active nodes, with 80\% of the first 100 nodes expected to be
+Substituting $N=2000$, $A=2$, $I=6$, $O=6$ into Equation~\ref{eq:prob} (chosen to match
+experimental values used in Section~\ref{sec:analysis})
+predicts only 160 active nodes, or 8\% of the total genome.  Positionally, 80\% of the first 100 nodes are expected to be
 active while just 4\% of the remaining 1900 nodes are expected to be active.
@@
 of possible nodes preceding it is reduced.  \emph{DAG} is biased in the reverse, such that $X$ will
-have a higher chance of connecting to $Y$ that $Y$ to $X$.  This is because $X$ can connect to fewer
+have a higher chance of connecting to $Y$ than $Y$ to $X$.  This is because $X$ can connect to fewer
 nodes, so each mutation has a higher chance of choosing $Y$.
@@
 
-The first problem, 3-bit parity, represents the most
+The first problem, 3-bit Parity, represents the most
 common application for
@@
 is worse than all non \emph{DAG} configuration on that problem, even though no statistical
-tests are presented here.  The control on multiply did by far the worst of all configurations,
+tests are presented here.  The control on Multiply did by far the worst of all configurations,
 with only \emph{Accumulate Normal} likely failing statistical significance.
@@
 reported for traditional CGP in~\cite{walker:2008:cgpmodules}
-for 3-bit parity required 6 times as many evaluations as \emph{Skip-Normal}.
+for 3-bit Parity required 6 times as many evaluations as \emph{Skip-Normal}.
 On the 3-bit Multiplier problem using a slightly different operator set,
@@
 In Section~\ref{sec:ordering} we used Equation~\ref{eq:prob} to suggest \emph{Normal}
-as a strong bias on the location of active nodes regardless of fitness landscape, and that
+has a strong bias on the location of active nodes regardless of fitness landscape, and that
 \emph{Reorder} and \emph{DAG} would likely have less or no bias.
@@
 The example inputs used in that section match those of \emph{Normal}'s tuned
-configuration on 3-bit multiply.  The predicted number of active nodes and their
+configuration on 3-bit Multiply.  The predicted number of active nodes and their
 location in the genome match almost exactly
@@
 they satisfy.
-\\
-\\
 \begin{enumerate}

### 2013-12-03 CGP Journal/response.txt
First full draft of response to reviewers

@@
-In [8], the authors use AND, OR, NAND, NOR for all of the other boolean problems tested (Even Parity, Digital Adder, and Digital Comparitor),
-but use AND, AND with one input inverted, XOR and OR on digital multiplier.
-The two papers [8] cites that performed research on this problem used AND, XOR.  They claim this makes the problem easier as
-human designed solutions use only AND, XOR.  Therefore eliminating XOR as a choice likely makes this problem even harder.
+First, we would like to thank both the reviewers and the editorial staff for their
+time in examining our work.  The comments we received were very helpful and
+have lead to significant improvements.
 
-It is possible to achieve O(1) amortized cost when performing mutations using the probability method.
-This can be done leveraging the binomial distribution, which in many languages is implemented to give
-the random number of trials k of N total which are true given probability P.  k genes can then
-be selected from the genome at random just as when using a mutation rate.  This operator is
-advantageous in that allows for a flexible number of mutating genes, up to and including
-mutating all genes in a single step.  If I were to extend Single to use a configurable
-number of genes, this would be my choice of method.
-
-Over four thousand runs would be required to properly tune and evaluate any additional
-problem.
-A major focus of our work (ordering effect on the location and number of active genes)
-is grounded in theory which makes no assumptions on the type of problem being solved.
-As the experimental results matched the theory, we have no reason to suspect our
-conclusions on these topics to change when the operators or data type is changed.
-Some conclusions that could change depending on data type: The relative effectiveness
-of the proposed modifications (IE, skip vs single, reorder vs normal, etc), the
-evolved genome sizes, and the charts in Table IV and Table V.
+Before covering individual points, we would like to respond to the two
+reviewers who suggested we could improve the support of our conclusions by
+including problems outside the binary domain.  We completely agree, and as
+with Reviewer 3 we feel like this is an excellent avenue for future work.  However,
+the focus of this work was to delve deeply into a small set of problems with
+the hope of developing the hypotheses about CGP's inherent behavior.  To that
+end we felt attempting to increase the breadth would limit our depth.  As was
+pointed out by two reviewers, our work required extensive experimentation
+to achieve the conclusions on our small set of problems.  More directly, to
+include a single new problem (boolean or otherwise) with which we
+could perform comparable analysis would require over four thousand additional runs
+to perform proper parameter tuning and data collection.
 
@@
 four problems with similar domains and the same operators, we are able to determine
-how our techniques scale with problem difficulty.
+how our techniques scale with problem difficulty.  Similarly, the analysis tools
+we developed to perform the comparison could rely on problem uniformity.
+Also, since the binary domain is
+the most studied and successful in CGP literature, detailed analysis and conclusions drawn about
+its behavior in this domain are likely to have the most impact on the field.
 
-While we agree it is desirable to have more problems, and more diverse problems,
-we were limited by how deep we wanted to go in analyzing our set of problems.
-As stated elsewhere, each problem required approximately four thousand runs to
-determine optimal genome size and mutation rate.  Similarly we wanted to delve
-deeply into behavior, which requires time to determine measures and discover patterns,
-not all of which are going to be domain independent.  As CGP's foundation is the
-binary domain, we felt this was the correct choice for such detail oriented work.
+A major focus of our work (ordering effect on the location and number of active genes)
+is grounded in theory which makes no assumptions on the type of problem being solved.
+Furthermore, the operators we investigate deal with domain independent problems.
+As the experimental results matched the theory, we have no reason to suspect our
+conclusions on these topics to change when the operators or data type is changed.
+However, we understand that the following could change depending on data type: The relative effectiveness
+of the proposed modifications (IE, skip vs single, reorder vs normal, etc), the
+evolved genome sizes, and the charts in Table IV and Table V.  Yet we do not feel
+these are the primary concern of the work, nor do we have any reason to believe
+the conclusions should only hold true for binary problems.
 
-Upon the suggestion of Reviewer 2, we further examined the statistical measures
+With that said, we would now like to provide point by point responses to the reviewer
+comments.
+
+Reviewer 1:
+"You may find it useful to at least mention other representations of CGP in
+which the dependency of node ordering has been reduced, or even, removed"
+Page 3, right column, last paragraph before Section Reorder: This paragraph was
+added with appropriate citations to a few versions of CGP which do not follow Equation 1.
+
+"When comparing your results with previous publications [8], why didn't you used the same function set?"
+In [8], the authors use AND, OR, NAND, NOR for all of the other boolean problems tested (Even Parity, Digital Adder, and Digital Comparitor),
+but use AND, AND with one input inverted, XOR and OR on digital multiplier.
+The two papers [8] cites that performed research on this problem used AND, XOR.  They claim this makes the problem easier as
+human designed solutions use only AND, XOR.  Therefore eliminating XOR as a choice likely makes this problem even harder.
+That said, our paper chose to keep the operators consistent across problems.  Converting our
+version to use their operators (either [8]s or the work cited by [8]) would require an infeasible
+amount of time to again perform tuning.  However, Page 8, left column we have added more
+information about this comparison.  Notably, we used our parameters with their operators
+and maintained Reorder as being statistically better.
+
+"It is unfortunate that all analysis was limited to boolean problems"
+Covered in general response.
+
+"[I]t may be helpful to make it clear that Skip, Accumulate etc. are your invention"
+For terseness we felt that as our abstract states our focus was on "creating methods"
+combined with fact that these methods both exist in previous published work, no modification
+was necessary.
+
+"[I]n many CGP implementations it is not common to exclude the gene's current value during mutation."
+Page 1, right column, last paragraph: We have correct this to point out both that it isn't
+so much common as unclearly defined and that our choice is in doing so is to allow
+for precise behavioral analysis.  While the distinction may result in minor differences in behavior,
+we see no reason for it to effect any conclusions drawn about CGP's aggregate behavior.
+
+"Once all of a *node's* dependencies..."
+Fixed.
+
+"References or at least a very brief description of the purpose of for
+Kruskal–Wallis (as for the Mann-Whitney U test) would aid those not familiar."
+Page 7, Footnote 3 and 4 have been added.
+
+Reviewer 2:
+"Please provide the reasons why the interaction is important? Also, please explain (briefly) how your goals can be achieved."
+Page 1, left column, last sentence as been added.
+
+"[W]hy only "one" active gene is mutated?"
+Page 3, left column, last paragraph before Section Genome Reordering: Added explanation that more than
+one could have been used, but one is sufficient and removes a parameter.
+
+"", and defined Normal as CGP's historical ordering method", do you mean the ordering method of the original CGP?"
+Yes.  Page 3, right column, paragraph starting with "To rectify these", last sentence clarifies our meaning.
+
+""While this increase ... drawback.", this sentence is not entirely clear to me. Please elaborate more and provide reasons for this statement."
+Page 4, right column, last paragraph before Section DAG.  Last few sentences expand on the idea.
+
+""mutating node" or "mutated node"?"
+Mutating is in fact intended here.  This section is discussing the process of performing
+a mutation on a node, so at this point in the description it is "mutating."
+
+"It would be nice to give some rationales why potential configurations run 5 times and later 4 times? Why are higher values used?"
+Page 6, right column, last full paragraph, starting from "We chose to run each initially 5 times"
+
+""we would suggest Skip be used over accumulate for any real world application." is quite a strong claim. Only binary problems are examined in the experiments, is it enough to support your claim?"
+Page 7, right column, last full paragraph: Added some language to clarify our reasoning.  From what we can tell
+the methods are different ways of doing the exact same thing.  Since Accumulate is more complicated, we suggest Skip.
+
+"If possible, please provide the average, standard deviation of "Evaluations to success" and the average running time of each configuration.
+We further examined the statistical measures
 of evaluations to success which we report in Table II.  As it is not safe to assume the number
@@
 As to the including the wall clock time for our experiments, we attempted to cover our
-reasoning for not doing this in the 4rd to last paragraph in the Results section.
+reasoning for not doing this starting on Page 8, right column, continuing to the end of the section.
 In order to provide a fair comparison of runtime, we would need to redo tuning
@@
 for experiments of this size.
+
+Reviewer 3:
+"Although I am not suggesting any more work be undertaken. It would have been nice
+to see more work done on harder problems. ... The paper has focused on Boolean problems and it would be interesting to
+carry out similar analysis in other domains. Of course, this remains for the future."
+Covered in general response.
+
+"Mutation is generally carried out in two ways [rate vs probability]...The two methods require markedly different amounts of computation"
+It is possible to achieve O(1) amortized cost when performing mutations using the probability method.
+This can be done leveraging the binomial distribution, which in many languages is implemented to give
+the random number of trials k of N total which are true given probability P.  k genes can then
+be selected from the genome at random just as when using a mutation rate.  This operator is
+advantageous in that allows for a flexible number of mutating genes, up to and including
+mutating all genes in a single step.  We alluded to this Page 9, left column, first full paragraph.
+Note, if we were to extend Single to use a configurable
+number of genes, this would be my choice of method.  As such we would disagree that the probability method
+by necessity requires more computation than rate.
+
+That said, for clarity we have corrected all references to mutation "rate" and replaced them with "probability."
+
+"Another quite minor point is that as stated in Miller's book (page 28) when a function gene is selected for mutation, no
+check is made as to whether the mutated function is the same as that before mutation"
+Page 1, right column, last paragraph: We have clarified this.  See response to similar comment by Reviewer 1.
+
+"than" should be "then"
+Fixed.
+
+"It is worth noting in the paper that this step does incur a computational cost."
+Page 2, left column, last full paragraph.
+
+"I think the paper would benefit from some
+remarks about the extra computational cost of carrying out the Accumulate strategy"
+Page 2, right column, last paragraph.
+
+"Since one knows what the parent's active genes are, one could change an active gene at random."
+Page 3, left column, last paragraph before Section Genome Ordering partially covers this.  In general
+we felt the existing arguments about the importance of neutral drift covered why inactive genes are
+also mutated.
+
+"...this representation likely..." -> "...this representation is likely..."
+Fixed.
+
+"convenience" -> "convenient"
+Page 4, left column, first paragraph: We decided the word convenient was not necessary and removed
+it entirely.
+
+"effecting" -> "affecting"
+Fixed.
+
+"It is probably a Latex issue, but the caption TABLE 1 merges with the bottom line of the table."
+The IEEE standard is to have captions above a table.  Our original submission had captions below
+the tables, which was why there was a problem.  Caption locations for all tables have been corrected.
+
+"making comparisons with modular CGP on 3-bit parity is of little value actually"
+That comparison has been removed as we felt it was not useful.
+
+"the are" -> "they are"
+Fixed.
+
+"If this is true, it may help to explain CGP uses such large genomes..."
+Fixed.
+
+"most effect" -> "Largest effect"?
+Fixed.

### 2013-12-05 CGP Journal/response.tex
Wording fixes.

--- 2013-12-03 CGP Journal/response.txt
@@
+\documentclass{letter}
+\usepackage{hyperref}
+\usepackage{verbatim}
+%\signature{Brian W. Goldman and William F. Punch}
+%\address{21 Bridge Street \\ Smallville \\ Dunwich DU3 4WE}
+\begin{document}
+
+\begin{letter}{}
+\opening{Greetings Editors and Reviewers:}
+
 First, we would like to thank both the reviewers and the editorial staff for their
@@
 including problems outside the binary domain.  We completely agree, and as
-with Reviewer 3 we feel like this is an excellent avenue for future work.  However,
-the focus of this work was to delve deeply into a small set of problems with
-the hope of developing the hypotheses about CGP's inherent behavior.  To that
-end we felt attempting to increase the breadth would limit our depth.  As was
+noted by Reviewer 3 we feel like this is an excellent avenue for future work.  However,
+the focus of this work was to delve deeply into a small set of problems to
+develop hypotheses about CGP's inherent behavior.  To that
+end we felt that increasing the paper's breadth would limit our depth.  As was
 pointed out by two reviewers, our work required extensive experimentation
-to achieve the conclusions on our small set of problems.  More directly, to
-include a single new problem (boolean or otherwise) with which we
-could perform comparable analysis would require over four thousand additional runs
-to perform proper parameter tuning and data collection.
+to achieve our conclusions on a small set of problems.  More directly, to
+include a single new problem (boolean or otherwise) with
+comparable analysis would require over four thousand additional runs.
+%including proper parameter tuning and data collection.
 
-The reason we chose all of our original problems as binary was to allow for the
+The reason we chose binary problems was to allow for the
 detailed analysis of behavior in the latter half of Section VI.  If we had done
@@
 to determine why we observed changes in behavior between problems.  By using
-four problems with similar domains and the same operators, we are able to determine
+four problems with similar domains and the same operators, we were able to determine
 how our techniques scale with problem difficulty.  Similarly, the analysis tools
-we developed to perform the comparison could rely on problem uniformity.
-Also, since the binary domain is
-the most studied and successful in CGP literature, detailed analysis and conclusions drawn about
-its behavior in this domain are likely to have the most impact on the field.
+we developed to perform the comparison allowed us to rely on problem uniformity.
+Finally, since the binary domain is
+the most studied and successful in CGP literature, detailed analysis and conclusions drawn
+are likely to have the most impact on the field.
 
+We would like to reiterate that our thesis is understanding
+CGP's problem solving behaviors and not necessarily to suggest any of our modifications
+as universal improvements.  We agree that the inclusion of wall time comparisons on a diverse
+set of hard real world problems can be very informative when defending new methods
+for optimization, but we strive more for comprehension than improvement.
+
+\begin{comment}
 A major focus of our work (ordering effect on the location and number of active genes)
@@
 the conclusions should only hold true for binary problems.
+\end{comment}
 
@@
 human designed solutions use only AND, XOR.  Therefore eliminating XOR as a choice likely makes this problem even harder.
-That said, our paper chose to keep the operators consistent across problems.  Converting our
-version to use their operators (either [8]s or the work cited by [8]) would require an infeasible
-amount of time to again perform tuning.  However, Page 8, left column we have added more
+That said, our paper chose to keep the operators consistent across problems.
+However, Page 8, left column we have added more
 information about this comparison.  Notably, we used our parameters with their operators
@@
 "[I]n many CGP implementations it is not common to exclude the gene's current value during mutation."
-Page 1, right column, last paragraph: We have correct this to point out both that it isn't
+Page 1, right column, last paragraph: We have corrected this to point out both that it isn't
 so much common as unclearly defined and that our choice is in doing so is to allow
@@
 In order to provide a fair comparison of runtime, we would need to redo tuning
-with a focus on minimizing wall clock time instead of number of evaluations, which
-would require another four thousand runs to time just one problem.  An additional
+with a focus on minimizing wall clock time instead of number of evaluations.
+An additional
 problem in reporting this value is hardware diversity.  To complete our experiments
-we utilized a computing cluster with diverse hardware quality.  To avoid this problem,
-we would need to recreate all experiments on homogeneous hardware, which is intractable
-for experiments of this size.
+we utilized a computing cluster with diverse hardware quality.
 
@@
 "making comparisons with modular CGP on 3-bit parity is of little value actually"
-That comparison has been removed as we felt it was not useful.
+That comparison has been removed as we agree it was not useful.
 
@@
 Fixed.
+
+\end{letter}
+
+\end{document}

### 2013-12-05 CGP Journal/response.tex
Formatting and typo fixing

@@
 \usepackage{verbatim}
-%\signature{Brian W. Goldman and William F. Punch}
+\usepackage[T1]{fontenc}
+\usepackage{fullpage}
+\signature{Brian W. Goldman and William F. Punch}
 %\address{21 Bridge Street \\ Smallville \\ Dunwich DU3 4WE}
@@
 Finally, since the binary domain is
-the most studied and successful in CGP literature, detailed analysis and conclusions drawn
+the most studied and successful in CGP literature, detailed analysis and conclusions drawn there
 are likely to have the most impact on the field.
@@
 
-Reviewer 1:
-"You may find it useful to at least mention other representations of CGP in
-which the dependency of node ordering has been reduced, or even, removed"
+\textbf{Reviewer 1:}
+
+\begin{quote}
+You may find it useful to at least mention other representations of CGP in
+which the dependency of node ordering has been reduced, or even, removed.
+\end{quote}
 Page 3, right column, last paragraph before Section Reorder: This paragraph was
@@
 
-"When comparing your results with previous publications [8], why didn't you used the same function set?"
+\begin{quote}
+When comparing your results with previous publications [8], why didn't you used the same function set?
+\end{quote}
 In [8], the authors use AND, OR, NAND, NOR for all of the other boolean problems tested (Even Parity, Digital Adder, and Digital Comparitor),
@@
 
-"It is unfortunate that all analysis was limited to boolean problems"
+\begin{quote}It is unfortunate that all analysis was limited to boolean problems\end{quote}
 Covered in general response.
 
-"[I]t may be helpful to make it clear that Skip, Accumulate etc. are your invention"
-For terseness we felt that as our abstract states our focus was on "creating methods"
+\begin{quote}[I]t may be helpful to make it clear that Skip, Accumulate etc. are your invention\end{quote}
+For terseness we felt that as our abstract states our focus was on ``creating methods''
 combined with fact that these methods both exist in previous published work, no modification
@@
 
-"[I]n many CGP implementations it is not common to exclude the gene's current value during mutation."
+\begin{quote}[I]n many CGP implementations it is not common to exclude the gene's current value during mutation.\end{quote}
 Page 1, right column, last paragraph: We have corrected this to point out both that it isn't
-so much common as unclearly defined and that our choice is in doing so is to allow
+so much common as unclearly defined and that our choice was made to allow
 for precise behavioral analysis.  While the distinction may result in minor differences in behavior,
@@
 
-"Once all of a *node's* dependencies..."
+\begin{quote}Once all of a *node's* dependencies...\end{quote}
 Fixed.
 
-"References or at least a very brief description of the purpose of for
-Kruskal–Wallis (as for the Mann-Whitney U test) would aid those not familiar."
+\begin{quote}References or at least a very brief description of the purpose of for
+Kruskal-Wallis (as for the Mann-Whitney U test) would aid those not familiar.\end{quote}
 Page 7, Footnote 3 and 4 have been added.
 
-Reviewer 2:
-"Please provide the reasons why the interaction is important? Also, please explain (briefly) how your goals can be achieved."
+\textbf{Reviewer 2:}
+
+\begin{quote}Please provide the reasons why the interaction is important? Also, please explain (briefly) how your goals can be achieved.\end{quote}
 Page 1, left column, last sentence as been added.
 
-"[W]hy only "one" active gene is mutated?"
+\begin{quote}[W]hy only ``one'' active gene is mutated?\end{quote}
 Page 3, left column, last paragraph before Section Genome Reordering: Added explanation that more than
@@
 
-"", and defined Normal as CGP's historical ordering method", do you mean the ordering method of the original CGP?"
-Yes.  Page 3, right column, paragraph starting with "To rectify these", last sentence clarifies our meaning.
+\begin{quote}``and defined Normal as CGP's historical ordering method'', do you mean the ordering method of the original CGP?\end{quote}
+Yes.  Page 3, right column, paragraph starting with ``To rectify these'', last sentence clarifies our meaning.
 
-""While this increase ... drawback.", this sentence is not entirely clear to me. Please elaborate more and provide reasons for this statement."
+\begin{quote}``While this increase ... drawback.'', this sentence is not entirely clear to me. Please elaborate more and provide reasons for this statement.\end{quote}
 Page 4, right column, last paragraph before Section DAG.  Last few sentences expand on the idea.
 
-""mutating node" or "mutated node"?"
+\begin{quote}``mutating node'' or ``mutated node''?\end{quote}
 Mutating is in fact intended here.  This section is discussing the process of performing
-a mutation on a node, so at this point in the description it is "mutating."
+a mutation on a node, so at this point in the description it is ``mutating.''
 
-"It would be nice to give some rationales why potential configurations run 5 times and later 4 times? Why are higher values used?"
-Page 6, right column, last full paragraph, starting from "We chose to run each initially 5 times"
+\begin{quote}It would be nice to give some rationales why potential configurations run 5 times and later 4 times? Why are higher values used?\end{quote}
+Page 6, right column, last full paragraph, starting from ``We chose to run each initially 5 times.''
 
-""we would suggest Skip be used over accumulate for any real world application." is quite a strong claim. Only binary problems are examined in the experiments, is it enough to support your claim?"
+\begin{quote}``we would suggest Skip be used over accumulate for any real world application.'' is quite a strong claim.
+Only binary problems are examined in the experiments, is it enough to support your claim?\end{quote}
 Page 7, right column, last full paragraph: Added some language to clarify our reasoning.  From what we can tell
@@
 
-"If possible, please provide the average, standard deviation of "Evaluations to success" and the average running time of each configuration.
+\begin{quote}If possible, please provide the average, standard deviation of ``Evaluations to success'' and the average running time of each configuration.\end{quote}
 We further examined the statistical measures
@@
 
-As to the including the wall clock time for our experiments, we attempted to cover our
+As to including the wall clock time for our experiments, we attempted to cover our
 reasoning for not doing this starting on Page 8, right column, continuing to the end of the section.
@@
 problem in reporting this value is hardware diversity.  To complete our experiments
-we utilized a computing cluster with diverse hardware quality.
+we utilized a computing cluster with diverse hardware quality, and running all experiments
+on uniform hardware is infeasible for us at this time.
 
-Reviewer 3:
-"Although I am not suggesting any more work be undertaken. It would have been nice
+\textbf{Reviewer 3:}
+
+\begin{quote}Although I am not suggesting any more work be undertaken. It would have been nice
 to see more work done on harder problems. ... The paper has focused on Boolean problems and it would be interesting to
-carry out similar analysis in other domains. Of course, this remains for the future."
+carry out similar analysis in other domains. Of course, this remains for the future.\end{quote}
 Covered in general response.
 
-"Mutation is generally carried out in two ways [rate vs probability]...The two methods require markedly different amounts of computation"
+\begin{quote}Mutation is generally carried out in two ways [rate vs probability]...The two methods require markedly
+different amounts of computation...\end{quote}
 It is possible to achieve O(1) amortized cost when performing mutations using the probability method.
@@
 Note, if we were to extend Single to use a configurable
-number of genes, this would be my choice of method.  As such we would disagree that the probability method
+number of genes, this would be our choice of method.  As such we would disagree that the probability method
 by necessity requires more computation than rate.
 
-That said, for clarity we have corrected all references to mutation "rate" and replaced them with "probability."
+That said, for clarity we have corrected all references to mutation ``rate'' and replaced them with ``probability.''
 
-"Another quite minor point is that as stated in Miller's book (page 28) when a function gene is selected for mutation, no
-check is made as to whether the mutated function is the same as that before mutation"
+\begin{quote}Another quite minor point is that as stated in Miller's book (page 28) when a function gene is selected for mutation, no
+check is made as to whether the mutated function is the same as that before mutation\end{quote}
 Page 1, right column, last paragraph: We have clarified this.  See response to similar comment by Reviewer 1.
 
-"than" should be "then"
+\begin{quote}``than'' should be ``then''\end{quote}
 Fixed.
 
-"It is worth noting in the paper that this step does incur a computational cost."
+\begin{quote}It is worth noting in the paper that [checking active genes are changed] does incur a computational cost.\end{quote}
 Page 2, left column, last full paragraph.
 
-"I think the paper would benefit from some
-remarks about the extra computational cost of carrying out the Accumulate strategy"
+\begin{quote}I think the paper would benefit from some
+remarks about the extra computational cost of carrying out the Accumulate strategy\end{quote}
 Page 2, right column, last paragraph.
 
-"Since one knows what the parent's active genes are, one could change an active gene at random."
+\begin{quote}Since one knows what the parent's active genes are, one could change an active gene at random.\end{quote}
 Page 3, left column, last paragraph before Section Genome Ordering partially covers this.  In general
@@
 
-"...this representation likely..." -> "...this representation is likely..."
+\begin{quote}``...this representation likely...'' -> ``...this representation is likely...''\end{quote}
 Fixed.
 
-"convenience" -> "convenient"
+\begin{quote}``convenience'' -> ``convenient''\end{quote}
 Page 4, left column, first paragraph: We decided the word convenient was not necessary and removed
@@
 
-"effecting" -> "affecting"
+\begin{quote}``effecting'' -> ``affecting''\end{quote}
 Fixed.
 
-"It is probably a Latex issue, but the caption TABLE 1 merges with the bottom line of the table."
+\begin{quote}It is probably a Latex issue, but the caption TABLE 1 merges with the bottom line of the table.\end{quote}
 The IEEE standard is to have captions above a table.  Our original submission had captions below
@@
 
-"making comparisons with modular CGP on 3-bit parity is of little value actually"
+\begin{quote}making comparisons with modular CGP on 3-bit parity is of little value actually\end{quote}
 That comparison has been removed as we agree it was not useful.
 
-"the are" -> "they are"
+\begin{quote}``the are'' -> ``they are''\end{quote}
 Fixed.
 
-"If this is true, it may help to explain CGP uses such large genomes..."
+\begin{quote}If this is true, it may help to explain CGP uses such large genomes...\end{quote}
 Fixed.
 
-"most effect" -> "Largest effect"?
+\begin{quote}``most effect'' -> ``Largest effect''?\end{quote}
 Fixed.
 
+\closing{Thank you again for your thoughtful consideration.}
 \end{letter}

### 2013-12-13 Parameterless Population Pyramid/main.tex
Added table and figure, more filled in sections.

@@
 {\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}
+
+\newcommand{\includegraphicswide}[1]
+{\includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{#1}}
 
@@
 \label{sec-alltogether}
-TODO Pieces working together.  HC removes pairwise noise, which allows crossover to detect
-clusters.  Levels allow realization of linkage at high levels, reduction in noise.
+The pieces of P3 were specifically chosen to interact in desirable ways.  First,
+the hill climber provides an efficient tool for discovering pairwise linkage, as
+search only ends when no single bit change can be made which improves fitness.
+Therefore crossover, which relies on pairwise linkage, has high quality information
+to leverage when determining clusters.
+
+Similarly, by starting from a random position for each new solution, P3 progressively reduces
+the chance of spurious linkage formation due to initialization.  Furthermore, these random
+restarts provide exploration to balance out the hill climber and crossover's extreme
+exploitation.
+
+Finally, the use of levels allows P3 to perform solution filtering without throwing
+away information.  Once found, a local optima is stored indefinitely and its linkage
+and genes will have impact on future crossovers.  However, by splitting solutions
+into levels, the linkage information and the donated genes come from similar solution
+qualities.  This prevents noise in gene linkage which only exists in low quality
+solutions from effecting cluster formation for high quality solutions.  It also
+reduces the chance for high quality solutions to simply overwrite new, unoptimized
+solutions, preserving search diversity.
 
@@
 LTGA works by iteratively improving a population of solutions in a generational manner.
-Each generation the half of the solutions (chosen using standard tournament selection)
+Each generation half of the solutions (chosen using binary tournament selection)
 are used to construct a linkage tree just as described in Section~\ref{sec-crossover}.
@@
 
-Unlike P3, LTGA does not make use of a hill climber, and there is evidence that
+Unlike P3, this variant of LTGA does not make use of a hill climber, and there is evidence that
 doing so results in decreased efficiency~\cite{bosman:2011:lsbbo}.  LTGA applies
@@
 is increased if the best offspring in a generation is not strictly better than its parent.
+If an offspring produced by crossover has identical genes to either of its parents,
+it is not evaluated.
 
 \section{Test Problems}
-TODO Mention that since all of our algorithms are variable order insensitive
-and have no bias toward creating 1s over 0s, all will have identical results if
-problem variables are randomly reordered or XOR'd.
+All four of our optimization techniques share some interesting properties for
+problem equivalence.  First, all are fitness scale invariant, in that changing
+the fitness values without changing the relation between solution fitnesses does
+not effect any of their search methods.  This is true because none of the methods
+uses the fitness values directly, only when comparing the relative fitness between
+two solutions using boolean operators.  This does not hold true for some optimization
+methods, for instance fitness proportional selection.
 
-TODO Chose only problems with known global optima to allow comparison on time to
-global optima.
+None of the methods have a bias toward creating specific gene values outside
+of the bias created by the problem itself.  This means all of the optimizers
+will have unchanged behavior if a problem's inputs are reflected or XOR'd with a fixed string.  As
+an example, the problem of maximize the number of ones in a solution will be treated identically
+to the problem of maximizing the number of zeros in a solution.  Both are also identical
+to the problem of minimizing the hamming distance to any arbitrary solution.
+
+All of the methods are gene order independent.  This means that the location
+of genes in the genome can be randomly shuffled without effecting search behavior.
+This is not true in many evolutionary systems due to crossover.  For example,
+when using one point crossover the distance between genes in the genome effects the
+probability they will be kept together through a crossover.
+
+Combined, these features mean that tests on apparently simple problems can be
+indicative of search behavior on a large class of actually complex problems.
+
+In choosing test problems, we decided to use only benchmarks with knowable global
+optima.  This allows us to measure how many evaulations each method requires to converge to the
+global optima, as opposed to just the highest fitness reached.  This is advantageous
+as three of the four algorithms have no hard termination condition.  The remaining
+algorithm, LTGA, uses a population parameter which can be increased to slow
+convergence with the potential increase in fitness reachable before convergence.
+Therefore it is difficult to say how and when to declare any of the techniques
+converged unless the global optimum is found.
+While in practice it is often more important to find a good solution quickly
+than to find the optimum after an exceedingly long time, using this as a measure
+for algorithm quality can lead to algorithms which prematurely converge.
 
@@
 In this problem the genome is broken up hierarchically into a complete binary tree,
-with each gene representing a single bit, and each internal node representing the union
-between all genes in that nodes subtree.  A node in the tree contributes to the fitness
+with each gene representing a leaf, and each internal node representing the union
+between all genes in that node's subtree.  A node in the tree contributes to the fitness
 only if all of the genes in its subset have the same value (IE are all 0 or are all 1).
@@
 Our final single instance problem is borrowed from the real valued optimization domain.
-The Discretized Rastrigin used the standard Rastrigin evaluation function (Equation~\ref{eq-rast}).
+The Discretized Rastrigin problem uses the standard Rastrigin evaluation function (Equation~\ref{eq-rast}).
 \begin{equation}
@@
 involve minimizing the interaction terms in a graph based on edge weights by changing
-the sign of graph nodes.  Similar to with NK landscapes, there is a subset of Ising problems
+the sign of graph nodes.  Similar to with NK landscapes, there is a polynomially solvable subset of Ising problems
 known as $2D\pm J$ Ising Spin Glasses.  This subset restricts graph interactions to a
@@
 meaning all problems have a known best fitness.  Note that its possible there to be
-other solutions that satisfy all clauses.
+other solutions that also satisfy all clauses.
 
 \section{Experimental Results}
+\begin{figure*}
+  \centering
+  \includegraphicswide{WithRast}
+  \caption{Test problems}
+  \label{fig-results}
+\end{figure*}
+
+\begin{table}
+	\centering
+	\begin{tabular}{|c|r|c|}
+	  \hline
+	                & \textbf{Median} & \textbf{Confidence Interval} \\ \hline
+    \multicolumn{3}{|c|}{\textbf{Deceptive Trap}, N=805, Speedup=2.936} \\ \hline
+\textbf{LTGA} & 1,807,127 & 1,805,039 .. 1,808,718 \\ \hline
+\textbf{P3} & 615,503 & 577,000 .. 653,504 \\ \hline\hline
+    \multicolumn{3}{|c|}{\textbf{Deceptive Step Trap}, N=805, Speedup=1.107} \\ \hline
+\textbf{LTGA} & 14,106,751 & 14,102,987 .. 14,109,910 \\ \hline
+\textbf{P3} & 12,742,411 & 11,951,148 .. 13,777,090 \\ \hline\hline
+    \multicolumn{3}{|c|}{\textbf{HIFF}, N=2048, Speedup=3.1278} \\ \hline
+\textbf{LTGA} & 1,847,077 & 1,845,662 .. 1,848,951 \\ \hline
+\textbf{P3} & 590,519 & 570,739 .. 608,160 \\ \hline\hline
+    \multicolumn{3}{|c|}{\textbf{Discretized Rastrigin}, N=800, Speedup=1.1854} \\ \hline
+\textbf{LTGA} & 244,086 & 242,815 .. 245,189 \\ \hline
+\textbf{P3} & 205,908 & 201,012 .. 210,913 \\ \hline\hline
+    \multicolumn{3}{|c|}{\textbf{Nearest Neighbor NK}, N=600, Speedup=4.408} \\ \hline
+\textbf{LTGA} & 37,620,027 & 37,244,617 .. 37,990,772 \\ \hline
+\textbf{P3} & 8,533,322 & 7,315,275 .. 9,481,271 \\ \hline\hline
+    \multicolumn{3}{|c|}{\textbf{Ising Spin Glass}, N=784, Speedup=20.403} \\ \hline
+\textbf{LTGA} & 12,677,619 & 12,601,917 .. 12,761,798 \\ \hline
+\textbf{P3} & 621,347 & 568,100 .. 667,236 \\ \hline\hline
+    \multicolumn{3}{|c|}{\textbf{MAX-SAT}, N=60, Speedup=942.50} \\ \hline
+\textbf{LTGA} & 21,130,960 & 17,576,126 .. 28,314,680 \\ \hline
+\textbf{P3} & 22,420 & 14,363 .. 29,161 \\ \hline
+
+  \end{tabular}
+	\caption{Results}
+	\label{table-results}
+\end{table}
+
 TODO Cover all 7 problems as a group, focusing on the two ``types'' of problems.
@@
 
-TODO Include graphic for 7 problem results.
-
-TODO Include table comparing results on largest version of each problem.
+TODO Compare LTGA results with previously published results as best as possible.
 

### 2013-12-17 Parameterless Population Pyramid/main.tex
Add it all, who cares.

@@
 \newcommand{\includegraphicswide}[1]
-{\includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{#1}}
+{\includegraphics[width=.9\textwidth,height=\textheight,keepaspectratio]{#1}}
 
@@
 
-\title{Parameterless Population Pyramid}
+\title{Parameter-less Population Pyramid}
 \subtitle{[Genetic Algorithms Track]}
@@
 \begin{abstract}
-TODO
+%Maximum 200 Words
+A major limiting factor to many evolutionary techniques is their need to be reconfigured
+to each new problem.  While some previous methods been able to remove or reduce
+the need for parameter tuning, many do so by trading peak efficiency for general applicability.
+The Parameter-less Population Pyramid (P3) strikes a balance between continuous integration
+of diversity with extremely exploitative elitist operators, allowing it to solve easy
+problems quickly and hard problems eventually.  When compared with three state of the art
+optimization techniques across seven different problem classes P3 always finds the optimum
+at least a constant factor \emph{\textbf{faster}} than the rest.  Even when optimally
+tuning the comparisons, P3 has a lower growth complexity when test problems come
+from a random distribution.  
+Beyond empirical results, we suggest avenues for theoretical analysis of P3's
+expected runtime.  In general, while these are only the first results for P3,
+we feel they are indicative wide applicability.
 \end{abstract}
@@
 
-\keywords{Linkage Learning, Local Search, TODO}
+\keywords{Linkage Learning, Local Search, Parameter-less}
 
 \section{Introduction}
-TODO Parameterless optimization, model building, etc.
+Providing a good solution in reasonable time to a large set of real world black box problems
+is a fundamental goal of evolutionary computation.  Achieving this goal requires an
+optimizer to make assumptions about the nature of the problem is is solving.
+%TODO NFL Citation?
+Often built into the design of a genetic algorithm (GA) are evolutionary parameters,
+which allow the user to tune the optimizer's assumptions to a specific problem.  These
+parameters can allow a GA to efficiently search a multitude of landscapes.  However,
+to do so the user must first determine how to correctly tune the parameters, which is
+a complex and time consuming process unto itself.
 
-TODO Talk about ALPS~\cite{hornby:2006:alps}.
+The problem of parameter configuration has given rise to a number of attempts at
+parameter-less optimization.  While many parameters (when using reasonable settings) are unlikely to effect the order
+of complexity of an algorithm, population size can.  More specifically, sufficiently small
+population sizes can cause exponential search times, with arbitrarily large population sizes
+similarly stalling search.  Even more critically, setting the population size correctly relies
+on difficult to discern features of each problem. In an early example, a crossover based GA was devised
+which used concurrent racing populations to remove all user specified parameters~\cite{harik:1999:parameterlessga}.
+This method of population racing has also been paired with more advanced evolutionary system,
+such as Hierarchical Bayesian Optimization~\cite{pelikan:2004:parameterlesshboa}. The claim was that while not optimal,
+this method ensured the GA would take no more than a logarithmic increase in evaluations.
+Furthermore, doing so would ensure that these optimizers could be generally applicable
+without problem specific tuning.  Experimental evidence supports this theory,
+with some evidence that in general the bound is closer to a constant factor worse
+than optimal.
 
-TODO Talk about LO-LIMD~\cite{posik:2011:parameterless} and how it was parameter-less,
-but is incapable of handling overlapping linkages.
+In this work we shall introduce the Parameter-less Population Pyramid (P3).  This iterative
+method for constructing a collection of populations requires no user parameters.
+It leverages entropy based linkage detection to learn how to efficiently mix
+solutions without problem specific information beyond the evaluation function.  Perhaps
+most importantly, unlike other parameter-less techniques it experimentally appears to be at least
+a constant factor \emph{\textbf{improvement}} over comparable optimally configured
+optimization methods.
 
@@
 define a crossover (Section~\ref{sec-crossover}), which improves \#4 to create
-\#5. As \#5 used information from $P_0$ is it added to the next level of the pyramid, $P_1$.
+\#5. As \#5 used information from $P_0$ it is added to the next level of the pyramid, $P_1$.
 Again P3 lacks information to continue, so \#6 is randomly generated and improved into
@@
 has a population of solutions, with each level representing something akin to different
-generations of evolution.  The exact relationship between the levels is described
-in Section~\ref{sec-alltogether}.
+generations of evolution.  More specifically, a solution's level is one higher than
+its highest parent, similar to the layering done in~\cite{hornby:2006:alps}.
 
@@
 and merge the two closest into a new cluster.  The process ends when there is only
-a single cluster containing all loci.  As our clustering algorithm provides the proper
-criteria this algorithm requires $O(N^2)$ total time to correctly construct the linkage tree
-from the frequency table~\cite{gronau:2007:upgma}.
+a single cluster containing all loci.  As our clustering algorithm is both convex and
+satisfies the swapping lemma~\cite{gronau:2007:upgma},
+this algorithm requires $O(N^2)$ total time to correctly construct the linkage tree
+from the frequency table.
 
@@
 of clusters used during crossover.  For instance, if $|P_i| = 1$ no clusters exist, and if $|P_i| = 2$
-at most two clusters exist: one where the solutions agree and one where they disagree.
+at most two clusters exist: one for all genes where the solutions agree and one for where they disagree.
 
@@
 Evaluate the candidate.  If the fitness of the candidate has been reduced, revert the candidate's
-gene values.  When choosing the donor P3 ensures there is at least one gene value
+gene values.  To clarify, changes are kept in the case when the fitness is unaffected.
+This allows the crossover to drift through fitness plateaus.
+
+When choosing the donor P3 ensures there is at least one gene value
 different between the donor and the candidate solution.  This can potentially require
-up to $O(|P_i|)$, but is on average more efficient than allowing donations that
+up to $O(|C_j||P_i|)$, but empirically appears more efficient than allowing donations that
 do not change any gene values.
@@
 are optimized before large changes are made to the genome, which helps preserve
-diversity.  As cluster building and application ordering are not dependent on
+diversity.  This also helps prevent genetic hitchhikers, which make interfere with
+entropy calculations.  As cluster building and application ordering are not dependent on
 the position of loci in the genome, there is no positional bias in the search points
@@
 restarts provide exploration to balance out the hill climber and crossover's extreme
-exploitation.
+exploitation.  This also allows the population to grow large enough to contain enough
+diversity to solve the problem without having to know the optimal size beforehand.
 
@@
 solutions, preserving search diversity.
+
+\begin{comment}
+A major advantage to the design of P3 is that it makes very weak assumptions about
+the search landscape.  In many evolutionary systems the crossover operator assumes
+that genes linked in the search landscape are collocated in the genome.  This means
+a problem changes difficulty depending on how the genome is set up, meaning the user
+has to know information about gene linkage when applying the algorithm.  All of P3's
+operations ignore gene ordering, removing that level of configuration from the user.
+\end{comment}
+
+\subsection{Contrasted With The Generational Model}
+A major advantage of P3's structure is that it avoids the need to set a fixed
+population size.  In a generational model, the diversity contained in the initial
+population represents a bottleneck for future generations.  This is true
+because in most systems, especially on deceptive problems, mutation rates are
+low enough as to make the creation of large, multi-gene substructures unlikely
+after initialization.  As such the generational model often results in a race between
+decreasing diversity and increasing fitness, with the hope that the global optimum
+will be found before too much diversity is lost.  As such, the population size
+must be set large enough to contain sufficient initial diversity and
+to maintain that diversity during search for just long enough for all of it to be
+leveraged.
+
+Premature convergence of a population leads to the generational model often having
+two termination conditions: out of time or out of diversity.  In the former, if
+a smaller population size had been used, more of the diversity could have been exploited
+meaning a better solution likely could have been reached in the same number of evaluations.
+In the latter, if a larger population size had been used more evaluations could have been
+performed increasing the likelihood of improving solution quality.  Often the only option
+is to restart, throwing away all of the previous search information.  In either case, unless
+you can set the population size correctly beforehand, the generational model will likely
+result in waste.
+
+As P3 has exploited all known information each time the top of the pyramid is reached,
+there is no fear of wasted evaluations on unused diversity.  Similarly, because P3
+does not throw away previously optimized solutions while adding diversity, it avoids
+waste caused by premature convergence.
+
+In general, P3 only makes two assumptions: local structure and non uniform gene interdependence.
+In order for P3 to be efficient there needs to be at least some correlation between a
+solution's fitness and the fitness of its single bit neighborhood.  This is required for
+the hill climber to function properly.  Under this assumption, each time the hill climber
+improves a solution it also improves the expected fitness of the neighborhood of the solution.
+Stated another way, this assumes that small changes in genotype result in small changes in phenotype.
+The non uniform gene interdependence assumption is required by the crossover operator to
+perform clustering.  Without this assumption, the clusters formed by crossover will
+all be due to spurious relations, and the donation of genes between solutions will be
+no better than random modification.  Both assumptions are likely to hold true for
+real world problems.  Also, many existing generational model systems have similar
+or even more restrictive assumptions about search landscape.
+
+\subsection{Order of Complexity}
+With the exception of crossover, P3 requires no more than $O(N)$ computation
+per evaluation, implying it is limited by the time required to evaluate a solution.
+In order to perform crossover, P3 must spend $O(N^2)$ time updating pairwise
+frequencies each time a solution is added to a level of the pyramid.  Similarly,
+it requires $O(N^2)$ time to rebuild the set of crossover clusters once the frequency
+table has been updated.  Na\"ively, this leads to the algorithm itself requiring $O(N^2)$ time.
+However, multiple evaluations can occur for each time a single individual is added
+to a population.
+
+Each time a random restart occurs, at minimum $N+1$ evaluations are performed.
+This means each time a solution is added to $P_0$, at least $O(N)$ evaluations
+must be performed.  This implies that the $N^2$ cost of the update to $P_0$ can
+be amortized over these evaluations, making this update $O(N)$ per evaluation.  Note
+that $N+1$ is a very weak lower bound, as this only occurs if the randomly generated
+individual is already at a local optimum.  Furthermore, only restarts that result
+in a unique individual cause updates to $P_0$.
+
+To cause an update to $P_i$, all of the crossovers discovered for $P_{i-1}$ must
+first be applied.  Each causes their own evaluation.  If P3 were to na\"ively use
+all crossovers and perform an evaluation each time, this would cause $O(N)$ evaluations.
+Again, this would result in $O(N)$ amortized cost per update to $P_i$.  However,
+Section~\ref{sec-crossover} discussed how P3 ignore some clusters and will skip
+an evaluation if no change to the solution was made by a crossover.  As such we currently
+are unable to state conclusively that updating $P_i$ will require $O(N)$ evaluations.  That
+said, empirically we expect it to be so.  Furthermore, it seems counterintuitive that
+intentionally wasting evaluations would be considered beneficial to runtime.
+
+There is one more operation in P3 that may not scale linearly with genome size per evaluation:
+finding a non-identical donor.  When performing crossover P3 iteratively searches a population
+for a random donor which contains genes which differ from the current solution for the current crossover
+cluster.  In the worst case, this operation could require $|C_j||P_i|$, where $|C_j|$ is the number
+of genes in the crossover and $|P_i|$ is the number of solutions in the population.  However, $|P_i|$ grows
+very slowly.  Also, in empirical evaluation, P3's wall clock time per evaluation appears to scale linearly with
+problem size.
+
+TODO Consider discussing memory usage.
 
@@
 this comparison algorithm should perform even better than P3.  Similar to P3 this
-algorithm is parameterless, makes no assumptions about gene ordering, and will
+algorithm is Parameter-less, makes no assumptions about gene ordering, and will
 find the global optima given a sufficient (potentially exponential) number of evaluations.
@@
 or the problem of premature convergence.
+
+There have been other model based optimization techniques that use little to
+no parameters.  Specifically LO-LIMD~\cite{posik:2011:parameterless} uses a
+population size of one and a parameter free algorithm to perform optimization.
+It relies on the existence of completely separable subproblems and noise free evaluation in order to determine
+if two genes are linked through perturbation of their values.  While effective on
+this class of problems, most real world problems do not fall into this category.
+In general, LTGA is the most powerful general optimizer with few parameters.
 
@@
 In choosing test problems, we decided to use only benchmarks with knowable global
-optima.  This allows us to measure how many evaulations each method requires to converge to the
+optima.  This allows us to measure how many evaluations each method requires to converge to the
 global optima, as opposed to just the highest fitness reached.  This is advantageous
@@
 converged unless the global optimum is found.
-While in practice it is often more important to find a good solution quickly
-than to find the optimum after an exceedingly long time, using this as a measure
-for algorithm quality can lead to algorithms which prematurely converge.
+
+In practice it is often more important to find a good solution quickly than find
+the best solution after an exceedingly high number of evaluations.  However, even
+a cursory consideration of P3's operators should convince you of its ability to
+find high quality solutions in very few evaluations.  This is specifically true of the iterative
+application of the first improvement hill climber and the exceptionally elitist
+crossover operator.  Therefore we feel it is more important to test P3's resistance
+to premature convergence, as that seems a more likely flaw than being slow to reach
+``good enough'' solutions.
 
@@
 there is one local optima (all bits set to 0) and one global optima (all bits set to 1).
+We chose to use a trap size of $k=7$.
 This problem requires exponential time to solve unless there is some method to effectively
@@
 addition of fitness plateaus of configurable size $s$.  This creates an enormous
-number of local optima and makes detection of linkages much more challenging.
+number of local optima and makes detection of linkages much more challenging.  We
+chose to use a trap size of $k=7$ with plateaus of size $s=2$.
 
@@
 used wrapping neighborhoods, such that genes near the end of the genome depend on
-genes at the start of the genome.
+genes at the start of the genome.  We chose to set $k=5$.
 
@@
 meaning all problems have a known best fitness.  Note that its possible there to be
-other solutions that also satisfy all clauses.
+other solutions that also satisfy all clauses.  We chose a clause to variable ratio
+of $r=4.27$.
 
@@
   \includegraphicswide{WithRast}
-  \caption{Test problems}
+  \caption{Comparison of the median number of evaluations to reach the global optimum for
+           the four different optimization methods on seven benchmark problems with respect
+           to problem size.  If the median run did not reach the global optimum no point
+           is drawn.  Results given on a log-log scale.}
   \label{fig-results}
@@
   \end{tabular}
-	\caption{Results}
+	\caption{Comparison of the median evaluations to success for LTGA and P3.  Includes
+        	the 95\% bootstrapped confidence interval around the median, as well as
+        	the speedup factor for P3 over LTGA.}
 	\label{table-results}
@@
 
-TODO Cover all 7 problems as a group, focusing on the two ``types'' of problems.
-Make it explicit that you outperform by a constant factor on the static problems
-and by an order of complexity on generated problems.
+In total, our experiments performed 100 runs of each optimization algorithm across
+7 benchmark problems, with each run limited to 100 million evaluations.
+In performing data collection, we performed over 150,000 runs, totaling approximately
+1.5 trillion evaluations.  Of these, approximately 100,000 runs were used to optimally
+set the population size of LTGA.  Note that these runs should be considered a training phase,
+as such all reported results for LTGA use independent runs to create a testing phase.
+This was the only algorithm tested that received such tuning.  As such, LTGA should have a significant
+advantage over the others in that it has pre-learned the correct population size.
+All of the code used to perform our experiments
+as well as individual run results and analysis scripts are available from our
+website\footnote{\url{https://name.of.site.removed/anonymous}}.  The results
+are summarized in Figure~\ref{fig-results} and Table~\ref{table-results}.
 
-TODO Mention 1.5 trillion evaluations and 150,000 runs (average 10 million evaluations per run).
+On all 7 problems, the First Improvement Hill Climber and the $(1+(\lambda,\lambda))$
+optimizers appear to have a higher growth complexity than P3.  On all of the problems
+except MAT-SAT they are also worse than LTGA.  This makes sense as all of the
+benchmarks are deceptive in some sense, and neither of these techniques is designed
+to handle deceptive landscapes.  The inability for the iterative hill climber to solve
+any of the problems indicates that P3 must be doing something more than just local search.
 
-TODO Compare LTGA results with previously published results as best as possible.
+Figure~\ref{fig-results} does provide one misleading
+exception for $(1+(\lambda,\lambda))$ on the deceptive trap problem, where it appears
+level out at high genome sizes.  This is entirely an artifact of the extremely high
+variance of this algorithm on this problem.  Note that many of the large problem
+sizes failed to find the optimum for the majority of runs (no dot drawn for that size).
 
-TODO Mention source code is available\footnote{\url{https://name.of.site.removed/anonymous}}.
+The behavior of P3 and LTGA on the four single instance problems appears very similar,
+with P3 generally being lower by a constant factor.  This makes sense as LTGA's
+tuning allows it to determine a single population size sufficient to solve that
+instance.  However, because it uses a fixed population size, it must set the that
+size large enough to ensure all random initializations of that population are
+likely to contain enough diversity to reach the global optimum.  By comparison, P3 is able to
+iteratively increase the population size, stopping once enough diversity has been generated.
+As such we would expect P3 to do better, but not a lot better.
+
+When applied to the randomly generated problem classes, P3 appears to outperform LTGA
+by an order of complexity.  We suspect some of the cause is LTGA's need to fix
+a single population size.  Specifically, during tuning LTGA must set its population
+size large enough to solve all of the random instances of that class.  Therefore
+all instances which could be solved more efficiently with a smaller population size
+are penalized.  P3 on the other hand grows with problem difficulty.  As such its
+median evaluation time better reflects the median difficult of a class, whereas LTGA's
+evaluation time reflects the maximum difficulty of a class.
+
+Of the runs where LTGA was successful in the median run, it failed to find the optimal solution
+109 times, or 0.8\%.  By comparison, P3 failed to find the optimum 3 times.  All three
+where on Nearest Neighbor NK, two at size 550 and one at size 600.  Also, P3 was able
+to reliably find the optimum on Nearest Neighbor NK $N=650$, $N=700$ and MAX-SAT $N=65$, $N=70$
+while LTGA was not.
+
+Table~\ref{table-results} provides a close up look at the largest problem sizes
+in which LTGA was able to find the global optimum reliably in less than 100 million
+evaluations.  For each, we performed a pairwise Mann-Whitney~U test, all resulting
+in p-values less than $10^-22$, with the exception of Deceptive Step Trap which
+resulted in $0.00025$.  As such these results are highly significant.  The amount
+of speedup ranged from 1.1 on Deceptive Step Trap to 942 on MAX-SAT, but in all
+cases P3 outperformed LTGA and all other comparison techniques.
+
+As a further verification that our tuning for P3 is valid, we refer
+to~\cite{thierens:2013:ltgahiff}, which tested the same variant of LTGA that we
+used.  In the section on Child Node filtering, they reported approximately 200,000
+mean evaluations to success on the HIFF problem using $N=512$.  Our result of
+239,365 is of comparable quality, with P3 solving the same problem using an average
+of 92,129 evaluations.  The discrepancy is likely due to the cited work using
+their tuning runs in their average, unrealistically allowing each run to use a different tuned population size.
 
 \section{Promising Theory}
-TODO Easy proof on unimodal (no plateaus) and all functions of unitation due to HC.
-Sketch of how to prove runtime on Deceptive Trap, HIFF, suggestions toward all
-possible non-overlapping functions.
-
 Due to P3's persistent integration of random solutions, it always has a non zero
@@
 of the hill climber.  Notably we expect a bound on Leading Ones of $O(N^2)$ time.
+%TODO For future reference, n / factorial(n) + sum(i**2 / factorial(i + 1) for i in range(n))
+%is the expected number of improvements made per iteration (converges to 1.718281828459045).
 
@@
 The only remaining portion of the proof is to determine how many evaluations are required to reach this point,
-which is relatively trivial.
+which should be relatively trivial.
 
-Using loose requirements, this can likely be extended to any arbitrary fully decomposable
+Using relaxed requirements, this can likely be extended to any arbitrary fully decomposable
 problem. Assuming that for sufficiently large population sizes the entropy between
@@
 \section{Conclusions and Future Work}
-TODO No parameters required, beats tuned LTGA.  Good at solving hard classes
-of problems, not just single instances of a problem as it scales to problem difficulty.
-Future work in theory.  Also readily applicable to real world problems as no
-problem information is required.
+While many parameter-less techniques trade the potential quality of optimal tuning
+for general effectiveness, P3 appears to robustly outperform all of the comparison
+state of the art techniques across a diverse set of benchmark problems.  This is
+because unlike other parameter-less techniques, P3 does not duplicate effort running
+multiple populations, nor does it throw away learned information like many evolutionary
+techniques.  By striking
+a balance between continuously adding diversity and the extreme exploitation provided
+by the multiple evaluation crossover, P3 is able to solve easy problems quickly
+and hard problems eventually without any problem specific tuning.
+
+Even when compared against LTGA, the closest relative to P3 in current literature
+and current state of the art linkage learning algorithm, P3 surpasses expectation.
+This is even after LTGA was extensively tuned to each problem class, an unrealistic
+proposition on any real world problem.  A likely advantage is that P3 is able to
+automatically scale to problem difficulty, adding diversity only as the problem
+requires.
+%TODO Consider reiterating stuff about gene location invariant, scaling invariant.
+
+P3's ability to efficiently optimize not just single problem instance classes,
+but randomly generated problem classes without parameter setting makes it an
+ideal candidate for real world application.  Its focus on exploiting all known
+information before spending time increasing diversity makes it a great ``stop anytime''
+algorithm.  In future work, we plan to investigate in more detail P3's time to
+``good enough'' solutions, not just optimal.
+
+While not yet complete, we feel that future work will be able to provide a strong
+theoretical foundation for the design of P3.  Here we have presented outlines
+for a number of expected runtime proofs, some of which apply to complex problems.
+Again, it is P3's mix of strongly elitist selection, paired with continuous integration
+of random information that allows for this analysis which cannot be done for many
+other state of the art techniques, such as LTGA.
+
+Currently, P3 is limited to single objective, noiseless, binary encoded problems.
+In future work we plan to extend (or show no modification is necessary) P3 to handle
+problems from these classes.  Doing so will help make P3 more applicable to a wider
+range of real world problems.
+
+Finally, as future work we plan to investigate the individual pieces of P3 for
+their usefulness.  While intuitively many of them fit together well, it is possible
+that slight modification will allow for even better synergy.  Conversely, the population
+pyramid scheme itself may combine well with other optimization techniques.
 

### 2013-12-17 Parameterless Population Pyramid/main.tex
First revision pass complete.

@@
 this class of problems, most real world problems do not fall into this category.
-In general, LTGA is the most powerful general optimizer with few parameters.
+%In general, LTGA is the most powerful general optimizer with few parameters.
 
@@
 
+In total, our experiments performed 100 runs of each optimization algorithm across
+7 benchmark problems, with each run limited to 100 million evaluations.
+In performing data collection, we performed over 150,000 runs, totaling approximately
+1.5 trillion evaluations.  Of these, approximately 100,000 runs were used to optimally
+set the population size of LTGA.  Note that these runs should be considered a training phase,
+and as such all reported results for LTGA use independent runs to create a testing phase.
+This was the only algorithm tested that received such tuning.  As such, LTGA should have a significant
+advantage over the others in that it has pre-learned the correct population size.
+All of the code used to perform our experiments
+as well as individual run results and analysis scripts are available from our
+website\footnote{\url{https://name.of.site.removed/anonymous}}.  The results
+are summarized in Figure~\ref{fig-results} and Table~\ref{table-results}.
+
+On all 7 problems, the First Improvement Hill Climber and the $(1+(\lambda,\lambda))$
+optimizers appear to have a higher growth complexity than P3.  On all of the problems
+except MAT-SAT they are also worse than LTGA.  This makes sense as all of the
+benchmarks are deceptive in some sense, and neither of these techniques is designed
+to handle deceptive landscapes.  The inability for the iterative hill climber to solve
+any of the problems indicates that P3 must be doing something more than just local search.
+
+Figure~\ref{fig-results} does provide one misleading
+exception for $(1+(\lambda,\lambda))$ on the deceptive trap problem, where it appears to
+level out at high genome sizes.  This is entirely an artifact of the extremely high
+variance of this algorithm on this problem.  Note that many of the large problem
+sizes failed to find the optimum for the majority of runs (no dot drawn for that size).
+
+The behavior of P3 and LTGA on the four single instance problems appears very similar,
+with P3 generally being lower by a constant factor.  This makes sense as LTGA's
+tuning allows it to determine a single population size sufficient to solve that
+instance.  However, because it uses a fixed population size, it must set the that
+size large enough to ensure all random initializations of that population are
+likely to contain enough diversity to reach the global optimum.  By comparison, P3 is able to
+iteratively increase the population size, stopping once enough diversity has been generated.
+As such we would expect P3 to do better, but not a lot better.
+
 \begin{table}
@@
 
-In total, our experiments performed 100 runs of each optimization algorithm across
-7 benchmark problems, with each run limited to 100 million evaluations.
-In performing data collection, we performed over 150,000 runs, totaling approximately
-1.5 trillion evaluations.  Of these, approximately 100,000 runs were used to optimally
-set the population size of LTGA.  Note that these runs should be considered a training phase,
-as such all reported results for LTGA use independent runs to create a testing phase.
-This was the only algorithm tested that received such tuning.  As such, LTGA should have a significant
-advantage over the others in that it has pre-learned the correct population size.
-All of the code used to perform our experiments
-as well as individual run results and analysis scripts are available from our
-website\footnote{\url{https://name.of.site.removed/anonymous}}.  The results
-are summarized in Figure~\ref{fig-results} and Table~\ref{table-results}.
-
-On all 7 problems, the First Improvement Hill Climber and the $(1+(\lambda,\lambda))$
-optimizers appear to have a higher growth complexity than P3.  On all of the problems
-except MAT-SAT they are also worse than LTGA.  This makes sense as all of the
-benchmarks are deceptive in some sense, and neither of these techniques is designed
-to handle deceptive landscapes.  The inability for the iterative hill climber to solve
-any of the problems indicates that P3 must be doing something more than just local search.
-
-Figure~\ref{fig-results} does provide one misleading
-exception for $(1+(\lambda,\lambda))$ on the deceptive trap problem, where it appears
-level out at high genome sizes.  This is entirely an artifact of the extremely high
-variance of this algorithm on this problem.  Note that many of the large problem
-sizes failed to find the optimum for the majority of runs (no dot drawn for that size).
-
-The behavior of P3 and LTGA on the four single instance problems appears very similar,
-with P3 generally being lower by a constant factor.  This makes sense as LTGA's
-tuning allows it to determine a single population size sufficient to solve that
-instance.  However, because it uses a fixed population size, it must set the that
-size large enough to ensure all random initializations of that population are
-likely to contain enough diversity to reach the global optimum.  By comparison, P3 is able to
-iteratively increase the population size, stopping once enough diversity has been generated.
-As such we would expect P3 to do better, but not a lot better.
-
 When applied to the randomly generated problem classes, P3 appears to outperform LTGA
@@
 evaluations.  For each, we performed a pairwise Mann-Whitney~U test, all resulting
-in p-values less than $10^-22$, with the exception of Deceptive Step Trap which
+in p-values less than $10^{-22}$, with the exception of Deceptive Step Trap which
 resulted in $0.00025$.  As such these results are highly significant.  The amount
@@
 random individual and optimize it to a local optima using the hill climber.  In
-$P_0$ all solutions will either have all 1s or all 0s.  As a result $D(C_i, C_j) = 0$
+$P_0$ each trap in all solutions will either have all 1s or all 0s.  As a result $D(C_i, C_j) = 0$
 if all $C_i$ and $C_j$ come from the same trap.  Therefore all crossover clusters

### 2014-01-03 Parameterless Population Pyramid/main.tex
punch pass 1

@@
 %Maximum 200 Words
-A major limiting factor to many evolutionary techniques is their need to be reconfigured
-to each new problem.  While some previous methods been able to remove or reduce
-the need for parameter tuning, many do so by trading peak efficiency for general applicability.
-The Parameter-less Population Pyramid (P3) strikes a balance between continuous integration
-of diversity with extremely exploitative elitist operators, allowing it to solve easy
-problems quickly and hard problems eventually.  When compared with three state of the art
-optimization techniques across seven different problem classes P3 always finds the optimum
-at least a constant factor \emph{\textbf{faster}} than the rest.  Even when optimally
-tuning the comparisons, P3 has a lower growth complexity when test problems come
-from a random distribution.  
-Beyond empirical results, we suggest avenues for theoretical analysis of P3's
-expected runtime.  In general, while these are only the first results for P3,
-we feel they are indicative wide applicability.
+  A major limiting factor to practical application of evolutionary
+  techniques is the tuning of their parameters which need to be
+  reconfigured to each new problem.  While some previous methods been
+  able to remove or reduce the need for parameter tuning, many do so
+  by trading peak efficiency for general applicability.  The
+  Parameter-less Population Pyramid (P3) is an evolutionary technique
+  requiring \textbf{no} parameters and still broadly applicable to
+  many problem types. It strikes a balance between continuous
+  integration of diversity with exploitative elitist operators,
+  allowing it to solve easy problems quickly and hard problems
+  eventually.  When compared with three state of the art optimization
+  techniques across seven different problem classes, P3 always finds
+  the optimum at least a constant factor \emph{\textbf{faster}} than
+  the rest.  Even when optimally tuning comparison technique
+  parameters, P3 has a lower growth complexity when test problems come
+  from a random distribution.  Beyond empirical results, we suggest
+  avenues for theoretical analysis of P3's expected runtime.  In
+  general, while these are only the first results for P3, we feel they
+  are indicative wide applicability.
 \end{abstract}
@@
 \section{Introduction}
-Providing a good solution in reasonable time to a large set of real world black box problems
-is a fundamental goal of evolutionary computation.  Achieving this goal requires an
-optimizer to make assumptions about the nature of the problem is is solving.
-%TODO NFL Citation?
-Often built into the design of a genetic algorithm (GA) are evolutionary parameters,
-which allow the user to tune the optimizer's assumptions to a specific problem.  These
-parameters can allow a GA to efficiently search a multitude of landscapes.  However,
-to do so the user must first determine how to correctly tune the parameters, which is
-a complex and time consuming process unto itself.
+Providing a good solution in reasonable time to a broad class of real
+world black box problems is a fundamental goal of evolutionary
+computation.  Achieving this goal often requires an optimizer to make
+assumptions about the nature of the problem is is solving as no
+optimizer can solve all problems equally well~\cite{Wolpert:1997:nfl}.
 
-The problem of parameter configuration has given rise to a number of attempts at
-parameter-less optimization.  While many parameters (when using reasonable settings) are unlikely to effect the order
-of complexity of an algorithm, population size can.  More specifically, sufficiently small
-population sizes can cause exponential search times, with arbitrarily large population sizes
-similarly stalling search.  Even more critically, setting the population size correctly relies
-on difficult to discern features of each problem. In an early example, a crossover based GA was devised
-which used concurrent racing populations to remove all user specified parameters~\cite{harik:1999:parameterlessga}.
-This method of population racing has also been paired with more advanced evolutionary system,
-such as Hierarchical Bayesian Optimization~\cite{pelikan:2004:parameterlesshboa}. The claim was that while not optimal,
-this method ensured the GA would take no more than a logarithmic increase in evaluations.
-Furthermore, doing so would ensure that these optimizers could be generally applicable
-without problem specific tuning.  Experimental evidence supports this theory,
-with some evidence that in general the bound is closer to a constant factor worse
-than optimal.
+Often built into the design of a genetic algorithm (GA) are
+evolutionary parameters which allow the user to tune the optimizer's
+assumptions concerning a specific problem.  These parameters can allow
+a GA to more efficiently search the problem's landscapes.  However, to
+do so the user must first determine how to correctly \textbf{tune} the
+parameters, typically a complex and time consuming process unto
+itself.
 
-In this work we shall introduce the Parameter-less Population Pyramid (P3).  This iterative
-method for constructing a collection of populations requires no user parameters.
-It leverages entropy based linkage detection to learn how to efficiently mix
-solutions without problem specific information beyond the evaluation function.  Perhaps
-most importantly, unlike other parameter-less techniques it experimentally appears to be at least
-a constant factor \emph{\textbf{improvement}} over comparable optimally configured
+Some considerable work has been devoted to understanding how to tune
+these parameters. Early work in Evolutionary Strategies introduced the
+concept of self-adaptive evolution which was subsequently introduced
+to genetic algorithms~\cite{Back:1992:selfadapt}. Here, each solution
+carries some of the parameters used and those parameters are adapted
+along with the solution they are co-encoded with. Other work introduced the concept
+of a \emph{meta} algorithm\cite{grefenstette:1986:optimalga}. Here a
+second optimizer runs ``on top'' of the actual problem-solving
+optimizer, the second slowly tuning the parameters of the first. Both
+of these approaches are rather slow, in particular the meta
+approach, as actual problem solving must occur at set levels to give
+feedback as to how to change the parameters (online or to the meta
+optimizer).
+
+Many of these parameters have ``reasonable'' settings that are
+unlikely to effect the order of complexity of an genetic algorithm,
+but population size is an exception. For example, a population sized
+too small can cause exponential search times, while arbitrarily large
+population sizes can lead to wasted evaluations and stalled search.
+More importantly, setting the population size correctly relies on
+difficult to discern features of each problem. Some research has been
+done on setting optimal population
+sizes. Goldberg~\cite{goldberg:1991:gasize} showed that small
+population sizes were not capable of finding good solutions, though
+some practitioners~\cite{haupt:2000:optimum} have found that small
+populations with a large mutation rate are effective on practical
+problems.
+
+All of these complications have lead to research on how to rid
+evolutionary/genetic algorithms of parameter settings of any kind.  In
+an early example, a crossover based GA was devised which used
+concurrent racing populations to remove all user specified
+parameters~\cite{harik:1999:parameterlessga}.  This method of
+population racing has also been paired with more advanced evolutionary
+system, such as Hierarchical Bayesian
+Optimization~\cite{pelikan:2004:parameterlesshboa}. The claim was that
+while not optimal, this method ensured the GA would take no more than
+a logarithmic increase in evaluations.  Furthermore, doing so would
+ensure that these optimizers could be generally applicable without
+problem specific tuning.  Experimental evidence supports this theory,
+with some evidence that in general the bound is closer to a constant
+factor worse than optimal.
+
+In this work we shall introduce the Parameter-less Population Pyramid
+(P3).  P3 is an iterative method for constructing a \textbf{collection} of
+populations that requires no user parameters.  It leverages entropy based
+linkage detection to learn how to efficiently mix solutions without
+problem specific information beyond the evaluation function.  Perhaps
+most importantly, unlike other parameter-less techniques it
+experimentally appears to be at least a constant factor
+\emph{\textbf{improvement}} over comparable optimally configured
 optimization methods.
@@
 \label{sec-p3}
-Before going into detail about how each aspect of P3 works, consider the example
-evolution presented in Figure~\ref{fig-p3}.  At the start of optimization there
-is no known problem information, so solution \#1 is randomly generated.  This
-is then brought to a local optima using a hill climber (Section~\ref{sec-hillclimber})
-creating solution \#2, which is added to the pyramid (Section~\ref{sec-pyramid}) level
-$P_0$.  As there is no further exploitable information, \#3 is randomly generated
-and optimized to create \#4, which is added to $P_0$.  \#2 and \#4 are used to
-define a crossover (Section~\ref{sec-crossover}), which improves \#4 to create
-\#5. As \#5 used information from $P_0$ it is added to the next level of the pyramid, $P_1$.
-Again P3 lacks information to continue, so \#6 is randomly generated and improved into
-\#7.  Unlike \#2 and \#4, \#7 is not added in this example because it happens to be identical to
-a solution already in the pyramid.  Even though \#7 is not unique, it still crosses with
-$P_0$, which does make a unique improvement \#8 that is added to $P_1$ and mixed with that
-level to create \#9.  As before, the process creates \#10, \#11, and \#12 using
-random generation, hill climbing, and crossover, respectively.  In this example \#13
-is not added to $P_2$ because it not a strict improvement in fitness over \#12.  There
-is not enough information in $P_2$ to improve \#13 further, so the process would begin
-yet again with a random solution.
+Before going into detail about how each aspect of P3 works, let us
+consider an example of P3 operation. Unlike many EC techniques, P3
+does not have a single population of solutions.  Instead P3 maintains
+a pyramid like structure, such that each level of the pyramid has a
+population of solutions, with each level representing something akin
+to different generations of evolution. Consider the example evolution
+presented in Figure~\ref{fig-p3}.  At the start of optimization there
+are no solutions so solution \#1 is randomly generated.  Solution \#1
+is brought to a local optima using a hill climber
+(Section~\ref{sec-hillclimber}) creating solution \#2, which is added
+to the population pyramid (Section~\ref{sec-pyramid}) at population
+level $P_0$.  With a population of 1 at $P_0$, little more can be done
+so solution \#3 is randomly generated and optimized via hill climbing
+to create solution \#4, which is likewise added to the population
+pyrimad at population level $P_0$.  With a population of two,
+crossover (Section~\ref{sec-crossover}) is applied to solutions \#2
+and \#4 creating solution \#5, which is an improvement. As \#5 used
+information from $P_0$ it is added to the next level of the pyramid at
+population level $P_1$.  Because of the crossover technique that P3
+uses, there is no more information to be gained from the present
+population. Thus, \#6 is randomly
+generated and improved into \#7.  Unlike \#2 and \#4, \#7 is not added
+in this example because it happens to be identical to a solution
+already in the pyramid.  Even though \#7 is not unique, it still
+crosses with $P_0$, which does make a unique improvement \#8 that is
+added to $P_1$ and mixed with that level to create \#9.  As before,
+the process creates \#10, \#11, and \#12 using random generation, hill
+climbing, and crossover, respectively.  In this example \#13 is not
+added to $P_2$ because it not a strict improvement in fitness over
+\#12.  There is not enough information in $P_2$ to improve \#13
+further, so the process would begin yet again with a random solution.
 
@@
 \label{sec-hillclimber}
-The terminology used to describe different methods of making minor modifications
-to a solution, evaluating the quality of those changes, and iteratively constructing
-new solutions, is full of duplicated, ambiguously defined terms.  As such we shall
-describe what is likely an existing algorithm and give it a descriptive name that
-may not match all of its previous incarnations.
+% The terminology used to describe different methods of making minor modifications
+% to a solution, evaluating the quality of those changes, and iteratively constructing
+% new solutions, is full of duplicated, ambiguously defined terms.  As such we shall
+% describe what is likely an existing algorithm and give it a descriptive name that
+% may not match all of its previous incarnations.
 
-For our purposes, we use the First Improvement Hill Climber (FIHC), a method of exploiting
-local information through a series of modify-evaluate-accept iterations, to start
-from randomly initialized solutions and end with single change local optimal results.
-FIHC uses the following algorithm.  Consider all gene loci $x_i$ in a random order.
-For a given $x_i$ try all possible alternative values for the gene $x_i$, evaluating
-solution quality with each.  If any alternative value is a strict fitness improvement
-over the current value for $x_i$, replace $x_i$ with the improved value.  If no
-change is made to the solution, stop, otherwise iterate.
+For our purposes, we use a First Improvement Hill Climber
+(FIHC)\footnote{We use the name FHIC for an algorithm that likely has
+  various names in the literature, but use it here with a descriptive
+  name for clarity.}, a method of exploiting local information through
+a series of modify-evaluate-accept iterations starting from a randomly
+initialized solution and ending with a optimum, either local or global.
+FIHC uses the following algorithm.  Consider all gene loci
+$x_i$ in a random order.  For a given $x_i$ try all possible
+alternative values for the gene $x_i$, evaluating solution quality
+with change.  If any alternative value is a strict fitness improvement
+over the current value for $x_i$, replace $x_i$ with the improved
+value.  Iteration of the process continues until no such change results in
+a strict fitness improvement.
 
@@
 which require $O(N^2)$ as changes are only accepted after all neighbor solutions
-are evaluated.  In general FIHC is designed to find optima using a minimum number
+are evaluated.  In general, FIHC is designed to find optima using a minimum number
 of evaluations.  Furthermore, because FIHC tests loci in a random order it does
@@
 \label{sec-pyramid}
-Unlike many EC techniques, P3 does not have a single population of solutions.
-Instead to maintains a pyramid like structure, such that each level of the pyramid
-has a population of solutions, with each level representing something akin to different
-generations of evolution.  More specifically, a solution's level is one higher than
-its highest parent, similar to the layering done in~\cite{hornby:2006:alps}.
+As mentioned, P3 maintains a pyramid like structure of populations
+with each level representing something akin to different generations
+of evolution. More specifically, a solution's level is one higher than
+its highest parent, similar to the layering done
+in~\cite{hornby:2006:alps}.
 
@@
 (Section~\ref{sec-ltga}),
-in that it uses gene value entropy to construct clusters of genes that should have their values
+P3 uses gene value entropy to construct clusters of genes that should have their values
 conserved during crossover.  There are two primary steps in
-defining this type of crossover: cluster creation and cluster usage.
+defining this type of crossover: \emph{cluster creation} and \emph{cluster usage}.
 
@@
 represent individual gene loci and each internal node represent the subset of loci created
-by joining the subsets encoded in its children.  These clusters represent detected linked
-loci in the genome, and are used when performing crossover.
+by joining the subsets encoded in its children.  These clusters represent linked
+loci that have been detected in the genome, and are used when performing crossover.
 Using the table of observed frequencies described in Section~\ref{sec-pyramid}, Equation~\ref{eq-distance}
@@
 
-Starting from the leaves, such that each loci appears exactly once, the binary tree
-can be constructed using an agglomerative method.  Each iteration determine the
-pairwise linkage between each cluster of loci using Equation~\ref{eq-distance},
-and merge the two closest into a new cluster.  The process ends when there is only
-a single cluster containing all loci.  As our clustering algorithm is both convex and
-satisfies the swapping lemma~\cite{gronau:2007:upgma},
-this algorithm requires $O(N^2)$ total time to correctly construct the linkage tree
+Starting from the leaves, the binary tree can be constructed using an
+agglomerative method.  Each
+iteration determines the pairwise linkage between each cluster of loci
+using Equation~\ref{eq-distance}, and merge the two closest into a new
+cluster.  The process ends when there is only a single cluster
+containing all loci.  As our clustering algorithm is both convex and
+satisfies the swapping lemma~\cite{gronau:2007:upgma}, this algorithm
+requires $O(N^2)$ total time to correctly construct the linkage tree
 from the frequency table.
 
-Not all of the clusters in the linkage tree are worth using during crossover.  The most
-obvious is a cluster containing all the loci.  If we performed a crossover such that all
-of the loci must remain together, no new individual would be created.  We further remove
-clusters that are completely contained within some other cluster, such that if
-$D(C_i,C_j) = 0$, $C_i$ and $C_j$ are not used when performing crossover.  $D(C_i,C_j)$ only
-returns $0$ when the loci in $C_i$ perfectly predict the bits in $C_j$ and vice versa,
-and from this we conclude that the settings for those two loci should never be disrupted.
-When used on particularly small populations this has the potential to greatly reduce the number
-of clusters used during crossover.  For instance, if $|P_i| = 1$ no clusters exist, and if $|P_i| = 2$
-at most two clusters exist: one for all genes where the solutions agree and one for where they disagree.
+Not all of  the resulting clusters in the linkage  tree are useful for
+crossover.  The most obvious useless  cluster is one that contains all
+the loci.   A crossover that preserves  all loci in  a solution cannot
+create  a  new  individual.   We  further  remove  clusters  that  are
+completely  contained   within  some  other  cluster,   such  that  if
+$D(C_i,C_j)  =  0$, $C_i$  and  $C_j$  are  not used  when  performing
+crossover.   $D(C_i,C_j)$ only  returns  $0$ when  the  loci in  $C_i$
+perfectly predict the  bits in $C_j$ and vice versa. From this we can
+conclude  that  the  settings  for  those two  loci  should  never  be
+disrupted.  When  used on particularly small populations  this has the
+potential  to  greatly  reduce  the  number of  clusters  used  during
+crossover.  For  instance, if  $|P_i| = 1$,  no clusters exist. If
+$|P_i| =  2$ at most two clusters  exist: one for all  genes where the
+solutions agree and one for where they disagree.
 
-Given the useful set of clusters, we can now define how they are used.  P3 uses
-crossover to recombine known information in a level of the pyramid with a candidate
-solution.  For each cluster, do the following.  Select a random donor from the population.
-Copy the donor's gene values for all loci in the cluster into the candidate solution.
-Evaluate the candidate.  If the fitness of the candidate has been reduced, revert the candidate's
-gene values.  To clarify, changes are kept in the case when the fitness is unaffected.
-This allows the crossover to drift through fitness plateaus.
+Given a useful set of clusters, we can now define how they are used.
+P3 uses crossover to recombine known information in a level of the
+pyramid with a candidate solution.  The algorithm is as follows. For
+each cluster select a random donor from the population.  Copy the
+donor's gene values for all loci in the cluster into the candidate
+solution.  Evaluate the candidate.  If the fitness of the candidate
+has been reduced, revert the candidate's gene values.  It is important
+to note that candidate changes are kept in the case when the fitness
+is \emph{unaffected}.  This allows crossover to drift, potentially
+moving through plateaus.
 
-When choosing the donor P3 ensures there is at least one gene value
-different between the donor and the candidate solution.  This can potentially require
-up to $O(|C_j||P_i|)$, but empirically appears more efficient than allowing donations that
-do not change any gene values.
+When choosing a donor, P3 ensures there is at least one gene value
+different between the donor and the candidate solution.  This can
+potentially require up to $O(|C_j||P_i|)$, but empirically appears
+more efficient than allowing donations that do not change any gene
+values.
 
@@
 with clusters of equal size ordered randomly.  This ensures smaller building blocks
-are optimized before large changes are made to the genome, which helps preserve
+are optimized before large changes are made to the genome, thus helping preserve
 diversity.  This also helps prevent genetic hitchhikers, which make interfere with
@@
 \label{sec-alltogether}
-The pieces of P3 were specifically chosen to interact in desirable ways.  First,
+The pieces of P3 were specifically chosen to provide important capabilities.  First,
 the hill climber provides an efficient tool for discovering pairwise linkage, as
@@
 
-Finally, the use of levels allows P3 to perform solution filtering without throwing
-away information.  Once found, a local optima is stored indefinitely and its linkage
-and genes will have impact on future crossovers.  However, by splitting solutions
-into levels, the linkage information and the donated genes come from similar solution
-qualities.  This prevents noise in gene linkage which only exists in low quality
-solutions from effecting cluster formation for high quality solutions.  It also
-reduces the chance for high quality solutions to simply overwrite new, unoptimized
-solutions, preserving search diversity.
+Finally, the use of levels allows P3 to perform solution filtering
+without throwing away information.  Once found, a local optima is
+stored indefinitely, preserving both its linkage and genes for future
+crossovers.  By splitting solutions into levels, the linkage
+information and the donated genes come from similar solution
+qualities.  This prevents noise in gene linkage which only exists in
+low quality solutions from effecting cluster formation for high
+quality solutions.  It also reduces the chance for high quality
+solutions to simply overwrite new, unoptimized solutions, preserving
+search diversity.
 
-\begin{comment}
-A major advantage to the design of P3 is that it makes very weak assumptions about
-the search landscape.  In many evolutionary systems the crossover operator assumes
-that genes linked in the search landscape are collocated in the genome.  This means
-a problem changes difficulty depending on how the genome is set up, meaning the user
-has to know information about gene linkage when applying the algorithm.  All of P3's
-operations ignore gene ordering, removing that level of configuration from the user.
-\end{comment}
+  A major advantage to the design of P3 is that it makes very weak
+  assumptions about the search landscape.  In many evolutionary
+  systems the crossover operator assumes that genes linked in the
+  search landscape are collocated in the genome.  This introduces an
+  initialization problem, changing the difficulty of the problem due
+  to a, potentially unknown, ordering required on the genome.  All of
+  P3's operations ignore gene ordering, removing that level of
+  configuration from the user.
 
 \subsection{Contrasted With The Generational Model}
-A major advantage of P3's structure is that it avoids the need to set a fixed
-population size.  In a generational model, the diversity contained in the initial
-population represents a bottleneck for future generations.  This is true
-because in most systems, especially on deceptive problems, mutation rates are
-low enough as to make the creation of large, multi-gene substructures unlikely
-after initialization.  As such the generational model often results in a race between
-decreasing diversity and increasing fitness, with the hope that the global optimum
-will be found before too much diversity is lost.  As such, the population size
-must be set large enough to contain sufficient initial diversity and
-to maintain that diversity during search for just long enough for all of it to be
-leveraged.
+A major advantage of P3's structure is that it avoids the need to set
+a fixed population size.  In a generational model, the diversity
+contained in the initial population represents a bottleneck for future
+generations.  This is true because in most systems, especially on
+deceptive problems, mutation rates are low enough as to make the
+creation of large, multi-gene substructures unlikely after
+initialization.  As such the generational model often results in a
+race between decreasing diversity and increasing fitness. As such, the
+population size must be set large enough to contain sufficient initial
+diversity and to maintain that diversity during search for just long
+enough for all of it to be leveraged.
 
-Premature convergence of a population leads to the generational model often having
-two termination conditions: out of time or out of diversity.  In the former, if
-a smaller population size had been used, more of the diversity could have been exploited
-meaning a better solution likely could have been reached in the same number of evaluations.
-In the latter, if a larger population size had been used more evaluations could have been
-performed increasing the likelihood of improving solution quality.  Often the only option
-is to restart, throwing away all of the previous search information.  In either case, unless
-you can set the population size correctly beforehand, the generational model will likely
-result in waste.
+Premature convergence of a population leads to the generational model
+often having two termination conditions: out of time or out of
+diversity.  In the former, if a smaller population size had been used,
+more of the diversity could have been exploited meaning a better
+solution likely could have been reached in the same number of
+evaluations.  In the latter, if a larger population size had been
+used, more evaluations could have been performed increasing the
+likelihood of improving solution quality.  In
+either case, unless the population size is correctly set, the
+generational model will likely result in waste.
 
 As P3 has exploited all known information each time the top of the pyramid is reached,
-there is no fear of wasted evaluations on unused diversity.  Similarly, because P3
+there no wasted evaluations on unused diversity.  Similarly, because P3
 does not throw away previously optimized solutions while adding diversity, it avoids
-waste caused by premature convergence.
+waste caused by premature convergence. It is this essential balance of
+evaluation to diverstiy that P3 focuses on.
 
@@
 perform clustering.  Without this assumption, the clusters formed by crossover will
-all be due to spurious relations, and the donation of genes between solutions will be
+detect spurious relations. The resulting donation of genes between solutions will be
 no better than random modification.  Both assumptions are likely to hold true for
-real world problems.  Also, many existing generational model systems have similar
-or even more restrictive assumptions about search landscape.
+real world problems and are similarly assumed by most generational models.
 
 \subsection{Order of Complexity}
-With the exception of crossover, P3 requires no more than $O(N)$ computation
+With the exception of crossover, P3 requires no more than $O(N)$ computations
 per evaluation, implying it is limited by the time required to evaluate a solution.
@@
 
-To cause an update to $P_i$, all of the crossovers discovered for $P_{i-1}$ must
-first be applied.  Each causes their own evaluation.  If P3 were to na\"ively use
-all crossovers and perform an evaluation each time, this would cause $O(N)$ evaluations.
-Again, this would result in $O(N)$ amortized cost per update to $P_i$.  However,
-Section~\ref{sec-crossover} discussed how P3 ignore some clusters and will skip
-an evaluation if no change to the solution was made by a crossover.  As such we currently
-are unable to state conclusively that updating $P_i$ will require $O(N)$ evaluations.  That
-said, empirically we expect it to be so.  Furthermore, it seems counterintuitive that
-intentionally wasting evaluations would be considered beneficial to runtime.
+To cause an update to $P_i$, all of the crossovers discovered for
+$P_{i-1}$ must first be applied.  Each causes their own evaluation.
+If P3 were to na\"ively use all crossovers and perform an evaluation
+each time, this would cause $O(N)$ evaluations.  Again, this would
+result in $O(N)$ amortized cost per update to $P_i$.  However,
+Section~\ref{sec-crossover} discussed how P3 will ignore some clusters
+and will also skip an evaluation if no change to the solution was made
+by a crossover.  As such we currently are unable to state conclusively
+that updating $P_i$ will require $O(N)$ evaluations.  That said,
+empirically it to be true.  Furthermore, it seems counterintuitive
+that intentionally wasting evaluations would be considered beneficial
+to runtime.
 
-There is one more operation in P3 that may not scale linearly with genome size per evaluation:
-finding a non-identical donor.  When performing crossover P3 iteratively searches a population
-for a random donor which contains genes which differ from the current solution for the current crossover
-cluster.  In the worst case, this operation could require $|C_j||P_i|$, where $|C_j|$ is the number
-of genes in the crossover and $|P_i|$ is the number of solutions in the population.  However, $|P_i|$ grows
-very slowly.  Also, in empirical evaluation, P3's wall clock time per evaluation appears to scale linearly with
-problem size.
+There is one more operation in P3 that may not scale linearly with
+genome size per evaluation: finding a non-identical donor.  When
+performing crossover P3 iteratively searches a population for a random
+donor which contains genes which differ from the current solution for
+the current crossover cluster.  In the worst case, this operation
+could require $|C_j||P_i|$, where $|C_j|$ is the number of genes in
+the crossover and $|P_i|$ is the number of solutions in the
+population.  However, $|P_i|$ grows very slowly.  Also, in empirical
+evaluation, P3's wall clock time per evaluation appears to scale
+linearly with problem size.
 
@@
 
-As P3 utilizes a hill climber to perform optimization, it is necessary to show P3's
-added complexity is able to outperform using the hill climber alone.  Therefore
-as the first comparison we define a complete optimization method based of off the
-First Improvement Hill Climber defined in Section~\ref{sec-hillclimber}.  This definition
-is extended to match P3's style of restarting. Whenever the hill climber
-obtains a solution that cannot be improved using a single bit flip, a new solution
-is randomly generated and optimized.
+As P3 utilizes a hill climber to perform optimization, it is necessary
+to show P3's added complexity is able to outperform an optimizer that
+uses a hill climber alone.  Therefore as the first comparison we
+define a complete optimization method based on the First Improvement
+Hill Climber defined in Section~\ref{sec-hillclimber}.  This
+definition is extended to match P3's style of restarting. Whenever the
+hill climber obtains a solution that cannot be improved using a single
+bit flip, a new solution is randomly generated and optimized.
 
-If P3's quality relies solely on starting at random points and applying hill climber,
-this comparison algorithm should perform even better than P3.  Similar to P3 this
-algorithm is Parameter-less, makes no assumptions about gene ordering, and will
-find the global optima given a sufficient (potentially exponential) number of evaluations.
+If P3's quality relies solely on starting at random points and
+applying FIHC, this comparison algorithm should perform even better
+than P3.  Like P3, FHIC-alone algorithm is Parameter-less, makes no
+assumptions about gene ordering, and will find the global optima given
+a sufficient (potentially exponential) number of evaluations.
 
@@
 LTGA works by iteratively improving a population of solutions in a generational manner.
-Each generation half of the solutions (chosen using binary tournament selection)
+Each generation, half of the solutions (chosen using binary tournament selection)
 are used to construct a linkage tree just as described in Section~\ref{sec-crossover}.
@@
 
-Unlike P3, this variant of LTGA does not make use of a hill climber, and there is evidence that
-doing so results in decreased efficiency~\cite{bosman:2011:lsbbo}.  LTGA applies
-clusters in least linked first order whereas P3 uses smallest first.  Also, LTGA
-does not search for donors until one containing different genes is found.  Instead
-if the donated genes are identical, it simply skips the evaluation.  LTGA
-requires a population size parameter, maintains a fixed population size across
-generations, and does not prevent duplicate individuals in a population.  LTGA
-does not add new genetic information after initialization, meaning that if
-required diversity is not contained in the original population or lost in future
+Unlike P3, this variant of LTGA does not make use of a hill climber,
+and there is evidence that doing so results in decreased
+efficiency~\cite{bosman:2011:lsbbo}.  LTGA applies clusters in least
+linked first order whereas P3 uses smallest first.  Also, LTGA does
+not search for donors until one containing different genes is found.
+Instead, if the donated genes are identical, it simply skips the
+evaluation.  LTGA requires a population size parameter, maintains a
+fixed population size across generations, and does not prevent
+duplicate individuals in a population.  LTGA does not add new genetic
+information after initialization, meaning that if required diversity
+is not contained in the original population or lost in future
 generations, LTGA will fail to find the global optimum.
 
-In order to set the population size parameter to solve a class of problems, we use
+Since LTGA requires a population size parameter, we thought it most fair to
+compare against LTGA with an optimized population size, even though P3
+requires no such tuning. To set the population size parameter, we use
 the bisection method~\cite{goldman:2012:ltga}, which determines the minimum population
@@
 is found or the evaluation limit is reached, we want to ensure LTGA has a similar requirement
-on its success.  Therefore using the Rule of Three~\cite{jovanovic:1997:ruleofthree} we
+on its success.  Therefore, using the Rule of Three~\cite{jovanovic:1997:ruleofthree}, we
 can provide a bound on the probability of failure $\frac{3}{k+1}$ with $95\%$ confidence,
@@
 but as those variants are not in active use, we did not use them for comparison.
-However, no LTGA variant has been able to remove the population size parameter
-or the problem of premature convergence.
+However, no LTGA variant has removed the population size parameter
+or fully addressed the problem of premature convergence.
 
@@
 As a final comparison we chose the $(1+(\lambda,\lambda))$ algorithm~\cite{doerr:2013:lambdalambda}
-which is currently the best theory supported crossover method.  Unlike P3 and LTGA,
+which is currently the best, theory supported, crossover method.  Unlike P3 and LTGA,
 $(1+(\lambda,\lambda))$ does not use a population of solutions or any model building
@@
 offspring, selecting the best to recombine with the original parent to produce $\lambda$
-more offspring using uniform crossover.  In the original work a method for controlling
+more offspring using uniform crossover.  In the original work, a method for controlling
 the value of $\lambda$ is provided based on offspring success, meaning there are no
@@
 will have unchanged behavior if a problem's inputs are reflected or XOR'd with a fixed string.  As
-an example, the problem of maximize the number of ones in a solution will be treated identically
-to the problem of maximizing the number of zeros in a solution.  Both are also identical
+an example, the problem of \emph{maximize the number of ones} in a solution will be treated identically
+to the problem of \emph{maximize the number of zeros} in a solution.  Both are also identical
 to the problem of minimizing the hamming distance to any arbitrary solution.
@@
 algorithm, LTGA, uses a population parameter which can be increased to slow
-convergence with the potential increase in fitness reachable before convergence.
+convergence while simultaneously increasing the fitness reachable before convergence.
 Therefore it is difficult to say how and when to declare any of the techniques
@@
 the best solution after an exceedingly high number of evaluations.  However, even
-a cursory consideration of P3's operators should convince you of its ability to
+a cursory consideration of P3's operators should convince one of its ability to
 find high quality solutions in very few evaluations.  This is specifically true of the iterative
@@
 However, when using a hill climber and linkage learning, the Deceptive Trap problem
-may be trivially solved~\cite{goldman:2012:ltga}.  As such the Deceptive Step Trap
-problem was defined, given in Equation~\ref{eq-deceptivesteptrap}.
+may be trivially solved~\cite{goldman:2012:ltga}.  Thus a more
+difficult problem,  Deceptive Step Trap
+problem was defined, and is given in Equation~\ref{eq-deceptivesteptrap}.
 \begin{equation}
@@
 
-While both of these problems have difficult to solve subproblems, there non-overlapping
-nature and single optimal setting may not capture aspects of real world problem solving.
+While both of these problems have difficult to solve subproblems, the non-overlapping
+nature of the solutions and their single optimal setting may not capture aspects of real world problem solving.
 Therefore the Hierarchical If and only If (HIFF) problem is often used to test an
-algorithms ability to handle these characteristics~\cite{thierens:2013:ltgahiff}.
+algorithm's ability to handle these characteristics~\cite{thierens:2013:ltgahiff}.
 In this problem the genome is broken up hierarchically into a complete binary tree,
@@
 optimizers appear to have a higher growth complexity than P3.  On all of the problems
-except MAT-SAT they are also worse than LTGA.  This makes sense as all of the
+except MAX-SAT they are also worse than LTGA.  This makes sense as all of the
 benchmarks are deceptive in some sense, and neither of these techniques is designed
@@
 \section{Promising Theory}
-Due to P3's persistent integration of random solutions, it always has a non zero
-probability of reaching the global optimum.  Because it meets this prerequisite
-(unlike LTGA), it is likely possible to derive expected running times for P3 on
-a number of problems.  While we do not present a rigorous solution to any particular
-problem here, we wish to illuminate potential avenues for future proofs.
+Due to P3's persistent integration of random solutions, it always has
+a non zero probability of reaching the global optimum.  Because of
+this (and unlike LTGA), it is possible to derive expected
+running times for P3 on a number of problems.  While we do not present
+a rigorous solution to any particular problem here, though we do wish to
+illuminate potential avenues for future proofs.
 

### 2014-01-05 Parameterless Population Pyramid/main.tex
Minor word updates

@@
 computation.  Achieving this goal often requires an optimizer to make
-assumptions about the nature of the problem is is solving as no
+assumptions about the nature of the problem it is solving as no
 optimizer can solve all problems equally well~\cite{Wolpert:1997:nfl}.
@@
 alternative values for the gene $x_i$, evaluating solution quality
-with change.  If any alternative value is a strict fitness improvement
+with each change.  If any alternative value is a strict fitness improvement
 over the current value for $x_i$, replace $x_i$ with the improved
@@
 iteration determines the pairwise linkage between each cluster of loci
-using Equation~\ref{eq-distance}, and merge the two closest into a new
+using Equation~\ref{eq-distance}, and merges the two closest into a new
 cluster.  The process ends when there is only a single cluster
@@
 As P3 has exploited all known information each time the top of the pyramid is reached,
-there no wasted evaluations on unused diversity.  Similarly, because P3
+no evaluations are wasted  on unused diversity.  Similarly, because P3
 does not throw away previously optimized solutions while adding diversity, it avoids
@@
 that updating $P_i$ will require $O(N)$ evaluations.  That said,
-empirically it to be true.  Furthermore, it seems counterintuitive
+empirically it appears to be true.  Furthermore, it seems counterintuitive
 that intentionally wasting evaluations would be considered beneficial
@@
 running times for P3 on a number of problems.  While we do not present
-a rigorous solution to any particular problem here, though we do wish to
+a rigorous solution to any particular problem here, we wish to
 illuminate potential avenues for future proofs.

### 2014-01-13 Parameterless Population Pyramid/main.tex
Some wordsmithing

@@
 %Maximum 200 Words
-  A major limiting factor to practical application of evolutionary
-  techniques is the tuning of their parameters which need to be
-  reconfigured to each new problem.  While some previous methods been
-  able to remove or reduce the need for parameter tuning, many do so
+  Real world applications of evolutionary techniques are often hindered by
+  the need to determine problem specific parameter settings.
+  While some previous methods have reduced or removed
+  the need for parameter tuning, many do so
   by trading peak efficiency for general applicability.  The
   Parameter-less Population Pyramid (P3) is an evolutionary technique
-  requiring \textbf{no} parameters and still broadly applicable to
-  many problem types. It strikes a balance between continuous
+  requiring \textbf{no} parameters which is still broadly applicable to
+  many problem types. P3 strikes a balance between continuous
   integration of diversity with exploitative elitist operators,
@@
   the rest.  Even when optimally tuning comparison technique
-  parameters, P3 has a lower growth complexity when test problems come
+  parameters, P3 has a lower evaluations to success growth complexity when test problems come
   from a random distribution.  Beyond empirical results, we suggest
@@
   general, while these are only the first results for P3, we feel they
-  are indicative wide applicability.
+  indicate wide applicability.
 \end{abstract}
@@
 evolutionary parameters which allow the user to tune the optimizer's
-assumptions concerning a specific problem.  These parameters can allow
-a GA to more efficiently search the problem's landscapes.  However, to
+assumptions to match a specific problem.  These parameters can allow
+a GA to more efficiently search the problem's landscape.  However, to
 do so the user must first determine how to correctly \textbf{tune} the
@@
 along with the solution they are co-encoded with. Other work introduced the concept
-of a \emph{meta} algorithm\cite{grefenstette:1986:optimalga}. Here a
+of a \emph{meta} algorithm~\cite{grefenstette:1986:optimalga}. Here a
 second optimizer runs ``on top'' of the actual problem-solving
@@
 
-Many of these parameters have ``reasonable'' settings that are
-unlikely to effect the order of complexity of an genetic algorithm,
-but population size is an exception. For example, a population sized
-too small can cause exponential search times, while arbitrarily large
+Many parameters have ``reasonable'' settings that are
+unlikely to effect the order of complexity of a genetic algorithm,
+with population size being a notable exception. For example, an overly small
+population can cause exponential search times, while arbitrarily large
 population sizes can lead to wasted evaluations and stalled search.
 More importantly, setting the population size correctly relies on
-difficult to discern features of each problem. Some research has been
-done on setting optimal population
-sizes. Goldberg~\cite{goldberg:1991:gasize} showed that small
+difficult to discern features of each problem. 
+Goldberg~\cite{goldberg:1991:gasize} showed that small
 population sizes were not capable of finding good solutions, though
 some practitioners~\cite{haupt:2000:optimum} have found that small
-populations with a large mutation rate are effective on practical
+populations with a large mutation rate are effective on some practical
 problems.
 
-All of these complications have lead to research on how to rid
-evolutionary/genetic algorithms of parameter settings of any kind.  In
+These complications have encouraged work to remove the need for
+operator defined evolutionary parameters. In
 an early example, a crossover based GA was devised which used
@@
 while not optimal, this method ensured the GA would take no more than
-a logarithmic increase in evaluations.  Furthermore, doing so would
+a logarithmic factor increase in evaluations beyond optimal.  Furthermore, doing so would
 ensure that these optimizers could be generally applicable without
@@
 to different generations of evolution. Consider the example evolution
-presented in Figure~\ref{fig-p3}.  At the start of optimization there
-are no solutions so solution \#1 is randomly generated.  Solution \#1
+presented in Figure~\ref{fig-p3}.  At the start of optimization P3
+has no information about the landscape so solution \#1 is randomly generated.  Solution \#1
 is brought to a local optima using a hill climber
@@
 to the population pyramid (Section~\ref{sec-pyramid}) at population
-level $P_0$.  With a population of 1 at $P_0$, little more can be done
+level $P_0$.  With a population of 1 at $P_0$, there is insufficient information to continue,
 so solution \#3 is randomly generated and optimized via hill climbing
@@
 climbing, and crossover, respectively.  In this example \#13 is not
-added to $P_2$ because it not a strict improvement in fitness over
+added to $P_2$ because it not a strict fitness improvement over
 \#12.  There is not enough information in $P_2$ to improve \#13
@@
 a series of modify-evaluate-accept iterations starting from a randomly
-initialized solution and ending with a optimum, either local or global.
+initialized solution and ending with an optimum, either local or global.
 FIHC uses the following algorithm.  Consider all gene loci
@@
 P3 uses a crossover method derived from that defined for LTGA~\cite{thierens:2013:ltgahiff}
-(Section~\ref{sec-ltga}),
+(Section~\ref{sec-ltga});
 P3 uses gene value entropy to construct clusters of genes that should have their values
@@
 crossover.   $D(C_i,C_j)$ only  returns  $0$ when  the  loci in  $C_i$
-perfectly predict the  bits in $C_j$ and vice versa. From this we can
+perfectly predict the gene values in $C_j$ and vice versa. From this we can
 conclude  that  the  settings  for  those two  loci  should  never  be
@@
 are optimized before large changes are made to the genome, thus helping preserve
-diversity.  This also helps prevent genetic hitchhikers, which make interfere with
+diversity.  This also helps prevent genetic hitchhikers, which may interfere with
 entropy calculations.  As cluster building and application ordering are not dependent on
-the position of loci in the genome, there is no positional bias in the search points
+the position of loci in the genome, this operator introduces no positional bias in the search points
 evaluated.
@@
 \label{sec-alltogether}
-The pieces of P3 were specifically chosen to provide important capabilities.  First,
+The pieces of P3 were specifically chosen to provide important complimentary capabilities.  First,
 the hill climber provides an efficient tool for discovering pairwise linkage, as
@@
 initialization.  As such the generational model often results in a
-race between decreasing diversity and increasing fitness. As such, the
+race between decreasing diversity and increasing fitness. This means the
 population size must be set large enough to contain sufficient initial
@@
 waste caused by premature convergence. It is this essential balance of
-evaluation to diverstiy that P3 focuses on.
+exploitation to diversity that P3 focuses on.
 
@@
 and will also skip an evaluation if no change to the solution was made
-by a crossover.  As such we currently are unable to state conclusively
+by a crossover.  As such we are currently unable to state conclusively
 that updating $P_i$ will require $O(N)$ evaluations.  That said,
@@
 
-TODO Consider discussing memory usage.
+%TODO Consider discussing memory usage.
 
@@
 \subsection{Random Restart First Improvement Hill Climber}
-TODO Need a citation either here or Section~\ref{sec-hillclimber}.
+%TODO Need a citation either here or Section~\ref{sec-hillclimber}.
 
@@
 applying FIHC, this comparison algorithm should perform even better
-than P3.  Like P3, FHIC-alone algorithm is Parameter-less, makes no
+than P3.  Like P3, the FHIC-alone algorithm is parameter-less, makes no
 assumptions about gene ordering, and will find the global optima given
@@
 compare against LTGA with an optimized population size, even though P3
-requires no such tuning. To set the population size parameter, we use
+receives no such tuning. To set the population size parameter, we use
 the bisection method~\cite{goldman:2012:ltga}, which determines the minimum population
@@
 
-Our final problem class is a subset of the Maximum Satisfiability (MAXSAT) problem.
+Our final problem class is a subset of the Maximum Satisfiability (MAX-SAT) problem.
 This problem consists of a series of disjunctive clauses, such that each clause contains
 three problem variables, some of which are negated. While
-MAXSAT requires exponential time to solve, we have devised a method of generating
+MAX-SAT requires exponential time to solve, we have devised a method of generating
 instances that ensures each has at least one solution that satisfies all clauses.
@@
 as well as individual run results and analysis scripts are available from our
-website\footnote{\url{https://name.of.site.removed/anonymous}}.  The results
+website.\footnote{\url{https://name.of.site.removed/anonymous}}  The results
 are summarized in Figure~\ref{fig-results} and Table~\ref{table-results}.
@@
 during crossover.  As crossover cannot decrease the fitness of a solution each
-crossover will have a non zero probability of improving subproblem solution quality.
-The probability of low fitness local optima for a subproblem reaching high levels
+crossover will have a greater than zero probability of improving subproblem solution quality.
+The probability of low fitness local optima for a subproblem reaching high pyramid levels
 is therefore low, which increases the probability of donors only having high quality
@@
 Currently, P3 is limited to single objective, noiseless, binary encoded problems.
-In future work we plan to extend (or show no modification is necessary) P3 to handle
+In future work we plan to extend (or show no modification is necessary for) P3 to handle
 problems from these classes.  Doing so will help make P3 more applicable to a wider

### 2014-01-15 Parameterless Population Pyramid/main.tex
abstract updated

@@
 %Maximum 200 Words
-  Real world applications of evolutionary techniques are often hindered by
-  the need to determine problem specific parameter settings.
-  While some previous methods have reduced or removed
-  the need for parameter tuning, many do so
-  by trading peak efficiency for general applicability.  The
-  Parameter-less Population Pyramid (P3) is an evolutionary technique
-  requiring \textbf{no} parameters which is still broadly applicable to
-  many problem types. P3 strikes a balance between continuous
-  integration of diversity with exploitative elitist operators,
-  allowing it to solve easy problems quickly and hard problems
-  eventually.  When compared with three state of the art optimization
-  techniques across seven different problem classes, P3 always finds
-  the optimum at least a constant factor \emph{\textbf{faster}} than
-  the rest.  Even when optimally tuning comparison technique
-  parameters, P3 has a lower evaluations to success growth complexity when test problems come
-  from a random distribution.  Beyond empirical results, we suggest
-  avenues for theoretical analysis of P3's expected runtime.  In
-  general, while these are only the first results for P3, we feel they
-  indicate wide applicability.
+  Real world applications of evolutionary techniques are often
+  hindered by the need to determine problem specific parameter
+  settings.  While some previous methods have reduced or removed the
+  need for parameter tuning, many do so by trading efficiency for
+  general applicability.  The Parameter-less Population Pyramid (P3)
+  is an evolutionary technique that both requires \textbf{no}
+  parameters and is broadly effective. P3 strikes a balance between
+  continuous integration of diversity with exploitative elitist
+  operators, allowing it to solve easy problems quickly and hard
+  problems eventually.  When compared with three optimally tuned,
+  state of the art optimization techniques, P3 always finds the
+  optimum at least a constant factor \emph{\textbf{faster}} as shown
+  on four benchmarks (Deceptive Trap, Deceptive Step Trap,
+  Hierarchical If and only If, Rastrigin).  More importantly, on three
+  randomized benchmarks (NK Landscapes, Ising Spin Glasses, MAX-SAT),
+  P3 has a lower order of computational complexity as measured by
+  evaluations.  Based on over 1 trillion evaluations, our results
+  suggest P3 has wide applicatiblity to a broad class of real world
+  problems.
 \end{abstract}

### 2014-01-15 Parameterless Population Pyramid/main.tex
Missing word
Added back in the "theory" part of the abstract.

@@
   state of the art optimization techniques, P3 always finds the
-  optimum at least a constant factor \emph{\textbf{faster}} as shown
-  on four benchmarks (Deceptive Trap, Deceptive Step Trap,
+  optimum at least a constant factor \emph{\textbf{faster}} across
+  four benchmarks (Deceptive Trap, Deceptive Step Trap,
   Hierarchical If and only If, Rastrigin).  More importantly, on three
@@
   P3 has a lower order of computational complexity as measured by
-  evaluations.  Based on over 1 trillion evaluations, our results
-  suggest P3 has wide applicatiblity to a broad class of real world
+  evaluations.
+  We also provide outlines for expected runtime analysis of P3,
+  setting the stage for future theory based conclusions.
+  Based on over 1 trillion evaluations, our results
+  suggest P3 has wide applicability to a broad class of
   problems.
@@
 climbing, and crossover, respectively.  In this example \#13 is not
-added to $P_2$ because it not a strict fitness improvement over
+added to $P_2$ because it is not a strict fitness improvement over
 \#12.  There is not enough information in $P_2$ to improve \#13

### 2014-01-17 Parameterless Population Pyramid/main.tex
Major wordsmithing to get down to 8 pages.

@@
   general applicability.  The Parameter-less Population Pyramid (P3)
-  is an evolutionary technique that both requires \textbf{no}
-  parameters and is broadly effective. P3 strikes a balance between
+  is an evolutionary technique that both requires \textbf{no parameters}
+  and is broadly effective. P3 strikes a balance between
   continuous integration of diversity with exploitative elitist
@@
   four benchmarks (Deceptive Trap, Deceptive Step Trap,
-  Hierarchical If and only If, Rastrigin).  More importantly, on three
+  HIFF, Rastrigin).  More importantly, on three
   randomized benchmarks (NK Landscapes, Ising Spin Glasses, MAX-SAT),
@@
 %\category{Computing Methodologies}{Artificial Intelligence}{Search Methodologies}
-\category{TODO}{TODO}{TODO}
+\category{I.2.8}{Artificial Intelligence}{Problem Solving, Control Methods, and Search}
 %A category including the fourth, optional field follows...
@@
 
-\terms{Algorithms}
+\terms{Algorithms, Performance, Experimentation}
 
@@
 world black box problems is a fundamental goal of evolutionary
-computation.  Achieving this goal often requires an optimizer to make
+computation.  Achieving this goal requires an optimizer to make
 assumptions about the nature of the problem it is solving as no
 optimizer can solve all problems equally well~\cite{Wolpert:1997:nfl}.
-
+Many genetic algorithms (GAs) introduce evolutionary parameters which provide
+user control of search assumptions.  These parameters can have enormous impact
+on search efficiency~\cite{goldberg:1991:gasize} and can have non-obvious optimal
+settings which require time consuming tuning to discover~\cite{grefenstette:1986:optimalga}.
+When applied to black box problems with non-trivial evaluation functions, the
+tuning process can be intractable, forcing users to rely on stock values or attempt
+a small number of test values.
+
+\begin{comment}
 Often built into the design of a genetic algorithm (GA) are
@@
 problems.
-
-These complications have encouraged work to remove the need for
-operator defined evolutionary parameters. In
+\end{comment}
+
+The problem of parameter tuning has encouraged work to remove parameters from
+user control in an effort to improve out-of-the-box GA quality. In
 an early example, a crossover based GA was devised which used
-concurrent racing populations to remove all user specified
-parameters~\cite{harik:1999:parameterlessga}.  This method of
-population racing has also been paired with more advanced evolutionary
+concurrent racing populations which removed all user specified
+parameters~\cite{harik:1999:parameterlessga}.
+Population racing has also been paired with more advanced evolutionary
 system, such as Hierarchical Bayesian
-Optimization~\cite{pelikan:2004:parameterlesshboa}. The claim was that
-while not optimal, this method ensured the GA would take no more than
-a logarithmic factor increase in evaluations beyond optimal.  Furthermore, doing so would
-ensure that these optimizers could be generally applicable without
-problem specific tuning.  Experimental evidence supports this theory,
-with some evidence that in general the bound is closer to a constant
+Optimization~\cite{pelikan:2004:parameterlesshboa}.
+Without the need to set a population size, racing ensures the GA will take no more than
+a logarithmic factor increase in evaluations beyond optimal.
+Some experimental evidence suggests the bound is closer to a constant
 factor worse than optimal.
@@
 In this work we shall introduce the Parameter-less Population Pyramid
-(P3).  P3 is an iterative method for constructing a \textbf{collection} of
+(P3).  P3 is an iterative method for constructing a collection of
 populations that requires no user parameters.  It leverages entropy based
@@
 experimentally appears to be at least a constant factor
-\emph{\textbf{improvement}} over comparable optimally configured
+\emph{\textbf{improvement}} over comparable, optimally configured,
 optimization methods.
@@
 \label{sec-p3}
-Before going into detail about how each aspect of P3 works, let us
-consider an example of P3 operation. Unlike many EC techniques, P3
+Unlike many EC techniques, P3
 does not have a single population of solutions.  Instead P3 maintains
@@
 to create solution \#4, which is likewise added to the population
-pyrimad at population level $P_0$.  With a population of two,
-crossover (Section~\ref{sec-crossover}) is applied to solutions \#2
-and \#4 creating solution \#5, which is an improvement. As \#5 used
+pyramid at population level $P_0$.
+Crossover (Section~\ref{sec-crossover}) is applied to solutions \#2
+and \#4 creating solution \#5. As \#5 used
 information from $P_0$ it is added to the next level of the pyramid at
@@
 
-For our purposes, we use a First Improvement Hill Climber
-(FIHC)\footnote{We use the name FHIC for an algorithm that likely has
-  various names in the literature, but use it here with a descriptive
-  name for clarity.}, a method of exploiting local information through
-a series of modify-evaluate-accept iterations starting from a randomly
-initialized solution and ending with an optimum, either local or global.
-FIHC uses the following algorithm.  Consider all gene loci
-$x_i$ in a random order.  For a given $x_i$ try all possible
-alternative values for the gene $x_i$, evaluating solution quality
-with each change.  If any alternative value is a strict fitness improvement
-over the current value for $x_i$, replace $x_i$ with the improved
+We use the First Improvement Hill Climber
+(FIHC),\footnote{We use the name FIHC for descriptive clarity.  This algorithm
+  likely already exists in the literature under various alternative names.} which acts as follows.
+%a method of exploiting local information through
+%a series of modify-evaluate-accept iterations starting from a randomly
+%initialized solution and ending with an optimum, either local or global.
+%FIHC uses the following algorithm.
+Consider all gene loci
+$x_i$ in a random order.  Evaluate the solution using all possible
+values of $x_i$. If any alternative value is a strict fitness improvement
+over the original value for $x_i$, replace $x_i$ with the improved
 value.  Iteration of the process continues until no such change results in
@@
 
+%TODO Consider removing lsbbo citation
 FIHC solves all functions of unitation in $O(N)$ time.  This is in contrast to
@@
 
-During cluster creation, P3 attempts to create a binary tree of clusters such that the leaves
+During cluster creation, P3 creates a binary tree of clusters such that the leaves
 represent individual gene loci and each internal node represent the subset of loci created
-by joining the subsets encoded in its children.  These clusters represent linked
-loci that have been detected in the genome, and are used when performing crossover.
+by joining the subsets encoded in its children.
+%These clusters represent linked
+%loci that have been detected in the genome, and are used when performing crossover.
 Using the table of observed frequencies described in Section~\ref{sec-pyramid}, Equation~\ref{eq-distance}
-efficiently estimates the linkage between clusters of loci using the average pairwise
+calculates the linkage ``distance'' between clusters of loci using the average pairwise
 entropy, as given in Equation~\ref{eq-entropy}.
@@
 
-Starting from the leaves, the binary tree can be constructed using an
-agglomerative method.  Each
-iteration determines the pairwise linkage between each cluster of loci
-using Equation~\ref{eq-distance}, and merges the two closest into a new
-cluster.  The process ends when there is only a single cluster
-containing all loci.  As our clustering algorithm is both convex and
-satisfies the swapping lemma~\cite{gronau:2007:upgma}, this algorithm
-requires $O(N^2)$ total time to correctly construct the linkage tree
-from the frequency table.
+Starting from the leaves, the binary tree is constructed agglomeratively. Each
+iteration the clusters $C_i$ and $C_j$ which minimize $D(C_i, C_j)$ are
+merged to form a new cluster $C_i \cup C_j$.
+The process ends when there is only a single cluster
+containing all loci.  This clustering algorithm
+requires $O(N^2)$ operations to construct the linkage tree
+from the frequency table~\cite{gronau:2007:upgma}.
 
@@
 create  a  new  individual.   We  further  remove  clusters  that  are
-completely  contained   within  some  other  cluster,   such  that  if
+completely  contained   within their direct superset,   such  that  if
 $D(C_i,C_j) = 0$, $C_i$  and  $C_j$ are  not used  when  performing
@@
 perfectly predict the gene values in $C_j$ and vice versa. From this we can
-conclude  that  the  settings  for  those two  loci  should  never  be
+conclude  that  the relative settings  for  those  loci  should  never  be
 disrupted.  When  used on particularly small populations  this has the
@@
 solutions agree and one for where they disagree.
-
-Given a useful set of clusters, we can now define how they are used.
+This means that Figure~\ref{fig-p3} is infeasible as crossover
+cannot make \#5 and \#9 unique from \{\#2, \#4\} and \{\#5, \#8\}, respectively.
+
 P3 uses crossover to recombine known information in a level of the
-pyramid with a candidate solution.  The algorithm is as follows. For
+pyramid with a candidate solution.  For
 each cluster select a random donor from the population.  Copy the
-donor's gene values for all loci in the cluster into the candidate
+donor's gene values for all loci in that cluster into the candidate
 solution.  Evaluate the candidate.  If the fitness of the candidate
@@
 is \emph{unaffected}.  This allows crossover to drift, potentially
-moving through plateaus.
+moving through fitness plateaus.
 
@@
 low quality solutions from effecting cluster formation for high
-quality solutions.  It also reduces the chance for high quality
-solutions to simply overwrite new, unoptimized solutions, preserving
+quality solutions.  It also reduces the chance that optimized
+solutions will overwrite newer solutions, preserving
 search diversity.
 
-  A major advantage to the design of P3 is that it makes very weak
+P3 makes very weak
 assumptions about the search landscape.  In many evolutionary
-  systems the crossover operator assumes that genes linked in the
-  search landscape are collocated in the genome.  This introduces an
-  initialization problem, changing the difficulty of the problem due
-  to a, potentially unknown, ordering required on the genome.  All of
+systems the crossover operator assumes linked genes
+are collocated in the genome.  This requires the user
+to configure genome structure based on gene linkage information they
+are unlikely to have. All of
 P3's operations ignore gene ordering, removing that level of
 configuration from the user.
+
+In general, P3 makes two assumptions about the search landscape:
+local structure and non uniform gene interdependence.
+For the hill climber to be effective, there should be at least some
+correlation between a solution's fitness and the fitness of its single
+bit neighborhood. Stated another way, this assumes that small changes
+in genotype result in small changes in phenotype. The non uniform gene
+interdependence assumption is required by the crossover operator to
+perform clustering.  Without this assumption, the clusters formed by
+crossover will detect spurious relations. The resulting donation of genes
+between solutions will be no better than random modification.  Both
+assumptions are likely to hold true for real world problems and are similarly
+assumed by most evolutionary optimizers.
 
@@
 generations.  This is true because in most systems, especially on
-deceptive problems, mutation rates are low enough as to make the
-creation of large, multi-gene substructures unlikely after
+deceptive problems, mutation rates are too low
+to create large, multi-gene substructures after
 initialization.  As such the generational model often results in a
-race between decreasing diversity and increasing fitness. This means the
-population size must be set large enough to contain sufficient initial
-diversity and to maintain that diversity during search for just long
-enough for all of it to be leveraged.
-
-Premature convergence of a population leads to the generational model
-often having two termination conditions: out of time or out of
+race between decreasing diversity and increasing fitness. 
+%This means the
+%population size must be set large enough to contain sufficient initial
+%diversity and to maintain that diversity during search for just long
+%enough for all of it to be leveraged.
+
+Premature convergence of a population causes the generational model
+to have two termination conditions: out of time or out of
 diversity.  In the former, if a smaller population size had been used,
-more of the diversity could have been exploited meaning a better
-solution likely could have been reached in the same number of
-evaluations.  In the latter, if a larger population size had been
+less evaluations could have been spent supporting diversity allowing more
+focus on exploitation.
+In the latter, if a larger population size had been
 used, more evaluations could have been performed increasing the
@@
 
-As P3 has exploited all known information each time the top of the pyramid is reached,
+In contrast, P3 has exploited all known information each time the top of the pyramid is reached;
 no evaluations are wasted  on unused diversity.  Similarly, because P3
 does not throw away previously optimized solutions while adding diversity, it avoids
-waste caused by premature convergence. It is this essential balance of
-exploitation to diversity that P3 focuses on.
-
-In general, P3 only makes two assumptions: local structure and non uniform gene interdependence.
-In order for P3 to be efficient there needs to be at least some correlation between a
-solution's fitness and the fitness of its single bit neighborhood.  This is required for
-the hill climber to function properly.  Under this assumption, each time the hill climber
-improves a solution it also improves the expected fitness of the neighborhood of the solution.
-Stated another way, this assumes that small changes in genotype result in small changes in phenotype.
-The non uniform gene interdependence assumption is required by the crossover operator to
-perform clustering.  Without this assumption, the clusters formed by crossover will
-detect spurious relations. The resulting donation of genes between solutions will be
-no better than random modification.  Both assumptions are likely to hold true for
-real world problems and are similarly assumed by most generational models.
-
-\subsection{Order of Complexity}
+waste caused by premature convergence.
+
+\subsection{Order of Complexity Per Evaluation}
 With the exception of crossover, P3 requires no more than $O(N)$ computations
 per evaluation, implying it is limited by the time required to evaluate a solution.
-In order to perform crossover, P3 must spend $O(N^2)$ time updating pairwise
+In order to perform crossover, P3 must spend $O(N^2)$ operations updating pairwise
 frequencies each time a solution is added to a level of the pyramid.  Similarly,
@@
 This means each time a solution is added to $P_0$, at least $O(N)$ evaluations
-must be performed.  This implies that the $N^2$ cost of the update to $P_0$ can
-be amortized over these evaluations, making this update $O(N)$ per evaluation.  Note
-that $N+1$ is a very weak lower bound, as this only occurs if the randomly generated
-individual is already at a local optimum.  Furthermore, only restarts that result
-in a unique individual cause updates to $P_0$.
+must be performed.  This implies that the $O(N^2)$ cost of the update to $P_0$ can
+be amortized over these evaluations, making this update $O(N)$ per evaluation.
+%Note
+%that $N+1$ is a very weak lower bound, as this only occurs if the randomly generated
+%individual is already at a local optimum.  Furthermore, only restarts that result
+%in a unique individual cause updates to $P_0$.
 
 To cause an update to $P_i$, all of the crossovers discovered for
-$P_{i-1}$ must first be applied.  Each causes their own evaluation.
+$P_{i-1}$ must first be applied, each causing their own evaluation.
 If P3 were to na\"ively use all crossovers and perform an evaluation
 each time, this would cause $O(N)$ evaluations.  Again, this would
-result in $O(N)$ amortized cost per update to $P_i$.  However,
+result in $O(N)$ amortized cost to update $P_i$.  However,
 Section~\ref{sec-crossover} discussed how P3 will ignore some clusters
 and will also skip an evaluation if no change to the solution was made
-by a crossover.  As such we are currently unable to state conclusively
-that updating $P_i$ will require $O(N)$ evaluations.  That said,
+by a cluster.  As such we are currently unable to state conclusively
+that updating $P_i$ will require $O(N)$ operations per evaluation.  That said,
 empirically it appears to be true.  Furthermore, it seems counterintuitive
@@
 define a complete optimization method based on the First Improvement
-Hill Climber defined in Section~\ref{sec-hillclimber}.  This
+Hill Climber from Section~\ref{sec-hillclimber}.  This
 definition is extended to match P3's style of restarting. Whenever the
@@
 
+To ensure maximal LTGA quality, we use the bisection method~\cite{goldman:2012:ltga}
+to minimally set LTGA's population size.  To be considered successful, a
+population size must result in 100 runs without premature convergence.
+According to the Rule of Three~\cite{jovanovic:1997:ruleofthree}, this bounds
+LTGA's failure rate to less than $\frac{3}{100+1} \approx 3\%$ with 95\% confidence.
+Note that runs used in setting the population are considered testing and were not
+used during analysis.  On randomized problems, training was performed on a disjoint
+set of problem instances.
+
+\begin{comment}
 Since LTGA requires a population size parameter, we thought it most fair to
@@
 classes, bisection is performed on a different set of instances than testing.
+\end{comment}
 
@@
 
+%TODO Consider removing this
 There have been other model based optimization techniques that use little to
@@
 As a final comparison we chose the $(1+(\lambda,\lambda))$ algorithm~\cite{doerr:2013:lambdalambda}
-which is currently the best, theory supported, crossover method.  Unlike P3 and LTGA,
+which is currently the best theory based crossover method.  Unlike P3 and LTGA,
 $(1+(\lambda,\lambda))$ does not use a population of solutions or any model building
@@
 when offspring of equal fitness are produced, we found this conflicted with our
-method of restarting.  As such we use the original control strategy, where $\lambda$
-is increased if the best offspring in a generation is not strictly better than its parent.
+method of restarting.  As such we use the original control strategy. %, where $\lambda$
+%is increased if the best offspring in a generation is not strictly better than its parent.
 If an offspring produced by crossover has identical genes to either of its parents,
@@
 All four of our optimization techniques share some interesting properties for
-problem equivalence.  First, all are fitness scale invariant, in that changing
-the fitness values without changing the relation between solution fitnesses does
-not effect any of their search methods.  This is true because none of the methods
+problem equivalence.  All are fitness scale invariant, in that changing
+the fitness values without changing the relation between solution fitnesses will not effect search.
+\begin{comment}This is true because none of the methods
 uses the fitness values directly, only when comparing the relative fitness between
@@
 methods, for instance fitness proportional selection.
-
-None of the methods have a bias toward creating specific gene values outside
-of the bias created by the problem itself.  This means all of the optimizers
+\end{comment}
+None of the methods have a bias toward creating specific gene values, in that
+XORing all solutions with a fixed string before evaluation will not effect search speed.
+\begin{comment}
+This means all of the optimizers
 will have unchanged behavior if a problem's inputs are reflected or XOR'd with a fixed string.  As
@@
 to the problem of minimizing the hamming distance to any arbitrary solution.
-
-All of the methods are gene order independent.  This means that the location
+\end{comment}
+All of the methods are gene order independent, in that the locations of genes
+in the genome can be shuffled without effecting search.
+\begin{comment}
+This means that the location
 of genes in the genome can be randomly shuffled without effecting search behavior.
@@
 probability they will be kept together through a crossover.
-
+\end{comment}
 Combined, these features mean that tests on apparently simple problems can be
@@
 
+\begin{comment}
 In practice it is often more important to find a good solution quickly than find
@@
 ``good enough'' solutions.
+\end{comment}
 
@@
 
-The classic example for this kind of test problem is the Deceptive Trap problem.
-For this benchmark the genome is broken up into non-overlapping problems of $k$
-genes, with the fitness of a single ``trap'' given by Equation~\ref{eq-deceptivetrap}.
+The classic example for this kind of test is the Deceptive Trap problem.
+For this benchmark the genome is broken up into non-overlapping ``traps'' of $k$
+genes, with the fitness of each trap given by Equation~\ref{eq-deceptivetrap}.
 \begin{equation}
@@
 \end{equation}
-In this equation $t$ represents the number of bits set in the trap.  For each trap
+In this equation $t$ represents the sum of gene values in the trap.  For each trap
 there is one local optima (all bits set to 0) and one global optima (all bits set to 1).
@@
 However, when using a hill climber and linkage learning, the Deceptive Trap problem
-may be trivially solved~\cite{goldman:2012:ltga}.  Thus a more
-difficult problem,  Deceptive Step Trap
+may be trivially solved~\cite{goldman:2012:ltga}.  Thus the more
+difficult Deceptive Step Trap
 problem was defined, and is given in Equation~\ref{eq-deceptivesteptrap}.
@@
 
+%TODO Could probably trim or cut pieces of this paragraph
 While both of these problems have difficult to solve subproblems, the non-overlapping
-nature of the solutions and their single optimal setting may not capture aspects of real world problem solving.
+nature of the solutions may not capture challenging aspects of real world problem solving.
 Therefore the Hierarchical If and only If (HIFF) problem is often used to test an
@@
 of specific instances.  This can improve the generality of results as experiments
-then predict the expected quality over the entire class even if only a subset
-of possible instances are actually used during experimentation.  Unlike single
+then predict the expected quality over the entire class.  Unlike single
 instance problems, the challenge with randomly generated problems is often how
@@
 
-Ising Spin Glasses are another large class of $NP-Hard$ problems.  These problems
-involve minimizing the interaction terms in a graph based on edge weights by changing
-the sign of graph nodes.  Similar to with NK landscapes, there is a polynomially solvable subset of Ising problems
+Ising Spin Glasses are another large class of NP-Hard problems: Given a graph
+describing the interactions between vertices, set the signs of all vertices to minimize
+interactions.
+Similar to NK landscapes, there is a polynomially solvable subset of Ising problems
 known as $2D\pm J$ Ising Spin Glasses.  This subset restricts graph interactions to a
@@
 
-Our final problem class is a subset of the Maximum Satisfiability (MAX-SAT) problem.
+Our final class is a subset of the Maximum Satisfiability (MAX-SAT) problem.
 This problem consists of a series of disjunctive clauses, such that each clause contains
@@
 MAX-SAT requires exponential time to solve, we have devised a method of generating
-instances that ensures each has at least one solution that satisfies all clauses.
-First, a random target solution is generated.  For each clause, set the sign of each
-variable in the clause such that that clause will be satisfied by the target solution.
-To ensure there is no bias to variable signs in the generated problem there is a $\frac{1}{6}$ probability
+only satisfiable instances.
+First, a random target solution is generated.  For each clause, set the signs of each
+term ensuring the target solution satisfies that clause.
+To ensure there is no bias to term signs in the generated problem there is a $\frac{1}{6}$ probability
 all clause signs match the target, a $\frac{1}{6}$ probability two of the three signs match,
-and a $\frac{4}{6}$ probability only one sign matches.  By setting the clause signs
-in this way we ensure the target solution is always able to satisfy all clauses,
-meaning all problems have a known best fitness.  Note that its possible there to be
+and a $\frac{4}{6}$ probability only one sign matches.  Note that its possible for there to be
 other solutions that also satisfy all clauses.  We chose a clause to variable ratio
@@
   \caption{Comparison of the median number of evaluations to reach the global optimum for
-           the four different optimization methods on seven benchmark problems with respect
+           the four different optimization methods with respect
            to problem size.  If the median run did not reach the global optimum no point
@@
 1.5 trillion evaluations.  Of these, approximately 100,000 runs were used to optimally
-set the population size of LTGA.  Note that these runs should be considered a training phase,
-and as such all reported results for LTGA use independent runs to create a testing phase.
-This was the only algorithm tested that received such tuning.  As such, LTGA should have a significant
+set the population size of LTGA.  Note that these runs are considered a training phase,
+and therefore all results for LTGA use independent runs to create a testing phase.
+This was the only algorithm that received tuning.  As such, LTGA should have a significant
 advantage over the others in that it has pre-learned the correct population size.
@@
 to handle deceptive landscapes.  The inability for the iterative hill climber to solve
-any of the problems indicates that P3 must be doing something more than just local search.
+any of the problems indicates that P3's success is more than just performing local search.
 
 Figure~\ref{fig-results} does provide one misleading
-exception for $(1+(\lambda,\lambda))$ on the deceptive trap problem, where it appears to
+exception for $(1+(\lambda,\lambda))$ on the Deceptive Trap problem, where it appears to
 level out at high genome sizes.  This is entirely an artifact of the extremely high
@@
 instance.  However, because it uses a fixed population size, it must set the that
-size large enough to ensure all random initializations of that population are
-likely to contain enough diversity to reach the global optimum.  By comparison, P3 is able to
+size large enough to ensure all random initializations of its population
+contain enough diversity to reach the global optimum.  By comparison, P3 is able to
 iteratively increase the population size, stopping once enough diversity has been generated.
@@
 are penalized.  P3 on the other hand grows with problem difficulty.  As such its
-median evaluation time better reflects the median difficult of a class, whereas LTGA's
+median evaluation time better reflects the median difficulty of a class, whereas LTGA's
 evaluation time reflects the maximum difficulty of a class.
 
-Of the runs where LTGA was successful in the median run, it failed to find the optimal solution
+Of the experiments where LTGA was successful in the median run, it failed to find the optimal solution
 109 times, or 0.8\%.  By comparison, P3 failed to find the optimum 3 times.  All three
@@
 
-Table~\ref{table-results} provides a close up look at the largest problem sizes
+Table~\ref{table-results} provides a detailed look at the largest problem sizes
 in which LTGA was able to find the global optimum reliably in less than 100 million
@@
 
-As a further verification that our tuning for P3 is valid, we refer
+As verification that our tuning for LTGA is valid, we refer
 to~\cite{thierens:2013:ltgahiff}, which tested the same variant of LTGA that we
@@
 search space without fitness plateaus will be solved by a single application
-of the hill climber.  Notably we expect a bound on Leading Ones of $O(N^2)$ time.
+of the hill climber.  Notably we expect a bound on Leading Ones of $O(N^2)$ evaluations.
 %TODO For future reference, n / factorial(n) + sum(i**2 / factorial(i + 1) for i in range(n))
@@
 
-Yet unlike many theory based algorithms, we believe P3 can achieve bounds on more
-complex and interesting landscapes.  Consider a sketch for a proof on the Deceptive
+Unlike many theory based algorithms, we believe P3 can achieve bounds on more
+complex and interesting landscapes.  Consider this proof sketch on the Deceptive
 Trap problem.  Each restart will require at most $2N$ evaluations to evaluate the
@@
 will move only complete traps between solutions, and once there is sufficient
-diversity in the population there will be a crossover for each individual trap.
+diversity in the population there will be a cluster for each individual trap.
 When performing a crossover of a single trap there can only be a single alternative
@@
 versa.  Therefore if $P_i$ contains the optimum solution for a trap in at least one
-individual and has a crossover that moves that trap, all solutions produced by crossing
+individual and has a cluster that identifies that trap, all solutions produced by crossing
 $P_i$ with a candidate solution will have the optimized value of that trap.  Extending to
@@
 version of a trap for all traps the very next crossover with $P_i$ will produce the global optimum.
-The only remaining portion of the proof is to determine how many evaluations are required to reach this point,
-which should be relatively trivial.
+The only remaining portion of the proof is to determine how many evaluations are required to reach this point.
 
@@
 
-This proof sketch can likely be extended to the HIFF problem, with minor modifications.
+This proof sketch can also be extended to the HIFF problem, with minor modifications.
 Each application of the hill climber ensures all subproblems of size 2 will be solved.
-Once $P_0$ has a sufficient number of solutions it will correctly cross all subproblems
+Once $P_0$ has a sufficient number of solutions it will correctly cross all subproblems of size 2
 independently, such that all future solutions crossed with $P_0$ will have all subproblems
@@
 This is even after LTGA was extensively tuned to each problem class, an unrealistic
-proposition on any real world problem.  A likely advantage is that P3 is able to
-automatically scale to problem difficulty, adding diversity only as the problem
-requires.
+proposition on any real world problem.
+%A likely advantage is that P3 is able to
+%automatically scale to problem difficulty, adding diversity only as the problem
+%requires.
 %TODO Consider reiterating stuff about gene location invariant, scaling invariant.
@@
 information before spending time increasing diversity makes it a great ``stop anytime''
-algorithm.  In future work, we plan to investigate in more detail P3's time to
-``good enough'' solutions, not just optimal.
+algorithm.
+%In future work, we plan to investigate in more detail P3's time to
+%``good enough'' solutions, not just optimal.
 
 While not yet complete, we feel that future work will be able to provide a strong
-theoretical foundation for the design of P3.  Here we have presented outlines
+theoretical foundation for the design of P3.
+\begin{comment}
+Here we have presented outlines
 for a number of expected runtime proofs, some of which apply to complex problems.
@@
 other state of the art techniques, such as LTGA.
-
-Currently, P3 is limited to single objective, noiseless, binary encoded problems.
+\end{comment}
+
+Currently, P3 is limited to single objective, noiseless, discrete valued problems.
 In future work we plan to extend (or show no modification is necessary for) P3 to handle
@@
 
+\begin{comment}
 Finally, as future work we plan to investigate the individual pieces of P3 for
@@
 pyramid scheme itself may combine well with other optimization techniques.
+\end{comment}
 

### 2014-01-20 Parameterless Population Pyramid/main.tex
Minor clarification
Minor fixes from Josh's suggestions

@@
 %TODO Consider removing lsbbo citation
-FIHC solves all functions of unitation in $O(N)$ time.  This is in contrast to
+FIHC solves all functions of unitation in $O(N)$ time, where $N$ is the number of
+genes.  This is in contrast to
 methods like Steepest Ascent Hill Climber~\cite{bosman:2011:lsbbo, goldman:2012:ltga}
@@
 Along with storing the solutions, each level also maintains a table of pairwise
-gene value frequencies.  This is also built constructively, such that initially
+gene value frequencies for use by crossover.  This is also built constructively, such that initially
 all frequencies are set to 0.  Each time a solution is added to a level, all
@@
 cannot make \#5 and \#9 unique from \{\#2, \#4\} and \{\#5, \#8\}, respectively.
+This inaccuracy was kept to make Figure~\ref{fig-p3} better express
+higher level concepts.
 

### 2014-01-20 Parameterless Population Pyramid/main.tex
Unitation -> linear

@@
 %TODO Consider removing lsbbo citation
-FIHC solves all functions of unitation in $O(N)$ time, where $N$ is the number of
+FIHC solves all linear functions in $O(N)$ time, where $N$ is the number of
 genes.  This is in contrast to
@@
 
-The traditional One Max class, as well as all functions of unitation, will be
+The traditional One Max class, as well as all linear functions, will be
 trivially solved by the hill climber in at most $N+1$ evaluations.  These come

### 2014-01-27 Parameterless Population Pyramid/main.tex
Minor tweaks.

@@
 a logarithmic factor increase in evaluations beyond optimal.
-Some experimental evidence suggests the bound is closer to a constant
-factor worse than optimal.
+While experimental evidence suggests racing is only a constant
+factor loss, this method explicitly trades efficiency for applicability.
 
@@
 to the population pyramid (Section~\ref{sec-pyramid}) at population
-level $P_0$.  With a population of 1 at $P_0$, there is insufficient information to continue,
+level $P_0$.  With a population of 1 at $P_0$, there is insufficient information to perform crossover,
 so solution \#3 is randomly generated and optimized via hill climbing
@@
 As mentioned, P3 maintains a pyramid like structure of populations
-with each level representing something akin to different generations
-of evolution. More specifically, a solution's level is one higher than
+with more optimized solutions in higher levels. More specifically, a solution's level is one higher than
 its highest parent, similar to the layering done
@@
 P3 uses gene value entropy to construct clusters of genes that should have their values
-conserved during crossover.  There are two primary steps in
+conserved during crossover.  There are two steps in
 defining this type of crossover: \emph{cluster creation} and \emph{cluster usage}.
@@
 P3 uses crossover to recombine known information in a level of the
-pyramid with a candidate solution.  For
+pyramid with a candidate solution as follows.  For
 each cluster select a random donor from the population.  Copy the
@@
 entropy calculations.  As cluster building and application ordering are not dependent on
-the position of loci in the genome, this operator introduces no positional bias in the search points
-evaluated.
+the position of loci in the genome, this operator introduces no positional bias on search.
 
@@
 are unlikely to have. All of
-P3's operations ignore gene ordering, removing that level of
-configuration from the user.
+P3's operations ignore gene ordering, allowing the user to ignore
+that aspect of configuration.
 
@@
 Each time a random restart occurs, at minimum $N+1$ evaluations are performed.
-This means each time a solution is added to $P_0$, at least $O(N)$ evaluations
+This means each time a solution is added to $P_0$, at least $\Omega(N)$ evaluations
 must be performed.  This implies that the $O(N^2)$ cost of the update to $P_0$ can
@@
 If P3 were to na\"ively use all crossovers and perform an evaluation
-each time, this would cause $O(N)$ evaluations.  Again, this would
+each time, this would cause $\Omega(N)$ evaluations.  Again, this would
 result in $O(N)$ amortized cost to update $P_i$.  However,
@@
 performing crossover P3 iteratively searches a population for a random
-donor which contains genes which differ from the current solution for
+donor with genes that differ from the current solution for
 the current crossover cluster.  In the worst case, this operation
@@
 \label{sec-ltga}
-The closest relative to P3 in existing literature is the Linkage Tree Genetic Algorithm
+The closest relative to P3 in the existing literature is the Linkage Tree Genetic Algorithm
 (LTGA).  There have been a number of different variations proposed to LTGA since
@@
 
-To ensure maximal LTGA quality, we use the bisection method~\cite{goldman:2012:ltga}
-to minimally set LTGA's population size.  To be considered successful, a
+To ensure LTGA is able to find the global optimum quickly, we use the bisection method~\cite{goldman:2012:ltga}
+to minimally set LTGA's population size parameter.  To be considered successful, a
 population size must result in 100 runs without premature convergence.
@@
 Other variants of LTGA have more closely resembled P3~\cite{goldman:2012:ltga},
-but as those variants are not in active use, we did not use them for comparison.
+but as those variants do not appear to be in active use, we did not use them for comparison.
 However, no LTGA variant has removed the population size parameter
 or fully addressed the problem of premature convergence.
-
 %TODO Consider removing this
-There have been other model based optimization techniques that use little to
+There have also been other model based optimization techniques that use few to
 no parameters.  Specifically LO-LIMD~\cite{posik:2011:parameterless} uses a
@@
 
-We also made minor modifications to selection and to prevent wasting evaluations.
+We also made minor modifications to improve selection and to prevent wasting evaluations, as follows.
 If the best offspring produced in a generation is a mutant offspring, select that
@@
 %is increased if the best offspring in a generation is not strictly better than its parent.
-If an offspring produced by crossover has identical genes to either of its parents,
+Finally, if an offspring produced by crossover has identical genes to either of its parents,
 it is not evaluated.
@@
 By enforcing only certain characteristics of a search space while allowing others
-to be generated randomly, we can perform tests on classes of problems instead
+to be generated randomly, we can perform tests on \emph{classes} of problems instead
 of specific instances.  This can improve the generality of results as experiments
@@
 term ensuring the target solution satisfies that clause.
-To ensure there is no bias to term signs in the generated problem there is a $\frac{1}{6}$ probability
+To ensure there is no bias to term signs in the generated problem, there is a $\frac{1}{6}$ probability
 all clause signs match the target, a $\frac{1}{6}$ probability two of the three signs match,
-and a $\frac{4}{6}$ probability only one sign matches.  Note that its possible for there to be
-other solutions that also satisfy all clauses.  We chose a clause to variable ratio
+and a $\frac{4}{6}$ probability only one sign matches.  Note that it's possible for there to be
+other solutions that also satisfy all clauses.  We chose the standard clause to variable ratio
 of $r=4.27$.
@@
            the four different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no point
-           is drawn.  Results given on a log-log scale.}
+           to problem size.  If the median run did not reach the global optimum no data element
+           is shown.  Results given on a log-log scale.}
   \label{fig-results}
@@
 benchmarks are deceptive in some sense, and neither of these techniques is designed
-to handle deceptive landscapes.  The inability for the iterative hill climber to solve
-any of the problems indicates that P3's success is more than just performing local search.
+to handle deceptive landscapes.  The inability for the iterative hill climber to quickly solve
+the problems indicates that P3's success is more than just performing local search.
 
@@
 variance of this algorithm on this problem.  Note that many of the large problem
-sizes failed to find the optimum for the majority of runs (no dot drawn for that size).
+sizes failed to find the optimum for the majority of runs (no data element shown for that size).
 
@@
 iteratively increase the population size, stopping once enough diversity has been generated.
-As such we would expect P3 to do better, but not a lot better.
+As such we would expect P3 to do better, but not a lot better than tuned LTGA.
 
@@
 resulted in $0.00025$.  As such these results are highly significant.  The amount
-of speedup ranged from 1.1 on Deceptive Step Trap to 942 on MAX-SAT, but in all
+of speedup as compared to optimally tuned LTGA ranged from 1.1 on Deceptive Step Trap to 942 on MAX-SAT, but in all
 cases P3 outperformed LTGA and all other comparison techniques.
@@
 of 92,129 evaluations.  The discrepancy is likely due to the cited work using
-their tuning runs in their average, unrealistically allowing each run to use a different tuned population size.
+their tuning runs in their average.
 
@@
 
-Using relaxed requirements, this can likely be extended to any arbitrary fully decomposable
+Using relaxed requirements, this can likely be extended to any arbitrary, fully decomposable,
 problem. Assuming that for sufficiently large population sizes the entropy between
@@
 is therefore low, which increases the probability of donors only having high quality
-solutions to subproblems.  Combined this is expected to create a runtime potentially
+solutions to subproblems.  Combined, this is expected to create a runtime potentially
 exponential in subproblem size, but polynomial in problem size.
 
-This proof sketch can also be extended to the HIFF problem, with minor modifications.
+This proof sketch can also be extended to the HIFF problem with minor modifications.
 Each application of the hill climber ensures all subproblems of size 2 will be solved.
@@
 While many parameter-less techniques trade the potential quality of optimal tuning
-for general effectiveness, P3 appears to robustly outperform all of the comparison
-state of the art techniques across a diverse set of benchmark problems.  This is
-because unlike other parameter-less techniques, P3 does not duplicate effort running
+for general effectiveness, P3 appears to robustly outperform all of the comparison,
+state of the art, techniques across a diverse set of benchmark problems.  This is
+because, unlike other parameter-less techniques, P3 does not duplicate effort running
 multiple populations, nor does it throw away learned information like many evolutionary
@@
 
-Even when compared against LTGA, the closest relative to P3 in current literature
-and current state of the art linkage learning algorithm, P3 surpasses expectation.
+Even when compared against LTGA, the closest relative to P3 in current literature,
+P3 performs better.
 This is even after LTGA was extensively tuned to each problem class, an unrealistic
@@
 P3's ability to efficiently optimize not just single problem instance classes,
-but randomly generated problem classes without parameter setting makes it an
-ideal candidate for real world application.  Its focus on exploiting all known
+but randomly generated problem classes, without parameter tuning makes it an
+ideal candidate for real world applications.  Its focus on exploiting all known
 information before spending time increasing diversity makes it a great ``stop anytime''
@@
 
-Currently, P3 is limited to single objective, noiseless, discrete valued problems.
+Currently, our results are limited to single objective, noiseless, discrete valued problems.
 In future work we plan to extend (or show no modification is necessary for) P3 to handle

### 2014-01-28 Parameterless Population Pyramid/main.tex
Word tweaks in abstract.

@@
   general applicability.  The Parameter-less Population Pyramid (P3)
-  is an evolutionary technique that both requires \textbf{no parameters}
-  and is broadly effective. P3 strikes a balance between
-  continuous integration of diversity with exploitative elitist
+  is an evolutionary technique that requires \textbf{no parameters}
+  and is still broadly effective. P3 strikes a balance between
+  continuous integration of diversity and exploitative elitist
   operators, allowing it to solve easy problems quickly and hard

### 2014-02-04 Beacon P3 Proposal/main.tex
Initial commit of BEACON funding proposal

% LaTeX Article Template - using defaults
\documentclass{article}
%\setlength{\topmargin}{-0.75in}         % move margin up 0.75in
%\setlength{\textwidth}{6.5in}      % text width 6.5in
%\setlength{\textheight}{9.25in}         % textlength 9.25in
%\setlength{\oddsidemargin}{0in}    % have left margin 1in over
%\setlength{\baselineskip}{0.1in}

\usepackage{graphicx} %images
\usepackage{listings} %code
\usepackage{amsmath} %pmatrix
\usepackage{comment} %multiline comments
\usepackage{algpseudocode} % creating proceedures
\usepackage{verbatim}
\usepackage{url}
\usepackage{fullpage}
\usepackage{epigraph}

\newcommand{\includegraphicsfit}[1]
{\includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{#1}}

\newcommand{\includegraphicssmall}[1]
{\includegraphics[width=.25\textwidth,height=.25\textheight,keepaspectratio]{#1}}

\def\signed#1{{\leavevmode\unskip\nobreak\hfil\penalty50\hskip2em
  \hbox{}\nobreak\hfil\raise-3pt\hbox{(#1)}%
  \parfillskip=0pt \finalhyphendemerits=0 \endgraf}}

\begin{document}
\begin{center}
\textbf{Real World Application of the Parameter-less Population Pyramid}
\end{center}

Real world applications of evolutionary optimization are often
hindered by the need to determine problem specific parameter
settings.  
Parameters can have a significant impact on search efficiency~\cite{goldberg:1991:gasize}
and finding good settings can be very time consuming~\cite{grefenstette:1986:optimalga}.
While some previous methods have reduced~\cite{Back:1992:selfadapt}
or removed~\cite{pelikan:2004:parameterlesshboa} the
need for parameter tuning, many do so by trading efficiency for
general applicability.  The Parameter-less Population Pyramid (P3)
is an evolutionary technique that requires \textbf{no parameters}
and is still broadly effective. P3 strikes a balance between
continuous integration of diversity and exploitative elitist
operators, allowing it to solve easy problems quickly and hard
problems eventually.  Work currently under review for publication
compared P3 with three optimally tuned state-of-the-art algorithms~\cite{thierens:2013:ltgahiff,doerr:2013:lambdalambda}.
Across seven problem classes P3 always found the
optimum at least a constant factor \emph{\textbf{faster}} than all comparisons.
More importantly, on the three most challenging problem classes,
P3 had a lower order of computational complexity as measured by
evaluations.  Based on over one trillion evaluations, these results
suggest P3 has wide applicability to a broad class of
problems.  The next step for P3 is to determine if those results carry over into real world
applications.  \textbf{Therefore, the primary focus of this project is to integrate
P3 into X-TOOLSS in order to (1) allow non-experts easy access to P3 and
(2) gather evidence of P3's ability to optimize real world problems.}

P3' driving principle is to make
minimal assumptions about the search landscape while maximally
exploiting gathered information.  As such, P3 does not use a fixed
population size.  Unlike most evolutionary optimization systems, P3 uses multiple
populations, with each population akin to a generation. 
The optimization process begins with a single randomly generated solution.
Having no other knowledge of the search landscape, P3 uses local search
to bring that solution to an optima.  This solution is then stored in $F_0$
(the first population).  Without making further assumptions
about the search landscape, P3 chooses another random solution to optimize using
local search, which is again stored in $F_0$.  Given two solutions in $F_0$, P3 can then
perform an entropy based optimal mixing crossover~\cite{bosman:2011:lsbbo}.  The
result of this mixing is then put into $F_1$.  As this new solution has nothing
to cross with, the process begins again at a random solution.  In this way new
solutions are added and crossed with existing levels of the population pyramid,
with new levels being created as needed.  This allows P3 to simultaneously exploit
all of its evaluations to produce a single high quality solution while avoiding convergence through
constant addition of new genetic material.

A challenge for all theory based optimization techniques is to make the transition
to solving meaningful problems.  While the field has progressed significantly beyond
its origins, many practitioners use outdated evolutionary systems~\cite{bechmann:2013:hummies}
or rely on methods tailored to their specific problem~\cite{izzo:2013:hummies}.
The fact that these suboptimal optimizers can already outperform human designers
suggests that bridging the gap between academia and industry may yield spectacular results.

There are two primary steps to making P3 an effective optimizer in practice: accessibility and adoption.
To achieve these goals we plan to integrate P3 into X-TOOLSS.\footnote{\url{http://nxt.ncat.edu/}}
X-TOOLSS is an open source optimization framework developed by Dr. Dozier's lab at North Carolina A\&T.
This framework consists of a number of canonical evolutionary optimization techniques, and is under
active use by engineers at NASA.  X-TOOLSS provides an easy to use interface which allows non-computer
scientists to apply evolutionary optimization to their domain specific problem.  Bundled with the
software are data collection and visualization tools which facilitate deeper understanding of
the optimization process and comparisons between optimization techniques.

To gauge the interest of X-TOOLSS users, with the help of Dr. Dozier, we asked Adam Burt, a NASA
Aerospace Engineer, to try out P3.  Here is his unedited response comparing P3 to his past work
using X-TOOLSS's basic evolutionary optimizers:

\begin{quote}
So I wanted to let you know that P3 appears to be doing incredibly well on the topology optimization problem.  Typically for 200ish elements it takes around 8-9000 function evaluations to find a good solution. P3 does it in roughly 600!!  I have already shared my enthusiasm with a co-worker of mine who has also done GA based topology optimization and he seemed fairly impressed as well.  I think you might really have something here.  I am going to do a few paper searches to see if I can find other authors who have done similar problems to see how many function evaluations they have on average, but I feel like 600 ish is very small for a GA.  My co-worker was also pretty excited about the prospect of a tuneless GA, I believe that will have a lot of benefit in the engineering application world. \signed{Adam Burt, NASA Aerospace Engineer}
\end{quote}

Adam has graciously worked with us to perform these preliminary experiments, but more general
release is currently hindered by the expert centric design of P3's software.  Even now that Adam
knows how to use our software, it requires him to have two separate tool sets for optimization.
Therefore we feel it is imperative to integrate P3 into X-TOOLSS to make it more user friendly,
allowing more engineers to benefit from P3's quality.

Beyond accessibility, X-TOOLSS provides an invaluable potential to P3 for academic data collection.
The framework's optimization analysis tools have been refined through continuous multi-user use, and
will likely lead to further insights into P3's optimization.  Furthermore, as X-TOOLSS has a number
of built in evolutionary optimizers, it provides a perfect device for gathering comparative results
as P3 is applied to new and challenging problems.  Once integrated, non-computer scientists can utilize
X-TOOLSS to compare P3 directly with existing evolutionary techniques.  This can provide the foundation
for collaborative research between industry and academia, with industry gaining high quality solutions
to challenging problems and academia gathering data on optimization techniques.

%\includegraphicsfit{WithRast}

\subsubsection*{Intellectual Merit}
%Proposed work is highly innovative and has potential for having a ground-breaking impact in its field
%Project aligns with BEACON mission and shows promise of opening new areas for multidisciplinary investigation
%Project appears very likely to generate successful (funded) proposals
%Degree of Multidisciplinarity of Project - No chance

All previous parameter-less methods in evolutionary computing perform at least a constant factor
worse than when using methods with optimal problem specific parameter tuning~\cite{pelikan:2004:parameterlesshboa}.
What makes P3 so interesting as that in all of our preliminary data it has performed at
least a constant factor \textbf{better} than optimally tuned state-of-the-art algorithms.
However, these results are limited to artificial problems.  In order to perform a complete analysis,
P3 needs to be applied to real world optimization problems.  Through X-TOOLSS and 
with the help of engineers at NASA, this
analysis can be performed with relative ease.

Once collected, P3's ability to tackle real problems will provide either further evidence for its usefulness
to people outside of the evolutionary optimization community or as a means for further improving P3.
In either case it will help drive the direction of future research in parameter-less evolutionary optimization.

\subsubsection*{Broader Impacts}
%Education - Not likely
%Project is organized around solution of a problem identified by a particular company/group of companies as important to them
%Some diversity reflected among project participants (current or named for future appointment under project)
%Project participants are from two or more BEACON institutions
Once P3 has been integrated into X-TOOLSS, its parameter-less nature will allow
end users (specifically NASA engineers) to try P3 on their optimization problems with ease.
Our preliminary results suggest P3 will produce better quality solutions in less time
than existing X-TOOLSS optimizers.

Beyond the direct benefits gained by NASA and X-TOOLSS users, better understanding of P3
can improve evolutionary optimization in general.  As mentioned previously, evolutionary
methods are already able to produce human competitive designs, with enough successes each
year to hold an annual competition.\footnote{\url{http://www.genetic-programming.org/combined.php}}
These results impact everything from space exploration to cancer screenings, software error correction
to protein structure prediction.

\newpage
\bibliographystyle{abbrv}
\bibliography{../main}
\end{document}

### 2014-02-07 Beacon P3 Proposal/main.tex
Added GE to BEACON proposal.

@@
 suggest P3 has wide applicability to a broad class of
-problems.  The next step for P3 is to determine if those results carry over into real world
-applications.  \textbf{Therefore, the primary focus of this project is to integrate
-P3 into X-TOOLSS in order to (1) allow non-experts easy access to P3 and
-(2) gather evidence of P3's ability to optimize real world problems.}
+problems.  \textbf{Therefore, the primary focus of this project is to integrate
+P3 into X-TOOLSS in order to (1) allow non-experts easy access to P3,
+(2) improve P3's ability to optimize real world problems and (3) expand
+P3 to new evolutionary domains utilizing Genetic Programming.}
 
@@
 for collaborative research between industry and academia, with industry gaining high quality solutions
-to challenging problems and academia gathering data on optimization techniques.
+to challenging problems and academia gathering data on how to improve optimization techniques.
 
 %\includegraphicsfit{WithRast}
+
+P3 is currently limited to evolving discrete value encoded linear genomes. 
+Therefore we propose to leverage Grammatical Evolution~\cite{oneill:2001:grammatical}
+to give P3 the ability to evolve entire programs.  Grammatical Evolution (GE) is a subset
+of Genetic Programming which utilizes a non-trivial genotype to phenotype mapping, converting
+binary strings into full fledged programs.  To apply GE, a domain expert first creates a
+grammar which defines the structure of all possible useful programs.  When evaluating a
+solution, bits are read from the individual to determine which grammar expansion rules to perform,
+deterministically creating a program.  This is an ideal transition for P3 as from an
+evolutionary standpoint GE applications will be no different than binary string applications as
+only the evaluation step will have to change.
 
@@
 In either case it will help drive the direction of future research in parameter-less evolutionary optimization.
+Expanding P3 to evolve entire programs will also provide interesting insight into the challenging domain
+of Genetic Programming.
 
@@
 These results impact everything from space exploration to cancer screenings, software error correction
-to protein structure prediction.
+to protein structure prediction.  Evolution of programs can help P3 create truly innovative solutions to complex problems.
 

### 2014-02-10 Beacon P3 Proposal/main.tex
update proposal

@@
 Across seven problem classes P3 always found the
-optimum at least a constant factor \emph{\textbf{faster}} than all comparisons.
+optimum at least a constant factor \emph{\textbf{faster}} than the
+comparison algorithms.
 More importantly, on the three most challenging problem classes,
@@
 suggest P3 has wide applicability to a broad class of
-problems.  \textbf{Therefore, the primary focus of this project is to integrate
-P3 into X-TOOLSS in order to (1) allow non-experts easy access to P3,
-(2) improve P3's ability to optimize real world problems and (3) expand
-P3 to new evolutionary domains utilizing Genetic Programming.}
+problems.  \textbf{Therefore, the primary focus of this project is to:
+(1) integrate P3 into X-TOOLSS so it may be applied to
+new, practical problems, (2) use feedback from P3 X-TOOLSS applications
+to improve P3 and (3) expand
+P3's representation to Genetic Programming and explore its
+applicability to different kinds of problems.}
 
@@
 
-A challenge for all theory based optimization techniques is to make the transition
+A challenge for many optimization techniques is to make the transition
 to solving meaningful problems.  While the field has progressed significantly beyond
@@
 
-Adam has graciously worked with us to perform these preliminary experiments, but more general
-release is currently hindered by the expert centric design of P3's software.  Even now that Adam
-knows how to use our software, it requires him to have two separate tool sets for optimization.
+Adam has graciously worked with us to perform these preliminary
+experiments, but P3's present form requires more work to setup a
+problem than is likely tolerable by a typical domain user. 
+%more general
+%release is currently hindered by the expert centric design of P3's software.  Even now that Adam
+%knows how to use our software, it requires him to have two separate tool sets for optimization.
 Therefore we feel it is imperative to integrate P3 into X-TOOLSS to make it more user friendly,
-allowing more engineers to benefit from P3's quality.
+allowing more engineers to benefit from P3's capabilities.
 
-Beyond accessibility, X-TOOLSS provides an invaluable potential to P3 for academic data collection.
+Beyond accessibility, X-TOOLSS provides an invaluable potential for data collection.
 The framework's optimization analysis tools have been refined through continuous multi-user use, and
-will likely lead to further insights into P3's optimization.  Furthermore, as X-TOOLSS has a number
+will likely lead to further insights into P3's performance and applicability.  Furthermore, as X-TOOLSS has a number
 of built in evolutionary optimizers, it provides a perfect device for gathering comparative results
-as P3 is applied to new and challenging problems.  Once integrated, non-computer scientists can utilize
-X-TOOLSS to compare P3 directly with existing evolutionary techniques.  This can provide the foundation
+as P3 is applied to new problems.  Once integrated, domain experts can utilize
+X-TOOLSS to compare P3 directly with existing evolutionary techniques.
+This will provide a foundation
 for collaborative research between industry and academia, with industry gaining high quality solutions
@@
 
-P3 is currently limited to evolving discrete value encoded linear genomes. 
-Therefore we propose to leverage Grammatical Evolution~\cite{oneill:2001:grammatical}
-to give P3 the ability to evolve entire programs.  Grammatical Evolution (GE) is a subset
+P3 is currently limited to evolving discrete value encoded linear
+genomes which also limits the range of problems the approach can address.
+We therefore also propose to examine ways to expand P3 into the
+Genetic Programming area. We plan to leverage Grammatical Evolution~\cite{oneill:2001:grammatical}
+to give P3 the ability to evolve programs.  Grammatical Evolution (GE) is a subset
 of Genetic Programming which utilizes a non-trivial genotype to phenotype mapping, converting
@@
 solution, bits are read from the individual to determine which grammar expansion rules to perform,
-deterministically creating a program.  This is an ideal transition for P3 as from an
-evolutionary standpoint GE applications will be no different than binary string applications as
-only the evaluation step will have to change.
+deterministically creating a program.  This is an ideal transition for
+P3 since GE applications will be appear as binary string
+applications requiring 
+only modificiation to the evaluation step.
 

### 2014-02-12 CV/resume.tex
Added resume

%TODO Add: Voted Leader of the year
%______________________________________________________________________________________________________________________
% @brief    LaTeX2e Resume based on the example from Kamil K Wojcicki
\documentclass[margin,line]{resume}
\usepackage{verbatim}

%______________________________________________________________________________________________________________________
\begin{document}
\name{\Large Brian W. Goldman}
\begin{resume}

    %__________________________________________________________________________________________________________________
    % Contact Information
    \section{\mysidestyle Contact\\Information}

    6195 Gossard                            \hfill phone: (314) 313-1281          \vspace{0mm}\\\vspace{0mm}%
    East Lansing, Michigan, 48823                          \hfill email: brianwgoldman@acm.org          %\vspace{0mm}%\\\vspace{0mm}%
    %__________________________________________________________________________________________________________________
    % Research Interests
    \section{\mysidestyle Research\\Interests}
    Evolutionary Computation, Genetic Programming, Artificial Intelligence, Machine Learning

    %__________________________________________________________________________________________________________________
    % Education
    \section{\mysidestyle Education}
    \textbf{Michigan State University}, East Lansing, Michigan \vspace{2mm}\\\vspace{1mm}%
    \textsl{Doctor of Philosophy in Computer Science \& Engineering} \hfill \textbf{ August 2012 -- Present}\vspace{-3mm}\\\vspace{-1mm}%
    \begin{list2}
        \item Thesis Advisor:  Dr. William F. Punch
        \item GPA: 4.0/4.0
        \item Expected Graduation: May 2015
    \end{list2}\vspace{-1.5mm}
    \textbf{Missouri S\&T}, Rolla, Missouri \vspace{2mm}\\\vspace{1mm}%
    \textsl{Master of Science in Computer Science} \hfill \textbf{ August 2010 -- May 2012}\vspace{-3mm}\\\vspace{-1mm}%
    \begin{list2}
        \item Thesis Advisor:  Dr. Daniel R. Tauritz
        \item GPA: 4.0/4.0
        \item Summa Cum Laude
    \end{list2}\vspace{-1.5mm}
    \textsl{Bachelor of Science in Computer Science} \hfill \textbf{ August 2006 -- July 2010}\vspace{-3mm}\\\vspace{-1mm}%
    \begin{list2}
        \item Minor in Mathematics
        \item GPA: 3.84/4.0 in major, 3.81/4.0 overall
        \item Summa Cum Laude
    \end{list2}\vspace{-1.5mm}

    %__________________________________________________________________________________________________________________
    % Research
    \section{\mysidestyle Research\\Experience}

    %----------------------CGP------------------
    \textsl{Cartesian Genetic Programming} \hfill \textbf{August 2012 -- August 2013}\\
    Examined how representation and evolutionary operators interacted to cause degenerate
    optimization behavior, leading to the proposal of improvements.  Presented at EuroGP 2013
    in Vienna Austria, GECCO 2013 in Amsterdam, The Netherlands.  Revised and resubmitted to
    IEEE Transactions on Evolutionary Computation, awaiting final review.

    \vspace{-2mm}
    B. W. Goldman and W. F. Punch,
    ``Length Bias and Search Limitations in Cartesian Genetic Programming,''
    \textsl{In Proceedings of the 15th Annual Conference on Genetic and Evolutionary Computation}, pp. 933--940, 2013

    \vspace{-2mm}
    B. W. Goldman and W. F. Punch,
    ``Reducing Wasted Evaluations in  Cartesian Genetic Programming,''
    \textsl{In Proceedings of the 16th European Conference on Genetic Programming}, pp. 61--72, 2012

    %---------------------Benchmarks--------------------------------
    \textsl{Improving Genetic Programming Research} \hfill \textbf{August 2012 -- December 2012}\\
    Utilizing a survey of genetic programming researches and in honor of the fields twentieth anniversary,
    proposed consensus suggestions for how to raise the quality of future publications.

    \vspace{-2mm}
    D. R. White, J. McDermott, M. Castelli, L. Manzoni, B. W. Goldman,
    G. Kronberger, W. Jaskowski, U.-M. O’Reilly, and S. Luke,
    ``Better GP Benchmarks: Community Survey Results and Proposals,''
    \textsl{Genetic Programming and Evolvable Machines}, vol. 14, no. 1, pp. 3--29, 2013.

    %----------------------LTGA-------------------------------
    \textsl{Linkage Tree Genetic Algorithm} \hfill \textbf{January 2012 -- May 2012}\\
    Performed a deep analysis of the multiple variations of this recently created automatic linkage
    discovery technique, with suggestions for improved methods of testing. Presented at GECCO 2012 in Philadelphia, Pennsylvania.

    \vspace{-2mm}
    B. W. Goldman and D. R. Tauritz,
    ``Linkage tree genetic algorithms: variants and analysis,''
    \textsl{In Proceedings of the 14th Annual Conference on Genetic and Evolutionary Computation}, pp. 625--632, 2012

    \textsl{Supportive Coevolution} \hfill \textbf{January 2011 -- May 2012}\\
    A new form of coevolution in which the primary species directly encodes prospective problem solutions,
    and is supported by other species tasked with dynamically evolving better ways to evolve the primary species.
    Presented at GECCO 2012 in Philadelphia, Pennsylvania.

    \vspace{-2mm}
    B. W. Goldman and D. R. Tauritz,
    ``Supportive Coevolution,''
    \textsl{In Proceedings of the 14th Annual Conference Companion on Genetic and Evolutionary Computation}, pp. 59--66, 2012.

    \pagebreak

    \textsl{Self-Configuring Crossover} \hfill \textbf{January 2010 -- June 2011}\\
    Dynamically evolving novel crossover operators during evolutionary algorithm optimization 
    using linear genetic programming and self-adaptation.  Presented at GECCO 2011 in Dublin, Ireland.

    \vspace{-2mm}
    B. W. Goldman and D. R. Tauritz,
    ``Self-Configuring Crossover,''
    \textsl{In Proceedings of the 13th Annual Conference Companion on Genetic and Evolutionary Computation}, pp. 575--582, 2011.

    \textsl{Dynamic Parameters} \hfill \textbf{January 2010 -- February 2011}\\
    Dynamically changing evolutionary algorithm parameters during run time can achieve better solution quality
    than optimally set static parameters for a wide variety of parameters.
    Presented at GECCO 2011 in Dublin, Ireland.

    \vspace{-2mm}    
    B. W. Goldman and D. R. Tauritz,
    ``Meta-Evolved Empirical Evidence of the Effectiveness of Dynamic Parameters'',
    \textsl{In Proceedings of the 13th Annual Conference Companion on Genetic and Evolutionary Computation}, pp. 155--156, 2011.

    %__________________________________________________________________________________________________________________
    % Mentoring Experience
    \section{\mysidestyle Mentoring\\Experience}

    \textsl{Extending Self-Configuration} \hfill \textbf{August 2012 -- present}\\
    Mentoring undergraduate student research project investigating ways to improve evolutionary
    self configuration and expanding to new operators.

    \vspace{-2mm}
    N. R. Kamrath, B. W. Goldman, and D. R. Tauritz,
    ``Using Supportive Coevolution to Evolve Self-Configuring Crossover,''
    \textsl{In Proceedings of the 15th Annual Conference Companion on Genetic and Evolutionary Computation}, pp. 1489--1496, 2013.

    %__________________________________________________________________________________________________________________
    % Teaching Experience
    \section{\mysidestyle Selected\\Teaching\\Experience}
    \textbf{Michigan State University}, East Lansing, Michigan \vspace{2mm}\\\vspace{1mm}%
    \textsl{CSE232: Introduction to Programming II} \hfill \textbf{August 2013 -- May 2014}\\
    Primary instructor for lab section,
    topics include the C++ programming language up to dynamic memory, templating, and user created data structures.

    \textbf{Missouri S\&T}, Rolla, Missouri \vspace{2mm}\\\vspace{1mm}%
    \begin{comment}
    \textsl{CS328: Object Oriented Numerical Methods} \hfill \textbf{January 2012 -- present}\\
    Grader for intensive C++ course,
    topics include advanced programming techniques and how to efficiently solve large systems of equations.
    \end{comment}
    \textsl{CS54: Introduction to C++ Lab} \hfill \textbf{January 2011 -- December 2011}\\
    Primary instructor for introductory computer science course,
    topics include programming basics up to inheritance, style, editors, and how to interact with Unix.
    \begin{comment}
    \textsl{CS387: Parallel Computing} \hfill \textbf{January 2011 -- May 2011}\\
    Teaching assistant for parallel computation course,
    topics include MPI, cluster computing, GPUs, Monte Carlo experiments, and high performance computing.
    \end{comment}

    \textsl{CS348: Evolutionary Computation} \hfill \textbf{August 2010 -- December 2010}\\
    Teaching assistant for evolutionary computation course,
    topics include genetic algorithms, evolutionary strategies, genetic programming, and learning classifier systems.

    %__________________________________________________________________________________________________________________
    % Professional Experience
    \section{\mysidestyle Recent\\Professional\\Experience}

    \textbf{Los Alamos National Laboratories}, Los Alamos, New Mexico \vspace{2mm}\\\vspace{1mm}%
    \textsl{Advanced Computing Solutions Program} \hfill \textbf{May 2012 -- August 2012}\\
    Investigated intrusion detection techniques using automated network behavior analysis.

    \textbf{Sandia National Laboratories}, Albuquerque, New Mexico \vspace{2mm}\\\vspace{1mm}%
    \textsl{Center for Cyber Defenders Graduate Student Intern} \hfill \textbf{May 2011 -- August 2011}\\
    Developed covert command and control communication methods, created security audit tools for a
    surveillance camera system, and designed an educational game to teach secure system design.

    \textbf{Dynetics}, Huntsville, Alabama \vspace{2mm}\\\vspace{1mm}%
    \textsl{Computer Analyst 3} \hfill \textbf{May 2010 -- August 2010}\\
    Worked with department of defense applications to improve unmanned aerial vehicle data analysis.

    \begin{comment}
    \textbf{AT\&T}, Saint Louis, Missouri \vspace{2mm}\\\vspace{1mm}%
    \textsl{Summer Technical Intern} \hfill \textbf{May 2008 -- August 2008}\\
    Developed automated service detection to ensure all pertinent applications installed on a server
    were restarted at server reboot.  Modernized legacy software performing phone number database queries.
    \end{comment}

    %__________________________________________________________________________________________________________________
    % Honors and Awards
    \section{\mysidestyle Honors and\\Activities} 
    Program Committee Member, GECCO 2012--present Genetic Algorithms Track \vspace{1mm}\\%
    Voted 2012 Leader of the Year, Missouri S\&T Computer Science Department \vspace{1mm}\\%
    %Chair of ACM SIG for AI competition game development (SIG-Game), 2011--present \vspace{1mm}\\%
    Google AI challenge, ranked 25$^{th}$ out of 4619 world wide, 6$^{th}$ in USA, 2010 \vspace{1mm}\\%
    Missouri S\&T Human versus Computer chess tournament, two first place AIs, 2009 and 2011 \vspace{1mm}\\%
\begin{comment}    
    ACM SIG-Game Competition testing and balancing using evolving AI, 2009--present \vspace{1mm}\\%
    ACM SIG-Game first place AI, 2008          \vspace{1mm}\\%
    University of Illinois at Urbana-Champaign MechMania AI competition second place AI, 2007  \vspace{1mm}\\%
    Missouri S\&T Alumni Scholarship 2006--2010             %\vspace{1mm}\\%

    %__________________________________________________________________________________________________________________
    % Computer Skills
    \section{\mysidestyle Programming} 
    Python, C, C++, C\#, Matlab, Linux shell scripting, \LaTeX, Java, CUDA.

%______________________________________________________________________________________________________________________
\section{\mysidestyle References}

\begin{tabular}{@{}p{6cm}p{6cm}}
\textbf{Dr. Daniel R. Tauritz} (Advisor)       &  \textbf{Professor Fikret Ercal}                   \\
Associate Professor                &  Professor                       \\
Department of Computer Science     & Department of Computer Science \\
Missouri S\&T                        &  Missouri S\&T                      \\
Rolla, Missouri                      &  Rolla, Missouri        \\
phone: \textsl{(573) 341-7218}    &  phone: \textsl{(573) 341-4857}     \\
e-mail: \textsl{tauritzd@mst.edu}   &  e-mail: \textsl{ercal@mst.edu}    \\
\end{tabular}

\begin{tabular}{@{}p{6cm}p{6cm}}
\textbf{Mr. Dan Thomsen}                \\%&  \textbf{Mr Sean Loye}                    \\
Senior Scientist                        \\      %&  Systems Engineer                         \\
Sandia National Laboratories            \\      %&  Hewlett Packard                          \\
Albuquerque, New Mexico                 \\ %&  Milton, Queensland, Australia            \\
phone: \textsl{(612) 789-0559}    \\%&  phone: \textsl{available on request}     \\
e-mail: \textsl{available on request}   \\%&  e-mail: \textsl{available on request}    \\
\end{tabular}

\end{comment}
%______________________________________________________________________________________________________________________
\end{resume}
\end{document}

%______________________________________________________________________________________________________________________
% EOF

### 2014-02-12 Beacon P3 Proposal/main.tex
Removed some verticle space to make it fit on 2 pages.
Minor word fixes to the BEACON proposal.

@@
 
-P3 is currently limited to evolving discrete value encoded linear
-genomes which also limits the range of problems the approach can address.
+P3 is designed to evolve discrete value encoded linear
+genomes which limits the range of problems this approach can address.
 We therefore also propose to examine ways to expand P3 into the
@@
 
+\vspace{-2.5mm}
 \subsubsection*{Intellectual Merit}
@@
 worse than when using methods with optimal problem specific parameter tuning~\cite{pelikan:2004:parameterlesshboa}.
-What makes P3 so interesting as that in all of our preliminary data it has performed at
+What makes P3 so interesting is that in all of our preliminary data it has performed at
 least a constant factor \textbf{better} than optimally tuned state-of-the-art algorithms.
@@
 
+\vspace{-2.5mm}
 \subsubsection*{Broader Impacts}

### 2014-03-18 Parameterless Population Pyramid/main.tex
Minor fixes to P3, added in high level pseudocode

@@
 \usepackage{verbatim}
+\usepackage[noend]{algpseudocode}
 \newcommand{\includegraphicsfit}[1]
@@
 Providing a good solution in reasonable time to a broad class of real
-world black box problems is a fundamental goal of evolutionary
+world black-box problems is a fundamental goal of evolutionary
 computation.  Achieving this goal requires an optimizer to make
@@
 on search efficiency~\cite{goldberg:1991:gasize} and can have non-obvious optimal
-settings which require time consuming tuning to discover~\cite{grefenstette:1986:optimalga}.
-When applied to black box problems with non-trivial evaluation functions, the
+settings which require time-consuming tuning to discover~\cite{grefenstette:1986:optimalga}.
+When applied to black-box problems with non-trivial evaluation functions, the
 tuning process can be intractable, forcing users to rely on stock values or attempt
@@
 user control in an effort to improve out-of-the-box GA quality. In
-an early example, a crossover based GA was devised which used
+an early example, a crossover-based GA was devised which used
 concurrent racing populations which removed all user specified
@@
 \label{sec-p3}
-Unlike many EC techniques, P3
+Unlike many GA techniques, P3
 does not have a single population of solutions.  Instead P3 maintains
-a pyramid like structure, such that each level of the pyramid has a
+a pyramid-like structure, such that each level of the pyramid has a
 population of solutions, with each level representing something akin
@@
 has no information about the landscape so solution \#1 is randomly generated.  Solution \#1
-is brought to a local optima using a hill climber
+is brought to a local optimum using a hill climber
 (Section~\ref{sec-hillclimber}) creating solution \#2, which is added
@@
 \begin{figure}
+  \begin{algorithmic}
+  \Procedure{Iterate-P3}{}
+    \State Create random solution
+    \State Apply hill climber (Section~\ref{sec-hillclimber})
+    \If{solution $\notin$ pyramid}
+      \State Add solution to $P_0$ (Section~\ref{sec-pyramid})
+    \EndIf
+
+    \ForAll{$P_i \in pyramid$}
+      \State Mix solution with $P_i$ (Section~\ref{sec-crossover})
+      \If{solution fitness has improved}
+        \If{solution $\notin$ pyramid}
+          \State Add solution to $P_{i+1}$
+        \EndIf
+      \EndIf
+    \EndFor
+  \EndProcedure
+\end{algorithmic}
+  \caption{One iteration of P3 optimization.}
+  \label{fig:psuedo}
+\end{figure}
+
+\begin{figure}
   \centering
@@
 
-The levels represent independent sets of unique solutions, such that
+The levels represent disjoint sets of unique solutions, such that
 $\forall_{i, j \in P, i \neq j} P_i \cap P_j = \emptyset$.  This relationship
@@
 Finally, the use of levels allows P3 to perform solution filtering
-without throwing away information.  Once found, a local optima is
+without throwing away information.  Once found, a local optimum is
 stored indefinitely, preserving both its linkage and genes for future
@@
 than P3.  Like P3, the FHIC-alone algorithm is parameter-less, makes no
-assumptions about gene ordering, and will find the global optima given
+assumptions about gene ordering, and will find the global optimum given
 a sufficient (potentially exponential) number of evaluations.
@@
 This algorithm is also the current state of the art, outperforming other algorithms
-in black box optimization across numerous benchmarks.
+in black-box optimization across numerous benchmarks.
 
@@
 the bisection method~\cite{goldman:2012:ltga}, which determines the minimum population
-size required to meet a specified success criteria.  As P3 will run until the global optima
+size required to meet a specified success criteria.  As P3 will run until the global optimum
 is found or the evaluation limit is reached, we want to ensure LTGA has a similar requirement
@@
 problem equivalence.  All are fitness scale invariant, in that changing
-the fitness values without changing the relation between solution fitnesses will not effect search.
+the fitness values without changing the relation between solution fitnesses will not affect search.
 \begin{comment}This is true because none of the methods
@@
 None of the methods have a bias toward creating specific gene values, in that
-XORing all solutions with a fixed string before evaluation will not effect search speed.
+XORing all solutions with a fixed string before evaluation will not affect search speed.
 \begin{comment}
@@
 In choosing test problems, we decided to use only benchmarks with knowable global
-optima.  This allows us to measure how many evaluations each method requires to converge to the
-global optima, as opposed to just the highest fitness reached.  This is advantageous
+optimum.  This allows us to measure how many evaluations each method requires to converge to the
+global optimum, as opposed to just the highest fitness reached.  This is advantageous
 as three of the four algorithms have no hard termination condition.  The remaining
@@
 In this equation $t$ represents the sum of gene values in the trap.  For each trap
-there is one local optima (all bits set to 0) and one global optima (all bits set to 1).
+there is one local optimum (all bits set to 0) and one global optimum (all bits set to 1).
 We chose to use a trap size of $k=7$.
@@
 This function creates a highly multimodal landscape which is completely separable
-and has a single global optima.  In order to apply our algorithms to this complex problem,
+and has a single global optimum.  In order to apply our algorithms to this complex problem,
 we discretized the representation using 10 bit gray coded numbers evenly distributed
@@
 Trap problem.  Each restart will require at most $2N$ evaluations to evaluate the
-random individual and optimize it to a local optima using the hill climber.  In
+random individual and optimize it to a local optimum using the hill climber.  In
 $P_0$ each trap in all solutions will either have all 1s or all 0s.  As a result $D(C_i, C_j) = 0$

### 2014-03-19 Parameterless Population Pyramid/main.tex
Made paper unanonymous, added clustering psuedocode.

@@
 %\end{comment}
-%\begin{comment}
+\begin{comment}
 \author{
@@
 
+\begin{figure}
+  \begin{algorithmic}
+  \Procedure{Cluster-Creation}{}
+    \State $unmerged \leftarrow \{\{i\} | 0 \le i < N\}$
+    \State $useful \leftarrow unmerged$
+    \While{$|unmerged|>0$}
+      \State $C_i, C_j \leftarrow \min_{C_i,C_j \in unmerged} D(C_i, C_j)$
+      \State $unmerged \leftarrow unmerged - \{C_i, C_j\} + \{C_i \cup C_j\}$
+      \State $useful \leftarrow \{C_i \cup C_j\}$
+      \If{$D(C_i, C_j) == 0$}
+        \State $useful \leftarrow useful - \{C_i, C_j\}$
+      \EndIf
+    \EndWhile
+    \State Order $useful$ based on cluster size, smallest first.
+    \State Remove largest cluster from $useful$.
+
+    \Return $useful$
+  \EndProcedure
+\end{algorithmic}
+  \caption{Algorithm describing how clusters are created using Equation~\ref{eq-distance}.}
+  \label{fig:cluster-creation}
+\end{figure}
+
 During cluster creation, P3 creates a binary tree of clusters such that the leaves
@@
 as well as individual run results and analysis scripts are available from our
-website.\footnote{\url{https://name.of.site.removed/anonymous}}  The results
+website.\footnote{\url{https://github.com/brianwgoldman/Parameter-less_Population_Pyramid}}  The results
 are summarized in Figure~\ref{fig-results} and Table~\ref{table-results}.

### 2014-03-22 Parameterless Population Pyramid/main.tex
Algorithms in, diagram out.  Fits to 8 pages!

@@
 population of solutions, with each level representing something akin
-to different generations of evolution. Consider the example evolution
+to different generations of evolution.  Figure~\ref{fig-p3} provides
+a high level summary of how P3 iteratively constructs these populations
+and utilizes them to improve randomly generated solutions through
+hill climbing and crossover.  While P3 can be applied to any discrete value
+representation, we will focus on the binary domain here for simplicity.
+\\
+\begin{comment}
+Consider the example evolution
 presented in Figure~\ref{fig-p3}.  At the start of optimization P3
@@
 further, so the process would begin yet again with a random solution.
+\end{comment}
 
@@
     \State Apply hill climber (Section~\ref{sec-hillclimber})
-    \If{solution $\notin$ pyramid}
+    \If{$\forall_{P_i\in pyramid}$ solution $\notin P_i$}
       \State Add solution to $P_0$ (Section~\ref{sec-pyramid})
@@
       \State Mix solution with $P_i$ (Section~\ref{sec-crossover})
-      \If{solution fitness has improved}
-        \If{solution $\notin$ pyramid}
+      \If{solution's fitness has improved}
+        \If{$\forall_{P_i\in pyramid}$ solution $\notin P_i$}
           \State Add solution to $P_{i+1}$
@@
 \end{algorithmic}
-  \caption{One iteration of P3 optimization.}
-  \label{fig:psuedo}
+  \caption{One iteration of P3 optimization. $pyramid$ is an
+           ordered set of populations.}
+  \label{fig-p3}
 \end{figure}
 
+\begin{comment}
 \begin{figure}
@@
 \end{figure}
+\end{comment}
 
@@
 constant in the number of solutions currently in the level.
+\\
 
@@
   \Procedure{Cluster-Creation}{}
-    \State $unmerged \leftarrow \{\{i\} | 0 \le i < N\}$
+    \State $unmerged \leftarrow \{\{0\}, \{1\}, \{2\}, \dots, \{N-1\}\}$
     \State $useful \leftarrow unmerged$
-    \While{$|unmerged|>0$}
+    \While{$|unmerged|>1$}
       \State $C_i, C_j \leftarrow \min_{C_i,C_j \in unmerged} D(C_i, C_j)$
       \State $unmerged \leftarrow unmerged - \{C_i, C_j\} + \{C_i \cup C_j\}$
-      \State $useful \leftarrow \{C_i \cup C_j\}$
-      \If{$D(C_i, C_j) == 0$}
+      \State $useful \leftarrow useful + \{C_i \cup C_j\}$
+      \If{$D(C_i, C_j) = 0$}
         \State $useful \leftarrow useful - \{C_i, C_j\}$
@@
 \end{algorithmic}
-  \caption{Algorithm describing how clusters are created using Equation~\ref{eq-distance}.}
-  \label{fig:cluster-creation}
+  \caption{Algorithm describing how clusters are created using Equation~\ref{eq-distance}
+           for each $P_i$. $unmerged$ and $useful$ are ordered sets of sets of gene loci.}
+  \label{fig-cluster-creation}
 \end{figure}
@@
 
+Figure~\ref{fig-cluster-creation} depicts the process of agglomeratively constructing
+the linkage tree.  Starting with each gene location in separate clusters, Equation~\ref{eq-entropy}
+is used to find which $C_i$ and $C_j$ in $unmerged$, the set of gene clusters, have the tightest
+linkage (minimum distance).  These are merged to form a new cluster $C_i \cup C_j$, and
+the separate pieces are removed from $unmerged$ while $useful$ tracks all clusters ever
+created in the hierarchy.  This process iterates until all locations have been merged
+into a single cluster.  With proper bookkeeping this algorithm will construct
+the linkage tree from the frequency table in $O(N^2)$~\cite{gronau:2007:upgma}.
+
+\begin{comment}
 Starting from the leaves, the binary tree is constructed agglomeratively. Each
@@
 from the frequency table~\cite{gronau:2007:upgma}.
+\end{comment}
 
@@
 solutions agree and one for where they disagree.
+\begin{comment}
 This means that Figure~\ref{fig-p3} is infeasible as crossover
@@
 higher level concepts.
+\end{comment}
+
+\begin{figure}
+  \begin{algorithmic}
+  \Procedure{Cluster-Usage}{}
+    \ForAll{$C_i \in useful$}
+      \ForAll{$d \in shuffled(P_i)$}
+        \State Copy $d$'s gene values for $C_i$ into solution
+        \If{solution was changed}
+          \If{solution's fitness decreased}
+            \State Revert changes
+          \EndIf
+          \State \textbf{break}
+        \EndIf
+      \EndFor
+    \EndFor
+  \EndProcedure
+\end{algorithmic}
+  \caption{Algorithm describing how clusters are used to perform crossover.}
+  \label{fig-cluster-usage}
+\end{figure}
 
 P3 uses crossover to recombine known information in a level of the
-pyramid with a candidate solution as follows.  For
+pyramid with a candidate solution as depicted in Figure~\ref{fig-cluster-usage}.
+For each cluster $C_i$ found by \Call{Cluster-Creation}{}, choose a random
+donor $d$ from the current level such that $d$'s gene values for at
+least one gene in $C_i$ differs from the candidate solution.
+Copy all of $d$'s gene values into the solution and evaluate the new
+solution.  If the fitness of the candidate
+has been reduced, revert the candidate's gene values.  It is important
+to note that candidate changes are kept in the case when the fitness
+is \emph{unaffected}.  This allows crossover to drift, potentially
+moving through fitness plateaus.
+
+\begin{comment}
+For
 each cluster select a random donor from the population.  Copy the
@@
 moving through fitness plateaus.
+\end{comment}
 

### 2014-03-26 CGP Journal/main.tex
First full draft of response to reviewers
Changes to emphasize "boolean" nature of experiments.

@@
 
-Our results show that most CGP techniques evolve genomes that
+%TODO mention this
+Our results on boolean problems show CGP evolves genomes that
 are highly inactive, very redundant, and full of seemingly useless constants.
@@
 neutrality~\cite{vassilev:2000:neutrality}, and structural bias~\cite{payne:2009:bias}.
-We shall focus on the interaction between mutation and genome ordering.
-%NEW
+We shall focus on the interaction between mutation and genome ordering, both
+%TODO final change
+generally and how they specifically apply to the boolean problem domain.
+%Done-NEW
 While these features of CGP have trivial definitions, we will argue, using
@@
 neutral drift.
-%NEW
+%Done-NEW
 In CGP mutation is done in a myriad of nearly synonymous ways.  Here we mutate
@@
 all possible functions, excluding the gene's current value.
-%NEW
+%Done-NEW
 This version of mutation was chosen to allow precise prediction of changes
@@
 
-%NEW
+%Done-NEW
 This technique incurs a minor increase in runtime, as each active gene in the
@@
 
-%NEW
+%Done-NEW
 Similar to \emph{Skip}, \emph{Accumulate} incurs a minor computation cost in
@@
 
-%NEW
+%Done-NEW
 Conceptually this method could be extended to mutate some number of active genes
@@
 created \emph{Reorder} and \emph{DAG} as methods for genome ordering.
-%NEW
+%Done-NEW
 For clarity, we shall refer to CGP's historical ordering method of enforcing
@@
 
-%NEW
+%DONE-NEW
 It is important to note that Equation~\ref{eq:prob} does not apply to all variants of CGP.
@@
 an individual can create is increased when using \emph{Reorder}.
-%/NEW
+%Done-NEW
 Therefore creating a specific offspring that improves upon the parent is
@@
 To achieve this, we incrementally determine
-%NEW No change was made to mutating node as that is in fact what we wanted
+%Done-NEW No change was made to mutating node as that is in fact what we wanted
 %/NEW
@@
 gathered before choosing between high quality configurations.
-%NEW
+%Done-NEW
 We chose to run each initially 5 times to reduce the chance of
@@
 were then tested using
-%NEW
+%Done-NEW
 Kruskal-Wallis\footnote{Kruskal-Wallis, a non-parametric equivalent of ANOVA, determines
@@
 to determine statistical significance using pairwise
-%NEW
+%Done-NEW
 Mann-Whitney~U\footnote{Mann-Whitney~U, a non-parametric equivalent of Student's t-test,
@@
 methods.
-%NEW
+%Done-NEW
 This supports the idea that the two are actually achieving synonymous behavior
@@
 than \emph{Skip}, we would suggest \emph{Skip} be used over \emph{Accumulate} for
-%NEW Slight change here, "real world" became "future"
+%Done-NEW Slight change here, "real world" became "future"
 any future applications.
@@
 as opposed to our \emph{Skip-Normal} requiring 600 thousand, also a factor of 6 improvement.
-%NEW
+%DONE-NEW
 To determine the effect the difference in function set had on results, we ran

### 2014-03-26 CGP Journal/main.tex
Condensing to fit onto 15 pages.

@@
 In total this creates 9 algorithm combinations.  Of these, our previous work
-has been reported showing the effectiveness of the duplication methods with \emph{Normal} ordering~\cite{goldman:2013:cgpwaste}
+%TODO Changed here
+has shown the effectiveness of the duplication methods with \emph{Normal} ordering~\cite{goldman:2013:cgpwaste}
 as well as analysis of all three ordering methods with \emph{Single}~\cite{goldman:2013:ordering}.
@@
 to be highly correlated.  This is apparent in the fact that the \textbf{No Reactivation}
-column in each table is almost perfectly sorted, even though rows were sorted using the
-\textbf{Active Unchanged} column.  Again the number of active nodes seems the likely
+%TODO Changed here
+column in each table is almost perfectly sorted, even though rows were sorted by the
+\textbf{Active Unchanged} column.  Again the number of active nodes is the likely
 cause.  Having fewer active nodes reduces the number of introns, making it more
@@
 an increasingly large number of offspring with no change in active node behavior,
-with little to no node reactivation, and a high chance of changing nodes before
+with little node reactivation, and a high chance of changing nodes before
 they are reactivated.  This suggests two important features: \emph{Skip} is likely
@@
 inactive nodes primarily to inject random behavior.  The former comes from the
-argument that a large part of the active unchanged individuals are also likely
+%TODO Changed here
+argument that most of the active unchanged individuals are also likely
 actively identical to their parents.  The latter is supported by the fact that
-\emph{Skip-Normal} so infrequently reactivates a node with tested behavior.  More
-often than not offspring connecting in inactive nodes are testing that node's behavior
-for the first time.
+\emph{Skip-Normal} infrequently reactivates a node with tested behavior with
+offspring frequently connecting in untested inactive nodes.
 

### 2014-03-26 CGP Journal/main.tex
A little more.

@@
 are satisfied, it can be put into \emph{addable}.  In this way, nodes are
-randomly removed from \emph{addable}, assigned the next possible location
+%TODO Change here
+randomly removed from \emph{addable}, assigned the next location
 in the genome, with new nodes added to \emph{addable} as they becomes semantically

### 2014-03-26 CGP Journal/main.tex
Finally down to 15 pages again!

@@
 being used to create output values referred to as ``inactive.''  Inactive nodes
-allow CGP individuals to drift genetically, as they can be mutated without effecting
-the fitness of the individual.  This genetic drift can then be incorporated, as
-mutation can change the DAG structure causing previously inactive nodes to become
+%TODO Changed here
+allow for genetic drift, as individuals can be mutated without effecting
+their fitness.  These mutations can then be incorporated, as future
+mutations can change the DAG structure causing previously inactive nodes to become
 active.  Previous work suggests CGP is most efficient when up to 95\% of the genome

### 2014-03-27 Parameterless Population Pyramid/main.tex
Updated copyright on P3
Removed "Genetic Algorithms Track" subtitle from P3.
Upated some formatting stuff in P3.

@@
 
-\documentclass{sig-alternate}
+\documentclass{sig-alternate-2013}
 %\linespread{2.5}
@@
 \usepackage{bm}
-\begin{document}
-%
-% --- Author Metadata here ---
-\conferenceinfo{GECCO'14,} {July 12--16, 2014, Vancouver, BC, Canada.}
-\CopyrightYear{2014}
-\crdata{978-1-4503-1963-8/13/07}
+
+\newfont{\mycrnotice}{ptmr8t at 7pt}
+\newfont{\myconfname}{ptmri8t at 7pt}
+\let\crnotice\mycrnotice%
+\let\confname\myconfname%
+
+\permission{Permission to make digital or hard copies of part or all of this work
+for personal or classroom use is granted without fee provided that copies are not
+made or distributed for profit or commercial advantage, and that copies bear this
+notice and the full citation on the first page. Copyrights for third-party
+components of this work must be honored. For all other uses, contact the
+owner/author(s). Copyright is held by the author/owner(s).}
+\conferenceinfo{GECCO'14,}{July 12--16, 2014, Vancouver, BC, Canada.}
+\copyrightetc{ACM \the\acmcopyr}
+\crdata{978-1-4503-1964-5/14/07. \\ http://dx.doi.org/10.1145/2576768.2598350}
+
 \clubpenalty=10000
 \widowpenalty = 10000
-%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
-%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
-% --- End of Author Metadata ---
-
+\begin{document}
 \title{Parameter-less Population Pyramid}
-\subtitle{[Genetic Algorithms Track]}
+%\subtitle{[Genetic Algorithms Track]}
 %
@@
 
-\keywords{Linkage Learning, Local Search, Parameter-less}
+\keywords{Linkage Learning; Local Search; Parameter-less}
 

### 2014-03-31 Parameterless Population Pyramid/main.tex
Fixed copywrite notice, take 2.

@@
 
-\permission{Permission to make digital or hard copies of part or all of this work
+\permission{Permission to make digital or hard copies of all or part of this work
 for personal or classroom use is granted without fee provided that copies are not
-made or distributed for profit or commercial advantage, and that copies bear this
-notice and the full citation on the first page. Copyrights for third-party
-components of this work must be honored. For all other uses, contact the
-owner/author(s). Copyright is held by the author/owner(s).}
+made or distributed for profit or commercial advantage and that copies bear this
+notice and the full citation on the first page. Copyrights for components of this
+work owned by others than ACM must be honored. Abstracting with credit is permitted.
+To copy otherwise, or republish, to post on servers or to redistribute to lists,
+requires prior specific permission and/or a fee. Request permissions from permissions@acm.org.}
 \conferenceinfo{GECCO'14,}{July 12--16, 2014, Vancouver, BC, Canada.}
-\copyrightetc{ACM \the\acmcopyr}
-\crdata{978-1-4503-1964-5/14/07. \\ http://dx.doi.org/10.1145/2576768.2598350}
+\copyrightetc{Copyright 2014 ACM \the\acmcopyr}
+\crdata{978-1-4503-2662-9/14/07\ ...\$15.00. \\ http://dx.doi.org/10.1145/2576768.2598350}
 

### 2014-05-13 CGP Journal/main.tex
Updated URL to be correct.
Renamed files to match IEEE requirements

@@
 11.5 billion evaluations were performed to gather our results.  The source code
-used to produce our results is available from our website.\footnote{\url{https://website.name.removed.com/AnonymousCGP}}
+used to produce our results is available from our
+website.\footnote{\url{https://github.com/brianwgoldman/Analysis-of-CGPs-Mechanisms}}
 
@@
     & \textbf{\em Normal} & \textbf{\em Reorder} & \textbf{\em DAG} \\ \hline
-    \thirdlabel{Skip} & \graphicthird{multiply_skip_normal} &
-                        \graphicthird{multiply_skip_reorder} &
-                        \graphicthird{multiply_skip_dag}\\ \hline
-    \thirdlabel{Accumulate} & \graphicthird{multiply_accumulate_normal} &
-                              \graphicthird{multiply_accumulate_reorder} &
-                              \graphicthird{multiply_accumulate_dag}\\ \hline
-    \thirdlabel{Single} & \graphicthird{multiply_single_normal} &
-                          \graphicthird{multiply_single_reorder} &
-                          \graphicthird{multiply_single_dag}\\ \hline
+    \thirdlabel{Skip} & \graphicthird{goldm00_multiply_skip_normal} &
+                        \graphicthird{goldm01_multiply_skip_reorder} &
+                        \graphicthird{goldm02_multiply_skip_dag}\\ \hline
+    \thirdlabel{Accumulate} & \graphicthird{goldm03_multiply_accumulate_normal} &
+                              \graphicthird{goldm04_multiply_accumulate_reorder} &
+                              \graphicthird{goldm05_multiply_accumulate_dag}\\ \hline
+    \thirdlabel{Single} & \graphicthird{goldm06_multiply_single_normal} &
+                          \graphicthird{goldm07_multiply_single_reorder} &
+                          \graphicthird{goldm08_multiply_single_dag}\\ \hline
 	\end{tabular}
@@
     & \textbf{\em Normal} & \textbf{\em Reorder} & \textbf{\em DAG} \\ \hline
-    \thirdlabel{Skip} & \graphicthird{bar_encode_skip_normal} &
-                        \graphicthird{bar_encode_skip_reorder} &
-                        \graphicthird{bar_encode_skip_dag}\\ \hline
-    \thirdlabel{Accumulate} & \graphicthird{bar_encode_accumulate_normal} &
-                              \graphicthird{bar_encode_accumulate_reorder} &
-                              \graphicthird{bar_encode_accumulate_dag}\\ \hline
-    \thirdlabel{Single} & \graphicthird{bar_encode_single_normal} &
-                          \graphicthird{bar_encode_single_reorder} &
-                          \graphicthird{bar_encode_single_dag}\\ \hline
+    \thirdlabel{Skip} & \graphicthird{goldm09_bar_encode_skip_normal} &
+                        \graphicthird{goldm10_bar_encode_skip_reorder} &
+                        \graphicthird{goldm11_bar_encode_skip_dag}\\ \hline
+    \thirdlabel{Accumulate} & \graphicthird{goldm12_bar_encode_accumulate_normal} &
+                              \graphicthird{goldm13_bar_encode_accumulate_reorder} &
+                              \graphicthird{goldm14_bar_encode_accumulate_dag}\\ \hline
+    \thirdlabel{Single} & \graphicthird{goldm15_bar_encode_single_normal} &
+                          \graphicthird{goldm16_bar_encode_single_reorder} &
+                          \graphicthird{goldm17_bar_encode_single_dag}\\ \hline
 	\end{tabular}
@@
     & \textbf{\em Normal} & \textbf{\em Reorder} & \textbf{\em DAG} \\ \hline
-    \thirdlabel{Skip} & \graphicthird{bar_multiply_skip_normal} &
-                        \graphicthird{bar_multiply_skip_reorder} &
-                        \graphicthird{bar_multiply_skip_dag}\\ \hline
-    \thirdlabel{Accumulate} & \graphicthird{bar_multiply_accumulate_normal} &
-                              \graphicthird{bar_multiply_accumulate_reorder} &
-                              \graphicthird{bar_multiply_accumulate_dag}\\ \hline
-    \thirdlabel{Single} & \graphicthird{bar_multiply_single_normal} &
-                          \graphicthird{bar_multiply_single_reorder} &
-                          \graphicthird{bar_multiply_single_dag}\\ \hline
+    \thirdlabel{Skip} & \graphicthird{goldm18_bar_multiply_skip_normal} &
+                        \graphicthird{goldm19_bar_multiply_skip_reorder} &
+                        \graphicthird{goldm20_bar_multiply_skip_dag}\\ \hline
+    \thirdlabel{Accumulate} & \graphicthird{goldm21_bar_multiply_accumulate_normal} &
+                              \graphicthird{goldm22_bar_multiply_accumulate_reorder} &
+                              \graphicthird{goldm23_bar_multiply_accumulate_dag}\\ \hline
+    \thirdlabel{Single} & \graphicthird{goldm24_bar_multiply_single_normal} &
+                          \graphicthird{goldm25_bar_multiply_single_reorder} &
+                          \graphicthird{goldm26_bar_multiply_single_dag}\\ \hline
 	\end{tabular}
@@
 
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{goldman_headshot}}]{Brian W. Goldman}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{goldm27_goldman_headshot}}]{Brian W. Goldman}
 received the B. S. degree and M.S. degree in computer science from the
@@
 % if you will not have a photo at all:
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{headPunch}}]{William F. Punch}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{goldm28_headPunch}}]{William F. Punch}
 Bill Punch received his Ph.D. from the Ohio State University and is currently an

### 2014-07-09 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Added P3 journal article outline

--- 2014-03-26 CGP Journal/main.tex
@@
 % Do not put math or special symbols in the title.
-\title{Analysis of Cartesian Genetic Programming's Evolutionary Mechanisms}
+\title{Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid}
 %
@@
 \begin{abstract}
-Understanding how search operators interact with solution representation is
-a critical step to improving search.
-In Cartesian Genetic Programming (CGP), and Genetic Programming (GP)
-in general, the complex genotype to phenotype map makes achieving this understanding a
-challenge.  By examining aspects such as tuned parameter values, the search quality of CGP variants at different
-problem difficulties, node behavior, and offspring replacement properties we
-seek to better understand the characteristics of CGP search.  Our focus is twofold:
-creating methods to prevent wasted CGP evaluations (\emph{Skip}, \emph{Accumulate},
-and \emph{Single}) and creating methods to overcome CGP's search limitations imposed by
-genome ordering (\emph{Reorder} and \emph{DAG}).
-
-%TODO mention this
-Our results on boolean problems show CGP evolves genomes that
-are highly inactive, very redundant, and full of seemingly useless constants.
-On some tested problems we found less than 1\% of the genome was actually required to
-encode the evolved solution.
-Furthermore, traditional CGP ordering results in large portions of the genome
-that are never used by any ancestor of the evolved solution.
-\emph{Reorder} and \emph{DAG} allow evolution to utilize the entire genome.
-More generally, our
-results suggest that \emph{Skip-Reorder} and \emph{Single-Reorder}
-are most likely to solve hard problems using the least number of evaluations and
-the least amount of time while better avoiding degenerate
-behavior.
-
+TODO Abstract
 \end{abstract}
@@
 \begin{IEEEkeywords}
-Cartesian genetic programming (CGP), Analysis
+TODO
 \end{IEEEkeywords}
@@
 % and "HIS" in caps to complete the first word.
-\IEEEPARstart{W}{hile effective} evolutionary methods are often simple to design
-and implement, discerning the reason that system
-leads to effective search can be challenging.
-This is exceptionally
-true for Genetic Programming (GP), which often employs both nontrivial
-genotype to phenotype maps and operators that act on genotypes without regard to
-phenotypic impact.  Yet understanding the root causes of evolutionary success
-and failure is critical to improving existing techniques, designing new techniques,
-and understanding how and where to apply each optimization system.
-
-There have been a number of previous studies into various aspects of GP evolution.
-For instance~\cite{daida3:2003:treebias} showed how a tree's shape effects its evolvability
-regardless of phenotype,
-resulting in the use of new grammar based operations~\cite{xuan:2006:grammar}.
-Analysis of the root cause of bloat has helped to inform methods for controlling
-bloat~\cite{luke:2006:bloat}.
-
-In this work we set out to develop a deeper understanding of what makes Cartesian
-Genetic Programming (CGP) an effective evolutionary optimizer.
-Previous studies of CGP's evolutionary mechanisms have
-focused on bloat~\cite{miller:2001:bloat},
-neutrality~\cite{vassilev:2000:neutrality}, and structural bias~\cite{payne:2009:bias}.
-We shall focus on the interaction between mutation and genome ordering, both
-%TODO final change
-generally and how they specifically apply to the boolean problem domain.
-%Done-NEW
-While these features of CGP have trivial definitions, we will argue, using
-experimentation and analysis of alternatives, that they have a great deal
-of impact on search success.
-%/NEW
-
-\begin{comment}
-In Section~\ref{sec:cgp} we provide
-an introduction to CGP, with sufficient background to put later sections into
-context.  Section~\ref{sec:duplicate} describes previous work on preventing mutation
-from creating identifiably wasted evaluations.
-Section~\ref{sec:ordering} discusses the impact of CGP's genome ordering,
-along with methods to both examine and bypass the limitations imposed by that ordering.
-Each technique is then rigorously tuned and qualitatively compared on a collection
-of benchmarks in Section~\ref{sec:quality}.  Section~\ref{sec:analysis} then
-provides a deep look into the details of CGP's search, and how each method impacts
-search, with final remarks given in Section~\ref{sec:conclusion}.
-\end{comment}
-
-\section{Cartesian Genetic Programming}
-\label{sec:cgp}
-Cartesian Genetic Programming (CGP) was originally proposed as a method for general
-genetic programming in~\cite{miller:2000:CGPorigin}.  While its name comes from
-its original application evolving circuits on a two dimensional grid, modern CGP can
-represent any directed acyclic graph (DAG), and has been utilized in applications such as binary
-circuits~\cite{walker:2008:cgpmodules},
-robot controllers~\cite{harding:2005:robots},
-neural networks~\cite{khan:2010:cgpann},
-image classifiers~\cite{harding:2012:mtcgp},
-and regression~\cite{harding:2009:smcgp}.
-
-CGP represents DAGs using a linear genome of integer values.  Each node in the
-DAG is encoded as a tuple of genes, with one gene specifying the function
-the node applies to its inputs, and the remaining genes expressing where the
-node takes input from.  Nodes can take input from either a problem input or
-any node preceding them in the linear genome.  Restricting connections in this way
-prevents the creation of cycles, while still allowing CGP to reuse values.  This is
-in contrast to tree based GP, which must duplicate functionality everywhere the same value is needed.
-To complete the representation, a set
-of extra genes are included at the end of the genome to specify which nodes or
-input locations to use as function outputs.
-
-As both output locations and information flow in the DAG are evolvable, often
-only a tiny fraction of the genome participates in creating the
-output values.  These nodes are referred to as ``active,'' with the nodes not
-being used to create output values referred to as ``inactive.''  Inactive nodes
-%TODO Changed here
-allow for genetic drift, as individuals can be mutated without effecting
-their fitness.  These mutations can then be incorporated, as future
-mutations can change the DAG structure causing previously inactive nodes to become
-active.  Previous work suggests CGP is most efficient when up to 95\% of the genome
-is inactive~\cite{miller:2006:redundancy}, while more recent work suggests this may
-be a result of hidden parsimony pressure in CGP~\cite{goldman:2013:ordering}.
-
-CGP uses very simple evolutionary mechanisms.  The most common evolutionary strategy
-is $\mu + \lambda$ where $\mu \leftarrow 1$ and $\lambda \leftarrow 4$.
-This means during each generation a single parent produces four offspring using
-mutation.  The best offspring then competes with the parent,\footnote{Ties are broken
-randomly.} with the offspring
-replacing the parent if it is no less fit.  This replacement strategy encourages
-neutral drift.
-%Done-NEW
-In CGP mutation is done in a myriad of nearly synonymous ways.  Here we mutate
-%/NEW
-each gene of each node at a set probability, where mutation
-involves changing a gene randomly to some different valid value.  For example,
-if a function gene is chosen for mutation, its new value is randomly chosen from
-all possible functions, excluding the gene's current value.
-%Done-NEW
-This version of mutation was chosen to allow precise prediction of changes
-in gene value necessarily in later sections.
-%/NEW
-Combined,
-this form of CGP can be viewed as a stochastic hill climber with neutrality.
-For a more in depth description of CGP, see~\cite{miller:2011:chapter2}.
-
-\section{Duplicate Evaluation Avoidance}
-\label{sec:duplicate}
-CGP's mutation operator is traditionally applied to both active and inactive genes uniformly.
-This has the
-potential to create offspring who's only mutated genes are in inactive nodes.
-These offspring are actively identical (contain identical active genes) to their
-parents, and therefore by definition have identical fitness to their parents.
-As such, re-evaluating these individuals is computationally wasteful as their fitness
-is known.
-Detecting duplicated individuals can be done without significant overhead since
-finding the set of active nodes is already done prior to
-evaluation~\cite{vasicek:2012:efficient}.
-Our previous work~\cite{goldman:2013:cgpwaste} has taken an initial look at the
-effect this waste can have, and proposed \emph{Skip}, \emph{Accumulate}, and
-\emph{Single} as methods for avoiding or preventing these duplicate evaluations.
-
-\begin{comment}
-The number of wasted evaluations is
-directly related to the mutation probability and the number of active genes.  With
-sufficiently high probabilities and enough active genes, the chance of
-creating an offspring which is actively identical to its parent is very low.
-Yet such a configuration may make it difficult for CGP to optimize, as mutation
-is likely to change many active genes in each application.  As was discussed
-in~\cite{goldman:2013:cgpwaste}, this can make normal CGP very sensitive to the
-mutation probability.
-\end{comment}
-
-\subsection{Skip}
-The \emph{Skip} method for avoiding duplicate evaluations involves the least
-amount of modification from CGP's normal behavior.  After an offspring is produced
-and its set of active nodes is determined but before it is evaluated, it is compared
-with its parent.  Each gene in each active node is compared for equivalence with
-the corresponding gene in the offspring's parent.  If all active genes are found to be equal
-then the offspring is not evaluated and is instead given the same fitness
-as its parent.
-Note that since inactive genes are not compared for equivalence it is still possible
-for offspring to genetically differ from their parents.
-
-Because \emph{Skip} does not modify any evolutionary mechanisms found in traditional CGP,
-it will produce identical individuals, with a potential reduction in evaluations.
-If the mutation probability is high enough relative
-to the number of active genes, \emph{Skip} will require the same number of evaluations as traditional CGP, since
-the probability of an offspring being actively identical to its parent is effectively zero.
-When the mutation probability is low enough relative to the number of active genes, some number of
-offspring will be actively identical to their parents, resulting in a reduction in evaluations
-but no change in the evolutionary trajectory.
-
-%Done-NEW
-This technique incurs a minor increase in runtime, as each active gene in the
-offspring must be compared against the parent's gene at the same locus.
-This operation is at most linear in genome size.  Yet
-on problems where evaluation is even somewhat expensive, the evaluations prevented
-by performing this check will result in a net reduction in runtime.
-%/NEW
-
-We would expect both intuitively and from our initial
-experimentation~\cite{goldman:2013:cgpwaste} that \emph{Skip} will be less sensitive
-to the mutation probability than normal CGP.  We would also expect \emph{Skip} to use a lower
-mutation probability than normal CGP when both are optimized to reduce evaluations.
-The former comes from the fact that in normal CGP any mutation probability that has a significant
-chance of creating offspring actively identical to their parents wastes
-evaluations.  In \emph{Skip} there are only two penalties for reducing the
-mutation probability.  From an evolutionary standpoint, if search becomes trapped in a local optima, low mutation
-probabilities may have increased difficulty escaping to find the global optimum.  From a performance standpoint,
-exceptionally low mutation probabilities may spend a prohibitive amount of time
-attempting to produce an evaluable offspring.
-Note that these penalties also exist in normal CGP.
-
-\subsection{Accumulate}
-Similar to \emph{Skip}, \emph{Accumulate} works by adding a step between offspring
-creation and offspring evaluation.  Instead of skipping evaluations when offspring
-are determined to be actively identical to their parents, \emph{Accumulate} enters
-into a cycle of repeated mutation until an individual worth evaluating is created.
-To understand this technique, consider an example where the parent $P$ produces
-the offspring $F_0$ using mutation. If $F_0$ is actively identical to $P$, $F_0$
-creates its own offspring $F_1$ using mutation.  This process of $F_i$ producing
-$F_{i+1}$ continues until $F_n$ is produced, where $F_n$ is not actively identical
-to $P$.  $F_n$ is then evaluated.  If $F_n$ is no worse than $P$, it replaces
-$F_0$, and evolution continues as though $P$ had produced $F_n$ directly.  If
-$F_n$ is worse than $P$, $F_{n-1}$ replaces $F_0$, as $F_{n-1}$ has accumulated
-the most mutations to inactive genes without reducing fitness.  Note that
-the mutation used each time is identical to normal CGP's mutation, such that
-$F_n$ will have one or more active genes different from $P$.  Furthermore, even
-though this technique produces $n+1$ individuals, only $F_n$ is evaluated.
-Viewed in another way, \emph{Accumulate} performs a micro evolution on each offspring
-before it is evaluated, with an emphasis on encouraging neutral drift.
-
-Previous experimentation has shown that \emph{Accumulate} acts very similarly to
-\emph{Skip}~\cite{goldman:2013:cgpwaste}, with the exception that \emph{Accumulate}
-favors lower mutation probabilities when tuned to reduce evaluations to success. While at first these algorithms may appear
-quite different, further consideration shows how they are similar.  In \emph{Skip}
-any actively identical offspring that is produced is likely to be selected as we expect mutations to
-active genes to more often reduce fitness than improve it.  In the next generation
-this actively identical offspring is then mutated again, with its lineage likely
-continuing until it finally does mutate one or more active genes.  At this point, if the
-mutant is better it replaces the parent, otherwise the parent, which has been accumulating
-mutations to inactive genes, is kept.  In this way \emph{Skip} mirrors \emph{Accumulate},
-except \emph{Skip} acts over multiple generations instead of compressed into a single generation.
-
-%Done-NEW
-Similar to \emph{Skip}, \emph{Accumulate} incurs a minor computation cost in
-order to prevent wasting evaluations.  Generating each $F_i$ requires at most
-$O(AN)$, where $A$ is the arity of the nodes and $N$ is the number of nodes.
-While \emph{Accumulate} may produce many $F_i$ in the process of creating
-a single offspring, conceptually \emph{Skip} using the same mutation rate
-will create a similar number of total individuals when performing the same
-number of evaluations.  As such, we suggest \emph{Accumulate}
-will take no more than a constant factor more runtime to produce the same number
-of evaluable individuals as \emph{Skip}.  As before, this increase is likely
-eclipsed when solving problems with sufficiently time consuming evaluation functions.
-%/NEW
-
-\subsection{Single}
-While \emph{Skip} and \emph{Accumulate} save evaluations by not evaluating offspring
-which are actively identical to their parents, \emph{Single} changes
-mutation to ensure that only evaluable offspring are created.  Instead
-of mutating each gene at a set probability, \emph{Single} mutates genes at
-random until exactly one active gene is mutated.
-
-This modification gives \emph{Single} three properties distinct from the other
-forms of duplicate avoidance.  First, \emph{Single} forces offspring to have
-an active gene which differs from their parent.  This limits drift as the
-mutated active gene must either be to an intron or represent another way to
-code a solution of equal quality.  Second, as a benefit of forced changes,
-\emph{Single} avoids the overhead of repeatedly generating individuals without
-creating an evaluable offspring.  Third, \emph{Single} does not require a mutation
-parameter, effectively setting the mutation probability to $\frac{1}{a}$ for active
-genes and $\frac{1}{a+1}$ for inactive genes, where $a$ is the number of active genes.
-
-In encodings without inactive genes, limiting mutation to changing exactly one gene
-could prevent an algorithm from escaping some types of local optima.  Yet because
-CGP allows for inactive genes, \emph{Single} is still able to escape most local
-optima using sufficient drift of inactive genes.  When a high percentage of
-the genome becomes active or when \emph{Single} is otherwise limited in its ability
-to drift due to lack of introns, it does have an increased potential to become stuck.
-As will be discussed in Section~\ref{sec:ordering} and was shown in~\cite{goldman:2013:ordering}
-this problem of highly active genomes is very unlikely.
-
-%Done-NEW
-Conceptually this method could be extended to mutate some number of active genes
-before terminating.  Doing so would reintroduce a parameter to govern how many
-active genes to mutate.  Our choice of a single mutation allows for the minimal
-amount of change to ensure the individual should be evaluated, while mutations
-made to inactive genes still give CGP sufficient power in each mutation to escape
-local optimums.
-%/NEW
-
-\section{Genome Ordering}
-\label{sec:ordering}
-As was discussed in Section~\ref{sec:cgp}, CGP uses node ordering in the genome
-to prevent cycles, by ensuring nodes only receive input from sources that
-precede them in the genome.  This restriction does not limit CGP's ability
-to represent DAGs, as all DAGs can be serialized to fit this requirement.
-However, this representation is likely having an impact on CGP's ability to evolve specific
-DAGs~\cite{goldman:2013:ordering}.
-
-Primarily, enforcing node ordering adds artificial limitations to CGP's ability to
-connect nodes.  Nodes which can be connected without creating a cycle may
-still be prevented from forming that connection because of random genome ordering.
-Similarly, it is impossible for adjacent nodes
-to ever be connected through an intermediate node.  As a result, useful structures
-in the genome may need to be evolved repeatedly in order to find the best
-location for that structure in the genome.
-
-Ordering, when combined with the mutation's uniform resetting of connection genes, may be the cause of CGP's
-immensely inactive genomes.  As was rigorously defined in our work~\cite{goldman:2013:ordering},
-the number of active genes in a genome is expected to scale logarithmically
-with genome size, independent of problem application.  Furthermore, the less
-active a genome, the lower the likelihood of structure ordering preventing structure
-connection.
-
-As an extension to the theory presented in~\cite{goldman:2013:ordering}, we have
-derived the formula which predicts the probability a node is active based on its
-index, given in Equation~\ref{eq:prob}.
-\begin{equation}
-p(i) \leftarrow 1 - (\frac{N+I-1}{N+I})^O \cdot \prod_{j=i+1}^{N} 1 - p(j) \cdot (1-(\frac{j+I-1}{j+I})^A)
-\label{eq:prob}
-\end{equation}
-In this equation, $i$ is the index of the working node, $N$ is the genome size, $I$ is the number of input locations,
-$O$ is the number of output locations, and $A$ is the arity of each node.
-This equation is constructed as a negation, such that we determine the probability
-the node is not active due to each potential source that could make it active.
-The product term calculates the probability that the node at index $j$ is connected to the
-node at index $i$, multiplied by the probability that the node at index $j$ is active.  This is
-negated to determine the probability node $i$ is not active because of node $j$.
-All together, the product term calculates the probability that no active node is
-connected to the node at index $i$.  Nodes can also become active if they are directly
-connected to by an output location.  As such the product is scaled by the probability
-that all output locations connect to inputs or nodes not at index $i$.  This
-formula assumes only operator bias, not selective pressure, is acting on the
-genome.
-
-\begin{comment}
-\begin{figure}
-\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{active_prob}
-\caption{Plot of Equation~\ref{eq:prob} describing the probability that a node
-         at a given index will be active.  $N=2000, A=2, I=6, O=6$.
-         Predicts approximately 160 total active nodes, with 80\% of the first
-         100 nodes active and 4\% of the remaining 1900 nodes active.}
-\label{fig:prob}
-\end{figure}
-
-From Figure~\ref{fig:prob} it is clear that nodes near the beginning of the genome
-are almost certainly going to be active, while nodes in the last three quarters
-of the genome are almost never going to be active.
-\end{comment}
-%NEW
-Substituting $N=2000$, $A=2$, $I=6$, $O=6$ into Equation~\ref{eq:prob} (chosen to match
-experimental values used in Section~\ref{sec:analysis})
-predicts only 160 active nodes, or 8\% of the total genome.  Positionally, 80\% of the first 100 nodes are expected to be
-active while just 4\% of the remaining 1900 nodes are expected to be active.
-%/NEW
-
-To rectify these issues and examine CGP without ordering bias, \cite{goldman:2013:ordering}
-created \emph{Reorder} and \emph{DAG} as methods for genome ordering.
-%Done-NEW
-For clarity, we shall refer to CGP's historical ordering method of enforcing
-only forward connections as \emph{Normal}.
-%/NEW
-
-%DONE-NEW
-It is important to note that Equation~\ref{eq:prob} does not apply to all variants of CGP.
-For instance, the early versions of CGP~\cite{miller:2011:chapter2} which utilize a
-levels back parameter $L$ will have
-a flat active probability for all nodes at least $L$ nodes from the start and $L$ nodes from the end
-of the genome.  Yet this does not remove the issue of artificially limiting connections,
-and in fact adds many new limitations.  More recently,~\cite{cai:2005:ircgp} proposed
-a method which reassembled the genome using semantic cues, which circumvents the
-forward only limitation in \emph{Normal}.
-%/NEW
-\subsection{Reorder}
-\label{sec:reorder}
-The \emph{Reorder} method shuffles an existing genome's node ordering without
-changing node behavior.  This is possible because for a given node, there can
-be a large number of other nodes which have no required ordering,
-as they neither take input (directly or indirectly) from the node nor provide their
-output (directly or indirectly) to the node.  In general, this process works by assigning nodes new
-locations in the genome at random once all of the nodes they take input from have been assigned
-locations earlier in the genome.
-
-The first step in performing \emph{Reorder} on a genome is to create
-data structures to store direct connection relationships.  These structures are used
-to determine in constant time all connections incident on a node, and 
-can be built in order $O(AN)$ time, where $A$ is the arity of the nodes
-and $N$ is the number of nodes.
-
-With this preprocessing complete, the \Call{RandomSerialization}{} algorithm given in Figure~\ref{fig:reorder}
-is used to assign each node a new location in the modified genome.
-It starts by constructing the \emph{addable} set, which contains all nodes who's direct
-dependencies have already been added.  Initially this contains only the input locations.
-Once a node is assigned a location, all of the nodes that depend on its output
-have that dependency marked as satisfied.  Once all of a node's dependencies
-are satisfied, it can be put into \emph{addable}.  In this way, nodes are
-%TODO Change here
-randomly removed from \emph{addable}, assigned the next location
-in the genome, with new nodes added to \emph{addable} as they becomes semantically
-viable.  Iteration ends when all nodes have been assigned new locations.
-\Call{RandomSerialization}{} requires $O(AN)$, as there can be only $N$ iterations,
-with each iteration updating a distinct subset of the $AN$ dependencies.
-
-\begin{figure}
-  \begin{algorithmic}
-    \Procedure{RandomSerialization}{}
-    \State $addable \leftarrow input\_locations$
-    \State $index \leftarrow |input\_locations|$
-    \State $new\_loc \leftarrow \emptyset$
-    \While{$|addable| > 0$}
-      \State $working \leftarrow randpop(addable)$
-      \If{$working \notin input\_locations$}
-        \State $new\_loc[working] \leftarrow index$
-        \State $index \leftarrow index + 1$
-      \EndIf
-      \ForAll{$link \in feeds\_to(working)$}
-        \State Satisfy $link$'s dependence on $working$
-        \If{$|unsatisfied(link)| = 0$}
-          \State $addable \leftarrow addable \cup \{link\}$
-        \EndIf
-      \EndFor
-    \EndWhile
-    \State\Return $new\_loc$
-    \EndProcedure
-  \end{algorithmic}
-  \caption{Algorithm which converts any DAG into a random serial ordering such
-           that all nodes take input from nodes that precede them in the genome.
-           Used by \emph{Reorder} to determine new node locations and \emph{DAG} to
-           determine node evaluation ordering.}
-  \label{fig:reorder}
-\end{figure}
-
-The final step in \emph{Reorder} is to use the list of new locations to convert
-the existing genome into the reordered genome.  This can be done
-by converting connection genes and output location genes using
-the $new\_loc$ map returned from \Call{RandomSerialization}{}.
-This also requires $O(AN)$ time, as all connection genes in all nodes must
-be converted.  As this is the final step unique to \emph{Reorder}, and
-no previous steps are of higher complexity, it is also the complexity of
-the \emph{Reorder} algorithm as a whole.  Furthermore, it is of the same complexity class
-as copying the entire genome, meaning it does not change CGP's overall complexity.
-
-\emph{Reorder} is used once each generation to shuffle the nodes of the parent's
-genome.  As the shuffling does not semantically change the parent, it does
-not require re-evaluation.  Furthermore, as shuffling does not change the requirement
-that nodes only depend on those preceding them in the genome, shuffling does not
-require any changes to any other CGP methods, such as evaluation and mutation.
-It is important to note, however, that the potential mutations that can be applied to the parent
-have changed.  Consider two nodes $X$ and $Y$ which are in the genome, and
-have no dependent relationship.  As $X$ does not take input from $Y$ and
-$Y$ does not take input from $X$, either can be placed before the other
-in the genome.  Without shuffling, the initial ordering of nodes is fixed.
-No operation in \emph{Normal} is capable of making $X$
-take input from $Y$.  \emph{Reorder}'s shuffling can allow
-subsequent mutations to make the connection.
-
-\emph{Reorder} has the potential to reduce node reinvention.  In \emph{Normal}, if
-$X$ preceded $Y$, mutation would have to recreate either $X$ or $Y$ in new locations
-in order to make the required connection.  \emph{Reorder} provides the opportunity
-to examine different orderings of the genome without such costs.
-
-The price \emph{Reorder} pays for this potential improvement is a larger mutational search
-space.  Though a shuffled individual is phenotypically unchanged,
-its mutational adjacency can be significantly different.  As such the variety of offspring
-an individual can create is increased when using \emph{Reorder}.
-%Done-NEW
-Therefore creating a specific offspring that improves upon the parent is
-less likely, in that the number of alternatives is increased.  Yet many
-of these additional alternatives may also be improvements upon the parent.  This may
-increase the likelihood of producing an offspring that improves
-upon the parent, allowing \emph{Reorder} to overcome this drawback.
-%/NEW
-
-\subsection{DAG}
-\label{sec:dag}
-\emph{Normal} and \emph{Reorder} use node location in a linear genome to create
-an efficient protection against cycles: nodes must connect forward.  \emph{DAG}
-removes this restriction and allows forward and backward connections, relying on other, more complex, methodologies to
-prevent cycles.  To accomplish this goal, modifications must be made to how CGP
-performs mutation and evaluation.
-
-As in \emph{Normal} and \emph{Reorder}, when a connection gene is chosen for mutation, its value is chosen randomly.
-However, the options to choose from are only limited to those connections which do not
-create cycles.
-To achieve this, we incrementally determine
-%Done-NEW No change was made to mutating node as that is in fact what we wanted
-%/NEW
-which nodes transitively depend on the mutating node and which do not, with
-information stored in a $known$ map data structure.  This data structure is initialized with
-the information that the node is dependent on itself and that all input locations
-are not dependent on the mutating node.  Nodes are then
-tested in a random order for dependence using the \Call{IsDependent}{} algorithm given in Figure~\ref{fig:dag}.
-The first node returned at the top level that is not dependent on the mutating node is then
-used as the new connection gene value.  As a result, mutating a connection gene
-cannot introduce a cycle.
-
-\begin{figure}
-  \begin{algorithmic}
-  \Procedure{IsDependent}{$working$, $known$}
-    \If{$working \in known$}
-      \State\Return{$known[working]$}
-    \EndIf
-    \ForAll{$link \in reads\_from(working)$}
-      \If{\Call{IsDependent}{$link$, $known$}}
-        \State $known[working] \leftarrow True$
-        \State\Return{$True$}
-      \EndIf
-    \EndFor
-    \State $known[working] \leftarrow False$
-    \State\Return{$False$}
-  \EndProcedure
-  \end{algorithmic}
-  \caption{Algorithm used by \emph{DAG} to determine valid connection gene values.}
-  \label{fig:dag}
-\end{figure}
-
-\Call{IsDependent}{} is a non-repeated recursive depth
-first search of the individual's DAG.  Search terminates as soon as a dependent
-node is found, and information from previous searches is memoized to prevent
-repeated search.  Each recursion level results in a node being added to $known$.
-As such initial calls will likely result in lots of recursion, but subsequent
-calls will find answers more quickly.  In the worst case, this algorithm may need
-to examine all nodes in the genome to determine if $working$ is dependent on the
-mutating node, giving \Call{IsDependent}{} a complexity of $O(AN)$.  In the worst case, this algorithm
-may be called on all possible nodes in the genome to check their dependence.
-Yet cumulatively those calls can only take $O(AN)$ time due to the non-repeating
-nature of the search.  As a result, the expected worst case run time for \emph{DAG}
-mutation is $O(mA^2N^2)$, where $m$ is the mutation probability.  We expect $mAN$
-connection genes to be mutated, each requiring a worst case of $N$ dependency checks
-each taking $A$ time.
-This is in contrast to \emph{Normal} and \emph{Reorder} which require $O(mAN)$ to perform
-mutation, as individual mutations can be performed in constant time.
-
-The method for evaluating \emph{DAG} individuals is very similar to the method for
-efficiently evaluating \emph{Normal} and \emph{Reorder} individuals.  As before,
-preprocessing is done to determine which nodes are active, and only those nodes are
-executed during evaluation.  Also as before, the process begins from the output locations,
-recursively following connection genes and marking nodes as active until the input locations
-are reached.  At this point, the process changes.  In \emph{DAG}, we must determine not only the set of
-nodes that are active, but also the order in which those nodes should be executed.
-
-\Call{RandomSerialization}{} can be reused to determine the order
-in which nodes should be executed during evaluation.  Instead of using $new\_loc$
-to reorder the genome, we can use this map to specify the order in which nodes should be
-executed.  For instance, if a node was given the new location of $X$, we know
-that once all nodes at locations preceding $X$ have been evaluated, the node at $X$
-can be evaluated.  Stripping the inactive nodes from $new\_loc$ and inverting the map
-results in an efficient, valid ordering in which to execute the nodes.
-
-The process of determining which nodes are active requires $O(AN)$ time, regardless
-of node ordering.  As was discussed in Section~\ref{sec:reorder},
-getting the $new\_loc$ map requires $O(AN)$ time.  Finally, converting the map
-into an ordering of the active nodes takes $O(N)$ time.  Each of these steps are
-sequential, meaning \emph{DAG} does not increase the runtime complexity of evaluation
-preprocessing.  Furthermore, once the preprocessing step is done, \emph{DAG} takes
-identical time to evaluate as the other methods, regardless of the number of
-input combinations.
-
-With these modifications, there is no longer any meaning to the location of a node
-in the genome.  In \emph{Normal} the number of valid
-values for a connection gene is related to the number of nodes
-preceding that gene, and independent of the number of nodes following that gene.
-In \emph{DAG} a connection gene's valid values only depends on the current solution
-represented by the individual, independent of the genes location in the genome.
-As such, there are fewer artificial limitations on how \emph{DAG} can modify
-solutions.
-
-While \emph{Reorder} and \emph{DAG} both reduce mutational limitations, \emph{DAG}'s changes are more sweeping.
-Though both can create any connection that does not form a cycle,
-\emph{Reorder} maintains a bias toward certain kinds of connections.
-Consider again two nodes $X$ and $Y$
-such that neither is dependent on the other.  If $X$ is transitively dependent on very
-few nodes, and $Y$ is transitively dependent on many nodes, \emph{Reorder} is more likely
-to shuffle $X$ to precede $Y$.  This is true because at any point during the construction
-of the new ordering, the probability
-that all of $X$'s dependencies have been added is
-higher than for $Y$'s dependencies.  As a result we expect $X$ to be put into $addable$ sooner,
-and therefore we expect it to be added sooner.  The greater the discrepancy in the dependency
-set sizes the less likely the order between the nodes will be reversed.
-Conversely, \emph{DAG} has no such bias.  It is always possible to mutate $X$ to connect to $Y$.
-
-The probability of connecting $X$ to $Y$ is also influenced by how many nodes transitively depend on $X$ and $Y$.
-For example, assume $X$ has more nodes transitively dependent on it than $Y$.
-In \emph{Reorder}, $X$ is again more likely to be placed before $Y$, this time because the number
-of possible nodes preceding it is reduced.  \emph{DAG} is biased in the reverse, such that $X$ will
-have a higher chance of connecting to $Y$ than $Y$ to $X$.  This is because $X$ can connect to fewer
-nodes, so each mutation has a higher chance of choosing $Y$.
-
-\section{Qualitative Comparisons}
-\label{sec:quality}
-
-\begin{table}
-	\centering
-	\caption{Short descriptions of each of the algorithm variants.}
-	\begin{tabular}{|c|p{6cm}|}
-	\hline
-  \multicolumn{2}{|c|}{\textbf{Duplication}} \\ \hline
-	\textbf{\emph{Skip}} & Whenever an actively identical offspring is detected, assign it its parent's fitness. \\ & \\
-	\textbf{\emph{Accumulate}} & Repeatedly perform full genome mutation until \textbf{one or more} active genes change. \\ & \\
-	\textbf{\emph{Single}} & Select and mutate individual genes until \textbf{exactly one} active gene changes. \\
-	\hline \multicolumn{2}{|c|}{\textbf{Ordering}} \\ \hline
-	\textbf{\emph{Normal}} & Nodes must take input from other nodes or inputs that precede them in the genome. \\ & \\
-	\textbf{\emph{Reorder}} & Each generation the parent's node ordering is shuffled without effecting node behavior. \\ & \\
-	\textbf{\emph{DAG}} & Connection genes can mutate to make any connection as long as no cycle is formed. \\
-	  \hline
-	\end{tabular}
-	\label{tab:glossary}
-\end{table}
-
-\begin{comment}
-Empirical results are an important step in confirming our theoretical examinations.
-Therefore we have tested the effectiveness
-of the described variants and their combination in problem solving.  Deeper analysis of
-evolutionary side effects are given in Section~\ref{sec:analysis}.
-Table~\ref{tab:glossary} gives a summary of what each variant means.
-\end{comment}
-
-All told we have proposed three new methods for avoiding duplicate evaluations
-(\emph{Skip}, \emph{Accumulate}, and \emph{Single}) and two new methods for
-genome ordering (\emph{Reorder} and \emph{DAG}).
-See Table~\ref{tab:glossary} for a short description each variant.
-As these two
-sets are non overlapping, and there is no intuitive groupings, we chose to test
-all possible combinations in our empirical analysis.  As a control we included \emph{Normal} genome ordering,
-as it provides insight to how each of the duplicate prevention techniques work
-if applied alone.  Similar measures are not needed in testing the genome ordering techniques,
-as \emph{Skip} has no impact on evolutionary search, and is strictly an improvement
-to using no duplication detection at all.
-
-In total this creates 9 algorithm combinations.  Of these, our previous work
-%TODO Changed here
-has shown the effectiveness of the duplication methods with \emph{Normal} ordering~\cite{goldman:2013:cgpwaste}
-as well as analysis of all three ordering methods with \emph{Single}~\cite{goldman:2013:ordering}.
-This means there remain 4 novel combinations to be tested.
-This is also the first time the ordering methods
-have been rigorously tested for quality, as much of the work in~\cite{goldman:2013:ordering}
-was analysis oriented.
-
-\subsection{Problem Set}
-In order to provide test landscapes for evolution, we chose problems common to
-the CGP literature and previous work with the duplication prevention and ordering
-techniques.  We have chosen 4 binary circuit problems, as the binary representation
-allows for some of the precise analysis performed in Section~\ref{sec:analysis}.
-For all problems, we used the function set \{AND, OR, NAND, NOR\}.
-To cover a range of different binary problems, each of the four chosen have
-different numbers of inputs, different number of outputs, and different levels
-of difficulty.
-
-The first problem, 3-bit Parity, represents the most
-common application for
-CGP~\cite{yu:2001:neutrality,miller:2006:redundancy,walker:2008:cgpmodules}.
-We include this problem purely for backward comparison, as we agree that in general
-it is too simple of a problem to merit its own conclusions~\cite{white:2013:bgpb}.
-Yet as part of the group it may help yield understanding about how each variant
-performs on simple problems.
-
-As representatives of binary problems with varying input and output sizes, we
-reuse the 16-4-bit Encode and 4-16-bit Decode proposed in~\cite{goldman:2013:cgpwaste}.
-In these problems, the circuit evolved must either convert a 4-bit encoded integer
-to a 1 on the corresponding output line (Decode), or take a 1 on one of the 16
-input lines and convert it back into a 4-bit encoded integer (Encode).  These
-problems share properties with the commonly used Multiplexer problem, but include
-multiple output lines, and can be evaluated quicker as each only requires 16 possible
-test points.
-
-Finally we chose the 3-bit Multiply to represent hard binary problems,
-as was suggested in~\cite{white:2013:bgpb} and used
-by~\cite{vassilev:2000:neutrality,miller:2006:redundancy,walker:2008:cgpmodules}.
-This problem is very difficult by comparison to the other problems, and has
-the largest number of test points.
-
-\subsection{Parameter Setting}
-\label{sec:parameter}
-To perform a fair comparison, we ensure proper parameter configuration
-to avoid potential bias that might benefit
-a specific method.  While tuning can lead to the alternative issue
-of methods only being effective after extensive problem specific tuning, we set out to use
-rough-grained values and provide each technique with equal tuning time to alleviate that bias.
-
-We focus on two parameter values in our configuration: mutation probability and genome size.
-While there are other potential configurable parameters (population size, offspring size, etc),
-we feel the CGP literature has converged on settings for those parameters (1, 4).
-Furthermore, as duplication detection explicitly deals with the relationship between
-mutation probability and number of active genes, and the number of active genes depends
-on the genome size and ordering method, mutation probability and genome size seem to be
-the most likely to impact results.
-
-To set parameter values, we started by defining a grid of parameter values:
-50, 100, 200, 500, 1000, 2000, 5000, 10000 for genome size and 0.05, 0.02, 0.01,
-0.005, 0.002, 0.001, 0.0005, 0.0002 for mutation probability.  Note that here a genome
-size of 50 means there are 50 nodes, each composed of multiple genes, plus the
-required genes to specify output locations.  Also note that a mutation probability of 0.05
-is used here to mean each gene has a 5\% probability of being mutated in a single
-application of mutation.  These ranges were chosen to completely cover the
-range of previously used parameter settings.
-The grid is formed by trying all possible mutation probabilities
-with all possible genome sizes, resulting in 64 potential configurations for each
-of the algorithm combinations.  Finally note that as \emph{Single} does not
-use a mutation probability, only the eight configurations with different genome sizes were
-used when \emph{Single} was employed.
-
-To choose which parameter configuration to use with each algorithm combination
-for each problem, we used an iterative process of performing runs, comparing medians, removing configurations,
-and repeating.  For each algorithm combination and problem, we first ran all
-potential parameter configurations 5 times (320 runs if not using \emph{Single}, 40 otherwise).
-The parameter configurations are then sorted based on their median.  The best half are then
-each run 4 more times, with the best half of that group chosen.  In this way, 32 configurations
-are run 5 times, 16 are run 9 times, 8 are run 13 times, 4 are run 17 times, 2 are run 21 times,
-and 2 are run 25 times.  For algorithm combinations using \emph{Single}, all 8 parameter configurations were run 17 times
-before any were removed.  This iterative reduction allows us to focus tuning on
-those parameter configurations most likely to be effective, with more information
-gathered before choosing between high quality configurations.
-%Done-NEW
-We chose to run each initially 5 times to reduce the chance of
-eliminating a configuration due to outliers.  We incrementally added 4
-to provide enough information to significantly update estimates,
-while keeping the number of runs odd, allowing the median to be a
-representative run and not an interpolation.  As a final consideration
-the number of experiments were set to make the total number of runs feasible.
-%/NEW
-In total this means
-15,696 complete runs are required to set the parameter configuration for all
-algorithm configurations for all problems.
-
-To ensure termination, all runs were limited to 10,000,000 evaluations, well beyond
-the expected time to completion.  As the median was used to compare configurations,
-selection should be insensitive to this limit unless a large
-portion of runs failed to optimize.
-
-To prevent undue biasing in the final results, these tuning runs were
-not used in the subsequent data analysis.  Effectively, the runs used to set
-the parameter configuration were considered
-the training runs for that configuration.  As a result, our final results were gathered
-from a completely independent set of runs to test the configuration's general characteristics.
-Each configuration was run 50 times to ensure statistical power.  All told over
-11.5 billion evaluations were performed to gather our results.  The source code
-used to produce our results is available from our website.\footnote{\url{https://website.name.removed.com/AnonymousCGP}}
-
-\subsection{Results}
-
-\begin{table*}
-	\centering
-	\caption{Results for algorithm configuration on each problem.  Highlighted row
-	is the control configuration \emph{Skip-Normal}.  MES=Median Evaluations to Success;
-	Confidence Interval=95\% bootstrapped confidence interval; Active=Median number of active nodes in evolved solution;
-	Used=Median number of nodes active in evolved solution after applying SIMPLIFY (See Section~\ref{sec:behavior});
-	p-value=Result from Mann-Whitney~U comparison with control.
-	}
-	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
-	  \hline
-	  \multicolumn{9}{|c|}{\textbf{3-bit Parity}, Kruskal-Wallis=$5.99\cdot 10^{-14}$} \\ \hline
-\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{Confidence Interval} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
-\emph{Accumulate} & \emph{Normal} & 10,000 & 0.02 & 682 & 445 .. 869 & 122 & 20 & 0.5602 \\ \hline
-\emph{Accumulate} & \emph{Reorder}&  5,000 & 0.01 & 796 & 576 .. 1,019 & 331 & 23 & 0.7174 \\ \hline
-\rowcolor{Gray}
-\emph{Skip} & \emph{Normal}       & 10,000 & 0.02 & 823 & 662 .. 1,034 & 119 & 20 & NA \\ \hline
-\emph{Skip} & \emph{Reorder}      &  2,000 & 0.01 & 870 & 708 .. 1,070 & 198 & 20 & 0.3125 \\ \hline
-\emph{Accumulate} & \emph{Dag}    &  5,000 & 0.01 & 964 & 759 .. 1,250 & 1,174 & 22 & 0.3310 \\ \hline
-\emph{Skip} & \emph{Dag}          & 10,000 & 0.02 & 1,042 & 734 .. 1,339 & 2,083 & 24 & 0.0248 \\ \hline
-\emph{Single} & \emph{Normal}     & 500 & NA & 1,214 & 885 .. 1,487 & 39 & 17 & 0.0046 \\ \hline
-\emph{Single} & \emph{Reorder}    & 200 & NA & 1,506 & 552 .. 2,122 & 57 & 18 & 0 \\ \hline
-\emph{Single} & \emph{Dag}        & 100 & NA & 2,569 & 2,109 .. 3,358 & 43 & 15 & 0 \\ \hline\hline
-
-	  \multicolumn{9}{|c|}{\textbf{16-4-bit Encode}, Kruskal-Wallis=$1.19\cdot 10^{-6}$} \\ \hline
-
-\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{Confidence Interval} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
-\rowcolor{Gray}
-\emph{Skip} & \emph{Normal}        & 10,000 & 0.005 & 16,226 & 9,315 .. 22,229  &   283 & 49 & NA \\ \hline
-\emph{Skip} & \emph{Reorder}       & 10,000 & 0.005 & 17,883 & 10,690 .. 23,999 & 2,070 & 62 & 0.6918 \\ \hline
-\emph{Accumulate} & \emph{Reorder} & 10,000 & 0.002 & 18,132 & 13,646 .. 22,983 & 2,027 & 64 & 0.8822 \\ \hline
-\emph{Accumulate} & \emph{Normal}  & 10,000 & 0.005 & 21,428 & 17,536 .. 25,948 &   279 & 49 & 0.2525 \\ \hline
-\emph{Accumulate} & \emph{Dag}     & 10,000 & 0.002 & 24,855 & 18,222 .. 33,069 & 4,059 & 69 & 0.0736 \\ \hline
-\emph{Skip} & \emph{Dag}           & 10,000 & 0.002 & 25,678 & 16,170 .. 33,422 & 4,178 & 62 & 0.0211 \\ \hline
-\emph{Single} & \emph{Reorder}     & 100    & NA    & 26,558 & 20,678 .. 31,613 &    52 & 32 & 0.0076 \\ \hline
-\emph{Single} & \emph{Normal}      & 2,000  & NA    & 31,057 & 24,145 .. 42,278 &   141 & 43 & 0.0085 \\ \hline
-\emph{Single} & \emph{Dag}         & 100    & NA    & 34,208 & 29,840 .. 38,035 &    59 & 34 & 0 \\ \hline\hline
-
-	  \multicolumn{9}{|c|}{\textbf{4-16-bit Decode}, Kruskal-Wallis=$1.02\cdot 10^{-36}$} \\ \hline
-
-\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{Confidence Interval} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
-\emph{Skip} & \emph{Reorder}       &  2,000 & 0.002 &  62,853 & 50,033 .. 72,267 &   757 & 114 & 0.5247 \\ \hline
-\emph{Single} & \emph{Reorder}     &    500 &    NA &  63,731 & 56,070 .. 70,706 &   245 &  97 & 0.3610 \\ \hline
-\emph{Accumulate} & \emph{Normal}  &  1,000 & 0.002 &  67,390 & 60,738 .. 74,624 &   211 &  87 & 0.8767 \\ \hline
-\rowcolor{Gray}
-\emph{Skip} & \emph{Normal}        &  1,000 & 0.002 &  68,231 & 63,266 .. 74,463 &   202 &  85 & NA \\ \hline
-\emph{Single} & \emph{Normal}      &  1,000 &    NA &  68,819 & 62,940 .. 74,939 &   203 &  85 & 0.7695 \\ \hline
-\emph{Accumulate} & \emph{Reorder} &  2,000 & 0.002 &  72,991 & 67,163 .. 78,131 &   760 & 115 & 0.1868 \\ \hline
-\emph{Skip} & \emph{Dag}           & 10,000 & 0.001 & 132,821 & 117,034 .. 155,471 & 4,983 & 125 & 0 \\ \hline
-\emph{Accumulate} & \emph{Dag}     &  5,000 & 0.002 & 133,188 & 121,369 .. 147,073 & 2,542 & 125 & 0 \\ \hline
-\emph{Single} & \emph{Dag}         &    500 &    NA & 160,696 & 142,001 .. 160,696 &   299 &  97 & 0 \\ \hline\hline
-
-	  \multicolumn{9}{|c|}{\textbf{3-bit Multiply}, Kruskal-Wallis=$5.14\cdot 10^{-18}$} \\ \hline
-
-\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{Confidence Interval} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
-\emph{Skip} & \emph{Reorder}      & 10,000 & 0.0005& 196,116 & 160,073 .. 223,500  & 3,175 & 484 & 0  \\ \hline
-\emph{Accumulate} & \emph{Reorder}& 10,000 & 0.001 & 199,193 & 123,496 .. 275,013 & 3,117 & 489 & 0  \\ \hline
-\emph{Single} & \emph{Dag}        &   200  & NA    & 239,844 & 204,347 .. 286,418 & 114 & 74 & 0 \\ \hline
-\emph{Single} & \emph{Reorder}    & 500    & NA    & 243,664 & 178,566 .. 301,246 & 266 & 131 & 0 \\ \hline
-\emph{Skip} & \emph{Dag}          & 2,000  & 0.002 & 369,931 & 234,885 .. 483,753 & 952 & 149 & 0.0100 \\ \hline
-\emph{Accumulate} & \emph{Dag}    & 2,000  & 0.002 & 379,202 & 255,165 .. 460,783 & 973 & 158 & 0.0053  \\ \hline
-\emph{Single} & \emph{Normal}     & 5,000  & NA    & 399,911 & 280,070 .. 516,034 & 273 & 112 & 0.0047 \\ \hline
-\emph{Accumulate} & \emph{Normal} & 2,000  & 0.002 & 483,938 & 308,049 .. 611,034 & 168 & 87 & 0.45 \\ \hline
-\rowcolor{Gray}
-\emph{Skip} & \emph{Normal}       & 2,000  & 0.002 & 657,121 & 469,287 .. 864,073 & 175 & 84 & NA \\ \hline
-
-	\end{tabular}
-	\label{tab:results}
-\end{table*}
-
-The results from running each tuned parameter configuration for each algorithm
-combination on each problem are summarized by Table~\ref{tab:results}.
-Rows for each problem are ordered by that algorithm combination's median evaluations to
-success (MES), with the best algorithm combination appearing at the top of each section.
-The highlighted rows mark \emph{Skip-Normal}, the control algorithm combination
-representing traditional CGP performance.  All configurations for each problem
-were then tested using
-%Done-NEW
-Kruskal-Wallis\footnote{Kruskal-Wallis, a non-parametric equivalent of ANOVA, determines
-if samples were drawn from the same distribution.  If this test is significant, at
-least one of the samples differs from the others.}
-%/NEW
-to determine if any configuration was
-statistically different from any other.  All problems were determined to be highly significant.
-As such, each algorithm combination was then compared with the control
-to determine statistical significance using pairwise
-%Done-NEW
-Mann-Whitney~U\footnote{Mann-Whitney~U, a non-parametric equivalent of Student's t-test,
-determines if two populations were drawn from the same distribution.  If this test is significant,
-the populations are not equal.}
-%/NEW
-tests. The
-resulting p-value for each test is reported in the table.
-
-As discussed in Section~\ref{sec:parameter}, the genome size and mutation probability
-for each algorithm combination was extensively optimized.  In line with previous
-research~\cite{miller:2006:redundancy}, this led to massive genome sizes for
-almost all algorithm combinations on all problems.  With the exception of 
-some combinations using \emph{Single}, all algorithm combinations used over
-1,000 nodes.  Half of all combinations without \emph{Single} used 10,000 nodes.
-As this was the maximum allowed by our tuning, these combinations may work better with even larger
-genomes.
-
-Of the 12 configurations including \emph{Single}, 11 used a genome size smaller
-than all configurations not using \emph{Single}.  The exception was \emph{Single-Normal}
-on Multiply.  No other algorithm had as much impact on genome size tuning as
-the inclusion or exclusion of \emph{Single}.  In our previous work~\cite{goldman:2013:ordering}
-it appeared that \emph{Reorder} and \emph{DAG} allowed CGP to use smaller genomes.
-In light of our current results, this conclusion should be amended.
-It appears that \emph{Single} is most effective when the genome size is small and
-that \emph{Reorder} and \emph{DAG} are best able to cope with smaller genome sizes.
-Independent of duplication method, the ordering methods do not seem to create any
-pressure on which genome size is chosen.
-
-Unlike genome size, all configurations choose approximately the same mutation probability
-for each problem.  With the exception of Parity, there appears
-to be a general consensus that each gene should be mutated with probability 0.002,
-independent of genome size.  Parity used a higher mutation probability, but this is likely
-because 3-bit Parity is by far the easiest problem tested and therefore likely benefits
-from higher exploration.
-
-On many of the problems, many of the configurations have little to no statistical
-difference from the control in their evaluations required to reach the global optimum.
-On Parity and Encode it is probably safe to conclude that \emph{Single} is worse than
-the control.  As these are the easiest problems tested, that difference may have
-little real world meaning.  \emph{DAG} is almost certainly worse than the control on
-Decode.  Its results are so much worse that we can also likely conclude that \emph{DAG}
-is worse than all non \emph{DAG} configuration on that problem, even though no statistical
-tests are presented here.  The control on Multiply did by far the worst of all configurations,
-with only \emph{Accumulate Normal} likely failing statistical significance.
-
-By ordering the algorithm combinations by median evaluations to success (MES) and examining that ordering across each
-problem, we can look for general trends.  From the table, it appears that on easier problems
-the method of duplication used is the strongest predictor of rank, while
-the ordering methods are more predictive on harder problems.  This comes from the
-observation that in Parity and Encode (Upper half of Table~\ref{tab:results})
-the first column is effectively sorted, while for Decode and Multiply
-(Lower half of Table~\ref{tab:results})
-the second column is sorted.
-Each table includes coincidentally one exception to these rules.  On Parity and
-Encode, pairings with \emph{DAG} are ranked lower than other ordering methods, so much
-so as to break the ordering once on both problems.  On Decode \emph{Accumulate-Reorder}
-ranks worse than expected.  Finally on Multiply \emph{Single-DAG} does far better
-than expected.
-
-On almost every problem, \emph{DAG} required
-more evaluations to success than \emph{Reorder}, and it only outperformed \emph{Normal}
-on Multiply.  Combined with its's algorithmic overhead described in
-Section~\ref{sec:dag}, \emph{DAG} is therefore not likely to be useful for anything
-but comparison purposes.
-
-On all four problems, \emph{Reorder} finds the solution in either comparable
-or significantly less evaluations than \emph{Normal} ordering, independent of
-the duplicate method in use.  Combined with the complexity analysis in
-Section~\ref{sec:reorder} which shows it is no more asymptotically complex,
-we would suggest \emph{Reorder} as an alternative to \emph{Normal} on hard
-problems.
-
-The difference between using \emph{Skip} and \emph{Accumulate} remains
-negligible~\cite{goldman:2013:cgpwaste}, even in combination with different ordering
-methods.
-%Done-NEW
-This supports the idea that the two are actually achieving synonymous behavior
-through different means.
-%/NEW
-As \emph{Accumulate} is far more complex in terms of algorithm function
-than \emph{Skip}, we would suggest \emph{Skip} be used over \emph{Accumulate} for
-%Done-NEW Slight change here, "real world" became "future"
-any future applications.
-
-As a final check that our results do not somehow unfairly handicap traditional CGP,
-we compared our median evaluations to success with previous publications.  The results
-reported for traditional CGP in~\cite{walker:2008:cgpmodules}
-for 3-bit Parity required 6 times as many evaluations as \emph{Skip-Normal}.
-On the 3-bit Multiplier problem using a slightly different operator set,
-the same source had traditional CGP requiring 4 million evaluations to success,
-as opposed to our \emph{Skip-Normal} requiring 600 thousand, also a factor of 6 improvement.
-%DONE-NEW
-To determine the effect the difference in function set had on results, we ran
-\emph{Skip-Normal} and \emph{Skip-Reorder} using their functions but our parameters.
-The former improved to 453,684 while the latter increased to 303,657, maintaining
-a statistically significant difference.  These
-results make sense as the operator set in~\cite{walker:2008:cgpmodules} is able
-to more compactly represent the solution than our set.  This puts less strain
-on \emph{Normal} to deal with complex solutions, while \emph{Reorder}'s extremely
-large genome size is likely made unnecessary.  Note that as we did not tune
-using their operator set, our results should be considered a lower bound on quality for
-using \emph{Skip-Normal} and \emph{Skip-Reorder} to solve 3-bit Multiplier
-with their operators.
-%/NEW
-
-Our parameter tuning focused on minimizing the number of evaluations each algorithm
-combination required to solve problems, as that is the primary measure of runtime
-for CGP and GP in general.  As such, comparing the true runtime of the tuned parameters
-is likely to give potentially misleading results.  In using profilers and examining
-general trends, one of the largest indicators of runtime per evaluation was genome size.
-This makes sense as each time an offspring is generated all of the parent's genes
-must first be copied.  Furthermore, all genetic operations in CGP scale with genome size.
-Also, larger genome sizes correlate with more active nodes,
-each of which must be executed on each input tested during individual evaluation.
-
-With these caveats, the observed runtime for each algorithm combination follows
-what we would expect given the number of evaluations performed and the genome size.
-When \emph{Reorder} and \emph{Normal} use the same genome sizes, they require
-nearly the same amount of time per evaluation, with \emph{Reorder}'s added
-steps and increased number of active nodes constituting a relatively small increase.
-In general this difference in time per evaluation is trumped by differences in
-actual number of evaluations, which remains the primary cause of differential total runtime.
-As predicted, \emph{DAG} does take longer than the other algorithms to perform each evaluation,
-reinforcing the conclusion that \emph{DAG} should not be used for applications.
-
-As our implementation of \emph{Skip} and \emph{Accumulate} required each gene
-to generate a random number to determine if it should mutate, these methods took
-significantly longer per evaluation than \emph{Single}.  Utilizing a binomial
-distribution, identical behavior could have been achieved in effectively constant
-instead of linear time.  Further skewing the comparison is that \emph{Single}'s
-tuning resulted in far smaller genomes, giving it a significant edge over the other
-techniques in runtime per evaluation.
-
-If we had tuned using runtime instead of evaluations, we would expect whichever
-technique is most effective at small genome sizes to obtain the best results.
-From theoretical work in~\cite{goldman:2013:ordering}, we would expect \emph{Reorder}
-to work better with this limitation than \emph{Normal}.  Assuming the tuned values
-we obtained are any gauge, \emph{Single-Reorder} will likely require the least
-amount of runtime when tuned to reduce runtime.
-
-\begin{comment}
-\subsection{Scalability}
-TODO Discuss using tuned parameters on different problem sizes and analyze
-results.  Include O(N) table similar to previous paper.
-
-TODO Include scaling comparison with~\cite{walker:2008:cgpmodules}.
-\end{comment}
-
-\section{Analysis of Evolutionary Mechanisms}
-\label{sec:analysis}
-Comparing how many evaluations each algorithm requires to solve benchmark
-problems provides very little understanding of why one is more efficient than another.
-Thus we must turn to a more detailed view
-to support theorized capabilities of each variant and
-to discover hidden aspects of CGP.
-To accomplish this task we have devised novel metrics to expose what makes a
-CGP run successful.
-
-\subsection{Never Active Nodes}
-In line with the suggestions made in~\cite{miller:2006:redundancy}, the evolved
-number of active nodes across all problems and all algorithm configurations
-in Table~\ref{tab:results}
-was significantly lower than the genome size.  With the exception of a few combinations
-with \emph{DAG} and \emph{Single}, the evolved genomes were over 50\% inactive.
-\emph{Skip-Normal}, our control representing normal CGP, had only 1\%, 3\%, 20\%,
-and 9\% of nodes in the genome active in the median runs of Parity, Encode, Decode,
-and Multiply, respectively.  These final results fall in line with the predictions
-made in~\cite{goldman:2013:ordering}, in that if the problems are ordered based on
-their number of output locations, their evolved percentage of active nodes is also ordered.
-Furthermore, if Equation~\ref{eq:prob} is used to calculate the expected number of
-active nodes for configurations using \emph{Normal} ordering, we find that in all
-cases the evolved number is only marginally higher than the expectation.
-
-With such a small percentage of the final genome active, what is the rest of the
-genome for?
-Just because a node is not active in the final solution does not mean it was never
-useful.  For a node to be truly of no use to search, we must consider nodes that
-were never active.  Never active nodes are nodes such that for all ancestors of the
-final solution (the line of descent), no parent had that node active.  Note that
-for \emph{Reorder} this demarcation of node activity follows the node through
-each shuffle, such that if a node was marked as never active before the genome
-was shuffled it is marked as never active after the shuffle, and vice versa.
-This is in contrast to marking locations as never active, which was not done in
-this study.
-
-\begin{table*}
-	\caption{Location of never active nodes across all 50 runs for all nine algorithm combinations on the Multiply problem.}
-	\centering
-  \begin{tabular}{c|c|c|c|}
-    \cline{2-4}
-    & \textbf{\em Normal} & \textbf{\em Reorder} & \textbf{\em DAG} \\ \hline
-    \thirdlabel{Skip} & \graphicthird{multiply_skip_normal} &
-                        \graphicthird{multiply_skip_reorder} &
-                        \graphicthird{multiply_skip_dag}\\ \hline
-    \thirdlabel{Accumulate} & \graphicthird{multiply_accumulate_normal} &
-                              \graphicthird{multiply_accumulate_reorder} &
-                              \graphicthird{multiply_accumulate_dag}\\ \hline
-    \thirdlabel{Single} & \graphicthird{multiply_single_normal} &
-                          \graphicthird{multiply_single_reorder} &
-                          \graphicthird{multiply_single_dag}\\ \hline
-	\end{tabular}
-	\label{tab:never_active}
-\end{table*}
-
-Table~\ref{tab:never_active} provides information about never active nodes for
-each algorithm combination on the 3-bit Multiply problem.  The horizontal axis
-in each of the nine plots shows different genome locations, while the vertical
-axis denotes the level of fitness the recorded individual obtained.  The shading
-of each coordinate gives the percentage of 50 runs in which the first individual of
-the given fitness level had no ancestor in which the node at a given location was active.
-While only Multiply is shown here, all of the
-other problems tested had similar behavior, with allowances for differences in
-number of evaluations performed.
-
-Examining the columns of Table~\ref{tab:never_active} makes it clear that
-each of the ordering methods had significantly different never active behavior.
-Looking at \emph{Normal} we see that only at very high fitness levels does evolution
-begin to activate nodes in the latter half of the genome.  Even by the final solution
-large percentages of the genome were never activated, with \emph{Skip}, \emph{Accumulate},
-and \emph{Single} having median percentages of never active nodes of 43.35\%, 42.05\%, and
-56.17\%, respectively.  This is in stark contrast to \emph{Reorder} and \emph{DAG},
-which never had a median higher than 0.05\%.
-
-Compared to the difference in columns, the difference in the rows of Table~\ref{tab:never_active}
-caused by the different duplication methods were almost non-existent.  Only
-\emph{Single} has a distinguishable behavior, but this is almost certainly the result
-of the genome size and not the duplication method.  \emph{Single-Normal} used
-a genome size 2.5 times larger than \emph{Normal} did with the other techniques,
-so the resulting increase in never active nodes makes sense.  When \emph{Single}
-was paired with \emph{Reorder} and \emph{DAG} the genome sizes were far smaller,
-resulting in noisier graphs with less never active nodes.
-
-\begin{comment}
-So now the question is why does \emph{Normal} use so little of its genome?  To
-answer this we must consider what can prevent a node from becoming active.  For
-a node to never be active, all tested offspring with mutations that result in that node being activated
-must have been less fit than their parent.  Even if those offspring are no more fit, the selection method
-used by CGP would have allowed that offspring to replace its parent, at least probabilistically.  This leads
-to two possible causes for a node never being active: bad genes or infrequent node sampling.
-
-The bad genes cause suggests the behavior encoded
-by a node is simply useless on the problem landscape.  If this was the root cause
-of never active nodes, we would expect \emph{Reorder} and \emph{DAG} to have
-similar problems, as neither technique explicitly changes the likelihood of node
-behaviors.  Furthermore, we would expect the location of never active nodes to be
-uniform in the genome, which is obviously not the case.  Finally, even if bad
-genes were the problem, we would expect the mutational drift allowed by CGP's
-selection method to change their bad genes and eventually allow them to be active.
-
-The infrequent node sampling cause suggests that few
-individuals are ever produced which have that node active, limiting the likelihood
-of those individuals being selected regardless of potential fitness.  Under this
-cause and combined with Equation~\ref{eq:prob}, we
-expect a strong positional bias in the location of the never active nodes;
-we would expect nodes near the end of the genome to be the most likely to be never active.
-Figure~\ref{fig:prob} shows the expected probability given by Equation~\ref{eq:prob},
-using the same genome size, arity, number of inputs, and number of outputs as our
-3-bit multiply.  This curve bears a striking resemblance to all three \emph{Normal}
-plots in Table~\ref{tab:never_active}.
-Finally, from~\cite{goldman:2013:ordering} we would expect \emph{Reorder} and \emph{DAG}
-to have less or no positional bias, which also matches the results given in Table~\ref{tab:never_active}.
-\end{comment}
-
-%The above paragraphs were replaced by the below paragraph.
-In Section~\ref{sec:ordering} we used Equation~\ref{eq:prob} to suggest \emph{Normal}
-has a strong bias on the location of active nodes regardless of fitness landscape, and that
-\emph{Reorder} and \emph{DAG} would likely have less or no bias.
-%NEW
-The example inputs used in that section match those of \emph{Normal}'s tuned
-configuration on 3-bit Multiply.  The predicted number of active nodes and their
-location in the genome match almost exactly
-the behavior displayed in all three \emph{Normal} plots in Table~\ref{tab:never_active}.
-%/NEW
-
-Examining the progression in fitness across the different ordering methods shows
-the effect of initialization on each.  All three order methods share the same initialization
-method, and as such all have very similar never active node behavior early in
-evolution.  This initialization similarity explains why \emph{DAG} has any positional
-bias at all, even though its evolutionary mechanisms have none.  Similarly \emph{Reorder}
-diverges from \emph{Normal} given sufficient time, eventually losing its positional
-bias.
-
-This early similarity may be part of the reason \emph{Normal} and \emph{Reorder}
-have such similar behavior on relatively easy problems.  After relatively few
-evaluations both have explored similar sections of their genome.  Yet on hard
-problems that require further exploration, \emph{Reorder} is able to do so, while
-\emph{Normal} remains positionally limited.
-
-\subsection{Node Behaviors}
-\label{sec:behavior}
-
-In order to understand how each node is being used by CGP,
-we first determine the semantics of each node~\cite{mcphee:2008:treesemantics}.
-By semantics we mean that for each input to an individual, each node
-will produce a single output.  These output values can be recorded and ordered.
-As the domain in use is binary, the outputs can be viewed as a bit string, where
-each bit indicates the value a node outputs when an individual is presented with a
-specific input.  Similar recording can be done for other domains, but binary
-allows for by far the simplest encoding.
-
-This semantics representation of a node is a complete description of its functionality.
-Any two nodes with an identical semantic representation have identical behavior,
-even if their method for encoding that behavior is drastically different.  From
-this perspective, we can view CGP evolution as attempting to evolve a node who's
-semantics match the desired output semantics.  Doing so allows
-us to examine how the evolutionary operators change the behavior of nodes,
-not just their gene values.
-
-Utilizing semantic analysis, we can extract the useful
-portions of an individual, a subset of the active nodes.  Consider two active
-nodes X and Y with the same semantics such that X proceeds Y in the genome.
-As CGP is a graph representation, only
-one of these nodes is actually necessary.  Any node that reads the output of Y
-can instead read the output of X, removing the need to include Y in the solution.
-
-\begin{figure}
-  \begin{algorithmic}
-  \Procedure{Simplify}{}
-    \State $reconnect \leftarrow \emptyset$
-    \ForAll{$node \in reversed(active)$}
-      \State $reconnect[semantics(node)] \leftarrow node$
-    \EndFor
-    \ForAll{$i \in input\_locations$}
-      \State $reconnect[semantics(i)] \leftarrow i$
-    \EndFor
-    \ForAll{$g \in genes$}
-      \If{$g$ is a connection gene}
-        \If{$semantics(g) \in reconnect$}
-          \State $g \leftarrow reconnect[semantics(g)]$
-        \EndIf
-      \EndIf
-    \EndFor
-  \EndProcedure
-  \end{algorithmic}
-  \caption{Algorithm to ensure each active node calculates
-           a unique semantic.}
-  \label{fig:simplify}
-\end{figure}
-
-The \Call{Simplify}{} algorithm given in Figure~\ref{fig:simplify} is designed
-to efficiently remove redundancy encoded in a genome.  This
-algorithm moves all connections to the first node that produces the desired
-semantic.  After application, and recalculation of which nodes are still
-active, each active node produces a unique semantic string.  Any node that is
-transitively dependent on its own semantic will always be removed from
-the active set.  In general, if a set of nodes all produce the same semantic,
-whichever is predicted to have the least transitive dependencies will be kept.  Note that this
-does not guarantee the minimum possible genome, as that would require
-exponential runtime.  The \Call{Simplify}{} algorithm requires $O(AN)$ time,
-since the final \textbf{for all} loop is the most expensive step.  As before, this means
-\Call{Simplify}{} is no more complex than copying a genome.
-
-\begin{table*}
-	\centering
-	\caption{Average node behavior for all nine algorithm combinations on the Encode problem.
-	         Excludes never active nodes.}
-  \begin{tabular}{c|c|c|c|}
-    \cline{2-4}
-    & \textbf{\em Normal} & \textbf{\em Reorder} & \textbf{\em DAG} \\ \hline
-    \thirdlabel{Skip} & \graphicthird{bar_encode_skip_normal} &
-                        \graphicthird{bar_encode_skip_reorder} &
-                        \graphicthird{bar_encode_skip_dag}\\ \hline
-    \thirdlabel{Accumulate} & \graphicthird{bar_encode_accumulate_normal} &
-                              \graphicthird{bar_encode_accumulate_reorder} &
-                              \graphicthird{bar_encode_accumulate_dag}\\ \hline
-    \thirdlabel{Single} & \graphicthird{bar_encode_single_normal} &
-                          \graphicthird{bar_encode_single_reorder} &
-                          \graphicthird{bar_encode_single_dag}\\ \hline
-	\end{tabular}
-	\label{tab:encode_behavior}
-\end{table*}
-
-\begin{table*}
-	\centering
-	\caption{Average node behavior for all nine algorithm combinations on the Multiply problem.
-	         Excludes never active nodes.}
-  \begin{tabular}{c|c|c|c|}
-    \cline{2-4}
-    & \textbf{\em Normal} & \textbf{\em Reorder} & \textbf{\em DAG} \\ \hline
-    \thirdlabel{Skip} & \graphicthird{bar_multiply_skip_normal} &
-                        \graphicthird{bar_multiply_skip_reorder} &
-                        \graphicthird{bar_multiply_skip_dag}\\ \hline
-    \thirdlabel{Accumulate} & \graphicthird{bar_multiply_accumulate_normal} &
-                              \graphicthird{bar_multiply_accumulate_reorder} &
-                              \graphicthird{bar_multiply_accumulate_dag}\\ \hline
-    \thirdlabel{Single} & \graphicthird{bar_multiply_single_normal} &
-                          \graphicthird{bar_multiply_single_reorder} &
-                          \graphicthird{bar_multiply_single_dag}\\ \hline
-	\end{tabular}
-	\label{tab:multiply_behavior}
-\end{table*}
-
-In CGP it is common to report what fraction of the genome is active, as this represents
-the easiest estimate of solution size.  By utilizing semantic comparison
-and the \Call{Simplify}{} algorithm, we can get a much better estimate of what
-purpose each node in the genome serves.
-Table~\ref{tab:results} provides two columns relevant to node behavior: \textbf{Active}
-and \textbf{Used}.  The former gives the traditional measure of the median amount of
-active nodes in the solution found.  The latter is the median number of nodes
-still active after applying the \Call{Simplify}{} algorithm.
-
-On all problems and for all algorithm combinations, the number of active nodes is
-far smaller than the genome size, and the number of used nodes
-is far smaller than the number of active nodes.  In some extreme cases less
-than 1\% of active nodes in the evolved solution are actually necessary.  Only the
-most extreme configurations have even 50\% of active nodes in use by the
-simplified solution.
-
-Table~\ref{tab:encode_behavior} and Table~\ref{tab:multiply_behavior} help illustrate
-the behavioral breakdown of nodes for each algorithm combination on the Encode
-and Multiply problems, respectively.  These two problems were chosen as they
-represent different ends of the quality spectrum for \emph{Skip-Normal} and
-because together they are sufficiently representative of behaviors seen on the other
-problems.
-
-Each diagram describes node behavior as falling into eight distinct categories
-split between active and inactive nodes.  Nodes are put into the first category
-they satisfy.
-\begin{enumerate}
-\item ``Constant'' - Independent of problem input, constant nodes produce a fixed output.
-\item ``Used'' - Nodes still active after applying \Call{Simplify}{}.
-\item ``Useful'' - Nodes with identical semantics to a ``Used'' node.
-\item ``Intron'' - Any node with semantics identical to a node that was
-                   active before applying \Call{Simplify}{}, but was inactive afterward.
-\item ``Explore'' - All remaining nodes.
-\end{enumerate}
-Used nodes are by definition only in the active portion of the genome.  Similarly
-Explore nodes can only be in the inactive portion of the genome.   Note that
-all never active nodes are ignored by these plots, as their behavior is meaningless.
-
-The first striking feature of these tables is how many nodes produce constant
-outputs.  As the instruction set used in our genomes never require constant values
-to produce any non constant behavior (TRUE NAND X $\Leftrightarrow$ X NAND X)
-and most operations with constants result in pass through nodes
-(FALSE OR X $\Leftrightarrow$ X), we would not expect evolution to proliferate
-these constant nodes.  If these constants are in fact of no use, future work
-could likely improve CGP's performance by reducing their prevalence.  Yet
-it is possible they serve an evolutionary function by allowing mutation more
-flexibility.  For instance, pass through nodes may be allowing mutation to create
-adjustments not easily done if this behavior was forbidden.
-
-As with previous analysis, \emph{Skip} and \emph{Accumulate} have nearly identical
-plots, regardless of ordering method and problem.  \emph{Single} again has
-a significantly different plot when paired with \emph{Reorder} or \emph{DAG},
-with some hits of a change on Encode with \emph{Normal}.  These combinations
-were also the most likely to have significantly smaller genome sizes.  From
-Table~\ref{tab:results} we see that the number of used nodes on each problem is relatively
-constant across the combinations even though the number of active nodes and the
-genome sizes changed significantly.  As such, the proportion of nodes in the
-used category will vary as these other features change.
-
-From Table~\ref{tab:encode_behavior} and Table~\ref{tab:multiply_behavior}, having
-the plurality of active nodes in the ``Used'' category appears to correlate with
-increase evaluations to success in Table~\ref{tab:results}.  Notice that in all \emph{Single} plots the
-tallest bar for active nodes is ``Used,'' and the only non \emph{Single} plots
-where this is true are \emph{Normal} on Multiply.  On Encode, \emph{Single}
-runs had the highest median evaluations to success, with \emph{Normal} runs
-doing similarly on Multiply.  This appears to imply that having duplication
-of node behavior in the active nodes, and not just inactive nodes, is beneficial
-to search.  If this is true, it may help explain why CGP uses such large genomes,
-as \emph{Normal} is only able to achieve enough active nodes to allow duplication
-when the genome size significantly exceeds the actual solution size.  Also
-the success of \emph{Reorder} and \emph{DAG} on Multiply may also be partially be attributable
-to this ability to increase the number of active nodes.
-
-One significant difference in behavior between the two problems is proportion of
-Explore nodes.  A potential explanation is that on 16-4-bit Encode there are only
-$2^{16} = 65,536$ unique semantics while on 3-bit Multiply there are $2^{64} = 1.845\cdot 10^{19}$
-unique semantics.  Combined with the fact that not all semantics are equally
-likely to be produced, the difference may just be that on Encode nodes are
-that much more likely to recreate existing behavior.  Another possibility is that
-CGP increases the semantic diversity of inactive nodes over evolutionary time.
-As such, the fact that Multiply required approximately an order of magnitude more
-evaluations to solve would lead to the discrepancy.
-
-\subsection{Parent Replacement}
-
-\begin{table}
-	\centering
-	\caption{Behavior of offspring replacement for each algorithm combination
-	         on each problem.
-	         Act-Un=Active Unchanged; No React=No Reactivation; React Changed=Reactivation Changed}
-	\begin{tabular}{|c|c|c|c|c|}
-	  \hline
-	  \multicolumn{5}{|c|}{\textbf{3-bit Parity}} \\ \hline
-%\textbf{Duplicate} & \textbf{Ordering} & \textbf{Active Unchanged} & \textbf{No Reactivation} & \textbf{Reactivated Changed} \\ \hline
-\textbf{Duplicate} & \textbf{Ordering} & \textbf{Act-Un} & \textbf{No React} & \textbf{React Changed} \\ \hline
-       \emph{Skip} &     \emph{DAG} & 0.00\% & 3.48\% & 55.63\% \\ \hline
- \emph{Accumulate} &     \emph{DAG} & 0.00\% & 8.48\% & 52.82\% \\ \hline
- \emph{Accumulate} & \emph{Reorder} & 2.09\% & 33.26\% & 58.46\% \\ \hline
-   \rowcolor{Gray}
-       \emph{Skip} &  \emph{Normal} & 2.25\% & 39.35\% & 67.17\% \\ \hline
- \emph{Accumulate} &  \emph{Normal} & 2.69\% & 39.99\% & 67.27\% \\ \hline
-       \emph{Skip} & \emph{Reorder} & 11.46\% & 53.99\% & 59.47\% \\ \hline
-     \emph{Single} & \emph{Reorder} & 24.94\% & 80.49\% & 60.63\% \\ \hline
-     \emph{Single} &  \emph{Normal} & 26.44\% & 83.55\% & 59.96\% \\ \hline
-     \emph{Single} &     \emph{DAG} & 27.80\% & 83.73\% & 59.92\% \\ \hline\hline
-
-  \multicolumn{5}{|c|}{\textbf{16-4-bit Encode}} \\ \hline
-
-%\textbf{Duplicate} & \textbf{Ordering} & \textbf{Active Unchanged} & \textbf{No Reactivation} & \textbf{Reactivated Changed} \\ \hline
-\textbf{Duplicate} & \textbf{Ordering} & \textbf{Act-Un} & \textbf{No React} & \textbf{React Changed} \\ \hline
- \emph{Accumulate} &     \emph{DAG} & 0.05\% & 4.06\% & 48.70\% \\ \hline
-       \emph{Skip} &     \emph{DAG} & 0.06\% & 3.85\% & 48.33\% \\ \hline
- \emph{Accumulate} & \emph{Reorder} & 1.77\% & 19.24\% & 55.47\% \\ \hline
-       \emph{Skip} & \emph{Reorder} & 2.02\% & 19.80\% & 56.15\% \\ \hline
-   \rowcolor{Gray}
-       \emph{Skip} &  \emph{Normal} & 14.16\% & 54.78\% & 64.14\% \\ \hline
- \emph{Accumulate} &  \emph{Normal} & 14.52\% & 54.79\% & 64.17\% \\ \hline
-     \emph{Single} &  \emph{Normal} & 33.77\% & 79.34\% & 59.86\% \\ \hline
-     \emph{Single} &     \emph{DAG} & 35.98\% & 80.91\% & 58.39\% \\ \hline
-     \emph{Single} & \emph{Reorder} & 39.95\% & 78.91\% & 59.72\% \\ \hline\hline
-
-  \multicolumn{5}{|c|}{\textbf{4-16-bit Decode}} \\ \hline
-
-%\textbf{Duplicate} & \textbf{Ordering} & \textbf{Active Unchanged} & \textbf{No Reactivation} & \textbf{Reactivated Changed} \\ \hline
-\textbf{Duplicate} & \textbf{Ordering} & \textbf{Act-Un} & \textbf{No React} & \textbf{React Changed} \\ \hline
- \emph{Accumulate} &     \emph{DAG} & 0.03\% & 10.10\% & 54.99\% \\ \hline
-       \emph{Skip} &     \emph{DAG} & 0.05\% & 10.21\% & 51.33\% \\ \hline
- \emph{Accumulate} & \emph{Reorder} & 9.98\% & 49.65\% & 70.57\% \\ \hline
-       \emph{Skip} & \emph{Reorder} & 12.53\% & 51.63\% & 70.24\% \\ \hline
-     \emph{Single} & \emph{Reorder} & 28.38\% & 79.73\% & 66.85\% \\ \hline
-     \emph{Single} &  \emph{Normal} & 29.43\% & 80.86\% & 65.85\% \\ \hline
-     \emph{Single} &     \emph{DAG} & 32.60\% & 81.73\% & 57.70\% \\ \hline
- \emph{Accumulate} &  \emph{Normal} & 51.11\% & 84.54\% & 72.57\% \\ \hline
-   \rowcolor{Gray}
-       \emph{Skip} &  \emph{Normal} & 65.47\% & 89.17\% & 72.21\% \\ \hline\hline
-
-  \multicolumn{5}{|c|}{\textbf{3-bit Multiply}} \\ \hline
-
-%\textbf{Duplicate} & \textbf{Ordering} & \textbf{Active Unchanged} & \textbf{No Reactivation} & \textbf{Reactivated Changed} \\ \hline
-\textbf{Duplicate} & \textbf{Ordering} & \textbf{Act-Un} & \textbf{No React} & \textbf{React Changed} \\ \hline
- \emph{Accumulate} & \emph{Reorder} & 1.14\% & 22.09\% & 71.75\% \\ \hline
- \emph{Accumulate} &     \emph{DAG} & 4.52\% & 45.38\% & 63.42\% \\ \hline
-       \emph{Skip} &     \emph{DAG} & 5.05\% & 45.58\% & 63.03\% \\ \hline
-       \emph{Skip} & \emph{Reorder} & 10.40\% & 47.67\% & 71.79\% \\ \hline
-     \emph{Single} &  \emph{Normal} & 18.85\% & 78.19\% & 70.11\% \\ \hline
-     \emph{Single} &     \emph{DAG} & 20.72\% & 80.20\% & 65.10\% \\ \hline
-     \emph{Single} & \emph{Reorder} & 22.14\% & 80.21\% & 69.76\% \\ \hline
- \emph{Accumulate} &  \emph{Normal} & 57.69\% & 87.27\% & 78.95\% \\ \hline
-   \rowcolor{Gray}
-       \emph{Skip} &  \emph{Normal} & 73.46\% & 92.04\% & 79.07\% \\ \hline
-	\end{tabular}
-	\label{tab:reactivate}
-\end{table}
-
-An important aspect of CGP is its replacement strategy that allows for neutral drift.
-In each generation the best offspring replaces its parent if it is no less fit.  Previous
-research has shown the negative impact removing this feature can have on
-performance~\cite{yu:2001:neutrality}. However, how it interacts with the different variants
-of CGP has not been investigated previously.
-
-Table~\ref{tab:reactivate} provides a
-novel look at CGP's replacement behavior.  Each time an offspring replaced a parent
-in 50 runs, we recorded how the behavior of each node differed between the
-parent and the offspring.
-
-The \textbf{Active Unchanged} column reports how often all active nodes in the offspring
-have identical semantics to the corresponding nodes in the offspring's parent.
-This goes beyond the concept of actively identical offspring and checks if the behavior
-of any active node was modified by the mutation.  The rows in each of the four
-tables are sorted using this column.  Smaller values indicate an algorithm
-combination that produced more changes in active node behavior.
-
-The \textbf{No Reactivation} column uses the word reactivation to refer to nodes
-that were active in some ancestor of the offspring, were inactive in the offspring's
-parent, but are now again active in the offspring.  Reactivation is a measure of how
-often CGP reuses previously useful active nodes.  Reported in the table is how
-frequently an offspring replaces its parent such that is has no reactivated nodes.  Smaller
-values indicate an algorithm combination that frequently reuses nodes, as opposed to
-including randomly generated nodes (never active nodes) or ignoring inactive nodes
-entirely.
-
-The \textbf{Reactivated Changed} column shows the percentage of reactivated nodes
-that have a different semantic from the last time they were active.  Conversely,
-reactivated nodes without a change mean that CGP has taken a previously active behavior,
-stored it in the inactive space, and reactivated it in later mutations.  Smaller values
-indicate an algorithm combination that more often preserve node behaviors while
-inactive.
-
-Combined these three measures provide some surprising insights.  Most notably
-is that \emph{Single}, a technique designed to force offspring to be actively different
-from their parents, produces offspring with no changes in active node behavior
-20\%-40\% of the time.  This further exposes CGP's ability to have genetically different
-yet phenotypically identical individuals.  Even more surprising is that \emph{Single}
-appears to increase the chance of producing offspring without changing active behavior.
-Somewhat unexpectedly \emph{Normal} appears to also reduce changes to active node behavior,
-causing the largest effect after \emph{Single}.
-
-The likely cause of both is the number
-of active nodes receiving mutations each generation.  Even though \emph{Single}
-forces a gene in an active node to be mutated, it limits mutation to only one change.
-If this one change does not change that node's behavior, the offspring will be marked
-as active unchanged.  Conversely, \emph{Skip} and \emph{Accumulate} can mutate any
-number of active genes each generation, improving their chance of a change in behavior.
-Similarly, even though \emph{Normal} makes no explicit change in mutation, it is characterized
-by a significantly smaller number of active nodes than \emph{Reorder} or \emph{DAG}.  More
-active nodes means more chances for mutation to change the behavior of at least one.
-
-Somewhat unintuitively, node reactivation and active node semantic changes appear
-to be highly correlated.  This is apparent in the fact that the \textbf{No Reactivation}
-%TODO Changed here
-column in each table is almost perfectly sorted, even though rows were sorted by the
-\textbf{Active Unchanged} column.  Again the number of active nodes is the likely
-cause.  Having fewer active nodes reduces the number of introns, making it more
-difficult for inactive nodes to become active.  Fewer active nodes also decreases
-how many inactive nodes could have ever been active.
-
-Potentially the most interesting results come from the control combination of
-\emph{Skip-Normal}.  As problem difficulty increases, \emph{Skip-Normal} produces
-an increasingly large number of offspring with no change in active node behavior,
-with little node reactivation, and a high chance of changing nodes before
-they are reactivated.  This suggests two important features: \emph{Skip} is likely
-saving a significant number of evaluations, and \emph{Normal} CGP is likely using
-inactive nodes primarily to inject random behavior.  The former comes from the
-%TODO Changed here
-argument that most of the active unchanged individuals are also likely
-actively identical to their parents.  The latter is supported by the fact that
-\emph{Skip-Normal} infrequently reactivates a node with tested behavior with
-offspring frequently connecting in untested inactive nodes.
-
-% An example of a floating figure using the graphicx package.
-% Note that \label must occur AFTER (or within) \caption.
-% For figures, \caption should occur after the \includegraphics.
-% Note that IEEEtran v1.7 and later has special internal code that
-% is designed to preserve the operation of \label within \caption
-% even when the captionsoff option is in effect. However, because
-% of issues like this, it may be the safest practice to put all your
-% \label just after \caption rather than within \caption{}.
-%
-% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
-% option should be used if it is desired that the figures are to be
-% displayed while in draft mode.
-%
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{myfigure}
-% where an .eps filename suffix will be assumed under latex, 
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-%\caption{Simulation Results.}
-%\label{fig_sim}
-%\end{figure}
-
-% Note that IEEE typically puts floats only at the top, even when this
-% results in a large percentage of a column being occupied by floats.
-
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command,
-% and the \label for the overall figure must come after \caption.
-% \hfil is used as a separator to get equal spacing.
-% Watch out that the combined width of all the subfigures on a 
-% line do not exceed the text width or a line break will occur.
-%
-%\begin{figure*}[!t]
-%\centering
-%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
-%\label{fig_second_case}}
-%\caption{Simulation results.}
-%\label{fig_sim}
-%\end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat[]), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
-
-% An example of a floating table. Note that, for IEEE style tables, the 
-% \caption command should come BEFORE the table. Table text will default to
-% \footnotesize as IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
-%\begin{table}[!t]
-%% increase table row spacing, adjust to taste
-%\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-%\caption{An Example of a Table}
-%\label{table_example}
-%\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{tabular}{|c||c|}
-%\hline
-%One & Two\\
-%\hline
-%Three & Four\\
-%\hline
-%\end{tabular}
-%\end{table}
-
-% Note that IEEE does not put floats in the very first column - or typically
-% anywhere on the first page for that matter. Also, in-text middle ("here")
-% positioning is not used. Most IEEE journals use top floats exclusively.
-% Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
-% floats. This can be corrected via the \fnbelowfloat command of the
-% stfloats package.
-
-\section{Conclusion}
-\label{sec:conclusion}
-\begin{comment}
-Even with CGP's relatively simple evolutionary mechanisms, fully understanding
-the effects those mechanisms have on search takes concerted effort.  By examining
-the effectiveness as well as the actual behavior of traditional CGP and
-a handful of variants we have helped advance this understanding.
-\end{comment}
-
-Generally, CGP appears to require the least number of evaluations when the genome
-size far exceeds what is necessary for the problem at hand.  When properly tuned,
-almost all variants had only a tiny fraction of the genome active in their final
-solution.  Furthermore, a significant portion of these active nodes were found
-to be redundant; removable using a very simple reduction algorithm.  The only
-algorithm combinations that preferred more reasonable genome sizes were those using \emph{Single},
-especially when paired with \emph{Reorder} or \emph{DAG}.  Yet we theorize this may
-have more to do with \emph{Single} scaling poorly with the number of active nodes
-than with it overcoming whatever causes traditional CGP to prefer enormous genomes.
-\begin{comment}
-While \emph{Reorder} and \emph{DAG} were able to use smaller genomes than \emph{Normal}
-when paired with \emph{Single}, this decrease in size was not observed in any other
-pairings.  As such, the suggestion in~\cite{goldman:2013:ordering} that these
-techniques generally reduce genome size should be discarded.  Instead it is likely
-more correct to say that \emph{Reorder} and \emph{DAG} are less detrimentally effected
-by reducing the genome size.
-\end{comment}
-
-Compared on median evaluations to success, \emph{Normal} and \emph{Reorder} are
-likely to be equally good for simple problems, with some suggestion that \emph{Reorder}
-performs better as problems become more difficult.  In general \emph{DAG} was found
-to be unnecessarily time consuming, as it required higher algorithmic complexity
-without any performance gain.  As such we would suggest \emph{Reorder} for general
-future CGP use.
-
-In line with the results from~\cite{goldman:2013:cgpwaste}, \emph{Skip} and
-\emph{Accumulate} continued to show no significant difference in effectiveness
-or behavior.  As \emph{Skip} is simpler to understand and implement, we suggest
-\emph{Accumulate} be ignored in all future CGP applications.  \emph{Single}'s
-behavior continues to be enough different from \emph{Skip} to warrant further
-study, especially with problems requiring higher numbers of active genes.  In
-general though, if evaluation time is significantly impacted by the number of
-active nodes or the total size of the genome, we suggest \emph{Single}.  While
-mutation probability appears relatively independent of problem and algorithm combination (all of our tuning
-resulted in 0.002 being optimal or close to it), \emph{Single} still has the
-advantage of not requiring this parameter at all.
-
-As to actual behavior, CGP genomes include a surprising amount of redundant and
-unused nodes.  As theoretically predicted in~\cite{goldman:2013:ordering},
-\emph{Normal} ordering leaves large sections of the genome completely unused
-while \emph{Reorder} and \emph{DAG} are more likely to utilize the whole genome.
-Further analysis showed that CGP, at least with our function set, produced a large
-number of constant nodes, such that their output is independent of problem
-inputs.  As the binary domain only has two constants (always true and always false),
-each genome at most should require two constant nodes.  Even so these behaviors
-were found duplicated in both the active and inactive portions of final solutions
-for all algorithm variants.  Also common was duplication of node behavior, even
-though CGP's encoding does not require any duplication to produce any value.  Yet
-attempting to prevent or remove this behavioral duplication may hinder CGP's evolutionary
-success, as it will limit neutral drift.
-
-CGP's use of neutrality has become one of its cornerstone arguments for its success.
-Yet the reason for neutrality's importance appears to be different from previous
-arguments.  In our testing of traditional CGP we found large sections of the
-genome were never used by any ancestor of the final solution.  Furthermore,
-offspring almost never include active nodes that were inactive in their direct
-parent, but active in a previous ancestor.  Even when reactivation does happen,
-the node has almost always changed behavior significantly from when it was active.
-As such we suggest the advantages of neutrality in regards to inactive nodes is
-not that it allows the storage of previously useful structures, but that it allows
-inactive nodes to be constantly changed facilitating mutational inclusion of what
-are effectively random nodes.  Under this theory, CGP would likely benefit from periodically resetting
-all inactive nodes.  Note that with \emph{Reorder}
-and \emph{DAG} these conclusions do not hold as a reasonable amount of node
-reactivation occurs.
-
-While this analysis was limited to boolean problems by necessity of some of
-the analysis methods, most of the conclusions have no obvious reason to be
-limited to the boolean domain.  Still, applying semantic analysis
-to detect changes in behavior and behavior duplication, along with looking at
-never active nodes and tuned parameter settings, on other domains will help
-solidify our conclusions.
-
+\IEEEPARstart{T}{TODO}
+
+~\cite{Wolpert:1997:nfl}
+
+~\cite{goldberg:1991:gasize} % Search efficiency
+
+~\cite{grefenstette:1986:optimalga} % Expensive to tune
+
+~\cite{harik:1999:parameterlessga}
+
+~\cite{posik:2011:parameterless} % Model based parameter-less, only noiseless non-overlapping
+
+~\cite{goldman:2014:p3}
+
+\section{Optimizers}
+
+\subsection{P3}
+~\cite{goldman:2014:p3}
+
+~\cite{hornby:2006:alps}
+
+~\cite{gronau:2007:upgma} % Efficient clustering
+
+\subsection{Hill Climber}
+~\cite{goldman:2014:p3}
+
+\subsection{LTGA}
+~\cite{thierens:2010:ltga}
+
+~\cite{thierens:2013:ltgahiff}
+
+~\cite{bosman:2011:lsbbo} % LTGA with hill climbing
+
+\subsection{hBOA}
+
+\subsection{Parameter-less hBOA}
+~\cite{pelikan:2004:parameterlesshboa}
+
+\subsection{$1+(\lambda, \lambda)$}
+~\cite{doerr:2013:lambdalambda}
+
+~\cite{goldman:2014:p3}
+
+\section{Problem Descriptions}
+\subsection{Single Instance Problems}
+~\cite{goldberg:1991:gasize} % Deceptive Trap
+
+~\cite{goldman:2012:ltga} % Deceptive Step Trap
+
+~\cite{thierens:2013:ltgahiff} % HIFF
+
+~\cite{goldman:2014:p3} % Rastrigin
+
+\subsection{Randomly Generated Problem Classes}
+
+~\cite{wright:2000:solvingnk}
+
+~\cite{saul:1994:spinglass}
+\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}
+
+~\cite{goldman:2014:p3} % MAXSAT
+
+\section{Comparison Algorithm Parameter Tuning}
+~\cite{goldman:2012:ltga} % Bisection with no parameters
+
+~\cite{goldman:2014:p3}
+
+~\cite{jovanovic:1997:ruleofthree}
+
+\section{Global Optimum}
+TODO Compare all 6 optimization techniques on time to reach global optimum
+
+TODO Table listing empirical $O(N)$ times for each technique on each problem. Include
+statistical comparison of log-log regression lines.
+
+TODO Consider a graphic showing 5\% and 95\% Quantiles to illustrate P3's scaling on problems
+
+\section{Stop Anytime}
+TODO Compare all 6 optimization techniques on fitnesses reached at different times
+during the run on problem sizes near 100 or 150
+
+TODO Compare all 4 optimization techniques on fitnesses reached at different times
+during the run on problem sizes near 500 or 750
+
+\section{Complexity Per Evaluation}
+TODO Include graphic of rebuilds per evaluation (decreasing with length)
+
+TODO Include graphic of donations per evaluation (sub-linear growth)
+
+TODO Include graphic comparing seconds per evaluation up to small problems (hboa super-linear)
+
+TODO Include graphic comparing seconds per evaluation for large problems (perhaps just 1 graph of P3 across all problems)
+
+\section{Inner Workings}
+TODO Include graphic of proportion of evaluations spend on hill climber (constant with length)
+
+TODO Include graphic of crossover success (constant with length)
+
+TODO Include graphic of Ising Spin Glass, NK, and DeceptiveStepTrap's pyramid shape and crossover success rates at each level
+
+~\cite{lobo:2011:dynamicpop}
+
+\section{Conclusions and Future Work}
 % if have a single appendix:
@@
 
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{goldman_headshot}}]{Brian W. Goldman}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{headGoldman}}]{Brian W. Goldman}
 received the B. S. degree and M.S. degree in computer science from the

### 2014-08-20 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Switched over to EC Journal style.

@@
-%% bare_jrnl.tex
-%% V1.4
-%% 2012/12/27
-%% by Michael Shell
-%% see http://www.michaelshell.org/
-%% for current contact information.
-%%
-%% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.8 or later) with an IEEE journal paper.
-%%
-%% Support sites:
-%% http://www.michaelshell.org/tex/ieeetran/
-%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
-%% and
-%% http://www.ieee.org/
-
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. IEEE's font choices can trigger bugs that do  ***
-% *** not appear when using other class files.                            ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
-
-%%*************************************************************************
-%% Legal Notice:
-%% This code is offered as-is without any warranty either expressed or
-%% implied; without even the implied warranty of MERCHANTABILITY or
-%% FITNESS FOR A PARTICULAR PURPOSE! 
-%% User assumes all risk.
-%% In no event shall IEEE or any contributor to this code be liable for
-%% any damages or losses, including, but not limited to, incidental,
-%% consequential, or any other damages, resulting from the use or misuse
-%% of any information contained here.
-%%
-%% All comments are the opinions of their respective authors and are not
-%% necessarily endorsed by the IEEE.
-%%
-%% This work is distributed under the LaTeX Project Public License (LPPL)
-%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
-%% distributed and modified. A copy of the LPPL, version 1.3, is included
-%% in the base LaTeX documentation of all distributions of LaTeX released
-%% 2003/12/01 or later.
-%% Retain all contribution notices and credits.
-%% ** Modified files should be clearly indicated as such, including  **
-%% ** renaming them and changing author support contact information. **
-%%
-%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
-%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex,
-%%                    bare_jrnl_transmag.tex
-%%*************************************************************************
-
-% Note that the a4paper option is mainly intended so that authors in
-% countries using A4 can easily print to A4 and see how their papers will
-% look in print - the typesetting of the document will not typically be
-% affected with changes in paper size (but the bottom and side margins will).
-% Use the testflow package mentioned above to verify correct handling of
-% both paper sizes by the user's LaTeX system.
-%
-% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
-% should be used if it is desired that the figures are to be displayed in
-% draft mode.
-%
-\documentclass[journal]{IEEEtran}
-%
-% If IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it like:
-% \documentclass[journal]{../sty/IEEEtran}
-
-%\linespread{2.5}
-
-% Some very useful LaTeX packages include:
-% (uncomment the ones you want to load)
-
-% *** MISC UTILITY PACKAGES ***
-%
-%\usepackage{ifpdf}
-% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
-% compilation based on whether the output is pdf or dvi.
-% usage:
-% \ifpdf
-%   % pdf code
-% \else
-%   % dvi code
-% \fi
-% The latest version of ifpdf.sty can be obtained from:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
-% Also, note that IEEEtran.cls V1.7 and later provides a builtin
-% \ifCLASSINFOpdf conditional that works the same way.
-% When switching from latex to pdflatex and vice-versa, the compiler may
-% have to be run twice to clear warning/error messages.
-
-% *** CITATION PACKAGES ***
-%
-%\usepackage{cite}
-% cite.sty was written by Donald Arseneau
-% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
-% \cite{} output to follow that of IEEE. Loading the cite package will
-% result in citation numbers being automatically sorted and properly
-% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
-% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
-% \cite will automatically add leading space, if needed. Use cite.sty's
-% noadjust option (cite.sty V3.8 and later) if you want to turn this off
-% such as if a citation ever needs to be enclosed in parenthesis.
-% cite.sty is already installed on most LaTeX systems. Be sure and use
-% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
-% not currently provide for hyperlinked citations.
-% The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
-% The documentation is contained in the cite.sty file itself.
-
-% *** GRAPHICS RELATED PACKAGES ***
-%
-\ifCLASSINFOpdf
-  % \usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../pdf/}{../jpeg/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  % \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.eps}
-\fi
-% graphicx was written by David Carlisle and Sebastian Rahtz. It is
-% required if you want graphics, photos, etc. graphicx.sty is already
-% installed on most LaTeX systems. The latest version and documentation
-% can be obtained at: 
-% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
-% Another good source of documentation is "Using Imported Graphics in
-% LaTeX2e" by Keith Reckdahl which can be found at:
-% http://www.ctan.org/tex-archive/info/epslatex/
-%
-% latex, and pdflatex in dvi mode, support graphics in encapsulated
-% postscript (.eps) format. pdflatex in pdf mode supports graphics
-% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
-% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
-% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
-% which can result in "jaggedy"/blurry rendering of lines and letters as
-% well as large increases in file sizes.
-%
-% You can find documentation about the pdfTeX application at:
-% http://www.tug.org/applications/pdftex
-
-% *** MATH PACKAGES ***
-%
-%\usepackage[cmex10]{amsmath}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics. If using
-% it, be sure to load this package with the cmex10 option to ensure that
-% only type 1 fonts will utilized at all point sizes. Without this option,
-% it is possible that some math symbols, particularly those within
-% footnotes, will be rendered in bitmap form which will result in a
-% document that can not be IEEE Xplore compliant!
-%
-% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
-%\interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
-
-% *** SPECIALIZED LIST PACKAGES ***
-%
-%\usepackage{algorithmic}
-\usepackage{algpseudocode}
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
-% There is also a support site at:
-% http://algorithms.berlios.de/index.html
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/
-
-% *** ALIGNMENT PACKAGES ***
-%
-%\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/tools/
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-% *** SUBFIGURE PACKAGES ***
-%\ifCLASSOPTIONcompsoc
-%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
-%\else
-%  \usepackage[caption=false,font=footnotesize]{subfig}
-%\fi
-% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
-% for subfigure.sty, the latter of which is no longer maintained and is
-% incompatible with some LaTeX packages including fixltx2e. However,
-% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
-% which will override IEEEtran.cls' handling of captions and this will result
-% in non-IEEE style figure/table captions. To prevent this problem, be sure
-% and invoke subfig.sty's "caption=false" package option (available since
-% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
-% handling of captions.
-% Note that the Computer Society format requires a larger sans serif font
-% than the serif footnote size font used in traditional IEEE formatting
-% and thus the need to invoke different subfig.sty package options depending
-% on whether compsoc mode has been enabled.
-%
-% The latest version and documentation of subfig.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
-
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure. The latest version and documentation can be found at:
-% http://www.ctan.org/tex-archive/macros/latex/base/
-
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
-% Do not use the stfloats baselinefloat ability as IEEE does not allow
-% \baselineskip to stretch. Authors submitting work to the IEEE should note
-% that IEEE rarely uses double column equations and that authors should try
-% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
-% packages (also by Sigitas Tolusis) as IEEE does not format its papers in
-% such ways.
-% Do not attempt to use stfloats with fixltx2e as they are incompatible.
-% Instead, use Morten Hogholm'a dblfloatfix which combines the features
-% of both fixltx2e and stfloats:
-%
-% \usepackage{dblfloatfix}
-% The latest version can be found at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/dblfloatfix/
-
-%\ifCLASSOPTIONcaptionsoff
-%  \usepackage[nomarkers]{endfloat}
-% \let\MYoriglatexcaption\caption
-% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
-%\fi
-% endfloat.sty was written by James Darrell McCauley, Jeff Goldberg and 
-% Axel Sommerfeldt. This package may be useful when used in conjunction with 
-% IEEEtran.cls'  captionsoff option. Some IEEE journals/societies require that
-% submissions have lists of figures/tables at the end of the paper and that
-% figures/tables without any captions are placed on a page by themselves at
-% the end of the document. If needed, the draftcls IEEEtran class option or
-% \CLASSINPUTbaselinestretch interface can be used to increase the line
-% spacing as well. Be sure and use the nomarkers option of endfloat to
-% prevent endfloat from "marking" where the figures would have been placed
-% in the text. The two hack lines of code above are a slight modification of
-% that suggested by in the endfloat docs (section 8.4.1) to ensure that
-% the full captions always appear in the list of figures/tables - even if
-% the user used the short optional argument of \caption[]{}.
-% IEEE papers do not typically make use of \caption[]'s optional argument,
-% so this should not be an issue. A similar trick can be used to disable
-% captions of packages such as subfig.sty that lack options to turn off
-% the subcaptions:
-% For subfig.sty:
-% \let\MYorigsubfloat\subfloat
-% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
-% However, the above trick will not work if both optional arguments of
-% the \subfloat command are used. Furthermore, there needs to be a
-% description of each subfigure *somewhere* and endfloat does not add
-% subfigure captions to its list of figures. Thus, the best approach is to
-% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
-% and instead reference/explain all the subfigures within the main caption.
-% The latest version of endfloat.sty and its documentation can obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/endfloat/
-%
-% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
-% later in the document, say, to conditionally put the References on a 
-% page by themselves.
-
-% *** PDF, URL AND HYPERLINK PACKAGES ***
-%
-%\usepackage{url}
-% url.sty was written by Donald Arseneau. It provides better support for
-% handling and breaking URLs. url.sty is already installed on most LaTeX
-% systems. The latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/url/
-% Basically, \url{my_url_here}.
-
-% *** Do not adjust lengths that control margins, column widths, etc. ***
-% *** Do not use packages that alter fonts (such as pslatex).         ***
-% There should be no need to do such things with IEEEtran.cls V1.6 and later.
-% (Unless specifically asked to do so by the journal or conference you plan
-% to submit to, of course. )
-
-% correct bad hyphenation here
-%\hyphenation{op-tical net-works semi-conduc-tor}
-
-\usepackage{verbatim}
-\usepackage{color, colortbl}
-\definecolor{Gray}{gray}{0.9}
-\usepackage{graphicx}
-\usepackage{rotating}
-\usepackage{multirow}
+\documentclass[twoside]{article}
+\usepackage{ecj,palatino,epsfig,latexsym,natbib}
 \usepackage{url}
 
-\newcommand{\graphicthird}[1]
-{\includegraphics[width=.27\textwidth,height=.27\textheight,keepaspectratio]{#1}}
+\newcommand{\includegraphicswide}[1]
+{\includegraphics[width=.9\textwidth,height=\textheight,keepaspectratio]{#1}}
 
-\newcommand{\thirdlabel}[1]
-{\multicolumn{1}{|c|}{\raisebox{.15\textwidth}{\rotatebox[origin=c]{90}{\textbf{\em #1}}}}}
+\newcommand{\includegraphicsfit}[1]
+{\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}
+
+%% do not add any other page- or text-size instruction here
+
+\parskip=0.00in
 
 \begin{document}
-%
-% paper title
-% can use linebreaks \\ within to get better formatting as desired
-% Do not put math or special symbols in the title.
-\title{Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid}
-%
-%
-% author names and IEEE memberships
-% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
-% a structure at a ~ so this keeps an author's name from being broken across
-% two lines.
-% use \thanks{} to gain access to the first footnote area
-% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
-% was not built to handle multiple paragraphs
-%
 
-%\author{Michael~Shell,~\IEEEmembership{Member,~IEEE,}
-%        John~Doe,~\IEEEmembership{Fellow,~OSA,}
-%        and~Jane~Doe,~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
-\author{Brian~W.~Goldman, William~F.~Punch}
-%\thanks{M. Shell is with the Department
-%of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
-%GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
-%\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
-%\thanks{Manuscript received April 19, 2005; revised December 27, 2012.}}
+\ecjHeader{x}{x}{xxx-xxx}{200X}{TODO 45-character paper description}{B. W. Goldman and W. F. Punch}
+\title{\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid}
 
-% note the % following the last \IEEEmembership and also \thanks - 
-% these prevent an unwanted space from occurring between the last author name
-% and the end of the author line. i.e., if you had this:
-% 
-% \author{....lastname \thanks{...} \thanks{...} }
-%                     ^------------^------------^----Do not want these spaces!
-%
-% a space would be appended to the last name and could cause every name on that
-% line to be shifted left slightly. This is one of those "LaTeX things". For
-% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
-% "AB" then you have to do: "\textbf{A}\textbf{B}"
-% \thanks is no different in this regard, so shield the last } of each \thanks
-% that ends a line with a % and do not let a space in before the next \thanks.
-% Spaces after \IEEEmembership other than the last one are OK (and needed) as
-% you are supposed to have spaces between the names. For what it is worth,
-% this is a minor point as most people would not even notice if the said evil
-% space somehow managed to creep in.
+\author{\name{\bf B. W. Goldman} \hfill \addr{brianwgoldman@acm.org}\\ 
+        \addr{Department of Computer Science and Engineering, Michigan State University, 
+        East Lansing, 48823, United States}
+\AND
+       \name{\bf W. F. Punch} \hfill \addr{punch@msu.edu}\\
+        \addr{Department of Computer Science and Engineering, Michigan State University, 
+        East Lansing, 48823, United States}
+}
 
-% The paper headers
-\markboth{IEEE Transactions on Evolutionary Computation,~Vol.~\#\#, No.~\#, Month~Year}%
-{IEEE Transactions on Evolutionary Computation,~Vol.~\#\#, No.~\#, Month~Year}
-% The only time the second header will appear is for the odd numbered pages
-% after the title page when using the twoside option.
-% 
-% *** Note that you probably will NOT want to include the author's ***
-% *** name in the headers of peer review papers.                   ***
-% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
-% you desire.
-
-% If you want to put a publisher's ID mark on the page you can do it like
-% this:
-%\IEEEpubid{0000--0000/00\$00.00~\copyright~2012 IEEE}
-% Remember, if you use this you must call \IEEEpubidadjcol in the second
-% column for its text to clear the IEEEpubid mark.
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% make the title area
 \maketitle
 
-% As a general rule, do not put math, special symbols or citations
-% in the abstract or keywords.
 \begin{abstract}
-TODO Abstract
+
+TODO About 200 words.
+
 \end{abstract}
 
-% Note that keywords are not normally used for peerreview papers.
-\begin{IEEEkeywords}
-TODO
-\end{IEEEkeywords}
+\begin{keywords}
 
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
-\IEEEpeerreviewmaketitle
+Genetic algorithms, 
+linkage learning,
+local search,
+parameter-less.
+
+\end{keywords}
 
 \section{Introduction}
-% The very first letter is a 2 line initial drop letter followed
-% by the rest of the first word in caps.
-% 
-% form to use if the first word consists of a single letter:
-% \IEEEPARstart{A}{demo} file is ....
-% 
-% form to use if you need the single drop letter followed by
-% normal text (unknown if ever used by IEEE):
-% \IEEEPARstart{A}{}demo file is ....
-% 
-% Some journals put the first two words in caps:
-% \IEEEPARstart{T}{his demo} file is ....
-% 
-% Here we have the typical use of a "T" for an initial drop letter
-% and "HIS" in caps to complete the first word.
-\IEEEPARstart{T}{TODO}
-
 ~\cite{Wolpert:1997:nfl}
@@
 \subsection{hBOA}
+
+~\cite{pelikan:2006:hboa}
 
@@
 \section{Global Optimum}
-TODO Compare all 6 optimization techniques on time to reach global optimum
+
+\begin{figure*}
+  \centering
+  \includegraphicsfit{evals-to-success}
+  \caption{Comparison of the median number of evaluations to reach the global optimum for
+           the six different optimization methods with respect
+           to problem size.  If the median run did not reach the global optimum no data element
+           is shown.  Results given on a log-log scale.}
+  \label{fig-evals-to-success}
+\end{figure*}
+
+TODO Compare all 6 optimization techniques on time to reach global optimum.  Figure~\ref{fig-evals-to-success}.
 
@@
 
-TODO Consider a graphic showing 5\% and 95\% Quantiles to illustrate P3's scaling on problems
+\begin{figure*}
+  \centering
+  \includegraphicsfit{evals-to-success-range}
+  \caption{Comparison of the upper and lower quartiles of evaluations required to reach the global optimum for
+           three different optimization methods with respect
+           to problem size.}
+  \label{fig-evals-to-success-range}
+\end{figure*}
+
+TODO Graphic showing quartiles to illustrate P3's scaling on problems
 
@@
 \section{Conclusions and Future Work}
-% if have a single appendix:
-%\appendix[Proof of the Zonklar Equations]
-% or
-%\appendix  % for no appendix heading
-% do not use \section anymore after \appendix, only \section*
-% is possibly needed
 
-% use appendices with more than one appendix
-% then use \section to start each appendix
-% you must declare a \section before using any
-% \subsection or using \label (\appendices by itself
-% starts a section numbered zero.)
-%
+\section{About Figures}
 
-\begin{comment}
-\appendices
-\section{Example Appendix}
-Appendix one text goes here.
+\begin{figure}[t]
+\begin{center}
+\centerline{ 
+% \psfig{file=yourfigure.eps,width=3.0truein}
+%  UNCOMMENT the above line and add yourfigure.eps
+%  AND delete or comment-out the framebox
+\framebox(375,50)
+}
+\end{center}
+\caption{This is a caption below a framebox where a
+figure might appear.  Use epsfig in the {\tt $\backslash$usepackage\{epsfig,ecj...\}} 
+command to help to insure that we can process your figure.}
+\label{graph1}
+\end{figure}
 
-% you can choose not to have a title for an appendix
-% if you want by leaving the argument blank
-\section{}
-Appendix two text goes here.
-% use section* for acknowledgement
-\section*{Acknowledgment}
+\small
 
-This material is based in part upon work supported by the National Science Foundation
-under Cooperative Agreement No. DBI-0939454. Any opinions, findings, and conclusions
-or recommendations expressed in this material are those of the author(s) and do not
-necessarily reflect the views of the National Science Foundation.
-\end{comment}
+\bibliographystyle{apalike}
+\bibliography{../main}
 
-% Can use something like this to put references on a page
-% by themselves when using endfloat and the captionsoff option.
-\ifCLASSOPTIONcaptionsoff
-  \newpage
-\fi
-
-% trigger a \newpage just before the given reference
-% number - used to balance the columns on the last page
-% adjust value as needed - may need to be readjusted if
-% the document is modified later
-%\IEEEtriggeratref{8}
-% The "triggered" command can be changed if desired:
-%\IEEEtriggercmd{\enlargethispage{-5in}}
-
-% references section
-
-% can use a bibliography generated by BibTeX as a .bbl file
-% BibTeX documentation can be easily obtained at:
-% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
-% The IEEEtran BibTeX style support page is at:
-% http://www.michaelshell.org/tex/ieeetran/bibtex/
-%\bibliographystyle{IEEEtran}
-% argument is your BibTeX string definitions and bibliography database(s)
-%\bibliography{IEEEabrv,../bib/paper}
-%
-% <OR> manually copy in the resultant .bbl file
-% set second argument of \begin to the number of references
-% (used to reserve space for the reference number labels box)
-%\begin{thebibliography}{1}
-
-%\bibitem{IEEEhowto:kopka}
-%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
-%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
-
-%\end{thebibliography}
-
-\bibliographystyle{IEEEtran}
-\bibliography{IEEEabrv,../main}
-
-% biography section
-% 
-% If you have an EPS/PDF photo (graphicx package needed) extra braces are
-% needed around the contents of the optional argument to biography to prevent
-% the LaTeX parser from getting confused when it sees the complicated
-% \includegraphics command within an optional argument. (You could create
-% your own custom macro containing the \includegraphics command to make things
-% simpler here.)
-%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
-% or if you just want to reserve a space for a photo:
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{headGoldman}}]{Brian W. Goldman}
-received the B. S. degree and M.S. degree in computer science from the
-Missouri University of Science and Technology, Rolla, in 2010
-and 2012 respectively.
-
-He is currently a Ph.D. student with the Department of Computer Science
-and Engineering at Michigan State University, East Lansing.  His current
-research interests include evolutionary computation and genetic programming.
-
-\end{IEEEbiography}
-
-% if you will not have a photo at all:
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{headPunch}}]{William F. Punch}
-Bill Punch received his Ph.D. from the Ohio State University and is currently an
-Associate Professor at the Computer Science and Engineering Department of Michigan
-State University. He is also the  director of the MSU High Performance Computing
-Center. His research interests are evolutionary computation, high performance
-computing and computing pedagogy. He co-directs the Genetic Algorithms Research
-and Application Group (GARAGe) and is on the executive committee of the NSF STC,
-BEACON, at MSU. His book with Rich Enbody titled "The Practice of Computing Using
-Python" is now in its second edition.
-\end{IEEEbiography}
-
-% insert where needed to balance the two columns on the last page with
-% biographies
-%\newpage
-
-% You can push biographies down or up by placing
-% a \vfill before or after them. The appropriate
-% use of \vfill depends on what kind of text is
-% on the last page and whether or not the columns
-% are being equalized.
-
-%\vfill
-
-% Can be used to pull up biographies so that the bottom of the last one
-% is flush with the other column.
-%\enlargethispage{-5in}
-
-% that's all folks
 \end{document}

### 2014-08-27 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Added introduction and figures.

@@
 \usepackage{url}
+\usepackage{verbatim}
+\usepackage{subcaption}
+\usepackage[noend]{algpseudocode}
 
@@
 \section{Introduction}
-~\cite{Wolpert:1997:nfl}
+A primary purpose of evolutionary optimization is to efficiently find good solutions
+to challenging real world problems with minimal prior knowledge about the problem itself.
+This driving goal has created search algorithms which can escape user bias to create
+truly novel results, sometimes publishable or patentable in their own right~\citep{kannappan:2014:humies}.
+While it is not possible for any algorithm to do better than random across all possible
+problems~\citep{Wolpert:1997:nfl}, effectiveness can be achieved by assuming the search
+landscape has structure and then biasing the algorithm toward exploiting those features.
 
-~\cite{goldberg:1991:gasize} % Search efficiency
+In evolutionary optimization, and genetic algorithms~(GAs) in particular, search is often
+biased through parameters. This is beneficial as it allows practitioners to inject their
+knowledge about the shape of the search landscape into the algorithm. However, it can require
+expert knowledge to leverage this feature to its fullest potential or require exceedingly expensive
+parameter tuning~\citep{grefenstette:1986:optimalga}. Furthermore, the quality of solutions found,
+and the speed at which they are found, in GAs is strongly tied to setting these parameters
+correctly~\citep{goldberg:1991:gasize}. Parameters such as population size, mutation rate, crossover
+rate, tournament size, etc can have no clear relationship to the problem being solved, meaning even
+domain experts may not understand how they will interact with the problem or each other.
 
-~\cite{grefenstette:1986:optimalga} % Expensive to tune
+As such there has been periodic efforts to reduce or remove the need for parameter tuning.
+\cite{Back:1992:selfadapt} introduced self-adaptive parameters, where parameter values
+were included in each solution's genome and underwent evolution. This allowed search
+to optimize some of its own parameters and therefore reduced the need for expert tuning.
+\cite{harik:1999:parameterlessga} was able to design an entirely parameter-less GA by
+designing control strategies which made on line modifications to each parameter based
+on how well search was progressing. Unfortunately these methods were provably less efficient
+than correctly setting the parameters by up to a logarithmic
+factor~\citep{pelikan:1999:worstparameter-less}.
 
-~\cite{harik:1999:parameterlessga}
+One area that has been very effective at reducing the number of algorithm parameters has been
+model based search. \cite{pelikan:2006:hboa}'s Hierarchical Bayesian Optimization
+Algorithm~(hBOA) and \cite{thierens:2010:ltga}'s Linkage Tree Genetic Algorithm~(LTGA)
+both only require a single parameter: population size. \cite{posik:2011:parameterless}
+leveraged these methods to create a fully parameter-less algorithm which was restricted
+only to order-k fully decomposable noiseless problems.
 
-~\cite{posik:2011:parameterless} % Model based parameter-less, only noiseless non-overlapping
+Most recently \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid~(P3).
+This method combines model based search with a local search operator using a pyramid structure
+of populations to achieve parameter-less optimization. Initial results suggest that unlike
+previous parameter-less methods, P3 is actually more efficient than current state-of-the-art
+parameterized search algorithms. In this work we shall extend these results to cover more
+comparison algorithms, results on both efficiency in reaching the global optimum and intermediate
+fitnesses, algorithm complexity analysis, and provide more in depth analysis of P3 itself.
+\begin{comment}
+Section~\ref{sec-optimizers}
+explains how each of these algorithms, including P3, perform search. Section~\ref{sec-problems}
+provides a description of each test problem. As hBOA and LTGA require a population size parameter
+Section~\ref{sec-tuning} provides our methodology to ensure each is optimally tuned to each problem.
+\end{comment}
 
+\section{Optimizers}
+\label{sec-optimizers}
+
+\subsection{Hill Climber}
 ~\cite{goldman:2014:p3}
 
-\section{Optimizers}
+\subsection{LTGA}
+Figure~\ref{fig-cluster-creation}. Figure~\ref{fig-cluster-usage}.
+\begin{figure}
+  \begin{algorithmic}
+  \Procedure{Cluster-Creation}{}
+    \State $unmerged \leftarrow \{\{0\}, \{1\}, \{2\}, \dots, \{N-1\}\}$
+    \State $useful \leftarrow unmerged$
+    \While{$|unmerged|>1$}
+      \State $C_i, C_j \leftarrow \min_{C_i,C_j \in unmerged} D(C_i, C_j)$
+      \State $unmerged \leftarrow unmerged - \{C_i, C_j\} + \{C_i \cup C_j\}$
+      \State $useful \leftarrow useful + \{C_i \cup C_j\}$
+      \If{$D(C_i, C_j) = 0$}
+        \State $useful \leftarrow useful - \{C_i, C_j\}$
+      \EndIf
+    \EndWhile
+    \State Order $useful$ based on cluster size, smallest first.
+    \State Remove largest cluster from $useful$.
+
+    \Return $useful$
+  \EndProcedure
+\end{algorithmic}
+  \caption{Algorithm describing how clusters are created using Equation~\ref{eq-distance}
+           for each $P_i$. $unmerged$ and $useful$ are ordered sets of sets of gene loci.}
+  \label{fig-cluster-creation}
+\end{figure}
+
+\begin{figure}
+  \begin{algorithmic}
+  \Procedure{Cluster-Usage}{}
+    \ForAll{$C_i \in useful$}
+      \ForAll{$d \in shuffled(P_i)$}
+        \State Copy $d$'s gene values for $C_i$ into solution
+        \If{solution was changed}
+          \If{solution's fitness decreased}
+            \State Revert changes
+          \EndIf
+          \State \textbf{break}
+        \EndIf
+      \EndFor
+    \EndFor
+  \EndProcedure
+\end{algorithmic}
+  \caption{Algorithm describing how clusters are used to perform crossover.}
+  \label{fig-cluster-usage}
+\end{figure}
+
+~\cite{thierens:2010:ltga}
+~\cite{thierens:2013:ltgahiff}
+
+\begin{equation}
+  D(C_i,C_j) = \frac{1}{\left | C_i \right |\cdot \left |C_j \right|}\sum_{c_i \in C_i}\sum_{c_j \in C_j}
+  2 - \frac{H(C_i) + H(C_j)}{H(C_i \cup C_j)}
+  \label{eq-distance}
+\end{equation}
+\begin{equation}
+  H(C) = -\sum_{s\in S} p_c(s)\log(p_c(s))
+  \label{eq-entropy}
+\end{equation}
+
+~\cite{bosman:2011:lsbbo} % LTGA with hill climbing
 
 \subsection{P3}
+Figure~\ref{fig-p3}.
+\begin{figure}
+  \begin{algorithmic}
+  \Procedure{Iterate-P3}{}
+    \State Create random solution
+    \State Apply hill climber (Section~\ref{sec-hillclimber})
+    \If{solution $\notin hashset$}
+      \State Add solution to $P_0$ (Section~\ref{sec-pyramid})
+      \State Add solution to $hashset$
+    \EndIf
+
+    \ForAll{$P_i \in pyramid$}
+      \State Mix solution with $P_i$ (Section~\ref{sec-crossover})
+      \If{solution's fitness has improved}
+        \If{solution $\notin hashset$}
+          \State Add solution to $P_{i+1}$
+          \State Add solution to $hashset$
+        \EndIf
+      \EndIf
+    \EndFor
+  \EndProcedure
+\end{algorithmic}
+  \caption{One iteration of P3 optimization. $pyramid$ is an
+           ordered set of populations and $hashset$ is a set
+           of all solutions in $pyramid$.}
+  \label{fig-p3}
+\end{figure}
+
 ~\cite{goldman:2014:p3}
@@
 ~\cite{gronau:2007:upgma} % Efficient clustering
-
-\subsection{Hill Climber}
-~\cite{goldman:2014:p3}
-
-\subsection{LTGA}
-~\cite{thierens:2010:ltga}
-
-~\cite{thierens:2013:ltgahiff}
-
-~\cite{bosman:2011:lsbbo} % LTGA with hill climbing
 
@@
 \section{Problem Descriptions}
+\label{sec-problems}
+
 \subsection{Single Instance Problems}
@@
 \section{Comparison Algorithm Parameter Tuning}
+\label{sec-tuning}
+
 ~\cite{goldman:2012:ltga} % Bisection with no parameters
@@
 \section{Global Optimum}
+\label{sec-optimum}
 
-\begin{figure*}
-  \centering
+\begin{figure}[t]
+  \begin{center}
   \includegraphicsfit{evals-to-success}
+  \end{center}
   \caption{Comparison of the median number of evaluations to reach the global optimum for
@@
   \label{fig-evals-to-success}
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{figure*}
-  \centering
+\begin{figure}[t]
+  \begin{center}
   \includegraphicsfit{evals-to-success-range}
-  \caption{Comparison of the upper and lower quartiles of evaluations required to reach the global optimum for
-           three different optimization methods with respect
-           to problem size.}
+  \end{center}
+  \caption{Comparison of the upper and lower quartiles of evaluations required
+           to reach the global optimum for P3 and LTGA with respect to problem size.}
   \label{fig-evals-to-success-range}
-\end{figure*}
+\end{figure}
 
-TODO Graphic showing quartiles to illustrate P3's scaling on problems
+TODO Graphic showing quartiles to illustrate P3's scaling on problems. Figure~\ref{fig-evals-to-success-range}.
 
 \section{Stop Anytime}
+\begin{figure}[t]
+  \begin{center}
+  \includegraphicsfit{fitness-over-time}
+  \end{center}
+  \caption{Compares the median best fitness reached during search for each of the six optimization methods.}
+  \label{fig-fitness-over-time}
+\end{figure}
+
 TODO Compare all 6 optimization techniques on fitnesses reached at different times
-during the run on problem sizes near 100 or 150
+during the run on problem sizes. Explain results same for different sizes. Figure~\ref{fig-fitness-over-time}.
 
-TODO Compare all 4 optimization techniques on fitnesses reached at different times
-during the run on problem sizes near 500 or 750
+TODO Include graphic for one problem where LTGA was tuned using a significantly higher failure rate.
 
-\section{Complexity Per Evaluation}
-TODO Include graphic of rebuilds per evaluation (decreasing with length)
+\begin{figure}[t]
+  \begin{centering}
+    \begin{subfigure}{.5\textwidth}
+      \begin{centering}
+        \includegraphicsfit{rebuilds}
+      \end{centering}
+      \caption{Rebuilds}
+      \label{fig-rebuilds}
+    \end{subfigure}%
+    \begin{subfigure}{.5\textwidth}
+      \begin{centering}
+        \includegraphicsfit{donations}
+      \end{centering}
+      \caption{Donations}
+      \label{fig-donations}
+    \end{subfigure}
+  \end{centering}
+  \caption{Computation costs incurred by model rebuilding (Figure~\ref{fig-rebuilds}) and
+           repeated donations (Figure~\ref{fig-donations}) per evaluation as problem size increases.}
+\end{figure}
 
-TODO Include graphic of donations per evaluation (sub-linear growth)
+TODO Include graphic of rebuilds per evaluation. Figure~\ref{fig-rebuilds}.
 
-TODO Include graphic comparing seconds per evaluation up to small problems (hboa super-linear)
+TODO Include graphic of donations per evaluation. Figure~\ref{fig-donations}.
 
-TODO Include graphic comparing seconds per evaluation for large problems (perhaps just 1 graph of P3 across all problems)
+TODO Include a table of $O(N)$ seconds per evaluation for all 4 model building techniques on all 7 problems.
 
 \section{Inner Workings}
-TODO Include graphic of proportion of evaluations spend on hill climber (constant with length)
 
-TODO Include graphic of crossover success (constant with length)
+\begin{figure}[t]
+  \begin{centering}
+    \begin{subfigure}{.5\textwidth}
+      \begin{centering}
+        \includegraphicsfit{cross}
+      \end{centering}
+      \caption{Crossover Proportion}
+      \label{fig-cross}
+    \end{subfigure}%
+    \begin{subfigure}{.5\textwidth}
+      \begin{centering}
+        \includegraphicsfit{cross-success}
+      \end{centering}
+      \caption{Crossover Success}
+      \label{fig-cross-success}
+    \end{subfigure}
+  \end{centering}
+  \caption{For each problem Figure~\ref{fig-cross} shows the proportion of P3 evaluations spend on crossovers
+           and Figure~\ref{fig-cross-success} shows the percentage of fitness improving crossover evaluations.}
+\end{figure}
 
-TODO Include graphic of Ising Spin Glass, NK, and DeceptiveStepTrap's pyramid shape and crossover success rates at each level
+TODO Include graphic of proportion of evaluations spent on crossover. Figure~\ref{fig-cross}.
+
+TODO Include graphic of crossover success Figure~\ref{fig-cross-success}.
+
+\begin{figure}[t]
+  \begin{centering}
+    \begin{subfigure}{.5\textwidth}
+      \begin{centering}
+        \includegraphicsfit{level-size}
+      \end{centering}
+      \caption{Population Size}
+      \label{fig-level-size}
+    \end{subfigure}%
+    \begin{subfigure}{.5\textwidth}
+      \begin{centering}
+        \includegraphicsfit{level-success}
+      \end{centering}
+      \caption{Crossover Success}
+      \label{fig-level-success}
+    \end{subfigure}
+  \end{centering}
+  \caption{For each problem Figure~\ref{fig-level-size} shows the number of solutions stored in each level of the pyramid
+           and Figure~\ref{fig-level-success} shows the percentage of fitness improving crossover evaluations at each level.}
+\end{figure}
+
+TODO Include graphic of pyramid shape (Figure~\ref{fig-level-size}) and crossover success rates (Figure~\ref{fig-level-success}) at each level.
 
@@
 
+~\cite{goldman:2011:dynamic-parameters}
+
 \section{Conclusions and Future Work}
-
-\section{About Figures}
-
-\begin{figure}[t]
-\begin{center}
-\centerline{ 
-% \psfig{file=yourfigure.eps,width=3.0truein}
-%  UNCOMMENT the above line and add yourfigure.eps
-%  AND delete or comment-out the framebox
-\framebox(375,50)
-}
-\end{center}
-\caption{This is a caption below a framebox where a
-figure might appear.  Use epsfig in the {\tt $\backslash$usepackage\{epsfig,ecj...\}} 
-command to help to insure that we can process your figure.}
-\label{graph1}
-\end{figure}
 

### 2014-08-28 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Hill climber section and algorithm.

@@
 \subsection{Hill Climber}
-~\cite{goldman:2014:p3}
+Perhaps the simplest black box search heuristic is stochastic local search, or hill climbing.
+This optimization technique focuses on improving a single solution until it reaches a local
+optimum. Here we use the first improvement hill climber defined by \cite{goldman:2014:p3}
+and given in Figure~\ref{fig-hc}. This algorithm works by flipping each bit in a random
+order, keeping modifications when fitness is improved, until doing so cannot result in
+further fitness improvements.
+
+The hill climber requires $O(1)$ operations per evaluation amortized cost. In order to
+terminate, it must perform at least $N$ evaluations, where $N$ is the number of bits
+in the solution, in order to ensure no single bit flip can result in a fitness improvement.
+As such any operation that happens only once per search can be amortized over at least $N$
+evaluations, covering the initialization of $options$.
+
+Assuming proper hashing data structures, determining if $index \notin tried$ should require only $O(1)$
+time. While this operation may be called without a directly related evaluation, amortized cost
+provides an upper bound. Consider that in order to add an $index$ into $tried$ an evaluation must
+be performed. Lets say this evaluation pays for two checks of that $index$, one of which was used
+during this loop. If no fitness improving move is found, the next time through the loop the second check
+is used. A third check is only necessary if no fitness improving move was found for any index.
+This is a contradiction as the loop terminates when no improving move was found. Therefore this check
+can only be called twice per evaluation.
+
+\begin{figure}
+  \begin{algorithmic}
+  \Procedure{Hill-Climber}{}
+    \State $options \leftarrow [0 \dots N-1]$
+    \State $tried \leftarrow \emptyset$
+    \While{$|tried| < |options|$}
+      \State $shuffle(options)$
+      \ForAll{$index \in options$}
+        \If{$index \notin tried$}
+          \State Flip bit $index$ in solution
+          \If{solution's fitness increased}
+            \State $tried \leftarrow \emptyset$
+          \Else
+            \State Revert change
+          \EndIf
+          \State $tried \leftarrow tried \cup \{index\}$
+        \EndIf
+      \EndFor
+    \EndWhile
+  \EndProcedure
+\end{algorithmic}
+  \caption{Hill climbing algorithm used to improve randomly generated solutions until no single
+           bit change results in a fitness improvement.}
+  \label{fig-hc}
+\end{figure}
 
@@
 
-TODO Compare all 6 optimization techniques on time to reach global optimum.  Figure~\ref{fig-evals-to-success}.
+TODO Compare all 6 optimization techniques on evaluations to reach global optimum.  Figure~\ref{fig-evals-to-success}.
 
@@
 
-TODO Include a table of $O(N)$ seconds per evaluation for all 4 model building techniques on all 7 problems.
+TODO Include a table of $O(N)$ seconds per evaluation and total seconds for all 4 model building techniques on all 7 problems.
 

### 2014-09-03 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Spelled out LTGA
More LTGA

@@
 \begin{figure}
-  \begin{algorithmic}
+  \begin{algorithmic}[1]
   \Procedure{Hill-Climber}{}
@@
     \While{$|tried| < |options|$}
-      \State $shuffle(options)$
-      \ForAll{$index \in options$}
+      \ForAll{$index \in shuffled(options)$}
         \If{$index \notin tried$}
@@
 
-\subsection{LTGA}
-Figure~\ref{fig-cluster-creation}. Figure~\ref{fig-cluster-usage}.
+\subsection{Linkage Tree Genetic Algorithm}
+~\cite{thierens:2010:ltga} introduced the Linkage Tree Genetic Algorithm (LTGA) which automatically
+detects and exploits problem epistasis by examining pairwise gene entropy. Due to its enhanced
+ability to preserve high fitness gene subsets, LTGA was able to outperform state of the art
+GAs across many benchmarks. Since its first proposal, many variants of LTGA have been
+proposed~\citep{goldman:2012:ltga}. Therefore for clarity we have chosen the version
+presented by \cite{thierens:2013:ltgahiff} as our model for comparison as it is both
+modern and appears to approach the consensus in the literature.
+
+LTGA's effectiveness comes from its method of performing crossover. Instead of blindly
+mixing genes between parents, LTGA attempts to preserve important interrelationships
+between genes. Before performing any crossovers in a generation, LTGA first builds
+a set of hierarchical gene clusters which are then used to dictate how genes are mixed
+during crossover.
+
 \begin{figure}
-  \begin{algorithmic}
+  \begin{algorithmic}[1]
   \Procedure{Cluster-Creation}{}
@@
     \EndWhile
-    \State Order $useful$ based on cluster size, smallest first.
-    \State Remove largest cluster from $useful$.
+    \State Order $useful$ based based on last merged first\label{fig-cluster-creation-ordering}
+    \State Remove largest cluster from $useful$
 
@@
 \end{algorithmic}
-  \caption{Algorithm describing how clusters are created using Equation~\ref{eq-distance}
-           for each $P_i$. $unmerged$ and $useful$ are ordered sets of sets of gene loci.}
+  \caption{Algorithm describing how LTGA creates clusters using Equation~\ref{eq-distance}
+           for a population. $unmerged$ and $useful$ are ordered sets of sets of gene loci.}
   \label{fig-cluster-creation}
@@
 
+Figure~\ref{fig-cluster-creation} provides the agglomerative method LTGA uses to create gene clusters.
+This algorithm will create a tree of sets, such that the leaves of the tree contain a single gene and
+nodes are the union of their children's sets. These sets are then used by crossover to specify epistatic
+relationships which should be preserved.
+The process begins by creating the set of sets $unmerged$ which tracks all top level clusters. Initially
+$unmerged$ contains single member sets for each gene. Each iteration the two sets with the minimum pairwise
+distance (given in Equation~\ref{eq-distance}) are merged to create a single cluster. This process is repeated
+until only a single set remains in $unmerged$ which contains all of the genes in the genome.
+
+\begin{equation}
+  D(C_i,C_j) = \frac{1}{\left | C_i \right |\cdot \left |C_j \right|}\sum_{c_i \in C_i}\sum_{c_j \in C_j}
+  2 - \frac{H(c_i) + H(c_j)}{H(c_i \cup c_j)}
+  \label{eq-distance}
+\end{equation}
+\begin{equation}
+  H(c) = -\sum_{s\in S} p_c(s)\log(p_c(s))
+  \label{eq-entropy}
+\end{equation}
+
+Throughout this process $useful$ tracks the set of all gene clusters which should be preserved for use by crossover.
+This set begins with all genes in separate clusters, and each time a new cluster is created it is added to $useful$.
+However, not all clusters are necessarily worth keeping. For instance, in all versions of LTGA the cluster
+containing all genes is removed from $useful$ as preserving all genes during crossover can only create clones.
+\cite{thierens:2013:ltgahiff} extended this removal to include any unsupported subsets. If the pairwise distance
+between two clusters is 0, this means there are no individuals in the population which disrupt the values stored in
+each cluster. Therefore when performing crossover there is no reason to believe a fitness improvement can be achieved
+by breaking the stored pattern. As such only clusters who's direct superset has non-zero distance are kept.
+
+\cite{thierens:2013:ltgahiff}'s version of LTGA does not use the entire population when determining pairwise entropy.
+Instead, binary tournament use used to select half of the population. This is done to ensure the model is built
+using only high quality solutions, even during the first generation.
+
+In order to efficiently perform clustering, a pairwise gene frequency table is constructed.
+Equation~\ref{eq-entropy} uses how frequently each of the four possible string values for each
+pair of genes occurs in the selected solutions. Extracting this information requires $O(\mu N^2)$
+time, where $\mu$ is the population size and $N$ is the genome size. The process of converting
+this pairwise frequency information into clusters can be achieved in $O(N^2)$ using the bookkeeping
+methods presented by \cite{gronau:2007:upgma}. This cost is performed only once per generation,
+and is then used to perform approximately $O(\mu N)$ crossover evaluations. As a result, the amortized cost of
+LTGA's model building is $O(N)$.
+
 \begin{figure}
-  \begin{algorithmic}
+  \begin{algorithmic}[1]
   \Procedure{Cluster-Usage}{}
     \ForAll{$C_i \in useful$}
-      \ForAll{$d \in shuffled(P_i)$}
+      %\ForAll{$d \in shuffled(P_i)$}
+        \State $d \leftarrow rand\_choice(P)$\label{fig-cluster-usage-donate}
         \State Copy $d$'s gene values for $C_i$ into solution
@@
           \EndIf
-          \State \textbf{break}
+          %\State \textbf{break}
         \EndIf
-      \EndFor
+      %\EndFor
     \EndFor
@@
 
-~\cite{thierens:2010:ltga}
-~\cite{thierens:2013:ltgahiff}
+Figure~\ref{fig-cluster-usage} describes how the identified clusters are used by crossover to preserve
+gene linkage while still exploring the search space. Unlike more traditional crossover methods, LTGA
+crosses each individual with the entire population. Also, to produce a single offspring, multiple evaluations
+of the fitness function are performed.
 
-\begin{equation}
-  D(C_i,C_j) = \frac{1}{\left | C_i \right |\cdot \left |C_j \right|}\sum_{c_i \in C_i}\sum_{c_j \in C_j}
-  2 - \frac{H(C_i) + H(C_j)}{H(C_i \cup C_j)}
-  \label{eq-distance}
-\end{equation}
-\begin{equation}
-  H(C) = -\sum_{s\in S} p_c(s)\log(p_c(s))
-  \label{eq-entropy}
-\end{equation}
+Each generation, each individual in the population undergoes crossover. In a single crossover event, each
+cluster of genes $C_i$ in $useful$ is applied in last merged first order as a crossover mask. A random donor $d$
+is chosen from the entire population (not just the model selected population), and $d$'s gene's for $C_i$ are copied
+into the working solution. If a modification is made, an evaluation is then performed. If the crossover
+resulted in no worse fitness than the changes are kept. This allows for neutral drift across plateaus.
+The resulting solution, which must be at least as fit as its parent, is then
+copied into the next generation.
+
+In total each individual can cause up to $|useful|$ evaluations. If all clusters were kept, even those deemed
+unhelpful, and all donations were evaluated, even those which did not change any genes, then \Call{Cluster-Usage}{}
+would perform exactly $2N-2$ evaluations for each of the $\mu$ in the population. This provides the amortizing evaluations
+required to make clustering only $O(N)$ operations per evaluation. However, but skipping some evaluations, its
+possible that clustering may be super-linear.
 
@@
 \begin{figure}
-  \begin{algorithmic}
+  \begin{algorithmic}[1]
   \Procedure{Iterate-P3}{}
     \State Create random solution
-    \State Apply hill climber (Section~\ref{sec-hillclimber})
+    \State Apply hill climber
     \If{solution $\notin hashset$}
-      \State Add solution to $P_0$ (Section~\ref{sec-pyramid})
+      \State Add solution to $P_0$
       \State Add solution to $hashset$
@@
     \ForAll{$P_i \in pyramid$}
-      \State Mix solution with $P_i$ (Section~\ref{sec-crossover})
+      \State Mix solution with $P_i$
       \If{solution's fitness has improved}
@@
 
+Figure~\ref{fig-cluster-creation} Line~\ref{fig-cluster-creation-ordering}.
+
+Figure~\ref{fig-cluster-usage} Line~\ref{fig-cluster-usage-donate}.
+
 ~\cite{hornby:2006:alps}
-
-~\cite{gronau:2007:upgma} % Efficient clustering
 
@@
 \subsection{$1+(\lambda, \lambda)$}
-~\cite{doerr:2013:lambdalambda}
+\cite{doerr:2013:lambdalambda} was the first genetic algorithm to provably show
+the advantages of performing crossover on simple landscapes. This comparatively simple
+algorithm maintains only a single individual and a self-controlled parameter $\lambda$.
 
-~\cite{goldman:2014:p3}
+Each iteration, a number of bits to flip is chosen from the binomial distribution $b\sim B(N, \frac{\lambda}{N})$.
+Next, $\lfloor\lambda\rfloor$ offspring are produced by flipping $b$ bits. The
+best mutant then produces $\lambda$ offspring via uniform crossover with the original parent, such that each gene comes from the
+mutant with probability $\frac{1}{\lambda}$. In the original algorithm the best
+offspring produced by crossover then replaces the original parent if its fitness is no worse.
+The $\lambda$ parameter, which is initialized to 1, is decreased if the offspring replaced
+its parent and decreased otherwise.
+
+The original formulation was designed specifically for unimodal landscapes and as such were
+not directly suitable for multimodal problems. \cite{goldman:2014:p3} extended the $1+(\lambda, \lambda)$
+to include random restarts. As search stagnates, the $\lambda$ parameter increases in value. Eventually
+this results in $\lambda \ge N$. This results in mutation flipping all bits of the individual, preventing
+useful exploration. Therefore whenever $\lambda \ge N$, search is restarted from a random solution with $\lambda$
+reset to 1.
+
+A few other efficiency modifications were also made. If there is a tie in crossover offspring fitness,
+whichever has a larger hamming distance from the parent is retained. This encourages drifting across plateaus.
+The ``mod'' control strategy proposed by \cite{doerr:2013:lambdalambda} was not used as it conflicted with
+the random restart strategy.
+If a crossover individual is identical to either of its parents, it is not evaluated.
+If mutation produces an offspring which is better than the best crossover offspring, it is used to compare
+against the original parent.
 

### 2014-09-05 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
P3 description

@@
 both only require a single parameter: population size. \cite{posik:2011:parameterless}
-leveraged these methods to create a fully parameter-less algorithm which was restricted
-only to order-k fully decomposable noiseless problems.
+leveraged model building to create a fully parameter-less algorithm, but it is restricted to
+only order-k fully decomposable noiseless problems.
 
@@
 
-~\cite{bosman:2011:lsbbo} % LTGA with hill climbing
+By design, LTGA only has a single parameter: population size. LTGA uses no mutation, and crossover is defined
+in terms of the clustering algorithm. Selection between generations is fully elitist and embedded in the crossover,
+with selection of model building solutions fixed to a binary tournament. Neither \Call{Cluster-Creation}{} nor
+\Call{Cluster-Usage}{} rely on parameter values. LTGA does not provide any method for controlling or setting
+the population size, and uses a fixed user specified size for each generation.
 
-\subsection{P3}
-Figure~\ref{fig-p3}.
+\subsection{Parameter-less Population Pyramid}
+\cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid (P3) as a method for
+performing optimization which did not require the user to provide any parameters. This is
+achieved by combining efficient local search with the model building methods of LTGA using
+an iteratively constructed hierarchy of populations.
+
+The high level algorithm of P3 is presented in Figure~\ref{fig-p3}. Unlike more traditional
+genetic algorithms, P3 does not follow a generational model. Instead, it maintains an iteratively
+expanding pyramid of expanding populations. Each iteration, a new random solution is generated.
+This solution is brought to a local optimum using the algorithm in Figure~\ref{fig-hc}. If that
+local optimum has not been added to the pyramid yet, the solution is added to the lowest
+population $P_0$.
+
+Next, the solution is iteratively improved by applying LTGA's crossover algorithm (Figure~\ref{fig-cluster-usage})
+with each population $P_i$ in the pyramid. If this process results in a strict fitness improvement and has
+created a solution not yet stored in the pyramid, it is added to the next highest pyramid level $P_{i+1}$.
+If $P_{i+1}$ does not yet exist, it is created. In this way populations in the pyramid expand over time,
+and the number of populations stored increases over time. Initially the pyramid contains no solutions
+or populations, meaning the user does not need to specify a population size.
+
 \begin{figure}
@@
 
-~\cite{goldman:2014:p3}
+To accommodate P3's unique population structure, some of LTGA's clustering procedures needed
+to be adapted. In LTGA
+clusters are identified at the start of each generation, with those clusters used
+to create all solutions produced in that generation. As P3 does not perform serial
+generations, this process had to be modified. Instead, P3 updates the model each
+time a solution is added to a population, the model is rebuilt. Furthermore, unlike
+LTGA all solutions in the population are used to generate the model, not just the
+winners of a binary tournament. This selection process is not necessary as the
+worst solutions in the pyramid are already high quality due to local search.
+Using local search in LTGA was examined by \cite{bosman:2011:lsbbo} and found
+to provide no significant improvement.
 
-Figure~\ref{fig-cluster-creation} Line~\ref{fig-cluster-creation-ordering}.
+Beyond the changes in population structuring, P3 also has slight modifications to
+LTGA's version of \Call{Cluster-Creation}{} and \Call{Cluster-Usage}{}.
+P3 changes Line~\ref{fig-cluster-creation-ordering} in Figure~\ref{fig-cluster-creation}
+from \emph{last merged first} ordering to \emph{smallest first} ordering.
+This method applies gene clusters during crossover based on how many genes
+are in each cluster\footnote{Ties are broken randomly.}, and not on how tightly linked those genes are.
+\cite{goldman:2012:ltga} found that this alternative was better at preserving
+diversity, and therefore required smaller populations, but was not integrated into
+our chosen canonical LTGA variant.
 
-Figure~\ref{fig-cluster-usage} Line~\ref{fig-cluster-usage-donate}.
+P3 also modified Line~\ref{fig-cluster-usage-donate} in Figure~\ref{fig-cluster-usage}.
+Instead of choosing a single genetic donor for each cluster, P3 iterates over the
+population in a random order until a solution in the population is found which
+has a least one gene different for that cluster of genes than the improving solution.
+This process increases the likelihood of an evaluation being performed for every cluster,
+and helps test rare gene patterns in the population.
+
+TODO Explain higher computational costs
 
 ~\cite{hornby:2006:alps}
+
+TODO Explain more than the sum of its parts
 

### 2014-09-09 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
More hBOA

@@
 
-TODO Explain higher computational costs
+In LTGA the cost of rebuilding the model is $O(\mu N^2)$ as it must collect pairwise
+gene frequency information for all $\mu$ solutions in the population. P3 does not store
+a single population, and it does not have a fixed $\mu$ size for any population. However,
+each time a solution is added to the population, it requires $O(N^2)$ time to update
+the table of pairwise frequencies and another $O(N^2)$ time to rebuild the linkage model.
+The model is then used immediately to perform up to one evaluation for each of the up
+to $2N-2$ clusters. Therefore similar to LTGA if no evaluation shortcuts were made P3
+will have an amortized cost of $O(N)$ modeling cost per evaluation. While P3 does
+rebuild the model more frequently per solution in the population, it also performs
+a number of local search evaluations which are quite efficient, meaning theoretical
+comparisons of their speed are difficult to perform. As a final note, P3's repeated
+attempts to find a useful donation make it less likely than LTGA to skip evaluations,
+but has an added cost to find these donations. Repeated donations could require as much
+as $O(\mu)$ donations per evaluation, but experimental evidence suggest that this
+operation actually saves more time than it costs by increasing the number of evaluations
+per model rebuild.
 
-~\cite{hornby:2006:alps}
+Each of the pieces of the P3 algorithm were selected not just for their stand alone efficacy,
+but for the ways in which they interact. By using the hill climber to ensure each solution
+is at a local optima, the underlying pairwise relationships in the problem are exposed. As
+a result detecting clusters for use by crossover is much more effective. The crossover operator
+is extremely elitist, as each gene donation must result in no fitness loss, and a solution must
+strictly improve to be added to the next level of the pyramid. This is balanced by continual
+integration of new random solutions. Furthermore, each random restart decreases the probability
+of spurious linkages caused by shared ancestry. This diversity is further preserved by applying
+gene clusters in smallest first order during crossover as this reduces the probability of genetic
+hitchhikers.
 
-TODO Explain more than the sum of its parts
+Other algorithms have proposed using multiple concurrent populations.
+\cite{hornby:2006:alps} had a hierarchy of populations with solutions periodically advancing
+upward. This allowed for continuous integration of diversity as the lowest population is reseeded
+with random solutions. However, this method resulted in increased parameterization as not only was
+a population size required, parameters for how frequently generations advanced between levels and
+how many total levels to have were added. \cite{harik:1999:parameterlessga} used multiple independent
+populations of different sizes as a method for removing the population size parameter, but
+doing was provably less efficient than using an optimal population size.
 
-\subsection{hBOA}
+\subsection{Hierarchical Bayesian Optimization Algorithm}
 
-~\cite{pelikan:2006:hboa}
+\cite{pelikan:2006:hboa} used statistical principles in combination with a decision tree structure
+to create the Hierarchical Bayesian Optimization Algorithm (hBOA). This method creates a model of
+epistatic relationships between genes which is then used to stochastically generate new solutions.
+Each generation a binary tournament with replacement is used to selection $\mu$ solutions from
+the population. These solutions are then used to build the model used to generate $\mu$ new
+solutions. The new solutions are then integrated into the population using restricted tournament
+replacement.
+
+Conceptually, the model built by hBOA is trying to infer rules of the form ``Given that this
+subset of genes are set to these values, how frequently is gene x set to value v?'' This can
+be represented using a decision forest, with each tree in the forest representing one gene
+in the solution. In the decision tree $T_i$, which is used to set the value of gene $x_i$,
+Each internal node represents previous decisions on how to set
+some other gene $x_j$, with the children of that node representing how the decision was made. The
+leaves of each tree give the probability that $x_i$ should be set to each possible value.
+Therefore in binary alphabets binary trees with a single probability $p$ stored at each leaf
+can be used, where $p$ is the probability $x_i$ should be set to one.
+
+The forest is constructed iteratively, with each tree initially containing a single leaf
+with each leaf storing a pointer to all selected solutions. Each iteration the algorithm considers
+all possible ways of splitting an existing leaf using another gene $x_j$, such that solutions in the
+leaf are moved to the newly created leaves based on their value for $x_j$. The general goal is to
+separate the solutions such that all solutions with $x_i = 0$ move to one leaf while solutions with
+$x_j = 1$ move to the other. Each possible split is scored using Equation~\ref{eq-hboa} with the best
+split kept. This process continues until none of the splits result in a higher quality. To favor compact
+models, Equation~\ref{eq-hboa} is scaled by Equation~\ref{eq-hboa-stop}, which provides increased cost
+for more total leaves.
+
+\begin{equation}
+  BDe(B) = p(B)\prod_{i=0}^{N}\prod_{l\in L_i} \frac{\Gamma(m'_i(l))}{\Gamma(m_i(l) + m'_i(l))}
+  \prod_{x_i}\frac{\Gamma(m_i(x_i, l) + m'_i(x_i,l))}{\Gamma(m'_i(x_i,l))}
+  \label{eq-hboa}
+\end{equation}
+
+\begin{equation}
+  p(B) = c2^{-0.5(\sum_i|L_i|)log_2\mu}
+  \label{eq-hboa-stop}
+\end{equation}
+
+While Equation~\ref{eq-hboa} is very complex and almost always results in near infinitesimal results, many
+of the terms can be canceled out as it is only important to know the relative quality between two models:
+$BDe(B) < BDe(B')$.
+The outermost product of Equation~\ref{eq-hboa} iterates over all trees in the forest. However, each split
+can modify only one of the trees and therefore the contribution of all others can be canceled. The middle
+product is across all leaves in the tree. Again since only one leaf can be changed, all other terms can
+be canceled. By convention hBOA uses uninformed Bayesian priors of $m'_i(l)= 2$ and $m'_i(x_i, l)=1$ for
+binary alphabets. As the $\Gamma(n) = (n-1)!$ this means the top term in the middle product and the bottom
+term in the third product reduce to 1. The only remaining terms are then $m_i(l)$ and $m_i(x_i, l)$ which
+represent the number of solutions which reached leaf $l$ and the number of solutions which reached leaf $l$
+with a specific value, respectively.
+
+Equation~\ref{eq-hboa-stop} can also be simplified when doing comparisons. The only difference for $p(B)$ and $p(B')$ is the total
+number of leaves ($\sum_i|L_i|$) has increased by 1. Therefore the ratio simplifies to $2^{0.5 log_2\mu}$ regardless of
+actual model size.
+
+All combined, this means the expression $BDe(B) < BDe(B')$ can be calculated using Equation~\ref{eq-hboa-final},
+where B' is different from B by exactly 1 split, such that $l$ was split to create $l'$ and $l''$. Only splits that
+satisfy this inequality are considered enough to keep. Furthermore, whichever split maximizes the right side of the
+Equation~\ref{eq-hboa-final} is chosen each iteration. Note that these factorials can be exceedingly large and
+therefore its imperative that implementations avoid rounding errors and overflows.
+
+\begin{equation}
+  2^{0.5 log_2\mu} < \frac{(m_i(l) + 1)!}{m_i(0, l)!m_i(1,l)!} \cdot
+  \frac{m_i(0, l')!m_i(1,l')!m_i(0, l'')!m_i(1,l'')!}{(m_i(l') + 1)!(m_i(l'') + 1)!}
+  \label{eq-hboa-final}
+\end{equation}
 
@@
 \subsection{$1+(\lambda, \lambda)$}
-\cite{doerr:2013:lambdalambda} was the first genetic algorithm to provably show
+\cite{doerr:2013:lambdalambda} presented the first genetic algorithm to provably show
 the advantages of performing crossover on simple landscapes. This comparatively simple

### 2014-09-09 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
First edit of first two sections.

@@
 In evolutionary optimization, and genetic algorithms~(GAs) in particular, search is often
-biased through parameters. This is beneficial as it allows practitioners to inject their
-knowledge about the shape of the search landscape into the algorithm. However, it can require
-expert knowledge to leverage this feature to its fullest potential or require exceedingly expensive
-parameter tuning~\citep{grefenstette:1986:optimalga}. Furthermore, the quality of solutions found,
-and the speed at which they are found, in GAs is strongly tied to setting these parameters
-correctly~\citep{goldberg:1991:gasize}. Parameters such as population size, mutation rate, crossover
+biased through parameters. This can be beneficial as it allows practitioners to inject their
+knowledge about the shape of the search landscape into the algorithm.
+However, the quality of solutions found, and the speed at which they are found, is strongly tied to setting these parameters
+correctly~\citep{goldberg:1991:gasize}. As such expert knowledge or exceedingly
+expensive parameter tuning~\citep{grefenstette:1986:optimalga} may be required to leverage
+this feature to its fullest potential. Furthermore,  Parameters such as population size, mutation rate, crossover
 rate, tournament size, etc can have no clear relationship to the problem being solved, meaning even
-domain experts may not understand how they will interact with the problem or each other.
+domain experts may not understand how they will interact with the problem or with each other.
 
 As such there has been periodic efforts to reduce or remove the need for parameter tuning.
-\cite{Back:1992:selfadapt} introduced self-adaptive parameters, where parameter values
+\cite{Back:1992:selfadapt} introduced self-adaptive parameters, in which parameter values
 were included in each solution's genome and underwent evolution. This allowed search
-to optimize some of its own parameters and therefore reduced the need for expert tuning.
+to optimize some of its own parameters resulting in a reduced the need for expert tuning.
 \cite{harik:1999:parameterlessga} was able to design an entirely parameter-less GA by
@@
 on how well search was progressing. Unfortunately these methods were provably less efficient
-than correctly setting the parameters by up to a logarithmic
-factor~\citep{pelikan:1999:worstparameter-less}.
+than directly setting the parameters to optimal values~\citep{pelikan:1999:worstparameter-less}.
 
@@
 Most recently \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid~(P3).
-This method combines model based search with a local search operator using a pyramid structure
-of populations to achieve parameter-less optimization. Initial results suggest that unlike
+This method uses a pyramid structure of populations to combine model based search with local search
+to achieve parameter-less optimization. Initial results suggest that unlike
 previous parameter-less methods, P3 is actually more efficient than current state-of-the-art
 parameterized search algorithms. In this work we shall extend these results to cover more
-comparison algorithms, results on both efficiency in reaching the global optimum and intermediate
+comparison algorithms, compare both efficiency in reaching the global optimum and intermediate
 fitnesses, algorithm complexity analysis, and provide more in depth analysis of P3 itself.
@@
 
-The hill climber requires $O(1)$ operations per evaluation amortized cost. In order to
-terminate, it must perform at least $N$ evaluations, where $N$ is the number of bits
-in the solution, in order to ensure no single bit flip can result in a fitness improvement.
+The hill climber requires an amortized cost of $O(1)$ operations per evaluation. In order to
+terminate, at least one evaluation must be performed for each of the $N$ bits in the solution.
 As such any operation that happens only once per search can be amortized over at least $N$
@@
 
-Assuming proper hashing data structures, determining if $index \notin tried$ should require only $O(1)$
+Using proper hashing data structures, determining if $index \notin tried$ requires $O(1)$
 time. While this operation may be called without a directly related evaluation, amortized cost
@@
 GAs across many benchmarks. Since its first proposal, many variants of LTGA have been
-proposed~\citep{goldman:2012:ltga}. Therefore for clarity we have chosen the version
-presented by \cite{thierens:2013:ltgahiff} as our model for comparison as it is both
+proposed~\citep{goldman:2012:ltga} so for clarity we have chosen the version
+presented by \cite{thierens:2013:ltgahiff} as our model. This variant was chosen as it is
 modern and appears to approach the consensus in the literature.
@@
 The process begins by creating the set of sets $unmerged$ which tracks all top level clusters. Initially
-$unmerged$ contains single member sets for each gene. Each iteration the two sets with the minimum pairwise
+$unmerged$ contains single member sets for each gene. Each iteration the two sets with the minimum average pairwise
 distance (given in Equation~\ref{eq-distance}) are merged to create a single cluster. This process is repeated
@@
 \cite{thierens:2013:ltgahiff} extended this removal to include any unsupported subsets. If the pairwise distance
-between two clusters is 0, this means there are no individuals in the population which disrupt the values stored in
-each cluster. Therefore when performing crossover there is no reason to believe a fitness improvement can be achieved
-by breaking the stored pattern. As such only clusters who's direct superset has non-zero distance are kept.
+between two clusters is 0, this means there are no individuals in the population which disrupt the relationships between the two
+clusters. Therefore when performing crossover there is no reason to believe a fitness improvement can be achieved
+by breaking the stored pattern. As such a cluster is only kept if its direct superset has a non-zero distance.
 
@@
 into the working solution. If a modification is made, an evaluation is then performed. If the crossover
-resulted in no worse fitness than the changes are kept. This allows for neutral drift across plateaus.
+resulted in no worse fitness then the changes are kept. This allows for neutral drift across plateaus.
 The resulting solution, which must be at least as fit as its parent, is then
@@
 unhelpful, and all donations were evaluated, even those which did not change any genes, then \Call{Cluster-Usage}{}
-would perform exactly $2N-2$ evaluations for each of the $\mu$ in the population. This provides the amortizing evaluations
-required to make clustering only $O(N)$ operations per evaluation. However, but skipping some evaluations, its
+would perform exactly $2N-2$ evaluations for each of the $\mu$ solutions in the population. This provides the amortizing evaluations
+required to make clustering only $O(N)$ operations per evaluation. However, by skipping some evaluations, its
 possible that clustering may be super-linear.
+
+LTGA has no explicit form of diversity control and has no method for introducing new genetic information once
+the population has converged. Therefore an LTGA run is considered converged once two consecutive populations
+contain the same unique solutions.
 
@@
 \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid (P3) as a method for
-performing optimization which did not require the user to provide any parameters. This is
+performing optimization which does not require the user to provide any parameters. This is
 achieved by combining efficient local search with the model building methods of LTGA using
@@
 The high level algorithm of P3 is presented in Figure~\ref{fig-p3}. Unlike more traditional
-genetic algorithms, P3 does not follow a generational model. Instead, it maintains an iteratively
+GAs, P3 does not follow a generational model. Instead, it maintains an iteratively
 expanding pyramid of expanding populations. Each iteration, a new random solution is generated.
 This solution is brought to a local optimum using the algorithm in Figure~\ref{fig-hc}. If that
-local optimum has not been added to the pyramid yet, the solution is added to the lowest
+local optimum has not yet been added to the pyramid, the solution is added to the lowest
 population $P_0$.
@@
 generations, this process had to be modified. Instead, P3 updates the model each
-time a solution is added to a population, the model is rebuilt. Furthermore, unlike
+time a solution is added to a population. Furthermore, unlike
 LTGA all solutions in the population are used to generate the model, not just the
@@
 \cite{goldman:2012:ltga} found that this alternative was better at preserving
-diversity, and therefore required smaller populations, but was not integrated into
+diversity, and therefore required smaller populations, but has  not been integrated into
 our chosen canonical LTGA variant.
@@
 population in a random order until a solution in the population is found which
-has a least one gene different for that cluster of genes than the improving solution.
+has a least one gene different for that cluster of genes from the improving solution.
 This process increases the likelihood of an evaluation being performed for every cluster,
@@
 but has an added cost to find these donations. Repeated donations could require as much
-as $O(\mu)$ donations per evaluation, but experimental evidence suggest that this
-operation actually saves more time than it costs by increasing the number of evaluations
+as $O(\mu)$ attempts per evaluation, but experimental evidence suggest that this
+operation actually saves more overhead than it costs by increasing the number of evaluations
 per model rebuild.
@@
 \cite{hornby:2006:alps} had a hierarchy of populations with solutions periodically advancing
-upward. This allowed for continuous integration of diversity as the lowest population is reseeded
+upward. This allows for continuous integration of diversity as the lowest population is reseeded
 with random solutions. However, this method resulted in increased parameterization as not only was
@@
 populations of different sizes as a method for removing the population size parameter, but
-doing was provably less efficient than using an optimal population size.
+doing was provably less efficient than using an optimal population size as no information is shared
+between the populations.
 
@@
 Each generation a binary tournament with replacement is used to selection $\mu$ solutions from
-the population. These solutions are then used to build the model used to generate $\mu$ new
+the population. These solutions are then used to build the model, which in turn is used to generate $\mu$ new
 solutions. The new solutions are then integrated into the population using restricted tournament
@@
 Conceptually, the model built by hBOA is trying to infer rules of the form ``Given that this
-subset of genes are set to these values, how frequently is gene x set to value v?'' This can
-be represented using a decision forest, with each tree in the forest representing one gene
+subset of genes are set to these values, how frequently is gene $x_i$ set to value v?'' This can
+be represented using a directed acyclic decision forest, with each tree in the forest representing one gene
 in the solution. In the decision tree $T_i$, which is used to set the value of gene $x_i$,
-Each internal node represents previous decisions on how to set
+each internal node represents previous decisions on how to set
 some other gene $x_j$, with the children of that node representing how the decision was made. The
@@
 The forest is constructed iteratively, with each tree initially containing a single leaf
-with each leaf storing a pointer to all selected solutions. Each iteration the algorithm considers
+and with each leaf storing a pointer for each selected solution. Each iteration the algorithm considers
 all possible ways of splitting an existing leaf using another gene $x_j$, such that solutions in the
@@
 separate the solutions such that all solutions with $x_i = 0$ move to one leaf while solutions with
-$x_j = 1$ move to the other. Each possible split is scored using Equation~\ref{eq-hboa} with the best
-split kept. This process continues until none of the splits result in a higher quality. To favor compact
+$x_j = 1$ move to the other. Whichever split maximizes Equation~\ref{eq-hboa} is then kept.
+This process continues until none of the splits result in a higher quality model than the current one. To favor compact
 models, Equation~\ref{eq-hboa} is scaled by Equation~\ref{eq-hboa-stop}, which provides increased cost
@@
 \begin{equation}
-  BDe(B) = p(B)\prod_{i=0}^{N}\prod_{l\in L_i} \frac{\Gamma(m'_i(l))}{\Gamma(m_i(l) + m'_i(l))}
+  BDe(B) = \prod_{i=0}^{N}\prod_{l\in L_i} \frac{\Gamma(m'_i(l))}{\Gamma(m_i(l) + m'_i(l))}
   \prod_{x_i}\frac{\Gamma(m_i(x_i, l) + m'_i(x_i,l))}{\Gamma(m'_i(x_i,l))}
@@
 of the terms can be canceled out as it is only important to know the relative quality between two models:
-$BDe(B) < BDe(B')$.
+$p(B)BDe(B) < p(B')BDe(B')$.
 The outermost product of Equation~\ref{eq-hboa} iterates over all trees in the forest. However, each split
@@
 be canceled. By convention hBOA uses uninformed Bayesian priors of $m'_i(l)= 2$ and $m'_i(x_i, l)=1$ for
-binary alphabets. As the $\Gamma(n) = (n-1)!$ this means the top term in the middle product and the bottom
+binary alphabets. As $\Gamma(n) = (n-1)!$ this means the top term in the middle product and the bottom
 term in the third product reduce to 1. The only remaining terms are then $m_i(l)$ and $m_i(x_i, l)$ which
@@
 number of leaves ($\sum_i|L_i|$) has increased by 1. Therefore the ratio simplifies to $2^{0.5 log_2\mu}$ regardless of
-actual model size.
+total model size.
 
@@
 where B' is different from B by exactly 1 split, such that $l$ was split to create $l'$ and $l''$. Only splits that
-satisfy this inequality are considered enough to keep. Furthermore, whichever split maximizes the right side of the
+satisfy this inequality are considered good enough to keep. Furthermore, whichever split maximizes the right side of the
 Equation~\ref{eq-hboa-final} is chosen each iteration. Note that these factorials can be exceedingly large and
@@
 
+Initially there are $\Theta(N^2)$ possible ways to split existing leaves, as each of the $N$ single node
+trees can be split by any of the other $N-1$ genes. Each iteration a new edge is added to the decision
+forest, meaning some of the previously tested splits cannot be used. For instance, if $T_i$, which is used
+to decide the value of $x_i$, is split using the value of $x_j$, $T_j$ can no longer be split using $x_i$.
+As a split creates two new leaves, $O(N)$ new potential splits must also be tested. Equation~\ref{eq-hboa-final}
+parses all solutions which reach a leaf to count gene frequencies, requiring $\mu$ time.
+The number of total leaves created depends heavily on the problem and $\mu$.
+However, assuming no splits are accepted or that the cost of testing all future splits is less than
+the initial $\Theta(N^2)$, constructing the model requires $\Omega(\mu N^2)$ time. Each
+model is used to generate $\mu$ solutions, leading to a cost per evaluation of $\Omega(N^2)$.
+
+To generate a solution, the value of each gene $x_i$ is set using its corresponding decision tree $T_i$. Because
+the forest is directed acyclic, there must be an ordering of $T_i$ such that before $T_i$ is executed all
+$x_j$ it uses to make decisions have already been set. As such, previous decisions made by other trees
+are used to follow each $T_i$ until a leaf is reached. The value of $x_i$ is then set based on the
+probably that other solutions reached that leaf with each value of $x_i$.
+
+To perform replacement, hBOA uses restricted tournament replacement. After each new solution is generated
+and evaluated, $w$ solutions are chosen at random from the population, where $w=\min\{N, \frac{\mu}{20}\}$.
+Whichever of the $w$ solutions is the most genetically similar to the offspring is compared with the offspring.
+If the fitness of the offspring is no worse, it replaces the genetically similar member of $w$, otherwise the
+offspring is discarded. This method is designed to preserve genetic diversity as only genetically similar
+solutions must compete on fitness.
+
+To model properly, hBOA is designed to work with large population sizes, resulting in a large number of
+evaluations per generation. As hBOA utilizes explicit diversity maintenance, standard methods for determining
+convergence are not considered very accurate. Therefore the authors suggest that an hBOA run should be
+terminated after performing generations equal to $N$.
+
+Like other model based techniques, hBOA has very few parameters. There is no mutation or crossover,
+and modeling does not rely on any explicit parameters. Solution selection, generation, and replacement
+are all derived from the population size, which must be set by the user.
+
 \subsection{Parameter-less hBOA}
-~\cite{pelikan:2004:parameterlesshboa}
+Using the methods first introduced by \cite{harik:1999:parameterlessga} for the Parameter-less GA,
+\cite{pelikan:2004:parameterlesshboa} created Parameter-less hBOA which automatically scales its
+population size to fit the problem. This is done by maintaining a list of concurrent populations
+using exponentially scaled population sizes.
+
+A run of Parameter-less hBOA starts with a single population of size $\mu_0$, conventionally set
+to $\mu_0=10$. After two generations are performed, a new population of size $\mu_1 = 2\mu_0$ is created
+and performs a generation. Evolution then continues with the $\mu_0$ population performing two generations
+for each one performed by $\mu_1$. Each time population $\mu_i$ performs its second generation a new population
+$\mu_{i+1}=2\mu_i$ is created, which performs generations half as often as $\mu_i$. In this way an infinite number of
+parallel population can be simulated, with each population receiving the same number of total evaluations.
+
+In all other aspects each population is identical to an hBOA population using a fixed $\mu$. No search information
+is shared among populations, and each search is independently terminated. As such Parameter-less hBOA cannot
+perform better than hBOA using the optimal population size for a given instance, as it must also spend evaluations
+on populations of different sizes. This inefficiency is bounded by a log multiple of the total number of
+evaluations~\citep{pelikan:1999:worstparameter-less}.
 
@@
 
-Each iteration, a number of bits to flip is chosen from the binomial distribution $b\sim B(N, \frac{\lambda}{N})$.
+Each iteration, the number of bits to flip is chosen from the binomial distribution $b\sim B(N, \frac{\lambda}{N})$.
 Next, $\lfloor\lambda\rfloor$ offspring are produced by flipping $b$ bits. The
@@
 The $\lambda$ parameter, which is initialized to 1, is decreased if the offspring replaced
-its parent and decreased otherwise.
+its parent and increased otherwise.
 
 The original formulation was designed specifically for unimodal landscapes and as such were
-not directly suitable for multimodal problems. \cite{goldman:2014:p3} extended the $1+(\lambda, \lambda)$
+not directly suitable for multimodal problems. \cite{goldman:2014:p3} extended $1+(\lambda, \lambda)$
 to include random restarts. As search stagnates, the $\lambda$ parameter increases in value. Eventually
-this results in $\lambda \ge N$. This results in mutation flipping all bits of the individual, preventing
-useful exploration. Therefore whenever $\lambda \ge N$, search is restarted from a random solution with $\lambda$
+this results in $\lambda \ge N$ causing mutation to always flip all bits of the individual.
+As this prevents any future improvement whenever $\lambda \ge N$ search is restarted from a random solution with $\lambda$
 reset to 1.

### 2014-09-10 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Added single instance problem section.

@@
 \label{sec-problems}
+TODO Consider adding paragraph about problem equivalence.
 
 \subsection{Single Instance Problems}
-~\cite{goldberg:1991:gasize} % Deceptive Trap
+Understanding how a stochastic search algorithm will behave on arbitrary and complex
+search landscapes can be exceedingly difficult. Therefore a common practice for
+algorithm understanding is to perform search on well defined, well understood
+landscapes. To be of interest these landscapes need to represent interesting
+and important challenges aspects of real world problems.
 
-~\cite{goldman:2012:ltga} % Deceptive Step Trap
+Perhaps the most common such landscape is the Deceptive Trap problem \citep{goldberg:1991:gasize}.
+In this landscape the genome is broken up into $k$ sized fully decomposable subproblems referred
+to as traps.
+Each subproblem is scored using Equation~\ref{eq-deceptive-trap}, where $t$ is the number of
+bits in the trap set to 1. To global optimum in each trap is a string of all 1s, while all
+other solutions lead to a local optima of all 0s. As a result this problem tests an algorithms
+ability to overcome $k$ sized deception and is commonly used to determine how effective crossover
+is at preserving building blocks. Any crossover event that mixes bits from different parents in
+the same trap will likely result in that trap being optimized to the local optima. For our experiments
+we chose $k=7$ to ensure highly deceptive traps.
 
-~\cite{thierens:2013:ltgahiff} % HIFF
+\begin{equation}
+   trap(t) = \left\{
+     \begin{array}{rl}
+       k-1-t, &  t<k\\
+       k,   &  t = k
+     \end{array}
+   \right.
+  \label{eq-deceptive-trap}
+\end{equation}
 
-~\cite{goldman:2014:p3} % Rastrigin
+\cite{goldman:2012:ltga} found that mixing local search with linkage learning rendered the Deceptive
+Trap problem trivial. This is because local search is able to optimize each trap to one of the two
+optima (all 1s or all 0s), and then linkage learning can have perfect knowledge of gene interactions.
+In order to make the problem more difficult yet still deceptive, the authors proposed the Deceptive
+Step Trap problem, given in Equation~\ref{eq-deceptive-step-trap}. This function modifies the results
+of Equation~\ref{eq-deceptive-trap} to include plateaus of size $s$, introducing an exponential number
+of local optima in each trap. As a result the Deceptive Step Trap is much more challenging for linkage
+learning techniques, while still being highly deceptive. We chose to use $s=2$ to create the maximum
+number of local optima.
+\begin{equation}
+   step\_trap(t) = \left \lfloor \frac{(k-s)\pmod{s} + trap(t)}{s} \right \rfloor
+  \label{eq-deceptive-step-trap}
+\end{equation}
+
+Another challenging aspect of landscapes can be higher order relationships. For these problems
+the are easily detected main effects as well as larger interrelationships. The Hierarchical If
+and only If (HIFF) problem is designed to capture the difficulties of this class of problem.
+In HIFF the genome is broken up into a complete binary tree, such that each gene appears in exactly one
+leaf and each internal node is the subset of genes contained in its children. If all genes represented
+in a node of the tree are set to the same value, they score equal to the size of the set. In this way
+small subsets lead toward solutions to larger subsets. However a node can score if all genes are either
+all 1s or all 0s, meaning that to solve higher order subproblems it is necessary to perform crossovers
+that preserve lower order solutions. This problem is a natural fit for LTGA as the linkage tree can
+perfectly duplicate the problem's true relationships \citep{thierens:2013:ltgahiff}.
+
+As a final class of well known problems, we have chosen to borrow the Rastrigin problem from real valued
+optization. This problem's landscape, determined by Equation~\ref{eq-rast}, is highly multimodal caused by
+the oscillating cosine function. \cite{goldman:2014:p3} proposed the Discretized Rastrigin problem, such
+that each floating point $x_i$ in Equation~\ref{eq-rast} is encoded using a 10 bit gray code.
+\begin{equation}
+  An + \sum_{i=1}^{n}\left [ x_i^2-A\cos (2\pi x_i) \right ] \forall x\in [-5.12,5.12]
+  \label{eq-rast}
+\end{equation}
 

### 2014-09-10 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Minor editing updates from Bill

@@
 expensive parameter tuning~\citep{grefenstette:1986:optimalga} may be required to leverage
-this feature to its fullest potential. Furthermore,  Parameters such as population size, mutation rate, crossover
+this feature to its fullest potential. Furthermore,  parameters such as population size, mutation rate, crossover
 rate, tournament size, etc can have no clear relationship to the problem being solved, meaning even
@@
 
-As such there has been periodic efforts to reduce or remove the need for parameter tuning.
+As such there have been periodic efforts to reduce or remove the need for parameter tuning.
 \cite{Back:1992:selfadapt} introduced self-adaptive parameters, in which parameter values
@@
 to optimize some of its own parameters resulting in a reduced the need for expert tuning.
-\cite{harik:1999:parameterlessga} was able to design an entirely parameter-less GA by
-designing control strategies which made on line modifications to each parameter based
+\cite{harik:1999:parameterlessga} were able to design an entirely parameter-less GA by
+designing control strategies which made on-line modifications to each parameter based
 on how well search was progressing. Unfortunately these methods were provably less efficient
@@
 leveraged model building to create a fully parameter-less algorithm, but it is restricted to
-only order-k fully decomposable noiseless problems.
+only order-k, fully decomposable, noiseless problems.
 
@@
 
-Using proper hashing data structures, determining if $index \notin tried$ requires $O(1)$
+With the use of proper hashing data structures, determining if $index \notin tried$ requires $O(1)$
 time. While this operation may be called without a directly related evaluation, amortized cost
@@
 The process begins by creating the set of sets $unmerged$ which tracks all top level clusters. Initially
-$unmerged$ contains single member sets for each gene. Each iteration the two sets with the minimum average pairwise
+$unmerged$ contains single member sets for each gene. After each iteration the two sets with the minimum average pairwise
 distance (given in Equation~\ref{eq-distance}) are merged to create a single cluster. This process is repeated
@@
 expanding pyramid of expanding populations. Each iteration, a new random solution is generated.
-This solution is brought to a local optimum using the algorithm in Figure~\ref{fig-hc}. If that
+This solution is brought to a local optimum using the hill climbing algorithm in Figure~\ref{fig-hc}. If that
 local optimum has not yet been added to the pyramid, the solution is added to the lowest
@@
 
-To accommodate P3's unique population structure, some of LTGA's clustering procedures needed
-to be adapted. In LTGA
+To accommodate P3's unique population structure, some of LTGA's clustering procedures were adapted. In LTGA,
 clusters are identified at the start of each generation, with those clusters used
@@
 The model is then used immediately to perform up to one evaluation for each of the up
-to $2N-2$ clusters. Therefore similar to LTGA if no evaluation shortcuts were made P3
-will have an amortized cost of $O(N)$ modeling cost per evaluation. While P3 does
+to $2N-2$ clusters. Just as in LTGA, if no evaluation shortcuts were made, P3
+has an amortized cost of $O(N)$ modeling cost per evaluation. While P3 does
 rebuild the model more frequently per solution in the population, it also performs
@@
 is at a local optima, the underlying pairwise relationships in the problem are exposed. As
-a result detecting clusters for use by crossover is much more effective. The crossover operator
+a result, detecting clusters for use by crossover is much more effective. The crossover operator
 is extremely elitist, as each gene donation must result in no fitness loss, and a solution must

### 2014-09-11 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Reordered optimizer section and broke into "Comparison Optimizers" and "P3" sections
Added Comparison Optimizers explanation

@@
 
-\section{Optimizers}
+\section{Comparison Optimizers}
 \label{sec-optimizers}
 
-\subsection{Hill Climber}
+In order to fully understand the effectiveness of a black box search algorithm, it is useful
+to compare it with other similar algorithms. Therefore here we preset five advanced algorithms with
+related features to P3. The Random Restart Hill Climber was chosen as an efficient form of repeated
+local search, and because it shows the advantages of performing the more complex portions of P3.
+The $1+(\lambda, \lambda)$ algorithm is the current best theory supported simple genetic algorithm
+and its method of crossover is in some sense a macro-mutation just as in P3. hBOA and Parameter-less
+hBOA are advanced model building search techniques which are very effective at learning complex
+problem structure, designed to achieve similar goals as P3's linkage learning but using very different
+methods. Finally LTGA represents the current state-of-the-art in black box search and is the origin
+of P3's linkage learning and crossover methods.
+
+Finally, only hBOA and LTGA require any parameters, with each of these only requiring a population
+size. This makes knowing the optimal behavior of these algorithms much more tractable.
+
+\subsection{Random Restart Hill Climber}
+\label{sec-hill-climber}
 Perhaps the simplest black box search heuristic is stochastic local search, or hill climbing.
@@
 further fitness improvements.
-
-The hill climber requires an amortized cost of $O(1)$ operations per evaluation. In order to
-terminate, at least one evaluation must be performed for each of the $N$ bits in the solution.
-As such any operation that happens only once per search can be amortized over at least $N$
-evaluations, covering the initialization of $options$.
-
-With the use of proper hashing data structures, determining if $index \notin tried$ requires $O(1)$
-time. While this operation may be called without a directly related evaluation, amortized cost
-provides an upper bound. Consider that in order to add an $index$ into $tried$ an evaluation must
-be performed. Lets say this evaluation pays for two checks of that $index$, one of which was used
-during this loop. If no fitness improving move is found, the next time through the loop the second check
-is used. A third check is only necessary if no fitness improving move was found for any index.
-This is a contradiction as the loop terminates when no improving move was found. Therefore this check
-can only be called twice per evaluation.
 
@@
 \end{figure}
+
+The hill climber requires an amortized cost of $O(1)$ operations per evaluation. In order to
+terminate, at least one evaluation must be performed for each of the $N$ bits in the solution.
+As such any operation that happens only once per search can be amortized over at least $N$
+evaluations, covering the initialization of $options$.
+
+With the use of proper hashing data structures, determining if $index \notin tried$ requires $O(1)$
+time. While this operation may be called without a directly related evaluation, amortized cost
+provides an upper bound. Consider that in order to add an $index$ into $tried$ an evaluation must
+be performed. Lets say this evaluation pays for two checks of that $index$, one of which was used
+during this loop. If no fitness improving move is found, the next time through the loop the second check
+is used. A third check is only necessary if no fitness improving move was found for any index.
+This is a contradiction as the loop terminates when no improving move was found. Therefore this check
+can only be called twice per evaluation.
+
+Due to its nature, this hill climber cannot escape basins of attraction. Once a solution is reached
+such that none of the single bit neighbors are fitness improvements, search stops. As a result to
+have any hope of solving multimodal problems this algorithm requires a restart mechanism. We have
+chosen here to naively restart search from a random solution whenever a local optima is found. This
+ensures that on all landscapes there is always a non-zero probability of search finding the global optimum.
+
+\subsection{$1+(\lambda, \lambda)$}
+\cite{doerr:2013:lambdalambda} presented the first genetic algorithm to provably show
+the advantages of performing crossover on simple landscapes. This comparatively simple
+algorithm maintains only a single individual and a self-controlled parameter $\lambda$.
+
+Each iteration, the number of bits to flip is chosen from the binomial distribution $b\sim B(N, \frac{\lambda}{N})$.
+Next, $\lfloor\lambda\rfloor$ offspring are produced by flipping $b$ bits. The
+best mutant then produces $\lambda$ offspring via uniform crossover with the original parent, such that each gene comes from the
+mutant with probability $\frac{1}{\lambda}$. In the original algorithm the best
+offspring produced by crossover then replaces the original parent if its fitness is no worse.
+The $\lambda$ parameter, which is initialized to 1, is decreased if the offspring replaced
+its parent and increased otherwise.
+
+The original formulation was designed specifically for unimodal landscapes and as such were
+not directly suitable for multimodal problems. \cite{goldman:2014:p3} extended $1+(\lambda, \lambda)$
+to include random restarts. As search stagnates, the $\lambda$ parameter increases in value. Eventually
+this results in $\lambda \ge N$ causing mutation to always flip all bits of the individual.
+As this prevents any future improvement whenever $\lambda \ge N$ search is restarted from a random solution with $\lambda$
+reset to 1.
+
+A few other efficiency modifications were also made. If there is a tie in crossover offspring fitness,
+whichever has a larger hamming distance from the parent is retained. This encourages drifting across plateaus.
+The ``mod'' control strategy proposed by \cite{doerr:2013:lambdalambda} was not used as it conflicted with
+the random restart strategy.
+If a crossover individual is identical to either of its parents, it is not evaluated.
+If mutation produces an offspring which is better than the best crossover offspring, it is used to compare
+against the original parent.
+
+\subsection{Hierarchical Bayesian Optimization Algorithm}
+
+\cite{pelikan:2006:hboa} used statistical principles in combination with a decision tree structure
+to create the Hierarchical Bayesian Optimization Algorithm (hBOA). This method creates a model of
+epistatic relationships between genes which is then used to stochastically generate new solutions.
+Each generation a binary tournament with replacement is used to selection $\mu$ solutions from
+the population. These solutions are then used to build the model, which in turn is used to generate $\mu$ new
+solutions. The new solutions are then integrated into the population using restricted tournament
+replacement.
+
+Conceptually, the model built by hBOA is trying to infer rules of the form ``Given that this
+subset of genes are set to these values, how frequently is gene $x_i$ set to value v?'' This can
+be represented using a directed acyclic decision forest, with each tree in the forest representing one gene
+in the solution. In the decision tree $T_i$, which is used to set the value of gene $x_i$,
+each internal node represents previous decisions on how to set
+some other gene $x_j$, with the children of that node representing how the decision was made. The
+leaves of each tree give the probability that $x_i$ should be set to each possible value.
+Therefore in binary alphabets binary trees with a single probability $p$ stored at each leaf
+can be used, where $p$ is the probability $x_i$ should be set to one.
+
+The forest is constructed iteratively, with each tree initially containing a single leaf
+and with each leaf storing a pointer for each selected solution. Each iteration the algorithm considers
+all possible ways of splitting an existing leaf using another gene $x_j$, such that solutions in the
+leaf are moved to the newly created leaves based on their value for $x_j$. The general goal is to
+separate the solutions such that all solutions with $x_i = 0$ move to one leaf while solutions with
+$x_j = 1$ move to the other. Whichever split maximizes Equation~\ref{eq-hboa} is then kept.
+This process continues until none of the splits result in a higher quality model than the current one. To favor compact
+models, Equation~\ref{eq-hboa} is scaled by Equation~\ref{eq-hboa-stop}, which provides increased cost
+for more total leaves.
+
+\begin{equation}
+  BDe(B) = \prod_{i=0}^{N}\prod_{l\in L_i} \frac{\Gamma(m'_i(l))}{\Gamma(m_i(l) + m'_i(l))}
+  \prod_{x_i}\frac{\Gamma(m_i(x_i, l) + m'_i(x_i,l))}{\Gamma(m'_i(x_i,l))}
+  \label{eq-hboa}
+\end{equation}
+
+\begin{equation}
+  p(B) = c2^{-0.5(\sum_i|L_i|)log_2\mu}
+  \label{eq-hboa-stop}
+\end{equation}
+
+While Equation~\ref{eq-hboa} is very complex and almost always results in near infinitesimal results, many
+of the terms can be canceled out as it is only important to know the relative quality between two models:
+$p(B)BDe(B) < p(B')BDe(B')$.
+The outermost product of Equation~\ref{eq-hboa} iterates over all trees in the forest. However, each split
+can modify only one of the trees and therefore the contribution of all others can be canceled. The middle
+product is across all leaves in the tree. Again since only one leaf can be changed, all other terms can
+be canceled. By convention hBOA uses uninformed Bayesian priors of $m'_i(l)= 2$ and $m'_i(x_i, l)=1$ for
+binary alphabets. As $\Gamma(n) = (n-1)!$ this means the top term in the middle product and the bottom
+term in the third product reduce to 1. The only remaining terms are then $m_i(l)$ and $m_i(x_i, l)$ which
+represent the number of solutions which reached leaf $l$ and the number of solutions which reached leaf $l$
+with a specific value, respectively.
+
+Equation~\ref{eq-hboa-stop} can also be simplified when doing comparisons. The only difference for $p(B)$ and $p(B')$ is the total
+number of leaves ($\sum_i|L_i|$) has increased by 1. Therefore the ratio simplifies to $2^{0.5 log_2\mu}$ regardless of
+total model size.
+
+All combined, this means the expression $BDe(B) < BDe(B')$ can be calculated using Equation~\ref{eq-hboa-final},
+where B' is different from B by exactly 1 split, such that $l$ was split to create $l'$ and $l''$. Only splits that
+satisfy this inequality are considered good enough to keep. Furthermore, whichever split maximizes the right side of the
+Equation~\ref{eq-hboa-final} is chosen each iteration. Note that these factorials can be exceedingly large and
+therefore its imperative that implementations avoid rounding errors and overflows.
+
+\begin{equation}
+  2^{0.5 log_2\mu} < \frac{(m_i(l) + 1)!}{m_i(0, l)!m_i(1,l)!} \cdot
+  \frac{m_i(0, l')!m_i(1,l')!m_i(0, l'')!m_i(1,l'')!}{(m_i(l') + 1)!(m_i(l'') + 1)!}
+  \label{eq-hboa-final}
+\end{equation}
+
+Initially there are $\Theta(N^2)$ possible ways to split existing leaves, as each of the $N$ single node
+trees can be split by any of the other $N-1$ genes. Each iteration a new edge is added to the decision
+forest, meaning some of the previously tested splits cannot be used. For instance, if $T_i$, which is used
+to decide the value of $x_i$, is split using the value of $x_j$, $T_j$ can no longer be split using $x_i$.
+As a split creates two new leaves, $O(N)$ new potential splits must also be tested. Equation~\ref{eq-hboa-final}
+parses all solutions which reach a leaf to count gene frequencies, requiring $\mu$ time.
+The number of total leaves created depends heavily on the problem and $\mu$.
+However, assuming no splits are accepted or that the cost of testing all future splits is less than
+the initial $\Theta(N^2)$, constructing the model requires $\Omega(\mu N^2)$ time. Each
+model is used to generate $\mu$ solutions, leading to a cost per evaluation of $\Omega(N^2)$.
+
+To generate a solution, the value of each gene $x_i$ is set using its corresponding decision tree $T_i$. Because
+the forest is directed acyclic, there must be an ordering of $T_i$ such that before $T_i$ is executed all
+$x_j$ it uses to make decisions have already been set. As such, previous decisions made by other trees
+are used to follow each $T_i$ until a leaf is reached. The value of $x_i$ is then set based on the
+probably that other solutions reached that leaf with each value of $x_i$.
+
+To perform replacement, hBOA uses restricted tournament replacement. After each new solution is generated
+and evaluated, $w$ solutions are chosen at random from the population, where $w=\min\{N, \frac{\mu}{20}\}$.
+Whichever of the $w$ solutions is the most genetically similar to the offspring is compared with the offspring.
+If the fitness of the offspring is no worse, it replaces the genetically similar member of $w$, otherwise the
+offspring is discarded. This method is designed to preserve genetic diversity as only genetically similar
+solutions must compete on fitness.
+
+To model properly, hBOA is designed to work with large population sizes, resulting in a large number of
+evaluations per generation. As hBOA utilizes explicit diversity maintenance, standard methods for determining
+convergence are not considered very accurate. Therefore the authors suggest that an hBOA run should be
+terminated after performing generations equal to $N$.
+
+Like other model based techniques, hBOA has very few parameters. There is no mutation or crossover,
+and modeling does not rely on any explicit parameters. Solution selection, generation, and replacement
+are all derived from the population size, which must be set by the user.
+
+\subsection{Parameter-less hBOA}
+Using the methods first introduced by \cite{harik:1999:parameterlessga} for the Parameter-less GA,
+\cite{pelikan:2004:parameterlesshboa} created Parameter-less hBOA which automatically scales its
+population size to fit the problem. This is done by maintaining a list of concurrent populations
+using exponentially scaled population sizes.
+
+A run of Parameter-less hBOA starts with a single population of size $\mu_0$, conventionally set
+to $\mu_0=10$. After two generations are performed, a new population of size $\mu_1 = 2\mu_0$ is created
+and performs a generation. Evolution then continues with the $\mu_0$ population performing two generations
+for each one performed by $\mu_1$. Each time population $\mu_i$ performs its second generation a new population
+$\mu_{i+1}=2\mu_i$ is created, which performs generations half as often as $\mu_i$. In this way an infinite number of
+parallel population can be simulated, with each population receiving the same number of total evaluations.
+
+In all other aspects each population is identical to an hBOA population using a fixed $\mu$. No search information
+is shared among populations, and each search is independently terminated. As such Parameter-less hBOA cannot
+perform better than hBOA using the optimal population size for a given instance, as it must also spend evaluations
+on populations of different sizes. This inefficiency is bounded by a log multiple of the total number of
+evaluations~\citep{pelikan:1999:worstparameter-less}.
 
@@
 
-\subsection{Parameter-less Population Pyramid}
+\section{Parameter-less Population Pyramid}
 \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid (P3) as a method for
@@
 between the populations.
-
-\subsection{Hierarchical Bayesian Optimization Algorithm}
-
-\cite{pelikan:2006:hboa} used statistical principles in combination with a decision tree structure
-to create the Hierarchical Bayesian Optimization Algorithm (hBOA). This method creates a model of
-epistatic relationships between genes which is then used to stochastically generate new solutions.
-Each generation a binary tournament with replacement is used to selection $\mu$ solutions from
-the population. These solutions are then used to build the model, which in turn is used to generate $\mu$ new
-solutions. The new solutions are then integrated into the population using restricted tournament
-replacement.
-
-Conceptually, the model built by hBOA is trying to infer rules of the form ``Given that this
-subset of genes are set to these values, how frequently is gene $x_i$ set to value v?'' This can
-be represented using a directed acyclic decision forest, with each tree in the forest representing one gene
-in the solution. In the decision tree $T_i$, which is used to set the value of gene $x_i$,
-each internal node represents previous decisions on how to set
-some other gene $x_j$, with the children of that node representing how the decision was made. The
-leaves of each tree give the probability that $x_i$ should be set to each possible value.
-Therefore in binary alphabets binary trees with a single probability $p$ stored at each leaf
-can be used, where $p$ is the probability $x_i$ should be set to one.
-
-The forest is constructed iteratively, with each tree initially containing a single leaf
-and with each leaf storing a pointer for each selected solution. Each iteration the algorithm considers
-all possible ways of splitting an existing leaf using another gene $x_j$, such that solutions in the
-leaf are moved to the newly created leaves based on their value for $x_j$. The general goal is to
-separate the solutions such that all solutions with $x_i = 0$ move to one leaf while solutions with
-$x_j = 1$ move to the other. Whichever split maximizes Equation~\ref{eq-hboa} is then kept.
-This process continues until none of the splits result in a higher quality model than the current one. To favor compact
-models, Equation~\ref{eq-hboa} is scaled by Equation~\ref{eq-hboa-stop}, which provides increased cost
-for more total leaves.
-
-\begin{equation}
-  BDe(B) = \prod_{i=0}^{N}\prod_{l\in L_i} \frac{\Gamma(m'_i(l))}{\Gamma(m_i(l) + m'_i(l))}
-  \prod_{x_i}\frac{\Gamma(m_i(x_i, l) + m'_i(x_i,l))}{\Gamma(m'_i(x_i,l))}
-  \label{eq-hboa}
-\end{equation}
-
-\begin{equation}
-  p(B) = c2^{-0.5(\sum_i|L_i|)log_2\mu}
-  \label{eq-hboa-stop}
-\end{equation}
-
-While Equation~\ref{eq-hboa} is very complex and almost always results in near infinitesimal results, many
-of the terms can be canceled out as it is only important to know the relative quality between two models:
-$p(B)BDe(B) < p(B')BDe(B')$.
-The outermost product of Equation~\ref{eq-hboa} iterates over all trees in the forest. However, each split
-can modify only one of the trees and therefore the contribution of all others can be canceled. The middle
-product is across all leaves in the tree. Again since only one leaf can be changed, all other terms can
-be canceled. By convention hBOA uses uninformed Bayesian priors of $m'_i(l)= 2$ and $m'_i(x_i, l)=1$ for
-binary alphabets. As $\Gamma(n) = (n-1)!$ this means the top term in the middle product and the bottom
-term in the third product reduce to 1. The only remaining terms are then $m_i(l)$ and $m_i(x_i, l)$ which
-represent the number of solutions which reached leaf $l$ and the number of solutions which reached leaf $l$
-with a specific value, respectively.
-
-Equation~\ref{eq-hboa-stop} can also be simplified when doing comparisons. The only difference for $p(B)$ and $p(B')$ is the total
-number of leaves ($\sum_i|L_i|$) has increased by 1. Therefore the ratio simplifies to $2^{0.5 log_2\mu}$ regardless of
-total model size.
-
-All combined, this means the expression $BDe(B) < BDe(B')$ can be calculated using Equation~\ref{eq-hboa-final},
-where B' is different from B by exactly 1 split, such that $l$ was split to create $l'$ and $l''$. Only splits that
-satisfy this inequality are considered good enough to keep. Furthermore, whichever split maximizes the right side of the
-Equation~\ref{eq-hboa-final} is chosen each iteration. Note that these factorials can be exceedingly large and
-therefore its imperative that implementations avoid rounding errors and overflows.
-
-\begin{equation}
-  2^{0.5 log_2\mu} < \frac{(m_i(l) + 1)!}{m_i(0, l)!m_i(1,l)!} \cdot
-  \frac{m_i(0, l')!m_i(1,l')!m_i(0, l'')!m_i(1,l'')!}{(m_i(l') + 1)!(m_i(l'') + 1)!}
-  \label{eq-hboa-final}
-\end{equation}
-
-Initially there are $\Theta(N^2)$ possible ways to split existing leaves, as each of the $N$ single node
-trees can be split by any of the other $N-1$ genes. Each iteration a new edge is added to the decision
-forest, meaning some of the previously tested splits cannot be used. For instance, if $T_i$, which is used
-to decide the value of $x_i$, is split using the value of $x_j$, $T_j$ can no longer be split using $x_i$.
-As a split creates two new leaves, $O(N)$ new potential splits must also be tested. Equation~\ref{eq-hboa-final}
-parses all solutions which reach a leaf to count gene frequencies, requiring $\mu$ time.
-The number of total leaves created depends heavily on the problem and $\mu$.
-However, assuming no splits are accepted or that the cost of testing all future splits is less than
-the initial $\Theta(N^2)$, constructing the model requires $\Omega(\mu N^2)$ time. Each
-model is used to generate $\mu$ solutions, leading to a cost per evaluation of $\Omega(N^2)$.
-
-To generate a solution, the value of each gene $x_i$ is set using its corresponding decision tree $T_i$. Because
-the forest is directed acyclic, there must be an ordering of $T_i$ such that before $T_i$ is executed all
-$x_j$ it uses to make decisions have already been set. As such, previous decisions made by other trees
-are used to follow each $T_i$ until a leaf is reached. The value of $x_i$ is then set based on the
-probably that other solutions reached that leaf with each value of $x_i$.
-
-To perform replacement, hBOA uses restricted tournament replacement. After each new solution is generated
-and evaluated, $w$ solutions are chosen at random from the population, where $w=\min\{N, \frac{\mu}{20}\}$.
-Whichever of the $w$ solutions is the most genetically similar to the offspring is compared with the offspring.
-If the fitness of the offspring is no worse, it replaces the genetically similar member of $w$, otherwise the
-offspring is discarded. This method is designed to preserve genetic diversity as only genetically similar
-solutions must compete on fitness.
-
-To model properly, hBOA is designed to work with large population sizes, resulting in a large number of
-evaluations per generation. As hBOA utilizes explicit diversity maintenance, standard methods for determining
-convergence are not considered very accurate. Therefore the authors suggest that an hBOA run should be
-terminated after performing generations equal to $N$.
-
-Like other model based techniques, hBOA has very few parameters. There is no mutation or crossover,
-and modeling does not rely on any explicit parameters. Solution selection, generation, and replacement
-are all derived from the population size, which must be set by the user.
-
-\subsection{Parameter-less hBOA}
-Using the methods first introduced by \cite{harik:1999:parameterlessga} for the Parameter-less GA,
-\cite{pelikan:2004:parameterlesshboa} created Parameter-less hBOA which automatically scales its
-population size to fit the problem. This is done by maintaining a list of concurrent populations
-using exponentially scaled population sizes.
-
-A run of Parameter-less hBOA starts with a single population of size $\mu_0$, conventionally set
-to $\mu_0=10$. After two generations are performed, a new population of size $\mu_1 = 2\mu_0$ is created
-and performs a generation. Evolution then continues with the $\mu_0$ population performing two generations
-for each one performed by $\mu_1$. Each time population $\mu_i$ performs its second generation a new population
-$\mu_{i+1}=2\mu_i$ is created, which performs generations half as often as $\mu_i$. In this way an infinite number of
-parallel population can be simulated, with each population receiving the same number of total evaluations.
-
-In all other aspects each population is identical to an hBOA population using a fixed $\mu$. No search information
-is shared among populations, and each search is independently terminated. As such Parameter-less hBOA cannot
-perform better than hBOA using the optimal population size for a given instance, as it must also spend evaluations
-on populations of different sizes. This inefficiency is bounded by a log multiple of the total number of
-evaluations~\citep{pelikan:1999:worstparameter-less}.
-
-\subsection{$1+(\lambda, \lambda)$}
-\cite{doerr:2013:lambdalambda} presented the first genetic algorithm to provably show
-the advantages of performing crossover on simple landscapes. This comparatively simple
-algorithm maintains only a single individual and a self-controlled parameter $\lambda$.
-
-Each iteration, the number of bits to flip is chosen from the binomial distribution $b\sim B(N, \frac{\lambda}{N})$.
-Next, $\lfloor\lambda\rfloor$ offspring are produced by flipping $b$ bits. The
-best mutant then produces $\lambda$ offspring via uniform crossover with the original parent, such that each gene comes from the
-mutant with probability $\frac{1}{\lambda}$. In the original algorithm the best
-offspring produced by crossover then replaces the original parent if its fitness is no worse.
-The $\lambda$ parameter, which is initialized to 1, is decreased if the offspring replaced
-its parent and increased otherwise.
-
-The original formulation was designed specifically for unimodal landscapes and as such were
-not directly suitable for multimodal problems. \cite{goldman:2014:p3} extended $1+(\lambda, \lambda)$
-to include random restarts. As search stagnates, the $\lambda$ parameter increases in value. Eventually
-this results in $\lambda \ge N$ causing mutation to always flip all bits of the individual.
-As this prevents any future improvement whenever $\lambda \ge N$ search is restarted from a random solution with $\lambda$
-reset to 1.
-
-A few other efficiency modifications were also made. If there is a tie in crossover offspring fitness,
-whichever has a larger hamming distance from the parent is retained. This encourages drifting across plateaus.
-The ``mod'' control strategy proposed by \cite{doerr:2013:lambdalambda} was not used as it conflicted with
-the random restart strategy.
-If a crossover individual is identical to either of its parents, it is not evaluated.
-If mutation produces an offspring which is better than the best crossover offspring, it is used to compare
-against the original parent.
 

### 2014-09-11 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Randomly generated problems

@@
 Finally, only hBOA and LTGA require any parameters, with each of these only requiring a population
-size. This makes knowing the optimal behavior of these algorithms much more tractable.
+size. This makes knowing the optimal behavior of these algorithms much more tractable. All of the
+algorithms are also gene order independent, fitness scale invariant, and unbiased. This means
+for any problem the order in which problem variables appear in the genome can be changed
+without changing the behavior of the search. The fitness can also be manipulated in any fashion
+as long as the rank ordering of solutions is unchanged. They are also unaffected by the meaning
+assigned to each bit, such that inverting a predetermined random subset of genes before evaluation
+will not impact search efficiency.
 
@@
 Using local search in LTGA was examined by \cite{bosman:2011:lsbbo} and found
-to provide no significant improvement.
+to provide no significant improvement. A likely cause was that that study applied
+local search to every solution, not just the initial population, resulting in significant overhead.
 
@@
 \label{sec-problems}
-TODO Consider adding paragraph about problem equivalence.
 
@@
 As a final class of well known problems, we have chosen to borrow the Rastrigin problem from real valued
-optization. This problem's landscape, determined by Equation~\ref{eq-rast}, is highly multimodal caused by
+optimization. This problem's landscape, determined by Equation~\ref{eq-rast}, is highly multimodal caused by
 the oscillating cosine function. \cite{goldman:2014:p3} proposed the Discretized Rastrigin problem, such
@@
 \subsection{Randomly Generated Problem Classes}
+While well defined landscapes can provide specific insights into how an algorithm works,
+their static nature can lead to misleading results. Specifically algorithm quality might
+be very fragile, only effective at searching well behaved landscapes. A more realistic
+test of an algorithm's black box effectiveness is to work with randomly generated instances
+drawn from a problem class. When tested over a sufficiently large sample it is then possible
+to draw more general conclusions about the algorithms effectiveness. The challenge with these
+landscapes is determining the global optimum to know if an algorithm was successful.
 
-~\cite{wright:2000:solvingnk}
+Perhaps the most common model for generating random rugged landscapes is the NK model. An
+NK Landscape determines the fitness of each gene based on epistatic relationships with $K$
+other genes in the genome. This fitness is specified using a randomly generated table of
+fitness values, were each possible combination of the $K+1$ genes is mapped to some floating
+point value $[0-1]$. In unrestricted NK landscapes the relationships between genes are also
+randomly chosen and as a result finding the global optimum is $NP$-Hard for $k>1$. However,
+if epistasis is instead set such that each gene depends on the $K$ directly following it
+in the genome the solution can be found in polynomial time~\citep{wright:2000:solvingnk}.
+These Nearest Neighbor NK landscapes are therefore ideal for search algorithm testing.
+For all of our experiments we fixed $K=5$ to ensure highly rugged landscapes.
 
-~\cite{saul:1994:spinglass}
-\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}
+\cite{saul:1994:spinglass} presents a combinatorial benchmark problem derived from physics:
+Ising Spin Glasses. A spin glass is defined by interaction terms between variables with
+the genome encoding the sign for each variable and a goal to minimize the total interaction value.
+Similar to NK Landscapes the general class is $NP$-Hard to optimize, but the $2D\pm J$ subset of
+Ising Spin Glasses can be polynomially solved\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}.
 
-~\cite{goldman:2014:p3} % MAXSAT
+As our final class of randomly generated problems we chose the Maximum Satisfiability (MAX-SAT) problem.
+Related to the more common 3-SAT problem, a MAX-SAT is defined by a set of three term clauses.
+Each term is a randomly chosen variable, which may also be negated. A clause scores if an and only if
+at least one term in the clause evaluates to true. In order to make MAX-SAT instances with a known global
+optimum \cite{goldman:2014:p3} proposed constructing clauses around a fixed solution. In this way the
+signs of the terms are set to ensure the target solution satisfies the clause. To ensure each
+problem is challenging chose a clause to variable ratio of $4.27$.
 

### 2014-09-11 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Tuning

@@
 \label{sec-tuning}
-
-~\cite{goldman:2012:ltga} % Bisection with no parameters
-
-~\cite{goldman:2014:p3}
-
-~\cite{jovanovic:1997:ruleofthree}
+While four of the six algorithms in our experiments do not require any user specified parameters,
+hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
+handicapped, we have extensively tuned each algorithm to each problem size for each problem to
+determine the optimal population size using the bisection method \citep{sastry:2001:bisection}.
+Extended by \cite{goldman:2012:ltga} the bisection sets the population size by iteratively
+doubling the population size until some success criteria is met and then performing bisection
+between the lowest successful and highest unsuccessful sizes. Thus the minimum population size
+which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
+of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
+above by $\frac{3}{r+1}$~\cite{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
+do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
 

### 2014-09-16 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Added table.

@@
 {\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}
+
+\usepackage{array}
+\newcolumntype{L}{>{\centering\arraybackslash}m{1.9cm}}
 
@@
 
+\begin{table}
+	\centering
+	\begin{tabular}{|c|r|r|r|r|r|r|r|r|}
+	  \cline{2-9}
+	  \multicolumn{1}{c|}{} & \multicolumn{2}{L|}{Deceptive Trap} & \multicolumn{2}{L|}{Deceptive Step Trap}
+	                        & \multicolumn{2}{L|}{HIFF} & \multicolumn{2}{L|}{Rastrigin} \\ \cline{2-9}
+	  \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{a} & \multicolumn{1}{c|}{b}
+	                        & \multicolumn{1}{c|}{a} & \multicolumn{1}{c|}{b}
+	                        & \multicolumn{1}{c|}{a} & \multicolumn{1}{c|}{b}
+	                        & \multicolumn{1}{c|}{a} & \multicolumn{1}{c|}{b} \\ \hline
+	                          % DT                    % DST                    % HIFF                  % Rast
+	  {Hill Climber}          &$<$0.00 &        11.10 &$<$0.00 &        10.60 &$<$0.00 &         9.45 &$<$0.00 &         5.50 \\ \hline
+	  {$1+(\lambda,\lambda)$} &   0.02 &         3.66 &$<$0.00 &         6.69 &$<$0.00 &         6.55 &$<$0.00 &         4.50 \\ \hline
+	  {hBOA}                  &  18.16 &         2.28 &  40.38 &         2.07 &   4.60 &         1.82 &   3.50 &         2.29 \\ \hline
+	  {Parameter-less hBOA}   &  16.92 &         2.72 &  28.71 &         2.59 &   4.94 &         2.25 &   3.55 &         2.61 \\ \hline
+	  {LTGA}                  & 324.88 & \textbf{1.28}& 165.27 &         1.72 &   7.02 &         1.66 &  21.05 & \textbf{1.43}\\ \hline
+	  {P3}                    &  68.80 &         1.38 & 354.69 & \textbf{1.56}&   4.06 & \textbf{1.59}&   5.29 &         1.59 \\ \hline
+  \end{tabular}
+	\caption{Results of fitting $y=ax^b$ model where $y$ is evaluations to success
+	         and $x$ is problem size. Bold entries are the statistically lowest growth rates.}
+	\label{table-bigo}
+\end{table}
+
 TODO Table listing empirical $O(N)$ times for each technique on each problem. Include

### 2014-09-16 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Pre Bill editting.

@@
 truly novel results, sometimes publishable or patentable in their own right~\citep{kannappan:2014:humies}.
-While it is not possible for any algorithm to do better than random across all possible
+While it is not possible for any algorithm to do better than random search across all possible
 problems~\citep{Wolpert:1997:nfl}, effectiveness can be achieved by assuming the search
-landscape has structure and then biasing the algorithm toward exploiting those features.
+landscape has structure and then biasing the algorithm toward exploiting that structure.
 
@@
 \cite{harik:1999:parameterlessga} were able to design an entirely parameter-less GA by
-designing control strategies which made on-line modifications to each parameter based
-on how well search was progressing. Unfortunately these methods were provably less efficient
+leveraging schema theory and parallel populations. Unfortunately these methods were provably less efficient
 than directly setting the parameters to optimal values~\citep{pelikan:1999:worstparameter-less}.
 
-One area that has been very effective at reducing the number of algorithm parameters has been
+One area that has been very effective at reducing the number of algorithm parameters is
 model based search. \cite{pelikan:2006:hboa}'s Hierarchical Bayesian Optimization
@@
 without changing the behavior of the search. The fitness can also be manipulated in any fashion
-as long as the rank ordering of solutions is unchanged. They are also unaffected by the meaning
+as long as the rank ordering of solutions is unchanged. These algorithms are also unaffected by the meaning
 assigned to each bit, such that inverting a predetermined random subset of genes before evaluation
@@
 
-Each iteration, the number of bits to flip is chosen from the binomial distribution $b\sim B(N, \frac{\lambda}{N})$.
+Each iteration, the number of bits to flip is chosen from the binomial distribution $b\sim B(N, \frac{\lambda}{N})$,
+where $N$ is the number of bits in the genome.
 Next, $\lfloor\lambda\rfloor$ offspring are produced by flipping $b$ bits. The
-best mutant then produces $\lambda$ offspring via uniform crossover with the original parent, such that each gene comes from the
+best mutant then produces $\lfloor\lambda\rfloor$ offspring via uniform crossover with the original parent, such that each gene comes from the
 mutant with probability $\frac{1}{\lambda}$. In the original algorithm the best
@@
 this results in $\lambda \ge N$ causing mutation to always flip all bits of the individual.
-As this prevents any future improvement whenever $\lambda \ge N$ search is restarted from a random solution with $\lambda$
+As this prevents any future improvement, whenever $\lambda \ge N$ search is restarted from a random solution with $\lambda$
 reset to 1.
@@
 be canceled. By convention hBOA uses uninformed Bayesian priors of $m'_i(l)= 2$ and $m'_i(x_i, l)=1$ for
-binary alphabets. As $\Gamma(n) = (n-1)!$ this means the top term in the middle product and the bottom
+binary alphabets. As $\Gamma(a) = (a-1)!$ this means the top term in the middle product and the bottom
 term in the third product reduce to 1. The only remaining terms are then $m_i(l)$ and $m_i(x_i, l)$ which
 represent the number of solutions which reached leaf $l$ and the number of solutions which reached leaf $l$
-with a specific value, respectively.
+with a specific value for $x_i$, respectively.
 
-Equation~\ref{eq-hboa-stop} can also be simplified when doing comparisons. The only difference for $p(B)$ and $p(B')$ is the total
-number of leaves ($\sum_i|L_i|$) has increased by 1. Therefore the ratio simplifies to $2^{0.5 log_2\mu}$ regardless of
+Equation~\ref{eq-hboa-stop} can also be simplified when doing comparisons. If model $B'$ has exactly one more
+leaf than model $B$ than $p(B)$ and $p(B')$ the ratio $\frac{p(B)}{p(B')}$ simplifies to $2^{0.5 log_2\mu}$ regardless of
 total model size.
 
-All combined, this means the expression $BDe(B) < BDe(B')$ can be calculated using Equation~\ref{eq-hboa-final},
-where B' is different from B by exactly 1 split, such that $l$ was split to create $l'$ and $l''$. Only splits that
+All combined, this means the expression $p(B)BDe(B) < p(B')BDe(B')$ can be calculated using Equation~\ref{eq-hboa-final},
+where $B'$ is different from $B$ by exactly 1 split, such that $l$ was split to create $l'$ and $l''$. Only splits that
 satisfy this inequality are considered good enough to keep. Furthermore, whichever split maximizes the right side of the
 Equation~\ref{eq-hboa-final} is chosen each iteration. Note that these factorials can be exceedingly large and
-therefore its imperative that implementations avoid rounding errors and overflows.
+therefore it is imperative that implementations avoid rounding errors and overflows.
 
@@
 \cite{thierens:2013:ltgahiff}'s version of LTGA does not use the entire population when determining pairwise entropy.
-Instead, binary tournament use used to select half of the population. This is done to ensure the model is built
+Instead, binary tournament is used to select half of the population. This is done to ensure the model is built
 using only high quality solutions, even during the first generation.
@@
 landscapes. To be of interest these landscapes need to represent interesting
-and important challenges aspects of real world problems.
+and important aspects of real world problems.
 
 Perhaps the most common such landscape is the Deceptive Trap problem \citep{goldberg:1991:gasize}.
-In this landscape the genome is broken up into $k$ sized fully decomposable subproblems referred
+In this landscape the genome is broken up into $k$ bit non-overlapping subproblems referred
 to as traps.
@@
 bits in the trap set to 1. To global optimum in each trap is a string of all 1s, while all
-other solutions lead to a local optima of all 0s. As a result this problem tests an algorithms
+other solutions lead to a local optima of all 0s. As a result this problem tests an algorithm's
 ability to overcome $k$ sized deception and is commonly used to determine how effective crossover
@@
 
-Another challenging aspect of landscapes can be higher order relationships. For these problems
-the are easily detected main effects as well as larger interrelationships. The Hierarchical If
+Another challenging aspect of landscapes can be higher order relationships. The Hierarchical If
 and only If (HIFF) problem is designed to capture the difficulties of this class of problem.
@@
 While well defined landscapes can provide specific insights into how an algorithm works,
-their static nature can lead to misleading results. Specifically algorithm quality might
-be very fragile, only effective at searching well behaved landscapes. A more realistic
+their static nature can create misleading results. Specifically algorithm quality might
+be very fragile such that it is only effective at searching well behaved landscapes. A more realistic
 test of an algorithm's black box effectiveness is to work with randomly generated instances
@@
 point value $[0-1]$. In unrestricted NK landscapes the relationships between genes are also
-randomly chosen and as a result finding the global optimum is $NP$-Hard for $k>1$. However,
+randomly chosen and as a result finding the global optimum is $NP$-Hard for $K>1$. However,
 if epistasis is instead set such that each gene depends on the $K$ directly following it
@@
 Similar to NK Landscapes the general class is $NP$-Hard to optimize, but the $2D\pm J$ subset of
-Ising Spin Glasses can be polynomially solved\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}.
+Ising Spin Glasses can be polynomially solved.\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}
 
 As our final class of randomly generated problems we chose the Maximum Satisfiability (MAX-SAT) problem.
-Related to the more common 3-SAT problem, a MAX-SAT is defined by a set of three term clauses.
+Related to the more common 3-SAT problem, a MAX-SAT instance is defined by a set of three term clauses.
 Each term is a randomly chosen variable, which may also be negated. A clause scores if an and only if
@@
 hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
-handicapped, we have extensively tuned each algorithm to each problem size for each problem to
-determine the optimal population size using the bisection method \citep{sastry:2001:bisection}.
-Extended by \cite{goldman:2012:ltga} the bisection sets the population size by iteratively
-doubling the population size until some success criteria is met and then performing bisection
+handicapped, we have extensively tuned each using the bisection method \citep{sastry:2001:bisection}
+to determine the optimal population size for each size of each problem.
+Extended by \cite{goldman:2012:ltga} this method iteratively
+doubles the population size until some success criteria is met and then performs bisection
 between the lowest successful and highest unsuccessful sizes. Thus the minimum population size
@@
 of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
-above by $\frac{3}{r+1}$~\cite{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
+above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
 do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.

### 2014-09-17 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Misedd this commit.

@@
 Equation~\ref{eq-hboa-stop} can also be simplified when doing comparisons. If model $B'$ has exactly one more
-leaf than model $B$ than $p(B)$ and $p(B')$ the ratio $\frac{p(B)}{p(B')}$ simplifies to $2^{0.5 log_2\mu}$ regardless of
+leaf than model $B$ than the ratio $\frac{p(B)}{p(B')}$ simplifies to $2^{0.5 log_2\mu}$ regardless of
 total model size.

### 2014-09-17 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Filled in some Global Optimum section

@@
 
+The goal of black box optimization is to quickly and reliably obtain high quality solutions. In this
+section we will compare how many evaluations each algorithm requires to find the global optimum on
+each problem.
+
+Figure~\ref{fig-evals-to-success} shows the median number of evaluations required by each of the six
+algorithms to find the global optimum for multiple sizes of each problem.  Each data point in
+Figure~\ref{fig-evals-to-success} represents the median of 100 runs, where unsuccessful runs
+are treated as requiring more evaluations than any successful run. If the median run was not successful
+no point is shown.
+
+The Random Restart Hill Climber and $1+(\lambda, \lambda)$ are both relatively effective on small problem
+sizes, but have difficulty scaling to larger problems.  Only on MAX-SAT are these optimizers competitive
+at larger tested problem sizes. However this may just be an artifact of that the largest tested MAX-SAT
+was an order of magnitude smaller than the largest size tested for most other problems. On the Deceptive
+Trap and Deceptive Step Trap problem $1+(\lambda, \lambda)$ has some other interesting results. For
+problems with 4 or less traps it performs significantly better than any other algorithm. This is likely
+due to an artifact of the landscape as $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
+flipping entire traps. For larger problems it becomes increasingly unlikely that mutation and crossover can
+create enough changes to an individual trap to overcome deception without negatively changing others. This
+behavior also causes high variance in success rate, as evident by the occasional successes on large problems.
+This ability to successfully integrate modifications of more than a single bit generally appears to lead $1+(\lambda, \lambda)$
+outperform the Hill Climber on all problems except Nearest Neighbor NK.
+
+As expected, Parameter-less hBOA was generally efficient than hBOA using an optimal population size. The slopes
+of the two lines are relatively similar across all problems, meeting with the theoretical expectation that
+Parameter-less hBOA is no more than a logarithmic factor worse than hBOA.  The only problem which reversed
+this trend was MAX-SAT, where Parameter-less hBOA actually outperformed hBOA. A potential explanation for
+this upset is that hBOA is using the optimal population size for the problem class, not for each instance.
+Therefore hBOA must use a population size sufficiently large to solve the hardest instances in the class.
+Parameter-less hBOA on the other hand is able to solve easy instances using much smaller population sizes.
+The net effect is therefore that by being more adaptable Parameter-less hBOA is able to achieve better median
+results.
+
+Considering how different hBOA and LTGA are in performing optimization, it is somewhat surprising how similar
+their results are on HIFF. hBOA utilizes Bayesian decision trees to stochastically generate solutions while
+LTGA uses a linkage tree to iteratively modify subsets of genes in an existing solution. On this problem both
+methods likely have the same two limiting factors: initial diversity and generational turnover. In order to
+ensure convergence to the global optima, both methods rely on initial populations which are sufficiently large.
+Both need to preserve enough diversity to ensure higher order building blocks can be created later. Both methods
+also build models once per generation. On HIFF detecting higher order relationships is very unlikely until lower
+order subproblems have been optimized. As a result both LTGA and hBOA are unlikely to solve more than one order
+of subproblem per generation. Therefore both require a number of generations based on the number of subproblem sizes
+in the problem.
+
+LTGA, as the current state-of-the-art, was generally the fastest to find the global optimum of all comparison
+algorithms. This is especially true on the four single instance problems where it often found the global
+optimum in a lower order of complexity.  However on Nearest Neighbor NK and Ising Spin Glasses LTGA appears
+to perform worse than hBOA.  This may be due to LTGA's inability to represent the overlapping nature of these problems.
+On MAX-SAT LTGA only outperforms hBOA. Again, the likely cause to this is LTGA's requirement to have a
+population size sufficiently large to solve the hardest problem in each class, reducing its effectiveness
+on instances of median difficulty.
+
+Across all problem and problem sizes P3 generally finds the global optimum using less evaluations than
+all of the comparison algorithms.  On Deceptive Trap P3 significantly outperforms both hill climbing
+and LTGA by combining the best features of each, using hill climbing to optimize each trap and crossover
+to perfectly mix traps. Deceptive Step Trap less amenable to local search with P3's quality effectively
+tied with both LTGA and hBOA. The slight improvement here may be due to P3's ability to solve some runs
+using smaller population sizes than LTGA and hBOA due to lucky initialization. Using local search on HIFF
+likely saves P3 a generation over LTGA and hBOA as all pairwise subproblems will already be solved. Combined
+with P3's ability to add diversity only as needed this likely explains why P3 is able to solve HIFF a constant
+multiple faster than either population based technique. P3 likely solves its median run of Nearest Neighbor
+NK and Ising Spin Glasses faster than both hBOA and LTGA again by using population sizes fit to each instance
+and not fit to the entire problem class. On MAX-SAT, where the most effective solvers are Hill Climber and
+$1+(\lambda, \lambda)$, P3 outperforms both likely due to its use of local search with a better ability to
+overcome deception.
+
+To more fully examine the hypothesis that its P3's ability to scale to differences in problem difficulty,
+Figure~\ref{fig-evals-to-success-range} shows the upper and lower quartiles for P3 and LTGA. On each problem
+LTGA has a much smaller difference between its best and worst runs. This makes sense as LTGA uses the same
+population size regardless of instance and progresses search generationally. In contrast P3 has a much
+higher split, with some runs finishing very quickly. This added variability is generally on only the lower
+end as P3's upper quartile is only worse than LTGA's upper quartile on Deceptive Step Trap and Nearest
+Neighbor NK. A possible explanation for this is that as P3 has not pre-learned the amount of diversity required
+to solve all instances does not take as risky behavior as LTGA.
+
 \begin{figure}[t]
@@
 \end{figure}
-
-TODO Compare all 6 optimization techniques on evaluations to reach global optimum.  Figure~\ref{fig-evals-to-success}.
 
@@
 \end{figure}
-
-TODO Graphic showing quartiles to illustrate P3's scaling on problems. Figure~\ref{fig-evals-to-success-range}.
 

### 2014-09-18 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Better section names.
Wrote a lot of the fitness over time section

@@
 
-\section{Global Optimum}
+\section{Finding the Global Optimum}
 \label{sec-optimum}
@@
 
+\begin{comment}
 \begin{table}
@@
 
+\end{comment}
+
 \begin{figure}[t]
@@
 
-\section{Stop Anytime}
+\section{Fitness Over Time}
 \begin{figure}[t]
@@
 
+For some applications, finding the global optimum is less important than finding good solutions quickly. Therefore
+we examine this behavior in Figure~\ref{fig-fitness-over-time}. At regular intervals during optimization
+Figure~\ref{fig-fitness-over-time} shows the median of the best fitnesses found by that point of search across
+100 runs. For each problem we show the largest problem size for which we were able to successfully set a population
+size for hBOA, but the trends shown are representative of all larger problem sizes. The maximum reporting interval
+is set to include the slowest P3 run to reach the global optimum.
+
+At almost every recording interval for every problem hBOA has the worst fitness. The only exceptions come when
+the total number of evaluations performed is significantly less than the genome size hBOA reaches better fitness
+than the Hill Climber, $1+(\lambda, \lambda)$, and P3. This early in search hBOA and LTGA are both still evaluating
+individuals in their randomly generated initial populations, while the three techniques worse than hBOA have started
+from a single solution and are attempting to optimize. Of the methods, hBOA uses by far the largest population sizes,
+about 55 times larger than LTGA. As such it makes sense that hBOA would take a long time to being to reach high
+quality solutions.
+
+Parameter-less hBOA generally reports higher fitness than hBOA when using the same number of total evaluations. This makes
+sense as Parameter-less hBOA's use of small populations allows it to begin to optimize long before hBOA has even evaluated
+its entire initial population. However, this trend is reversed after a sufficient number of evaluations, as is apparent
+on Deceptive Step Trap and Ising Spin Glasses. This makes sense as hBOA's population is set to quickly find the global
+optimum.
+
+Early in optimization P3 and the Random Restart Hill Climber have effectively identical quality. This makes sense
+as until P3 performs two restarts it is performing the same evaluations as the Hill Climber. Once P3 begins
+performing crossover it immediately improves over the Hill Climber. As such P3 is likely better than a simple
+hill climber regardless of how long each technique is run and irrespective of how high quality the solution
+found has to be.
+
+Perhaps the most striking result is the quality of $1+(\lambda, \lambda)$. Until quite far into search this
+method performs better than both LTGA and hBOA. Given sufficient evaluations $1+(\lambda, \lambda)$ outperforms
+Random Restart Hill Climbing on all 7 problems. For brief periods in the middle of search it performs the
+best of all techniques on the Deceptive Step Trap, HIFF, Ising Spin Glass, and MAX-SAT problems. A likely
+explanation is $1+(\lambda, \lambda)$'s ability to efficiently incorporate gene modifications of larger
+than one bit. This allows it to overcome the plateaus in Deceptive Step Trap, solve medium sized subproblems
+in HIFF, flip the signs on multiple adjacent bits in Ising Spin Glass, and cross plateaus in MAX-SAT. However,
+this method is not enough to quickly reach the global optima in many of these problems which causes it to eventually
+be overtaken by the model building techniques.
+
+TODO Talk about the shape of P3 and LTGA, specifically the jumps.
+
+\begin{figure}[t]
+  \begin{centering}
+    \begin{subfigure}{.5\textwidth}
+      \begin{centering}
+        \includegraphicsfit{small-pop-dst}
+      \end{centering}
+      \caption{Deceptive Step Trap 203}
+      \label{fig-small-pop-dst}
+    \end{subfigure}%
+    \begin{subfigure}{.5\textwidth}
+      \begin{centering}
+        \includegraphicsfit{small-pop-nk}
+      \end{centering}
+      \caption{Nearest Neighbor NK 200}
+      \label{fig-small-pop-nk}
+    \end{subfigure}
+  \end{centering}
+  \caption{Comparison of how reducing LTGA's population size effects
+           the median best fitness reached during search.}
+\end{figure}
+
 TODO Compare all 6 optimization techniques on fitnesses reached at different times
-during the run on problem sizes. Explain results same for different sizes. Figure~\ref{fig-fitness-over-time}.
+during the run on problem sizes. Explain results same for different sizes.
 
@@
 
+\section{Computational Expenses}
 \begin{figure}[t]

### 2014-09-22 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Wrote the rest of the 'over time' section

@@
 
-TODO Talk about the shape of P3 and LTGA, specifically the jumps.
+For all problems, LTGA lags behind P3 in fitness improvement, but has a similar overall shape. On every problem except
+Rastrigin, LTGA has a significant early delay in fitness improvement, followed by near instantaneous improvement, plateau,
+and followed by a final period of improvement which results in the global optimum. This early period corresponds with
+initialization of the population, with the first fitness gain achieved immediately upon completing the first generation.
+This makes sense as LTGA's mixing strategy performs a sort of local search when the model has not yet detected gene epistasis.
+This brings LTGA's fitness to almost the same level as the techniques explicitly performing local search, but is not quite
+as effective at reaching local optima. Subsequent generations then make minor fitness improvements, but generally stagnate
+while the model remains uninformed and high quality building blocks in the population are still relatively rare. Once the model
+becomes accurate enough and the probability of a crossover using high quality genetic material increases sufficiently, LTGA
+enters a second period of rapid improvement.
+
+This behavior is easiest to understand on the two Trap problems. LTGA's first generation will push individual traps toward a
+local optima, but it requires more than a single generation to do so completely. Until that occurs, the model is likely behaving
+no better than random as there is little apparent linkage between bits. Once the model begins to identify individual traps crossover
+can begin to increase the frequency of higher fitness local optima. This process begins slowly as low fitness local optima are more
+likely to be chosen for donation than high fitness local optima due to their frequency in the population. The process becomes self
+catalysing as increased frequency of the global optimum trap genes means increased likelihood of the global optimum being spread by crossover.
+
+An almost identical behavior exists in P3, most strikingly in the Deceptive Trap. When solving
+Deceptive Trap, P3 immediately achieves a high fitness, equaled only by the Random Restart Hill Climber in quality. This rise is
+nothing more than bringing every trap to one of the two local optima, with the majority of traps set suboptimally. $1+(\lambda, \lambda)$
+is slightly less efficient about performing single bit improvements, but is the next quickest to reach this plateau. While the hill climber
+remains stuck, P3's use of crossover slowly begins to combine optimal trap values found by multiple independent restarts. Unlike LTGA which
+performs this improvement all at once, P3 integrates all optimal trap values immediately, waiting only on random restarts to find
+the optimal trap setting.
+
+It is important to note that while P3 generally achieved the best fitness across all time points, hBOA and LTGA were tuned specifically
+to find the global optimum and are not necessarily using the optimal population size for finding intermediate fitnesses. Consider
+that for both reducing the population size would almost certainly improve their fitness for evaluation steps less than their population
+size. To examine the effect of population size we tested LTGA using one tenth of the population size required for reliably finding the
+global optimum, with the results given in Figure~\ref{fig-small-pop}. The two problems shown are representative of the behavior of using
+a smaller population size on the other five problems. Reducing the population size resulted in LTGA's initial fitness improvement
+occurring earlier, which makes sense as the first generation is completed much more quickly. On Deceptive Step Trap this means that
+LTGA leaps ahead of P3 due to its ability to overcome the two bit fitness plateaus. Across all tested problems using the smaller
+population size reduced the quality of LTGA's first fitness plateau, likely due to missing required diversity to reach higher quality.
+The second period of improvement also comes earlier and reaches a lower quality fitness. On Deceptive Step Trap this prevents all 100
+runs from reaching the global optimum, down from 100 successful for the full population size. On Nearest Neighbor NK this drops the number
+of successful runs from 98 to 68. As a result we conclude that while reducing the population size of LTGA may improve its quality
+at early points during optimization, doing so reduces the robustness of the final solution found. This is in contrast to P3
+which balances high quality intermediate fitness without trading away eventual optimality.
 
@@
            the median best fitness reached during search.}
+  \label{fig-small-pop}
 \end{figure}
-
-TODO Compare all 6 optimization techniques on fitnesses reached at different times
-during the run on problem sizes. Explain results same for different sizes.
-
-TODO Include graphic for one problem where LTGA was tuned using a significantly higher failure rate.
 

### 2014-09-22 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Added seconds to success figure.
By removing the [t] after figures I was able to better space figures in the paper.
Moved small population figure.
Filled in large bits of computational costs section.

@@
 \section{Parameter-less Population Pyramid}
+\label{sec-p3}
 \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid (P3) as a method for
@@
 
+\begin{figure}
+  \begin{center}
+  \includegraphicsfit{evals-to-success}
+  \end{center}
+  \caption{Comparison of the median number of evaluations to reach the global optimum for
+           the six different optimization methods with respect
+           to problem size.  If the median run did not reach the global optimum no data element
+           is shown.  Results given on a log-log scale.}
+  \label{fig-evals-to-success}
+\end{figure}
+
 The goal of black box optimization is to quickly and reliably obtain high quality solutions. In this
@@
 on instances of median difficulty.
+
+\begin{figure}
+  \begin{center}
+  \includegraphicsfit{evals-to-success-range}
+  \end{center}
+  \caption{Comparison of the upper and lower quartiles of evaluations required
+           to reach the global optimum for P3 and LTGA with respect to problem size.}
+  \label{fig-evals-to-success-range}
+\end{figure}
 
@@
 to solve all instances does not take as risky behavior as LTGA.
-
-\begin{figure}[t]
-  \begin{center}
-  \includegraphicsfit{evals-to-success}
-  \end{center}
-  \caption{Comparison of the median number of evaluations to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-evals-to-success}
-\end{figure}
 
@@
 
-\begin{figure}[t]
-  \begin{center}
-  \includegraphicsfit{evals-to-success-range}
-  \end{center}
-  \caption{Comparison of the upper and lower quartiles of evaluations required
-           to reach the global optimum for P3 and LTGA with respect to problem size.}
-  \label{fig-evals-to-success-range}
-\end{figure}
-
 \section{Fitness Over Time}
-\begin{figure}[t]
+\begin{figure}
   \begin{center}
@@
 the optimal trap setting.
+
+\begin{figure}[t]
+  \begin{centering}
+    \begin{subfigure}{.5\textwidth}
+      \begin{centering}
+        \includegraphicsfit{small-pop-dst}
+      \end{centering}
+      \caption{Deceptive Step Trap 203}
+      \label{fig-small-pop-dst}
+    \end{subfigure}%
+    \begin{subfigure}{.5\textwidth}
+      \begin{centering}
+        \includegraphicsfit{small-pop-nk}
+      \end{centering}
+      \caption{Nearest Neighbor NK 200}
+      \label{fig-small-pop-nk}
+    \end{subfigure}
+  \end{centering}
+  \caption{Comparison of how reducing LTGA's population size effects
+           the median best fitness reached during search.}
+  \label{fig-small-pop}
+\end{figure}
 
@@
 
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{small-pop-dst}
-      \end{centering}
-      \caption{Deceptive Step Trap 203}
-      \label{fig-small-pop-dst}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{small-pop-nk}
-      \end{centering}
-      \caption{Nearest Neighbor NK 200}
-      \label{fig-small-pop-nk}
-    \end{subfigure}
-  \end{centering}
-  \caption{Comparison of how reducing LTGA's population size effects
-           the median best fitness reached during search.}
-  \label{fig-small-pop}
-\end{figure}
-
 \section{Computational Expenses}
@@
            repeated donations (Figure~\ref{fig-donations}) per evaluation as problem size increases.}
-\end{figure}
-
-TODO Include graphic of rebuilds per evaluation. Figure~\ref{fig-rebuilds}.
-
-TODO Include graphic of donations per evaluation. Figure~\ref{fig-donations}.
-
-TODO Include a table of $O(N)$ seconds per evaluation and total seconds for all 4 model building techniques on all 7 problems.
+  \label{fig-costs}
+\end{figure}
+
+When discussing the asymptotic complexity of P3 in Section~\ref{sec-p3}, two aspects eluded precise
+analysis: how expensive is model rebuilding and how many gene donations are made. Figure~\ref{fig-costs}
+provides some insight into how often these two aspects of the algorithm are utilized.
+
+Figure~\ref{fig-rebuilds} reports in an algorithmic sense how expensive model rebuilding for search.
+In order to calculate this value we recorded how many times search rebuilt the model during each run.
+Under the minimal assumption that an evaluation costs $O(N)$ time and that model rebuilds cost $O(N^2)$,
+Figure~\ref{fig-rebuilds} shows the ratio cost of model rebuilding per evaluation. If the cost of model building
+scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically linear.
+For Nearest Neighbor NK, Ising Spin Glasses, and Rastrigin this appears to be the case. For both Trap problems
+and HIFF there appears to be slow growth in the ratio. The problem sizes used for MAX-SAT were not sufficient
+to accurately gauge the asymptotic behavior. Together this suggests that while the cost of building the model
+is almost linear per evaluation, it can grow slowly. However, even in the worst case (HIFF) this growth was
+bounded to no more than twice the algorithmic cost of an evaluation.
+
+When applying a crossover subset, P3 tries random donors from the population until one is found with at least
+one bit different from the improving solution. In theory this can result in up to $O(\mu)$ operations.
+Figure~\ref{fig-donations} examines the observed average number of donations per evaluation performed.
+Ising Spin Glass, HIFF, and Rastrigin all achieve effectively constant behavior here, implying repeated
+donation does not impact the asymptotic runtime of P3. Both Trap functions and Nearest Neighbor NK all
+increase in number of donations as problem size increases, potentially increasing algorithmic costs. An
+important note again is that each donation may range in size from a single bit up to $N-1$, with smaller
+donations far more likely to result in repeated attempts. As such this may cause some super-linear growth
+in P3, but its unlikely to be very high.
+
+\begin{figure}
+  \begin{center}
+  \includegraphicsfit{seconds-to-success}
+  \end{center}
+  \caption{Comparison of the median number of seconds to reach the global optimum for
+           the six different optimization methods with respect
+           to problem size.  If the median run did not reach the global optimum no data element
+           is shown.  Results given on a log-log scale.}
+  \label{fig-seconds-to-success}
+\end{figure}
+
+To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}. Due to the scale
+of our experiments, diverse hardware potentially running multiple processes in parallel were used
+in order to collect all runs. As such these timings should be considered very skeptically as a number
+of sources may contribute to biased error in the results. Under those caveats, the results do suggest
+that any increase in P3's algorithmic complexity is offset by its efficient use of evaluations. On
+Deceptive Trap, Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT it still appears to find the global
+optimum faster than all comparison techniques. On HIFF LTGA is able to match P3's performance almost exactly,
+and on Rastrigin P3 falls behind only on moderate sized problems. Deceptive Step Trap yields the worst performance,
+with LTGA using optimal population sizes having a significantly lower growth curve than P3. This mirrors
+the results from Figure~\ref{fig-evals-to-success} under the assumption that P3's computational time per evaluation is higher
+than LTGA but by no more than a constant factor.
+
+When considering wall clock time hBOA and Parameter-less hBOA perform significantly worse than when using
+evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while under reasonable
+assumptions P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
+hBOA goes from being slightly more efficient than LTGA in terms of evaluations to significantly worse in terms of seconds.
+As P3 and LTGA require a similar asymptotic complexity per
+evaluation as the Hill Climber and $1+(\lambda, \lambda)$, no similar change in ordering occurs.
+
+These results suggest that even for minimally expensive fitness functions P3 can find the global optimum
+at least as quickly in seconds as its optimally tuned competitors. However, due to the potential noise in our data collection
+we are hesitant to draw too strong a conclusion.
 

### 2014-09-23 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Fixed dash in self-catalysing

@@
 can begin to increase the frequency of higher fitness local optima. This process begins slowly as low fitness local optima are more
-likely to be chosen for donation than high fitness local optima due to their frequency in the population. The process becomes self
-catalysing as increased frequency of the global optimum trap genes means increased likelihood of the global optimum being spread by crossover.
+likely to be chosen for donation than high fitness local optima due to their frequency in the population. The process becomes
+self-catalysing as increased frequency of the global optimum trap genes means increased likelihood of the global optimum being spread by crossover.
 

### 2014-09-23 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Crossover rates and successes.

@@
 \section{Fitness Over Time}
+\label{sec-overtime}
 \begin{figure}
@@
 at least as quickly in seconds as its optimally tuned competitors. However, due to the potential noise in our data collection
-we are hesitant to draw too strong a conclusion.
+we are hesitant to draw too strong of a conclusion.
 
@@
 
-TODO Include graphic of proportion of evaluations spent on crossover. Figure~\ref{fig-cross}.
-
-TODO Include graphic of crossover success Figure~\ref{fig-cross-success}.
+While analysis of optimization speed is useful from a practitioner standpoint, doing so provides very
+little insight into algorithm behavior. To better understand how P3 works in detail we present here
+a look at some internal features specific to P3.
+
+P3 performs optimization using a mixture of hill climbing and crossover. Exactly how much time the algorithm
+dedicates to each is not clear from the algorithm description alone. As such, Figure~\ref{fig-cross} shows
+that the relationship between the two is strongly problem dependent but relatively independent of problem size.
+In general the trend appears to suggest that problems which can be more tightly represented by crossover tend
+to require less crossover evaluations. Deceptive Trap and Rastrigin spent the least evaluations on crossover, using
+15\% and 20\% respectively on the largest problem sizes. Both of these problems are perfectly separable and have relatively
+few local optima in each subproblem. HIFF and Ising Spin Glasses use crossover about 60\%. The HIFF structure can
+be perfectly learned by the linkage tree, but the higher order relationships require multiple stages of crossover to optimize.
+Ising Spin Glasses have an overlapping structure, making them impossible to perfectly represent in a single linkage tree, but
+that overlap is relatively minor in comparison with Nearest Neighbor NK, which uses 81\%. Deceptive Step Trap spends the highest
+percentage of evaluations on crossover, 92\%. Unlike all of the other problems, the value taken by an individual bit in Deceptive
+Step Trap has very little meaning. This is because any trap value is either a local optima or 1 bit flip away from being a local optima.
+This has a combined result of meaning it takes very few local search evaluations to find a local optima but very many crossover operations
+to proliferate the global optima. Epistatic relationships are also going to be very weak and even small population sizes are likely
+to mark all clusters as $useful$.
+
+Solutions are only added to the next highest pyramid level if at least one crossover donation resulted in a fitness
+improvement. Figure~\ref{fig-cross-success} examines the percentage of crossover donations that resulted in a fitness
+improvement. The ordering in Figure~\ref{fig-cross-success} is an almost perfect inversion of Figure~\ref{fig-cross},
+implying that the more successful crossover is at achieving fitness improving donations the lower the proportion of evaluations P3 will
+spend performing crossover. P3 achieves an impressive 30\% success rate for crossover donations on the largest Deceptive Trap
+problem sizes. This makes sense as crossover is able to immediately and perfectly identify entire traps such that all donations
+always contain only complete traps. Furthermore, by forcing crossover to attempt donations of genes different from the current solution,
+donations containing a single trap will have a very high success rate, always improving suboptimal traps which are far more frequently
+created by local search that optimal traps. Rastrigin and HIFF have relatively high success rates of 11\% and 7\%, respectively. Again
+this is likely due to accurate modeling and relatively low diversity of building block alternatives.
+Ising Spin Glasses and Deceptive Step Trap both have low success rates near 0.01\%. This follows as both are difficult to model with
+a linkage tree. In comparison to Figure~\ref{fig-cross-success} Deceptive Step Trap performs better than expected for crossover.
+However, the reason for this is likely that even exceedingly uninformed crossover can make improvements to Deceptive Step Trap as
+flipping two bits in the same trap will move between two local optima. Nearest Neighbor NK had the least successful crossovers,
+with only 0.007\% successful on the largest tested problems. In comparison to Ising Spin Glass, NK's overlap is far more extensive.
+As a result it makes sense that individual crossovers are going to be very unlikely to be fitness improvements.
+
+In comparison with LTGA, the crossover success rates of P3 are lower but similar in problem ordering. This is almost certainly
+a consequence of P3's use of local search. Counter intuitively the use of hill climbing on the initial population is likely reducing
+P3's crossover success not because it reduces model quality or the donation pool, but because it is much more challenging to improve
+locally optimal solutions than randomly generated ones. LTGA's crossover benefits from application to unoptimized solutions, which
+makes its aggregate crossover success incomparable to P3's.
+
+While the frequency of crossover changes per problem, and the likelihood of an individual crossover making a fitness improvement can be
+quite low, the results from Section~\ref{sec-optimum} and Section~\ref{sec-overtime} suggest its still critical to optimization. Without
+crossover P3's performance would be identical to the Random Restart Hill Climber, which was unable to solve even moderately sized problems
+and quickly fell behind P3 in intermediate fitness quality. Therefore even the infrequently successful crossover donation are
+critical to success. This does however suggest a potential avenue for future improvement as different modeling and donation algorithms
+may improve P3's crossover success.
 

### 2014-09-23 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Pyramid shape.

@@
 
-TODO Include graphic of pyramid shape (Figure~\ref{fig-level-size}) and crossover success rates (Figure~\ref{fig-level-success}) at each level.
-
-~\cite{lobo:2011:dynamicpop}
-
-~\cite{goldman:2011:dynamic-parameters}
+Another feature unique to P3 worth investigating is the shape and size of the population pyramid
+constructed for each problem. Figure~\ref{fig-level-size} shows the number of solutions stored at
+each level of the pyramid for the largest tested problem size for each problem. Each point is the
+median size across 100 runs. If a run did not store any solutions at a level it is treated as 0,
+and no point is drawn if the median run had 0 solutions stored at that level. This behavior is
+representative pyramid shape across all problem sizes tested.
+
+With the exception of the dip in Deceptive Step Trap, all of the pyramids show a monotonic reduction
+in size as the level increases. This makes sense from an algorithmic standpoint as solutions are
+added to the first level as long as they are unique local optima. For each subsequent level a
+solution must be unique and a strict fitness improvement over its previously added version.
+As a result it makes sense that it is generally less likely for a solution to be added to
+higher levels than lower ones. This shape is also in many ways desirable from a population
+genetics standpoint. When using a generational model \cite{lobo:2011:dynamicpop} found theoretical
+evidence that the optimal population size reduces each generation with \cite{goldman:2011:dynamic-parameters}
+providing supporting empirical data. In P3, levels of the pyramid are akin to generations as solutions
+stored in each level have undergone the same number of variation and selection events,
+with higher levels representing more of each. Therefore it is possible that P3's
+pyramid shape is related to its ability to outperform competitors using an optimal
+fixed population size. Subsequently smaller levels are able to focus search,
+preserving only the diversity required by that stage of search. In comparison,
+LTGA and hBOA set the population size at each generation equal to that required
+to ensure sufficient initial diversity.
+
+The number of solutions stored at the fourth level of Deceptive Step Trap is
+significantly lower than that of the third or fifth levels, breaking the decreasing
+trend of the other six problems. This behavior is rooted in the peculiar nature
+of this landscape, and is also present in all other problem sizes tested.
+After local search, all traps in all solutions have a total
+number of 1 bits equal to \{0, 1, 3, 5, 7\} as these correspond to the local
+optima when using $k=7$ and $s=2$. Crossover is very likely to overcome the
+two bit plateaus, and as a result solutions in the second level are unlikely
+to contain the lowest fitness local optima of 5 set bits. Similarly the third
+level has very few traps set to the next worst local optima of 3 bits. As a result,
+most solutions which reach the third level can only be improved by overcoming deception
+and replacing the second best local optima of zero or one bits set to having all seven bits set.
+At this point in search the population has not significantly increased the frequency of optimal
+trap settings and with 8 ways to represent suboptimal local optima linkage is also likely to be
+inaccurate. As a result our expectation is that when performing crossover with the third level
+of the pyramid very few fitness improvements will actually be made, meaning
+few solutions will be added to the fourth level. Those solutions that are added must, by definition,
+contain a higher frequency of optimal trap settings to have achieved a fitness improvement. As
+a result the crossover success rate and modeling quality at level four is likely to be much better
+than at level three, increasing the likelihood of solutions being improved by crossover.
+
+When examining the crossover success rate shown in Figure~\ref{fig-level-success} these suppositions are
+supported. The success rate for Deceptive Step Trap decreases for the first three levels, with the third level
+successful only 0.0004\% of the time. This quickly rebounds at higher levels. On all other problems, the
+crossover success rate follows a trajectory closer expectation. Low levels are more successful than medium
+levels as it is easier to make improvements to less optimized solutions. This trend reverses somewhat at
+higher levels as solution filtering allows for highly accurate models and high frequencies of high quality
+building blocks. The highest level on most problems has a very low crossover success rate, likely because
+so few solutions other than the global optimum every cross with that level.
 

### 2014-09-23 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Conclusion and future work.

@@
 \section{Conclusions and Future Work}
+The Parameter-less Population Pyramid (P3) is a recently introduced method for performing black box
+optimization. P3's primary innovation is the replacement of the generational model with a pyramid of populations.
+This pyramid is constructed iteratively, with both the number of levels and the number of solutions stored
+at each level growing as search progresses. P3 uses a model based method learns a linkage tree from gene epistasis as a method for
+performing effective crossover. Combined with a simple hill climber, P3's design contains
+many synergistic features.
+
+Across a large number of problems and problem sizes P3 required less evaluations to reach the global optimum
+than optimally tuned state of the art competitors. On single instance problems P3's improvement was by a
+constant factor, while for the three problem classes P3's improvement increased with problem size. This
+quality extends to intermediate points during evolution, with P3 generally reaching at least as high
+of fitness as the competitive techniques when using the same number of evaluations. While P3 does require
+modeling overhead, the expense of this overhead is approximately linear with respect to genome size. There
+is some evidence to suggest that even when compared on wall clock time, P3 performs at least as well as
+the best comparison techniques. All of these achievements are made without any problem specific parameter
+tuning, making P3 easier to apply to new domains that its two closets competitors in quality.
+
+P3's quality is likely due to a number of desirable traits. First, mixing local search with model based
+crossover lets search focus on properly mixing high quality solutions. Second, by adding diversity only
+as necessary P3 tends to use the minimal amount of random initialization, unlike other techniques which must
+overcompensate with larger population sizes on single instance problems and consider the worst instance
+when solving problem classes. Third, by heavily exploiting existing diversity before adding more P3 is able
+to reach high quality intermediate fitnesses quickly without prematurely converging. Fourth, the very
+nature of the pyramid's shape may be allowing search to preserve a desirable proportion of diversity at
+each fitness level, similar to a generational model using a decreasing population size.
+
+There are a number of meaningful avenues for future P3 experimentation. Perhaps the most
+pressing for practitioner acceptance is to apply P3 to real world problems and compare
+its results with other black box or even problem specific heuristics. While parameter-less,
+P3 is currently limited to discrete, fixed length genomes evaluated using single objective
+fitness. These limitations can likely be relaxed with future work to make P3 more widely applicable.
+While asymptotically linear, P3's modeling techniques and local search methods are likely going
+to be prohibitively expensive for genome sizes in the hundreds of thousands or millions of genes,
+and our work suggests the inability of the model to capture overlapping linkage may be hindering
+search efficiency. Overcoming these limitations by using a new modeling technique may allow
+the pyramid model even greater flexibility. Similarly, while P3 is able to overcome low
+order deception via linkage learning the iterative improvement method by which crossovers
+are made may mislead search on landscapes with higher order deception.
 

### 2014-09-23 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Added abstract and website (fake).

@@
 \begin{abstract}
-
-TODO About 200 words.
+%Should be about 200 words
+The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
+evolutionary optimization without requiring problem specific user specified parameters.
+This is achieved by combining local search with advanced crossover to iteratively construct
+and exploit multiple populations simultaneously, with each population akin to a generation
+in traditional search.
+
+When compared with five advanced search algorithms across dozens of problem sizes for
+seven different benchmarks P3 reliably required at least a constant factor less evaluations to find the global optimum
+than all tested competitors. P3 also in general maintains the highest fitness reached when comparing
+algorithms based on different number of evaluations performed. Some evidence suggests this performance
+also extends to considering total number seconds to find the global optimum.
+We investigate the causes of this efficiency and provide some evidence that P3's ability to
+add diversity only as needed may be contributing to its success.
+
+This quality is made even more striking in that unlike P3, the best comparison algorithms received extensive
+problem specific tuning to achieve their performance. As a result we conclude that in contrast to
+previous parameter-less techniques P3 is more efficient than existing parameterized methods for
+black box optimization.
 
@@
 
+Our implementations of all of these algorithms as well as all of the raw results and processing scripts
+are available from our website.\footnote{\url{https://github.com/brianwgoldman/FastEfficientP3}}
 \subsection{Random Restart Hill Climber}

### 2014-09-24 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Filled in header.
Major editting pre-Bill. Added in more numbers when discussing figures.

@@
 
-\ecjHeader{x}{x}{xxx-xxx}{200X}{TODO 45-character paper description}{B. W. Goldman and W. F. Punch}
+% 45-character limit on header
+\ecjHeader{x}{x}{xxx-xxx}{201X}{Parameter-less Population Pyramid}{B. W. Goldman and W. F. Punch}
 \title{\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid}
@@
 The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
-evolutionary optimization without requiring problem specific user specified parameters.
+evolutionary optimization without requiring problem specific parameters.
 This is achieved by combining local search with advanced crossover to iteratively construct
@@
 seven different benchmarks P3 reliably required at least a constant factor less evaluations to find the global optimum
-than all tested competitors. P3 also in general maintains the highest fitness reached when comparing
-algorithms based on different number of evaluations performed. Some evidence suggests this performance
-also extends to considering total number seconds to find the global optimum.
+than all tested competitors. When compared using fixed numbers of evaluations P3
+generally finds the highest quality solutions. Some evidence also suggests this performance
+extends to considering total number seconds to find the global optimum even when using efficient fitness functions.
 We investigate the causes of this efficiency and provide some evidence that P3's ability to
@@
 do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
+As bisection can make infinitely large population sizes, any run which had not found the global
+optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.
 
@@
 are treated as requiring more evaluations than any successful run. If the median run was not successful
-no point is shown.
+no point is shown. The maximum problem size used for each problem was set to be the largest size we could
+feasibly determine an optimal population size for LTGA. For many larger problems results are not shown
+for hBOA due to the extreme computational cost required to optimally set the population size.
 
@@
 sizes, but have difficulty scaling to larger problems.  Only on MAX-SAT are these optimizers competitive
-at larger tested problem sizes. However this may just be an artifact of that the largest tested MAX-SAT
+at larger tested problem sizes. However, this may just be an artifact as the largest tested MAX-SAT
 was an order of magnitude smaller than the largest size tested for most other problems. On the Deceptive
@@
 problems with 4 or less traps it performs significantly better than any other algorithm. This is likely
-due to an artifact of the landscape as $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
+because $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
 flipping entire traps. For larger problems it becomes increasingly unlikely that mutation and crossover can
@@
 This ability to successfully integrate modifications of more than a single bit generally appears to lead $1+(\lambda, \lambda)$
-outperform the Hill Climber on all problems except Nearest Neighbor NK.
-
-As expected, Parameter-less hBOA was generally efficient than hBOA using an optimal population size. The slopes
+to outperform the Hill Climber on all problems except Nearest Neighbor NK.
+
+As expected, Parameter-less hBOA was generally less efficient than hBOA using an optimal population size. The slopes
 of the two lines are relatively similar across all problems, meeting with the theoretical expectation that
@@
 
-LTGA, as the current state-of-the-art, was generally the fastest to find the global optimum of all comparison
-algorithms. This is especially true on the four single instance problems where it often found the global
-optimum in a lower order of complexity.  However on Nearest Neighbor NK and Ising Spin Glasses LTGA appears
+LTGA, as the current state-of-the-art, was generally at least as fast to find the global optimum as hBOA.
+However on Nearest Neighbor NK and Ising Spin Glasses LTGA appears
 to perform worse than hBOA.  This may be due to LTGA's inability to represent the overlapping nature of these problems.
-On MAX-SAT LTGA only outperforms hBOA. Again, the likely cause to this is LTGA's requirement to have a
+On MAX-SAT LTGA only outperforms hBOA and performs worse than the other 4 techniques.
+Again, the likely cause to this is LTGA's requirement to have a
 population size sufficiently large to solve the hardest problem in each class, reducing its effectiveness
@@
 
-Across all problem and problem sizes P3 generally finds the global optimum using less evaluations than
-all of the comparison algorithms.  On Deceptive Trap P3 significantly outperforms both hill climbing
+Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
+The largest of these used 49 bits, with P3 achieving the best results on all 92 larger
+configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the 5, 4, and 1 smallest Deceptive
+Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the 3 smallest Nearest Neighbor NK instances
+and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
+on 5 smallest Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
+
+On Deceptive Trap P3 significantly outperforms both hill climbing
 and LTGA by combining the best features of each, using hill climbing to optimize each trap and crossover
-to perfectly mix traps. Deceptive Step Trap less amenable to local search with P3's quality effectively
+to perfectly mix traps. Deceptive Step Trap is less amenable to local search with P3's quality effectively
 tied with both LTGA and hBOA. The slight improvement here may be due to P3's ability to solve some runs
@@
 
-To more fully examine the hypothesis that its P3's ability to scale to differences in problem difficulty,
+To more fully examine the hypothesis that some of P3's success is do to its ability to scale to differences in problem difficulty,
 Figure~\ref{fig-evals-to-success-range} shows the upper and lower quartiles for P3 and LTGA. On each problem
@@
 Neighbor NK. A possible explanation for this is that as P3 has not pre-learned the amount of diversity required
-to solve all instances does not take as risky behavior as LTGA.
+to solve all instances does not take as risky of behavior as LTGA.
 
@@
 
-At almost every recording interval for every problem hBOA has the worst fitness. The only exceptions come when
-the total number of evaluations performed is significantly less than the genome size hBOA reaches better fitness
+% 181 total
+% 126 has P3 best
+% 27 LTGA
+% 20 hboa
+% 50 lambdalambda
+% hc 9
+% phboa 18
+
+At 63\% of recording intervals hBOA has the worst fitness of any solver. Many of the exception occur when
+the total number of evaluations performed is significantly less than the genome size and hBOA reaches better fitness
 than the Hill Climber, $1+(\lambda, \lambda)$, and P3. This early in search hBOA and LTGA are both still evaluating
@@
 from a single solution and are attempting to optimize. Of the methods, hBOA uses by far the largest population sizes,
-about 55 times larger than LTGA. As such it makes sense that hBOA would take a long time to being to reach high
+about 55 times larger than LTGA on average for problems shown. As such it makes sense that hBOA would take a long time to reach high
 quality solutions.
 
-Parameter-less hBOA generally reports higher fitness than hBOA when using the same number of total evaluations. This makes
+Parameter-less hBOA reported equal or higher fitness than hBOA when using the same number of total evaluations in 68\% of intervals. This makes
 sense as Parameter-less hBOA's use of small populations allows it to begin to optimize long before hBOA has even evaluated
@@
 
-Early in optimization P3 and the Random Restart Hill Climber have effectively identical quality. This makes sense
-as until P3 performs two restarts it is performing the same evaluations as the Hill Climber. Once P3 begins
-performing crossover it immediately improves over the Hill Climber. As such P3 is likely better than a simple
+Early in optimization P3 and the Random Restart Hill Climber have effectively identical quality. This is because
+P3 performs the same evaluations as the Hill Climber for the first two restarts. Once P3 begins
+performing crossover it immediately improves over the Hill Climber. In 95\% of intervals P3 had a fitness at
+least as high as Hill Climbing. As such P3 is likely better than a simple
 hill climber regardless of how long each technique is run and irrespective of how high quality the solution
@@
 For all problems, LTGA lags behind P3 in fitness improvement, but has a similar overall shape. On every problem except
-Rastrigin, LTGA has a significant early delay in fitness improvement, followed by near instantaneous improvement, plateau,
-and followed by a final period of improvement which results in the global optimum. This early period corresponds with
+Rastrigin, LTGA has five distinct periods: fitness plateau, near instantaneous improvement, fitness plateau,
+and improvement to global optimum. The early period corresponds with
 initialization of the population, with the first fitness gain achieved immediately upon completing the first generation.
-This makes sense as LTGA's mixing strategy performs a sort of local search when the model has not yet detected gene epistasis.
-This brings LTGA's fitness to almost the same level as the techniques explicitly performing local search, but is not quite
+This can be explained as LTGA's mixing strategy performs a sort of local search when the model has not yet detected gene epistasis.
+This brings LTGA's fitness to almost the same level as the techniques explicitly performing local search, but it is not quite
 as effective at reaching local optima. Subsequent generations then make minor fitness improvements, but generally stagnate
@@
 
-It is important to note that while P3 generally achieved the best fitness across all time points, hBOA and LTGA were tuned specifically
+It is important to note that while P3 achieved the best fitness across 70\% of time points, hBOA and LTGA were tuned specifically
 to find the global optimum and are not necessarily using the optimal population size for finding intermediate fitnesses. Consider
@@
 global optimum, with the results given in Figure~\ref{fig-small-pop}. The two problems shown are representative of the behavior of using
-a smaller population size on the other five problems. Reducing the population size resulted in LTGA's initial fitness improvement
-occurring earlier, which makes sense as the first generation is completed much more quickly. On Deceptive Step Trap this means that
+a smaller population size on the other five problems. In aggregate, P3 still achieved a fitness at
+least as high as LTGA using the smaller population size at 80\% of intervals.
+Reducing the population size resulted in LTGA's initial fitness improvement
+occurring earlier, as the first generation is completed much more quickly. On Deceptive Step Trap this means that
 LTGA leaps ahead of P3 due to its ability to overcome the two bit fitness plateaus. Across all tested problems using the smaller
 population size reduced the quality of LTGA's first fitness plateau, likely due to missing required diversity to reach higher quality.
-The second period of improvement also comes earlier and reaches a lower quality fitness. On Deceptive Step Trap this prevents all 100
-runs from reaching the global optimum, down from 100 successful for the full population size. On Nearest Neighbor NK this drops the number
+The second period of improvement also comes earlier and reaches a lower quality fitness. On Deceptive Step Trap 0
+runs reach the global optimum, down from 100 successful for the full population size. On Nearest Neighbor NK this drops the number
 of successful runs from 98 to 68. As a result we conclude that while reducing the population size of LTGA may improve its quality
@@
 
-Figure~\ref{fig-rebuilds} reports in an algorithmic sense how expensive model rebuilding for search.
+Figure~\ref{fig-rebuilds} reports in an algorithmic sense how expensive model rebuilding is for search.
 In order to calculate this value we recorded how many times search rebuilt the model during each run.
-Under the minimal assumption that an evaluation costs $O(N)$ time and that model rebuilds cost $O(N^2)$,
+Under the minimal assumption that an evaluation costs $\Omega(N)$ time and that model rebuilds cost $O(N^2)$,
 Figure~\ref{fig-rebuilds} shows the ratio cost of model rebuilding per evaluation. If the cost of model building
-scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically linear.
+scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically constant.
 For Nearest Neighbor NK, Ising Spin Glasses, and Rastrigin this appears to be the case. For both Trap problems
@@
 increase in number of donations as problem size increases, potentially increasing algorithmic costs. An
-important note again is that each donation may range in size from a single bit up to $N-1$, with smaller
+important note is that each donation may range in size from a single bit up to $N-1$, with smaller
 donations far more likely to result in repeated attempts. As such this may cause some super-linear growth
@@
 of our experiments, diverse hardware potentially running multiple processes in parallel were used
-in order to collect all runs. As such these timings should be considered very skeptically as a number
+in order to collect our results. As such these timings should be considered very skeptically as a number
 of sources may contribute to biased error in the results. Under those caveats, the results do suggest
@@
 the results from Figure~\ref{fig-evals-to-success} under the assumption that P3's computational time per evaluation is higher
-than LTGA but by no more than a constant factor.
+than LTGA but by no more than a constant factor. In aggregate P3 found the global optimum the fastest on 72 of 130 configurations,
+with LTGA, $1+(\lambda, \lambda)$, Hill Climber, hBOA outperforming P3 on 44, 18, 10, and 1 configurations, respectively.
 
@@
 15\% and 20\% respectively on the largest problem sizes. Both of these problems are perfectly separable and have relatively
-few local optima in each subproblem. HIFF and Ising Spin Glasses use crossover about 60\%. The HIFF structure can
+few local optima in each subproblem. HIFF and Ising Spin Glasses use crossover for about 60\% of evaluations. The HIFF structure can
 be perfectly learned by the linkage tree, but the higher order relationships require multiple stages of crossover to optimize.
@@
 In comparison with LTGA, the crossover success rates of P3 are lower but similar in problem ordering. This is almost certainly
-a consequence of P3's use of local search. Counter intuitively the use of hill climbing on the initial population is likely reducing
-P3's crossover success not because it reduces model quality or the donation pool, but because it is much more challenging to improve
+a consequence of P3's use of local search. Counterintuitively the use of hill climbing on the initial population is likely reducing
+P3's crossover success, not because it reduces model quality or the donation pool, but because it is much more challenging to improve
 locally optimal solutions than randomly generated ones. LTGA's crossover benefits from application to unoptimized solutions, which
@@
 median size across 100 runs. If a run did not store any solutions at a level it is treated as 0,
-and no point is drawn if the median run had 0 solutions stored at that level. This behavior is
-representative pyramid shape across all problem sizes tested.
+and no point is drawn if the median run had 0 solutions stored at that level. While pyramid size
+is effected by problem size, the overall shape is not. As such the behavior shown in Figure~\ref{fig-level-size}
+is representative of that for all other tested problem sizes.
 
@@
 genetics standpoint. When using a generational model \cite{lobo:2011:dynamicpop} found theoretical
-evidence that the optimal population size reduces each generation with \cite{goldman:2011:dynamic-parameters}
-providing supporting empirical data. In P3, levels of the pyramid are akin to generations as solutions
+evidence that the optimal population size reduces each generation and \cite{goldman:2011:dynamic-parameters}
+made similar empirically based conclusions. In P3, levels of the pyramid are akin to generations as solutions
 stored in each level have undergone the same number of variation and selection events,
-with higher levels representing more of each. Therefore it is possible that P3's
+with higher levels representing more of both. Therefore it is possible that P3's
 pyramid shape is related to its ability to outperform competitors using an optimal
@@
 contain a higher frequency of optimal trap settings to have achieved a fitness improvement. As
-a result the crossover success rate and modeling quality at level four is likely to be much better
+a result the crossover success rate and modeling quality at level four is likely much better
 than at level three, increasing the likelihood of solutions being improved by crossover.
@@
 successful only 0.0004\% of the time. This quickly rebounds at higher levels. On all other problems, the
-crossover success rate follows a trajectory closer expectation. Low levels are more successful than medium
+crossover success rate follows a trajectory closer to expectation. Low levels are more successful than medium
 levels as it is easier to make improvements to less optimized solutions. This trend reverses somewhat at
@@
 This pyramid is constructed iteratively, with both the number of levels and the number of solutions stored
-at each level growing as search progresses. P3 uses a model based method learns a linkage tree from gene epistasis as a method for
-performing effective crossover. Combined with a simple hill climber, P3's design contains
+at each level growing as search progresses. P3 uses a model based crossover method
+which learns a linkage tree from gene epistasis. Combined with a simple hill climber, P3's design contains
 many synergistic features.
@@
 order deception via linkage learning the iterative improvement method by which crossovers
-are made may mislead search on landscapes with higher order deception.
+are made may mislead search on landscapes with higher order deception. Even without
+these improvements our results suggest that P3 will be highly efficient at finding
+global optima on black box problems without any problem specific tuning.
 

### 2014-10-02 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Simple edits requested by Bill.

@@
 The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
-evolutionary optimization without requiring problem specific parameters.
+evolutionary optimization without requiring any user specified parameters.
 This is achieved by combining local search with advanced crossover to iteratively construct
 and exploit multiple populations simultaneously, with each population akin to a generation
-in traditional search.
+in traditional evolutionary computation.
 
 When compared with five advanced search algorithms across dozens of problem sizes for
-seven different benchmarks P3 reliably required at least a constant factor less evaluations to find the global optimum
-than all tested competitors. When compared using fixed numbers of evaluations P3
-generally finds the highest quality solutions. Some evidence also suggests this performance
-extends to considering total number seconds to find the global optimum even when using efficient fitness functions.
+seven different benchmarks, P3 reliably required at least a constant factor less evaluations to find the global optimum
+than all tested competitors. When compared using fixed numbers of evaluations, P3
+generally finds the highest quality solutions. Some evidence also suggests that P3
+is better when compared using the number seconds to find the global optimum even when using efficient fitness functions.
 We investigate the causes of this efficiency and provide some evidence that P3's ability to
@@
 This quality is made even more striking in that unlike P3, the best comparison algorithms received extensive
-problem specific tuning to achieve their performance. As a result we conclude that in contrast to
-previous parameter-less techniques P3 is more efficient than existing parameterized methods for
-black box optimization.
+problem specific tuning to achieve their performance. As a result we conclude that, in contrast to
+previous parameter-less techniques, P3 is an efficient, general, parameter-less approach to black box
+optimization which is more effective than existing state-of-the-art techniques.
 
@@
 However, the quality of solutions found, and the speed at which they are found, is strongly tied to setting these parameters
-correctly~\citep{goldberg:1991:gasize}. As such expert knowledge or exceedingly
+correctly~\citep{goldberg:1991:gasize}. As such, expert knowledge or exceedingly
 expensive parameter tuning~\citep{grefenstette:1986:optimalga} may be required to leverage
 this feature to its fullest potential. Furthermore,  parameters such as population size, mutation rate, crossover
-rate, tournament size, etc can have no clear relationship to the problem being solved, meaning even
+rate, tournament size, etc. may have no clear relationship to the problem being solved, meaning even
 domain experts may not understand how they will interact with the problem or with each other.
 
-As such there have been periodic efforts to reduce or remove the need for parameter tuning.
+There have been periodic efforts to reduce or remove the need for parameter tuning.
 \cite{Back:1992:selfadapt} introduced self-adaptive parameters, in which parameter values
-were included in each solution's genome and underwent evolution. This allowed search
+were included in each solution's genome and themselves underwent evolution. This allowed search
 to optimize some of its own parameters resulting in a reduced the need for expert tuning.
@@
 Algorithm~(hBOA) and \cite{thierens:2010:ltga}'s Linkage Tree Genetic Algorithm~(LTGA)
-both only require a single parameter: population size. \cite{posik:2011:parameterless}
+both require only a single parameter: population size. \cite{posik:2011:parameterless}
 leveraged model building to create a fully parameter-less algorithm, but it is restricted to
@@
 
-In order to fully understand the effectiveness of a black box search algorithm, it is useful
-to compare it with other similar algorithms. Therefore here we preset five advanced algorithms with
-related features to P3. The Random Restart Hill Climber was chosen as an efficient form of repeated
-local search, and because it shows the advantages of performing the more complex portions of P3.
+In order to fully understand the effectiveness of P3, we compare it with five advanced algorithms which
+have related features to P3. The Random Restart Hill Climber~\citep{goldman:2014:p3} was chosen as an efficient form of repeated
+local search. As P3 combines this hill climber with crossover,
+comparing with local search alone shows the advantages of performing the other portions of P3.
 The $1+(\lambda, \lambda)$ algorithm is the current best theory supported simple genetic algorithm
@@
 size. This makes knowing the optimal behavior of these algorithms much more tractable. All of the
-algorithms are also gene order independent, fitness scale invariant, and unbiased. This means
-for any problem the order in which problem variables appear in the genome can be changed
+algorithms are also gene order independent, fitness scale invariant, and unbiased. This means,
+for any problem, the order in which problem variables appear in the genome can be changed
 without changing the behavior of the search. The fitness can also be manipulated in any fashion
@@
   \Procedure{Hill-Climber}{}
-    \State $options \leftarrow [0 \dots N-1]$
+    \State $options \leftarrow [0 \dots N-1]$\label{fig-hc-options}
     \State $tried \leftarrow \emptyset$
@@
 As such any operation that happens only once per search can be amortized over at least $N$
-evaluations, covering the initialization of $options$.
+evaluations, covering the initialization of $options$ on Line~\ref{fig-hc-options}.
 
@@
 This is a contradiction as the loop terminates when no improving move was found. Therefore this check
-can only be called twice per evaluation.
+is called at most twice per evaluation.
 
 Due to its nature, this hill climber cannot escape basins of attraction. Once a solution is reached
-such that none of the single bit neighbors are fitness improvements, search stops. As a result to
-have any hope of solving multimodal problems this algorithm requires a restart mechanism. We have
+such that none of the single bit neighbors are fitness improvements, search stops. Thus
+this algorithm requires a restart mechanism to solve multimodal problems. We have
 chosen here to naively restart search from a random solution whenever a local optima is found. This
@@
 and evaluated, $w$ solutions are chosen at random from the population, where $w=\min\{N, \frac{\mu}{20}\}$.
-Whichever of the $w$ solutions is the most genetically similar to the offspring is compared with the offspring.
+Whichever of the $w$ solutions is the most genetically similar to the offspring compares its fitness with the offspring.
 If the fitness of the offspring is no worse, it replaces the genetically similar member of $w$, otherwise the
@@
 ability to preserve high fitness gene subsets, LTGA was able to outperform state of the art
-GAs across many benchmarks. Since its first proposal, many variants of LTGA have been
+GAs across many benchmarks. Since its introduction, many variants of LTGA have been
 proposed~\citep{goldman:2012:ltga} so for clarity we have chosen the version
@@
 Figure~\ref{fig-cluster-creation} provides the agglomerative method LTGA uses to create gene clusters.
-This algorithm will create a tree of sets, such that the leaves of the tree contain a single gene and
-nodes are the union of their children's sets. These sets are then used by crossover to specify epistatic
+This algorithm creates a tree of sets using pairwise gene entropy, such that the leaves of the tree contain a single gene and
+internal nodes are the union of their children's sets. These sets are then used by crossover to specify epistatic
 relationships which should be preserved.
@@
 by breaking the stored pattern. As such a cluster is only kept if its direct superset has a non-zero distance.
+As a final step Line~\ref{fig-cluster-creation-ordering} reorders $useful$ such that clusters appear in reversed
+order from which they were added to $useful$. Thus the most linked clusters and those containing single genes
+appear at the end of the returned list.
 
@@
 
-In order to efficiently perform clustering, a pairwise gene frequency table is constructed.
-Equation~\ref{eq-entropy} uses how frequently each of the four possible string values for each
-pair of genes occurs in the selected solutions. Extracting this information requires $O(\mu N^2)$
+In order to efficiently perform clustering, a pairwise gene frequency table is constructed from the selected solutions.
+To calculate Equation~\ref{eq-distance}, Equation~\ref{eq-entropy} is called for each
+gene ($H(c_i)$) and pair of genes ($H(c_i \cup c_j)$). Extracting this information requires $O(\mu N^2)$
 time, where $\mu$ is the population size and $N$ is the genome size. The process of converting
@@
 Each generation, each individual in the population undergoes crossover. In a single crossover event, each
-cluster of genes $C_i$ in $useful$ is applied in last merged first order as a crossover mask. A random donor $d$
+cluster of genes $C_i$ in $useful$ is applied as a crossover mask. A random donor $d$
 is chosen from the entire population (not just the model selected population), and $d$'s gene's for $C_i$ are copied
@@
 LTGA all solutions in the population are used to generate the model, not just the
-winners of a binary tournament. This selection process is not necessary as the
+winners of a binary tournament. We can do this because even the
 worst solutions in the pyramid are already high quality due to local search.
@@
 diversity, and therefore required smaller populations, but has  not been integrated into
-our chosen canonical LTGA variant.
+the canonical LTGA variant we used.
 
@@
 in a node of the tree are set to the same value, they score equal to the size of the set. In this way
-small subsets lead toward solutions to larger subsets. However a node can score if all genes are either
+small subsets lead toward solutions to larger subsets. However, a node can score if all genes are either
 all 1s or all 0s, meaning that to solve higher order subproblems it is necessary to perform crossovers
@@
 While well defined landscapes can provide specific insights into how an algorithm works,
-their static nature can create misleading results. Specifically algorithm quality might
+their static nature can be misleading. Specifically, algorithm quality might
 be very fragile such that it is only effective at searching well behaved landscapes. A more realistic
@@
 to draw more general conclusions about the algorithms effectiveness. The challenge with these
-landscapes is determining the global optimum to know if an algorithm was successful.
+landscapes is determining the global optimum to gauge if an algorithm was successful.
 
@@
 randomly chosen and as a result finding the global optimum is $NP$-Hard for $K>1$. However,
-if epistasis is instead set such that each gene depends on the $K$ directly following it
+if epistasis is set such that each gene depends on the $K$ directly following it
 in the genome the solution can be found in polynomial time~\citep{wright:2000:solvingnk}.
@@
 \cite{saul:1994:spinglass} presents a combinatorial benchmark problem derived from physics:
-Ising Spin Glasses. A spin glass is defined by interaction terms between variables with
-the genome encoding the sign for each variable and a goal to minimize the total interaction value.
-Similar to NK Landscapes the general class is $NP$-Hard to optimize, but the $2D\pm J$ subset of
+Ising Spin Glasses. A spin glass is defined by a weighted graph of interaction terms between vertices.
+Each gene assigns a value to each vertex, with the fitness calculated by Equation~\ref{eq-ising}.
+In this equation, $E$ is the set of all edges, $e_{ij}$ is the edge weight connecting vertex $i$ to vertex $j$, and
+$x_i$ and $x_j$ are the gene values for vertex $i$ and $j$. Optimal fitness is when this sum is minimized.
+\begin{equation}
+\sum_{e_{ij} \in E} x_ie_{ij}x_j
+  \label{eq-ising}
+\end{equation}
+Similar to NK Landscapes, the general class is $NP$-Hard to optimize, but the $2D\pm J$ subset of
 Ising Spin Glasses can be polynomially solved.\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}
+In this subset the graph is restricted to be a two dimensional torus, edge weights are randomly set
+to either -1 or 1, and vertex values must be -1 or 1.
 
@@
 at least one term in the clause evaluates to true. In order to make MAX-SAT instances with a known global
-optimum \cite{goldman:2014:p3} proposed constructing clauses around a fixed solution. In this way the
+optimum, \cite{goldman:2014:p3} proposed constructing clauses around a fixed solution. In this way the
 signs of the terms are set to ensure the target solution satisfies the clause. To ensure each
@@
 to determine the optimal population size for each size of each problem.
-Extended by \cite{goldman:2012:ltga} this method iteratively
+Extended by \cite{goldman:2012:ltga}, this method iteratively
 doubles the population size until some success criteria is met and then performs bisection
@@
 this trend was MAX-SAT, where Parameter-less hBOA actually outperformed hBOA. A potential explanation for
-this upset is that hBOA is using the optimal population size for the problem class, not for each instance.
+this is that hBOA is using the optimal population size for the problem class, not for each instance.
 Therefore hBOA must use a population size sufficiently large to solve the hardest instances in the class.
@@
 The largest of these used 49 bits, with P3 achieving the best results on all 92 larger
-configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the 5, 4, and 1 smallest Deceptive
-Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the 3 smallest Nearest Neighbor NK instances
+configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
+Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
 and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
-on 5 smallest Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
+on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
 
@@
 we examine this behavior in Figure~\ref{fig-fitness-over-time}. At regular intervals during optimization
-Figure~\ref{fig-fitness-over-time} shows the median of the best fitnesses found by that point of search across
+Figure~\ref{fig-fitness-over-time} shows the median of the best fitnesses found at that point of search across
 100 runs. For each problem we show the largest problem size for which we were able to successfully set a population
@@
 
-At 63\% of recording intervals hBOA has the worst fitness of any solver. Many of the exception occur when
-the total number of evaluations performed is significantly less than the genome size and hBOA reaches better fitness
-than the Hill Climber, $1+(\lambda, \lambda)$, and P3. This early in search hBOA and LTGA are both still evaluating
+At 63\% of the recording intervals hBOA has the worst fitness of any solver. Many of the exceptions occur when
+the total number of evaluations performed is significantly less than the genome size. Under these circumstances hBOA reaches better fitness
+than the Hill Climber, $1+(\lambda, \lambda)$, and P3. Early in search hBOA and LTGA are both still evaluating
 individuals in their randomly generated initial populations, while the three techniques worse than hBOA have started
@@
 local optima, but it requires more than a single generation to do so completely. Until that occurs, the model is likely behaving
-no better than random as there is little apparent linkage between bits. Once the model begins to identify individual traps crossover
+no better than random as there is little apparent linkage between bits. Once the model begins to identify individual traps, crossover
 can begin to increase the frequency of higher fitness local optima. This process begins slowly as low fitness local optima are more
@@
 Ising Spin Glasses have an overlapping structure, making them impossible to perfectly represent in a single linkage tree, but
-that overlap is relatively minor in comparison with Nearest Neighbor NK, which uses 81\%. Deceptive Step Trap spends the highest
+that overlap is relatively minor in comparison with Nearest Neighbor NK, which uses 81\% of evaluations on crossover. Deceptive Step Trap spends the highest
 percentage of evaluations on crossover, 92\%. Unlike all of the other problems, the value taken by an individual bit in Deceptive
@@
 with only 0.007\% successful on the largest tested problems. In comparison to Ising Spin Glass, NK's overlap is far more extensive.
-As a result it makes sense that individual crossovers are going to be very unlikely to be fitness improvements.
+As a result it makes sense that individual crossovers are very unlikely to be fitness improvements.
 
@@
 While the frequency of crossover changes per problem, and the likelihood of an individual crossover making a fitness improvement can be
-quite low, the results from Section~\ref{sec-optimum} and Section~\ref{sec-overtime} suggest its still critical to optimization. Without
+quite low, the results from Section~\ref{sec-optimum} and Section~\ref{sec-overtime} suggest it's still critical to optimization. Without
 crossover P3's performance would be identical to the Random Restart Hill Climber, which was unable to solve even moderately sized problems

### 2014-10-03 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Population size figures and explanation.

@@
 
-Our implementations of all of these algorithms as well as all of the raw results and processing scripts
+Our implementations of all of these algorithms as well as all of the population size information, raw results,
+and processing scripts
 are available from our website.\footnote{\url{https://github.com/brianwgoldman/FastEfficientP3}}
@@
 \label{sec-tuning}
+
+\begin{figure}[t]
+  \begin{centering}
+    \begin{subfigure}{.5\textwidth}
+      \begin{centering}
+        \includegraphicsfit{pop-hboa}
+      \end{centering}
+      \caption{Tuned Population Sizes for hBOA}
+      \label{fig-small-pop-dst}
+    \end{subfigure}%
+    \begin{subfigure}{.5\textwidth}
+      \begin{centering}
+        \includegraphicsfit{pop-ltga}
+      \end{centering}
+      \caption{Tuned Population Sizes for LTGA}
+      \label{fig-small-pop-nk}
+    \end{subfigure}
+  \end{centering}
+  \caption{Optimal population sizes found using bisection on each size
+           for each problem.}
+  \label{fig-pop-sizes}
+\end{figure}
+
 While four of the six algorithms in our experiments do not require any user specified parameters,
@@
 
+Figure~\ref{fig-pop-sizes} shows the results from performing bisection. In general hBOA
+required population sizes which were at least an order of magnitude larger than LTGA. Due to
+runtime and memory usage finding the optimal value for hBOA was much less tractable than LTGA
+for moderate to large problem sizes. LTGA's population size also grew significantly slower
+than hBOA's as problem size increased, especially on the two Trap problems and Rastrigin.
+Both algorithms were very ineffective on MAX-SAT, neither able to tune to problems sizes over 60 bits.
+This is likely due to the fact that some randomly generated MAX-SAT landscapes are very flat
+and highly deceptive.
 \section{Finding the Global Optimum}

### 2014-10-04 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Removed commented out table.

@@
 
-\begin{comment}
-\begin{table}
-	\centering
-	\begin{tabular}{|c|r|r|r|r|r|r|r|r|}
-	  \cline{2-9}
-	  \multicolumn{1}{c|}{} & \multicolumn{2}{L|}{Deceptive Trap} & \multicolumn{2}{L|}{Deceptive Step Trap}
-	                        & \multicolumn{2}{L|}{HIFF} & \multicolumn{2}{L|}{Rastrigin} \\ \cline{2-9}
-	  \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{a} & \multicolumn{1}{c|}{b}
-	                        & \multicolumn{1}{c|}{a} & \multicolumn{1}{c|}{b}
-	                        & \multicolumn{1}{c|}{a} & \multicolumn{1}{c|}{b}
-	                        & \multicolumn{1}{c|}{a} & \multicolumn{1}{c|}{b} \\ \hline
-	                          % DT                    % DST                    % HIFF                  % Rast
-	  {Hill Climber}          &$<$0.00 &        11.10 &$<$0.00 &        10.60 &$<$0.00 &         9.45 &$<$0.00 &         5.50 \\ \hline
-	  {$1+(\lambda,\lambda)$} &   0.02 &         3.66 &$<$0.00 &         6.69 &$<$0.00 &         6.55 &$<$0.00 &         4.50 \\ \hline
-	  {hBOA}                  &  18.16 &         2.28 &  40.38 &         2.07 &   4.60 &         1.82 &   3.50 &         2.29 \\ \hline
-	  {Parameter-less hBOA}   &  16.92 &         2.72 &  28.71 &         2.59 &   4.94 &         2.25 &   3.55 &         2.61 \\ \hline
-	  {LTGA}                  & 324.88 & \textbf{1.28}& 165.27 &         1.72 &   7.02 &         1.66 &  21.05 & \textbf{1.43}\\ \hline
-	  {P3}                    &  68.80 &         1.38 & 354.69 & \textbf{1.56}&   4.06 & \textbf{1.59}&   5.29 &         1.59 \\ \hline
-  \end{tabular}
-	\caption{Results of fitting $y=ax^b$ model where $y$ is evaluations to success
-	         and $x$ is problem size. Bold entries are the statistically lowest growth rates.}
-	\label{table-bigo}
-\end{table}
-
-TODO Table listing empirical $O(N)$ times for each technique on each problem. Include
-statistical comparison of log-log regression lines.
-
-\end{comment}
-
 \section{Fitness Over Time}

### 2014-10-06 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Sections 6-9 reorganized.

@@
 
+\subsection{Quantitative Comparison}
+Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
+The largest of these used 49 bits, with P3 achieving the best results on all 92 larger
+configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
+Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
+and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
+on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
+
+\subsection{Local Search}
 The Random Restart Hill Climber and $1+(\lambda, \lambda)$ are both relatively effective on small problem
-sizes, but have difficulty scaling to larger problems.  Only on MAX-SAT are these optimizers competitive
-at larger tested problem sizes. However, this may just be an artifact as the largest tested MAX-SAT
-was an order of magnitude smaller than the largest size tested for most other problems. On the Deceptive
-Trap and Deceptive Step Trap problem $1+(\lambda, \lambda)$ has some other interesting results. For
-problems with 4 or less traps it performs significantly better than any other algorithm. This is likely
+sizes. This is especially true for the three randomly generated problem problem classes. These problems
+may contain relatively few local optima or just be exceptionally difficult for the model based algorithms.
+For Deceptive Trap and Deceptive Step Trap using 4 or less traps, $1+(\lambda, \lambda)$ performs significantly
+better than any other algorithm. This is likely
 because $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
-flipping entire traps. For larger problems it becomes increasingly unlikely that mutation and crossover can
-create enough changes to an individual trap to overcome deception without negatively changing others. This
-behavior also causes high variance in success rate, as evident by the occasional successes on large problems.
+flipping entire traps.
 This ability to successfully integrate modifications of more than a single bit generally appears to lead $1+(\lambda, \lambda)$
@@
 
-As expected, Parameter-less hBOA was generally less efficient than hBOA using an optimal population size. The slopes
-of the two lines are relatively similar across all problems, meeting with the theoretical expectation that
-Parameter-less hBOA is no more than a logarithmic factor worse than hBOA.  The only problem which reversed
-this trend was MAX-SAT, where Parameter-less hBOA actually outperformed hBOA. A potential explanation for
-this is that hBOA is using the optimal population size for the problem class, not for each instance.
-Therefore hBOA must use a population size sufficiently large to solve the hardest instances in the class.
-Parameter-less hBOA on the other hand is able to solve easy instances using much smaller population sizes.
-The net effect is therefore that by being more adaptable Parameter-less hBOA is able to achieve better median
-results.
+On larger problem sizes, the ability for local search to reach the global optimum quickly diminishes.
+Only on MAX-SAT are these optimizers competitive
+at larger tested problem sizes. However, this may just be an artifact as the largest tested MAX-SAT
+was an order of magnitude smaller than the largest size tested for most other problems. As the problem
+size increases the number of local optima increases exponentially, which explains why Random Restart Hill
+Climbing was unable to scale.
+For larger problems it becomes increasingly unlikely that $1+(\lambda, \lambda)$'s mutation and crossover can
+create enough changes to the correct parts of an individual required to overcome deception without negatively changing others. This
+behavior also causes high variance in success rate, as evident by the occasional successes on large Deceptive Trap problems.
+
+By integrating local search, P3 gains most of the benefits. Unlike the other model based methods, P3 generally outperforms
+both Random Restart Hill Climber and $1+(\lambda, \lambda)$ on small problem sizes. On larger problem sizes
+P3's ability to perform targeted crossover quickly allows it to surpass $1+(\lambda, \lambda)$'s uninformed
+crossover. This targeted crossover may also explain why P3 is able to outperform $1+(\lambda, \lambda)$ on MAX-SAT.
+
+Local search may also account for some of P3's ability to surpass LTGA and hBOA on larger problem sizes.
+On Deceptive Trap P3's use of hill climbing ensures all traps are immediately optimized, allowing for perfect
+linkage detection and high quality donation. Deceptive Step Trap, which is less amenable to local search, does
+not show a similar improvement over the other model building techniques and performs only slightly better.
+Using local search on HIFF likely saves P3 a generation over LTGA and hBOA as all pairwise subproblems will already be solved,
+accounting for some of its improvement.
+
+\subsection{Model Building}
+Only techniques which explicitly built models of the search landscape were able to solve the largest problem
+instances. On single instance problems LTGA was generally more effective than hBOA, with hBOA outperforming
+LTGA on Nearest Neighbor NK and Ising Spin Glasses. This may be caused by the differences in modeling method:
+the single instance problems can all be perfectly represented with a linkage tree while none of the randomly generated
+problem classes can.
 
@@
 
-LTGA, as the current state-of-the-art, was generally at least as fast to find the global optimum as hBOA.
-However on Nearest Neighbor NK and Ising Spin Glasses LTGA appears
-to perform worse than hBOA.  This may be due to LTGA's inability to represent the overlapping nature of these problems.
-On MAX-SAT LTGA only outperforms hBOA and performs worse than the other 4 techniques.
-Again, the likely cause to this is LTGA's requirement to have a
-population size sufficiently large to solve the hardest problem in each class, reducing its effectiveness
-on instances of median difficulty.
-
+Similar to hBOA and LTGA, P3's model building allows it to scale to large problem sizes. As P3 utilizes the same
+model building method as LTGA, we should expect the effects of model building to be similar between those two algorithms.
+Indeed on the four problems where a linkage tree can perfectly represent gene epistasis, P3 has very similar
+scalability (slope) as LTGA, in most cases performing better only by a constant factor.
+
+\subsection{Population Sizing}
 \begin{figure}
@@
 
-Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
-The largest of these used 49 bits, with P3 achieving the best results on all 92 larger
-configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
-Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
-and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
-on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
-
-On Deceptive Trap P3 significantly outperforms both hill climbing
-and LTGA by combining the best features of each, using hill climbing to optimize each trap and crossover
-to perfectly mix traps. Deceptive Step Trap is less amenable to local search with P3's quality effectively
-tied with both LTGA and hBOA. The slight improvement here may be due to P3's ability to solve some runs
-using smaller population sizes than LTGA and hBOA due to lucky initialization. Using local search on HIFF
-likely saves P3 a generation over LTGA and hBOA as all pairwise subproblems will already be solved. Combined
-with P3's ability to add diversity only as needed this likely explains why P3 is able to solve HIFF a constant
-multiple faster than either population based technique. P3 likely solves its median run of Nearest Neighbor
-NK and Ising Spin Glasses faster than both hBOA and LTGA again by using population sizes fit to each instance
-and not fit to the entire problem class. On MAX-SAT, where the most effective solvers are Hill Climber and
-$1+(\lambda, \lambda)$, P3 outperforms both likely due to its use of local search with a better ability to
-overcome deception.
+As expected, Parameter-less hBOA was generally less efficient than hBOA using an optimal population size. The slopes
+of the two lines are relatively similar across all problems, meeting with the theoretical expectation that
+Parameter-less hBOA is no more than a logarithmic factor worse than hBOA.  The only problem which reversed
+this trend was MAX-SAT, where Parameter-less hBOA actually outperformed hBOA and LTGA. A potential explanation for
+this is that hBOA and LTGA are using the optimal population size for the entire problem class, not for each instance.
+Therefore both must use a population size sufficiently large to solve the hardest instances in the class.
+Parameter-less hBOA, on the other hand, is able to solve easy instances using much smaller population sizes.
+The net effect is therefore that by being more adaptable Parameter-less hBOA is able to achieve better results
+on instances of median difficulty.
+
+Scaling population size to the problem instance instead of the problem class may explain some of why P3 outperforms
+LTGA and hBOA on large problem sizes. P3 solves its median run of Nearest Neighbor
+NK and Ising Spin Glasses faster than both hBOA and LTGA.
 
@@
 end as P3's upper quartile is only worse than LTGA's upper quartile on Deceptive Step Trap and Nearest
-Neighbor NK. A possible explanation for this is that as P3 has not pre-learned the amount of diversity required
-to solve all instances does not take as risky of behavior as LTGA.
+Neighbor NK. On these problems the difference in upper quartile is very small, suggesting that even in the worst
+case P3 is no worse than LTGA.
 
@@
 % 181 total
-% 126 has P3 best
+% 121 has P3 best
+%   9 hc
+%  50 lambdalambda
+%  20 hboa
+%  18 phboa
 %  27 LTGA
-% 20 hboa
-% 50 lambdalambda
-% hc 9
-% phboa 18
-
-At 63\% of the recording intervals hBOA has the worst fitness of any solver. Many of the exceptions occur when
-the total number of evaluations performed is significantly less than the genome size. Under these circumstances hBOA reaches better fitness
-than the Hill Climber, $1+(\lambda, \lambda)$, and P3. Early in search hBOA and LTGA are both still evaluating
-individuals in their randomly generated initial populations, while the three techniques worse than hBOA have started
-from a single solution and are attempting to optimize. Of the methods, hBOA uses by far the largest population sizes,
-about 55 times larger than LTGA on average for problems shown. As such it makes sense that hBOA would take a long time to reach high
-quality solutions.
-
-Parameter-less hBOA reported equal or higher fitness than hBOA when using the same number of total evaluations in 68\% of intervals. This makes
-sense as Parameter-less hBOA's use of small populations allows it to begin to optimize long before hBOA has even evaluated
-its entire initial population. However, this trend is reversed after a sufficient number of evaluations, as is apparent
-on Deceptive Step Trap and Ising Spin Glasses. This makes sense as hBOA's population is set to quickly find the global
-optimum.
-
+
+\subsection{Quantitative Comparison}
+Of 181 sample points, P3 had the highest median fitness in 121. In pairwise competition, $1+(\lambda, \lambda)$
+was the most likely to outperform P3, doing so on 50 sample points. LTGA, hBOA, and Parameter-less hBOA
+were the next best, outperforming P3 on 27, 20, and 18 sample points. Random Restart Hill Climbing almost
+never outperformed P3, doing so only 9 times.
+
+\subsection{Local Search}
+Perhaps the most striking result is the quality of $1+(\lambda, \lambda)$. Until quite far into search this
+method performs better than both LTGA and hBOA. Given sufficient evaluations $1+(\lambda, \lambda)$ outperforms
+Random Restart Hill Climbing on all 7 problems. For brief periods in the middle of search it performs the
+best of all techniques on both Trap functions, HIFF, Ising Spin Glass, and MAX-SAT problems. A likely
+explanation is $1+(\lambda, \lambda)$'s ability to efficiently incorporate gene modifications of larger
+than one bit. This allows it to overcome the deception and plateaus in Deceptive Trap and Deceptive Step Trap, solve medium sized subproblems
+in HIFF, flip the signs on multiple adjacent bits in Ising Spin Glass, and cross plateaus in MAX-SAT. However,
+this method is not enough to quickly reach the global optima in many of these problems which causes it to eventually
+be overtaken by the model building techniques.
+
+The integration of hill climbing into P3 makes it strictly better than using hill climbing alone.
 Early in optimization P3 and the Random Restart Hill Climber have effectively identical quality. This is because
@@
 
-Perhaps the most striking result is the quality of $1+(\lambda, \lambda)$. Until quite far into search this
-method performs better than both LTGA and hBOA. Given sufficient evaluations $1+(\lambda, \lambda)$ outperforms
-Random Restart Hill Climbing on all 7 problems. For brief periods in the middle of search it performs the
-best of all techniques on the Deceptive Step Trap, HIFF, Ising Spin Glass, and MAX-SAT problems. A likely
-explanation is $1+(\lambda, \lambda)$'s ability to efficiently incorporate gene modifications of larger
-than one bit. This allows it to overcome the plateaus in Deceptive Step Trap, solve medium sized subproblems
-in HIFF, flip the signs on multiple adjacent bits in Ising Spin Glass, and cross plateaus in MAX-SAT. However,
-this method is not enough to quickly reach the global optima in many of these problems which causes it to eventually
-be overtaken by the model building techniques.
-
-For all problems, LTGA lags behind P3 in fitness improvement, but has a similar overall shape. On every problem except
-Rastrigin, LTGA has five distinct periods: fitness plateau, near instantaneous improvement, fitness plateau,
+\subsection{Model Building}
+Both hBOA and LTGA are marked by periods of little improvement and rapid improvement. Initially this is due to
+population size effects (Section~\ref{sec-pop-sizing}). In hBOA this is taken to the extreme, with all fitness
+improvement made at the very end of search. In both cases this is caused by model building. Before the model
+is accurate little improvement is made. Once it is accurate, fitness improves dramatically.
+
+On every problem LTGA has five distinct periods: fitness plateau, near instantaneous improvement, fitness plateau,
 and improvement to global optimum. The early period corresponds with
@@
 
+\subsection{Population Sizing}
+\label{sec-pop-sizing}
+
+At 58\% of the recording intervals hBOA has the worst fitness of any solver. Many of the exceptions occur when
+the total number of evaluations performed is significantly less than the genome size. Under these circumstances hBOA reaches better fitness
+than the Random Restart Hill Climber, $1+(\lambda, \lambda)$, and P3. Early in search hBOA and LTGA are both still evaluating
+individuals in their randomly generated initial populations, while the three techniques worse than hBOA have started
+from a single solution and are attempting to optimize. Of the methods, hBOA uses by far the largest population sizes,
+about 64 times larger than LTGA on average for problems shown. As such it makes sense that hBOA would take a long time to reach high
+quality solutions.
+
+Parameter-less hBOA reported equal or higher fitness than hBOA when using the same number of total evaluations in 62\% of intervals. This makes
+sense as Parameter-less hBOA's use of small populations allows it to begin to optimize long before hBOA has even evaluated
+its entire initial population. However, this trend is reversed after a sufficient number of evaluations, as is apparent
+on Deceptive Step Trap and Ising Spin Glasses. This makes sense as hBOA's population is set to quickly find the global
+optimum.
+
 \begin{figure}[t]
@@
 
-It is important to note that while P3 achieved the best fitness across 70\% of time points, hBOA and LTGA were tuned specifically
+It is important to note that hBOA and LTGA were tuned specifically
 to find the global optimum and are not necessarily using the optimal population size for finding intermediate fitnesses. Consider
@@
 occurring earlier, as the first generation is completed much more quickly. On Deceptive Step Trap this means that
-LTGA leaps ahead of P3 due to its ability to overcome the two bit fitness plateaus. Across all tested problems using the smaller
+LTGA leaps ahead of P3 due to its ability to overcome the two bit fitness plateaus.
+%
+Across all tested problems using the smaller
 population size reduced the quality of LTGA's first fitness plateau, likely due to missing required diversity to reach higher quality.
@@
 \section{Computational Expenses}
+While it is common in evolutionary computation to assume the evaluation function will dominate algorithm
+complexity, in some domains this will not be true. Model based methods are especially likely to violate
+this norm. Therefore in order to assess P3's quality in solving problems with efficient fitness functions,
+we provide data on both its algorithmic complexity and wall clock time.
+
+\subsection{Operation Counting}
 \begin{figure}[t]
@@
 
+\subsection{Wall Clock Performance}
 \begin{figure}
@@
 
+\subsection{Crossover}
 P3 performs optimization using a mixture of hill climbing and crossover. Exactly how much time the algorithm
@@
 
+\subsection{Pyramid}
 \begin{figure}[t]

### 2014-10-06 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Minor edits.

@@
 may contain relatively few local optima or just be exceptionally difficult for the model based algorithms.
-For Deceptive Trap and Deceptive Step Trap using 4 or less traps, $1+(\lambda, \lambda)$ performs significantly
+On Deceptive Trap and Deceptive Step Trap using 4 or less traps, $1+(\lambda, \lambda)$ performs significantly
 better than any other algorithm. This is likely
@@
 Climbing was unable to scale.
-For larger problems it becomes increasingly unlikely that $1+(\lambda, \lambda)$'s mutation and crossover can
-create enough changes to the correct parts of an individual required to overcome deception without negatively changing others. This
-behavior also causes high variance in success rate, as evident by the occasional successes on large Deceptive Trap problems.
-
-By integrating local search, P3 gains most of the benefits. Unlike the other model based methods, P3 generally outperforms
+For larger problems it also becomes increasingly unlikely for $1+(\lambda, \lambda)$ to make the right combination
+of changes required to reach the global optimum. This
+behavior causes high variance in success rate, as evident by the occasional successes on large Deceptive Trap problems.
+
+P3's integration of local search gives it similar benefits. Unlike the other model based methods, P3 generally outperforms
 both Random Restart Hill Climber and $1+(\lambda, \lambda)$ on small problem sizes. On larger problem sizes
@@
 \subsection{Model Building}
-Only techniques which explicitly built models of the search landscape were able to solve the largest problem
+Only techniques which explicitly built models of gene epistasis were able to solve the largest problem
 instances. On single instance problems LTGA was generally more effective than hBOA, with hBOA outperforming
 LTGA on Nearest Neighbor NK and Ising Spin Glasses. This may be caused by the differences in modeling method:
-the single instance problems can all be perfectly represented with a linkage tree while none of the randomly generated
-problem classes can.
+unlike the single instance problems, gene epistasis in the randomly generated
+problem classes cannot be be perfectly represented with a linkage tree.
 
@@
 Scaling population size to the problem instance instead of the problem class may explain some of why P3 outperforms
-LTGA and hBOA on large problem sizes. P3 solves its median run of Nearest Neighbor
-NK and Ising Spin Glasses faster than both hBOA and LTGA.
-
-To more fully examine the hypothesis that some of P3's success is do to its ability to scale to differences in problem difficulty,
+LTGA and hBOA on large problem sizes.
+To more fully examine this hypothesis,
 Figure~\ref{fig-evals-to-success-range} shows the upper and lower quartiles for P3 and LTGA. On each problem
@@
 Figure~\ref{fig-fitness-over-time} shows the median of the best fitnesses found at that point of search across
-100 runs. For each problem we show the largest problem size for which we were able to successfully set a population
-size for hBOA, but the trends shown are representative of all larger problem sizes. The maximum reporting interval
+100 runs. For each problem we show the largest problem size for which we were able to successfully gather results
+for all six algorithms, but the trends shown are representative of all larger problem sizes. The maximum reporting interval
 is set to include the slowest P3 run to reach the global optimum.
@@
 Random Restart Hill Climbing on all 7 problems. For brief periods in the middle of search it performs the
-best of all techniques on both Trap functions, HIFF, Ising Spin Glass, and MAX-SAT problems. A likely
+best of all techniques on Deceptive Trap, Deceptive Step Trap, HIFF, Ising Spin Glass, and MAX-SAT problems. A likely
 explanation is $1+(\lambda, \lambda)$'s ability to efficiently incorporate gene modifications of larger
@@
 no better than random as there is little apparent linkage between bits. Once the model begins to identify individual traps, crossover
-can begin to increase the frequency of higher fitness local optima. This process begins slowly as low fitness local optima are more
+can begin to increase the frequency of higher fitness trap settings. This process begins slowly as low fitness local optima are more
 likely to be chosen for donation than high fitness local optima due to their frequency in the population. The process becomes
@@
 nothing more than bringing every trap to one of the two local optima, with the majority of traps set suboptimally. $1+(\lambda, \lambda)$
-is slightly less efficient about performing single bit improvements, but is the next quickest to reach this plateau. While the hill climber
-remains stuck, P3's use of crossover slowly begins to combine optimal trap values found by multiple independent restarts. Unlike LTGA which
+is slightly less efficient about performing single bit improvements, but is the next quickest to reach this plateau.
+While the Random Restart Hill Climber
+remains stuck, P3's use of crossover slowly begins to combine optimal trap values found by multiple independent restarts. Unlike LTGA, which
 performs this improvement all at once, P3 integrates all optimal trap values immediately, waiting only on random restarts to find
@@
 the best comparison techniques. All of these achievements are made without any problem specific parameter
-tuning, making P3 easier to apply to new domains that its two closets competitors in quality.
+tuning, making P3 easier to apply to new domains than its two closets competitors in quality.
 
@@
 the pyramid model even greater flexibility. Similarly, while P3 is able to overcome low
-order deception via linkage learning the iterative improvement method by which crossovers
+order deception via linkage learning, the iterative improvement method by which crossovers
 are made may mislead search on landscapes with higher order deception. Even without

### 2014-10-06 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Cumulative wall clock time.

@@
 
-To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}. Due to the scale
-of our experiments, diverse hardware potentially running multiple processes in parallel were used
-in order to collect our results. As such these timings should be considered very skeptically as a number
+To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}. In total these runs
+took over 6 computing years to complete. We were therefore forced to use
+diverse hardware potentially running multiple processes in parallel.
+As such these timings should be considered very skeptically as a number
 of sources may contribute to biased error in the results. Under those caveats, the results do suggest

### 2014-10-06 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
New Abstract.

@@
 evolutionary optimization without requiring any user specified parameters.
+P3's primary innovation is to replace the generational model with the pyramid model,
+in which multiple populations are iteratively created and expanded. In combination
+with local search and advanced crossover P3 scales to problem difficulty, exploiting
+previously learned information before adding more diversity.
+
+Across seven problems, each tested using on average 18 problem sizes, P3 outperformed
+all five advanced comparison algorithms. This improvement includes requiring less evaluations
+and less wall clock seconds to find the global optimum and better fitness when using
+the same number of evaluations. From algorithm analysis and comparison we find P3's
+effectiveness is likely due to its ability to properly maintain, add, and exploit diversity.
+
+Unlike the best comparison algorithms, P3 was able to achieve this quality without any
+problem specific tuning. Thus, unlike previous parameter-less methods, P3 does not
+sacrifice quality for applicability. Therefore we conclude that
+P3 is an efficient, general, parameter-less approach to black box
+optimization which is more effective than existing state-of-the-art techniques.
+
+\begin{comment}
+The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
+evolutionary optimization without requiring any user specified parameters.
 This is achieved by combining local search with advanced crossover to iteratively construct
@@
 optimization which is more effective than existing state-of-the-art techniques.
-
+\end{comment}
 \end{abstract}
@@
 detects and exploits problem epistasis by examining pairwise gene entropy. Due to its enhanced
-ability to preserve high fitness gene subsets, LTGA was able to outperform state of the art
+ability to preserve high fitness gene subsets, LTGA was able to outperform state-of-the-art
 GAs across many benchmarks. Since its introduction, many variants of LTGA have been
@@
       \caption{Tuned Population Sizes for hBOA}
-      \label{fig-small-pop-dst}
+      \label{fig-pop-hboa}
     \end{subfigure}%
@@
       \caption{Tuned Population Sizes for LTGA}
-      \label{fig-small-pop-nk}
+      \label{fig-pop-ltga}
     \end{subfigure}
@@
 Across a large number of problems and problem sizes P3 required less evaluations to reach the global optimum
-than optimally tuned state of the art competitors. On single instance problems P3's improvement was by a
+than optimally tuned state-of-the-art competitors. On single instance problems P3's improvement was by a
 constant factor, while for the three problem classes P3's improvement increased with problem size. This

### 2014-10-08 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Edits before reordering results section (again).

@@
 evolutionary optimization without requiring any user specified parameters.
-P3's primary innovation is to replace the generational model with the pyramid model,
-in which multiple populations are iteratively created and expanded. In combination
+P3's primary innovation is to replace the generational model with a pyramid of
+multiple populations that are iteratively created and expanded. In combination
 with local search and advanced crossover P3 scales to problem difficulty, exploiting
@@
 and less wall clock seconds to find the global optimum and better fitness when using
-the same number of evaluations. From algorithm analysis and comparison we find P3's
-effectiveness is likely due to its ability to properly maintain, add, and exploit diversity.
+the same number of evaluations. Using both algorithm analysis and comparison we find P3's
+effectiveness is due to its ability to properly maintain, add, and exploit diversity.
 
@@
 
-\begin{comment}
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
-evolutionary optimization without requiring any user specified parameters.
-This is achieved by combining local search with advanced crossover to iteratively construct
-and exploit multiple populations simultaneously, with each population akin to a generation
-in traditional evolutionary computation.
-
-When compared with five advanced search algorithms across dozens of problem sizes for
-seven different benchmarks, P3 reliably required at least a constant factor less evaluations to find the global optimum
-than all tested competitors. When compared using fixed numbers of evaluations, P3
-generally finds the highest quality solutions. Some evidence also suggests that P3
-is better when compared using the number seconds to find the global optimum even when using efficient fitness functions.
-We investigate the causes of this efficiency and provide some evidence that P3's ability to
-add diversity only as needed may be contributing to its success.
-
-This quality is made even more striking in that unlike P3, the best comparison algorithms received extensive
-problem specific tuning to achieve their performance. As a result we conclude that, in contrast to
-previous parameter-less techniques, P3 is an efficient, general, parameter-less approach to black box
-optimization which is more effective than existing state-of-the-art techniques.
-\end{comment}
 \end{abstract}
@@
 However, the quality of solutions found, and the speed at which they are found, is strongly tied to setting these parameters
-correctly~\citep{goldberg:1991:gasize}. As such, expert knowledge or exceedingly
+correctly~\citep{goldberg:1991:gasize}. As such, either expert knowledge or exceedingly
 expensive parameter tuning~\citep{grefenstette:1986:optimalga} may be required to leverage
@@
 rate, tournament size, etc. may have no clear relationship to the problem being solved, meaning even
-domain experts may not understand how they will interact with the problem or with each other.
+domain experts may not understand how the parameters will interact with the problem or with each other.
 
@@
 previous parameter-less methods, P3 is actually more efficient than current state-of-the-art
-parameterized search algorithms. In this work we shall extend these results to cover more
-comparison algorithms, compare both efficiency in reaching the global optimum and intermediate
-fitnesses, algorithm complexity analysis, and provide more in depth analysis of P3 itself.
-\begin{comment}
-Section~\ref{sec-optimizers}
-explains how each of these algorithms, including P3, perform search. Section~\ref{sec-problems}
-provides a description of each test problem. As hBOA and LTGA require a population size parameter
-Section~\ref{sec-tuning} provides our methodology to ensure each is optimally tuned to each problem.
-\end{comment}
+parameterized search algorithms. In this work we shall: extend these results to cover more
+comparison algorithms; compare both efficiency in reaching the global optimum and intermediate
+fitnesses; algorithm complexity analysis; and provide more in depth analysis of P3 itself.
 
@@
 local search. As P3 combines this hill climber with crossover,
-comparing with local search alone shows the advantages of performing the other portions of P3.
+comparing with local search alone shows the advantages of P3's overall approach.
 The $1+(\lambda, \lambda)$ algorithm is the current best theory supported simple genetic algorithm
@@
 and given in Figure~\ref{fig-hc}. This algorithm works by flipping each bit in a random
-order, keeping modifications when fitness is improved, until doing so cannot result in
+order, keeping modifications when fitness is improved, until single bit flips cannot result in
 further fitness improvements.
@@
       \ForAll{$index \in shuffled(options)$}
-        \If{$index \notin tried$}
+        \If{$index \notin tried$}\label{fig-hc-tried}
           \State Flip bit $index$ in solution
@@
 evaluations, covering the initialization of $options$ on Line~\ref{fig-hc-options}.
-
-With the use of proper hashing data structures, determining if $index \notin tried$ requires $O(1)$
-time. While this operation may be called without a directly related evaluation, amortized cost
-provides an upper bound. Consider that in order to add an $index$ into $tried$ an evaluation must
-be performed. Lets say this evaluation pays for two checks of that $index$, one of which was used
-during this loop. If no fitness improving move is found, the next time through the loop the second check
-is used. A third check is only necessary if no fitness improving move was found for any index.
-This is a contradiction as the loop terminates when no improving move was found. Therefore this check
-is called at most twice per evaluation.
+Line~\ref{fig-hc-tried}, which prevents wasted evaluations, can be called at most twice
+per evaluation: once to add $index$ into $tried$ and once to prevent $index$ from being unnecessarily
+evaluated again. The only way three or more calls could happen is if
+no fitness improvement was made for the entire previous iteration, which contradicts the loop invariant.
 
@@
 epistatic relationships between genes which is then used to stochastically generate new solutions.
-Each generation a binary tournament with replacement is used to selection $\mu$ solutions from
+Each generation a binary tournament with replacement is used to select $\mu$ solutions from
 the population. These solutions are then used to build the model, which in turn is used to generate $\mu$ new
@@
 leaves of each tree give the probability that $x_i$ should be set to each possible value.
-Therefore in binary alphabets binary trees with a single probability $p$ stored at each leaf
-can be used, where $p$ is the probability $x_i$ should be set to one.
 
@@
 separate the solutions such that all solutions with $x_i = 0$ move to one leaf while solutions with
-$x_j = 1$ move to the other. Whichever split maximizes Equation~\ref{eq-hboa} is then kept.
-This process continues until none of the splits result in a higher quality model than the current one. To favor compact
-models, Equation~\ref{eq-hboa} is scaled by Equation~\ref{eq-hboa-stop}, which provides increased cost
-for more total leaves.
-
-\begin{equation}
-  BDe(B) = \prod_{i=0}^{N}\prod_{l\in L_i} \frac{\Gamma(m'_i(l))}{\Gamma(m_i(l) + m'_i(l))}
-  \prod_{x_i}\frac{\Gamma(m_i(x_i, l) + m'_i(x_i,l))}{\Gamma(m'_i(x_i,l))}
-  \label{eq-hboa}
-\end{equation}
-
-\begin{equation}
-  p(B) = c2^{-0.5(\sum_i|L_i|)log_2\mu}
-  \label{eq-hboa-stop}
-\end{equation}
-
-While Equation~\ref{eq-hboa} is very complex and almost always results in near infinitesimal results, many
-of the terms can be canceled out as it is only important to know the relative quality between two models:
-$p(B)BDe(B) < p(B')BDe(B')$.
-The outermost product of Equation~\ref{eq-hboa} iterates over all trees in the forest. However, each split
-can modify only one of the trees and therefore the contribution of all others can be canceled. The middle
-product is across all leaves in the tree. Again since only one leaf can be changed, all other terms can
-be canceled. By convention hBOA uses uninformed Bayesian priors of $m'_i(l)= 2$ and $m'_i(x_i, l)=1$ for
-binary alphabets. As $\Gamma(a) = (a-1)!$ this means the top term in the middle product and the bottom
-term in the third product reduce to 1. The only remaining terms are then $m_i(l)$ and $m_i(x_i, l)$ which
-represent the number of solutions which reached leaf $l$ and the number of solutions which reached leaf $l$
-with a specific value for $x_i$, respectively.
-
-Equation~\ref{eq-hboa-stop} can also be simplified when doing comparisons. If model $B'$ has exactly one more
-leaf than model $B$ than the ratio $\frac{p(B)}{p(B')}$ simplifies to $2^{0.5 log_2\mu}$ regardless of
-total model size.
-
-All combined, this means the expression $p(B)BDe(B) < p(B')BDe(B')$ can be calculated using Equation~\ref{eq-hboa-final},
-where $B'$ is different from $B$ by exactly 1 split, such that $l$ was split to create $l'$ and $l''$. Only splits that
-satisfy this inequality are considered good enough to keep. Furthermore, whichever split maximizes the right side of the
-Equation~\ref{eq-hboa-final} is chosen each iteration. Note that these factorials can be exceedingly large and
-therefore it is imperative that implementations avoid rounding errors and overflows.
+$x_j = 1$ move to the other.
+
+This goal is formalized using model scoring from Bayesian statistics. In its raw form this
+almost always creates near infinitesimal results, calculating fractions which include factorials of $\mu$
+and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{sec-appendix} we
+derived the simplified Equation~\ref{eq-hboa-final}. If no proposed split satisfies the inequality, iteration stops.
+If multiple splits do, whichever maximizes the right side is chosen.
 
@@
 To perform replacement, hBOA uses restricted tournament replacement. After each new solution is generated
-and evaluated, $w$ solutions are chosen at random from the population, where $w=\min\{N, \frac{\mu}{20}\}$.
-Whichever of the $w$ solutions is the most genetically similar to the offspring compares its fitness with the offspring.
-If the fitness of the offspring is no worse, it replaces the genetically similar member of $w$, otherwise the
+and evaluated, a set of $w$ solutions are chosen at random from the population, where $w=\min\{N, \frac{\mu}{20}\}$.
+From this set the solution which is most genetically similar to the offspring is chosen. If the offspring
+is at least as fit as the chosen solution, it replaces the chosen solution in the population. Otherwise the
 offspring is discarded. This method is designed to preserve genetic diversity as only genetically similar
@@
 proposed~\citep{goldman:2012:ltga} so for clarity we have chosen the version
-presented by \cite{thierens:2013:ltgahiff} as our model. This variant was chosen as it is
-modern and appears to approach the consensus in the literature.
+presented by \cite{thierens:2013:ltgahiff} as our model.
 
@@
 
-To accommodate P3's unique population structure, some of LTGA's clustering procedures were adapted. In LTGA,
-clusters are identified at the start of each generation, with those clusters used
-to create all solutions produced in that generation. As P3 does not perform serial
-generations, this process had to be modified. Instead, P3 updates the model each
+To accommodate P3's unique population structure, some of LTGA's clustering procedures were modified. In LTGA,
+clusters are identified at the start of each generation and are used
+to create all offspring in that generation. As P3 does not perform serial
+generations, P3 instead updates the model each
 time a solution is added to a population. Furthermore, unlike
@@
 
-Beyond the changes in population structuring, P3 also has slight modifications to
+Beyond the changes in population structuring, P3 modifies
 LTGA's version of \Call{Cluster-Creation}{} and \Call{Cluster-Usage}{}.
@@
 \cite{goldman:2012:ltga} found that this alternative was better at preserving
-diversity, and therefore required smaller populations, but has  not been integrated into
-the canonical LTGA variant we used.
+diversity, and therefore required smaller populations, but has not yet been integrated into
+canonical LTGA.
 
@@
 with random solutions. However, this method resulted in increased parameterization as not only was
-a population size required, parameters for how frequently generations advanced between levels and
+a population size required, but also parameters for how frequently generations advanced between levels and
 how many total levels to have were added. \cite{harik:1999:parameterlessga} used multiple independent
@@
 Each subproblem is scored using Equation~\ref{eq-deceptive-trap}, where $t$ is the number of
-bits in the trap set to 1. To global optimum in each trap is a string of all 1s, while all
-other solutions lead to a local optima of all 0s. As a result this problem tests an algorithm's
+bits in the trap set to 1. The global optimum in each trap is a string of all 1s, while all
+other solutions lead to a local optima of all 0s. This problem tests an algorithm's
 ability to overcome $k$ sized deception and is commonly used to determine how effective crossover
@@
 optima (all 1s or all 0s), and then linkage learning can have perfect knowledge of gene interactions.
-In order to make the problem more difficult yet still deceptive, the authors proposed the Deceptive
+In order to make the problem more difficult, the authors proposed the Deceptive
 Step Trap problem, given in Equation~\ref{eq-deceptive-step-trap}. This function modifies the results
@@
 signs of the terms are set to ensure the target solution satisfies the clause. To ensure each
-problem is challenging chose a clause to variable ratio of $4.27$.
+problem is challenging we chose a clause to variable ratio of $4.27$.
 
@@
 hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
-handicapped, we have extensively tuned each using the bisection method \citep{sastry:2001:bisection}
-to determine the optimal population size for each size of each problem.
+handicapped, we extensively tuned each using the bisection method \citep{sastry:2001:bisection}
+to determine the optimal population size for each problem size.
 Extended by \cite{goldman:2012:ltga}, this method iteratively
 doubles the population size until some success criteria is met and then performs bisection
-between the lowest successful and highest unsuccessful sizes. Thus the minimum population size
+between the lowest successful and highest unsuccessful sizes. In this way the minimum population size
 which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
@@
 required population sizes which were at least an order of magnitude larger than LTGA. Due to
-runtime and memory usage finding the optimal value for hBOA was much less tractable than LTGA
+runtime and memory overhead, finding the optimal value for hBOA was much less tractable than LTGA
 for moderate to large problem sizes. LTGA's population size also grew significantly slower
@@
 This is likely due to the fact that some randomly generated MAX-SAT landscapes are very flat
-and highly deceptive.
+and highly deceptive~\cite{rana:1998:gamaxsat}.
 \section{Finding the Global Optimum}
@@
 
-The goal of black box optimization is to quickly and reliably obtain high quality solutions. In this
-section we will compare how many evaluations each algorithm requires to find the global optimum on
-each problem.
-
 Figure~\ref{fig-evals-to-success} shows the median number of evaluations required by each of the six
@@
 no point is shown. The maximum problem size used for each problem was set to be the largest size we could
-feasibly determine an optimal population size for LTGA. For many larger problems results are not shown
+feasibly determine LTGA's optimal population size. For many larger problems results are not shown
 for hBOA due to the extreme computational cost required to optimally set the population size.
@@
 Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
-The largest of these used 49 bits, with P3 achieving the best results on all 92 larger
+The largest problem size for any problem where P3 was not the most efficient used 49 bits, with P3 achieving the best results on all 92 larger
 configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
@@
 
-As expected, Parameter-less hBOA was generally less efficient than hBOA using an optimal population size. The slopes
-of the two lines are relatively similar across all problems, meeting with the theoretical expectation that
+As expected, Parameter-less hBOA was generally less efficient than optimally tuned hBOA. The slopes
+of the two lines are relatively similar across all problems, affirming the expectation that
 Parameter-less hBOA is no more than a logarithmic factor worse than hBOA.  The only problem which reversed
@@
 
+\appendix
+\section{hBOA formula simplification}
+\label{sec-appendix}
+To measure the quality of a decision forest, hBOA applies Equation~\ref{eq-hboa}.
+To favor compact
+models, Equation~\ref{eq-hboa} is scaled by Equation~\ref{eq-hboa-stop}, which provides increased cost
+for more total leaves.
+This quality is used for two purposes: comparison of potential changes from the existing model
+and comparison of that change with the existing model. The basis for our simplication
+is to rearrange $p(B)BDe(B) < p(B')BDe(B')$ to be $\frac{p(B)}{p(B')} < \frac{BDe(B')}{BDe(B)}$.
+
+\begin{equation}
+  BDe(B) = \prod_{i=0}^{N}\prod_{l\in L_i} \frac{\Gamma(m'_i(l))}{\Gamma(m_i(l) + m'_i(l))}
+  \prod_{x_i}\frac{\Gamma(m_i(x_i, l) + m'_i(x_i,l))}{\Gamma(m'_i(x_i,l))}
+  \label{eq-hboa}
+\end{equation}
+
+\begin{equation}
+  p(B) = c2^{-0.5(\sum_i|L_i|)log_2\mu}
+  \label{eq-hboa-stop}
+\end{equation}
+
+The outermost product of Equation~\ref{eq-hboa} iterates over all trees in the forest. However, each split
+can modify only one of the trees and therefore the contribution of all others can be canceled. The middle
+product is across all leaves in the tree. Again since only one leaf can be changed, all other terms can
+be canceled. By convention hBOA uses uninformed Bayesian priors of $m'_i(l)= 2$ and $m'_i(x_i, l)=1$ for
+binary alphabets. As $\Gamma(a) = (a-1)!$ this means the top term in the middle product and the bottom
+term in the third product reduce to 1. The only remaining terms are then $m_i(l)$ and $m_i(x_i, l)$ which
+represent the number of solutions which reached leaf $l$ and the number of solutions which reached leaf $l$
+with a specific value for $x_i$, respectively.
+
+Equation~\ref{eq-hboa-stop} can also be simplified when doing comparisons. If model $B'$ has exactly one more
+leaf than model $B$ then the ratio $\frac{p(B)}{p(B')}$ simplifies to $2^{0.5 log_2\mu}$ regardless of
+total model size.
+
+The resulting simplifications create Equation~\ref{eq-hboa-final},
+where $B'$ is different from $B$ by exactly 1 split, such that $l$ was split to create $l'$ and $l''$.
+The best split is whichever maximizes its improvement over $B$, which is equal to the right side
+of the inequality. Note that these factorials can be exceedingly large and
+therefore it is imperative that implementations avoid rounding errors and overflows.
+
 \end{document}

### 2014-10-09 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Less weak language.

@@
 correctly~\citep{goldberg:1991:gasize}. As such, either expert knowledge or exceedingly
-expensive parameter tuning~\citep{grefenstette:1986:optimalga} may be required to leverage
+expensive parameter tuning~\citep{grefenstette:1986:optimalga} are required to leverage
 this feature to its fullest potential. Furthermore,  parameters such as population size, mutation rate, crossover
-rate, tournament size, etc. may have no clear relationship to the problem being solved, meaning even
+rate, tournament size, etc. usually have no clear relationship to the problem being solved, meaning even
 domain experts may not understand how the parameters will interact with the problem or with each other.
@@
 are used to follow each $T_i$ until a leaf is reached. The value of $x_i$ is then set based on the
-probably that other solutions reached that leaf with each value of $x_i$.
+probability that other solutions reached that leaf with each value of $x_i$.
 
@@
 On Deceptive Trap and Deceptive Step Trap using 4 or less traps, $1+(\lambda, \lambda)$ performs significantly
-better than any other algorithm. This is likely
+better than any other algorithm. We believe this is
 because $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
 flipping entire traps.
-This ability to successfully integrate modifications of more than a single bit generally appears to lead $1+(\lambda, \lambda)$
-to outperform the Hill Climber on all problems except Nearest Neighbor NK.
+This ability also leads $1+(\lambda, \lambda)$
+to outperform the Random Restart Hill Climber on all problems except Nearest Neighbor NK.
 
@@
 Only on MAX-SAT are these optimizers competitive
-at larger tested problem sizes. However, this may just be an artifact as the largest tested MAX-SAT
+at larger tested problem sizes. However, we believe this is because the largest tested MAX-SAT
 was an order of magnitude smaller than the largest size tested for most other problems. As the problem
@@
 
-P3's integration of local search gives it similar benefits. Unlike the other model based methods, P3 generally outperforms
-both Random Restart Hill Climber and $1+(\lambda, \lambda)$ on small problem sizes. On larger problem sizes
-P3's ability to perform targeted crossover quickly allows it to surpass $1+(\lambda, \lambda)$'s uninformed
-crossover. This targeted crossover may also explain why P3 is able to outperform $1+(\lambda, \lambda)$ on MAX-SAT.
-
-Local search may also account for some of P3's ability to surpass LTGA and hBOA on larger problem sizes.
-On Deceptive Trap P3's use of hill climbing ensures all traps are immediately optimized, allowing for perfect
-linkage detection and high quality donation. Deceptive Step Trap, which is less amenable to local search, does
-not show a similar improvement over the other model building techniques and performs only slightly better.
-Using local search on HIFF likely saves P3 a generation over LTGA and hBOA as all pairwise subproblems will already be solved,
-accounting for some of its improvement.
-
 \subsection{Model Building}
 Only techniques which explicitly built models of gene epistasis were able to solve the largest problem
-instances. On single instance problems LTGA was generally more effective than hBOA, with hBOA outperforming
+instances. On single instance problems LTGA was more effective than hBOA, with hBOA outperforming
 LTGA on Nearest Neighbor NK and Ising Spin Glasses. This may be caused by the differences in modeling method:
@@
 
+%TODO Simplify this paragraph
 Considering how different hBOA and LTGA are in performing optimization, it is somewhat surprising how similar
-their results are on HIFF. hBOA utilizes Bayesian decision trees to stochastically generate solutions while
-LTGA uses a linkage tree to iteratively modify subsets of genes in an existing solution. On this problem both
-methods likely have the same two limiting factors: initial diversity and generational turnover. In order to
-ensure convergence to the global optima, both methods rely on initial populations which are sufficiently large.
-Both need to preserve enough diversity to ensure higher order building blocks can be created later. Both methods
-also build models once per generation. On HIFF detecting higher order relationships is very unlikely until lower
-order subproblems have been optimized. As a result both LTGA and hBOA are unlikely to solve more than one order
-of subproblem per generation. Therefore both require a number of generations based on the number of subproblem sizes
-in the problem.
-
-Similar to hBOA and LTGA, P3's model building allows it to scale to large problem sizes. As P3 utilizes the same
-model building method as LTGA, we should expect the effects of model building to be similar between those two algorithms.
-Indeed on the four problems where a linkage tree can perfectly represent gene epistasis, P3 has very similar
-scalability (slope) as LTGA, in most cases performing better only by a constant factor.
-
-\subsection{Population Sizing}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{evals-to-success-range}
-  \end{center}
-  \caption{Comparison of the upper and lower quartiles of evaluations required
-           to reach the global optimum for P3 and LTGA with respect to problem size.}
-  \label{fig-evals-to-success-range}
-\end{figure}
-
-As expected, Parameter-less hBOA was generally less efficient than optimally tuned hBOA. The slopes
-of the two lines are relatively similar across all problems, affirming the expectation that
-Parameter-less hBOA is no more than a logarithmic factor worse than hBOA.  The only problem which reversed
-this trend was MAX-SAT, where Parameter-less hBOA actually outperformed hBOA and LTGA. A potential explanation for
-this is that hBOA and LTGA are using the optimal population size for the entire problem class, not for each instance.
-Therefore both must use a population size sufficiently large to solve the hardest instances in the class.
-Parameter-less hBOA, on the other hand, is able to solve easy instances using much smaller population sizes.
-The net effect is therefore that by being more adaptable Parameter-less hBOA is able to achieve better results
-on instances of median difficulty.
-
-Scaling population size to the problem instance instead of the problem class may explain some of why P3 outperforms
-LTGA and hBOA on large problem sizes.
-To more fully examine this hypothesis,
-Figure~\ref{fig-evals-to-success-range} shows the upper and lower quartiles for P3 and LTGA. On each problem
-LTGA has a much smaller difference between its best and worst runs. This makes sense as LTGA uses the same
-population size regardless of instance and progresses search generationally. In contrast P3 has a much
-higher split, with some runs finishing very quickly. This added variability is generally on only the lower
-end as P3's upper quartile is only worse than LTGA's upper quartile on Deceptive Step Trap and Nearest
-Neighbor NK. On these problems the difference in upper quartile is very small, suggesting that even in the worst
-case P3 is no worse than LTGA.
+their results are on HIFF.
+Both techniques rely on populations large enough to support the diversity required to reach the global optimum
+and to model epistasis. Both techniques also rebuild models once per generation. As the subproblems of HIFF
+are nested, it is unlikely that either technique can accurately model higher order epistasis before solving
+lower order subproblems. Therefore both methods require one generation per subproblem order.
+
+\subsection{P3}
+Unlike the other model based methods, P3 generally outperforms
+both Random Restart Hill Climber and $1+(\lambda, \lambda)$ even on small problem sizes.
+Unlike the other local search methods, P3 outperforms LTGA and hBOA even on large problem sizes.
+This implies that P3 is gaining the benefits of each, leveraging local search to solve easy
+problems and model building to solve harder ones.
+
+Furthermore, the interaction between these two optimization tools explains some of the reason P3
+outperforms each method alone.
+On Deceptive Trap P3's use of hill climbing ensures all traps are immediately optimized, allowing for perfect
+linkage detection and high quality donation.
+On HIFF local search solves all pairwise subproblems, saving P3 a generation over LTGA and hBOA.
+In comparison P3 is only a slight improvement on Deceptive Step Trap, which is less amenable to local search.
 
@@
 Perhaps the most striking result is the quality of $1+(\lambda, \lambda)$. Until quite far into search this
-method performs better than both LTGA and hBOA. Given sufficient evaluations $1+(\lambda, \lambda)$ outperforms
+method performs better than both LTGA and hBOA. Given sufficient evaluations $1+(\lambda, \lambda)$ also outperforms
 Random Restart Hill Climbing on all 7 problems. For brief periods in the middle of search it performs the
-best of all techniques on Deceptive Trap, Deceptive Step Trap, HIFF, Ising Spin Glass, and MAX-SAT problems. A likely
-explanation is $1+(\lambda, \lambda)$'s ability to efficiently incorporate gene modifications of larger
-than one bit. This allows it to overcome the deception and plateaus in Deceptive Trap and Deceptive Step Trap, solve medium sized subproblems
+best of all techniques on Deceptive Trap, Deceptive Step Trap, HIFF, Ising Spin Glass, and MAX-SAT problems.
+$1+(\lambda, \lambda)$'s ability to efficiently incorporate gene modifications of larger
+than one bit allows it to overcome the deception and plateaus in Deceptive Trap and Deceptive Step Trap, solve medium sized subproblems
 in HIFF, flip the signs on multiple adjacent bits in Ising Spin Glass, and cross plateaus in MAX-SAT. However,
@@
 
-The integration of hill climbing into P3 makes it strictly better than using hill climbing alone.
-Early in optimization P3 and the Random Restart Hill Climber have effectively identical quality. This is because
-P3 performs the same evaluations as the Hill Climber for the first two restarts. Once P3 begins
-performing crossover it immediately improves over the Hill Climber. In 95\% of intervals P3 had a fitness at
-least as high as Hill Climbing. As such P3 is likely better than a simple
-hill climber regardless of how long each technique is run and irrespective of how high quality the solution
-found has to be.
-
 \subsection{Model Building}
-Both hBOA and LTGA are marked by periods of little improvement and rapid improvement. Initially this is due to
-population size effects (Section~\ref{sec-pop-sizing}). In hBOA this is taken to the extreme, with all fitness
+Both hBOA and LTGA are marked by periods of little improvement followed by rapid improvement.
+In hBOA this is taken to the extreme, with all fitness
 improvement made at the very end of search. In both cases this is caused by model building. Before the model
@@
 initialization of the population, with the first fitness gain achieved immediately upon completing the first generation.
-This can be explained as LTGA's mixing strategy performs a sort of local search when the model has not yet detected gene epistasis.
-This brings LTGA's fitness to almost the same level as the techniques explicitly performing local search, but it is not quite
-as effective at reaching local optima. Subsequent generations then make minor fitness improvements, but generally stagnate
-while the model remains uninformed and high quality building blocks in the population are still relatively rare. Once the model
-becomes accurate enough and the probability of a crossover using high quality genetic material increases sufficiently, LTGA
+When using an inaccurate model, LTGA's mixing strategy performs a sort of less effective local search.
+Subsequent generations then make only minor fitness improvements. Once the model
+becomes accurate and the probability of a crossover using high quality genetic material increases sufficiently, LTGA
 enters a second period of rapid improvement.
 
+\begin{comment}
 This behavior is easiest to understand on the two Trap problems. LTGA's first generation will push individual traps toward a
@@
 self-catalysing as increased frequency of the global optimum trap genes means increased likelihood of the global optimum being spread by crossover.
-
-An almost identical behavior exists in P3, most strikingly in the Deceptive Trap. When solving
-Deceptive Trap, P3 immediately achieves a high fitness, equaled only by the Random Restart Hill Climber in quality. This rise is
-nothing more than bringing every trap to one of the two local optima, with the majority of traps set suboptimally. $1+(\lambda, \lambda)$
-is slightly less efficient about performing single bit improvements, but is the next quickest to reach this plateau.
-While the Random Restart Hill Climber
-remains stuck, P3's use of crossover slowly begins to combine optimal trap values found by multiple independent restarts. Unlike LTGA, which
-performs this improvement all at once, P3 integrates all optimal trap values immediately, waiting only on random restarts to find
-the optimal trap setting.
-
-\subsection{Population Sizing}
-\label{sec-pop-sizing}
-
-At 58\% of the recording intervals hBOA has the worst fitness of any solver. Many of the exceptions occur when
-the total number of evaluations performed is significantly less than the genome size. Under these circumstances hBOA reaches better fitness
-than the Random Restart Hill Climber, $1+(\lambda, \lambda)$, and P3. Early in search hBOA and LTGA are both still evaluating
-individuals in their randomly generated initial populations, while the three techniques worse than hBOA have started
-from a single solution and are attempting to optimize. Of the methods, hBOA uses by far the largest population sizes,
-about 64 times larger than LTGA on average for problems shown. As such it makes sense that hBOA would take a long time to reach high
-quality solutions.
-
-Parameter-less hBOA reported equal or higher fitness than hBOA when using the same number of total evaluations in 62\% of intervals. This makes
-sense as Parameter-less hBOA's use of small populations allows it to begin to optimize long before hBOA has even evaluated
-its entire initial population. However, this trend is reversed after a sufficient number of evaluations, as is apparent
-on Deceptive Step Trap and Ising Spin Glasses. This makes sense as hBOA's population is set to quickly find the global
-optimum.
+\end{comment}
+
+\subsection{P3}
+The integration of hill climbing into P3 makes it strictly better than using hill climbing alone.
+Early in optimization P3 and the Random Restart Hill Climber have effectively identical quality. This is because
+P3 performs the same evaluations as the Hill Climber for the first two restarts. Once P3 begins
+performing crossover it immediately improves over the Hill Climber. In 95\% of intervals P3 had a fitness at
+least as high as Hill Climbing. As such P3 is better than a simple
+hill climber regardless of how long each technique is run and irrespective of how high quality the solution
+found has to be.
+
+Unlike the model based methods, which struggle until model accuracy improves, P3's iterative solution
+integration makes it improve much sooner. This behavior exists in most problems, but is easies to understand on Deceptive Trap.
+On this problem, P3 immediately brings all traps to local optima, equaled only by the Random Restart Hill Climber in quality.
+In comparison LTGA must evaluate the entire population and perform multiple generations to reach similar quality.
+P3 is able to immediately integrate optimal versions of each
+trap into a single individual as they are found by local search, resulting in smoother fitness improvement than LTGA.
+
+\section{Computational Expenses}
+While it is common in evolutionary computation to assume the evaluation function will dominate algorithm
+complexity, in some domains this will not be true. Model based methods are especially likely to violate
+this norm. Therefore in order to assess P3's quality in solving problems with efficient fitness functions,
+we provide data on both its algorithmic complexity and wall clock time.
+
+\subsection{Operation Counting}
+\begin{figure}[t]
+  \begin{centering}
+    \begin{subfigure}{.5\textwidth}
+      \begin{centering}
+        \includegraphicsfit{rebuilds}
+      \end{centering}
+      \caption{Rebuilds}
+      \label{fig-rebuilds}
+    \end{subfigure}%
+    \begin{subfigure}{.5\textwidth}
+      \begin{centering}
+        \includegraphicsfit{donations}
+      \end{centering}
+      \caption{Donations}
+      \label{fig-donations}
+    \end{subfigure}
+  \end{centering}
+  \caption{Computation costs incurred by model rebuilding (Figure~\ref{fig-rebuilds}) and
+           repeated donations (Figure~\ref{fig-donations}) per evaluation as problem size increases.}
+  \label{fig-costs}
+\end{figure}
+
+When discussing the asymptotic complexity of P3 in Section~\ref{sec-p3}, two aspects eluded precise
+analysis: how expensive is model rebuilding and how many gene donations are made. Figure~\ref{fig-costs}
+provides some insight into how often these two aspects of the algorithm are utilized.
+
+Figure~\ref{fig-rebuilds} reports in an algorithmic sense how expensive model rebuilding is for search.
+In order to calculate this value we recorded how many times search rebuilt the model during each run.
+Under the minimal assumption that an evaluation costs $\Omega(N)$ time and that model rebuilds cost $O(N^2)$,
+Figure~\ref{fig-rebuilds} shows the ratio cost of model rebuilding per evaluation. If the cost of model building
+scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically constant.
+For Nearest Neighbor NK, Ising Spin Glasses, and Rastrigin this is the case. For both Trap problems
+and HIFF there is slow growth in the ratio. The problem sizes used for MAX-SAT were not sufficient
+to accurately gauge the asymptotic behavior. Together this suggests that while the cost of building the model
+is almost linear per evaluation, it can grow slowly. However, even in the worst case (HIFF) this growth was
+no more than twice the algorithmic cost of an evaluation even using 2048 bits.
+
+When applying a crossover subset, P3 tries random donors from the population until one is found with at least
+one bit different from the improving solution. In theory this can result in up to $O(\mu)$ operations.
+Figure~\ref{fig-donations} examines the observed average number of donations per evaluation performed.
+Ising Spin Glass, HIFF, and Rastrigin all achieve effectively constant behavior here, implying repeated
+donation does not impact the asymptotic runtime of P3. Both Trap functions and Nearest Neighbor NK all
+increase in number of donations as problem size increases, potentially increasing algorithmic costs. An
+important note is that each donation may range in size from a single bit up to $N-1$, with smaller
+donations far more likely to result in repeated attempts. As such this may cause some super-linear growth
+in P3, but its unlikely to be very high.
+
+\subsection{Wall Clock Performance}
+\begin{figure}
+  \begin{center}
+  \includegraphicsfit{seconds-to-success}
+  \end{center}
+  \caption{Comparison of the median number of seconds to reach the global optimum for
+           the six different optimization methods with respect
+           to problem size.  If the median run did not reach the global optimum no data element
+           is shown.  Results given on a log-log scale.}
+  \label{fig-seconds-to-success}
+\end{figure}
+
+To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}. In total these runs
+took over 6 computing years to complete. We were therefore forced to use
+diverse hardware potentially running multiple processes in parallel.
+As such these timings should be considered very skeptically as a number
+of sources may contribute to biased error in the results.
+
+\subsection{Model Building}
+When considering wall clock time hBOA and Parameter-less hBOA perform significantly worse than when using
+evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while under reasonable
+assumptions P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
+hBOA goes from being slightly more efficient than LTGA in terms of evaluations to significantly worse in terms of seconds.
+As P3 and LTGA require a similar asymptotic complexity per
+evaluation as the Hill Climber and $1+(\lambda, \lambda)$, no similar change in ordering occurs.
+
+\subsection{P3}
+Any increase in P3's algorithmic complexity is offset by its efficient use of evaluations.
+In aggregate P3 found the global optimum the fastest on 72 of 130 configurations,
+with LTGA, $1+(\lambda, \lambda)$, Hill Climber, hBOA outperforming P3 on 44, 18, 10, and 1 configurations, respectively.
+On
+Deceptive Trap, Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT it still finds the global
+optimum faster than all comparison techniques. LTGA is able to match P3's performance almost exactly on HIFF,
+and on Rastrigin P3 falls behind only on moderate sized problems. Deceptive Step Trap yields the worst performance,
+with LTGA using optimal population sizes having a significantly lower growth curve than P3. This mirrors
+the results from Figure~\ref{fig-evals-to-success} under the assumption that P3's computational time per evaluation is higher
+than LTGA but by no more than a constant factor.
+
+\section{Population Sizing}
+A major advantage to P3 is that it does not require the user to set a population size parameter.
+Beyond making P3 easier to apply, this also conveys two additional advantages: diversity scaled
+to initialization and no need to sacrifice intermediate fitness for eventual optimality.
+
+\subsection{Problem Instance versus Problem Class}
+Our procedure for tuning LTGA and hBOA outlined in Section~\ref{sec-tuning} involved finding
+the optimal population size for each class of problem. For black box optimization this is
+realistically the best either algorithm could hope for in practice. Tuning to a problem
+instance or population initialization when solving real world problems is pointless as
+there is no reason to repeat search once the instance has been solved.
+This limitation does not exist in parameter-less methods, which scale their diversity
+based on the problem instance without needing to solve the instance repeatedly.
+
+To achieve high success rates on randomly generated problem classes LTGA and hBOA must use a population size
+large enough to solve the hardest instances in that class. Therefore these methods will have
+population sizes larger than necessary to solve the easiest instances in the class. Even
+on single instance problems, both methods will require population sizes large enough to
+ensure the worst random initialization is diverse enough to solve the problem, which may be
+much larger than the best random initialization.
+
+\begin{figure}
+  \begin{center}
+  \includegraphicsfit{evals-to-success-range}
+  \end{center}
+  \caption{Comparison of the upper and lower quartiles of evaluations required
+           to reach the global optimum for P3 and LTGA with respect to problem size.}
+  \label{fig-evals-to-success-range}
+\end{figure}
+
+Figure~\ref{fig-evals-to-success-range}
+highlights how this can effect the required number of evaluations to reach the global optimum, showing
+the upper and lower quartiles for P3 and LTGA. On each problem
+LTGA has a much smaller difference between its best and worst runs. This makes sense as LTGA uses the same
+population size regardless of instance and progresses search generationally. In contrast P3 has a much
+higher split, with many runs finishing very quickly. This added variability is also generally on the lower
+quartile, with P3's upper quartile only worse than LTGA's on Deceptive Step Trap and Nearest
+Neighbor NK. This supports the hypothesis that P3 is able to scale its diversity not just
+to the problem class, but to the problem instance or even problem initialization, something
+wholly infeasible for tuned population sizing to do.
+
+This tuning distinction is also apparent when comparing Parameter-less hBOA with hBOA in Figure~\ref{fig-evals-to-success}.
+While generally performing worse than hBOA, the difference between the two algorithms is smallest
+on randomly generated problem classes. On MAX-SAT Parameter-less hBOA actually outperformed both hBOA and LTGA,
+likely due to its ability to scale diversity to the problem instance instead of the entire problem class.
+
+\subsection{Fast versus Optimal}
+\label{sec-fast-vs-optimal}
+
+In Section~\ref{sec-overtime} we examined intermediate fitness qualities of LTGA and hBOA when using
+population sizes tuned to reach the global optimum. As a result, both were exceptionally ineffective
+at quickly reaching high quality solutions. This is because unlike P3, these methods have an explicit
+trade off between optimal performance and intermediate performance caused by their population size parameter.
+
+At 58\% of the recording intervals hBOA has the worst fitness of any solver. Most of the exceptions
+occur when hBOA is still evaluating its initial population, allowing this random search to temporarily surpass the
+local search methods. After $N$ evaluations, however, hBOA and LTGA both fall behind until their models begin
+to improve.
+Parameter-less hBOA reaches intermediate fitnesses faster, doing so on 62\% of intervals, as its models begin
+to optimize earlier than hBOA.  However, this trend is reversed after a sufficient number of evaluations, as is apparent
+on Deceptive Step Trap and Ising Spin Glasses, as hBOA's tuned population overtakes Parameter-less hBOA's parallel populations.
 
@@
 
+Figure~\ref{fig-small-pop} examines the effect of population size on LTGA's intermediate fitness by examining the effect
+of reducing LTGA's population size to one tenth of the tuned value. The two problems shown are representative of the behavior of using
+a smaller population size on the other five problems. Reducing the population size caused LTGA to improve earlier but plateau
+at lower fitnesses. This caused LTGA's success rate to drop from 100 to 0 on Deceptive Step Trap and from 98 to 68 on Nearest
+Neighbor NK.
+Even when using a reduced population size for LTGA, P3 still achieved a fitness at
+least as high as LTGA at 80\% of intervals.
+
+\begin{comment}
 It is important to note that hBOA and LTGA were tuned specifically
@@
 size. To examine the effect of population size we tested LTGA using one tenth of the population size required for reliably finding the
-global optimum, with the results given in Figure~\ref{fig-small-pop}. The two problems shown are representative of the behavior of using
-a smaller population size on the other five problems. In aggregate, P3 still achieved a fitness at
-least as high as LTGA using the smaller population size at 80\% of intervals.
+global optimum, with the results given in Figure~\ref{fig-small-pop}.
 Reducing the population size resulted in LTGA's initial fitness improvement
@@
 which balances high quality intermediate fitness without trading away eventual optimality.
-
-\section{Computational Expenses}
-While it is common in evolutionary computation to assume the evaluation function will dominate algorithm
-complexity, in some domains this will not be true. Model based methods are especially likely to violate
-this norm. Therefore in order to assess P3's quality in solving problems with efficient fitness functions,
-we provide data on both its algorithmic complexity and wall clock time.
-
-\subsection{Operation Counting}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{rebuilds}
-      \end{centering}
-      \caption{Rebuilds}
-      \label{fig-rebuilds}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{donations}
-      \end{centering}
-      \caption{Donations}
-      \label{fig-donations}
-    \end{subfigure}
-  \end{centering}
-  \caption{Computation costs incurred by model rebuilding (Figure~\ref{fig-rebuilds}) and
-           repeated donations (Figure~\ref{fig-donations}) per evaluation as problem size increases.}
-  \label{fig-costs}
-\end{figure}
-
-When discussing the asymptotic complexity of P3 in Section~\ref{sec-p3}, two aspects eluded precise
-analysis: how expensive is model rebuilding and how many gene donations are made. Figure~\ref{fig-costs}
-provides some insight into how often these two aspects of the algorithm are utilized.
-
-Figure~\ref{fig-rebuilds} reports in an algorithmic sense how expensive model rebuilding is for search.
-In order to calculate this value we recorded how many times search rebuilt the model during each run.
-Under the minimal assumption that an evaluation costs $\Omega(N)$ time and that model rebuilds cost $O(N^2)$,
-Figure~\ref{fig-rebuilds} shows the ratio cost of model rebuilding per evaluation. If the cost of model building
-scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically constant.
-For Nearest Neighbor NK, Ising Spin Glasses, and Rastrigin this appears to be the case. For both Trap problems
-and HIFF there appears to be slow growth in the ratio. The problem sizes used for MAX-SAT were not sufficient
-to accurately gauge the asymptotic behavior. Together this suggests that while the cost of building the model
-is almost linear per evaluation, it can grow slowly. However, even in the worst case (HIFF) this growth was
-bounded to no more than twice the algorithmic cost of an evaluation.
-
-When applying a crossover subset, P3 tries random donors from the population until one is found with at least
-one bit different from the improving solution. In theory this can result in up to $O(\mu)$ operations.
-Figure~\ref{fig-donations} examines the observed average number of donations per evaluation performed.
-Ising Spin Glass, HIFF, and Rastrigin all achieve effectively constant behavior here, implying repeated
-donation does not impact the asymptotic runtime of P3. Both Trap functions and Nearest Neighbor NK all
-increase in number of donations as problem size increases, potentially increasing algorithmic costs. An
-important note is that each donation may range in size from a single bit up to $N-1$, with smaller
-donations far more likely to result in repeated attempts. As such this may cause some super-linear growth
-in P3, but its unlikely to be very high.
-
-\subsection{Wall Clock Performance}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{seconds-to-success}
-  \end{center}
-  \caption{Comparison of the median number of seconds to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-seconds-to-success}
-\end{figure}
-
-To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}. In total these runs
-took over 6 computing years to complete. We were therefore forced to use
-diverse hardware potentially running multiple processes in parallel.
-As such these timings should be considered very skeptically as a number
-of sources may contribute to biased error in the results. Under those caveats, the results do suggest
-that any increase in P3's algorithmic complexity is offset by its efficient use of evaluations. On
-Deceptive Trap, Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT it still appears to find the global
-optimum faster than all comparison techniques. On HIFF LTGA is able to match P3's performance almost exactly,
-and on Rastrigin P3 falls behind only on moderate sized problems. Deceptive Step Trap yields the worst performance,
-with LTGA using optimal population sizes having a significantly lower growth curve than P3. This mirrors
-the results from Figure~\ref{fig-evals-to-success} under the assumption that P3's computational time per evaluation is higher
-than LTGA but by no more than a constant factor. In aggregate P3 found the global optimum the fastest on 72 of 130 configurations,
-with LTGA, $1+(\lambda, \lambda)$, Hill Climber, hBOA outperforming P3 on 44, 18, 10, and 1 configurations, respectively.
-
-When considering wall clock time hBOA and Parameter-less hBOA perform significantly worse than when using
-evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while under reasonable
-assumptions P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
-hBOA goes from being slightly more efficient than LTGA in terms of evaluations to significantly worse in terms of seconds.
-As P3 and LTGA require a similar asymptotic complexity per
-evaluation as the Hill Climber and $1+(\lambda, \lambda)$, no similar change in ordering occurs.
-
-These results suggest that even for minimally expensive fitness functions P3 can find the global optimum
-at least as quickly in seconds as its optimally tuned competitors. However, due to the potential noise in our data collection
-we are hesitant to draw too strong of a conclusion.
+\end{comment}
 
@@
 that the relationship between the two is strongly problem dependent but relatively independent of problem size.
-In general the trend appears to suggest that problems which can be more tightly represented by crossover tend
-to require less crossover evaluations. Deceptive Trap and Rastrigin spent the least evaluations on crossover, using
+In general, problems which can be more accurately represented by crossover
+required less crossover evaluations. Deceptive Trap and Rastrigin spent the least evaluations on crossover, using
 15\% and 20\% respectively on the largest problem sizes. Both of these problems are perfectly separable and have relatively
@@
 a linkage tree. In comparison to Figure~\ref{fig-cross-success} Deceptive Step Trap performs better than expected for crossover.
-However, the reason for this is likely that even exceedingly uninformed crossover can make improvements to Deceptive Step Trap as
+However, the reason for this is that even exceedingly uninformed crossover can make improvements to Deceptive Step Trap as
 flipping two bits in the same trap will move between two local optima. Nearest Neighbor NK had the least successful crossovers,
@@
 In comparison with LTGA, the crossover success rates of P3 are lower but similar in problem ordering. This is almost certainly
-a consequence of P3's use of local search. Counterintuitively the use of hill climbing on the initial population is likely reducing
+a consequence of P3's use of local search. Counterintuitively the use of hill climbing on the initial population reduces
 P3's crossover success, not because it reduces model quality or the donation pool, but because it is much more challenging to improve
@@
 While the frequency of crossover changes per problem, and the likelihood of an individual crossover making a fitness improvement can be
-quite low, the results from Section~\ref{sec-optimum} and Section~\ref{sec-overtime} suggest it's still critical to optimization. Without
+quite low, the results from Section~\ref{sec-optimum} and Section~\ref{sec-overtime} show it's still critical to optimization. Without
 crossover P3's performance would be identical to the Random Restart Hill Climber, which was unable to solve even moderately sized problems
 and quickly fell behind P3 in intermediate fitness quality. Therefore even the infrequently successful crossover donation are
-critical to success. This does however suggest a potential avenue for future improvement as different modeling and donation algorithms
-may improve P3's crossover success.
+critical to success. This does however suggest a potential avenue for future improvement by using more successful
+modeling and donation algorithms.
 
@@
 solution must be unique and a strict fitness improvement over its previously added version.
-As a result it makes sense that it is generally less likely for a solution to be added to
+As a result solutions are less likely to be added to
 higher levels than lower ones. This shape is also in many ways desirable from a population
@@
 optima when using $k=7$ and $s=2$. Crossover is very likely to overcome the
-two bit plateaus, and as a result solutions in the second level are unlikely
-to contain the lowest fitness local optima of 5 set bits. Similarly the third
+two bit plateaus, and as a result solutions in the second level generally do not
+contain the lowest fitness local optima of 5 set bits. Similarly the third
 level has very few traps set to the next worst local optima of 3 bits. As a result,
@@
 At this point in search the population has not significantly increased the frequency of optimal
-trap settings and with 8 ways to represent suboptimal local optima linkage is also likely to be
+trap settings and with 8 ways to represent suboptimal local optima linkage is generally
 inaccurate. As a result our expectation is that when performing crossover with the third level
@@
 contain a higher frequency of optimal trap settings to have achieved a fitness improvement. As
-a result the crossover success rate and modeling quality at level four is likely much better
-than at level three, increasing the likelihood of solutions being improved by crossover.
+a result the crossover success rate and modeling quality at level four is much better
+than at level three, increasing the frequency of solutions being improved by crossover.
 
@@
 higher levels as solution filtering allows for highly accurate models and high frequencies of high quality
-building blocks. The highest level on most problems has a very low crossover success rate, likely because
+building blocks. The highest level on most problems has a very low crossover success rate because
 so few solutions other than the global optimum every cross with that level.
@@
 modeling overhead, the expense of this overhead is approximately linear with respect to genome size. There
-is some evidence to suggest that even when compared on wall clock time, P3 performs at least as well as
+is some evidence that even when compared on wall clock time, P3 performs at least as well as
 the best comparison techniques. All of these achievements are made without any problem specific parameter
@@
 
-P3's quality is likely due to a number of desirable traits. First, mixing local search with model based
+P3's quality is due to a number of desirable traits. First, mixing local search with model based
 crossover lets search focus on properly mixing high quality solutions. Second, by adding diversity only
@@
 to reach high quality intermediate fitnesses quickly without prematurely converging. Fourth, the very
-nature of the pyramid's shape may be allowing search to preserve a desirable proportion of diversity at
+nature of the pyramid's shape allows search to preserve a desirable proportion of diversity at
 each fitness level, similar to a generational model using a decreasing population size.
@@
 P3 is currently limited to discrete, fixed length genomes evaluated using single objective
-fitness. These limitations can likely be relaxed with future work to make P3 more widely applicable.
+fitness. These limitations can be relaxed with future work to make P3 more widely applicable.
 While asymptotically linear, P3's modeling techniques and local search methods are likely going
 to be prohibitively expensive for genome sizes in the hundreds of thousands or millions of genes,
-and our work suggests the inability of the model to capture overlapping linkage may be hindering
+and the inability of the model to capture overlapping linkage may be hindering
 search efficiency. Overcoming these limitations by using a new modeling technique may allow
@@
 are made may mislead search on landscapes with higher order deception. Even without
-these improvements our results suggest that P3 will be highly efficient at finding
+these improvements our results show P3 is highly efficient at finding
 global optima on black box problems without any problem specific tuning.
@@
 
+\normalsize
 \appendix
-\section{hBOA formula simplification}
+\section{hBOA Simplification}
 \label{sec-appendix}
@@
 and comparison of that change with the existing model. The basis for our simplication
-is to rearrange $p(B)BDe(B) < p(B')BDe(B')$ to be $\frac{p(B)}{p(B')} < \frac{BDe(B')}{BDe(B)}$.
+is to rearrange $p(B)BDe(B) < p(B')BDe(B')$ to be $\frac{p(B)}{p(B')}<\frac{BDe(B')}{BDe(B)}$.
 

### 2014-10-10 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Major editting and reworking.

@@
 This is likely due to the fact that some randomly generated MAX-SAT landscapes are very flat
-and highly deceptive~\cite{rana:1998:gamaxsat}.
+and highly deceptive~\citep{rana:1998:gamaxsat}.
 \section{Finding the Global Optimum}
@@
 Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
-The largest problem size for any problem where P3 was not the most efficient used 49 bits, with P3 achieving the best results on all 92 larger
+The largest problem size for any problem where P3 was not the most efficient has 49 bits, with P3 achieving the best results on all 92 larger
 configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
@@
 
-%TODO Simplify this paragraph
 Considering how different hBOA and LTGA are in performing optimization, it is somewhat surprising how similar
 their results are on HIFF.
-Both techniques rely on populations large enough to support the diversity required to reach the global optimum
-and to model epistasis. Both techniques also rebuild models once per generation. As the subproblems of HIFF
+However, both techniques rely on populations large enough to support the diversity required to reach the global optimum
+and to model epistasis. Both techniques also only rebuild models once per generation. As the subproblems of HIFF
 are nested, it is unlikely that either technique can accurately model higher order epistasis before solving
@@
 Unlike the model based methods, which struggle until model accuracy improves, P3's iterative solution
-integration makes it improve much sooner. This behavior exists in most problems, but is easies to understand on Deceptive Trap.
+integration makes it improve much sooner. This behavior exists in most problems, but is easiest to understand on Deceptive Trap.
 On this problem, P3 immediately brings all traps to local optima, equaled only by the Random Restart Hill Climber in quality.
@@
 
-To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}. In total these runs
+To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}.
+Similar to Figure~\ref{fig-evals-to-success}, each point represents the median
+of 100 runs, with unsuccessful runs treated as slower than successful runs.
+In total these runs
 took over 6 computing years to complete. We were therefore forced to use
@@
 As such these timings should be considered very skeptically as a number
-of sources may contribute to biased error in the results.
-
-\subsection{Model Building}
+of sources may contribute to biased error in the results. However, due to
+the size of the dataset and the general regularity in the results general conclusions
+can still be made.
+
+\subsubsection{Model Building}
 When considering wall clock time hBOA and Parameter-less hBOA perform significantly worse than when using
@@
 
-\subsection{P3}
+\subsubsection{P3}
 Any increase in P3's algorithmic complexity is offset by its efficient use of evaluations.
@@
 Our procedure for tuning LTGA and hBOA outlined in Section~\ref{sec-tuning} involved finding
-the optimal population size for each class of problem. For black box optimization this is
-realistically the best either algorithm could hope for in practice. Tuning to a problem
-instance or population initialization when solving real world problems is pointless as
-there is no reason to repeat search once the instance has been solved.
+the optimal population size for each class of problem. For real world black box optimization this is
+realistically the best either algorithm could hope for as tuning to a problem
+instance or population initialization involves repeatedly solving the problem being tuned.
 This limitation does not exist in parameter-less methods, which scale their diversity
-based on the problem instance without needing to solve the instance repeatedly.
+based on the problem instance without needing to solve that instance repeatedly.
 
@@
 While generally performing worse than hBOA, the difference between the two algorithms is smallest
-on randomly generated problem classes. On MAX-SAT Parameter-less hBOA actually outperformed both hBOA and LTGA,
+on randomly generated problem classes. On MAX-SAT, Parameter-less hBOA actually outperformed both hBOA and LTGA,
 likely due to its ability to scale diversity to the problem instance instead of the entire problem class.
@@
 to improve.
-Parameter-less hBOA reaches intermediate fitnesses faster, doing so on 62\% of intervals, as its models begin
-to optimize earlier than hBOA.  However, this trend is reversed after a sufficient number of evaluations, as is apparent
+Parameter-less hBOA reaches intermediate fitnesses faster than hBOA, doing so on 62\% of intervals, as its models begin
+to optimize earlier than hBOA.  However, this trend is reversed after a sufficient number of evaluations, most clearly
 on Deceptive Step Trap and Ising Spin Glasses, as hBOA's tuned population overtakes Parameter-less hBOA's parallel populations.
@@
 \subsection{Crossover}
+Figure~\ref{fig-cross} shows the proportion of evaluations P3 spends on crossover, as opposed to hill climbing,
+and Figure~\ref{fig-cross-success} shows what percentage of crossover evaluations resulted in a fitness improvement.
+Together these figures provide some insight into the role of crossover within P3. The behavior
+for each is clearly problem dependent and generally asymptoticly stable as problem size increases.
+
+When solving problems where epistasis can be effectively detected and represented by a linkage tree, P3 tends to spend
+less evaluations performing crossover and each crossover is more likely to be successful.
+Deceptive Trap and Rastrigin are the easiest problems to model epistasis, with local search quickly
+reducing pairwise entropy in each. These are also the problems where P3 uses the least evaluations
+on crossover and has the highest success rates for crossover. At the other extreme are Nearest Neighbor
+NK and Ising Spin Glasses, which both have overlapping linkage not representable by a linkage tree.
+These problems have the highest crossover usage and lowest crossover success of any problem except Deceptive
+Step Trap.  While Deceptive Step Trap's epistasis can be accurately modeled by a linkage tree, the
+exponential number of plateaus makes detecting gene linkage very challenging.
+
+\begin{comment}
 P3 performs optimization using a mixture of hill climbing and crossover. Exactly how much time the algorithm
@@
 As a result it makes sense that individual crossovers are very unlikely to be fitness improvements.
-
-In comparison with LTGA, the crossover success rates of P3 are lower but similar in problem ordering. This is almost certainly
-a consequence of P3's use of local search. Counterintuitively the use of hill climbing on the initial population reduces
+\end{comment}
+
+When compared with LTGA, P3's crossover success rates are lower but similar in problem ordering.
+Counterintuitively the use of hill climbing on the initial population reduces
 P3's crossover success, not because it reduces model quality or the donation pool, but because it is much more challenging to improve
@@
 
-While the frequency of crossover changes per problem, and the likelihood of an individual crossover making a fitness improvement can be
-quite low, the results from Section~\ref{sec-optimum} and Section~\ref{sec-overtime} show it's still critical to optimization. Without
+Even when crossover success rates are quite low, such as Nearest Neighbor NK's 0.007\% success,
+the results from Section~\ref{sec-optimum} and Section~\ref{sec-overtime} show it's still critical to optimization. Without
 crossover P3's performance would be identical to the Random Restart Hill Climber, which was unable to solve even moderately sized problems
@@
 With the exception of the dip in Deceptive Step Trap, all of the pyramids show a monotonic reduction
-in size as the level increases. This makes sense from an algorithmic standpoint as solutions are
-added to the first level as long as they are unique local optima. For each subsequent level a
-solution must be unique and a strict fitness improvement over its previously added version.
-As a result solutions are less likely to be added to
-higher levels than lower ones. This shape is also in many ways desirable from a population
-genetics standpoint. When using a generational model \cite{lobo:2011:dynamicpop} found theoretical
-evidence that the optimal population size reduces each generation and \cite{goldman:2011:dynamic-parameters}
-made similar empirically based conclusions. In P3, levels of the pyramid are akin to generations as solutions
-stored in each level have undergone the same number of variation and selection events,
-with higher levels representing more of both. Therefore it is possible that P3's
-pyramid shape is related to its ability to outperform competitors using an optimal
-fixed population size. Subsequently smaller levels are able to focus search,
-preserving only the diversity required by that stage of search. In comparison,
-LTGA and hBOA set the population size at each generation equal to that required
-to ensure sufficient initial diversity.
-
+in size as the level increases. This is because a solution must be a strict fitness improvement
+over its previously added version to be added to a higher level, which becomes less likely
+each time the solution improves. \cite{lobo:2011:dynamicpop} found theoretical evidence
+and \cite{goldman:2011:dynamic-parameters} found empirical evidence that the optimal
+population size decreases each generation of traditional evolutionary search.
+By decreasing in size, P3 implicitly stores more diversity in low levels and focuses
+search on high quality solutions at higher levels. In comparison,
+LTGA and hBOA suboptimally use a fixed population size at each generation.
+
+Figure~\ref{fig-level-success} examines how crossover success changes at different
+levels of the pyramid. At low levels, success gets progressively lower
+as solution quality increases faster than the model's ability to improve solutions.
+At higher levels modeling becomes more accurate and donations contain higher frequencies
+of high quality building blocks, resulting in increased crossover success. The
+highest level of most problems has a low crossover success rate, as solutions
+crossing with that level have already been improved by previous operations to
+the point where the only improvement would be to create the global optimum, which can
+only happen once.
+
+\subsection{Deceptive Step Trap}
 The number of solutions stored at the fourth level of Deceptive Step Trap is
 significantly lower than that of the third or fifth levels, breaking the decreasing
-trend of the other six problems. This behavior is rooted in the peculiar nature
-of this landscape, and is also present in all other problem sizes tested.
+trend of the other six problems. Figure~\ref{fig-level-success} has a similar
+aberration, with crossover success dropping to  0.0004\% before rebounding and
+following the more common trajectory. This behavior exists in all other problem
+sizes tested, with the dips occurring at exactly the same level.
+
+This behavior is rooted in the peculiar nature
+of this landscape.
 After local search, all traps in all solutions have a total
@@
 two bit plateaus, and as a result solutions in the second level generally do not
-contain the lowest fitness local optima of 5 set bits. Similarly the third
-level has very few traps set to the next worst local optima of 3 bits. As a result,
-most solutions which reach the third level can only be improved by overcoming deception
-and replacing the second best local optima of zero or one bits set to having all seven bits set.
-At this point in search the population has not significantly increased the frequency of optimal
-trap settings and with 8 ways to represent suboptimal local optima linkage is generally
-inaccurate. As a result our expectation is that when performing crossover with the third level
-of the pyramid very few fitness improvements will actually be made, meaning
-few solutions will be added to the fourth level. Those solutions that are added must, by definition,
-contain a higher frequency of optimal trap settings to have achieved a fitness improvement. As
-a result the crossover success rate and modeling quality at level four is much better
-than at level three, increasing the frequency of solutions being improved by crossover.
-
-When examining the crossover success rate shown in Figure~\ref{fig-level-success} these suppositions are
-supported. The success rate for Deceptive Step Trap decreases for the first three levels, with the third level
-successful only 0.0004\% of the time. This quickly rebounds at higher levels. On all other problems, the
-crossover success rate follows a trajectory closer to expectation. Low levels are more successful than medium
-levels as it is easier to make improvements to less optimized solutions. This trend reverses somewhat at
-higher levels as solution filtering allows for highly accurate models and high frequencies of high quality
-building blocks. The highest level on most problems has a very low crossover success rate because
-so few solutions other than the global optimum every cross with that level.
+contain the lowest fitness local optima (5 bits set) and the third
+level has very few traps set to the next worst local optima (3 bits set). As a result,
+solutions which reach the third level can only be improved by replacing the deceptive
+local optima (0 and 1 bits set) with the global optimum (7 bits set).
+The global optimum is very rare in the population, and with 8 ways to represent local optima
+linkage learning is inaccurate. Therefore it is very unlikely for crossover to be successful,
+meaning few solutions will be added to the fourth level. Solutions that do improve by
+definition must have a higher frequency of optimal trap settings, meaning level four's model
+will be more accurate and donations are more likely to contain optimal trap values.
+Thus the level size and crossover success rates increase after contracting around level four.
 
@@
 order deception via linkage learning, the iterative improvement method by which crossovers
-are made may mislead search on landscapes with higher order deception. Even without
+are made may mislead search on landscapes with higher order deception.
+
+However, even without
 these improvements our results show P3 is highly efficient at finding
@@
 The best split is whichever maximizes its improvement over $B$, which is equal to the right side
-of the inequality. Note that these factorials can be exceedingly large and
+of the inequality. Note that these factorials can still be exceedingly large and
 therefore it is imperative that implementations avoid rounding errors and overflows.

### 2014-10-10 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Last commit before sending to Bill

@@
 
-\subsection{Deceptive Step Trap}
+\subsubsection{Deceptive Step Trap}
 The number of solutions stored at the fourth level of Deceptive Step Trap is

### 2014-10-14 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
All edits complete.

@@
 almost always creates near infinitesimal results, calculating fractions which include factorials of $\mu$
-and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{sec-appendix} we
-derived the simplified Equation~\ref{eq-hboa-final}. If no proposed split satisfies the inequality, iteration stops.
+and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{sec-appendix}, we
+derived a simplified form shown in Equation~\ref{eq-hboa-final}. If no proposed split satisfies the inequality, iteration stops.
 If multiple splits do, whichever maximizes the right side is chosen.
@@
 
-To model properly, hBOA is designed to work with large population sizes, resulting in a large number of
+hBOA is designed to work with large population sizes, resulting in a large number of
 evaluations per generation. As hBOA utilizes explicit diversity maintenance, standard methods for determining
@@
 Random Restart Hill Climbing on all 7 problems. For brief periods in the middle of search it performs the
-best of all techniques on Deceptive Trap, Deceptive Step Trap, HIFF, Ising Spin Glass, and MAX-SAT problems.
+best of all techniques on: Deceptive Trap; Deceptive Step Trap; HIFF; Ising Spin Glass; and MAX-SAT problems.
 $1+(\lambda, \lambda)$'s ability to efficiently incorporate gene modifications of larger
@@
 in HIFF, flip the signs on multiple adjacent bits in Ising Spin Glass, and cross plateaus in MAX-SAT. However,
-this method is not enough to quickly reach the global optima in many of these problems which causes it to eventually
+this method is slow in reaching the global optima in many of these problems which causes it to eventually
 be overtaken by the model building techniques.
@@
 Unlike the model based methods, which struggle until model accuracy improves, P3's iterative solution
-integration makes it improve much sooner. This behavior exists in most problems, but is easiest to understand on Deceptive Trap.
+integration allows it to improve much more quickly. This behavior exists in most problems, but is easiest to understand on Deceptive Trap.
 On this problem, P3 immediately brings all traps to local optima, equaled only by the Random Restart Hill Climber in quality.
@@
 complexity, in some domains this will not be true. Model based methods are especially likely to violate
-this norm. Therefore in order to assess P3's quality in solving problems with efficient fitness functions,
+this norm. Therefore, in order to assess P3's quality in solving problems with efficient fitness functions,
 we provide data on both its algorithmic complexity and wall clock time.
@@
 of sources may contribute to biased error in the results. However, due to
-the size of the dataset and the general regularity in the results general conclusions
+the size of the dataset and the general regularity in the results some general conclusions
 can still be made.
@@
 \subsubsection{Model Building}
-When considering wall clock time hBOA and Parameter-less hBOA perform significantly worse than when using
-evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while under reasonable
-assumptions P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
+hBOA and Parameter-less hBOA perform significantly worse when using wall clock time as the unit of comparison than when using
+evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while, under reasonable
+assumptions, P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
 hBOA goes from being slightly more efficient than LTGA in terms of evaluations to significantly worse in terms of seconds.
@@
 On
-Deceptive Trap, Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT it still finds the global
+Deceptive Trap, Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT P3 still finds the global
 optimum faster than all comparison techniques. LTGA is able to match P3's performance almost exactly on HIFF,
@@
 
-To achieve high success rates on randomly generated problem classes LTGA and hBOA must use a population size
+To achieve high success rates on randomly generated problem classes, LTGA and hBOA must use a population size
 large enough to solve the hardest instances in that class. Therefore these methods will have
@@
 When compared with LTGA, P3's crossover success rates are lower but similar in problem ordering.
-Counterintuitively the use of hill climbing on the initial population reduces
+Counterintuitively, the use of hill climbing on the initial population reduces
 P3's crossover success, not because it reduces model quality or the donation pool, but because it is much more challenging to improve
@@
 the results from Section~\ref{sec-optimum} and Section~\ref{sec-overtime} show it's still critical to optimization. Without
-crossover P3's performance would be identical to the Random Restart Hill Climber, which was unable to solve even moderately sized problems
-and quickly fell behind P3 in intermediate fitness quality. Therefore even the infrequently successful crossover donation are
-critical to success. This does however suggest a potential avenue for future improvement by using more successful
+crossover, P3's performance would be identical to the Random Restart Hill Climber, which was unable to solve even moderately sized problems
+and quickly fell behind P3 in intermediate fitness quality. Therefore even infrequently successful crossover donations are
+critical to success. This does, however, suggest a potential avenue for future improvement by using more successful
 modeling and donation algorithms.
@@
 
-Another feature unique to P3 worth investigating is the shape and size of the population pyramid
+Another feature unique to P3 is the shape and size of the population pyramid
 constructed for each problem. Figure~\ref{fig-level-size} shows the number of solutions stored at
-each level of the pyramid for the largest tested problem size for each problem. Each point is the
-median size across 100 runs. If a run did not store any solutions at a level it is treated as 0,
-and no point is drawn if the median run had 0 solutions stored at that level. While pyramid size
+each level of the pyramid for the largest tested problem sizes. Each point is the
+median size across 100 runs. If a run did not store any solutions at a level it is treated as 0.
+No point is drawn if the median run had 0 solutions stored at that level. While pyramid size
 is effected by problem size, the overall shape is not. As such the behavior shown in Figure~\ref{fig-level-size}
@@
 in size as the level increases. This is because a solution must be a strict fitness improvement
-over its previously added version to be added to a higher level, which becomes less likely
+over its previous version to be added to a higher level, which becomes less likely
 each time the solution improves. \cite{lobo:2011:dynamicpop} found theoretical evidence

### 2014-10-15 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Stray leading space removed.

@@
 \subsection{Linkage Tree Genetic Algorithm}
-~\cite{thierens:2010:ltga} introduced the Linkage Tree Genetic Algorithm (LTGA) which automatically
+\cite{thierens:2010:ltga} introduced the Linkage Tree Genetic Algorithm (LTGA) which automatically
 detects and exploits problem epistasis by examining pairwise gene entropy. Due to its enhanced

### 2014-10-15 Comprehensive/abstract.tex
Initial commit of Comprehensive Report

--- 2014-10-15 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
@@
-\documentclass[twoside]{article}
-\usepackage{ecj,palatino,epsfig,latexsym,natbib}
-\usepackage{url}
-\usepackage{verbatim}
-\usepackage{subcaption}
-\usepackage[noend]{algpseudocode}
+%%% Abstract:
+\thispagestyle{empty} \setcounter{page}{2}
+\begin{doublespace}
+\begin{centering}
+ABSTRACT\\ %
+Analyzing the Parameter-less Population Pyramid \\ %
+By \\ %
+Brian W. Goldman\\ %
+\ \\
+\end{centering}
 
-\newcommand{\includegraphicswide}[1]
-{\includegraphics[width=.9\textwidth,height=\textheight,keepaspectratio]{#1}}
-
-\newcommand{\includegraphicsfit}[1]
-{\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}
-
-\usepackage{array}
-\newcolumntype{L}{>{\centering\arraybackslash}m{1.9cm}}
-
-%% do not add any other page- or text-size instruction here
-
-\parskip=0.00in
-
-\begin{document}
-
-% 45-character limit on header
-\ecjHeader{x}{x}{xxx-xxx}{201X}{Parameter-less Population Pyramid}{B. W. Goldman and W. F. Punch}
-\title{\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid}
-
-\author{\name{\bf B. W. Goldman} \hfill \addr{brianwgoldman@acm.org}\\ 
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-\AND
-       \name{\bf W. F. Punch} \hfill \addr{punch@msu.edu}\\
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-}
-
-\maketitle
-
-\begin{abstract}
-%Should be about 200 words
 The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
@@
 
-\end{abstract}
-
-\begin{keywords}
-
-Genetic algorithms, 
-linkage learning,
-local search,
-parameter-less.
-
-\end{keywords}
-
-\section{Introduction}
-A primary purpose of evolutionary optimization is to efficiently find good solutions
-to challenging real world problems with minimal prior knowledge about the problem itself.
-This driving goal has created search algorithms which can escape user bias to create
-truly novel results, sometimes publishable or patentable in their own right~\citep{kannappan:2014:humies}.
-While it is not possible for any algorithm to do better than random search across all possible
-problems~\citep{Wolpert:1997:nfl}, effectiveness can be achieved by assuming the search
-landscape has structure and then biasing the algorithm toward exploiting that structure.
-
-In evolutionary optimization, and genetic algorithms~(GAs) in particular, search is often
-biased through parameters. This can be beneficial as it allows practitioners to inject their
-knowledge about the shape of the search landscape into the algorithm.
-However, the quality of solutions found, and the speed at which they are found, is strongly tied to setting these parameters
-correctly~\citep{goldberg:1991:gasize}. As such, either expert knowledge or exceedingly
-expensive parameter tuning~\citep{grefenstette:1986:optimalga} are required to leverage
-this feature to its fullest potential. Furthermore,  parameters such as population size, mutation rate, crossover
-rate, tournament size, etc. usually have no clear relationship to the problem being solved, meaning even
-domain experts may not understand how the parameters will interact with the problem or with each other.
-
-There have been periodic efforts to reduce or remove the need for parameter tuning.
-\cite{Back:1992:selfadapt} introduced self-adaptive parameters, in which parameter values
-were included in each solution's genome and themselves underwent evolution. This allowed search
-to optimize some of its own parameters resulting in a reduced the need for expert tuning.
-\cite{harik:1999:parameterlessga} were able to design an entirely parameter-less GA by
-leveraging schema theory and parallel populations. Unfortunately these methods were provably less efficient
-than directly setting the parameters to optimal values~\citep{pelikan:1999:worstparameter-less}.
-
-One area that has been very effective at reducing the number of algorithm parameters is
-model based search. \cite{pelikan:2006:hboa}'s Hierarchical Bayesian Optimization
-Algorithm~(hBOA) and \cite{thierens:2010:ltga}'s Linkage Tree Genetic Algorithm~(LTGA)
-both require only a single parameter: population size. \cite{posik:2011:parameterless}
-leveraged model building to create a fully parameter-less algorithm, but it is restricted to
-only order-k, fully decomposable, noiseless problems.
-
-Most recently \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid~(P3).
-This method uses a pyramid structure of populations to combine model based search with local search
-to achieve parameter-less optimization. Initial results suggest that unlike
-previous parameter-less methods, P3 is actually more efficient than current state-of-the-art
-parameterized search algorithms. In this work we shall: extend these results to cover more
-comparison algorithms; compare both efficiency in reaching the global optimum and intermediate
-fitnesses; algorithm complexity analysis; and provide more in depth analysis of P3 itself.
-
-\section{Comparison Optimizers}
-\label{sec-optimizers}
-
-In order to fully understand the effectiveness of P3, we compare it with five advanced algorithms which
-have related features to P3. The Random Restart Hill Climber~\citep{goldman:2014:p3} was chosen as an efficient form of repeated
-local search. As P3 combines this hill climber with crossover,
-comparing with local search alone shows the advantages of P3's overall approach.
-The $1+(\lambda, \lambda)$ algorithm is the current best theory supported simple genetic algorithm
-and its method of crossover is in some sense a macro-mutation just as in P3. hBOA and Parameter-less
-hBOA are advanced model building search techniques which are very effective at learning complex
-problem structure, designed to achieve similar goals as P3's linkage learning but using very different
-methods. Finally LTGA represents the current state-of-the-art in black box search and is the origin
-of P3's linkage learning and crossover methods.
-
-Finally, only hBOA and LTGA require any parameters, with each of these only requiring a population
-size. This makes knowing the optimal behavior of these algorithms much more tractable. All of the
-algorithms are also gene order independent, fitness scale invariant, and unbiased. This means,
-for any problem, the order in which problem variables appear in the genome can be changed
-without changing the behavior of the search. The fitness can also be manipulated in any fashion
-as long as the rank ordering of solutions is unchanged. These algorithms are also unaffected by the meaning
-assigned to each bit, such that inverting a predetermined random subset of genes before evaluation
-will not impact search efficiency.
-
-Our implementations of all of these algorithms as well as all of the population size information, raw results,
-and processing scripts
-are available from our website.\footnote{\url{https://github.com/brianwgoldman/FastEfficientP3}}
-\subsection{Random Restart Hill Climber}
-\label{sec-hill-climber}
-Perhaps the simplest black box search heuristic is stochastic local search, or hill climbing.
-This optimization technique focuses on improving a single solution until it reaches a local
-optimum. Here we use the first improvement hill climber defined by \cite{goldman:2014:p3}
-and given in Figure~\ref{fig-hc}. This algorithm works by flipping each bit in a random
-order, keeping modifications when fitness is improved, until single bit flips cannot result in
-further fitness improvements.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Hill-Climber}{}
-    \State $options \leftarrow [0 \dots N-1]$\label{fig-hc-options}
-    \State $tried \leftarrow \emptyset$
-    \While{$|tried| < |options|$}
-      \ForAll{$index \in shuffled(options)$}
-        \If{$index \notin tried$}\label{fig-hc-tried}
-          \State Flip bit $index$ in solution
-          \If{solution's fitness increased}
-            \State $tried \leftarrow \emptyset$
-          \Else
-            \State Revert change
-          \EndIf
-          \State $tried \leftarrow tried \cup \{index\}$
-        \EndIf
-      \EndFor
-    \EndWhile
-  \EndProcedure
-\end{algorithmic}
-  \caption{Hill climbing algorithm used to improve randomly generated solutions until no single
-           bit change results in a fitness improvement.}
-  \label{fig-hc}
-\end{figure}
-
-The hill climber requires an amortized cost of $O(1)$ operations per evaluation. In order to
-terminate, at least one evaluation must be performed for each of the $N$ bits in the solution.
-As such any operation that happens only once per search can be amortized over at least $N$
-evaluations, covering the initialization of $options$ on Line~\ref{fig-hc-options}.
-Line~\ref{fig-hc-tried}, which prevents wasted evaluations, can be called at most twice
-per evaluation: once to add $index$ into $tried$ and once to prevent $index$ from being unnecessarily
-evaluated again. The only way three or more calls could happen is if
-no fitness improvement was made for the entire previous iteration, which contradicts the loop invariant.
-
-Due to its nature, this hill climber cannot escape basins of attraction. Once a solution is reached
-such that none of the single bit neighbors are fitness improvements, search stops. Thus
-this algorithm requires a restart mechanism to solve multimodal problems. We have
-chosen here to naively restart search from a random solution whenever a local optima is found. This
-ensures that on all landscapes there is always a non-zero probability of search finding the global optimum.
-
-\subsection{$1+(\lambda, \lambda)$}
-\cite{doerr:2013:lambdalambda} presented the first genetic algorithm to provably show
-the advantages of performing crossover on simple landscapes. This comparatively simple
-algorithm maintains only a single individual and a self-controlled parameter $\lambda$.
-
-Each iteration, the number of bits to flip is chosen from the binomial distribution $b\sim B(N, \frac{\lambda}{N})$,
-where $N$ is the number of bits in the genome.
-Next, $\lfloor\lambda\rfloor$ offspring are produced by flipping $b$ bits. The
-best mutant then produces $\lfloor\lambda\rfloor$ offspring via uniform crossover with the original parent, such that each gene comes from the
-mutant with probability $\frac{1}{\lambda}$. In the original algorithm the best
-offspring produced by crossover then replaces the original parent if its fitness is no worse.
-The $\lambda$ parameter, which is initialized to 1, is decreased if the offspring replaced
-its parent and increased otherwise.
-
-The original formulation was designed specifically for unimodal landscapes and as such were
-not directly suitable for multimodal problems. \cite{goldman:2014:p3} extended $1+(\lambda, \lambda)$
-to include random restarts. As search stagnates, the $\lambda$ parameter increases in value. Eventually
-this results in $\lambda \ge N$ causing mutation to always flip all bits of the individual.
-As this prevents any future improvement, whenever $\lambda \ge N$ search is restarted from a random solution with $\lambda$
-reset to 1.
-
-A few other efficiency modifications were also made. If there is a tie in crossover offspring fitness,
-whichever has a larger hamming distance from the parent is retained. This encourages drifting across plateaus.
-The ``mod'' control strategy proposed by \cite{doerr:2013:lambdalambda} was not used as it conflicted with
-the random restart strategy.
-If a crossover individual is identical to either of its parents, it is not evaluated.
-If mutation produces an offspring which is better than the best crossover offspring, it is used to compare
-against the original parent.
-
-\subsection{Hierarchical Bayesian Optimization Algorithm}
-
-\cite{pelikan:2006:hboa} used statistical principles in combination with a decision tree structure
-to create the Hierarchical Bayesian Optimization Algorithm (hBOA). This method creates a model of
-epistatic relationships between genes which is then used to stochastically generate new solutions.
-Each generation a binary tournament with replacement is used to select $\mu$ solutions from
-the population. These solutions are then used to build the model, which in turn is used to generate $\mu$ new
-solutions. The new solutions are then integrated into the population using restricted tournament
-replacement.
-
-Conceptually, the model built by hBOA is trying to infer rules of the form ``Given that this
-subset of genes are set to these values, how frequently is gene $x_i$ set to value v?'' This can
-be represented using a directed acyclic decision forest, with each tree in the forest representing one gene
-in the solution. In the decision tree $T_i$, which is used to set the value of gene $x_i$,
-each internal node represents previous decisions on how to set
-some other gene $x_j$, with the children of that node representing how the decision was made. The
-leaves of each tree give the probability that $x_i$ should be set to each possible value.
-
-The forest is constructed iteratively, with each tree initially containing a single leaf
-and with each leaf storing a pointer for each selected solution. Each iteration the algorithm considers
-all possible ways of splitting an existing leaf using another gene $x_j$, such that solutions in the
-leaf are moved to the newly created leaves based on their value for $x_j$. The general goal is to
-separate the solutions such that all solutions with $x_i = 0$ move to one leaf while solutions with
-$x_j = 1$ move to the other.
-
-This goal is formalized using model scoring from Bayesian statistics. In its raw form this
-almost always creates near infinitesimal results, calculating fractions which include factorials of $\mu$
-and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{sec-appendix}, we
-derived a simplified form shown in Equation~\ref{eq-hboa-final}. If no proposed split satisfies the inequality, iteration stops.
-If multiple splits do, whichever maximizes the right side is chosen.
-
-\begin{equation}
-  2^{0.5 log_2\mu} < \frac{(m_i(l) + 1)!}{m_i(0, l)!m_i(1,l)!} \cdot
-  \frac{m_i(0, l')!m_i(1,l')!m_i(0, l'')!m_i(1,l'')!}{(m_i(l') + 1)!(m_i(l'') + 1)!}
-  \label{eq-hboa-final}
-\end{equation}
-
-Initially there are $\Theta(N^2)$ possible ways to split existing leaves, as each of the $N$ single node
-trees can be split by any of the other $N-1$ genes. Each iteration a new edge is added to the decision
-forest, meaning some of the previously tested splits cannot be used. For instance, if $T_i$, which is used
-to decide the value of $x_i$, is split using the value of $x_j$, $T_j$ can no longer be split using $x_i$.
-As a split creates two new leaves, $O(N)$ new potential splits must also be tested. Equation~\ref{eq-hboa-final}
-parses all solutions which reach a leaf to count gene frequencies, requiring $\mu$ time.
-The number of total leaves created depends heavily on the problem and $\mu$.
-However, assuming no splits are accepted or that the cost of testing all future splits is less than
-the initial $\Theta(N^2)$, constructing the model requires $\Omega(\mu N^2)$ time. Each
-model is used to generate $\mu$ solutions, leading to a cost per evaluation of $\Omega(N^2)$.
-
-To generate a solution, the value of each gene $x_i$ is set using its corresponding decision tree $T_i$. Because
-the forest is directed acyclic, there must be an ordering of $T_i$ such that before $T_i$ is executed all
-$x_j$ it uses to make decisions have already been set. As such, previous decisions made by other trees
-are used to follow each $T_i$ until a leaf is reached. The value of $x_i$ is then set based on the
-probability that other solutions reached that leaf with each value of $x_i$.
-
-To perform replacement, hBOA uses restricted tournament replacement. After each new solution is generated
-and evaluated, a set of $w$ solutions are chosen at random from the population, where $w=\min\{N, \frac{\mu}{20}\}$.
-From this set the solution which is most genetically similar to the offspring is chosen. If the offspring
-is at least as fit as the chosen solution, it replaces the chosen solution in the population. Otherwise the
-offspring is discarded. This method is designed to preserve genetic diversity as only genetically similar
-solutions must compete on fitness.
-
-hBOA is designed to work with large population sizes, resulting in a large number of
-evaluations per generation. As hBOA utilizes explicit diversity maintenance, standard methods for determining
-convergence are not considered very accurate. Therefore the authors suggest that an hBOA run should be
-terminated after performing generations equal to $N$.
-
-Like other model based techniques, hBOA has very few parameters. There is no mutation or crossover,
-and modeling does not rely on any explicit parameters. Solution selection, generation, and replacement
-are all derived from the population size, which must be set by the user.
-
-\subsection{Parameter-less hBOA}
-Using the methods first introduced by \cite{harik:1999:parameterlessga} for the Parameter-less GA,
-\cite{pelikan:2004:parameterlesshboa} created Parameter-less hBOA which automatically scales its
-population size to fit the problem. This is done by maintaining a list of concurrent populations
-using exponentially scaled population sizes.
-
-A run of Parameter-less hBOA starts with a single population of size $\mu_0$, conventionally set
-to $\mu_0=10$. After two generations are performed, a new population of size $\mu_1 = 2\mu_0$ is created
-and performs a generation. Evolution then continues with the $\mu_0$ population performing two generations
-for each one performed by $\mu_1$. Each time population $\mu_i$ performs its second generation a new population
-$\mu_{i+1}=2\mu_i$ is created, which performs generations half as often as $\mu_i$. In this way an infinite number of
-parallel population can be simulated, with each population receiving the same number of total evaluations.
-
-In all other aspects each population is identical to an hBOA population using a fixed $\mu$. No search information
-is shared among populations, and each search is independently terminated. As such Parameter-less hBOA cannot
-perform better than hBOA using the optimal population size for a given instance, as it must also spend evaluations
-on populations of different sizes. This inefficiency is bounded by a log multiple of the total number of
-evaluations~\citep{pelikan:1999:worstparameter-less}.
-
-\subsection{Linkage Tree Genetic Algorithm}
-\cite{thierens:2010:ltga} introduced the Linkage Tree Genetic Algorithm (LTGA) which automatically
-detects and exploits problem epistasis by examining pairwise gene entropy. Due to its enhanced
-ability to preserve high fitness gene subsets, LTGA was able to outperform state-of-the-art
-GAs across many benchmarks. Since its introduction, many variants of LTGA have been
-proposed~\citep{goldman:2012:ltga} so for clarity we have chosen the version
-presented by \cite{thierens:2013:ltgahiff} as our model.
-
-LTGA's effectiveness comes from its method of performing crossover. Instead of blindly
-mixing genes between parents, LTGA attempts to preserve important interrelationships
-between genes. Before performing any crossovers in a generation, LTGA first builds
-a set of hierarchical gene clusters which are then used to dictate how genes are mixed
-during crossover.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Cluster-Creation}{}
-    \State $unmerged \leftarrow \{\{0\}, \{1\}, \{2\}, \dots, \{N-1\}\}$
-    \State $useful \leftarrow unmerged$
-    \While{$|unmerged|>1$}
-      \State $C_i, C_j \leftarrow \min_{C_i,C_j \in unmerged} D(C_i, C_j)$
-      \State $unmerged \leftarrow unmerged - \{C_i, C_j\} + \{C_i \cup C_j\}$
-      \State $useful \leftarrow useful + \{C_i \cup C_j\}$
-      \If{$D(C_i, C_j) = 0$}
-        \State $useful \leftarrow useful - \{C_i, C_j\}$
-      \EndIf
-    \EndWhile
-    \State Order $useful$ based based on last merged first\label{fig-cluster-creation-ordering}
-    \State Remove largest cluster from $useful$
-
-    \Return $useful$
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm describing how LTGA creates clusters using Equation~\ref{eq-distance}
-           for a population. $unmerged$ and $useful$ are ordered sets of sets of gene loci.}
-  \label{fig-cluster-creation}
-\end{figure}
-
-Figure~\ref{fig-cluster-creation} provides the agglomerative method LTGA uses to create gene clusters.
-This algorithm creates a tree of sets using pairwise gene entropy, such that the leaves of the tree contain a single gene and
-internal nodes are the union of their children's sets. These sets are then used by crossover to specify epistatic
-relationships which should be preserved.
-The process begins by creating the set of sets $unmerged$ which tracks all top level clusters. Initially
-$unmerged$ contains single member sets for each gene. After each iteration the two sets with the minimum average pairwise
-distance (given in Equation~\ref{eq-distance}) are merged to create a single cluster. This process is repeated
-until only a single set remains in $unmerged$ which contains all of the genes in the genome.
-
-\begin{equation}
-  D(C_i,C_j) = \frac{1}{\left | C_i \right |\cdot \left |C_j \right|}\sum_{c_i \in C_i}\sum_{c_j \in C_j}
-  2 - \frac{H(c_i) + H(c_j)}{H(c_i \cup c_j)}
-  \label{eq-distance}
-\end{equation}
-\begin{equation}
-  H(c) = -\sum_{s\in S} p_c(s)\log(p_c(s))
-  \label{eq-entropy}
-\end{equation}
-
-Throughout this process $useful$ tracks the set of all gene clusters which should be preserved for use by crossover.
-This set begins with all genes in separate clusters, and each time a new cluster is created it is added to $useful$.
-However, not all clusters are necessarily worth keeping. For instance, in all versions of LTGA the cluster
-containing all genes is removed from $useful$ as preserving all genes during crossover can only create clones.
-\cite{thierens:2013:ltgahiff} extended this removal to include any unsupported subsets. If the pairwise distance
-between two clusters is 0, this means there are no individuals in the population which disrupt the relationships between the two
-clusters. Therefore when performing crossover there is no reason to believe a fitness improvement can be achieved
-by breaking the stored pattern. As such a cluster is only kept if its direct superset has a non-zero distance.
-As a final step Line~\ref{fig-cluster-creation-ordering} reorders $useful$ such that clusters appear in reversed
-order from which they were added to $useful$. Thus the most linked clusters and those containing single genes
-appear at the end of the returned list.
-
-\cite{thierens:2013:ltgahiff}'s version of LTGA does not use the entire population when determining pairwise entropy.
-Instead, binary tournament is used to select half of the population. This is done to ensure the model is built
-using only high quality solutions, even during the first generation.
-
-In order to efficiently perform clustering, a pairwise gene frequency table is constructed from the selected solutions.
-To calculate Equation~\ref{eq-distance}, Equation~\ref{eq-entropy} is called for each
-gene ($H(c_i)$) and pair of genes ($H(c_i \cup c_j)$). Extracting this information requires $O(\mu N^2)$
-time, where $\mu$ is the population size and $N$ is the genome size. The process of converting
-this pairwise frequency information into clusters can be achieved in $O(N^2)$ using the bookkeeping
-methods presented by \cite{gronau:2007:upgma}. This cost is performed only once per generation,
-and is then used to perform approximately $O(\mu N)$ crossover evaluations. As a result, the amortized cost of
-LTGA's model building is $O(N)$.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Cluster-Usage}{}
-    \ForAll{$C_i \in useful$}
-      %\ForAll{$d \in shuffled(P_i)$}
-        \State $d \leftarrow rand\_choice(P)$\label{fig-cluster-usage-donate}
-        \State Copy $d$'s gene values for $C_i$ into solution
-        \If{solution was changed}
-          \If{solution's fitness decreased}
-            \State Revert changes
-          \EndIf
-          %\State \textbf{break}
-        \EndIf
-      %\EndFor
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm describing how clusters are used to perform crossover.}
-  \label{fig-cluster-usage}
-\end{figure}
-
-Figure~\ref{fig-cluster-usage} describes how the identified clusters are used by crossover to preserve
-gene linkage while still exploring the search space. Unlike more traditional crossover methods, LTGA
-crosses each individual with the entire population. Also, to produce a single offspring, multiple evaluations
-of the fitness function are performed.
-
-Each generation, each individual in the population undergoes crossover. In a single crossover event, each
-cluster of genes $C_i$ in $useful$ is applied as a crossover mask. A random donor $d$
-is chosen from the entire population (not just the model selected population), and $d$'s gene's for $C_i$ are copied
-into the working solution. If a modification is made, an evaluation is then performed. If the crossover
-resulted in no worse fitness then the changes are kept. This allows for neutral drift across plateaus.
-The resulting solution, which must be at least as fit as its parent, is then
-copied into the next generation.
-
-In total each individual can cause up to $|useful|$ evaluations. If all clusters were kept, even those deemed
-unhelpful, and all donations were evaluated, even those which did not change any genes, then \Call{Cluster-Usage}{}
-would perform exactly $2N-2$ evaluations for each of the $\mu$ solutions in the population. This provides the amortizing evaluations
-required to make clustering only $O(N)$ operations per evaluation. However, by skipping some evaluations, its
-possible that clustering may be super-linear.
-
-LTGA has no explicit form of diversity control and has no method for introducing new genetic information once
-the population has converged. Therefore an LTGA run is considered converged once two consecutive populations
-contain the same unique solutions.
-
-By design, LTGA only has a single parameter: population size. LTGA uses no mutation, and crossover is defined
-in terms of the clustering algorithm. Selection between generations is fully elitist and embedded in the crossover,
-with selection of model building solutions fixed to a binary tournament. Neither \Call{Cluster-Creation}{} nor
-\Call{Cluster-Usage}{} rely on parameter values. LTGA does not provide any method for controlling or setting
-the population size, and uses a fixed user specified size for each generation.
-
-\section{Parameter-less Population Pyramid}
-\label{sec-p3}
-\cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid (P3) as a method for
-performing optimization which does not require the user to provide any parameters. This is
-achieved by combining efficient local search with the model building methods of LTGA using
-an iteratively constructed hierarchy of populations.
-
-The high level algorithm of P3 is presented in Figure~\ref{fig-p3}. Unlike more traditional
-GAs, P3 does not follow a generational model. Instead, it maintains an iteratively
-expanding pyramid of expanding populations. Each iteration, a new random solution is generated.
-This solution is brought to a local optimum using the hill climbing algorithm in Figure~\ref{fig-hc}. If that
-local optimum has not yet been added to the pyramid, the solution is added to the lowest
-population $P_0$.
-
-Next, the solution is iteratively improved by applying LTGA's crossover algorithm (Figure~\ref{fig-cluster-usage})
-with each population $P_i$ in the pyramid. If this process results in a strict fitness improvement and has
-created a solution not yet stored in the pyramid, it is added to the next highest pyramid level $P_{i+1}$.
-If $P_{i+1}$ does not yet exist, it is created. In this way populations in the pyramid expand over time,
-and the number of populations stored increases over time. Initially the pyramid contains no solutions
-or populations, meaning the user does not need to specify a population size.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Iterate-P3}{}
-    \State Create random solution
-    \State Apply hill climber
-    \If{solution $\notin hashset$}
-      \State Add solution to $P_0$
-      \State Add solution to $hashset$
-    \EndIf
-
-    \ForAll{$P_i \in pyramid$}
-      \State Mix solution with $P_i$
-      \If{solution's fitness has improved}
-        \If{solution $\notin hashset$}
-          \State Add solution to $P_{i+1}$
-          \State Add solution to $hashset$
-        \EndIf
-      \EndIf
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-  \caption{One iteration of P3 optimization. $pyramid$ is an
-           ordered set of populations and $hashset$ is a set
-           of all solutions in $pyramid$.}
-  \label{fig-p3}
-\end{figure}
-
-To accommodate P3's unique population structure, some of LTGA's clustering procedures were modified. In LTGA,
-clusters are identified at the start of each generation and are used
-to create all offspring in that generation. As P3 does not perform serial
-generations, P3 instead updates the model each
-time a solution is added to a population. Furthermore, unlike
-LTGA all solutions in the population are used to generate the model, not just the
-winners of a binary tournament. We can do this because even the
-worst solutions in the pyramid are already high quality due to local search.
-Using local search in LTGA was examined by \cite{bosman:2011:lsbbo} and found
-to provide no significant improvement. A likely cause was that that study applied
-local search to every solution, not just the initial population, resulting in significant overhead.
-
-Beyond the changes in population structuring, P3 modifies
-LTGA's version of \Call{Cluster-Creation}{} and \Call{Cluster-Usage}{}.
-P3 changes Line~\ref{fig-cluster-creation-ordering} in Figure~\ref{fig-cluster-creation}
-from \emph{last merged first} ordering to \emph{smallest first} ordering.
-This method applies gene clusters during crossover based on how many genes
-are in each cluster\footnote{Ties are broken randomly.}, and not on how tightly linked those genes are.
-\cite{goldman:2012:ltga} found that this alternative was better at preserving
-diversity, and therefore required smaller populations, but has not yet been integrated into
-canonical LTGA.
-
-P3 also modified Line~\ref{fig-cluster-usage-donate} in Figure~\ref{fig-cluster-usage}.
-Instead of choosing a single genetic donor for each cluster, P3 iterates over the
-population in a random order until a solution in the population is found which
-has a least one gene different for that cluster of genes from the improving solution.
-This process increases the likelihood of an evaluation being performed for every cluster,
-and helps test rare gene patterns in the population.
-
-In LTGA the cost of rebuilding the model is $O(\mu N^2)$ as it must collect pairwise
-gene frequency information for all $\mu$ solutions in the population. P3 does not store
-a single population, and it does not have a fixed $\mu$ size for any population. However,
-each time a solution is added to the population, it requires $O(N^2)$ time to update
-the table of pairwise frequencies and another $O(N^2)$ time to rebuild the linkage model.
-The model is then used immediately to perform up to one evaluation for each of the up
-to $2N-2$ clusters. Just as in LTGA, if no evaluation shortcuts were made, P3
-has an amortized cost of $O(N)$ modeling cost per evaluation. While P3 does
-rebuild the model more frequently per solution in the population, it also performs
-a number of local search evaluations which are quite efficient, meaning theoretical
-comparisons of their speed are difficult to perform. As a final note, P3's repeated
-attempts to find a useful donation make it less likely than LTGA to skip evaluations,
-but has an added cost to find these donations. Repeated donations could require as much
-as $O(\mu)$ attempts per evaluation, but experimental evidence suggest that this
-operation actually saves more overhead than it costs by increasing the number of evaluations
-per model rebuild.
-
-Each of the pieces of the P3 algorithm were selected not just for their stand alone efficacy,
-but for the ways in which they interact. By using the hill climber to ensure each solution
-is at a local optima, the underlying pairwise relationships in the problem are exposed. As
-a result, detecting clusters for use by crossover is much more effective. The crossover operator
-is extremely elitist, as each gene donation must result in no fitness loss, and a solution must
-strictly improve to be added to the next level of the pyramid. This is balanced by continual
-integration of new random solutions. Furthermore, each random restart decreases the probability
-of spurious linkages caused by shared ancestry. This diversity is further preserved by applying
-gene clusters in smallest first order during crossover as this reduces the probability of genetic
-hitchhikers.
-
-Other algorithms have proposed using multiple concurrent populations.
-\cite{hornby:2006:alps} had a hierarchy of populations with solutions periodically advancing
-upward. This allows for continuous integration of diversity as the lowest population is reseeded
-with random solutions. However, this method resulted in increased parameterization as not only was
-a population size required, but also parameters for how frequently generations advanced between levels and
-how many total levels to have were added. \cite{harik:1999:parameterlessga} used multiple independent
-populations of different sizes as a method for removing the population size parameter, but
-doing was provably less efficient than using an optimal population size as no information is shared
-between the populations.
-
-\section{Problem Descriptions}
-\label{sec-problems}
-
-\subsection{Single Instance Problems}
-Understanding how a stochastic search algorithm will behave on arbitrary and complex
-search landscapes can be exceedingly difficult. Therefore a common practice for
-algorithm understanding is to perform search on well defined, well understood
-landscapes. To be of interest these landscapes need to represent interesting
-and important aspects of real world problems.
-
-Perhaps the most common such landscape is the Deceptive Trap problem \citep{goldberg:1991:gasize}.
-In this landscape the genome is broken up into $k$ bit non-overlapping subproblems referred
-to as traps.
-Each subproblem is scored using Equation~\ref{eq-deceptive-trap}, where $t$ is the number of
-bits in the trap set to 1. The global optimum in each trap is a string of all 1s, while all
-other solutions lead to a local optima of all 0s. This problem tests an algorithm's
-ability to overcome $k$ sized deception and is commonly used to determine how effective crossover
-is at preserving building blocks. Any crossover event that mixes bits from different parents in
-the same trap will likely result in that trap being optimized to the local optima. For our experiments
-we chose $k=7$ to ensure highly deceptive traps.
-
-\begin{equation}
-   trap(t) = \left\{
-     \begin{array}{rl}
-       k-1-t, &  t<k\\
-       k,   &  t = k
-     \end{array}
-   \right.
-  \label{eq-deceptive-trap}
-\end{equation}
-
-\cite{goldman:2012:ltga} found that mixing local search with linkage learning rendered the Deceptive
-Trap problem trivial. This is because local search is able to optimize each trap to one of the two
-optima (all 1s or all 0s), and then linkage learning can have perfect knowledge of gene interactions.
-In order to make the problem more difficult, the authors proposed the Deceptive
-Step Trap problem, given in Equation~\ref{eq-deceptive-step-trap}. This function modifies the results
-of Equation~\ref{eq-deceptive-trap} to include plateaus of size $s$, introducing an exponential number
-of local optima in each trap. As a result the Deceptive Step Trap is much more challenging for linkage
-learning techniques, while still being highly deceptive. We chose to use $s=2$ to create the maximum
-number of local optima.
-\begin{equation}
-   step\_trap(t) = \left \lfloor \frac{(k-s)\pmod{s} + trap(t)}{s} \right \rfloor
-  \label{eq-deceptive-step-trap}
-\end{equation}
-
-Another challenging aspect of landscapes can be higher order relationships. The Hierarchical If
-and only If (HIFF) problem is designed to capture the difficulties of this class of problem.
-In HIFF the genome is broken up into a complete binary tree, such that each gene appears in exactly one
-leaf and each internal node is the subset of genes contained in its children. If all genes represented
-in a node of the tree are set to the same value, they score equal to the size of the set. In this way
-small subsets lead toward solutions to larger subsets. However, a node can score if all genes are either
-all 1s or all 0s, meaning that to solve higher order subproblems it is necessary to perform crossovers
-that preserve lower order solutions. This problem is a natural fit for LTGA as the linkage tree can
-perfectly duplicate the problem's true relationships \citep{thierens:2013:ltgahiff}.
-
-As a final class of well known problems, we have chosen to borrow the Rastrigin problem from real valued
-optimization. This problem's landscape, determined by Equation~\ref{eq-rast}, is highly multimodal caused by
-the oscillating cosine function. \cite{goldman:2014:p3} proposed the Discretized Rastrigin problem, such
-that each floating point $x_i$ in Equation~\ref{eq-rast} is encoded using a 10 bit gray code.
-\begin{equation}
-  An + \sum_{i=1}^{n}\left [ x_i^2-A\cos (2\pi x_i) \right ] \forall x\in [-5.12,5.12]
-  \label{eq-rast}
-\end{equation}
-
-\subsection{Randomly Generated Problem Classes}
-While well defined landscapes can provide specific insights into how an algorithm works,
-their static nature can be misleading. Specifically, algorithm quality might
-be very fragile such that it is only effective at searching well behaved landscapes. A more realistic
-test of an algorithm's black box effectiveness is to work with randomly generated instances
-drawn from a problem class. When tested over a sufficiently large sample it is then possible
-to draw more general conclusions about the algorithms effectiveness. The challenge with these
-landscapes is determining the global optimum to gauge if an algorithm was successful.
-
-Perhaps the most common model for generating random rugged landscapes is the NK model. An
-NK Landscape determines the fitness of each gene based on epistatic relationships with $K$
-other genes in the genome. This fitness is specified using a randomly generated table of
-fitness values, were each possible combination of the $K+1$ genes is mapped to some floating
-point value $[0-1]$. In unrestricted NK landscapes the relationships between genes are also
-randomly chosen and as a result finding the global optimum is $NP$-Hard for $K>1$. However,
-if epistasis is set such that each gene depends on the $K$ directly following it
-in the genome the solution can be found in polynomial time~\citep{wright:2000:solvingnk}.
-These Nearest Neighbor NK landscapes are therefore ideal for search algorithm testing.
-For all of our experiments we fixed $K=5$ to ensure highly rugged landscapes.
-
-\cite{saul:1994:spinglass} presents a combinatorial benchmark problem derived from physics:
-Ising Spin Glasses. A spin glass is defined by a weighted graph of interaction terms between vertices.
-Each gene assigns a value to each vertex, with the fitness calculated by Equation~\ref{eq-ising}.
-In this equation, $E$ is the set of all edges, $e_{ij}$ is the edge weight connecting vertex $i$ to vertex $j$, and
-$x_i$ and $x_j$ are the gene values for vertex $i$ and $j$. Optimal fitness is when this sum is minimized.
-\begin{equation}
-\sum_{e_{ij} \in E} x_ie_{ij}x_j
-  \label{eq-ising}
-\end{equation}
-Similar to NK Landscapes, the general class is $NP$-Hard to optimize, but the $2D\pm J$ subset of
-Ising Spin Glasses can be polynomially solved.\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}
-In this subset the graph is restricted to be a two dimensional torus, edge weights are randomly set
-to either -1 or 1, and vertex values must be -1 or 1.
-
-As our final class of randomly generated problems we chose the Maximum Satisfiability (MAX-SAT) problem.
-Related to the more common 3-SAT problem, a MAX-SAT instance is defined by a set of three term clauses.
-Each term is a randomly chosen variable, which may also be negated. A clause scores if an and only if
-at least one term in the clause evaluates to true. In order to make MAX-SAT instances with a known global
-optimum, \cite{goldman:2014:p3} proposed constructing clauses around a fixed solution. In this way the
-signs of the terms are set to ensure the target solution satisfies the clause. To ensure each
-problem is challenging we chose a clause to variable ratio of $4.27$.
-
-\section{Comparison Algorithm Parameter Tuning}
-\label{sec-tuning}
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-hboa}
-      \end{centering}
-      \caption{Tuned Population Sizes for hBOA}
-      \label{fig-pop-hboa}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-ltga}
-      \end{centering}
-      \caption{Tuned Population Sizes for LTGA}
-      \label{fig-pop-ltga}
-    \end{subfigure}
-  \end{centering}
-  \caption{Optimal population sizes found using bisection on each size
-           for each problem.}
-  \label{fig-pop-sizes}
-\end{figure}
-
-While four of the six algorithms in our experiments do not require any user specified parameters,
-hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
-handicapped, we extensively tuned each using the bisection method \citep{sastry:2001:bisection}
-to determine the optimal population size for each problem size.
-Extended by \cite{goldman:2012:ltga}, this method iteratively
-doubles the population size until some success criteria is met and then performs bisection
-between the lowest successful and highest unsuccessful sizes. In this way the minimum population size
-which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
-of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
-above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
-do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
-As bisection can make infinitely large population sizes, any run which had not found the global
-optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.
-
-Figure~\ref{fig-pop-sizes} shows the results from performing bisection. In general hBOA
-required population sizes which were at least an order of magnitude larger than LTGA. Due to
-runtime and memory overhead, finding the optimal value for hBOA was much less tractable than LTGA
-for moderate to large problem sizes. LTGA's population size also grew significantly slower
-than hBOA's as problem size increased, especially on the two Trap problems and Rastrigin.
-Both algorithms were very ineffective on MAX-SAT, neither able to tune to problems sizes over 60 bits.
-This is likely due to the fact that some randomly generated MAX-SAT landscapes are very flat
-and highly deceptive~\citep{rana:1998:gamaxsat}.
-\section{Finding the Global Optimum}
-\label{sec-optimum}
-
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{evals-to-success}
-  \end{center}
-  \caption{Comparison of the median number of evaluations to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-evals-to-success}
-\end{figure}
-
-Figure~\ref{fig-evals-to-success} shows the median number of evaluations required by each of the six
-algorithms to find the global optimum for multiple sizes of each problem.  Each data point in
-Figure~\ref{fig-evals-to-success} represents the median of 100 runs, where unsuccessful runs
-are treated as requiring more evaluations than any successful run. If the median run was not successful
-no point is shown. The maximum problem size used for each problem was set to be the largest size we could
-feasibly determine LTGA's optimal population size. For many larger problems results are not shown
-for hBOA due to the extreme computational cost required to optimally set the population size.
-
-\subsection{Quantitative Comparison}
-Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
-The largest problem size for any problem where P3 was not the most efficient has 49 bits, with P3 achieving the best results on all 92 larger
-configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
-Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
-and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
-on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
-
-\subsection{Local Search}
-The Random Restart Hill Climber and $1+(\lambda, \lambda)$ are both relatively effective on small problem
-sizes. This is especially true for the three randomly generated problem problem classes. These problems
-may contain relatively few local optima or just be exceptionally difficult for the model based algorithms.
-On Deceptive Trap and Deceptive Step Trap using 4 or less traps, $1+(\lambda, \lambda)$ performs significantly
-better than any other algorithm. We believe this is
-because $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
-flipping entire traps.
-This ability also leads $1+(\lambda, \lambda)$
-to outperform the Random Restart Hill Climber on all problems except Nearest Neighbor NK.
-
-On larger problem sizes, the ability for local search to reach the global optimum quickly diminishes.
-Only on MAX-SAT are these optimizers competitive
-at larger tested problem sizes. However, we believe this is because the largest tested MAX-SAT
-was an order of magnitude smaller than the largest size tested for most other problems. As the problem
-size increases the number of local optima increases exponentially, which explains why Random Restart Hill
-Climbing was unable to scale.
-For larger problems it also becomes increasingly unlikely for $1+(\lambda, \lambda)$ to make the right combination
-of changes required to reach the global optimum. This
-behavior causes high variance in success rate, as evident by the occasional successes on large Deceptive Trap problems.
-
-\subsection{Model Building}
-Only techniques which explicitly built models of gene epistasis were able to solve the largest problem
-instances. On single instance problems LTGA was more effective than hBOA, with hBOA outperforming
-LTGA on Nearest Neighbor NK and Ising Spin Glasses. This may be caused by the differences in modeling method:
-unlike the single instance problems, gene epistasis in the randomly generated
-problem classes cannot be be perfectly represented with a linkage tree.
-
-Considering how different hBOA and LTGA are in performing optimization, it is somewhat surprising how similar
-their results are on HIFF.
-However, both techniques rely on populations large enough to support the diversity required to reach the global optimum
-and to model epistasis. Both techniques also only rebuild models once per generation. As the subproblems of HIFF
-are nested, it is unlikely that either technique can accurately model higher order epistasis before solving
-lower order subproblems. Therefore both methods require one generation per subproblem order.
-
-\subsection{P3}
-Unlike the other model based methods, P3 generally outperforms
-both Random Restart Hill Climber and $1+(\lambda, \lambda)$ even on small problem sizes.
-Unlike the other local search methods, P3 outperforms LTGA and hBOA even on large problem sizes.
-This implies that P3 is gaining the benefits of each, leveraging local search to solve easy
-problems and model building to solve harder ones.
-
-Furthermore, the interaction between these two optimization tools explains some of the reason P3
-outperforms each method alone.
-On Deceptive Trap P3's use of hill climbing ensures all traps are immediately optimized, allowing for perfect
-linkage detection and high quality donation.
-On HIFF local search solves all pairwise subproblems, saving P3 a generation over LTGA and hBOA.
-In comparison P3 is only a slight improvement on Deceptive Step Trap, which is less amenable to local search.
-
-\section{Fitness Over Time}
-\label{sec-overtime}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{fitness-over-time}
-  \end{center}
-  \caption{Compares the median best fitness reached during search for each of the six optimization methods.}
-  \label{fig-fitness-over-time}
-\end{figure}
-
-For some applications, finding the global optimum is less important than finding good solutions quickly. Therefore
-we examine this behavior in Figure~\ref{fig-fitness-over-time}. At regular intervals during optimization
-Figure~\ref{fig-fitness-over-time} shows the median of the best fitnesses found at that point of search across
-100 runs. For each problem we show the largest problem size for which we were able to successfully gather results
-for all six algorithms, but the trends shown are representative of all larger problem sizes. The maximum reporting interval
-is set to include the slowest P3 run to reach the global optimum.
-
-% 181 total
-% 121 has P3 best
-%   9 hc
-%  50 lambdalambda
-%  20 hboa
-%  18 phboa
-%  27 LTGA
-
-\subsection{Quantitative Comparison}
-Of 181 sample points, P3 had the highest median fitness in 121. In pairwise competition, $1+(\lambda, \lambda)$
-was the most likely to outperform P3, doing so on 50 sample points. LTGA, hBOA, and Parameter-less hBOA
-were the next best, outperforming P3 on 27, 20, and 18 sample points. Random Restart Hill Climbing almost
-never outperformed P3, doing so only 9 times.
-
-\subsection{Local Search}
-Perhaps the most striking result is the quality of $1+(\lambda, \lambda)$. Until quite far into search this
-method performs better than both LTGA and hBOA. Given sufficient evaluations $1+(\lambda, \lambda)$ also outperforms
-Random Restart Hill Climbing on all 7 problems. For brief periods in the middle of search it performs the
-best of all techniques on: Deceptive Trap; Deceptive Step Trap; HIFF; Ising Spin Glass; and MAX-SAT problems.
-$1+(\lambda, \lambda)$'s ability to efficiently incorporate gene modifications of larger
-than one bit allows it to overcome the deception and plateaus in Deceptive Trap and Deceptive Step Trap, solve medium sized subproblems
-in HIFF, flip the signs on multiple adjacent bits in Ising Spin Glass, and cross plateaus in MAX-SAT. However,
-this method is slow in reaching the global optima in many of these problems which causes it to eventually
-be overtaken by the model building techniques.
-
-\subsection{Model Building}
-Both hBOA and LTGA are marked by periods of little improvement followed by rapid improvement.
-In hBOA this is taken to the extreme, with all fitness
-improvement made at the very end of search. In both cases this is caused by model building. Before the model
-is accurate little improvement is made. Once it is accurate, fitness improves dramatically.
-
-On every problem LTGA has five distinct periods: fitness plateau, near instantaneous improvement, fitness plateau,
-and improvement to global optimum. The early period corresponds with
-initialization of the population, with the first fitness gain achieved immediately upon completing the first generation.
-When using an inaccurate model, LTGA's mixing strategy performs a sort of less effective local search.
-Subsequent generations then make only minor fitness improvements. Once the model
-becomes accurate and the probability of a crossover using high quality genetic material increases sufficiently, LTGA
-enters a second period of rapid improvement.
-
-\begin{comment}
-This behavior is easiest to understand on the two Trap problems. LTGA's first generation will push individual traps toward a
-local optima, but it requires more than a single generation to do so completely. Until that occurs, the model is likely behaving
-no better than random as there is little apparent linkage between bits. Once the model begins to identify individual traps, crossover
-can begin to increase the frequency of higher fitness trap settings. This process begins slowly as low fitness local optima are more
-likely to be chosen for donation than high fitness local optima due to their frequency in the population. The process becomes
-self-catalysing as increased frequency of the global optimum trap genes means increased likelihood of the global optimum being spread by crossover.
-\end{comment}
-
-\subsection{P3}
-The integration of hill climbing into P3 makes it strictly better than using hill climbing alone.
-Early in optimization P3 and the Random Restart Hill Climber have effectively identical quality. This is because
-P3 performs the same evaluations as the Hill Climber for the first two restarts. Once P3 begins
-performing crossover it immediately improves over the Hill Climber. In 95\% of intervals P3 had a fitness at
-least as high as Hill Climbing. As such P3 is better than a simple
-hill climber regardless of how long each technique is run and irrespective of how high quality the solution
-found has to be.
-
-Unlike the model based methods, which struggle until model accuracy improves, P3's iterative solution
-integration allows it to improve much more quickly. This behavior exists in most problems, but is easiest to understand on Deceptive Trap.
-On this problem, P3 immediately brings all traps to local optima, equaled only by the Random Restart Hill Climber in quality.
-In comparison LTGA must evaluate the entire population and perform multiple generations to reach similar quality.
-P3 is able to immediately integrate optimal versions of each
-trap into a single individual as they are found by local search, resulting in smoother fitness improvement than LTGA.
-
-\section{Computational Expenses}
-While it is common in evolutionary computation to assume the evaluation function will dominate algorithm
-complexity, in some domains this will not be true. Model based methods are especially likely to violate
-this norm. Therefore, in order to assess P3's quality in solving problems with efficient fitness functions,
-we provide data on both its algorithmic complexity and wall clock time.
-
-\subsection{Operation Counting}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{rebuilds}
-      \end{centering}
-      \caption{Rebuilds}
-      \label{fig-rebuilds}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{donations}
-      \end{centering}
-      \caption{Donations}
-      \label{fig-donations}
-    \end{subfigure}
-  \end{centering}
-  \caption{Computation costs incurred by model rebuilding (Figure~\ref{fig-rebuilds}) and
-           repeated donations (Figure~\ref{fig-donations}) per evaluation as problem size increases.}
-  \label{fig-costs}
-\end{figure}
-
-When discussing the asymptotic complexity of P3 in Section~\ref{sec-p3}, two aspects eluded precise
-analysis: how expensive is model rebuilding and how many gene donations are made. Figure~\ref{fig-costs}
-provides some insight into how often these two aspects of the algorithm are utilized.
-
-Figure~\ref{fig-rebuilds} reports in an algorithmic sense how expensive model rebuilding is for search.
-In order to calculate this value we recorded how many times search rebuilt the model during each run.
-Under the minimal assumption that an evaluation costs $\Omega(N)$ time and that model rebuilds cost $O(N^2)$,
-Figure~\ref{fig-rebuilds} shows the ratio cost of model rebuilding per evaluation. If the cost of model building
-scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically constant.
-For Nearest Neighbor NK, Ising Spin Glasses, and Rastrigin this is the case. For both Trap problems
-and HIFF there is slow growth in the ratio. The problem sizes used for MAX-SAT were not sufficient
-to accurately gauge the asymptotic behavior. Together this suggests that while the cost of building the model
-is almost linear per evaluation, it can grow slowly. However, even in the worst case (HIFF) this growth was
-no more than twice the algorithmic cost of an evaluation even using 2048 bits.
-
-When applying a crossover subset, P3 tries random donors from the population until one is found with at least
-one bit different from the improving solution. In theory this can result in up to $O(\mu)$ operations.
-Figure~\ref{fig-donations} examines the observed average number of donations per evaluation performed.
-Ising Spin Glass, HIFF, and Rastrigin all achieve effectively constant behavior here, implying repeated
-donation does not impact the asymptotic runtime of P3. Both Trap functions and Nearest Neighbor NK all
-increase in number of donations as problem size increases, potentially increasing algorithmic costs. An
-important note is that each donation may range in size from a single bit up to $N-1$, with smaller
-donations far more likely to result in repeated attempts. As such this may cause some super-linear growth
-in P3, but its unlikely to be very high.
-
-\subsection{Wall Clock Performance}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{seconds-to-success}
-  \end{center}
-  \caption{Comparison of the median number of seconds to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-seconds-to-success}
-\end{figure}
-
-To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}.
-Similar to Figure~\ref{fig-evals-to-success}, each point represents the median
-of 100 runs, with unsuccessful runs treated as slower than successful runs.
-In total these runs
-took over 6 computing years to complete. We were therefore forced to use
-diverse hardware potentially running multiple processes in parallel.
-As such these timings should be considered very skeptically as a number
-of sources may contribute to biased error in the results. However, due to
-the size of the dataset and the general regularity in the results some general conclusions
-can still be made.
-
-\subsubsection{Model Building}
-hBOA and Parameter-less hBOA perform significantly worse when using wall clock time as the unit of comparison than when using
-evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while, under reasonable
-assumptions, P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
-hBOA goes from being slightly more efficient than LTGA in terms of evaluations to significantly worse in terms of seconds.
-As P3 and LTGA require a similar asymptotic complexity per
-evaluation as the Hill Climber and $1+(\lambda, \lambda)$, no similar change in ordering occurs.
-
-\subsubsection{P3}
-Any increase in P3's algorithmic complexity is offset by its efficient use of evaluations.
-In aggregate P3 found the global optimum the fastest on 72 of 130 configurations,
-with LTGA, $1+(\lambda, \lambda)$, Hill Climber, hBOA outperforming P3 on 44, 18, 10, and 1 configurations, respectively.
-On
-Deceptive Trap, Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT P3 still finds the global
-optimum faster than all comparison techniques. LTGA is able to match P3's performance almost exactly on HIFF,
-and on Rastrigin P3 falls behind only on moderate sized problems. Deceptive Step Trap yields the worst performance,
-with LTGA using optimal population sizes having a significantly lower growth curve than P3. This mirrors
-the results from Figure~\ref{fig-evals-to-success} under the assumption that P3's computational time per evaluation is higher
-than LTGA but by no more than a constant factor.
-
-\section{Population Sizing}
-A major advantage to P3 is that it does not require the user to set a population size parameter.
-Beyond making P3 easier to apply, this also conveys two additional advantages: diversity scaled
-to initialization and no need to sacrifice intermediate fitness for eventual optimality.
-
-\subsection{Problem Instance versus Problem Class}
-Our procedure for tuning LTGA and hBOA outlined in Section~\ref{sec-tuning} involved finding
-the optimal population size for each class of problem. For real world black box optimization this is
-realistically the best either algorithm could hope for as tuning to a problem
-instance or population initialization involves repeatedly solving the problem being tuned.
-This limitation does not exist in parameter-less methods, which scale their diversity
-based on the problem instance without needing to solve that instance repeatedly.
-
-To achieve high success rates on randomly generated problem classes, LTGA and hBOA must use a population size
-large enough to solve the hardest instances in that class. Therefore these methods will have
-population sizes larger than necessary to solve the easiest instances in the class. Even
-on single instance problems, both methods will require population sizes large enough to
-ensure the worst random initialization is diverse enough to solve the problem, which may be
-much larger than the best random initialization.
-
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{evals-to-success-range}
-  \end{center}
-  \caption{Comparison of the upper and lower quartiles of evaluations required
-           to reach the global optimum for P3 and LTGA with respect to problem size.}
-  \label{fig-evals-to-success-range}
-\end{figure}
-
-Figure~\ref{fig-evals-to-success-range}
-highlights how this can effect the required number of evaluations to reach the global optimum, showing
-the upper and lower quartiles for P3 and LTGA. On each problem
-LTGA has a much smaller difference between its best and worst runs. This makes sense as LTGA uses the same
-population size regardless of instance and progresses search generationally. In contrast P3 has a much
-higher split, with many runs finishing very quickly. This added variability is also generally on the lower
-quartile, with P3's upper quartile only worse than LTGA's on Deceptive Step Trap and Nearest
-Neighbor NK. This supports the hypothesis that P3 is able to scale its diversity not just
-to the problem class, but to the problem instance or even problem initialization, something
-wholly infeasible for tuned population sizing to do.
-
-This tuning distinction is also apparent when comparing Parameter-less hBOA with hBOA in Figure~\ref{fig-evals-to-success}.
-While generally performing worse than hBOA, the difference between the two algorithms is smallest
-on randomly generated problem classes. On MAX-SAT, Parameter-less hBOA actually outperformed both hBOA and LTGA,
-likely due to its ability to scale diversity to the problem instance instead of the entire problem class.
-
-\subsection{Fast versus Optimal}
-\label{sec-fast-vs-optimal}
-
-In Section~\ref{sec-overtime} we examined intermediate fitness qualities of LTGA and hBOA when using
-population sizes tuned to reach the global optimum. As a result, both were exceptionally ineffective
-at quickly reaching high quality solutions. This is because unlike P3, these methods have an explicit
-trade off between optimal performance and intermediate performance caused by their population size parameter.
-
-At 58\% of the recording intervals hBOA has the worst fitness of any solver. Most of the exceptions
-occur when hBOA is still evaluating its initial population, allowing this random search to temporarily surpass the
-local search methods. After $N$ evaluations, however, hBOA and LTGA both fall behind until their models begin
-to improve.
-Parameter-less hBOA reaches intermediate fitnesses faster than hBOA, doing so on 62\% of intervals, as its models begin
-to optimize earlier than hBOA.  However, this trend is reversed after a sufficient number of evaluations, most clearly
-on Deceptive Step Trap and Ising Spin Glasses, as hBOA's tuned population overtakes Parameter-less hBOA's parallel populations.
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{small-pop-dst}
-      \end{centering}
-      \caption{Deceptive Step Trap 203}
-      \label{fig-small-pop-dst}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{small-pop-nk}
-      \end{centering}
-      \caption{Nearest Neighbor NK 200}
-      \label{fig-small-pop-nk}
-    \end{subfigure}
-  \end{centering}
-  \caption{Comparison of how reducing LTGA's population size effects
-           the median best fitness reached during search.}
-  \label{fig-small-pop}
-\end{figure}
-
-Figure~\ref{fig-small-pop} examines the effect of population size on LTGA's intermediate fitness by examining the effect
-of reducing LTGA's population size to one tenth of the tuned value. The two problems shown are representative of the behavior of using
-a smaller population size on the other five problems. Reducing the population size caused LTGA to improve earlier but plateau
-at lower fitnesses. This caused LTGA's success rate to drop from 100 to 0 on Deceptive Step Trap and from 98 to 68 on Nearest
-Neighbor NK.
-Even when using a reduced population size for LTGA, P3 still achieved a fitness at
-least as high as LTGA at 80\% of intervals.
-
-\begin{comment}
-It is important to note that hBOA and LTGA were tuned specifically
-to find the global optimum and are not necessarily using the optimal population size for finding intermediate fitnesses. Consider
-that for both reducing the population size would almost certainly improve their fitness for evaluation steps less than their population
-size. To examine the effect of population size we tested LTGA using one tenth of the population size required for reliably finding the
-global optimum, with the results given in Figure~\ref{fig-small-pop}.
-Reducing the population size resulted in LTGA's initial fitness improvement
-occurring earlier, as the first generation is completed much more quickly. On Deceptive Step Trap this means that
-LTGA leaps ahead of P3 due to its ability to overcome the two bit fitness plateaus.
-%
-Across all tested problems using the smaller
-population size reduced the quality of LTGA's first fitness plateau, likely due to missing required diversity to reach higher quality.
-The second period of improvement also comes earlier and reaches a lower quality fitness. On Deceptive Step Trap 0
-runs reach the global optimum, down from 100 successful for the full population size. On Nearest Neighbor NK this drops the number
-of successful runs from 98 to 68. As a result we conclude that while reducing the population size of LTGA may improve its quality
-at early points during optimization, doing so reduces the robustness of the final solution found. This is in contrast to P3
-which balances high quality intermediate fitness without trading away eventual optimality.
-\end{comment}
-
-\section{Inner Workings}
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{cross}
-      \end{centering}
-      \caption{Crossover Proportion}
-      \label{fig-cross}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{cross-success}
-      \end{centering}
-      \caption{Crossover Success}
-      \label{fig-cross-success}
-    \end{subfigure}
-  \end{centering}
-  \caption{For each problem Figure~\ref{fig-cross} shows the proportion of P3 evaluations spend on crossovers
-           and Figure~\ref{fig-cross-success} shows the percentage of fitness improving crossover evaluations.}
-\end{figure}
-
-While analysis of optimization speed is useful from a practitioner standpoint, doing so provides very
-little insight into algorithm behavior. To better understand how P3 works in detail we present here
-a look at some internal features specific to P3.
-
-\subsection{Crossover}
-Figure~\ref{fig-cross} shows the proportion of evaluations P3 spends on crossover, as opposed to hill climbing,
-and Figure~\ref{fig-cross-success} shows what percentage of crossover evaluations resulted in a fitness improvement.
-Together these figures provide some insight into the role of crossover within P3. The behavior
-for each is clearly problem dependent and generally asymptoticly stable as problem size increases.
-
-When solving problems where epistasis can be effectively detected and represented by a linkage tree, P3 tends to spend
-less evaluations performing crossover and each crossover is more likely to be successful.
-Deceptive Trap and Rastrigin are the easiest problems to model epistasis, with local search quickly
-reducing pairwise entropy in each. These are also the problems where P3 uses the least evaluations
-on crossover and has the highest success rates for crossover. At the other extreme are Nearest Neighbor
-NK and Ising Spin Glasses, which both have overlapping linkage not representable by a linkage tree.
-These problems have the highest crossover usage and lowest crossover success of any problem except Deceptive
-Step Trap.  While Deceptive Step Trap's epistasis can be accurately modeled by a linkage tree, the
-exponential number of plateaus makes detecting gene linkage very challenging.
-
-\begin{comment}
-P3 performs optimization using a mixture of hill climbing and crossover. Exactly how much time the algorithm
-dedicates to each is not clear from the algorithm description alone. As such, Figure~\ref{fig-cross} shows
-that the relationship between the two is strongly problem dependent but relatively independent of problem size.
-In general, problems which can be more accurately represented by crossover
-required less crossover evaluations. Deceptive Trap and Rastrigin spent the least evaluations on crossover, using
-15\% and 20\% respectively on the largest problem sizes. Both of these problems are perfectly separable and have relatively
-few local optima in each subproblem. HIFF and Ising Spin Glasses use crossover for about 60\% of evaluations. The HIFF structure can
-be perfectly learned by the linkage tree, but the higher order relationships require multiple stages of crossover to optimize.
-Ising Spin Glasses have an overlapping structure, making them impossible to perfectly represent in a single linkage tree, but
-that overlap is relatively minor in comparison with Nearest Neighbor NK, which uses 81\% of evaluations on crossover. Deceptive Step Trap spends the highest
-percentage of evaluations on crossover, 92\%. Unlike all of the other problems, the value taken by an individual bit in Deceptive
-Step Trap has very little meaning. This is because any trap value is either a local optima or 1 bit flip away from being a local optima.
-This has a combined result of meaning it takes very few local search evaluations to find a local optima but very many crossover operations
-to proliferate the global optima. Epistatic relationships are also going to be very weak and even small population sizes are likely
-to mark all clusters as $useful$.
-
-Solutions are only added to the next highest pyramid level if at least one crossover donation resulted in a fitness
-improvement. Figure~\ref{fig-cross-success} examines the percentage of crossover donations that resulted in a fitness
-improvement. The ordering in Figure~\ref{fig-cross-success} is an almost perfect inversion of Figure~\ref{fig-cross},
-implying that the more successful crossover is at achieving fitness improving donations the lower the proportion of evaluations P3 will
-spend performing crossover. P3 achieves an impressive 30\% success rate for crossover donations on the largest Deceptive Trap
-problem sizes. This makes sense as crossover is able to immediately and perfectly identify entire traps such that all donations
-always contain only complete traps. Furthermore, by forcing crossover to attempt donations of genes different from the current solution,
-donations containing a single trap will have a very high success rate, always improving suboptimal traps which are far more frequently
-created by local search that optimal traps. Rastrigin and HIFF have relatively high success rates of 11\% and 7\%, respectively. Again
-this is likely due to accurate modeling and relatively low diversity of building block alternatives.
-Ising Spin Glasses and Deceptive Step Trap both have low success rates near 0.01\%. This follows as both are difficult to model with
-a linkage tree. In comparison to Figure~\ref{fig-cross-success} Deceptive Step Trap performs better than expected for crossover.
-However, the reason for this is that even exceedingly uninformed crossover can make improvements to Deceptive Step Trap as
-flipping two bits in the same trap will move between two local optima. Nearest Neighbor NK had the least successful crossovers,
-with only 0.007\% successful on the largest tested problems. In comparison to Ising Spin Glass, NK's overlap is far more extensive.
-As a result it makes sense that individual crossovers are very unlikely to be fitness improvements.
-\end{comment}
-
-When compared with LTGA, P3's crossover success rates are lower but similar in problem ordering.
-Counterintuitively, the use of hill climbing on the initial population reduces
-P3's crossover success, not because it reduces model quality or the donation pool, but because it is much more challenging to improve
-locally optimal solutions than randomly generated ones. LTGA's crossover benefits from application to unoptimized solutions, which
-makes its aggregate crossover success incomparable to P3's.
-
-Even when crossover success rates are quite low, such as Nearest Neighbor NK's 0.007\% success,
-the results from Section~\ref{sec-optimum} and Section~\ref{sec-overtime} show it's still critical to optimization. Without
-crossover, P3's performance would be identical to the Random Restart Hill Climber, which was unable to solve even moderately sized problems
-and quickly fell behind P3 in intermediate fitness quality. Therefore even infrequently successful crossover donations are
-critical to success. This does, however, suggest a potential avenue for future improvement by using more successful
-modeling and donation algorithms.
-
-\subsection{Pyramid}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{level-size}
-      \end{centering}
-      \caption{Population Size}
-      \label{fig-level-size}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{level-success}
-      \end{centering}
-      \caption{Crossover Success}
-      \label{fig-level-success}
-    \end{subfigure}
-  \end{centering}
-  \caption{For each problem Figure~\ref{fig-level-size} shows the number of solutions stored in each level of the pyramid
-           and Figure~\ref{fig-level-success} shows the percentage of fitness improving crossover evaluations at each level.}
-\end{figure}
-
-Another feature unique to P3 is the shape and size of the population pyramid
-constructed for each problem. Figure~\ref{fig-level-size} shows the number of solutions stored at
-each level of the pyramid for the largest tested problem sizes. Each point is the
-median size across 100 runs. If a run did not store any solutions at a level it is treated as 0.
-No point is drawn if the median run had 0 solutions stored at that level. While pyramid size
-is effected by problem size, the overall shape is not. As such the behavior shown in Figure~\ref{fig-level-size}
-is representative of that for all other tested problem sizes.
-
-With the exception of the dip in Deceptive Step Trap, all of the pyramids show a monotonic reduction
-in size as the level increases. This is because a solution must be a strict fitness improvement
-over its previous version to be added to a higher level, which becomes less likely
-each time the solution improves. \cite{lobo:2011:dynamicpop} found theoretical evidence
-and \cite{goldman:2011:dynamic-parameters} found empirical evidence that the optimal
-population size decreases each generation of traditional evolutionary search.
-By decreasing in size, P3 implicitly stores more diversity in low levels and focuses
-search on high quality solutions at higher levels. In comparison,
-LTGA and hBOA suboptimally use a fixed population size at each generation.
-
-Figure~\ref{fig-level-success} examines how crossover success changes at different
-levels of the pyramid. At low levels, success gets progressively lower
-as solution quality increases faster than the model's ability to improve solutions.
-At higher levels modeling becomes more accurate and donations contain higher frequencies
-of high quality building blocks, resulting in increased crossover success. The
-highest level of most problems has a low crossover success rate, as solutions
-crossing with that level have already been improved by previous operations to
-the point where the only improvement would be to create the global optimum, which can
-only happen once.
-
-\subsubsection{Deceptive Step Trap}
-The number of solutions stored at the fourth level of Deceptive Step Trap is
-significantly lower than that of the third or fifth levels, breaking the decreasing
-trend of the other six problems. Figure~\ref{fig-level-success} has a similar
-aberration, with crossover success dropping to  0.0004\% before rebounding and
-following the more common trajectory. This behavior exists in all other problem
-sizes tested, with the dips occurring at exactly the same level.
-
-This behavior is rooted in the peculiar nature
-of this landscape.
-After local search, all traps in all solutions have a total
-number of 1 bits equal to \{0, 1, 3, 5, 7\} as these correspond to the local
-optima when using $k=7$ and $s=2$. Crossover is very likely to overcome the
-two bit plateaus, and as a result solutions in the second level generally do not
-contain the lowest fitness local optima (5 bits set) and the third
-level has very few traps set to the next worst local optima (3 bits set). As a result,
-solutions which reach the third level can only be improved by replacing the deceptive
-local optima (0 and 1 bits set) with the global optimum (7 bits set).
-The global optimum is very rare in the population, and with 8 ways to represent local optima
-linkage learning is inaccurate. Therefore it is very unlikely for crossover to be successful,
-meaning few solutions will be added to the fourth level. Solutions that do improve by
-definition must have a higher frequency of optimal trap settings, meaning level four's model
-will be more accurate and donations are more likely to contain optimal trap values.
-Thus the level size and crossover success rates increase after contracting around level four.
-
-\section{Conclusions and Future Work}
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing black box
-optimization. P3's primary innovation is the replacement of the generational model with a pyramid of populations.
-This pyramid is constructed iteratively, with both the number of levels and the number of solutions stored
-at each level growing as search progresses. P3 uses a model based crossover method
-which learns a linkage tree from gene epistasis. Combined with a simple hill climber, P3's design contains
-many synergistic features.
-
-Across a large number of problems and problem sizes P3 required less evaluations to reach the global optimum
-than optimally tuned state-of-the-art competitors. On single instance problems P3's improvement was by a
-constant factor, while for the three problem classes P3's improvement increased with problem size. This
-quality extends to intermediate points during evolution, with P3 generally reaching at least as high
-of fitness as the competitive techniques when using the same number of evaluations. While P3 does require
-modeling overhead, the expense of this overhead is approximately linear with respect to genome size. There
-is some evidence that even when compared on wall clock time, P3 performs at least as well as
-the best comparison techniques. All of these achievements are made without any problem specific parameter
-tuning, making P3 easier to apply to new domains than its two closets competitors in quality.
-
-P3's quality is due to a number of desirable traits. First, mixing local search with model based
-crossover lets search focus on properly mixing high quality solutions. Second, by adding diversity only
-as necessary P3 tends to use the minimal amount of random initialization, unlike other techniques which must
-overcompensate with larger population sizes on single instance problems and consider the worst instance
-when solving problem classes. Third, by heavily exploiting existing diversity before adding more P3 is able
-to reach high quality intermediate fitnesses quickly without prematurely converging. Fourth, the very
-nature of the pyramid's shape allows search to preserve a desirable proportion of diversity at
-each fitness level, similar to a generational model using a decreasing population size.
-
-There are a number of meaningful avenues for future P3 experimentation. Perhaps the most
-pressing for practitioner acceptance is to apply P3 to real world problems and compare
-its results with other black box or even problem specific heuristics. While parameter-less,
-P3 is currently limited to discrete, fixed length genomes evaluated using single objective
-fitness. These limitations can be relaxed with future work to make P3 more widely applicable.
-While asymptotically linear, P3's modeling techniques and local search methods are likely going
-to be prohibitively expensive for genome sizes in the hundreds of thousands or millions of genes,
-and the inability of the model to capture overlapping linkage may be hindering
-search efficiency. Overcoming these limitations by using a new modeling technique may allow
-the pyramid model even greater flexibility. Similarly, while P3 is able to overcome low
-order deception via linkage learning, the iterative improvement method by which crossovers
-are made may mislead search on landscapes with higher order deception.
-
-However, even without
-these improvements our results show P3 is highly efficient at finding
-global optima on black box problems without any problem specific tuning.
-
-\small
-
-\bibliographystyle{apalike}
-\bibliography{../main}
-
-\normalsize
-\appendix
-\section{hBOA Simplification}
-\label{sec-appendix}
-To measure the quality of a decision forest, hBOA applies Equation~\ref{eq-hboa}.
-To favor compact
-models, Equation~\ref{eq-hboa} is scaled by Equation~\ref{eq-hboa-stop}, which provides increased cost
-for more total leaves.
-This quality is used for two purposes: comparison of potential changes from the existing model
-and comparison of that change with the existing model. The basis for our simplication
-is to rearrange $p(B)BDe(B) < p(B')BDe(B')$ to be $\frac{p(B)}{p(B')}<\frac{BDe(B')}{BDe(B)}$.
-
-\begin{equation}
-  BDe(B) = \prod_{i=0}^{N}\prod_{l\in L_i} \frac{\Gamma(m'_i(l))}{\Gamma(m_i(l) + m'_i(l))}
-  \prod_{x_i}\frac{\Gamma(m_i(x_i, l) + m'_i(x_i,l))}{\Gamma(m'_i(x_i,l))}
-  \label{eq-hboa}
-\end{equation}
-
-\begin{equation}
-  p(B) = c2^{-0.5(\sum_i|L_i|)log_2\mu}
-  \label{eq-hboa-stop}
-\end{equation}
-
-The outermost product of Equation~\ref{eq-hboa} iterates over all trees in the forest. However, each split
-can modify only one of the trees and therefore the contribution of all others can be canceled. The middle
-product is across all leaves in the tree. Again since only one leaf can be changed, all other terms can
-be canceled. By convention hBOA uses uninformed Bayesian priors of $m'_i(l)= 2$ and $m'_i(x_i, l)=1$ for
-binary alphabets. As $\Gamma(a) = (a-1)!$ this means the top term in the middle product and the bottom
-term in the third product reduce to 1. The only remaining terms are then $m_i(l)$ and $m_i(x_i, l)$ which
-represent the number of solutions which reached leaf $l$ and the number of solutions which reached leaf $l$
-with a specific value for $x_i$, respectively.
-
-Equation~\ref{eq-hboa-stop} can also be simplified when doing comparisons. If model $B'$ has exactly one more
-leaf than model $B$ then the ratio $\frac{p(B)}{p(B')}$ simplifies to $2^{0.5 log_2\mu}$ regardless of
-total model size.
-
-The resulting simplifications create Equation~\ref{eq-hboa-final},
-where $B'$ is different from $B$ by exactly 1 split, such that $l$ was split to create $l'$ and $l''$.
-The best split is whichever maximizes its improvement over $B$, which is equal to the right side
-of the inequality. Note that these factorials can still be exceedingly large and
-therefore it is imperative that implementations avoid rounding errors and overflows.
-
-\end{document}
+\end{doublespace}
+\newpage

### 2014-10-15 Comprehensive/appendix-hboa.tex
Initial commit of Comprehensive Report

--- 2014-10-15 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
@@
-\documentclass[twoside]{article}
-\usepackage{ecj,palatino,epsfig,latexsym,natbib}
-\usepackage{url}
-\usepackage{verbatim}
-\usepackage{subcaption}
-\usepackage[noend]{algpseudocode}
-
-\newcommand{\includegraphicswide}[1]
-{\includegraphics[width=.9\textwidth,height=\textheight,keepaspectratio]{#1}}
-
-\newcommand{\includegraphicsfit}[1]
-{\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}
-
-\usepackage{array}
-\newcolumntype{L}{>{\centering\arraybackslash}m{1.9cm}}
-
-%% do not add any other page- or text-size instruction here
-
-\parskip=0.00in
-
-\begin{document}
-
-% 45-character limit on header
-\ecjHeader{x}{x}{xxx-xxx}{201X}{Parameter-less Population Pyramid}{B. W. Goldman and W. F. Punch}
-\title{\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid}
-
-\author{\name{\bf B. W. Goldman} \hfill \addr{brianwgoldman@acm.org}\\ 
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-\AND
-       \name{\bf W. F. Punch} \hfill \addr{punch@msu.edu}\\
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-}
-
-\maketitle
-
-\begin{abstract}
-%Should be about 200 words
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
-evolutionary optimization without requiring any user specified parameters.
-P3's primary innovation is to replace the generational model with a pyramid of
-multiple populations that are iteratively created and expanded. In combination
-with local search and advanced crossover P3 scales to problem difficulty, exploiting
-previously learned information before adding more diversity.
-
-Across seven problems, each tested using on average 18 problem sizes, P3 outperformed
-all five advanced comparison algorithms. This improvement includes requiring less evaluations
-and less wall clock seconds to find the global optimum and better fitness when using
-the same number of evaluations. Using both algorithm analysis and comparison we find P3's
-effectiveness is due to its ability to properly maintain, add, and exploit diversity.
-
-Unlike the best comparison algorithms, P3 was able to achieve this quality without any
-problem specific tuning. Thus, unlike previous parameter-less methods, P3 does not
-sacrifice quality for applicability. Therefore we conclude that
-P3 is an efficient, general, parameter-less approach to black box
-optimization which is more effective than existing state-of-the-art techniques.
-
-\end{abstract}
-
-\begin{keywords}
-
-Genetic algorithms, 
-linkage learning,
-local search,
-parameter-less.
-
-\end{keywords}
-
-\section{Introduction}
-A primary purpose of evolutionary optimization is to efficiently find good solutions
-to challenging real world problems with minimal prior knowledge about the problem itself.
-This driving goal has created search algorithms which can escape user bias to create
-truly novel results, sometimes publishable or patentable in their own right~\citep{kannappan:2014:humies}.
-While it is not possible for any algorithm to do better than random search across all possible
-problems~\citep{Wolpert:1997:nfl}, effectiveness can be achieved by assuming the search
-landscape has structure and then biasing the algorithm toward exploiting that structure.
-
-In evolutionary optimization, and genetic algorithms~(GAs) in particular, search is often
-biased through parameters. This can be beneficial as it allows practitioners to inject their
-knowledge about the shape of the search landscape into the algorithm.
-However, the quality of solutions found, and the speed at which they are found, is strongly tied to setting these parameters
-correctly~\citep{goldberg:1991:gasize}. As such, either expert knowledge or exceedingly
-expensive parameter tuning~\citep{grefenstette:1986:optimalga} are required to leverage
-this feature to its fullest potential. Furthermore,  parameters such as population size, mutation rate, crossover
-rate, tournament size, etc. usually have no clear relationship to the problem being solved, meaning even
-domain experts may not understand how the parameters will interact with the problem or with each other.
-
-There have been periodic efforts to reduce or remove the need for parameter tuning.
-\cite{Back:1992:selfadapt} introduced self-adaptive parameters, in which parameter values
-were included in each solution's genome and themselves underwent evolution. This allowed search
-to optimize some of its own parameters resulting in a reduced the need for expert tuning.
-\cite{harik:1999:parameterlessga} were able to design an entirely parameter-less GA by
-leveraging schema theory and parallel populations. Unfortunately these methods were provably less efficient
-than directly setting the parameters to optimal values~\citep{pelikan:1999:worstparameter-less}.
-
-One area that has been very effective at reducing the number of algorithm parameters is
-model based search. \cite{pelikan:2006:hboa}'s Hierarchical Bayesian Optimization
-Algorithm~(hBOA) and \cite{thierens:2010:ltga}'s Linkage Tree Genetic Algorithm~(LTGA)
-both require only a single parameter: population size. \cite{posik:2011:parameterless}
-leveraged model building to create a fully parameter-less algorithm, but it is restricted to
-only order-k, fully decomposable, noiseless problems.
-
-Most recently \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid~(P3).
-This method uses a pyramid structure of populations to combine model based search with local search
-to achieve parameter-less optimization. Initial results suggest that unlike
-previous parameter-less methods, P3 is actually more efficient than current state-of-the-art
-parameterized search algorithms. In this work we shall: extend these results to cover more
-comparison algorithms; compare both efficiency in reaching the global optimum and intermediate
-fitnesses; algorithm complexity analysis; and provide more in depth analysis of P3 itself.
-
-\section{Comparison Optimizers}
-\label{sec-optimizers}
-
-In order to fully understand the effectiveness of P3, we compare it with five advanced algorithms which
-have related features to P3. The Random Restart Hill Climber~\citep{goldman:2014:p3} was chosen as an efficient form of repeated
-local search. As P3 combines this hill climber with crossover,
-comparing with local search alone shows the advantages of P3's overall approach.
-The $1+(\lambda, \lambda)$ algorithm is the current best theory supported simple genetic algorithm
-and its method of crossover is in some sense a macro-mutation just as in P3. hBOA and Parameter-less
-hBOA are advanced model building search techniques which are very effective at learning complex
-problem structure, designed to achieve similar goals as P3's linkage learning but using very different
-methods. Finally LTGA represents the current state-of-the-art in black box search and is the origin
-of P3's linkage learning and crossover methods.
-
-Finally, only hBOA and LTGA require any parameters, with each of these only requiring a population
-size. This makes knowing the optimal behavior of these algorithms much more tractable. All of the
-algorithms are also gene order independent, fitness scale invariant, and unbiased. This means,
-for any problem, the order in which problem variables appear in the genome can be changed
-without changing the behavior of the search. The fitness can also be manipulated in any fashion
-as long as the rank ordering of solutions is unchanged. These algorithms are also unaffected by the meaning
-assigned to each bit, such that inverting a predetermined random subset of genes before evaluation
-will not impact search efficiency.
-
-Our implementations of all of these algorithms as well as all of the population size information, raw results,
-and processing scripts
-are available from our website.\footnote{\url{https://github.com/brianwgoldman/FastEfficientP3}}
-\subsection{Random Restart Hill Climber}
-\label{sec-hill-climber}
-Perhaps the simplest black box search heuristic is stochastic local search, or hill climbing.
-This optimization technique focuses on improving a single solution until it reaches a local
-optimum. Here we use the first improvement hill climber defined by \cite{goldman:2014:p3}
-and given in Figure~\ref{fig-hc}. This algorithm works by flipping each bit in a random
-order, keeping modifications when fitness is improved, until single bit flips cannot result in
-further fitness improvements.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Hill-Climber}{}
-    \State $options \leftarrow [0 \dots N-1]$\label{fig-hc-options}
-    \State $tried \leftarrow \emptyset$
-    \While{$|tried| < |options|$}
-      \ForAll{$index \in shuffled(options)$}
-        \If{$index \notin tried$}\label{fig-hc-tried}
-          \State Flip bit $index$ in solution
-          \If{solution's fitness increased}
-            \State $tried \leftarrow \emptyset$
-          \Else
-            \State Revert change
-          \EndIf
-          \State $tried \leftarrow tried \cup \{index\}$
-        \EndIf
-      \EndFor
-    \EndWhile
-  \EndProcedure
-\end{algorithmic}
-  \caption{Hill climbing algorithm used to improve randomly generated solutions until no single
-           bit change results in a fitness improvement.}
-  \label{fig-hc}
-\end{figure}
-
-The hill climber requires an amortized cost of $O(1)$ operations per evaluation. In order to
-terminate, at least one evaluation must be performed for each of the $N$ bits in the solution.
-As such any operation that happens only once per search can be amortized over at least $N$
-evaluations, covering the initialization of $options$ on Line~\ref{fig-hc-options}.
-Line~\ref{fig-hc-tried}, which prevents wasted evaluations, can be called at most twice
-per evaluation: once to add $index$ into $tried$ and once to prevent $index$ from being unnecessarily
-evaluated again. The only way three or more calls could happen is if
-no fitness improvement was made for the entire previous iteration, which contradicts the loop invariant.
-
-Due to its nature, this hill climber cannot escape basins of attraction. Once a solution is reached
-such that none of the single bit neighbors are fitness improvements, search stops. Thus
-this algorithm requires a restart mechanism to solve multimodal problems. We have
-chosen here to naively restart search from a random solution whenever a local optima is found. This
-ensures that on all landscapes there is always a non-zero probability of search finding the global optimum.
-
-\subsection{$1+(\lambda, \lambda)$}
-\cite{doerr:2013:lambdalambda} presented the first genetic algorithm to provably show
-the advantages of performing crossover on simple landscapes. This comparatively simple
-algorithm maintains only a single individual and a self-controlled parameter $\lambda$.
-
-Each iteration, the number of bits to flip is chosen from the binomial distribution $b\sim B(N, \frac{\lambda}{N})$,
-where $N$ is the number of bits in the genome.
-Next, $\lfloor\lambda\rfloor$ offspring are produced by flipping $b$ bits. The
-best mutant then produces $\lfloor\lambda\rfloor$ offspring via uniform crossover with the original parent, such that each gene comes from the
-mutant with probability $\frac{1}{\lambda}$. In the original algorithm the best
-offspring produced by crossover then replaces the original parent if its fitness is no worse.
-The $\lambda$ parameter, which is initialized to 1, is decreased if the offspring replaced
-its parent and increased otherwise.
-
-The original formulation was designed specifically for unimodal landscapes and as such were
-not directly suitable for multimodal problems. \cite{goldman:2014:p3} extended $1+(\lambda, \lambda)$
-to include random restarts. As search stagnates, the $\lambda$ parameter increases in value. Eventually
-this results in $\lambda \ge N$ causing mutation to always flip all bits of the individual.
-As this prevents any future improvement, whenever $\lambda \ge N$ search is restarted from a random solution with $\lambda$
-reset to 1.
-
-A few other efficiency modifications were also made. If there is a tie in crossover offspring fitness,
-whichever has a larger hamming distance from the parent is retained. This encourages drifting across plateaus.
-The ``mod'' control strategy proposed by \cite{doerr:2013:lambdalambda} was not used as it conflicted with
-the random restart strategy.
-If a crossover individual is identical to either of its parents, it is not evaluated.
-If mutation produces an offspring which is better than the best crossover offspring, it is used to compare
-against the original parent.
-
-\subsection{Hierarchical Bayesian Optimization Algorithm}
-
-\cite{pelikan:2006:hboa} used statistical principles in combination with a decision tree structure
-to create the Hierarchical Bayesian Optimization Algorithm (hBOA). This method creates a model of
-epistatic relationships between genes which is then used to stochastically generate new solutions.
-Each generation a binary tournament with replacement is used to select $\mu$ solutions from
-the population. These solutions are then used to build the model, which in turn is used to generate $\mu$ new
-solutions. The new solutions are then integrated into the population using restricted tournament
-replacement.
-
-Conceptually, the model built by hBOA is trying to infer rules of the form ``Given that this
-subset of genes are set to these values, how frequently is gene $x_i$ set to value v?'' This can
-be represented using a directed acyclic decision forest, with each tree in the forest representing one gene
-in the solution. In the decision tree $T_i$, which is used to set the value of gene $x_i$,
-each internal node represents previous decisions on how to set
-some other gene $x_j$, with the children of that node representing how the decision was made. The
-leaves of each tree give the probability that $x_i$ should be set to each possible value.
-
-The forest is constructed iteratively, with each tree initially containing a single leaf
-and with each leaf storing a pointer for each selected solution. Each iteration the algorithm considers
-all possible ways of splitting an existing leaf using another gene $x_j$, such that solutions in the
-leaf are moved to the newly created leaves based on their value for $x_j$. The general goal is to
-separate the solutions such that all solutions with $x_i = 0$ move to one leaf while solutions with
-$x_j = 1$ move to the other.
-
-This goal is formalized using model scoring from Bayesian statistics. In its raw form this
-almost always creates near infinitesimal results, calculating fractions which include factorials of $\mu$
-and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{sec-appendix}, we
-derived a simplified form shown in Equation~\ref{eq-hboa-final}. If no proposed split satisfies the inequality, iteration stops.
-If multiple splits do, whichever maximizes the right side is chosen.
-
-\begin{equation}
-  2^{0.5 log_2\mu} < \frac{(m_i(l) + 1)!}{m_i(0, l)!m_i(1,l)!} \cdot
-  \frac{m_i(0, l')!m_i(1,l')!m_i(0, l'')!m_i(1,l'')!}{(m_i(l') + 1)!(m_i(l'') + 1)!}
-  \label{eq-hboa-final}
-\end{equation}
-
-Initially there are $\Theta(N^2)$ possible ways to split existing leaves, as each of the $N$ single node
-trees can be split by any of the other $N-1$ genes. Each iteration a new edge is added to the decision
-forest, meaning some of the previously tested splits cannot be used. For instance, if $T_i$, which is used
-to decide the value of $x_i$, is split using the value of $x_j$, $T_j$ can no longer be split using $x_i$.
-As a split creates two new leaves, $O(N)$ new potential splits must also be tested. Equation~\ref{eq-hboa-final}
-parses all solutions which reach a leaf to count gene frequencies, requiring $\mu$ time.
-The number of total leaves created depends heavily on the problem and $\mu$.
-However, assuming no splits are accepted or that the cost of testing all future splits is less than
-the initial $\Theta(N^2)$, constructing the model requires $\Omega(\mu N^2)$ time. Each
-model is used to generate $\mu$ solutions, leading to a cost per evaluation of $\Omega(N^2)$.
-
-To generate a solution, the value of each gene $x_i$ is set using its corresponding decision tree $T_i$. Because
-the forest is directed acyclic, there must be an ordering of $T_i$ such that before $T_i$ is executed all
-$x_j$ it uses to make decisions have already been set. As such, previous decisions made by other trees
-are used to follow each $T_i$ until a leaf is reached. The value of $x_i$ is then set based on the
-probability that other solutions reached that leaf with each value of $x_i$.
-
-To perform replacement, hBOA uses restricted tournament replacement. After each new solution is generated
-and evaluated, a set of $w$ solutions are chosen at random from the population, where $w=\min\{N, \frac{\mu}{20}\}$.
-From this set the solution which is most genetically similar to the offspring is chosen. If the offspring
-is at least as fit as the chosen solution, it replaces the chosen solution in the population. Otherwise the
-offspring is discarded. This method is designed to preserve genetic diversity as only genetically similar
-solutions must compete on fitness.
-
-hBOA is designed to work with large population sizes, resulting in a large number of
-evaluations per generation. As hBOA utilizes explicit diversity maintenance, standard methods for determining
-convergence are not considered very accurate. Therefore the authors suggest that an hBOA run should be
-terminated after performing generations equal to $N$.
-
-Like other model based techniques, hBOA has very few parameters. There is no mutation or crossover,
-and modeling does not rely on any explicit parameters. Solution selection, generation, and replacement
-are all derived from the population size, which must be set by the user.
-
-\subsection{Parameter-less hBOA}
-Using the methods first introduced by \cite{harik:1999:parameterlessga} for the Parameter-less GA,
-\cite{pelikan:2004:parameterlesshboa} created Parameter-less hBOA which automatically scales its
-population size to fit the problem. This is done by maintaining a list of concurrent populations
-using exponentially scaled population sizes.
-
-A run of Parameter-less hBOA starts with a single population of size $\mu_0$, conventionally set
-to $\mu_0=10$. After two generations are performed, a new population of size $\mu_1 = 2\mu_0$ is created
-and performs a generation. Evolution then continues with the $\mu_0$ population performing two generations
-for each one performed by $\mu_1$. Each time population $\mu_i$ performs its second generation a new population
-$\mu_{i+1}=2\mu_i$ is created, which performs generations half as often as $\mu_i$. In this way an infinite number of
-parallel population can be simulated, with each population receiving the same number of total evaluations.
-
-In all other aspects each population is identical to an hBOA population using a fixed $\mu$. No search information
-is shared among populations, and each search is independently terminated. As such Parameter-less hBOA cannot
-perform better than hBOA using the optimal population size for a given instance, as it must also spend evaluations
-on populations of different sizes. This inefficiency is bounded by a log multiple of the total number of
-evaluations~\citep{pelikan:1999:worstparameter-less}.
-
-\subsection{Linkage Tree Genetic Algorithm}
-\cite{thierens:2010:ltga} introduced the Linkage Tree Genetic Algorithm (LTGA) which automatically
-detects and exploits problem epistasis by examining pairwise gene entropy. Due to its enhanced
-ability to preserve high fitness gene subsets, LTGA was able to outperform state-of-the-art
-GAs across many benchmarks. Since its introduction, many variants of LTGA have been
-proposed~\citep{goldman:2012:ltga} so for clarity we have chosen the version
-presented by \cite{thierens:2013:ltgahiff} as our model.
-
-LTGA's effectiveness comes from its method of performing crossover. Instead of blindly
-mixing genes between parents, LTGA attempts to preserve important interrelationships
-between genes. Before performing any crossovers in a generation, LTGA first builds
-a set of hierarchical gene clusters which are then used to dictate how genes are mixed
-during crossover.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Cluster-Creation}{}
-    \State $unmerged \leftarrow \{\{0\}, \{1\}, \{2\}, \dots, \{N-1\}\}$
-    \State $useful \leftarrow unmerged$
-    \While{$|unmerged|>1$}
-      \State $C_i, C_j \leftarrow \min_{C_i,C_j \in unmerged} D(C_i, C_j)$
-      \State $unmerged \leftarrow unmerged - \{C_i, C_j\} + \{C_i \cup C_j\}$
-      \State $useful \leftarrow useful + \{C_i \cup C_j\}$
-      \If{$D(C_i, C_j) = 0$}
-        \State $useful \leftarrow useful - \{C_i, C_j\}$
-      \EndIf
-    \EndWhile
-    \State Order $useful$ based based on last merged first\label{fig-cluster-creation-ordering}
-    \State Remove largest cluster from $useful$
-
-    \Return $useful$
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm describing how LTGA creates clusters using Equation~\ref{eq-distance}
-           for a population. $unmerged$ and $useful$ are ordered sets of sets of gene loci.}
-  \label{fig-cluster-creation}
-\end{figure}
-
-Figure~\ref{fig-cluster-creation} provides the agglomerative method LTGA uses to create gene clusters.
-This algorithm creates a tree of sets using pairwise gene entropy, such that the leaves of the tree contain a single gene and
-internal nodes are the union of their children's sets. These sets are then used by crossover to specify epistatic
-relationships which should be preserved.
-The process begins by creating the set of sets $unmerged$ which tracks all top level clusters. Initially
-$unmerged$ contains single member sets for each gene. After each iteration the two sets with the minimum average pairwise
-distance (given in Equation~\ref{eq-distance}) are merged to create a single cluster. This process is repeated
-until only a single set remains in $unmerged$ which contains all of the genes in the genome.
-
-\begin{equation}
-  D(C_i,C_j) = \frac{1}{\left | C_i \right |\cdot \left |C_j \right|}\sum_{c_i \in C_i}\sum_{c_j \in C_j}
-  2 - \frac{H(c_i) + H(c_j)}{H(c_i \cup c_j)}
-  \label{eq-distance}
-\end{equation}
-\begin{equation}
-  H(c) = -\sum_{s\in S} p_c(s)\log(p_c(s))
-  \label{eq-entropy}
-\end{equation}
-
-Throughout this process $useful$ tracks the set of all gene clusters which should be preserved for use by crossover.
-This set begins with all genes in separate clusters, and each time a new cluster is created it is added to $useful$.
-However, not all clusters are necessarily worth keeping. For instance, in all versions of LTGA the cluster
-containing all genes is removed from $useful$ as preserving all genes during crossover can only create clones.
-\cite{thierens:2013:ltgahiff} extended this removal to include any unsupported subsets. If the pairwise distance
-between two clusters is 0, this means there are no individuals in the population which disrupt the relationships between the two
-clusters. Therefore when performing crossover there is no reason to believe a fitness improvement can be achieved
-by breaking the stored pattern. As such a cluster is only kept if its direct superset has a non-zero distance.
-As a final step Line~\ref{fig-cluster-creation-ordering} reorders $useful$ such that clusters appear in reversed
-order from which they were added to $useful$. Thus the most linked clusters and those containing single genes
-appear at the end of the returned list.
-
-\cite{thierens:2013:ltgahiff}'s version of LTGA does not use the entire population when determining pairwise entropy.
-Instead, binary tournament is used to select half of the population. This is done to ensure the model is built
-using only high quality solutions, even during the first generation.
-
-In order to efficiently perform clustering, a pairwise gene frequency table is constructed from the selected solutions.
-To calculate Equation~\ref{eq-distance}, Equation~\ref{eq-entropy} is called for each
-gene ($H(c_i)$) and pair of genes ($H(c_i \cup c_j)$). Extracting this information requires $O(\mu N^2)$
-time, where $\mu$ is the population size and $N$ is the genome size. The process of converting
-this pairwise frequency information into clusters can be achieved in $O(N^2)$ using the bookkeeping
-methods presented by \cite{gronau:2007:upgma}. This cost is performed only once per generation,
-and is then used to perform approximately $O(\mu N)$ crossover evaluations. As a result, the amortized cost of
-LTGA's model building is $O(N)$.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Cluster-Usage}{}
-    \ForAll{$C_i \in useful$}
-      %\ForAll{$d \in shuffled(P_i)$}
-        \State $d \leftarrow rand\_choice(P)$\label{fig-cluster-usage-donate}
-        \State Copy $d$'s gene values for $C_i$ into solution
-        \If{solution was changed}
-          \If{solution's fitness decreased}
-            \State Revert changes
-          \EndIf
-          %\State \textbf{break}
-        \EndIf
-      %\EndFor
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm describing how clusters are used to perform crossover.}
-  \label{fig-cluster-usage}
-\end{figure}
-
-Figure~\ref{fig-cluster-usage} describes how the identified clusters are used by crossover to preserve
-gene linkage while still exploring the search space. Unlike more traditional crossover methods, LTGA
-crosses each individual with the entire population. Also, to produce a single offspring, multiple evaluations
-of the fitness function are performed.
-
-Each generation, each individual in the population undergoes crossover. In a single crossover event, each
-cluster of genes $C_i$ in $useful$ is applied as a crossover mask. A random donor $d$
-is chosen from the entire population (not just the model selected population), and $d$'s gene's for $C_i$ are copied
-into the working solution. If a modification is made, an evaluation is then performed. If the crossover
-resulted in no worse fitness then the changes are kept. This allows for neutral drift across plateaus.
-The resulting solution, which must be at least as fit as its parent, is then
-copied into the next generation.
-
-In total each individual can cause up to $|useful|$ evaluations. If all clusters were kept, even those deemed
-unhelpful, and all donations were evaluated, even those which did not change any genes, then \Call{Cluster-Usage}{}
-would perform exactly $2N-2$ evaluations for each of the $\mu$ solutions in the population. This provides the amortizing evaluations
-required to make clustering only $O(N)$ operations per evaluation. However, by skipping some evaluations, its
-possible that clustering may be super-linear.
-
-LTGA has no explicit form of diversity control and has no method for introducing new genetic information once
-the population has converged. Therefore an LTGA run is considered converged once two consecutive populations
-contain the same unique solutions.
-
-By design, LTGA only has a single parameter: population size. LTGA uses no mutation, and crossover is defined
-in terms of the clustering algorithm. Selection between generations is fully elitist and embedded in the crossover,
-with selection of model building solutions fixed to a binary tournament. Neither \Call{Cluster-Creation}{} nor
-\Call{Cluster-Usage}{} rely on parameter values. LTGA does not provide any method for controlling or setting
-the population size, and uses a fixed user specified size for each generation.
-
-\section{Parameter-less Population Pyramid}
-\label{sec-p3}
-\cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid (P3) as a method for
-performing optimization which does not require the user to provide any parameters. This is
-achieved by combining efficient local search with the model building methods of LTGA using
-an iteratively constructed hierarchy of populations.
-
-The high level algorithm of P3 is presented in Figure~\ref{fig-p3}. Unlike more traditional
-GAs, P3 does not follow a generational model. Instead, it maintains an iteratively
-expanding pyramid of expanding populations. Each iteration, a new random solution is generated.
-This solution is brought to a local optimum using the hill climbing algorithm in Figure~\ref{fig-hc}. If that
-local optimum has not yet been added to the pyramid, the solution is added to the lowest
-population $P_0$.
-
-Next, the solution is iteratively improved by applying LTGA's crossover algorithm (Figure~\ref{fig-cluster-usage})
-with each population $P_i$ in the pyramid. If this process results in a strict fitness improvement and has
-created a solution not yet stored in the pyramid, it is added to the next highest pyramid level $P_{i+1}$.
-If $P_{i+1}$ does not yet exist, it is created. In this way populations in the pyramid expand over time,
-and the number of populations stored increases over time. Initially the pyramid contains no solutions
-or populations, meaning the user does not need to specify a population size.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Iterate-P3}{}
-    \State Create random solution
-    \State Apply hill climber
-    \If{solution $\notin hashset$}
-      \State Add solution to $P_0$
-      \State Add solution to $hashset$
-    \EndIf
-
-    \ForAll{$P_i \in pyramid$}
-      \State Mix solution with $P_i$
-      \If{solution's fitness has improved}
-        \If{solution $\notin hashset$}
-          \State Add solution to $P_{i+1}$
-          \State Add solution to $hashset$
-        \EndIf
-      \EndIf
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-  \caption{One iteration of P3 optimization. $pyramid$ is an
-           ordered set of populations and $hashset$ is a set
-           of all solutions in $pyramid$.}
-  \label{fig-p3}
-\end{figure}
-
-To accommodate P3's unique population structure, some of LTGA's clustering procedures were modified. In LTGA,
-clusters are identified at the start of each generation and are used
-to create all offspring in that generation. As P3 does not perform serial
-generations, P3 instead updates the model each
-time a solution is added to a population. Furthermore, unlike
-LTGA all solutions in the population are used to generate the model, not just the
-winners of a binary tournament. We can do this because even the
-worst solutions in the pyramid are already high quality due to local search.
-Using local search in LTGA was examined by \cite{bosman:2011:lsbbo} and found
-to provide no significant improvement. A likely cause was that that study applied
-local search to every solution, not just the initial population, resulting in significant overhead.
-
-Beyond the changes in population structuring, P3 modifies
-LTGA's version of \Call{Cluster-Creation}{} and \Call{Cluster-Usage}{}.
-P3 changes Line~\ref{fig-cluster-creation-ordering} in Figure~\ref{fig-cluster-creation}
-from \emph{last merged first} ordering to \emph{smallest first} ordering.
-This method applies gene clusters during crossover based on how many genes
-are in each cluster\footnote{Ties are broken randomly.}, and not on how tightly linked those genes are.
-\cite{goldman:2012:ltga} found that this alternative was better at preserving
-diversity, and therefore required smaller populations, but has not yet been integrated into
-canonical LTGA.
-
-P3 also modified Line~\ref{fig-cluster-usage-donate} in Figure~\ref{fig-cluster-usage}.
-Instead of choosing a single genetic donor for each cluster, P3 iterates over the
-population in a random order until a solution in the population is found which
-has a least one gene different for that cluster of genes from the improving solution.
-This process increases the likelihood of an evaluation being performed for every cluster,
-and helps test rare gene patterns in the population.
-
-In LTGA the cost of rebuilding the model is $O(\mu N^2)$ as it must collect pairwise
-gene frequency information for all $\mu$ solutions in the population. P3 does not store
-a single population, and it does not have a fixed $\mu$ size for any population. However,
-each time a solution is added to the population, it requires $O(N^2)$ time to update
-the table of pairwise frequencies and another $O(N^2)$ time to rebuild the linkage model.
-The model is then used immediately to perform up to one evaluation for each of the up
-to $2N-2$ clusters. Just as in LTGA, if no evaluation shortcuts were made, P3
-has an amortized cost of $O(N)$ modeling cost per evaluation. While P3 does
-rebuild the model more frequently per solution in the population, it also performs
-a number of local search evaluations which are quite efficient, meaning theoretical
-comparisons of their speed are difficult to perform. As a final note, P3's repeated
-attempts to find a useful donation make it less likely than LTGA to skip evaluations,
-but has an added cost to find these donations. Repeated donations could require as much
-as $O(\mu)$ attempts per evaluation, but experimental evidence suggest that this
-operation actually saves more overhead than it costs by increasing the number of evaluations
-per model rebuild.
-
-Each of the pieces of the P3 algorithm were selected not just for their stand alone efficacy,
-but for the ways in which they interact. By using the hill climber to ensure each solution
-is at a local optima, the underlying pairwise relationships in the problem are exposed. As
-a result, detecting clusters for use by crossover is much more effective. The crossover operator
-is extremely elitist, as each gene donation must result in no fitness loss, and a solution must
-strictly improve to be added to the next level of the pyramid. This is balanced by continual
-integration of new random solutions. Furthermore, each random restart decreases the probability
-of spurious linkages caused by shared ancestry. This diversity is further preserved by applying
-gene clusters in smallest first order during crossover as this reduces the probability of genetic
-hitchhikers.
-
-Other algorithms have proposed using multiple concurrent populations.
-\cite{hornby:2006:alps} had a hierarchy of populations with solutions periodically advancing
-upward. This allows for continuous integration of diversity as the lowest population is reseeded
-with random solutions. However, this method resulted in increased parameterization as not only was
-a population size required, but also parameters for how frequently generations advanced between levels and
-how many total levels to have were added. \cite{harik:1999:parameterlessga} used multiple independent
-populations of different sizes as a method for removing the population size parameter, but
-doing was provably less efficient than using an optimal population size as no information is shared
-between the populations.
-
-\section{Problem Descriptions}
-\label{sec-problems}
-
-\subsection{Single Instance Problems}
-Understanding how a stochastic search algorithm will behave on arbitrary and complex
-search landscapes can be exceedingly difficult. Therefore a common practice for
-algorithm understanding is to perform search on well defined, well understood
-landscapes. To be of interest these landscapes need to represent interesting
-and important aspects of real world problems.
-
-Perhaps the most common such landscape is the Deceptive Trap problem \citep{goldberg:1991:gasize}.
-In this landscape the genome is broken up into $k$ bit non-overlapping subproblems referred
-to as traps.
-Each subproblem is scored using Equation~\ref{eq-deceptive-trap}, where $t$ is the number of
-bits in the trap set to 1. The global optimum in each trap is a string of all 1s, while all
-other solutions lead to a local optima of all 0s. This problem tests an algorithm's
-ability to overcome $k$ sized deception and is commonly used to determine how effective crossover
-is at preserving building blocks. Any crossover event that mixes bits from different parents in
-the same trap will likely result in that trap being optimized to the local optima. For our experiments
-we chose $k=7$ to ensure highly deceptive traps.
-
-\begin{equation}
-   trap(t) = \left\{
-     \begin{array}{rl}
-       k-1-t, &  t<k\\
-       k,   &  t = k
-     \end{array}
-   \right.
-  \label{eq-deceptive-trap}
-\end{equation}
-
-\cite{goldman:2012:ltga} found that mixing local search with linkage learning rendered the Deceptive
-Trap problem trivial. This is because local search is able to optimize each trap to one of the two
-optima (all 1s or all 0s), and then linkage learning can have perfect knowledge of gene interactions.
-In order to make the problem more difficult, the authors proposed the Deceptive
-Step Trap problem, given in Equation~\ref{eq-deceptive-step-trap}. This function modifies the results
-of Equation~\ref{eq-deceptive-trap} to include plateaus of size $s$, introducing an exponential number
-of local optima in each trap. As a result the Deceptive Step Trap is much more challenging for linkage
-learning techniques, while still being highly deceptive. We chose to use $s=2$ to create the maximum
-number of local optima.
-\begin{equation}
-   step\_trap(t) = \left \lfloor \frac{(k-s)\pmod{s} + trap(t)}{s} \right \rfloor
-  \label{eq-deceptive-step-trap}
-\end{equation}
-
-Another challenging aspect of landscapes can be higher order relationships. The Hierarchical If
-and only If (HIFF) problem is designed to capture the difficulties of this class of problem.
-In HIFF the genome is broken up into a complete binary tree, such that each gene appears in exactly one
-leaf and each internal node is the subset of genes contained in its children. If all genes represented
-in a node of the tree are set to the same value, they score equal to the size of the set. In this way
-small subsets lead toward solutions to larger subsets. However, a node can score if all genes are either
-all 1s or all 0s, meaning that to solve higher order subproblems it is necessary to perform crossovers
-that preserve lower order solutions. This problem is a natural fit for LTGA as the linkage tree can
-perfectly duplicate the problem's true relationships \citep{thierens:2013:ltgahiff}.
-
-As a final class of well known problems, we have chosen to borrow the Rastrigin problem from real valued
-optimization. This problem's landscape, determined by Equation~\ref{eq-rast}, is highly multimodal caused by
-the oscillating cosine function. \cite{goldman:2014:p3} proposed the Discretized Rastrigin problem, such
-that each floating point $x_i$ in Equation~\ref{eq-rast} is encoded using a 10 bit gray code.
-\begin{equation}
-  An + \sum_{i=1}^{n}\left [ x_i^2-A\cos (2\pi x_i) \right ] \forall x\in [-5.12,5.12]
-  \label{eq-rast}
-\end{equation}
-
-\subsection{Randomly Generated Problem Classes}
-While well defined landscapes can provide specific insights into how an algorithm works,
-their static nature can be misleading. Specifically, algorithm quality might
-be very fragile such that it is only effective at searching well behaved landscapes. A more realistic
-test of an algorithm's black box effectiveness is to work with randomly generated instances
-drawn from a problem class. When tested over a sufficiently large sample it is then possible
-to draw more general conclusions about the algorithms effectiveness. The challenge with these
-landscapes is determining the global optimum to gauge if an algorithm was successful.
-
-Perhaps the most common model for generating random rugged landscapes is the NK model. An
-NK Landscape determines the fitness of each gene based on epistatic relationships with $K$
-other genes in the genome. This fitness is specified using a randomly generated table of
-fitness values, were each possible combination of the $K+1$ genes is mapped to some floating
-point value $[0-1]$. In unrestricted NK landscapes the relationships between genes are also
-randomly chosen and as a result finding the global optimum is $NP$-Hard for $K>1$. However,
-if epistasis is set such that each gene depends on the $K$ directly following it
-in the genome the solution can be found in polynomial time~\citep{wright:2000:solvingnk}.
-These Nearest Neighbor NK landscapes are therefore ideal for search algorithm testing.
-For all of our experiments we fixed $K=5$ to ensure highly rugged landscapes.
-
-\cite{saul:1994:spinglass} presents a combinatorial benchmark problem derived from physics:
-Ising Spin Glasses. A spin glass is defined by a weighted graph of interaction terms between vertices.
-Each gene assigns a value to each vertex, with the fitness calculated by Equation~\ref{eq-ising}.
-In this equation, $E$ is the set of all edges, $e_{ij}$ is the edge weight connecting vertex $i$ to vertex $j$, and
-$x_i$ and $x_j$ are the gene values for vertex $i$ and $j$. Optimal fitness is when this sum is minimized.
-\begin{equation}
-\sum_{e_{ij} \in E} x_ie_{ij}x_j
-  \label{eq-ising}
-\end{equation}
-Similar to NK Landscapes, the general class is $NP$-Hard to optimize, but the $2D\pm J$ subset of
-Ising Spin Glasses can be polynomially solved.\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}
-In this subset the graph is restricted to be a two dimensional torus, edge weights are randomly set
-to either -1 or 1, and vertex values must be -1 or 1.
-
-As our final class of randomly generated problems we chose the Maximum Satisfiability (MAX-SAT) problem.
-Related to the more common 3-SAT problem, a MAX-SAT instance is defined by a set of three term clauses.
-Each term is a randomly chosen variable, which may also be negated. A clause scores if an and only if
-at least one term in the clause evaluates to true. In order to make MAX-SAT instances with a known global
-optimum, \cite{goldman:2014:p3} proposed constructing clauses around a fixed solution. In this way the
-signs of the terms are set to ensure the target solution satisfies the clause. To ensure each
-problem is challenging we chose a clause to variable ratio of $4.27$.
-
-\section{Comparison Algorithm Parameter Tuning}
-\label{sec-tuning}
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-hboa}
-      \end{centering}
-      \caption{Tuned Population Sizes for hBOA}
-      \label{fig-pop-hboa}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-ltga}
-      \end{centering}
-      \caption{Tuned Population Sizes for LTGA}
-      \label{fig-pop-ltga}
-    \end{subfigure}
-  \end{centering}
-  \caption{Optimal population sizes found using bisection on each size
-           for each problem.}
-  \label{fig-pop-sizes}
-\end{figure}
-
-While four of the six algorithms in our experiments do not require any user specified parameters,
-hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
-handicapped, we extensively tuned each using the bisection method \citep{sastry:2001:bisection}
-to determine the optimal population size for each problem size.
-Extended by \cite{goldman:2012:ltga}, this method iteratively
-doubles the population size until some success criteria is met and then performs bisection
-between the lowest successful and highest unsuccessful sizes. In this way the minimum population size
-which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
-of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
-above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
-do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
-As bisection can make infinitely large population sizes, any run which had not found the global
-optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.
-
-Figure~\ref{fig-pop-sizes} shows the results from performing bisection. In general hBOA
-required population sizes which were at least an order of magnitude larger than LTGA. Due to
-runtime and memory overhead, finding the optimal value for hBOA was much less tractable than LTGA
-for moderate to large problem sizes. LTGA's population size also grew significantly slower
-than hBOA's as problem size increased, especially on the two Trap problems and Rastrigin.
-Both algorithms were very ineffective on MAX-SAT, neither able to tune to problems sizes over 60 bits.
-This is likely due to the fact that some randomly generated MAX-SAT landscapes are very flat
-and highly deceptive~\citep{rana:1998:gamaxsat}.
-\section{Finding the Global Optimum}
-\label{sec-optimum}
-
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{evals-to-success}
-  \end{center}
-  \caption{Comparison of the median number of evaluations to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-evals-to-success}
-\end{figure}
-
-Figure~\ref{fig-evals-to-success} shows the median number of evaluations required by each of the six
-algorithms to find the global optimum for multiple sizes of each problem.  Each data point in
-Figure~\ref{fig-evals-to-success} represents the median of 100 runs, where unsuccessful runs
-are treated as requiring more evaluations than any successful run. If the median run was not successful
-no point is shown. The maximum problem size used for each problem was set to be the largest size we could
-feasibly determine LTGA's optimal population size. For many larger problems results are not shown
-for hBOA due to the extreme computational cost required to optimally set the population size.
-
-\subsection{Quantitative Comparison}
-Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
-The largest problem size for any problem where P3 was not the most efficient has 49 bits, with P3 achieving the best results on all 92 larger
-configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
-Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
-and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
-on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
-
-\subsection{Local Search}
-The Random Restart Hill Climber and $1+(\lambda, \lambda)$ are both relatively effective on small problem
-sizes. This is especially true for the three randomly generated problem problem classes. These problems
-may contain relatively few local optima or just be exceptionally difficult for the model based algorithms.
-On Deceptive Trap and Deceptive Step Trap using 4 or less traps, $1+(\lambda, \lambda)$ performs significantly
-better than any other algorithm. We believe this is
-because $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
-flipping entire traps.
-This ability also leads $1+(\lambda, \lambda)$
-to outperform the Random Restart Hill Climber on all problems except Nearest Neighbor NK.
-
-On larger problem sizes, the ability for local search to reach the global optimum quickly diminishes.
-Only on MAX-SAT are these optimizers competitive
-at larger tested problem sizes. However, we believe this is because the largest tested MAX-SAT
-was an order of magnitude smaller than the largest size tested for most other problems. As the problem
-size increases the number of local optima increases exponentially, which explains why Random Restart Hill
-Climbing was unable to scale.
-For larger problems it also becomes increasingly unlikely for $1+(\lambda, \lambda)$ to make the right combination
-of changes required to reach the global optimum. This
-behavior causes high variance in success rate, as evident by the occasional successes on large Deceptive Trap problems.
-
-\subsection{Model Building}
-Only techniques which explicitly built models of gene epistasis were able to solve the largest problem
-instances. On single instance problems LTGA was more effective than hBOA, with hBOA outperforming
-LTGA on Nearest Neighbor NK and Ising Spin Glasses. This may be caused by the differences in modeling method:
-unlike the single instance problems, gene epistasis in the randomly generated
-problem classes cannot be be perfectly represented with a linkage tree.
-
-Considering how different hBOA and LTGA are in performing optimization, it is somewhat surprising how similar
-their results are on HIFF.
-However, both techniques rely on populations large enough to support the diversity required to reach the global optimum
-and to model epistasis. Both techniques also only rebuild models once per generation. As the subproblems of HIFF
-are nested, it is unlikely that either technique can accurately model higher order epistasis before solving
-lower order subproblems. Therefore both methods require one generation per subproblem order.
-
-\subsection{P3}
-Unlike the other model based methods, P3 generally outperforms
-both Random Restart Hill Climber and $1+(\lambda, \lambda)$ even on small problem sizes.
-Unlike the other local search methods, P3 outperforms LTGA and hBOA even on large problem sizes.
-This implies that P3 is gaining the benefits of each, leveraging local search to solve easy
-problems and model building to solve harder ones.
-
-Furthermore, the interaction between these two optimization tools explains some of the reason P3
-outperforms each method alone.
-On Deceptive Trap P3's use of hill climbing ensures all traps are immediately optimized, allowing for perfect
-linkage detection and high quality donation.
-On HIFF local search solves all pairwise subproblems, saving P3 a generation over LTGA and hBOA.
-In comparison P3 is only a slight improvement on Deceptive Step Trap, which is less amenable to local search.
-
-\section{Fitness Over Time}
-\label{sec-overtime}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{fitness-over-time}
-  \end{center}
-  \caption{Compares the median best fitness reached during search for each of the six optimization methods.}
-  \label{fig-fitness-over-time}
-\end{figure}
-
-For some applications, finding the global optimum is less important than finding good solutions quickly. Therefore
-we examine this behavior in Figure~\ref{fig-fitness-over-time}. At regular intervals during optimization
-Figure~\ref{fig-fitness-over-time} shows the median of the best fitnesses found at that point of search across
-100 runs. For each problem we show the largest problem size for which we were able to successfully gather results
-for all six algorithms, but the trends shown are representative of all larger problem sizes. The maximum reporting interval
-is set to include the slowest P3 run to reach the global optimum.
-
-% 181 total
-% 121 has P3 best
-%   9 hc
-%  50 lambdalambda
-%  20 hboa
-%  18 phboa
-%  27 LTGA
-
-\subsection{Quantitative Comparison}
-Of 181 sample points, P3 had the highest median fitness in 121. In pairwise competition, $1+(\lambda, \lambda)$
-was the most likely to outperform P3, doing so on 50 sample points. LTGA, hBOA, and Parameter-less hBOA
-were the next best, outperforming P3 on 27, 20, and 18 sample points. Random Restart Hill Climbing almost
-never outperformed P3, doing so only 9 times.
-
-\subsection{Local Search}
-Perhaps the most striking result is the quality of $1+(\lambda, \lambda)$. Until quite far into search this
-method performs better than both LTGA and hBOA. Given sufficient evaluations $1+(\lambda, \lambda)$ also outperforms
-Random Restart Hill Climbing on all 7 problems. For brief periods in the middle of search it performs the
-best of all techniques on: Deceptive Trap; Deceptive Step Trap; HIFF; Ising Spin Glass; and MAX-SAT problems.
-$1+(\lambda, \lambda)$'s ability to efficiently incorporate gene modifications of larger
-than one bit allows it to overcome the deception and plateaus in Deceptive Trap and Deceptive Step Trap, solve medium sized subproblems
-in HIFF, flip the signs on multiple adjacent bits in Ising Spin Glass, and cross plateaus in MAX-SAT. However,
-this method is slow in reaching the global optima in many of these problems which causes it to eventually
-be overtaken by the model building techniques.
-
-\subsection{Model Building}
-Both hBOA and LTGA are marked by periods of little improvement followed by rapid improvement.
-In hBOA this is taken to the extreme, with all fitness
-improvement made at the very end of search. In both cases this is caused by model building. Before the model
-is accurate little improvement is made. Once it is accurate, fitness improves dramatically.
-
-On every problem LTGA has five distinct periods: fitness plateau, near instantaneous improvement, fitness plateau,
-and improvement to global optimum. The early period corresponds with
-initialization of the population, with the first fitness gain achieved immediately upon completing the first generation.
-When using an inaccurate model, LTGA's mixing strategy performs a sort of less effective local search.
-Subsequent generations then make only minor fitness improvements. Once the model
-becomes accurate and the probability of a crossover using high quality genetic material increases sufficiently, LTGA
-enters a second period of rapid improvement.
-
-\begin{comment}
-This behavior is easiest to understand on the two Trap problems. LTGA's first generation will push individual traps toward a
-local optima, but it requires more than a single generation to do so completely. Until that occurs, the model is likely behaving
-no better than random as there is little apparent linkage between bits. Once the model begins to identify individual traps, crossover
-can begin to increase the frequency of higher fitness trap settings. This process begins slowly as low fitness local optima are more
-likely to be chosen for donation than high fitness local optima due to their frequency in the population. The process becomes
-self-catalysing as increased frequency of the global optimum trap genes means increased likelihood of the global optimum being spread by crossover.
-\end{comment}
-
-\subsection{P3}
-The integration of hill climbing into P3 makes it strictly better than using hill climbing alone.
-Early in optimization P3 and the Random Restart Hill Climber have effectively identical quality. This is because
-P3 performs the same evaluations as the Hill Climber for the first two restarts. Once P3 begins
-performing crossover it immediately improves over the Hill Climber. In 95\% of intervals P3 had a fitness at
-least as high as Hill Climbing. As such P3 is better than a simple
-hill climber regardless of how long each technique is run and irrespective of how high quality the solution
-found has to be.
-
-Unlike the model based methods, which struggle until model accuracy improves, P3's iterative solution
-integration allows it to improve much more quickly. This behavior exists in most problems, but is easiest to understand on Deceptive Trap.
-On this problem, P3 immediately brings all traps to local optima, equaled only by the Random Restart Hill Climber in quality.
-In comparison LTGA must evaluate the entire population and perform multiple generations to reach similar quality.
-P3 is able to immediately integrate optimal versions of each
-trap into a single individual as they are found by local search, resulting in smoother fitness improvement than LTGA.
-
-\section{Computational Expenses}
-While it is common in evolutionary computation to assume the evaluation function will dominate algorithm
-complexity, in some domains this will not be true. Model based methods are especially likely to violate
-this norm. Therefore, in order to assess P3's quality in solving problems with efficient fitness functions,
-we provide data on both its algorithmic complexity and wall clock time.
-
-\subsection{Operation Counting}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{rebuilds}
-      \end{centering}
-      \caption{Rebuilds}
-      \label{fig-rebuilds}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{donations}
-      \end{centering}
-      \caption{Donations}
-      \label{fig-donations}
-    \end{subfigure}
-  \end{centering}
-  \caption{Computation costs incurred by model rebuilding (Figure~\ref{fig-rebuilds}) and
-           repeated donations (Figure~\ref{fig-donations}) per evaluation as problem size increases.}
-  \label{fig-costs}
-\end{figure}
-
-When discussing the asymptotic complexity of P3 in Section~\ref{sec-p3}, two aspects eluded precise
-analysis: how expensive is model rebuilding and how many gene donations are made. Figure~\ref{fig-costs}
-provides some insight into how often these two aspects of the algorithm are utilized.
-
-Figure~\ref{fig-rebuilds} reports in an algorithmic sense how expensive model rebuilding is for search.
-In order to calculate this value we recorded how many times search rebuilt the model during each run.
-Under the minimal assumption that an evaluation costs $\Omega(N)$ time and that model rebuilds cost $O(N^2)$,
-Figure~\ref{fig-rebuilds} shows the ratio cost of model rebuilding per evaluation. If the cost of model building
-scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically constant.
-For Nearest Neighbor NK, Ising Spin Glasses, and Rastrigin this is the case. For both Trap problems
-and HIFF there is slow growth in the ratio. The problem sizes used for MAX-SAT were not sufficient
-to accurately gauge the asymptotic behavior. Together this suggests that while the cost of building the model
-is almost linear per evaluation, it can grow slowly. However, even in the worst case (HIFF) this growth was
-no more than twice the algorithmic cost of an evaluation even using 2048 bits.
-
-When applying a crossover subset, P3 tries random donors from the population until one is found with at least
-one bit different from the improving solution. In theory this can result in up to $O(\mu)$ operations.
-Figure~\ref{fig-donations} examines the observed average number of donations per evaluation performed.
-Ising Spin Glass, HIFF, and Rastrigin all achieve effectively constant behavior here, implying repeated
-donation does not impact the asymptotic runtime of P3. Both Trap functions and Nearest Neighbor NK all
-increase in number of donations as problem size increases, potentially increasing algorithmic costs. An
-important note is that each donation may range in size from a single bit up to $N-1$, with smaller
-donations far more likely to result in repeated attempts. As such this may cause some super-linear growth
-in P3, but its unlikely to be very high.
-
-\subsection{Wall Clock Performance}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{seconds-to-success}
-  \end{center}
-  \caption{Comparison of the median number of seconds to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-seconds-to-success}
-\end{figure}
-
-To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}.
-Similar to Figure~\ref{fig-evals-to-success}, each point represents the median
-of 100 runs, with unsuccessful runs treated as slower than successful runs.
-In total these runs
-took over 6 computing years to complete. We were therefore forced to use
-diverse hardware potentially running multiple processes in parallel.
-As such these timings should be considered very skeptically as a number
-of sources may contribute to biased error in the results. However, due to
-the size of the dataset and the general regularity in the results some general conclusions
-can still be made.
-
-\subsubsection{Model Building}
-hBOA and Parameter-less hBOA perform significantly worse when using wall clock time as the unit of comparison than when using
-evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while, under reasonable
-assumptions, P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
-hBOA goes from being slightly more efficient than LTGA in terms of evaluations to significantly worse in terms of seconds.
-As P3 and LTGA require a similar asymptotic complexity per
-evaluation as the Hill Climber and $1+(\lambda, \lambda)$, no similar change in ordering occurs.
-
-\subsubsection{P3}
-Any increase in P3's algorithmic complexity is offset by its efficient use of evaluations.
-In aggregate P3 found the global optimum the fastest on 72 of 130 configurations,
-with LTGA, $1+(\lambda, \lambda)$, Hill Climber, hBOA outperforming P3 on 44, 18, 10, and 1 configurations, respectively.
-On
-Deceptive Trap, Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT P3 still finds the global
-optimum faster than all comparison techniques. LTGA is able to match P3's performance almost exactly on HIFF,
-and on Rastrigin P3 falls behind only on moderate sized problems. Deceptive Step Trap yields the worst performance,
-with LTGA using optimal population sizes having a significantly lower growth curve than P3. This mirrors
-the results from Figure~\ref{fig-evals-to-success} under the assumption that P3's computational time per evaluation is higher
-than LTGA but by no more than a constant factor.
-
-\section{Population Sizing}
-A major advantage to P3 is that it does not require the user to set a population size parameter.
-Beyond making P3 easier to apply, this also conveys two additional advantages: diversity scaled
-to initialization and no need to sacrifice intermediate fitness for eventual optimality.
-
-\subsection{Problem Instance versus Problem Class}
-Our procedure for tuning LTGA and hBOA outlined in Section~\ref{sec-tuning} involved finding
-the optimal population size for each class of problem. For real world black box optimization this is
-realistically the best either algorithm could hope for as tuning to a problem
-instance or population initialization involves repeatedly solving the problem being tuned.
-This limitation does not exist in parameter-less methods, which scale their diversity
-based on the problem instance without needing to solve that instance repeatedly.
-
-To achieve high success rates on randomly generated problem classes, LTGA and hBOA must use a population size
-large enough to solve the hardest instances in that class. Therefore these methods will have
-population sizes larger than necessary to solve the easiest instances in the class. Even
-on single instance problems, both methods will require population sizes large enough to
-ensure the worst random initialization is diverse enough to solve the problem, which may be
-much larger than the best random initialization.
-
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{evals-to-success-range}
-  \end{center}
-  \caption{Comparison of the upper and lower quartiles of evaluations required
-           to reach the global optimum for P3 and LTGA with respect to problem size.}
-  \label{fig-evals-to-success-range}
-\end{figure}
-
-Figure~\ref{fig-evals-to-success-range}
-highlights how this can effect the required number of evaluations to reach the global optimum, showing
-the upper and lower quartiles for P3 and LTGA. On each problem
-LTGA has a much smaller difference between its best and worst runs. This makes sense as LTGA uses the same
-population size regardless of instance and progresses search generationally. In contrast P3 has a much
-higher split, with many runs finishing very quickly. This added variability is also generally on the lower
-quartile, with P3's upper quartile only worse than LTGA's on Deceptive Step Trap and Nearest
-Neighbor NK. This supports the hypothesis that P3 is able to scale its diversity not just
-to the problem class, but to the problem instance or even problem initialization, something
-wholly infeasible for tuned population sizing to do.
-
-This tuning distinction is also apparent when comparing Parameter-less hBOA with hBOA in Figure~\ref{fig-evals-to-success}.
-While generally performing worse than hBOA, the difference between the two algorithms is smallest
-on randomly generated problem classes. On MAX-SAT, Parameter-less hBOA actually outperformed both hBOA and LTGA,
-likely due to its ability to scale diversity to the problem instance instead of the entire problem class.
-
-\subsection{Fast versus Optimal}
-\label{sec-fast-vs-optimal}
-
-In Section~\ref{sec-overtime} we examined intermediate fitness qualities of LTGA and hBOA when using
-population sizes tuned to reach the global optimum. As a result, both were exceptionally ineffective
-at quickly reaching high quality solutions. This is because unlike P3, these methods have an explicit
-trade off between optimal performance and intermediate performance caused by their population size parameter.
-
-At 58\% of the recording intervals hBOA has the worst fitness of any solver. Most of the exceptions
-occur when hBOA is still evaluating its initial population, allowing this random search to temporarily surpass the
-local search methods. After $N$ evaluations, however, hBOA and LTGA both fall behind until their models begin
-to improve.
-Parameter-less hBOA reaches intermediate fitnesses faster than hBOA, doing so on 62\% of intervals, as its models begin
-to optimize earlier than hBOA.  However, this trend is reversed after a sufficient number of evaluations, most clearly
-on Deceptive Step Trap and Ising Spin Glasses, as hBOA's tuned population overtakes Parameter-less hBOA's parallel populations.
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{small-pop-dst}
-      \end{centering}
-      \caption{Deceptive Step Trap 203}
-      \label{fig-small-pop-dst}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{small-pop-nk}
-      \end{centering}
-      \caption{Nearest Neighbor NK 200}
-      \label{fig-small-pop-nk}
-    \end{subfigure}
-  \end{centering}
-  \caption{Comparison of how reducing LTGA's population size effects
-           the median best fitness reached during search.}
-  \label{fig-small-pop}
-\end{figure}
-
-Figure~\ref{fig-small-pop} examines the effect of population size on LTGA's intermediate fitness by examining the effect
-of reducing LTGA's population size to one tenth of the tuned value. The two problems shown are representative of the behavior of using
-a smaller population size on the other five problems. Reducing the population size caused LTGA to improve earlier but plateau
-at lower fitnesses. This caused LTGA's success rate to drop from 100 to 0 on Deceptive Step Trap and from 98 to 68 on Nearest
-Neighbor NK.
-Even when using a reduced population size for LTGA, P3 still achieved a fitness at
-least as high as LTGA at 80\% of intervals.
-
-\begin{comment}
-It is important to note that hBOA and LTGA were tuned specifically
-to find the global optimum and are not necessarily using the optimal population size for finding intermediate fitnesses. Consider
-that for both reducing the population size would almost certainly improve their fitness for evaluation steps less than their population
-size. To examine the effect of population size we tested LTGA using one tenth of the population size required for reliably finding the
-global optimum, with the results given in Figure~\ref{fig-small-pop}.
-Reducing the population size resulted in LTGA's initial fitness improvement
-occurring earlier, as the first generation is completed much more quickly. On Deceptive Step Trap this means that
-LTGA leaps ahead of P3 due to its ability to overcome the two bit fitness plateaus.
-%
-Across all tested problems using the smaller
-population size reduced the quality of LTGA's first fitness plateau, likely due to missing required diversity to reach higher quality.
-The second period of improvement also comes earlier and reaches a lower quality fitness. On Deceptive Step Trap 0
-runs reach the global optimum, down from 100 successful for the full population size. On Nearest Neighbor NK this drops the number
-of successful runs from 98 to 68. As a result we conclude that while reducing the population size of LTGA may improve its quality
-at early points during optimization, doing so reduces the robustness of the final solution found. This is in contrast to P3
-which balances high quality intermediate fitness without trading away eventual optimality.
-\end{comment}
-
-\section{Inner Workings}
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{cross}
-      \end{centering}
-      \caption{Crossover Proportion}
-      \label{fig-cross}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{cross-success}
-      \end{centering}
-      \caption{Crossover Success}
-      \label{fig-cross-success}
-    \end{subfigure}
-  \end{centering}
-  \caption{For each problem Figure~\ref{fig-cross} shows the proportion of P3 evaluations spend on crossovers
-           and Figure~\ref{fig-cross-success} shows the percentage of fitness improving crossover evaluations.}
-\end{figure}
-
-While analysis of optimization speed is useful from a practitioner standpoint, doing so provides very
-little insight into algorithm behavior. To better understand how P3 works in detail we present here
-a look at some internal features specific to P3.
-
-\subsection{Crossover}
-Figure~\ref{fig-cross} shows the proportion of evaluations P3 spends on crossover, as opposed to hill climbing,
-and Figure~\ref{fig-cross-success} shows what percentage of crossover evaluations resulted in a fitness improvement.
-Together these figures provide some insight into the role of crossover within P3. The behavior
-for each is clearly problem dependent and generally asymptoticly stable as problem size increases.
-
-When solving problems where epistasis can be effectively detected and represented by a linkage tree, P3 tends to spend
-less evaluations performing crossover and each crossover is more likely to be successful.
-Deceptive Trap and Rastrigin are the easiest problems to model epistasis, with local search quickly
-reducing pairwise entropy in each. These are also the problems where P3 uses the least evaluations
-on crossover and has the highest success rates for crossover. At the other extreme are Nearest Neighbor
-NK and Ising Spin Glasses, which both have overlapping linkage not representable by a linkage tree.
-These problems have the highest crossover usage and lowest crossover success of any problem except Deceptive
-Step Trap.  While Deceptive Step Trap's epistasis can be accurately modeled by a linkage tree, the
-exponential number of plateaus makes detecting gene linkage very challenging.
-
-\begin{comment}
-P3 performs optimization using a mixture of hill climbing and crossover. Exactly how much time the algorithm
-dedicates to each is not clear from the algorithm description alone. As such, Figure~\ref{fig-cross} shows
-that the relationship between the two is strongly problem dependent but relatively independent of problem size.
-In general, problems which can be more accurately represented by crossover
-required less crossover evaluations. Deceptive Trap and Rastrigin spent the least evaluations on crossover, using
-15\% and 20\% respectively on the largest problem sizes. Both of these problems are perfectly separable and have relatively
-few local optima in each subproblem. HIFF and Ising Spin Glasses use crossover for about 60\% of evaluations. The HIFF structure can
-be perfectly learned by the linkage tree, but the higher order relationships require multiple stages of crossover to optimize.
-Ising Spin Glasses have an overlapping structure, making them impossible to perfectly represent in a single linkage tree, but
-that overlap is relatively minor in comparison with Nearest Neighbor NK, which uses 81\% of evaluations on crossover. Deceptive Step Trap spends the highest
-percentage of evaluations on crossover, 92\%. Unlike all of the other problems, the value taken by an individual bit in Deceptive
-Step Trap has very little meaning. This is because any trap value is either a local optima or 1 bit flip away from being a local optima.
-This has a combined result of meaning it takes very few local search evaluations to find a local optima but very many crossover operations
-to proliferate the global optima. Epistatic relationships are also going to be very weak and even small population sizes are likely
-to mark all clusters as $useful$.
-
-Solutions are only added to the next highest pyramid level if at least one crossover donation resulted in a fitness
-improvement. Figure~\ref{fig-cross-success} examines the percentage of crossover donations that resulted in a fitness
-improvement. The ordering in Figure~\ref{fig-cross-success} is an almost perfect inversion of Figure~\ref{fig-cross},
-implying that the more successful crossover is at achieving fitness improving donations the lower the proportion of evaluations P3 will
-spend performing crossover. P3 achieves an impressive 30\% success rate for crossover donations on the largest Deceptive Trap
-problem sizes. This makes sense as crossover is able to immediately and perfectly identify entire traps such that all donations
-always contain only complete traps. Furthermore, by forcing crossover to attempt donations of genes different from the current solution,
-donations containing a single trap will have a very high success rate, always improving suboptimal traps which are far more frequently
-created by local search that optimal traps. Rastrigin and HIFF have relatively high success rates of 11\% and 7\%, respectively. Again
-this is likely due to accurate modeling and relatively low diversity of building block alternatives.
-Ising Spin Glasses and Deceptive Step Trap both have low success rates near 0.01\%. This follows as both are difficult to model with
-a linkage tree. In comparison to Figure~\ref{fig-cross-success} Deceptive Step Trap performs better than expected for crossover.
-However, the reason for this is that even exceedingly uninformed crossover can make improvements to Deceptive Step Trap as
-flipping two bits in the same trap will move between two local optima. Nearest Neighbor NK had the least successful crossovers,
-with only 0.007\% successful on the largest tested problems. In comparison to Ising Spin Glass, NK's overlap is far more extensive.
-As a result it makes sense that individual crossovers are very unlikely to be fitness improvements.
-\end{comment}
-
-When compared with LTGA, P3's crossover success rates are lower but similar in problem ordering.
-Counterintuitively, the use of hill climbing on the initial population reduces
-P3's crossover success, not because it reduces model quality or the donation pool, but because it is much more challenging to improve
-locally optimal solutions than randomly generated ones. LTGA's crossover benefits from application to unoptimized solutions, which
-makes its aggregate crossover success incomparable to P3's.
-
-Even when crossover success rates are quite low, such as Nearest Neighbor NK's 0.007\% success,
-the results from Section~\ref{sec-optimum} and Section~\ref{sec-overtime} show it's still critical to optimization. Without
-crossover, P3's performance would be identical to the Random Restart Hill Climber, which was unable to solve even moderately sized problems
-and quickly fell behind P3 in intermediate fitness quality. Therefore even infrequently successful crossover donations are
-critical to success. This does, however, suggest a potential avenue for future improvement by using more successful
-modeling and donation algorithms.
-
-\subsection{Pyramid}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{level-size}
-      \end{centering}
-      \caption{Population Size}
-      \label{fig-level-size}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{level-success}
-      \end{centering}
-      \caption{Crossover Success}
-      \label{fig-level-success}
-    \end{subfigure}
-  \end{centering}
-  \caption{For each problem Figure~\ref{fig-level-size} shows the number of solutions stored in each level of the pyramid
-           and Figure~\ref{fig-level-success} shows the percentage of fitness improving crossover evaluations at each level.}
-\end{figure}
-
-Another feature unique to P3 is the shape and size of the population pyramid
-constructed for each problem. Figure~\ref{fig-level-size} shows the number of solutions stored at
-each level of the pyramid for the largest tested problem sizes. Each point is the
-median size across 100 runs. If a run did not store any solutions at a level it is treated as 0.
-No point is drawn if the median run had 0 solutions stored at that level. While pyramid size
-is effected by problem size, the overall shape is not. As such the behavior shown in Figure~\ref{fig-level-size}
-is representative of that for all other tested problem sizes.
-
-With the exception of the dip in Deceptive Step Trap, all of the pyramids show a monotonic reduction
-in size as the level increases. This is because a solution must be a strict fitness improvement
-over its previous version to be added to a higher level, which becomes less likely
-each time the solution improves. \cite{lobo:2011:dynamicpop} found theoretical evidence
-and \cite{goldman:2011:dynamic-parameters} found empirical evidence that the optimal
-population size decreases each generation of traditional evolutionary search.
-By decreasing in size, P3 implicitly stores more diversity in low levels and focuses
-search on high quality solutions at higher levels. In comparison,
-LTGA and hBOA suboptimally use a fixed population size at each generation.
-
-Figure~\ref{fig-level-success} examines how crossover success changes at different
-levels of the pyramid. At low levels, success gets progressively lower
-as solution quality increases faster than the model's ability to improve solutions.
-At higher levels modeling becomes more accurate and donations contain higher frequencies
-of high quality building blocks, resulting in increased crossover success. The
-highest level of most problems has a low crossover success rate, as solutions
-crossing with that level have already been improved by previous operations to
-the point where the only improvement would be to create the global optimum, which can
-only happen once.
-
-\subsubsection{Deceptive Step Trap}
-The number of solutions stored at the fourth level of Deceptive Step Trap is
-significantly lower than that of the third or fifth levels, breaking the decreasing
-trend of the other six problems. Figure~\ref{fig-level-success} has a similar
-aberration, with crossover success dropping to  0.0004\% before rebounding and
-following the more common trajectory. This behavior exists in all other problem
-sizes tested, with the dips occurring at exactly the same level.
-
-This behavior is rooted in the peculiar nature
-of this landscape.
-After local search, all traps in all solutions have a total
-number of 1 bits equal to \{0, 1, 3, 5, 7\} as these correspond to the local
-optima when using $k=7$ and $s=2$. Crossover is very likely to overcome the
-two bit plateaus, and as a result solutions in the second level generally do not
-contain the lowest fitness local optima (5 bits set) and the third
-level has very few traps set to the next worst local optima (3 bits set). As a result,
-solutions which reach the third level can only be improved by replacing the deceptive
-local optima (0 and 1 bits set) with the global optimum (7 bits set).
-The global optimum is very rare in the population, and with 8 ways to represent local optima
-linkage learning is inaccurate. Therefore it is very unlikely for crossover to be successful,
-meaning few solutions will be added to the fourth level. Solutions that do improve by
-definition must have a higher frequency of optimal trap settings, meaning level four's model
-will be more accurate and donations are more likely to contain optimal trap values.
-Thus the level size and crossover success rates increase after contracting around level four.
-
-\section{Conclusions and Future Work}
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing black box
-optimization. P3's primary innovation is the replacement of the generational model with a pyramid of populations.
-This pyramid is constructed iteratively, with both the number of levels and the number of solutions stored
-at each level growing as search progresses. P3 uses a model based crossover method
-which learns a linkage tree from gene epistasis. Combined with a simple hill climber, P3's design contains
-many synergistic features.
-
-Across a large number of problems and problem sizes P3 required less evaluations to reach the global optimum
-than optimally tuned state-of-the-art competitors. On single instance problems P3's improvement was by a
-constant factor, while for the three problem classes P3's improvement increased with problem size. This
-quality extends to intermediate points during evolution, with P3 generally reaching at least as high
-of fitness as the competitive techniques when using the same number of evaluations. While P3 does require
-modeling overhead, the expense of this overhead is approximately linear with respect to genome size. There
-is some evidence that even when compared on wall clock time, P3 performs at least as well as
-the best comparison techniques. All of these achievements are made without any problem specific parameter
-tuning, making P3 easier to apply to new domains than its two closets competitors in quality.
-
-P3's quality is due to a number of desirable traits. First, mixing local search with model based
-crossover lets search focus on properly mixing high quality solutions. Second, by adding diversity only
-as necessary P3 tends to use the minimal amount of random initialization, unlike other techniques which must
-overcompensate with larger population sizes on single instance problems and consider the worst instance
-when solving problem classes. Third, by heavily exploiting existing diversity before adding more P3 is able
-to reach high quality intermediate fitnesses quickly without prematurely converging. Fourth, the very
-nature of the pyramid's shape allows search to preserve a desirable proportion of diversity at
-each fitness level, similar to a generational model using a decreasing population size.
-
-There are a number of meaningful avenues for future P3 experimentation. Perhaps the most
-pressing for practitioner acceptance is to apply P3 to real world problems and compare
-its results with other black box or even problem specific heuristics. While parameter-less,
-P3 is currently limited to discrete, fixed length genomes evaluated using single objective
-fitness. These limitations can be relaxed with future work to make P3 more widely applicable.
-While asymptotically linear, P3's modeling techniques and local search methods are likely going
-to be prohibitively expensive for genome sizes in the hundreds of thousands or millions of genes,
-and the inability of the model to capture overlapping linkage may be hindering
-search efficiency. Overcoming these limitations by using a new modeling technique may allow
-the pyramid model even greater flexibility. Similarly, while P3 is able to overcome low
-order deception via linkage learning, the iterative improvement method by which crossovers
-are made may mislead search on landscapes with higher order deception.
-
-However, even without
-these improvements our results show P3 is highly efficient at finding
-global optima on black box problems without any problem specific tuning.
-
-\small
-
-\bibliographystyle{apalike}
-\bibliography{../main}
-
-\normalsize
-\appendix
-\section{hBOA Simplification}
-\label{sec-appendix}
+\chapter{hBOA Simplification}
+\label{chap-appendix}
 To measure the quality of a decision forest, hBOA applies Equation~\ref{eq-hboa}.
@@
 therefore it is imperative that implementations avoid rounding errors and overflows.
-
-\end{document}

### 2014-10-15 Comprehensive/chapter-benchmarking.tex
Initial commit of Comprehensive Report

--- 2014-10-15 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
@@
-\documentclass[twoside]{article}
-\usepackage{ecj,palatino,epsfig,latexsym,natbib}
-\usepackage{url}
-\usepackage{verbatim}
-\usepackage{subcaption}
-\usepackage[noend]{algpseudocode}
+\chapter{Benchmarking P3}
+TODO Explain the purpose of this chapter
 
-\newcommand{\includegraphicswide}[1]
-{\includegraphics[width=.9\textwidth,height=\textheight,keepaspectratio]{#1}}
-
-\newcommand{\includegraphicsfit}[1]
-{\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}
-
-\usepackage{array}
-\newcolumntype{L}{>{\centering\arraybackslash}m{1.9cm}}
-
-%% do not add any other page- or text-size instruction here
-
-\parskip=0.00in
-
-\begin{document}
-
-% 45-character limit on header
-\ecjHeader{x}{x}{xxx-xxx}{201X}{Parameter-less Population Pyramid}{B. W. Goldman and W. F. Punch}
-\title{\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid}
-
-\author{\name{\bf B. W. Goldman} \hfill \addr{brianwgoldman@acm.org}\\ 
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-\AND
-       \name{\bf W. F. Punch} \hfill \addr{punch@msu.edu}\\
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-}
-
-\maketitle
-
-\begin{abstract}
-%Should be about 200 words
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
-evolutionary optimization without requiring any user specified parameters.
-P3's primary innovation is to replace the generational model with a pyramid of
-multiple populations that are iteratively created and expanded. In combination
-with local search and advanced crossover P3 scales to problem difficulty, exploiting
-previously learned information before adding more diversity.
-
-Across seven problems, each tested using on average 18 problem sizes, P3 outperformed
-all five advanced comparison algorithms. This improvement includes requiring less evaluations
-and less wall clock seconds to find the global optimum and better fitness when using
-the same number of evaluations. Using both algorithm analysis and comparison we find P3's
-effectiveness is due to its ability to properly maintain, add, and exploit diversity.
-
-Unlike the best comparison algorithms, P3 was able to achieve this quality without any
-problem specific tuning. Thus, unlike previous parameter-less methods, P3 does not
-sacrifice quality for applicability. Therefore we conclude that
-P3 is an efficient, general, parameter-less approach to black box
-optimization which is more effective than existing state-of-the-art techniques.
-
-\end{abstract}
-
-\begin{keywords}
-
-Genetic algorithms, 
-linkage learning,
-local search,
-parameter-less.
-
-\end{keywords}
-
-\section{Introduction}
-A primary purpose of evolutionary optimization is to efficiently find good solutions
-to challenging real world problems with minimal prior knowledge about the problem itself.
-This driving goal has created search algorithms which can escape user bias to create
-truly novel results, sometimes publishable or patentable in their own right~\citep{kannappan:2014:humies}.
-While it is not possible for any algorithm to do better than random search across all possible
-problems~\citep{Wolpert:1997:nfl}, effectiveness can be achieved by assuming the search
-landscape has structure and then biasing the algorithm toward exploiting that structure.
-
-In evolutionary optimization, and genetic algorithms~(GAs) in particular, search is often
-biased through parameters. This can be beneficial as it allows practitioners to inject their
-knowledge about the shape of the search landscape into the algorithm.
-However, the quality of solutions found, and the speed at which they are found, is strongly tied to setting these parameters
-correctly~\citep{goldberg:1991:gasize}. As such, either expert knowledge or exceedingly
-expensive parameter tuning~\citep{grefenstette:1986:optimalga} are required to leverage
-this feature to its fullest potential. Furthermore,  parameters such as population size, mutation rate, crossover
-rate, tournament size, etc. usually have no clear relationship to the problem being solved, meaning even
-domain experts may not understand how the parameters will interact with the problem or with each other.
-
-There have been periodic efforts to reduce or remove the need for parameter tuning.
-\cite{Back:1992:selfadapt} introduced self-adaptive parameters, in which parameter values
-were included in each solution's genome and themselves underwent evolution. This allowed search
-to optimize some of its own parameters resulting in a reduced the need for expert tuning.
-\cite{harik:1999:parameterlessga} were able to design an entirely parameter-less GA by
-leveraging schema theory and parallel populations. Unfortunately these methods were provably less efficient
-than directly setting the parameters to optimal values~\citep{pelikan:1999:worstparameter-less}.
-
-One area that has been very effective at reducing the number of algorithm parameters is
-model based search. \cite{pelikan:2006:hboa}'s Hierarchical Bayesian Optimization
-Algorithm~(hBOA) and \cite{thierens:2010:ltga}'s Linkage Tree Genetic Algorithm~(LTGA)
-both require only a single parameter: population size. \cite{posik:2011:parameterless}
-leveraged model building to create a fully parameter-less algorithm, but it is restricted to
-only order-k, fully decomposable, noiseless problems.
-
-Most recently \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid~(P3).
-This method uses a pyramid structure of populations to combine model based search with local search
-to achieve parameter-less optimization. Initial results suggest that unlike
-previous parameter-less methods, P3 is actually more efficient than current state-of-the-art
-parameterized search algorithms. In this work we shall: extend these results to cover more
-comparison algorithms; compare both efficiency in reaching the global optimum and intermediate
-fitnesses; algorithm complexity analysis; and provide more in depth analysis of P3 itself.
-
-\section{Comparison Optimizers}
-\label{sec-optimizers}
-
-In order to fully understand the effectiveness of P3, we compare it with five advanced algorithms which
-have related features to P3. The Random Restart Hill Climber~\citep{goldman:2014:p3} was chosen as an efficient form of repeated
-local search. As P3 combines this hill climber with crossover,
-comparing with local search alone shows the advantages of P3's overall approach.
-The $1+(\lambda, \lambda)$ algorithm is the current best theory supported simple genetic algorithm
-and its method of crossover is in some sense a macro-mutation just as in P3. hBOA and Parameter-less
-hBOA are advanced model building search techniques which are very effective at learning complex
-problem structure, designed to achieve similar goals as P3's linkage learning but using very different
-methods. Finally LTGA represents the current state-of-the-art in black box search and is the origin
-of P3's linkage learning and crossover methods.
-
-Finally, only hBOA and LTGA require any parameters, with each of these only requiring a population
-size. This makes knowing the optimal behavior of these algorithms much more tractable. All of the
-algorithms are also gene order independent, fitness scale invariant, and unbiased. This means,
-for any problem, the order in which problem variables appear in the genome can be changed
-without changing the behavior of the search. The fitness can also be manipulated in any fashion
-as long as the rank ordering of solutions is unchanged. These algorithms are also unaffected by the meaning
-assigned to each bit, such that inverting a predetermined random subset of genes before evaluation
-will not impact search efficiency.
-
-Our implementations of all of these algorithms as well as all of the population size information, raw results,
-and processing scripts
-are available from our website.\footnote{\url{https://github.com/brianwgoldman/FastEfficientP3}}
-\subsection{Random Restart Hill Climber}
-\label{sec-hill-climber}
-Perhaps the simplest black box search heuristic is stochastic local search, or hill climbing.
-This optimization technique focuses on improving a single solution until it reaches a local
-optimum. Here we use the first improvement hill climber defined by \cite{goldman:2014:p3}
-and given in Figure~\ref{fig-hc}. This algorithm works by flipping each bit in a random
-order, keeping modifications when fitness is improved, until single bit flips cannot result in
-further fitness improvements.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Hill-Climber}{}
-    \State $options \leftarrow [0 \dots N-1]$\label{fig-hc-options}
-    \State $tried \leftarrow \emptyset$
-    \While{$|tried| < |options|$}
-      \ForAll{$index \in shuffled(options)$}
-        \If{$index \notin tried$}\label{fig-hc-tried}
-          \State Flip bit $index$ in solution
-          \If{solution's fitness increased}
-            \State $tried \leftarrow \emptyset$
-          \Else
-            \State Revert change
-          \EndIf
-          \State $tried \leftarrow tried \cup \{index\}$
-        \EndIf
-      \EndFor
-    \EndWhile
-  \EndProcedure
-\end{algorithmic}
-  \caption{Hill climbing algorithm used to improve randomly generated solutions until no single
-           bit change results in a fitness improvement.}
-  \label{fig-hc}
-\end{figure}
-
-The hill climber requires an amortized cost of $O(1)$ operations per evaluation. In order to
-terminate, at least one evaluation must be performed for each of the $N$ bits in the solution.
-As such any operation that happens only once per search can be amortized over at least $N$
-evaluations, covering the initialization of $options$ on Line~\ref{fig-hc-options}.
-Line~\ref{fig-hc-tried}, which prevents wasted evaluations, can be called at most twice
-per evaluation: once to add $index$ into $tried$ and once to prevent $index$ from being unnecessarily
-evaluated again. The only way three or more calls could happen is if
-no fitness improvement was made for the entire previous iteration, which contradicts the loop invariant.
-
-Due to its nature, this hill climber cannot escape basins of attraction. Once a solution is reached
-such that none of the single bit neighbors are fitness improvements, search stops. Thus
-this algorithm requires a restart mechanism to solve multimodal problems. We have
-chosen here to naively restart search from a random solution whenever a local optima is found. This
-ensures that on all landscapes there is always a non-zero probability of search finding the global optimum.
-
-\subsection{$1+(\lambda, \lambda)$}
-\cite{doerr:2013:lambdalambda} presented the first genetic algorithm to provably show
-the advantages of performing crossover on simple landscapes. This comparatively simple
-algorithm maintains only a single individual and a self-controlled parameter $\lambda$.
-
-Each iteration, the number of bits to flip is chosen from the binomial distribution $b\sim B(N, \frac{\lambda}{N})$,
-where $N$ is the number of bits in the genome.
-Next, $\lfloor\lambda\rfloor$ offspring are produced by flipping $b$ bits. The
-best mutant then produces $\lfloor\lambda\rfloor$ offspring via uniform crossover with the original parent, such that each gene comes from the
-mutant with probability $\frac{1}{\lambda}$. In the original algorithm the best
-offspring produced by crossover then replaces the original parent if its fitness is no worse.
-The $\lambda$ parameter, which is initialized to 1, is decreased if the offspring replaced
-its parent and increased otherwise.
-
-The original formulation was designed specifically for unimodal landscapes and as such were
-not directly suitable for multimodal problems. \cite{goldman:2014:p3} extended $1+(\lambda, \lambda)$
-to include random restarts. As search stagnates, the $\lambda$ parameter increases in value. Eventually
-this results in $\lambda \ge N$ causing mutation to always flip all bits of the individual.
-As this prevents any future improvement, whenever $\lambda \ge N$ search is restarted from a random solution with $\lambda$
-reset to 1.
-
-A few other efficiency modifications were also made. If there is a tie in crossover offspring fitness,
-whichever has a larger hamming distance from the parent is retained. This encourages drifting across plateaus.
-The ``mod'' control strategy proposed by \cite{doerr:2013:lambdalambda} was not used as it conflicted with
-the random restart strategy.
-If a crossover individual is identical to either of its parents, it is not evaluated.
-If mutation produces an offspring which is better than the best crossover offspring, it is used to compare
-against the original parent.
-
-\subsection{Hierarchical Bayesian Optimization Algorithm}
-
-\cite{pelikan:2006:hboa} used statistical principles in combination with a decision tree structure
-to create the Hierarchical Bayesian Optimization Algorithm (hBOA). This method creates a model of
-epistatic relationships between genes which is then used to stochastically generate new solutions.
-Each generation a binary tournament with replacement is used to select $\mu$ solutions from
-the population. These solutions are then used to build the model, which in turn is used to generate $\mu$ new
-solutions. The new solutions are then integrated into the population using restricted tournament
-replacement.
-
-Conceptually, the model built by hBOA is trying to infer rules of the form ``Given that this
-subset of genes are set to these values, how frequently is gene $x_i$ set to value v?'' This can
-be represented using a directed acyclic decision forest, with each tree in the forest representing one gene
-in the solution. In the decision tree $T_i$, which is used to set the value of gene $x_i$,
-each internal node represents previous decisions on how to set
-some other gene $x_j$, with the children of that node representing how the decision was made. The
-leaves of each tree give the probability that $x_i$ should be set to each possible value.
-
-The forest is constructed iteratively, with each tree initially containing a single leaf
-and with each leaf storing a pointer for each selected solution. Each iteration the algorithm considers
-all possible ways of splitting an existing leaf using another gene $x_j$, such that solutions in the
-leaf are moved to the newly created leaves based on their value for $x_j$. The general goal is to
-separate the solutions such that all solutions with $x_i = 0$ move to one leaf while solutions with
-$x_j = 1$ move to the other.
-
-This goal is formalized using model scoring from Bayesian statistics. In its raw form this
-almost always creates near infinitesimal results, calculating fractions which include factorials of $\mu$
-and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{sec-appendix}, we
-derived a simplified form shown in Equation~\ref{eq-hboa-final}. If no proposed split satisfies the inequality, iteration stops.
-If multiple splits do, whichever maximizes the right side is chosen.
-
-\begin{equation}
-  2^{0.5 log_2\mu} < \frac{(m_i(l) + 1)!}{m_i(0, l)!m_i(1,l)!} \cdot
-  \frac{m_i(0, l')!m_i(1,l')!m_i(0, l'')!m_i(1,l'')!}{(m_i(l') + 1)!(m_i(l'') + 1)!}
-  \label{eq-hboa-final}
-\end{equation}
-
-Initially there are $\Theta(N^2)$ possible ways to split existing leaves, as each of the $N$ single node
-trees can be split by any of the other $N-1$ genes. Each iteration a new edge is added to the decision
-forest, meaning some of the previously tested splits cannot be used. For instance, if $T_i$, which is used
-to decide the value of $x_i$, is split using the value of $x_j$, $T_j$ can no longer be split using $x_i$.
-As a split creates two new leaves, $O(N)$ new potential splits must also be tested. Equation~\ref{eq-hboa-final}
-parses all solutions which reach a leaf to count gene frequencies, requiring $\mu$ time.
-The number of total leaves created depends heavily on the problem and $\mu$.
-However, assuming no splits are accepted or that the cost of testing all future splits is less than
-the initial $\Theta(N^2)$, constructing the model requires $\Omega(\mu N^2)$ time. Each
-model is used to generate $\mu$ solutions, leading to a cost per evaluation of $\Omega(N^2)$.
-
-To generate a solution, the value of each gene $x_i$ is set using its corresponding decision tree $T_i$. Because
-the forest is directed acyclic, there must be an ordering of $T_i$ such that before $T_i$ is executed all
-$x_j$ it uses to make decisions have already been set. As such, previous decisions made by other trees
-are used to follow each $T_i$ until a leaf is reached. The value of $x_i$ is then set based on the
-probability that other solutions reached that leaf with each value of $x_i$.
-
-To perform replacement, hBOA uses restricted tournament replacement. After each new solution is generated
-and evaluated, a set of $w$ solutions are chosen at random from the population, where $w=\min\{N, \frac{\mu}{20}\}$.
-From this set the solution which is most genetically similar to the offspring is chosen. If the offspring
-is at least as fit as the chosen solution, it replaces the chosen solution in the population. Otherwise the
-offspring is discarded. This method is designed to preserve genetic diversity as only genetically similar
-solutions must compete on fitness.
-
-hBOA is designed to work with large population sizes, resulting in a large number of
-evaluations per generation. As hBOA utilizes explicit diversity maintenance, standard methods for determining
-convergence are not considered very accurate. Therefore the authors suggest that an hBOA run should be
-terminated after performing generations equal to $N$.
-
-Like other model based techniques, hBOA has very few parameters. There is no mutation or crossover,
-and modeling does not rely on any explicit parameters. Solution selection, generation, and replacement
-are all derived from the population size, which must be set by the user.
-
-\subsection{Parameter-less hBOA}
-Using the methods first introduced by \cite{harik:1999:parameterlessga} for the Parameter-less GA,
-\cite{pelikan:2004:parameterlesshboa} created Parameter-less hBOA which automatically scales its
-population size to fit the problem. This is done by maintaining a list of concurrent populations
-using exponentially scaled population sizes.
-
-A run of Parameter-less hBOA starts with a single population of size $\mu_0$, conventionally set
-to $\mu_0=10$. After two generations are performed, a new population of size $\mu_1 = 2\mu_0$ is created
-and performs a generation. Evolution then continues with the $\mu_0$ population performing two generations
-for each one performed by $\mu_1$. Each time population $\mu_i$ performs its second generation a new population
-$\mu_{i+1}=2\mu_i$ is created, which performs generations half as often as $\mu_i$. In this way an infinite number of
-parallel population can be simulated, with each population receiving the same number of total evaluations.
-
-In all other aspects each population is identical to an hBOA population using a fixed $\mu$. No search information
-is shared among populations, and each search is independently terminated. As such Parameter-less hBOA cannot
-perform better than hBOA using the optimal population size for a given instance, as it must also spend evaluations
-on populations of different sizes. This inefficiency is bounded by a log multiple of the total number of
-evaluations~\citep{pelikan:1999:worstparameter-less}.
-
-\subsection{Linkage Tree Genetic Algorithm}
-\cite{thierens:2010:ltga} introduced the Linkage Tree Genetic Algorithm (LTGA) which automatically
-detects and exploits problem epistasis by examining pairwise gene entropy. Due to its enhanced
-ability to preserve high fitness gene subsets, LTGA was able to outperform state-of-the-art
-GAs across many benchmarks. Since its introduction, many variants of LTGA have been
-proposed~\citep{goldman:2012:ltga} so for clarity we have chosen the version
-presented by \cite{thierens:2013:ltgahiff} as our model.
-
-LTGA's effectiveness comes from its method of performing crossover. Instead of blindly
-mixing genes between parents, LTGA attempts to preserve important interrelationships
-between genes. Before performing any crossovers in a generation, LTGA first builds
-a set of hierarchical gene clusters which are then used to dictate how genes are mixed
-during crossover.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Cluster-Creation}{}
-    \State $unmerged \leftarrow \{\{0\}, \{1\}, \{2\}, \dots, \{N-1\}\}$
-    \State $useful \leftarrow unmerged$
-    \While{$|unmerged|>1$}
-      \State $C_i, C_j \leftarrow \min_{C_i,C_j \in unmerged} D(C_i, C_j)$
-      \State $unmerged \leftarrow unmerged - \{C_i, C_j\} + \{C_i \cup C_j\}$
-      \State $useful \leftarrow useful + \{C_i \cup C_j\}$
-      \If{$D(C_i, C_j) = 0$}
-        \State $useful \leftarrow useful - \{C_i, C_j\}$
-      \EndIf
-    \EndWhile
-    \State Order $useful$ based based on last merged first\label{fig-cluster-creation-ordering}
-    \State Remove largest cluster from $useful$
-
-    \Return $useful$
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm describing how LTGA creates clusters using Equation~\ref{eq-distance}
-           for a population. $unmerged$ and $useful$ are ordered sets of sets of gene loci.}
-  \label{fig-cluster-creation}
-\end{figure}
-
-Figure~\ref{fig-cluster-creation} provides the agglomerative method LTGA uses to create gene clusters.
-This algorithm creates a tree of sets using pairwise gene entropy, such that the leaves of the tree contain a single gene and
-internal nodes are the union of their children's sets. These sets are then used by crossover to specify epistatic
-relationships which should be preserved.
-The process begins by creating the set of sets $unmerged$ which tracks all top level clusters. Initially
-$unmerged$ contains single member sets for each gene. After each iteration the two sets with the minimum average pairwise
-distance (given in Equation~\ref{eq-distance}) are merged to create a single cluster. This process is repeated
-until only a single set remains in $unmerged$ which contains all of the genes in the genome.
-
-\begin{equation}
-  D(C_i,C_j) = \frac{1}{\left | C_i \right |\cdot \left |C_j \right|}\sum_{c_i \in C_i}\sum_{c_j \in C_j}
-  2 - \frac{H(c_i) + H(c_j)}{H(c_i \cup c_j)}
-  \label{eq-distance}
-\end{equation}
-\begin{equation}
-  H(c) = -\sum_{s\in S} p_c(s)\log(p_c(s))
-  \label{eq-entropy}
-\end{equation}
-
-Throughout this process $useful$ tracks the set of all gene clusters which should be preserved for use by crossover.
-This set begins with all genes in separate clusters, and each time a new cluster is created it is added to $useful$.
-However, not all clusters are necessarily worth keeping. For instance, in all versions of LTGA the cluster
-containing all genes is removed from $useful$ as preserving all genes during crossover can only create clones.
-\cite{thierens:2013:ltgahiff} extended this removal to include any unsupported subsets. If the pairwise distance
-between two clusters is 0, this means there are no individuals in the population which disrupt the relationships between the two
-clusters. Therefore when performing crossover there is no reason to believe a fitness improvement can be achieved
-by breaking the stored pattern. As such a cluster is only kept if its direct superset has a non-zero distance.
-As a final step Line~\ref{fig-cluster-creation-ordering} reorders $useful$ such that clusters appear in reversed
-order from which they were added to $useful$. Thus the most linked clusters and those containing single genes
-appear at the end of the returned list.
-
-\cite{thierens:2013:ltgahiff}'s version of LTGA does not use the entire population when determining pairwise entropy.
-Instead, binary tournament is used to select half of the population. This is done to ensure the model is built
-using only high quality solutions, even during the first generation.
-
-In order to efficiently perform clustering, a pairwise gene frequency table is constructed from the selected solutions.
-To calculate Equation~\ref{eq-distance}, Equation~\ref{eq-entropy} is called for each
-gene ($H(c_i)$) and pair of genes ($H(c_i \cup c_j)$). Extracting this information requires $O(\mu N^2)$
-time, where $\mu$ is the population size and $N$ is the genome size. The process of converting
-this pairwise frequency information into clusters can be achieved in $O(N^2)$ using the bookkeeping
-methods presented by \cite{gronau:2007:upgma}. This cost is performed only once per generation,
-and is then used to perform approximately $O(\mu N)$ crossover evaluations. As a result, the amortized cost of
-LTGA's model building is $O(N)$.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Cluster-Usage}{}
-    \ForAll{$C_i \in useful$}
-      %\ForAll{$d \in shuffled(P_i)$}
-        \State $d \leftarrow rand\_choice(P)$\label{fig-cluster-usage-donate}
-        \State Copy $d$'s gene values for $C_i$ into solution
-        \If{solution was changed}
-          \If{solution's fitness decreased}
-            \State Revert changes
-          \EndIf
-          %\State \textbf{break}
-        \EndIf
-      %\EndFor
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm describing how clusters are used to perform crossover.}
-  \label{fig-cluster-usage}
-\end{figure}
-
-Figure~\ref{fig-cluster-usage} describes how the identified clusters are used by crossover to preserve
-gene linkage while still exploring the search space. Unlike more traditional crossover methods, LTGA
-crosses each individual with the entire population. Also, to produce a single offspring, multiple evaluations
-of the fitness function are performed.
-
-Each generation, each individual in the population undergoes crossover. In a single crossover event, each
-cluster of genes $C_i$ in $useful$ is applied as a crossover mask. A random donor $d$
-is chosen from the entire population (not just the model selected population), and $d$'s gene's for $C_i$ are copied
-into the working solution. If a modification is made, an evaluation is then performed. If the crossover
-resulted in no worse fitness then the changes are kept. This allows for neutral drift across plateaus.
-The resulting solution, which must be at least as fit as its parent, is then
-copied into the next generation.
-
-In total each individual can cause up to $|useful|$ evaluations. If all clusters were kept, even those deemed
-unhelpful, and all donations were evaluated, even those which did not change any genes, then \Call{Cluster-Usage}{}
-would perform exactly $2N-2$ evaluations for each of the $\mu$ solutions in the population. This provides the amortizing evaluations
-required to make clustering only $O(N)$ operations per evaluation. However, by skipping some evaluations, its
-possible that clustering may be super-linear.
-
-LTGA has no explicit form of diversity control and has no method for introducing new genetic information once
-the population has converged. Therefore an LTGA run is considered converged once two consecutive populations
-contain the same unique solutions.
-
-By design, LTGA only has a single parameter: population size. LTGA uses no mutation, and crossover is defined
-in terms of the clustering algorithm. Selection between generations is fully elitist and embedded in the crossover,
-with selection of model building solutions fixed to a binary tournament. Neither \Call{Cluster-Creation}{} nor
-\Call{Cluster-Usage}{} rely on parameter values. LTGA does not provide any method for controlling or setting
-the population size, and uses a fixed user specified size for each generation.
-
-\section{Parameter-less Population Pyramid}
-\label{sec-p3}
-\cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid (P3) as a method for
-performing optimization which does not require the user to provide any parameters. This is
-achieved by combining efficient local search with the model building methods of LTGA using
-an iteratively constructed hierarchy of populations.
-
-The high level algorithm of P3 is presented in Figure~\ref{fig-p3}. Unlike more traditional
-GAs, P3 does not follow a generational model. Instead, it maintains an iteratively
-expanding pyramid of expanding populations. Each iteration, a new random solution is generated.
-This solution is brought to a local optimum using the hill climbing algorithm in Figure~\ref{fig-hc}. If that
-local optimum has not yet been added to the pyramid, the solution is added to the lowest
-population $P_0$.
-
-Next, the solution is iteratively improved by applying LTGA's crossover algorithm (Figure~\ref{fig-cluster-usage})
-with each population $P_i$ in the pyramid. If this process results in a strict fitness improvement and has
-created a solution not yet stored in the pyramid, it is added to the next highest pyramid level $P_{i+1}$.
-If $P_{i+1}$ does not yet exist, it is created. In this way populations in the pyramid expand over time,
-and the number of populations stored increases over time. Initially the pyramid contains no solutions
-or populations, meaning the user does not need to specify a population size.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Iterate-P3}{}
-    \State Create random solution
-    \State Apply hill climber
-    \If{solution $\notin hashset$}
-      \State Add solution to $P_0$
-      \State Add solution to $hashset$
-    \EndIf
-
-    \ForAll{$P_i \in pyramid$}
-      \State Mix solution with $P_i$
-      \If{solution's fitness has improved}
-        \If{solution $\notin hashset$}
-          \State Add solution to $P_{i+1}$
-          \State Add solution to $hashset$
-        \EndIf
-      \EndIf
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-  \caption{One iteration of P3 optimization. $pyramid$ is an
-           ordered set of populations and $hashset$ is a set
-           of all solutions in $pyramid$.}
-  \label{fig-p3}
-\end{figure}
-
-To accommodate P3's unique population structure, some of LTGA's clustering procedures were modified. In LTGA,
-clusters are identified at the start of each generation and are used
-to create all offspring in that generation. As P3 does not perform serial
-generations, P3 instead updates the model each
-time a solution is added to a population. Furthermore, unlike
-LTGA all solutions in the population are used to generate the model, not just the
-winners of a binary tournament. We can do this because even the
-worst solutions in the pyramid are already high quality due to local search.
-Using local search in LTGA was examined by \cite{bosman:2011:lsbbo} and found
-to provide no significant improvement. A likely cause was that that study applied
-local search to every solution, not just the initial population, resulting in significant overhead.
-
-Beyond the changes in population structuring, P3 modifies
-LTGA's version of \Call{Cluster-Creation}{} and \Call{Cluster-Usage}{}.
-P3 changes Line~\ref{fig-cluster-creation-ordering} in Figure~\ref{fig-cluster-creation}
-from \emph{last merged first} ordering to \emph{smallest first} ordering.
-This method applies gene clusters during crossover based on how many genes
-are in each cluster\footnote{Ties are broken randomly.}, and not on how tightly linked those genes are.
-\cite{goldman:2012:ltga} found that this alternative was better at preserving
-diversity, and therefore required smaller populations, but has not yet been integrated into
-canonical LTGA.
-
-P3 also modified Line~\ref{fig-cluster-usage-donate} in Figure~\ref{fig-cluster-usage}.
-Instead of choosing a single genetic donor for each cluster, P3 iterates over the
-population in a random order until a solution in the population is found which
-has a least one gene different for that cluster of genes from the improving solution.
-This process increases the likelihood of an evaluation being performed for every cluster,
-and helps test rare gene patterns in the population.
-
-In LTGA the cost of rebuilding the model is $O(\mu N^2)$ as it must collect pairwise
-gene frequency information for all $\mu$ solutions in the population. P3 does not store
-a single population, and it does not have a fixed $\mu$ size for any population. However,
-each time a solution is added to the population, it requires $O(N^2)$ time to update
-the table of pairwise frequencies and another $O(N^2)$ time to rebuild the linkage model.
-The model is then used immediately to perform up to one evaluation for each of the up
-to $2N-2$ clusters. Just as in LTGA, if no evaluation shortcuts were made, P3
-has an amortized cost of $O(N)$ modeling cost per evaluation. While P3 does
-rebuild the model more frequently per solution in the population, it also performs
-a number of local search evaluations which are quite efficient, meaning theoretical
-comparisons of their speed are difficult to perform. As a final note, P3's repeated
-attempts to find a useful donation make it less likely than LTGA to skip evaluations,
-but has an added cost to find these donations. Repeated donations could require as much
-as $O(\mu)$ attempts per evaluation, but experimental evidence suggest that this
-operation actually saves more overhead than it costs by increasing the number of evaluations
-per model rebuild.
-
-Each of the pieces of the P3 algorithm were selected not just for their stand alone efficacy,
-but for the ways in which they interact. By using the hill climber to ensure each solution
-is at a local optima, the underlying pairwise relationships in the problem are exposed. As
-a result, detecting clusters for use by crossover is much more effective. The crossover operator
-is extremely elitist, as each gene donation must result in no fitness loss, and a solution must
-strictly improve to be added to the next level of the pyramid. This is balanced by continual
-integration of new random solutions. Furthermore, each random restart decreases the probability
-of spurious linkages caused by shared ancestry. This diversity is further preserved by applying
-gene clusters in smallest first order during crossover as this reduces the probability of genetic
-hitchhikers.
-
-Other algorithms have proposed using multiple concurrent populations.
-\cite{hornby:2006:alps} had a hierarchy of populations with solutions periodically advancing
-upward. This allows for continuous integration of diversity as the lowest population is reseeded
-with random solutions. However, this method resulted in increased parameterization as not only was
-a population size required, but also parameters for how frequently generations advanced between levels and
-how many total levels to have were added. \cite{harik:1999:parameterlessga} used multiple independent
-populations of different sizes as a method for removing the population size parameter, but
-doing was provably less efficient than using an optimal population size as no information is shared
-between the populations.
-
-\section{Problem Descriptions}
-\label{sec-problems}
-
-\subsection{Single Instance Problems}
-Understanding how a stochastic search algorithm will behave on arbitrary and complex
-search landscapes can be exceedingly difficult. Therefore a common practice for
-algorithm understanding is to perform search on well defined, well understood
-landscapes. To be of interest these landscapes need to represent interesting
-and important aspects of real world problems.
-
-Perhaps the most common such landscape is the Deceptive Trap problem \citep{goldberg:1991:gasize}.
-In this landscape the genome is broken up into $k$ bit non-overlapping subproblems referred
-to as traps.
-Each subproblem is scored using Equation~\ref{eq-deceptive-trap}, where $t$ is the number of
-bits in the trap set to 1. The global optimum in each trap is a string of all 1s, while all
-other solutions lead to a local optima of all 0s. This problem tests an algorithm's
-ability to overcome $k$ sized deception and is commonly used to determine how effective crossover
-is at preserving building blocks. Any crossover event that mixes bits from different parents in
-the same trap will likely result in that trap being optimized to the local optima. For our experiments
-we chose $k=7$ to ensure highly deceptive traps.
-
-\begin{equation}
-   trap(t) = \left\{
-     \begin{array}{rl}
-       k-1-t, &  t<k\\
-       k,   &  t = k
-     \end{array}
-   \right.
-  \label{eq-deceptive-trap}
-\end{equation}
-
-\cite{goldman:2012:ltga} found that mixing local search with linkage learning rendered the Deceptive
-Trap problem trivial. This is because local search is able to optimize each trap to one of the two
-optima (all 1s or all 0s), and then linkage learning can have perfect knowledge of gene interactions.
-In order to make the problem more difficult, the authors proposed the Deceptive
-Step Trap problem, given in Equation~\ref{eq-deceptive-step-trap}. This function modifies the results
-of Equation~\ref{eq-deceptive-trap} to include plateaus of size $s$, introducing an exponential number
-of local optima in each trap. As a result the Deceptive Step Trap is much more challenging for linkage
-learning techniques, while still being highly deceptive. We chose to use $s=2$ to create the maximum
-number of local optima.
-\begin{equation}
-   step\_trap(t) = \left \lfloor \frac{(k-s)\pmod{s} + trap(t)}{s} \right \rfloor
-  \label{eq-deceptive-step-trap}
-\end{equation}
-
-Another challenging aspect of landscapes can be higher order relationships. The Hierarchical If
-and only If (HIFF) problem is designed to capture the difficulties of this class of problem.
-In HIFF the genome is broken up into a complete binary tree, such that each gene appears in exactly one
-leaf and each internal node is the subset of genes contained in its children. If all genes represented
-in a node of the tree are set to the same value, they score equal to the size of the set. In this way
-small subsets lead toward solutions to larger subsets. However, a node can score if all genes are either
-all 1s or all 0s, meaning that to solve higher order subproblems it is necessary to perform crossovers
-that preserve lower order solutions. This problem is a natural fit for LTGA as the linkage tree can
-perfectly duplicate the problem's true relationships \citep{thierens:2013:ltgahiff}.
-
-As a final class of well known problems, we have chosen to borrow the Rastrigin problem from real valued
-optimization. This problem's landscape, determined by Equation~\ref{eq-rast}, is highly multimodal caused by
-the oscillating cosine function. \cite{goldman:2014:p3} proposed the Discretized Rastrigin problem, such
-that each floating point $x_i$ in Equation~\ref{eq-rast} is encoded using a 10 bit gray code.
-\begin{equation}
-  An + \sum_{i=1}^{n}\left [ x_i^2-A\cos (2\pi x_i) \right ] \forall x\in [-5.12,5.12]
-  \label{eq-rast}
-\end{equation}
-
-\subsection{Randomly Generated Problem Classes}
-While well defined landscapes can provide specific insights into how an algorithm works,
-their static nature can be misleading. Specifically, algorithm quality might
-be very fragile such that it is only effective at searching well behaved landscapes. A more realistic
-test of an algorithm's black box effectiveness is to work with randomly generated instances
-drawn from a problem class. When tested over a sufficiently large sample it is then possible
-to draw more general conclusions about the algorithms effectiveness. The challenge with these
-landscapes is determining the global optimum to gauge if an algorithm was successful.
-
-Perhaps the most common model for generating random rugged landscapes is the NK model. An
-NK Landscape determines the fitness of each gene based on epistatic relationships with $K$
-other genes in the genome. This fitness is specified using a randomly generated table of
-fitness values, were each possible combination of the $K+1$ genes is mapped to some floating
-point value $[0-1]$. In unrestricted NK landscapes the relationships between genes are also
-randomly chosen and as a result finding the global optimum is $NP$-Hard for $K>1$. However,
-if epistasis is set such that each gene depends on the $K$ directly following it
-in the genome the solution can be found in polynomial time~\citep{wright:2000:solvingnk}.
-These Nearest Neighbor NK landscapes are therefore ideal for search algorithm testing.
-For all of our experiments we fixed $K=5$ to ensure highly rugged landscapes.
-
-\cite{saul:1994:spinglass} presents a combinatorial benchmark problem derived from physics:
-Ising Spin Glasses. A spin glass is defined by a weighted graph of interaction terms between vertices.
-Each gene assigns a value to each vertex, with the fitness calculated by Equation~\ref{eq-ising}.
-In this equation, $E$ is the set of all edges, $e_{ij}$ is the edge weight connecting vertex $i$ to vertex $j$, and
-$x_i$ and $x_j$ are the gene values for vertex $i$ and $j$. Optimal fitness is when this sum is minimized.
-\begin{equation}
-\sum_{e_{ij} \in E} x_ie_{ij}x_j
-  \label{eq-ising}
-\end{equation}
-Similar to NK Landscapes, the general class is $NP$-Hard to optimize, but the $2D\pm J$ subset of
-Ising Spin Glasses can be polynomially solved.\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}
-In this subset the graph is restricted to be a two dimensional torus, edge weights are randomly set
-to either -1 or 1, and vertex values must be -1 or 1.
-
-As our final class of randomly generated problems we chose the Maximum Satisfiability (MAX-SAT) problem.
-Related to the more common 3-SAT problem, a MAX-SAT instance is defined by a set of three term clauses.
-Each term is a randomly chosen variable, which may also be negated. A clause scores if an and only if
-at least one term in the clause evaluates to true. In order to make MAX-SAT instances with a known global
-optimum, \cite{goldman:2014:p3} proposed constructing clauses around a fixed solution. In this way the
-signs of the terms are set to ensure the target solution satisfies the clause. To ensure each
-problem is challenging we chose a clause to variable ratio of $4.27$.
-
-\section{Comparison Algorithm Parameter Tuning}
-\label{sec-tuning}
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-hboa}
-      \end{centering}
-      \caption{Tuned Population Sizes for hBOA}
-      \label{fig-pop-hboa}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-ltga}
-      \end{centering}
-      \caption{Tuned Population Sizes for LTGA}
-      \label{fig-pop-ltga}
-    \end{subfigure}
-  \end{centering}
-  \caption{Optimal population sizes found using bisection on each size
-           for each problem.}
-  \label{fig-pop-sizes}
-\end{figure}
-
-While four of the six algorithms in our experiments do not require any user specified parameters,
-hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
-handicapped, we extensively tuned each using the bisection method \citep{sastry:2001:bisection}
-to determine the optimal population size for each problem size.
-Extended by \cite{goldman:2012:ltga}, this method iteratively
-doubles the population size until some success criteria is met and then performs bisection
-between the lowest successful and highest unsuccessful sizes. In this way the minimum population size
-which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
-of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
-above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
-do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
-As bisection can make infinitely large population sizes, any run which had not found the global
-optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.
-
-Figure~\ref{fig-pop-sizes} shows the results from performing bisection. In general hBOA
-required population sizes which were at least an order of magnitude larger than LTGA. Due to
-runtime and memory overhead, finding the optimal value for hBOA was much less tractable than LTGA
-for moderate to large problem sizes. LTGA's population size also grew significantly slower
-than hBOA's as problem size increased, especially on the two Trap problems and Rastrigin.
-Both algorithms were very ineffective on MAX-SAT, neither able to tune to problems sizes over 60 bits.
-This is likely due to the fact that some randomly generated MAX-SAT landscapes are very flat
-and highly deceptive~\citep{rana:1998:gamaxsat}.
+%%%%%%%%%%%%%%%%% Global Optimum %%%%%%%%%%%%%%%%%%%%
 \section{Finding the Global Optimum}
@@
 
+%%%%%%%%%%%%%%%%%%% Fitness Over Time %%%%%%%%%%%%%%%%%%%%%
 \section{Fitness Over Time}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%% Computational Expenses %%%%%%%%%%%%%%%%%%%%
 \section{Computational Expenses}
@@
 
-When discussing the asymptotic complexity of P3 in Section~\ref{sec-p3}, two aspects eluded precise
+When discussing the asymptotic complexity of P3 in Chapter~\ref{chap-p3}, two aspects eluded precise
 analysis: how expensive is model rebuilding and how many gene donations are made. Figure~\ref{fig-costs}
@@
 than LTGA but by no more than a constant factor.
-
-\section{Population Sizing}
-A major advantage to P3 is that it does not require the user to set a population size parameter.
-Beyond making P3 easier to apply, this also conveys two additional advantages: diversity scaled
-to initialization and no need to sacrifice intermediate fitness for eventual optimality.
-
-\subsection{Problem Instance versus Problem Class}
-Our procedure for tuning LTGA and hBOA outlined in Section~\ref{sec-tuning} involved finding
-the optimal population size for each class of problem. For real world black box optimization this is
-realistically the best either algorithm could hope for as tuning to a problem
-instance or population initialization involves repeatedly solving the problem being tuned.
-This limitation does not exist in parameter-less methods, which scale their diversity
-based on the problem instance without needing to solve that instance repeatedly.
-
-To achieve high success rates on randomly generated problem classes, LTGA and hBOA must use a population size
-large enough to solve the hardest instances in that class. Therefore these methods will have
-population sizes larger than necessary to solve the easiest instances in the class. Even
-on single instance problems, both methods will require population sizes large enough to
-ensure the worst random initialization is diverse enough to solve the problem, which may be
-much larger than the best random initialization.
-
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{evals-to-success-range}
-  \end{center}
-  \caption{Comparison of the upper and lower quartiles of evaluations required
-           to reach the global optimum for P3 and LTGA with respect to problem size.}
-  \label{fig-evals-to-success-range}
-\end{figure}
-
-Figure~\ref{fig-evals-to-success-range}
-highlights how this can effect the required number of evaluations to reach the global optimum, showing
-the upper and lower quartiles for P3 and LTGA. On each problem
-LTGA has a much smaller difference between its best and worst runs. This makes sense as LTGA uses the same
-population size regardless of instance and progresses search generationally. In contrast P3 has a much
-higher split, with many runs finishing very quickly. This added variability is also generally on the lower
-quartile, with P3's upper quartile only worse than LTGA's on Deceptive Step Trap and Nearest
-Neighbor NK. This supports the hypothesis that P3 is able to scale its diversity not just
-to the problem class, but to the problem instance or even problem initialization, something
-wholly infeasible for tuned population sizing to do.
-
-This tuning distinction is also apparent when comparing Parameter-less hBOA with hBOA in Figure~\ref{fig-evals-to-success}.
-While generally performing worse than hBOA, the difference between the two algorithms is smallest
-on randomly generated problem classes. On MAX-SAT, Parameter-less hBOA actually outperformed both hBOA and LTGA,
-likely due to its ability to scale diversity to the problem instance instead of the entire problem class.
-
-\subsection{Fast versus Optimal}
-\label{sec-fast-vs-optimal}
-
-In Section~\ref{sec-overtime} we examined intermediate fitness qualities of LTGA and hBOA when using
-population sizes tuned to reach the global optimum. As a result, both were exceptionally ineffective
-at quickly reaching high quality solutions. This is because unlike P3, these methods have an explicit
-trade off between optimal performance and intermediate performance caused by their population size parameter.
-
-At 58\% of the recording intervals hBOA has the worst fitness of any solver. Most of the exceptions
-occur when hBOA is still evaluating its initial population, allowing this random search to temporarily surpass the
-local search methods. After $N$ evaluations, however, hBOA and LTGA both fall behind until their models begin
-to improve.
-Parameter-less hBOA reaches intermediate fitnesses faster than hBOA, doing so on 62\% of intervals, as its models begin
-to optimize earlier than hBOA.  However, this trend is reversed after a sufficient number of evaluations, most clearly
-on Deceptive Step Trap and Ising Spin Glasses, as hBOA's tuned population overtakes Parameter-less hBOA's parallel populations.
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{small-pop-dst}
-      \end{centering}
-      \caption{Deceptive Step Trap 203}
-      \label{fig-small-pop-dst}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{small-pop-nk}
-      \end{centering}
-      \caption{Nearest Neighbor NK 200}
-      \label{fig-small-pop-nk}
-    \end{subfigure}
-  \end{centering}
-  \caption{Comparison of how reducing LTGA's population size effects
-           the median best fitness reached during search.}
-  \label{fig-small-pop}
-\end{figure}
-
-Figure~\ref{fig-small-pop} examines the effect of population size on LTGA's intermediate fitness by examining the effect
-of reducing LTGA's population size to one tenth of the tuned value. The two problems shown are representative of the behavior of using
-a smaller population size on the other five problems. Reducing the population size caused LTGA to improve earlier but plateau
-at lower fitnesses. This caused LTGA's success rate to drop from 100 to 0 on Deceptive Step Trap and from 98 to 68 on Nearest
-Neighbor NK.
-Even when using a reduced population size for LTGA, P3 still achieved a fitness at
-least as high as LTGA at 80\% of intervals.
-
-\begin{comment}
-It is important to note that hBOA and LTGA were tuned specifically
-to find the global optimum and are not necessarily using the optimal population size for finding intermediate fitnesses. Consider
-that for both reducing the population size would almost certainly improve their fitness for evaluation steps less than their population
-size. To examine the effect of population size we tested LTGA using one tenth of the population size required for reliably finding the
-global optimum, with the results given in Figure~\ref{fig-small-pop}.
-Reducing the population size resulted in LTGA's initial fitness improvement
-occurring earlier, as the first generation is completed much more quickly. On Deceptive Step Trap this means that
-LTGA leaps ahead of P3 due to its ability to overcome the two bit fitness plateaus.
-%
-Across all tested problems using the smaller
-population size reduced the quality of LTGA's first fitness plateau, likely due to missing required diversity to reach higher quality.
-The second period of improvement also comes earlier and reaches a lower quality fitness. On Deceptive Step Trap 0
-runs reach the global optimum, down from 100 successful for the full population size. On Nearest Neighbor NK this drops the number
-of successful runs from 98 to 68. As a result we conclude that while reducing the population size of LTGA may improve its quality
-at early points during optimization, doing so reduces the robustness of the final solution found. This is in contrast to P3
-which balances high quality intermediate fitness without trading away eventual optimality.
-\end{comment}
-
-\section{Inner Workings}
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{cross}
-      \end{centering}
-      \caption{Crossover Proportion}
-      \label{fig-cross}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{cross-success}
-      \end{centering}
-      \caption{Crossover Success}
-      \label{fig-cross-success}
-    \end{subfigure}
-  \end{centering}
-  \caption{For each problem Figure~\ref{fig-cross} shows the proportion of P3 evaluations spend on crossovers
-           and Figure~\ref{fig-cross-success} shows the percentage of fitness improving crossover evaluations.}
-\end{figure}
-
-While analysis of optimization speed is useful from a practitioner standpoint, doing so provides very
-little insight into algorithm behavior. To better understand how P3 works in detail we present here
-a look at some internal features specific to P3.
-
-\subsection{Crossover}
-Figure~\ref{fig-cross} shows the proportion of evaluations P3 spends on crossover, as opposed to hill climbing,
-and Figure~\ref{fig-cross-success} shows what percentage of crossover evaluations resulted in a fitness improvement.
-Together these figures provide some insight into the role of crossover within P3. The behavior
-for each is clearly problem dependent and generally asymptoticly stable as problem size increases.
-
-When solving problems where epistasis can be effectively detected and represented by a linkage tree, P3 tends to spend
-less evaluations performing crossover and each crossover is more likely to be successful.
-Deceptive Trap and Rastrigin are the easiest problems to model epistasis, with local search quickly
-reducing pairwise entropy in each. These are also the problems where P3 uses the least evaluations
-on crossover and has the highest success rates for crossover. At the other extreme are Nearest Neighbor
-NK and Ising Spin Glasses, which both have overlapping linkage not representable by a linkage tree.
-These problems have the highest crossover usage and lowest crossover success of any problem except Deceptive
-Step Trap.  While Deceptive Step Trap's epistasis can be accurately modeled by a linkage tree, the
-exponential number of plateaus makes detecting gene linkage very challenging.
-
-\begin{comment}
-P3 performs optimization using a mixture of hill climbing and crossover. Exactly how much time the algorithm
-dedicates to each is not clear from the algorithm description alone. As such, Figure~\ref{fig-cross} shows
-that the relationship between the two is strongly problem dependent but relatively independent of problem size.
-In general, problems which can be more accurately represented by crossover
-required less crossover evaluations. Deceptive Trap and Rastrigin spent the least evaluations on crossover, using
-15\% and 20\% respectively on the largest problem sizes. Both of these problems are perfectly separable and have relatively
-few local optima in each subproblem. HIFF and Ising Spin Glasses use crossover for about 60\% of evaluations. The HIFF structure can
-be perfectly learned by the linkage tree, but the higher order relationships require multiple stages of crossover to optimize.
-Ising Spin Glasses have an overlapping structure, making them impossible to perfectly represent in a single linkage tree, but
-that overlap is relatively minor in comparison with Nearest Neighbor NK, which uses 81\% of evaluations on crossover. Deceptive Step Trap spends the highest
-percentage of evaluations on crossover, 92\%. Unlike all of the other problems, the value taken by an individual bit in Deceptive
-Step Trap has very little meaning. This is because any trap value is either a local optima or 1 bit flip away from being a local optima.
-This has a combined result of meaning it takes very few local search evaluations to find a local optima but very many crossover operations
-to proliferate the global optima. Epistatic relationships are also going to be very weak and even small population sizes are likely
-to mark all clusters as $useful$.
-
-Solutions are only added to the next highest pyramid level if at least one crossover donation resulted in a fitness
-improvement. Figure~\ref{fig-cross-success} examines the percentage of crossover donations that resulted in a fitness
-improvement. The ordering in Figure~\ref{fig-cross-success} is an almost perfect inversion of Figure~\ref{fig-cross},
-implying that the more successful crossover is at achieving fitness improving donations the lower the proportion of evaluations P3 will
-spend performing crossover. P3 achieves an impressive 30\% success rate for crossover donations on the largest Deceptive Trap
-problem sizes. This makes sense as crossover is able to immediately and perfectly identify entire traps such that all donations
-always contain only complete traps. Furthermore, by forcing crossover to attempt donations of genes different from the current solution,
-donations containing a single trap will have a very high success rate, always improving suboptimal traps which are far more frequently
-created by local search that optimal traps. Rastrigin and HIFF have relatively high success rates of 11\% and 7\%, respectively. Again
-this is likely due to accurate modeling and relatively low diversity of building block alternatives.
-Ising Spin Glasses and Deceptive Step Trap both have low success rates near 0.01\%. This follows as both are difficult to model with
-a linkage tree. In comparison to Figure~\ref{fig-cross-success} Deceptive Step Trap performs better than expected for crossover.
-However, the reason for this is that even exceedingly uninformed crossover can make improvements to Deceptive Step Trap as
-flipping two bits in the same trap will move between two local optima. Nearest Neighbor NK had the least successful crossovers,
-with only 0.007\% successful on the largest tested problems. In comparison to Ising Spin Glass, NK's overlap is far more extensive.
-As a result it makes sense that individual crossovers are very unlikely to be fitness improvements.
-\end{comment}
-
-When compared with LTGA, P3's crossover success rates are lower but similar in problem ordering.
-Counterintuitively, the use of hill climbing on the initial population reduces
-P3's crossover success, not because it reduces model quality or the donation pool, but because it is much more challenging to improve
-locally optimal solutions than randomly generated ones. LTGA's crossover benefits from application to unoptimized solutions, which
-makes its aggregate crossover success incomparable to P3's.
-
-Even when crossover success rates are quite low, such as Nearest Neighbor NK's 0.007\% success,
-the results from Section~\ref{sec-optimum} and Section~\ref{sec-overtime} show it's still critical to optimization. Without
-crossover, P3's performance would be identical to the Random Restart Hill Climber, which was unable to solve even moderately sized problems
-and quickly fell behind P3 in intermediate fitness quality. Therefore even infrequently successful crossover donations are
-critical to success. This does, however, suggest a potential avenue for future improvement by using more successful
-modeling and donation algorithms.
-
-\subsection{Pyramid}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{level-size}
-      \end{centering}
-      \caption{Population Size}
-      \label{fig-level-size}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{level-success}
-      \end{centering}
-      \caption{Crossover Success}
-      \label{fig-level-success}
-    \end{subfigure}
-  \end{centering}
-  \caption{For each problem Figure~\ref{fig-level-size} shows the number of solutions stored in each level of the pyramid
-           and Figure~\ref{fig-level-success} shows the percentage of fitness improving crossover evaluations at each level.}
-\end{figure}
-
-Another feature unique to P3 is the shape and size of the population pyramid
-constructed for each problem. Figure~\ref{fig-level-size} shows the number of solutions stored at
-each level of the pyramid for the largest tested problem sizes. Each point is the
-median size across 100 runs. If a run did not store any solutions at a level it is treated as 0.
-No point is drawn if the median run had 0 solutions stored at that level. While pyramid size
-is effected by problem size, the overall shape is not. As such the behavior shown in Figure~\ref{fig-level-size}
-is representative of that for all other tested problem sizes.
-
-With the exception of the dip in Deceptive Step Trap, all of the pyramids show a monotonic reduction
-in size as the level increases. This is because a solution must be a strict fitness improvement
-over its previous version to be added to a higher level, which becomes less likely
-each time the solution improves. \cite{lobo:2011:dynamicpop} found theoretical evidence
-and \cite{goldman:2011:dynamic-parameters} found empirical evidence that the optimal
-population size decreases each generation of traditional evolutionary search.
-By decreasing in size, P3 implicitly stores more diversity in low levels and focuses
-search on high quality solutions at higher levels. In comparison,
-LTGA and hBOA suboptimally use a fixed population size at each generation.
-
-Figure~\ref{fig-level-success} examines how crossover success changes at different
-levels of the pyramid. At low levels, success gets progressively lower
-as solution quality increases faster than the model's ability to improve solutions.
-At higher levels modeling becomes more accurate and donations contain higher frequencies
-of high quality building blocks, resulting in increased crossover success. The
-highest level of most problems has a low crossover success rate, as solutions
-crossing with that level have already been improved by previous operations to
-the point where the only improvement would be to create the global optimum, which can
-only happen once.
-
-\subsubsection{Deceptive Step Trap}
-The number of solutions stored at the fourth level of Deceptive Step Trap is
-significantly lower than that of the third or fifth levels, breaking the decreasing
-trend of the other six problems. Figure~\ref{fig-level-success} has a similar
-aberration, with crossover success dropping to  0.0004\% before rebounding and
-following the more common trajectory. This behavior exists in all other problem
-sizes tested, with the dips occurring at exactly the same level.
-
-This behavior is rooted in the peculiar nature
-of this landscape.
-After local search, all traps in all solutions have a total
-number of 1 bits equal to \{0, 1, 3, 5, 7\} as these correspond to the local
-optima when using $k=7$ and $s=2$. Crossover is very likely to overcome the
-two bit plateaus, and as a result solutions in the second level generally do not
-contain the lowest fitness local optima (5 bits set) and the third
-level has very few traps set to the next worst local optima (3 bits set). As a result,
-solutions which reach the third level can only be improved by replacing the deceptive
-local optima (0 and 1 bits set) with the global optimum (7 bits set).
-The global optimum is very rare in the population, and with 8 ways to represent local optima
-linkage learning is inaccurate. Therefore it is very unlikely for crossover to be successful,
-meaning few solutions will be added to the fourth level. Solutions that do improve by
-definition must have a higher frequency of optimal trap settings, meaning level four's model
-will be more accurate and donations are more likely to contain optimal trap values.
-Thus the level size and crossover success rates increase after contracting around level four.
-
-\section{Conclusions and Future Work}
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing black box
-optimization. P3's primary innovation is the replacement of the generational model with a pyramid of populations.
-This pyramid is constructed iteratively, with both the number of levels and the number of solutions stored
-at each level growing as search progresses. P3 uses a model based crossover method
-which learns a linkage tree from gene epistasis. Combined with a simple hill climber, P3's design contains
-many synergistic features.
-
-Across a large number of problems and problem sizes P3 required less evaluations to reach the global optimum
-than optimally tuned state-of-the-art competitors. On single instance problems P3's improvement was by a
-constant factor, while for the three problem classes P3's improvement increased with problem size. This
-quality extends to intermediate points during evolution, with P3 generally reaching at least as high
-of fitness as the competitive techniques when using the same number of evaluations. While P3 does require
-modeling overhead, the expense of this overhead is approximately linear with respect to genome size. There
-is some evidence that even when compared on wall clock time, P3 performs at least as well as
-the best comparison techniques. All of these achievements are made without any problem specific parameter
-tuning, making P3 easier to apply to new domains than its two closets competitors in quality.
-
-P3's quality is due to a number of desirable traits. First, mixing local search with model based
-crossover lets search focus on properly mixing high quality solutions. Second, by adding diversity only
-as necessary P3 tends to use the minimal amount of random initialization, unlike other techniques which must
-overcompensate with larger population sizes on single instance problems and consider the worst instance
-when solving problem classes. Third, by heavily exploiting existing diversity before adding more P3 is able
-to reach high quality intermediate fitnesses quickly without prematurely converging. Fourth, the very
-nature of the pyramid's shape allows search to preserve a desirable proportion of diversity at
-each fitness level, similar to a generational model using a decreasing population size.
-
-There are a number of meaningful avenues for future P3 experimentation. Perhaps the most
-pressing for practitioner acceptance is to apply P3 to real world problems and compare
-its results with other black box or even problem specific heuristics. While parameter-less,
-P3 is currently limited to discrete, fixed length genomes evaluated using single objective
-fitness. These limitations can be relaxed with future work to make P3 more widely applicable.
-While asymptotically linear, P3's modeling techniques and local search methods are likely going
-to be prohibitively expensive for genome sizes in the hundreds of thousands or millions of genes,
-and the inability of the model to capture overlapping linkage may be hindering
-search efficiency. Overcoming these limitations by using a new modeling technique may allow
-the pyramid model even greater flexibility. Similarly, while P3 is able to overcome low
-order deception via linkage learning, the iterative improvement method by which crossovers
-are made may mislead search on landscapes with higher order deception.
-
-However, even without
-these improvements our results show P3 is highly efficient at finding
-global optima on black box problems without any problem specific tuning.
-
-\small
-
-\bibliographystyle{apalike}
-\bibliography{../main}
-
-\normalsize
-\appendix
-\section{hBOA Simplification}
-\label{sec-appendix}
-To measure the quality of a decision forest, hBOA applies Equation~\ref{eq-hboa}.
-To favor compact
-models, Equation~\ref{eq-hboa} is scaled by Equation~\ref{eq-hboa-stop}, which provides increased cost
-for more total leaves.
-This quality is used for two purposes: comparison of potential changes from the existing model
-and comparison of that change with the existing model. The basis for our simplication
-is to rearrange $p(B)BDe(B) < p(B')BDe(B')$ to be $\frac{p(B)}{p(B')}<\frac{BDe(B')}{BDe(B)}$.
-
-\begin{equation}
-  BDe(B) = \prod_{i=0}^{N}\prod_{l\in L_i} \frac{\Gamma(m'_i(l))}{\Gamma(m_i(l) + m'_i(l))}
-  \prod_{x_i}\frac{\Gamma(m_i(x_i, l) + m'_i(x_i,l))}{\Gamma(m'_i(x_i,l))}
-  \label{eq-hboa}
-\end{equation}
-
-\begin{equation}
-  p(B) = c2^{-0.5(\sum_i|L_i|)log_2\mu}
-  \label{eq-hboa-stop}
-\end{equation}
-
-The outermost product of Equation~\ref{eq-hboa} iterates over all trees in the forest. However, each split
-can modify only one of the trees and therefore the contribution of all others can be canceled. The middle
-product is across all leaves in the tree. Again since only one leaf can be changed, all other terms can
-be canceled. By convention hBOA uses uninformed Bayesian priors of $m'_i(l)= 2$ and $m'_i(x_i, l)=1$ for
-binary alphabets. As $\Gamma(a) = (a-1)!$ this means the top term in the middle product and the bottom
-term in the third product reduce to 1. The only remaining terms are then $m_i(l)$ and $m_i(x_i, l)$ which
-represent the number of solutions which reached leaf $l$ and the number of solutions which reached leaf $l$
-with a specific value for $x_i$, respectively.
-
-Equation~\ref{eq-hboa-stop} can also be simplified when doing comparisons. If model $B'$ has exactly one more
-leaf than model $B$ then the ratio $\frac{p(B)}{p(B')}$ simplifies to $2^{0.5 log_2\mu}$ regardless of
-total model size.
-
-The resulting simplifications create Equation~\ref{eq-hboa-final},
-where $B'$ is different from $B$ by exactly 1 split, such that $l$ was split to create $l'$ and $l''$.
-The best split is whichever maximizes its improvement over $B$, which is equal to the right side
-of the inequality. Note that these factorials can still be exceedingly large and
-therefore it is imperative that implementations avoid rounding errors and overflows.
-
-\end{document}

### 2014-10-15 Comprehensive/chapter-comparison_algorithms.tex
Initial commit of Comprehensive Report

--- 2014-10-15 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
@@
-\documentclass[twoside]{article}
-\usepackage{ecj,palatino,epsfig,latexsym,natbib}
-\usepackage{url}
-\usepackage{verbatim}
-\usepackage{subcaption}
-\usepackage[noend]{algpseudocode}
-
-\newcommand{\includegraphicswide}[1]
-{\includegraphics[width=.9\textwidth,height=\textheight,keepaspectratio]{#1}}
-
-\newcommand{\includegraphicsfit}[1]
-{\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}
-
-\usepackage{array}
-\newcolumntype{L}{>{\centering\arraybackslash}m{1.9cm}}
-
-%% do not add any other page- or text-size instruction here
-
-\parskip=0.00in
-
-\begin{document}
-
-% 45-character limit on header
-\ecjHeader{x}{x}{xxx-xxx}{201X}{Parameter-less Population Pyramid}{B. W. Goldman and W. F. Punch}
-\title{\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid}
-
-\author{\name{\bf B. W. Goldman} \hfill \addr{brianwgoldman@acm.org}\\ 
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-\AND
-       \name{\bf W. F. Punch} \hfill \addr{punch@msu.edu}\\
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-}
-
-\maketitle
-
-\begin{abstract}
-%Should be about 200 words
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
-evolutionary optimization without requiring any user specified parameters.
-P3's primary innovation is to replace the generational model with a pyramid of
-multiple populations that are iteratively created and expanded. In combination
-with local search and advanced crossover P3 scales to problem difficulty, exploiting
-previously learned information before adding more diversity.
-
-Across seven problems, each tested using on average 18 problem sizes, P3 outperformed
-all five advanced comparison algorithms. This improvement includes requiring less evaluations
-and less wall clock seconds to find the global optimum and better fitness when using
-the same number of evaluations. Using both algorithm analysis and comparison we find P3's
-effectiveness is due to its ability to properly maintain, add, and exploit diversity.
-
-Unlike the best comparison algorithms, P3 was able to achieve this quality without any
-problem specific tuning. Thus, unlike previous parameter-less methods, P3 does not
-sacrifice quality for applicability. Therefore we conclude that
-P3 is an efficient, general, parameter-less approach to black box
-optimization which is more effective than existing state-of-the-art techniques.
-
-\end{abstract}
-
-\begin{keywords}
-
-Genetic algorithms, 
-linkage learning,
-local search,
-parameter-less.
-
-\end{keywords}
-
-\section{Introduction}
-A primary purpose of evolutionary optimization is to efficiently find good solutions
-to challenging real world problems with minimal prior knowledge about the problem itself.
-This driving goal has created search algorithms which can escape user bias to create
-truly novel results, sometimes publishable or patentable in their own right~\citep{kannappan:2014:humies}.
-While it is not possible for any algorithm to do better than random search across all possible
-problems~\citep{Wolpert:1997:nfl}, effectiveness can be achieved by assuming the search
-landscape has structure and then biasing the algorithm toward exploiting that structure.
-
-In evolutionary optimization, and genetic algorithms~(GAs) in particular, search is often
-biased through parameters. This can be beneficial as it allows practitioners to inject their
-knowledge about the shape of the search landscape into the algorithm.
-However, the quality of solutions found, and the speed at which they are found, is strongly tied to setting these parameters
-correctly~\citep{goldberg:1991:gasize}. As such, either expert knowledge or exceedingly
-expensive parameter tuning~\citep{grefenstette:1986:optimalga} are required to leverage
-this feature to its fullest potential. Furthermore,  parameters such as population size, mutation rate, crossover
-rate, tournament size, etc. usually have no clear relationship to the problem being solved, meaning even
-domain experts may not understand how the parameters will interact with the problem or with each other.
-
-There have been periodic efforts to reduce or remove the need for parameter tuning.
-\cite{Back:1992:selfadapt} introduced self-adaptive parameters, in which parameter values
-were included in each solution's genome and themselves underwent evolution. This allowed search
-to optimize some of its own parameters resulting in a reduced the need for expert tuning.
-\cite{harik:1999:parameterlessga} were able to design an entirely parameter-less GA by
-leveraging schema theory and parallel populations. Unfortunately these methods were provably less efficient
-than directly setting the parameters to optimal values~\citep{pelikan:1999:worstparameter-less}.
-
-One area that has been very effective at reducing the number of algorithm parameters is
-model based search. \cite{pelikan:2006:hboa}'s Hierarchical Bayesian Optimization
-Algorithm~(hBOA) and \cite{thierens:2010:ltga}'s Linkage Tree Genetic Algorithm~(LTGA)
-both require only a single parameter: population size. \cite{posik:2011:parameterless}
-leveraged model building to create a fully parameter-less algorithm, but it is restricted to
-only order-k, fully decomposable, noiseless problems.
-
-Most recently \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid~(P3).
-This method uses a pyramid structure of populations to combine model based search with local search
-to achieve parameter-less optimization. Initial results suggest that unlike
-previous parameter-less methods, P3 is actually more efficient than current state-of-the-art
-parameterized search algorithms. In this work we shall: extend these results to cover more
-comparison algorithms; compare both efficiency in reaching the global optimum and intermediate
-fitnesses; algorithm complexity analysis; and provide more in depth analysis of P3 itself.
-
-\section{Comparison Optimizers}
-\label{sec-optimizers}
+\chapter{Comparison Optimizers}
+\label{chap-optimizers}
 
@@
 are available from our website.\footnote{\url{https://github.com/brianwgoldman/FastEfficientP3}}
-\subsection{Random Restart Hill Climber}
+
+%%%%%%%%%% HC %%%%%%%%%%%%%%%%
+\section{Random Restart Hill Climber}
 \label{sec-hill-climber}
@@
 
-\subsection{$1+(\lambda, \lambda)$}
+%%%%%%%%%%%% Lambda Lambda %%%%%%%%%%%%%%%%
+\section{$1+(\lambda, \lambda)$}
 \cite{doerr:2013:lambdalambda} presented the first genetic algorithm to provably show
@@
 
-\subsection{Hierarchical Bayesian Optimization Algorithm}
+%%%%%%%%%%%%%%%%%% HBOA %%%%%%%%%%%%%%%%%%%%
+\section{Hierarchical Bayesian Optimization Algorithm}
 
@@
 almost always creates near infinitesimal results, calculating fractions which include factorials of $\mu$
-and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{sec-appendix}, we
+and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{appendix-hboa}, we
 derived a simplified form shown in Equation~\ref{eq-hboa-final}. If no proposed split satisfies the inequality, iteration stops.
@@
 
-\subsection{Parameter-less hBOA}
+%%%%%%%%%%%%%%% Parameter-less HBOA %%%%%%%%%%%%%%%%%%%%%%
+\section{Parameter-less hBOA}
 Using the methods first introduced by \cite{harik:1999:parameterlessga} for the Parameter-less GA,
@@
 
-\subsection{Linkage Tree Genetic Algorithm}
-\cite{thierens:2010:ltga} introduced the Linkage Tree Genetic Algorithm (LTGA) which automatically
+%%%%%%%%%%%%%%%%%%%% LTGA %%%%%%%%%%%%%%%%%%%%%%%%
+\section{Linkage Tree Genetic Algorithm}
+~\cite{thierens:2010:ltga} introduced the Linkage Tree Genetic Algorithm (LTGA) which automatically
 detects and exploits problem epistasis by examining pairwise gene entropy. Due to its enhanced
@@
 
-\section{Parameter-less Population Pyramid}
-\label{sec-p3}
-\cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid (P3) as a method for
-performing optimization which does not require the user to provide any parameters. This is
-achieved by combining efficient local search with the model building methods of LTGA using
-an iteratively constructed hierarchy of populations.
-
-The high level algorithm of P3 is presented in Figure~\ref{fig-p3}. Unlike more traditional
-GAs, P3 does not follow a generational model. Instead, it maintains an iteratively
-expanding pyramid of expanding populations. Each iteration, a new random solution is generated.
-This solution is brought to a local optimum using the hill climbing algorithm in Figure~\ref{fig-hc}. If that
-local optimum has not yet been added to the pyramid, the solution is added to the lowest
-population $P_0$.
-
-Next, the solution is iteratively improved by applying LTGA's crossover algorithm (Figure~\ref{fig-cluster-usage})
-with each population $P_i$ in the pyramid. If this process results in a strict fitness improvement and has
-created a solution not yet stored in the pyramid, it is added to the next highest pyramid level $P_{i+1}$.
-If $P_{i+1}$ does not yet exist, it is created. In this way populations in the pyramid expand over time,
-and the number of populations stored increases over time. Initially the pyramid contains no solutions
-or populations, meaning the user does not need to specify a population size.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Iterate-P3}{}
-    \State Create random solution
-    \State Apply hill climber
-    \If{solution $\notin hashset$}
-      \State Add solution to $P_0$
-      \State Add solution to $hashset$
-    \EndIf
-
-    \ForAll{$P_i \in pyramid$}
-      \State Mix solution with $P_i$
-      \If{solution's fitness has improved}
-        \If{solution $\notin hashset$}
-          \State Add solution to $P_{i+1}$
-          \State Add solution to $hashset$
-        \EndIf
-      \EndIf
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-  \caption{One iteration of P3 optimization. $pyramid$ is an
-           ordered set of populations and $hashset$ is a set
-           of all solutions in $pyramid$.}
-  \label{fig-p3}
-\end{figure}
-
-To accommodate P3's unique population structure, some of LTGA's clustering procedures were modified. In LTGA,
-clusters are identified at the start of each generation and are used
-to create all offspring in that generation. As P3 does not perform serial
-generations, P3 instead updates the model each
-time a solution is added to a population. Furthermore, unlike
-LTGA all solutions in the population are used to generate the model, not just the
-winners of a binary tournament. We can do this because even the
-worst solutions in the pyramid are already high quality due to local search.
-Using local search in LTGA was examined by \cite{bosman:2011:lsbbo} and found
-to provide no significant improvement. A likely cause was that that study applied
-local search to every solution, not just the initial population, resulting in significant overhead.
-
-Beyond the changes in population structuring, P3 modifies
-LTGA's version of \Call{Cluster-Creation}{} and \Call{Cluster-Usage}{}.
-P3 changes Line~\ref{fig-cluster-creation-ordering} in Figure~\ref{fig-cluster-creation}
-from \emph{last merged first} ordering to \emph{smallest first} ordering.
-This method applies gene clusters during crossover based on how many genes
-are in each cluster\footnote{Ties are broken randomly.}, and not on how tightly linked those genes are.
-\cite{goldman:2012:ltga} found that this alternative was better at preserving
-diversity, and therefore required smaller populations, but has not yet been integrated into
-canonical LTGA.
-
-P3 also modified Line~\ref{fig-cluster-usage-donate} in Figure~\ref{fig-cluster-usage}.
-Instead of choosing a single genetic donor for each cluster, P3 iterates over the
-population in a random order until a solution in the population is found which
-has a least one gene different for that cluster of genes from the improving solution.
-This process increases the likelihood of an evaluation being performed for every cluster,
-and helps test rare gene patterns in the population.
-
-In LTGA the cost of rebuilding the model is $O(\mu N^2)$ as it must collect pairwise
-gene frequency information for all $\mu$ solutions in the population. P3 does not store
-a single population, and it does not have a fixed $\mu$ size for any population. However,
-each time a solution is added to the population, it requires $O(N^2)$ time to update
-the table of pairwise frequencies and another $O(N^2)$ time to rebuild the linkage model.
-The model is then used immediately to perform up to one evaluation for each of the up
-to $2N-2$ clusters. Just as in LTGA, if no evaluation shortcuts were made, P3
-has an amortized cost of $O(N)$ modeling cost per evaluation. While P3 does
-rebuild the model more frequently per solution in the population, it also performs
-a number of local search evaluations which are quite efficient, meaning theoretical
-comparisons of their speed are difficult to perform. As a final note, P3's repeated
-attempts to find a useful donation make it less likely than LTGA to skip evaluations,
-but has an added cost to find these donations. Repeated donations could require as much
-as $O(\mu)$ attempts per evaluation, but experimental evidence suggest that this
-operation actually saves more overhead than it costs by increasing the number of evaluations
-per model rebuild.
-
-Each of the pieces of the P3 algorithm were selected not just for their stand alone efficacy,
-but for the ways in which they interact. By using the hill climber to ensure each solution
-is at a local optima, the underlying pairwise relationships in the problem are exposed. As
-a result, detecting clusters for use by crossover is much more effective. The crossover operator
-is extremely elitist, as each gene donation must result in no fitness loss, and a solution must
-strictly improve to be added to the next level of the pyramid. This is balanced by continual
-integration of new random solutions. Furthermore, each random restart decreases the probability
-of spurious linkages caused by shared ancestry. This diversity is further preserved by applying
-gene clusters in smallest first order during crossover as this reduces the probability of genetic
-hitchhikers.
-
-Other algorithms have proposed using multiple concurrent populations.
-\cite{hornby:2006:alps} had a hierarchy of populations with solutions periodically advancing
-upward. This allows for continuous integration of diversity as the lowest population is reseeded
-with random solutions. However, this method resulted in increased parameterization as not only was
-a population size required, but also parameters for how frequently generations advanced between levels and
-how many total levels to have were added. \cite{harik:1999:parameterlessga} used multiple independent
-populations of different sizes as a method for removing the population size parameter, but
-doing was provably less efficient than using an optimal population size as no information is shared
-between the populations.
-
-\section{Problem Descriptions}
-\label{sec-problems}
-
-\subsection{Single Instance Problems}
-Understanding how a stochastic search algorithm will behave on arbitrary and complex
-search landscapes can be exceedingly difficult. Therefore a common practice for
-algorithm understanding is to perform search on well defined, well understood
-landscapes. To be of interest these landscapes need to represent interesting
-and important aspects of real world problems.
-
-Perhaps the most common such landscape is the Deceptive Trap problem \citep{goldberg:1991:gasize}.
-In this landscape the genome is broken up into $k$ bit non-overlapping subproblems referred
-to as traps.
-Each subproblem is scored using Equation~\ref{eq-deceptive-trap}, where $t$ is the number of
-bits in the trap set to 1. The global optimum in each trap is a string of all 1s, while all
-other solutions lead to a local optima of all 0s. This problem tests an algorithm's
-ability to overcome $k$ sized deception and is commonly used to determine how effective crossover
-is at preserving building blocks. Any crossover event that mixes bits from different parents in
-the same trap will likely result in that trap being optimized to the local optima. For our experiments
-we chose $k=7$ to ensure highly deceptive traps.
-
-\begin{equation}
-   trap(t) = \left\{
-     \begin{array}{rl}
-       k-1-t, &  t<k\\
-       k,   &  t = k
-     \end{array}
-   \right.
-  \label{eq-deceptive-trap}
-\end{equation}
-
-\cite{goldman:2012:ltga} found that mixing local search with linkage learning rendered the Deceptive
-Trap problem trivial. This is because local search is able to optimize each trap to one of the two
-optima (all 1s or all 0s), and then linkage learning can have perfect knowledge of gene interactions.
-In order to make the problem more difficult, the authors proposed the Deceptive
-Step Trap problem, given in Equation~\ref{eq-deceptive-step-trap}. This function modifies the results
-of Equation~\ref{eq-deceptive-trap} to include plateaus of size $s$, introducing an exponential number
-of local optima in each trap. As a result the Deceptive Step Trap is much more challenging for linkage
-learning techniques, while still being highly deceptive. We chose to use $s=2$ to create the maximum
-number of local optima.
-\begin{equation}
-   step\_trap(t) = \left \lfloor \frac{(k-s)\pmod{s} + trap(t)}{s} \right \rfloor
-  \label{eq-deceptive-step-trap}
-\end{equation}
-
-Another challenging aspect of landscapes can be higher order relationships. The Hierarchical If
-and only If (HIFF) problem is designed to capture the difficulties of this class of problem.
-In HIFF the genome is broken up into a complete binary tree, such that each gene appears in exactly one
-leaf and each internal node is the subset of genes contained in its children. If all genes represented
-in a node of the tree are set to the same value, they score equal to the size of the set. In this way
-small subsets lead toward solutions to larger subsets. However, a node can score if all genes are either
-all 1s or all 0s, meaning that to solve higher order subproblems it is necessary to perform crossovers
-that preserve lower order solutions. This problem is a natural fit for LTGA as the linkage tree can
-perfectly duplicate the problem's true relationships \citep{thierens:2013:ltgahiff}.
-
-As a final class of well known problems, we have chosen to borrow the Rastrigin problem from real valued
-optimization. This problem's landscape, determined by Equation~\ref{eq-rast}, is highly multimodal caused by
-the oscillating cosine function. \cite{goldman:2014:p3} proposed the Discretized Rastrigin problem, such
-that each floating point $x_i$ in Equation~\ref{eq-rast} is encoded using a 10 bit gray code.
-\begin{equation}
-  An + \sum_{i=1}^{n}\left [ x_i^2-A\cos (2\pi x_i) \right ] \forall x\in [-5.12,5.12]
-  \label{eq-rast}
-\end{equation}
-
-\subsection{Randomly Generated Problem Classes}
-While well defined landscapes can provide specific insights into how an algorithm works,
-their static nature can be misleading. Specifically, algorithm quality might
-be very fragile such that it is only effective at searching well behaved landscapes. A more realistic
-test of an algorithm's black box effectiveness is to work with randomly generated instances
-drawn from a problem class. When tested over a sufficiently large sample it is then possible
-to draw more general conclusions about the algorithms effectiveness. The challenge with these
-landscapes is determining the global optimum to gauge if an algorithm was successful.
-
-Perhaps the most common model for generating random rugged landscapes is the NK model. An
-NK Landscape determines the fitness of each gene based on epistatic relationships with $K$
-other genes in the genome. This fitness is specified using a randomly generated table of
-fitness values, were each possible combination of the $K+1$ genes is mapped to some floating
-point value $[0-1]$. In unrestricted NK landscapes the relationships between genes are also
-randomly chosen and as a result finding the global optimum is $NP$-Hard for $K>1$. However,
-if epistasis is set such that each gene depends on the $K$ directly following it
-in the genome the solution can be found in polynomial time~\citep{wright:2000:solvingnk}.
-These Nearest Neighbor NK landscapes are therefore ideal for search algorithm testing.
-For all of our experiments we fixed $K=5$ to ensure highly rugged landscapes.
-
-\cite{saul:1994:spinglass} presents a combinatorial benchmark problem derived from physics:
-Ising Spin Glasses. A spin glass is defined by a weighted graph of interaction terms between vertices.
-Each gene assigns a value to each vertex, with the fitness calculated by Equation~\ref{eq-ising}.
-In this equation, $E$ is the set of all edges, $e_{ij}$ is the edge weight connecting vertex $i$ to vertex $j$, and
-$x_i$ and $x_j$ are the gene values for vertex $i$ and $j$. Optimal fitness is when this sum is minimized.
-\begin{equation}
-\sum_{e_{ij} \in E} x_ie_{ij}x_j
-  \label{eq-ising}
-\end{equation}
-Similar to NK Landscapes, the general class is $NP$-Hard to optimize, but the $2D\pm J$ subset of
-Ising Spin Glasses can be polynomially solved.\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}
-In this subset the graph is restricted to be a two dimensional torus, edge weights are randomly set
-to either -1 or 1, and vertex values must be -1 or 1.
-
-As our final class of randomly generated problems we chose the Maximum Satisfiability (MAX-SAT) problem.
-Related to the more common 3-SAT problem, a MAX-SAT instance is defined by a set of three term clauses.
-Each term is a randomly chosen variable, which may also be negated. A clause scores if an and only if
-at least one term in the clause evaluates to true. In order to make MAX-SAT instances with a known global
-optimum, \cite{goldman:2014:p3} proposed constructing clauses around a fixed solution. In this way the
-signs of the terms are set to ensure the target solution satisfies the clause. To ensure each
-problem is challenging we chose a clause to variable ratio of $4.27$.
-
+%%%%%%%%%%%%%%%% Tuning %%%%%%%%%%%%%%%%%%%%%
 \section{Comparison Algorithm Parameter Tuning}
@@
 and highly deceptive~\citep{rana:1998:gamaxsat}.
-\section{Finding the Global Optimum}
-\label{sec-optimum}
-
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{evals-to-success}
-  \end{center}
-  \caption{Comparison of the median number of evaluations to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-evals-to-success}
-\end{figure}
-
-Figure~\ref{fig-evals-to-success} shows the median number of evaluations required by each of the six
-algorithms to find the global optimum for multiple sizes of each problem.  Each data point in
-Figure~\ref{fig-evals-to-success} represents the median of 100 runs, where unsuccessful runs
-are treated as requiring more evaluations than any successful run. If the median run was not successful
-no point is shown. The maximum problem size used for each problem was set to be the largest size we could
-feasibly determine LTGA's optimal population size. For many larger problems results are not shown
-for hBOA due to the extreme computational cost required to optimally set the population size.
-
-\subsection{Quantitative Comparison}
-Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
-The largest problem size for any problem where P3 was not the most efficient has 49 bits, with P3 achieving the best results on all 92 larger
-configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
-Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
-and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
-on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
-
-\subsection{Local Search}
-The Random Restart Hill Climber and $1+(\lambda, \lambda)$ are both relatively effective on small problem
-sizes. This is especially true for the three randomly generated problem problem classes. These problems
-may contain relatively few local optima or just be exceptionally difficult for the model based algorithms.
-On Deceptive Trap and Deceptive Step Trap using 4 or less traps, $1+(\lambda, \lambda)$ performs significantly
-better than any other algorithm. We believe this is
-because $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
-flipping entire traps.
-This ability also leads $1+(\lambda, \lambda)$
-to outperform the Random Restart Hill Climber on all problems except Nearest Neighbor NK.
-
-On larger problem sizes, the ability for local search to reach the global optimum quickly diminishes.
-Only on MAX-SAT are these optimizers competitive
-at larger tested problem sizes. However, we believe this is because the largest tested MAX-SAT
-was an order of magnitude smaller than the largest size tested for most other problems. As the problem
-size increases the number of local optima increases exponentially, which explains why Random Restart Hill
-Climbing was unable to scale.
-For larger problems it also becomes increasingly unlikely for $1+(\lambda, \lambda)$ to make the right combination
-of changes required to reach the global optimum. This
-behavior causes high variance in success rate, as evident by the occasional successes on large Deceptive Trap problems.
-
-\subsection{Model Building}
-Only techniques which explicitly built models of gene epistasis were able to solve the largest problem
-instances. On single instance problems LTGA was more effective than hBOA, with hBOA outperforming
-LTGA on Nearest Neighbor NK and Ising Spin Glasses. This may be caused by the differences in modeling method:
-unlike the single instance problems, gene epistasis in the randomly generated
-problem classes cannot be be perfectly represented with a linkage tree.
-
-Considering how different hBOA and LTGA are in performing optimization, it is somewhat surprising how similar
-their results are on HIFF.
-However, both techniques rely on populations large enough to support the diversity required to reach the global optimum
-and to model epistasis. Both techniques also only rebuild models once per generation. As the subproblems of HIFF
-are nested, it is unlikely that either technique can accurately model higher order epistasis before solving
-lower order subproblems. Therefore both methods require one generation per subproblem order.
-
-\subsection{P3}
-Unlike the other model based methods, P3 generally outperforms
-both Random Restart Hill Climber and $1+(\lambda, \lambda)$ even on small problem sizes.
-Unlike the other local search methods, P3 outperforms LTGA and hBOA even on large problem sizes.
-This implies that P3 is gaining the benefits of each, leveraging local search to solve easy
-problems and model building to solve harder ones.
-
-Furthermore, the interaction between these two optimization tools explains some of the reason P3
-outperforms each method alone.
-On Deceptive Trap P3's use of hill climbing ensures all traps are immediately optimized, allowing for perfect
-linkage detection and high quality donation.
-On HIFF local search solves all pairwise subproblems, saving P3 a generation over LTGA and hBOA.
-In comparison P3 is only a slight improvement on Deceptive Step Trap, which is less amenable to local search.
-
-\section{Fitness Over Time}
-\label{sec-overtime}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{fitness-over-time}
-  \end{center}
-  \caption{Compares the median best fitness reached during search for each of the six optimization methods.}
-  \label{fig-fitness-over-time}
-\end{figure}
-
-For some applications, finding the global optimum is less important than finding good solutions quickly. Therefore
-we examine this behavior in Figure~\ref{fig-fitness-over-time}. At regular intervals during optimization
-Figure~\ref{fig-fitness-over-time} shows the median of the best fitnesses found at that point of search across
-100 runs. For each problem we show the largest problem size for which we were able to successfully gather results
-for all six algorithms, but the trends shown are representative of all larger problem sizes. The maximum reporting interval
-is set to include the slowest P3 run to reach the global optimum.
-
-% 181 total
-% 121 has P3 best
-%   9 hc
-%  50 lambdalambda
-%  20 hboa
-%  18 phboa
-%  27 LTGA
-
-\subsection{Quantitative Comparison}
-Of 181 sample points, P3 had the highest median fitness in 121. In pairwise competition, $1+(\lambda, \lambda)$
-was the most likely to outperform P3, doing so on 50 sample points. LTGA, hBOA, and Parameter-less hBOA
-were the next best, outperforming P3 on 27, 20, and 18 sample points. Random Restart Hill Climbing almost
-never outperformed P3, doing so only 9 times.
-
-\subsection{Local Search}
-Perhaps the most striking result is the quality of $1+(\lambda, \lambda)$. Until quite far into search this
-method performs better than both LTGA and hBOA. Given sufficient evaluations $1+(\lambda, \lambda)$ also outperforms
-Random Restart Hill Climbing on all 7 problems. For brief periods in the middle of search it performs the
-best of all techniques on: Deceptive Trap; Deceptive Step Trap; HIFF; Ising Spin Glass; and MAX-SAT problems.
-$1+(\lambda, \lambda)$'s ability to efficiently incorporate gene modifications of larger
-than one bit allows it to overcome the deception and plateaus in Deceptive Trap and Deceptive Step Trap, solve medium sized subproblems
-in HIFF, flip the signs on multiple adjacent bits in Ising Spin Glass, and cross plateaus in MAX-SAT. However,
-this method is slow in reaching the global optima in many of these problems which causes it to eventually
-be overtaken by the model building techniques.
-
-\subsection{Model Building}
-Both hBOA and LTGA are marked by periods of little improvement followed by rapid improvement.
-In hBOA this is taken to the extreme, with all fitness
-improvement made at the very end of search. In both cases this is caused by model building. Before the model
-is accurate little improvement is made. Once it is accurate, fitness improves dramatically.
-
-On every problem LTGA has five distinct periods: fitness plateau, near instantaneous improvement, fitness plateau,
-and improvement to global optimum. The early period corresponds with
-initialization of the population, with the first fitness gain achieved immediately upon completing the first generation.
-When using an inaccurate model, LTGA's mixing strategy performs a sort of less effective local search.
-Subsequent generations then make only minor fitness improvements. Once the model
-becomes accurate and the probability of a crossover using high quality genetic material increases sufficiently, LTGA
-enters a second period of rapid improvement.
-
-\begin{comment}
-This behavior is easiest to understand on the two Trap problems. LTGA's first generation will push individual traps toward a
-local optima, but it requires more than a single generation to do so completely. Until that occurs, the model is likely behaving
-no better than random as there is little apparent linkage between bits. Once the model begins to identify individual traps, crossover
-can begin to increase the frequency of higher fitness trap settings. This process begins slowly as low fitness local optima are more
-likely to be chosen for donation than high fitness local optima due to their frequency in the population. The process becomes
-self-catalysing as increased frequency of the global optimum trap genes means increased likelihood of the global optimum being spread by crossover.
-\end{comment}
-
-\subsection{P3}
-The integration of hill climbing into P3 makes it strictly better than using hill climbing alone.
-Early in optimization P3 and the Random Restart Hill Climber have effectively identical quality. This is because
-P3 performs the same evaluations as the Hill Climber for the first two restarts. Once P3 begins
-performing crossover it immediately improves over the Hill Climber. In 95\% of intervals P3 had a fitness at
-least as high as Hill Climbing. As such P3 is better than a simple
-hill climber regardless of how long each technique is run and irrespective of how high quality the solution
-found has to be.
-
-Unlike the model based methods, which struggle until model accuracy improves, P3's iterative solution
-integration allows it to improve much more quickly. This behavior exists in most problems, but is easiest to understand on Deceptive Trap.
-On this problem, P3 immediately brings all traps to local optima, equaled only by the Random Restart Hill Climber in quality.
-In comparison LTGA must evaluate the entire population and perform multiple generations to reach similar quality.
-P3 is able to immediately integrate optimal versions of each
-trap into a single individual as they are found by local search, resulting in smoother fitness improvement than LTGA.
-
-\section{Computational Expenses}
-While it is common in evolutionary computation to assume the evaluation function will dominate algorithm
-complexity, in some domains this will not be true. Model based methods are especially likely to violate
-this norm. Therefore, in order to assess P3's quality in solving problems with efficient fitness functions,
-we provide data on both its algorithmic complexity and wall clock time.
-
-\subsection{Operation Counting}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{rebuilds}
-      \end{centering}
-      \caption{Rebuilds}
-      \label{fig-rebuilds}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{donations}
-      \end{centering}
-      \caption{Donations}
-      \label{fig-donations}
-    \end{subfigure}
-  \end{centering}
-  \caption{Computation costs incurred by model rebuilding (Figure~\ref{fig-rebuilds}) and
-           repeated donations (Figure~\ref{fig-donations}) per evaluation as problem size increases.}
-  \label{fig-costs}
-\end{figure}
-
-When discussing the asymptotic complexity of P3 in Section~\ref{sec-p3}, two aspects eluded precise
-analysis: how expensive is model rebuilding and how many gene donations are made. Figure~\ref{fig-costs}
-provides some insight into how often these two aspects of the algorithm are utilized.
-
-Figure~\ref{fig-rebuilds} reports in an algorithmic sense how expensive model rebuilding is for search.
-In order to calculate this value we recorded how many times search rebuilt the model during each run.
-Under the minimal assumption that an evaluation costs $\Omega(N)$ time and that model rebuilds cost $O(N^2)$,
-Figure~\ref{fig-rebuilds} shows the ratio cost of model rebuilding per evaluation. If the cost of model building
-scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically constant.
-For Nearest Neighbor NK, Ising Spin Glasses, and Rastrigin this is the case. For both Trap problems
-and HIFF there is slow growth in the ratio. The problem sizes used for MAX-SAT were not sufficient
-to accurately gauge the asymptotic behavior. Together this suggests that while the cost of building the model
-is almost linear per evaluation, it can grow slowly. However, even in the worst case (HIFF) this growth was
-no more than twice the algorithmic cost of an evaluation even using 2048 bits.
-
-When applying a crossover subset, P3 tries random donors from the population until one is found with at least
-one bit different from the improving solution. In theory this can result in up to $O(\mu)$ operations.
-Figure~\ref{fig-donations} examines the observed average number of donations per evaluation performed.
-Ising Spin Glass, HIFF, and Rastrigin all achieve effectively constant behavior here, implying repeated
-donation does not impact the asymptotic runtime of P3. Both Trap functions and Nearest Neighbor NK all
-increase in number of donations as problem size increases, potentially increasing algorithmic costs. An
-important note is that each donation may range in size from a single bit up to $N-1$, with smaller
-donations far more likely to result in repeated attempts. As such this may cause some super-linear growth
-in P3, but its unlikely to be very high.
-
-\subsection{Wall Clock Performance}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{seconds-to-success}
-  \end{center}
-  \caption{Comparison of the median number of seconds to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-seconds-to-success}
-\end{figure}
-
-To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}.
-Similar to Figure~\ref{fig-evals-to-success}, each point represents the median
-of 100 runs, with unsuccessful runs treated as slower than successful runs.
-In total these runs
-took over 6 computing years to complete. We were therefore forced to use
-diverse hardware potentially running multiple processes in parallel.
-As such these timings should be considered very skeptically as a number
-of sources may contribute to biased error in the results. However, due to
-the size of the dataset and the general regularity in the results some general conclusions
-can still be made.
-
-\subsubsection{Model Building}
-hBOA and Parameter-less hBOA perform significantly worse when using wall clock time as the unit of comparison than when using
-evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while, under reasonable
-assumptions, P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
-hBOA goes from being slightly more efficient than LTGA in terms of evaluations to significantly worse in terms of seconds.
-As P3 and LTGA require a similar asymptotic complexity per
-evaluation as the Hill Climber and $1+(\lambda, \lambda)$, no similar change in ordering occurs.
-
-\subsubsection{P3}
-Any increase in P3's algorithmic complexity is offset by its efficient use of evaluations.
-In aggregate P3 found the global optimum the fastest on 72 of 130 configurations,
-with LTGA, $1+(\lambda, \lambda)$, Hill Climber, hBOA outperforming P3 on 44, 18, 10, and 1 configurations, respectively.
-On
-Deceptive Trap, Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT P3 still finds the global
-optimum faster than all comparison techniques. LTGA is able to match P3's performance almost exactly on HIFF,
-and on Rastrigin P3 falls behind only on moderate sized problems. Deceptive Step Trap yields the worst performance,
-with LTGA using optimal population sizes having a significantly lower growth curve than P3. This mirrors
-the results from Figure~\ref{fig-evals-to-success} under the assumption that P3's computational time per evaluation is higher
-than LTGA but by no more than a constant factor.
-
-\section{Population Sizing}
-A major advantage to P3 is that it does not require the user to set a population size parameter.
-Beyond making P3 easier to apply, this also conveys two additional advantages: diversity scaled
-to initialization and no need to sacrifice intermediate fitness for eventual optimality.
-
-\subsection{Problem Instance versus Problem Class}
-Our procedure for tuning LTGA and hBOA outlined in Section~\ref{sec-tuning} involved finding
-the optimal population size for each class of problem. For real world black box optimization this is
-realistically the best either algorithm could hope for as tuning to a problem
-instance or population initialization involves repeatedly solving the problem being tuned.
-This limitation does not exist in parameter-less methods, which scale their diversity
-based on the problem instance without needing to solve that instance repeatedly.
-
-To achieve high success rates on randomly generated problem classes, LTGA and hBOA must use a population size
-large enough to solve the hardest instances in that class. Therefore these methods will have
-population sizes larger than necessary to solve the easiest instances in the class. Even
-on single instance problems, both methods will require population sizes large enough to
-ensure the worst random initialization is diverse enough to solve the problem, which may be
-much larger than the best random initialization.
-
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{evals-to-success-range}
-  \end{center}
-  \caption{Comparison of the upper and lower quartiles of evaluations required
-           to reach the global optimum for P3 and LTGA with respect to problem size.}
-  \label{fig-evals-to-success-range}
-\end{figure}
-
-Figure~\ref{fig-evals-to-success-range}
-highlights how this can effect the required number of evaluations to reach the global optimum, showing
-the upper and lower quartiles for P3 and LTGA. On each problem
-LTGA has a much smaller difference between its best and worst runs. This makes sense as LTGA uses the same
-population size regardless of instance and progresses search generationally. In contrast P3 has a much
-higher split, with many runs finishing very quickly. This added variability is also generally on the lower
-quartile, with P3's upper quartile only worse than LTGA's on Deceptive Step Trap and Nearest
-Neighbor NK. This supports the hypothesis that P3 is able to scale its diversity not just
-to the problem class, but to the problem instance or even problem initialization, something
-wholly infeasible for tuned population sizing to do.
-
-This tuning distinction is also apparent when comparing Parameter-less hBOA with hBOA in Figure~\ref{fig-evals-to-success}.
-While generally performing worse than hBOA, the difference between the two algorithms is smallest
-on randomly generated problem classes. On MAX-SAT, Parameter-less hBOA actually outperformed both hBOA and LTGA,
-likely due to its ability to scale diversity to the problem instance instead of the entire problem class.
-
-\subsection{Fast versus Optimal}
-\label{sec-fast-vs-optimal}
-
-In Section~\ref{sec-overtime} we examined intermediate fitness qualities of LTGA and hBOA when using
-population sizes tuned to reach the global optimum. As a result, both were exceptionally ineffective
-at quickly reaching high quality solutions. This is because unlike P3, these methods have an explicit
-trade off between optimal performance and intermediate performance caused by their population size parameter.
-
-At 58\% of the recording intervals hBOA has the worst fitness of any solver. Most of the exceptions
-occur when hBOA is still evaluating its initial population, allowing this random search to temporarily surpass the
-local search methods. After $N$ evaluations, however, hBOA and LTGA both fall behind until their models begin
-to improve.
-Parameter-less hBOA reaches intermediate fitnesses faster than hBOA, doing so on 62\% of intervals, as its models begin
-to optimize earlier than hBOA.  However, this trend is reversed after a sufficient number of evaluations, most clearly
-on Deceptive Step Trap and Ising Spin Glasses, as hBOA's tuned population overtakes Parameter-less hBOA's parallel populations.
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{small-pop-dst}
-      \end{centering}
-      \caption{Deceptive Step Trap 203}
-      \label{fig-small-pop-dst}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{small-pop-nk}
-      \end{centering}
-      \caption{Nearest Neighbor NK 200}
-      \label{fig-small-pop-nk}
-    \end{subfigure}
-  \end{centering}
-  \caption{Comparison of how reducing LTGA's population size effects
-           the median best fitness reached during search.}
-  \label{fig-small-pop}
-\end{figure}
-
-Figure~\ref{fig-small-pop} examines the effect of population size on LTGA's intermediate fitness by examining the effect
-of reducing LTGA's population size to one tenth of the tuned value. The two problems shown are representative of the behavior of using
-a smaller population size on the other five problems. Reducing the population size caused LTGA to improve earlier but plateau
-at lower fitnesses. This caused LTGA's success rate to drop from 100 to 0 on Deceptive Step Trap and from 98 to 68 on Nearest
-Neighbor NK.
-Even when using a reduced population size for LTGA, P3 still achieved a fitness at
-least as high as LTGA at 80\% of intervals.
-
-\begin{comment}
-It is important to note that hBOA and LTGA were tuned specifically
-to find the global optimum and are not necessarily using the optimal population size for finding intermediate fitnesses. Consider
-that for both reducing the population size would almost certainly improve their fitness for evaluation steps less than their population
-size. To examine the effect of population size we tested LTGA using one tenth of the population size required for reliably finding the
-global optimum, with the results given in Figure~\ref{fig-small-pop}.
-Reducing the population size resulted in LTGA's initial fitness improvement
-occurring earlier, as the first generation is completed much more quickly. On Deceptive Step Trap this means that
-LTGA leaps ahead of P3 due to its ability to overcome the two bit fitness plateaus.
-%
-Across all tested problems using the smaller
-population size reduced the quality of LTGA's first fitness plateau, likely due to missing required diversity to reach higher quality.
-The second period of improvement also comes earlier and reaches a lower quality fitness. On Deceptive Step Trap 0
-runs reach the global optimum, down from 100 successful for the full population size. On Nearest Neighbor NK this drops the number
-of successful runs from 98 to 68. As a result we conclude that while reducing the population size of LTGA may improve its quality
-at early points during optimization, doing so reduces the robustness of the final solution found. This is in contrast to P3
-which balances high quality intermediate fitness without trading away eventual optimality.
-\end{comment}
-
-\section{Inner Workings}
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{cross}
-      \end{centering}
-      \caption{Crossover Proportion}
-      \label{fig-cross}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{cross-success}
-      \end{centering}
-      \caption{Crossover Success}
-      \label{fig-cross-success}
-    \end{subfigure}
-  \end{centering}
-  \caption{For each problem Figure~\ref{fig-cross} shows the proportion of P3 evaluations spend on crossovers
-           and Figure~\ref{fig-cross-success} shows the percentage of fitness improving crossover evaluations.}
-\end{figure}
-
-While analysis of optimization speed is useful from a practitioner standpoint, doing so provides very
-little insight into algorithm behavior. To better understand how P3 works in detail we present here
-a look at some internal features specific to P3.
-
-\subsection{Crossover}
-Figure~\ref{fig-cross} shows the proportion of evaluations P3 spends on crossover, as opposed to hill climbing,
-and Figure~\ref{fig-cross-success} shows what percentage of crossover evaluations resulted in a fitness improvement.
-Together these figures provide some insight into the role of crossover within P3. The behavior
-for each is clearly problem dependent and generally asymptoticly stable as problem size increases.
-
-When solving problems where epistasis can be effectively detected and represented by a linkage tree, P3 tends to spend
-less evaluations performing crossover and each crossover is more likely to be successful.
-Deceptive Trap and Rastrigin are the easiest problems to model epistasis, with local search quickly
-reducing pairwise entropy in each. These are also the problems where P3 uses the least evaluations
-on crossover and has the highest success rates for crossover. At the other extreme are Nearest Neighbor
-NK and Ising Spin Glasses, which both have overlapping linkage not representable by a linkage tree.
-These problems have the highest crossover usage and lowest crossover success of any problem except Deceptive
-Step Trap.  While Deceptive Step Trap's epistasis can be accurately modeled by a linkage tree, the
-exponential number of plateaus makes detecting gene linkage very challenging.
-
-\begin{comment}
-P3 performs optimization using a mixture of hill climbing and crossover. Exactly how much time the algorithm
-dedicates to each is not clear from the algorithm description alone. As such, Figure~\ref{fig-cross} shows
-that the relationship between the two is strongly problem dependent but relatively independent of problem size.
-In general, problems which can be more accurately represented by crossover
-required less crossover evaluations. Deceptive Trap and Rastrigin spent the least evaluations on crossover, using
-15\% and 20\% respectively on the largest problem sizes. Both of these problems are perfectly separable and have relatively
-few local optima in each subproblem. HIFF and Ising Spin Glasses use crossover for about 60\% of evaluations. The HIFF structure can
-be perfectly learned by the linkage tree, but the higher order relationships require multiple stages of crossover to optimize.
-Ising Spin Glasses have an overlapping structure, making them impossible to perfectly represent in a single linkage tree, but
-that overlap is relatively minor in comparison with Nearest Neighbor NK, which uses 81\% of evaluations on crossover. Deceptive Step Trap spends the highest
-percentage of evaluations on crossover, 92\%. Unlike all of the other problems, the value taken by an individual bit in Deceptive
-Step Trap has very little meaning. This is because any trap value is either a local optima or 1 bit flip away from being a local optima.
-This has a combined result of meaning it takes very few local search evaluations to find a local optima but very many crossover operations
-to proliferate the global optima. Epistatic relationships are also going to be very weak and even small population sizes are likely
-to mark all clusters as $useful$.
-
-Solutions are only added to the next highest pyramid level if at least one crossover donation resulted in a fitness
-improvement. Figure~\ref{fig-cross-success} examines the percentage of crossover donations that resulted in a fitness
-improvement. The ordering in Figure~\ref{fig-cross-success} is an almost perfect inversion of Figure~\ref{fig-cross},
-implying that the more successful crossover is at achieving fitness improving donations the lower the proportion of evaluations P3 will
-spend performing crossover. P3 achieves an impressive 30\% success rate for crossover donations on the largest Deceptive Trap
-problem sizes. This makes sense as crossover is able to immediately and perfectly identify entire traps such that all donations
-always contain only complete traps. Furthermore, by forcing crossover to attempt donations of genes different from the current solution,
-donations containing a single trap will have a very high success rate, always improving suboptimal traps which are far more frequently
-created by local search that optimal traps. Rastrigin and HIFF have relatively high success rates of 11\% and 7\%, respectively. Again
-this is likely due to accurate modeling and relatively low diversity of building block alternatives.
-Ising Spin Glasses and Deceptive Step Trap both have low success rates near 0.01\%. This follows as both are difficult to model with
-a linkage tree. In comparison to Figure~\ref{fig-cross-success} Deceptive Step Trap performs better than expected for crossover.
-However, the reason for this is that even exceedingly uninformed crossover can make improvements to Deceptive Step Trap as
-flipping two bits in the same trap will move between two local optima. Nearest Neighbor NK had the least successful crossovers,
-with only 0.007\% successful on the largest tested problems. In comparison to Ising Spin Glass, NK's overlap is far more extensive.
-As a result it makes sense that individual crossovers are very unlikely to be fitness improvements.
-\end{comment}
-
-When compared with LTGA, P3's crossover success rates are lower but similar in problem ordering.
-Counterintuitively, the use of hill climbing on the initial population reduces
-P3's crossover success, not because it reduces model quality or the donation pool, but because it is much more challenging to improve
-locally optimal solutions than randomly generated ones. LTGA's crossover benefits from application to unoptimized solutions, which
-makes its aggregate crossover success incomparable to P3's.
-
-Even when crossover success rates are quite low, such as Nearest Neighbor NK's 0.007\% success,
-the results from Section~\ref{sec-optimum} and Section~\ref{sec-overtime} show it's still critical to optimization. Without
-crossover, P3's performance would be identical to the Random Restart Hill Climber, which was unable to solve even moderately sized problems
-and quickly fell behind P3 in intermediate fitness quality. Therefore even infrequently successful crossover donations are
-critical to success. This does, however, suggest a potential avenue for future improvement by using more successful
-modeling and donation algorithms.
-
-\subsection{Pyramid}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{level-size}
-      \end{centering}
-      \caption{Population Size}
-      \label{fig-level-size}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{level-success}
-      \end{centering}
-      \caption{Crossover Success}
-      \label{fig-level-success}
-    \end{subfigure}
-  \end{centering}
-  \caption{For each problem Figure~\ref{fig-level-size} shows the number of solutions stored in each level of the pyramid
-           and Figure~\ref{fig-level-success} shows the percentage of fitness improving crossover evaluations at each level.}
-\end{figure}
-
-Another feature unique to P3 is the shape and size of the population pyramid
-constructed for each problem. Figure~\ref{fig-level-size} shows the number of solutions stored at
-each level of the pyramid for the largest tested problem sizes. Each point is the
-median size across 100 runs. If a run did not store any solutions at a level it is treated as 0.
-No point is drawn if the median run had 0 solutions stored at that level. While pyramid size
-is effected by problem size, the overall shape is not. As such the behavior shown in Figure~\ref{fig-level-size}
-is representative of that for all other tested problem sizes.
-
-With the exception of the dip in Deceptive Step Trap, all of the pyramids show a monotonic reduction
-in size as the level increases. This is because a solution must be a strict fitness improvement
-over its previous version to be added to a higher level, which becomes less likely
-each time the solution improves. \cite{lobo:2011:dynamicpop} found theoretical evidence
-and \cite{goldman:2011:dynamic-parameters} found empirical evidence that the optimal
-population size decreases each generation of traditional evolutionary search.
-By decreasing in size, P3 implicitly stores more diversity in low levels and focuses
-search on high quality solutions at higher levels. In comparison,
-LTGA and hBOA suboptimally use a fixed population size at each generation.
-
-Figure~\ref{fig-level-success} examines how crossover success changes at different
-levels of the pyramid. At low levels, success gets progressively lower
-as solution quality increases faster than the model's ability to improve solutions.
-At higher levels modeling becomes more accurate and donations contain higher frequencies
-of high quality building blocks, resulting in increased crossover success. The
-highest level of most problems has a low crossover success rate, as solutions
-crossing with that level have already been improved by previous operations to
-the point where the only improvement would be to create the global optimum, which can
-only happen once.
-
-\subsubsection{Deceptive Step Trap}
-The number of solutions stored at the fourth level of Deceptive Step Trap is
-significantly lower than that of the third or fifth levels, breaking the decreasing
-trend of the other six problems. Figure~\ref{fig-level-success} has a similar
-aberration, with crossover success dropping to  0.0004\% before rebounding and
-following the more common trajectory. This behavior exists in all other problem
-sizes tested, with the dips occurring at exactly the same level.
-
-This behavior is rooted in the peculiar nature
-of this landscape.
-After local search, all traps in all solutions have a total
-number of 1 bits equal to \{0, 1, 3, 5, 7\} as these correspond to the local
-optima when using $k=7$ and $s=2$. Crossover is very likely to overcome the
-two bit plateaus, and as a result solutions in the second level generally do not
-contain the lowest fitness local optima (5 bits set) and the third
-level has very few traps set to the next worst local optima (3 bits set). As a result,
-solutions which reach the third level can only be improved by replacing the deceptive
-local optima (0 and 1 bits set) with the global optimum (7 bits set).
-The global optimum is very rare in the population, and with 8 ways to represent local optima
-linkage learning is inaccurate. Therefore it is very unlikely for crossover to be successful,
-meaning few solutions will be added to the fourth level. Solutions that do improve by
-definition must have a higher frequency of optimal trap settings, meaning level four's model
-will be more accurate and donations are more likely to contain optimal trap values.
-Thus the level size and crossover success rates increase after contracting around level four.
-
-\section{Conclusions and Future Work}
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing black box
-optimization. P3's primary innovation is the replacement of the generational model with a pyramid of populations.
-This pyramid is constructed iteratively, with both the number of levels and the number of solutions stored
-at each level growing as search progresses. P3 uses a model based crossover method
-which learns a linkage tree from gene epistasis. Combined with a simple hill climber, P3's design contains
-many synergistic features.
-
-Across a large number of problems and problem sizes P3 required less evaluations to reach the global optimum
-than optimally tuned state-of-the-art competitors. On single instance problems P3's improvement was by a
-constant factor, while for the three problem classes P3's improvement increased with problem size. This
-quality extends to intermediate points during evolution, with P3 generally reaching at least as high
-of fitness as the competitive techniques when using the same number of evaluations. While P3 does require
-modeling overhead, the expense of this overhead is approximately linear with respect to genome size. There
-is some evidence that even when compared on wall clock time, P3 performs at least as well as
-the best comparison techniques. All of these achievements are made without any problem specific parameter
-tuning, making P3 easier to apply to new domains than its two closets competitors in quality.
-
-P3's quality is due to a number of desirable traits. First, mixing local search with model based
-crossover lets search focus on properly mixing high quality solutions. Second, by adding diversity only
-as necessary P3 tends to use the minimal amount of random initialization, unlike other techniques which must
-overcompensate with larger population sizes on single instance problems and consider the worst instance
-when solving problem classes. Third, by heavily exploiting existing diversity before adding more P3 is able
-to reach high quality intermediate fitnesses quickly without prematurely converging. Fourth, the very
-nature of the pyramid's shape allows search to preserve a desirable proportion of diversity at
-each fitness level, similar to a generational model using a decreasing population size.
-
-There are a number of meaningful avenues for future P3 experimentation. Perhaps the most
-pressing for practitioner acceptance is to apply P3 to real world problems and compare
-its results with other black box or even problem specific heuristics. While parameter-less,
-P3 is currently limited to discrete, fixed length genomes evaluated using single objective
-fitness. These limitations can be relaxed with future work to make P3 more widely applicable.
-While asymptotically linear, P3's modeling techniques and local search methods are likely going
-to be prohibitively expensive for genome sizes in the hundreds of thousands or millions of genes,
-and the inability of the model to capture overlapping linkage may be hindering
-search efficiency. Overcoming these limitations by using a new modeling technique may allow
-the pyramid model even greater flexibility. Similarly, while P3 is able to overcome low
-order deception via linkage learning, the iterative improvement method by which crossovers
-are made may mislead search on landscapes with higher order deception.
-
-However, even without
-these improvements our results show P3 is highly efficient at finding
-global optima on black box problems without any problem specific tuning.
-
-\small
-
-\bibliographystyle{apalike}
-\bibliography{../main}
-
-\normalsize
-\appendix
-\section{hBOA Simplification}
-\label{sec-appendix}
-To measure the quality of a decision forest, hBOA applies Equation~\ref{eq-hboa}.
-To favor compact
-models, Equation~\ref{eq-hboa} is scaled by Equation~\ref{eq-hboa-stop}, which provides increased cost
-for more total leaves.
-This quality is used for two purposes: comparison of potential changes from the existing model
-and comparison of that change with the existing model. The basis for our simplication
-is to rearrange $p(B)BDe(B) < p(B')BDe(B')$ to be $\frac{p(B)}{p(B')}<\frac{BDe(B')}{BDe(B)}$.
-
-\begin{equation}
-  BDe(B) = \prod_{i=0}^{N}\prod_{l\in L_i} \frac{\Gamma(m'_i(l))}{\Gamma(m_i(l) + m'_i(l))}
-  \prod_{x_i}\frac{\Gamma(m_i(x_i, l) + m'_i(x_i,l))}{\Gamma(m'_i(x_i,l))}
-  \label{eq-hboa}
-\end{equation}
-
-\begin{equation}
-  p(B) = c2^{-0.5(\sum_i|L_i|)log_2\mu}
-  \label{eq-hboa-stop}
-\end{equation}
-
-The outermost product of Equation~\ref{eq-hboa} iterates over all trees in the forest. However, each split
-can modify only one of the trees and therefore the contribution of all others can be canceled. The middle
-product is across all leaves in the tree. Again since only one leaf can be changed, all other terms can
-be canceled. By convention hBOA uses uninformed Bayesian priors of $m'_i(l)= 2$ and $m'_i(x_i, l)=1$ for
-binary alphabets. As $\Gamma(a) = (a-1)!$ this means the top term in the middle product and the bottom
-term in the third product reduce to 1. The only remaining terms are then $m_i(l)$ and $m_i(x_i, l)$ which
-represent the number of solutions which reached leaf $l$ and the number of solutions which reached leaf $l$
-with a specific value for $x_i$, respectively.
-
-Equation~\ref{eq-hboa-stop} can also be simplified when doing comparisons. If model $B'$ has exactly one more
-leaf than model $B$ then the ratio $\frac{p(B)}{p(B')}$ simplifies to $2^{0.5 log_2\mu}$ regardless of
-total model size.
-
-The resulting simplifications create Equation~\ref{eq-hboa-final},
-where $B'$ is different from $B$ by exactly 1 split, such that $l$ was split to create $l'$ and $l''$.
-The best split is whichever maximizes its improvement over $B$, which is equal to the right side
-of the inequality. Note that these factorials can still be exceedingly large and
-therefore it is imperative that implementations avoid rounding errors and overflows.
-
-\end{document}

### 2014-10-15 Comprehensive/chapter-graybox.tex
Initial commit of Comprehensive Report

\chapter{New Problem Domain: Gray Box Optimization}

\section{Problems in this domain}

\section{Formal Requirements}

\section{Efficient Local Search}

\section{Efficient Hamming Search}

\section{Memetic Algorithms}

### 2014-10-15 Comprehensive/chapter-internal.tex
Initial commit of Comprehensive Report

--- 2014-10-15 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
@@
-\documentclass[twoside]{article}
-\usepackage{ecj,palatino,epsfig,latexsym,natbib}
-\usepackage{url}
-\usepackage{verbatim}
-\usepackage{subcaption}
-\usepackage[noend]{algpseudocode}
+\chapter{Internal Workings}
+TODO Setup why this is interesting
 
-\newcommand{\includegraphicswide}[1]
-{\includegraphics[width=.9\textwidth,height=\textheight,keepaspectratio]{#1}}
-
-\newcommand{\includegraphicsfit}[1]
-{\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}
-
-\usepackage{array}
-\newcolumntype{L}{>{\centering\arraybackslash}m{1.9cm}}
-
-%% do not add any other page- or text-size instruction here
-
-\parskip=0.00in
-
-\begin{document}
-
-% 45-character limit on header
-\ecjHeader{x}{x}{xxx-xxx}{201X}{Parameter-less Population Pyramid}{B. W. Goldman and W. F. Punch}
-\title{\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid}
-
-\author{\name{\bf B. W. Goldman} \hfill \addr{brianwgoldman@acm.org}\\ 
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-\AND
-       \name{\bf W. F. Punch} \hfill \addr{punch@msu.edu}\\
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-}
-
-\maketitle
-
-\begin{abstract}
-%Should be about 200 words
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
-evolutionary optimization without requiring any user specified parameters.
-P3's primary innovation is to replace the generational model with a pyramid of
-multiple populations that are iteratively created and expanded. In combination
-with local search and advanced crossover P3 scales to problem difficulty, exploiting
-previously learned information before adding more diversity.
-
-Across seven problems, each tested using on average 18 problem sizes, P3 outperformed
-all five advanced comparison algorithms. This improvement includes requiring less evaluations
-and less wall clock seconds to find the global optimum and better fitness when using
-the same number of evaluations. Using both algorithm analysis and comparison we find P3's
-effectiveness is due to its ability to properly maintain, add, and exploit diversity.
-
-Unlike the best comparison algorithms, P3 was able to achieve this quality without any
-problem specific tuning. Thus, unlike previous parameter-less methods, P3 does not
-sacrifice quality for applicability. Therefore we conclude that
-P3 is an efficient, general, parameter-less approach to black box
-optimization which is more effective than existing state-of-the-art techniques.
-
-\end{abstract}
-
-\begin{keywords}
-
-Genetic algorithms, 
-linkage learning,
-local search,
-parameter-less.
-
-\end{keywords}
-
-\section{Introduction}
-A primary purpose of evolutionary optimization is to efficiently find good solutions
-to challenging real world problems with minimal prior knowledge about the problem itself.
-This driving goal has created search algorithms which can escape user bias to create
-truly novel results, sometimes publishable or patentable in their own right~\citep{kannappan:2014:humies}.
-While it is not possible for any algorithm to do better than random search across all possible
-problems~\citep{Wolpert:1997:nfl}, effectiveness can be achieved by assuming the search
-landscape has structure and then biasing the algorithm toward exploiting that structure.
-
-In evolutionary optimization, and genetic algorithms~(GAs) in particular, search is often
-biased through parameters. This can be beneficial as it allows practitioners to inject their
-knowledge about the shape of the search landscape into the algorithm.
-However, the quality of solutions found, and the speed at which they are found, is strongly tied to setting these parameters
-correctly~\citep{goldberg:1991:gasize}. As such, either expert knowledge or exceedingly
-expensive parameter tuning~\citep{grefenstette:1986:optimalga} are required to leverage
-this feature to its fullest potential. Furthermore,  parameters such as population size, mutation rate, crossover
-rate, tournament size, etc. usually have no clear relationship to the problem being solved, meaning even
-domain experts may not understand how the parameters will interact with the problem or with each other.
-
-There have been periodic efforts to reduce or remove the need for parameter tuning.
-\cite{Back:1992:selfadapt} introduced self-adaptive parameters, in which parameter values
-were included in each solution's genome and themselves underwent evolution. This allowed search
-to optimize some of its own parameters resulting in a reduced the need for expert tuning.
-\cite{harik:1999:parameterlessga} were able to design an entirely parameter-less GA by
-leveraging schema theory and parallel populations. Unfortunately these methods were provably less efficient
-than directly setting the parameters to optimal values~\citep{pelikan:1999:worstparameter-less}.
-
-One area that has been very effective at reducing the number of algorithm parameters is
-model based search. \cite{pelikan:2006:hboa}'s Hierarchical Bayesian Optimization
-Algorithm~(hBOA) and \cite{thierens:2010:ltga}'s Linkage Tree Genetic Algorithm~(LTGA)
-both require only a single parameter: population size. \cite{posik:2011:parameterless}
-leveraged model building to create a fully parameter-less algorithm, but it is restricted to
-only order-k, fully decomposable, noiseless problems.
-
-Most recently \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid~(P3).
-This method uses a pyramid structure of populations to combine model based search with local search
-to achieve parameter-less optimization. Initial results suggest that unlike
-previous parameter-less methods, P3 is actually more efficient than current state-of-the-art
-parameterized search algorithms. In this work we shall: extend these results to cover more
-comparison algorithms; compare both efficiency in reaching the global optimum and intermediate
-fitnesses; algorithm complexity analysis; and provide more in depth analysis of P3 itself.
-
-\section{Comparison Optimizers}
-\label{sec-optimizers}
-
-In order to fully understand the effectiveness of P3, we compare it with five advanced algorithms which
-have related features to P3. The Random Restart Hill Climber~\citep{goldman:2014:p3} was chosen as an efficient form of repeated
-local search. As P3 combines this hill climber with crossover,
-comparing with local search alone shows the advantages of P3's overall approach.
-The $1+(\lambda, \lambda)$ algorithm is the current best theory supported simple genetic algorithm
-and its method of crossover is in some sense a macro-mutation just as in P3. hBOA and Parameter-less
-hBOA are advanced model building search techniques which are very effective at learning complex
-problem structure, designed to achieve similar goals as P3's linkage learning but using very different
-methods. Finally LTGA represents the current state-of-the-art in black box search and is the origin
-of P3's linkage learning and crossover methods.
-
-Finally, only hBOA and LTGA require any parameters, with each of these only requiring a population
-size. This makes knowing the optimal behavior of these algorithms much more tractable. All of the
-algorithms are also gene order independent, fitness scale invariant, and unbiased. This means,
-for any problem, the order in which problem variables appear in the genome can be changed
-without changing the behavior of the search. The fitness can also be manipulated in any fashion
-as long as the rank ordering of solutions is unchanged. These algorithms are also unaffected by the meaning
-assigned to each bit, such that inverting a predetermined random subset of genes before evaluation
-will not impact search efficiency.
-
-Our implementations of all of these algorithms as well as all of the population size information, raw results,
-and processing scripts
-are available from our website.\footnote{\url{https://github.com/brianwgoldman/FastEfficientP3}}
-\subsection{Random Restart Hill Climber}
-\label{sec-hill-climber}
-Perhaps the simplest black box search heuristic is stochastic local search, or hill climbing.
-This optimization technique focuses on improving a single solution until it reaches a local
-optimum. Here we use the first improvement hill climber defined by \cite{goldman:2014:p3}
-and given in Figure~\ref{fig-hc}. This algorithm works by flipping each bit in a random
-order, keeping modifications when fitness is improved, until single bit flips cannot result in
-further fitness improvements.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Hill-Climber}{}
-    \State $options \leftarrow [0 \dots N-1]$\label{fig-hc-options}
-    \State $tried \leftarrow \emptyset$
-    \While{$|tried| < |options|$}
-      \ForAll{$index \in shuffled(options)$}
-        \If{$index \notin tried$}\label{fig-hc-tried}
-          \State Flip bit $index$ in solution
-          \If{solution's fitness increased}
-            \State $tried \leftarrow \emptyset$
-          \Else
-            \State Revert change
-          \EndIf
-          \State $tried \leftarrow tried \cup \{index\}$
-        \EndIf
-      \EndFor
-    \EndWhile
-  \EndProcedure
-\end{algorithmic}
-  \caption{Hill climbing algorithm used to improve randomly generated solutions until no single
-           bit change results in a fitness improvement.}
-  \label{fig-hc}
-\end{figure}
-
-The hill climber requires an amortized cost of $O(1)$ operations per evaluation. In order to
-terminate, at least one evaluation must be performed for each of the $N$ bits in the solution.
-As such any operation that happens only once per search can be amortized over at least $N$
-evaluations, covering the initialization of $options$ on Line~\ref{fig-hc-options}.
-Line~\ref{fig-hc-tried}, which prevents wasted evaluations, can be called at most twice
-per evaluation: once to add $index$ into $tried$ and once to prevent $index$ from being unnecessarily
-evaluated again. The only way three or more calls could happen is if
-no fitness improvement was made for the entire previous iteration, which contradicts the loop invariant.
-
-Due to its nature, this hill climber cannot escape basins of attraction. Once a solution is reached
-such that none of the single bit neighbors are fitness improvements, search stops. Thus
-this algorithm requires a restart mechanism to solve multimodal problems. We have
-chosen here to naively restart search from a random solution whenever a local optima is found. This
-ensures that on all landscapes there is always a non-zero probability of search finding the global optimum.
-
-\subsection{$1+(\lambda, \lambda)$}
-\cite{doerr:2013:lambdalambda} presented the first genetic algorithm to provably show
-the advantages of performing crossover on simple landscapes. This comparatively simple
-algorithm maintains only a single individual and a self-controlled parameter $\lambda$.
-
-Each iteration, the number of bits to flip is chosen from the binomial distribution $b\sim B(N, \frac{\lambda}{N})$,
-where $N$ is the number of bits in the genome.
-Next, $\lfloor\lambda\rfloor$ offspring are produced by flipping $b$ bits. The
-best mutant then produces $\lfloor\lambda\rfloor$ offspring via uniform crossover with the original parent, such that each gene comes from the
-mutant with probability $\frac{1}{\lambda}$. In the original algorithm the best
-offspring produced by crossover then replaces the original parent if its fitness is no worse.
-The $\lambda$ parameter, which is initialized to 1, is decreased if the offspring replaced
-its parent and increased otherwise.
-
-The original formulation was designed specifically for unimodal landscapes and as such were
-not directly suitable for multimodal problems. \cite{goldman:2014:p3} extended $1+(\lambda, \lambda)$
-to include random restarts. As search stagnates, the $\lambda$ parameter increases in value. Eventually
-this results in $\lambda \ge N$ causing mutation to always flip all bits of the individual.
-As this prevents any future improvement, whenever $\lambda \ge N$ search is restarted from a random solution with $\lambda$
-reset to 1.
-
-A few other efficiency modifications were also made. If there is a tie in crossover offspring fitness,
-whichever has a larger hamming distance from the parent is retained. This encourages drifting across plateaus.
-The ``mod'' control strategy proposed by \cite{doerr:2013:lambdalambda} was not used as it conflicted with
-the random restart strategy.
-If a crossover individual is identical to either of its parents, it is not evaluated.
-If mutation produces an offspring which is better than the best crossover offspring, it is used to compare
-against the original parent.
-
-\subsection{Hierarchical Bayesian Optimization Algorithm}
-
-\cite{pelikan:2006:hboa} used statistical principles in combination with a decision tree structure
-to create the Hierarchical Bayesian Optimization Algorithm (hBOA). This method creates a model of
-epistatic relationships between genes which is then used to stochastically generate new solutions.
-Each generation a binary tournament with replacement is used to select $\mu$ solutions from
-the population. These solutions are then used to build the model, which in turn is used to generate $\mu$ new
-solutions. The new solutions are then integrated into the population using restricted tournament
-replacement.
-
-Conceptually, the model built by hBOA is trying to infer rules of the form ``Given that this
-subset of genes are set to these values, how frequently is gene $x_i$ set to value v?'' This can
-be represented using a directed acyclic decision forest, with each tree in the forest representing one gene
-in the solution. In the decision tree $T_i$, which is used to set the value of gene $x_i$,
-each internal node represents previous decisions on how to set
-some other gene $x_j$, with the children of that node representing how the decision was made. The
-leaves of each tree give the probability that $x_i$ should be set to each possible value.
-
-The forest is constructed iteratively, with each tree initially containing a single leaf
-and with each leaf storing a pointer for each selected solution. Each iteration the algorithm considers
-all possible ways of splitting an existing leaf using another gene $x_j$, such that solutions in the
-leaf are moved to the newly created leaves based on their value for $x_j$. The general goal is to
-separate the solutions such that all solutions with $x_i = 0$ move to one leaf while solutions with
-$x_j = 1$ move to the other.
-
-This goal is formalized using model scoring from Bayesian statistics. In its raw form this
-almost always creates near infinitesimal results, calculating fractions which include factorials of $\mu$
-and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{sec-appendix}, we
-derived a simplified form shown in Equation~\ref{eq-hboa-final}. If no proposed split satisfies the inequality, iteration stops.
-If multiple splits do, whichever maximizes the right side is chosen.
-
-\begin{equation}
-  2^{0.5 log_2\mu} < \frac{(m_i(l) + 1)!}{m_i(0, l)!m_i(1,l)!} \cdot
-  \frac{m_i(0, l')!m_i(1,l')!m_i(0, l'')!m_i(1,l'')!}{(m_i(l') + 1)!(m_i(l'') + 1)!}
-  \label{eq-hboa-final}
-\end{equation}
-
-Initially there are $\Theta(N^2)$ possible ways to split existing leaves, as each of the $N$ single node
-trees can be split by any of the other $N-1$ genes. Each iteration a new edge is added to the decision
-forest, meaning some of the previously tested splits cannot be used. For instance, if $T_i$, which is used
-to decide the value of $x_i$, is split using the value of $x_j$, $T_j$ can no longer be split using $x_i$.
-As a split creates two new leaves, $O(N)$ new potential splits must also be tested. Equation~\ref{eq-hboa-final}
-parses all solutions which reach a leaf to count gene frequencies, requiring $\mu$ time.
-The number of total leaves created depends heavily on the problem and $\mu$.
-However, assuming no splits are accepted or that the cost of testing all future splits is less than
-the initial $\Theta(N^2)$, constructing the model requires $\Omega(\mu N^2)$ time. Each
-model is used to generate $\mu$ solutions, leading to a cost per evaluation of $\Omega(N^2)$.
-
-To generate a solution, the value of each gene $x_i$ is set using its corresponding decision tree $T_i$. Because
-the forest is directed acyclic, there must be an ordering of $T_i$ such that before $T_i$ is executed all
-$x_j$ it uses to make decisions have already been set. As such, previous decisions made by other trees
-are used to follow each $T_i$ until a leaf is reached. The value of $x_i$ is then set based on the
-probability that other solutions reached that leaf with each value of $x_i$.
-
-To perform replacement, hBOA uses restricted tournament replacement. After each new solution is generated
-and evaluated, a set of $w$ solutions are chosen at random from the population, where $w=\min\{N, \frac{\mu}{20}\}$.
-From this set the solution which is most genetically similar to the offspring is chosen. If the offspring
-is at least as fit as the chosen solution, it replaces the chosen solution in the population. Otherwise the
-offspring is discarded. This method is designed to preserve genetic diversity as only genetically similar
-solutions must compete on fitness.
-
-hBOA is designed to work with large population sizes, resulting in a large number of
-evaluations per generation. As hBOA utilizes explicit diversity maintenance, standard methods for determining
-convergence are not considered very accurate. Therefore the authors suggest that an hBOA run should be
-terminated after performing generations equal to $N$.
-
-Like other model based techniques, hBOA has very few parameters. There is no mutation or crossover,
-and modeling does not rely on any explicit parameters. Solution selection, generation, and replacement
-are all derived from the population size, which must be set by the user.
-
-\subsection{Parameter-less hBOA}
-Using the methods first introduced by \cite{harik:1999:parameterlessga} for the Parameter-less GA,
-\cite{pelikan:2004:parameterlesshboa} created Parameter-less hBOA which automatically scales its
-population size to fit the problem. This is done by maintaining a list of concurrent populations
-using exponentially scaled population sizes.
-
-A run of Parameter-less hBOA starts with a single population of size $\mu_0$, conventionally set
-to $\mu_0=10$. After two generations are performed, a new population of size $\mu_1 = 2\mu_0$ is created
-and performs a generation. Evolution then continues with the $\mu_0$ population performing two generations
-for each one performed by $\mu_1$. Each time population $\mu_i$ performs its second generation a new population
-$\mu_{i+1}=2\mu_i$ is created, which performs generations half as often as $\mu_i$. In this way an infinite number of
-parallel population can be simulated, with each population receiving the same number of total evaluations.
-
-In all other aspects each population is identical to an hBOA population using a fixed $\mu$. No search information
-is shared among populations, and each search is independently terminated. As such Parameter-less hBOA cannot
-perform better than hBOA using the optimal population size for a given instance, as it must also spend evaluations
-on populations of different sizes. This inefficiency is bounded by a log multiple of the total number of
-evaluations~\citep{pelikan:1999:worstparameter-less}.
-
-\subsection{Linkage Tree Genetic Algorithm}
-\cite{thierens:2010:ltga} introduced the Linkage Tree Genetic Algorithm (LTGA) which automatically
-detects and exploits problem epistasis by examining pairwise gene entropy. Due to its enhanced
-ability to preserve high fitness gene subsets, LTGA was able to outperform state-of-the-art
-GAs across many benchmarks. Since its introduction, many variants of LTGA have been
-proposed~\citep{goldman:2012:ltga} so for clarity we have chosen the version
-presented by \cite{thierens:2013:ltgahiff} as our model.
-
-LTGA's effectiveness comes from its method of performing crossover. Instead of blindly
-mixing genes between parents, LTGA attempts to preserve important interrelationships
-between genes. Before performing any crossovers in a generation, LTGA first builds
-a set of hierarchical gene clusters which are then used to dictate how genes are mixed
-during crossover.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Cluster-Creation}{}
-    \State $unmerged \leftarrow \{\{0\}, \{1\}, \{2\}, \dots, \{N-1\}\}$
-    \State $useful \leftarrow unmerged$
-    \While{$|unmerged|>1$}
-      \State $C_i, C_j \leftarrow \min_{C_i,C_j \in unmerged} D(C_i, C_j)$
-      \State $unmerged \leftarrow unmerged - \{C_i, C_j\} + \{C_i \cup C_j\}$
-      \State $useful \leftarrow useful + \{C_i \cup C_j\}$
-      \If{$D(C_i, C_j) = 0$}
-        \State $useful \leftarrow useful - \{C_i, C_j\}$
-      \EndIf
-    \EndWhile
-    \State Order $useful$ based based on last merged first\label{fig-cluster-creation-ordering}
-    \State Remove largest cluster from $useful$
-
-    \Return $useful$
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm describing how LTGA creates clusters using Equation~\ref{eq-distance}
-           for a population. $unmerged$ and $useful$ are ordered sets of sets of gene loci.}
-  \label{fig-cluster-creation}
-\end{figure}
-
-Figure~\ref{fig-cluster-creation} provides the agglomerative method LTGA uses to create gene clusters.
-This algorithm creates a tree of sets using pairwise gene entropy, such that the leaves of the tree contain a single gene and
-internal nodes are the union of their children's sets. These sets are then used by crossover to specify epistatic
-relationships which should be preserved.
-The process begins by creating the set of sets $unmerged$ which tracks all top level clusters. Initially
-$unmerged$ contains single member sets for each gene. After each iteration the two sets with the minimum average pairwise
-distance (given in Equation~\ref{eq-distance}) are merged to create a single cluster. This process is repeated
-until only a single set remains in $unmerged$ which contains all of the genes in the genome.
-
-\begin{equation}
-  D(C_i,C_j) = \frac{1}{\left | C_i \right |\cdot \left |C_j \right|}\sum_{c_i \in C_i}\sum_{c_j \in C_j}
-  2 - \frac{H(c_i) + H(c_j)}{H(c_i \cup c_j)}
-  \label{eq-distance}
-\end{equation}
-\begin{equation}
-  H(c) = -\sum_{s\in S} p_c(s)\log(p_c(s))
-  \label{eq-entropy}
-\end{equation}
-
-Throughout this process $useful$ tracks the set of all gene clusters which should be preserved for use by crossover.
-This set begins with all genes in separate clusters, and each time a new cluster is created it is added to $useful$.
-However, not all clusters are necessarily worth keeping. For instance, in all versions of LTGA the cluster
-containing all genes is removed from $useful$ as preserving all genes during crossover can only create clones.
-\cite{thierens:2013:ltgahiff} extended this removal to include any unsupported subsets. If the pairwise distance
-between two clusters is 0, this means there are no individuals in the population which disrupt the relationships between the two
-clusters. Therefore when performing crossover there is no reason to believe a fitness improvement can be achieved
-by breaking the stored pattern. As such a cluster is only kept if its direct superset has a non-zero distance.
-As a final step Line~\ref{fig-cluster-creation-ordering} reorders $useful$ such that clusters appear in reversed
-order from which they were added to $useful$. Thus the most linked clusters and those containing single genes
-appear at the end of the returned list.
-
-\cite{thierens:2013:ltgahiff}'s version of LTGA does not use the entire population when determining pairwise entropy.
-Instead, binary tournament is used to select half of the population. This is done to ensure the model is built
-using only high quality solutions, even during the first generation.
-
-In order to efficiently perform clustering, a pairwise gene frequency table is constructed from the selected solutions.
-To calculate Equation~\ref{eq-distance}, Equation~\ref{eq-entropy} is called for each
-gene ($H(c_i)$) and pair of genes ($H(c_i \cup c_j)$). Extracting this information requires $O(\mu N^2)$
-time, where $\mu$ is the population size and $N$ is the genome size. The process of converting
-this pairwise frequency information into clusters can be achieved in $O(N^2)$ using the bookkeeping
-methods presented by \cite{gronau:2007:upgma}. This cost is performed only once per generation,
-and is then used to perform approximately $O(\mu N)$ crossover evaluations. As a result, the amortized cost of
-LTGA's model building is $O(N)$.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Cluster-Usage}{}
-    \ForAll{$C_i \in useful$}
-      %\ForAll{$d \in shuffled(P_i)$}
-        \State $d \leftarrow rand\_choice(P)$\label{fig-cluster-usage-donate}
-        \State Copy $d$'s gene values for $C_i$ into solution
-        \If{solution was changed}
-          \If{solution's fitness decreased}
-            \State Revert changes
-          \EndIf
-          %\State \textbf{break}
-        \EndIf
-      %\EndFor
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm describing how clusters are used to perform crossover.}
-  \label{fig-cluster-usage}
-\end{figure}
-
-Figure~\ref{fig-cluster-usage} describes how the identified clusters are used by crossover to preserve
-gene linkage while still exploring the search space. Unlike more traditional crossover methods, LTGA
-crosses each individual with the entire population. Also, to produce a single offspring, multiple evaluations
-of the fitness function are performed.
-
-Each generation, each individual in the population undergoes crossover. In a single crossover event, each
-cluster of genes $C_i$ in $useful$ is applied as a crossover mask. A random donor $d$
-is chosen from the entire population (not just the model selected population), and $d$'s gene's for $C_i$ are copied
-into the working solution. If a modification is made, an evaluation is then performed. If the crossover
-resulted in no worse fitness then the changes are kept. This allows for neutral drift across plateaus.
-The resulting solution, which must be at least as fit as its parent, is then
-copied into the next generation.
-
-In total each individual can cause up to $|useful|$ evaluations. If all clusters were kept, even those deemed
-unhelpful, and all donations were evaluated, even those which did not change any genes, then \Call{Cluster-Usage}{}
-would perform exactly $2N-2$ evaluations for each of the $\mu$ solutions in the population. This provides the amortizing evaluations
-required to make clustering only $O(N)$ operations per evaluation. However, by skipping some evaluations, its
-possible that clustering may be super-linear.
-
-LTGA has no explicit form of diversity control and has no method for introducing new genetic information once
-the population has converged. Therefore an LTGA run is considered converged once two consecutive populations
-contain the same unique solutions.
-
-By design, LTGA only has a single parameter: population size. LTGA uses no mutation, and crossover is defined
-in terms of the clustering algorithm. Selection between generations is fully elitist and embedded in the crossover,
-with selection of model building solutions fixed to a binary tournament. Neither \Call{Cluster-Creation}{} nor
-\Call{Cluster-Usage}{} rely on parameter values. LTGA does not provide any method for controlling or setting
-the population size, and uses a fixed user specified size for each generation.
-
-\section{Parameter-less Population Pyramid}
-\label{sec-p3}
-\cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid (P3) as a method for
-performing optimization which does not require the user to provide any parameters. This is
-achieved by combining efficient local search with the model building methods of LTGA using
-an iteratively constructed hierarchy of populations.
-
-The high level algorithm of P3 is presented in Figure~\ref{fig-p3}. Unlike more traditional
-GAs, P3 does not follow a generational model. Instead, it maintains an iteratively
-expanding pyramid of expanding populations. Each iteration, a new random solution is generated.
-This solution is brought to a local optimum using the hill climbing algorithm in Figure~\ref{fig-hc}. If that
-local optimum has not yet been added to the pyramid, the solution is added to the lowest
-population $P_0$.
-
-Next, the solution is iteratively improved by applying LTGA's crossover algorithm (Figure~\ref{fig-cluster-usage})
-with each population $P_i$ in the pyramid. If this process results in a strict fitness improvement and has
-created a solution not yet stored in the pyramid, it is added to the next highest pyramid level $P_{i+1}$.
-If $P_{i+1}$ does not yet exist, it is created. In this way populations in the pyramid expand over time,
-and the number of populations stored increases over time. Initially the pyramid contains no solutions
-or populations, meaning the user does not need to specify a population size.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Iterate-P3}{}
-    \State Create random solution
-    \State Apply hill climber
-    \If{solution $\notin hashset$}
-      \State Add solution to $P_0$
-      \State Add solution to $hashset$
-    \EndIf
-
-    \ForAll{$P_i \in pyramid$}
-      \State Mix solution with $P_i$
-      \If{solution's fitness has improved}
-        \If{solution $\notin hashset$}
-          \State Add solution to $P_{i+1}$
-          \State Add solution to $hashset$
-        \EndIf
-      \EndIf
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-  \caption{One iteration of P3 optimization. $pyramid$ is an
-           ordered set of populations and $hashset$ is a set
-           of all solutions in $pyramid$.}
-  \label{fig-p3}
-\end{figure}
-
-To accommodate P3's unique population structure, some of LTGA's clustering procedures were modified. In LTGA,
-clusters are identified at the start of each generation and are used
-to create all offspring in that generation. As P3 does not perform serial
-generations, P3 instead updates the model each
-time a solution is added to a population. Furthermore, unlike
-LTGA all solutions in the population are used to generate the model, not just the
-winners of a binary tournament. We can do this because even the
-worst solutions in the pyramid are already high quality due to local search.
-Using local search in LTGA was examined by \cite{bosman:2011:lsbbo} and found
-to provide no significant improvement. A likely cause was that that study applied
-local search to every solution, not just the initial population, resulting in significant overhead.
-
-Beyond the changes in population structuring, P3 modifies
-LTGA's version of \Call{Cluster-Creation}{} and \Call{Cluster-Usage}{}.
-P3 changes Line~\ref{fig-cluster-creation-ordering} in Figure~\ref{fig-cluster-creation}
-from \emph{last merged first} ordering to \emph{smallest first} ordering.
-This method applies gene clusters during crossover based on how many genes
-are in each cluster\footnote{Ties are broken randomly.}, and not on how tightly linked those genes are.
-\cite{goldman:2012:ltga} found that this alternative was better at preserving
-diversity, and therefore required smaller populations, but has not yet been integrated into
-canonical LTGA.
-
-P3 also modified Line~\ref{fig-cluster-usage-donate} in Figure~\ref{fig-cluster-usage}.
-Instead of choosing a single genetic donor for each cluster, P3 iterates over the
-population in a random order until a solution in the population is found which
-has a least one gene different for that cluster of genes from the improving solution.
-This process increases the likelihood of an evaluation being performed for every cluster,
-and helps test rare gene patterns in the population.
-
-In LTGA the cost of rebuilding the model is $O(\mu N^2)$ as it must collect pairwise
-gene frequency information for all $\mu$ solutions in the population. P3 does not store
-a single population, and it does not have a fixed $\mu$ size for any population. However,
-each time a solution is added to the population, it requires $O(N^2)$ time to update
-the table of pairwise frequencies and another $O(N^2)$ time to rebuild the linkage model.
-The model is then used immediately to perform up to one evaluation for each of the up
-to $2N-2$ clusters. Just as in LTGA, if no evaluation shortcuts were made, P3
-has an amortized cost of $O(N)$ modeling cost per evaluation. While P3 does
-rebuild the model more frequently per solution in the population, it also performs
-a number of local search evaluations which are quite efficient, meaning theoretical
-comparisons of their speed are difficult to perform. As a final note, P3's repeated
-attempts to find a useful donation make it less likely than LTGA to skip evaluations,
-but has an added cost to find these donations. Repeated donations could require as much
-as $O(\mu)$ attempts per evaluation, but experimental evidence suggest that this
-operation actually saves more overhead than it costs by increasing the number of evaluations
-per model rebuild.
-
-Each of the pieces of the P3 algorithm were selected not just for their stand alone efficacy,
-but for the ways in which they interact. By using the hill climber to ensure each solution
-is at a local optima, the underlying pairwise relationships in the problem are exposed. As
-a result, detecting clusters for use by crossover is much more effective. The crossover operator
-is extremely elitist, as each gene donation must result in no fitness loss, and a solution must
-strictly improve to be added to the next level of the pyramid. This is balanced by continual
-integration of new random solutions. Furthermore, each random restart decreases the probability
-of spurious linkages caused by shared ancestry. This diversity is further preserved by applying
-gene clusters in smallest first order during crossover as this reduces the probability of genetic
-hitchhikers.
-
-Other algorithms have proposed using multiple concurrent populations.
-\cite{hornby:2006:alps} had a hierarchy of populations with solutions periodically advancing
-upward. This allows for continuous integration of diversity as the lowest population is reseeded
-with random solutions. However, this method resulted in increased parameterization as not only was
-a population size required, but also parameters for how frequently generations advanced between levels and
-how many total levels to have were added. \cite{harik:1999:parameterlessga} used multiple independent
-populations of different sizes as a method for removing the population size parameter, but
-doing was provably less efficient than using an optimal population size as no information is shared
-between the populations.
-
-\section{Problem Descriptions}
-\label{sec-problems}
-
-\subsection{Single Instance Problems}
-Understanding how a stochastic search algorithm will behave on arbitrary and complex
-search landscapes can be exceedingly difficult. Therefore a common practice for
-algorithm understanding is to perform search on well defined, well understood
-landscapes. To be of interest these landscapes need to represent interesting
-and important aspects of real world problems.
-
-Perhaps the most common such landscape is the Deceptive Trap problem \citep{goldberg:1991:gasize}.
-In this landscape the genome is broken up into $k$ bit non-overlapping subproblems referred
-to as traps.
-Each subproblem is scored using Equation~\ref{eq-deceptive-trap}, where $t$ is the number of
-bits in the trap set to 1. The global optimum in each trap is a string of all 1s, while all
-other solutions lead to a local optima of all 0s. This problem tests an algorithm's
-ability to overcome $k$ sized deception and is commonly used to determine how effective crossover
-is at preserving building blocks. Any crossover event that mixes bits from different parents in
-the same trap will likely result in that trap being optimized to the local optima. For our experiments
-we chose $k=7$ to ensure highly deceptive traps.
-
-\begin{equation}
-   trap(t) = \left\{
-     \begin{array}{rl}
-       k-1-t, &  t<k\\
-       k,   &  t = k
-     \end{array}
-   \right.
-  \label{eq-deceptive-trap}
-\end{equation}
-
-\cite{goldman:2012:ltga} found that mixing local search with linkage learning rendered the Deceptive
-Trap problem trivial. This is because local search is able to optimize each trap to one of the two
-optima (all 1s or all 0s), and then linkage learning can have perfect knowledge of gene interactions.
-In order to make the problem more difficult, the authors proposed the Deceptive
-Step Trap problem, given in Equation~\ref{eq-deceptive-step-trap}. This function modifies the results
-of Equation~\ref{eq-deceptive-trap} to include plateaus of size $s$, introducing an exponential number
-of local optima in each trap. As a result the Deceptive Step Trap is much more challenging for linkage
-learning techniques, while still being highly deceptive. We chose to use $s=2$ to create the maximum
-number of local optima.
-\begin{equation}
-   step\_trap(t) = \left \lfloor \frac{(k-s)\pmod{s} + trap(t)}{s} \right \rfloor
-  \label{eq-deceptive-step-trap}
-\end{equation}
-
-Another challenging aspect of landscapes can be higher order relationships. The Hierarchical If
-and only If (HIFF) problem is designed to capture the difficulties of this class of problem.
-In HIFF the genome is broken up into a complete binary tree, such that each gene appears in exactly one
-leaf and each internal node is the subset of genes contained in its children. If all genes represented
-in a node of the tree are set to the same value, they score equal to the size of the set. In this way
-small subsets lead toward solutions to larger subsets. However, a node can score if all genes are either
-all 1s or all 0s, meaning that to solve higher order subproblems it is necessary to perform crossovers
-that preserve lower order solutions. This problem is a natural fit for LTGA as the linkage tree can
-perfectly duplicate the problem's true relationships \citep{thierens:2013:ltgahiff}.
-
-As a final class of well known problems, we have chosen to borrow the Rastrigin problem from real valued
-optimization. This problem's landscape, determined by Equation~\ref{eq-rast}, is highly multimodal caused by
-the oscillating cosine function. \cite{goldman:2014:p3} proposed the Discretized Rastrigin problem, such
-that each floating point $x_i$ in Equation~\ref{eq-rast} is encoded using a 10 bit gray code.
-\begin{equation}
-  An + \sum_{i=1}^{n}\left [ x_i^2-A\cos (2\pi x_i) \right ] \forall x\in [-5.12,5.12]
-  \label{eq-rast}
-\end{equation}
-
-\subsection{Randomly Generated Problem Classes}
-While well defined landscapes can provide specific insights into how an algorithm works,
-their static nature can be misleading. Specifically, algorithm quality might
-be very fragile such that it is only effective at searching well behaved landscapes. A more realistic
-test of an algorithm's black box effectiveness is to work with randomly generated instances
-drawn from a problem class. When tested over a sufficiently large sample it is then possible
-to draw more general conclusions about the algorithms effectiveness. The challenge with these
-landscapes is determining the global optimum to gauge if an algorithm was successful.
-
-Perhaps the most common model for generating random rugged landscapes is the NK model. An
-NK Landscape determines the fitness of each gene based on epistatic relationships with $K$
-other genes in the genome. This fitness is specified using a randomly generated table of
-fitness values, were each possible combination of the $K+1$ genes is mapped to some floating
-point value $[0-1]$. In unrestricted NK landscapes the relationships between genes are also
-randomly chosen and as a result finding the global optimum is $NP$-Hard for $K>1$. However,
-if epistasis is set such that each gene depends on the $K$ directly following it
-in the genome the solution can be found in polynomial time~\citep{wright:2000:solvingnk}.
-These Nearest Neighbor NK landscapes are therefore ideal for search algorithm testing.
-For all of our experiments we fixed $K=5$ to ensure highly rugged landscapes.
-
-\cite{saul:1994:spinglass} presents a combinatorial benchmark problem derived from physics:
-Ising Spin Glasses. A spin glass is defined by a weighted graph of interaction terms between vertices.
-Each gene assigns a value to each vertex, with the fitness calculated by Equation~\ref{eq-ising}.
-In this equation, $E$ is the set of all edges, $e_{ij}$ is the edge weight connecting vertex $i$ to vertex $j$, and
-$x_i$ and $x_j$ are the gene values for vertex $i$ and $j$. Optimal fitness is when this sum is minimized.
-\begin{equation}
-\sum_{e_{ij} \in E} x_ie_{ij}x_j
-  \label{eq-ising}
-\end{equation}
-Similar to NK Landscapes, the general class is $NP$-Hard to optimize, but the $2D\pm J$ subset of
-Ising Spin Glasses can be polynomially solved.\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}
-In this subset the graph is restricted to be a two dimensional torus, edge weights are randomly set
-to either -1 or 1, and vertex values must be -1 or 1.
-
-As our final class of randomly generated problems we chose the Maximum Satisfiability (MAX-SAT) problem.
-Related to the more common 3-SAT problem, a MAX-SAT instance is defined by a set of three term clauses.
-Each term is a randomly chosen variable, which may also be negated. A clause scores if an and only if
-at least one term in the clause evaluates to true. In order to make MAX-SAT instances with a known global
-optimum, \cite{goldman:2014:p3} proposed constructing clauses around a fixed solution. In this way the
-signs of the terms are set to ensure the target solution satisfies the clause. To ensure each
-problem is challenging we chose a clause to variable ratio of $4.27$.
-
-\section{Comparison Algorithm Parameter Tuning}
-\label{sec-tuning}
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-hboa}
-      \end{centering}
-      \caption{Tuned Population Sizes for hBOA}
-      \label{fig-pop-hboa}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-ltga}
-      \end{centering}
-      \caption{Tuned Population Sizes for LTGA}
-      \label{fig-pop-ltga}
-    \end{subfigure}
-  \end{centering}
-  \caption{Optimal population sizes found using bisection on each size
-           for each problem.}
-  \label{fig-pop-sizes}
-\end{figure}
-
-While four of the six algorithms in our experiments do not require any user specified parameters,
-hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
-handicapped, we extensively tuned each using the bisection method \citep{sastry:2001:bisection}
-to determine the optimal population size for each problem size.
-Extended by \cite{goldman:2012:ltga}, this method iteratively
-doubles the population size until some success criteria is met and then performs bisection
-between the lowest successful and highest unsuccessful sizes. In this way the minimum population size
-which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
-of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
-above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
-do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
-As bisection can make infinitely large population sizes, any run which had not found the global
-optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.
-
-Figure~\ref{fig-pop-sizes} shows the results from performing bisection. In general hBOA
-required population sizes which were at least an order of magnitude larger than LTGA. Due to
-runtime and memory overhead, finding the optimal value for hBOA was much less tractable than LTGA
-for moderate to large problem sizes. LTGA's population size also grew significantly slower
-than hBOA's as problem size increased, especially on the two Trap problems and Rastrigin.
-Both algorithms were very ineffective on MAX-SAT, neither able to tune to problems sizes over 60 bits.
-This is likely due to the fact that some randomly generated MAX-SAT landscapes are very flat
-and highly deceptive~\citep{rana:1998:gamaxsat}.
-\section{Finding the Global Optimum}
-\label{sec-optimum}
-
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{evals-to-success}
-  \end{center}
-  \caption{Comparison of the median number of evaluations to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-evals-to-success}
-\end{figure}
-
-Figure~\ref{fig-evals-to-success} shows the median number of evaluations required by each of the six
-algorithms to find the global optimum for multiple sizes of each problem.  Each data point in
-Figure~\ref{fig-evals-to-success} represents the median of 100 runs, where unsuccessful runs
-are treated as requiring more evaluations than any successful run. If the median run was not successful
-no point is shown. The maximum problem size used for each problem was set to be the largest size we could
-feasibly determine LTGA's optimal population size. For many larger problems results are not shown
-for hBOA due to the extreme computational cost required to optimally set the population size.
-
-\subsection{Quantitative Comparison}
-Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
-The largest problem size for any problem where P3 was not the most efficient has 49 bits, with P3 achieving the best results on all 92 larger
-configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
-Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
-and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
-on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
-
-\subsection{Local Search}
-The Random Restart Hill Climber and $1+(\lambda, \lambda)$ are both relatively effective on small problem
-sizes. This is especially true for the three randomly generated problem problem classes. These problems
-may contain relatively few local optima or just be exceptionally difficult for the model based algorithms.
-On Deceptive Trap and Deceptive Step Trap using 4 or less traps, $1+(\lambda, \lambda)$ performs significantly
-better than any other algorithm. We believe this is
-because $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
-flipping entire traps.
-This ability also leads $1+(\lambda, \lambda)$
-to outperform the Random Restart Hill Climber on all problems except Nearest Neighbor NK.
-
-On larger problem sizes, the ability for local search to reach the global optimum quickly diminishes.
-Only on MAX-SAT are these optimizers competitive
-at larger tested problem sizes. However, we believe this is because the largest tested MAX-SAT
-was an order of magnitude smaller than the largest size tested for most other problems. As the problem
-size increases the number of local optima increases exponentially, which explains why Random Restart Hill
-Climbing was unable to scale.
-For larger problems it also becomes increasingly unlikely for $1+(\lambda, \lambda)$ to make the right combination
-of changes required to reach the global optimum. This
-behavior causes high variance in success rate, as evident by the occasional successes on large Deceptive Trap problems.
-
-\subsection{Model Building}
-Only techniques which explicitly built models of gene epistasis were able to solve the largest problem
-instances. On single instance problems LTGA was more effective than hBOA, with hBOA outperforming
-LTGA on Nearest Neighbor NK and Ising Spin Glasses. This may be caused by the differences in modeling method:
-unlike the single instance problems, gene epistasis in the randomly generated
-problem classes cannot be be perfectly represented with a linkage tree.
-
-Considering how different hBOA and LTGA are in performing optimization, it is somewhat surprising how similar
-their results are on HIFF.
-However, both techniques rely on populations large enough to support the diversity required to reach the global optimum
-and to model epistasis. Both techniques also only rebuild models once per generation. As the subproblems of HIFF
-are nested, it is unlikely that either technique can accurately model higher order epistasis before solving
-lower order subproblems. Therefore both methods require one generation per subproblem order.
-
-\subsection{P3}
-Unlike the other model based methods, P3 generally outperforms
-both Random Restart Hill Climber and $1+(\lambda, \lambda)$ even on small problem sizes.
-Unlike the other local search methods, P3 outperforms LTGA and hBOA even on large problem sizes.
-This implies that P3 is gaining the benefits of each, leveraging local search to solve easy
-problems and model building to solve harder ones.
-
-Furthermore, the interaction between these two optimization tools explains some of the reason P3
-outperforms each method alone.
-On Deceptive Trap P3's use of hill climbing ensures all traps are immediately optimized, allowing for perfect
-linkage detection and high quality donation.
-On HIFF local search solves all pairwise subproblems, saving P3 a generation over LTGA and hBOA.
-In comparison P3 is only a slight improvement on Deceptive Step Trap, which is less amenable to local search.
-
-\section{Fitness Over Time}
-\label{sec-overtime}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{fitness-over-time}
-  \end{center}
-  \caption{Compares the median best fitness reached during search for each of the six optimization methods.}
-  \label{fig-fitness-over-time}
-\end{figure}
-
-For some applications, finding the global optimum is less important than finding good solutions quickly. Therefore
-we examine this behavior in Figure~\ref{fig-fitness-over-time}. At regular intervals during optimization
-Figure~\ref{fig-fitness-over-time} shows the median of the best fitnesses found at that point of search across
-100 runs. For each problem we show the largest problem size for which we were able to successfully gather results
-for all six algorithms, but the trends shown are representative of all larger problem sizes. The maximum reporting interval
-is set to include the slowest P3 run to reach the global optimum.
-
-% 181 total
-% 121 has P3 best
-%   9 hc
-%  50 lambdalambda
-%  20 hboa
-%  18 phboa
-%  27 LTGA
-
-\subsection{Quantitative Comparison}
-Of 181 sample points, P3 had the highest median fitness in 121. In pairwise competition, $1+(\lambda, \lambda)$
-was the most likely to outperform P3, doing so on 50 sample points. LTGA, hBOA, and Parameter-less hBOA
-were the next best, outperforming P3 on 27, 20, and 18 sample points. Random Restart Hill Climbing almost
-never outperformed P3, doing so only 9 times.
-
-\subsection{Local Search}
-Perhaps the most striking result is the quality of $1+(\lambda, \lambda)$. Until quite far into search this
-method performs better than both LTGA and hBOA. Given sufficient evaluations $1+(\lambda, \lambda)$ also outperforms
-Random Restart Hill Climbing on all 7 problems. For brief periods in the middle of search it performs the
-best of all techniques on: Deceptive Trap; Deceptive Step Trap; HIFF; Ising Spin Glass; and MAX-SAT problems.
-$1+(\lambda, \lambda)$'s ability to efficiently incorporate gene modifications of larger
-than one bit allows it to overcome the deception and plateaus in Deceptive Trap and Deceptive Step Trap, solve medium sized subproblems
-in HIFF, flip the signs on multiple adjacent bits in Ising Spin Glass, and cross plateaus in MAX-SAT. However,
-this method is slow in reaching the global optima in many of these problems which causes it to eventually
-be overtaken by the model building techniques.
-
-\subsection{Model Building}
-Both hBOA and LTGA are marked by periods of little improvement followed by rapid improvement.
-In hBOA this is taken to the extreme, with all fitness
-improvement made at the very end of search. In both cases this is caused by model building. Before the model
-is accurate little improvement is made. Once it is accurate, fitness improves dramatically.
-
-On every problem LTGA has five distinct periods: fitness plateau, near instantaneous improvement, fitness plateau,
-and improvement to global optimum. The early period corresponds with
-initialization of the population, with the first fitness gain achieved immediately upon completing the first generation.
-When using an inaccurate model, LTGA's mixing strategy performs a sort of less effective local search.
-Subsequent generations then make only minor fitness improvements. Once the model
-becomes accurate and the probability of a crossover using high quality genetic material increases sufficiently, LTGA
-enters a second period of rapid improvement.
-
-\begin{comment}
-This behavior is easiest to understand on the two Trap problems. LTGA's first generation will push individual traps toward a
-local optima, but it requires more than a single generation to do so completely. Until that occurs, the model is likely behaving
-no better than random as there is little apparent linkage between bits. Once the model begins to identify individual traps, crossover
-can begin to increase the frequency of higher fitness trap settings. This process begins slowly as low fitness local optima are more
-likely to be chosen for donation than high fitness local optima due to their frequency in the population. The process becomes
-self-catalysing as increased frequency of the global optimum trap genes means increased likelihood of the global optimum being spread by crossover.
-\end{comment}
-
-\subsection{P3}
-The integration of hill climbing into P3 makes it strictly better than using hill climbing alone.
-Early in optimization P3 and the Random Restart Hill Climber have effectively identical quality. This is because
-P3 performs the same evaluations as the Hill Climber for the first two restarts. Once P3 begins
-performing crossover it immediately improves over the Hill Climber. In 95\% of intervals P3 had a fitness at
-least as high as Hill Climbing. As such P3 is better than a simple
-hill climber regardless of how long each technique is run and irrespective of how high quality the solution
-found has to be.
-
-Unlike the model based methods, which struggle until model accuracy improves, P3's iterative solution
-integration allows it to improve much more quickly. This behavior exists in most problems, but is easiest to understand on Deceptive Trap.
-On this problem, P3 immediately brings all traps to local optima, equaled only by the Random Restart Hill Climber in quality.
-In comparison LTGA must evaluate the entire population and perform multiple generations to reach similar quality.
-P3 is able to immediately integrate optimal versions of each
-trap into a single individual as they are found by local search, resulting in smoother fitness improvement than LTGA.
-
-\section{Computational Expenses}
-While it is common in evolutionary computation to assume the evaluation function will dominate algorithm
-complexity, in some domains this will not be true. Model based methods are especially likely to violate
-this norm. Therefore, in order to assess P3's quality in solving problems with efficient fitness functions,
-we provide data on both its algorithmic complexity and wall clock time.
-
-\subsection{Operation Counting}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{rebuilds}
-      \end{centering}
-      \caption{Rebuilds}
-      \label{fig-rebuilds}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{donations}
-      \end{centering}
-      \caption{Donations}
-      \label{fig-donations}
-    \end{subfigure}
-  \end{centering}
-  \caption{Computation costs incurred by model rebuilding (Figure~\ref{fig-rebuilds}) and
-           repeated donations (Figure~\ref{fig-donations}) per evaluation as problem size increases.}
-  \label{fig-costs}
-\end{figure}
-
-When discussing the asymptotic complexity of P3 in Section~\ref{sec-p3}, two aspects eluded precise
-analysis: how expensive is model rebuilding and how many gene donations are made. Figure~\ref{fig-costs}
-provides some insight into how often these two aspects of the algorithm are utilized.
-
-Figure~\ref{fig-rebuilds} reports in an algorithmic sense how expensive model rebuilding is for search.
-In order to calculate this value we recorded how many times search rebuilt the model during each run.
-Under the minimal assumption that an evaluation costs $\Omega(N)$ time and that model rebuilds cost $O(N^2)$,
-Figure~\ref{fig-rebuilds} shows the ratio cost of model rebuilding per evaluation. If the cost of model building
-scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically constant.
-For Nearest Neighbor NK, Ising Spin Glasses, and Rastrigin this is the case. For both Trap problems
-and HIFF there is slow growth in the ratio. The problem sizes used for MAX-SAT were not sufficient
-to accurately gauge the asymptotic behavior. Together this suggests that while the cost of building the model
-is almost linear per evaluation, it can grow slowly. However, even in the worst case (HIFF) this growth was
-no more than twice the algorithmic cost of an evaluation even using 2048 bits.
-
-When applying a crossover subset, P3 tries random donors from the population until one is found with at least
-one bit different from the improving solution. In theory this can result in up to $O(\mu)$ operations.
-Figure~\ref{fig-donations} examines the observed average number of donations per evaluation performed.
-Ising Spin Glass, HIFF, and Rastrigin all achieve effectively constant behavior here, implying repeated
-donation does not impact the asymptotic runtime of P3. Both Trap functions and Nearest Neighbor NK all
-increase in number of donations as problem size increases, potentially increasing algorithmic costs. An
-important note is that each donation may range in size from a single bit up to $N-1$, with smaller
-donations far more likely to result in repeated attempts. As such this may cause some super-linear growth
-in P3, but its unlikely to be very high.
-
-\subsection{Wall Clock Performance}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{seconds-to-success}
-  \end{center}
-  \caption{Comparison of the median number of seconds to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-seconds-to-success}
-\end{figure}
-
-To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}.
-Similar to Figure~\ref{fig-evals-to-success}, each point represents the median
-of 100 runs, with unsuccessful runs treated as slower than successful runs.
-In total these runs
-took over 6 computing years to complete. We were therefore forced to use
-diverse hardware potentially running multiple processes in parallel.
-As such these timings should be considered very skeptically as a number
-of sources may contribute to biased error in the results. However, due to
-the size of the dataset and the general regularity in the results some general conclusions
-can still be made.
-
-\subsubsection{Model Building}
-hBOA and Parameter-less hBOA perform significantly worse when using wall clock time as the unit of comparison than when using
-evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while, under reasonable
-assumptions, P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
-hBOA goes from being slightly more efficient than LTGA in terms of evaluations to significantly worse in terms of seconds.
-As P3 and LTGA require a similar asymptotic complexity per
-evaluation as the Hill Climber and $1+(\lambda, \lambda)$, no similar change in ordering occurs.
-
-\subsubsection{P3}
-Any increase in P3's algorithmic complexity is offset by its efficient use of evaluations.
-In aggregate P3 found the global optimum the fastest on 72 of 130 configurations,
-with LTGA, $1+(\lambda, \lambda)$, Hill Climber, hBOA outperforming P3 on 44, 18, 10, and 1 configurations, respectively.
-On
-Deceptive Trap, Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT P3 still finds the global
-optimum faster than all comparison techniques. LTGA is able to match P3's performance almost exactly on HIFF,
-and on Rastrigin P3 falls behind only on moderate sized problems. Deceptive Step Trap yields the worst performance,
-with LTGA using optimal population sizes having a significantly lower growth curve than P3. This mirrors
-the results from Figure~\ref{fig-evals-to-success} under the assumption that P3's computational time per evaluation is higher
-than LTGA but by no more than a constant factor.
-
+%%%%%%%%%%%%% Population Sizing %%%%%%%%%%%%%
 \section{Population Sizing}
@@
 
-\begin{comment}
-It is important to note that hBOA and LTGA were tuned specifically
-to find the global optimum and are not necessarily using the optimal population size for finding intermediate fitnesses. Consider
-that for both reducing the population size would almost certainly improve their fitness for evaluation steps less than their population
-size. To examine the effect of population size we tested LTGA using one tenth of the population size required for reliably finding the
-global optimum, with the results given in Figure~\ref{fig-small-pop}.
-Reducing the population size resulted in LTGA's initial fitness improvement
-occurring earlier, as the first generation is completed much more quickly. On Deceptive Step Trap this means that
-LTGA leaps ahead of P3 due to its ability to overcome the two bit fitness plateaus.
-%
-Across all tested problems using the smaller
-population size reduced the quality of LTGA's first fitness plateau, likely due to missing required diversity to reach higher quality.
-The second period of improvement also comes earlier and reaches a lower quality fitness. On Deceptive Step Trap 0
-runs reach the global optimum, down from 100 successful for the full population size. On Nearest Neighbor NK this drops the number
-of successful runs from 98 to 68. As a result we conclude that while reducing the population size of LTGA may improve its quality
-at early points during optimization, doing so reduces the robustness of the final solution found. This is in contrast to P3
-which balances high quality intermediate fitness without trading away eventual optimality.
-\end{comment}
-
-\section{Inner Workings}
+%%%%%%%%%%%%%% P3 Inner Workings %%%%%%%%%%%%%%%%%%%%%%%
+\section{Inner Workings Specific to P3}
 
@@
 exponential number of plateaus makes detecting gene linkage very challenging.
-
-\begin{comment}
-P3 performs optimization using a mixture of hill climbing and crossover. Exactly how much time the algorithm
-dedicates to each is not clear from the algorithm description alone. As such, Figure~\ref{fig-cross} shows
-that the relationship between the two is strongly problem dependent but relatively independent of problem size.
-In general, problems which can be more accurately represented by crossover
-required less crossover evaluations. Deceptive Trap and Rastrigin spent the least evaluations on crossover, using
-15\% and 20\% respectively on the largest problem sizes. Both of these problems are perfectly separable and have relatively
-few local optima in each subproblem. HIFF and Ising Spin Glasses use crossover for about 60\% of evaluations. The HIFF structure can
-be perfectly learned by the linkage tree, but the higher order relationships require multiple stages of crossover to optimize.
-Ising Spin Glasses have an overlapping structure, making them impossible to perfectly represent in a single linkage tree, but
-that overlap is relatively minor in comparison with Nearest Neighbor NK, which uses 81\% of evaluations on crossover. Deceptive Step Trap spends the highest
-percentage of evaluations on crossover, 92\%. Unlike all of the other problems, the value taken by an individual bit in Deceptive
-Step Trap has very little meaning. This is because any trap value is either a local optima or 1 bit flip away from being a local optima.
-This has a combined result of meaning it takes very few local search evaluations to find a local optima but very many crossover operations
-to proliferate the global optima. Epistatic relationships are also going to be very weak and even small population sizes are likely
-to mark all clusters as $useful$.
-
-Solutions are only added to the next highest pyramid level if at least one crossover donation resulted in a fitness
-improvement. Figure~\ref{fig-cross-success} examines the percentage of crossover donations that resulted in a fitness
-improvement. The ordering in Figure~\ref{fig-cross-success} is an almost perfect inversion of Figure~\ref{fig-cross},
-implying that the more successful crossover is at achieving fitness improving donations the lower the proportion of evaluations P3 will
-spend performing crossover. P3 achieves an impressive 30\% success rate for crossover donations on the largest Deceptive Trap
-problem sizes. This makes sense as crossover is able to immediately and perfectly identify entire traps such that all donations
-always contain only complete traps. Furthermore, by forcing crossover to attempt donations of genes different from the current solution,
-donations containing a single trap will have a very high success rate, always improving suboptimal traps which are far more frequently
-created by local search that optimal traps. Rastrigin and HIFF have relatively high success rates of 11\% and 7\%, respectively. Again
-this is likely due to accurate modeling and relatively low diversity of building block alternatives.
-Ising Spin Glasses and Deceptive Step Trap both have low success rates near 0.01\%. This follows as both are difficult to model with
-a linkage tree. In comparison to Figure~\ref{fig-cross-success} Deceptive Step Trap performs better than expected for crossover.
-However, the reason for this is that even exceedingly uninformed crossover can make improvements to Deceptive Step Trap as
-flipping two bits in the same trap will move between two local optima. Nearest Neighbor NK had the least successful crossovers,
-with only 0.007\% successful on the largest tested problems. In comparison to Ising Spin Glass, NK's overlap is far more extensive.
-As a result it makes sense that individual crossovers are very unlikely to be fitness improvements.
-\end{comment}
 
@@
 Thus the level size and crossover success rates increase after contracting around level four.
-
-\section{Conclusions and Future Work}
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing black box
-optimization. P3's primary innovation is the replacement of the generational model with a pyramid of populations.
-This pyramid is constructed iteratively, with both the number of levels and the number of solutions stored
-at each level growing as search progresses. P3 uses a model based crossover method
-which learns a linkage tree from gene epistasis. Combined with a simple hill climber, P3's design contains
-many synergistic features.
-
-Across a large number of problems and problem sizes P3 required less evaluations to reach the global optimum
-than optimally tuned state-of-the-art competitors. On single instance problems P3's improvement was by a
-constant factor, while for the three problem classes P3's improvement increased with problem size. This
-quality extends to intermediate points during evolution, with P3 generally reaching at least as high
-of fitness as the competitive techniques when using the same number of evaluations. While P3 does require
-modeling overhead, the expense of this overhead is approximately linear with respect to genome size. There
-is some evidence that even when compared on wall clock time, P3 performs at least as well as
-the best comparison techniques. All of these achievements are made without any problem specific parameter
-tuning, making P3 easier to apply to new domains than its two closets competitors in quality.
-
-P3's quality is due to a number of desirable traits. First, mixing local search with model based
-crossover lets search focus on properly mixing high quality solutions. Second, by adding diversity only
-as necessary P3 tends to use the minimal amount of random initialization, unlike other techniques which must
-overcompensate with larger population sizes on single instance problems and consider the worst instance
-when solving problem classes. Third, by heavily exploiting existing diversity before adding more P3 is able
-to reach high quality intermediate fitnesses quickly without prematurely converging. Fourth, the very
-nature of the pyramid's shape allows search to preserve a desirable proportion of diversity at
-each fitness level, similar to a generational model using a decreasing population size.
-
-There are a number of meaningful avenues for future P3 experimentation. Perhaps the most
-pressing for practitioner acceptance is to apply P3 to real world problems and compare
-its results with other black box or even problem specific heuristics. While parameter-less,
-P3 is currently limited to discrete, fixed length genomes evaluated using single objective
-fitness. These limitations can be relaxed with future work to make P3 more widely applicable.
-While asymptotically linear, P3's modeling techniques and local search methods are likely going
-to be prohibitively expensive for genome sizes in the hundreds of thousands or millions of genes,
-and the inability of the model to capture overlapping linkage may be hindering
-search efficiency. Overcoming these limitations by using a new modeling technique may allow
-the pyramid model even greater flexibility. Similarly, while P3 is able to overcome low
-order deception via linkage learning, the iterative improvement method by which crossovers
-are made may mislead search on landscapes with higher order deception.
-
-However, even without
-these improvements our results show P3 is highly efficient at finding
-global optima on black box problems without any problem specific tuning.
-
-\small
-
-\bibliographystyle{apalike}
-\bibliography{../main}
-
-\normalsize
-\appendix
-\section{hBOA Simplification}
-\label{sec-appendix}
-To measure the quality of a decision forest, hBOA applies Equation~\ref{eq-hboa}.
-To favor compact
-models, Equation~\ref{eq-hboa} is scaled by Equation~\ref{eq-hboa-stop}, which provides increased cost
-for more total leaves.
-This quality is used for two purposes: comparison of potential changes from the existing model
-and comparison of that change with the existing model. The basis for our simplication
-is to rearrange $p(B)BDe(B) < p(B')BDe(B')$ to be $\frac{p(B)}{p(B')}<\frac{BDe(B')}{BDe(B)}$.
-
-\begin{equation}
-  BDe(B) = \prod_{i=0}^{N}\prod_{l\in L_i} \frac{\Gamma(m'_i(l))}{\Gamma(m_i(l) + m'_i(l))}
-  \prod_{x_i}\frac{\Gamma(m_i(x_i, l) + m'_i(x_i,l))}{\Gamma(m'_i(x_i,l))}
-  \label{eq-hboa}
-\end{equation}
-
-\begin{equation}
-  p(B) = c2^{-0.5(\sum_i|L_i|)log_2\mu}
-  \label{eq-hboa-stop}
-\end{equation}
-
-The outermost product of Equation~\ref{eq-hboa} iterates over all trees in the forest. However, each split
-can modify only one of the trees and therefore the contribution of all others can be canceled. The middle
-product is across all leaves in the tree. Again since only one leaf can be changed, all other terms can
-be canceled. By convention hBOA uses uninformed Bayesian priors of $m'_i(l)= 2$ and $m'_i(x_i, l)=1$ for
-binary alphabets. As $\Gamma(a) = (a-1)!$ this means the top term in the middle product and the bottom
-term in the third product reduce to 1. The only remaining terms are then $m_i(l)$ and $m_i(x_i, l)$ which
-represent the number of solutions which reached leaf $l$ and the number of solutions which reached leaf $l$
-with a specific value for $x_i$, respectively.
-
-Equation~\ref{eq-hboa-stop} can also be simplified when doing comparisons. If model $B'$ has exactly one more
-leaf than model $B$ then the ratio $\frac{p(B)}{p(B')}$ simplifies to $2^{0.5 log_2\mu}$ regardless of
-total model size.
-
-The resulting simplifications create Equation~\ref{eq-hboa-final},
-where $B'$ is different from $B$ by exactly 1 split, such that $l$ was split to create $l'$ and $l''$.
-The best split is whichever maximizes its improvement over $B$, which is equal to the right side
-of the inequality. Note that these factorials can still be exceedingly large and
-therefore it is imperative that implementations avoid rounding errors and overflows.
-
-\end{document}

### 2014-10-15 Comprehensive/chapter-introduction.tex
Initial commit of Comprehensive Report

--- 2014-10-15 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
@@
-\documentclass[twoside]{article}
-\usepackage{ecj,palatino,epsfig,latexsym,natbib}
-\usepackage{url}
-\usepackage{verbatim}
-\usepackage{subcaption}
-\usepackage[noend]{algpseudocode}
-
-\newcommand{\includegraphicswide}[1]
-{\includegraphics[width=.9\textwidth,height=\textheight,keepaspectratio]{#1}}
-
-\newcommand{\includegraphicsfit}[1]
-{\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}
-
-\usepackage{array}
-\newcolumntype{L}{>{\centering\arraybackslash}m{1.9cm}}
-
-%% do not add any other page- or text-size instruction here
-
-\parskip=0.00in
-
-\begin{document}
-
-% 45-character limit on header
-\ecjHeader{x}{x}{xxx-xxx}{201X}{Parameter-less Population Pyramid}{B. W. Goldman and W. F. Punch}
-\title{\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid}
-
-\author{\name{\bf B. W. Goldman} \hfill \addr{brianwgoldman@acm.org}\\ 
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-\AND
-       \name{\bf W. F. Punch} \hfill \addr{punch@msu.edu}\\
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-}
-
-\maketitle
-
-\begin{abstract}
-%Should be about 200 words
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
-evolutionary optimization without requiring any user specified parameters.
-P3's primary innovation is to replace the generational model with a pyramid of
-multiple populations that are iteratively created and expanded. In combination
-with local search and advanced crossover P3 scales to problem difficulty, exploiting
-previously learned information before adding more diversity.
-
-Across seven problems, each tested using on average 18 problem sizes, P3 outperformed
-all five advanced comparison algorithms. This improvement includes requiring less evaluations
-and less wall clock seconds to find the global optimum and better fitness when using
-the same number of evaluations. Using both algorithm analysis and comparison we find P3's
-effectiveness is due to its ability to properly maintain, add, and exploit diversity.
-
-Unlike the best comparison algorithms, P3 was able to achieve this quality without any
-problem specific tuning. Thus, unlike previous parameter-less methods, P3 does not
-sacrifice quality for applicability. Therefore we conclude that
-P3 is an efficient, general, parameter-less approach to black box
-optimization which is more effective than existing state-of-the-art techniques.
-
-\end{abstract}
-
-\begin{keywords}
-
-Genetic algorithms, 
-linkage learning,
-local search,
-parameter-less.
-
-\end{keywords}
-
-\section{Introduction}
+\chapter{Introduction}
+\label{chap-introduction}
+%TODO Expand motivation and assume less reader knowledge
 A primary purpose of evolutionary optimization is to efficiently find good solutions
@@
 This driving goal has created search algorithms which can escape user bias to create
-truly novel results, sometimes publishable or patentable in their own right~\citep{kannappan:2014:humies}.
+truly novel results, sometimes publishable or patentable in their own right~\cite{kannappan:2014:humies}.
 While it is not possible for any algorithm to do better than random search across all possible
-problems~\citep{Wolpert:1997:nfl}, effectiveness can be achieved by assuming the search
+problems~\cite{Wolpert:1997:nfl}, effectiveness can be achieved by assuming the search
 landscape has structure and then biasing the algorithm toward exploiting that structure.
@@
 However, the quality of solutions found, and the speed at which they are found, is strongly tied to setting these parameters
-correctly~\citep{goldberg:1991:gasize}. As such, either expert knowledge or exceedingly
-expensive parameter tuning~\citep{grefenstette:1986:optimalga} are required to leverage
+correctly~\cite{goldberg:1991:gasize}. As such, either expert knowledge or exceedingly
+expensive parameter tuning~\cite{grefenstette:1986:optimalga} are required to leverage
 this feature to its fullest potential. Furthermore,  parameters such as population size, mutation rate, crossover
@@
 leveraging schema theory and parallel populations. Unfortunately these methods were provably less efficient
-than directly setting the parameters to optimal values~\citep{pelikan:1999:worstparameter-less}.
+than directly setting the parameters to optimal values~\cite{pelikan:1999:worstparameter-less}.
 
@@
 fitnesses; algorithm complexity analysis; and provide more in depth analysis of P3 itself.
-
-\section{Comparison Optimizers}
-\label{sec-optimizers}
-
-In order to fully understand the effectiveness of P3, we compare it with five advanced algorithms which
-have related features to P3. The Random Restart Hill Climber~\citep{goldman:2014:p3} was chosen as an efficient form of repeated
-local search. As P3 combines this hill climber with crossover,
-comparing with local search alone shows the advantages of P3's overall approach.
-The $1+(\lambda, \lambda)$ algorithm is the current best theory supported simple genetic algorithm
-and its method of crossover is in some sense a macro-mutation just as in P3. hBOA and Parameter-less
-hBOA are advanced model building search techniques which are very effective at learning complex
-problem structure, designed to achieve similar goals as P3's linkage learning but using very different
-methods. Finally LTGA represents the current state-of-the-art in black box search and is the origin
-of P3's linkage learning and crossover methods.
-
-Finally, only hBOA and LTGA require any parameters, with each of these only requiring a population
-size. This makes knowing the optimal behavior of these algorithms much more tractable. All of the
-algorithms are also gene order independent, fitness scale invariant, and unbiased. This means,
-for any problem, the order in which problem variables appear in the genome can be changed
-without changing the behavior of the search. The fitness can also be manipulated in any fashion
-as long as the rank ordering of solutions is unchanged. These algorithms are also unaffected by the meaning
-assigned to each bit, such that inverting a predetermined random subset of genes before evaluation
-will not impact search efficiency.
-
-Our implementations of all of these algorithms as well as all of the population size information, raw results,
-and processing scripts
-are available from our website.\footnote{\url{https://github.com/brianwgoldman/FastEfficientP3}}
-\subsection{Random Restart Hill Climber}
-\label{sec-hill-climber}
-Perhaps the simplest black box search heuristic is stochastic local search, or hill climbing.
-This optimization technique focuses on improving a single solution until it reaches a local
-optimum. Here we use the first improvement hill climber defined by \cite{goldman:2014:p3}
-and given in Figure~\ref{fig-hc}. This algorithm works by flipping each bit in a random
-order, keeping modifications when fitness is improved, until single bit flips cannot result in
-further fitness improvements.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Hill-Climber}{}
-    \State $options \leftarrow [0 \dots N-1]$\label{fig-hc-options}
-    \State $tried \leftarrow \emptyset$
-    \While{$|tried| < |options|$}
-      \ForAll{$index \in shuffled(options)$}
-        \If{$index \notin tried$}\label{fig-hc-tried}
-          \State Flip bit $index$ in solution
-          \If{solution's fitness increased}
-            \State $tried \leftarrow \emptyset$
-          \Else
-            \State Revert change
-          \EndIf
-          \State $tried \leftarrow tried \cup \{index\}$
-        \EndIf
-      \EndFor
-    \EndWhile
-  \EndProcedure
-\end{algorithmic}
-  \caption{Hill climbing algorithm used to improve randomly generated solutions until no single
-           bit change results in a fitness improvement.}
-  \label{fig-hc}
-\end{figure}
-
-The hill climber requires an amortized cost of $O(1)$ operations per evaluation. In order to
-terminate, at least one evaluation must be performed for each of the $N$ bits in the solution.
-As such any operation that happens only once per search can be amortized over at least $N$
-evaluations, covering the initialization of $options$ on Line~\ref{fig-hc-options}.
-Line~\ref{fig-hc-tried}, which prevents wasted evaluations, can be called at most twice
-per evaluation: once to add $index$ into $tried$ and once to prevent $index$ from being unnecessarily
-evaluated again. The only way three or more calls could happen is if
-no fitness improvement was made for the entire previous iteration, which contradicts the loop invariant.
-
-Due to its nature, this hill climber cannot escape basins of attraction. Once a solution is reached
-such that none of the single bit neighbors are fitness improvements, search stops. Thus
-this algorithm requires a restart mechanism to solve multimodal problems. We have
-chosen here to naively restart search from a random solution whenever a local optima is found. This
-ensures that on all landscapes there is always a non-zero probability of search finding the global optimum.
-
-\subsection{$1+(\lambda, \lambda)$}
-\cite{doerr:2013:lambdalambda} presented the first genetic algorithm to provably show
-the advantages of performing crossover on simple landscapes. This comparatively simple
-algorithm maintains only a single individual and a self-controlled parameter $\lambda$.
-
-Each iteration, the number of bits to flip is chosen from the binomial distribution $b\sim B(N, \frac{\lambda}{N})$,
-where $N$ is the number of bits in the genome.
-Next, $\lfloor\lambda\rfloor$ offspring are produced by flipping $b$ bits. The
-best mutant then produces $\lfloor\lambda\rfloor$ offspring via uniform crossover with the original parent, such that each gene comes from the
-mutant with probability $\frac{1}{\lambda}$. In the original algorithm the best
-offspring produced by crossover then replaces the original parent if its fitness is no worse.
-The $\lambda$ parameter, which is initialized to 1, is decreased if the offspring replaced
-its parent and increased otherwise.
-
-The original formulation was designed specifically for unimodal landscapes and as such were
-not directly suitable for multimodal problems. \cite{goldman:2014:p3} extended $1+(\lambda, \lambda)$
-to include random restarts. As search stagnates, the $\lambda$ parameter increases in value. Eventually
-this results in $\lambda \ge N$ causing mutation to always flip all bits of the individual.
-As this prevents any future improvement, whenever $\lambda \ge N$ search is restarted from a random solution with $\lambda$
-reset to 1.
-
-A few other efficiency modifications were also made. If there is a tie in crossover offspring fitness,
-whichever has a larger hamming distance from the parent is retained. This encourages drifting across plateaus.
-The ``mod'' control strategy proposed by \cite{doerr:2013:lambdalambda} was not used as it conflicted with
-the random restart strategy.
-If a crossover individual is identical to either of its parents, it is not evaluated.
-If mutation produces an offspring which is better than the best crossover offspring, it is used to compare
-against the original parent.
-
-\subsection{Hierarchical Bayesian Optimization Algorithm}
-
-\cite{pelikan:2006:hboa} used statistical principles in combination with a decision tree structure
-to create the Hierarchical Bayesian Optimization Algorithm (hBOA). This method creates a model of
-epistatic relationships between genes which is then used to stochastically generate new solutions.
-Each generation a binary tournament with replacement is used to select $\mu$ solutions from
-the population. These solutions are then used to build the model, which in turn is used to generate $\mu$ new
-solutions. The new solutions are then integrated into the population using restricted tournament
-replacement.
-
-Conceptually, the model built by hBOA is trying to infer rules of the form ``Given that this
-subset of genes are set to these values, how frequently is gene $x_i$ set to value v?'' This can
-be represented using a directed acyclic decision forest, with each tree in the forest representing one gene
-in the solution. In the decision tree $T_i$, which is used to set the value of gene $x_i$,
-each internal node represents previous decisions on how to set
-some other gene $x_j$, with the children of that node representing how the decision was made. The
-leaves of each tree give the probability that $x_i$ should be set to each possible value.
-
-The forest is constructed iteratively, with each tree initially containing a single leaf
-and with each leaf storing a pointer for each selected solution. Each iteration the algorithm considers
-all possible ways of splitting an existing leaf using another gene $x_j$, such that solutions in the
-leaf are moved to the newly created leaves based on their value for $x_j$. The general goal is to
-separate the solutions such that all solutions with $x_i = 0$ move to one leaf while solutions with
-$x_j = 1$ move to the other.
-
-This goal is formalized using model scoring from Bayesian statistics. In its raw form this
-almost always creates near infinitesimal results, calculating fractions which include factorials of $\mu$
-and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{sec-appendix}, we
-derived a simplified form shown in Equation~\ref{eq-hboa-final}. If no proposed split satisfies the inequality, iteration stops.
-If multiple splits do, whichever maximizes the right side is chosen.
-
-\begin{equation}
-  2^{0.5 log_2\mu} < \frac{(m_i(l) + 1)!}{m_i(0, l)!m_i(1,l)!} \cdot
-  \frac{m_i(0, l')!m_i(1,l')!m_i(0, l'')!m_i(1,l'')!}{(m_i(l') + 1)!(m_i(l'') + 1)!}
-  \label{eq-hboa-final}
-\end{equation}
-
-Initially there are $\Theta(N^2)$ possible ways to split existing leaves, as each of the $N$ single node
-trees can be split by any of the other $N-1$ genes. Each iteration a new edge is added to the decision
-forest, meaning some of the previously tested splits cannot be used. For instance, if $T_i$, which is used
-to decide the value of $x_i$, is split using the value of $x_j$, $T_j$ can no longer be split using $x_i$.
-As a split creates two new leaves, $O(N)$ new potential splits must also be tested. Equation~\ref{eq-hboa-final}
-parses all solutions which reach a leaf to count gene frequencies, requiring $\mu$ time.
-The number of total leaves created depends heavily on the problem and $\mu$.
-However, assuming no splits are accepted or that the cost of testing all future splits is less than
-the initial $\Theta(N^2)$, constructing the model requires $\Omega(\mu N^2)$ time. Each
-model is used to generate $\mu$ solutions, leading to a cost per evaluation of $\Omega(N^2)$.
-
-To generate a solution, the value of each gene $x_i$ is set using its corresponding decision tree $T_i$. Because
-the forest is directed acyclic, there must be an ordering of $T_i$ such that before $T_i$ is executed all
-$x_j$ it uses to make decisions have already been set. As such, previous decisions made by other trees
-are used to follow each $T_i$ until a leaf is reached. The value of $x_i$ is then set based on the
-probability that other solutions reached that leaf with each value of $x_i$.
-
-To perform replacement, hBOA uses restricted tournament replacement. After each new solution is generated
-and evaluated, a set of $w$ solutions are chosen at random from the population, where $w=\min\{N, \frac{\mu}{20}\}$.
-From this set the solution which is most genetically similar to the offspring is chosen. If the offspring
-is at least as fit as the chosen solution, it replaces the chosen solution in the population. Otherwise the
-offspring is discarded. This method is designed to preserve genetic diversity as only genetically similar
-solutions must compete on fitness.
-
-hBOA is designed to work with large population sizes, resulting in a large number of
-evaluations per generation. As hBOA utilizes explicit diversity maintenance, standard methods for determining
-convergence are not considered very accurate. Therefore the authors suggest that an hBOA run should be
-terminated after performing generations equal to $N$.
-
-Like other model based techniques, hBOA has very few parameters. There is no mutation or crossover,
-and modeling does not rely on any explicit parameters. Solution selection, generation, and replacement
-are all derived from the population size, which must be set by the user.
-
-\subsection{Parameter-less hBOA}
-Using the methods first introduced by \cite{harik:1999:parameterlessga} for the Parameter-less GA,
-\cite{pelikan:2004:parameterlesshboa} created Parameter-less hBOA which automatically scales its
-population size to fit the problem. This is done by maintaining a list of concurrent populations
-using exponentially scaled population sizes.
-
-A run of Parameter-less hBOA starts with a single population of size $\mu_0$, conventionally set
-to $\mu_0=10$. After two generations are performed, a new population of size $\mu_1 = 2\mu_0$ is created
-and performs a generation. Evolution then continues with the $\mu_0$ population performing two generations
-for each one performed by $\mu_1$. Each time population $\mu_i$ performs its second generation a new population
-$\mu_{i+1}=2\mu_i$ is created, which performs generations half as often as $\mu_i$. In this way an infinite number of
-parallel population can be simulated, with each population receiving the same number of total evaluations.
-
-In all other aspects each population is identical to an hBOA population using a fixed $\mu$. No search information
-is shared among populations, and each search is independently terminated. As such Parameter-less hBOA cannot
-perform better than hBOA using the optimal population size for a given instance, as it must also spend evaluations
-on populations of different sizes. This inefficiency is bounded by a log multiple of the total number of
-evaluations~\citep{pelikan:1999:worstparameter-less}.
-
-\subsection{Linkage Tree Genetic Algorithm}
-\cite{thierens:2010:ltga} introduced the Linkage Tree Genetic Algorithm (LTGA) which automatically
-detects and exploits problem epistasis by examining pairwise gene entropy. Due to its enhanced
-ability to preserve high fitness gene subsets, LTGA was able to outperform state-of-the-art
-GAs across many benchmarks. Since its introduction, many variants of LTGA have been
-proposed~\citep{goldman:2012:ltga} so for clarity we have chosen the version
-presented by \cite{thierens:2013:ltgahiff} as our model.
-
-LTGA's effectiveness comes from its method of performing crossover. Instead of blindly
-mixing genes between parents, LTGA attempts to preserve important interrelationships
-between genes. Before performing any crossovers in a generation, LTGA first builds
-a set of hierarchical gene clusters which are then used to dictate how genes are mixed
-during crossover.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Cluster-Creation}{}
-    \State $unmerged \leftarrow \{\{0\}, \{1\}, \{2\}, \dots, \{N-1\}\}$
-    \State $useful \leftarrow unmerged$
-    \While{$|unmerged|>1$}
-      \State $C_i, C_j \leftarrow \min_{C_i,C_j \in unmerged} D(C_i, C_j)$
-      \State $unmerged \leftarrow unmerged - \{C_i, C_j\} + \{C_i \cup C_j\}$
-      \State $useful \leftarrow useful + \{C_i \cup C_j\}$
-      \If{$D(C_i, C_j) = 0$}
-        \State $useful \leftarrow useful - \{C_i, C_j\}$
-      \EndIf
-    \EndWhile
-    \State Order $useful$ based based on last merged first\label{fig-cluster-creation-ordering}
-    \State Remove largest cluster from $useful$
-
-    \Return $useful$
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm describing how LTGA creates clusters using Equation~\ref{eq-distance}
-           for a population. $unmerged$ and $useful$ are ordered sets of sets of gene loci.}
-  \label{fig-cluster-creation}
-\end{figure}
-
-Figure~\ref{fig-cluster-creation} provides the agglomerative method LTGA uses to create gene clusters.
-This algorithm creates a tree of sets using pairwise gene entropy, such that the leaves of the tree contain a single gene and
-internal nodes are the union of their children's sets. These sets are then used by crossover to specify epistatic
-relationships which should be preserved.
-The process begins by creating the set of sets $unmerged$ which tracks all top level clusters. Initially
-$unmerged$ contains single member sets for each gene. After each iteration the two sets with the minimum average pairwise
-distance (given in Equation~\ref{eq-distance}) are merged to create a single cluster. This process is repeated
-until only a single set remains in $unmerged$ which contains all of the genes in the genome.
-
-\begin{equation}
-  D(C_i,C_j) = \frac{1}{\left | C_i \right |\cdot \left |C_j \right|}\sum_{c_i \in C_i}\sum_{c_j \in C_j}
-  2 - \frac{H(c_i) + H(c_j)}{H(c_i \cup c_j)}
-  \label{eq-distance}
-\end{equation}
-\begin{equation}
-  H(c) = -\sum_{s\in S} p_c(s)\log(p_c(s))
-  \label{eq-entropy}
-\end{equation}
-
-Throughout this process $useful$ tracks the set of all gene clusters which should be preserved for use by crossover.
-This set begins with all genes in separate clusters, and each time a new cluster is created it is added to $useful$.
-However, not all clusters are necessarily worth keeping. For instance, in all versions of LTGA the cluster
-containing all genes is removed from $useful$ as preserving all genes during crossover can only create clones.
-\cite{thierens:2013:ltgahiff} extended this removal to include any unsupported subsets. If the pairwise distance
-between two clusters is 0, this means there are no individuals in the population which disrupt the relationships between the two
-clusters. Therefore when performing crossover there is no reason to believe a fitness improvement can be achieved
-by breaking the stored pattern. As such a cluster is only kept if its direct superset has a non-zero distance.
-As a final step Line~\ref{fig-cluster-creation-ordering} reorders $useful$ such that clusters appear in reversed
-order from which they were added to $useful$. Thus the most linked clusters and those containing single genes
-appear at the end of the returned list.
-
-\cite{thierens:2013:ltgahiff}'s version of LTGA does not use the entire population when determining pairwise entropy.
-Instead, binary tournament is used to select half of the population. This is done to ensure the model is built
-using only high quality solutions, even during the first generation.
-
-In order to efficiently perform clustering, a pairwise gene frequency table is constructed from the selected solutions.
-To calculate Equation~\ref{eq-distance}, Equation~\ref{eq-entropy} is called for each
-gene ($H(c_i)$) and pair of genes ($H(c_i \cup c_j)$). Extracting this information requires $O(\mu N^2)$
-time, where $\mu$ is the population size and $N$ is the genome size. The process of converting
-this pairwise frequency information into clusters can be achieved in $O(N^2)$ using the bookkeeping
-methods presented by \cite{gronau:2007:upgma}. This cost is performed only once per generation,
-and is then used to perform approximately $O(\mu N)$ crossover evaluations. As a result, the amortized cost of
-LTGA's model building is $O(N)$.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Cluster-Usage}{}
-    \ForAll{$C_i \in useful$}
-      %\ForAll{$d \in shuffled(P_i)$}
-        \State $d \leftarrow rand\_choice(P)$\label{fig-cluster-usage-donate}
-        \State Copy $d$'s gene values for $C_i$ into solution
-        \If{solution was changed}
-          \If{solution's fitness decreased}
-            \State Revert changes
-          \EndIf
-          %\State \textbf{break}
-        \EndIf
-      %\EndFor
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm describing how clusters are used to perform crossover.}
-  \label{fig-cluster-usage}
-\end{figure}
-
-Figure~\ref{fig-cluster-usage} describes how the identified clusters are used by crossover to preserve
-gene linkage while still exploring the search space. Unlike more traditional crossover methods, LTGA
-crosses each individual with the entire population. Also, to produce a single offspring, multiple evaluations
-of the fitness function are performed.
-
-Each generation, each individual in the population undergoes crossover. In a single crossover event, each
-cluster of genes $C_i$ in $useful$ is applied as a crossover mask. A random donor $d$
-is chosen from the entire population (not just the model selected population), and $d$'s gene's for $C_i$ are copied
-into the working solution. If a modification is made, an evaluation is then performed. If the crossover
-resulted in no worse fitness then the changes are kept. This allows for neutral drift across plateaus.
-The resulting solution, which must be at least as fit as its parent, is then
-copied into the next generation.
-
-In total each individual can cause up to $|useful|$ evaluations. If all clusters were kept, even those deemed
-unhelpful, and all donations were evaluated, even those which did not change any genes, then \Call{Cluster-Usage}{}
-would perform exactly $2N-2$ evaluations for each of the $\mu$ solutions in the population. This provides the amortizing evaluations
-required to make clustering only $O(N)$ operations per evaluation. However, by skipping some evaluations, its
-possible that clustering may be super-linear.
-
-LTGA has no explicit form of diversity control and has no method for introducing new genetic information once
-the population has converged. Therefore an LTGA run is considered converged once two consecutive populations
-contain the same unique solutions.
-
-By design, LTGA only has a single parameter: population size. LTGA uses no mutation, and crossover is defined
-in terms of the clustering algorithm. Selection between generations is fully elitist and embedded in the crossover,
-with selection of model building solutions fixed to a binary tournament. Neither \Call{Cluster-Creation}{} nor
-\Call{Cluster-Usage}{} rely on parameter values. LTGA does not provide any method for controlling or setting
-the population size, and uses a fixed user specified size for each generation.
-
-\section{Parameter-less Population Pyramid}
-\label{sec-p3}
-\cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid (P3) as a method for
-performing optimization which does not require the user to provide any parameters. This is
-achieved by combining efficient local search with the model building methods of LTGA using
-an iteratively constructed hierarchy of populations.
-
-The high level algorithm of P3 is presented in Figure~\ref{fig-p3}. Unlike more traditional
-GAs, P3 does not follow a generational model. Instead, it maintains an iteratively
-expanding pyramid of expanding populations. Each iteration, a new random solution is generated.
-This solution is brought to a local optimum using the hill climbing algorithm in Figure~\ref{fig-hc}. If that
-local optimum has not yet been added to the pyramid, the solution is added to the lowest
-population $P_0$.
-
-Next, the solution is iteratively improved by applying LTGA's crossover algorithm (Figure~\ref{fig-cluster-usage})
-with each population $P_i$ in the pyramid. If this process results in a strict fitness improvement and has
-created a solution not yet stored in the pyramid, it is added to the next highest pyramid level $P_{i+1}$.
-If $P_{i+1}$ does not yet exist, it is created. In this way populations in the pyramid expand over time,
-and the number of populations stored increases over time. Initially the pyramid contains no solutions
-or populations, meaning the user does not need to specify a population size.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Iterate-P3}{}
-    \State Create random solution
-    \State Apply hill climber
-    \If{solution $\notin hashset$}
-      \State Add solution to $P_0$
-      \State Add solution to $hashset$
-    \EndIf
-
-    \ForAll{$P_i \in pyramid$}
-      \State Mix solution with $P_i$
-      \If{solution's fitness has improved}
-        \If{solution $\notin hashset$}
-          \State Add solution to $P_{i+1}$
-          \State Add solution to $hashset$
-        \EndIf
-      \EndIf
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-  \caption{One iteration of P3 optimization. $pyramid$ is an
-           ordered set of populations and $hashset$ is a set
-           of all solutions in $pyramid$.}
-  \label{fig-p3}
-\end{figure}
-
-To accommodate P3's unique population structure, some of LTGA's clustering procedures were modified. In LTGA,
-clusters are identified at the start of each generation and are used
-to create all offspring in that generation. As P3 does not perform serial
-generations, P3 instead updates the model each
-time a solution is added to a population. Furthermore, unlike
-LTGA all solutions in the population are used to generate the model, not just the
-winners of a binary tournament. We can do this because even the
-worst solutions in the pyramid are already high quality due to local search.
-Using local search in LTGA was examined by \cite{bosman:2011:lsbbo} and found
-to provide no significant improvement. A likely cause was that that study applied
-local search to every solution, not just the initial population, resulting in significant overhead.
-
-Beyond the changes in population structuring, P3 modifies
-LTGA's version of \Call{Cluster-Creation}{} and \Call{Cluster-Usage}{}.
-P3 changes Line~\ref{fig-cluster-creation-ordering} in Figure~\ref{fig-cluster-creation}
-from \emph{last merged first} ordering to \emph{smallest first} ordering.
-This method applies gene clusters during crossover based on how many genes
-are in each cluster\footnote{Ties are broken randomly.}, and not on how tightly linked those genes are.
-\cite{goldman:2012:ltga} found that this alternative was better at preserving
-diversity, and therefore required smaller populations, but has not yet been integrated into
-canonical LTGA.
-
-P3 also modified Line~\ref{fig-cluster-usage-donate} in Figure~\ref{fig-cluster-usage}.
-Instead of choosing a single genetic donor for each cluster, P3 iterates over the
-population in a random order until a solution in the population is found which
-has a least one gene different for that cluster of genes from the improving solution.
-This process increases the likelihood of an evaluation being performed for every cluster,
-and helps test rare gene patterns in the population.
-
-In LTGA the cost of rebuilding the model is $O(\mu N^2)$ as it must collect pairwise
-gene frequency information for all $\mu$ solutions in the population. P3 does not store
-a single population, and it does not have a fixed $\mu$ size for any population. However,
-each time a solution is added to the population, it requires $O(N^2)$ time to update
-the table of pairwise frequencies and another $O(N^2)$ time to rebuild the linkage model.
-The model is then used immediately to perform up to one evaluation for each of the up
-to $2N-2$ clusters. Just as in LTGA, if no evaluation shortcuts were made, P3
-has an amortized cost of $O(N)$ modeling cost per evaluation. While P3 does
-rebuild the model more frequently per solution in the population, it also performs
-a number of local search evaluations which are quite efficient, meaning theoretical
-comparisons of their speed are difficult to perform. As a final note, P3's repeated
-attempts to find a useful donation make it less likely than LTGA to skip evaluations,
-but has an added cost to find these donations. Repeated donations could require as much
-as $O(\mu)$ attempts per evaluation, but experimental evidence suggest that this
-operation actually saves more overhead than it costs by increasing the number of evaluations
-per model rebuild.
-
-Each of the pieces of the P3 algorithm were selected not just for their stand alone efficacy,
-but for the ways in which they interact. By using the hill climber to ensure each solution
-is at a local optima, the underlying pairwise relationships in the problem are exposed. As
-a result, detecting clusters for use by crossover is much more effective. The crossover operator
-is extremely elitist, as each gene donation must result in no fitness loss, and a solution must
-strictly improve to be added to the next level of the pyramid. This is balanced by continual
-integration of new random solutions. Furthermore, each random restart decreases the probability
-of spurious linkages caused by shared ancestry. This diversity is further preserved by applying
-gene clusters in smallest first order during crossover as this reduces the probability of genetic
-hitchhikers.
-
-Other algorithms have proposed using multiple concurrent populations.
-\cite{hornby:2006:alps} had a hierarchy of populations with solutions periodically advancing
-upward. This allows for continuous integration of diversity as the lowest population is reseeded
-with random solutions. However, this method resulted in increased parameterization as not only was
-a population size required, but also parameters for how frequently generations advanced between levels and
-how many total levels to have were added. \cite{harik:1999:parameterlessga} used multiple independent
-populations of different sizes as a method for removing the population size parameter, but
-doing was provably less efficient than using an optimal population size as no information is shared
-between the populations.
-
-\section{Problem Descriptions}
-\label{sec-problems}
-
-\subsection{Single Instance Problems}
-Understanding how a stochastic search algorithm will behave on arbitrary and complex
-search landscapes can be exceedingly difficult. Therefore a common practice for
-algorithm understanding is to perform search on well defined, well understood
-landscapes. To be of interest these landscapes need to represent interesting
-and important aspects of real world problems.
-
-Perhaps the most common such landscape is the Deceptive Trap problem \citep{goldberg:1991:gasize}.
-In this landscape the genome is broken up into $k$ bit non-overlapping subproblems referred
-to as traps.
-Each subproblem is scored using Equation~\ref{eq-deceptive-trap}, where $t$ is the number of
-bits in the trap set to 1. The global optimum in each trap is a string of all 1s, while all
-other solutions lead to a local optima of all 0s. This problem tests an algorithm's
-ability to overcome $k$ sized deception and is commonly used to determine how effective crossover
-is at preserving building blocks. Any crossover event that mixes bits from different parents in
-the same trap will likely result in that trap being optimized to the local optima. For our experiments
-we chose $k=7$ to ensure highly deceptive traps.
-
-\begin{equation}
-   trap(t) = \left\{
-     \begin{array}{rl}
-       k-1-t, &  t<k\\
-       k,   &  t = k
-     \end{array}
-   \right.
-  \label{eq-deceptive-trap}
-\end{equation}
-
-\cite{goldman:2012:ltga} found that mixing local search with linkage learning rendered the Deceptive
-Trap problem trivial. This is because local search is able to optimize each trap to one of the two
-optima (all 1s or all 0s), and then linkage learning can have perfect knowledge of gene interactions.
-In order to make the problem more difficult, the authors proposed the Deceptive
-Step Trap problem, given in Equation~\ref{eq-deceptive-step-trap}. This function modifies the results
-of Equation~\ref{eq-deceptive-trap} to include plateaus of size $s$, introducing an exponential number
-of local optima in each trap. As a result the Deceptive Step Trap is much more challenging for linkage
-learning techniques, while still being highly deceptive. We chose to use $s=2$ to create the maximum
-number of local optima.
-\begin{equation}
-   step\_trap(t) = \left \lfloor \frac{(k-s)\pmod{s} + trap(t)}{s} \right \rfloor
-  \label{eq-deceptive-step-trap}
-\end{equation}
-
-Another challenging aspect of landscapes can be higher order relationships. The Hierarchical If
-and only If (HIFF) problem is designed to capture the difficulties of this class of problem.
-In HIFF the genome is broken up into a complete binary tree, such that each gene appears in exactly one
-leaf and each internal node is the subset of genes contained in its children. If all genes represented
-in a node of the tree are set to the same value, they score equal to the size of the set. In this way
-small subsets lead toward solutions to larger subsets. However, a node can score if all genes are either
-all 1s or all 0s, meaning that to solve higher order subproblems it is necessary to perform crossovers
-that preserve lower order solutions. This problem is a natural fit for LTGA as the linkage tree can
-perfectly duplicate the problem's true relationships \citep{thierens:2013:ltgahiff}.
-
-As a final class of well known problems, we have chosen to borrow the Rastrigin problem from real valued
-optimization. This problem's landscape, determined by Equation~\ref{eq-rast}, is highly multimodal caused by
-the oscillating cosine function. \cite{goldman:2014:p3} proposed the Discretized Rastrigin problem, such
-that each floating point $x_i$ in Equation~\ref{eq-rast} is encoded using a 10 bit gray code.
-\begin{equation}
-  An + \sum_{i=1}^{n}\left [ x_i^2-A\cos (2\pi x_i) \right ] \forall x\in [-5.12,5.12]
-  \label{eq-rast}
-\end{equation}
-
-\subsection{Randomly Generated Problem Classes}
-While well defined landscapes can provide specific insights into how an algorithm works,
-their static nature can be misleading. Specifically, algorithm quality might
-be very fragile such that it is only effective at searching well behaved landscapes. A more realistic
-test of an algorithm's black box effectiveness is to work with randomly generated instances
-drawn from a problem class. When tested over a sufficiently large sample it is then possible
-to draw more general conclusions about the algorithms effectiveness. The challenge with these
-landscapes is determining the global optimum to gauge if an algorithm was successful.
-
-Perhaps the most common model for generating random rugged landscapes is the NK model. An
-NK Landscape determines the fitness of each gene based on epistatic relationships with $K$
-other genes in the genome. This fitness is specified using a randomly generated table of
-fitness values, were each possible combination of the $K+1$ genes is mapped to some floating
-point value $[0-1]$. In unrestricted NK landscapes the relationships between genes are also
-randomly chosen and as a result finding the global optimum is $NP$-Hard for $K>1$. However,
-if epistasis is set such that each gene depends on the $K$ directly following it
-in the genome the solution can be found in polynomial time~\citep{wright:2000:solvingnk}.
-These Nearest Neighbor NK landscapes are therefore ideal for search algorithm testing.
-For all of our experiments we fixed $K=5$ to ensure highly rugged landscapes.
-
-\cite{saul:1994:spinglass} presents a combinatorial benchmark problem derived from physics:
-Ising Spin Glasses. A spin glass is defined by a weighted graph of interaction terms between vertices.
-Each gene assigns a value to each vertex, with the fitness calculated by Equation~\ref{eq-ising}.
-In this equation, $E$ is the set of all edges, $e_{ij}$ is the edge weight connecting vertex $i$ to vertex $j$, and
-$x_i$ and $x_j$ are the gene values for vertex $i$ and $j$. Optimal fitness is when this sum is minimized.
-\begin{equation}
-\sum_{e_{ij} \in E} x_ie_{ij}x_j
-  \label{eq-ising}
-\end{equation}
-Similar to NK Landscapes, the general class is $NP$-Hard to optimize, but the $2D\pm J$ subset of
-Ising Spin Glasses can be polynomially solved.\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}
-In this subset the graph is restricted to be a two dimensional torus, edge weights are randomly set
-to either -1 or 1, and vertex values must be -1 or 1.
-
-As our final class of randomly generated problems we chose the Maximum Satisfiability (MAX-SAT) problem.
-Related to the more common 3-SAT problem, a MAX-SAT instance is defined by a set of three term clauses.
-Each term is a randomly chosen variable, which may also be negated. A clause scores if an and only if
-at least one term in the clause evaluates to true. In order to make MAX-SAT instances with a known global
-optimum, \cite{goldman:2014:p3} proposed constructing clauses around a fixed solution. In this way the
-signs of the terms are set to ensure the target solution satisfies the clause. To ensure each
-problem is challenging we chose a clause to variable ratio of $4.27$.
-
-\section{Comparison Algorithm Parameter Tuning}
-\label{sec-tuning}
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-hboa}
-      \end{centering}
-      \caption{Tuned Population Sizes for hBOA}
-      \label{fig-pop-hboa}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-ltga}
-      \end{centering}
-      \caption{Tuned Population Sizes for LTGA}
-      \label{fig-pop-ltga}
-    \end{subfigure}
-  \end{centering}
-  \caption{Optimal population sizes found using bisection on each size
-           for each problem.}
-  \label{fig-pop-sizes}
-\end{figure}
-
-While four of the six algorithms in our experiments do not require any user specified parameters,
-hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
-handicapped, we extensively tuned each using the bisection method \citep{sastry:2001:bisection}
-to determine the optimal population size for each problem size.
-Extended by \cite{goldman:2012:ltga}, this method iteratively
-doubles the population size until some success criteria is met and then performs bisection
-between the lowest successful and highest unsuccessful sizes. In this way the minimum population size
-which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
-of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
-above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
-do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
-As bisection can make infinitely large population sizes, any run which had not found the global
-optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.
-
-Figure~\ref{fig-pop-sizes} shows the results from performing bisection. In general hBOA
-required population sizes which were at least an order of magnitude larger than LTGA. Due to
-runtime and memory overhead, finding the optimal value for hBOA was much less tractable than LTGA
-for moderate to large problem sizes. LTGA's population size also grew significantly slower
-than hBOA's as problem size increased, especially on the two Trap problems and Rastrigin.
-Both algorithms were very ineffective on MAX-SAT, neither able to tune to problems sizes over 60 bits.
-This is likely due to the fact that some randomly generated MAX-SAT landscapes are very flat
-and highly deceptive~\citep{rana:1998:gamaxsat}.
-\section{Finding the Global Optimum}
-\label{sec-optimum}
-
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{evals-to-success}
-  \end{center}
-  \caption{Comparison of the median number of evaluations to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-evals-to-success}
-\end{figure}
-
-Figure~\ref{fig-evals-to-success} shows the median number of evaluations required by each of the six
-algorithms to find the global optimum for multiple sizes of each problem.  Each data point in
-Figure~\ref{fig-evals-to-success} represents the median of 100 runs, where unsuccessful runs
-are treated as requiring more evaluations than any successful run. If the median run was not successful
-no point is shown. The maximum problem size used for each problem was set to be the largest size we could
-feasibly determine LTGA's optimal population size. For many larger problems results are not shown
-for hBOA due to the extreme computational cost required to optimally set the population size.
-
-\subsection{Quantitative Comparison}
-Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
-The largest problem size for any problem where P3 was not the most efficient has 49 bits, with P3 achieving the best results on all 92 larger
-configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
-Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
-and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
-on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
-
-\subsection{Local Search}
-The Random Restart Hill Climber and $1+(\lambda, \lambda)$ are both relatively effective on small problem
-sizes. This is especially true for the three randomly generated problem problem classes. These problems
-may contain relatively few local optima or just be exceptionally difficult for the model based algorithms.
-On Deceptive Trap and Deceptive Step Trap using 4 or less traps, $1+(\lambda, \lambda)$ performs significantly
-better than any other algorithm. We believe this is
-because $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
-flipping entire traps.
-This ability also leads $1+(\lambda, \lambda)$
-to outperform the Random Restart Hill Climber on all problems except Nearest Neighbor NK.
-
-On larger problem sizes, the ability for local search to reach the global optimum quickly diminishes.
-Only on MAX-SAT are these optimizers competitive
-at larger tested problem sizes. However, we believe this is because the largest tested MAX-SAT
-was an order of magnitude smaller than the largest size tested for most other problems. As the problem
-size increases the number of local optima increases exponentially, which explains why Random Restart Hill
-Climbing was unable to scale.
-For larger problems it also becomes increasingly unlikely for $1+(\lambda, \lambda)$ to make the right combination
-of changes required to reach the global optimum. This
-behavior causes high variance in success rate, as evident by the occasional successes on large Deceptive Trap problems.
-
-\subsection{Model Building}
-Only techniques which explicitly built models of gene epistasis were able to solve the largest problem
-instances. On single instance problems LTGA was more effective than hBOA, with hBOA outperforming
-LTGA on Nearest Neighbor NK and Ising Spin Glasses. This may be caused by the differences in modeling method:
-unlike the single instance problems, gene epistasis in the randomly generated
-problem classes cannot be be perfectly represented with a linkage tree.
-
-Considering how different hBOA and LTGA are in performing optimization, it is somewhat surprising how similar
-their results are on HIFF.
-However, both techniques rely on populations large enough to support the diversity required to reach the global optimum
-and to model epistasis. Both techniques also only rebuild models once per generation. As the subproblems of HIFF
-are nested, it is unlikely that either technique can accurately model higher order epistasis before solving
-lower order subproblems. Therefore both methods require one generation per subproblem order.
-
-\subsection{P3}
-Unlike the other model based methods, P3 generally outperforms
-both Random Restart Hill Climber and $1+(\lambda, \lambda)$ even on small problem sizes.
-Unlike the other local search methods, P3 outperforms LTGA and hBOA even on large problem sizes.
-This implies that P3 is gaining the benefits of each, leveraging local search to solve easy
-problems and model building to solve harder ones.
-
-Furthermore, the interaction between these two optimization tools explains some of the reason P3
-outperforms each method alone.
-On Deceptive Trap P3's use of hill climbing ensures all traps are immediately optimized, allowing for perfect
-linkage detection and high quality donation.
-On HIFF local search solves all pairwise subproblems, saving P3 a generation over LTGA and hBOA.
-In comparison P3 is only a slight improvement on Deceptive Step Trap, which is less amenable to local search.
-
-\section{Fitness Over Time}
-\label{sec-overtime}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{fitness-over-time}
-  \end{center}
-  \caption{Compares the median best fitness reached during search for each of the six optimization methods.}
-  \label{fig-fitness-over-time}
-\end{figure}
-
-For some applications, finding the global optimum is less important than finding good solutions quickly. Therefore
-we examine this behavior in Figure~\ref{fig-fitness-over-time}. At regular intervals during optimization
-Figure~\ref{fig-fitness-over-time} shows the median of the best fitnesses found at that point of search across
-100 runs. For each problem we show the largest problem size for which we were able to successfully gather results
-for all six algorithms, but the trends shown are representative of all larger problem sizes. The maximum reporting interval
-is set to include the slowest P3 run to reach the global optimum.
-
-% 181 total
-% 121 has P3 best
-%   9 hc
-%  50 lambdalambda
-%  20 hboa
-%  18 phboa
-%  27 LTGA
-
-\subsection{Quantitative Comparison}
-Of 181 sample points, P3 had the highest median fitness in 121. In pairwise competition, $1+(\lambda, \lambda)$
-was the most likely to outperform P3, doing so on 50 sample points. LTGA, hBOA, and Parameter-less hBOA
-were the next best, outperforming P3 on 27, 20, and 18 sample points. Random Restart Hill Climbing almost
-never outperformed P3, doing so only 9 times.
-
-\subsection{Local Search}
-Perhaps the most striking result is the quality of $1+(\lambda, \lambda)$. Until quite far into search this
-method performs better than both LTGA and hBOA. Given sufficient evaluations $1+(\lambda, \lambda)$ also outperforms
-Random Restart Hill Climbing on all 7 problems. For brief periods in the middle of search it performs the
-best of all techniques on: Deceptive Trap; Deceptive Step Trap; HIFF; Ising Spin Glass; and MAX-SAT problems.
-$1+(\lambda, \lambda)$'s ability to efficiently incorporate gene modifications of larger
-than one bit allows it to overcome the deception and plateaus in Deceptive Trap and Deceptive Step Trap, solve medium sized subproblems
-in HIFF, flip the signs on multiple adjacent bits in Ising Spin Glass, and cross plateaus in MAX-SAT. However,
-this method is slow in reaching the global optima in many of these problems which causes it to eventually
-be overtaken by the model building techniques.
-
-\subsection{Model Building}
-Both hBOA and LTGA are marked by periods of little improvement followed by rapid improvement.
-In hBOA this is taken to the extreme, with all fitness
-improvement made at the very end of search. In both cases this is caused by model building. Before the model
-is accurate little improvement is made. Once it is accurate, fitness improves dramatically.
-
-On every problem LTGA has five distinct periods: fitness plateau, near instantaneous improvement, fitness plateau,
-and improvement to global optimum. The early period corresponds with
-initialization of the population, with the first fitness gain achieved immediately upon completing the first generation.
-When using an inaccurate model, LTGA's mixing strategy performs a sort of less effective local search.
-Subsequent generations then make only minor fitness improvements. Once the model
-becomes accurate and the probability of a crossover using high quality genetic material increases sufficiently, LTGA
-enters a second period of rapid improvement.
-
-\begin{comment}
-This behavior is easiest to understand on the two Trap problems. LTGA's first generation will push individual traps toward a
-local optima, but it requires more than a single generation to do so completely. Until that occurs, the model is likely behaving
-no better than random as there is little apparent linkage between bits. Once the model begins to identify individual traps, crossover
-can begin to increase the frequency of higher fitness trap settings. This process begins slowly as low fitness local optima are more
-likely to be chosen for donation than high fitness local optima due to their frequency in the population. The process becomes
-self-catalysing as increased frequency of the global optimum trap genes means increased likelihood of the global optimum being spread by crossover.
-\end{comment}
-
-\subsection{P3}
-The integration of hill climbing into P3 makes it strictly better than using hill climbing alone.
-Early in optimization P3 and the Random Restart Hill Climber have effectively identical quality. This is because
-P3 performs the same evaluations as the Hill Climber for the first two restarts. Once P3 begins
-performing crossover it immediately improves over the Hill Climber. In 95\% of intervals P3 had a fitness at
-least as high as Hill Climbing. As such P3 is better than a simple
-hill climber regardless of how long each technique is run and irrespective of how high quality the solution
-found has to be.
-
-Unlike the model based methods, which struggle until model accuracy improves, P3's iterative solution
-integration allows it to improve much more quickly. This behavior exists in most problems, but is easiest to understand on Deceptive Trap.
-On this problem, P3 immediately brings all traps to local optima, equaled only by the Random Restart Hill Climber in quality.
-In comparison LTGA must evaluate the entire population and perform multiple generations to reach similar quality.
-P3 is able to immediately integrate optimal versions of each
-trap into a single individual as they are found by local search, resulting in smoother fitness improvement than LTGA.
-
-\section{Computational Expenses}
-While it is common in evolutionary computation to assume the evaluation function will dominate algorithm
-complexity, in some domains this will not be true. Model based methods are especially likely to violate
-this norm. Therefore, in order to assess P3's quality in solving problems with efficient fitness functions,
-we provide data on both its algorithmic complexity and wall clock time.
-
-\subsection{Operation Counting}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{rebuilds}
-      \end{centering}
-      \caption{Rebuilds}
-      \label{fig-rebuilds}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{donations}
-      \end{centering}
-      \caption{Donations}
-      \label{fig-donations}
-    \end{subfigure}
-  \end{centering}
-  \caption{Computation costs incurred by model rebuilding (Figure~\ref{fig-rebuilds}) and
-           repeated donations (Figure~\ref{fig-donations}) per evaluation as problem size increases.}
-  \label{fig-costs}
-\end{figure}
-
-When discussing the asymptotic complexity of P3 in Section~\ref{sec-p3}, two aspects eluded precise
-analysis: how expensive is model rebuilding and how many gene donations are made. Figure~\ref{fig-costs}
-provides some insight into how often these two aspects of the algorithm are utilized.
-
-Figure~\ref{fig-rebuilds} reports in an algorithmic sense how expensive model rebuilding is for search.
-In order to calculate this value we recorded how many times search rebuilt the model during each run.
-Under the minimal assumption that an evaluation costs $\Omega(N)$ time and that model rebuilds cost $O(N^2)$,
-Figure~\ref{fig-rebuilds} shows the ratio cost of model rebuilding per evaluation. If the cost of model building
-scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically constant.
-For Nearest Neighbor NK, Ising Spin Glasses, and Rastrigin this is the case. For both Trap problems
-and HIFF there is slow growth in the ratio. The problem sizes used for MAX-SAT were not sufficient
-to accurately gauge the asymptotic behavior. Together this suggests that while the cost of building the model
-is almost linear per evaluation, it can grow slowly. However, even in the worst case (HIFF) this growth was
-no more than twice the algorithmic cost of an evaluation even using 2048 bits.
-
-When applying a crossover subset, P3 tries random donors from the population until one is found with at least
-one bit different from the improving solution. In theory this can result in up to $O(\mu)$ operations.
-Figure~\ref{fig-donations} examines the observed average number of donations per evaluation performed.
-Ising Spin Glass, HIFF, and Rastrigin all achieve effectively constant behavior here, implying repeated
-donation does not impact the asymptotic runtime of P3. Both Trap functions and Nearest Neighbor NK all
-increase in number of donations as problem size increases, potentially increasing algorithmic costs. An
-important note is that each donation may range in size from a single bit up to $N-1$, with smaller
-donations far more likely to result in repeated attempts. As such this may cause some super-linear growth
-in P3, but its unlikely to be very high.
-
-\subsection{Wall Clock Performance}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{seconds-to-success}
-  \end{center}
-  \caption{Comparison of the median number of seconds to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-seconds-to-success}
-\end{figure}
-
-To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}.
-Similar to Figure~\ref{fig-evals-to-success}, each point represents the median
-of 100 runs, with unsuccessful runs treated as slower than successful runs.
-In total these runs
-took over 6 computing years to complete. We were therefore forced to use
-diverse hardware potentially running multiple processes in parallel.
-As such these timings should be considered very skeptically as a number
-of sources may contribute to biased error in the results. However, due to
-the size of the dataset and the general regularity in the results some general conclusions
-can still be made.
-
-\subsubsection{Model Building}
-hBOA and Parameter-less hBOA perform significantly worse when using wall clock time as the unit of comparison than when using
-evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while, under reasonable
-assumptions, P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
-hBOA goes from being slightly more efficient than LTGA in terms of evaluations to significantly worse in terms of seconds.
-As P3 and LTGA require a similar asymptotic complexity per
-evaluation as the Hill Climber and $1+(\lambda, \lambda)$, no similar change in ordering occurs.
-
-\subsubsection{P3}
-Any increase in P3's algorithmic complexity is offset by its efficient use of evaluations.
-In aggregate P3 found the global optimum the fastest on 72 of 130 configurations,
-with LTGA, $1+(\lambda, \lambda)$, Hill Climber, hBOA outperforming P3 on 44, 18, 10, and 1 configurations, respectively.
-On
-Deceptive Trap, Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT P3 still finds the global
-optimum faster than all comparison techniques. LTGA is able to match P3's performance almost exactly on HIFF,
-and on Rastrigin P3 falls behind only on moderate sized problems. Deceptive Step Trap yields the worst performance,
-with LTGA using optimal population sizes having a significantly lower growth curve than P3. This mirrors
-the results from Figure~\ref{fig-evals-to-success} under the assumption that P3's computational time per evaluation is higher
-than LTGA but by no more than a constant factor.
-
-\section{Population Sizing}
-A major advantage to P3 is that it does not require the user to set a population size parameter.
-Beyond making P3 easier to apply, this also conveys two additional advantages: diversity scaled
-to initialization and no need to sacrifice intermediate fitness for eventual optimality.
-
-\subsection{Problem Instance versus Problem Class}
-Our procedure for tuning LTGA and hBOA outlined in Section~\ref{sec-tuning} involved finding
-the optimal population size for each class of problem. For real world black box optimization this is
-realistically the best either algorithm could hope for as tuning to a problem
-instance or population initialization involves repeatedly solving the problem being tuned.
-This limitation does not exist in parameter-less methods, which scale their diversity
-based on the problem instance without needing to solve that instance repeatedly.
-
-To achieve high success rates on randomly generated problem classes, LTGA and hBOA must use a population size
-large enough to solve the hardest instances in that class. Therefore these methods will have
-population sizes larger than necessary to solve the easiest instances in the class. Even
-on single instance problems, both methods will require population sizes large enough to
-ensure the worst random initialization is diverse enough to solve the problem, which may be
-much larger than the best random initialization.
-
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{evals-to-success-range}
-  \end{center}
-  \caption{Comparison of the upper and lower quartiles of evaluations required
-           to reach the global optimum for P3 and LTGA with respect to problem size.}
-  \label{fig-evals-to-success-range}
-\end{figure}
-
-Figure~\ref{fig-evals-to-success-range}
-highlights how this can effect the required number of evaluations to reach the global optimum, showing
-the upper and lower quartiles for P3 and LTGA. On each problem
-LTGA has a much smaller difference between its best and worst runs. This makes sense as LTGA uses the same
-population size regardless of instance and progresses search generationally. In contrast P3 has a much
-higher split, with many runs finishing very quickly. This added variability is also generally on the lower
-quartile, with P3's upper quartile only worse than LTGA's on Deceptive Step Trap and Nearest
-Neighbor NK. This supports the hypothesis that P3 is able to scale its diversity not just
-to the problem class, but to the problem instance or even problem initialization, something
-wholly infeasible for tuned population sizing to do.
-
-This tuning distinction is also apparent when comparing Parameter-less hBOA with hBOA in Figure~\ref{fig-evals-to-success}.
-While generally performing worse than hBOA, the difference between the two algorithms is smallest
-on randomly generated problem classes. On MAX-SAT, Parameter-less hBOA actually outperformed both hBOA and LTGA,
-likely due to its ability to scale diversity to the problem instance instead of the entire problem class.
-
-\subsection{Fast versus Optimal}
-\label{sec-fast-vs-optimal}
-
-In Section~\ref{sec-overtime} we examined intermediate fitness qualities of LTGA and hBOA when using
-population sizes tuned to reach the global optimum. As a result, both were exceptionally ineffective
-at quickly reaching high quality solutions. This is because unlike P3, these methods have an explicit
-trade off between optimal performance and intermediate performance caused by their population size parameter.
-
-At 58\% of the recording intervals hBOA has the worst fitness of any solver. Most of the exceptions
-occur when hBOA is still evaluating its initial population, allowing this random search to temporarily surpass the
-local search methods. After $N$ evaluations, however, hBOA and LTGA both fall behind until their models begin
-to improve.
-Parameter-less hBOA reaches intermediate fitnesses faster than hBOA, doing so on 62\% of intervals, as its models begin
-to optimize earlier than hBOA.  However, this trend is reversed after a sufficient number of evaluations, most clearly
-on Deceptive Step Trap and Ising Spin Glasses, as hBOA's tuned population overtakes Parameter-less hBOA's parallel populations.
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{small-pop-dst}
-      \end{centering}
-      \caption{Deceptive Step Trap 203}
-      \label{fig-small-pop-dst}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{small-pop-nk}
-      \end{centering}
-      \caption{Nearest Neighbor NK 200}
-      \label{fig-small-pop-nk}
-    \end{subfigure}
-  \end{centering}
-  \caption{Comparison of how reducing LTGA's population size effects
-           the median best fitness reached during search.}
-  \label{fig-small-pop}
-\end{figure}
-
-Figure~\ref{fig-small-pop} examines the effect of population size on LTGA's intermediate fitness by examining the effect
-of reducing LTGA's population size to one tenth of the tuned value. The two problems shown are representative of the behavior of using
-a smaller population size on the other five problems. Reducing the population size caused LTGA to improve earlier but plateau
-at lower fitnesses. This caused LTGA's success rate to drop from 100 to 0 on Deceptive Step Trap and from 98 to 68 on Nearest
-Neighbor NK.
-Even when using a reduced population size for LTGA, P3 still achieved a fitness at
-least as high as LTGA at 80\% of intervals.
-
-\begin{comment}
-It is important to note that hBOA and LTGA were tuned specifically
-to find the global optimum and are not necessarily using the optimal population size for finding intermediate fitnesses. Consider
-that for both reducing the population size would almost certainly improve their fitness for evaluation steps less than their population
-size. To examine the effect of population size we tested LTGA using one tenth of the population size required for reliably finding the
-global optimum, with the results given in Figure~\ref{fig-small-pop}.
-Reducing the population size resulted in LTGA's initial fitness improvement
-occurring earlier, as the first generation is completed much more quickly. On Deceptive Step Trap this means that
-LTGA leaps ahead of P3 due to its ability to overcome the two bit fitness plateaus.
-%
-Across all tested problems using the smaller
-population size reduced the quality of LTGA's first fitness plateau, likely due to missing required diversity to reach higher quality.
-The second period of improvement also comes earlier and reaches a lower quality fitness. On Deceptive Step Trap 0
-runs reach the global optimum, down from 100 successful for the full population size. On Nearest Neighbor NK this drops the number
-of successful runs from 98 to 68. As a result we conclude that while reducing the population size of LTGA may improve its quality
-at early points during optimization, doing so reduces the robustness of the final solution found. This is in contrast to P3
-which balances high quality intermediate fitness without trading away eventual optimality.
-\end{comment}
-
-\section{Inner Workings}
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{cross}
-      \end{centering}
-      \caption{Crossover Proportion}
-      \label{fig-cross}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{cross-success}
-      \end{centering}
-      \caption{Crossover Success}
-      \label{fig-cross-success}
-    \end{subfigure}
-  \end{centering}
-  \caption{For each problem Figure~\ref{fig-cross} shows the proportion of P3 evaluations spend on crossovers
-           and Figure~\ref{fig-cross-success} shows the percentage of fitness improving crossover evaluations.}
-\end{figure}
-
-While analysis of optimization speed is useful from a practitioner standpoint, doing so provides very
-little insight into algorithm behavior. To better understand how P3 works in detail we present here
-a look at some internal features specific to P3.
-
-\subsection{Crossover}
-Figure~\ref{fig-cross} shows the proportion of evaluations P3 spends on crossover, as opposed to hill climbing,
-and Figure~\ref{fig-cross-success} shows what percentage of crossover evaluations resulted in a fitness improvement.
-Together these figures provide some insight into the role of crossover within P3. The behavior
-for each is clearly problem dependent and generally asymptoticly stable as problem size increases.
-
-When solving problems where epistasis can be effectively detected and represented by a linkage tree, P3 tends to spend
-less evaluations performing crossover and each crossover is more likely to be successful.
-Deceptive Trap and Rastrigin are the easiest problems to model epistasis, with local search quickly
-reducing pairwise entropy in each. These are also the problems where P3 uses the least evaluations
-on crossover and has the highest success rates for crossover. At the other extreme are Nearest Neighbor
-NK and Ising Spin Glasses, which both have overlapping linkage not representable by a linkage tree.
-These problems have the highest crossover usage and lowest crossover success of any problem except Deceptive
-Step Trap.  While Deceptive Step Trap's epistasis can be accurately modeled by a linkage tree, the
-exponential number of plateaus makes detecting gene linkage very challenging.
-
-\begin{comment}
-P3 performs optimization using a mixture of hill climbing and crossover. Exactly how much time the algorithm
-dedicates to each is not clear from the algorithm description alone. As such, Figure~\ref{fig-cross} shows
-that the relationship between the two is strongly problem dependent but relatively independent of problem size.
-In general, problems which can be more accurately represented by crossover
-required less crossover evaluations. Deceptive Trap and Rastrigin spent the least evaluations on crossover, using
-15\% and 20\% respectively on the largest problem sizes. Both of these problems are perfectly separable and have relatively
-few local optima in each subproblem. HIFF and Ising Spin Glasses use crossover for about 60\% of evaluations. The HIFF structure can
-be perfectly learned by the linkage tree, but the higher order relationships require multiple stages of crossover to optimize.
-Ising Spin Glasses have an overlapping structure, making them impossible to perfectly represent in a single linkage tree, but
-that overlap is relatively minor in comparison with Nearest Neighbor NK, which uses 81\% of evaluations on crossover. Deceptive Step Trap spends the highest
-percentage of evaluations on crossover, 92\%. Unlike all of the other problems, the value taken by an individual bit in Deceptive
-Step Trap has very little meaning. This is because any trap value is either a local optima or 1 bit flip away from being a local optima.
-This has a combined result of meaning it takes very few local search evaluations to find a local optima but very many crossover operations
-to proliferate the global optima. Epistatic relationships are also going to be very weak and even small population sizes are likely
-to mark all clusters as $useful$.
-
-Solutions are only added to the next highest pyramid level if at least one crossover donation resulted in a fitness
-improvement. Figure~\ref{fig-cross-success} examines the percentage of crossover donations that resulted in a fitness
-improvement. The ordering in Figure~\ref{fig-cross-success} is an almost perfect inversion of Figure~\ref{fig-cross},
-implying that the more successful crossover is at achieving fitness improving donations the lower the proportion of evaluations P3 will
-spend performing crossover. P3 achieves an impressive 30\% success rate for crossover donations on the largest Deceptive Trap
-problem sizes. This makes sense as crossover is able to immediately and perfectly identify entire traps such that all donations
-always contain only complete traps. Furthermore, by forcing crossover to attempt donations of genes different from the current solution,
-donations containing a single trap will have a very high success rate, always improving suboptimal traps which are far more frequently
-created by local search that optimal traps. Rastrigin and HIFF have relatively high success rates of 11\% and 7\%, respectively. Again
-this is likely due to accurate modeling and relatively low diversity of building block alternatives.
-Ising Spin Glasses and Deceptive Step Trap both have low success rates near 0.01\%. This follows as both are difficult to model with
-a linkage tree. In comparison to Figure~\ref{fig-cross-success} Deceptive Step Trap performs better than expected for crossover.
-However, the reason for this is that even exceedingly uninformed crossover can make improvements to Deceptive Step Trap as
-flipping two bits in the same trap will move between two local optima. Nearest Neighbor NK had the least successful crossovers,
-with only 0.007\% successful on the largest tested problems. In comparison to Ising Spin Glass, NK's overlap is far more extensive.
-As a result it makes sense that individual crossovers are very unlikely to be fitness improvements.
-\end{comment}
-
-When compared with LTGA, P3's crossover success rates are lower but similar in problem ordering.
-Counterintuitively, the use of hill climbing on the initial population reduces
-P3's crossover success, not because it reduces model quality or the donation pool, but because it is much more challenging to improve
-locally optimal solutions than randomly generated ones. LTGA's crossover benefits from application to unoptimized solutions, which
-makes its aggregate crossover success incomparable to P3's.
-
-Even when crossover success rates are quite low, such as Nearest Neighbor NK's 0.007\% success,
-the results from Section~\ref{sec-optimum} and Section~\ref{sec-overtime} show it's still critical to optimization. Without
-crossover, P3's performance would be identical to the Random Restart Hill Climber, which was unable to solve even moderately sized problems
-and quickly fell behind P3 in intermediate fitness quality. Therefore even infrequently successful crossover donations are
-critical to success. This does, however, suggest a potential avenue for future improvement by using more successful
-modeling and donation algorithms.
-
-\subsection{Pyramid}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{level-size}
-      \end{centering}
-      \caption{Population Size}
-      \label{fig-level-size}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{level-success}
-      \end{centering}
-      \caption{Crossover Success}
-      \label{fig-level-success}
-    \end{subfigure}
-  \end{centering}
-  \caption{For each problem Figure~\ref{fig-level-size} shows the number of solutions stored in each level of the pyramid
-           and Figure~\ref{fig-level-success} shows the percentage of fitness improving crossover evaluations at each level.}
-\end{figure}
-
-Another feature unique to P3 is the shape and size of the population pyramid
-constructed for each problem. Figure~\ref{fig-level-size} shows the number of solutions stored at
-each level of the pyramid for the largest tested problem sizes. Each point is the
-median size across 100 runs. If a run did not store any solutions at a level it is treated as 0.
-No point is drawn if the median run had 0 solutions stored at that level. While pyramid size
-is effected by problem size, the overall shape is not. As such the behavior shown in Figure~\ref{fig-level-size}
-is representative of that for all other tested problem sizes.
-
-With the exception of the dip in Deceptive Step Trap, all of the pyramids show a monotonic reduction
-in size as the level increases. This is because a solution must be a strict fitness improvement
-over its previous version to be added to a higher level, which becomes less likely
-each time the solution improves. \cite{lobo:2011:dynamicpop} found theoretical evidence
-and \cite{goldman:2011:dynamic-parameters} found empirical evidence that the optimal
-population size decreases each generation of traditional evolutionary search.
-By decreasing in size, P3 implicitly stores more diversity in low levels and focuses
-search on high quality solutions at higher levels. In comparison,
-LTGA and hBOA suboptimally use a fixed population size at each generation.
-
-Figure~\ref{fig-level-success} examines how crossover success changes at different
-levels of the pyramid. At low levels, success gets progressively lower
-as solution quality increases faster than the model's ability to improve solutions.
-At higher levels modeling becomes more accurate and donations contain higher frequencies
-of high quality building blocks, resulting in increased crossover success. The
-highest level of most problems has a low crossover success rate, as solutions
-crossing with that level have already been improved by previous operations to
-the point where the only improvement would be to create the global optimum, which can
-only happen once.
-
-\subsubsection{Deceptive Step Trap}
-The number of solutions stored at the fourth level of Deceptive Step Trap is
-significantly lower than that of the third or fifth levels, breaking the decreasing
-trend of the other six problems. Figure~\ref{fig-level-success} has a similar
-aberration, with crossover success dropping to  0.0004\% before rebounding and
-following the more common trajectory. This behavior exists in all other problem
-sizes tested, with the dips occurring at exactly the same level.
-
-This behavior is rooted in the peculiar nature
-of this landscape.
-After local search, all traps in all solutions have a total
-number of 1 bits equal to \{0, 1, 3, 5, 7\} as these correspond to the local
-optima when using $k=7$ and $s=2$. Crossover is very likely to overcome the
-two bit plateaus, and as a result solutions in the second level generally do not
-contain the lowest fitness local optima (5 bits set) and the third
-level has very few traps set to the next worst local optima (3 bits set). As a result,
-solutions which reach the third level can only be improved by replacing the deceptive
-local optima (0 and 1 bits set) with the global optimum (7 bits set).
-The global optimum is very rare in the population, and with 8 ways to represent local optima
-linkage learning is inaccurate. Therefore it is very unlikely for crossover to be successful,
-meaning few solutions will be added to the fourth level. Solutions that do improve by
-definition must have a higher frequency of optimal trap settings, meaning level four's model
-will be more accurate and donations are more likely to contain optimal trap values.
-Thus the level size and crossover success rates increase after contracting around level four.
-
-\section{Conclusions and Future Work}
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing black box
-optimization. P3's primary innovation is the replacement of the generational model with a pyramid of populations.
-This pyramid is constructed iteratively, with both the number of levels and the number of solutions stored
-at each level growing as search progresses. P3 uses a model based crossover method
-which learns a linkage tree from gene epistasis. Combined with a simple hill climber, P3's design contains
-many synergistic features.
-
-Across a large number of problems and problem sizes P3 required less evaluations to reach the global optimum
-than optimally tuned state-of-the-art competitors. On single instance problems P3's improvement was by a
-constant factor, while for the three problem classes P3's improvement increased with problem size. This
-quality extends to intermediate points during evolution, with P3 generally reaching at least as high
-of fitness as the competitive techniques when using the same number of evaluations. While P3 does require
-modeling overhead, the expense of this overhead is approximately linear with respect to genome size. There
-is some evidence that even when compared on wall clock time, P3 performs at least as well as
-the best comparison techniques. All of these achievements are made without any problem specific parameter
-tuning, making P3 easier to apply to new domains than its two closets competitors in quality.
-
-P3's quality is due to a number of desirable traits. First, mixing local search with model based
-crossover lets search focus on properly mixing high quality solutions. Second, by adding diversity only
-as necessary P3 tends to use the minimal amount of random initialization, unlike other techniques which must
-overcompensate with larger population sizes on single instance problems and consider the worst instance
-when solving problem classes. Third, by heavily exploiting existing diversity before adding more P3 is able
-to reach high quality intermediate fitnesses quickly without prematurely converging. Fourth, the very
-nature of the pyramid's shape allows search to preserve a desirable proportion of diversity at
-each fitness level, similar to a generational model using a decreasing population size.
-
-There are a number of meaningful avenues for future P3 experimentation. Perhaps the most
-pressing for practitioner acceptance is to apply P3 to real world problems and compare
-its results with other black box or even problem specific heuristics. While parameter-less,
-P3 is currently limited to discrete, fixed length genomes evaluated using single objective
-fitness. These limitations can be relaxed with future work to make P3 more widely applicable.
-While asymptotically linear, P3's modeling techniques and local search methods are likely going
-to be prohibitively expensive for genome sizes in the hundreds of thousands or millions of genes,
-and the inability of the model to capture overlapping linkage may be hindering
-search efficiency. Overcoming these limitations by using a new modeling technique may allow
-the pyramid model even greater flexibility. Similarly, while P3 is able to overcome low
-order deception via linkage learning, the iterative improvement method by which crossovers
-are made may mislead search on landscapes with higher order deception.
-
-However, even without
-these improvements our results show P3 is highly efficient at finding
-global optima on black box problems without any problem specific tuning.
-
-\small
-
-\bibliographystyle{apalike}
-\bibliography{../main}
-
-\normalsize
-\appendix
-\section{hBOA Simplification}
-\label{sec-appendix}
-To measure the quality of a decision forest, hBOA applies Equation~\ref{eq-hboa}.
-To favor compact
-models, Equation~\ref{eq-hboa} is scaled by Equation~\ref{eq-hboa-stop}, which provides increased cost
-for more total leaves.
-This quality is used for two purposes: comparison of potential changes from the existing model
-and comparison of that change with the existing model. The basis for our simplication
-is to rearrange $p(B)BDe(B) < p(B')BDe(B')$ to be $\frac{p(B)}{p(B')}<\frac{BDe(B')}{BDe(B)}$.
-
-\begin{equation}
-  BDe(B) = \prod_{i=0}^{N}\prod_{l\in L_i} \frac{\Gamma(m'_i(l))}{\Gamma(m_i(l) + m'_i(l))}
-  \prod_{x_i}\frac{\Gamma(m_i(x_i, l) + m'_i(x_i,l))}{\Gamma(m'_i(x_i,l))}
-  \label{eq-hboa}
-\end{equation}
-
-\begin{equation}
-  p(B) = c2^{-0.5(\sum_i|L_i|)log_2\mu}
-  \label{eq-hboa-stop}
-\end{equation}
-
-The outermost product of Equation~\ref{eq-hboa} iterates over all trees in the forest. However, each split
-can modify only one of the trees and therefore the contribution of all others can be canceled. The middle
-product is across all leaves in the tree. Again since only one leaf can be changed, all other terms can
-be canceled. By convention hBOA uses uninformed Bayesian priors of $m'_i(l)= 2$ and $m'_i(x_i, l)=1$ for
-binary alphabets. As $\Gamma(a) = (a-1)!$ this means the top term in the middle product and the bottom
-term in the third product reduce to 1. The only remaining terms are then $m_i(l)$ and $m_i(x_i, l)$ which
-represent the number of solutions which reached leaf $l$ and the number of solutions which reached leaf $l$
-with a specific value for $x_i$, respectively.
-
-Equation~\ref{eq-hboa-stop} can also be simplified when doing comparisons. If model $B'$ has exactly one more
-leaf than model $B$ then the ratio $\frac{p(B)}{p(B')}$ simplifies to $2^{0.5 log_2\mu}$ regardless of
-total model size.
-
-The resulting simplifications create Equation~\ref{eq-hboa-final},
-where $B'$ is different from $B$ by exactly 1 split, such that $l$ was split to create $l'$ and $l''$.
-The best split is whichever maximizes its improvement over $B$, which is equal to the right side
-of the inequality. Note that these factorials can still be exceedingly large and
-therefore it is imperative that implementations avoid rounding errors and overflows.
-
-\end{document}

### 2014-10-15 Comprehensive/chapter-p3.tex
Initial commit of Comprehensive Report

--- 2014-10-15 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
@@
-\documentclass[twoside]{article}
-\usepackage{ecj,palatino,epsfig,latexsym,natbib}
-\usepackage{url}
-\usepackage{verbatim}
-\usepackage{subcaption}
-\usepackage[noend]{algpseudocode}
-
-\newcommand{\includegraphicswide}[1]
-{\includegraphics[width=.9\textwidth,height=\textheight,keepaspectratio]{#1}}
-
-\newcommand{\includegraphicsfit}[1]
-{\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}
-
-\usepackage{array}
-\newcolumntype{L}{>{\centering\arraybackslash}m{1.9cm}}
-
-%% do not add any other page- or text-size instruction here
-
-\parskip=0.00in
-
-\begin{document}
-
-% 45-character limit on header
-\ecjHeader{x}{x}{xxx-xxx}{201X}{Parameter-less Population Pyramid}{B. W. Goldman and W. F. Punch}
-\title{\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid}
-
-\author{\name{\bf B. W. Goldman} \hfill \addr{brianwgoldman@acm.org}\\ 
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-\AND
-       \name{\bf W. F. Punch} \hfill \addr{punch@msu.edu}\\
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-}
-
-\maketitle
-
-\begin{abstract}
-%Should be about 200 words
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
-evolutionary optimization without requiring any user specified parameters.
-P3's primary innovation is to replace the generational model with a pyramid of
-multiple populations that are iteratively created and expanded. In combination
-with local search and advanced crossover P3 scales to problem difficulty, exploiting
-previously learned information before adding more diversity.
-
-Across seven problems, each tested using on average 18 problem sizes, P3 outperformed
-all five advanced comparison algorithms. This improvement includes requiring less evaluations
-and less wall clock seconds to find the global optimum and better fitness when using
-the same number of evaluations. Using both algorithm analysis and comparison we find P3's
-effectiveness is due to its ability to properly maintain, add, and exploit diversity.
-
-Unlike the best comparison algorithms, P3 was able to achieve this quality without any
-problem specific tuning. Thus, unlike previous parameter-less methods, P3 does not
-sacrifice quality for applicability. Therefore we conclude that
-P3 is an efficient, general, parameter-less approach to black box
-optimization which is more effective than existing state-of-the-art techniques.
-
-\end{abstract}
-
-\begin{keywords}
-
-Genetic algorithms, 
-linkage learning,
-local search,
-parameter-less.
-
-\end{keywords}
-
-\section{Introduction}
-A primary purpose of evolutionary optimization is to efficiently find good solutions
-to challenging real world problems with minimal prior knowledge about the problem itself.
-This driving goal has created search algorithms which can escape user bias to create
-truly novel results, sometimes publishable or patentable in their own right~\citep{kannappan:2014:humies}.
-While it is not possible for any algorithm to do better than random search across all possible
-problems~\citep{Wolpert:1997:nfl}, effectiveness can be achieved by assuming the search
-landscape has structure and then biasing the algorithm toward exploiting that structure.
-
-In evolutionary optimization, and genetic algorithms~(GAs) in particular, search is often
-biased through parameters. This can be beneficial as it allows practitioners to inject their
-knowledge about the shape of the search landscape into the algorithm.
-However, the quality of solutions found, and the speed at which they are found, is strongly tied to setting these parameters
-correctly~\citep{goldberg:1991:gasize}. As such, either expert knowledge or exceedingly
-expensive parameter tuning~\citep{grefenstette:1986:optimalga} are required to leverage
-this feature to its fullest potential. Furthermore,  parameters such as population size, mutation rate, crossover
-rate, tournament size, etc. usually have no clear relationship to the problem being solved, meaning even
-domain experts may not understand how the parameters will interact with the problem or with each other.
-
-There have been periodic efforts to reduce or remove the need for parameter tuning.
-\cite{Back:1992:selfadapt} introduced self-adaptive parameters, in which parameter values
-were included in each solution's genome and themselves underwent evolution. This allowed search
-to optimize some of its own parameters resulting in a reduced the need for expert tuning.
-\cite{harik:1999:parameterlessga} were able to design an entirely parameter-less GA by
-leveraging schema theory and parallel populations. Unfortunately these methods were provably less efficient
-than directly setting the parameters to optimal values~\citep{pelikan:1999:worstparameter-less}.
-
-One area that has been very effective at reducing the number of algorithm parameters is
-model based search. \cite{pelikan:2006:hboa}'s Hierarchical Bayesian Optimization
-Algorithm~(hBOA) and \cite{thierens:2010:ltga}'s Linkage Tree Genetic Algorithm~(LTGA)
-both require only a single parameter: population size. \cite{posik:2011:parameterless}
-leveraged model building to create a fully parameter-less algorithm, but it is restricted to
-only order-k, fully decomposable, noiseless problems.
-
-Most recently \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid~(P3).
-This method uses a pyramid structure of populations to combine model based search with local search
-to achieve parameter-less optimization. Initial results suggest that unlike
-previous parameter-less methods, P3 is actually more efficient than current state-of-the-art
-parameterized search algorithms. In this work we shall: extend these results to cover more
-comparison algorithms; compare both efficiency in reaching the global optimum and intermediate
-fitnesses; algorithm complexity analysis; and provide more in depth analysis of P3 itself.
-
-\section{Comparison Optimizers}
-\label{sec-optimizers}
-
-In order to fully understand the effectiveness of P3, we compare it with five advanced algorithms which
-have related features to P3. The Random Restart Hill Climber~\citep{goldman:2014:p3} was chosen as an efficient form of repeated
-local search. As P3 combines this hill climber with crossover,
-comparing with local search alone shows the advantages of P3's overall approach.
-The $1+(\lambda, \lambda)$ algorithm is the current best theory supported simple genetic algorithm
-and its method of crossover is in some sense a macro-mutation just as in P3. hBOA and Parameter-less
-hBOA are advanced model building search techniques which are very effective at learning complex
-problem structure, designed to achieve similar goals as P3's linkage learning but using very different
-methods. Finally LTGA represents the current state-of-the-art in black box search and is the origin
-of P3's linkage learning and crossover methods.
-
-Finally, only hBOA and LTGA require any parameters, with each of these only requiring a population
-size. This makes knowing the optimal behavior of these algorithms much more tractable. All of the
-algorithms are also gene order independent, fitness scale invariant, and unbiased. This means,
-for any problem, the order in which problem variables appear in the genome can be changed
-without changing the behavior of the search. The fitness can also be manipulated in any fashion
-as long as the rank ordering of solutions is unchanged. These algorithms are also unaffected by the meaning
-assigned to each bit, such that inverting a predetermined random subset of genes before evaluation
-will not impact search efficiency.
-
-Our implementations of all of these algorithms as well as all of the population size information, raw results,
-and processing scripts
-are available from our website.\footnote{\url{https://github.com/brianwgoldman/FastEfficientP3}}
-\subsection{Random Restart Hill Climber}
-\label{sec-hill-climber}
-Perhaps the simplest black box search heuristic is stochastic local search, or hill climbing.
-This optimization technique focuses on improving a single solution until it reaches a local
-optimum. Here we use the first improvement hill climber defined by \cite{goldman:2014:p3}
-and given in Figure~\ref{fig-hc}. This algorithm works by flipping each bit in a random
-order, keeping modifications when fitness is improved, until single bit flips cannot result in
-further fitness improvements.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Hill-Climber}{}
-    \State $options \leftarrow [0 \dots N-1]$\label{fig-hc-options}
-    \State $tried \leftarrow \emptyset$
-    \While{$|tried| < |options|$}
-      \ForAll{$index \in shuffled(options)$}
-        \If{$index \notin tried$}\label{fig-hc-tried}
-          \State Flip bit $index$ in solution
-          \If{solution's fitness increased}
-            \State $tried \leftarrow \emptyset$
-          \Else
-            \State Revert change
-          \EndIf
-          \State $tried \leftarrow tried \cup \{index\}$
-        \EndIf
-      \EndFor
-    \EndWhile
-  \EndProcedure
-\end{algorithmic}
-  \caption{Hill climbing algorithm used to improve randomly generated solutions until no single
-           bit change results in a fitness improvement.}
-  \label{fig-hc}
-\end{figure}
-
-The hill climber requires an amortized cost of $O(1)$ operations per evaluation. In order to
-terminate, at least one evaluation must be performed for each of the $N$ bits in the solution.
-As such any operation that happens only once per search can be amortized over at least $N$
-evaluations, covering the initialization of $options$ on Line~\ref{fig-hc-options}.
-Line~\ref{fig-hc-tried}, which prevents wasted evaluations, can be called at most twice
-per evaluation: once to add $index$ into $tried$ and once to prevent $index$ from being unnecessarily
-evaluated again. The only way three or more calls could happen is if
-no fitness improvement was made for the entire previous iteration, which contradicts the loop invariant.
-
-Due to its nature, this hill climber cannot escape basins of attraction. Once a solution is reached
-such that none of the single bit neighbors are fitness improvements, search stops. Thus
-this algorithm requires a restart mechanism to solve multimodal problems. We have
-chosen here to naively restart search from a random solution whenever a local optima is found. This
-ensures that on all landscapes there is always a non-zero probability of search finding the global optimum.
-
-\subsection{$1+(\lambda, \lambda)$}
-\cite{doerr:2013:lambdalambda} presented the first genetic algorithm to provably show
-the advantages of performing crossover on simple landscapes. This comparatively simple
-algorithm maintains only a single individual and a self-controlled parameter $\lambda$.
-
-Each iteration, the number of bits to flip is chosen from the binomial distribution $b\sim B(N, \frac{\lambda}{N})$,
-where $N$ is the number of bits in the genome.
-Next, $\lfloor\lambda\rfloor$ offspring are produced by flipping $b$ bits. The
-best mutant then produces $\lfloor\lambda\rfloor$ offspring via uniform crossover with the original parent, such that each gene comes from the
-mutant with probability $\frac{1}{\lambda}$. In the original algorithm the best
-offspring produced by crossover then replaces the original parent if its fitness is no worse.
-The $\lambda$ parameter, which is initialized to 1, is decreased if the offspring replaced
-its parent and increased otherwise.
-
-The original formulation was designed specifically for unimodal landscapes and as such were
-not directly suitable for multimodal problems. \cite{goldman:2014:p3} extended $1+(\lambda, \lambda)$
-to include random restarts. As search stagnates, the $\lambda$ parameter increases in value. Eventually
-this results in $\lambda \ge N$ causing mutation to always flip all bits of the individual.
-As this prevents any future improvement, whenever $\lambda \ge N$ search is restarted from a random solution with $\lambda$
-reset to 1.
-
-A few other efficiency modifications were also made. If there is a tie in crossover offspring fitness,
-whichever has a larger hamming distance from the parent is retained. This encourages drifting across plateaus.
-The ``mod'' control strategy proposed by \cite{doerr:2013:lambdalambda} was not used as it conflicted with
-the random restart strategy.
-If a crossover individual is identical to either of its parents, it is not evaluated.
-If mutation produces an offspring which is better than the best crossover offspring, it is used to compare
-against the original parent.
-
-\subsection{Hierarchical Bayesian Optimization Algorithm}
-
-\cite{pelikan:2006:hboa} used statistical principles in combination with a decision tree structure
-to create the Hierarchical Bayesian Optimization Algorithm (hBOA). This method creates a model of
-epistatic relationships between genes which is then used to stochastically generate new solutions.
-Each generation a binary tournament with replacement is used to select $\mu$ solutions from
-the population. These solutions are then used to build the model, which in turn is used to generate $\mu$ new
-solutions. The new solutions are then integrated into the population using restricted tournament
-replacement.
-
-Conceptually, the model built by hBOA is trying to infer rules of the form ``Given that this
-subset of genes are set to these values, how frequently is gene $x_i$ set to value v?'' This can
-be represented using a directed acyclic decision forest, with each tree in the forest representing one gene
-in the solution. In the decision tree $T_i$, which is used to set the value of gene $x_i$,
-each internal node represents previous decisions on how to set
-some other gene $x_j$, with the children of that node representing how the decision was made. The
-leaves of each tree give the probability that $x_i$ should be set to each possible value.
-
-The forest is constructed iteratively, with each tree initially containing a single leaf
-and with each leaf storing a pointer for each selected solution. Each iteration the algorithm considers
-all possible ways of splitting an existing leaf using another gene $x_j$, such that solutions in the
-leaf are moved to the newly created leaves based on their value for $x_j$. The general goal is to
-separate the solutions such that all solutions with $x_i = 0$ move to one leaf while solutions with
-$x_j = 1$ move to the other.
-
-This goal is formalized using model scoring from Bayesian statistics. In its raw form this
-almost always creates near infinitesimal results, calculating fractions which include factorials of $\mu$
-and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{sec-appendix}, we
-derived a simplified form shown in Equation~\ref{eq-hboa-final}. If no proposed split satisfies the inequality, iteration stops.
-If multiple splits do, whichever maximizes the right side is chosen.
-
-\begin{equation}
-  2^{0.5 log_2\mu} < \frac{(m_i(l) + 1)!}{m_i(0, l)!m_i(1,l)!} \cdot
-  \frac{m_i(0, l')!m_i(1,l')!m_i(0, l'')!m_i(1,l'')!}{(m_i(l') + 1)!(m_i(l'') + 1)!}
-  \label{eq-hboa-final}
-\end{equation}
-
-Initially there are $\Theta(N^2)$ possible ways to split existing leaves, as each of the $N$ single node
-trees can be split by any of the other $N-1$ genes. Each iteration a new edge is added to the decision
-forest, meaning some of the previously tested splits cannot be used. For instance, if $T_i$, which is used
-to decide the value of $x_i$, is split using the value of $x_j$, $T_j$ can no longer be split using $x_i$.
-As a split creates two new leaves, $O(N)$ new potential splits must also be tested. Equation~\ref{eq-hboa-final}
-parses all solutions which reach a leaf to count gene frequencies, requiring $\mu$ time.
-The number of total leaves created depends heavily on the problem and $\mu$.
-However, assuming no splits are accepted or that the cost of testing all future splits is less than
-the initial $\Theta(N^2)$, constructing the model requires $\Omega(\mu N^2)$ time. Each
-model is used to generate $\mu$ solutions, leading to a cost per evaluation of $\Omega(N^2)$.
-
-To generate a solution, the value of each gene $x_i$ is set using its corresponding decision tree $T_i$. Because
-the forest is directed acyclic, there must be an ordering of $T_i$ such that before $T_i$ is executed all
-$x_j$ it uses to make decisions have already been set. As such, previous decisions made by other trees
-are used to follow each $T_i$ until a leaf is reached. The value of $x_i$ is then set based on the
-probability that other solutions reached that leaf with each value of $x_i$.
-
-To perform replacement, hBOA uses restricted tournament replacement. After each new solution is generated
-and evaluated, a set of $w$ solutions are chosen at random from the population, where $w=\min\{N, \frac{\mu}{20}\}$.
-From this set the solution which is most genetically similar to the offspring is chosen. If the offspring
-is at least as fit as the chosen solution, it replaces the chosen solution in the population. Otherwise the
-offspring is discarded. This method is designed to preserve genetic diversity as only genetically similar
-solutions must compete on fitness.
-
-hBOA is designed to work with large population sizes, resulting in a large number of
-evaluations per generation. As hBOA utilizes explicit diversity maintenance, standard methods for determining
-convergence are not considered very accurate. Therefore the authors suggest that an hBOA run should be
-terminated after performing generations equal to $N$.
-
-Like other model based techniques, hBOA has very few parameters. There is no mutation or crossover,
-and modeling does not rely on any explicit parameters. Solution selection, generation, and replacement
-are all derived from the population size, which must be set by the user.
-
-\subsection{Parameter-less hBOA}
-Using the methods first introduced by \cite{harik:1999:parameterlessga} for the Parameter-less GA,
-\cite{pelikan:2004:parameterlesshboa} created Parameter-less hBOA which automatically scales its
-population size to fit the problem. This is done by maintaining a list of concurrent populations
-using exponentially scaled population sizes.
-
-A run of Parameter-less hBOA starts with a single population of size $\mu_0$, conventionally set
-to $\mu_0=10$. After two generations are performed, a new population of size $\mu_1 = 2\mu_0$ is created
-and performs a generation. Evolution then continues with the $\mu_0$ population performing two generations
-for each one performed by $\mu_1$. Each time population $\mu_i$ performs its second generation a new population
-$\mu_{i+1}=2\mu_i$ is created, which performs generations half as often as $\mu_i$. In this way an infinite number of
-parallel population can be simulated, with each population receiving the same number of total evaluations.
-
-In all other aspects each population is identical to an hBOA population using a fixed $\mu$. No search information
-is shared among populations, and each search is independently terminated. As such Parameter-less hBOA cannot
-perform better than hBOA using the optimal population size for a given instance, as it must also spend evaluations
-on populations of different sizes. This inefficiency is bounded by a log multiple of the total number of
-evaluations~\citep{pelikan:1999:worstparameter-less}.
-
-\subsection{Linkage Tree Genetic Algorithm}
-\cite{thierens:2010:ltga} introduced the Linkage Tree Genetic Algorithm (LTGA) which automatically
-detects and exploits problem epistasis by examining pairwise gene entropy. Due to its enhanced
-ability to preserve high fitness gene subsets, LTGA was able to outperform state-of-the-art
-GAs across many benchmarks. Since its introduction, many variants of LTGA have been
-proposed~\citep{goldman:2012:ltga} so for clarity we have chosen the version
-presented by \cite{thierens:2013:ltgahiff} as our model.
-
-LTGA's effectiveness comes from its method of performing crossover. Instead of blindly
-mixing genes between parents, LTGA attempts to preserve important interrelationships
-between genes. Before performing any crossovers in a generation, LTGA first builds
-a set of hierarchical gene clusters which are then used to dictate how genes are mixed
-during crossover.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Cluster-Creation}{}
-    \State $unmerged \leftarrow \{\{0\}, \{1\}, \{2\}, \dots, \{N-1\}\}$
-    \State $useful \leftarrow unmerged$
-    \While{$|unmerged|>1$}
-      \State $C_i, C_j \leftarrow \min_{C_i,C_j \in unmerged} D(C_i, C_j)$
-      \State $unmerged \leftarrow unmerged - \{C_i, C_j\} + \{C_i \cup C_j\}$
-      \State $useful \leftarrow useful + \{C_i \cup C_j\}$
-      \If{$D(C_i, C_j) = 0$}
-        \State $useful \leftarrow useful - \{C_i, C_j\}$
-      \EndIf
-    \EndWhile
-    \State Order $useful$ based based on last merged first\label{fig-cluster-creation-ordering}
-    \State Remove largest cluster from $useful$
-
-    \Return $useful$
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm describing how LTGA creates clusters using Equation~\ref{eq-distance}
-           for a population. $unmerged$ and $useful$ are ordered sets of sets of gene loci.}
-  \label{fig-cluster-creation}
-\end{figure}
-
-Figure~\ref{fig-cluster-creation} provides the agglomerative method LTGA uses to create gene clusters.
-This algorithm creates a tree of sets using pairwise gene entropy, such that the leaves of the tree contain a single gene and
-internal nodes are the union of their children's sets. These sets are then used by crossover to specify epistatic
-relationships which should be preserved.
-The process begins by creating the set of sets $unmerged$ which tracks all top level clusters. Initially
-$unmerged$ contains single member sets for each gene. After each iteration the two sets with the minimum average pairwise
-distance (given in Equation~\ref{eq-distance}) are merged to create a single cluster. This process is repeated
-until only a single set remains in $unmerged$ which contains all of the genes in the genome.
-
-\begin{equation}
-  D(C_i,C_j) = \frac{1}{\left | C_i \right |\cdot \left |C_j \right|}\sum_{c_i \in C_i}\sum_{c_j \in C_j}
-  2 - \frac{H(c_i) + H(c_j)}{H(c_i \cup c_j)}
-  \label{eq-distance}
-\end{equation}
-\begin{equation}
-  H(c) = -\sum_{s\in S} p_c(s)\log(p_c(s))
-  \label{eq-entropy}
-\end{equation}
-
-Throughout this process $useful$ tracks the set of all gene clusters which should be preserved for use by crossover.
-This set begins with all genes in separate clusters, and each time a new cluster is created it is added to $useful$.
-However, not all clusters are necessarily worth keeping. For instance, in all versions of LTGA the cluster
-containing all genes is removed from $useful$ as preserving all genes during crossover can only create clones.
-\cite{thierens:2013:ltgahiff} extended this removal to include any unsupported subsets. If the pairwise distance
-between two clusters is 0, this means there are no individuals in the population which disrupt the relationships between the two
-clusters. Therefore when performing crossover there is no reason to believe a fitness improvement can be achieved
-by breaking the stored pattern. As such a cluster is only kept if its direct superset has a non-zero distance.
-As a final step Line~\ref{fig-cluster-creation-ordering} reorders $useful$ such that clusters appear in reversed
-order from which they were added to $useful$. Thus the most linked clusters and those containing single genes
-appear at the end of the returned list.
-
-\cite{thierens:2013:ltgahiff}'s version of LTGA does not use the entire population when determining pairwise entropy.
-Instead, binary tournament is used to select half of the population. This is done to ensure the model is built
-using only high quality solutions, even during the first generation.
-
-In order to efficiently perform clustering, a pairwise gene frequency table is constructed from the selected solutions.
-To calculate Equation~\ref{eq-distance}, Equation~\ref{eq-entropy} is called for each
-gene ($H(c_i)$) and pair of genes ($H(c_i \cup c_j)$). Extracting this information requires $O(\mu N^2)$
-time, where $\mu$ is the population size and $N$ is the genome size. The process of converting
-this pairwise frequency information into clusters can be achieved in $O(N^2)$ using the bookkeeping
-methods presented by \cite{gronau:2007:upgma}. This cost is performed only once per generation,
-and is then used to perform approximately $O(\mu N)$ crossover evaluations. As a result, the amortized cost of
-LTGA's model building is $O(N)$.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Cluster-Usage}{}
-    \ForAll{$C_i \in useful$}
-      %\ForAll{$d \in shuffled(P_i)$}
-        \State $d \leftarrow rand\_choice(P)$\label{fig-cluster-usage-donate}
-        \State Copy $d$'s gene values for $C_i$ into solution
-        \If{solution was changed}
-          \If{solution's fitness decreased}
-            \State Revert changes
-          \EndIf
-          %\State \textbf{break}
-        \EndIf
-      %\EndFor
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm describing how clusters are used to perform crossover.}
-  \label{fig-cluster-usage}
-\end{figure}
-
-Figure~\ref{fig-cluster-usage} describes how the identified clusters are used by crossover to preserve
-gene linkage while still exploring the search space. Unlike more traditional crossover methods, LTGA
-crosses each individual with the entire population. Also, to produce a single offspring, multiple evaluations
-of the fitness function are performed.
-
-Each generation, each individual in the population undergoes crossover. In a single crossover event, each
-cluster of genes $C_i$ in $useful$ is applied as a crossover mask. A random donor $d$
-is chosen from the entire population (not just the model selected population), and $d$'s gene's for $C_i$ are copied
-into the working solution. If a modification is made, an evaluation is then performed. If the crossover
-resulted in no worse fitness then the changes are kept. This allows for neutral drift across plateaus.
-The resulting solution, which must be at least as fit as its parent, is then
-copied into the next generation.
-
-In total each individual can cause up to $|useful|$ evaluations. If all clusters were kept, even those deemed
-unhelpful, and all donations were evaluated, even those which did not change any genes, then \Call{Cluster-Usage}{}
-would perform exactly $2N-2$ evaluations for each of the $\mu$ solutions in the population. This provides the amortizing evaluations
-required to make clustering only $O(N)$ operations per evaluation. However, by skipping some evaluations, its
-possible that clustering may be super-linear.
-
-LTGA has no explicit form of diversity control and has no method for introducing new genetic information once
-the population has converged. Therefore an LTGA run is considered converged once two consecutive populations
-contain the same unique solutions.
-
-By design, LTGA only has a single parameter: population size. LTGA uses no mutation, and crossover is defined
-in terms of the clustering algorithm. Selection between generations is fully elitist and embedded in the crossover,
-with selection of model building solutions fixed to a binary tournament. Neither \Call{Cluster-Creation}{} nor
-\Call{Cluster-Usage}{} rely on parameter values. LTGA does not provide any method for controlling or setting
-the population size, and uses a fixed user specified size for each generation.
-
-\section{Parameter-less Population Pyramid}
-\label{sec-p3}
+\chapter{Parameter-less Population Pyramid}
+\label{chap-p3}
 \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid (P3) as a method for
@@
 between the populations.
-
-\section{Problem Descriptions}
-\label{sec-problems}
-
-\subsection{Single Instance Problems}
-Understanding how a stochastic search algorithm will behave on arbitrary and complex
-search landscapes can be exceedingly difficult. Therefore a common practice for
-algorithm understanding is to perform search on well defined, well understood
-landscapes. To be of interest these landscapes need to represent interesting
-and important aspects of real world problems.
-
-Perhaps the most common such landscape is the Deceptive Trap problem \citep{goldberg:1991:gasize}.
-In this landscape the genome is broken up into $k$ bit non-overlapping subproblems referred
-to as traps.
-Each subproblem is scored using Equation~\ref{eq-deceptive-trap}, where $t$ is the number of
-bits in the trap set to 1. The global optimum in each trap is a string of all 1s, while all
-other solutions lead to a local optima of all 0s. This problem tests an algorithm's
-ability to overcome $k$ sized deception and is commonly used to determine how effective crossover
-is at preserving building blocks. Any crossover event that mixes bits from different parents in
-the same trap will likely result in that trap being optimized to the local optima. For our experiments
-we chose $k=7$ to ensure highly deceptive traps.
-
-\begin{equation}
-   trap(t) = \left\{
-     \begin{array}{rl}
-       k-1-t, &  t<k\\
-       k,   &  t = k
-     \end{array}
-   \right.
-  \label{eq-deceptive-trap}
-\end{equation}
-
-\cite{goldman:2012:ltga} found that mixing local search with linkage learning rendered the Deceptive
-Trap problem trivial. This is because local search is able to optimize each trap to one of the two
-optima (all 1s or all 0s), and then linkage learning can have perfect knowledge of gene interactions.
-In order to make the problem more difficult, the authors proposed the Deceptive
-Step Trap problem, given in Equation~\ref{eq-deceptive-step-trap}. This function modifies the results
-of Equation~\ref{eq-deceptive-trap} to include plateaus of size $s$, introducing an exponential number
-of local optima in each trap. As a result the Deceptive Step Trap is much more challenging for linkage
-learning techniques, while still being highly deceptive. We chose to use $s=2$ to create the maximum
-number of local optima.
-\begin{equation}
-   step\_trap(t) = \left \lfloor \frac{(k-s)\pmod{s} + trap(t)}{s} \right \rfloor
-  \label{eq-deceptive-step-trap}
-\end{equation}
-
-Another challenging aspect of landscapes can be higher order relationships. The Hierarchical If
-and only If (HIFF) problem is designed to capture the difficulties of this class of problem.
-In HIFF the genome is broken up into a complete binary tree, such that each gene appears in exactly one
-leaf and each internal node is the subset of genes contained in its children. If all genes represented
-in a node of the tree are set to the same value, they score equal to the size of the set. In this way
-small subsets lead toward solutions to larger subsets. However, a node can score if all genes are either
-all 1s or all 0s, meaning that to solve higher order subproblems it is necessary to perform crossovers
-that preserve lower order solutions. This problem is a natural fit for LTGA as the linkage tree can
-perfectly duplicate the problem's true relationships \citep{thierens:2013:ltgahiff}.
-
-As a final class of well known problems, we have chosen to borrow the Rastrigin problem from real valued
-optimization. This problem's landscape, determined by Equation~\ref{eq-rast}, is highly multimodal caused by
-the oscillating cosine function. \cite{goldman:2014:p3} proposed the Discretized Rastrigin problem, such
-that each floating point $x_i$ in Equation~\ref{eq-rast} is encoded using a 10 bit gray code.
-\begin{equation}
-  An + \sum_{i=1}^{n}\left [ x_i^2-A\cos (2\pi x_i) \right ] \forall x\in [-5.12,5.12]
-  \label{eq-rast}
-\end{equation}
-
-\subsection{Randomly Generated Problem Classes}
-While well defined landscapes can provide specific insights into how an algorithm works,
-their static nature can be misleading. Specifically, algorithm quality might
-be very fragile such that it is only effective at searching well behaved landscapes. A more realistic
-test of an algorithm's black box effectiveness is to work with randomly generated instances
-drawn from a problem class. When tested over a sufficiently large sample it is then possible
-to draw more general conclusions about the algorithms effectiveness. The challenge with these
-landscapes is determining the global optimum to gauge if an algorithm was successful.
-
-Perhaps the most common model for generating random rugged landscapes is the NK model. An
-NK Landscape determines the fitness of each gene based on epistatic relationships with $K$
-other genes in the genome. This fitness is specified using a randomly generated table of
-fitness values, were each possible combination of the $K+1$ genes is mapped to some floating
-point value $[0-1]$. In unrestricted NK landscapes the relationships between genes are also
-randomly chosen and as a result finding the global optimum is $NP$-Hard for $K>1$. However,
-if epistasis is set such that each gene depends on the $K$ directly following it
-in the genome the solution can be found in polynomial time~\citep{wright:2000:solvingnk}.
-These Nearest Neighbor NK landscapes are therefore ideal for search algorithm testing.
-For all of our experiments we fixed $K=5$ to ensure highly rugged landscapes.
-
-\cite{saul:1994:spinglass} presents a combinatorial benchmark problem derived from physics:
-Ising Spin Glasses. A spin glass is defined by a weighted graph of interaction terms between vertices.
-Each gene assigns a value to each vertex, with the fitness calculated by Equation~\ref{eq-ising}.
-In this equation, $E$ is the set of all edges, $e_{ij}$ is the edge weight connecting vertex $i$ to vertex $j$, and
-$x_i$ and $x_j$ are the gene values for vertex $i$ and $j$. Optimal fitness is when this sum is minimized.
-\begin{equation}
-\sum_{e_{ij} \in E} x_ie_{ij}x_j
-  \label{eq-ising}
-\end{equation}
-Similar to NK Landscapes, the general class is $NP$-Hard to optimize, but the $2D\pm J$ subset of
-Ising Spin Glasses can be polynomially solved.\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}
-In this subset the graph is restricted to be a two dimensional torus, edge weights are randomly set
-to either -1 or 1, and vertex values must be -1 or 1.
-
-As our final class of randomly generated problems we chose the Maximum Satisfiability (MAX-SAT) problem.
-Related to the more common 3-SAT problem, a MAX-SAT instance is defined by a set of three term clauses.
-Each term is a randomly chosen variable, which may also be negated. A clause scores if an and only if
-at least one term in the clause evaluates to true. In order to make MAX-SAT instances with a known global
-optimum, \cite{goldman:2014:p3} proposed constructing clauses around a fixed solution. In this way the
-signs of the terms are set to ensure the target solution satisfies the clause. To ensure each
-problem is challenging we chose a clause to variable ratio of $4.27$.
-
-\section{Comparison Algorithm Parameter Tuning}
-\label{sec-tuning}
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-hboa}
-      \end{centering}
-      \caption{Tuned Population Sizes for hBOA}
-      \label{fig-pop-hboa}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-ltga}
-      \end{centering}
-      \caption{Tuned Population Sizes for LTGA}
-      \label{fig-pop-ltga}
-    \end{subfigure}
-  \end{centering}
-  \caption{Optimal population sizes found using bisection on each size
-           for each problem.}
-  \label{fig-pop-sizes}
-\end{figure}
-
-While four of the six algorithms in our experiments do not require any user specified parameters,
-hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
-handicapped, we extensively tuned each using the bisection method \citep{sastry:2001:bisection}
-to determine the optimal population size for each problem size.
-Extended by \cite{goldman:2012:ltga}, this method iteratively
-doubles the population size until some success criteria is met and then performs bisection
-between the lowest successful and highest unsuccessful sizes. In this way the minimum population size
-which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
-of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
-above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
-do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
-As bisection can make infinitely large population sizes, any run which had not found the global
-optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.
-
-Figure~\ref{fig-pop-sizes} shows the results from performing bisection. In general hBOA
-required population sizes which were at least an order of magnitude larger than LTGA. Due to
-runtime and memory overhead, finding the optimal value for hBOA was much less tractable than LTGA
-for moderate to large problem sizes. LTGA's population size also grew significantly slower
-than hBOA's as problem size increased, especially on the two Trap problems and Rastrigin.
-Both algorithms were very ineffective on MAX-SAT, neither able to tune to problems sizes over 60 bits.
-This is likely due to the fact that some randomly generated MAX-SAT landscapes are very flat
-and highly deceptive~\citep{rana:1998:gamaxsat}.
-\section{Finding the Global Optimum}
-\label{sec-optimum}
-
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{evals-to-success}
-  \end{center}
-  \caption{Comparison of the median number of evaluations to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-evals-to-success}
-\end{figure}
-
-Figure~\ref{fig-evals-to-success} shows the median number of evaluations required by each of the six
-algorithms to find the global optimum for multiple sizes of each problem.  Each data point in
-Figure~\ref{fig-evals-to-success} represents the median of 100 runs, where unsuccessful runs
-are treated as requiring more evaluations than any successful run. If the median run was not successful
-no point is shown. The maximum problem size used for each problem was set to be the largest size we could
-feasibly determine LTGA's optimal population size. For many larger problems results are not shown
-for hBOA due to the extreme computational cost required to optimally set the population size.
-
-\subsection{Quantitative Comparison}
-Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
-The largest problem size for any problem where P3 was not the most efficient has 49 bits, with P3 achieving the best results on all 92 larger
-configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
-Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
-and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
-on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
-
-\subsection{Local Search}
-The Random Restart Hill Climber and $1+(\lambda, \lambda)$ are both relatively effective on small problem
-sizes. This is especially true for the three randomly generated problem problem classes. These problems
-may contain relatively few local optima or just be exceptionally difficult for the model based algorithms.
-On Deceptive Trap and Deceptive Step Trap using 4 or less traps, $1+(\lambda, \lambda)$ performs significantly
-better than any other algorithm. We believe this is
-because $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
-flipping entire traps.
-This ability also leads $1+(\lambda, \lambda)$
-to outperform the Random Restart Hill Climber on all problems except Nearest Neighbor NK.
-
-On larger problem sizes, the ability for local search to reach the global optimum quickly diminishes.
-Only on MAX-SAT are these optimizers competitive
-at larger tested problem sizes. However, we believe this is because the largest tested MAX-SAT
-was an order of magnitude smaller than the largest size tested for most other problems. As the problem
-size increases the number of local optima increases exponentially, which explains why Random Restart Hill
-Climbing was unable to scale.
-For larger problems it also becomes increasingly unlikely for $1+(\lambda, \lambda)$ to make the right combination
-of changes required to reach the global optimum. This
-behavior causes high variance in success rate, as evident by the occasional successes on large Deceptive Trap problems.
-
-\subsection{Model Building}
-Only techniques which explicitly built models of gene epistasis were able to solve the largest problem
-instances. On single instance problems LTGA was more effective than hBOA, with hBOA outperforming
-LTGA on Nearest Neighbor NK and Ising Spin Glasses. This may be caused by the differences in modeling method:
-unlike the single instance problems, gene epistasis in the randomly generated
-problem classes cannot be be perfectly represented with a linkage tree.
-
-Considering how different hBOA and LTGA are in performing optimization, it is somewhat surprising how similar
-their results are on HIFF.
-However, both techniques rely on populations large enough to support the diversity required to reach the global optimum
-and to model epistasis. Both techniques also only rebuild models once per generation. As the subproblems of HIFF
-are nested, it is unlikely that either technique can accurately model higher order epistasis before solving
-lower order subproblems. Therefore both methods require one generation per subproblem order.
-
-\subsection{P3}
-Unlike the other model based methods, P3 generally outperforms
-both Random Restart Hill Climber and $1+(\lambda, \lambda)$ even on small problem sizes.
-Unlike the other local search methods, P3 outperforms LTGA and hBOA even on large problem sizes.
-This implies that P3 is gaining the benefits of each, leveraging local search to solve easy
-problems and model building to solve harder ones.
-
-Furthermore, the interaction between these two optimization tools explains some of the reason P3
-outperforms each method alone.
-On Deceptive Trap P3's use of hill climbing ensures all traps are immediately optimized, allowing for perfect
-linkage detection and high quality donation.
-On HIFF local search solves all pairwise subproblems, saving P3 a generation over LTGA and hBOA.
-In comparison P3 is only a slight improvement on Deceptive Step Trap, which is less amenable to local search.
-
-\section{Fitness Over Time}
-\label{sec-overtime}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{fitness-over-time}
-  \end{center}
-  \caption{Compares the median best fitness reached during search for each of the six optimization methods.}
-  \label{fig-fitness-over-time}
-\end{figure}
-
-For some applications, finding the global optimum is less important than finding good solutions quickly. Therefore
-we examine this behavior in Figure~\ref{fig-fitness-over-time}. At regular intervals during optimization
-Figure~\ref{fig-fitness-over-time} shows the median of the best fitnesses found at that point of search across
-100 runs. For each problem we show the largest problem size for which we were able to successfully gather results
-for all six algorithms, but the trends shown are representative of all larger problem sizes. The maximum reporting interval
-is set to include the slowest P3 run to reach the global optimum.
-
-% 181 total
-% 121 has P3 best
-%   9 hc
-%  50 lambdalambda
-%  20 hboa
-%  18 phboa
-%  27 LTGA
-
-\subsection{Quantitative Comparison}
-Of 181 sample points, P3 had the highest median fitness in 121. In pairwise competition, $1+(\lambda, \lambda)$
-was the most likely to outperform P3, doing so on 50 sample points. LTGA, hBOA, and Parameter-less hBOA
-were the next best, outperforming P3 on 27, 20, and 18 sample points. Random Restart Hill Climbing almost
-never outperformed P3, doing so only 9 times.
-
-\subsection{Local Search}
-Perhaps the most striking result is the quality of $1+(\lambda, \lambda)$. Until quite far into search this
-method performs better than both LTGA and hBOA. Given sufficient evaluations $1+(\lambda, \lambda)$ also outperforms
-Random Restart Hill Climbing on all 7 problems. For brief periods in the middle of search it performs the
-best of all techniques on: Deceptive Trap; Deceptive Step Trap; HIFF; Ising Spin Glass; and MAX-SAT problems.
-$1+(\lambda, \lambda)$'s ability to efficiently incorporate gene modifications of larger
-than one bit allows it to overcome the deception and plateaus in Deceptive Trap and Deceptive Step Trap, solve medium sized subproblems
-in HIFF, flip the signs on multiple adjacent bits in Ising Spin Glass, and cross plateaus in MAX-SAT. However,
-this method is slow in reaching the global optima in many of these problems which causes it to eventually
-be overtaken by the model building techniques.
-
-\subsection{Model Building}
-Both hBOA and LTGA are marked by periods of little improvement followed by rapid improvement.
-In hBOA this is taken to the extreme, with all fitness
-improvement made at the very end of search. In both cases this is caused by model building. Before the model
-is accurate little improvement is made. Once it is accurate, fitness improves dramatically.
-
-On every problem LTGA has five distinct periods: fitness plateau, near instantaneous improvement, fitness plateau,
-and improvement to global optimum. The early period corresponds with
-initialization of the population, with the first fitness gain achieved immediately upon completing the first generation.
-When using an inaccurate model, LTGA's mixing strategy performs a sort of less effective local search.
-Subsequent generations then make only minor fitness improvements. Once the model
-becomes accurate and the probability of a crossover using high quality genetic material increases sufficiently, LTGA
-enters a second period of rapid improvement.
-
-\begin{comment}
-This behavior is easiest to understand on the two Trap problems. LTGA's first generation will push individual traps toward a
-local optima, but it requires more than a single generation to do so completely. Until that occurs, the model is likely behaving
-no better than random as there is little apparent linkage between bits. Once the model begins to identify individual traps, crossover
-can begin to increase the frequency of higher fitness trap settings. This process begins slowly as low fitness local optima are more
-likely to be chosen for donation than high fitness local optima due to their frequency in the population. The process becomes
-self-catalysing as increased frequency of the global optimum trap genes means increased likelihood of the global optimum being spread by crossover.
-\end{comment}
-
-\subsection{P3}
-The integration of hill climbing into P3 makes it strictly better than using hill climbing alone.
-Early in optimization P3 and the Random Restart Hill Climber have effectively identical quality. This is because
-P3 performs the same evaluations as the Hill Climber for the first two restarts. Once P3 begins
-performing crossover it immediately improves over the Hill Climber. In 95\% of intervals P3 had a fitness at
-least as high as Hill Climbing. As such P3 is better than a simple
-hill climber regardless of how long each technique is run and irrespective of how high quality the solution
-found has to be.
-
-Unlike the model based methods, which struggle until model accuracy improves, P3's iterative solution
-integration allows it to improve much more quickly. This behavior exists in most problems, but is easiest to understand on Deceptive Trap.
-On this problem, P3 immediately brings all traps to local optima, equaled only by the Random Restart Hill Climber in quality.
-In comparison LTGA must evaluate the entire population and perform multiple generations to reach similar quality.
-P3 is able to immediately integrate optimal versions of each
-trap into a single individual as they are found by local search, resulting in smoother fitness improvement than LTGA.
-
-\section{Computational Expenses}
-While it is common in evolutionary computation to assume the evaluation function will dominate algorithm
-complexity, in some domains this will not be true. Model based methods are especially likely to violate
-this norm. Therefore, in order to assess P3's quality in solving problems with efficient fitness functions,
-we provide data on both its algorithmic complexity and wall clock time.
-
-\subsection{Operation Counting}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{rebuilds}
-      \end{centering}
-      \caption{Rebuilds}
-      \label{fig-rebuilds}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{donations}
-      \end{centering}
-      \caption{Donations}
-      \label{fig-donations}
-    \end{subfigure}
-  \end{centering}
-  \caption{Computation costs incurred by model rebuilding (Figure~\ref{fig-rebuilds}) and
-           repeated donations (Figure~\ref{fig-donations}) per evaluation as problem size increases.}
-  \label{fig-costs}
-\end{figure}
-
-When discussing the asymptotic complexity of P3 in Section~\ref{sec-p3}, two aspects eluded precise
-analysis: how expensive is model rebuilding and how many gene donations are made. Figure~\ref{fig-costs}
-provides some insight into how often these two aspects of the algorithm are utilized.
-
-Figure~\ref{fig-rebuilds} reports in an algorithmic sense how expensive model rebuilding is for search.
-In order to calculate this value we recorded how many times search rebuilt the model during each run.
-Under the minimal assumption that an evaluation costs $\Omega(N)$ time and that model rebuilds cost $O(N^2)$,
-Figure~\ref{fig-rebuilds} shows the ratio cost of model rebuilding per evaluation. If the cost of model building
-scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically constant.
-For Nearest Neighbor NK, Ising Spin Glasses, and Rastrigin this is the case. For both Trap problems
-and HIFF there is slow growth in the ratio. The problem sizes used for MAX-SAT were not sufficient
-to accurately gauge the asymptotic behavior. Together this suggests that while the cost of building the model
-is almost linear per evaluation, it can grow slowly. However, even in the worst case (HIFF) this growth was
-no more than twice the algorithmic cost of an evaluation even using 2048 bits.
-
-When applying a crossover subset, P3 tries random donors from the population until one is found with at least
-one bit different from the improving solution. In theory this can result in up to $O(\mu)$ operations.
-Figure~\ref{fig-donations} examines the observed average number of donations per evaluation performed.
-Ising Spin Glass, HIFF, and Rastrigin all achieve effectively constant behavior here, implying repeated
-donation does not impact the asymptotic runtime of P3. Both Trap functions and Nearest Neighbor NK all
-increase in number of donations as problem size increases, potentially increasing algorithmic costs. An
-important note is that each donation may range in size from a single bit up to $N-1$, with smaller
-donations far more likely to result in repeated attempts. As such this may cause some super-linear growth
-in P3, but its unlikely to be very high.
-
-\subsection{Wall Clock Performance}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{seconds-to-success}
-  \end{center}
-  \caption{Comparison of the median number of seconds to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-seconds-to-success}
-\end{figure}
-
-To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}.
-Similar to Figure~\ref{fig-evals-to-success}, each point represents the median
-of 100 runs, with unsuccessful runs treated as slower than successful runs.
-In total these runs
-took over 6 computing years to complete. We were therefore forced to use
-diverse hardware potentially running multiple processes in parallel.
-As such these timings should be considered very skeptically as a number
-of sources may contribute to biased error in the results. However, due to
-the size of the dataset and the general regularity in the results some general conclusions
-can still be made.
-
-\subsubsection{Model Building}
-hBOA and Parameter-less hBOA perform significantly worse when using wall clock time as the unit of comparison than when using
-evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while, under reasonable
-assumptions, P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
-hBOA goes from being slightly more efficient than LTGA in terms of evaluations to significantly worse in terms of seconds.
-As P3 and LTGA require a similar asymptotic complexity per
-evaluation as the Hill Climber and $1+(\lambda, \lambda)$, no similar change in ordering occurs.
-
-\subsubsection{P3}
-Any increase in P3's algorithmic complexity is offset by its efficient use of evaluations.
-In aggregate P3 found the global optimum the fastest on 72 of 130 configurations,
-with LTGA, $1+(\lambda, \lambda)$, Hill Climber, hBOA outperforming P3 on 44, 18, 10, and 1 configurations, respectively.
-On
-Deceptive Trap, Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT P3 still finds the global
-optimum faster than all comparison techniques. LTGA is able to match P3's performance almost exactly on HIFF,
-and on Rastrigin P3 falls behind only on moderate sized problems. Deceptive Step Trap yields the worst performance,
-with LTGA using optimal population sizes having a significantly lower growth curve than P3. This mirrors
-the results from Figure~\ref{fig-evals-to-success} under the assumption that P3's computational time per evaluation is higher
-than LTGA but by no more than a constant factor.
-
-\section{Population Sizing}
-A major advantage to P3 is that it does not require the user to set a population size parameter.
-Beyond making P3 easier to apply, this also conveys two additional advantages: diversity scaled
-to initialization and no need to sacrifice intermediate fitness for eventual optimality.
-
-\subsection{Problem Instance versus Problem Class}
-Our procedure for tuning LTGA and hBOA outlined in Section~\ref{sec-tuning} involved finding
-the optimal population size for each class of problem. For real world black box optimization this is
-realistically the best either algorithm could hope for as tuning to a problem
-instance or population initialization involves repeatedly solving the problem being tuned.
-This limitation does not exist in parameter-less methods, which scale their diversity
-based on the problem instance without needing to solve that instance repeatedly.
-
-To achieve high success rates on randomly generated problem classes, LTGA and hBOA must use a population size
-large enough to solve the hardest instances in that class. Therefore these methods will have
-population sizes larger than necessary to solve the easiest instances in the class. Even
-on single instance problems, both methods will require population sizes large enough to
-ensure the worst random initialization is diverse enough to solve the problem, which may be
-much larger than the best random initialization.
-
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{evals-to-success-range}
-  \end{center}
-  \caption{Comparison of the upper and lower quartiles of evaluations required
-           to reach the global optimum for P3 and LTGA with respect to problem size.}
-  \label{fig-evals-to-success-range}
-\end{figure}
-
-Figure~\ref{fig-evals-to-success-range}
-highlights how this can effect the required number of evaluations to reach the global optimum, showing
-the upper and lower quartiles for P3 and LTGA. On each problem
-LTGA has a much smaller difference between its best and worst runs. This makes sense as LTGA uses the same
-population size regardless of instance and progresses search generationally. In contrast P3 has a much
-higher split, with many runs finishing very quickly. This added variability is also generally on the lower
-quartile, with P3's upper quartile only worse than LTGA's on Deceptive Step Trap and Nearest
-Neighbor NK. This supports the hypothesis that P3 is able to scale its diversity not just
-to the problem class, but to the problem instance or even problem initialization, something
-wholly infeasible for tuned population sizing to do.
-
-This tuning distinction is also apparent when comparing Parameter-less hBOA with hBOA in Figure~\ref{fig-evals-to-success}.
-While generally performing worse than hBOA, the difference between the two algorithms is smallest
-on randomly generated problem classes. On MAX-SAT, Parameter-less hBOA actually outperformed both hBOA and LTGA,
-likely due to its ability to scale diversity to the problem instance instead of the entire problem class.
-
-\subsection{Fast versus Optimal}
-\label{sec-fast-vs-optimal}
-
-In Section~\ref{sec-overtime} we examined intermediate fitness qualities of LTGA and hBOA when using
-population sizes tuned to reach the global optimum. As a result, both were exceptionally ineffective
-at quickly reaching high quality solutions. This is because unlike P3, these methods have an explicit
-trade off between optimal performance and intermediate performance caused by their population size parameter.
-
-At 58\% of the recording intervals hBOA has the worst fitness of any solver. Most of the exceptions
-occur when hBOA is still evaluating its initial population, allowing this random search to temporarily surpass the
-local search methods. After $N$ evaluations, however, hBOA and LTGA both fall behind until their models begin
-to improve.
-Parameter-less hBOA reaches intermediate fitnesses faster than hBOA, doing so on 62\% of intervals, as its models begin
-to optimize earlier than hBOA.  However, this trend is reversed after a sufficient number of evaluations, most clearly
-on Deceptive Step Trap and Ising Spin Glasses, as hBOA's tuned population overtakes Parameter-less hBOA's parallel populations.
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{small-pop-dst}
-      \end{centering}
-      \caption{Deceptive Step Trap 203}
-      \label{fig-small-pop-dst}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{small-pop-nk}
-      \end{centering}
-      \caption{Nearest Neighbor NK 200}
-      \label{fig-small-pop-nk}
-    \end{subfigure}
-  \end{centering}
-  \caption{Comparison of how reducing LTGA's population size effects
-           the median best fitness reached during search.}
-  \label{fig-small-pop}
-\end{figure}
-
-Figure~\ref{fig-small-pop} examines the effect of population size on LTGA's intermediate fitness by examining the effect
-of reducing LTGA's population size to one tenth of the tuned value. The two problems shown are representative of the behavior of using
-a smaller population size on the other five problems. Reducing the population size caused LTGA to improve earlier but plateau
-at lower fitnesses. This caused LTGA's success rate to drop from 100 to 0 on Deceptive Step Trap and from 98 to 68 on Nearest
-Neighbor NK.
-Even when using a reduced population size for LTGA, P3 still achieved a fitness at
-least as high as LTGA at 80\% of intervals.
-
-\begin{comment}
-It is important to note that hBOA and LTGA were tuned specifically
-to find the global optimum and are not necessarily using the optimal population size for finding intermediate fitnesses. Consider
-that for both reducing the population size would almost certainly improve their fitness for evaluation steps less than their population
-size. To examine the effect of population size we tested LTGA using one tenth of the population size required for reliably finding the
-global optimum, with the results given in Figure~\ref{fig-small-pop}.
-Reducing the population size resulted in LTGA's initial fitness improvement
-occurring earlier, as the first generation is completed much more quickly. On Deceptive Step Trap this means that
-LTGA leaps ahead of P3 due to its ability to overcome the two bit fitness plateaus.
-%
-Across all tested problems using the smaller
-population size reduced the quality of LTGA's first fitness plateau, likely due to missing required diversity to reach higher quality.
-The second period of improvement also comes earlier and reaches a lower quality fitness. On Deceptive Step Trap 0
-runs reach the global optimum, down from 100 successful for the full population size. On Nearest Neighbor NK this drops the number
-of successful runs from 98 to 68. As a result we conclude that while reducing the population size of LTGA may improve its quality
-at early points during optimization, doing so reduces the robustness of the final solution found. This is in contrast to P3
-which balances high quality intermediate fitness without trading away eventual optimality.
-\end{comment}
-
-\section{Inner Workings}
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{cross}
-      \end{centering}
-      \caption{Crossover Proportion}
-      \label{fig-cross}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{cross-success}
-      \end{centering}
-      \caption{Crossover Success}
-      \label{fig-cross-success}
-    \end{subfigure}
-  \end{centering}
-  \caption{For each problem Figure~\ref{fig-cross} shows the proportion of P3 evaluations spend on crossovers
-           and Figure~\ref{fig-cross-success} shows the percentage of fitness improving crossover evaluations.}
-\end{figure}
-
-While analysis of optimization speed is useful from a practitioner standpoint, doing so provides very
-little insight into algorithm behavior. To better understand how P3 works in detail we present here
-a look at some internal features specific to P3.
-
-\subsection{Crossover}
-Figure~\ref{fig-cross} shows the proportion of evaluations P3 spends on crossover, as opposed to hill climbing,
-and Figure~\ref{fig-cross-success} shows what percentage of crossover evaluations resulted in a fitness improvement.
-Together these figures provide some insight into the role of crossover within P3. The behavior
-for each is clearly problem dependent and generally asymptoticly stable as problem size increases.
-
-When solving problems where epistasis can be effectively detected and represented by a linkage tree, P3 tends to spend
-less evaluations performing crossover and each crossover is more likely to be successful.
-Deceptive Trap and Rastrigin are the easiest problems to model epistasis, with local search quickly
-reducing pairwise entropy in each. These are also the problems where P3 uses the least evaluations
-on crossover and has the highest success rates for crossover. At the other extreme are Nearest Neighbor
-NK and Ising Spin Glasses, which both have overlapping linkage not representable by a linkage tree.
-These problems have the highest crossover usage and lowest crossover success of any problem except Deceptive
-Step Trap.  While Deceptive Step Trap's epistasis can be accurately modeled by a linkage tree, the
-exponential number of plateaus makes detecting gene linkage very challenging.
-
-\begin{comment}
-P3 performs optimization using a mixture of hill climbing and crossover. Exactly how much time the algorithm
-dedicates to each is not clear from the algorithm description alone. As such, Figure~\ref{fig-cross} shows
-that the relationship between the two is strongly problem dependent but relatively independent of problem size.
-In general, problems which can be more accurately represented by crossover
-required less crossover evaluations. Deceptive Trap and Rastrigin spent the least evaluations on crossover, using
-15\% and 20\% respectively on the largest problem sizes. Both of these problems are perfectly separable and have relatively
-few local optima in each subproblem. HIFF and Ising Spin Glasses use crossover for about 60\% of evaluations. The HIFF structure can
-be perfectly learned by the linkage tree, but the higher order relationships require multiple stages of crossover to optimize.
-Ising Spin Glasses have an overlapping structure, making them impossible to perfectly represent in a single linkage tree, but
-that overlap is relatively minor in comparison with Nearest Neighbor NK, which uses 81\% of evaluations on crossover. Deceptive Step Trap spends the highest
-percentage of evaluations on crossover, 92\%. Unlike all of the other problems, the value taken by an individual bit in Deceptive
-Step Trap has very little meaning. This is because any trap value is either a local optima or 1 bit flip away from being a local optima.
-This has a combined result of meaning it takes very few local search evaluations to find a local optima but very many crossover operations
-to proliferate the global optima. Epistatic relationships are also going to be very weak and even small population sizes are likely
-to mark all clusters as $useful$.
-
-Solutions are only added to the next highest pyramid level if at least one crossover donation resulted in a fitness
-improvement. Figure~\ref{fig-cross-success} examines the percentage of crossover donations that resulted in a fitness
-improvement. The ordering in Figure~\ref{fig-cross-success} is an almost perfect inversion of Figure~\ref{fig-cross},
-implying that the more successful crossover is at achieving fitness improving donations the lower the proportion of evaluations P3 will
-spend performing crossover. P3 achieves an impressive 30\% success rate for crossover donations on the largest Deceptive Trap
-problem sizes. This makes sense as crossover is able to immediately and perfectly identify entire traps such that all donations
-always contain only complete traps. Furthermore, by forcing crossover to attempt donations of genes different from the current solution,
-donations containing a single trap will have a very high success rate, always improving suboptimal traps which are far more frequently
-created by local search that optimal traps. Rastrigin and HIFF have relatively high success rates of 11\% and 7\%, respectively. Again
-this is likely due to accurate modeling and relatively low diversity of building block alternatives.
-Ising Spin Glasses and Deceptive Step Trap both have low success rates near 0.01\%. This follows as both are difficult to model with
-a linkage tree. In comparison to Figure~\ref{fig-cross-success} Deceptive Step Trap performs better than expected for crossover.
-However, the reason for this is that even exceedingly uninformed crossover can make improvements to Deceptive Step Trap as
-flipping two bits in the same trap will move between two local optima. Nearest Neighbor NK had the least successful crossovers,
-with only 0.007\% successful on the largest tested problems. In comparison to Ising Spin Glass, NK's overlap is far more extensive.
-As a result it makes sense that individual crossovers are very unlikely to be fitness improvements.
-\end{comment}
-
-When compared with LTGA, P3's crossover success rates are lower but similar in problem ordering.
-Counterintuitively, the use of hill climbing on the initial population reduces
-P3's crossover success, not because it reduces model quality or the donation pool, but because it is much more challenging to improve
-locally optimal solutions than randomly generated ones. LTGA's crossover benefits from application to unoptimized solutions, which
-makes its aggregate crossover success incomparable to P3's.
-
-Even when crossover success rates are quite low, such as Nearest Neighbor NK's 0.007\% success,
-the results from Section~\ref{sec-optimum} and Section~\ref{sec-overtime} show it's still critical to optimization. Without
-crossover, P3's performance would be identical to the Random Restart Hill Climber, which was unable to solve even moderately sized problems
-and quickly fell behind P3 in intermediate fitness quality. Therefore even infrequently successful crossover donations are
-critical to success. This does, however, suggest a potential avenue for future improvement by using more successful
-modeling and donation algorithms.
-
-\subsection{Pyramid}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{level-size}
-      \end{centering}
-      \caption{Population Size}
-      \label{fig-level-size}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{level-success}
-      \end{centering}
-      \caption{Crossover Success}
-      \label{fig-level-success}
-    \end{subfigure}
-  \end{centering}
-  \caption{For each problem Figure~\ref{fig-level-size} shows the number of solutions stored in each level of the pyramid
-           and Figure~\ref{fig-level-success} shows the percentage of fitness improving crossover evaluations at each level.}
-\end{figure}
-
-Another feature unique to P3 is the shape and size of the population pyramid
-constructed for each problem. Figure~\ref{fig-level-size} shows the number of solutions stored at
-each level of the pyramid for the largest tested problem sizes. Each point is the
-median size across 100 runs. If a run did not store any solutions at a level it is treated as 0.
-No point is drawn if the median run had 0 solutions stored at that level. While pyramid size
-is effected by problem size, the overall shape is not. As such the behavior shown in Figure~\ref{fig-level-size}
-is representative of that for all other tested problem sizes.
-
-With the exception of the dip in Deceptive Step Trap, all of the pyramids show a monotonic reduction
-in size as the level increases. This is because a solution must be a strict fitness improvement
-over its previous version to be added to a higher level, which becomes less likely
-each time the solution improves. \cite{lobo:2011:dynamicpop} found theoretical evidence
-and \cite{goldman:2011:dynamic-parameters} found empirical evidence that the optimal
-population size decreases each generation of traditional evolutionary search.
-By decreasing in size, P3 implicitly stores more diversity in low levels and focuses
-search on high quality solutions at higher levels. In comparison,
-LTGA and hBOA suboptimally use a fixed population size at each generation.
-
-Figure~\ref{fig-level-success} examines how crossover success changes at different
-levels of the pyramid. At low levels, success gets progressively lower
-as solution quality increases faster than the model's ability to improve solutions.
-At higher levels modeling becomes more accurate and donations contain higher frequencies
-of high quality building blocks, resulting in increased crossover success. The
-highest level of most problems has a low crossover success rate, as solutions
-crossing with that level have already been improved by previous operations to
-the point where the only improvement would be to create the global optimum, which can
-only happen once.
-
-\subsubsection{Deceptive Step Trap}
-The number of solutions stored at the fourth level of Deceptive Step Trap is
-significantly lower than that of the third or fifth levels, breaking the decreasing
-trend of the other six problems. Figure~\ref{fig-level-success} has a similar
-aberration, with crossover success dropping to  0.0004\% before rebounding and
-following the more common trajectory. This behavior exists in all other problem
-sizes tested, with the dips occurring at exactly the same level.
-
-This behavior is rooted in the peculiar nature
-of this landscape.
-After local search, all traps in all solutions have a total
-number of 1 bits equal to \{0, 1, 3, 5, 7\} as these correspond to the local
-optima when using $k=7$ and $s=2$. Crossover is very likely to overcome the
-two bit plateaus, and as a result solutions in the second level generally do not
-contain the lowest fitness local optima (5 bits set) and the third
-level has very few traps set to the next worst local optima (3 bits set). As a result,
-solutions which reach the third level can only be improved by replacing the deceptive
-local optima (0 and 1 bits set) with the global optimum (7 bits set).
-The global optimum is very rare in the population, and with 8 ways to represent local optima
-linkage learning is inaccurate. Therefore it is very unlikely for crossover to be successful,
-meaning few solutions will be added to the fourth level. Solutions that do improve by
-definition must have a higher frequency of optimal trap settings, meaning level four's model
-will be more accurate and donations are more likely to contain optimal trap values.
-Thus the level size and crossover success rates increase after contracting around level four.
-
-\section{Conclusions and Future Work}
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing black box
-optimization. P3's primary innovation is the replacement of the generational model with a pyramid of populations.
-This pyramid is constructed iteratively, with both the number of levels and the number of solutions stored
-at each level growing as search progresses. P3 uses a model based crossover method
-which learns a linkage tree from gene epistasis. Combined with a simple hill climber, P3's design contains
-many synergistic features.
-
-Across a large number of problems and problem sizes P3 required less evaluations to reach the global optimum
-than optimally tuned state-of-the-art competitors. On single instance problems P3's improvement was by a
-constant factor, while for the three problem classes P3's improvement increased with problem size. This
-quality extends to intermediate points during evolution, with P3 generally reaching at least as high
-of fitness as the competitive techniques when using the same number of evaluations. While P3 does require
-modeling overhead, the expense of this overhead is approximately linear with respect to genome size. There
-is some evidence that even when compared on wall clock time, P3 performs at least as well as
-the best comparison techniques. All of these achievements are made without any problem specific parameter
-tuning, making P3 easier to apply to new domains than its two closets competitors in quality.
-
-P3's quality is due to a number of desirable traits. First, mixing local search with model based
-crossover lets search focus on properly mixing high quality solutions. Second, by adding diversity only
-as necessary P3 tends to use the minimal amount of random initialization, unlike other techniques which must
-overcompensate with larger population sizes on single instance problems and consider the worst instance
-when solving problem classes. Third, by heavily exploiting existing diversity before adding more P3 is able
-to reach high quality intermediate fitnesses quickly without prematurely converging. Fourth, the very
-nature of the pyramid's shape allows search to preserve a desirable proportion of diversity at
-each fitness level, similar to a generational model using a decreasing population size.
-
-There are a number of meaningful avenues for future P3 experimentation. Perhaps the most
-pressing for practitioner acceptance is to apply P3 to real world problems and compare
-its results with other black box or even problem specific heuristics. While parameter-less,
-P3 is currently limited to discrete, fixed length genomes evaluated using single objective
-fitness. These limitations can be relaxed with future work to make P3 more widely applicable.
-While asymptotically linear, P3's modeling techniques and local search methods are likely going
-to be prohibitively expensive for genome sizes in the hundreds of thousands or millions of genes,
-and the inability of the model to capture overlapping linkage may be hindering
-search efficiency. Overcoming these limitations by using a new modeling technique may allow
-the pyramid model even greater flexibility. Similarly, while P3 is able to overcome low
-order deception via linkage learning, the iterative improvement method by which crossovers
-are made may mislead search on landscapes with higher order deception.
-
-However, even without
-these improvements our results show P3 is highly efficient at finding
-global optima on black box problems without any problem specific tuning.
-
-\small
-
-\bibliographystyle{apalike}
-\bibliography{../main}
-
-\normalsize
-\appendix
-\section{hBOA Simplification}
-\label{sec-appendix}
-To measure the quality of a decision forest, hBOA applies Equation~\ref{eq-hboa}.
-To favor compact
-models, Equation~\ref{eq-hboa} is scaled by Equation~\ref{eq-hboa-stop}, which provides increased cost
-for more total leaves.
-This quality is used for two purposes: comparison of potential changes from the existing model
-and comparison of that change with the existing model. The basis for our simplication
-is to rearrange $p(B)BDe(B) < p(B')BDe(B')$ to be $\frac{p(B)}{p(B')}<\frac{BDe(B')}{BDe(B)}$.
-
-\begin{equation}
-  BDe(B) = \prod_{i=0}^{N}\prod_{l\in L_i} \frac{\Gamma(m'_i(l))}{\Gamma(m_i(l) + m'_i(l))}
-  \prod_{x_i}\frac{\Gamma(m_i(x_i, l) + m'_i(x_i,l))}{\Gamma(m'_i(x_i,l))}
-  \label{eq-hboa}
-\end{equation}
-
-\begin{equation}
-  p(B) = c2^{-0.5(\sum_i|L_i|)log_2\mu}
-  \label{eq-hboa-stop}
-\end{equation}
-
-The outermost product of Equation~\ref{eq-hboa} iterates over all trees in the forest. However, each split
-can modify only one of the trees and therefore the contribution of all others can be canceled. The middle
-product is across all leaves in the tree. Again since only one leaf can be changed, all other terms can
-be canceled. By convention hBOA uses uninformed Bayesian priors of $m'_i(l)= 2$ and $m'_i(x_i, l)=1$ for
-binary alphabets. As $\Gamma(a) = (a-1)!$ this means the top term in the middle product and the bottom
-term in the third product reduce to 1. The only remaining terms are then $m_i(l)$ and $m_i(x_i, l)$ which
-represent the number of solutions which reached leaf $l$ and the number of solutions which reached leaf $l$
-with a specific value for $x_i$, respectively.
-
-Equation~\ref{eq-hboa-stop} can also be simplified when doing comparisons. If model $B'$ has exactly one more
-leaf than model $B$ then the ratio $\frac{p(B)}{p(B')}$ simplifies to $2^{0.5 log_2\mu}$ regardless of
-total model size.
-
-The resulting simplifications create Equation~\ref{eq-hboa-final},
-where $B'$ is different from $B$ by exactly 1 split, such that $l$ was split to create $l'$ and $l''$.
-The best split is whichever maximizes its improvement over $B$, which is equal to the right side
-of the inequality. Note that these factorials can still be exceedingly large and
-therefore it is imperative that implementations avoid rounding errors and overflows.
-
-\end{document}

### 2014-10-15 Comprehensive/chapter-problems.tex
Initial commit of Comprehensive Report

--- 2014-10-15 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
@@
-\documentclass[twoside]{article}
-\usepackage{ecj,palatino,epsfig,latexsym,natbib}
-\usepackage{url}
-\usepackage{verbatim}
-\usepackage{subcaption}
-\usepackage[noend]{algpseudocode}
+\chapter{Problem Descriptions}
+\label{chap-problems}
+TODO Details here
 
-\newcommand{\includegraphicswide}[1]
-{\includegraphics[width=.9\textwidth,height=\textheight,keepaspectratio]{#1}}
-
-\newcommand{\includegraphicsfit}[1]
-{\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}
-
-\usepackage{array}
-\newcolumntype{L}{>{\centering\arraybackslash}m{1.9cm}}
-
-%% do not add any other page- or text-size instruction here
-
-\parskip=0.00in
-
-\begin{document}
-
-% 45-character limit on header
-\ecjHeader{x}{x}{xxx-xxx}{201X}{Parameter-less Population Pyramid}{B. W. Goldman and W. F. Punch}
-\title{\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid}
-
-\author{\name{\bf B. W. Goldman} \hfill \addr{brianwgoldman@acm.org}\\ 
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-\AND
-       \name{\bf W. F. Punch} \hfill \addr{punch@msu.edu}\\
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-}
-
-\maketitle
-
-\begin{abstract}
-%Should be about 200 words
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
-evolutionary optimization without requiring any user specified parameters.
-P3's primary innovation is to replace the generational model with a pyramid of
-multiple populations that are iteratively created and expanded. In combination
-with local search and advanced crossover P3 scales to problem difficulty, exploiting
-previously learned information before adding more diversity.
-
-Across seven problems, each tested using on average 18 problem sizes, P3 outperformed
-all five advanced comparison algorithms. This improvement includes requiring less evaluations
-and less wall clock seconds to find the global optimum and better fitness when using
-the same number of evaluations. Using both algorithm analysis and comparison we find P3's
-effectiveness is due to its ability to properly maintain, add, and exploit diversity.
-
-Unlike the best comparison algorithms, P3 was able to achieve this quality without any
-problem specific tuning. Thus, unlike previous parameter-less methods, P3 does not
-sacrifice quality for applicability. Therefore we conclude that
-P3 is an efficient, general, parameter-less approach to black box
-optimization which is more effective than existing state-of-the-art techniques.
-
-\end{abstract}
-
-\begin{keywords}
-
-Genetic algorithms, 
-linkage learning,
-local search,
-parameter-less.
-
-\end{keywords}
-
-\section{Introduction}
-A primary purpose of evolutionary optimization is to efficiently find good solutions
-to challenging real world problems with minimal prior knowledge about the problem itself.
-This driving goal has created search algorithms which can escape user bias to create
-truly novel results, sometimes publishable or patentable in their own right~\citep{kannappan:2014:humies}.
-While it is not possible for any algorithm to do better than random search across all possible
-problems~\citep{Wolpert:1997:nfl}, effectiveness can be achieved by assuming the search
-landscape has structure and then biasing the algorithm toward exploiting that structure.
-
-In evolutionary optimization, and genetic algorithms~(GAs) in particular, search is often
-biased through parameters. This can be beneficial as it allows practitioners to inject their
-knowledge about the shape of the search landscape into the algorithm.
-However, the quality of solutions found, and the speed at which they are found, is strongly tied to setting these parameters
-correctly~\citep{goldberg:1991:gasize}. As such, either expert knowledge or exceedingly
-expensive parameter tuning~\citep{grefenstette:1986:optimalga} are required to leverage
-this feature to its fullest potential. Furthermore,  parameters such as population size, mutation rate, crossover
-rate, tournament size, etc. usually have no clear relationship to the problem being solved, meaning even
-domain experts may not understand how the parameters will interact with the problem or with each other.
-
-There have been periodic efforts to reduce or remove the need for parameter tuning.
-\cite{Back:1992:selfadapt} introduced self-adaptive parameters, in which parameter values
-were included in each solution's genome and themselves underwent evolution. This allowed search
-to optimize some of its own parameters resulting in a reduced the need for expert tuning.
-\cite{harik:1999:parameterlessga} were able to design an entirely parameter-less GA by
-leveraging schema theory and parallel populations. Unfortunately these methods were provably less efficient
-than directly setting the parameters to optimal values~\citep{pelikan:1999:worstparameter-less}.
-
-One area that has been very effective at reducing the number of algorithm parameters is
-model based search. \cite{pelikan:2006:hboa}'s Hierarchical Bayesian Optimization
-Algorithm~(hBOA) and \cite{thierens:2010:ltga}'s Linkage Tree Genetic Algorithm~(LTGA)
-both require only a single parameter: population size. \cite{posik:2011:parameterless}
-leveraged model building to create a fully parameter-less algorithm, but it is restricted to
-only order-k, fully decomposable, noiseless problems.
-
-Most recently \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid~(P3).
-This method uses a pyramid structure of populations to combine model based search with local search
-to achieve parameter-less optimization. Initial results suggest that unlike
-previous parameter-less methods, P3 is actually more efficient than current state-of-the-art
-parameterized search algorithms. In this work we shall: extend these results to cover more
-comparison algorithms; compare both efficiency in reaching the global optimum and intermediate
-fitnesses; algorithm complexity analysis; and provide more in depth analysis of P3 itself.
-
-\section{Comparison Optimizers}
-\label{sec-optimizers}
-
-In order to fully understand the effectiveness of P3, we compare it with five advanced algorithms which
-have related features to P3. The Random Restart Hill Climber~\citep{goldman:2014:p3} was chosen as an efficient form of repeated
-local search. As P3 combines this hill climber with crossover,
-comparing with local search alone shows the advantages of P3's overall approach.
-The $1+(\lambda, \lambda)$ algorithm is the current best theory supported simple genetic algorithm
-and its method of crossover is in some sense a macro-mutation just as in P3. hBOA and Parameter-less
-hBOA are advanced model building search techniques which are very effective at learning complex
-problem structure, designed to achieve similar goals as P3's linkage learning but using very different
-methods. Finally LTGA represents the current state-of-the-art in black box search and is the origin
-of P3's linkage learning and crossover methods.
-
-Finally, only hBOA and LTGA require any parameters, with each of these only requiring a population
-size. This makes knowing the optimal behavior of these algorithms much more tractable. All of the
-algorithms are also gene order independent, fitness scale invariant, and unbiased. This means,
-for any problem, the order in which problem variables appear in the genome can be changed
-without changing the behavior of the search. The fitness can also be manipulated in any fashion
-as long as the rank ordering of solutions is unchanged. These algorithms are also unaffected by the meaning
-assigned to each bit, such that inverting a predetermined random subset of genes before evaluation
-will not impact search efficiency.
-
-Our implementations of all of these algorithms as well as all of the population size information, raw results,
-and processing scripts
-are available from our website.\footnote{\url{https://github.com/brianwgoldman/FastEfficientP3}}
-\subsection{Random Restart Hill Climber}
-\label{sec-hill-climber}
-Perhaps the simplest black box search heuristic is stochastic local search, or hill climbing.
-This optimization technique focuses on improving a single solution until it reaches a local
-optimum. Here we use the first improvement hill climber defined by \cite{goldman:2014:p3}
-and given in Figure~\ref{fig-hc}. This algorithm works by flipping each bit in a random
-order, keeping modifications when fitness is improved, until single bit flips cannot result in
-further fitness improvements.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Hill-Climber}{}
-    \State $options \leftarrow [0 \dots N-1]$\label{fig-hc-options}
-    \State $tried \leftarrow \emptyset$
-    \While{$|tried| < |options|$}
-      \ForAll{$index \in shuffled(options)$}
-        \If{$index \notin tried$}\label{fig-hc-tried}
-          \State Flip bit $index$ in solution
-          \If{solution's fitness increased}
-            \State $tried \leftarrow \emptyset$
-          \Else
-            \State Revert change
-          \EndIf
-          \State $tried \leftarrow tried \cup \{index\}$
-        \EndIf
-      \EndFor
-    \EndWhile
-  \EndProcedure
-\end{algorithmic}
-  \caption{Hill climbing algorithm used to improve randomly generated solutions until no single
-           bit change results in a fitness improvement.}
-  \label{fig-hc}
-\end{figure}
-
-The hill climber requires an amortized cost of $O(1)$ operations per evaluation. In order to
-terminate, at least one evaluation must be performed for each of the $N$ bits in the solution.
-As such any operation that happens only once per search can be amortized over at least $N$
-evaluations, covering the initialization of $options$ on Line~\ref{fig-hc-options}.
-Line~\ref{fig-hc-tried}, which prevents wasted evaluations, can be called at most twice
-per evaluation: once to add $index$ into $tried$ and once to prevent $index$ from being unnecessarily
-evaluated again. The only way three or more calls could happen is if
-no fitness improvement was made for the entire previous iteration, which contradicts the loop invariant.
-
-Due to its nature, this hill climber cannot escape basins of attraction. Once a solution is reached
-such that none of the single bit neighbors are fitness improvements, search stops. Thus
-this algorithm requires a restart mechanism to solve multimodal problems. We have
-chosen here to naively restart search from a random solution whenever a local optima is found. This
-ensures that on all landscapes there is always a non-zero probability of search finding the global optimum.
-
-\subsection{$1+(\lambda, \lambda)$}
-\cite{doerr:2013:lambdalambda} presented the first genetic algorithm to provably show
-the advantages of performing crossover on simple landscapes. This comparatively simple
-algorithm maintains only a single individual and a self-controlled parameter $\lambda$.
-
-Each iteration, the number of bits to flip is chosen from the binomial distribution $b\sim B(N, \frac{\lambda}{N})$,
-where $N$ is the number of bits in the genome.
-Next, $\lfloor\lambda\rfloor$ offspring are produced by flipping $b$ bits. The
-best mutant then produces $\lfloor\lambda\rfloor$ offspring via uniform crossover with the original parent, such that each gene comes from the
-mutant with probability $\frac{1}{\lambda}$. In the original algorithm the best
-offspring produced by crossover then replaces the original parent if its fitness is no worse.
-The $\lambda$ parameter, which is initialized to 1, is decreased if the offspring replaced
-its parent and increased otherwise.
-
-The original formulation was designed specifically for unimodal landscapes and as such were
-not directly suitable for multimodal problems. \cite{goldman:2014:p3} extended $1+(\lambda, \lambda)$
-to include random restarts. As search stagnates, the $\lambda$ parameter increases in value. Eventually
-this results in $\lambda \ge N$ causing mutation to always flip all bits of the individual.
-As this prevents any future improvement, whenever $\lambda \ge N$ search is restarted from a random solution with $\lambda$
-reset to 1.
-
-A few other efficiency modifications were also made. If there is a tie in crossover offspring fitness,
-whichever has a larger hamming distance from the parent is retained. This encourages drifting across plateaus.
-The ``mod'' control strategy proposed by \cite{doerr:2013:lambdalambda} was not used as it conflicted with
-the random restart strategy.
-If a crossover individual is identical to either of its parents, it is not evaluated.
-If mutation produces an offspring which is better than the best crossover offspring, it is used to compare
-against the original parent.
-
-\subsection{Hierarchical Bayesian Optimization Algorithm}
-
-\cite{pelikan:2006:hboa} used statistical principles in combination with a decision tree structure
-to create the Hierarchical Bayesian Optimization Algorithm (hBOA). This method creates a model of
-epistatic relationships between genes which is then used to stochastically generate new solutions.
-Each generation a binary tournament with replacement is used to select $\mu$ solutions from
-the population. These solutions are then used to build the model, which in turn is used to generate $\mu$ new
-solutions. The new solutions are then integrated into the population using restricted tournament
-replacement.
-
-Conceptually, the model built by hBOA is trying to infer rules of the form ``Given that this
-subset of genes are set to these values, how frequently is gene $x_i$ set to value v?'' This can
-be represented using a directed acyclic decision forest, with each tree in the forest representing one gene
-in the solution. In the decision tree $T_i$, which is used to set the value of gene $x_i$,
-each internal node represents previous decisions on how to set
-some other gene $x_j$, with the children of that node representing how the decision was made. The
-leaves of each tree give the probability that $x_i$ should be set to each possible value.
-
-The forest is constructed iteratively, with each tree initially containing a single leaf
-and with each leaf storing a pointer for each selected solution. Each iteration the algorithm considers
-all possible ways of splitting an existing leaf using another gene $x_j$, such that solutions in the
-leaf are moved to the newly created leaves based on their value for $x_j$. The general goal is to
-separate the solutions such that all solutions with $x_i = 0$ move to one leaf while solutions with
-$x_j = 1$ move to the other.
-
-This goal is formalized using model scoring from Bayesian statistics. In its raw form this
-almost always creates near infinitesimal results, calculating fractions which include factorials of $\mu$
-and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{sec-appendix}, we
-derived a simplified form shown in Equation~\ref{eq-hboa-final}. If no proposed split satisfies the inequality, iteration stops.
-If multiple splits do, whichever maximizes the right side is chosen.
-
-\begin{equation}
-  2^{0.5 log_2\mu} < \frac{(m_i(l) + 1)!}{m_i(0, l)!m_i(1,l)!} \cdot
-  \frac{m_i(0, l')!m_i(1,l')!m_i(0, l'')!m_i(1,l'')!}{(m_i(l') + 1)!(m_i(l'') + 1)!}
-  \label{eq-hboa-final}
-\end{equation}
-
-Initially there are $\Theta(N^2)$ possible ways to split existing leaves, as each of the $N$ single node
-trees can be split by any of the other $N-1$ genes. Each iteration a new edge is added to the decision
-forest, meaning some of the previously tested splits cannot be used. For instance, if $T_i$, which is used
-to decide the value of $x_i$, is split using the value of $x_j$, $T_j$ can no longer be split using $x_i$.
-As a split creates two new leaves, $O(N)$ new potential splits must also be tested. Equation~\ref{eq-hboa-final}
-parses all solutions which reach a leaf to count gene frequencies, requiring $\mu$ time.
-The number of total leaves created depends heavily on the problem and $\mu$.
-However, assuming no splits are accepted or that the cost of testing all future splits is less than
-the initial $\Theta(N^2)$, constructing the model requires $\Omega(\mu N^2)$ time. Each
-model is used to generate $\mu$ solutions, leading to a cost per evaluation of $\Omega(N^2)$.
-
-To generate a solution, the value of each gene $x_i$ is set using its corresponding decision tree $T_i$. Because
-the forest is directed acyclic, there must be an ordering of $T_i$ such that before $T_i$ is executed all
-$x_j$ it uses to make decisions have already been set. As such, previous decisions made by other trees
-are used to follow each $T_i$ until a leaf is reached. The value of $x_i$ is then set based on the
-probability that other solutions reached that leaf with each value of $x_i$.
-
-To perform replacement, hBOA uses restricted tournament replacement. After each new solution is generated
-and evaluated, a set of $w$ solutions are chosen at random from the population, where $w=\min\{N, \frac{\mu}{20}\}$.
-From this set the solution which is most genetically similar to the offspring is chosen. If the offspring
-is at least as fit as the chosen solution, it replaces the chosen solution in the population. Otherwise the
-offspring is discarded. This method is designed to preserve genetic diversity as only genetically similar
-solutions must compete on fitness.
-
-hBOA is designed to work with large population sizes, resulting in a large number of
-evaluations per generation. As hBOA utilizes explicit diversity maintenance, standard methods for determining
-convergence are not considered very accurate. Therefore the authors suggest that an hBOA run should be
-terminated after performing generations equal to $N$.
-
-Like other model based techniques, hBOA has very few parameters. There is no mutation or crossover,
-and modeling does not rely on any explicit parameters. Solution selection, generation, and replacement
-are all derived from the population size, which must be set by the user.
-
-\subsection{Parameter-less hBOA}
-Using the methods first introduced by \cite{harik:1999:parameterlessga} for the Parameter-less GA,
-\cite{pelikan:2004:parameterlesshboa} created Parameter-less hBOA which automatically scales its
-population size to fit the problem. This is done by maintaining a list of concurrent populations
-using exponentially scaled population sizes.
-
-A run of Parameter-less hBOA starts with a single population of size $\mu_0$, conventionally set
-to $\mu_0=10$. After two generations are performed, a new population of size $\mu_1 = 2\mu_0$ is created
-and performs a generation. Evolution then continues with the $\mu_0$ population performing two generations
-for each one performed by $\mu_1$. Each time population $\mu_i$ performs its second generation a new population
-$\mu_{i+1}=2\mu_i$ is created, which performs generations half as often as $\mu_i$. In this way an infinite number of
-parallel population can be simulated, with each population receiving the same number of total evaluations.
-
-In all other aspects each population is identical to an hBOA population using a fixed $\mu$. No search information
-is shared among populations, and each search is independently terminated. As such Parameter-less hBOA cannot
-perform better than hBOA using the optimal population size for a given instance, as it must also spend evaluations
-on populations of different sizes. This inefficiency is bounded by a log multiple of the total number of
-evaluations~\citep{pelikan:1999:worstparameter-less}.
-
-\subsection{Linkage Tree Genetic Algorithm}
-\cite{thierens:2010:ltga} introduced the Linkage Tree Genetic Algorithm (LTGA) which automatically
-detects and exploits problem epistasis by examining pairwise gene entropy. Due to its enhanced
-ability to preserve high fitness gene subsets, LTGA was able to outperform state-of-the-art
-GAs across many benchmarks. Since its introduction, many variants of LTGA have been
-proposed~\citep{goldman:2012:ltga} so for clarity we have chosen the version
-presented by \cite{thierens:2013:ltgahiff} as our model.
-
-LTGA's effectiveness comes from its method of performing crossover. Instead of blindly
-mixing genes between parents, LTGA attempts to preserve important interrelationships
-between genes. Before performing any crossovers in a generation, LTGA first builds
-a set of hierarchical gene clusters which are then used to dictate how genes are mixed
-during crossover.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Cluster-Creation}{}
-    \State $unmerged \leftarrow \{\{0\}, \{1\}, \{2\}, \dots, \{N-1\}\}$
-    \State $useful \leftarrow unmerged$
-    \While{$|unmerged|>1$}
-      \State $C_i, C_j \leftarrow \min_{C_i,C_j \in unmerged} D(C_i, C_j)$
-      \State $unmerged \leftarrow unmerged - \{C_i, C_j\} + \{C_i \cup C_j\}$
-      \State $useful \leftarrow useful + \{C_i \cup C_j\}$
-      \If{$D(C_i, C_j) = 0$}
-        \State $useful \leftarrow useful - \{C_i, C_j\}$
-      \EndIf
-    \EndWhile
-    \State Order $useful$ based based on last merged first\label{fig-cluster-creation-ordering}
-    \State Remove largest cluster from $useful$
-
-    \Return $useful$
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm describing how LTGA creates clusters using Equation~\ref{eq-distance}
-           for a population. $unmerged$ and $useful$ are ordered sets of sets of gene loci.}
-  \label{fig-cluster-creation}
-\end{figure}
-
-Figure~\ref{fig-cluster-creation} provides the agglomerative method LTGA uses to create gene clusters.
-This algorithm creates a tree of sets using pairwise gene entropy, such that the leaves of the tree contain a single gene and
-internal nodes are the union of their children's sets. These sets are then used by crossover to specify epistatic
-relationships which should be preserved.
-The process begins by creating the set of sets $unmerged$ which tracks all top level clusters. Initially
-$unmerged$ contains single member sets for each gene. After each iteration the two sets with the minimum average pairwise
-distance (given in Equation~\ref{eq-distance}) are merged to create a single cluster. This process is repeated
-until only a single set remains in $unmerged$ which contains all of the genes in the genome.
-
-\begin{equation}
-  D(C_i,C_j) = \frac{1}{\left | C_i \right |\cdot \left |C_j \right|}\sum_{c_i \in C_i}\sum_{c_j \in C_j}
-  2 - \frac{H(c_i) + H(c_j)}{H(c_i \cup c_j)}
-  \label{eq-distance}
-\end{equation}
-\begin{equation}
-  H(c) = -\sum_{s\in S} p_c(s)\log(p_c(s))
-  \label{eq-entropy}
-\end{equation}
-
-Throughout this process $useful$ tracks the set of all gene clusters which should be preserved for use by crossover.
-This set begins with all genes in separate clusters, and each time a new cluster is created it is added to $useful$.
-However, not all clusters are necessarily worth keeping. For instance, in all versions of LTGA the cluster
-containing all genes is removed from $useful$ as preserving all genes during crossover can only create clones.
-\cite{thierens:2013:ltgahiff} extended this removal to include any unsupported subsets. If the pairwise distance
-between two clusters is 0, this means there are no individuals in the population which disrupt the relationships between the two
-clusters. Therefore when performing crossover there is no reason to believe a fitness improvement can be achieved
-by breaking the stored pattern. As such a cluster is only kept if its direct superset has a non-zero distance.
-As a final step Line~\ref{fig-cluster-creation-ordering} reorders $useful$ such that clusters appear in reversed
-order from which they were added to $useful$. Thus the most linked clusters and those containing single genes
-appear at the end of the returned list.
-
-\cite{thierens:2013:ltgahiff}'s version of LTGA does not use the entire population when determining pairwise entropy.
-Instead, binary tournament is used to select half of the population. This is done to ensure the model is built
-using only high quality solutions, even during the first generation.
-
-In order to efficiently perform clustering, a pairwise gene frequency table is constructed from the selected solutions.
-To calculate Equation~\ref{eq-distance}, Equation~\ref{eq-entropy} is called for each
-gene ($H(c_i)$) and pair of genes ($H(c_i \cup c_j)$). Extracting this information requires $O(\mu N^2)$
-time, where $\mu$ is the population size and $N$ is the genome size. The process of converting
-this pairwise frequency information into clusters can be achieved in $O(N^2)$ using the bookkeeping
-methods presented by \cite{gronau:2007:upgma}. This cost is performed only once per generation,
-and is then used to perform approximately $O(\mu N)$ crossover evaluations. As a result, the amortized cost of
-LTGA's model building is $O(N)$.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Cluster-Usage}{}
-    \ForAll{$C_i \in useful$}
-      %\ForAll{$d \in shuffled(P_i)$}
-        \State $d \leftarrow rand\_choice(P)$\label{fig-cluster-usage-donate}
-        \State Copy $d$'s gene values for $C_i$ into solution
-        \If{solution was changed}
-          \If{solution's fitness decreased}
-            \State Revert changes
-          \EndIf
-          %\State \textbf{break}
-        \EndIf
-      %\EndFor
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm describing how clusters are used to perform crossover.}
-  \label{fig-cluster-usage}
-\end{figure}
-
-Figure~\ref{fig-cluster-usage} describes how the identified clusters are used by crossover to preserve
-gene linkage while still exploring the search space. Unlike more traditional crossover methods, LTGA
-crosses each individual with the entire population. Also, to produce a single offspring, multiple evaluations
-of the fitness function are performed.
-
-Each generation, each individual in the population undergoes crossover. In a single crossover event, each
-cluster of genes $C_i$ in $useful$ is applied as a crossover mask. A random donor $d$
-is chosen from the entire population (not just the model selected population), and $d$'s gene's for $C_i$ are copied
-into the working solution. If a modification is made, an evaluation is then performed. If the crossover
-resulted in no worse fitness then the changes are kept. This allows for neutral drift across plateaus.
-The resulting solution, which must be at least as fit as its parent, is then
-copied into the next generation.
-
-In total each individual can cause up to $|useful|$ evaluations. If all clusters were kept, even those deemed
-unhelpful, and all donations were evaluated, even those which did not change any genes, then \Call{Cluster-Usage}{}
-would perform exactly $2N-2$ evaluations for each of the $\mu$ solutions in the population. This provides the amortizing evaluations
-required to make clustering only $O(N)$ operations per evaluation. However, by skipping some evaluations, its
-possible that clustering may be super-linear.
-
-LTGA has no explicit form of diversity control and has no method for introducing new genetic information once
-the population has converged. Therefore an LTGA run is considered converged once two consecutive populations
-contain the same unique solutions.
-
-By design, LTGA only has a single parameter: population size. LTGA uses no mutation, and crossover is defined
-in terms of the clustering algorithm. Selection between generations is fully elitist and embedded in the crossover,
-with selection of model building solutions fixed to a binary tournament. Neither \Call{Cluster-Creation}{} nor
-\Call{Cluster-Usage}{} rely on parameter values. LTGA does not provide any method for controlling or setting
-the population size, and uses a fixed user specified size for each generation.
-
-\section{Parameter-less Population Pyramid}
-\label{sec-p3}
-\cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid (P3) as a method for
-performing optimization which does not require the user to provide any parameters. This is
-achieved by combining efficient local search with the model building methods of LTGA using
-an iteratively constructed hierarchy of populations.
-
-The high level algorithm of P3 is presented in Figure~\ref{fig-p3}. Unlike more traditional
-GAs, P3 does not follow a generational model. Instead, it maintains an iteratively
-expanding pyramid of expanding populations. Each iteration, a new random solution is generated.
-This solution is brought to a local optimum using the hill climbing algorithm in Figure~\ref{fig-hc}. If that
-local optimum has not yet been added to the pyramid, the solution is added to the lowest
-population $P_0$.
-
-Next, the solution is iteratively improved by applying LTGA's crossover algorithm (Figure~\ref{fig-cluster-usage})
-with each population $P_i$ in the pyramid. If this process results in a strict fitness improvement and has
-created a solution not yet stored in the pyramid, it is added to the next highest pyramid level $P_{i+1}$.
-If $P_{i+1}$ does not yet exist, it is created. In this way populations in the pyramid expand over time,
-and the number of populations stored increases over time. Initially the pyramid contains no solutions
-or populations, meaning the user does not need to specify a population size.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Iterate-P3}{}
-    \State Create random solution
-    \State Apply hill climber
-    \If{solution $\notin hashset$}
-      \State Add solution to $P_0$
-      \State Add solution to $hashset$
-    \EndIf
-
-    \ForAll{$P_i \in pyramid$}
-      \State Mix solution with $P_i$
-      \If{solution's fitness has improved}
-        \If{solution $\notin hashset$}
-          \State Add solution to $P_{i+1}$
-          \State Add solution to $hashset$
-        \EndIf
-      \EndIf
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-  \caption{One iteration of P3 optimization. $pyramid$ is an
-           ordered set of populations and $hashset$ is a set
-           of all solutions in $pyramid$.}
-  \label{fig-p3}
-\end{figure}
-
-To accommodate P3's unique population structure, some of LTGA's clustering procedures were modified. In LTGA,
-clusters are identified at the start of each generation and are used
-to create all offspring in that generation. As P3 does not perform serial
-generations, P3 instead updates the model each
-time a solution is added to a population. Furthermore, unlike
-LTGA all solutions in the population are used to generate the model, not just the
-winners of a binary tournament. We can do this because even the
-worst solutions in the pyramid are already high quality due to local search.
-Using local search in LTGA was examined by \cite{bosman:2011:lsbbo} and found
-to provide no significant improvement. A likely cause was that that study applied
-local search to every solution, not just the initial population, resulting in significant overhead.
-
-Beyond the changes in population structuring, P3 modifies
-LTGA's version of \Call{Cluster-Creation}{} and \Call{Cluster-Usage}{}.
-P3 changes Line~\ref{fig-cluster-creation-ordering} in Figure~\ref{fig-cluster-creation}
-from \emph{last merged first} ordering to \emph{smallest first} ordering.
-This method applies gene clusters during crossover based on how many genes
-are in each cluster\footnote{Ties are broken randomly.}, and not on how tightly linked those genes are.
-\cite{goldman:2012:ltga} found that this alternative was better at preserving
-diversity, and therefore required smaller populations, but has not yet been integrated into
-canonical LTGA.
-
-P3 also modified Line~\ref{fig-cluster-usage-donate} in Figure~\ref{fig-cluster-usage}.
-Instead of choosing a single genetic donor for each cluster, P3 iterates over the
-population in a random order until a solution in the population is found which
-has a least one gene different for that cluster of genes from the improving solution.
-This process increases the likelihood of an evaluation being performed for every cluster,
-and helps test rare gene patterns in the population.
-
-In LTGA the cost of rebuilding the model is $O(\mu N^2)$ as it must collect pairwise
-gene frequency information for all $\mu$ solutions in the population. P3 does not store
-a single population, and it does not have a fixed $\mu$ size for any population. However,
-each time a solution is added to the population, it requires $O(N^2)$ time to update
-the table of pairwise frequencies and another $O(N^2)$ time to rebuild the linkage model.
-The model is then used immediately to perform up to one evaluation for each of the up
-to $2N-2$ clusters. Just as in LTGA, if no evaluation shortcuts were made, P3
-has an amortized cost of $O(N)$ modeling cost per evaluation. While P3 does
-rebuild the model more frequently per solution in the population, it also performs
-a number of local search evaluations which are quite efficient, meaning theoretical
-comparisons of their speed are difficult to perform. As a final note, P3's repeated
-attempts to find a useful donation make it less likely than LTGA to skip evaluations,
-but has an added cost to find these donations. Repeated donations could require as much
-as $O(\mu)$ attempts per evaluation, but experimental evidence suggest that this
-operation actually saves more overhead than it costs by increasing the number of evaluations
-per model rebuild.
-
-Each of the pieces of the P3 algorithm were selected not just for their stand alone efficacy,
-but for the ways in which they interact. By using the hill climber to ensure each solution
-is at a local optima, the underlying pairwise relationships in the problem are exposed. As
-a result, detecting clusters for use by crossover is much more effective. The crossover operator
-is extremely elitist, as each gene donation must result in no fitness loss, and a solution must
-strictly improve to be added to the next level of the pyramid. This is balanced by continual
-integration of new random solutions. Furthermore, each random restart decreases the probability
-of spurious linkages caused by shared ancestry. This diversity is further preserved by applying
-gene clusters in smallest first order during crossover as this reduces the probability of genetic
-hitchhikers.
-
-Other algorithms have proposed using multiple concurrent populations.
-\cite{hornby:2006:alps} had a hierarchy of populations with solutions periodically advancing
-upward. This allows for continuous integration of diversity as the lowest population is reseeded
-with random solutions. However, this method resulted in increased parameterization as not only was
-a population size required, but also parameters for how frequently generations advanced between levels and
-how many total levels to have were added. \cite{harik:1999:parameterlessga} used multiple independent
-populations of different sizes as a method for removing the population size parameter, but
-doing was provably less efficient than using an optimal population size as no information is shared
-between the populations.
-
-\section{Problem Descriptions}
-\label{sec-problems}
-
-\subsection{Single Instance Problems}
+\section{Single Instance Problems}
 Understanding how a stochastic search algorithm will behave on arbitrary and complex
@@
 
-\subsection{Randomly Generated Problem Classes}
+%%%%%%%%%%%%%%%%%% Randomly Generated %%%%%%%%%%%%%%%%%%%%%%
+\section{Randomly Generated Problem Classes}
 While well defined landscapes can provide specific insights into how an algorithm works,
@@
 problem is challenging we chose a clause to variable ratio of $4.27$.
-
-\section{Comparison Algorithm Parameter Tuning}
-\label{sec-tuning}
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-hboa}
-      \end{centering}
-      \caption{Tuned Population Sizes for hBOA}
-      \label{fig-pop-hboa}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-ltga}
-      \end{centering}
-      \caption{Tuned Population Sizes for LTGA}
-      \label{fig-pop-ltga}
-    \end{subfigure}
-  \end{centering}
-  \caption{Optimal population sizes found using bisection on each size
-           for each problem.}
-  \label{fig-pop-sizes}
-\end{figure}
-
-While four of the six algorithms in our experiments do not require any user specified parameters,
-hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
-handicapped, we extensively tuned each using the bisection method \citep{sastry:2001:bisection}
-to determine the optimal population size for each problem size.
-Extended by \cite{goldman:2012:ltga}, this method iteratively
-doubles the population size until some success criteria is met and then performs bisection
-between the lowest successful and highest unsuccessful sizes. In this way the minimum population size
-which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
-of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
-above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
-do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
-As bisection can make infinitely large population sizes, any run which had not found the global
-optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.
-
-Figure~\ref{fig-pop-sizes} shows the results from performing bisection. In general hBOA
-required population sizes which were at least an order of magnitude larger than LTGA. Due to
-runtime and memory overhead, finding the optimal value for hBOA was much less tractable than LTGA
-for moderate to large problem sizes. LTGA's population size also grew significantly slower
-than hBOA's as problem size increased, especially on the two Trap problems and Rastrigin.
-Both algorithms were very ineffective on MAX-SAT, neither able to tune to problems sizes over 60 bits.
-This is likely due to the fact that some randomly generated MAX-SAT landscapes are very flat
-and highly deceptive~\citep{rana:1998:gamaxsat}.
-\section{Finding the Global Optimum}
-\label{sec-optimum}
-
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{evals-to-success}
-  \end{center}
-  \caption{Comparison of the median number of evaluations to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-evals-to-success}
-\end{figure}
-
-Figure~\ref{fig-evals-to-success} shows the median number of evaluations required by each of the six
-algorithms to find the global optimum for multiple sizes of each problem.  Each data point in
-Figure~\ref{fig-evals-to-success} represents the median of 100 runs, where unsuccessful runs
-are treated as requiring more evaluations than any successful run. If the median run was not successful
-no point is shown. The maximum problem size used for each problem was set to be the largest size we could
-feasibly determine LTGA's optimal population size. For many larger problems results are not shown
-for hBOA due to the extreme computational cost required to optimally set the population size.
-
-\subsection{Quantitative Comparison}
-Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
-The largest problem size for any problem where P3 was not the most efficient has 49 bits, with P3 achieving the best results on all 92 larger
-configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
-Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
-and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
-on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
-
-\subsection{Local Search}
-The Random Restart Hill Climber and $1+(\lambda, \lambda)$ are both relatively effective on small problem
-sizes. This is especially true for the three randomly generated problem problem classes. These problems
-may contain relatively few local optima or just be exceptionally difficult for the model based algorithms.
-On Deceptive Trap and Deceptive Step Trap using 4 or less traps, $1+(\lambda, \lambda)$ performs significantly
-better than any other algorithm. We believe this is
-because $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
-flipping entire traps.
-This ability also leads $1+(\lambda, \lambda)$
-to outperform the Random Restart Hill Climber on all problems except Nearest Neighbor NK.
-
-On larger problem sizes, the ability for local search to reach the global optimum quickly diminishes.
-Only on MAX-SAT are these optimizers competitive
-at larger tested problem sizes. However, we believe this is because the largest tested MAX-SAT
-was an order of magnitude smaller than the largest size tested for most other problems. As the problem
-size increases the number of local optima increases exponentially, which explains why Random Restart Hill
-Climbing was unable to scale.
-For larger problems it also becomes increasingly unlikely for $1+(\lambda, \lambda)$ to make the right combination
-of changes required to reach the global optimum. This
-behavior causes high variance in success rate, as evident by the occasional successes on large Deceptive Trap problems.
-
-\subsection{Model Building}
-Only techniques which explicitly built models of gene epistasis were able to solve the largest problem
-instances. On single instance problems LTGA was more effective than hBOA, with hBOA outperforming
-LTGA on Nearest Neighbor NK and Ising Spin Glasses. This may be caused by the differences in modeling method:
-unlike the single instance problems, gene epistasis in the randomly generated
-problem classes cannot be be perfectly represented with a linkage tree.
-
-Considering how different hBOA and LTGA are in performing optimization, it is somewhat surprising how similar
-their results are on HIFF.
-However, both techniques rely on populations large enough to support the diversity required to reach the global optimum
-and to model epistasis. Both techniques also only rebuild models once per generation. As the subproblems of HIFF
-are nested, it is unlikely that either technique can accurately model higher order epistasis before solving
-lower order subproblems. Therefore both methods require one generation per subproblem order.
-
-\subsection{P3}
-Unlike the other model based methods, P3 generally outperforms
-both Random Restart Hill Climber and $1+(\lambda, \lambda)$ even on small problem sizes.
-Unlike the other local search methods, P3 outperforms LTGA and hBOA even on large problem sizes.
-This implies that P3 is gaining the benefits of each, leveraging local search to solve easy
-problems and model building to solve harder ones.
-
-Furthermore, the interaction between these two optimization tools explains some of the reason P3
-outperforms each method alone.
-On Deceptive Trap P3's use of hill climbing ensures all traps are immediately optimized, allowing for perfect
-linkage detection and high quality donation.
-On HIFF local search solves all pairwise subproblems, saving P3 a generation over LTGA and hBOA.
-In comparison P3 is only a slight improvement on Deceptive Step Trap, which is less amenable to local search.
-
-\section{Fitness Over Time}
-\label{sec-overtime}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{fitness-over-time}
-  \end{center}
-  \caption{Compares the median best fitness reached during search for each of the six optimization methods.}
-  \label{fig-fitness-over-time}
-\end{figure}
-
-For some applications, finding the global optimum is less important than finding good solutions quickly. Therefore
-we examine this behavior in Figure~\ref{fig-fitness-over-time}. At regular intervals during optimization
-Figure~\ref{fig-fitness-over-time} shows the median of the best fitnesses found at that point of search across
-100 runs. For each problem we show the largest problem size for which we were able to successfully gather results
-for all six algorithms, but the trends shown are representative of all larger problem sizes. The maximum reporting interval
-is set to include the slowest P3 run to reach the global optimum.
-
-% 181 total
-% 121 has P3 best
-%   9 hc
-%  50 lambdalambda
-%  20 hboa
-%  18 phboa
-%  27 LTGA
-
-\subsection{Quantitative Comparison}
-Of 181 sample points, P3 had the highest median fitness in 121. In pairwise competition, $1+(\lambda, \lambda)$
-was the most likely to outperform P3, doing so on 50 sample points. LTGA, hBOA, and Parameter-less hBOA
-were the next best, outperforming P3 on 27, 20, and 18 sample points. Random Restart Hill Climbing almost
-never outperformed P3, doing so only 9 times.
-
-\subsection{Local Search}
-Perhaps the most striking result is the quality of $1+(\lambda, \lambda)$. Until quite far into search this
-method performs better than both LTGA and hBOA. Given sufficient evaluations $1+(\lambda, \lambda)$ also outperforms
-Random Restart Hill Climbing on all 7 problems. For brief periods in the middle of search it performs the
-best of all techniques on: Deceptive Trap; Deceptive Step Trap; HIFF; Ising Spin Glass; and MAX-SAT problems.
-$1+(\lambda, \lambda)$'s ability to efficiently incorporate gene modifications of larger
-than one bit allows it to overcome the deception and plateaus in Deceptive Trap and Deceptive Step Trap, solve medium sized subproblems
-in HIFF, flip the signs on multiple adjacent bits in Ising Spin Glass, and cross plateaus in MAX-SAT. However,
-this method is slow in reaching the global optima in many of these problems which causes it to eventually
-be overtaken by the model building techniques.
-
-\subsection{Model Building}
-Both hBOA and LTGA are marked by periods of little improvement followed by rapid improvement.
-In hBOA this is taken to the extreme, with all fitness
-improvement made at the very end of search. In both cases this is caused by model building. Before the model
-is accurate little improvement is made. Once it is accurate, fitness improves dramatically.
-
-On every problem LTGA has five distinct periods: fitness plateau, near instantaneous improvement, fitness plateau,
-and improvement to global optimum. The early period corresponds with
-initialization of the population, with the first fitness gain achieved immediately upon completing the first generation.
-When using an inaccurate model, LTGA's mixing strategy performs a sort of less effective local search.
-Subsequent generations then make only minor fitness improvements. Once the model
-becomes accurate and the probability of a crossover using high quality genetic material increases sufficiently, LTGA
-enters a second period of rapid improvement.
-
-\begin{comment}
-This behavior is easiest to understand on the two Trap problems. LTGA's first generation will push individual traps toward a
-local optima, but it requires more than a single generation to do so completely. Until that occurs, the model is likely behaving
-no better than random as there is little apparent linkage between bits. Once the model begins to identify individual traps, crossover
-can begin to increase the frequency of higher fitness trap settings. This process begins slowly as low fitness local optima are more
-likely to be chosen for donation than high fitness local optima due to their frequency in the population. The process becomes
-self-catalysing as increased frequency of the global optimum trap genes means increased likelihood of the global optimum being spread by crossover.
-\end{comment}
-
-\subsection{P3}
-The integration of hill climbing into P3 makes it strictly better than using hill climbing alone.
-Early in optimization P3 and the Random Restart Hill Climber have effectively identical quality. This is because
-P3 performs the same evaluations as the Hill Climber for the first two restarts. Once P3 begins
-performing crossover it immediately improves over the Hill Climber. In 95\% of intervals P3 had a fitness at
-least as high as Hill Climbing. As such P3 is better than a simple
-hill climber regardless of how long each technique is run and irrespective of how high quality the solution
-found has to be.
-
-Unlike the model based methods, which struggle until model accuracy improves, P3's iterative solution
-integration allows it to improve much more quickly. This behavior exists in most problems, but is easiest to understand on Deceptive Trap.
-On this problem, P3 immediately brings all traps to local optima, equaled only by the Random Restart Hill Climber in quality.
-In comparison LTGA must evaluate the entire population and perform multiple generations to reach similar quality.
-P3 is able to immediately integrate optimal versions of each
-trap into a single individual as they are found by local search, resulting in smoother fitness improvement than LTGA.
-
-\section{Computational Expenses}
-While it is common in evolutionary computation to assume the evaluation function will dominate algorithm
-complexity, in some domains this will not be true. Model based methods are especially likely to violate
-this norm. Therefore, in order to assess P3's quality in solving problems with efficient fitness functions,
-we provide data on both its algorithmic complexity and wall clock time.
-
-\subsection{Operation Counting}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{rebuilds}
-      \end{centering}
-      \caption{Rebuilds}
-      \label{fig-rebuilds}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{donations}
-      \end{centering}
-      \caption{Donations}
-      \label{fig-donations}
-    \end{subfigure}
-  \end{centering}
-  \caption{Computation costs incurred by model rebuilding (Figure~\ref{fig-rebuilds}) and
-           repeated donations (Figure~\ref{fig-donations}) per evaluation as problem size increases.}
-  \label{fig-costs}
-\end{figure}
-
-When discussing the asymptotic complexity of P3 in Section~\ref{sec-p3}, two aspects eluded precise
-analysis: how expensive is model rebuilding and how many gene donations are made. Figure~\ref{fig-costs}
-provides some insight into how often these two aspects of the algorithm are utilized.
-
-Figure~\ref{fig-rebuilds} reports in an algorithmic sense how expensive model rebuilding is for search.
-In order to calculate this value we recorded how many times search rebuilt the model during each run.
-Under the minimal assumption that an evaluation costs $\Omega(N)$ time and that model rebuilds cost $O(N^2)$,
-Figure~\ref{fig-rebuilds} shows the ratio cost of model rebuilding per evaluation. If the cost of model building
-scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically constant.
-For Nearest Neighbor NK, Ising Spin Glasses, and Rastrigin this is the case. For both Trap problems
-and HIFF there is slow growth in the ratio. The problem sizes used for MAX-SAT were not sufficient
-to accurately gauge the asymptotic behavior. Together this suggests that while the cost of building the model
-is almost linear per evaluation, it can grow slowly. However, even in the worst case (HIFF) this growth was
-no more than twice the algorithmic cost of an evaluation even using 2048 bits.
-
-When applying a crossover subset, P3 tries random donors from the population until one is found with at least
-one bit different from the improving solution. In theory this can result in up to $O(\mu)$ operations.
-Figure~\ref{fig-donations} examines the observed average number of donations per evaluation performed.
-Ising Spin Glass, HIFF, and Rastrigin all achieve effectively constant behavior here, implying repeated
-donation does not impact the asymptotic runtime of P3. Both Trap functions and Nearest Neighbor NK all
-increase in number of donations as problem size increases, potentially increasing algorithmic costs. An
-important note is that each donation may range in size from a single bit up to $N-1$, with smaller
-donations far more likely to result in repeated attempts. As such this may cause some super-linear growth
-in P3, but its unlikely to be very high.
-
-\subsection{Wall Clock Performance}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{seconds-to-success}
-  \end{center}
-  \caption{Comparison of the median number of seconds to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-seconds-to-success}
-\end{figure}
-
-To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}.
-Similar to Figure~\ref{fig-evals-to-success}, each point represents the median
-of 100 runs, with unsuccessful runs treated as slower than successful runs.
-In total these runs
-took over 6 computing years to complete. We were therefore forced to use
-diverse hardware potentially running multiple processes in parallel.
-As such these timings should be considered very skeptically as a number
-of sources may contribute to biased error in the results. However, due to
-the size of the dataset and the general regularity in the results some general conclusions
-can still be made.
-
-\subsubsection{Model Building}
-hBOA and Parameter-less hBOA perform significantly worse when using wall clock time as the unit of comparison than when using
-evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while, under reasonable
-assumptions, P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
-hBOA goes from being slightly more efficient than LTGA in terms of evaluations to significantly worse in terms of seconds.
-As P3 and LTGA require a similar asymptotic complexity per
-evaluation as the Hill Climber and $1+(\lambda, \lambda)$, no similar change in ordering occurs.
-
-\subsubsection{P3}
-Any increase in P3's algorithmic complexity is offset by its efficient use of evaluations.
-In aggregate P3 found the global optimum the fastest on 72 of 130 configurations,
-with LTGA, $1+(\lambda, \lambda)$, Hill Climber, hBOA outperforming P3 on 44, 18, 10, and 1 configurations, respectively.
-On
-Deceptive Trap, Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT P3 still finds the global
-optimum faster than all comparison techniques. LTGA is able to match P3's performance almost exactly on HIFF,
-and on Rastrigin P3 falls behind only on moderate sized problems. Deceptive Step Trap yields the worst performance,
-with LTGA using optimal population sizes having a significantly lower growth curve than P3. This mirrors
-the results from Figure~\ref{fig-evals-to-success} under the assumption that P3's computational time per evaluation is higher
-than LTGA but by no more than a constant factor.
-
-\section{Population Sizing}
-A major advantage to P3 is that it does not require the user to set a population size parameter.
-Beyond making P3 easier to apply, this also conveys two additional advantages: diversity scaled
-to initialization and no need to sacrifice intermediate fitness for eventual optimality.
-
-\subsection{Problem Instance versus Problem Class}
-Our procedure for tuning LTGA and hBOA outlined in Section~\ref{sec-tuning} involved finding
-the optimal population size for each class of problem. For real world black box optimization this is
-realistically the best either algorithm could hope for as tuning to a problem
-instance or population initialization involves repeatedly solving the problem being tuned.
-This limitation does not exist in parameter-less methods, which scale their diversity
-based on the problem instance without needing to solve that instance repeatedly.
-
-To achieve high success rates on randomly generated problem classes, LTGA and hBOA must use a population size
-large enough to solve the hardest instances in that class. Therefore these methods will have
-population sizes larger than necessary to solve the easiest instances in the class. Even
-on single instance problems, both methods will require population sizes large enough to
-ensure the worst random initialization is diverse enough to solve the problem, which may be
-much larger than the best random initialization.
-
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{evals-to-success-range}
-  \end{center}
-  \caption{Comparison of the upper and lower quartiles of evaluations required
-           to reach the global optimum for P3 and LTGA with respect to problem size.}
-  \label{fig-evals-to-success-range}
-\end{figure}
-
-Figure~\ref{fig-evals-to-success-range}
-highlights how this can effect the required number of evaluations to reach the global optimum, showing
-the upper and lower quartiles for P3 and LTGA. On each problem
-LTGA has a much smaller difference between its best and worst runs. This makes sense as LTGA uses the same
-population size regardless of instance and progresses search generationally. In contrast P3 has a much
-higher split, with many runs finishing very quickly. This added variability is also generally on the lower
-quartile, with P3's upper quartile only worse than LTGA's on Deceptive Step Trap and Nearest
-Neighbor NK. This supports the hypothesis that P3 is able to scale its diversity not just
-to the problem class, but to the problem instance or even problem initialization, something
-wholly infeasible for tuned population sizing to do.
-
-This tuning distinction is also apparent when comparing Parameter-less hBOA with hBOA in Figure~\ref{fig-evals-to-success}.
-While generally performing worse than hBOA, the difference between the two algorithms is smallest
-on randomly generated problem classes. On MAX-SAT, Parameter-less hBOA actually outperformed both hBOA and LTGA,
-likely due to its ability to scale diversity to the problem instance instead of the entire problem class.
-
-\subsection{Fast versus Optimal}
-\label{sec-fast-vs-optimal}
-
-In Section~\ref{sec-overtime} we examined intermediate fitness qualities of LTGA and hBOA when using
-population sizes tuned to reach the global optimum. As a result, both were exceptionally ineffective
-at quickly reaching high quality solutions. This is because unlike P3, these methods have an explicit
-trade off between optimal performance and intermediate performance caused by their population size parameter.
-
-At 58\% of the recording intervals hBOA has the worst fitness of any solver. Most of the exceptions
-occur when hBOA is still evaluating its initial population, allowing this random search to temporarily surpass the
-local search methods. After $N$ evaluations, however, hBOA and LTGA both fall behind until their models begin
-to improve.
-Parameter-less hBOA reaches intermediate fitnesses faster than hBOA, doing so on 62\% of intervals, as its models begin
-to optimize earlier than hBOA.  However, this trend is reversed after a sufficient number of evaluations, most clearly
-on Deceptive Step Trap and Ising Spin Glasses, as hBOA's tuned population overtakes Parameter-less hBOA's parallel populations.
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{small-pop-dst}
-      \end{centering}
-      \caption{Deceptive Step Trap 203}
-      \label{fig-small-pop-dst}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{small-pop-nk}
-      \end{centering}
-      \caption{Nearest Neighbor NK 200}
-      \label{fig-small-pop-nk}
-    \end{subfigure}
-  \end{centering}
-  \caption{Comparison of how reducing LTGA's population size effects
-           the median best fitness reached during search.}
-  \label{fig-small-pop}
-\end{figure}
-
-Figure~\ref{fig-small-pop} examines the effect of population size on LTGA's intermediate fitness by examining the effect
-of reducing LTGA's population size to one tenth of the tuned value. The two problems shown are representative of the behavior of using
-a smaller population size on the other five problems. Reducing the population size caused LTGA to improve earlier but plateau
-at lower fitnesses. This caused LTGA's success rate to drop from 100 to 0 on Deceptive Step Trap and from 98 to 68 on Nearest
-Neighbor NK.
-Even when using a reduced population size for LTGA, P3 still achieved a fitness at
-least as high as LTGA at 80\% of intervals.
-
-\begin{comment}
-It is important to note that hBOA and LTGA were tuned specifically
-to find the global optimum and are not necessarily using the optimal population size for finding intermediate fitnesses. Consider
-that for both reducing the population size would almost certainly improve their fitness for evaluation steps less than their population
-size. To examine the effect of population size we tested LTGA using one tenth of the population size required for reliably finding the
-global optimum, with the results given in Figure~\ref{fig-small-pop}.
-Reducing the population size resulted in LTGA's initial fitness improvement
-occurring earlier, as the first generation is completed much more quickly. On Deceptive Step Trap this means that
-LTGA leaps ahead of P3 due to its ability to overcome the two bit fitness plateaus.
-%
-Across all tested problems using the smaller
-population size reduced the quality of LTGA's first fitness plateau, likely due to missing required diversity to reach higher quality.
-The second period of improvement also comes earlier and reaches a lower quality fitness. On Deceptive Step Trap 0
-runs reach the global optimum, down from 100 successful for the full population size. On Nearest Neighbor NK this drops the number
-of successful runs from 98 to 68. As a result we conclude that while reducing the population size of LTGA may improve its quality
-at early points during optimization, doing so reduces the robustness of the final solution found. This is in contrast to P3
-which balances high quality intermediate fitness without trading away eventual optimality.
-\end{comment}
-
-\section{Inner Workings}
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{cross}
-      \end{centering}
-      \caption{Crossover Proportion}
-      \label{fig-cross}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{cross-success}
-      \end{centering}
-      \caption{Crossover Success}
-      \label{fig-cross-success}
-    \end{subfigure}
-  \end{centering}
-  \caption{For each problem Figure~\ref{fig-cross} shows the proportion of P3 evaluations spend on crossovers
-           and Figure~\ref{fig-cross-success} shows the percentage of fitness improving crossover evaluations.}
-\end{figure}
-
-While analysis of optimization speed is useful from a practitioner standpoint, doing so provides very
-little insight into algorithm behavior. To better understand how P3 works in detail we present here
-a look at some internal features specific to P3.
-
-\subsection{Crossover}
-Figure~\ref{fig-cross} shows the proportion of evaluations P3 spends on crossover, as opposed to hill climbing,
-and Figure~\ref{fig-cross-success} shows what percentage of crossover evaluations resulted in a fitness improvement.
-Together these figures provide some insight into the role of crossover within P3. The behavior
-for each is clearly problem dependent and generally asymptoticly stable as problem size increases.
-
-When solving problems where epistasis can be effectively detected and represented by a linkage tree, P3 tends to spend
-less evaluations performing crossover and each crossover is more likely to be successful.
-Deceptive Trap and Rastrigin are the easiest problems to model epistasis, with local search quickly
-reducing pairwise entropy in each. These are also the problems where P3 uses the least evaluations
-on crossover and has the highest success rates for crossover. At the other extreme are Nearest Neighbor
-NK and Ising Spin Glasses, which both have overlapping linkage not representable by a linkage tree.
-These problems have the highest crossover usage and lowest crossover success of any problem except Deceptive
-Step Trap.  While Deceptive Step Trap's epistasis can be accurately modeled by a linkage tree, the
-exponential number of plateaus makes detecting gene linkage very challenging.
-
-\begin{comment}
-P3 performs optimization using a mixture of hill climbing and crossover. Exactly how much time the algorithm
-dedicates to each is not clear from the algorithm description alone. As such, Figure~\ref{fig-cross} shows
-that the relationship between the two is strongly problem dependent but relatively independent of problem size.
-In general, problems which can be more accurately represented by crossover
-required less crossover evaluations. Deceptive Trap and Rastrigin spent the least evaluations on crossover, using
-15\% and 20\% respectively on the largest problem sizes. Both of these problems are perfectly separable and have relatively
-few local optima in each subproblem. HIFF and Ising Spin Glasses use crossover for about 60\% of evaluations. The HIFF structure can
-be perfectly learned by the linkage tree, but the higher order relationships require multiple stages of crossover to optimize.
-Ising Spin Glasses have an overlapping structure, making them impossible to perfectly represent in a single linkage tree, but
-that overlap is relatively minor in comparison with Nearest Neighbor NK, which uses 81\% of evaluations on crossover. Deceptive Step Trap spends the highest
-percentage of evaluations on crossover, 92\%. Unlike all of the other problems, the value taken by an individual bit in Deceptive
-Step Trap has very little meaning. This is because any trap value is either a local optima or 1 bit flip away from being a local optima.
-This has a combined result of meaning it takes very few local search evaluations to find a local optima but very many crossover operations
-to proliferate the global optima. Epistatic relationships are also going to be very weak and even small population sizes are likely
-to mark all clusters as $useful$.
-
-Solutions are only added to the next highest pyramid level if at least one crossover donation resulted in a fitness
-improvement. Figure~\ref{fig-cross-success} examines the percentage of crossover donations that resulted in a fitness
-improvement. The ordering in Figure~\ref{fig-cross-success} is an almost perfect inversion of Figure~\ref{fig-cross},
-implying that the more successful crossover is at achieving fitness improving donations the lower the proportion of evaluations P3 will
-spend performing crossover. P3 achieves an impressive 30\% success rate for crossover donations on the largest Deceptive Trap
-problem sizes. This makes sense as crossover is able to immediately and perfectly identify entire traps such that all donations
-always contain only complete traps. Furthermore, by forcing crossover to attempt donations of genes different from the current solution,
-donations containing a single trap will have a very high success rate, always improving suboptimal traps which are far more frequently
-created by local search that optimal traps. Rastrigin and HIFF have relatively high success rates of 11\% and 7\%, respectively. Again
-this is likely due to accurate modeling and relatively low diversity of building block alternatives.
-Ising Spin Glasses and Deceptive Step Trap both have low success rates near 0.01\%. This follows as both are difficult to model with
-a linkage tree. In comparison to Figure~\ref{fig-cross-success} Deceptive Step Trap performs better than expected for crossover.
-However, the reason for this is that even exceedingly uninformed crossover can make improvements to Deceptive Step Trap as
-flipping two bits in the same trap will move between two local optima. Nearest Neighbor NK had the least successful crossovers,
-with only 0.007\% successful on the largest tested problems. In comparison to Ising Spin Glass, NK's overlap is far more extensive.
-As a result it makes sense that individual crossovers are very unlikely to be fitness improvements.
-\end{comment}
-
-When compared with LTGA, P3's crossover success rates are lower but similar in problem ordering.
-Counterintuitively, the use of hill climbing on the initial population reduces
-P3's crossover success, not because it reduces model quality or the donation pool, but because it is much more challenging to improve
-locally optimal solutions than randomly generated ones. LTGA's crossover benefits from application to unoptimized solutions, which
-makes its aggregate crossover success incomparable to P3's.
-
-Even when crossover success rates are quite low, such as Nearest Neighbor NK's 0.007\% success,
-the results from Section~\ref{sec-optimum} and Section~\ref{sec-overtime} show it's still critical to optimization. Without
-crossover, P3's performance would be identical to the Random Restart Hill Climber, which was unable to solve even moderately sized problems
-and quickly fell behind P3 in intermediate fitness quality. Therefore even infrequently successful crossover donations are
-critical to success. This does, however, suggest a potential avenue for future improvement by using more successful
-modeling and donation algorithms.
-
-\subsection{Pyramid}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{level-size}
-      \end{centering}
-      \caption{Population Size}
-      \label{fig-level-size}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{level-success}
-      \end{centering}
-      \caption{Crossover Success}
-      \label{fig-level-success}
-    \end{subfigure}
-  \end{centering}
-  \caption{For each problem Figure~\ref{fig-level-size} shows the number of solutions stored in each level of the pyramid
-           and Figure~\ref{fig-level-success} shows the percentage of fitness improving crossover evaluations at each level.}
-\end{figure}
-
-Another feature unique to P3 is the shape and size of the population pyramid
-constructed for each problem. Figure~\ref{fig-level-size} shows the number of solutions stored at
-each level of the pyramid for the largest tested problem sizes. Each point is the
-median size across 100 runs. If a run did not store any solutions at a level it is treated as 0.
-No point is drawn if the median run had 0 solutions stored at that level. While pyramid size
-is effected by problem size, the overall shape is not. As such the behavior shown in Figure~\ref{fig-level-size}
-is representative of that for all other tested problem sizes.
-
-With the exception of the dip in Deceptive Step Trap, all of the pyramids show a monotonic reduction
-in size as the level increases. This is because a solution must be a strict fitness improvement
-over its previous version to be added to a higher level, which becomes less likely
-each time the solution improves. \cite{lobo:2011:dynamicpop} found theoretical evidence
-and \cite{goldman:2011:dynamic-parameters} found empirical evidence that the optimal
-population size decreases each generation of traditional evolutionary search.
-By decreasing in size, P3 implicitly stores more diversity in low levels and focuses
-search on high quality solutions at higher levels. In comparison,
-LTGA and hBOA suboptimally use a fixed population size at each generation.
-
-Figure~\ref{fig-level-success} examines how crossover success changes at different
-levels of the pyramid. At low levels, success gets progressively lower
-as solution quality increases faster than the model's ability to improve solutions.
-At higher levels modeling becomes more accurate and donations contain higher frequencies
-of high quality building blocks, resulting in increased crossover success. The
-highest level of most problems has a low crossover success rate, as solutions
-crossing with that level have already been improved by previous operations to
-the point where the only improvement would be to create the global optimum, which can
-only happen once.
-
-\subsubsection{Deceptive Step Trap}
-The number of solutions stored at the fourth level of Deceptive Step Trap is
-significantly lower than that of the third or fifth levels, breaking the decreasing
-trend of the other six problems. Figure~\ref{fig-level-success} has a similar
-aberration, with crossover success dropping to  0.0004\% before rebounding and
-following the more common trajectory. This behavior exists in all other problem
-sizes tested, with the dips occurring at exactly the same level.
-
-This behavior is rooted in the peculiar nature
-of this landscape.
-After local search, all traps in all solutions have a total
-number of 1 bits equal to \{0, 1, 3, 5, 7\} as these correspond to the local
-optima when using $k=7$ and $s=2$. Crossover is very likely to overcome the
-two bit plateaus, and as a result solutions in the second level generally do not
-contain the lowest fitness local optima (5 bits set) and the third
-level has very few traps set to the next worst local optima (3 bits set). As a result,
-solutions which reach the third level can only be improved by replacing the deceptive
-local optima (0 and 1 bits set) with the global optimum (7 bits set).
-The global optimum is very rare in the population, and with 8 ways to represent local optima
-linkage learning is inaccurate. Therefore it is very unlikely for crossover to be successful,
-meaning few solutions will be added to the fourth level. Solutions that do improve by
-definition must have a higher frequency of optimal trap settings, meaning level four's model
-will be more accurate and donations are more likely to contain optimal trap values.
-Thus the level size and crossover success rates increase after contracting around level four.
-
-\section{Conclusions and Future Work}
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing black box
-optimization. P3's primary innovation is the replacement of the generational model with a pyramid of populations.
-This pyramid is constructed iteratively, with both the number of levels and the number of solutions stored
-at each level growing as search progresses. P3 uses a model based crossover method
-which learns a linkage tree from gene epistasis. Combined with a simple hill climber, P3's design contains
-many synergistic features.
-
-Across a large number of problems and problem sizes P3 required less evaluations to reach the global optimum
-than optimally tuned state-of-the-art competitors. On single instance problems P3's improvement was by a
-constant factor, while for the three problem classes P3's improvement increased with problem size. This
-quality extends to intermediate points during evolution, with P3 generally reaching at least as high
-of fitness as the competitive techniques when using the same number of evaluations. While P3 does require
-modeling overhead, the expense of this overhead is approximately linear with respect to genome size. There
-is some evidence that even when compared on wall clock time, P3 performs at least as well as
-the best comparison techniques. All of these achievements are made without any problem specific parameter
-tuning, making P3 easier to apply to new domains than its two closets competitors in quality.
-
-P3's quality is due to a number of desirable traits. First, mixing local search with model based
-crossover lets search focus on properly mixing high quality solutions. Second, by adding diversity only
-as necessary P3 tends to use the minimal amount of random initialization, unlike other techniques which must
-overcompensate with larger population sizes on single instance problems and consider the worst instance
-when solving problem classes. Third, by heavily exploiting existing diversity before adding more P3 is able
-to reach high quality intermediate fitnesses quickly without prematurely converging. Fourth, the very
-nature of the pyramid's shape allows search to preserve a desirable proportion of diversity at
-each fitness level, similar to a generational model using a decreasing population size.
-
-There are a number of meaningful avenues for future P3 experimentation. Perhaps the most
-pressing for practitioner acceptance is to apply P3 to real world problems and compare
-its results with other black box or even problem specific heuristics. While parameter-less,
-P3 is currently limited to discrete, fixed length genomes evaluated using single objective
-fitness. These limitations can be relaxed with future work to make P3 more widely applicable.
-While asymptotically linear, P3's modeling techniques and local search methods are likely going
-to be prohibitively expensive for genome sizes in the hundreds of thousands or millions of genes,
-and the inability of the model to capture overlapping linkage may be hindering
-search efficiency. Overcoming these limitations by using a new modeling technique may allow
-the pyramid model even greater flexibility. Similarly, while P3 is able to overcome low
-order deception via linkage learning, the iterative improvement method by which crossovers
-are made may mislead search on landscapes with higher order deception.
-
-However, even without
-these improvements our results show P3 is highly efficient at finding
-global optima on black box problems without any problem specific tuning.
-
-\small
-
-\bibliographystyle{apalike}
-\bibliography{../main}
-
-\normalsize
-\appendix
-\section{hBOA Simplification}
-\label{sec-appendix}
-To measure the quality of a decision forest, hBOA applies Equation~\ref{eq-hboa}.
-To favor compact
-models, Equation~\ref{eq-hboa} is scaled by Equation~\ref{eq-hboa-stop}, which provides increased cost
-for more total leaves.
-This quality is used for two purposes: comparison of potential changes from the existing model
-and comparison of that change with the existing model. The basis for our simplication
-is to rearrange $p(B)BDe(B) < p(B')BDe(B')$ to be $\frac{p(B)}{p(B')}<\frac{BDe(B')}{BDe(B)}$.
-
-\begin{equation}
-  BDe(B) = \prod_{i=0}^{N}\prod_{l\in L_i} \frac{\Gamma(m'_i(l))}{\Gamma(m_i(l) + m'_i(l))}
-  \prod_{x_i}\frac{\Gamma(m_i(x_i, l) + m'_i(x_i,l))}{\Gamma(m'_i(x_i,l))}
-  \label{eq-hboa}
-\end{equation}
-
-\begin{equation}
-  p(B) = c2^{-0.5(\sum_i|L_i|)log_2\mu}
-  \label{eq-hboa-stop}
-\end{equation}
-
-The outermost product of Equation~\ref{eq-hboa} iterates over all trees in the forest. However, each split
-can modify only one of the trees and therefore the contribution of all others can be canceled. The middle
-product is across all leaves in the tree. Again since only one leaf can be changed, all other terms can
-be canceled. By convention hBOA uses uninformed Bayesian priors of $m'_i(l)= 2$ and $m'_i(x_i, l)=1$ for
-binary alphabets. As $\Gamma(a) = (a-1)!$ this means the top term in the middle product and the bottom
-term in the third product reduce to 1. The only remaining terms are then $m_i(l)$ and $m_i(x_i, l)$ which
-represent the number of solutions which reached leaf $l$ and the number of solutions which reached leaf $l$
-with a specific value for $x_i$, respectively.
-
-Equation~\ref{eq-hboa-stop} can also be simplified when doing comparisons. If model $B'$ has exactly one more
-leaf than model $B$ then the ratio $\frac{p(B)}{p(B')}$ simplifies to $2^{0.5 log_2\mu}$ regardless of
-total model size.
-
-The resulting simplifications create Equation~\ref{eq-hboa-final},
-where $B'$ is different from $B$ by exactly 1 split, such that $l$ was split to create $l'$ and $l''$.
-The best split is whichever maximizes its improvement over $B$, which is equal to the right side
-of the inequality. Note that these factorials can still be exceedingly large and
-therefore it is imperative that implementations avoid rounding errors and overflows.
-
-\end{document}

### 2014-10-15 Comprehensive/chapter-proposed.tex
Initial commit of Comprehensive Report

\chapter{Proposed Research}

\section{Simple Memetic Algorithm Using Efficient Search}

\section{Developing Efficient Linkage Aware Crossover}

\section{Comparison of Gray Box Methods}

### 2014-10-24 Comprehensive/chapter-graybox.tex
Comprehensive Chapters 7 and 8 first drafts.

@@
 \chapter{New Problem Domain: Gray Box Optimization}
+\label{chap-graybox}
+All of the work presented so far has focused on the black box optimization
+domain. These problems are characterized by a complete lack of available
+problem specific information. All the algorithm has to perform search
+is a ``black box'' which measures the quality of arbitrary proposed
+solutions. Optimizers which are successful in this domain are very general
+as there are minimal requirements to apply them to a new problem.
+
+However, for many real world applications there is potentially more information
+available. The other end of the spectrum is white box optimization, in which
+the algorithm knows the exact problem class it is trying to solve. This domain
+is dominated by problem specific search
+heuristics~\cite{een:2004:satsolver,sherrington:1975:isingsolver,wright:2000:solvingnk} which
+leverage all aspects of the problem to achieve efficiency. This makes such
+algorithms very specific, such that outside of their target domain they
+either cannot be applied or have no expectation of quality. As such,
+each must also be designed by hand for each new problem class, requiring
+deep understanding of the problem and the time to develop the algorithm.
+
+In between those two domains there is another: gray box optimization. In this
+domain some features general to multiple problem classes are exploited, beyond
+just an evaluation function. The goal is to create optimization methods which
+benefits from these features without those methods becoming specialized to
+a small set of problem classes.
+
+In addition to the black box evaluation function, we shall
+add in two more exploitable features accessible to search algorithms:
+known variable epistasis and partial function evaluation.
 
 \section{Problems in this domain}
+To understand the types of problems in this domain, lets first examine
+a very simple artificial problem we'll call 3-Equal. The quality
+of a solution is determined by how often three consecutive bits are set
+to the same value. As a result the maximum fitness is $N$ (all zero or all one) and the minimum fitness
+is 0.
+
+The 3-Equal problem has both known epistasis and partial function evaluation. From the definition
+we know that each bit is epistatically linked to the two variables that precede it and the two
+that follow it. There are no other non-linear relationships with that bit. This also means
+it is possible to evaluate the fitness contribution of that bit only knowing the value of at
+most four others, regardless of the size of $N$. As we'll discuss in Section~\ref{sec-whitley-ls}
+this has enormous implications on search efficiency.
+
+Every problem discussed in Chapter~\ref{chap-problems} except HIFF fit into this domain. Deceptive
+Trap and Deceptive Step Trap both can have known epistasis (which bits are in which traps) and
+partial reevaluation (score a single trap), the same as Rastrigin (separable). All NK problems
+where $K \ll N $ have a knowable epistasis table and the fitness contribution of each is subfunction
+calculable without evaluating the entire string. Similarly, MAX-SAT's clause list specifies epistasis,
+with each clause independently evaluable. Ising Spin Glasses are even simpler, with each edge in the
+graph evaluable using only two problem variables.
+
+Beyond MAX-SAT and Ising Spin Glasses, which are interesting real world problems unto themselves,
+many $NP$-Hard real world problems can be expressed in these two requirements. This is especially true
+of graph problems.
+
+\begin{itemize}
+\item \textbf{Dominating set}: Find a minimum vertex
+set such that all vertices are either in the set or adjacent to something in the set.
+The fitness of a vertex is determined by if it is in the set (add to set size) and if
+it is dominated (add to undominated size). Epistasis is the adjacency set for each
+vertex, plus itself. Used in ad hoc wireless sensor networks~\cite{todo-dominating-set}.
+
+\item \textbf{Vertex Cover}: Find a minimum vertex
+set such that all edges are incident on at least one vertex in the set.
+Epistasis and fitness are calculated using each vertex (add to set size)
+and each edge (add to uncovered size), independently. Used in computation biology~\cite{todo-vertex-cover}.
+
+\item \textbf{Max-Cut}: Find the set of vertices which maximize the number of edges
+incident on exactly one vertex in the set. Epistasis and fitness are calculated
+using each edge independently. Used in VLSI design~\cite{todo-max-cut}.
+
+\item \textbf{Set Cover}: Given a universe of elements and a set of subsets of that universe,
+find a minimum set of subsets whose union recreates the entire universe. Each element in
+the universe is epistatically linked with subsets that contain that element (add to uncovered size)
+and with each subset also contributing directly to fitness (add to set size). Used in
+computer vision~\cite{todo-set-cover}.
+
+\item \textbf{Zero-One Linear Programming}: Given a set of constraints, maximize an
+objective function, all of which are linear combinations of variables. Each constraint
+creates an epistatic relationship between the variables in that constraint, and fitness
+is a linear combination of variables allowing for partial evaluation. Used in power systems~\cite{todo-linear-programming}.
+\end{itemize}
 
 \section{Formal Requirements}
+In order to draw conclusions about search efficiency, it is necessary to make the features
+of the target problem domain more explicit.
+
+The overall quality of a solution must be equal to the sum of applying all subfunctions to that
+solution. As a consequence each subfunction must be independently evaluable.
+The mapping of variables to subfunctions (epistasis) must also be known, and
+the maximum number of variables participating in each subfunction is $k$.
+To achieve maximum efficiency $k$ should be constant in regards to problem size. The cost
+of evaluating a subfunction should also be bounded by some function $b(k)$. For example, on MAX-SAT
+$k$ is the clause size, with $b(k)$ linear in clause size, and the fitness equal to the sum of
+all clause terms. On NK, $k=K+1$ as each subfunction in NK depends on a variable and $K$ others.
+
+As the mapping of variables to subfunctions is known, it is also possible to calculate $c$, which
+is the maximum number of subfunctions any variable participates in. For best efficiency $c$ should
+be constant in regards to problem size. For example for all Nearest Neighbor NK landscapes all variables
+participate in exactly $c=k=K+1$ subfunctions. Randomly generated MAX-SAT instances have no guarantee
+that $c$ is constant, but in general the expected number of subfunctions in which a variable appears is set to 12.81.
+Assuming that all subfunctions use $k$ variables and each variable appears in $c$ subfunctions
+provides a bound on the total number of subfunctions of $\frac{cN}{k}$.
 
 \section{Efficient Local Search}
+\label{sec-whitley-ls}
+Section~\ref{sec-hill-climber} presented an efficient method for performing local search for the black
+box domain. Each potential bit flip, referred to here as a move, requires the entire solution to be reevaluated, which takes $\Omega(N)$ time.
+Furthermore, each time a fitness improving move is made all previously tested moves must be tested again. As a result
+improving a random solution to be a local optimum using this algorithm requires between $\Omega(N^2)$ and $O(IN^2)$, where $I$ is the number
+of improving moves. The lower bound is achieved when the while loop in Figure~\ref{fig-hc} runs a constant number of times.
+For instance on One Max, Deceptive Trap, Deceptive Step Trap, and HIFF all possible single bit improvements are found during
+the first time through the loop. The worst case is when each loop is expected to only make a single improving move, which causes
+all $N$ moves to be reevaluated once per improvement, such as on the Leading Ones problem.
 
-\section{Efficient Hamming Search}
+Local search in the gray box domain can be significantly more efficient~\cite{whitley:2013:greedy}. This is due
+to two consequences of the domain: the fitness effect of making a move can be calculated in $O(1)$ time, not $O(N)$,
+and the number of moves which must be reevaluated per improvement is $O(1)$, not $O(N)$. This results in local
+search requiring $O(N+I)$ time. If $I$ is within a constant factor of $N$, this means generating random local
+optima is at most a constant factor slower than generating random solutions in the search space.
 
+\begin{figure}
+  \begin{algorithmic}[1]
+  \Procedure{InitializeDelta}{$solution$}
+    \State $fitness \leftarrow 0$
+    \State $\forall_{m} delta[m] \leftarrow 0$
+    \ForAll{$s \in subfunctions$}
+      \State $pre\_move \leftarrow f_s(solution)$
+      \State $fitness \leftarrow fitness + pre\_move$
+      \ForAll{$m \in effected\_moves(s)$}
+        \State $post\_move \leftarrow f_s(solution \oplus m)$
+        \State $delta[m] \leftarrow delta[m] + (post\_move - pre\_move)$
+      \EndFor
+    \EndFor
+  \EndProcedure
+\end{algorithmic}
+  \caption{Algorithm used to efficiently determine the change in fitness
+           associated with each potential move from a given solution.}
+  \label{fig-initialize-delta}
+\end{figure}
+
+To achieve this performance, this local search technique begins by determining the change in fitness
+caused by making each potential move $m$, denoted as $delta[m]$. Figure~\ref{fig-initialize-delta}
+calculates the $delta$ for each $m$ starting at a given solution, as well as the fitness of the solution.
+Here, $f_s$ evaluates
+the subfunction $s$ on the given solution, and $solution \oplus m$ is the result of making move $m$ on $solution$.
+For each of the $\frac{cN}{k}$ subfunctions \Call{InitializeDelta}{} must determine the fitness of that
+subfunction before and after making each of $k$ moves which overlap that subfunction. Combined this
+results in less than $cN2b(k)$ operations, which is $O(N)$ assuming $c$ and $b(k)$ do not grow with $N$.
+Also, this procedure calls $f_s$ only $k+1$ times more than is required to find the fitness of the solution itself.
+
+When performing hill climbing, only moves $m$ such that $delta[m] > 0$ are fitness improving moves. Initially
+all moves are considered $open$, meaning that they could potentially be fitness improvements. Each iteration
+a random move $m$ is chosen from $open$, and $delta[m]$ is checked. If $m$ is a fitness improving move,
+\Call{MakeMove}{$m$}, shown in Figure~\ref{fig-make-move}, is called.
+
+\begin{figure}
+  \begin{algorithmic}[1]
+  \Procedure{MakeMove}{$m$}
+    \State $fitness \leftarrow fitness + delta[m]$
+    \ForAll{$s \in effected\_subfunctions(m)$}
+      \State $pre\_both \leftarrow f_s(solution)$
+      \State $just\_m \leftarrow f_s(solution \oplus m)$
+      \ForAll{$m' \in effected\_moves(s)$}
+        \State $just\_m' \leftarrow f_s(solution \oplus m')$
+        \State $post\_both \leftarrow f_s(solution \oplus m \oplus m')$
+        \State $delta[m'] \leftarrow delta[m'] - (just\_m' - pre\_both) + (post\_both - just\_m)$
+        \State $open(m')$
+      \EndFor
+    \EndFor
+    \State $close(m)$
+    \State $solution \leftarrow solution \oplus m$
+  \EndProcedure
+\end{algorithmic}
+  \caption{Algorithm for updating stored information related to a solution when
+           making a move.}
+  \label{fig-make-move}
+\end{figure}
+
+\Call{MakeMove}{} updates the fitness and $delta$ values to reflect the change in the solution. The fitness
+of the solution after making the move does not require any calls to $f_s$ as $delta[m]$ already stores the
+change in fitness. This process requires updating all the $delta$ values for the $k$ moves which interact with each of the $c$ subfunctions
+effected by $m$. This update replaces outdated information which used the original solution ($just\_m' - pre\_both$)
+with how much the move improves over the new solution ($post\_both - just\_m$). As $delta[m']$ has updated, $m'$ could
+potentially become a fitness improving move and is therefore added into $open$. As $m$ was just flipped,
+it cannot be a fitness improvement and is therefore removed from $open$.  In total this requires less than
+$ck4b(k)$ time, which is $O(1)$ assuming $c$, $k$, and $b(k)$ do not scale with $N$.
+
+Each time an improving move is found at most $ck$ additional moves are added to $open$. If $open$ ever
+becomes empty a local optimum has been reached. Therefore the number of times $delta[m]$ is checked
+is no more than $N + Ick$. As a result the total cost of improving a random solution until it reaches
+a global optimum is only $O(N+I)$.
+
+As an interesting addition, this method can actually be used to efficiently perform approximate steepest ascent hill
+climbing~\cite{whitley:2013:greedy}. Instead of choosing moves from $open$ at random, the moves are binned
+using their $delta$ values. Moves are then chosen randomly from the highest quality non-empty bin, and changes
+in $delta$ can cause moves to change bins. Assuming the total number of bins does not increase with $N$, both steps
+can be done in $O(1)$ time. However, at least for MAX-SAT, there is evidence that when used in combination with subsequent
+search heuristics the less greedy first improvement algorithm was more effective.
+
+\subsection{Preliminary Empirical Comparison}
+\begin{figure}[t]
+  \begin{centering}
+    \begin{subfigure}{.5\textwidth}
+      \begin{centering}
+        \includegraphicsfit{hc-time}
+      \end{centering}
+      \caption{Seconds per local optima.}
+      \label{fig-hc-time}
+    \end{subfigure}%
+    \begin{subfigure}{.5\textwidth}
+      \begin{centering}
+        \includegraphicsfit{hc-radius}
+      \end{centering}
+      \caption{Moves in Hamming Ball}
+      \label{fig-hc-radius}
+    \end{subfigure}
+  \end{centering}
+  \caption{Comparison of performing hill climbing
+           using the black box and gray box domains on Nearest Neighbor NK with $K=5$.
+           Figure~\ref{fig-hc-time} shows the time required to bring a random
+           solution to a local optima, averaged over 900 restarts across 30 instances.
+           Figure~\ref{fig-hc-radius} shows the number of moves which must be
+           tested by hill climbing when $N=500$ as $r$ increases.}
+  \label{fig-hc-compare}
+\end{figure}
+
+Figure~\ref{fig-hc-time} compares the black box hill climber presented in Section~\ref{sec-hill-climber}
+with the gray box hill climber presented in Section~\ref{sec-whitley-ls}.  Each algorithm was
+used 30 times on each of 30 instances to improve random solutions until they reached a local optimum.
+Both methods use first improvement hill climbing and achieved statistically indistinguishable fitness results.
+However, the amount of time each used is drastically different. On the largest tested size of $N=8,000$ the gray box
+hill climber was 800 times faster than using the black box method. This improvement was so extreme
+that the gray box hill climber is able to find local optima on $N=8,000$ problems faster than the black box
+method can optimize $N=400$ problems.  When compared asymptotically the black box optimizer empirically requires
+$O(N^{2.14})$ time while the gray box requires $O(N)$. This matches the theoretical conclusions for both algorithms.
+
+\section{Efficient Hamming Ball Search}
+Another consequence of the gray box domain is that increasing the neighborhood of local search becomes
+more tractable~\cite{chicano:2014:ball}.  Instead of improving solutions until no single bit flip
+is a fitness improvement, this can be expanded to no flip of $r$ or less bits can improve the solution.
+
+In a black box setting, verifying that no $r$ bit flip can improve a solution requires testing all
+$n \choose r$ neighbors. This quickly becomes intractable as $r$ increases. However in the gray
+box domain not all combinations need to be tested. Consider that if two variables do not participate
+in the same subfunction, the relationship between their effects is by definition additive. As such
+there is no way for flipping both together to be a fitness improvement without flipping one of them
+being a fitness improvement. Therefore it is not necessary to try all possible $r$ sized subsets
+of the solution.
+
+Consider the 3-Equal problem.  If there are at least two loci between $x_i$ and $x_j$ they do
+not share a common subfunction. Therefore, the change in fitness resulting from flipping
+both is equal to $delta[x_i] + delta[x_j]$. As such if $delta[x_i] \leq 0$ and $delta[x_j] \leq 0$
+then $delta[x_i,x_j] \leq 0$. Now consider a solution such that the first half is set to 0
+and the second half is set to 1. Even though $c=3$ there is no way to improve this solution without
+simultaneously flipping all 0's to 1's or vice versa. Any smaller flip will not
+be a fitness improvement.
+
+To determine which of the $n \choose r$ flips which must be checked, consider a graph where
+each vertex is a variable in the solution. An edge exists between two vertices if and only
+if the corresponding variables participate in at least one subfunction together. Restated,
+there is only an edge if the two variables have a non-linear relationship. The maximum
+degree of a vertex in this graph is $c(k-1)$, making it sparse for sufficiently large $N$.
+If a subset of vertices is connected then it is possible that flipping all of those
+variables together will result in a fitness improvement even when flipping any subset
+of the subset will not. However, if the subset is not connected then each component
+of the subset can be tested independently.
+
+\begin{figure}
+  \begin{algorithmic}[1]
+  \Procedure{ConnectedSubgraphs}{$v$, $subset$, $closed$, $open$}
+    \State $new\_subset \leftarrow subset \cup \{v\}$
+    \State $found \leftarrow [new\_subset]$
+    \If{$|new\_subset| \geq r$}
+      \Return $found$
+    \EndIf
+    \State $closed\_here \leftarrow \emptyset$
+    \State $open' \leftarrow open \cup adjacent(v)$
+    \ForAll{$v' \in open'$}
+      \If{$v' \notin closed$}
+        \State $closed\_here \leftarrow closed\_here \cup \{v'\}$
+        \State $closed \leftarrow closed \cup \{v'\}$
+        \State $recurse \leftarrow $\Call{ConnectedSubgraphs}{$v'$, $new\_subset$, $closed$, $open'$}
+        \State $found \leftarrow found + recurse$
+      \EndIf
+    \EndFor
+    \State $closed \leftarrow closed - closed\_here$
+    \State \Return $found$
+  \EndProcedure
+\end{algorithmic}
+  \caption{Algorithm to recursively find all connected induced subgraphs of size $r$ or less which
+           contain $subset \cup \{v\}$ and no additional vertices in $closed$.}
+  \label{fig-connected-subgraphs}
+\end{figure}
+
+The problem of finding which subsets must be checked is therefore equivalent to
+finding all induced subgraphs of $r$ or less vertices which are connected. For sparse
+graphs this can be efficiently calculated using \Call{ConnectedSubgraphs}{} outlined
+in Figure~\ref{fig-connected-subgraphs}. To our knowledge this is a novel algorithm
+as previous work did not specify how connected subgraphs were found.
+
+$subset$ and $open$ are both empty for the first top level call of \Call{ConnectedSubgraphs}{}, and
+$closed$ contains only $v$.
+The result of that call is to find all induced connected subgraphs which contain $v$.
+Subsequent top level calls expand $closed$ to also include their $v$.
+Each recursive level explores how $new\_subset$ can be expanded by including all $v'$ which are adjacent
+to at least one vertex in $new\_subset$. Once all subsets including $new\_subset \cup v'$ have been found,
+$v'$ is closed to prevent duplicates. However, when recursion ends $closed$ is returned to its previous state.
+The maximum recursive depth for this algorithm is $r$ as each level increases $|subset|$ by 1. Each level
+makes at most $|open'|$ recursive calls, with $|open'| \leq |open| + c(k-1)$. Finally there must be a top level
+call with each of the $N$ variables. This creates a loose bound of $O(r!(c(k-1))^rN)$, which is $O(N)$ under the
+assumption that $r$ and $c$ do not increase with $N$. The bound is loose because some portion of recursions
+are prevented by $closed$ and because overlapping adjacency will reduce the rate at which $|open'|$ increases.
+
+As this algorithm finds all connected subsets in $O(N)$ time for a fixed $r$, the number of moves
+which must be tested to determine if a solution is an $r$-bit local optima is $O(N)$. This means that
+while on Nearest Neighbor NK with $N=8000$, $K=5$, and $r=3$ the black box method would require 85 billion
+checks, the gray box method requires only 248,000. Even when allowing connections to be completely random,
+which results in $c$ increasing with $N$, the gray box method still only requires approximately 375 million checks,
+two and a half orders of magnitude less than assuming a black box.
+
+From these conclusions it is possible to modify the hill climber presented in Section~\ref{sec-whitley-ls} to
+efficiently find $r$-bit local optima~\cite{chicano:2014:ball}. The only change is that instead of having a move
+and $delta$ for each bit, there must be a move and delta for each connected induced subgraph of $r$ or less bits.
+The efficiency analysis is unchanged with the exception that the constant increases exponentially with $r$ as
+the number of $delta$ values which must be updated each move increases. Still, with $c$, $r$, and $k$ constant
+with respect to $N$, it is possible to find $r$-bit hamming ball local optima in $O(N+I)$ time.
+
+\subsection{Preliminary Empirical Comparison}
+Figure~\ref{fig-hc-radius} shows how the number of moves increases as the radius of the hamming ball increases.
+Even on this relatively small problem ($N=500$) the black box method quickly becomes intractable. The gray box
+method is able to use less moves to find 4-bit local optima than the black box needs to find 2-bit local optima,
+and when solving 6-bit local optima the black box method requires 653,635 times as many moves as gray box. However,
+gray box still increases exponentially as $r$ increases, limiting how large $r$ can grow before becoming intractable.
+This relationship depends heavily on $c$ and how variables are distributed among subfunctions.
+
+\begin{comment}
 \section{Memetic Algorithms}
+~\cite{chen:2011:memetic}
+
+Local search, even $r$-bit hamming ball search, is not sufficient to robustly find global optimums on
+highly multi-modal problems. This limitation was shown in Figure~\ref{fig-evals-to-success} with hill climbing
+unable to reach the global optima on almost any large size problem.
+\end{comment}

### 2014-10-24 Comprehensive/chapter-proposed.tex
Comprehensive Chapters 7 and 8 first drafts.

@@
 \chapter{Proposed Research}
+The Parameter-less Population Pyramid (P3) was shown to be very effective in Chapter~\ref{chap-benchmarking}
+when applied to black box optimization. Some of this effectiveness in due to P3's integration of
+efficient local search and its ability to automatically detect and exploit variable epistasis.
+In Chapter~\ref{chap-graybox} we presented the gray box domain, which contains many interesting
+real world problems, and the improvements that can be made to local search in this domain.
+This presents the following questions:
+
+\begin{itemize}
+\item How can hamming ball local search be integrated into a global optimization method?
+\item What improvements can be made to P3's crossover operator when linkage is known?
+\item How effective is a gray box P3 algorithm in comparison with other gray box techniques?
+\end{itemize}
 
 \section{Simple Memetic Algorithm Using Efficient Search}
+%8.22154, 24.0066, 24.0077, 24.0016, 24.0108, 24.0032, 24.0095, 24.0035, 24.0102, 24.0071, 24.0078, 24.0061, 24.0023, 24.0055, 24.0016, 24.0022, 24.0094, 24.0045, 24.0064, 24.003, 24.0109, 24.0099, 24.0038, 4.78113, 5.47084, 24.0015, 24.0088, 24.0066, 24.0019, 20.6173, 24.0023, 18.9085, 24.0063, 24.0001, 24.0041, 24.0065, 24.002, 24.0017, 24.0029, 24.0096, 24.001, 24.0103, 24.0076, 24.0064, 2.73054, 24.0035, 24.0086, 24.0013, 24.0057, 24.0112
+No current published research has examined how effective hamming ball hill climbing is at reaching the global
+optimum. However, our preliminary experiments have found that even using large $r$ values when solving relatively small $N$
+problems, hamming ball hill climbing is not sufficient to find the global optimum. When applied to 50 instances
+of $N=200$, $K=5$, $r=4$, and limited to using less time than the slowest P3 run to solve that class (24 minutes),
+the hamming ball hill climber only found the global optimum 6 times.
+
+Therefore as a first step in exploring this new domain we propose to explore simple memetic algorithms~\cite{chen:2011:memetic}
+which make use of this hill climber. The goal is to determine if combination with methods like uniform crossover
+can result in improved search quality.
+
+\begin{figure}
+  \begin{algorithmic}[1]
+  \Procedure{Iterate-SUT}{}
+    \State Create random solution
+    \State Apply hamming ball hill climber to solution
+    \ForAll{$T_i \in T$}
+      \If{$T_i$ is empty}
+        \State $T_i \leftarrow$ solution
+        \State \Return
+      \EndIf
+      \State Cross solution with $T_i$ to create $2^i$ offspring
+      \State Apply hamming ball hill climber to each offspring
+      \State solution $\leftarrow$ best of offspring, solution, and $T_i$
+      \State $T_i \leftarrow$ empty
+    \EndFor
+    \State Add solution to end of $T$
+  \EndProcedure
+\end{algorithmic}
+  \caption{One iteration of SUT optimization. $T$ is an
+           ordered list of solution, each of which could be empty,
+           awaiting a crossover partner.}
+  \label{fig-sut}
+\end{figure}
+
+We shall tentatively refer to the first proposed algorithm of this type as the Simple Uniform crossover Tournament (SUT)
+algorithm. Conceptually, SUT is meant to be the simplest way to combine uniform crossover with hamming ball hill climbing.
+Solutions compete in an elimination tournament. Each contest in the tournament uses crossover between the solutions, with each
+offspring subsequently brought to a local optimum using hamming ball hill climbing. The best solution of the parents and offspring then
+advances up the tournament.
+
+Similar to P3, this algorithm is entirely iterative and requires no parameters. Figure~\ref{fig-sut} provides the high level algorithm
+for SUT. Each iteration starts by generating a random solution and improving it using the hamming ball hill climber. It then
+iteratively climbs the tournament $T$, competing with each $T_i$ if $T_i$ has a stored solution. If $T_i$ is empty the solution
+is stored until its competitor is found, and if the solution is not stored before reaching the top of the current bracket, a new
+level is added. Adding a new level means that a complete binary tree of crossovers has been created, allowing this algorithm
+to simulate an elimination tournament without knowing the size of the tournament beforehand. Furthermore, this algorithm should
+never need to be restarted from scratch as new diversity is added and improved before being biased by previously found solutions.
+
+When competing two solutions, crossover is used $2^i$ times, where $i$ is $[0..|T|-1]$. This means that the total number of crossovers
+performed at each level is equal. For example, to have a single competition at $T_2$ requires two competitions at $T_1$ and four
+competitions at $T_0$, with each level performing 4 crossovers.
+Each crossover produces a single offspring using uniform crossover, which is then improved using
+the hamming ball hill climber. The best of all offspring and the two parents then compete on fitness, with the best advancing
+up the tournament.
+
+\subsection{Preliminary Empirical Comparison}
+%1.30066, 0.668016, 20.0872, 24.015, 6.84026, 0.620081, 1.43271, 10.9687, 0.625069, 24.0347, 2.84006, 1.57277, 2.81627, 7.82749, 0.63878, 1.52631, 1.31218, 1.41851, 0.650093, 2.79497, 3.01983, 0.28769, 24.0449, 0.292457, 0.316901, 2.957, 1.35011, 0.636912, 2.8645, 0.578051, 1.30444, 0.274158, 3.16518, 0.141998, 1.34107, 13.8565, 0.579429, 0.345084, 0.650234, 13.4685, 3.04961, 0.670062, 1.53499, 19.3596, 2.76979, 1.50853, 1.3763, 1.4152, 12.3199, 2.13644
+When applied to the same problem and using the same $r$, SUT significantly outperforms the hamming ball hill climber.
+On Nearest Neighbor NK $N=200$, $K=5$, $r=4$, SUT was successful 47 of 50 times when limited to 24 minutes of computation.
+Its median time to success was 1.47 minutes, which is on par with P3's black box median time of 1.27 minutes.
+There is a good chance that this similarity will change on larger problems. Its possible that SUT will not
+continue to be able to find the global optimum as $N$ increases. It is also possible that P3's use of complete solution
+reevaluation and modeling may cause its runtime to grow faster than SUT. Further experimentation is required to know
+with certainty which effect will dominate search time.
 
 \section{Developing Efficient Linkage Aware Crossover}
+The reason for recording variable entropy in P3 is to detect unknown variable linage in the black box domain. However,
+the gray box domain makes high level variable linkage fully known. Therefore by leveraging this information it should
+be possible to more efficiently construct crossovers which are even more accurate than the inferred black box ones.
 
+There are a number of possible avenues for future research here. First, should the goal be to derive a single
+static linkage model for crossover or multiple random approximations? It may be possible to compute for
+a given problem the set of all useful crossovers which best balance exploitation with exploration. As
+this only needs to be done once per problem, the algorithm for creating the set can be relatively time consuming.
+The drawback of such a design is that the set may be prohibitively large, require an accurate understanding of what
+the best crossovers are, and repeated use of a static crossover set may adversely effect search.
+
+Repeatedly generating new random crossover sets overcomes many of these issues. An individual set can be relatively small
+as it does not need to contain all possibly useful crossovers. Simultaneously random generation allows the definition for
+what types of crossovers should be used to be less rigorous, favoring instead a more inclusive definition. The drawback
+is that generating random sets may be prohibitively expensive, it may unnecessarily create useless crossovers, and fail
+to create difficult to detect but important crossovers.
+
+Independent of if static or random crossovers are used, there are important research questions related to how variable epistasis
+should best be leveraged in crossover. One possibility is that crossover should aim to preserve build blocks. For instance P3's
+crossover attempts to cross entire traps on Deceptive Trap, as mixing values within a trap does not lead to the global optimum.
+Under this principle it seems reasonable to assert that crossovers which ensure variables which share a non-linear relationship
+are crossed as a group are better than those that do not. The task then is to decide how to handle subfunction overlap to allow
+for sufficient exploration.
+
+Conversely, it may be that intentionally mixing values within subfunctions leads to better exploration. When using a hamming
+ball hill climber as a repair function, taking genes from multiple solutions for a subfunction may be beneficial. Consider
+that on Deceptive Trap were $r\geq trap\_size/2$ the best crossover can be to split every trap in half. While the immediate
+offspring may have low fitness, the hill climber is then able to reach either parent's values for the trap: If the parents
+disagree for a trap, the hill climber can flip enough bits in a single move to return to either's value. Furthermore, it
+also allows the hill climber to search for new high quality ways of solving that subfunction.
+
+As an initial experiment, we have focused on recreating the P3 style of linkage tree without using entropy tables.
+Figure~\ref{fig-sfx-tree} provides an algorithm for using subfunction information to create a linkage tree.
+Initially all variables start in their own cluster. Each subfunction is then used in a random order such that
+if that subfunction overlaps multiple top level clusters it merges them to create a new cluster.
+
+\begin{figure}
+  \begin{algorithmic}[1]
+  \Procedure{SubfunctionTree}{}
+    \State $clusters \leftarrow [\{0\}, \{1\}, \{2\}, \dots, \{N-1\}]$
+    \State $cluster\_number \leftarrow [0 .. N-1]$
+    \ForAll{$subfunction \in shuffled(subfunctions)$}
+      \State $to\_merge \leftarrow \emptyset$
+      \ForAll{$b \in subfunction$}
+        \State $to\_merge \leftarrow to\_merge \cup \{cluster\_number[b]\}$
+      \EndFor
+      \If{$|to\_merge| > 0$}
+        \State $new\_cluster \leftarrow \emptyset$
+        \ForAll{$i \in to\_merge$}
+          \State $new\_cluster \leftarrow new\_cluster \cup clusters[i]$
+        \EndFor
+        \ForAll{$b \in new\_cluster$}
+          \State $cluster\_number[b] \leftarrow |clusters|$
+        \EndFor
+        \State $clusters \leftarrow clusters + new\_cluster$
+      \EndIf
+    \EndFor
+    \State Remove first $N~clusters$ and any cluster containing all variables
+  \EndProcedure
+\end{algorithmic}
+  \caption{Algorithm used to convert a list of subfunctions into linkage tree clusters.}
+  \label{fig-sfx-tree}
+\end{figure}
+
+The computational cost of this algorithm is $O(|cluster|)$ for each $cluster$ created.
+Assuming that all variables participate in at least one subfunction, this algorithm will create
+$\Omega(N)$ clusters. This is a safe assumption as in the gray box domain variables that do not
+effect fitness have no reason to appear in the solution.
+In order to complete,
+each bit of each subfunction must be referenced, requiring $O(Nc)$ time. As such this cost can
+be amortized over the $\Omega(N)$ clusters to become $O(1)$.
+Each time a new cluster is created
+the $cluster\_number$ of all bits in that cluster must be updated, requiring $O(|cluster|)$ time.
+Therefore using this algorithm detecting clusters is only a constant factor slower than using those clusters
+to perform donations.
+
+This methods has many desirable properties beyond efficiency. All crossovers always move at least one entire subfunctions,
+and clusters always differ by at least one entire subfunction. The probability two variables are linked in a
+cluster increases as the number of subfunctions they share increases. This extends to indirect relationships, such that
+the more paths of a given length between two variables the more likely they are to be kept in the same cluster.
+Yet because this process is stochastic, any connected subset of subfunctions can be a cluster.
+
+\subsection{Preliminary Empirical Comparison}
+We tested P3 using the gray box \Call{SubfunctionTree}{} in place of the black box \Call{Cluster-Creation}{}
+and using the gray box hamming ball hill climber instead of black box hill climbing. As an added efficiency
+instead of recalculating all $delta$ values and assuming all moves are open each time a crossover donation is made,
+only $delta$s and moves effected by the donation are updated. This means that hill climbing requires $O(|cluster|+I)$
+time.
+
+%0.0939752, 0.154129, 0.392246, 0.721644, 0.110155, 0.241687, 0.321031, 0.172651, 0.567039, 0.480878, 0.109734, 0.176248, 0.343879, 0.301121, 0.202199, 0.17966, 0.0330468, 0.117303, 0.0885791, 0.0588664, 0.594459, 0.101318, 1.86743, 0.0350668, 0.598237, 0.388794, 0.208924, 0.240061, 0.103984, 0.21943, 0.238308, 1.10106, 0.142684, 0.332108, 0.336863, 0.49791, 0.119978, 0.155924, 0.243874, 0.142354, 0.371339, 0.0854216, 0.168894, 0.419617, 0.128611, 1.9637, 0.288967, 0.142157, 0.103391, 0.307311
+On Nearest Neighbor NK
+$N=200$, $K=5$, $r=4$ this found the global optimum in all 50 runs, with a median time to success of 0.23 minutes. This makes it the
+most successful gray box tested so far, and 6 times faster than applying black box P3. Also, unlike the other
+methods decreasing $r$ actually makes it more efficient, achieving a median time to success of 0.013 minutes (0.78 seconds) with $r=1$,
+97 times faster than black box P3.
+%1.30066, 0.668016, 20.0872, 24.015, 6.84026, 0.620081, 1.43271, 10.9687, 0.625069, 24.0347, 2.84006, 1.57277, 2.81627, 7.82749, 0.63878, 1.52631, 1.31218, 1.41851, 0.650093, 2.79497, 3.01983, 0.28769, 24.0449, 0.292457, 0.316901, 2.957, 1.35011, 0.636912, 2.8645, 0.578051, 1.30444, 0.274158, 3.16518, 0.141998, 1.34107, 13.8565, 0.579429, 0.345084, 0.650234, 13.4685, 3.04961, 0.670062, 1.53499, 19.3596, 2.76979, 1.50853, 1.3763, 1.4152, 12.3199, 2.13644
 \section{Comparison of Gray Box Methods}
+Key comparisons we plan to examine:
+\begin{itemize}
+\item How does the combination of uniform crossover with hamming ball hill climbing compare to hill climbing alone?
+\item Does preserving subfunctions intact during crossover improve search efficiency?
+\item How does black box P3 compare with gray box P3?
+\item How do the more generic gray box methods compare with problem specific methods for solving real world problems?
+\end{itemize}
+
+Current research on these gray box methods have been very preliminary: a single size of a single problem class using
+a single $r$, $k$, and $c$. To fully understand how to effectively perform search in this new domain, much more extensive testing
+is necessary.
+
+Just as when testing P3 in the black box domain, an important step will be to identify a representative set
+of gray box problems. Planned test problems are NK landscapes, Ising Spin Glasses, and MAX-SAT. Beyond testing the
+problems with known global optimum, we plan to try Unrestricted NK and Ising Spin Glasses with different linkage models.
+This is critical as the behavior of many gray box algorithms depend on the nature of the epistasis graph. In these domains
+the focus will shift from finding the global optimum (which is unknown) to finding the best solution in a fixed amount of time.
+
+Beyond the randomly generated instances, we plan to seek out real world problems which can be expressed in this domain.
+This will help solidify the conclusions about the applicability of these methods outside of the optimization community.
+Similarly, we plan to compare the results of our gray box methods with algorithms specifically designed for each problem
+domain. This provides an interesting contention between the time required to develop problem specific solvers and the
+quality of using gray box methods.
+
+As the hamming ball hill climber has a user specified parameter $r$, we will need to explore this parameter space.
+Specifically we will want to ensure all comparisons use an appropriate value for $r$ which does not scale with
+$N$. This process will likely involve brute force testing all small $r$ values for moderately large problems in each class
+and then choosing the one which performs best. That $r$ value will then be used for all problem sizes for that problem class.
+
+Another area of focus will be in how $N$ effects the relative effectiveness of the methods. One of the goals of moving
+to a gray box domain is how it impacts Big-$O$ complexity. As a result we plan to empirically verify the scalability
+of each algorithm, as well as the wall clock time to find the global optimum. This is necessary as unlike the black box
+domain where evaluation is considered arbitrarily expensive, evaluation in the gray box domain is perfectly efficient.

### 2014-10-27 Comprehensive/chapter-comparison_algorithms.tex
Added references for Gray Box applications.

@@
 almost always creates near infinitesimal results, calculating fractions which include factorials of $\mu$
-and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{appendix-hboa}, we
+and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{chap-appendix-hboa}, we
 derived a simplified form shown in Equation~\ref{eq-hboa-final}. If no proposed split satisfies the inequality, iteration stops.

### 2014-10-27 Comprehensive/chapter-graybox.tex
Added references for Gray Box applications.

@@
 it is dominated (add to undominated size). Epistasis is the adjacency set for each
-vertex, plus itself. Used in ad hoc wireless sensor networks~\cite{todo-dominating-set}.
+vertex, plus itself. Used in wireless sensor networks~\cite{zhao:2013:dominatingset}.
 
@@
 Epistasis and fitness are calculated using each vertex (add to set size)
-and each edge (add to uncovered size), independently. Used in computation biology~\cite{todo-vertex-cover}.
+and each edge (add to uncovered size), independently. Used in network security~\cite{richter:2007:vertexcover}.
 
@@
 incident on exactly one vertex in the set. Epistasis and fitness are calculated
-using each edge independently. Used in VLSI design~\cite{todo-max-cut}.
+using each edge independently. Used in VLSI design~\cite{festa:2002:maxcut}.
 
@@
 and with each subset also contributing directly to fitness (add to set size). Used in
-computer vision~\cite{todo-set-cover}.
+computational biology~\cite{painsky:2014:setcover}.
 
@@
 creates an epistatic relationship between the variables in that constraint, and fitness
-is a linear combination of variables allowing for partial evaluation. Used in power systems~\cite{todo-linear-programming}.
+is a linear combination of variables allowing for partial evaluation. Used in power systems~\cite{arroyo:2000:linearprogram}.
 \end{itemize}

### 2014-10-27 CV/resume.tex
CV should now have all updated information.

@@
     \section{\mysidestyle Research\\Interests}
-    Evolutionary Computation, Genetic Programming, Artificial Intelligence, Machine Learning
+    Evolutionary Computation, Artificial Intelligence, Machine Learning
 
@@
         \item GPA: 4.0/4.0
-        \item Expected Graduation: May 2015
+        \item ABD, Expected Graduation: June 2015
     \end{list2}\vspace{-1.5mm}
@@
     \section{\mysidestyle Research\\Experience}
+    %----------------------P3-------------------
+    \textsl{Parameter-less Population Pyramid} \hfill \textbf{August 2013 -- Present}\\
+    Combined efficient local search with statistical linkage learning to create a
+    parameter-less black box optimization algorithm which outperforms optimally tuned
+    state-of-the-art competitors. Presented at GECCO 2014, winning Best Paper in Genetic
+    Algorithms Track. Submitted to Evolutionary Computation Journal.
+
+    \vspace{-2mm}
+    B. W. Goldman and W. F. Punch,
+    ``Parameter-less Population Pyramid,''
+    \textsl{In Proceedings of the 16th Annual Conference on Genetic and Evolutionary Computation}, pp. 785--792, 2014
 
@@
     optimization behavior, leading to the proposal of improvements.  Presented at EuroGP 2013
-    in Vienna Austria, GECCO 2013 in Amsterdam, The Netherlands.  Revised and resubmitted to
-    IEEE Transactions on Evolutionary Computation, awaiting final review.
+    in Vienna Austria, GECCO 2013 in Amsterdam, The Netherlands.
+
+    \vspace{-2mm}
+    B. W. Goldman and W. F. Punch,
+    ``Analysis of Cartesian Genetic Programming's Evolutionary Mechanisms,''
+    \textsl{IEEE Transactions on Evolutionary Computation}, In Press
 
@@
     \textsl{In Proceedings of the 16th European Conference on Genetic Programming}, pp. 61--72, 2012
+
+    %--------------------Avida---------------------------------------
+    \textsl{Artificial Life} \hfill \textbf{March 2013 -- March 2014}\\
+    Used Avida, a widely acclaimed platform for digital evolution, to study the evolution
+    of prey mimicry of poisonous species when under predation.
+
+    \vspace{-2mm}
+    K. D. S. Lehmann,  B. W. Goldman, I. Dworkin, D. M. Bryson, A. P. Wagner,
+    ``From Cues to Signals: Evolution of Interspecific Communication via Aposematism and Mimicry in a Predator-Prey System,''
+    \textsl{PloS one}, vol. 9, no. 3, e91783, 2014.
+
+    \pagebreak
 
@@
 
-    \pagebreak
-
     \textsl{Self-Configuring Crossover} \hfill \textbf{January 2010 -- June 2011}\\
@@
 
-    \textsl{Extending Self-Configuration} \hfill \textbf{August 2012 -- present}\\
+    \textsl{Extending Self-Configuration} \hfill \textbf{August 2012 -- May 2014}\\
     Mentoring undergraduate student research project investigating ways to improve evolutionary
@@
     \textbf{Michigan State University}, East Lansing, Michigan \vspace{2mm}\\\vspace{1mm}%
+    \textsl{CSE232: Introduction to Programming II} \hfill \textbf{May 2014 -- July 2014}\\
+    Lead instructor for course, in charge of lecturing, creating course content, and assessments.
+    Topics include the C++ programming language up to dynamic memory, templating, and user created data structures.
+
     \textsl{CSE232: Introduction to Programming II} \hfill \textbf{August 2013 -- May 2014}\\
@@
     topics include the C++ programming language up to dynamic memory, templating, and user created data structures.
+
+    \pagebreak
 
@@
     \section{\mysidestyle Honors and\\Activities}
+    ``Best Paper - Genentic Algorithms Track'' for GECCO 2014 ``Parameter-less Population Pyramid''\\% 
     Program Committee Member, GECCO 2012--present Genetic Algorithms Track \vspace{1mm}\\%

### 2014-10-27 CV/resume.tex
IEEE reviewer

@@
     ``Best Paper - Genentic Algorithms Track'' for GECCO 2014 ``Parameter-less Population Pyramid''\\%
+    IEEE Transactions on Evolutionary Computation Peer Reviewer\\%
     Program Committee Member, GECCO 2012--present Genetic Algorithms Track \vspace{1mm}\\%

### 2014-10-27 CV/cv.tex
Added "cv" in new formatting.

--- 2014-10-27 CV/resume.tex
@@
     % Research
-    \section{\mysidestyle Research\\Experience}
-    %----------------------P3-------------------
-    \textsl{Parameter-less Population Pyramid} \hfill \textbf{August 2013 -- Present}\\
-    Combined efficient local search with statistical linkage learning to create a
-    parameter-less black box optimization algorithm which outperforms optimally tuned
-    state-of-the-art competitors. Presented at GECCO 2014, winning Best Paper in Genetic
-    Algorithms Track. Submitted to Evolutionary Computation Journal.
+    \section{\mysidestyle Peer Reviewed\\Journal Articles}
 
@@
     B. W. Goldman and W. F. Punch,
-    ``Parameter-less Population Pyramid,''
-    \textsl{In Proceedings of the 16th Annual Conference on Genetic and Evolutionary Computation}, pp. 785--792, 2014
-
-    %----------------------CGP------------------
-    \textsl{Cartesian Genetic Programming} \hfill \textbf{August 2012 -- August 2013}\\
-    Examined how representation and evolutionary operators interacted to cause degenerate
-    optimization behavior, leading to the proposal of improvements.  Presented at EuroGP 2013
-    in Vienna Austria, GECCO 2013 in Amsterdam, The Netherlands.
+    ``Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid,''
+    \textsl{Evolutionary Computation Journal}, In Review
 
@@
     \vspace{-2mm}
-    B. W. Goldman and W. F. Punch,
-    ``Length Bias and Search Limitations in Cartesian Genetic Programming,''
-    \textsl{In Proceedings of the 15th Annual Conference on Genetic and Evolutionary Computation}, pp. 933--940, 2013
-
-    \vspace{-2mm}
-    B. W. Goldman and W. F. Punch,
-    ``Reducing Wasted Evaluations in  Cartesian Genetic Programming,''
-    \textsl{In Proceedings of the 16th European Conference on Genetic Programming}, pp. 61--72, 2012
-
-    %--------------------Avida---------------------------------------
-    \textsl{Artificial Life} \hfill \textbf{March 2013 -- March 2014}\\
-    Used Avida, a widely acclaimed platform for digital evolution, to study the evolution
-    of prey mimicry of poisonous species when under predation.
-
-    \vspace{-2mm}
     K. D. S. Lehmann,  B. W. Goldman, I. Dworkin, D. M. Bryson, A. P. Wagner,
@@
     \textsl{PloS one}, vol. 9, no. 3, e91783, 2014.
-
-    \pagebreak
-
-    %---------------------Benchmarks--------------------------------
-    \textsl{Improving Genetic Programming Research} \hfill \textbf{August 2012 -- December 2012}\\
-    Utilizing a survey of genetic programming researches and in honor of the fields twentieth anniversary,
-    proposed consensus suggestions for how to raise the quality of future publications.
 
@@
 
-    %----------------------LTGA-------------------------------
-    \textsl{Linkage Tree Genetic Algorithm} \hfill \textbf{January 2012 -- May 2012}\\
-    Performed a deep analysis of the multiple variations of this recently created automatic linkage
-    discovery technique, with suggestions for improved methods of testing. Presented at GECCO 2012 in Philadelphia, Pennsylvania.
+    \section{\mysidestyle Peer Reviewed\\Conference Papers}
+    \vspace{-2mm}
+    B. W. Goldman and W. F. Punch,
+    ``Parameter-less Population Pyramid,''
+    \textsl{GECCO}, pp. 785--792, 2014
+    \textbf{Won Best Paper in Track}
+
+    \vspace{-2mm}
+    B. W. Goldman and W. F. Punch,
+    ``Length Bias and Search Limitations in Cartesian Genetic Programming,''
+    \textsl{GECCO}, pp. 933--940, 2013
+
+    \vspace{-2mm}
+    N. R. Kamrath, B. W. Goldman, and D. R. Tauritz,
+    ``Using Supportive Coevolution to Evolve Self-Configuring Crossover,''
+    \textsl{GECCO}, pp. 1489--1496, 2013.
+
+    \vspace{-2mm}
+    B. W. Goldman and W. F. Punch,
+    ``Reducing Wasted Evaluations in  Cartesian Genetic Programming,''
+    \textsl{EuroGP}, pp. 61--72, 2013
 
@@
     ``Linkage tree genetic algorithms: variants and analysis,''
-    \textsl{In Proceedings of the 14th Annual Conference on Genetic and Evolutionary Computation}, pp. 625--632, 2012
-
-    \textsl{Supportive Coevolution} \hfill \textbf{January 2011 -- May 2012}\\
-    A new form of coevolution in which the primary species directly encodes prospective problem solutions,
-    and is supported by other species tasked with dynamically evolving better ways to evolve the primary species.
-    Presented at GECCO 2012 in Philadelphia, Pennsylvania.
+    \textsl{GECCO}, pp. 625--632, 2012
 
@@
     ``Supportive Coevolution,''
-    \textsl{In Proceedings of the 14th Annual Conference Companion on Genetic and Evolutionary Computation}, pp. 59--66, 2012.
-
-    \textsl{Self-Configuring Crossover} \hfill \textbf{January 2010 -- June 2011}\\
-    Dynamically evolving novel crossover operators during evolutionary algorithm optimization 
-    using linear genetic programming and self-adaptation.  Presented at GECCO 2011 in Dublin, Ireland.
+    \textsl{GECCO}, pp. 59--66, 2012.
 
@@
     ``Self-Configuring Crossover,''
-    \textsl{In Proceedings of the 13th Annual Conference Companion on Genetic and Evolutionary Computation}, pp. 575--582, 2011.
+    \textsl{GECO}, pp. 575--582, 2011.
 
-    \textsl{Dynamic Parameters} \hfill \textbf{January 2010 -- February 2011}\\
-    Dynamically changing evolutionary algorithm parameters during run time can achieve better solution quality
-    than optimally set static parameters for a wide variety of parameters.
-    Presented at GECCO 2011 in Dublin, Ireland.
-
+    \section{\mysidestyle Peer Reviewed\\Poster Papers}
     \vspace{-2mm}    
@@
     ``Meta-Evolved Empirical Evidence of the Effectiveness of Dynamic Parameters'',
-    \textsl{In Proceedings of the 13th Annual Conference Companion on Genetic and Evolutionary Computation}, pp. 155--156, 2011.
+    \textsl{GECCO}, pp. 155--156, 2011.
 
-    %__________________________________________________________________________________________________________________
-    % Mentoring Experience
-    \section{\mysidestyle Mentoring\\Experience}
-
-    \textsl{Extending Self-Configuration} \hfill \textbf{August 2012 -- May 2014}\\
-    Mentoring undergraduate student research project investigating ways to improve evolutionary
-    self configuration and expanding to new operators.
-
-    \vspace{-2mm}
-    N. R. Kamrath, B. W. Goldman, and D. R. Tauritz,
-    ``Using Supportive Coevolution to Evolve Self-Configuring Crossover,''
-    \textsl{In Proceedings of the 15th Annual Conference Companion on Genetic and Evolutionary Computation}, pp. 1489--1496, 2013.
+    \pagebreak
 
@@
     % Teaching Experience
-    \section{\mysidestyle Selected\\Teaching\\Experience}
+    \section{\mysidestyle Teaching\\Experience}
     \textbf{Michigan State University}, East Lansing, Michigan \vspace{2mm}\\\vspace{1mm}%
@@
 
-    \pagebreak
-
     \textbf{Missouri S\&T}, Rolla, Missouri \vspace{2mm}\\\vspace{1mm}%
-    \begin{comment}
-    \textsl{CS328: Object Oriented Numerical Methods} \hfill \textbf{January 2012 -- present}\\
+    \textsl{CS328: Object Oriented Numerical Methods} \hfill \textbf{January 2012 -- May 2012}\\
     Grader for intensive C++ course,
     topics include advanced programming techniques and how to efficiently solve large systems of equations.
-    \end{comment}
+
     \textsl{CS54: Introduction to C++ Lab} \hfill \textbf{January 2011 -- December 2011}\\
@@
     topics include programming basics up to inheritance, style, editors, and how to interact with Unix.
-    \begin{comment}
+
     \textsl{CS387: Parallel Computing} \hfill \textbf{January 2011 -- May 2011}\\
@@
     topics include MPI, cluster computing, GPUs, Monte Carlo experiments, and high performance computing.
-    \end{comment}
 
@@
     Voted 2012 Leader of the Year, Missouri S\&T Computer Science Department \vspace{1mm}\\%
-    %Chair of ACM SIG for AI competition game development (SIG-Game), 2011--present \vspace{1mm}\\%
+    %Chair of ACM SIG for AI competition game development (SIG-Game), 2011--2012 \vspace{1mm}\\%
     Google AI challenge, ranked 25$^{th}$ out of 4619 world wide, 6$^{th}$ in USA, 2010 \vspace{1mm}\\%
@@
 \begin{comment}    
-    ACM SIG-Game Competition testing and balancing using evolving AI, 2009--present \vspace{1mm}\\%
+    ACM SIG-Game Competition testing and balancing using evolving AI, 2009--2012 \vspace{1mm}\\%
     ACM SIG-Game first place AI, 2008          \vspace{1mm}\\%

### 2014-10-28 Comprehensive/abstract.tex
Comprehensive Chapters 7 and 8 first drafts.
Updated Comprehensive with Bills first suggestions

@@
 Brian W. Goldman\\ %
-\ \\
+%\ \\
 \end{centering}
@@
 
+We propose to extend P3 to the ``gray box'' problem domain, which includes many
+interesting real world problems and allows for a number of increases in search efficiency.
+Specifically, local search becomes a factor of problem size faster and more linkage information
+is accessible to crossover. We plan to develop a P3 variant to exploit these
+advantages and then compare it with state of the art black box, gray box, and
+white box methods for solving challenging real world problems.
 \end{doublespace}

### 2014-10-28 Comprehensive/chapter-proposed.tex
Updated Comprehensive with Bills first suggestions

@@
 problems, hamming ball hill climbing is not sufficient to find the global optimum. When applied to 50 instances
-of $N=200$, $K=5$, $r=4$, and limited to using less time than the slowest P3 run to solve that class (24 minutes),
+of Nearest Neighbor NK using $N=200$, $K=5$, $r=4$, and limited to using less time than the slowest P3 run to solve that class (24 minutes),
 the hamming ball hill climber only found the global optimum 6 times.
@@
 The computational cost of this algorithm is $O(|cluster|)$ for each $cluster$ created.
-Assuming that all variables participate in at least one subfunction, this algorithm will create
-$\Omega(N)$ clusters. This is a safe assumption as in the gray box domain variables that do not
-effect fitness have no reason to appear in the solution.
+Assuming that the problem is not separable, this algorithm will create
+$\Omega(N)$ clusters. This is a safe assumption in the gray box domain as it is more
+efficient to solve each component independently.
 In order to complete,
@@
 
-This methods has many desirable properties beyond efficiency. All crossovers always move at least one entire subfunctions,
+This methods has many desirable properties beyond efficiency. All crossovers always move at least one entire subfunction,
 and clusters always differ by at least one entire subfunction. The probability two variables are linked in a
@@
 %0.0939752, 0.154129, 0.392246, 0.721644, 0.110155, 0.241687, 0.321031, 0.172651, 0.567039, 0.480878, 0.109734, 0.176248, 0.343879, 0.301121, 0.202199, 0.17966, 0.0330468, 0.117303, 0.0885791, 0.0588664, 0.594459, 0.101318, 1.86743, 0.0350668, 0.598237, 0.388794, 0.208924, 0.240061, 0.103984, 0.21943, 0.238308, 1.10106, 0.142684, 0.332108, 0.336863, 0.49791, 0.119978, 0.155924, 0.243874, 0.142354, 0.371339, 0.0854216, 0.168894, 0.419617, 0.128611, 1.9637, 0.288967, 0.142157, 0.103391, 0.307311
-On Nearest Neighbor NK
+On Nearest Neighbor NK using
 $N=200$, $K=5$, $r=4$ this found the global optimum in all 50 runs, with a median time to success of 0.23 minutes. This makes it the
@@
 \section{Comparison of Gray Box Methods}
-Key comparisons we plan to examine:
-\begin{itemize}
-\item How does the combination of uniform crossover with hamming ball hill climbing compare to hill climbing alone?
-\item Does preserving subfunctions intact during crossover improve search efficiency?
-\item How does black box P3 compare with gray box P3?
-\item How do the more generic gray box methods compare with problem specific methods for solving real world problems?
-\end{itemize}
 
@@
 domain where evaluation is considered arbitrarily expensive, evaluation in the gray box domain is perfectly efficient.
+
+\section{Summary of Proposed Research Questions}
+\begin{itemize}
+\item How does the combination of uniform crossover with hamming ball hill climbing compare to hill climbing alone?
+\item Does preserving subfunctions intact during crossover improve search efficiency?
+\item How does black box P3 compare with gray box P3?
+\item How do the more generic gray box methods compare with problem specific methods for solving real world problems?
+\end{itemize}

### 2014-10-28 Comprehensive/abstract.tex
Abstract rework.

--- 2014-10-15 Comprehensive/abstract.tex
@@
 
+We propose to extend this research to include the ``gray box'' problem domain, which includes many
+interesting real world problems. We plan to develop a variant of P3 which exploits
+features of this domain, such as efficient local search and known variable linkage.
+Furthermore, we want to examine how ``hamming ball search'' can be integrated into
+global optimization, both through simple crossover and as part of P3.
+Finally, we plan compare this P3 variant with state of the art black box, gray box, and
+white box methods for solving challenging real world problems.
 \end{doublespace}

### 2014-11-17 CV/cv.tex
Updated CV

@@
-%TODO Add: Voted Leader of the year
 %______________________________________________________________________________________________________________________
@@
 
-    \vspace{-2mm}
+    \begin{enumerate}
+    \item
     B. W. Goldman and W. F. Punch,
     ``Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid,''
-    \textsl{Evolutionary Computation Journal}, In Review
+    \textbf{Invited Paper}, \textsl{Evolutionary Computation Journal}, In Review
 
-    \vspace{-2mm}
+    \item
     B. W. Goldman and W. F. Punch,
@@
 
-    \vspace{-2mm}
+    \item
     K. D. S. Lehmann,  B. W. Goldman, I. Dworkin, D. M. Bryson, A. P. Wagner,
@@
 
-    \vspace{-2mm}
+    \item
     D. R. White, J. McDermott, M. Castelli, L. Manzoni, B. W. Goldman,
@@
     \textsl{Genetic Programming and Evolvable Machines}, vol. 14, no. 1, pp. 3--29, 2013.    
+    \end{enumerate}
 
     \section{\mysidestyle Peer Reviewed\\Conference Papers}
-    \vspace{-2mm}
+    \begin{enumerate}
+
+    \item
     B. W. Goldman and W. F. Punch,
@@
 
-    \vspace{-2mm}
+    \item
     B. W. Goldman and W. F. Punch,
@@
 
-    \vspace{-2mm}
+    \item
     N. R. Kamrath, B. W. Goldman, and D. R. Tauritz,
@@
 
-    \vspace{-2mm}
+    \item
     B. W. Goldman and D. R. Tauritz,
@@
     \textsl{GECO}, pp. 575--582, 2011.
-
+    \end{enumerate}
     \section{\mysidestyle Peer Reviewed\\Poster Papers}
-    \vspace{-2mm}    
+    \begin{enumerate}
+    \item
     B. W. Goldman and D. R. Tauritz,
@@
     \textsl{GECCO}, pp. 155--156, 2011.
-
+    \end{enumerate}
     \pagebreak
@@
     \textsl{CSE232: Introduction to Programming II} \hfill \textbf{May 2014 -- July 2014}\\
-    Lead instructor for course, in charge of lecturing, creating course content, and assessments.
-    Topics include the C++ programming language up to dynamic memory, templating, and user created data structures.
+    Lead instructor for course. In charge of lecturing, creating course content, and assessments
+    for 37 students. Managed two graduate assistants who taught lab sections and graded assignments.
+    Received 3.81/4 student evaluation.
+    Topics include the C++ programming language from basics up to dynamic memory, templating, and user created data structures.
 
@@
     Primary instructor for lab section,
-    topics include the C++ programming language up to dynamic memory, templating, and user created data structures.
+    topics include the C++ programming language from basics up to dynamic memory, templating, and user created data structures.
 
@@
     \section{\mysidestyle Honors and\\Activities}
-    ``Best Paper - Genentic Algorithms Track'' for GECCO 2014 ``Parameter-less Population Pyramid''\\%
-    IEEE Transactions on Evolutionary Computation Peer Reviewer\\%
+    Best Paper - Genentic Algorithms Track for GECCO 2014 ``Parameter-less Population Pyramid''\\%
+    Peer Reviewer, IEEE Transactions on Evolutionary Computation Journal 2014--present\\%
+    Founder and organizer, Computational Evolution Discussion Group October 2013--present\\%
     Program Committee Member, GECCO 2012--present Genetic Algorithms Track \vspace{1mm}\\%

### 2014-11-19 CV/cv.tex
Added Optimization to CV research interests, changed order
Added more activities to CV

@@
     \section{\mysidestyle Research\\Interests}
-    Evolutionary Computation, Artificial Intelligence, Machine Learning
+    Optimization, Artificial Intelligence, Evolutionary Computation, Machine Learning
 
@@
     Best Paper - Genentic Algorithms Track for GECCO 2014 ``Parameter-less Population Pyramid''\\%
+    Co-chair, GECCO 2015 Combinatorial Black Box Optimization Competition\\%
     Peer Reviewer, IEEE Transactions on Evolutionary Computation Journal 2014--present\\%
@@
     Voted 2012 Leader of the Year, Missouri S\&T Computer Science Department \vspace{1mm}\\%
-    %Chair of ACM SIG for AI competition game development (SIG-Game), 2011--2012 \vspace{1mm}\\%
+    Chair of ACM SIG for AI competition game development (SIG-Game), 2011--2012 \vspace{1mm}\\%
     Google AI challenge, ranked 25$^{th}$ out of 4619 world wide, 6$^{th}$ in USA, 2010 \vspace{1mm}\\%

### 2014-11-21 CV/cv.tex
Impact factors and acceptance rates

@@
     ``Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid,''
-    \textbf{Invited Paper}, \textsl{Evolutionary Computation Journal}, In Review
+    \textbf{Invited Paper}, \textsl{Evolutionary Computation Journal} (Impact Factor 3.733), In Review
 
@@
     ``Analysis of Cartesian Genetic Programming's Evolutionary Mechanisms,''
-    \textsl{IEEE Transactions on Evolutionary Computation}, In Press
+    \textsl{IEEE Transactions on Evolutionary Computation} (Impact Factor 5.545), In Press
 
@@
     ``From Cues to Signals: Evolution of Interspecific Communication via Aposematism and Mimicry in a Predator-Prey System,''
-    \textsl{PloS one}, vol. 9, no. 3, e91783, 2014.
+    \textsl{PloS one} (Impact Factor 3.534), vol. 9, no. 3, e91783, 2014.
 
@@
     ``Parameter-less Population Pyramid,''
-    \textsl{GECCO}, pp. 785--792, 2014
+    \textsl{GECCO} (Acceptance Rate 33\%), pp. 785--792, 2014
     \textbf{Won Best Paper in Track}
@@
     ``Length Bias and Search Limitations in Cartesian Genetic Programming,''
-    \textsl{GECCO}, pp. 933--940, 2013
+    \textsl{GECCO} (Acceptance Rate 36\%), pp. 933--940, 2013
 
@@
     ``Using Supportive Coevolution to Evolve Self-Configuring Crossover,''
-    \textsl{GECCO}, pp. 1489--1496, 2013.
+    \textsl{GECCO} (Acceptance Rate 36\%), pp. 1489--1496, 2013.
 
@@
     ``Reducing Wasted Evaluations in  Cartesian Genetic Programming,''
-    \textsl{EuroGP}, pp. 61--72, 2013
+    \textsl{EuroGP} (Acceptance Rate 38\%), pp. 61--72, 2013
 
@@
     ``Linkage tree genetic algorithms: variants and analysis,''
-    \textsl{GECCO}, pp. 625--632, 2012
+    \textsl{GECCO} (Acceptance Rate 37\%), pp. 625--632, 2012
 
@@
     ``Supportive Coevolution,''
-    \textsl{GECCO}, pp. 59--66, 2012.
+    \textsl{GECCO} (Acceptance Rate 37\%), pp. 59--66, 2012.
 
@@
     ``Self-Configuring Crossover,''
-    \textsl{GECO}, pp. 575--582, 2011.
+    \textsl{GECCO} (Acceptance Rate 38\%), pp. 575--582, 2011.
     \end{enumerate}
+
+    \pagebreak
+
     \section{\mysidestyle Peer Reviewed\\Poster Papers}
@@
     \end{enumerate}
-    \pagebreak
 
@@
     Peer Reviewer, IEEE Transactions on Evolutionary Computation Journal 2014--present\\%
+    Peer Reviewer, Genetic Programming and Evolvable Machines 2014--present\\%
     Founder and organizer, Computational Evolution Discussion Group October 2013--present\\%

### 2015-01-07 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
Started Gray Box P3 paper

\documentclass{sig-alternate}
%\linespread{2.5}
\usepackage{url}
\usepackage{verbatim}
\usepackage[noend]{algpseudocode}
\newcommand{\includegraphicsfit}[1]
{\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}

\newcommand{\includegraphicswide}[1]
{\includegraphics[width=.9\textwidth,height=\textheight,keepaspectratio]{#1}}

\newcommand{\BigO}[1]{$\mathcal{O}{(#1)}$}

\usepackage{bm}

\newfont{\mycrnotice}{ptmr8t at 7pt}
\newfont{\myconfname}{ptmri8t at 7pt}
\let\crnotice\mycrnotice%
\let\confname\myconfname%

\permission{Permission to make digital or hard copies of all or part of this work
for personal or classroom use is granted without fee provided that copies are not
made or distributed for profit or commercial advantage and that copies bear this
notice and the full citation on the first page. Copyrights for components of this
work owned by others than ACM must be honored. Abstracting with credit is permitted.
To copy otherwise, or republish, to post on servers or to redistribute to lists,
requires prior specific permission and/or a fee. Request permissions from permissions@acm.org.}
\conferenceinfo{GECCO'15,} {July 11-15, 2015, Madrid, Spain.}
\CopyrightYear{2015}
\crdata{TBA}
\clubpenalty=10000
\widowpenalty = 10000

\begin{document}
\title{Gray Box Optimization using the Parameter-less Population Pyramid}
%\subtitle{[Genetic Algorithms Track]}

\numberofauthors{2} %
\begin{comment}
\author{
% 1st. author
\alignauthor
Brian W. Goldman\\
       \affaddr{BEACON Center for the Study of Evolution in Action}\\
       \affaddr{Michigan State University, U.S.A.}\\
       \email{brianwgoldman@acm.org}
% 2nd. author
\alignauthor
William F. Punch\\
       \affaddr{BEACON Center for the Study of Evolution in Action}\\
       \affaddr{Michigan State University, U.S.A.}\\
       \email{punch@msu.edu}
}
%\end{comment}
%\begin{comment}
\author{
% 1st. author
\alignauthor
Anonymous\\
       \affaddr{Group}\\
       \affaddr{Group}\\
       \affaddr{Organization}\\
       \email{email@site.com}
% 2nd. author
\alignauthor
Anonymous\\
       \affaddr{Group}\\
       \affaddr{Group}\\
       \affaddr{Organization}\\
       \email{email@site.com}
}
%\end{comment}

\maketitle
\begin{abstract}
%Maximum 200 Words
TODO
\end{abstract}

% A category with the (minimum) three required fields
%\category{Computing Methodologies}{Artificial Intelligence}{Search Methodologies}
\category{I.2.8}{Artificial Intelligence}{Problem Solving, Control Methods, and Search}
%A category including the fourth, optional field follows...
%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

\terms{Algorithms, Performance, Experimentation}

\keywords{TODO; Local Search; Parameter-less}

\section{Introduction}
~\cite{chen:2011:memetic}

~\cite{whitley:2013:greedy}~\cite{chicano:2014:ball}

~\cite{goldman:2014:p3}

\section{Gray Box Domain}
\label{sec-gray-box}
In black box optimization, the only information available is a specification
for what constitutes a valid solution and a function which returns the quality
of a given solution. While these very simple requirements allow almost any
optimization task to fit into this domain, it may prevent additional knowledge
about a problem from being provided to search algorithms. At the other end of
the spectrum is white box optimization, such that everything about the problem
is known allowing search techniques to specialize at the cost of generality.
In between these two extremes is gray box optimization, such that some additional
information about the problem is known, but the interface is still general enough
to allow optimization algorithms to solve a variety of problems.

Here we shall consider the gray box optimization domain with the following
characteristics. The quality of a solution is determined by the summation
of subfunctions. Each subfunction uses at most $k$ variables in the solution,
such that as the total number of solution variables $N$ increases, $k$ remains constant.
Each subfunction must be independently evaluable, and the variables each uses is known.

While more restrictive than black box optimization, this domain is very general.
MAX-SAT, and by extension all NP-hard problems, can be represented
as a summation of subfunctions to evaluate each clause~\cite{whitley:2013:greedy}.
Previous approximation studies of Vertex Cover~\cite{oliveto:2009:vertexcover},
Set Cover~\cite{yu:2010:setcover}, MAX-CUT~\cite{festa:2002:maxcut}, and
Ising Spin Glasses~\cite{pelikan:2003:hboaising} all work with problems which fit
in the gray box domain. These relate back to problems of real world interest such
as network security, computational biology, VLSI design, and statistical physics.

In order to obtain efficiency bounds for optimization algorithms, the following
assertions are included. The total number of subfunctions grows no more than
linearly with $N$, which holds true for randomized MAX-SAT, mesh spin glasses, and NK Landscapes.
Furthermore, each variable must participate in at most $c$ subfunctions, which holds true
for mesh spin glasses and Nearest Neighbor NK. If violated these assertions do not
necessarily prevent the gray box optimization algorithms from solving the problem, but may result
in slower than predicted performance.

\section{Hamming Ball Hill Climbing}
The first algorithm to rigorously exploit the features of the gray box domain
described in Section~\ref{sec-gray-box} was \cite{whitley:2013:greedy}. In this study
a hill climber was developed which found the approximate next best improving move in \BigO{1} time.
This is achieved by first determining the fitness effect $delta$ of making each move, and then updating
only effected move each time a change to the solution is made. Due to the requirements of the domain,
at most \BigO{ck} moves can be effected, with each requiring at most \BigO{1} subfunction calls
to update their $delta$. Combined with move binning, this technique is able to perform both
first improvement and next best hill climbing in \BigO{I+N}, where $I$ is the number of improving
moves required to reach a local optimum. In comparison, the black box approach for these searches can require
\BigO{IN^2}.

By again leveraging the properties of this domain, \cite{chicano:2014:ball} extended their
local search algorithm to find $r$-bit local optima. Termed the Hamming Ball Hill Climber,
this method efficiently searches the space of all moves which flip up to $r$ bits in the
solution. While na\"ively this create \BigO{N^r} moves, in the gray box domain only
\BigO{(3ck)^rN} must be checked. This is due to the bounded relationships between variables.
Consider that if $i$ and $j$ do not share a subfunction the effect of flipping both variables
must be equal to the sum of flipping both separately. Therefore only variables which share
a subfunction must be flipped together when checking $r=2$. The problem of determining
which variables must be flipped together is equivalent to finding the connected induced
subgraphs of $G$ with $r$ or fewer vertices, such that each vertex in $G$ is a variable
and an edge exists between two vertices if and only if those variables share a subfunction.
In \cite{chicano:2014:ball} an upper bound is derived for how many moves this could create,
but no algorithm was given to efficiently discover these moves.

Under the assertion that $r \ll N$ and that $r$ does not increase with $N$, the Hamming
Ball Hill Climber requires only \BigO{I+N} time to find $r$-bit local optimum starting
from a random solution. As $I$ is likely in \BigO{N} this means finding random $r$-bit local optima
is no more than a constant amount slower than generating random solutions.

\subsection{Novel r-order Subgraphs Algorithm}
\begin{figure}
  \begin{algorithmic}[1]
  \Procedure{CISG}{$v$, $subset$, $closed$, $open$}
    \State $subset' \leftarrow subset \cup \{v\}$
    \State $found \leftarrow [subset']$
    \If{$|subset'| \geq r$}
      \Return $found$
    \EndIf
    \State $closed\_here \leftarrow \emptyset$
    \State $open' \leftarrow open \cup adjacent(v)$
    \ForAll{$v' \in open'$ such that $v' \notin closed$}
        \State $closed\_here \leftarrow closed\_here \cup \{v'\}$
        \State $closed \leftarrow closed \cup \{v'\}$
        \State $recurse \leftarrow $\Call{CISG}{$v'$, $subset'$, $closed$, $open'$}
        \State $found \leftarrow found + recurse$
    \EndFor
    \State $closed \leftarrow closed - closed\_here$
    \State \Return $found$
  \EndProcedure
\end{algorithmic}
  \caption{Algorithm to recursively find all connected induced subgraphs of size $r$ or fewer which
           contain $subset \cup \{v\}$ and no additional vertices in $closed$.}
  \label{fig-connected-subgraphs}
\end{figure}

In order to determine which moves in the hamming ball must be evaluated,
we developed \Call{CISG}{} given in Figure~\ref{fig-connected-subgraphs}
to discover all Connected Induced Subgraphs of $r$ or fewer vertices.
\Call{CISG}{} is a recursive algorithm which finds all subgraphs
which contain a given $subset$ and a given vertex $v$, while excluding
any other vertices added to $closed$. To find all subgraphs, \Call{CISG}{}
is called once for each vertex in the graph, such that $closed$ contains
all previously searched vertices, and $subset=open=\emptyset$. In the initial
call all desired subgraphs which contain $v$ are found, which is why subsequent
top level calls include $v$ in $closed$ to prevent duplicate subgraphs from being
returned.

At each recursive level \Call{CISG}{} expands $open$ to include any vertices
adjacent to $v$ in the graph. By construction this means that $open$ contains
all possible ways of adding a single vertex to the current $subset$. As $v'$
are tested they are temporarily added to $closed$ to prevent recursive calls
from creating duplicates.

When applied to the sparse graphs inherent in the gray box domain, this algorithm
requires \BigO{r!(c(k-1))^rN} time, which reduces to \BigO{N}. The time spent
in each call is dominated by the loop over $open'$. In the worst case, $open'$
increases in size by the full adjacency of $v$, which is bounded by $c(k-1)$.
This creates a worse case complexity for a single top level call of
$\prod_{i}^{r} ic(k-1) = r!(c(k-1))^r$. This must be called once for each
of the $N$ variables resulting in \BigO{r!(c(k-1))^rN}.

\section{Tournament Uniform Crossover: TUX}
\begin{figure}
  \begin{algorithmic}[1]
  \Procedure{Iterate-TUX}{}
    \State Create random solution
    \State Apply hamming ball hill climber to solution
    \ForAll{$T_i \in T$}
      \If{$T_i$ is empty}
        \State $T_i \leftarrow$ solution
        \State \Return
      \EndIf
      \State Cross solution with $T_i$ to create $2^i$ offspring
      \State Apply hamming ball hill climber to each offspring
      \State solution $\leftarrow$ best of offspring, solution, and $T_i$
      \State $T_i \leftarrow$ empty
    \EndFor
    \State Add solution to end of $T$
  \EndProcedure
\end{algorithmic}
  \caption{One iteration of TUX optimization. $T$ is an
           ordered list of solution, each of which could be empty,
           awaiting a crossover partner.}
  \label{fig-TUX}
\end{figure}

\section{Parameter-less Population Pyramid: P3}

\subsection{Origins}
~\cite{goldman:2014:p3}
\begin{figure}
  \begin{algorithmic}
  \Procedure{Iterate-P3}{}
    \State Create random solution
    \State Apply hill climber
    \If{solution $\notin hashset$}
      \State Add solution to $P_0$
      \State Add solution to $hashset$
    \EndIf

    \ForAll{$P_i \in pyramid$}
      \State Mix solution with $P_i$
      \If{solution's fitness has improved}
        \If{solution $\notin hashset$}
          \State Add solution to $P_{i+1}$
          \State Add solution to $hashset$
        \EndIf
      \EndIf
    \EndFor
  \EndProcedure
\end{algorithmic}
  \caption{One iteration of P3 optimization. $pyramid$ is an
           ordered set of populations and $hashset$ is a set
           of all solutions in $pyramid$.}
  \label{fig-p3}
\end{figure}

\subsection{Gray Box Specialization}
\begin{figure}
  \begin{algorithmic}[1]
  \Procedure{SubfunctionTree}{}
    \State $clusters \leftarrow [\{0\}, \{1\}, \{2\}, \dots, \{N-1\}]$
    \State $cluster\_number \leftarrow [0 .. N-1]$
    \ForAll{$subfunction \in shuffled(subfunctions)$}
      \State $to\_merge \leftarrow \emptyset$
      \ForAll{$b \in subfunction$}
        \State $to\_merge \leftarrow to\_merge \cup \{cluster\_number[b]\}$
      \EndFor
      \If{$|to\_merge| > 1$}
        \State $new\_cluster \leftarrow \emptyset$
        \ForAll{$i \in to\_merge$}
          \State $new\_cluster \leftarrow new\_cluster \cup clusters[i]$
        \EndFor
        \ForAll{$b \in new\_cluster$}
          \State $cluster\_number[b] \leftarrow |clusters|$
        \EndFor
        \State $clusters \leftarrow clusters + new\_cluster$
      \EndIf
    \EndFor
    \State Remove first $N~clusters$
    \State Remove any cluster containing all variables
  \EndProcedure
\end{algorithmic}
  \caption{Algorithm used to convert a list of subfunctions into linkage tree clusters.}
  \label{fig-sfx-tree}
\end{figure}

TODO State that unlike BBP3, GBP3 performs hill climbing after each donation
\subsection{Time Complexity}

\section{Experimentation}
~\cite{wright:2000:solvingnk}

\subsection{The Effect of Radius}
TODO Plot for single K value with x=radius, y=fitness, color=solver

TODO Plot for just Pyramid with x=radius, y=seconds to optimum, color=k

TODO Explain k==1

\subsection{Fitness Over Time}
TODO Plot for fixed N, K, and R with x=seconds, y=fitness, color=solver

TODO Explain best Pyramid R versus best TUX R.

\subsection{Scalability}
TODO Plot for seconds to global optimum as N increases, include Black Box P3

TODO Consider O(N) statistics

\section{Conclusions and Future Work}
%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\bibliography{../main}  % sigproc.bib is the name of the Bibliography in this case
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references
%
% ACM needs 'a single self-contained file'!
%
%APPENDICES are optional
\balancecolumns
% That's all folks!
\end{document}

### 2015-01-08 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
More Gray Box P3

@@
 in slower than predicted performance.
+We also include the trivial requirements that all problem variables must participate
+in at least one subfunction and that the problem is not separable. Any problem which
+violates these requirements can be restated either using less variables or as multiple separate problems.
 
@@
 In \cite{chicano:2014:ball} an upper bound is derived for how many moves this could create,
-but no algorithm was given to efficiently discover these moves.
+but no algorithm was given to efficiently discover these moves. This process was not included
+in their timing calculations has it only has to be performed once per problem.
 
@@
 is no more than a constant amount slower than generating random solutions.
+
+While these techniques make it possible to choose the best improving move just as efficiently
+as choosing any random improving move, the findings of~\cite{whitley:2013:greedy} suggest that,
+especially when paired with subsequent search heuristics, using random improving moves is more effective.
+In \cite{chicano:2014:ball} this decision was amended slightly to include the requirement that smaller
+distance moves should be chosen before larger distance moves.
 
@@
       \EndIf
-      \State Cross solution with $T_i$ to create $2^i$ offspring
+      \State Cross solution with $T_i$ to create $2^{i+1}$ offspring
       \State Apply hamming ball hill climber to each offspring
@@
 
+Hamming Ball Hill Climbing is not sufficient to efficiently find the global optimum
+on problems with even moderate epistasis~\cite{chicano:2014:ball}. This is because
+like all random restart hill climbers, it relies on random initialization to fall
+inside the global optimum's basin of attraction.
+
+To remedy this limitation, we set out to develop a minimally complex memetic
+algorithm to help increase this probability. Figure~\ref{fig-TUX} presents
+the Tournament Uniform Crossover (TUX) algorithm, which combines simplistic
+selection with equal probability uniform crossover to generate starting
+solutions likely to be in the global optimum's basin of attraction.
+
+Conceptually TUX iteratively builds a structure similar to a single elimination
+bracket for solutions. Each "match" in the tournament takes in two candidate solutions,
+produces offspring via uniform crossover, applies hill climbing to each offspring, with
+the "winner" being the best of all those solutions. The tournament "bracket" is constructed
+iteratively, storing a single list of solutions $T$, such that $|T|$ is equal to the height
+of the tournament. This is possible because when a sub-bracket is complete only a single solution
+emerges and solutions only need to be stored until their partner is found.
+TUX is fully elitist but does not prematurely converge. This is because search continuously
+integrates new randomly generated solutions through other parts of the bracket. Whenever the
+top of the current bracket is reached, TUX doubles the size of the virtual tournament.
+
+When crossing solutions at $T_i$, TUX produces $2^{i+1}$ offspring. This relationship
+ensures that in total all levels of the tournament, including random initialization,
+perform the same number of hill climbing steps. It also shifts the focus of search toward
+areas expected to be of higher fitness. The expectation is also that it becomes progressively
+harder to improve solutions the higher up the tournament you advance, so more attempts
+are necessary to create new useful solutions.
+
+The primary advantages of TUX is that it does not introduce any new parameters (still
+requires an $r$ for the hill climber) and is relatively simple to implement. Even so
+it allows for learning from previous local optima and as Section~\ref{sec-experiments}
+will show it is quite effective at optimization.
+
 \section{Parameter-less Population Pyramid: P3}
-
 \subsection{Origins}
-~\cite{goldman:2014:p3}
 \begin{figure}
@@
 \end{figure}
+
+The Parameter-less Population Pyramid (P3) is a recently introduced
+black box optimization method which requires no problem specific configuration~\cite{goldman:2014:p3}.
+Unlike previous parameter-less methods it was shown to be more efficient than
+competing state-of-the-art algorithms across a number of problem classes.
+
+P3 combines hill climbing and model building using a novel, iteratively constructed, pyramid
+of populations. Figure~\ref{fig-p3} provides the high level description of how the algorithm
+works. In order to perform solution mixing, P3 learns a linkage tree from pairwise variable
+entropy which is then used to perform global optima mixing~\cite{thierens:2011:gomea}. In global
+optimal mixing a series of variable subsets are used to copy genetic information from multiple
+sources into a single solution. Each time variables are donated, the modified solution is evaluated,
+with the change being reverted if solution quality did not improve. In this way P3 performs
+multiple evaluations per mixing event, with a single solution receiving genes from randomly chosen
+donors in the pyramid level.
+Due to this model building, P3 requires \BigO{N} amortized time per evaluation and \BigO{N^2}
+memory usage.
 
@@
 
+P3 represents a natural method for integrating the Hamming Ball Hill Climber into
+a global optimization algorithm as P3 already utilizes local search. However, there
+are also a number of ways in which P3 can be made more efficient by leveraging the
+additional information available in the gray box domain.
+
+First and foremost, linkage learning is no longer necessary. By definition
+the direct non-linear relationships between variables are known. As a result
+P3 no longer needs to store the pairwise frequency information which leads
+the black box version to require \BigO{N^2} memory. Instead the subfunction
+information itself can be used to construct the linkage clusters.
+Figure~\ref{fig-sfx-tree} provides an efficient method for creating a linkage
+tree similar to that used in black box P3 given subfunction information.
+Each variable starts in its own cluster. Each subfunction in a random order
+is used to determine how to link existing clusters. If a subfunction overlaps
+multiple top level clusters, all overlapped clusters are merged to create a new
+top level cluster. This process creates a linkage tree similar to that learned
+for P3, with the difference that a node in the tree can have up to $k$ children.
+
+\Call{SubfunctionTree}{} requires \BigO{|cluster|} time to construct each $cluster$
+in the linkage tree. In the gray box domain the total number of subfunctions is required
+to be $\Theta(N)$ in order to ensure the problem is not separable. As all variables must
+participate in a subfunction, at least $\Omega(N)$ clusters must be created. Therefore
+the only cost which cannot be amortized over the clusters to \BigO{1} is the creation
+of each new $cluster$, leading to a time of \BigO{|cluster|} per $cluster$.
+
+TODO Efficiency comparison of cluster creation / usage, creates new clusters every time
+
+TODO Random subsample of clusters, not smallest first
+
+TODO Doesn't recalculate entire fitness, just tracks delta changes.
+
 TODO State that unlike BBP3, GBP3 performs hill climbing after each donation
-\subsection{Time Complexity}
+
+TODO Compare P3 storage with TUX to highlight full population versus single solution mixing
+
+TODO Overall Time Complexity
 
 \section{Experimentation}
+\label{sec-experiments}
 ~\cite{wright:2000:solvingnk}

### 2015-01-09 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
More Gray Box

@@
 multiple evaluations per mixing event, with a single solution receiving genes from randomly chosen
-donors in the pyramid level.
+donors in the pyramid level. This also means that unlike TUX, which uses information from only two
+solutions to perform crossover, P3 uses information from an entire population.
 Due to this model building, P3 requires \BigO{N} amortized time per evaluation and \BigO{N^2}
@@
 
-TODO Efficiency comparison of cluster creation / usage, creates new clusters every time
+This method of constructing clusters has a number of desirable properties beyond its
+efficiency. As all non-linear relationships are denoted by subfunctions, mixing
+complete subfunctions between individual has a high likelihood of preserving quality.
+Each cluster created contains at least one entire subfunction, and two
+clusters must differ by at least one entire subfunction. The probability two
+variables appear in the same cluster increases with the number of subfunctions
+they share. The more paths that exist between two variables in the subfunction
+connectivity graph, the more likely those variables will appear in the same cluster.
 
-TODO Random subsample of clusters, not smallest first
+Due to this change, a few compensatory adjustments to P3's algorithm were required.
+First, in black box P3 the number of clusters discovered by linkage learning is
+limited by the number of solutions at that level of the pyramid. To duplicate this
+behavior, after calling \Call{SubfunctionTree}{} a random subset of the discovered
+clusters is chosen, such that the size is limited to the number of solutions
+stored at the current level of the pyramid. As the clusters produced do not
+rely on the solutions stored in the pyramid, clusters are regenerated before each
+mixing event, unlike black box P3 in which they are regenerated only when the stored
+solutions change. Finally, unlike previous work, the clusters are applied in random order
+to avoid the time required to sort them based on length.
 
-TODO Doesn't recalculate entire fitness, just tracks delta changes.
-
-TODO State that unlike BBP3, GBP3 performs hill climbing after each donation
-
-TODO Compare P3 storage with TUX to highlight full population versus single solution mixing
+Another efficiency gain possible due to the gray box domain is that each time a
+donation is made, only the effected part of the solution's fitness needs to be
+recalculated. As a result the number of subfunction evaluations required
+to determine the change in fitness is only \BigO{|cluster|}. This also
+allows for gray box P3 to efficiently reapply hill climbing after each
+donation as only effected moves need to be rechecked. As a result a donation,
+and its resulting modifications from hill climbing, are kept only if the
+new local optima is at least as fit as the solution before the donation occurred.
+All combined a single donation plus returning to a local optimum requires \BigO{|cluster| + I} time,
+while just the donation in black box P3 requires \BigO{N}.
 

### 2015-01-12 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
Gray Box up to results

@@
 \section{Hamming Ball Hill Climbing}
+\label{sec-hamming}
 The first algorithm to rigorously exploit the features of the gray box domain
@@
 
-TODO Overall Time Complexity
-
 \section{Experimentation}
 \label{sec-experiments}
-~\cite{wright:2000:solvingnk}
+To compare these gray box optimization techniques we have chosen
+NKq-Landscapes~\cite{chicano:2014:ball}. NKq-Landscapes create
+a collection of randomly generated problem instances given a
+higher level problem class description. Each instance is described
+by a series of $N$ subfunctions, each corresponding to a variable
+in the solution. This subfunction uses its variable and $K$ other
+variables in the solution to calculate a fitness value. Fitness values
+are represented as a randomly generated lookup table, such that table
+entries are integers in the range $[0..q-1]$. As each subfunction reads
+$K+1$ variables, the table's size and $q$ are set to $2^{K+1}$. The quality
+of a solution is equal to the sum of the values returned by these subfunctions.
+
+In this work we shall consider two methods for choosing the $K$ variables
+each $x_i$'s subfunction depends on: Nearest Neighbor NKq and Unrestricted NKq.
+In Nearest Neighbor NKq each variable depends on the $K$ which immediately follow
+it in the genome, with variables at the end of the solution wrapping to depend on the
+starting variables. Landscapes of this form can be solved in polynomial
+time~\cite{wright:2000:solvingnk}, allowing comparisons of how quickly each optimization
+algorithm can find the global optima. Nearest Neighbor NKq also ensures that $c=k=K+1$
+and that both $c$ and $k$ do not increase as $N$ increases, ensuring that the efficiency
+conclusions made in Section~\ref{sec-hamming} are applicable.
+
+Unrestricted NKq landscapes draw the $K$ dependencies at random. For $K > 1$ it is
+NP-Hard to find the global optimum of these landscapes. This also means that while
+$k$ remains fixed, the maximum number of subfunctions a variable appears in ($c$) can increase
+as $N$ increases. As a result some of the efficiency claims in Section~\ref{sec-hamming}
+may not be applicable.
+
+For each $N$ and $K$ tested, we generated 50 Nearest Neighbor NKq and Unrestricted NKq instances.
+Each method was run once on each instance, and limited to 3 hours of computation and 4 GB of memory.
+Runs were performed on 2.5GHz Intel Xeon E5-2670v2 processors using the C++11 code provided
+from our website.\footnote{\url{https://github.com/name_of_repository/anonymous_name_of_project}}
+Each time the run achieved a new best fitness we record current amount of processing time used.
+Timing includes the discovery of subgraphs to allow for comparison between different radius values.
+When reporting the ``best'' fitness for an instance we mean the best fitness found by any method before
+the time limit is reached. On all Nearest Neighbor NKq instances the ``best'' fitness
+is the global optimum, verified using dynamic programming.
 
 \subsection{The Effect of Radius}
+The only algorithm parameter in the Hamming Ball Hill Climber, TUX, and Gray Box P3
+is the radius of the hamming ball.
+
 TODO Plot for single K value with x=radius, y=fitness, color=solver

### 2015-01-13 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
Editting pass

@@
 about a problem from being provided to search algorithms. At the other end of
-the spectrum is white box optimization, such that everything about the problem
+the spectrum is white box optimization, in which everything about the problem
 is known allowing search techniques to specialize at the cost of generality.
-In between these two extremes is gray box optimization, such that some additional
-information about the problem is known, but the interface is still general enough
-to allow optimization algorithms to solve a variety of problems.
+Gray box optimization exists between these two extremes, exposing additional
+information which is available for a variety of problems.
 
@@
 In order to obtain efficiency bounds for optimization algorithms, the following
-assertions are included. The total number of subfunctions grows no more than
+assertions are also included. The total number of subfunctions grows no more than
 linearly with $N$, which holds true for randomized MAX-SAT, mesh spin glasses, and NK Landscapes.
@@
 This is achieved by first determining the fitness effect $delta$ of making each move, and then updating
-only effected move each time a change to the solution is made. Due to the requirements of the domain,
+only effected moves each time a change to the solution is made. Due to the requirements of the domain,
 at most \BigO{ck} moves can be effected, with each requiring at most \BigO{1} subfunction calls
@@
 
-By again leveraging the properties of this domain, \cite{chicano:2014:ball} extended their
-local search algorithm to find $r$-bit local optima. Termed the Hamming Ball Hill Climber,
-this method efficiently searches the space of all moves which flip up to $r$ bits in the
-solution. While na\"ively this create \BigO{N^r} moves, in the gray box domain only
+By again leveraging the properties of this domain, the Hamming Ball Hill Climber~\cite{chicano:2014:ball} extends this
+method to find $r$-bit local optima. While na\"ively this would require testing \BigO{N^r} moves, in the gray box domain only
 \BigO{(3ck)^rN} must be checked. This is due to the bounded relationships between variables.
@@
 but no algorithm was given to efficiently discover these moves. This process was not included
-in their timing calculations has it only has to be performed once per problem.
+in their timing calculations as it only has to be performed once per problem.
 
@@
 In \cite{chicano:2014:ball} this decision was amended slightly to include the requirement that smaller
-distance moves should be chosen before larger distance moves.
+distance moves are chosen before larger distance moves.
 
@@
   \begin{algorithmic}[1]
+  \Procedure{ConnectedInducedSubgraphs}{}
+    \State $closed \leftarrow \emptyset$
+    \State $found \leftarrow []$
+    \ForAll{$v \in V$}
+      \State $closed \leftarrow closed \cup \{v\}$
+      \State $found \leftarrow found + $\Call{CISG}{$v$, $\emptyset$, $closed$, $\emptyset$}
+    \EndFor
+    \State \Return $found$
+  \EndProcedure
   \Procedure{CISG}{$v$, $subset$, $closed$, $open$}
@@
 \end{algorithmic}
-  \caption{Algorithm to recursively find all connected induced subgraphs of size $r$ or fewer which
-           contain $subset \cup \{v\}$ and no additional vertices in $closed$.}
+  \caption{Algorithm to recursively find all connected induced subgraphs of size $r$ or fewer.}
   \label{fig-connected-subgraphs}
@@
 In order to determine which moves in the hamming ball must be evaluated,
-we developed \Call{CISG}{} given in Figure~\ref{fig-connected-subgraphs}
-to discover all Connected Induced Subgraphs of $r$ or fewer vertices.
-\Call{CISG}{} is a recursive algorithm which finds all subgraphs
+we developed \Call{ConnectedInducedSubgraphs}{} given in Figure~\ref{fig-connected-subgraphs}.
+\Call{CISG}{} is a recursive helper function which finds all subgraphs
 which contain a given $subset$ and a given vertex $v$, while excluding
@@
 all previously searched vertices, and $subset=open=\emptyset$. In the initial
-call all desired subgraphs which contain $v$ are found, which is why subsequent
-top level calls include $v$ in $closed$ to prevent duplicate subgraphs from being
-returned.
+call all desired subgraphs which contain $v$ are found, which is why $v$ remains
+in $closed$ to prevent duplicate subgraphs from being returned.
 
@@
 adjacent to $v$ in the graph. By construction this means that $open$ contains
-all possible ways of adding a single vertex to the current $subset$. As $v'$
+all possible ways of adding a single vertex to the current $subset'$. As $v'$
 are tested they are temporarily added to $closed$ to prevent recursive calls
@@
   \Procedure{Iterate-TUX}{}
-    \State Create random solution
-    \State Apply hamming ball hill climber to solution
+    \State Create random $solution$
+    \State Hamming Ball Hill Climb $solution$
     \ForAll{$T_i \in T$}
       \If{$T_i$ is empty}
-        \State $T_i \leftarrow$ solution
+        \State $T_i \leftarrow solution$
         \State \Return
       \EndIf
-      \State Cross solution with $T_i$ to create $2^{i+1}$ offspring
-      \State Apply hamming ball hill climber to each offspring
-      \State solution $\leftarrow$ best of offspring, solution, and $T_i$
+      \State Cross $solution$ with $T_i$ to create $2^{i+1}$ offspring
+      \State Hamming Ball Hill Climb each offspring
+      \State $solution \leftarrow$ best of offspring, $solution$, and $T_i$
       \State $T_i \leftarrow$ empty
     \EndFor
-    \State Add solution to end of $T$
+    \State Add $solution$ to end of $T$
   \EndProcedure
@@
 Conceptually TUX iteratively builds a structure similar to a single elimination
-bracket for solutions. Each "match" in the tournament takes in two candidate solutions,
+bracket for solutions. Each ``match'' in the tournament takes in two candidate solutions,
 produces offspring via uniform crossover, applies hill climbing to each offspring, with
-the "winner" being the best of all those solutions. The tournament "bracket" is constructed
+the ``winner'' being the best of all those solutions. The tournament ``bracket'' is constructed
 iteratively, storing a single list of solutions $T$, such that $|T|$ is equal to the height
@@
 entropy which is then used to perform global optima mixing~\cite{thierens:2011:gomea}. In global
-optimal mixing a series of variable subsets are used to copy genetic information from multiple
+optimal mixing these learned clusters of variables are used to donate genetic information from multiple
 sources into a single solution. Each time variables are donated, the modified solution is evaluated,
-with the change being reverted if solution quality did not improve. In this way P3 performs
+with the change being reverted if solution quality decreased. In this way P3 performs
 multiple evaluations per mixing event, with a single solution receiving genes from randomly chosen
@@
 to be $\Theta(N)$ in order to ensure the problem is not separable. As all variables must
-participate in a subfunction, at least $\Omega(N)$ clusters must be created. Therefore
+participate in a subfunction, $\Theta(N)$ clusters must be created. Therefore
 the only cost which cannot be amortized over the clusters to \BigO{1} is the creation
@@
 efficiency. As all non-linear relationships are denoted by subfunctions, mixing
-complete subfunctions between individual has a high likelihood of preserving quality.
-Each cluster created contains at least one entire subfunction, and two
+complete subfunctions between individuals has a high likelihood of preserving quality.
+Each cluster contains at least one entire subfunction, and two
 clusters must differ by at least one entire subfunction. The probability two
@@
 In this work we shall consider two methods for choosing the $K$ variables
-each $x_i$'s subfunction depends on: Nearest Neighbor NKq and Unrestricted NKq.
-In Nearest Neighbor NKq each variable depends on the $K$ which immediately follow
-it in the genome, with variables at the end of the solution wrapping to depend on the
-starting variables. Landscapes of this form can be solved in polynomial
+each subfunction depends on: Nearest Neighbor NKq and Unrestricted NKq.
+In Nearest Neighbor NKq each variable depends on the $K$ variables which immediately follow
+it in the solution, with dependencies wrapping around the end of the solution.
+Landscapes of this form can be solved in polynomial
 time~\cite{wright:2000:solvingnk}, allowing comparisons of how quickly each optimization
 algorithm can find the global optima. Nearest Neighbor NKq also ensures that $c=k=K+1$
-and that both $c$ and $k$ do not increase as $N$ increases, ensuring that the efficiency
+and that both $c$ and $k$ do not increase as $N$ increases, meaning the efficiency
 conclusions made in Section~\ref{sec-hamming} are applicable.
@@
 the time limit is reached. On all Nearest Neighbor NKq instances the ``best'' fitness
-is the global optimum, verified using dynamic programming.
+is also the global optimum, verified using dynamic programming.
 

### 2015-01-20 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
Figures and edits

@@
 \newcommand{\includegraphicsfit}[1]
-{\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}
+{\includegraphics[width=\columnwidth,height=.4\textheight,keepaspectratio]{#1}}
 
@@
 \section{Introduction}
-~\cite{chen:2011:memetic}
 
-~\cite{whitley:2013:greedy}~\cite{chicano:2014:ball}
+Recent work~\cite{whitley:2013:greedy,chicano:2014:ball} has shown that by
+providing local search with more than just an evaluation function can lead
+to extraordinary improvements search power and efficiency. These improvements
+require information easily derived for a broad range of interesting NP-Hard
+problems.
 
-~\cite{goldman:2014:p3}
+Here we set out to extend these benefits to global optimization by integrating
+this local search with additional heuristics~\cite{chen:2011:memetic}.
+We introduce a novel algorithm designed to combine local
+and global search as simply as possible. To test the true power of this
+process we also extend the Parameter-less Population Pyramid~\cite{goldman:2014:p3},
+current state-of-the-art in black box optimization. This ``Gray Box P3''
+utilizes both efficient local search and the additional problem information
+to drastically improve performance and reduce memory usage.
 
@@
 Gray box optimization exists between these two extremes, exposing additional
-information which is available for a variety of problems.
+information to be exploited by search and which is available for a variety of problems.
 
 Here we shall consider the gray box optimization domain with the following
-characteristics. The quality of a solution is determined by the summation
-of subfunctions. Each subfunction uses at most $k$ variables in the solution,
+characteristics. The function which evaluates the quality of a solution is determined by the summation
+of subfunctions. Each subfunction uses at most $k$ variables in the given solution,
 such that as the total number of solution variables $N$ increases, $k$ remains constant.
-Each subfunction must be independently evaluable, and the variables each uses is known.
+Each subfunction must be independently evaluable, and the variables each uses are known.
 
-While more restrictive than black box optimization, this domain is very general.
+While more restrictive than black box optimization, this domain is still very general.
 MAX-SAT, and by extension all NP-hard problems, can be represented
@@
 linearly with $N$, which holds true for randomized MAX-SAT, mesh spin glasses, and NK Landscapes.
-Furthermore, each variable must participate in at most $c$ subfunctions, which holds true
-for mesh spin glasses and Nearest Neighbor NK. If violated these assertions do not
+Furthermore, each variable participates in at most $c$ subfunctions, which holds true
+for mesh spin glasses and Nearest Neighbor NK. If violated, these assertions do not
 necessarily prevent the gray box optimization algorithms from solving the problem, but may result
@@
 
-\section{Hamming Ball Hill Climbing}
+\section{Hamming-Ball Hill Climbing}
 \label{sec-hamming}
@@
 
-By again leveraging the properties of this domain, the Hamming Ball Hill Climber~\cite{chicano:2014:ball} extends this
+By again leveraging the properties of this domain, the Hamming-Ball Hill Climber~\cite{chicano:2014:ball} extends this
 method to find $r$-bit local optima. While na\"ively this would require testing \BigO{N^r} moves, in the gray box domain only
 \BigO{(3ck)^rN} must be checked. This is due to the bounded relationships between variables.
-Consider that if $i$ and $j$ do not share a subfunction the effect of flipping both variables
+If $i$ and $j$ do not share a subfunction, the effect of flipping both variables
 must be equal to the sum of flipping both separately. Therefore only variables which share
@@
 
-In order to determine which moves in the hamming ball must be evaluated,
+In order to determine which moves in the hamming-ball must be evaluated,
 we developed \Call{ConnectedInducedSubgraphs}{} given in Figure~\ref{fig-connected-subgraphs}.
@@
 \section{Tournament Uniform Crossover: TUX}
+\label{sec-tux}
 \begin{figure}
@@
     \State Create random $solution$
-    \State Hamming Ball Hill Climb $solution$
+    \State Hamming-Ball Hill Climb $solution$
     \ForAll{$T_i \in T$}
@@
       \State Cross $solution$ with $T_i$ to create $2^{i+1}$ offspring
-      \State Hamming Ball Hill Climb each offspring
+      \State Hamming-Ball Hill Climb each offspring
       \State $solution \leftarrow$ best of offspring, $solution$, and $T_i$
@@
 
-Hamming Ball Hill Climbing is not sufficient to efficiently find the global optimum
-on problems with even moderate epistasis~\cite{chicano:2014:ball}. This is because
+Hamming-Ball Hill Climbing is not sufficient to efficiently find the global optimum
+on problems with even moderate epistasis~\cite{chicano:2014:ball}. This is because,
 like all random restart hill climbers, it relies on random initialization to fall
@@
 
-P3 represents a natural method for integrating the Hamming Ball Hill Climber into
+P3 represents a natural method for integrating the Hamming-Ball Hill Climber into
 a global optimization algorithm as P3 already utilizes local search. However, there
@@
 \subsection{The Effect of Radius}
-The only algorithm parameter in the Hamming Ball Hill Climber, TUX, and Gray Box P3
-is the radius of the hamming ball.
+The only algorithm parameter in the Hamming-Ball Hill Climber, TUX, and Gray Box P3
+is the radius of the hamming-ball.
+
+\begin{figure}
+  \centering
+  \includegraphicsfit{fitness-nn}
+  \caption{Nearest Neighbor NKq}
+  \label{fig-fitness-nn}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphicsfit{fitness-un}
+  \caption{Unrestricted NKq}
+  \label{fig-fitness-un}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphicsfit{p3-seconds}
+  \caption{Timing}
+  \label{fig-p3-seconds}
+\end{figure}
 
@@
 TODO Plot for seconds to global optimum as N increases, include Black Box P3
+
+\begin{figure}
+  \centering
+  \includegraphicsfit{length}
+  \caption{Length}
+  \label{fig-length}
+\end{figure}
 

### 2015-01-20 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
Abstract draft

@@
 %Maximum 200 Words
-TODO
+
+For many challenging problems, the non-linear relationships between variables
+are known and bounded. Previous work has developed local search operators
+which require a linear factor less time than black box methods. Furthermore,
+this Hamming-Ball Hill Climber is capable of efficiently finding $r$-bit local optima.
+In this paper we extend this operator to perform efficient global optimization.
+We create a simple memetic algorithm to demonstrate how even trivial
+extensions can improve search results. We then modify the Parameter-less
+Population Pyramid (P3), shown to be highly effective for black box optimization,
+to leverage the additional information available in the gray box domain.
+
+We provide experimental evidence on both NKq-Landscapes and Ising Spin Glasses
+that Gray Box P3 is effective at finding the global optima even for problems with
+thousands of variables. This capability is complemented by its efficiency, with
+running time and memory usage decreased by a linear factor from Black Box P3.
+On NKq this results in a x374 speedup for problems with at least 1000 variables.
+While local search introduces a parameter, we experimentally show this parameter
+can likely be fixed to 1 without significant loss in search quality.
+
 \end{abstract}

### 2015-01-21 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
Figures, dashes, abstract update.

--- 2015-01-20 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
@@
 \newcommand{\includegraphicsfit}[1]
-{\includegraphics[width=\columnwidth,height=.4\textheight,keepaspectratio]{#1}}
+{\includegraphics[width=.9\columnwidth,height=.4\textheight,keepaspectratio]{#1}}
 
@@
 \begin{document}
-\title{Gray Box Optimization using the Parameter-less Population Pyramid}
+\title{Gray-Box Optimization using the Parameter-less Population Pyramid}
 %\subtitle{[Genetic Algorithms Track]}
@@
 %Maximum 200 Words
-TODO
+
+Unlike black-box optimization problems, gray-box optimization problems have known,
+limited-cardinality relationships between variables. Such problems are
+well-known and exist in many real world problems (network security, computational biology, VLSI design, and statistical physics). These
+problems are amenable to local search using a Hamming-Ball Hill Climber
+(HBHC) that can efficiently find r-bit local optimal. We show how a
+simple memetic algorithm in conjunction with HBHC can find \emph{global}
+optima for gray-box problems. Finally a gray-box version of P3, which
+utilizes the information about variable relationships given by the gray-box
+problem itself, is the most efficient of the examined algorithms.
+
+We provide experimental evidence on both NKq-Landscapes and Ising Spin Glasses
+that Gray-Box P3 is effective at finding the global optima even for problems with
+thousands of variables. This capability is complemented by its efficiency, with
+running time and memory usage decreased by a linear factor from Black-Box P3.
+On NKq this results in a x375 speedup for problems with at least 1000 variables.
+While local search introduces a parameter, we experimentally show this parameter
+can likely be fixed to 1 without significant loss in search quality.
+
 \end{abstract}
@@
 process we also extend the Parameter-less Population Pyramid~\cite{goldman:2014:p3},
-current state-of-the-art in black box optimization. This ``Gray Box P3''
+current state-of-the-art in black-box optimization. This ``Gray-Box P3''
 utilizes both efficient local search and the additional problem information
@@
 
-\section{Gray Box Domain}
+\section{Gray-Box Domain}
 \label{sec-gray-box}
-In black box optimization, the only information available is a specification
+In black-box optimization, the only information available is a specification
 for what constitutes a valid solution and a function which returns the quality
@@
 is known allowing search techniques to specialize at the cost of generality.
-Gray box optimization exists between these two extremes, exposing additional
+Gray-box optimization exists between these two extremes, exposing additional
 information to be exploited by search and which is available for a variety of problems.
 
-Here we shall consider the gray box optimization domain with the following
+Here we shall consider the gray-box optimization domain with the following
 characteristics. The function which evaluates the quality of a solution is determined by the summation
@@
 
-While more restrictive than black box optimization, this domain is still very general.
+While more restrictive than black-box optimization, this domain is still very general.
 MAX-SAT, and by extension all NP-hard problems, can be represented
@@
 Ising Spin Glasses~\cite{pelikan:2003:hboaising} all work with problems which fit
-in the gray box domain. These relate back to problems of real world interest such
+in the gray-box domain. These relate back to problems of real world interest such
 as network security, computational biology, VLSI design, and statistical physics.
@@
 for mesh spin glasses and Nearest Neighbor NK. If violated, these assertions do not
-necessarily prevent the gray box optimization algorithms from solving the problem, but may result
+necessarily prevent the gray-box optimization algorithms from solving the problem, but may result
 in slower than predicted performance.
@@
 \label{sec-hamming}
-The first algorithm to rigorously exploit the features of the gray box domain
+The first algorithm to rigorously exploit the features of the gray-box domain
 described in Section~\ref{sec-gray-box} was \cite{whitley:2013:greedy}. In this study
@@
 first improvement and next best hill climbing in \BigO{I+N}, where $I$ is the number of improving
-moves required to reach a local optimum. In comparison, the black box approach for these searches can require
+moves required to reach a local optimum. In comparison, the black-box approach for these searches can require
 \BigO{IN^2}.
@@
 By again leveraging the properties of this domain, the Hamming-Ball Hill Climber~\cite{chicano:2014:ball} extends this
-method to find $r$-bit local optima. While na\"ively this would require testing \BigO{N^r} moves, in the gray box domain only
+method to find $r$-bit local optima. While na\"ively this would require testing \BigO{N^r} moves, in the gray-box domain only
 \BigO{(3ck)^rN} must be checked. This is due to the bounded relationships between variables.
@@
 
-When applied to the sparse graphs inherent in the gray box domain, this algorithm
+When applied to the sparse graphs inherent in the gray-box domain, this algorithm
 requires \BigO{r!(c(k-1))^rN} time, which reduces to \BigO{N}. The time spent
@@
 The Parameter-less Population Pyramid (P3) is a recently introduced
-black box optimization method which requires no problem specific configuration~\cite{goldman:2014:p3}.
+black-box optimization method which requires no problem specific configuration~\cite{goldman:2014:p3}.
 Unlike previous parameter-less methods it was shown to be more efficient than
@@
 
-\subsection{Gray Box Specialization}
+\subsection{Gray-Box Specialization}
+\begin{comment}
 \begin{figure}
@@
 are also a number of ways in which P3 can be made more efficient by leveraging the
-additional information available in the gray box domain.
+additional information available in the gray-box domain.
 
@@
 P3 no longer needs to store the pairwise frequency information which leads
-the black box version to require \BigO{N^2} memory. Instead the subfunction
-information itself can be used to construct the linkage clusters.
+the black-box version to require \BigO{N^2} memory. Instead,
+we have developed a method for creating a linkage tree which learns clusters
+from the same graph defined in Section~\ref{sec-hamming}. The goal is for each
+cluster to be a connected induced subgraph, with the size of clusters
+mirroring those produced by the agglomerative linkage learning process normally
+used with P3. To form a single cluster, a random graph search is performed from
+a random starting vertex until a desired number of unique vertices have been explored.
+Cluster sizes are set by recursively such that for every cluster created of size $l>1$
+clusters of size $a$ and $l-a$ are also created, such that $a$ is chosen uniformly
+from the range $[1..l-1]$, with the largest clusters created by similarly splitting $l=N$.
+
+This linking algorithm has a number of useful properties. First, it creates exactly
+$2N-2$ clusters, distributed in size similarly to the black-box clustering algorithm.
+Performing a random graph search to find $l$ unique vertices requires \BigO{lck} time,
+meaning cluster creation is optimally efficient.
+The cluster splitting process has identical properties as random pivot quicksort,
+meaning the sum of cluster sizes is \BigO{N\log N} in the average case.
+This efficiency allows new clusters to be created before every mixing event,
+unlike black-box P3 where they are created only when new solutions are added to the population.
+For simplicity the clusters are shuffled after the are created, not sorted on size like in
+black-box P3.
+
+Beyond efficiency, there are good reasons to believe these clusters will be useful
+to search. The closer two variables are in the dependency graph, the more likely
+they are to appear in the same cluster. All variables on average are expected to appear
+in at least one cluster, but variables which are central in the graph will appear in more
+clusters than those on the periphery. If all else is equal, the more paths of a given
+size between two variables, the higher the probability of them being in the same cluster.
+Unlike black-box P3, this linkage tree does not require clusters to be nested, allowing
+more diversity in the types of clusters appearing in a single tree.
+
+In effect the clusters are sampling moves which the Hamming-Ball Hill Climber
+would make if $r \ge l$. As clusters are used to move values between solutions, these
+moves are restricted to moving in the direction of previously found high quality moves.
+As a result if the density of high quality solutions is higher between good solutions than
+outside of good solutions, this method will perform better than random search or sampling from
+all possible moves of size $l$.
+
+\begin{comment}
 Figure~\ref{fig-sfx-tree} provides an efficient method for creating a linkage
-tree similar to that used in black box P3 given subfunction information.
+tree similar to that used in black-box P3 given subfunction information.
 Each variable starts in its own cluster. Each subfunction in a random order
@@
 \Call{SubfunctionTree}{} requires \BigO{|cluster|} time to construct each $cluster$
-in the linkage tree. In the gray box domain the total number of subfunctions is required
+in the linkage tree. In the gray-box domain the total number of subfunctions is required
 to be $\Theta(N)$ in order to ensure the problem is not separable. As all variables must
@@
 connectivity graph, the more likely those variables will appear in the same cluster.
+\end{comment}
 
-Due to this change, a few compensatory adjustments to P3's algorithm were required.
-First, in black box P3 the number of clusters discovered by linkage learning is
-limited by the number of solutions at that level of the pyramid. To duplicate this
-behavior, after calling \Call{SubfunctionTree}{} a random subset of the discovered
-clusters is chosen, such that the size is limited to the number of solutions
-stored at the current level of the pyramid. As the clusters produced do not
-rely on the solutions stored in the pyramid, clusters are regenerated before each
-mixing event, unlike black box P3 in which they are regenerated only when the stored
-solutions change. Finally, unlike previous work, the clusters are applied in random order
-to avoid the time required to sort them based on length.
-
-Another efficiency gain possible due to the gray box domain is that each time a
+Another efficiency gain possible due to the gray-box domain is that each time a
 donation is made, only the effected part of the solution's fitness needs to be
 recalculated. As a result the number of subfunction evaluations required
-to determine the change in fitness is only \BigO{|cluster|}. This also
-allows for gray box P3 to efficiently reapply hill climbing after each
+to determine the change in fitness is only \BigO{l}. This also
+allows for gray-box P3 to efficiently reapply hill climbing after each
 donation as only effected moves need to be rechecked. As a result a donation,
@@
 new local optima is at least as fit as the solution before the donation occurred.
-All combined a single donation plus returning to a local optimum requires \BigO{|cluster| + I} time,
-while just the donation in black box P3 requires \BigO{N}.
+All combined a single donation plus returning to a local optimum requires \BigO{l + I} time,
+while just the donation in black-box P3 requires \BigO{N}.
 
@@
 \label{sec-experiments}
-To compare these gray box optimization techniques we have chosen
-NKq-Landscapes~\cite{chicano:2014:ball}. NKq-Landscapes create
+To compare these gray-box optimization techniques we have chosen
+NKq-Landscapes~\cite{chicano:2014:ball} and Ising Spin Glasses~\cite{saul:1994:spinglass}. NKq-Landscapes create
 a collection of randomly generated problem instances given a
@@
 
-For each $N$ and $K$ tested, we generated 50 Nearest Neighbor NKq and Unrestricted NKq instances.
+Ising Spin Glasses are a type of MAX-CUT problem relevant in statistical physics. Each spin glass
+encodes spins (vertices) and their relationships (edges) with the goal of assigning each spin a direction
+which minimizes relationship energy. Just as with NKq-Landscapes, Ising Spin Glasses as a whole
+are NP-Hard, but the $2D\pm J$ subset is polynomially solvable~\cite{saul:1994:spinglass}\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}. In this
+subset the graph is a 2D toroidal grid, with each edge's weight either -1 or 1. In gray-box terms
+problems of this subset have $k=2$ and $c=4$ regardless of $N$. While Ising Spin Glasses are traditionally
+a minimization problem, here we present them as a maximization problem by negating the normal fitness.
+
+For each problem class tested, we generated 50 instances.
 Each method was run once on each instance, and limited to 3 hours of computation and 4 GB of memory.
-Runs were performed on 2.5GHz Intel Xeon E5-2670v2 processors using the C++11 code provided
+Runs were performed on 2.5GHz Intel Xeon E5-2670v2 processors using the C++11 code available
 from our website.\footnote{\url{https://github.com/name_of_repository/anonymous_name_of_project}}
@@
 the time limit is reached. On all Nearest Neighbor NKq instances the ``best'' fitness
-is also the global optimum, verified using dynamic programming.
+is also the global optimum, verified using dynamic programming. The same is true of all Ising Spin Glass
+problems except for 6 of 50 on N=6084.
+
+All figures report the median run for either percentage error or seconds to reach the best. A run's percentage error is
+equal to its how much less its fitness is than the best, divided by the best. When reporting seconds
+to reach the best fitness, any run which did not find the best fitness is treated as slower than any run that did.
+If the median run was unsuccessful, no data point is drawn.
 
 \subsection{The Effect of Radius}
-The only algorithm parameter in the Hamming-Ball Hill Climber, TUX, and Gray Box P3
-is the radius of the hamming-ball.
+The only algorithm parameter in the Hamming-Ball Hill Climber, TUX, and Gray-Box P3
+is the radius of the hamming-ball. Therefore our first experiments are designed to determine
+the effect of this parameter on solution quality.
 
-\begin{figure}
+\begin{figure*}
   \centering
-  \includegraphicsfit{fitness-nn}
-  \caption{Nearest Neighbor NKq}
-  \label{fig-fitness-nn}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphicsfit{fitness-un}
-  \caption{Unrestricted NKq}
-  \label{fig-fitness-un}
-\end{figure}
+  \includegraphicswide{fitness}
+  \caption{Comparison of how radius effects solution quality. For NKq-Landscapes $N=6000$ and $K=4$ and
+  for Ising Spin Glasses $N=6084$.}
+  \label{fig-fitness}
+\end{figure*}
 
@@
   \includegraphicsfit{p3-seconds}
-  \caption{Timing}
+  \caption{Median time required for Gray-Box P3 to reach the global optimum of Nearest Neighbor NKq instances with $N=6000$.}
   \label{fig-p3-seconds}
 \end{figure}
-
-TODO Plot for single K value with x=radius, y=fitness, color=solver
-
-TODO Plot for just Pyramid with x=radius, y=seconds to optimum, color=k
 
@@
 \subsection{Fitness Over Time}
+
+\begin{figure*}
+  \centering
+  \includegraphicswide{solver-over}
+  \caption{solver-over.}
+  \label{fig-solver-over}
+\end{figure*}
+
+\begin{figure*}
+  \centering
+  \includegraphicswide{radius-over}
+  \caption{radius-over.}
+  \label{fig-radius-over}
+\end{figure*}
+
 TODO Plot for fixed N, K, and R with x=seconds, y=fitness, color=solver
@@
 \subsection{Scalability}
-TODO Plot for seconds to global optimum as N increases, include Black Box P3
+TODO Plot for seconds to global optimum as N increases, include Black-Box P3
 
@@
   \centering
-  \includegraphicsfit{length}
-  \caption{Length}
-  \label{fig-length}
+  \includegraphicsfit{length-nn}
+  \caption{Length-nn}
+  \label{fig-length-nn}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphicsfit{length-is}
+  \caption{Length-is}
+  \label{fig-length-is}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphicsfit{length-un}
+  \caption{Length-un}
+  \label{fig-length-un}
 \end{figure}

### 2015-01-21 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
Abstract to 197 words

@@
 Unlike black-box optimization problems, gray-box optimization problems have known,
-limited-cardinality relationships between variables. Such problems are
-well-known and exist in many real world problems (network security, computational biology, VLSI design, and statistical physics). These
-problems are amenable to local search using a Hamming-Ball Hill Climber
-(HBHC) that can efficiently find r-bit local optimal. We show how a
+limited-cardinality non-linear relationships between variables. This domain contains many
+real world problems with examples in network security, computational biology, VLSI design, and statistical physics.
+Leveraging these restrictions, the Hamming-Ball Hill Climber
+(HBHC) can efficiently find r-bit local optimal. We show how a
 simple memetic algorithm in conjunction with HBHC can find \emph{global}
-optima for gray-box problems. Finally a gray-box version of P3, which
-utilizes the information about variable relationships given by the gray-box
-problem itself, is the most efficient of the examined algorithms.
+optima for gray-box problems. We also develop a gray-box version of the Parameter-less
+Population Pyramid (P3) utilizing both the HBHC and the known information about variable relationships
+which outperforms all of the examined algorithms
 
@@
 thousands of variables. This capability is complemented by its efficiency, with
-running time and memory usage decreased by a linear factor from Black-Box P3.
+running time and memory usage decreased by up to a linear factor from Black-Box P3.
 On NKq this results in a x375 speedup for problems with at least 1000 variables.
@@
 can likely be fixed to 1 without significant loss in search quality.
-
 \end{abstract}

### 2015-01-26 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
Brain dump of the remaining file.

@@
 Population Pyramid (P3) utilizing both the HBHC and the known information about variable relationships
-which outperforms all of the examined algorithms
+which outperforms all of the examined algorithms.
 
@@
 \keywords{TODO; Local Search; Parameter-less}
+
+\newpage
 
@@
 
-By again leveraging the properties of this domain, the Hamming-Ball Hill Climber~\cite{chicano:2014:ball} extends this
+By again leveraging the properties of this domain, the Hamming-Ball Hill Climber (HBHC)~\cite{chicano:2014:ball} extends this
 method to find $r$-bit local optima. While na\"ively this would require testing \BigO{N^r} moves, in the gray-box domain only
@@
 
-P3 represents a natural method for integrating the Hamming-Ball Hill Climber into
+P3 represents a natural method for integrating the HBHC into
 a global optimization algorithm as P3 already utilizes local search. However, there
@@
 
-In effect the clusters are sampling moves which the Hamming-Ball Hill Climber
+In effect the clusters are sampling moves which the HBHC
 would make if $r \ge l$. As clusters are used to move values between solutions, these
@@
 \subsection{The Effect of Radius}
-The only algorithm parameter in the Hamming-Ball Hill Climber, TUX, and Gray-Box P3
+\label{sec-radius}
+\begin{figure*}
+  \centering
+  \includegraphicswide{fitness}
+  \caption{Comparison of how radius effects solution quality at termination. For NKq-Landscapes $N=6000$ and $K=4$ and
+  for Ising Spin Glasses $N=6084$.}
+  \label{fig-fitness}
+\end{figure*}
+
+The only algorithm parameter in the HBHC, TUX, and Gray-Box P3
 is the radius of the hamming-ball. Therefore our first experiments are designed to determine
@@
 
-\begin{figure*}
-  \centering
-  \includegraphicswide{fitness}
-  \caption{Comparison of how radius effects solution quality. For NKq-Landscapes $N=6000$ and $K=4$ and
-  for Ising Spin Glasses $N=6084$.}
-  \label{fig-fitness}
-\end{figure*}
+Figure~\ref{fig-fitness} shows the effect on final solution fitness as $r$ increases. As expected
+from~\cite{chicano:2014:ball} the HBHC obtains higher quality as $r$ increases with the magnitude
+of the improvement decreasing. TUX has a similar relationship and
+outperforms HBHC on all three problems for all $r$ values. Regardless of $r$, Gray-Box P3 outperforms both, with almost
+all $r$ values reaching the same best fitness. For Nearest Neighbor NKq Gray-Box P3 finds the global
+optimum in every run for $r < 4$, with only a single unsuccessful run at $r=4$.
 
@@
 
-TODO Explain k==1
+To further examine the effect of $r$ on Gray-Box-P3 Figure~\ref{fig-p3-seconds} shows the median number of
+seconds required to reach the global optimum. Setting $r=1$ was the most efficient configuration for all $K > 1$,
+supporting the trend that Gray-Box P3 works best with small $r$ values. With $K=1$, the landscape is smooth enough
+that with a sufficiently high $r$ the HBHC is able to find the global optimum.
 
@@
   \includegraphicswide{solver-over}
-  \caption{solver-over.}
+  \caption{Comparison of solution quality during optimization on a log-log scale for different algorithms. For NKq-Landscapes $N=6000$ and $K=4$ and
+  for Ising Spin Glasses $N=6084$. Each algorithm uses its best found $r$ value.}
   \label{fig-solver-over}
 \end{figure*}
+
+In optimization its sometimes more important how quickly an algorithm reaches high quality solutions than when
+it reaches the global optimum. Figure~\ref{fig-solver-over} shows how solution quality progresses during
+search for each algorithm. HBHC and TUX have significant early delays caused by their high $r$ values. Larger
+$r$'s require a large amount of initial partial evaluation before performing hill climbing. Once that process
+is complete HBHC effectively stalls, with TUX improving at a faster rate. Both are eclipsed by Gray-Box P3,
+which quickly descends to the global optimum, outperforming HBHC and TUX at every point.
 
@@
   \includegraphicswide{radius-over}
-  \caption{radius-over.}
+  \caption{Comparison of Gray-Box P3's solution quality during optimization on a log-log scale for different $r$ values.
+  For NKq-Landscapes $N=6000$ and $K=4$ and for Ising Spin Glasses $N=6084$.}
   \label{fig-radius-over}
@@
 
-TODO Plot for fixed N, K, and R with x=seconds, y=fitness, color=solver
-
-TODO Explain best Pyramid R versus best TUX R.
+To further illustrate the effect of $r$ on Gray-Box P3, we provide Figure~\ref{fig-radius-over}. On Nearest Neighbor NKq
+and Ising Spin Glasses, increasing $r$ does not change the shape of the curve. Instead the quality reached is simply time shifted,
+such that given more time higher $r$ values will reach the same quality. As a result for these problems we conclude
+that searching for an optimal value of $r$ is almost certainly going to be more expensive than simply fixing $r=1$.
+On Unrestricted NKq this relationship is less certain, with $r=1$ potentially having a different, and worse, shape than $r>1$.
+However, due to memory and time restrictions it is difficult to know if this trend continues.
 
 \subsection{Scalability}
-TODO Plot for seconds to global optimum as N increases, include Black-Box P3
-
 \begin{figure}
@@
   \includegraphicsfit{length-nn}
-  \caption{Length-nn}
+  \caption{Comparison of how each algorithm's time required to reach the best fitness found scales with problem size
+  on Nearest Neighbor NKq with $K=4$.}
   \label{fig-length-nn}
@@
   \includegraphicsfit{length-is}
-  \caption{Length-is}
+  \caption{Comparison of how each algorithm's time required to find the best fitness found scales with problem size
+  on Ising Spin Glass.}
   \label{fig-length-is}
 \end{figure}
+
+Perhaps the most critical test of an optimization algorithm's quality is
+how it scales as problem difficulty increases. To test this behavior,
+we ran all three algorithms using the best found $r$ values in Section~\ref{sec-radius}
+varying $N$ from 200 to 10,000 for NKq and 196 to 6084 for Ising Spin Glass.
+In these plots we also include the black-box version of P3 to show the efficiency
+gains available for using gray-box information.
+
+Figure~\ref{fig-length-nn} and Figure~\ref{fig-length-is} show how long each
+algorithm required to reach the best found on Nearest Neighbor NKq
+and Ising Spin Glass, respectively. For Nearest Neighbor NKq the best found
+is the global optimum for all runs of all lengths, while Ising Spin Glass
+the best found by any method was worse than the global optimum in 7 runs of
+$N=4096$ and 21 runs of $N=6084$. The median run of the HBHC was unable to reach the best fitness
+for any problems tested using more than 200 bits. TUX performed somewhat better, reaching
+the best fitness on problem sizes up to $N=800$ and $N=625$ for Nearest Neighbor NKq and Ising
+Spin Glass, respectively. Black-Box P3, which does not utilize partial reevaluation or the HBHC,
+was able to consistently reach the best fitness until it it hit the memory limit on $N=2000$ for
+NKq and $N=2916$ for Ising Spin Glass. This is because unlike the other methods, Black-Box P3
+requires \BigO{N^2} memory.
+
+For all sizes of both problems, Gray-Box P3 was the fastest to reach the best fitness. On
+Nearest Neighbor NKq the improvement is very significant, with no alternative finishing within
+two orders of magnitude. On its largest successful problem size, the mean time to completion for
+Black-Box P3 was 375 times slower than Gray-Box P3. This is especially impressive considering
+previous work has shown Black-Box P3 is faster to reach the global optimum than other leading black-box
+methods. Applying regression to both lines, we estimate that Black-Box P3's time to global optimum
+on Nearest Neighbor NKq is \BigO{N^{2.75}} while Gray-Box P3's is \BigO{N^{1.98}}.
+
+The results on Ising Spin Glass are similar, with a less extreme difference between
+Black-Box P3 and Gray-Box P3. In general Gray-Box is the fastest technique to
+find the global optimum by an order of magnitude, with Black-Box P3's mean run finishing
+4.6 times slower than Gray-Box on $N=2025$. The regression line suggests that while
+Gray-Box P3 scales at \BigO{N^{3.35}}, Black-Box P3 scales at \BigO{N^{3.05}}.
 
@@
   \includegraphicsfit{length-un}
-  \caption{Length-un}
+  \caption{Relative qualities of each method as problem size increases on Unrestricted NKq
+  with $K=4$.}
   \label{fig-length-un}
@@
 
-TODO Consider O(N) statistics
+As Unrestricted NKq does not have a known global optimum, and the different algorithms
+rarely found the same best fitness, Figure~\ref{fig-length-un} compares the median error
+for each technique at termination. Gray-Box P3 in general finds the best fitness,
+with TUX occasionally performing better. When $N>2000$, Gray-Box P3 finds better quality
+solutions that all other methods for every single instance.  HBHC and Black-Box P3 only reach similar
+qualities as TUX and Gray-Box P3 when $N=200$, doing so in 4 and 7 runs, respectively. As the problem
+size increase TUX begins to fall behind Gray-Box P3, with the HBHC stabilizing at about 2.5\% worse than
+the best found. Here Black-Box P3 performs worse than the other techniques,
+falling further behind as the problem size increases.
 
-\section{Conclusions and Future Work}
+\section{Discussion and Conclusions}
+In line with previous work, we have found that HBHC cannot effectively find global optima
+on problems with even moderate epistasis. Furthermore, in general is obtains almost no improvement
+in fitness after only a few restarts. We designed TUX as a simplistic way of choosing
+restart locations based on previously found local optima. Unlike HBHC alone, TUX
+was able to continue improving given more time, finding global optima on problems three
+times as large.
+
+TUX's effectiveness is likely due to the HBHC acting as a super repair operator for uniform crossover.
+Given a sufficiently large $r$ the HBHC can return sections of the crossover offspring to either
+parents original version of a given subfunction. The HBHC is elitist meaning there is a bias toward
+returning to the better of the two parent's versions. Furthermore, by being so disruptive,
+uniform crossover potentially allows for the HBHC to find unrelated improvements.
+
+While TUX improves over plain HBHC, Gray-Box P3 is required to perform truly successful
+global optimization. Gray-Box P3 replaces na\"{i}ve local search the HBHC and utilizes
+known non-linear relationships instead of statistical linkage learning. On NKq-Landscapes this drastically
+improves search effectiveness. A major source of this improvement is likely how difficult it is
+for Black-Box P3 to learn linkage relationships on these landscapes. Furthermore, Gray-Box P3
+can perform partial reevaluation and efficient hill climbing during the mixing phase.
+
+Gray-Box P3's success is somewhat muted by less extreme improvement on Ising Spin Glass.
+While it still outperformed all competitors, Black-Box P3 may actually scale better to larger
+problems. One explanation for this deviation is that Ising Spin Glass problems require more
+exploration of equal fitness plateaus. For instance in Figure~\ref{fig-solver-over} and Figure~\ref{fig-radius-over}
+there is a significant pause in improvement when Gray-Box P3 reaches the second best fitness in the landscape.
+Nothing in its design suggests that Gray-Box P3 should be more effective at neutral drift. Another potential
+issue is that on these landscapes the importance of each non-linear relationship may be detectably unequal.
+As a result the Black-Box linkage learning may be able to better cluster variables which having meaningful
+impact on fitness while Gray-Box assumes all are equally important. A useful direction for future work
+would be to explore methods of performing efficient learning on top of the known variable iterations.
+
+Somewhat surprising is the difference in behavior between the polynomially solvable problems
+and Unrestricted NKq. While Black-Box P3 performed very well in the former, it was the least
+successful in the latter. The optimal radius for Gray-Box P3 also shifted from 1 to 2. One
+potential explanation is the change in topological bottlenecking. In both Ising Spin Glass
+and Nearest Neighbor NKq the number of variables reachable from any one variable in $r$
+or less steps is significantly lower than the worst case. This explains why on Unrestricted NKq
+even moderately high $r$ values hit our memory limit, and may be causing Black-Box P3 to have
+difficulty learning clusters.
+
 %

### 2015-01-28 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
Wording fixes and commas

@@
 running time and memory usage decreased by up to a linear factor from Black-Box P3.
-On NKq this results in a x375 speedup for problems with at least 1000 variables.
+On NKq this results in a x375 speedup for problems with at least 1,000 variables.
 While local search introduces a parameter, we experimentally show this parameter
@@
 
-\keywords{TODO; Local Search; Parameter-less}
+\keywords{Local Search; Global Search; Parameter-less}
 
@@
 of populations. Figure~\ref{fig-p3} provides the high level description of how the algorithm
-works. In order to perform solution mixing, P3 learns a linkage tree from pairwise variable
-entropy which is then used to perform global optima mixing~\cite{thierens:2011:gomea}. In global
-optimal mixing these learned clusters of variables are used to donate genetic information from multiple
-sources into a single solution. Each time variables are donated, the modified solution is evaluated,
+works. To exploit information from multiple solutions, P3 performs global optimal mixing~\cite{thierens:2011:gomea}.
+Each time a solution is added to a level of the pyramid $P_i$, linkage learning is performed to
+create a collection of variable clusters. These clusters, collectively referred to as a linkage tree,
+are learned from the pairwise variable entropy of $P_i$. When a solution is mixed with $P_i$
+each cluster is used to donate variable values from a random donor in $P_i$ into the solution.
+After each donation the modified solution is evaluated,
 with the change being reverted if solution quality decreased. In this way P3 performs
-multiple evaluations per mixing event, with a single solution receiving genes from randomly chosen
-donors in the pyramid level. This also means that unlike TUX, which uses information from only two
+multiple evaluations per mixing event. This also means that unlike TUX, which uses information from only two
 solutions to perform crossover, P3 uses information from an entire population.
@@
 a random starting vertex until a desired number of unique vertices have been explored.
-Cluster sizes are set by recursively such that for every cluster created of size $l>1$
-clusters of size $a$ and $l-a$ are also created, such that $a$ is chosen uniformly
-from the range $[1..l-1]$, with the largest clusters created by similarly splitting $l=N$.
+Cluster sizes are set recursively. For each cluster of size $l>1$
+a cluster of size $a$ and a cluster of size $l-a$ are also created, with $a$ chosen uniformly
+from the range $[1..l-1]$. This recursive process begins by similarly splitting $l=N$.
 
@@
 This efficiency allows new clusters to be created before every mixing event,
-unlike black-box P3 where they are created only when new solutions are added to the population.
+unlike Black-Box P3 where they are created only when new solutions are added to the population.
 For simplicity the clusters are shuffled after the are created, not sorted on size like in
-black-box P3.
+Black-Box P3.
 
@@
 clusters than those on the periphery. If all else is equal, the more paths of a given
-size between two variables, the higher the probability of them being in the same cluster.
-Unlike black-box P3, this linkage tree does not require clusters to be nested, allowing
+length between two variables, the higher the probability of them being in the same cluster.
+Unlike Black-Box P3, this linkage tree does not require clusters to be nested, allowing
 more diversity in the types of clusters appearing in a single tree.
@@
 would make if $r \ge l$. As clusters are used to move values between solutions, these
-moves are restricted to moving in the direction of previously found high quality moves.
+moves are restricted to moving in the direction of previously found high quality solutions.
 As a result if the density of high quality solutions is higher between good solutions than
@@
 Figure~\ref{fig-sfx-tree} provides an efficient method for creating a linkage
-tree similar to that used in black-box P3 given subfunction information.
+tree similar to that used in Black-Box P3 given subfunction information.
 Each variable starts in its own cluster. Each subfunction in a random order
@@
 to determine the change in fitness is only \BigO{l}. This also
-allows for gray-box P3 to efficiently reapply hill climbing after each
+allows for Gray-Box P3 to efficiently reapply hill climbing after each
 donation as only effected moves need to be rechecked. As a result a donation,
@@
 All combined a single donation plus returning to a local optimum requires \BigO{l + I} time,
-while just the donation in black-box P3 requires \BigO{N}.
+while just the donation in Black-Box P3 requires \BigO{N}.
 
@@
 is also the global optimum, verified using dynamic programming. The same is true of all Ising Spin Glass
-problems except for 6 of 50 on N=6084.
+problems except for 6 of 50 on N=6,084.
 
 All figures report the median run for either percentage error or seconds to reach the best. A run's percentage error is
-equal to its how much less its fitness is than the best, divided by the best. When reporting seconds
+equal to how much less its fitness is than the best, divided by the best. When reporting seconds
 to reach the best fitness, any run which did not find the best fitness is treated as slower than any run that did.
@@
   \includegraphicswide{fitness}
-  \caption{Comparison of how radius effects solution quality at termination. For NKq-Landscapes $N=6000$ and $K=4$ and
-  for Ising Spin Glasses $N=6084$.}
+  \caption{Comparison of how radius effects solution quality at termination. For NKq-Landscapes $N=6,000$ and $K=4$ and
+  for Ising Spin Glasses $N=6,084$.}
   \label{fig-fitness}
@@
   \includegraphicsfit{p3-seconds}
-  \caption{Median time required for Gray-Box P3 to reach the global optimum of Nearest Neighbor NKq instances with $N=6000$.}
+  \caption{Median time required for Gray-Box P3 to reach the global optimum of Nearest Neighbor NKq instances with $N=6,000$.}
   \label{fig-p3-seconds}
@@
   \includegraphicswide{solver-over}
-  \caption{Comparison of solution quality during optimization on a log-log scale for different algorithms. For NKq-Landscapes $N=6000$ and $K=4$ and
-  for Ising Spin Glasses $N=6084$. Each algorithm uses its best found $r$ value.}
+  \caption{Comparison of solution quality during optimization on a log-log scale for different algorithms. For NKq-Landscapes $N=6,000$ and $K=4$ and
+  for Ising Spin Glasses $N=6,084$. Each algorithm uses its best found $r$ value.}
   \label{fig-solver-over}
@@
   \caption{Comparison of Gray-Box P3's solution quality during optimization on a log-log scale for different $r$ values.
-  For NKq-Landscapes $N=6000$ and $K=4$ and for Ising Spin Glasses $N=6084$.}
+  For NKq-Landscapes $N=6,000$ and $K=4$ and for Ising Spin Glasses $N=6,084$.}
   \label{fig-radius-over}
@@
 we ran all three algorithms using the best found $r$ values in Section~\ref{sec-radius}
-varying $N$ from 200 to 10,000 for NKq and 196 to 6084 for Ising Spin Glass.
+varying $N$ from 200 to 10,000 for NKq and 196 to 6,084 for Ising Spin Glass.
 In these plots we also include the black-box version of P3 to show the efficiency
@@
 the best found by any method was worse than the global optimum in 7 runs of
-$N=4096$ and 21 runs of $N=6084$. The median run of the HBHC was unable to reach the best fitness
+$N=4,096$ and 21 runs of $N=6,084$. The median run of the HBHC was unable to reach the best fitness
 for any problems tested using more than 200 bits. TUX performed somewhat better, reaching
@@
 Spin Glass, respectively. Black-Box P3, which does not utilize partial reevaluation or the HBHC,
-was able to consistently reach the best fitness until it it hit the memory limit on $N=2000$ for
-NKq and $N=2916$ for Ising Spin Glass. This is because unlike the other methods, Black-Box P3
+was able to consistently reach the best fitness until it it hit the memory limit on $N=2,000$ for
+NKq and $N=2,916$ for Ising Spin Glass. This limitation is because unlike the other methods, Black-Box P3
 requires \BigO{N^2} memory.
@@
 previous work has shown Black-Box P3 is faster to reach the global optimum than other leading black-box
-methods. Applying regression to both lines, we estimate that Black-Box P3's time to global optimum
+methods. Applying regression, we estimate that Black-Box P3's time to global optimum
 on Nearest Neighbor NKq is \BigO{N^{2.75}} while Gray-Box P3's is \BigO{N^{1.98}}.
@@
 find the global optimum by an order of magnitude, with Black-Box P3's mean run finishing
-4.6 times slower than Gray-Box on $N=2025$. The regression line suggests that while
+4.6 times slower than Gray-Box on $N=2,025$. The regression line suggests that while
 Gray-Box P3 scales at \BigO{N^{3.35}}, Black-Box P3 scales at \BigO{N^{3.05}}.
@@
 for each technique at termination. Gray-Box P3 in general finds the best fitness,
-with TUX occasionally performing better. When $N>2000$, Gray-Box P3 finds better quality
+with TUX occasionally performing better. When $N>2,000$, Gray-Box P3 finds better quality
 solutions that all other methods for every single instance.  HBHC and Black-Box P3 only reach similar
@@
 In line with previous work, we have found that HBHC cannot effectively find global optima
-on problems with even moderate epistasis. Furthermore, in general is obtains almost no improvement
+on problems with even moderate epistasis. In general it also obtains almost no improvement
 in fitness after only a few restarts. We designed TUX as a simplistic way of choosing
@@
 Given a sufficiently large $r$ the HBHC can return sections of the crossover offspring to either
-parents original version of a given subfunction. The HBHC is elitist meaning there is a bias toward
+parents' original version of a given subfunction. The HBHC is elitist meaning there is a bias toward
 returning to the better of the two parent's versions. Furthermore, by being so disruptive,
-uniform crossover potentially allows for the HBHC to find unrelated improvements.
+uniform crossover potentially allows for the HBHC to also find unrelated improvements.
 
 While TUX improves over plain HBHC, Gray-Box P3 is required to perform truly successful
-global optimization. Gray-Box P3 replaces na\"{i}ve local search the HBHC and utilizes
+global optimization. Gray-Box P3 replaces na\"{i}ve local search with the HBHC and utilizes
 known non-linear relationships instead of statistical linkage learning. On NKq-Landscapes this drastically
@@
 impact on fitness while Gray-Box assumes all are equally important. A useful direction for future work
-would be to explore methods of performing efficient learning on top of the known variable iterations.
+would be to explore methods of performing efficient learning on top of the known variable interactions.
 
@@
 or less steps is significantly lower than the worst case. This explains why on Unrestricted NKq
-even moderately high $r$ values hit our memory limit, and may be causing Black-Box P3 to have
-difficulty learning clusters.
+even moderately high $r$ values hit our memory limit.
+For Black-Box P3 this may also be causing increased difficulty in linkage learning as variables
+become indirectly dependent on much larger sets.
+
+While the inclusion of HBHC into Gray-Box P3 introduces a parameter value, it requires
+trivial configuration. In the worst case there may be a handful of $r$ values to test.
+Furthermore, our evidence suggests setting $r=1$ is quite powerful, with higher values
+likely to be only a time shift in quality. This is in contrast to $r$'s role in HBHC,
+where low $r$ values are never expected to reach the same quality has higher $r$ values.
+Therefore we conclude that Gray-Box P3 maintains the out-of-the-box quality of Black-Box
+P3, while drastically improving efficiency for this new domain of problems.
 

### 2015-01-28 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
Revisted the first half

@@
 
-Recent work~\cite{whitley:2013:greedy,chicano:2014:ball} has shown that by
-providing local search with more than just an evaluation function can lead
-to extraordinary improvements search power and efficiency. These improvements
+Recent work~\cite{whitley:2013:greedy,chicano:2014:ball} has shown that
+providing local search with more than just an evaluation function can
+extraordinarily improve search power and efficiency. These improvements
 require information easily derived for a broad range of interesting NP-Hard
@@
 and global search as simply as possible. To test the true power of this
-process we also extend the Parameter-less Population Pyramid~\cite{goldman:2014:p3},
+process we also extend the Parameter-less Population Pyramid (P3)~\cite{goldman:2014:p3}, the
 current state-of-the-art in black-box optimization. This ``Gray-Box P3''
@@
 Gray-box optimization exists between these two extremes, exposing additional
-information to be exploited by search and which is available for a variety of problems.
+information which is available for a variety of problems to be exploited by search.
 
@@
 a hill climber was developed which found the approximate next best improving move in \BigO{1} time.
-This is achieved by first determining the fitness effect $delta$ of making each move, and then updating
+This is achieved by first determining the fitness effect of making each move, and then updating
 only effected moves each time a change to the solution is made. Due to the requirements of the domain,
 at most \BigO{ck} moves can be effected, with each requiring at most \BigO{1} subfunction calls
-to update their $delta$. Combined with move binning, this technique is able to perform both
+to update their fitness effects. Combined with move binning, this technique is able to perform both
 first improvement and next best hill climbing in \BigO{I+N}, where $I$ is the number of improving
@@
 At each recursive level \Call{CISG}{} expands $open$ to include any vertices
-adjacent to $v$ in the graph. By construction this means that $open$ contains
+adjacent to $v$ in the graph. By construction this means that $open'$ contains
 all possible ways of adding a single vertex to the current $subset'$. As $v'$
@@
   \caption{One iteration of TUX optimization. $T$ is an
-           ordered list of solution, each of which could be empty,
+           ordered list of solutions, each of which could be empty,
            awaiting a crossover partner.}
@@
 
-Hamming-Ball Hill Climbing is not sufficient to efficiently find the global optimum
+Hamming-Ball Hill Climbing is not sufficient to efficiently find the global optima
 on problems with even moderate epistasis~\cite{chicano:2014:ball}. This is because,
 like all random restart hill climbers, it relies on random initialization to fall
-inside the global optimum's basin of attraction.
+inside the global optima's basin of attraction.
 
@@
 selection with equal probability uniform crossover to generate starting
-solutions likely to be in the global optimum's basin of attraction.
+solutions likely to be in the global optima's basin of attraction.
 
@@
 iteratively, storing a single list of solutions $T$, such that $|T|$ is equal to the height
-of the tournament. This is possible because when a sub-bracket is complete only a single solution
+of the tournament. Iterative construction is possible because when a sub-bracket is complete only a single solution
 emerges and solutions only need to be stored until their partner is found.
@@
 In effect the clusters are sampling moves which the HBHC
-would make if $r \ge l$. As clusters are used to move values between solutions, these
-moves are restricted to moving in the direction of previously found high quality solutions.
-As a result if the density of high quality solutions is higher between good solutions than
-outside of good solutions, this method will perform better than random search or sampling from
-all possible moves of size $l$.
+would make if $r \ge l$. For any solution in the search space which is not globally optimal, there
+must exist some move which will improve its quality. However, this move may
+be arbitrarily large and it is intractable to test all possible moves of even moderate size.
+By sampling from all possible large moves, we can maintain tractability while gaining
+a potentially non-zero probability of improvement.
+As clusters are used to donate values between solutions, these
+moves are always in the direction of previously found high quality solutions.
+This assumes that the density of high quality solutions is higher between good solutions than
+outside of good solutions.
 
@@
 and Unrestricted NKq. While Black-Box P3 performed very well in the former, it was the least
-successful in the latter. The optimal radius for Gray-Box P3 also shifted from 1 to 2. One
-potential explanation is the change in topological bottlenecking. In both Ising Spin Glass
-and Nearest Neighbor NKq the number of variables reachable from any one variable in $r$
-or less steps is significantly lower than the worst case. This explains why on Unrestricted NKq
+successful in the latter. The optimal radius for Gray-Box P3 also shifted from 1 to 2.
+A potential explanation is that in both Ising Spin Glass
+and Nearest Neighbor NKq the number of unique variables $r$ steps
+or less steps away from a given variable is significantly lower than the worst case. This explains why on Unrestricted NKq
 even moderately high $r$ values hit our memory limit.
 For Black-Box P3 this may also be causing increased difficulty in linkage learning as variables
-become indirectly dependent on much larger sets.
+become indirectly dependent on much larger sets. Furthermore, Black-Box
+P3 may be benefiting from an increased rate of duplicate dependencies on Nearest Neighbor NKq
+not present in Unrestricted NKq.
 

### 2015-01-30 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
Minor grammar fixes

@@
 Leveraging these restrictions, the Hamming-Ball Hill Climber
-(HBHC) can efficiently find r-bit local optimal. We show how a
+(HBHC) can efficiently find r-bit local optima. We show how a
 simple memetic algorithm in conjunction with HBHC can find \emph{global}
@@
 about a problem from being provided to search algorithms. At the other end of
-the spectrum is white box optimization, in which everything about the problem
+the spectrum is white-box optimization, in which everything about the problem
 is known allowing search techniques to specialize at the cost of generality.
@@
 Here we shall consider the gray-box optimization domain with the following
-characteristics. The function which evaluates the quality of a solution is determined by the summation
+characteristics: The function which evaluates the quality of a solution is determined by the summation
 of subfunctions. Each subfunction uses at most $k$ variables in the given solution,
@@
 in at least one subfunction and that the problem is not separable. Any problem which
-violates these requirements can be restated either using less variables or as multiple separate problems.
+violates these requirements can be restated either using fewer variables or as multiple separate problems.
 
@@
 described in Section~\ref{sec-gray-box} was \cite{whitley:2013:greedy}. In this study
-a hill climber was developed which found the approximate next best improving move in \BigO{1} time.
+a hill climber was developed that found the approximate next-best move in \BigO{1} time.
 This is achieved by first determining the fitness effect of making each move, and then updating
@@
 If $i$ and $j$ do not share a subfunction, the effect of flipping both variables
-must be equal to the sum of flipping both separately. Therefore only variables which share
+must be equal to the sum of flipping both separately. Therefore, only variables that share
 a subfunction must be flipped together when checking $r=2$. The problem of determining
@@
 
-Under the assertion that $r \ll N$ and that $r$ does not increase with $N$, the Hamming
-Ball Hill Climber requires only \BigO{I+N} time to find $r$-bit local optimum starting
+Under the assertion that $r \ll N$ and that $r$ does not increase with $N$, the
+Hamming-Ball Hill Climber requires only \BigO{I+N} time to find $r$-bit local optima starting
 from a random solution. As $I$ is likely in \BigO{N} this means finding random $r$-bit local optima
@@
 
-While these techniques make it possible to choose the best improving move just as efficiently
+While these techniques make it possible to choose the best move just as efficiently
 as choosing any random improving move, the findings of~\cite{whitley:2013:greedy} suggest that,
@@
 adjacent to $v$ in the graph. By construction this means that $open'$ contains
-all possible ways of adding a single vertex to the current $subset'$. As $v'$
-are tested they are temporarily added to $closed$ to prevent recursive calls
+all possible ways of adding a single vertex to the current $subset'$. As each $v'$
+is tested it is temporarily added to $closed$ to prevent recursive calls
 from creating duplicates.
@@
 All figures report the median run for either percentage error or seconds to reach the best. A run's percentage error is
-equal to how much less its fitness is than the best, divided by the best. When reporting seconds
+equal to the difference between its fitness and the best, divided by the best. When reporting seconds
 to reach the best fitness, any run which did not find the best fitness is treated as slower than any run that did.
@@
 and Nearest Neighbor NKq the number of unique variables $r$ steps
-or less steps away from a given variable is significantly lower than the worst case. This explains why on Unrestricted NKq
+or fewer steps away from a given variable is significantly lower than the worst case. This explains why on Unrestricted NKq
 even moderately high $r$ values hit our memory limit.

### 2015-02-02 CV/cv.tex
Minor CV update to ordering
Updated CV for journal acceptance

@@
     \section{\mysidestyle Research\\Interests}
-    Optimization, Artificial Intelligence, Evolutionary Computation, Machine Learning
+    Optimization, Machine Learning, Artificial Intelligence, Evolutionary Computation
 
@@
     ``Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid,''
-    \textbf{Invited Paper}, \textsl{Evolutionary Computation Journal} (Impact Factor 3.733), In Review
+    \textbf{Invited Paper}, \textsl{Evolutionary Computation Journal} (Impact Factor 3.733), Accepted With Revisions
 
@@
     \textsl{GECCO} (Acceptance Rate 33\%), pp. 785--792, 2014
-    \textbf{Won Best Paper in Track}
+    \textbf{Best paper award winner}
 

### 2015-02-02 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
Statistical Tests

@@
 Recent work~\cite{whitley:2013:greedy,chicano:2014:ball} has shown that
-providing local search with more than just an evaluation function can
+providing local search with more problem specific information than just an evaluation function can
 extraordinarily improve search power and efficiency. These improvements
@@
 
-Here we set out to extend these benefits to global optimization by integrating
+Here we extend those benefits to global optimization by integrating
 this local search with additional heuristics~\cite{chen:2011:memetic}.
@@
 such that as the total number of solution variables $N$ increases, $k$ remains constant.
-Each subfunction must be independently evaluable, and the variables each uses are known.
+Each subfunction is independently evaluable and the variables each subfunction uses are known.
 
@@
 
-The primary advantages of TUX is that it does not introduce any new parameters (still
+The primary advantages of TUX are that it does not introduce any new parameters (though it still
 requires an $r$ for the hill climber) and is relatively simple to implement. Even so
@@
 a cluster of size $a$ and a cluster of size $l-a$ are also created, with $a$ chosen uniformly
-from the range $[1..l-1]$. This recursive process begins by similarly splitting $l=N$.
+from the range $[1..l-1]$. This recursive process begins by initializing $l=N$ and splitting
+to form the first two clusters.
 
@@
 unlike Black-Box P3 where they are created only when new solutions are added to the population.
-For simplicity the clusters are shuffled after the are created, not sorted on size like in
+For simplicity the clusters are shuffled after they are created, not sorted on size like in
 Black-Box P3.
@@
 Beyond efficiency, there are good reasons to believe these clusters will be useful
-to search. The closer two variables are in the dependency graph, the more likely
+for search. The closer two variables are in the dependency graph, the more likely
 they are to appear in the same cluster. All variables on average are expected to appear
@@
 allows for Gray-Box P3 to efficiently reapply hill climbing after each
-donation as only effected moves need to be rechecked. As a result a donation,
-and its resulting modifications from hill climbing, are kept only if the
+donation as only effected moves need to be rechecked. As a result, a donation
+and its resulting modifications from hill climbing are kept only if the
 new local optima is at least as fit as the solution before the donation occurred.
@@
 
-\section{Experimentation}
+\section{Experiments}
 \label{sec-experiments}
@@
 each subfunction depends on: Nearest Neighbor NKq and Unrestricted NKq.
-In Nearest Neighbor NKq each variable depends on the $K$ variables which immediately follow
+In Nearest Neighbor NKq each variable depends on the $K$ variables which sequentially follow
 it in the solution, with dependencies wrapping around the end of the solution.
@@
 are NP-Hard, but the $2D\pm J$ subset is polynomially solvable~\cite{saul:1994:spinglass}\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}. In this
-subset the graph is a 2D toroidal grid, with each edge's weight either -1 or 1. In gray-box terms
+subset the graph is a 2D toroidal grid, with each edge's weight either -1 or 1. In gray-box terms,
 problems of this subset have $k=2$ and $c=4$ regardless of $N$. While Ising Spin Glasses are traditionally
@@
 from our website.\footnote{\url{https://github.com/name_of_repository/anonymous_name_of_project}}
-Each time the run achieved a new best fitness we record current amount of processing time used.
+Each time the run achieved a new best fitness we recorded the current amount of processing time used.
 Timing includes the discovery of subgraphs to allow for comparison between different radius values.
@@
 the time limit is reached. On all Nearest Neighbor NKq instances the ``best'' fitness
-is also the global optimum, verified using dynamic programming. The same is true of all Ising Spin Glass
-problems except for 6 of 50 on N=6,084.
+is also the global optimum, verified using dynamic programming. For Ising Spin Glass the ``best'' fitness
+was the global optimum 44 of 50 times.
 
@@
 outperforms HBHC on all three problems for all $r$ values. Regardless of $r$, Gray-Box P3 outperforms both, with almost
-all $r$ values reaching the same best fitness. For Nearest Neighbor NKq Gray-Box P3 finds the global
+all $r$ values reaching the same best fitness. For Nearest Neighbor NKq, Gray-Box P3 finds the global
 optimum in every run for $r < 4$, with only a single unsuccessful run at $r=4$.
@@
 
-In optimization its sometimes more important how quickly an algorithm reaches high quality solutions than when
-it reaches the global optimum. Figure~\ref{fig-solver-over} shows how solution quality progresses during
+Reaching high quality solutions quickly can sometimes be more important than reaching the global optimum
+eventually. Figure~\ref{fig-solver-over} shows how solution quality progresses during
 search for each algorithm. HBHC and TUX have significant early delays caused by their high $r$ values. Larger
@@
 
-To further illustrate the effect of $r$ on Gray-Box P3, we provide Figure~\ref{fig-radius-over}. On Nearest Neighbor NKq
+Figure~\ref{fig-radius-over} further illustrates the effect of $r$ on Gray-Box P3. On Nearest Neighbor NKq
 and Ising Spin Glasses, increasing $r$ does not change the shape of the curve. Instead the quality reached is simply time shifted,
-such that given more time higher $r$ values will reach the same quality. As a result for these problems we conclude
-that searching for an optimal value of $r$ is almost certainly going to be more expensive than simply fixing $r=1$.
+such that given more time higher $r$ values will reach the same quality. As a result, for these problems we conclude
+that higher $r$ values simply add more expense for no overall gain.
 On Unrestricted NKq this relationship is less certain, with $r=1$ potentially having a different, and worse, shape than $r>1$.
@@
 how it scales as problem difficulty increases. To test this behavior,
-we ran all three algorithms using the best found $r$ values in Section~\ref{sec-radius}
+we ran all three algorithms using the optimal $r$ values determined experimentally in Section~\ref{sec-radius}
 varying $N$ from 200 to 10,000 for NKq and 196 to 6,084 for Ising Spin Glass.
@@
 Figure~\ref{fig-length-nn} and Figure~\ref{fig-length-is} show how long each
-algorithm required to reach the best found on Nearest Neighbor NKq
-and Ising Spin Glass, respectively. For Nearest Neighbor NKq the best found
-is the global optimum for all runs of all lengths, while Ising Spin Glass
-the best found by any method was worse than the global optimum in 7 runs of
+algorithm required to reach the best overall quality found on Nearest Neighbor NKq
+and Ising Spin Glasses, respectively. For Nearest Neighbor NKq the best found
+is the global optimum for all runs of all lengths, while for Ising Spin Glasses
+the best quality found by any method was worse than the global optimum in 7 runs of
 $N=4,096$ and 21 runs of $N=6,084$. The median run of the HBHC was unable to reach the best fitness
@@
 was able to consistently reach the best fitness until it it hit the memory limit on $N=2,000$ for
-NKq and $N=2,916$ for Ising Spin Glass. This limitation is because unlike the other methods, Black-Box P3
-requires \BigO{N^2} memory.
+NKq and $N=2,916$ for Ising Spin Glass. This limitation is due to Black-Box P3's
+\BigO{N^2} memory requirements.
 
@@
 two orders of magnitude. On its largest successful problem size, the mean time to completion for
-Black-Box P3 was 375 times slower than Gray-Box P3. This is especially impressive considering
+Black-Box P3 was 375 times slower than Gray-Box P3. Using the Mann-Whitney test to compare the two run times
+results in $p < 10^{-16}$. This is especially impressive considering
 previous work has shown Black-Box P3 is faster to reach the global optimum than other leading black-box
@@
 find the global optimum by an order of magnitude, with Black-Box P3's mean run finishing
-4.6 times slower than Gray-Box on $N=2,025$. The regression line suggests that while
+4.6 times slower than Gray-Box on $N=2,025$. Using the Mann-Whitney test to compare the two run times
+results in $p < 10^{-14}$. The regression line suggests that while
 Gray-Box P3 scales at \BigO{N^{3.35}}, Black-Box P3 scales at \BigO{N^{3.05}}.
@@
 with TUX occasionally performing better. When $N>2,000$, Gray-Box P3 finds better quality
-solutions that all other methods for every single instance.  HBHC and Black-Box P3 only reach similar
+solutions that all other methods for every instance.  Using the Mann-Whitney test to compare the fitness
+of Gray-Box P3 and TUX when $N=10,000$ results in $p < 10^{-16}$. HBHC and Black-Box P3 only reach similar
 qualities as TUX and Gray-Box P3 when $N=200$, doing so in 4 and 7 runs, respectively. As the problem
@@
 
-Gray-Box P3's success is somewhat muted by less extreme improvement on Ising Spin Glass.
+Gray-Box P3's success is less dramatic on Ising Spin Glasses.
 While it still outperformed all competitors, Black-Box P3 may actually scale better to larger

### 2015-02-02 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
Removed last sentence of abstract

@@
 On NKq this results in a x375 speedup for problems with at least 1,000 variables.
-While local search introduces a parameter, we experimentally show this parameter
-can likely be fixed to 1 without significant loss in search quality.
+%While local search introduces a parameter, we experimentally show this parameter
+%can likely be fixed to 1 without significant loss in search quality.
 \end{abstract}
@@
 
-\newpage
+%\newpage
 
@@
 
+Figure~\ref{fig-radius-over} further illustrates the effect of $r$ on Gray-Box P3. On Nearest Neighbor NKq
+and Ising Spin Glasses, increasing $r$ does not change the shape of the curve. Instead the quality reached is simply time shifted,
+such that given more time higher $r$ values will reach the same quality. As a result, for these problems we conclude
+that higher $r$ values simply add more expense for no overall gain.
+On Unrestricted NKq this relationship is less certain, with $r=1$ potentially having a different, and worse, shape than $r>1$.
+However, due to memory and time restrictions it is difficult to know if this trend continues.
+
 \begin{figure*}
@@
 
-Figure~\ref{fig-radius-over} further illustrates the effect of $r$ on Gray-Box P3. On Nearest Neighbor NKq
-and Ising Spin Glasses, increasing $r$ does not change the shape of the curve. Instead the quality reached is simply time shifted,
-such that given more time higher $r$ values will reach the same quality. As a result, for these problems we conclude
-that higher $r$ values simply add more expense for no overall gain.
-On Unrestricted NKq this relationship is less certain, with $r=1$ potentially having a different, and worse, shape than $r>1$.
-However, due to memory and time restrictions it is difficult to know if this trend continues.
-
 \subsection{Scalability}
-\begin{figure}
-  \centering
-  \includegraphicsfit{length-nn}
-  \caption{Comparison of how each algorithm's time required to reach the best fitness found scales with problem size
-  on Nearest Neighbor NKq with $K=4$.}
-  \label{fig-length-nn}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphicsfit{length-is}
-  \caption{Comparison of how each algorithm's time required to find the best fitness found scales with problem size
-  on Ising Spin Glass.}
-  \label{fig-length-is}
-\end{figure}
-
 Perhaps the most critical test of an optimization algorithm's quality is
@@
 
+\begin{figure}
+  \centering
+  \includegraphicsfit{length-nn}
+  \caption{Comparison of how each algorithm's time required to reach the best fitness found scales with problem size
+  on Nearest Neighbor NKq with $K=4$.}
+  \label{fig-length-nn}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphicsfit{length-is}
+  \caption{Comparison of how each algorithm's time required to find the best fitness found scales with problem size
+  on Ising Spin Glass.}
+  \label{fig-length-is}
+\end{figure}
+
 For all sizes of both problems, Gray-Box P3 was the fastest to reach the best fitness. On
@@
 
-\begin{figure}
-  \centering
-  \includegraphicsfit{length-un}
-  \caption{Relative qualities of each method as problem size increases on Unrestricted NKq
-  with $K=4$.}
-  \label{fig-length-un}
-\end{figure}
-
 As Unrestricted NKq does not have a known global optimum, and the different algorithms
@@
 falling further behind as the problem size increases.
+
+\begin{figure}
+  \centering
+  \includegraphicsfit{length-un}
+  \caption{Relative qualities of each method as problem size increases on Unrestricted NKq
+  with $K=4$.}
+  \label{fig-length-un}
+\end{figure}
 

### 2015-02-02 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
"Dependency" graph
Minor wording fixes. Switched to 375x from x375.

@@
 simple memetic algorithm in conjunction with HBHC can find \emph{global}
-optima for gray-box problems. We also develop a gray-box version of the Parameter-less
-Population Pyramid (P3) utilizing both the HBHC and the known information about variable relationships
+optima for some gray-box problems. We also develop a gray-box version of the Parameter-less
+Population Pyramid (P3), utilizing both the HBHC and the known information about variable relationships,
 which outperforms all of the examined algorithms.
@@
 running time and memory usage decreased by up to a linear factor from Black-Box P3.
-On NKq this results in a x375 speedup for problems with at least 1,000 variables.
+On NKq this results in a 375x speedup for problems with at least 1,000 variables.
 %While local search introduces a parameter, we experimentally show this parameter
@@
 the spectrum is white-box optimization, in which everything about the problem
-is known allowing search techniques to specialize at the cost of generality.
+is known. Search techniques in this domain specialize at the cost of generality.
 Gray-box optimization exists between these two extremes, exposing additional
@@
 In order to obtain efficiency bounds for optimization algorithms, the following
-assertions are also included. The total number of subfunctions grows no more than
+assertions are also included: The total number of subfunctions grows no more than
 linearly with $N$, which holds true for randomized MAX-SAT, mesh spin glasses, and NK Landscapes.
@@
 in slower than predicted performance.
+%TODO Are these necessary?
 We also include the trivial requirements that all problem variables must participate
@@
 to update their fitness effects. Combined with move binning, this technique is able to perform both
-first improvement and next best hill climbing in \BigO{I+N}, where $I$ is the number of improving
+first improvement and next-best hill climbing in \BigO{I+N}, where $I$ is the number of improving
 moves required to reach a local optimum. In comparison, the black-box approach for these searches can require
@@
 but no algorithm was given to efficiently discover these moves. This process was not included
-in their timing calculations as it only has to be performed once per problem.
+in their timing calculations as it only has to be performed once per problem instance.
 
@@
 When applied to the sparse graphs inherent in the gray-box domain, this algorithm
-requires \BigO{r!(c(k-1))^rN} time, which reduces to \BigO{N}. The time spent
+requires \BigO{r!(ck)^rN} time, which reduces to \BigO{N}. The time spent
 in each call is dominated by the loop over $open'$. In the worst case, $open'$
-increases in size by the full adjacency of $v$, which is bounded by $c(k-1)$.
+increases in size by the full adjacency of $v$, which is bounded by $ck$.
 This creates a worse case complexity for a single top level call of
-$\prod_{i}^{r} ic(k-1) = r!(c(k-1))^r$. This must be called once for each
-of the $N$ variables resulting in \BigO{r!(c(k-1))^rN}.
+$\prod_{i}^{r} ick = r!(ck)^r$. This must be called once for each
+of the $N$ variables resulting in \BigO{r!(ck)^rN}.
 
@@
 we have developed a method for creating a linkage tree which learns clusters
-from the same graph defined in Section~\ref{sec-hamming}. The goal is for each
+from the same dependency graph defined in Section~\ref{sec-hamming}. The goal is for each
 cluster to be a connected induced subgraph, with the size of clusters
@@
 
-Unrestricted NKq landscapes draw the $K$ dependencies at random. For $K > 1$ it is
+Unrestricted NKq landscapes draw the $K$ dependencies at random without replacement. For $K > 1$ it is
 NP-Hard to find the global optimum of these landscapes. This also means that while
@@
 
-To further examine the effect of $r$ on Gray-Box-P3 Figure~\ref{fig-p3-seconds} shows the median number of
+To further examine the effect of $r$ on Gray-Box-P3, Figure~\ref{fig-p3-seconds} shows the median number of
 seconds required to reach the global optimum. Setting $r=1$ was the most efficient configuration for all $K > 1$,
@@
 how it scales as problem difficulty increases. To test this behavior,
-we ran all three algorithms using the optimal $r$ values determined experimentally in Section~\ref{sec-radius}
+we ran all three algorithms using the optimal $r$ values determined experimentally in Section~\ref{sec-radius},
 varying $N$ from 200 to 10,000 for NKq and 196 to 6,084 for Ising Spin Glass.
@@
 for any problems tested using more than 200 bits. TUX performed somewhat better, reaching
-the best fitness on problem sizes up to $N=800$ and $N=625$ for Nearest Neighbor NKq and Ising
+the best fitness more than half of the time on problem sizes up to $N=800$ and $N=625$ for Nearest Neighbor NKq and Ising
 Spin Glass, respectively. Black-Box P3, which does not utilize partial reevaluation or the HBHC,
@@
 two orders of magnitude. On its largest successful problem size, the mean time to completion for
-Black-Box P3 was 375 times slower than Gray-Box P3. Using the Mann-Whitney test to compare the two run times
+Black-Box P3 was 375 times slower than Gray-Box P3. Using the Mann-Whitney test to compare their run times
 results in $p < 10^{-16}$. This is especially impressive considering
@@
 find the global optimum by an order of magnitude, with Black-Box P3's mean run finishing
-4.6 times slower than Gray-Box on $N=2,025$. Using the Mann-Whitney test to compare the two run times
+4.6 times slower than Gray-Box on $N=2,025$. Using the Mann-Whitney test to compare their run times
 results in $p < 10^{-14}$. The regression line suggests that while
@@
 While it still outperformed all competitors, Black-Box P3 may actually scale better to larger
-problems. One explanation for this deviation is that Ising Spin Glass problems require more
+problems. One explanation for this deviation is that Ising Spin Glasses require more
 exploration of equal fitness plateaus. For instance in Figure~\ref{fig-solver-over} and Figure~\ref{fig-radius-over}

### 2015-02-04 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
Updated abstract

@@
 %Maximum 200 Words
-
 Unlike black-box optimization problems, gray-box optimization problems have known,
-limited-cardinality non-linear relationships between variables. This domain contains many
-real world problems with examples in network security, computational biology, VLSI design, and statistical physics.
+limited, non-linear relationships between variables. Though more restrictive, gray-box problems include many
+real-world applications in network security, computational biology, VLSI design, and statistical physics.
 Leveraging these restrictions, the Hamming-Ball Hill Climber
-(HBHC) can efficiently find r-bit local optima. We show how a
-simple memetic algorithm in conjunction with HBHC can find \emph{global}
-optima for some gray-box problems. We also develop a gray-box version of the Parameter-less
-Population Pyramid (P3), utilizing both the HBHC and the known information about variable relationships,
-which outperforms all of the examined algorithms.
+(HBHC) can efficiently find r-bit local optima. We show how
+1) a simple memetic algorithm in conjunction with HBHC can find \emph{global}
+optima for some gray-box problems and 2) a gray-box version of the Parameter-less
+Population Pyramid (P3), utilizing both the HBHC and the known
+information about variable relationships, outperforms all of the
+examined algorithms.
 

### 2015-02-17 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/reviews.tex
Added Bill's pass of the review document

\documentclass[]{article}
\usepackage{hyperref}

\begin{document}

We would like to first take the opportunity to thank our reviewers for their thorough and insightful comments. It is for this reason, the review process and diligent reviewers, that papers are of such high quality in Evolutionary Computation.

We believe that the comments can be broken into three classes:
\begin{itemize}
\item Errors in the paper that need correcting
\item Minor fixes/modifications that we can implement.
\item Items of disagreement
\end{itemize}

\section{Errors}
\begin{quote}
 When testing the LTGA outside P3 however, the code is still recomputing the entropies to update the mutual information matrix. This is highly inefficient, the original LTGA only has to recompute the mutual information matrix once in each generation, not whenever a better solution is found. The experiments with LTGA should be run with the original LTGA implementation, not the modified P3 version. The code can be found at \url{http://homepages.cwi.nl/~bosman/source_code.php}
\end{quote}

This was indeed an error on our part. Unfortunately the source code was not made available to us, despite a request, and the code was not yet released to the public until after the paper was submitted. The fix, a two line modification, does not affect any of the evaluation counts but does indeed modify our timed comparisons (wall clock time experiments). Subsequently, after repairing the entropy calculation, we were able to reach timing parity with Dr. Bosman's implementation. We have rerun all of our timing results and <TODO updated our findings accordingly>. In general we find that for single instance problems, optimally tuned LTGA (where we of course do not count the evaluations or timing necessary to find the optimal population size) finds the global optimum using less seconds than P3. However, on randomly generated problem classes P3's reduction in evaluations overcomes its increased CPU usage per evaluation. We have updated our figures and discussion to reflect these changes.

\section{Fixes}
\subsection{Reviewer 1}
\begin{itemize}

\item
\begin{quote}
P3 seems to be very noisy in its results: the paper only reports median results after 100 runs. More information about variance and outliers (e.g. through boxplots) should be shown for at least one hard problem, for instance spin glasses. 
\end{quote}

In the original submission Section 9.1 explicitly discusses the variance in outcomes, and Figure 10 shows the upper and lower quartiles (just as boxplots do) for all 7 problems. We have, however, added a boxplot in this section to reflect this idea more clearly

\item
\begin{quote}
P3 stores all improving solutions ever found in a huge hash table. It should be reported how many solutions are typically stored and again what is the variance in these numbers. We have found that this varies a lot. 
\end{quote}

While the original work did not explictly report the aggregate value, you can estimate it in the original submission using the area under each curve in Figure 13a. We have added text to the Figure caption to reflect this

\item
\begin{quote}
My main point would be that their paper shows that the EA community should rethink its population management scheme and not blindly follow the fixed population size generational scheme.
\end{quote}

We agree, though our overall point is more general in that we are trying to get the community to rethink \textbf{all} parameter settings. In the original submission Section 9, Section 10.2,  Figure 11, and Figure 13a explicitly discuss the benefits we believe P3 achieves due to side stepping the fixed size generational scheme. We also had previously mentioned it in the third paragraph of the conclusion. Nonetheless, we have now added some additional references and discussion expanding those sections

\item
\begin{quote}
The abstract states that P3 requires "less wall clock seconds" and "is more effective than existing state-of-the-art techniques".
\end{quote}

This has been updated based on the new timing results

\item
\begin{quote}
 Self-adaptation has been around since the '70's in Evolution Strategies.
\end{quote}

Quite true. We have updated to more appropriate references.

\item
\begin{quote}
RRHC is used very often in the past, you cannot refer to it in this way.
\end{quote}

We have updated the text to clarify that we are referring to a specific RRHC and not suggesting it was the first RRHC

\item
\begin{quote}
"Doerr et al. (2013) presented the first genetic algorithm to provably show the advantage of performing crossover on simple landscapes."
I understand some people have a very restricted view on provably showing something but this statement is arguably wrong.
\end{quote}

We did not intend to be offensive and have cleaned up that text.

\item
\begin{quote}
It is not updating the model, it is completely rebuilding an $O(L^2)$ linkage tree model 
\end{quote}

Replaced "updating" with "rebuilding"

\item
\begin{quote}
"Furthermore, unlike LTGA all solutions in the population are used to generate the model, not just the winners of a binary tournament".
This is not a major design choice, this only influences the selection pressure. Actually, the most recent papers report all experiments with LTGA using the entire population instead of a set of selected solutions.
\end{quote}

As discussed in Section 2.5 we are aware that there are many variants of LTGA. However, for clarity and to perform experiments we had to choose a reference version, and Thierens and Bosman (2013) uses binary tournament. We updated the wording to indicate there are other versions.

\item
\begin{quote}
Fig. 4: ITERATE-P3, line 12: the model rebuilding should be stated here.
\end{quote}

Though we have stated previously that adding to the pyramid triggers modelbuilding, we have added text to the figure to restate it.

\item
\begin{quote}
We have seen problems where smallest first is not better. Basically any a priori ordering represents a bias to the LTGA search algorithm.
\end{quote}

Our argument is not that smallest first is the \textit{best}, only that it has desirable diversity preservation features which are problem independent.

\item
\begin{quote}
 "This function modifies the results of Equation 4 to include plateaus of size s, introducing an exponential number of local optima in each trap."
Exponential number ? Clarify with a small example of Deceptive Step Trap.
\end{quote}

We have added the following text for clarity. 
For k=7 and s=2, all traps with 0, 1, 3, 5, and 7 bits set are local optima. This means that half of all ways to set the trap are 1 bit local optima. This holds true as trap size increases, meaning the number of local optima grows at $2^{k-1}+1$

\item
\begin{quote}
HIFF: needs a reference to Watson
\end{quote}

Done

\item
\begin{quote}
Equation (6): what is A ?
\end{quote}

A= 10 (updated in paper)

\item
\begin{quote}
"To ensure each problem is challenging we chose a clause to variable ratio of 4.27"
Reference to support this number.
\end{quote}

Added

\item
\begin{quote}
Figure 8: This is a confusing figure: please elaborate.
\end{quote}

It is a busy figure. We have added more about the figure and its conclusions to the text.

\item
\begin{quote}
 "An important note is that each donation may range in size from a single bit up to N-1, with smaller donations far more likely to result in repeated attempts. As such this may cause some super-linear growth in P3, but its unlikely to be very high."
Speculative ?
\end{quote}

Due to the tree structure there are more small clusters than large ones. The nested nature of clusters means that if most solutions agree for a large cluster, at least that many solutions must agree for both of its subset clusters. As to the super-linear growth, we draw that conclusion from Figure 8b and its discussion.

\item
\begin{quote}
" In total these runs took over 6 computing years to complete."
\end{quote}

The quote is correct, but being taken slightly out of context. This is the report for \textbf{all} runs. In our original submission 95\% of all computing time was used by hBOA and Parameter-less hBOA. The other 4 algorithms combined took less than 100 days to complete. After fixing our LTGA inefficiency and rerunning on uniform, modern hardware, the non hBOA algorithms took in total 31 days of computing to complete 124 trillion evaluations.

We have updated the text to distinguish the different algorithm classes.

\item
\begin{quote}
How can this be only a constant factor, when P3 uses a $O(L^2)$ operation each time an improved solution is found (= thousands of times) versus only a dozen $O(L^2)$ operations for LTGA ???
\end{quote}

This is covered in the original text, Section 2.4 and 2.5. However, to review, the run time bottleneck for LTGA is extracting pairwise frequency information which requires $O(mu*N^2)$ time per generation. This happens approximately once every $mu*N$ evaluations, meaning the time per evaluation is $(mu*N^2) / (N^2) = O(N)$. In P3 both entropy extraction and model rebuilding is $O(N^2)$ and must be done every $O(N)$ evaluations, meaning the time per evaluation is $N^2 / N = O(N)$. Therefore, excluding caveats mentioned in the paper, the two techniques should be within a constant factor runtime of each other. This relationship in general holds true for our updated experiments.

\item
\begin{quote}
 "There is some evidence that even when compared on wall clock time, P3 performs at least as well as the best comparison techniques"
\end{quote}

Fixed based on the algorithmic fix. Now ``comparable'' instead of ``better''

\item
\begin{quote}
"While asymptotically linear"   Linear in what ?
\end{quote}

Fixed

\item
\begin{quote}
 Do not refer to TRs, refer to actual publications.
\end{quote}

Removed

\item
\begin{quote}
Add a reference, important modifications of LTGA you use were first published in:...
\end{quote}

Done
\end{itemize}

\subsection{Reviewer 2}
\begin{itemize}

\item
\begin{quote}
My only gripe is that all algorithms are compared according to their median results and it would be nice to see at least a discussion of what this implies in terms of statistical significance.
\end{quote}

Added explanation for why medians are the correct choice here. Added statistical tests to data from Figure 6.

\item
\begin{quote}
Last sentence of section 1, penultimate clause doesn't have  verb in it. I suggest you rephrase as " ...; analyse the algorithm complexity; 
\end{quote}

Fixed

\item
\begin{quote}
Around equation (1) you need to state what m(i), m(i,l), l' and l'' denote rather than leaving the reader to find those in the appendix.
\end{quote}

\item
\begin{quote}
Section 8.2.1 uses the phrase "significantly worse".  This suggests that you have some specific confidence in the results according to some hypothesis testing method.  Please clarify or rephrase.
\end{quote}

You are correct, you cannot test that statistically. Rephrased to avoid that impression.

\end{itemize}

\section{Disagreements}

There is some disagreement, reflected in multiple comments, that local search is either unnecessary or an unfair advantage when used in P3. Our view is this.

We have explained that, for the particular situation in the bottom pyramid level, before a randomly generated solution is added to that level, that solution should be hill climbed. Our reasoning is straightforward; a randomly generated solution is unlikely to have any useful linkage information to be incorporated into that level. Once hill-climbed, that solution is more likely to provide useful linkage information to that bottom level. There is no guarantee that \textbf{every} such hill-climbed solution will be improved, and to the extent that it is not the erroneous linkage information added at the bottom level will need to be repaired at higher levels.  The bottom line is that local search, in this particular role, is a fundamental aspect of P3 and P3's performance as we see it.

When we count evaluations and wall-clock time, we include the local search performance. That is, we are explicitly not allowing for partial evaluations, and count every local search evaluation P3 performs just like any other evaluation. Thus our timings and evaluations are fair and accurate.

Our job, as algorithm designers, is to compare our approach against the best possible competing algorithms and report how our approach fares. We have taken, to the best of our ability using available published resources, those best algorithms and  P3 and compare them each against a rigourous set of problems. In fact, we feel we went quite a bit out of our way to \textbf{tune} these algorithms so that we may compare on the best possible footing, in other words to be fair.

Our jobs is not to speculate on possible changes to these algorithms and how, given those changes, the ``new and improved'' algorithms might perform. If subsequent researchers wish to do so, we welcome their work. In fact, we hope that this paper inspires others to go in new directions (paramaterless algorithms, local search integration regardless of problem, etc.)

\end{document}

### 2015-02-17 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Part of the response to reviewers

@@
 comparison algorithms; compare both efficiency in reaching the global optimum and intermediate
-fitnesses; algorithm complexity analysis; and provide more in depth analysis of P3 itself.
+fitnesses; analyze algorithm complexity; and provide more in depth analysis of P3 itself.
 
@@
 and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{sec-appendix}, we
-derived a simplified form shown in Equation~\ref{eq-hboa-final}. If no proposed split satisfies the inequality, iteration stops.
+derived a simplified form shown in Equation~\ref{eq-hboa-final}.
+Here $l$ is a leaf in tree $i$, with $l'$ and $l''$ the results of splitting $l$. $m_i(l)$ is the number
+of solutions which reach $l$ and $m_i(x_i, l)$ is the number of solutions which reach $l$ with the given value
+for $x_i$.
+If no proposed split satisfies the inequality, iteration stops.
 If multiple splits do, whichever maximizes the right side is chosen.
@@
     \If{solution $\notin hashset$}
-      \State Add solution to $P_0$
+      \State Add solution to $P_0$ and rebuild model
       \State Add solution to $hashset$
@@
         \If{solution $\notin hashset$}
-          \State Add solution to $P_{i+1}$
+          \State Add solution to $P_{i+1}$ and rebuild model
           \State Add solution to $hashset$
@@
 on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
+Pairwise comparison of LTGA and P3 on the largest problem size using the Mann-Whitney~U test results in
+$p < 10^{-5}$ for all problems.
 
@@
 \subsubsection{Model Building}
-hBOA and Parameter-less hBOA perform significantly worse when using wall clock time as the unit of comparison than when using
+hBOA and Parameter-less hBOA perform much worse when using wall clock time as the unit of comparison than when using
 evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while, under reasonable
 assumptions, P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
-hBOA goes from being slightly more efficient than LTGA in terms of evaluations to significantly worse in terms of seconds.
+hBOA goes from being slightly more efficient than LTGA in terms of evaluations to three orders of magnitude worse in terms of seconds.
 As P3 and LTGA require a similar asymptotic complexity per
@@
 \section{Population Sizing}
+\begin{figure}[t]
+  \begin{centering}
+    \begin{subfigure}{.5\textwidth}
+      \begin{centering}
+        \includegraphicsfit{pop-p3}
+      \end{centering}
+      \caption{As problem size increases}
+      \label{fig-pop-p3}
+    \end{subfigure}%
+    \begin{subfigure}{.5\textwidth}
+      \begin{centering}
+        \includegraphicsfit{solutions-stored}
+      \end{centering}
+      \caption{Largest problem size}
+      \label{fig-solutions-stored}
+    \end{subfigure}
+  \end{centering}
+  \caption{The total number of solutions stored by P3 when the global optimum is found.
+           In Figure~\ref{fig-solutions-stored} the red ``+'' indicates LTGA's tuned population size.}
+  \label{fig-p3-storage}
+\end{figure}
+
 A major advantage to P3 is that it does not require the user to set a population size parameter.
@@
   \begin{center}
-  \includegraphicsfit{evals-to-success-range}
+  \includegraphicsfit{evals-to-success-boxplot}
   \end{center}
@@
            to reach the global optimum for P3 and LTGA with respect to problem size.}
-  \label{fig-evals-to-success-range}
+  \label{fig-evals-to-success-boxplot}
 \end{figure}

### 2015-02-17 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/reviews.tex
Part of the response to reviewers

@@
 
+Done
+
 \item
@@
 
-We have explained that, for the particular situation in the bottom pyramid level, before a randomly generated solution is added to that level, that solution should be hill climbed.
-Our reasoning is straightforward; a randomly generated solution is unlikely to have any useful linkage information to be incorporated into that level.
+We have explained that, for the particular situation in the bottom pyramid level,
+before a randomly generated solution is added to that level, that solution should be hill climbed.
+Our reasoning is straightforward; a randomly generated solution is unlikely to
+have any useful linkage information to be incorporated into that level.
 Once hill-climbed, that solution is more likely to provide useful linkage information to that bottom level.
-There is no guarantee that \textbf{every} such hill-climbed solution will be improved, and to the extent that it is not the erroneous linkage information added at the bottom level will need to be repaired at higher levels.
-The bottom line is that local search, in this particular role, is a fundamental aspect of P3 and P3's performance as we see it.
+This is true even for problems with large fitness plateaus, such as Deceptive Step Trap.
+As such we feel the application of local search should be considered the standard approach
+with P3, only to be removed under exceptional circumstances.
 
@@
 
-Our job, as algorithm designers, is to compare our approach against the best possible competing algorithms and report how our approach fares.
-We have taken, to the best of our ability using available published resources, those best algorithms and  P3 and compare them each against a rigourous set of problems.
-In fact, we feel we went quite a bit out of our way to \textbf{tune} these algorithms so that we may compare on the best possible footing, in other words to be fair.
+Our job, as algorithm designers, is to compare our approach against the best possible competing algorithms
+and report how our approach fares. We have taken, to the best of our ability using available published resources,
+those best algorithms and compared them with P3 on a rigorous set of problems.
+We went out of our way to tune these algorithms
+so that we may compare on the best possible footing, giving them an advantage not available to P3.
 
-Our jobs is not to speculate on possible changes to these algorithms and how, given those changes, the ``new and improved'' algorithms might perform.
-If subsequent researchers wish to do so, we welcome their work.
-In fact, we hope that this paper inspires others to go in new directions (paramaterless algorithms, local search integration regardless of problem, etc.)
+However, adding initial local search as a tunable parameter to hboa and LTGA when partial evaluations are not available is not part
+of existing work for either method. Doing so would represent novel research outside the scope of our topic, and provide those techniques
+with even more problem specific configuration. We welcome future work by other researchers exploring the possible effects of combining
+local search with their techniques.
 

### 2015-02-18 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Most reviewer changes made.

@@
 all five advanced comparison algorithms. This improvement includes requiring less evaluations
-and less wall clock seconds to find the global optimum and better fitness when using
+to find the global optimum and better fitness when using
 the same number of evaluations. Using both algorithm analysis and comparison we find P3's
@@
 There have been periodic efforts to reduce or remove the need for parameter tuning.
-\cite{Back:1992:selfadapt} introduced self-adaptive parameters, in which parameter values
+\cite{rechenberg:1973:es} introduced self-adaptive parameters, in which parameter values
 were included in each solution's genome and themselves underwent evolution. This allowed search
@@
 In order to fully understand the effectiveness of P3, we compare it with five advanced algorithms which
-have related features to P3. The Random Restart Hill Climber~\citep{goldman:2014:p3} was chosen as an efficient form of repeated
+have related features to P3. The Random Restart Hill Climber defined by ~\cite{goldman:2014:p3} was chosen as an efficient form of repeated
 local search. As P3 combines this hill climber with crossover,
 comparing with local search alone shows the advantages of P3's overall approach.
-The $1+(\lambda, \lambda)$ algorithm is the current best theory supported simple genetic algorithm
+The $1+(\lambda, \lambda)$ algorithm~\citep{doerr:2013:lambdalambda} is the current best theory supported simple genetic algorithm
 and its method of crossover is in some sense a macro-mutation just as in P3. hBOA and Parameter-less
@@
 \cite{doerr:2013:lambdalambda} presented the first genetic algorithm to provably show
-the advantages of performing crossover on simple landscapes. This comparatively simple
+the advantages of performing crossover on One Max. This comparatively simple
 algorithm maintains only a single individual and a self-controlled parameter $\lambda$.
@@
 GAs across many benchmarks. Since its introduction, many variants of LTGA have been
-proposed~\citep{goldman:2012:ltga} so for clarity we have chosen the version
+proposed~\citep{thierens:2011:gomea,goldman:2012:ltga} so for clarity we have chosen the version
 presented by \cite{thierens:2013:ltgahiff} as our model.
@@
 to create all offspring in that generation. As P3 does not perform serial
-generations, P3 instead updates the model each
+generations, P3 instead rebuilds the model each
 time a solution is added to a population. Furthermore, unlike
-LTGA all solutions in the population are used to generate the model, not just the
+our chosen variant of LTGA, all solutions in the population are used to generate the model, not just the
 winners of a binary tournament. We can do this because even the
@@
 This method applies gene clusters during crossover based on how many genes
-are in each cluster\footnote{Ties are broken randomly.}, and not on how tightly linked those genes are.
+are in each cluster,\footnote{Ties are broken randomly.} and not on how tightly linked those genes are.
 \cite{goldman:2012:ltga} found that this alternative was better at preserving
-diversity, and therefore required smaller populations, but has not yet been integrated into
-canonical LTGA.
+diversity, and therefore required smaller populations.
 
@@
 of Equation~\ref{eq-deceptive-trap} to include plateaus of size $s$, introducing an exponential number
-of local optima in each trap. As a result the Deceptive Step Trap is much more challenging for linkage
-learning techniques, while still being highly deceptive. We chose to use $s=2$ to create the maximum
-number of local optima.
+of local optima in each trap.
+With $k=7$ and $s=2$, as used in our experiments, all traps with 0, 1, 3, 5, and 7 bits set are local optima.
+This means that half of all ways to set the trap are 1 bit local optima.
+More generally, the number of local optima grows at $\Theta(2^{k-1})$.
+As a result, the Deceptive Step Trap is much more challenging for linkage
+learning techniques, while still being highly deceptive.
 \begin{equation}
@@
 Another challenging aspect of landscapes can be higher order relationships. The Hierarchical If
-and only If (HIFF) problem is designed to capture the difficulties of this class of problem.
+and only If (HIFF) problem~\citep{watson:1998:hiff} is designed to capture the difficulties of this class of problem.
 In HIFF the genome is broken up into a complete binary tree, such that each gene appears in exactly one
@@
 \begin{equation}
-  An + \sum_{i=1}^{n}\left [ x_i^2-A\cos (2\pi x_i) \right ] \forall x\in [-5.12,5.12]
+  10n + \sum_{i=1}^{n}\left [ x_i^2-10\cos (2\pi x_i) \right ] \forall x\in [-5.12,5.12]
   \label{eq-rast}
@@
 signs of the terms are set to ensure the target solution satisfies the clause. To ensure each
-problem is challenging we chose a clause to variable ratio of $4.27$.
+problem is challenging we chose a clause to variable ratio of $4.27$~\citep{selman:1996:maxsat}.
 
@@
 are treated as requiring more evaluations than any successful run. If the median run was not successful
-no point is shown. The maximum problem size used for each problem was set to be the largest size we could
+no point is shown.
+Medians are used as the data is not normally distributed, and because it allows for more meaningful comparison
+between techniques with different success rates.
+The maximum problem size used for each problem was set to be the largest size we could
 feasibly determine LTGA's optimal population size. For many larger problems results are not shown
@@
 increase in number of donations as problem size increases, potentially increasing algorithmic costs. An
-important note is that each donation may range in size from a single bit up to $N-1$, with smaller
-donations far more likely to result in repeated attempts. As such this may cause some super-linear growth
+important note is that each donation may range in size from a single bit up to $N-1$. However,
+repeated donation attempts are far more likely to happen with smaller clusters.
+As such this may cause some super-linear growth
 in P3, but its unlikely to be very high.
@@
 of 100 runs, with unsuccessful runs treated as slower than successful runs.
+\begin{comment}
 In total these runs
@@
 can still be made.
+\end{comment}
+These results were collected using 2.5GHz Intel Xeon E5-2670v2 processors.
 
@@
 \subsubsection{P3}
-Any increase in P3's algorithmic complexity is offset by its efficient use of evaluations.
-In aggregate P3 found the global optimum the fastest on 72 of 130 configurations,
-with LTGA, $1+(\lambda, \lambda)$, Hill Climber, hBOA outperforming P3 on 44, 18, 10, and 1 configurations, respectively.
-On
-Deceptive Trap, Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT P3 still finds the global
-optimum faster than all comparison techniques. LTGA is able to match P3's performance almost exactly on HIFF,
-and on Rastrigin P3 falls behind only on moderate sized problems. Deceptive Step Trap yields the worst performance,
-with LTGA using optimal population sizes having a significantly lower growth curve than P3. This mirrors
-the results from Figure~\ref{fig-evals-to-success} under the assumption that P3's computational time per evaluation is higher
-than LTGA but by no more than a constant factor.
+When LTGA is optimally tuned to a single instance problem with an efficient evaluation function
+it can find the global optimum faster than P3 in terms of wall clock time. However, on randomly
+generated problem classes P3's efficient use of evaluations is enough to overtake LTGA.
+
+On the four single instance problems LTGA not only finds the global optimum using less wall clock time,
+the factor speedup increases as problem length does. Na\"{i}vely this suggests LTGA is achieving
+a lower order of complexity. However, for these experiments LTGA is growing at sub-linear time per
+evaluation, which is not asymptotically stable due to (at minimum) the time required to perform an evaluation.
+We suspect that the true cause is that $N$ is small enough to be overshadowed by lower order polynomial terms.
+For example, LTGA requires $O(N/\mu)$ time per evaluation to rebuild the linkage
+model from the frequency table. As a result, for small $\mu$ model building, and not extracting pairwise
+frequency, can dominate runtime.
+
+When applied to randomly generated problem classes, the differences in P3 and LTGA's evaluation complexity
+dominates runtime complexity. Similar to with Figure~\ref{fig-evals-to-success}, the amount of speedup P3
+achieves over LTGA increases with problem size on Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT.
+
+Across both types of problems we find that P3's time per evaluation grows approximately linearly. As such,
+we conclude that P3 requires asymptotically similar amounts of time per evaluation as the other efficient techniques.
 
@@
 to initialization and no need to sacrifice intermediate fitness for eventual optimality.
+
+Figure~\ref{fig-pop-p3} shows how the number of total solutions stored in the pyramid changes
+as problem size increases, similar to Figure~\ref{fig-pop-sizes} for hBOA and LTGA's tuned population sizes.
+As expected, the number of concurrently stored solutions increases as problem difficult increases, with the
+exact behavior dependent on the problem landscape. Figure~\ref{fig-solutions-stored} examines how the number
+of solutions stored is distributed on the largest problem sizes. Here we see that the behavior depends
+on the type of problem. On single instance problems P3's stored variance is relatively low, and generally
+higher than optimally tuned LTGA's population size. On randomly generated problem classes P3 has a much
+higher variance in stored solutions, but in general requires smaller sizes than LTGA.
 
@@
   \end{center}
-  \caption{Comparison of the upper and lower quartiles of evaluations required
-           to reach the global optimum for P3 and LTGA with respect to problem size.}
+  \caption{Distribution of evaluations required to reach the global optimum for
+           P3 and LTGA on the largest size of each problem.}
   \label{fig-evals-to-success-boxplot}
@@
 
-Figure~\ref{fig-evals-to-success-range}
+Figure~\ref{fig-evals-to-success-boxplot}
 highlights how this can effect the required number of evaluations to reach the global optimum, showing
-the upper and lower quartiles for P3 and LTGA. On each problem
+the distribution of results when solving the largest size of each problem. On each problem
 LTGA has a much smaller difference between its best and worst runs. This makes sense as LTGA uses the same
 population size regardless of instance and progresses search generationally. In contrast P3 has a much
-higher split, with many runs finishing very quickly. This added variability is also generally on the lower
-quartile, with P3's upper quartile only worse than LTGA's on Deceptive Step Trap and Nearest
-Neighbor NK. This supports the hypothesis that P3 is able to scale its diversity not just
+higher split, with many runs finishing very quickly.
+On all problems except Deceptive Step Trap, P3's upper quartile is lower than LTGA's lower quartile.
+Furthermore, on Deceptive Trap, HIFF, and Ising Spin Glasses, P3's worst run is better than LTGA's best run.
+For Nearest Neighbor NK, most of P3's runs finish much faster than the fastest LTGA runs. However, some
+of P3's outliers take approximately as long as LTGA's tuned performance.
+This supports the hypothesis that P3 is able to scale its diversity not just
 to the problem class, but to the problem instance or even problem initialization, something
@@
 modeling overhead, the expense of this overhead is approximately linear with respect to genome size. There
-is some evidence that even when compared on wall clock time, P3 performs at least as well as
+is some evidence that even when compared on wall clock time, P3 performs on par with
 the best comparison techniques. All of these achievements are made without any problem specific parameter
@@
 fitness. These limitations can be relaxed with future work to make P3 more widely applicable.
-While asymptotically linear, P3's modeling techniques and local search methods are likely going
+While asymptotically linear in problem size, P3's modeling techniques and local search methods are likely going
 to be prohibitively expensive for genome sizes in the hundreds of thousands or millions of genes,
@@
 global optima on black box problems without any problem specific tuning.
+
+\section{Acknowledgements}
+This material is based in part upon work supported by the National Science Foundation
+under Cooperative Agreement No. DBI-0939454. Any opinions, findings, and conclusions
+or recommendations expressed in this material are those of the author(s) and do not
+necessarily reflect the views of the National Science Foundation.
 

### 2015-02-18 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/reviews.tex
Most reviewer changes made.

@@
 
-This was indeed an error on our part.
-Unfortunately the source code was not made available to us, despite a request, and the code was not yet released to the public until after the paper was submitted.
-The fix, a two line modification, does not affect any of the evaluation counts but does indeed modify our timed comparisons (wall clock time experiments).
-Subsequently, after repairing the entropy calculation, we were able to reach timing parity with Dr.
-Bosman's implementation.
-We have rerun all of our timing results and <TODO updated our findings accordingly>.
-In general we find that for single instance problems, optimally tuned LTGA (where we of course do not count the evaluations or timing necessary to find the optimal population size) finds the global optimum using less seconds than P3.
-However, on randomly generated problem classes P3's reduction in evaluations overcomes its increased CPU usage per evaluation.
+This was indeed an error on our part. Unfortunately the source code was not made available to us, despite a request,
+and the code was not yet released to the public until after the paper was submitted. The fix,
+a two line modification, does not affect any of the evaluation counts but does indeed modify our
+timed comparisons (wall clock time experiments). Subsequently, after moving the entropy calculation,
+we were able to reach timing parity with Dr. Bosman's implementation. We have rerun all of our timing results
+and updated our findings accordingly. In general we find that for single instance problems,
+optimally tuned LTGA %(where we of course do not count the evaluations or timing necessary to find the optimal population size)
+finds the global optimum using less seconds than P3. However, on randomly generated problem classes,
+P3's reduction in evaluations overcomes its increased CPU usage per evaluation.
 We have updated our figures and discussion to reflect these changes.
@@
 
-In the original submission Section 9.1 explicitly discusses the variance in outcomes, and Figure 10 shows the upper and lower quartiles (just as boxplots do) for all 7 problems.
-We have, however, added a boxplot in this section to reflect this idea more clearly
+In the original submission Section 9.1 explicitly discussed the variance in outcomes,
+and Figure 10 showed the upper and lower quartiles (just as boxplots do) for all 7 problems.
+We have, however, replaced Figure 10 (Now Figure 11) with a boxplot to show outliers.
 
@@
 
-While the original work did not explictly report the aggregate value, you can estimate it in the original submission using the area under each curve in Figure 13a.
-We have added text to the Figure caption to reflect this
+While the original work did not explicitly report the aggregate value,
+you can estimate it in the original submission using the area under each curve in Figure 13a.
+We have added a new figure (Figure 10, parts A and B) to more clearly show total number of solutions stored.
 
@@
 Nonetheless, we have now added some additional references and discussion expanding those sections
+%TODO Add references and discussion
+
+\item
+\begin{quote}
+Only strictly better solutions are excepted: this is a big problem for LTGA with (at least) MAXSAT
+\end{quote}
+
+We agree that it is important for neutral drift to occur during mixing. Our original submission included that feature,
+as stated in Figure 3 line 6 and near the end of Section 2.5. We believe the source of this confusion is that P3
+requires strict fitness improvement when adding solutions to the pyramid, as stated in Figure 4 line 9 and near the beginning
+of Section 3.
 
@@
 
-This has been updated based on the new timing results
+The abstract no longer mentions wall clock time.
 
@@
 
-Our argument is not that smallest first is the \textit{best}, only that it has desirable diversity preservation features which are problem independent.
+Our argument is not that smallest first is the \textit{best}, only that it has desirable diversity preservation features which are problem independent. We have attempted to clarify this point.
 
@@
 We have added the following text for clarity. 
-For k=7 and s=2, all traps with 0, 1, 3, 5, and 7 bits set are local optima.
+With $k=7$ and $s=2$, as used in our experiments, all traps with 0, 1, 3, 5, and 7 bits set are local optima.
 This means that half of all ways to set the trap are 1 bit local optima.
-This holds true as trap size increases, meaning the number of local optima grows at $2^{k-1}+1$
+More generally, the number of local optima grows at $\Theta(2^{k-1})$.
 
@@
 It is a busy figure. We have added more about the figure and its conclusions to the text.
+%TODO add explanation
 
@@
 As to the super-linear growth, we draw that conclusion from Figure 8b and its discussion.
+We reworded this statement slightly to be more precise.
 
@@
 The other 4 algorithms combined took less than 100 days to complete.
-After fixing our LTGA inefficiency and rerunning on uniform, modern hardware, the non hBOA algorithms took in total 31 days of computing to complete 124 trillion evaluations.
+After fixing our LTGA inefficiency and rerunning on uniform, modern hardware, the non hBOA algorithms took in total 31 days of computing to complete 124 billion evaluations.
 
-We have updated the text to distinguish the different algorithm classes.
+In rerunning our experiments we chose to ensure all experiments were performed in identical hardware. Unfortunately this means we are unable
+to complete the largest problem sizes of hBOA and Parameter-less hBOA. This section has been updated accordingly.
 
@@
 
-This is covered in the original text, Section 2.4 and 2.5.
-However, to review, the run time bottleneck for LTGA is extracting pairwise frequency information which requires $O(mu*N^2)$ time per generation.
-This happens approximately once every $mu*N$ evaluations, meaning the time per evaluation is $(mu*N^2) / (N^2) = O(N)$.
-In P3 both entropy extraction and model rebuilding is $O(N^2)$ and must be done every $O(N)$ evaluations, meaning the time per evaluation is $N^2 / N = O(N)$.
-Therefore, excluding caveats mentioned in the paper, the two techniques should be within a constant factor runtime of each other.
+This is covered in the original text, Section 2.4 and 2.5. However, to review,
+the run time bottleneck for LTGA is extracting pairwise frequency information which
+requires $O(\mu N^2)$ time per generation. This happens approximately once every $\mu N$ evaluations,
+meaning the time per evaluation is $(\mu N^2) / (\mu N) = O(N)$.
+In P3, both entropy extraction and model rebuilding are $O(N^2)$ and must be done every $N$ evaluations,
+meaning the time per evaluation is $N^2 / N = O(N)$. Therefore, excluding caveats mentioned in the paper,
+the two techniques should be within a constant factor runtime of each other.
 This relationship in general holds true for our updated experiments.
@@
 
-Fixed based on the algorithmic fix. Now ``comparable'' instead of ``better''
+Fixed based on the algorithmic fix. Now ``on par'' instead of ``better.''
 
@@
 
-Removed
+Replaced.
 
@@
 
-Added explanation for why medians are the correct choice here. Added statistical tests to data from Figure 6.
+Median is the non-parametric alternative to mean, which requires data to be normally distributed.
+Also, medians allow us to incorporate failures into the statistic. For example, if an algorithm fails due to
+exceeding the evaluation limit, as long as the median run is successful the median is unaffected. If mean was used
+it becomes very difficult to handle failures of that kind. We have added some short statements about this to the text.
+
+We have also added statistical tests to data from Figure 6 to Section 6.1.
 

### 2015-02-18 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Pre my full review

@@
 domain experts may not understand how the parameters will interact with the problem or with each other.
+To further complicate matters, there is mounting evidence that parameter values should change
+during search~\citep{goldman:2011:dynamic-parameters,laporte:2014:adaptivepop}.
 
@@
   \end{centering}
-  \caption{Computation costs incurred by model rebuilding (Figure~\ref{fig-rebuilds}) and
+  \caption{Estimated computation costs incurred by model rebuilding (Figure~\ref{fig-rebuilds}) and
            repeated donations (Figure~\ref{fig-donations}) per evaluation as problem size increases.}
@@
 In order to calculate this value we recorded how many times search rebuilt the model during each run.
-Under the minimal assumption that an evaluation costs $\Omega(N)$ time and that model rebuilds cost $O(N^2)$,
-Figure~\ref{fig-rebuilds} shows the ratio cost of model rebuilding per evaluation. If the cost of model building
+Figure~\ref{fig-rebuilds} shows the estimated ratio of model rebuilding cost ($N^2$ per rebuild) over evaluation cost ($N$ per evaluation).
+If the cost of model building
 scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically constant.
@@
 
-Figure~\ref{fig-small-pop} examines the effect of population size on LTGA's intermediate fitness by examining the effect
-of reducing LTGA's population size to one tenth of the tuned value. The two problems shown are representative of the behavior of using
+Figure~\ref{fig-small-pop} examines the effect of population size on LTGA's intermediate fitness by
+reducing LTGA's population size to one tenth of the tuned value. The two problems shown are representative of the behavior of using
 a smaller population size on the other five problems. Reducing the population size caused LTGA to improve earlier but plateau

### 2015-02-18 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/reviews.tex
Pre my full review

@@
 
-We agree, though our overall point is more general in that we are trying to get the community to rethink \textbf{all} parameter settings.
-In the original submission Section 9, Section 10.2,  Figure 11, and Figure 13a explicitly discuss the benefits we believe P3 achieves due to side stepping the fixed size generational scheme.
+We agree, though our overall point is more general in that we are trying to get the community to rethink
+\textbf{all} parameter settings. In the original submission Section 9, Section 10.2,
+Figure 11, and Figure 13a explicitly discuss the benefits we believe P3 achieves due
+to side stepping the fixed size generational scheme.
 We also had previously mentioned it in the third paragraph of the conclusion.
-Nonetheless, we have now added some additional references and discussion expanding those sections
-%TODO Add references and discussion
+Nonetheless, we have now added some additional references and discussion to the introduction to indicate its importance.
 
@@
 
-It is a busy figure. We have added more about the figure and its conclusions to the text.
-%TODO add explanation
+It is a busy figure. We have worked to clarify the text describing the figure.
 
@@
 However, adding initial local search as a tunable parameter to hboa and LTGA when partial evaluations are not available is not part
-of existing work for either method. Doing so would represent novel research outside the scope of our topic, and provide those techniques
+of existing work for either method. Doing so would also provide those techniques
 with even more problem specific configuration. We welcome future work by other researchers exploring the possible effects of combining

### 2015-02-19 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Finished my pass of the paper

@@
 were included in each solution's genome and themselves underwent evolution. This allowed search
-to optimize some of its own parameters resulting in a reduced the need for expert tuning.
+to optimize some of its own parameters, resulting in a reduced need for expert tuning.
 \cite{harik:1999:parameterlessga} were able to design an entirely parameter-less GA by
@@
 
-Finally, only hBOA and LTGA require any parameters, with each of these only requiring a population
+Only hBOA and LTGA require any parameters, with each of these only requiring a population
 size. This makes knowing the optimal behavior of these algorithms much more tractable. All of the
@@
 this algorithm requires a restart mechanism to solve multimodal problems. We have
-chosen here to naively restart search from a random solution whenever a local optima is found. This
+chosen here to na\"{i}vely restart search from a random solution whenever a local optima is found. This
 ensures that on all landscapes there is always a non-zero probability of search finding the global optimum.
@@
 \Call{Cluster-Usage}{} rely on parameter values. LTGA does not provide any method for controlling or setting
-the population size, and uses a fixed user specified size for each generation.
+the population size, relying instead on a fixed user specified size.
 
@@
 with random solutions. However, this method resulted in increased parameterization as not only was
-a population size required, but also parameters for how frequently generations advanced between levels and
-how many total levels to have were added. \cite{harik:1999:parameterlessga} used multiple independent
+a population size required, but also new parameters for how frequently generations advanced between levels and
+how many total levels to have. \cite{harik:1999:parameterlessga} used multiple independent
 populations of different sizes as a method for removing the population size parameter, but
@@
 if epistasis is set such that each gene depends on the $K$ directly following it
-in the genome the solution can be found in polynomial time~\citep{wright:2000:solvingnk}.
+in the genome, the solution can be found in polynomial time~\citep{wright:2000:solvingnk}.
 These Nearest Neighbor NK landscapes are therefore ideal for search algorithm testing.
@@
 \label{sec-tuning}
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-hboa}
-      \end{centering}
-      \caption{Tuned Population Sizes for hBOA}
-      \label{fig-pop-hboa}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-ltga}
-      \end{centering}
-      \caption{Tuned Population Sizes for LTGA}
-      \label{fig-pop-ltga}
-    \end{subfigure}
-  \end{centering}
-  \caption{Optimal population sizes found using bisection on each size
-           for each problem.}
-  \label{fig-pop-sizes}
-\end{figure}
-
 While four of the six algorithms in our experiments do not require any user specified parameters,
@@
 
+\begin{figure}[t]
+  \begin{centering}
+    \begin{subfigure}{.5\textwidth}
+      \begin{centering}
+        \includegraphicsfit{pop-hboa}
+      \end{centering}
+      \caption{Tuned Population Sizes for hBOA}
+      \label{fig-pop-hboa}
+    \end{subfigure}%
+    \begin{subfigure}{.5\textwidth}
+      \begin{centering}
+        \includegraphicsfit{pop-ltga}
+      \end{centering}
+      \caption{Tuned Population Sizes for LTGA}
+      \label{fig-pop-ltga}
+    \end{subfigure}
+  \end{centering}
+  \caption{Optimal population sizes found using bisection on each size
+           for each problem.}
+  \label{fig-pop-sizes}
+\end{figure}
+
 Figure~\ref{fig-pop-sizes} shows the results from performing bisection. In general hBOA
@@
 on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
+The likelihood that P3 would achieve these pairwise results assuming its median result is actually worse is $p < 10^{-15}$
+according to the binomial test.
 Pairwise comparison of LTGA and P3 on the largest problem size using the Mann-Whitney~U test results in
@@
 The Random Restart Hill Climber and $1+(\lambda, \lambda)$ are both relatively effective on small problem
-sizes. This is especially true for the three randomly generated problem problem classes. These problems
+sizes. This is especially true for the three randomly generated problem classes. These problems
 may contain relatively few local optima or just be exceptionally difficult for the model based algorithms.
@@
 was the most likely to outperform P3, doing so on 50 sample points. LTGA, hBOA, and Parameter-less hBOA
-were the next best, outperforming P3 on 27, 20, and 18 sample points. Random Restart Hill Climbing almost
+were the next best, outperforming P3 on 27, 20, and 18 sample points, respectively. Random Restart Hill Climbing almost
 never outperformed P3, doing so only 9 times.
+The likelihood that P3 would achieve these pairwise results assuming its median result is actually worse is $p < 10^{-9}$
+according to the binomial test.
 
@@
 is accurate little improvement is made. Once it is accurate, fitness improves dramatically.
+
+At 58\% of the recording intervals hBOA has the worst fitness of any solver. Most of the exceptions
+occur when hBOA is still evaluating its initial population, allowing this random search to temporarily surpass the
+local search methods. After $N$ evaluations, however, hBOA and LTGA both fall behind until their models begin
+to improve.
+Parameter-less hBOA reaches intermediate fitnesses faster than hBOA, doing so on 62\% of intervals, as its models begin
+to optimize earlier than hBOA.  However, this trend is reversed after a sufficient number of evaluations, most clearly
+on Deceptive Step Trap and Ising Spin Glasses, as hBOA's tuned population overtakes Parameter-less hBOA's parallel populations.
 
@@
 
-At 58\% of the recording intervals hBOA has the worst fitness of any solver. Most of the exceptions
-occur when hBOA is still evaluating its initial population, allowing this random search to temporarily surpass the
-local search methods. After $N$ evaluations, however, hBOA and LTGA both fall behind until their models begin
-to improve.
-Parameter-less hBOA reaches intermediate fitnesses faster than hBOA, doing so on 62\% of intervals, as its models begin
-to optimize earlier than hBOA.  However, this trend is reversed after a sufficient number of evaluations, most clearly
-on Deceptive Step Trap and Ising Spin Glasses, as hBOA's tuned population overtakes Parameter-less hBOA's parallel populations.
-
 \begin{figure}[t]
@@
 least as high as LTGA at 80\% of intervals.
+The likelihood that P3 would achieve these pairwise results assuming its median result is actually worse is $p < 10^{-15}$
+according to the binomial test.
 
@@
 After local search, all traps in all solutions have a total
-number of 1 bits equal to \{0, 1, 3, 5, 7\} as these correspond to the local
+number of 1 bits equal to either 0, 1, 3, 5, or 7 as these correspond to the local
 optima when using $k=7$ and $s=2$. Crossover is very likely to overcome the
@@
 than optimally tuned state-of-the-art competitors. On single instance problems P3's improvement was by a
-constant factor, while for the three problem classes P3's improvement increased with problem size. This
+constant factor, while for the three randomly generated problem classes P3's improvement increased with problem size. This
 quality extends to intermediate points during evolution, with P3 generally reaching at least as high

### 2015-02-19 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/reviews.tex
Finished my pass of the paper

@@
 \begin{itemize}
-\item Errors in the paper that need correcting
-\item Minor fixes/modifications that we can implement.
+\item Errors in the paper that have been corrected.
+\item Minor fixes/modifications that we have addressed.
 \item Items of disagreement
@@
 This was indeed an error on our part. Unfortunately the source code was not made available to us, despite a request,
-and the code was not yet released to the public until after the paper was submitted. The fix,
+and the code was not released to the public until after the paper was submitted. The fix,
 a two line modification, does not affect any of the evaluation counts but does indeed modify our
-timed comparisons (wall clock time experiments). Subsequently, after moving the entropy calculation,
+wall clock time experiments. Subsequently, after moving the entropy calculation,
 we were able to reach timing parity with Dr. Bosman's implementation. We have rerun all of our timing results
@@
 We also had previously mentioned it in the third paragraph of the conclusion.
-Nonetheless, we have now added some additional references and discussion to the introduction to indicate its importance.
+Nonetheless, we have now added some additional references and discussion to the introduction to elevate its importance.
 
@@
 We agree that it is important for neutral drift to occur during mixing. Our original submission included that feature,
-as stated in Figure 3 line 6 and near the end of Section 2.5. We believe the source of this confusion is that P3
+as stated in Figure 3 line 6 and near the end of Section 2.5. We believe the source of the reviewer's confusion is that P3
 requires strict fitness improvement when adding solutions to the pyramid, as stated in Figure 4 line 9 and near the beginning
@@
 \begin{quote}
-The abstract states that P3 requires "less wall clock seconds" and "is more effective than existing state-of-the-art techniques".
+The abstract states that P3 requires ``less wall clock seconds'' and ``is more effective than existing state-of-the-art techniques''.
 \end{quote}
@@
 \begin{quote}
-"Doerr et al. (2013) presented the first genetic algorithm to provably show the advantage of performing crossover on simple landscapes."
+``Doerr et al. (2013) presented the first genetic algorithm to provably show the advantage of performing crossover on simple landscapes.''
 I understand some people have a very restricted view on provably showing something but this statement is arguably wrong.
@@
 
-We did not intend to be offensive and have cleaned up that text.
+We have made the claim more precise to better reflect their actual accomplishment.
+
+\item
+\begin{quote}
+typo:  ``in a reduced the need for''
+\end{quote}
+
+Fixed
 
@@
 
-Replaced "updating" with "rebuilding"
+Replaced ``updating'' with ``rebuilding''
 
@@
 \begin{quote}
-"Furthermore, unlike LTGA all solutions in the population are used to generate the model, not just the winners of a binary tournament".
+``Furthermore, unlike LTGA all solutions in the population are used to generate the model, not just the winners of a binary tournament''.
 This is not a major design choice, this only influences the selection pressure.
@@
 
-Though we have stated previously that adding to the pyramid triggers modelbuilding, we have added text to the figure to restate it.
+Though we have stated previously that adding to the pyramid triggers model building, we have added text to the figure to restate it.
 
@@
 \begin{quote}
- "This function modifies the results of Equation 4 to include plateaus of size s, introducing an exponential number of local optima in each trap."
+ ``This function modifies the results of Equation 4 to include plateaus of size s, introducing an exponential number of local optima in each trap.''
 Exponential number ? Clarify with a small example of Deceptive Step Trap.
@@
 We have added the following text for clarity. 
-With $k=7$ and $s=2$, as used in our experiments, all traps with 0, 1, 3, 5, and 7 bits set are local optima.
+``With $k=7$ and $s=2$, as used in our experiments, all traps with 0, 1, 3, 5, and 7 bits set are local optima.
 This means that half of all ways to set the trap are 1 bit local optima.
-More generally, the number of local optima grows at $\Theta(2^{k-1})$.
+More generally, the number of local optima grows at $\Theta(2^{k-1})$.''
 
@@
 \begin{quote}
-"To ensure each problem is challenging we chose a clause to variable ratio of 4.27"
+``To ensure each problem is challenging we chose a clause to variable ratio of 4.27''
 Reference to support this number.
@@
 \begin{quote}
- "An important note is that each donation may range in size from a single bit up to N-1, with smaller donations far more likely to result in repeated attempts.
-As such this may cause some super-linear growth in P3, but its unlikely to be very high."
+``An important note is that each donation may range in size from a single bit up to N-1, with smaller donations far more likely to result in repeated attempts.
+As such this may cause some super-linear growth in P3, but its unlikely to be very high.''
 Speculative ?
@@
 \begin{quote}
-" In total these runs took over 6 computing years to complete."
+``In total these runs took over 6 computing years to complete.''
 \end{quote}
@@
 This is covered in the original text, Section 2.4 and 2.5. However, to review,
-the run time bottleneck for LTGA is extracting pairwise frequency information which
+the runtime bottleneck for LTGA is extracting pairwise frequency information which
 requires $O(\mu N^2)$ time per generation. This happens approximately once every $\mu N$ evaluations,
@@
 \begin{quote}
- "There is some evidence that even when compared on wall clock time, P3 performs at least as well as the best comparison techniques"
+``There is some evidence that even when compared on wall clock time, P3 performs at least as well as the best comparison techniques''
 \end{quote}
 
-Fixed based on the algorithmic fix. Now ``on par'' instead of ``better.''
+Fixed based on the corrected code. Now ``on par'' instead of ``better.''
 
@@
 \begin{quote}
-"While asymptotically linear"   Linear in what ?
+``While asymptotically linear''   Linear in what ?
 \end{quote}
@@
 
-We have also added statistical tests to data from Figure 6 to Section 6.1.
+We have also added statistical tests of data from Figure 6, Figure 7, and Figure 12 to Section 6.1, Section 7.1, and Section 9.2.
 
@@
 \begin{quote}
-Last sentence of section 1, penultimate clause doesn't have  verb in it. I suggest you rephrase as " ...; analyse the algorithm complexity; 
+Last sentence of section 1, penultimate clause doesn't have  verb in it. I suggest you rephrase as `` ...; analyse the algorithm complexity;''
 \end{quote}
@@
 \begin{quote}
-Section 8.2.1 uses the phrase "significantly worse".  This suggests that you have some specific confidence in the results according to some hypothesis testing method.  Please clarify or rephrase.
+Section 8.2.1 uses the phrase ``significantly worse''.  This suggests that you have some specific confidence in the results according to some hypothesis testing method.  Please clarify or rephrase.
 \end{quote}
@@
 with even more problem specific configuration. We welcome future work by other researchers exploring the possible effects of combining
-local search with their techniques.
+local search with their techniques in this fashion.
 

### 2015-02-23 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Minor fixes from Bill's pass

@@
 This solution is brought to a local optimum using the hill climbing algorithm in Figure~\ref{fig-hc}. If that
-local optimum has not yet been added to the pyramid, the solution is added to the lowest
+local optimum has not yet been added to any level of the pyramid, the solution is added to the lowest
 population $P_0$.
@@
 Each of the pieces of the P3 algorithm were selected not just for their stand alone efficacy,
-but for the ways in which they interact. By using the hill climber to ensure each solution
-is at a local optima, the underlying pairwise relationships in the problem are exposed. As
+but for the ways in which they interact. By using the hill climber to optimize randomly
+generated solutions, the underlying pairwise relationships in the problem are exposed. As
 a result, detecting clusters for use by crossover is much more effective. The crossover operator

### 2015-03-03 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
LTGA and hBOA with local search.

@@
     \If{solution $\notin hashset$}
-      \State Add solution to $P_0$ and rebuild model
+      \State Add solution to $P_0$ and rebuild $P_0$'s model
       \State Add solution to $hashset$
@@
         \If{solution $\notin hashset$}
-          \State Add solution to $P_{i+1}$ and rebuild model
+          \State Add solution to $P_{i+1}$ and rebuild $P_{i+1}$'s model
           \State Add solution to $hashset$
@@
 and highly deceptive~\citep{rana:1998:gamaxsat}.
+
+While not currently treated as a parameter, we also performed preliminary tests of
+integrating hill climbing into LTGA and hBOA. To match P3, we applied first improvement
+hill climbing to each algorithm's initial population.
+We then performed bisection on the modified algorithms for the largest problem sizes
+where hBOA without hill climbing was effective. We found that in general both methods
+performed worse when combined with hill climbing, in some cases up to an order of
+magnitude worse. There were three exceptions: both improved on MAX-SAT and
+hBOA improved on Rastrigin. In all cases the inclusion of hill climbing did
+not result in either algorithm outperforming P3 in terms of evaluations
+required to reach the global optimum. As such, all further experiments
+use the unmodified published versions.
+
 \section{Finding the Global Optimum}

### 2015-03-03 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/reviews2.tex
Adding the second response to the editor.

\documentclass[]{article}
\usepackage{hyperref}

\begin{document}

\begin{quote}
So, I would leave it with the authors. If they are willing to include results on hBOA and/or LTGA with local search, that’s fantastic.
If they conclude that they would rather not spend the extra effort, then I would ask them to at least comment on the issue in the paper.
\end{quote}

\begin{quote}
You could formulate it very cautiously, clearly saying that these are just preliminary observations and that on some problems local search does help (MAXSAT).
\end{quote}

In Section 5 Comparison Algorithm Parameter Tuning, paragraph at the start of page 13: We now explain how
we performed preliminary tests of using local search with hBOA and LTGA, but found it to generally not helpful. We
highlight the three cases in which local search improved results.

The only other modification to the paper was in Figure 4, top of page 9: Line 5 and 11 now indicate that a specific
level's model is being rebuilt, clarifying that there is a different model for each level.
\end{document}

### 2015-03-25 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
Some minor fixes + quartile bars on graphs

@@
     \State $closed \leftarrow \emptyset$
-    \State $found \leftarrow []$
+    \State $found \leftarrow [~]$
     \ForAll{$v \in V$}
@@
 
-\section{Tournament Uniform Crossover: TUX}
+\section{Tournament Uniform\\ Crossover: TUX}
 \label{sec-tux}
@@
 
-\section{Parameter-less Population Pyramid: P3}
+\section{Parameter-less Population\\ Pyramid: P3}
 \subsection{Origins}
 \begin{figure}
-  \begin{algorithmic}
+  \begin{algorithmic}[1]
   \Procedure{Iterate-P3}{}
-    \State Create random solution
+    \State Create random $solution$
     \State Apply hill climber
-    \If{solution $\notin hashset$}
-      \State Add solution to $P_0$
-      \State Add solution to $hashset$
+    \If{$solution \notin hashset$}
+      \State Add $solution$ to $P_0$
+      \State Add $solution$ to $hashset$
     \EndIf
@@
     \ForAll{$P_i \in pyramid$}
-      \State Mix solution with $P_i$
-      \If{solution's fitness has improved}
-        \If{solution $\notin hashset$}
-          \State Add solution to $P_{i+1}$
-          \State Add solution to $hashset$
+      \State Mix $solution$ with $P_i$
+      \If{$solution$'s fitness has improved}
+        \If{$solution \notin hashset$}
+          \State Add $solution$ to $P_{i+1}$
+          \State Add $solution$ to $hashset$
         \EndIf
@@
 Runs were performed on 2.5GHz Intel Xeon E5-2670v2 processors using the C++11 code available
-from our website.\footnote{\url{https://github.com/name_of_repository/anonymous_name_of_project}}
+from our website.\footnote{\url{https://github.com/brianwgoldman/GrayBoxOptimization}}
 Each time the run achieved a new best fitness we recorded the current amount of processing time used.
@@
 is also the global optimum, verified using dynamic programming. For Ising Spin Glass the ``best'' fitness
-was the global optimum 44 of 50 times.
+was the global optimum 44 out of 50 times.
 

### 2015-04-01 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
Made changes from reviews

@@
 Leveraging these restrictions, the Hamming-Ball Hill Climber
-(HBHC) can efficiently find r-bit local optima. We show how
+(HBHC) can efficiently find high quality local optima. We show how
 1) a simple memetic algorithm in conjunction with HBHC can find \emph{global}
@@
 information about variable relationships, outperforms all of the
-examined algorithms.
+examined algorithms. While HBHC's inclusion into P3 adds a parameter,
+we show experimentally it can be fixed to 1 without adversely effecting search.
 
-We provide experimental evidence on both NKq-Landscapes and Ising Spin Glasses
+We provide experimental evidence on NKq-Landscapes and Ising Spin Glasses
 that Gray-Box P3 is effective at finding the global optima even for problems with
@@
 
-%\newpage
+\newpage
 
@@
 providing local search with more problem specific information than just an evaluation function can
-extraordinarily improve search power and efficiency. These improvements
+greatly improve search power and efficiency. These improvements
 require information easily derived for a broad range of interesting NP-Hard
@@
 
-Here we extend those benefits to global optimization by integrating
+Here we extend those benefits to global optimization on binary problems by integrating
 this local search with additional heuristics~\cite{chen:2011:memetic}.
@@
 process we also extend the Parameter-less Population Pyramid (P3)~\cite{goldman:2014:p3}, the
-current state-of-the-art in black-box optimization. This ``Gray-Box P3''
+current state-of-the-art in black-box optimization~\cite{goldman:2015:fastp3}. This ``Gray-Box P3''
 utilizes both efficient local search and the additional problem information
@@
 of a given solution. While these very simple requirements allow almost any
-optimization task to fit into this domain, it may prevent additional knowledge
-about a problem from being provided to search algorithms. At the other end of
+optimization task to fit into this domain, many problems have additional knowledge
+which could be used by search algorithms. At the other end of
 the spectrum is white-box optimization, in which everything about the problem
@@
 Here we shall consider the gray-box optimization domain with the following
-characteristics: The function which evaluates the quality of a solution is determined by the summation
+characteristics~\cite{chicano:2014:ball}: The function which evaluates the quality of a solution is determined by the summation
 of subfunctions. Each subfunction uses at most $k$ variables in the given solution,
-such that as the total number of solution variables $N$ increases, $k$ remains constant.
+such that $k$ remains constant as the number of problem variables $N$ increases.
 Each subfunction is independently evaluable and the variables each subfunction uses are known.
@@
 MAX-SAT, and by extension all NP-hard problems, can be represented
-as a summation of subfunctions to evaluate each clause~\cite{whitley:2013:greedy}.
+as a summation of subfunctions which evaluate each clause~\cite{whitley:2013:greedy}.
 Previous approximation studies of Vertex Cover~\cite{oliveto:2009:vertexcover},
@@
 In order to obtain efficiency bounds for optimization algorithms, the following
-assertions are also included: The total number of subfunctions grows no more than
+assertions are also included~\cite{chicano:2014:ball}: The total number of subfunctions grows no more than
 linearly with $N$, which holds true for randomized MAX-SAT, mesh spin glasses, and NK Landscapes.
@@
 %TODO Are these necessary?
+\begin{comment}
 We also include the trivial requirements that all problem variables must participate
@@
 violates these requirements can be restated either using fewer variables or as multiple separate problems.
+\end{comment}
 
@@
 described in Section~\ref{sec-gray-box} was \cite{whitley:2013:greedy}. In this study
-a hill climber was developed that found the approximate next-best move in \BigO{1} time.
+a hill climber was developed that found the approximate next-best move in \BigO{1} time given
+\BigO{N} setup.
 This is achieved by first determining the fitness effect of making each move, and then updating
@@
 By again leveraging the properties of this domain, the Hamming-Ball Hill Climber (HBHC)~\cite{chicano:2014:ball} extends this
-method to find $r$-bit local optima. While na\"ively this would require testing \BigO{N^r} moves, in the gray-box domain only
+method to find solutions that cannot be improved by flipping $r$
+or fewer bits called $r$-bit local optima.
+While na\"ively this would require testing \BigO{N^r} moves, in the gray-box domain only
 \BigO{(3ck)^rN} must be checked. This is due to the bounded relationships between variables.
@@
 is tested it is temporarily added to $closed$ to prevent recursive calls
-from creating duplicates.
+from adding it again to $subset'$.
 
@@
   \caption{One iteration of TUX optimization. $T$ is an
-           ordered list of solutions, each of which could be empty,
+           ordered list of solutions, each position of which could be empty,
            awaiting a crossover partner.}
@@
 the Tournament Uniform Crossover (TUX) algorithm, which combines simplistic
-selection with equal probability uniform crossover to generate starting
+selection with equal probability uniform crossover, the most basic unbiased crossover, to generate starting
 solutions likely to be in the global optima's basin of attraction.
@@
 the ``winner'' being the best of all those solutions. The tournament ``bracket'' is constructed
-iteratively, storing a single list of solutions $T$, such that $|T|$ is equal to the height
+iteratively, storing an initially empty list of solutions $T$, such that $|T|$ is equal to the height
 of the tournament. Iterative construction is possible because when a sub-bracket is complete only a single solution
@@
 The Parameter-less Population Pyramid (P3) is a recently introduced
-black-box optimization method which requires no problem specific configuration~\cite{goldman:2014:p3}.
+black-box optimization method which requires no problem specific configuration~\cite{goldman:2014:p3,goldman:2015:fastp3}.
 Unlike previous parameter-less methods it was shown to be more efficient than
@@
 P3 represents a natural method for integrating the HBHC into
-a global optimization algorithm as P3 already utilizes local search. However, there
+a global optimization algorithm as P3 already utilizes local search.
+While HBHC's inclusion adds a parameter, Section~\ref{sec-radius} and
+Section~\ref{sec-over-time} provide evidence that $r$ can be fixed to 1, preserving
+the parameter-less nature of P3.
+Beyond using HBHC, there
 are also a number of ways in which P3 can be made more efficient by leveraging the
@@
 in at least one cluster, but variables which are central in the graph will appear in more
-clusters than those on the periphery. If all else is equal, the more paths of a given
+clusters than those on the periphery. The more paths of a given
 length between two variables, the higher the probability of them being in the same cluster.
@@
 moves are always in the direction of previously found high quality solutions.
-This assumes that the density of high quality solutions is higher between good solutions than
-outside of good solutions.
+This assumes that the density of high quality solutions is higher than average between good solutions.
 
@@
 are NP-Hard, but the $2D\pm J$ subset is polynomially solvable~\cite{saul:1994:spinglass}\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}. In this
-subset the graph is a 2D toroidal grid, with each edge's weight either -1 or 1. In gray-box terms,
-problems of this subset have $k=2$ and $c=4$ regardless of $N$. While Ising Spin Glasses are traditionally
-a minimization problem, here we present them as a maximization problem by negating the normal fitness.
+subset the graph is a 2D toroidal grid, with edge weights of $\pm 1$. In gray-box terms,
+problems of this subset have $k=2$ and $c=4$ regardless of $N$.
+%While Ising Spin Glasses are traditionally
+%a minimization problem, here we present them as a maximization problem by negating the normal fitness.
 
 For each problem class tested, we generated 50 instances.
+Extreme problem sizes were chosen to stress each algorithm.
 Each method was run once on each instance, and limited to 3 hours of computation and 4 GB of memory.
@@
 
-All figures report the median run for either percentage error or seconds to reach the best. A run's percentage error is
+All figures report the median, upper and lower quartiles for either percentage error or seconds to reach the best. A run's percentage error is
 equal to the difference between its fitness and the best, divided by the best. When reporting seconds
@@
   \caption{Comparison of how radius effects solution quality at termination. For NKq-Landscapes $N=6,000$ and $K=4$ and
-  for Ising Spin Glasses $N=6,084$.}
+  for Ising Spin Glasses $N=6,084$. Range of radius values limited by memory constraints.}
   \label{fig-fitness}
@@
 Figure~\ref{fig-fitness} shows the effect on final solution fitness as $r$ increases. As expected
-from~\cite{chicano:2014:ball} the HBHC obtains higher quality as $r$ increases with the magnitude
+from~\cite{chicano:2014:ball} the HBHC obtains higher quality as $r$ increases, with the magnitude
 of the improvement decreasing. TUX has a similar relationship and
@@
   \includegraphicsfit{p3-seconds}
-  \caption{Median time required for Gray-Box P3 to reach the global optimum of Nearest Neighbor NKq instances with $N=6,000$.}
+  \caption{Time required for Gray-Box P3 to reach the global optimum of Nearest Neighbor NKq instances with $N=6,000$.}
   \label{fig-p3-seconds}
@@
 
-To further examine the effect of $r$ on Gray-Box-P3, Figure~\ref{fig-p3-seconds} shows the median number of
+To further examine the effect of $r$ on Gray-Box-P3, Figure~\ref{fig-p3-seconds} shows the number of
 seconds required to reach the global optimum. Setting $r=1$ was the most efficient configuration for all $K > 1$,
@@
 \subsection{Fitness Over Time}
-
+\label{sec-over-time}
 \begin{figure*}
@@
 search for each algorithm. HBHC and TUX have significant early delays caused by their high $r$ values. Larger
-$r$'s require a large amount of initial partial evaluation before performing hill climbing. Once that process
-is complete HBHC effectively stalls, with TUX improving at a faster rate. Both are eclipsed by Gray-Box P3,
-which quickly descends to the global optimum, outperforming HBHC and TUX at every point.
+$r$'s require a large amount of initial partial evaluation before performing hill climbing. After one full iteration
+HBHC effectively stalls, with TUX continuing to improve. Both are eclipsed by Gray-Box P3,
+which quickly descends to the global optimum, outperforming HBHC and TUX at every time point.
 
@@
   \caption{Comparison of how each algorithm's time required to reach the best fitness found scales with problem size
-  on Nearest Neighbor NKq with $K=4$.}
+  on Nearest Neighbor NKq with $K=4$. With $N=1000$ Gray-Box P3 is 375x faster than Black-Box P3.}
   \label{fig-length-nn}
@@
   \caption{Comparison of how each algorithm's time required to find the best fitness found scales with problem size
-  on Ising Spin Glass.}
+  on Ising Spin Glass. With $N=2025$ Gray-Box P3 is 4.6x faster than Black-Box P3.}
   \label{fig-length-is}
@@
 previous work has shown Black-Box P3 is faster to reach the global optimum than other leading black-box
-methods. Applying regression, we estimate that Black-Box P3's time to global optimum
+methods~\cite{goldman:2015:fastp3}. Applying regression, we estimate that Black-Box P3's time to global optimum
 on Nearest Neighbor NKq is \BigO{N^{2.75}} while Gray-Box P3's is \BigO{N^{1.98}}.
@@
 As Unrestricted NKq does not have a known global optimum, and the different algorithms
-rarely found the same best fitness, Figure~\ref{fig-length-un} compares the median error
+rarely found the same best fitness, Figure~\ref{fig-length-un} compares the error
 for each technique at termination. Gray-Box P3 in general finds the best fitness,
@@
 A potential explanation is that in both Ising Spin Glass
-and Nearest Neighbor NKq the number of unique variables $r$ steps
+and Nearest Neighbor NKq the number of unique variables $r$
 or fewer steps away from a given variable is significantly lower than the worst case. This explains why on Unrestricted NKq
@@
 
-While the inclusion of HBHC into Gray-Box P3 introduces a parameter value, it requires
+While the inclusion of HBHC into Gray-Box P3 introduces a parameter, it requires
 trivial configuration. In the worst case there may be a handful of $r$ values to test.

### 2015-04-01 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
update to get to length for submission

@@
 issue is that on these landscapes the importance of each non-linear relationship may be detectably unequal.
-As a result the Black-Box linkage learning may be able to better cluster variables which having meaningful
+As a result Black-Box linkage learning may better cluster variables which having meaningful
 impact on fitness while Gray-Box assumes all are equally important. A useful direction for future work

### 2015-04-02 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
Fixed copyright and updated style file

@@
 
-\permission{Permission to make digital or hard copies of all or part of this work
-for personal or classroom use is granted without fee provided that copies are not
-made or distributed for profit or commercial advantage and that copies bear this
-notice and the full citation on the first page. Copyrights for components of this
-work owned by others than ACM must be honored. Abstracting with credit is permitted.
-To copy otherwise, or republish, to post on servers or to redistribute to lists,
-requires prior specific permission and/or a fee. Request permissions from permissions@acm.org.}
-\conferenceinfo{GECCO'15,} {July 11-15, 2015, Madrid, Spain.}
-\CopyrightYear{2015}
-\crdata{TBA}
+\setcopyright{acmcopyright}
+\conferenceinfo{GECCO '15,}{July 11 - 16, 2015, Madrid, Spain}
+\isbn{978-1-4503-3472-3/15/07}\acmPrice{\$15.00}
+
+\doi{http://dx.doi.org/10.1145/2739480.2754775}
 \clubpenalty=10000

### 2015-04-02 Gray Box Optimization using the Parameter-less Population Pyramid/make_open_fonts.sh

@@ -1,0 +1,12 @@
+cat $1 |\
+sed 's+Times-Bold+NimbusSanL-Bold+g' |\
+sed 's+Times-Roman+NimbusSanL-Regu+g' |\
+sed 's+Times+NimbusSanL-Regu+g' |\
+sed 's+Helvetica-BoldOblique+NimbusSanL-BoldItal+g' |\
+sed 's+Helvetica-Oblique+NimbusSanL-ReguItal+g' |\
+sed 's+Helvetica-Bold+NimbusSanL-Bold+g' |\
+sed 's+Helvetica-Bold-iso+NimbusSanL-Bold+g' |\
+sed 's+Helvetica+NimbusSanL-Regu+g' |\
+sed 's+Helvetica-iso+NimbusSanL-Regu+g' |\
+sed 's+Symbol+StandardSymL+g' > $1.tmp
+mv $1.tmp $1

### 2015-04-02 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
Made github link point to "release" section

@@
 Runs were performed on 2.5GHz Intel Xeon E5-2670v2 processors using the C++11 code available
-from our website.\footnote{\url{https://github.com/brianwgoldman/GrayBoxOptimization}}
+from our website.\footnote{\url{https://github.com/brianwgoldman/GrayBoxOptimization/releases}}
 Each time the run achieved a new best fitness we recorded the current amount of processing time used.

### 2015-04-13 Dissertation/chapter-conclusions.tex
Copied Comprehensive to Disseration

--- 2015-02-19 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
@@
-\documentclass[twoside]{article}
-\usepackage{ecj,palatino,epsfig,latexsym,natbib}
-\usepackage{url}
-\usepackage{verbatim}
-\usepackage{subcaption}
-\usepackage[noend]{algpseudocode}
-
-\newcommand{\includegraphicswide}[1]
-{\includegraphics[width=.9\textwidth,height=\textheight,keepaspectratio]{#1}}
-
-\newcommand{\includegraphicsfit}[1]
-{\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}
-
-\usepackage{array}
-\newcolumntype{L}{>{\centering\arraybackslash}m{1.9cm}}
-
-%% do not add any other page- or text-size instruction here
-
-\parskip=0.00in
-
-\begin{document}
-
-% 45-character limit on header
-\ecjHeader{x}{x}{xxx-xxx}{201X}{Parameter-less Population Pyramid}{B. W. Goldman and W. F. Punch}
-\title{\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid}
-
-\author{\name{\bf B. W. Goldman} \hfill \addr{brianwgoldman@acm.org}\\ 
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-\AND
-       \name{\bf W. F. Punch} \hfill \addr{punch@msu.edu}\\
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-}
-
-\maketitle
-
-\begin{abstract}
-%Should be about 200 words
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
-evolutionary optimization without requiring any user specified parameters.
-P3's primary innovation is to replace the generational model with a pyramid of
-multiple populations that are iteratively created and expanded. In combination
-with local search and advanced crossover P3 scales to problem difficulty, exploiting
-previously learned information before adding more diversity.
-
-Across seven problems, each tested using on average 18 problem sizes, P3 outperformed
-all five advanced comparison algorithms. This improvement includes requiring less evaluations
-to find the global optimum and better fitness when using
-the same number of evaluations. Using both algorithm analysis and comparison we find P3's
-effectiveness is due to its ability to properly maintain, add, and exploit diversity.
-
-Unlike the best comparison algorithms, P3 was able to achieve this quality without any
-problem specific tuning. Thus, unlike previous parameter-less methods, P3 does not
-sacrifice quality for applicability. Therefore we conclude that
-P3 is an efficient, general, parameter-less approach to black box
-optimization which is more effective than existing state-of-the-art techniques.
-
-\end{abstract}
-
-\begin{keywords}
-
-Genetic algorithms, 
-linkage learning,
-local search,
-parameter-less.
-
-\end{keywords}
-
-\section{Introduction}
-A primary purpose of evolutionary optimization is to efficiently find good solutions
-to challenging real world problems with minimal prior knowledge about the problem itself.
-This driving goal has created search algorithms which can escape user bias to create
-truly novel results, sometimes publishable or patentable in their own right~\citep{kannappan:2014:humies}.
-While it is not possible for any algorithm to do better than random search across all possible
-problems~\citep{Wolpert:1997:nfl}, effectiveness can be achieved by assuming the search
-landscape has structure and then biasing the algorithm toward exploiting that structure.
-
-In evolutionary optimization, and genetic algorithms~(GAs) in particular, search is often
-biased through parameters. This can be beneficial as it allows practitioners to inject their
-knowledge about the shape of the search landscape into the algorithm.
-However, the quality of solutions found, and the speed at which they are found, is strongly tied to setting these parameters
-correctly~\citep{goldberg:1991:gasize}. As such, either expert knowledge or exceedingly
-expensive parameter tuning~\citep{grefenstette:1986:optimalga} are required to leverage
-this feature to its fullest potential. Furthermore,  parameters such as population size, mutation rate, crossover
-rate, tournament size, etc. usually have no clear relationship to the problem being solved, meaning even
-domain experts may not understand how the parameters will interact with the problem or with each other.
-To further complicate matters, there is mounting evidence that parameter values should change
-during search~\citep{goldman:2011:dynamic-parameters,laporte:2014:adaptivepop}.
-
-There have been periodic efforts to reduce or remove the need for parameter tuning.
-\cite{rechenberg:1973:es} introduced self-adaptive parameters, in which parameter values
-were included in each solution's genome and themselves underwent evolution. This allowed search
-to optimize some of its own parameters, resulting in a reduced need for expert tuning.
-\cite{harik:1999:parameterlessga} were able to design an entirely parameter-less GA by
-leveraging schema theory and parallel populations. Unfortunately these methods were provably less efficient
-than directly setting the parameters to optimal values~\citep{pelikan:1999:worstparameter-less}.
-
-One area that has been very effective at reducing the number of algorithm parameters is
-model based search. \cite{pelikan:2006:hboa}'s Hierarchical Bayesian Optimization
-Algorithm~(hBOA) and \cite{thierens:2010:ltga}'s Linkage Tree Genetic Algorithm~(LTGA)
-both require only a single parameter: population size. \cite{posik:2011:parameterless}
-leveraged model building to create a fully parameter-less algorithm, but it is restricted to
-only order-k, fully decomposable, noiseless problems.
-
-Most recently \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid~(P3).
-This method uses a pyramid structure of populations to combine model based search with local search
-to achieve parameter-less optimization. Initial results suggest that unlike
-previous parameter-less methods, P3 is actually more efficient than current state-of-the-art
-parameterized search algorithms. In this work we shall: extend these results to cover more
-comparison algorithms; compare both efficiency in reaching the global optimum and intermediate
-fitnesses; analyze algorithm complexity; and provide more in depth analysis of P3 itself.
-
-\section{Comparison Optimizers}
-\label{sec-optimizers}
-
-In order to fully understand the effectiveness of P3, we compare it with five advanced algorithms which
-have related features to P3. The Random Restart Hill Climber defined by ~\cite{goldman:2014:p3} was chosen as an efficient form of repeated
-local search. As P3 combines this hill climber with crossover,
-comparing with local search alone shows the advantages of P3's overall approach.
-The $1+(\lambda, \lambda)$ algorithm~\citep{doerr:2013:lambdalambda} is the current best theory supported simple genetic algorithm
-and its method of crossover is in some sense a macro-mutation just as in P3. hBOA and Parameter-less
-hBOA are advanced model building search techniques which are very effective at learning complex
-problem structure, designed to achieve similar goals as P3's linkage learning but using very different
-methods. Finally LTGA represents the current state-of-the-art in black box search and is the origin
-of P3's linkage learning and crossover methods.
-
-Only hBOA and LTGA require any parameters, with each of these only requiring a population
-size. This makes knowing the optimal behavior of these algorithms much more tractable. All of the
-algorithms are also gene order independent, fitness scale invariant, and unbiased. This means,
-for any problem, the order in which problem variables appear in the genome can be changed
-without changing the behavior of the search. The fitness can also be manipulated in any fashion
-as long as the rank ordering of solutions is unchanged. These algorithms are also unaffected by the meaning
-assigned to each bit, such that inverting a predetermined random subset of genes before evaluation
-will not impact search efficiency.
-
-Our implementations of all of these algorithms as well as all of the population size information, raw results,
-and processing scripts
-are available from our website.\footnote{\url{https://github.com/brianwgoldman/FastEfficientP3}}
-\subsection{Random Restart Hill Climber}
-\label{sec-hill-climber}
-Perhaps the simplest black box search heuristic is stochastic local search, or hill climbing.
-This optimization technique focuses on improving a single solution until it reaches a local
-optimum. Here we use the first improvement hill climber defined by \cite{goldman:2014:p3}
-and given in Figure~\ref{fig-hc}. This algorithm works by flipping each bit in a random
-order, keeping modifications when fitness is improved, until single bit flips cannot result in
-further fitness improvements.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Hill-Climber}{}
-    \State $options \leftarrow [0 \dots N-1]$\label{fig-hc-options}
-    \State $tried \leftarrow \emptyset$
-    \While{$|tried| < |options|$}
-      \ForAll{$index \in shuffled(options)$}
-        \If{$index \notin tried$}\label{fig-hc-tried}
-          \State Flip bit $index$ in solution
-          \If{solution's fitness increased}
-            \State $tried \leftarrow \emptyset$
-          \Else
-            \State Revert change
-          \EndIf
-          \State $tried \leftarrow tried \cup \{index\}$
-        \EndIf
-      \EndFor
-    \EndWhile
-  \EndProcedure
-\end{algorithmic}
-  \caption{Hill climbing algorithm used to improve randomly generated solutions until no single
-           bit change results in a fitness improvement.}
-  \label{fig-hc}
-\end{figure}
-
-The hill climber requires an amortized cost of $O(1)$ operations per evaluation. In order to
-terminate, at least one evaluation must be performed for each of the $N$ bits in the solution.
-As such any operation that happens only once per search can be amortized over at least $N$
-evaluations, covering the initialization of $options$ on Line~\ref{fig-hc-options}.
-Line~\ref{fig-hc-tried}, which prevents wasted evaluations, can be called at most twice
-per evaluation: once to add $index$ into $tried$ and once to prevent $index$ from being unnecessarily
-evaluated again. The only way three or more calls could happen is if
-no fitness improvement was made for the entire previous iteration, which contradicts the loop invariant.
-
-Due to its nature, this hill climber cannot escape basins of attraction. Once a solution is reached
-such that none of the single bit neighbors are fitness improvements, search stops. Thus
-this algorithm requires a restart mechanism to solve multimodal problems. We have
-chosen here to na\"{i}vely restart search from a random solution whenever a local optima is found. This
-ensures that on all landscapes there is always a non-zero probability of search finding the global optimum.
-
-\subsection{$1+(\lambda, \lambda)$}
-\cite{doerr:2013:lambdalambda} presented the first genetic algorithm to provably show
-the advantages of performing crossover on One Max. This comparatively simple
-algorithm maintains only a single individual and a self-controlled parameter $\lambda$.
-
-Each iteration, the number of bits to flip is chosen from the binomial distribution $b\sim B(N, \frac{\lambda}{N})$,
-where $N$ is the number of bits in the genome.
-Next, $\lfloor\lambda\rfloor$ offspring are produced by flipping $b$ bits. The
-best mutant then produces $\lfloor\lambda\rfloor$ offspring via uniform crossover with the original parent, such that each gene comes from the
-mutant with probability $\frac{1}{\lambda}$. In the original algorithm the best
-offspring produced by crossover then replaces the original parent if its fitness is no worse.
-The $\lambda$ parameter, which is initialized to 1, is decreased if the offspring replaced
-its parent and increased otherwise.
-
-The original formulation was designed specifically for unimodal landscapes and as such were
-not directly suitable for multimodal problems. \cite{goldman:2014:p3} extended $1+(\lambda, \lambda)$
-to include random restarts. As search stagnates, the $\lambda$ parameter increases in value. Eventually
-this results in $\lambda \ge N$ causing mutation to always flip all bits of the individual.
-As this prevents any future improvement, whenever $\lambda \ge N$ search is restarted from a random solution with $\lambda$
-reset to 1.
-
-A few other efficiency modifications were also made. If there is a tie in crossover offspring fitness,
-whichever has a larger hamming distance from the parent is retained. This encourages drifting across plateaus.
-The ``mod'' control strategy proposed by \cite{doerr:2013:lambdalambda} was not used as it conflicted with
-the random restart strategy.
-If a crossover individual is identical to either of its parents, it is not evaluated.
-If mutation produces an offspring which is better than the best crossover offspring, it is used to compare
-against the original parent.
-
-\subsection{Hierarchical Bayesian Optimization Algorithm}
-
-\cite{pelikan:2006:hboa} used statistical principles in combination with a decision tree structure
-to create the Hierarchical Bayesian Optimization Algorithm (hBOA). This method creates a model of
-epistatic relationships between genes which is then used to stochastically generate new solutions.
-Each generation a binary tournament with replacement is used to select $\mu$ solutions from
-the population. These solutions are then used to build the model, which in turn is used to generate $\mu$ new
-solutions. The new solutions are then integrated into the population using restricted tournament
-replacement.
-
-Conceptually, the model built by hBOA is trying to infer rules of the form ``Given that this
-subset of genes are set to these values, how frequently is gene $x_i$ set to value v?'' This can
-be represented using a directed acyclic decision forest, with each tree in the forest representing one gene
-in the solution. In the decision tree $T_i$, which is used to set the value of gene $x_i$,
-each internal node represents previous decisions on how to set
-some other gene $x_j$, with the children of that node representing how the decision was made. The
-leaves of each tree give the probability that $x_i$ should be set to each possible value.
-
-The forest is constructed iteratively, with each tree initially containing a single leaf
-and with each leaf storing a pointer for each selected solution. Each iteration the algorithm considers
-all possible ways of splitting an existing leaf using another gene $x_j$, such that solutions in the
-leaf are moved to the newly created leaves based on their value for $x_j$. The general goal is to
-separate the solutions such that all solutions with $x_i = 0$ move to one leaf while solutions with
-$x_j = 1$ move to the other.
-
-This goal is formalized using model scoring from Bayesian statistics. In its raw form this
-almost always creates near infinitesimal results, calculating fractions which include factorials of $\mu$
-and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{sec-appendix}, we
-derived a simplified form shown in Equation~\ref{eq-hboa-final}.
-Here $l$ is a leaf in tree $i$, with $l'$ and $l''$ the results of splitting $l$. $m_i(l)$ is the number
-of solutions which reach $l$ and $m_i(x_i, l)$ is the number of solutions which reach $l$ with the given value
-for $x_i$.
-If no proposed split satisfies the inequality, iteration stops.
-If multiple splits do, whichever maximizes the right side is chosen.
-
-\begin{equation}
-  2^{0.5 log_2\mu} < \frac{(m_i(l) + 1)!}{m_i(0, l)!m_i(1,l)!} \cdot
-  \frac{m_i(0, l')!m_i(1,l')!m_i(0, l'')!m_i(1,l'')!}{(m_i(l') + 1)!(m_i(l'') + 1)!}
-  \label{eq-hboa-final}
-\end{equation}
-
-Initially there are $\Theta(N^2)$ possible ways to split existing leaves, as each of the $N$ single node
-trees can be split by any of the other $N-1$ genes. Each iteration a new edge is added to the decision
-forest, meaning some of the previously tested splits cannot be used. For instance, if $T_i$, which is used
-to decide the value of $x_i$, is split using the value of $x_j$, $T_j$ can no longer be split using $x_i$.
-As a split creates two new leaves, $O(N)$ new potential splits must also be tested. Equation~\ref{eq-hboa-final}
-parses all solutions which reach a leaf to count gene frequencies, requiring $\mu$ time.
-The number of total leaves created depends heavily on the problem and $\mu$.
-However, assuming no splits are accepted or that the cost of testing all future splits is less than
-the initial $\Theta(N^2)$, constructing the model requires $\Omega(\mu N^2)$ time. Each
-model is used to generate $\mu$ solutions, leading to a cost per evaluation of $\Omega(N^2)$.
-
-To generate a solution, the value of each gene $x_i$ is set using its corresponding decision tree $T_i$. Because
-the forest is directed acyclic, there must be an ordering of $T_i$ such that before $T_i$ is executed all
-$x_j$ it uses to make decisions have already been set. As such, previous decisions made by other trees
-are used to follow each $T_i$ until a leaf is reached. The value of $x_i$ is then set based on the
-probability that other solutions reached that leaf with each value of $x_i$.
-
-To perform replacement, hBOA uses restricted tournament replacement. After each new solution is generated
-and evaluated, a set of $w$ solutions are chosen at random from the population, where $w=\min\{N, \frac{\mu}{20}\}$.
-From this set the solution which is most genetically similar to the offspring is chosen. If the offspring
-is at least as fit as the chosen solution, it replaces the chosen solution in the population. Otherwise the
-offspring is discarded. This method is designed to preserve genetic diversity as only genetically similar
-solutions must compete on fitness.
-
-hBOA is designed to work with large population sizes, resulting in a large number of
-evaluations per generation. As hBOA utilizes explicit diversity maintenance, standard methods for determining
-convergence are not considered very accurate. Therefore the authors suggest that an hBOA run should be
-terminated after performing generations equal to $N$.
-
-Like other model based techniques, hBOA has very few parameters. There is no mutation or crossover,
-and modeling does not rely on any explicit parameters. Solution selection, generation, and replacement
-are all derived from the population size, which must be set by the user.
-
-\subsection{Parameter-less hBOA}
-Using the methods first introduced by \cite{harik:1999:parameterlessga} for the Parameter-less GA,
-\cite{pelikan:2004:parameterlesshboa} created Parameter-less hBOA which automatically scales its
-population size to fit the problem. This is done by maintaining a list of concurrent populations
-using exponentially scaled population sizes.
-
-A run of Parameter-less hBOA starts with a single population of size $\mu_0$, conventionally set
-to $\mu_0=10$. After two generations are performed, a new population of size $\mu_1 = 2\mu_0$ is created
-and performs a generation. Evolution then continues with the $\mu_0$ population performing two generations
-for each one performed by $\mu_1$. Each time population $\mu_i$ performs its second generation a new population
-$\mu_{i+1}=2\mu_i$ is created, which performs generations half as often as $\mu_i$. In this way an infinite number of
-parallel population can be simulated, with each population receiving the same number of total evaluations.
-
-In all other aspects each population is identical to an hBOA population using a fixed $\mu$. No search information
-is shared among populations, and each search is independently terminated. As such Parameter-less hBOA cannot
-perform better than hBOA using the optimal population size for a given instance, as it must also spend evaluations
-on populations of different sizes. This inefficiency is bounded by a log multiple of the total number of
-evaluations~\citep{pelikan:1999:worstparameter-less}.
-
-\subsection{Linkage Tree Genetic Algorithm}
-\cite{thierens:2010:ltga} introduced the Linkage Tree Genetic Algorithm (LTGA) which automatically
-detects and exploits problem epistasis by examining pairwise gene entropy. Due to its enhanced
-ability to preserve high fitness gene subsets, LTGA was able to outperform state-of-the-art
-GAs across many benchmarks. Since its introduction, many variants of LTGA have been
-proposed~\citep{thierens:2011:gomea,goldman:2012:ltga} so for clarity we have chosen the version
-presented by \cite{thierens:2013:ltgahiff} as our model.
-
-LTGA's effectiveness comes from its method of performing crossover. Instead of blindly
-mixing genes between parents, LTGA attempts to preserve important interrelationships
-between genes. Before performing any crossovers in a generation, LTGA first builds
-a set of hierarchical gene clusters which are then used to dictate how genes are mixed
-during crossover.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Cluster-Creation}{}
-    \State $unmerged \leftarrow \{\{0\}, \{1\}, \{2\}, \dots, \{N-1\}\}$
-    \State $useful \leftarrow unmerged$
-    \While{$|unmerged|>1$}
-      \State $C_i, C_j \leftarrow \min_{C_i,C_j \in unmerged} D(C_i, C_j)$
-      \State $unmerged \leftarrow unmerged - \{C_i, C_j\} + \{C_i \cup C_j\}$
-      \State $useful \leftarrow useful + \{C_i \cup C_j\}$
-      \If{$D(C_i, C_j) = 0$}
-        \State $useful \leftarrow useful - \{C_i, C_j\}$
-      \EndIf
-    \EndWhile
-    \State Order $useful$ based based on last merged first\label{fig-cluster-creation-ordering}
-    \State Remove largest cluster from $useful$
-
-    \Return $useful$
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm describing how LTGA creates clusters using Equation~\ref{eq-distance}
-           for a population. $unmerged$ and $useful$ are ordered sets of sets of gene loci.}
-  \label{fig-cluster-creation}
-\end{figure}
-
-Figure~\ref{fig-cluster-creation} provides the agglomerative method LTGA uses to create gene clusters.
-This algorithm creates a tree of sets using pairwise gene entropy, such that the leaves of the tree contain a single gene and
-internal nodes are the union of their children's sets. These sets are then used by crossover to specify epistatic
-relationships which should be preserved.
-The process begins by creating the set of sets $unmerged$ which tracks all top level clusters. Initially
-$unmerged$ contains single member sets for each gene. After each iteration the two sets with the minimum average pairwise
-distance (given in Equation~\ref{eq-distance}) are merged to create a single cluster. This process is repeated
-until only a single set remains in $unmerged$ which contains all of the genes in the genome.
-
-\begin{equation}
-  D(C_i,C_j) = \frac{1}{\left | C_i \right |\cdot \left |C_j \right|}\sum_{c_i \in C_i}\sum_{c_j \in C_j}
-  2 - \frac{H(c_i) + H(c_j)}{H(c_i \cup c_j)}
-  \label{eq-distance}
-\end{equation}
-\begin{equation}
-  H(c) = -\sum_{s\in S} p_c(s)\log(p_c(s))
-  \label{eq-entropy}
-\end{equation}
-
-Throughout this process $useful$ tracks the set of all gene clusters which should be preserved for use by crossover.
-This set begins with all genes in separate clusters, and each time a new cluster is created it is added to $useful$.
-However, not all clusters are necessarily worth keeping. For instance, in all versions of LTGA the cluster
-containing all genes is removed from $useful$ as preserving all genes during crossover can only create clones.
-\cite{thierens:2013:ltgahiff} extended this removal to include any unsupported subsets. If the pairwise distance
-between two clusters is 0, this means there are no individuals in the population which disrupt the relationships between the two
-clusters. Therefore when performing crossover there is no reason to believe a fitness improvement can be achieved
-by breaking the stored pattern. As such a cluster is only kept if its direct superset has a non-zero distance.
-As a final step Line~\ref{fig-cluster-creation-ordering} reorders $useful$ such that clusters appear in reversed
-order from which they were added to $useful$. Thus the most linked clusters and those containing single genes
-appear at the end of the returned list.
-
-\cite{thierens:2013:ltgahiff}'s version of LTGA does not use the entire population when determining pairwise entropy.
-Instead, binary tournament is used to select half of the population. This is done to ensure the model is built
-using only high quality solutions, even during the first generation.
-
-In order to efficiently perform clustering, a pairwise gene frequency table is constructed from the selected solutions.
-To calculate Equation~\ref{eq-distance}, Equation~\ref{eq-entropy} is called for each
-gene ($H(c_i)$) and pair of genes ($H(c_i \cup c_j)$). Extracting this information requires $O(\mu N^2)$
-time, where $\mu$ is the population size and $N$ is the genome size. The process of converting
-this pairwise frequency information into clusters can be achieved in $O(N^2)$ using the bookkeeping
-methods presented by \cite{gronau:2007:upgma}. This cost is performed only once per generation,
-and is then used to perform approximately $O(\mu N)$ crossover evaluations. As a result, the amortized cost of
-LTGA's model building is $O(N)$.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Cluster-Usage}{}
-    \ForAll{$C_i \in useful$}
-      %\ForAll{$d \in shuffled(P_i)$}
-        \State $d \leftarrow rand\_choice(P)$\label{fig-cluster-usage-donate}
-        \State Copy $d$'s gene values for $C_i$ into solution
-        \If{solution was changed}
-          \If{solution's fitness decreased}
-            \State Revert changes
-          \EndIf
-          %\State \textbf{break}
-        \EndIf
-      %\EndFor
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm describing how clusters are used to perform crossover.}
-  \label{fig-cluster-usage}
-\end{figure}
-
-Figure~\ref{fig-cluster-usage} describes how the identified clusters are used by crossover to preserve
-gene linkage while still exploring the search space. Unlike more traditional crossover methods, LTGA
-crosses each individual with the entire population. Also, to produce a single offspring, multiple evaluations
-of the fitness function are performed.
-
-Each generation, each individual in the population undergoes crossover. In a single crossover event, each
-cluster of genes $C_i$ in $useful$ is applied as a crossover mask. A random donor $d$
-is chosen from the entire population (not just the model selected population), and $d$'s gene's for $C_i$ are copied
-into the working solution. If a modification is made, an evaluation is then performed. If the crossover
-resulted in no worse fitness then the changes are kept. This allows for neutral drift across plateaus.
-The resulting solution, which must be at least as fit as its parent, is then
-copied into the next generation.
-
-In total each individual can cause up to $|useful|$ evaluations. If all clusters were kept, even those deemed
-unhelpful, and all donations were evaluated, even those which did not change any genes, then \Call{Cluster-Usage}{}
-would perform exactly $2N-2$ evaluations for each of the $\mu$ solutions in the population. This provides the amortizing evaluations
-required to make clustering only $O(N)$ operations per evaluation. However, by skipping some evaluations, its
-possible that clustering may be super-linear.
-
-LTGA has no explicit form of diversity control and has no method for introducing new genetic information once
-the population has converged. Therefore an LTGA run is considered converged once two consecutive populations
-contain the same unique solutions.
-
-By design, LTGA only has a single parameter: population size. LTGA uses no mutation, and crossover is defined
-in terms of the clustering algorithm. Selection between generations is fully elitist and embedded in the crossover,
-with selection of model building solutions fixed to a binary tournament. Neither \Call{Cluster-Creation}{} nor
-\Call{Cluster-Usage}{} rely on parameter values. LTGA does not provide any method for controlling or setting
-the population size, relying instead on a fixed user specified size.
-
-\section{Parameter-less Population Pyramid}
-\label{sec-p3}
-\cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid (P3) as a method for
-performing optimization which does not require the user to provide any parameters. This is
-achieved by combining efficient local search with the model building methods of LTGA using
-an iteratively constructed hierarchy of populations.
-
-The high level algorithm of P3 is presented in Figure~\ref{fig-p3}. Unlike more traditional
-GAs, P3 does not follow a generational model. Instead, it maintains an iteratively
-expanding pyramid of expanding populations. Each iteration, a new random solution is generated.
-This solution is brought to a local optimum using the hill climbing algorithm in Figure~\ref{fig-hc}. If that
-local optimum has not yet been added to the pyramid, the solution is added to the lowest
-population $P_0$.
-
-Next, the solution is iteratively improved by applying LTGA's crossover algorithm (Figure~\ref{fig-cluster-usage})
-with each population $P_i$ in the pyramid. If this process results in a strict fitness improvement and has
-created a solution not yet stored in the pyramid, it is added to the next highest pyramid level $P_{i+1}$.
-If $P_{i+1}$ does not yet exist, it is created. In this way populations in the pyramid expand over time,
-and the number of populations stored increases over time. Initially the pyramid contains no solutions
-or populations, meaning the user does not need to specify a population size.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Iterate-P3}{}
-    \State Create random solution
-    \State Apply hill climber
-    \If{solution $\notin hashset$}
-      \State Add solution to $P_0$ and rebuild model
-      \State Add solution to $hashset$
-    \EndIf
-
-    \ForAll{$P_i \in pyramid$}
-      \State Mix solution with $P_i$
-      \If{solution's fitness has improved}
-        \If{solution $\notin hashset$}
-          \State Add solution to $P_{i+1}$ and rebuild model
-          \State Add solution to $hashset$
-        \EndIf
-      \EndIf
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-  \caption{One iteration of P3 optimization. $pyramid$ is an
-           ordered set of populations and $hashset$ is a set
-           of all solutions in $pyramid$.}
-  \label{fig-p3}
-\end{figure}
-
-To accommodate P3's unique population structure, some of LTGA's clustering procedures were modified. In LTGA,
-clusters are identified at the start of each generation and are used
-to create all offspring in that generation. As P3 does not perform serial
-generations, P3 instead rebuilds the model each
-time a solution is added to a population. Furthermore, unlike
-our chosen variant of LTGA, all solutions in the population are used to generate the model, not just the
-winners of a binary tournament. We can do this because even the
-worst solutions in the pyramid are already high quality due to local search.
-Using local search in LTGA was examined by \cite{bosman:2011:lsbbo} and found
-to provide no significant improvement. A likely cause was that that study applied
-local search to every solution, not just the initial population, resulting in significant overhead.
-
-Beyond the changes in population structuring, P3 modifies
-LTGA's version of \Call{Cluster-Creation}{} and \Call{Cluster-Usage}{}.
-P3 changes Line~\ref{fig-cluster-creation-ordering} in Figure~\ref{fig-cluster-creation}
-from \emph{last merged first} ordering to \emph{smallest first} ordering.
-This method applies gene clusters during crossover based on how many genes
-are in each cluster,\footnote{Ties are broken randomly.} and not on how tightly linked those genes are.
-\cite{goldman:2012:ltga} found that this alternative was better at preserving
-diversity, and therefore required smaller populations.
-
-P3 also modified Line~\ref{fig-cluster-usage-donate} in Figure~\ref{fig-cluster-usage}.
-Instead of choosing a single genetic donor for each cluster, P3 iterates over the
-population in a random order until a solution in the population is found which
-has a least one gene different for that cluster of genes from the improving solution.
-This process increases the likelihood of an evaluation being performed for every cluster,
-and helps test rare gene patterns in the population.
-
-In LTGA the cost of rebuilding the model is $O(\mu N^2)$ as it must collect pairwise
-gene frequency information for all $\mu$ solutions in the population. P3 does not store
-a single population, and it does not have a fixed $\mu$ size for any population. However,
-each time a solution is added to the population, it requires $O(N^2)$ time to update
-the table of pairwise frequencies and another $O(N^2)$ time to rebuild the linkage model.
-The model is then used immediately to perform up to one evaluation for each of the up
-to $2N-2$ clusters. Just as in LTGA, if no evaluation shortcuts were made, P3
-has an amortized cost of $O(N)$ modeling cost per evaluation. While P3 does
-rebuild the model more frequently per solution in the population, it also performs
-a number of local search evaluations which are quite efficient, meaning theoretical
-comparisons of their speed are difficult to perform. As a final note, P3's repeated
-attempts to find a useful donation make it less likely than LTGA to skip evaluations,
-but has an added cost to find these donations. Repeated donations could require as much
-as $O(\mu)$ attempts per evaluation, but experimental evidence suggest that this
-operation actually saves more overhead than it costs by increasing the number of evaluations
-per model rebuild.
-
-Each of the pieces of the P3 algorithm were selected not just for their stand alone efficacy,
-but for the ways in which they interact. By using the hill climber to ensure each solution
-is at a local optima, the underlying pairwise relationships in the problem are exposed. As
-a result, detecting clusters for use by crossover is much more effective. The crossover operator
-is extremely elitist, as each gene donation must result in no fitness loss, and a solution must
-strictly improve to be added to the next level of the pyramid. This is balanced by continual
-integration of new random solutions. Furthermore, each random restart decreases the probability
-of spurious linkages caused by shared ancestry. This diversity is further preserved by applying
-gene clusters in smallest first order during crossover as this reduces the probability of genetic
-hitchhikers.
-
-Other algorithms have proposed using multiple concurrent populations.
-\cite{hornby:2006:alps} had a hierarchy of populations with solutions periodically advancing
-upward. This allows for continuous integration of diversity as the lowest population is reseeded
-with random solutions. However, this method resulted in increased parameterization as not only was
-a population size required, but also new parameters for how frequently generations advanced between levels and
-how many total levels to have. \cite{harik:1999:parameterlessga} used multiple independent
-populations of different sizes as a method for removing the population size parameter, but
-doing was provably less efficient than using an optimal population size as no information is shared
-between the populations.
-
-\section{Problem Descriptions}
-\label{sec-problems}
-
-\subsection{Single Instance Problems}
-Understanding how a stochastic search algorithm will behave on arbitrary and complex
-search landscapes can be exceedingly difficult. Therefore a common practice for
-algorithm understanding is to perform search on well defined, well understood
-landscapes. To be of interest these landscapes need to represent interesting
-and important aspects of real world problems.
-
-Perhaps the most common such landscape is the Deceptive Trap problem \citep{goldberg:1991:gasize}.
-In this landscape the genome is broken up into $k$ bit non-overlapping subproblems referred
-to as traps.
-Each subproblem is scored using Equation~\ref{eq-deceptive-trap}, where $t$ is the number of
-bits in the trap set to 1. The global optimum in each trap is a string of all 1s, while all
-other solutions lead to a local optima of all 0s. This problem tests an algorithm's
-ability to overcome $k$ sized deception and is commonly used to determine how effective crossover
-is at preserving building blocks. Any crossover event that mixes bits from different parents in
-the same trap will likely result in that trap being optimized to the local optima. For our experiments
-we chose $k=7$ to ensure highly deceptive traps.
-
-\begin{equation}
-   trap(t) = \left\{
-     \begin{array}{rl}
-       k-1-t, &  t<k\\
-       k,   &  t = k
-     \end{array}
-   \right.
-  \label{eq-deceptive-trap}
-\end{equation}
-
-\cite{goldman:2012:ltga} found that mixing local search with linkage learning rendered the Deceptive
-Trap problem trivial. This is because local search is able to optimize each trap to one of the two
-optima (all 1s or all 0s), and then linkage learning can have perfect knowledge of gene interactions.
-In order to make the problem more difficult, the authors proposed the Deceptive
-Step Trap problem, given in Equation~\ref{eq-deceptive-step-trap}. This function modifies the results
-of Equation~\ref{eq-deceptive-trap} to include plateaus of size $s$, introducing an exponential number
-of local optima in each trap.
-With $k=7$ and $s=2$, as used in our experiments, all traps with 0, 1, 3, 5, and 7 bits set are local optima.
-This means that half of all ways to set the trap are 1 bit local optima.
-More generally, the number of local optima grows at $\Theta(2^{k-1})$.
-As a result, the Deceptive Step Trap is much more challenging for linkage
-learning techniques, while still being highly deceptive.
-\begin{equation}
-   step\_trap(t) = \left \lfloor \frac{(k-s)\pmod{s} + trap(t)}{s} \right \rfloor
-  \label{eq-deceptive-step-trap}
-\end{equation}
-
-Another challenging aspect of landscapes can be higher order relationships. The Hierarchical If
-and only If (HIFF) problem~\citep{watson:1998:hiff} is designed to capture the difficulties of this class of problem.
-In HIFF the genome is broken up into a complete binary tree, such that each gene appears in exactly one
-leaf and each internal node is the subset of genes contained in its children. If all genes represented
-in a node of the tree are set to the same value, they score equal to the size of the set. In this way
-small subsets lead toward solutions to larger subsets. However, a node can score if all genes are either
-all 1s or all 0s, meaning that to solve higher order subproblems it is necessary to perform crossovers
-that preserve lower order solutions. This problem is a natural fit for LTGA as the linkage tree can
-perfectly duplicate the problem's true relationships \citep{thierens:2013:ltgahiff}.
-
-As a final class of well known problems, we have chosen to borrow the Rastrigin problem from real valued
-optimization. This problem's landscape, determined by Equation~\ref{eq-rast}, is highly multimodal caused by
-the oscillating cosine function. \cite{goldman:2014:p3} proposed the Discretized Rastrigin problem, such
-that each floating point $x_i$ in Equation~\ref{eq-rast} is encoded using a 10 bit gray code.
-\begin{equation}
-  10n + \sum_{i=1}^{n}\left [ x_i^2-10\cos (2\pi x_i) \right ] \forall x\in [-5.12,5.12]
-  \label{eq-rast}
-\end{equation}
-
-\subsection{Randomly Generated Problem Classes}
-While well defined landscapes can provide specific insights into how an algorithm works,
-their static nature can be misleading. Specifically, algorithm quality might
-be very fragile such that it is only effective at searching well behaved landscapes. A more realistic
-test of an algorithm's black box effectiveness is to work with randomly generated instances
-drawn from a problem class. When tested over a sufficiently large sample it is then possible
-to draw more general conclusions about the algorithms effectiveness. The challenge with these
-landscapes is determining the global optimum to gauge if an algorithm was successful.
-
-Perhaps the most common model for generating random rugged landscapes is the NK model. An
-NK Landscape determines the fitness of each gene based on epistatic relationships with $K$
-other genes in the genome. This fitness is specified using a randomly generated table of
-fitness values, were each possible combination of the $K+1$ genes is mapped to some floating
-point value $[0-1]$. In unrestricted NK landscapes the relationships between genes are also
-randomly chosen and as a result finding the global optimum is $NP$-Hard for $K>1$. However,
-if epistasis is set such that each gene depends on the $K$ directly following it
-in the genome, the solution can be found in polynomial time~\citep{wright:2000:solvingnk}.
-These Nearest Neighbor NK landscapes are therefore ideal for search algorithm testing.
-For all of our experiments we fixed $K=5$ to ensure highly rugged landscapes.
-
-\cite{saul:1994:spinglass} presents a combinatorial benchmark problem derived from physics:
-Ising Spin Glasses. A spin glass is defined by a weighted graph of interaction terms between vertices.
-Each gene assigns a value to each vertex, with the fitness calculated by Equation~\ref{eq-ising}.
-In this equation, $E$ is the set of all edges, $e_{ij}$ is the edge weight connecting vertex $i$ to vertex $j$, and
-$x_i$ and $x_j$ are the gene values for vertex $i$ and $j$. Optimal fitness is when this sum is minimized.
-\begin{equation}
-\sum_{e_{ij} \in E} x_ie_{ij}x_j
-  \label{eq-ising}
-\end{equation}
-Similar to NK Landscapes, the general class is $NP$-Hard to optimize, but the $2D\pm J$ subset of
-Ising Spin Glasses can be polynomially solved.\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}
-In this subset the graph is restricted to be a two dimensional torus, edge weights are randomly set
-to either -1 or 1, and vertex values must be -1 or 1.
-
-As our final class of randomly generated problems we chose the Maximum Satisfiability (MAX-SAT) problem.
-Related to the more common 3-SAT problem, a MAX-SAT instance is defined by a set of three term clauses.
-Each term is a randomly chosen variable, which may also be negated. A clause scores if an and only if
-at least one term in the clause evaluates to true. In order to make MAX-SAT instances with a known global
-optimum, \cite{goldman:2014:p3} proposed constructing clauses around a fixed solution. In this way the
-signs of the terms are set to ensure the target solution satisfies the clause. To ensure each
-problem is challenging we chose a clause to variable ratio of $4.27$~\citep{selman:1996:maxsat}.
-
-\section{Comparison Algorithm Parameter Tuning}
-\label{sec-tuning}
-While four of the six algorithms in our experiments do not require any user specified parameters,
-hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
-handicapped, we extensively tuned each using the bisection method \citep{sastry:2001:bisection}
-to determine the optimal population size for each problem size.
-Extended by \cite{goldman:2012:ltga}, this method iteratively
-doubles the population size until some success criteria is met and then performs bisection
-between the lowest successful and highest unsuccessful sizes. In this way the minimum population size
-which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
-of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
-above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
-do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
-As bisection can make infinitely large population sizes, any run which had not found the global
-optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-hboa}
-      \end{centering}
-      \caption{Tuned Population Sizes for hBOA}
-      \label{fig-pop-hboa}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-ltga}
-      \end{centering}
-      \caption{Tuned Population Sizes for LTGA}
-      \label{fig-pop-ltga}
-    \end{subfigure}
-  \end{centering}
-  \caption{Optimal population sizes found using bisection on each size
-           for each problem.}
-  \label{fig-pop-sizes}
-\end{figure}
-
-Figure~\ref{fig-pop-sizes} shows the results from performing bisection. In general hBOA
-required population sizes which were at least an order of magnitude larger than LTGA. Due to
-runtime and memory overhead, finding the optimal value for hBOA was much less tractable than LTGA
-for moderate to large problem sizes. LTGA's population size also grew significantly slower
-than hBOA's as problem size increased, especially on the two Trap problems and Rastrigin.
-Both algorithms were very ineffective on MAX-SAT, neither able to tune to problems sizes over 60 bits.
-This is likely due to the fact that some randomly generated MAX-SAT landscapes are very flat
-and highly deceptive~\citep{rana:1998:gamaxsat}.
-\section{Finding the Global Optimum}
-\label{sec-optimum}
-
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{evals-to-success}
-  \end{center}
-  \caption{Comparison of the median number of evaluations to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-evals-to-success}
-\end{figure}
-
-Figure~\ref{fig-evals-to-success} shows the median number of evaluations required by each of the six
-algorithms to find the global optimum for multiple sizes of each problem. Each data point in
-Figure~\ref{fig-evals-to-success} represents the median of 100 runs, where unsuccessful runs
-are treated as requiring more evaluations than any successful run. If the median run was not successful
-no point is shown.
-Medians are used as the data is not normally distributed, and because it allows for more meaningful comparison
-between techniques with different success rates.
-The maximum problem size used for each problem was set to be the largest size we could
-feasibly determine LTGA's optimal population size. For many larger problems results are not shown
-for hBOA due to the extreme computational cost required to optimally set the population size.
-
-\subsection{Quantitative Comparison}
-Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
-The largest problem size for any problem where P3 was not the most efficient has 49 bits, with P3 achieving the best results on all 92 larger
-configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
-Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
-and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
-on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
-The likelihood that P3 would achieve these pairwise results assuming its median result is actually worse is $p < 10^{-15}$
-according to the binomial test.
-Pairwise comparison of LTGA and P3 on the largest problem size using the Mann-Whitney~U test results in
-$p < 10^{-5}$ for all problems.
-
-\subsection{Local Search}
-The Random Restart Hill Climber and $1+(\lambda, \lambda)$ are both relatively effective on small problem
-sizes. This is especially true for the three randomly generated problem classes. These problems
-may contain relatively few local optima or just be exceptionally difficult for the model based algorithms.
-On Deceptive Trap and Deceptive Step Trap using 4 or less traps, $1+(\lambda, \lambda)$ performs significantly
-better than any other algorithm. We believe this is
-because $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
-flipping entire traps.
-This ability also leads $1+(\lambda, \lambda)$
-to outperform the Random Restart Hill Climber on all problems except Nearest Neighbor NK.
-
-On larger problem sizes, the ability for local search to reach the global optimum quickly diminishes.
-Only on MAX-SAT are these optimizers competitive
-at larger tested problem sizes. However, we believe this is because the largest tested MAX-SAT
-was an order of magnitude smaller than the largest size tested for most other problems. As the problem
-size increases the number of local optima increases exponentially, which explains why Random Restart Hill
-Climbing was unable to scale.
-For larger problems it also becomes increasingly unlikely for $1+(\lambda, \lambda)$ to make the right combination
-of changes required to reach the global optimum. This
-behavior causes high variance in success rate, as evident by the occasional successes on large Deceptive Trap problems.
-
-\subsection{Model Building}
-Only techniques which explicitly built models of gene epistasis were able to solve the largest problem
-instances. On single instance problems LTGA was more effective than hBOA, with hBOA outperforming
-LTGA on Nearest Neighbor NK and Ising Spin Glasses. This may be caused by the differences in modeling method:
-unlike the single instance problems, gene epistasis in the randomly generated
-problem classes cannot be be perfectly represented with a linkage tree.
-
-Considering how different hBOA and LTGA are in performing optimization, it is somewhat surprising how similar
-their results are on HIFF.
-However, both techniques rely on populations large enough to support the diversity required to reach the global optimum
-and to model epistasis. Both techniques also only rebuild models once per generation. As the subproblems of HIFF
-are nested, it is unlikely that either technique can accurately model higher order epistasis before solving
-lower order subproblems. Therefore both methods require one generation per subproblem order.
-
-\subsection{P3}
-Unlike the other model based methods, P3 generally outperforms
-both Random Restart Hill Climber and $1+(\lambda, \lambda)$ even on small problem sizes.
-Unlike the other local search methods, P3 outperforms LTGA and hBOA even on large problem sizes.
-This implies that P3 is gaining the benefits of each, leveraging local search to solve easy
-problems and model building to solve harder ones.
-
-Furthermore, the interaction between these two optimization tools explains some of the reason P3
-outperforms each method alone.
-On Deceptive Trap P3's use of hill climbing ensures all traps are immediately optimized, allowing for perfect
-linkage detection and high quality donation.
-On HIFF local search solves all pairwise subproblems, saving P3 a generation over LTGA and hBOA.
-In comparison P3 is only a slight improvement on Deceptive Step Trap, which is less amenable to local search.
-
-\section{Fitness Over Time}
-\label{sec-overtime}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{fitness-over-time}
-  \end{center}
-  \caption{Compares the median best fitness reached during search for each of the six optimization methods.}
-  \label{fig-fitness-over-time}
-\end{figure}
-
-For some applications, finding the global optimum is less important than finding good solutions quickly. Therefore
-we examine this behavior in Figure~\ref{fig-fitness-over-time}. At regular intervals during optimization
-Figure~\ref{fig-fitness-over-time} shows the median of the best fitnesses found at that point of search across
-100 runs. For each problem we show the largest problem size for which we were able to successfully gather results
-for all six algorithms, but the trends shown are representative of all larger problem sizes. The maximum reporting interval
-is set to include the slowest P3 run to reach the global optimum.
-
-% 181 total
-% 121 has P3 best
-%   9 hc
-%  50 lambdalambda
-%  20 hboa
-%  18 phboa
-%  27 LTGA
-
-\subsection{Quantitative Comparison}
-Of 181 sample points, P3 had the highest median fitness in 121. In pairwise competition, $1+(\lambda, \lambda)$
-was the most likely to outperform P3, doing so on 50 sample points. LTGA, hBOA, and Parameter-less hBOA
-were the next best, outperforming P3 on 27, 20, and 18 sample points, respectively. Random Restart Hill Climbing almost
-never outperformed P3, doing so only 9 times.
-The likelihood that P3 would achieve these pairwise results assuming its median result is actually worse is $p < 10^{-9}$
-according to the binomial test.
-
-\subsection{Local Search}
-Perhaps the most striking result is the quality of $1+(\lambda, \lambda)$. Until quite far into search this
-method performs better than both LTGA and hBOA. Given sufficient evaluations $1+(\lambda, \lambda)$ also outperforms
-Random Restart Hill Climbing on all 7 problems. For brief periods in the middle of search it performs the
-best of all techniques on: Deceptive Trap; Deceptive Step Trap; HIFF; Ising Spin Glass; and MAX-SAT problems.
-$1+(\lambda, \lambda)$'s ability to efficiently incorporate gene modifications of larger
-than one bit allows it to overcome the deception and plateaus in Deceptive Trap and Deceptive Step Trap, solve medium sized subproblems
-in HIFF, flip the signs on multiple adjacent bits in Ising Spin Glass, and cross plateaus in MAX-SAT. However,
-this method is slow in reaching the global optima in many of these problems which causes it to eventually
-be overtaken by the model building techniques.
-
-\subsection{Model Building}
-Both hBOA and LTGA are marked by periods of little improvement followed by rapid improvement.
-In hBOA this is taken to the extreme, with all fitness
-improvement made at the very end of search. In both cases this is caused by model building. Before the model
-is accurate little improvement is made. Once it is accurate, fitness improves dramatically.
-
-At 58\% of the recording intervals hBOA has the worst fitness of any solver. Most of the exceptions
-occur when hBOA is still evaluating its initial population, allowing this random search to temporarily surpass the
-local search methods. After $N$ evaluations, however, hBOA and LTGA both fall behind until their models begin
-to improve.
-Parameter-less hBOA reaches intermediate fitnesses faster than hBOA, doing so on 62\% of intervals, as its models begin
-to optimize earlier than hBOA.  However, this trend is reversed after a sufficient number of evaluations, most clearly
-on Deceptive Step Trap and Ising Spin Glasses, as hBOA's tuned population overtakes Parameter-less hBOA's parallel populations.
-
-On every problem LTGA has five distinct periods: fitness plateau, near instantaneous improvement, fitness plateau,
-and improvement to global optimum. The early period corresponds with
-initialization of the population, with the first fitness gain achieved immediately upon completing the first generation.
-When using an inaccurate model, LTGA's mixing strategy performs a sort of less effective local search.
-Subsequent generations then make only minor fitness improvements. Once the model
-becomes accurate and the probability of a crossover using high quality genetic material increases sufficiently, LTGA
-enters a second period of rapid improvement.
-
-\begin{comment}
-This behavior is easiest to understand on the two Trap problems. LTGA's first generation will push individual traps toward a
-local optima, but it requires more than a single generation to do so completely. Until that occurs, the model is likely behaving
-no better than random as there is little apparent linkage between bits. Once the model begins to identify individual traps, crossover
-can begin to increase the frequency of higher fitness trap settings. This process begins slowly as low fitness local optima are more
-likely to be chosen for donation than high fitness local optima due to their frequency in the population. The process becomes
-self-catalysing as increased frequency of the global optimum trap genes means increased likelihood of the global optimum being spread by crossover.
-\end{comment}
-
-\subsection{P3}
-The integration of hill climbing into P3 makes it strictly better than using hill climbing alone.
-Early in optimization P3 and the Random Restart Hill Climber have effectively identical quality. This is because
-P3 performs the same evaluations as the Hill Climber for the first two restarts. Once P3 begins
-performing crossover it immediately improves over the Hill Climber. In 95\% of intervals P3 had a fitness at
-least as high as Hill Climbing. As such P3 is better than a simple
-hill climber regardless of how long each technique is run and irrespective of how high quality the solution
-found has to be.
-
-Unlike the model based methods, which struggle until model accuracy improves, P3's iterative solution
-integration allows it to improve much more quickly. This behavior exists in most problems, but is easiest to understand on Deceptive Trap.
-On this problem, P3 immediately brings all traps to local optima, equaled only by the Random Restart Hill Climber in quality.
-In comparison LTGA must evaluate the entire population and perform multiple generations to reach similar quality.
-P3 is able to immediately integrate optimal versions of each
-trap into a single individual as they are found by local search, resulting in smoother fitness improvement than LTGA.
-
-\section{Computational Expenses}
-While it is common in evolutionary computation to assume the evaluation function will dominate algorithm
-complexity, in some domains this will not be true. Model based methods are especially likely to violate
-this norm. Therefore, in order to assess P3's quality in solving problems with efficient fitness functions,
-we provide data on both its algorithmic complexity and wall clock time.
-
-\subsection{Operation Counting}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{rebuilds}
-      \end{centering}
-      \caption{Rebuilds}
-      \label{fig-rebuilds}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{donations}
-      \end{centering}
-      \caption{Donations}
-      \label{fig-donations}
-    \end{subfigure}
-  \end{centering}
-  \caption{Estimated computation costs incurred by model rebuilding (Figure~\ref{fig-rebuilds}) and
-           repeated donations (Figure~\ref{fig-donations}) per evaluation as problem size increases.}
-  \label{fig-costs}
-\end{figure}
-
-When discussing the asymptotic complexity of P3 in Section~\ref{sec-p3}, two aspects eluded precise
-analysis: how expensive is model rebuilding and how many gene donations are made. Figure~\ref{fig-costs}
-provides some insight into how often these two aspects of the algorithm are utilized.
-
-Figure~\ref{fig-rebuilds} reports in an algorithmic sense how expensive model rebuilding is for search.
-In order to calculate this value we recorded how many times search rebuilt the model during each run.
-Figure~\ref{fig-rebuilds} shows the estimated ratio of model rebuilding cost ($N^2$ per rebuild) over evaluation cost ($N$ per evaluation).
-If the cost of model building
-scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically constant.
-For Nearest Neighbor NK, Ising Spin Glasses, and Rastrigin this is the case. For both Trap problems
-and HIFF there is slow growth in the ratio. The problem sizes used for MAX-SAT were not sufficient
-to accurately gauge the asymptotic behavior. Together this suggests that while the cost of building the model
-is almost linear per evaluation, it can grow slowly. However, even in the worst case (HIFF) this growth was
-no more than twice the algorithmic cost of an evaluation even using 2048 bits.
-
-When applying a crossover subset, P3 tries random donors from the population until one is found with at least
-one bit different from the improving solution. In theory this can result in up to $O(\mu)$ operations.
-Figure~\ref{fig-donations} examines the observed average number of donations per evaluation performed.
-Ising Spin Glass, HIFF, and Rastrigin all achieve effectively constant behavior here, implying repeated
-donation does not impact the asymptotic runtime of P3. Both Trap functions and Nearest Neighbor NK all
-increase in number of donations as problem size increases, potentially increasing algorithmic costs. An
-important note is that each donation may range in size from a single bit up to $N-1$. However,
-repeated donation attempts are far more likely to happen with smaller clusters.
-As such this may cause some super-linear growth
-in P3, but its unlikely to be very high.
-
-\subsection{Wall Clock Performance}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{seconds-to-success}
-  \end{center}
-  \caption{Comparison of the median number of seconds to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-seconds-to-success}
-\end{figure}
-
-To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}.
-Similar to Figure~\ref{fig-evals-to-success}, each point represents the median
-of 100 runs, with unsuccessful runs treated as slower than successful runs.
-\begin{comment}
-In total these runs
-took over 6 computing years to complete. We were therefore forced to use
-diverse hardware potentially running multiple processes in parallel.
-As such these timings should be considered very skeptically as a number
-of sources may contribute to biased error in the results. However, due to
-the size of the dataset and the general regularity in the results some general conclusions
-can still be made.
-\end{comment}
-These results were collected using 2.5GHz Intel Xeon E5-2670v2 processors.
-
-\subsubsection{Model Building}
-hBOA and Parameter-less hBOA perform much worse when using wall clock time as the unit of comparison than when using
-evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while, under reasonable
-assumptions, P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
-hBOA goes from being slightly more efficient than LTGA in terms of evaluations to three orders of magnitude worse in terms of seconds.
-As P3 and LTGA require a similar asymptotic complexity per
-evaluation as the Hill Climber and $1+(\lambda, \lambda)$, no similar change in ordering occurs.
-
-\subsubsection{P3}
-When LTGA is optimally tuned to a single instance problem with an efficient evaluation function
-it can find the global optimum faster than P3 in terms of wall clock time. However, on randomly
-generated problem classes P3's efficient use of evaluations is enough to overtake LTGA.
-
-On the four single instance problems LTGA not only finds the global optimum using less wall clock time,
-the factor speedup increases as problem length does. Na\"{i}vely this suggests LTGA is achieving
-a lower order of complexity. However, for these experiments LTGA is growing at sub-linear time per
-evaluation, which is not asymptotically stable due to (at minimum) the time required to perform an evaluation.
-We suspect that the true cause is that $N$ is small enough to be overshadowed by lower order polynomial terms.
-For example, LTGA requires $O(N/\mu)$ time per evaluation to rebuild the linkage
-model from the frequency table. As a result, for small $\mu$ model building, and not extracting pairwise
-frequency, can dominate runtime.
-
-When applied to randomly generated problem classes, the differences in P3 and LTGA's evaluation complexity
-dominates runtime complexity. Similar to with Figure~\ref{fig-evals-to-success}, the amount of speedup P3
-achieves over LTGA increases with problem size on Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT.
-
-Across both types of problems we find that P3's time per evaluation grows approximately linearly. As such,
-we conclude that P3 requires asymptotically similar amounts of time per evaluation as the other efficient techniques.
-
-\section{Population Sizing}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-p3}
-      \end{centering}
-      \caption{As problem size increases}
-      \label{fig-pop-p3}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{solutions-stored}
-      \end{centering}
-      \caption{Largest problem size}
-      \label{fig-solutions-stored}
-    \end{subfigure}
-  \end{centering}
-  \caption{The total number of solutions stored by P3 when the global optimum is found.
-           In Figure~\ref{fig-solutions-stored} the red ``+'' indicates LTGA's tuned population size.}
-  \label{fig-p3-storage}
-\end{figure}
-
-A major advantage to P3 is that it does not require the user to set a population size parameter.
-Beyond making P3 easier to apply, this also conveys two additional advantages: diversity scaled
-to initialization and no need to sacrifice intermediate fitness for eventual optimality.
-
-Figure~\ref{fig-pop-p3} shows how the number of total solutions stored in the pyramid changes
-as problem size increases, similar to Figure~\ref{fig-pop-sizes} for hBOA and LTGA's tuned population sizes.
-As expected, the number of concurrently stored solutions increases as problem difficult increases, with the
-exact behavior dependent on the problem landscape. Figure~\ref{fig-solutions-stored} examines how the number
-of solutions stored is distributed on the largest problem sizes. Here we see that the behavior depends
-on the type of problem. On single instance problems P3's stored variance is relatively low, and generally
-higher than optimally tuned LTGA's population size. On randomly generated problem classes P3 has a much
-higher variance in stored solutions, but in general requires smaller sizes than LTGA.
-
-\subsection{Problem Instance versus Problem Class}
-Our procedure for tuning LTGA and hBOA outlined in Section~\ref{sec-tuning} involved finding
-the optimal population size for each class of problem. For real world black box optimization this is
-realistically the best either algorithm could hope for as tuning to a problem
-instance or population initialization involves repeatedly solving the problem being tuned.
-This limitation does not exist in parameter-less methods, which scale their diversity
-based on the problem instance without needing to solve that instance repeatedly.
-
-To achieve high success rates on randomly generated problem classes, LTGA and hBOA must use a population size
-large enough to solve the hardest instances in that class. Therefore these methods will have
-population sizes larger than necessary to solve the easiest instances in the class. Even
-on single instance problems, both methods will require population sizes large enough to
-ensure the worst random initialization is diverse enough to solve the problem, which may be
-much larger than the best random initialization.
-
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{evals-to-success-boxplot}
-  \end{center}
-  \caption{Distribution of evaluations required to reach the global optimum for
-           P3 and LTGA on the largest size of each problem.}
-  \label{fig-evals-to-success-boxplot}
-\end{figure}
-
-Figure~\ref{fig-evals-to-success-boxplot}
-highlights how this can effect the required number of evaluations to reach the global optimum, showing
-the distribution of results when solving the largest size of each problem. On each problem
-LTGA has a much smaller difference between its best and worst runs. This makes sense as LTGA uses the same
-population size regardless of instance and progresses search generationally. In contrast P3 has a much
-higher split, with many runs finishing very quickly.
-On all problems except Deceptive Step Trap, P3's upper quartile is lower than LTGA's lower quartile.
-Furthermore, on Deceptive Trap, HIFF, and Ising Spin Glasses, P3's worst run is better than LTGA's best run.
-For Nearest Neighbor NK, most of P3's runs finish much faster than the fastest LTGA runs. However, some
-of P3's outliers take approximately as long as LTGA's tuned performance.
-This supports the hypothesis that P3 is able to scale its diversity not just
-to the problem class, but to the problem instance or even problem initialization, something
-wholly infeasible for tuned population sizing to do.
-
-This tuning distinction is also apparent when comparing Parameter-less hBOA with hBOA in Figure~\ref{fig-evals-to-success}.
-While generally performing worse than hBOA, the difference between the two algorithms is smallest
-on randomly generated problem classes. On MAX-SAT, Parameter-less hBOA actually outperformed both hBOA and LTGA,
-likely due to its ability to scale diversity to the problem instance instead of the entire problem class.
-
-\subsection{Fast versus Optimal}
-\label{sec-fast-vs-optimal}
-
-In Section~\ref{sec-overtime} we examined intermediate fitness qualities of LTGA and hBOA when using
-population sizes tuned to reach the global optimum. As a result, both were exceptionally ineffective
-at quickly reaching high quality solutions. This is because unlike P3, these methods have an explicit
-trade off between optimal performance and intermediate performance caused by their population size parameter.
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{small-pop-dst}
-      \end{centering}
-      \caption{Deceptive Step Trap 203}
-      \label{fig-small-pop-dst}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{small-pop-nk}
-      \end{centering}
-      \caption{Nearest Neighbor NK 200}
-      \label{fig-small-pop-nk}
-    \end{subfigure}
-  \end{centering}
-  \caption{Comparison of how reducing LTGA's population size effects
-           the median best fitness reached during search.}
-  \label{fig-small-pop}
-\end{figure}
-
-Figure~\ref{fig-small-pop} examines the effect of population size on LTGA's intermediate fitness by
-reducing LTGA's population size to one tenth of the tuned value. The two problems shown are representative of the behavior of using
-a smaller population size on the other five problems. Reducing the population size caused LTGA to improve earlier but plateau
-at lower fitnesses. This caused LTGA's success rate to drop from 100 to 0 on Deceptive Step Trap and from 98 to 68 on Nearest
-Neighbor NK.
-Even when using a reduced population size for LTGA, P3 still achieved a fitness at
-least as high as LTGA at 80\% of intervals.
-The likelihood that P3 would achieve these pairwise results assuming its median result is actually worse is $p < 10^{-15}$
-according to the binomial test.
-
-\begin{comment}
-It is important to note that hBOA and LTGA were tuned specifically
-to find the global optimum and are not necessarily using the optimal population size for finding intermediate fitnesses. Consider
-that for both reducing the population size would almost certainly improve their fitness for evaluation steps less than their population
-size. To examine the effect of population size we tested LTGA using one tenth of the population size required for reliably finding the
-global optimum, with the results given in Figure~\ref{fig-small-pop}.
-Reducing the population size resulted in LTGA's initial fitness improvement
-occurring earlier, as the first generation is completed much more quickly. On Deceptive Step Trap this means that
-LTGA leaps ahead of P3 due to its ability to overcome the two bit fitness plateaus.
-%
-Across all tested problems using the smaller
-population size reduced the quality of LTGA's first fitness plateau, likely due to missing required diversity to reach higher quality.
-The second period of improvement also comes earlier and reaches a lower quality fitness. On Deceptive Step Trap 0
-runs reach the global optimum, down from 100 successful for the full population size. On Nearest Neighbor NK this drops the number
-of successful runs from 98 to 68. As a result we conclude that while reducing the population size of LTGA may improve its quality
-at early points during optimization, doing so reduces the robustness of the final solution found. This is in contrast to P3
-which balances high quality intermediate fitness without trading away eventual optimality.
-\end{comment}
-
-\section{Inner Workings}
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{cross}
-      \end{centering}
-      \caption{Crossover Proportion}
-      \label{fig-cross}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{cross-success}
-      \end{centering}
-      \caption{Crossover Success}
-      \label{fig-cross-success}
-    \end{subfigure}
-  \end{centering}
-  \caption{For each problem Figure~\ref{fig-cross} shows the proportion of P3 evaluations spend on crossovers
-           and Figure~\ref{fig-cross-success} shows the percentage of fitness improving crossover evaluations.}
-\end{figure}
-
-While analysis of optimization speed is useful from a practitioner standpoint, doing so provides very
-little insight into algorithm behavior. To better understand how P3 works in detail we present here
-a look at some internal features specific to P3.
-
-\subsection{Crossover}
-Figure~\ref{fig-cross} shows the proportion of evaluations P3 spends on crossover, as opposed to hill climbing,
-and Figure~\ref{fig-cross-success} shows what percentage of crossover evaluations resulted in a fitness improvement.
-Together these figures provide some insight into the role of crossover within P3. The behavior
-for each is clearly problem dependent and generally asymptoticly stable as problem size increases.
-
-When solving problems where epistasis can be effectively detected and represented by a linkage tree, P3 tends to spend
-less evaluations performing crossover and each crossover is more likely to be successful.
-Deceptive Trap and Rastrigin are the easiest problems to model epistasis, with local search quickly
-reducing pairwise entropy in each. These are also the problems where P3 uses the least evaluations
-on crossover and has the highest success rates for crossover. At the other extreme are Nearest Neighbor
-NK and Ising Spin Glasses, which both have overlapping linkage not representable by a linkage tree.
-These problems have the highest crossover usage and lowest crossover success of any problem except Deceptive
-Step Trap.  While Deceptive Step Trap's epistasis can be accurately modeled by a linkage tree, the
-exponential number of plateaus makes detecting gene linkage very challenging.
-
-\begin{comment}
-P3 performs optimization using a mixture of hill climbing and crossover. Exactly how much time the algorithm
-dedicates to each is not clear from the algorithm description alone. As such, Figure~\ref{fig-cross} shows
-that the relationship between the two is strongly problem dependent but relatively independent of problem size.
-In general, problems which can be more accurately represented by crossover
-required less crossover evaluations. Deceptive Trap and Rastrigin spent the least evaluations on crossover, using
-15\% and 20\% respectively on the largest problem sizes. Both of these problems are perfectly separable and have relatively
-few local optima in each subproblem. HIFF and Ising Spin Glasses use crossover for about 60\% of evaluations. The HIFF structure can
-be perfectly learned by the linkage tree, but the higher order relationships require multiple stages of crossover to optimize.
-Ising Spin Glasses have an overlapping structure, making them impossible to perfectly represent in a single linkage tree, but
-that overlap is relatively minor in comparison with Nearest Neighbor NK, which uses 81\% of evaluations on crossover. Deceptive Step Trap spends the highest
-percentage of evaluations on crossover, 92\%. Unlike all of the other problems, the value taken by an individual bit in Deceptive
-Step Trap has very little meaning. This is because any trap value is either a local optima or 1 bit flip away from being a local optima.
-This has a combined result of meaning it takes very few local search evaluations to find a local optima but very many crossover operations
-to proliferate the global optima. Epistatic relationships are also going to be very weak and even small population sizes are likely
-to mark all clusters as $useful$.
-
-Solutions are only added to the next highest pyramid level if at least one crossover donation resulted in a fitness
-improvement. Figure~\ref{fig-cross-success} examines the percentage of crossover donations that resulted in a fitness
-improvement. The ordering in Figure~\ref{fig-cross-success} is an almost perfect inversion of Figure~\ref{fig-cross},
-implying that the more successful crossover is at achieving fitness improving donations the lower the proportion of evaluations P3 will
-spend performing crossover. P3 achieves an impressive 30\% success rate for crossover donations on the largest Deceptive Trap
-problem sizes. This makes sense as crossover is able to immediately and perfectly identify entire traps such that all donations
-always contain only complete traps. Furthermore, by forcing crossover to attempt donations of genes different from the current solution,
-donations containing a single trap will have a very high success rate, always improving suboptimal traps which are far more frequently
-created by local search that optimal traps. Rastrigin and HIFF have relatively high success rates of 11\% and 7\%, respectively. Again
-this is likely due to accurate modeling and relatively low diversity of building block alternatives.
-Ising Spin Glasses and Deceptive Step Trap both have low success rates near 0.01\%. This follows as both are difficult to model with
-a linkage tree. In comparison to Figure~\ref{fig-cross-success} Deceptive Step Trap performs better than expected for crossover.
-However, the reason for this is that even exceedingly uninformed crossover can make improvements to Deceptive Step Trap as
-flipping two bits in the same trap will move between two local optima. Nearest Neighbor NK had the least successful crossovers,
-with only 0.007\% successful on the largest tested problems. In comparison to Ising Spin Glass, NK's overlap is far more extensive.
-As a result it makes sense that individual crossovers are very unlikely to be fitness improvements.
-\end{comment}
-
-When compared with LTGA, P3's crossover success rates are lower but similar in problem ordering.
-Counterintuitively, the use of hill climbing on the initial population reduces
-P3's crossover success, not because it reduces model quality or the donation pool, but because it is much more challenging to improve
-locally optimal solutions than randomly generated ones. LTGA's crossover benefits from application to unoptimized solutions, which
-makes its aggregate crossover success incomparable to P3's.
-
-Even when crossover success rates are quite low, such as Nearest Neighbor NK's 0.007\% success,
-the results from Section~\ref{sec-optimum} and Section~\ref{sec-overtime} show it's still critical to optimization. Without
-crossover, P3's performance would be identical to the Random Restart Hill Climber, which was unable to solve even moderately sized problems
-and quickly fell behind P3 in intermediate fitness quality. Therefore even infrequently successful crossover donations are
-critical to success. This does, however, suggest a potential avenue for future improvement by using more successful
-modeling and donation algorithms.
-
-\subsection{Pyramid}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{level-size}
-      \end{centering}
-      \caption{Population Size}
-      \label{fig-level-size}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{level-success}
-      \end{centering}
-      \caption{Crossover Success}
-      \label{fig-level-success}
-    \end{subfigure}
-  \end{centering}
-  \caption{For each problem Figure~\ref{fig-level-size} shows the number of solutions stored in each level of the pyramid
-           and Figure~\ref{fig-level-success} shows the percentage of fitness improving crossover evaluations at each level.}
-\end{figure}
-
-Another feature unique to P3 is the shape and size of the population pyramid
-constructed for each problem. Figure~\ref{fig-level-size} shows the number of solutions stored at
-each level of the pyramid for the largest tested problem sizes. Each point is the
-median size across 100 runs. If a run did not store any solutions at a level it is treated as 0.
-No point is drawn if the median run had 0 solutions stored at that level. While pyramid size
-is effected by problem size, the overall shape is not. As such the behavior shown in Figure~\ref{fig-level-size}
-is representative of that for all other tested problem sizes.
-
-With the exception of the dip in Deceptive Step Trap, all of the pyramids show a monotonic reduction
-in size as the level increases. This is because a solution must be a strict fitness improvement
-over its previous version to be added to a higher level, which becomes less likely
-each time the solution improves. \cite{lobo:2011:dynamicpop} found theoretical evidence
-and \cite{goldman:2011:dynamic-parameters} found empirical evidence that the optimal
-population size decreases each generation of traditional evolutionary search.
-By decreasing in size, P3 implicitly stores more diversity in low levels and focuses
-search on high quality solutions at higher levels. In comparison,
-LTGA and hBOA suboptimally use a fixed population size at each generation.
-
-Figure~\ref{fig-level-success} examines how crossover success changes at different
-levels of the pyramid. At low levels, success gets progressively lower
-as solution quality increases faster than the model's ability to improve solutions.
-At higher levels modeling becomes more accurate and donations contain higher frequencies
-of high quality building blocks, resulting in increased crossover success. The
-highest level of most problems has a low crossover success rate, as solutions
-crossing with that level have already been improved by previous operations to
-the point where the only improvement would be to create the global optimum, which can
-only happen once.
-
-\subsubsection{Deceptive Step Trap}
-The number of solutions stored at the fourth level of Deceptive Step Trap is
-significantly lower than that of the third or fifth levels, breaking the decreasing
-trend of the other six problems. Figure~\ref{fig-level-success} has a similar
-aberration, with crossover success dropping to  0.0004\% before rebounding and
-following the more common trajectory. This behavior exists in all other problem
-sizes tested, with the dips occurring at exactly the same level.
-
-This behavior is rooted in the peculiar nature
-of this landscape.
-After local search, all traps in all solutions have a total
-number of 1 bits equal to either 0, 1, 3, 5, or 7 as these correspond to the local
-optima when using $k=7$ and $s=2$. Crossover is very likely to overcome the
-two bit plateaus, and as a result solutions in the second level generally do not
-contain the lowest fitness local optima (5 bits set) and the third
-level has very few traps set to the next worst local optima (3 bits set). As a result,
-solutions which reach the third level can only be improved by replacing the deceptive
-local optima (0 and 1 bits set) with the global optimum (7 bits set).
-The global optimum is very rare in the population, and with 8 ways to represent local optima
-linkage learning is inaccurate. Therefore it is very unlikely for crossover to be successful,
-meaning few solutions will be added to the fourth level. Solutions that do improve by
-definition must have a higher frequency of optimal trap settings, meaning level four's model
-will be more accurate and donations are more likely to contain optimal trap values.
-Thus the level size and crossover success rates increase after contracting around level four.
-
-\section{Conclusions and Future Work}
+\chapter{Conclusions and Future Work}
 The Parameter-less Population Pyramid (P3) is a recently introduced method for performing black box
@@
 than optimally tuned state-of-the-art competitors. On single instance problems P3's improvement was by a
-constant factor, while for the three randomly generated problem classes P3's improvement increased with problem size. This
+constant factor, while for the three problem classes P3's improvement increased with problem size. This
 quality extends to intermediate points during evolution, with P3 generally reaching at least as high
@@
 modeling overhead, the expense of this overhead is approximately linear with respect to genome size. There
-is some evidence that even when compared on wall clock time, P3 performs on par with
+is some evidence that even when compared on wall clock time, P3 performs at least as well as
 the best comparison techniques. All of these achievements are made without any problem specific parameter
@@
 fitness. These limitations can be relaxed with future work to make P3 more widely applicable.
-While asymptotically linear in problem size, P3's modeling techniques and local search methods are likely going
+While asymptotically linear, P3's modeling techniques and local search methods are likely going
 to be prohibitively expensive for genome sizes in the hundreds of thousands or millions of genes,
@@
 global optima on black box problems without any problem specific tuning.
-
-\section{Acknowledgements}
-This material is based in part upon work supported by the National Science Foundation
-under Cooperative Agreement No. DBI-0939454. Any opinions, findings, and conclusions
-or recommendations expressed in this material are those of the author(s) and do not
-necessarily reflect the views of the National Science Foundation.
-
-\small
-
-\bibliographystyle{apalike}
-\bibliography{../main}
-
-\normalsize
-\appendix
-\section{hBOA Simplification}
-\label{sec-appendix}
-To measure the quality of a decision forest, hBOA applies Equation~\ref{eq-hboa}.
-To favor compact
-models, Equation~\ref{eq-hboa} is scaled by Equation~\ref{eq-hboa-stop}, which provides increased cost
-for more total leaves.
-This quality is used for two purposes: comparison of potential changes from the existing model
-and comparison of that change with the existing model. The basis for our simplication
-is to rearrange $p(B)BDe(B) < p(B')BDe(B')$ to be $\frac{p(B)}{p(B')}<\frac{BDe(B')}{BDe(B)}$.
-
-\begin{equation}
-  BDe(B) = \prod_{i=0}^{N}\prod_{l\in L_i} \frac{\Gamma(m'_i(l))}{\Gamma(m_i(l) + m'_i(l))}
-  \prod_{x_i}\frac{\Gamma(m_i(x_i, l) + m'_i(x_i,l))}{\Gamma(m'_i(x_i,l))}
-  \label{eq-hboa}
-\end{equation}
-
-\begin{equation}
-  p(B) = c2^{-0.5(\sum_i|L_i|)log_2\mu}
-  \label{eq-hboa-stop}
-\end{equation}
-
-The outermost product of Equation~\ref{eq-hboa} iterates over all trees in the forest. However, each split
-can modify only one of the trees and therefore the contribution of all others can be canceled. The middle
-product is across all leaves in the tree. Again since only one leaf can be changed, all other terms can
-be canceled. By convention hBOA uses uninformed Bayesian priors of $m'_i(l)= 2$ and $m'_i(x_i, l)=1$ for
-binary alphabets. As $\Gamma(a) = (a-1)!$ this means the top term in the middle product and the bottom
-term in the third product reduce to 1. The only remaining terms are then $m_i(l)$ and $m_i(x_i, l)$ which
-represent the number of solutions which reached leaf $l$ and the number of solutions which reached leaf $l$
-with a specific value for $x_i$, respectively.
-
-Equation~\ref{eq-hboa-stop} can also be simplified when doing comparisons. If model $B'$ has exactly one more
-leaf than model $B$ then the ratio $\frac{p(B)}{p(B')}$ simplifies to $2^{0.5 log_2\mu}$ regardless of
-total model size.
-
-The resulting simplifications create Equation~\ref{eq-hboa-final},
-where $B'$ is different from $B$ by exactly 1 split, such that $l$ was split to create $l'$ and $l''$.
-The best split is whichever maximizes its improvement over $B$, which is equal to the right side
-of the inequality. Note that these factorials can still be exceedingly large and
-therefore it is imperative that implementations avoid rounding errors and overflows.
-
-\end{document}

### 2015-04-13 Dissertation/chapter-graybox.tex
Copied Comprehensive to Disseration

--- 2014-10-27 Comprehensive/chapter-graybox.tex
@@
 \subsection{Preliminary Empirical Comparison}
+Figure~\ref{fig-hc-time} compares the black box hill climber presented in Section~\ref{sec-hill-climber}
+with the gray box hill climber presented in Section~\ref{sec-whitley-ls}.  Each algorithm was
+used 30 times on each of 30 instances to improve random solutions until they reached a local optimum.
+Both methods use first improvement hill climbing and achieved statistically indistinguishable fitness results.
+However, the amount of time each used is drastically different. On the largest tested size of $N=8,000$ the gray box
+hill climber was 800 times faster than using the black box method. This improvement was so extreme
+that the gray box hill climber is able to find local optima on $N=8,000$ problems faster than the black box
+method can optimize $N=400$ problems.  When compared asymptotically the black box optimizer empirically requires
+$O(N^{2.14})$ time while the gray box requires $O(N)$. This matches the theoretical conclusions for both algorithms.
+
+\section{Efficient Hamming Ball Search}
+Another consequence of the gray box domain is that increasing the neighborhood of local search becomes
+more tractable~\cite{chicano:2014:ball}.  Instead of improving solutions until no single bit flip
+is a fitness improvement, this can be expanded to no flip of $r$ or less bits can improve the solution.
+
+In a black box setting, verifying that no $r$ bit flip can improve a solution requires testing all
+$n \choose r$ neighbors. This quickly becomes intractable as $r$ increases. However in the gray
+box domain not all combinations need to be tested. Consider that if two variables do not participate
+in the same subfunction, the relationship between their effects is by definition additive. As such
+there is no way for flipping both together to be a fitness improvement without flipping one of them
+being a fitness improvement. Therefore it is not necessary to try all possible $r$ sized subsets
+of the solution.
+
 \begin{figure}[t]
@@
 \end{figure}
-
-Figure~\ref{fig-hc-time} compares the black box hill climber presented in Section~\ref{sec-hill-climber}
-with the gray box hill climber presented in Section~\ref{sec-whitley-ls}.  Each algorithm was
-used 30 times on each of 30 instances to improve random solutions until they reached a local optimum.
-Both methods use first improvement hill climbing and achieved statistically indistinguishable fitness results.
-However, the amount of time each used is drastically different. On the largest tested size of $N=8,000$ the gray box
-hill climber was 800 times faster than using the black box method. This improvement was so extreme
-that the gray box hill climber is able to find local optima on $N=8,000$ problems faster than the black box
-method can optimize $N=400$ problems.  When compared asymptotically the black box optimizer empirically requires
-$O(N^{2.14})$ time while the gray box requires $O(N)$. This matches the theoretical conclusions for both algorithms.
-
-\section{Efficient Hamming Ball Search}
-Another consequence of the gray box domain is that increasing the neighborhood of local search becomes
-more tractable~\cite{chicano:2014:ball}.  Instead of improving solutions until no single bit flip
-is a fitness improvement, this can be expanded to no flip of $r$ or less bits can improve the solution.
-
-In a black box setting, verifying that no $r$ bit flip can improve a solution requires testing all
-$n \choose r$ neighbors. This quickly becomes intractable as $r$ increases. However in the gray
-box domain not all combinations need to be tested. Consider that if two variables do not participate
-in the same subfunction, the relationship between their effects is by definition additive. As such
-there is no way for flipping both together to be a fitness improvement without flipping one of them
-being a fitness improvement. Therefore it is not necessary to try all possible $r$ sized subsets
-of the solution.
 

### 2015-04-13 Bongard/letter.tex
Added letter to Josh Bongard as example

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Plain Cover Letter
% LaTeX Template
% Version 1.0 (28/5/13)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% Rensselaer Polytechnic Institute 
% http://www.rpi.edu/dept/arc/training/latex/resumes/
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[11pt]{letter} % Default font size of the document, change to 10pt to fit more text

\usepackage{newcent} % Default font is the New Century Schoolbook PostScript font 
%\usepackage{helvet} % Uncomment this (while commenting the above line) to use the Helvetica font

% Margins
\topmargin=-1in % Moves the top of the document 1 inch above the default
\textheight=8.5in % Total height of the text on the page before text goes on to the next page, this can be increased in a longer letter
\oddsidemargin=-10pt % Position of the left margin, can be negative or positive if you want more or less room
\textwidth=6.5in % Total width of the text, increase this if the left margin was decreased and vice-versa

\let\raggedleft\raggedright % Pushes the date (at the top) to the left, comment this line to have the date on the right

\begin{document}

%----------------------------------------------------------------------------------------
%	ADDRESSEE SECTION
%----------------------------------------------------------------------------------------

\begin{letter}{
Josh Bongard \\
Morphology, Evolution \& Cognition Laboratory \\
University of Vermont}

%----------------------------------------------------------------------------------------
%	YOUR NAME & ADDRESS SECTION
%----------------------------------------------------------------------------------------

\begin{center}
\large\bf Brian W. Goldman \\ % Your name
%\vspace{20pt} \hrule height 1pt % If you would like a horizontal line separating the name from the address, uncomment the line to the left of this text
brianwgoldman@acm.org \\ (314) 313-1281 \\ Department of Computer Science and Engineering \\ Michigan State University % Your address and phone number
\end{center} 
\vfill

\signature{Brian W. Goldman} % Your name for the signature at the bottom

%----------------------------------------------------------------------------------------
%	LETTER CONTENT SECTION
%----------------------------------------------------------------------------------------

\opening{Dear Dr. Bongard,}

I am writing to apply for the Postdoctoral Associate position in your laboratory.
I am currently an all but dissertation Ph.D. student in BEACON and the Department of Computer Science and Engineering at Michigan State University,
and my defense is scheduled for July 1st, 2015. My primary research is in evolutionary computation,
focusing on advanced optimization methods.
In pursuing my research I have produced four journal articles and seven peer reviewed conference papers,
the most recent of which winning Best Paper in the Genetic Algorithms track of GECCO 2014.
%My publications have already received some attention from the field, with 54 citations and an h-index of 4.

My work in Genetic Programming (GP) has focused on how to improve the quality of evolutionary search.
Working with Cartesian Genetic Programming (CGP), a variant of GP which allows for
directed acyclic graphs and multiple program outputs, I discovered flaws in how the genetic operators interacted
with the genetic architecture. This lead me to propose modified operators which significantly improved search
efficiency and solution compactness.

An issue facing the field of GP is its reliance on simplistic and outdated benchmark problems. As part of a multi-institutional
collaboration I helped propose new tools for comparing proposed algorithms. By testing GP methods on more challenging
and more diverse problems we hope to direct the field toward even more success in tackling real-world applications.

More recently I have worked to develop the Parameter-less Population Pyramid (P3) boolean optimization algorithm. P3
utilizes advanced statistical modeling techniques to automatically detect epistatic relationships between the problem variables,
allowing for highly efficient crossover. In combination with local search and a novel population layering strategy, P3 is able
to outperform current state-of-the-art, perfectly tuned, optimization algorithms without receiving any problem specific configuration.
This work won Best Paper in the Genetic Algorithm's track of GECCO 2014, and lead to a collaboration with NASA mechanical engineers
to perform finite element design.

Through graduate coursework I have experience with many Machine Learning tools, such as Bayesian classifiers,
Support Vector Machines, Neural Networks, Random Forests, and Particle Swarm Optimization.
I received perfect grades in Pattern Recognition, Data Mining, Evolutionary Computation, Computational Intelligence,
and Artificial Intelligence.

I am very interested in the opportunity to apply my knowledge to the critically important domain of climate change offered by your
position. I believe this represents an excellent opportunity for new research
in advanced modeling which will in turn be applicable to a broad range of applications.

\closing{Thank you for your consideration,}

%\encl{Curriculum vitae, employment form} % List your enclosed documents here, comment this out to get rid of the "encl:"

%----------------------------------------------------------------------------------------

\end{letter}

\end{document}

### 2015-04-15 Dissertation/chapter-benchmarking.tex
Copied Comprehensive to Disseration
Major update from journal article polishing

--- 2014-10-15 Comprehensive/chapter-benchmarking.tex
@@
 \chapter{Benchmarking P3}
-TODO Explain the purpose of this chapter
+\label{chap-benchmarking}
 
@@
 on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
+The likelihood that P3 would achieve these pairwise results assuming its median result is actually worse is $p < 10^{-15}$
+according to the binomial test.
+Pairwise comparison of LTGA and P3 on the largest problem size using the Mann-Whitney~U test results in
+$p < 10^{-5}$ for all problems.
 
@@
 can still be made.
+\end{comment}
+These results were collected using 2.5GHz Intel Xeon E5-2670v2 processors.
 
 \subsubsection{Model Building}
-hBOA and Parameter-less hBOA perform significantly worse when using wall clock time as the unit of comparison than when using
+hBOA and Parameter-less hBOA perform much worse when using wall clock time as the unit of comparison than when using
 evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while, under reasonable
 assumptions, P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
-hBOA goes from being slightly more efficient than LTGA in terms of evaluations to significantly worse in terms of seconds.
+hBOA goes from being slightly more efficient than LTGA in terms of evaluations to three orders of magnitude worse in terms of seconds.
 As P3 and LTGA require a similar asymptotic complexity per
@@
 \subsubsection{P3}
-Any increase in P3's algorithmic complexity is offset by its efficient use of evaluations.
-In aggregate P3 found the global optimum the fastest on 72 of 130 configurations,
-with LTGA, $1+(\lambda, \lambda)$, Hill Climber, hBOA outperforming P3 on 44, 18, 10, and 1 configurations, respectively.
-On
-Deceptive Trap, Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT P3 still finds the global
-optimum faster than all comparison techniques. LTGA is able to match P3's performance almost exactly on HIFF,
-and on Rastrigin P3 falls behind only on moderate sized problems. Deceptive Step Trap yields the worst performance,
-with LTGA using optimal population sizes having a significantly lower growth curve than P3. This mirrors
-the results from Figure~\ref{fig-evals-to-success} under the assumption that P3's computational time per evaluation is higher
-than LTGA but by no more than a constant factor.
+When LTGA is optimally tuned to a single instance problem with an efficient evaluation function
+it can find the global optimum faster than P3 in terms of wall clock time. However, on randomly
+generated problem classes P3's efficient use of evaluations is enough to overtake LTGA.
+
+On the four single instance problems LTGA not only finds the global optimum using less wall clock time,
+the factor speedup increases as problem length does. Na\"{i}vely this suggests LTGA is achieving
+a lower order of complexity. However, for these experiments LTGA is growing at sub-linear time per
+evaluation, which is not asymptotically stable due to (at minimum) the time required to perform an evaluation.
+We suspect that the true cause is that $N$ is small enough to be overshadowed by lower order polynomial terms.
+For example, LTGA requires $O(N/\mu)$ time per evaluation to rebuild the linkage
+model from the frequency table. As a result, for small $\mu$ model building, and not extracting pairwise
+frequency, can dominate runtime.
+
+When applied to randomly generated problem classes, the differences in P3 and LTGA's evaluation complexity
+dominates runtime complexity. Similar to with Figure~\ref{fig-evals-to-success}, the amount of speedup P3
+achieves over LTGA increases with problem size on Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT.
+
+Across both types of problems we find that P3's time per evaluation grows approximately linearly. As such,
+we conclude that P3 requires asymptotically similar amounts of time per evaluation as the other efficient techniques.

### 2015-04-15 Dissertation/chapter-comparison_algorithms.tex
Major update from journal article polishing

--- 2015-03-03 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
@@
-\documentclass[twoside]{article}
-\usepackage{ecj,palatino,epsfig,latexsym,natbib}
-\usepackage{url}
-\usepackage{verbatim}
-\usepackage{subcaption}
-\usepackage[noend]{algpseudocode}
-
-\newcommand{\includegraphicswide}[1]
-{\includegraphics[width=.9\textwidth,height=\textheight,keepaspectratio]{#1}}
-
-\newcommand{\includegraphicsfit}[1]
-{\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}
-
-\usepackage{array}
-\newcolumntype{L}{>{\centering\arraybackslash}m{1.9cm}}
-
-%% do not add any other page- or text-size instruction here
-
-\parskip=0.00in
-
-\begin{document}
-
-% 45-character limit on header
-\ecjHeader{x}{x}{xxx-xxx}{201X}{Parameter-less Population Pyramid}{B. W. Goldman and W. F. Punch}
-\title{\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid}
-
-\author{\name{\bf B. W. Goldman} \hfill \addr{brianwgoldman@acm.org}\\ 
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-\AND
-       \name{\bf W. F. Punch} \hfill \addr{punch@msu.edu}\\
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-}
-
-\maketitle
-
-\begin{abstract}
-%Should be about 200 words
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
-evolutionary optimization without requiring any user specified parameters.
-P3's primary innovation is to replace the generational model with a pyramid of
-multiple populations that are iteratively created and expanded. In combination
-with local search and advanced crossover P3 scales to problem difficulty, exploiting
-previously learned information before adding more diversity.
-
-Across seven problems, each tested using on average 18 problem sizes, P3 outperformed
-all five advanced comparison algorithms. This improvement includes requiring less evaluations
-to find the global optimum and better fitness when using
-the same number of evaluations. Using both algorithm analysis and comparison we find P3's
-effectiveness is due to its ability to properly maintain, add, and exploit diversity.
-
-Unlike the best comparison algorithms, P3 was able to achieve this quality without any
-problem specific tuning. Thus, unlike previous parameter-less methods, P3 does not
-sacrifice quality for applicability. Therefore we conclude that
-P3 is an efficient, general, parameter-less approach to black box
-optimization which is more effective than existing state-of-the-art techniques.
-
-\end{abstract}
-
-\begin{keywords}
-
-Genetic algorithms, 
-linkage learning,
-local search,
-parameter-less.
-
-\end{keywords}
-
-\section{Introduction}
-A primary purpose of evolutionary optimization is to efficiently find good solutions
-to challenging real world problems with minimal prior knowledge about the problem itself.
-This driving goal has created search algorithms which can escape user bias to create
-truly novel results, sometimes publishable or patentable in their own right~\citep{kannappan:2014:humies}.
-While it is not possible for any algorithm to do better than random search across all possible
-problems~\citep{Wolpert:1997:nfl}, effectiveness can be achieved by assuming the search
-landscape has structure and then biasing the algorithm toward exploiting that structure.
-
-In evolutionary optimization, and genetic algorithms~(GAs) in particular, search is often
-biased through parameters. This can be beneficial as it allows practitioners to inject their
-knowledge about the shape of the search landscape into the algorithm.
-However, the quality of solutions found, and the speed at which they are found, is strongly tied to setting these parameters
-correctly~\citep{goldberg:1991:gasize}. As such, either expert knowledge or exceedingly
-expensive parameter tuning~\citep{grefenstette:1986:optimalga} are required to leverage
-this feature to its fullest potential. Furthermore,  parameters such as population size, mutation rate, crossover
-rate, tournament size, etc. usually have no clear relationship to the problem being solved, meaning even
-domain experts may not understand how the parameters will interact with the problem or with each other.
-To further complicate matters, there is mounting evidence that parameter values should change
-during search~\citep{goldman:2011:dynamic-parameters,laporte:2014:adaptivepop}.
-
-There have been periodic efforts to reduce or remove the need for parameter tuning.
-\cite{rechenberg:1973:es} introduced self-adaptive parameters, in which parameter values
-were included in each solution's genome and themselves underwent evolution. This allowed search
-to optimize some of its own parameters, resulting in a reduced need for expert tuning.
-\cite{harik:1999:parameterlessga} were able to design an entirely parameter-less GA by
-leveraging schema theory and parallel populations. Unfortunately these methods were provably less efficient
-than directly setting the parameters to optimal values~\citep{pelikan:1999:worstparameter-less}.
-
-One area that has been very effective at reducing the number of algorithm parameters is
-model based search. \cite{pelikan:2006:hboa}'s Hierarchical Bayesian Optimization
-Algorithm~(hBOA) and \cite{thierens:2010:ltga}'s Linkage Tree Genetic Algorithm~(LTGA)
-both require only a single parameter: population size. \cite{posik:2011:parameterless}
-leveraged model building to create a fully parameter-less algorithm, but it is restricted to
-only order-k, fully decomposable, noiseless problems.
-
-Most recently \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid~(P3).
-This method uses a pyramid structure of populations to combine model based search with local search
-to achieve parameter-less optimization. Initial results suggest that unlike
-previous parameter-less methods, P3 is actually more efficient than current state-of-the-art
-parameterized search algorithms. In this work we shall: extend these results to cover more
-comparison algorithms; compare both efficiency in reaching the global optimum and intermediate
-fitnesses; analyze algorithm complexity; and provide more in depth analysis of P3 itself.
-
-\section{Comparison Optimizers}
-\label{sec-optimizers}
+\chapter{Comparison Optimizers}
+\label{chap-optimizers}
 
@@
 
-\subsection{Linkage Tree Genetic Algorithm}
+\section{Linkage Tree Genetic Algorithm}
 \cite{thierens:2010:ltga} introduced the Linkage Tree Genetic Algorithm (LTGA) which automatically
@@
 
-\section{Parameter-less Population Pyramid}
-\label{sec-p3}
-\cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid (P3) as a method for
-performing optimization which does not require the user to provide any parameters. This is
-achieved by combining efficient local search with the model building methods of LTGA using
-an iteratively constructed hierarchy of populations.
-
-The high level algorithm of P3 is presented in Figure~\ref{fig-p3}. Unlike more traditional
-GAs, P3 does not follow a generational model. Instead, it maintains an iteratively
-expanding pyramid of expanding populations. Each iteration, a new random solution is generated.
-This solution is brought to a local optimum using the hill climbing algorithm in Figure~\ref{fig-hc}. If that
-local optimum has not yet been added to any level of the pyramid, the solution is added to the lowest
-population $P_0$.
-
-Next, the solution is iteratively improved by applying LTGA's crossover algorithm (Figure~\ref{fig-cluster-usage})
-with each population $P_i$ in the pyramid. If this process results in a strict fitness improvement and has
-created a solution not yet stored in the pyramid, it is added to the next highest pyramid level $P_{i+1}$.
-If $P_{i+1}$ does not yet exist, it is created. In this way populations in the pyramid expand over time,
-and the number of populations stored increases over time. Initially the pyramid contains no solutions
-or populations, meaning the user does not need to specify a population size.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Iterate-P3}{}
-    \State Create random solution
-    \State Apply hill climber
-    \If{solution $\notin hashset$}
-      \State Add solution to $P_0$ and rebuild $P_0$'s model
-      \State Add solution to $hashset$
-    \EndIf
-
-    \ForAll{$P_i \in pyramid$}
-      \State Mix solution with $P_i$
-      \If{solution's fitness has improved}
-        \If{solution $\notin hashset$}
-          \State Add solution to $P_{i+1}$ and rebuild $P_{i+1}$'s model
-          \State Add solution to $hashset$
-        \EndIf
-      \EndIf
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-  \caption{One iteration of P3 optimization. $pyramid$ is an
-           ordered set of populations and $hashset$ is a set
-           of all solutions in $pyramid$.}
-  \label{fig-p3}
-\end{figure}
-
-To accommodate P3's unique population structure, some of LTGA's clustering procedures were modified. In LTGA,
-clusters are identified at the start of each generation and are used
-to create all offspring in that generation. As P3 does not perform serial
-generations, P3 instead rebuilds the model each
-time a solution is added to a population. Furthermore, unlike
-our chosen variant of LTGA, all solutions in the population are used to generate the model, not just the
-winners of a binary tournament. We can do this because even the
-worst solutions in the pyramid are already high quality due to local search.
-Using local search in LTGA was examined by \cite{bosman:2011:lsbbo} and found
-to provide no significant improvement. A likely cause was that that study applied
-local search to every solution, not just the initial population, resulting in significant overhead.
-
-Beyond the changes in population structuring, P3 modifies
-LTGA's version of \Call{Cluster-Creation}{} and \Call{Cluster-Usage}{}.
-P3 changes Line~\ref{fig-cluster-creation-ordering} in Figure~\ref{fig-cluster-creation}
-from \emph{last merged first} ordering to \emph{smallest first} ordering.
-This method applies gene clusters during crossover based on how many genes
-are in each cluster,\footnote{Ties are broken randomly.} and not on how tightly linked those genes are.
-\cite{goldman:2012:ltga} found that this alternative was better at preserving
-diversity, and therefore required smaller populations.
-
-P3 also modified Line~\ref{fig-cluster-usage-donate} in Figure~\ref{fig-cluster-usage}.
-Instead of choosing a single genetic donor for each cluster, P3 iterates over the
-population in a random order until a solution in the population is found which
-has a least one gene different for that cluster of genes from the improving solution.
-This process increases the likelihood of an evaluation being performed for every cluster,
-and helps test rare gene patterns in the population.
-
-In LTGA the cost of rebuilding the model is $O(\mu N^2)$ as it must collect pairwise
-gene frequency information for all $\mu$ solutions in the population. P3 does not store
-a single population, and it does not have a fixed $\mu$ size for any population. However,
-each time a solution is added to the population, it requires $O(N^2)$ time to update
-the table of pairwise frequencies and another $O(N^2)$ time to rebuild the linkage model.
-The model is then used immediately to perform up to one evaluation for each of the up
-to $2N-2$ clusters. Just as in LTGA, if no evaluation shortcuts were made, P3
-has an amortized cost of $O(N)$ modeling cost per evaluation. While P3 does
-rebuild the model more frequently per solution in the population, it also performs
-a number of local search evaluations which are quite efficient, meaning theoretical
-comparisons of their speed are difficult to perform. As a final note, P3's repeated
-attempts to find a useful donation make it less likely than LTGA to skip evaluations,
-but has an added cost to find these donations. Repeated donations could require as much
-as $O(\mu)$ attempts per evaluation, but experimental evidence suggest that this
-operation actually saves more overhead than it costs by increasing the number of evaluations
-per model rebuild.
-
-Each of the pieces of the P3 algorithm were selected not just for their stand alone efficacy,
-but for the ways in which they interact. By using the hill climber to optimize randomly
-generated solutions, the underlying pairwise relationships in the problem are exposed. As
-a result, detecting clusters for use by crossover is much more effective. The crossover operator
-is extremely elitist, as each gene donation must result in no fitness loss, and a solution must
-strictly improve to be added to the next level of the pyramid. This is balanced by continual
-integration of new random solutions. Furthermore, each random restart decreases the probability
-of spurious linkages caused by shared ancestry. This diversity is further preserved by applying
-gene clusters in smallest first order during crossover as this reduces the probability of genetic
-hitchhikers.
-
-Other algorithms have proposed using multiple concurrent populations.
-\cite{hornby:2006:alps} had a hierarchy of populations with solutions periodically advancing
-upward. This allows for continuous integration of diversity as the lowest population is reseeded
-with random solutions. However, this method resulted in increased parameterization as not only was
-a population size required, but also new parameters for how frequently generations advanced between levels and
-how many total levels to have. \cite{harik:1999:parameterlessga} used multiple independent
-populations of different sizes as a method for removing the population size parameter, but
-doing was provably less efficient than using an optimal population size as no information is shared
-between the populations.
-
-\section{Problem Descriptions}
-\label{sec-problems}
-
-\subsection{Single Instance Problems}
-Understanding how a stochastic search algorithm will behave on arbitrary and complex
-search landscapes can be exceedingly difficult. Therefore a common practice for
-algorithm understanding is to perform search on well defined, well understood
-landscapes. To be of interest these landscapes need to represent interesting
-and important aspects of real world problems.
-
-Perhaps the most common such landscape is the Deceptive Trap problem \citep{goldberg:1991:gasize}.
-In this landscape the genome is broken up into $k$ bit non-overlapping subproblems referred
-to as traps.
-Each subproblem is scored using Equation~\ref{eq-deceptive-trap}, where $t$ is the number of
-bits in the trap set to 1. The global optimum in each trap is a string of all 1s, while all
-other solutions lead to a local optima of all 0s. This problem tests an algorithm's
-ability to overcome $k$ sized deception and is commonly used to determine how effective crossover
-is at preserving building blocks. Any crossover event that mixes bits from different parents in
-the same trap will likely result in that trap being optimized to the local optima. For our experiments
-we chose $k=7$ to ensure highly deceptive traps.
-
-\begin{equation}
-   trap(t) = \left\{
-     \begin{array}{rl}
-       k-1-t, &  t<k\\
-       k,   &  t = k
-     \end{array}
-   \right.
-  \label{eq-deceptive-trap}
-\end{equation}
-
-\cite{goldman:2012:ltga} found that mixing local search with linkage learning rendered the Deceptive
-Trap problem trivial. This is because local search is able to optimize each trap to one of the two
-optima (all 1s or all 0s), and then linkage learning can have perfect knowledge of gene interactions.
-In order to make the problem more difficult, the authors proposed the Deceptive
-Step Trap problem, given in Equation~\ref{eq-deceptive-step-trap}. This function modifies the results
-of Equation~\ref{eq-deceptive-trap} to include plateaus of size $s$, introducing an exponential number
-of local optima in each trap.
-With $k=7$ and $s=2$, as used in our experiments, all traps with 0, 1, 3, 5, and 7 bits set are local optima.
-This means that half of all ways to set the trap are 1 bit local optima.
-More generally, the number of local optima grows at $\Theta(2^{k-1})$.
-As a result, the Deceptive Step Trap is much more challenging for linkage
-learning techniques, while still being highly deceptive.
-\begin{equation}
-   step\_trap(t) = \left \lfloor \frac{(k-s)\pmod{s} + trap(t)}{s} \right \rfloor
-  \label{eq-deceptive-step-trap}
-\end{equation}
-
-Another challenging aspect of landscapes can be higher order relationships. The Hierarchical If
-and only If (HIFF) problem~\citep{watson:1998:hiff} is designed to capture the difficulties of this class of problem.
-In HIFF the genome is broken up into a complete binary tree, such that each gene appears in exactly one
-leaf and each internal node is the subset of genes contained in its children. If all genes represented
-in a node of the tree are set to the same value, they score equal to the size of the set. In this way
-small subsets lead toward solutions to larger subsets. However, a node can score if all genes are either
-all 1s or all 0s, meaning that to solve higher order subproblems it is necessary to perform crossovers
-that preserve lower order solutions. This problem is a natural fit for LTGA as the linkage tree can
-perfectly duplicate the problem's true relationships \citep{thierens:2013:ltgahiff}.
-
-As a final class of well known problems, we have chosen to borrow the Rastrigin problem from real valued
-optimization. This problem's landscape, determined by Equation~\ref{eq-rast}, is highly multimodal caused by
-the oscillating cosine function. \cite{goldman:2014:p3} proposed the Discretized Rastrigin problem, such
-that each floating point $x_i$ in Equation~\ref{eq-rast} is encoded using a 10 bit gray code.
-\begin{equation}
-  10n + \sum_{i=1}^{n}\left [ x_i^2-10\cos (2\pi x_i) \right ] \forall x\in [-5.12,5.12]
-  \label{eq-rast}
-\end{equation}
-
-\subsection{Randomly Generated Problem Classes}
-While well defined landscapes can provide specific insights into how an algorithm works,
-their static nature can be misleading. Specifically, algorithm quality might
-be very fragile such that it is only effective at searching well behaved landscapes. A more realistic
-test of an algorithm's black box effectiveness is to work with randomly generated instances
-drawn from a problem class. When tested over a sufficiently large sample it is then possible
-to draw more general conclusions about the algorithms effectiveness. The challenge with these
-landscapes is determining the global optimum to gauge if an algorithm was successful.
-
-Perhaps the most common model for generating random rugged landscapes is the NK model. An
-NK Landscape determines the fitness of each gene based on epistatic relationships with $K$
-other genes in the genome. This fitness is specified using a randomly generated table of
-fitness values, were each possible combination of the $K+1$ genes is mapped to some floating
-point value $[0-1]$. In unrestricted NK landscapes the relationships between genes are also
-randomly chosen and as a result finding the global optimum is $NP$-Hard for $K>1$. However,
-if epistasis is set such that each gene depends on the $K$ directly following it
-in the genome, the solution can be found in polynomial time~\citep{wright:2000:solvingnk}.
-These Nearest Neighbor NK landscapes are therefore ideal for search algorithm testing.
-For all of our experiments we fixed $K=5$ to ensure highly rugged landscapes.
-
-\cite{saul:1994:spinglass} presents a combinatorial benchmark problem derived from physics:
-Ising Spin Glasses. A spin glass is defined by a weighted graph of interaction terms between vertices.
-Each gene assigns a value to each vertex, with the fitness calculated by Equation~\ref{eq-ising}.
-In this equation, $E$ is the set of all edges, $e_{ij}$ is the edge weight connecting vertex $i$ to vertex $j$, and
-$x_i$ and $x_j$ are the gene values for vertex $i$ and $j$. Optimal fitness is when this sum is minimized.
-\begin{equation}
-\sum_{e_{ij} \in E} x_ie_{ij}x_j
-  \label{eq-ising}
-\end{equation}
-Similar to NK Landscapes, the general class is $NP$-Hard to optimize, but the $2D\pm J$ subset of
-Ising Spin Glasses can be polynomially solved.\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}
-In this subset the graph is restricted to be a two dimensional torus, edge weights are randomly set
-to either -1 or 1, and vertex values must be -1 or 1.
-
-As our final class of randomly generated problems we chose the Maximum Satisfiability (MAX-SAT) problem.
-Related to the more common 3-SAT problem, a MAX-SAT instance is defined by a set of three term clauses.
-Each term is a randomly chosen variable, which may also be negated. A clause scores if an and only if
-at least one term in the clause evaluates to true. In order to make MAX-SAT instances with a known global
-optimum, \cite{goldman:2014:p3} proposed constructing clauses around a fixed solution. In this way the
-signs of the terms are set to ensure the target solution satisfies the clause. To ensure each
-problem is challenging we chose a clause to variable ratio of $4.27$~\citep{selman:1996:maxsat}.
-
+%%%%%%%%%%%%%%%% Tuning %%%%%%%%%%%%%%%%%%%%%
 \section{Comparison Algorithm Parameter Tuning}
 \label{sec-tuning}
-While four of the six algorithms in our experiments do not require any user specified parameters,
-hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
-handicapped, we extensively tuned each using the bisection method \citep{sastry:2001:bisection}
-to determine the optimal population size for each problem size.
-Extended by \cite{goldman:2012:ltga}, this method iteratively
-doubles the population size until some success criteria is met and then performs bisection
-between the lowest successful and highest unsuccessful sizes. In this way the minimum population size
-which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
-of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
-above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
-do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
-As bisection can make infinitely large population sizes, any run which had not found the global
-optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.
 
@@
 
+While four of the six algorithms in our experiments do not require any user specified parameters,
+hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
+handicapped, we extensively tuned each using the bisection method \citep{sastry:2001:bisection}
+to determine the optimal population size for each problem size.
+Extended by \cite{goldman:2012:ltga}, this method iteratively
+doubles the population size until some success criteria is met and then performs bisection
+between the lowest successful and highest unsuccessful sizes. In this way the minimum population size
+which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
+of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
+above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
+do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
+As bisection can make infinitely large population sizes, any run which had not found the global
+optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.
+
 Figure~\ref{fig-pop-sizes} shows the results from performing bisection. In general hBOA
@@
 use the unmodified published versions.
-
-\section{Finding the Global Optimum}
-\label{sec-optimum}
-
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{evals-to-success}
-  \end{center}
-  \caption{Comparison of the median number of evaluations to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-evals-to-success}
-\end{figure}
-
-Figure~\ref{fig-evals-to-success} shows the median number of evaluations required by each of the six
-algorithms to find the global optimum for multiple sizes of each problem. Each data point in
-Figure~\ref{fig-evals-to-success} represents the median of 100 runs, where unsuccessful runs
-are treated as requiring more evaluations than any successful run. If the median run was not successful
-no point is shown.
-Medians are used as the data is not normally distributed, and because it allows for more meaningful comparison
-between techniques with different success rates.
-The maximum problem size used for each problem was set to be the largest size we could
-feasibly determine LTGA's optimal population size. For many larger problems results are not shown
-for hBOA due to the extreme computational cost required to optimally set the population size.
-
-\subsection{Quantitative Comparison}
-Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
-The largest problem size for any problem where P3 was not the most efficient has 49 bits, with P3 achieving the best results on all 92 larger
-configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
-Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
-and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
-on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
-The likelihood that P3 would achieve these pairwise results assuming its median result is actually worse is $p < 10^{-15}$
-according to the binomial test.
-Pairwise comparison of LTGA and P3 on the largest problem size using the Mann-Whitney~U test results in
-$p < 10^{-5}$ for all problems.
-
-\subsection{Local Search}
-The Random Restart Hill Climber and $1+(\lambda, \lambda)$ are both relatively effective on small problem
-sizes. This is especially true for the three randomly generated problem classes. These problems
-may contain relatively few local optima or just be exceptionally difficult for the model based algorithms.
-On Deceptive Trap and Deceptive Step Trap using 4 or less traps, $1+(\lambda, \lambda)$ performs significantly
-better than any other algorithm. We believe this is
-because $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
-flipping entire traps.
-This ability also leads $1+(\lambda, \lambda)$
-to outperform the Random Restart Hill Climber on all problems except Nearest Neighbor NK.
-
-On larger problem sizes, the ability for local search to reach the global optimum quickly diminishes.
-Only on MAX-SAT are these optimizers competitive
-at larger tested problem sizes. However, we believe this is because the largest tested MAX-SAT
-was an order of magnitude smaller than the largest size tested for most other problems. As the problem
-size increases the number of local optima increases exponentially, which explains why Random Restart Hill
-Climbing was unable to scale.
-For larger problems it also becomes increasingly unlikely for $1+(\lambda, \lambda)$ to make the right combination
-of changes required to reach the global optimum. This
-behavior causes high variance in success rate, as evident by the occasional successes on large Deceptive Trap problems.
-
-\subsection{Model Building}
-Only techniques which explicitly built models of gene epistasis were able to solve the largest problem
-instances. On single instance problems LTGA was more effective than hBOA, with hBOA outperforming
-LTGA on Nearest Neighbor NK and Ising Spin Glasses. This may be caused by the differences in modeling method:
-unlike the single instance problems, gene epistasis in the randomly generated
-problem classes cannot be be perfectly represented with a linkage tree.
-
-Considering how different hBOA and LTGA are in performing optimization, it is somewhat surprising how similar
-their results are on HIFF.
-However, both techniques rely on populations large enough to support the diversity required to reach the global optimum
-and to model epistasis. Both techniques also only rebuild models once per generation. As the subproblems of HIFF
-are nested, it is unlikely that either technique can accurately model higher order epistasis before solving
-lower order subproblems. Therefore both methods require one generation per subproblem order.
-
-\subsection{P3}
-Unlike the other model based methods, P3 generally outperforms
-both Random Restart Hill Climber and $1+(\lambda, \lambda)$ even on small problem sizes.
-Unlike the other local search methods, P3 outperforms LTGA and hBOA even on large problem sizes.
-This implies that P3 is gaining the benefits of each, leveraging local search to solve easy
-problems and model building to solve harder ones.
-
-Furthermore, the interaction between these two optimization tools explains some of the reason P3
-outperforms each method alone.
-On Deceptive Trap P3's use of hill climbing ensures all traps are immediately optimized, allowing for perfect
-linkage detection and high quality donation.
-On HIFF local search solves all pairwise subproblems, saving P3 a generation over LTGA and hBOA.
-In comparison P3 is only a slight improvement on Deceptive Step Trap, which is less amenable to local search.
-
-\section{Fitness Over Time}
-\label{sec-overtime}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{fitness-over-time}
-  \end{center}
-  \caption{Compares the median best fitness reached during search for each of the six optimization methods.}
-  \label{fig-fitness-over-time}
-\end{figure}
-
-For some applications, finding the global optimum is less important than finding good solutions quickly. Therefore
-we examine this behavior in Figure~\ref{fig-fitness-over-time}. At regular intervals during optimization
-Figure~\ref{fig-fitness-over-time} shows the median of the best fitnesses found at that point of search across
-100 runs. For each problem we show the largest problem size for which we were able to successfully gather results
-for all six algorithms, but the trends shown are representative of all larger problem sizes. The maximum reporting interval
-is set to include the slowest P3 run to reach the global optimum.
-
-% 181 total
-% 121 has P3 best
-%   9 hc
-%  50 lambdalambda
-%  20 hboa
-%  18 phboa
-%  27 LTGA
-
-\subsection{Quantitative Comparison}
-Of 181 sample points, P3 had the highest median fitness in 121. In pairwise competition, $1+(\lambda, \lambda)$
-was the most likely to outperform P3, doing so on 50 sample points. LTGA, hBOA, and Parameter-less hBOA
-were the next best, outperforming P3 on 27, 20, and 18 sample points, respectively. Random Restart Hill Climbing almost
-never outperformed P3, doing so only 9 times.
-The likelihood that P3 would achieve these pairwise results assuming its median result is actually worse is $p < 10^{-9}$
-according to the binomial test.
-
-\subsection{Local Search}
-Perhaps the most striking result is the quality of $1+(\lambda, \lambda)$. Until quite far into search this
-method performs better than both LTGA and hBOA. Given sufficient evaluations $1+(\lambda, \lambda)$ also outperforms
-Random Restart Hill Climbing on all 7 problems. For brief periods in the middle of search it performs the
-best of all techniques on: Deceptive Trap; Deceptive Step Trap; HIFF; Ising Spin Glass; and MAX-SAT problems.
-$1+(\lambda, \lambda)$'s ability to efficiently incorporate gene modifications of larger
-than one bit allows it to overcome the deception and plateaus in Deceptive Trap and Deceptive Step Trap, solve medium sized subproblems
-in HIFF, flip the signs on multiple adjacent bits in Ising Spin Glass, and cross plateaus in MAX-SAT. However,
-this method is slow in reaching the global optima in many of these problems which causes it to eventually
-be overtaken by the model building techniques.
-
-\subsection{Model Building}
-Both hBOA and LTGA are marked by periods of little improvement followed by rapid improvement.
-In hBOA this is taken to the extreme, with all fitness
-improvement made at the very end of search. In both cases this is caused by model building. Before the model
-is accurate little improvement is made. Once it is accurate, fitness improves dramatically.
-
-At 58\% of the recording intervals hBOA has the worst fitness of any solver. Most of the exceptions
-occur when hBOA is still evaluating its initial population, allowing this random search to temporarily surpass the
-local search methods. After $N$ evaluations, however, hBOA and LTGA both fall behind until their models begin
-to improve.
-Parameter-less hBOA reaches intermediate fitnesses faster than hBOA, doing so on 62\% of intervals, as its models begin
-to optimize earlier than hBOA.  However, this trend is reversed after a sufficient number of evaluations, most clearly
-on Deceptive Step Trap and Ising Spin Glasses, as hBOA's tuned population overtakes Parameter-less hBOA's parallel populations.
-
-On every problem LTGA has five distinct periods: fitness plateau, near instantaneous improvement, fitness plateau,
-and improvement to global optimum. The early period corresponds with
-initialization of the population, with the first fitness gain achieved immediately upon completing the first generation.
-When using an inaccurate model, LTGA's mixing strategy performs a sort of less effective local search.
-Subsequent generations then make only minor fitness improvements. Once the model
-becomes accurate and the probability of a crossover using high quality genetic material increases sufficiently, LTGA
-enters a second period of rapid improvement.
-
-\begin{comment}
-This behavior is easiest to understand on the two Trap problems. LTGA's first generation will push individual traps toward a
-local optima, but it requires more than a single generation to do so completely. Until that occurs, the model is likely behaving
-no better than random as there is little apparent linkage between bits. Once the model begins to identify individual traps, crossover
-can begin to increase the frequency of higher fitness trap settings. This process begins slowly as low fitness local optima are more
-likely to be chosen for donation than high fitness local optima due to their frequency in the population. The process becomes
-self-catalysing as increased frequency of the global optimum trap genes means increased likelihood of the global optimum being spread by crossover.
-\end{comment}
-
-\subsection{P3}
-The integration of hill climbing into P3 makes it strictly better than using hill climbing alone.
-Early in optimization P3 and the Random Restart Hill Climber have effectively identical quality. This is because
-P3 performs the same evaluations as the Hill Climber for the first two restarts. Once P3 begins
-performing crossover it immediately improves over the Hill Climber. In 95\% of intervals P3 had a fitness at
-least as high as Hill Climbing. As such P3 is better than a simple
-hill climber regardless of how long each technique is run and irrespective of how high quality the solution
-found has to be.
-
-Unlike the model based methods, which struggle until model accuracy improves, P3's iterative solution
-integration allows it to improve much more quickly. This behavior exists in most problems, but is easiest to understand on Deceptive Trap.
-On this problem, P3 immediately brings all traps to local optima, equaled only by the Random Restart Hill Climber in quality.
-In comparison LTGA must evaluate the entire population and perform multiple generations to reach similar quality.
-P3 is able to immediately integrate optimal versions of each
-trap into a single individual as they are found by local search, resulting in smoother fitness improvement than LTGA.
-
-\section{Computational Expenses}
-While it is common in evolutionary computation to assume the evaluation function will dominate algorithm
-complexity, in some domains this will not be true. Model based methods are especially likely to violate
-this norm. Therefore, in order to assess P3's quality in solving problems with efficient fitness functions,
-we provide data on both its algorithmic complexity and wall clock time.
-
-\subsection{Operation Counting}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{rebuilds}
-      \end{centering}
-      \caption{Rebuilds}
-      \label{fig-rebuilds}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{donations}
-      \end{centering}
-      \caption{Donations}
-      \label{fig-donations}
-    \end{subfigure}
-  \end{centering}
-  \caption{Estimated computation costs incurred by model rebuilding (Figure~\ref{fig-rebuilds}) and
-           repeated donations (Figure~\ref{fig-donations}) per evaluation as problem size increases.}
-  \label{fig-costs}
-\end{figure}
-
-When discussing the asymptotic complexity of P3 in Section~\ref{sec-p3}, two aspects eluded precise
-analysis: how expensive is model rebuilding and how many gene donations are made. Figure~\ref{fig-costs}
-provides some insight into how often these two aspects of the algorithm are utilized.
-
-Figure~\ref{fig-rebuilds} reports in an algorithmic sense how expensive model rebuilding is for search.
-In order to calculate this value we recorded how many times search rebuilt the model during each run.
-Figure~\ref{fig-rebuilds} shows the estimated ratio of model rebuilding cost ($N^2$ per rebuild) over evaluation cost ($N$ per evaluation).
-If the cost of model building
-scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically constant.
-For Nearest Neighbor NK, Ising Spin Glasses, and Rastrigin this is the case. For both Trap problems
-and HIFF there is slow growth in the ratio. The problem sizes used for MAX-SAT were not sufficient
-to accurately gauge the asymptotic behavior. Together this suggests that while the cost of building the model
-is almost linear per evaluation, it can grow slowly. However, even in the worst case (HIFF) this growth was
-no more than twice the algorithmic cost of an evaluation even using 2048 bits.
-
-When applying a crossover subset, P3 tries random donors from the population until one is found with at least
-one bit different from the improving solution. In theory this can result in up to $O(\mu)$ operations.
-Figure~\ref{fig-donations} examines the observed average number of donations per evaluation performed.
-Ising Spin Glass, HIFF, and Rastrigin all achieve effectively constant behavior here, implying repeated
-donation does not impact the asymptotic runtime of P3. Both Trap functions and Nearest Neighbor NK all
-increase in number of donations as problem size increases, potentially increasing algorithmic costs. An
-important note is that each donation may range in size from a single bit up to $N-1$. However,
-repeated donation attempts are far more likely to happen with smaller clusters.
-As such this may cause some super-linear growth
-in P3, but its unlikely to be very high.
-
-\subsection{Wall Clock Performance}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{seconds-to-success}
-  \end{center}
-  \caption{Comparison of the median number of seconds to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-seconds-to-success}
-\end{figure}
-
-To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}.
-Similar to Figure~\ref{fig-evals-to-success}, each point represents the median
-of 100 runs, with unsuccessful runs treated as slower than successful runs.
-\begin{comment}
-In total these runs
-took over 6 computing years to complete. We were therefore forced to use
-diverse hardware potentially running multiple processes in parallel.
-As such these timings should be considered very skeptically as a number
-of sources may contribute to biased error in the results. However, due to
-the size of the dataset and the general regularity in the results some general conclusions
-can still be made.
-\end{comment}
-These results were collected using 2.5GHz Intel Xeon E5-2670v2 processors.
-
-\subsubsection{Model Building}
-hBOA and Parameter-less hBOA perform much worse when using wall clock time as the unit of comparison than when using
-evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while, under reasonable
-assumptions, P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
-hBOA goes from being slightly more efficient than LTGA in terms of evaluations to three orders of magnitude worse in terms of seconds.
-As P3 and LTGA require a similar asymptotic complexity per
-evaluation as the Hill Climber and $1+(\lambda, \lambda)$, no similar change in ordering occurs.
-
-\subsubsection{P3}
-When LTGA is optimally tuned to a single instance problem with an efficient evaluation function
-it can find the global optimum faster than P3 in terms of wall clock time. However, on randomly
-generated problem classes P3's efficient use of evaluations is enough to overtake LTGA.
-
-On the four single instance problems LTGA not only finds the global optimum using less wall clock time,
-the factor speedup increases as problem length does. Na\"{i}vely this suggests LTGA is achieving
-a lower order of complexity. However, for these experiments LTGA is growing at sub-linear time per
-evaluation, which is not asymptotically stable due to (at minimum) the time required to perform an evaluation.
-We suspect that the true cause is that $N$ is small enough to be overshadowed by lower order polynomial terms.
-For example, LTGA requires $O(N/\mu)$ time per evaluation to rebuild the linkage
-model from the frequency table. As a result, for small $\mu$ model building, and not extracting pairwise
-frequency, can dominate runtime.
-
-When applied to randomly generated problem classes, the differences in P3 and LTGA's evaluation complexity
-dominates runtime complexity. Similar to with Figure~\ref{fig-evals-to-success}, the amount of speedup P3
-achieves over LTGA increases with problem size on Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT.
-
-Across both types of problems we find that P3's time per evaluation grows approximately linearly. As such,
-we conclude that P3 requires asymptotically similar amounts of time per evaluation as the other efficient techniques.
-
-\section{Population Sizing}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-p3}
-      \end{centering}
-      \caption{As problem size increases}
-      \label{fig-pop-p3}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{solutions-stored}
-      \end{centering}
-      \caption{Largest problem size}
-      \label{fig-solutions-stored}
-    \end{subfigure}
-  \end{centering}
-  \caption{The total number of solutions stored by P3 when the global optimum is found.
-           In Figure~\ref{fig-solutions-stored} the red ``+'' indicates LTGA's tuned population size.}
-  \label{fig-p3-storage}
-\end{figure}
-
-A major advantage to P3 is that it does not require the user to set a population size parameter.
-Beyond making P3 easier to apply, this also conveys two additional advantages: diversity scaled
-to initialization and no need to sacrifice intermediate fitness for eventual optimality.
-
-Figure~\ref{fig-pop-p3} shows how the number of total solutions stored in the pyramid changes
-as problem size increases, similar to Figure~\ref{fig-pop-sizes} for hBOA and LTGA's tuned population sizes.
-As expected, the number of concurrently stored solutions increases as problem difficult increases, with the
-exact behavior dependent on the problem landscape. Figure~\ref{fig-solutions-stored} examines how the number
-of solutions stored is distributed on the largest problem sizes. Here we see that the behavior depends
-on the type of problem. On single instance problems P3's stored variance is relatively low, and generally
-higher than optimally tuned LTGA's population size. On randomly generated problem classes P3 has a much
-higher variance in stored solutions, but in general requires smaller sizes than LTGA.
-
-\subsection{Problem Instance versus Problem Class}
-Our procedure for tuning LTGA and hBOA outlined in Section~\ref{sec-tuning} involved finding
-the optimal population size for each class of problem. For real world black box optimization this is
-realistically the best either algorithm could hope for as tuning to a problem
-instance or population initialization involves repeatedly solving the problem being tuned.
-This limitation does not exist in parameter-less methods, which scale their diversity
-based on the problem instance without needing to solve that instance repeatedly.
-
-To achieve high success rates on randomly generated problem classes, LTGA and hBOA must use a population size
-large enough to solve the hardest instances in that class. Therefore these methods will have
-population sizes larger than necessary to solve the easiest instances in the class. Even
-on single instance problems, both methods will require population sizes large enough to
-ensure the worst random initialization is diverse enough to solve the problem, which may be
-much larger than the best random initialization.
-
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{evals-to-success-boxplot}
-  \end{center}
-  \caption{Distribution of evaluations required to reach the global optimum for
-           P3 and LTGA on the largest size of each problem.}
-  \label{fig-evals-to-success-boxplot}
-\end{figure}
-
-Figure~\ref{fig-evals-to-success-boxplot}
-highlights how this can effect the required number of evaluations to reach the global optimum, showing
-the distribution of results when solving the largest size of each problem. On each problem
-LTGA has a much smaller difference between its best and worst runs. This makes sense as LTGA uses the same
-population size regardless of instance and progresses search generationally. In contrast P3 has a much
-higher split, with many runs finishing very quickly.
-On all problems except Deceptive Step Trap, P3's upper quartile is lower than LTGA's lower quartile.
-Furthermore, on Deceptive Trap, HIFF, and Ising Spin Glasses, P3's worst run is better than LTGA's best run.
-For Nearest Neighbor NK, most of P3's runs finish much faster than the fastest LTGA runs. However, some
-of P3's outliers take approximately as long as LTGA's tuned performance.
-This supports the hypothesis that P3 is able to scale its diversity not just
-to the problem class, but to the problem instance or even problem initialization, something
-wholly infeasible for tuned population sizing to do.
-
-This tuning distinction is also apparent when comparing Parameter-less hBOA with hBOA in Figure~\ref{fig-evals-to-success}.
-While generally performing worse than hBOA, the difference between the two algorithms is smallest
-on randomly generated problem classes. On MAX-SAT, Parameter-less hBOA actually outperformed both hBOA and LTGA,
-likely due to its ability to scale diversity to the problem instance instead of the entire problem class.
-
-\subsection{Fast versus Optimal}
-\label{sec-fast-vs-optimal}
-
-In Section~\ref{sec-overtime} we examined intermediate fitness qualities of LTGA and hBOA when using
-population sizes tuned to reach the global optimum. As a result, both were exceptionally ineffective
-at quickly reaching high quality solutions. This is because unlike P3, these methods have an explicit
-trade off between optimal performance and intermediate performance caused by their population size parameter.
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{small-pop-dst}
-      \end{centering}
-      \caption{Deceptive Step Trap 203}
-      \label{fig-small-pop-dst}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{small-pop-nk}
-      \end{centering}
-      \caption{Nearest Neighbor NK 200}
-      \label{fig-small-pop-nk}
-    \end{subfigure}
-  \end{centering}
-  \caption{Comparison of how reducing LTGA's population size effects
-           the median best fitness reached during search.}
-  \label{fig-small-pop}
-\end{figure}
-
-Figure~\ref{fig-small-pop} examines the effect of population size on LTGA's intermediate fitness by
-reducing LTGA's population size to one tenth of the tuned value. The two problems shown are representative of the behavior of using
-a smaller population size on the other five problems. Reducing the population size caused LTGA to improve earlier but plateau
-at lower fitnesses. This caused LTGA's success rate to drop from 100 to 0 on Deceptive Step Trap and from 98 to 68 on Nearest
-Neighbor NK.
-Even when using a reduced population size for LTGA, P3 still achieved a fitness at
-least as high as LTGA at 80\% of intervals.
-The likelihood that P3 would achieve these pairwise results assuming its median result is actually worse is $p < 10^{-15}$
-according to the binomial test.
-
-\begin{comment}
-It is important to note that hBOA and LTGA were tuned specifically
-to find the global optimum and are not necessarily using the optimal population size for finding intermediate fitnesses. Consider
-that for both reducing the population size would almost certainly improve their fitness for evaluation steps less than their population
-size. To examine the effect of population size we tested LTGA using one tenth of the population size required for reliably finding the
-global optimum, with the results given in Figure~\ref{fig-small-pop}.
-Reducing the population size resulted in LTGA's initial fitness improvement
-occurring earlier, as the first generation is completed much more quickly. On Deceptive Step Trap this means that
-LTGA leaps ahead of P3 due to its ability to overcome the two bit fitness plateaus.
-%
-Across all tested problems using the smaller
-population size reduced the quality of LTGA's first fitness plateau, likely due to missing required diversity to reach higher quality.
-The second period of improvement also comes earlier and reaches a lower quality fitness. On Deceptive Step Trap 0
-runs reach the global optimum, down from 100 successful for the full population size. On Nearest Neighbor NK this drops the number
-of successful runs from 98 to 68. As a result we conclude that while reducing the population size of LTGA may improve its quality
-at early points during optimization, doing so reduces the robustness of the final solution found. This is in contrast to P3
-which balances high quality intermediate fitness without trading away eventual optimality.
-\end{comment}
-
-\section{Inner Workings}
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{cross}
-      \end{centering}
-      \caption{Crossover Proportion}
-      \label{fig-cross}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{cross-success}
-      \end{centering}
-      \caption{Crossover Success}
-      \label{fig-cross-success}
-    \end{subfigure}
-  \end{centering}
-  \caption{For each problem Figure~\ref{fig-cross} shows the proportion of P3 evaluations spend on crossovers
-           and Figure~\ref{fig-cross-success} shows the percentage of fitness improving crossover evaluations.}
-\end{figure}
-
-While analysis of optimization speed is useful from a practitioner standpoint, doing so provides very
-little insight into algorithm behavior. To better understand how P3 works in detail we present here
-a look at some internal features specific to P3.
-
-\subsection{Crossover}
-Figure~\ref{fig-cross} shows the proportion of evaluations P3 spends on crossover, as opposed to hill climbing,
-and Figure~\ref{fig-cross-success} shows what percentage of crossover evaluations resulted in a fitness improvement.
-Together these figures provide some insight into the role of crossover within P3. The behavior
-for each is clearly problem dependent and generally asymptoticly stable as problem size increases.
-
-When solving problems where epistasis can be effectively detected and represented by a linkage tree, P3 tends to spend
-less evaluations performing crossover and each crossover is more likely to be successful.
-Deceptive Trap and Rastrigin are the easiest problems to model epistasis, with local search quickly
-reducing pairwise entropy in each. These are also the problems where P3 uses the least evaluations
-on crossover and has the highest success rates for crossover. At the other extreme are Nearest Neighbor
-NK and Ising Spin Glasses, which both have overlapping linkage not representable by a linkage tree.
-These problems have the highest crossover usage and lowest crossover success of any problem except Deceptive
-Step Trap.  While Deceptive Step Trap's epistasis can be accurately modeled by a linkage tree, the
-exponential number of plateaus makes detecting gene linkage very challenging.
-
-\begin{comment}
-P3 performs optimization using a mixture of hill climbing and crossover. Exactly how much time the algorithm
-dedicates to each is not clear from the algorithm description alone. As such, Figure~\ref{fig-cross} shows
-that the relationship between the two is strongly problem dependent but relatively independent of problem size.
-In general, problems which can be more accurately represented by crossover
-required less crossover evaluations. Deceptive Trap and Rastrigin spent the least evaluations on crossover, using
-15\% and 20\% respectively on the largest problem sizes. Both of these problems are perfectly separable and have relatively
-few local optima in each subproblem. HIFF and Ising Spin Glasses use crossover for about 60\% of evaluations. The HIFF structure can
-be perfectly learned by the linkage tree, but the higher order relationships require multiple stages of crossover to optimize.
-Ising Spin Glasses have an overlapping structure, making them impossible to perfectly represent in a single linkage tree, but
-that overlap is relatively minor in comparison with Nearest Neighbor NK, which uses 81\% of evaluations on crossover. Deceptive Step Trap spends the highest
-percentage of evaluations on crossover, 92\%. Unlike all of the other problems, the value taken by an individual bit in Deceptive
-Step Trap has very little meaning. This is because any trap value is either a local optima or 1 bit flip away from being a local optima.
-This has a combined result of meaning it takes very few local search evaluations to find a local optima but very many crossover operations
-to proliferate the global optima. Epistatic relationships are also going to be very weak and even small population sizes are likely
-to mark all clusters as $useful$.
-
-Solutions are only added to the next highest pyramid level if at least one crossover donation resulted in a fitness
-improvement. Figure~\ref{fig-cross-success} examines the percentage of crossover donations that resulted in a fitness
-improvement. The ordering in Figure~\ref{fig-cross-success} is an almost perfect inversion of Figure~\ref{fig-cross},
-implying that the more successful crossover is at achieving fitness improving donations the lower the proportion of evaluations P3 will
-spend performing crossover. P3 achieves an impressive 30\% success rate for crossover donations on the largest Deceptive Trap
-problem sizes. This makes sense as crossover is able to immediately and perfectly identify entire traps such that all donations
-always contain only complete traps. Furthermore, by forcing crossover to attempt donations of genes different from the current solution,
-donations containing a single trap will have a very high success rate, always improving suboptimal traps which are far more frequently
-created by local search that optimal traps. Rastrigin and HIFF have relatively high success rates of 11\% and 7\%, respectively. Again
-this is likely due to accurate modeling and relatively low diversity of building block alternatives.
-Ising Spin Glasses and Deceptive Step Trap both have low success rates near 0.01\%. This follows as both are difficult to model with
-a linkage tree. In comparison to Figure~\ref{fig-cross-success} Deceptive Step Trap performs better than expected for crossover.
-However, the reason for this is that even exceedingly uninformed crossover can make improvements to Deceptive Step Trap as
-flipping two bits in the same trap will move between two local optima. Nearest Neighbor NK had the least successful crossovers,
-with only 0.007\% successful on the largest tested problems. In comparison to Ising Spin Glass, NK's overlap is far more extensive.
-As a result it makes sense that individual crossovers are very unlikely to be fitness improvements.
-\end{comment}
-
-When compared with LTGA, P3's crossover success rates are lower but similar in problem ordering.
-Counterintuitively, the use of hill climbing on the initial population reduces
-P3's crossover success, not because it reduces model quality or the donation pool, but because it is much more challenging to improve
-locally optimal solutions than randomly generated ones. LTGA's crossover benefits from application to unoptimized solutions, which
-makes its aggregate crossover success incomparable to P3's.
-
-Even when crossover success rates are quite low, such as Nearest Neighbor NK's 0.007\% success,
-the results from Section~\ref{sec-optimum} and Section~\ref{sec-overtime} show it's still critical to optimization. Without
-crossover, P3's performance would be identical to the Random Restart Hill Climber, which was unable to solve even moderately sized problems
-and quickly fell behind P3 in intermediate fitness quality. Therefore even infrequently successful crossover donations are
-critical to success. This does, however, suggest a potential avenue for future improvement by using more successful
-modeling and donation algorithms.
-
-\subsection{Pyramid}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{level-size}
-      \end{centering}
-      \caption{Population Size}
-      \label{fig-level-size}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{level-success}
-      \end{centering}
-      \caption{Crossover Success}
-      \label{fig-level-success}
-    \end{subfigure}
-  \end{centering}
-  \caption{For each problem Figure~\ref{fig-level-size} shows the number of solutions stored in each level of the pyramid
-           and Figure~\ref{fig-level-success} shows the percentage of fitness improving crossover evaluations at each level.}
-\end{figure}
-
-Another feature unique to P3 is the shape and size of the population pyramid
-constructed for each problem. Figure~\ref{fig-level-size} shows the number of solutions stored at
-each level of the pyramid for the largest tested problem sizes. Each point is the
-median size across 100 runs. If a run did not store any solutions at a level it is treated as 0.
-No point is drawn if the median run had 0 solutions stored at that level. While pyramid size
-is effected by problem size, the overall shape is not. As such the behavior shown in Figure~\ref{fig-level-size}
-is representative of that for all other tested problem sizes.
-
-With the exception of the dip in Deceptive Step Trap, all of the pyramids show a monotonic reduction
-in size as the level increases. This is because a solution must be a strict fitness improvement
-over its previous version to be added to a higher level, which becomes less likely
-each time the solution improves. \cite{lobo:2011:dynamicpop} found theoretical evidence
-and \cite{goldman:2011:dynamic-parameters} found empirical evidence that the optimal
-population size decreases each generation of traditional evolutionary search.
-By decreasing in size, P3 implicitly stores more diversity in low levels and focuses
-search on high quality solutions at higher levels. In comparison,
-LTGA and hBOA suboptimally use a fixed population size at each generation.
-
-Figure~\ref{fig-level-success} examines how crossover success changes at different
-levels of the pyramid. At low levels, success gets progressively lower
-as solution quality increases faster than the model's ability to improve solutions.
-At higher levels modeling becomes more accurate and donations contain higher frequencies
-of high quality building blocks, resulting in increased crossover success. The
-highest level of most problems has a low crossover success rate, as solutions
-crossing with that level have already been improved by previous operations to
-the point where the only improvement would be to create the global optimum, which can
-only happen once.
-
-\subsubsection{Deceptive Step Trap}
-The number of solutions stored at the fourth level of Deceptive Step Trap is
-significantly lower than that of the third or fifth levels, breaking the decreasing
-trend of the other six problems. Figure~\ref{fig-level-success} has a similar
-aberration, with crossover success dropping to  0.0004\% before rebounding and
-following the more common trajectory. This behavior exists in all other problem
-sizes tested, with the dips occurring at exactly the same level.
-
-This behavior is rooted in the peculiar nature
-of this landscape.
-After local search, all traps in all solutions have a total
-number of 1 bits equal to either 0, 1, 3, 5, or 7 as these correspond to the local
-optima when using $k=7$ and $s=2$. Crossover is very likely to overcome the
-two bit plateaus, and as a result solutions in the second level generally do not
-contain the lowest fitness local optima (5 bits set) and the third
-level has very few traps set to the next worst local optima (3 bits set). As a result,
-solutions which reach the third level can only be improved by replacing the deceptive
-local optima (0 and 1 bits set) with the global optimum (7 bits set).
-The global optimum is very rare in the population, and with 8 ways to represent local optima
-linkage learning is inaccurate. Therefore it is very unlikely for crossover to be successful,
-meaning few solutions will be added to the fourth level. Solutions that do improve by
-definition must have a higher frequency of optimal trap settings, meaning level four's model
-will be more accurate and donations are more likely to contain optimal trap values.
-Thus the level size and crossover success rates increase after contracting around level four.
-
-\section{Conclusions and Future Work}
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing black box
-optimization. P3's primary innovation is the replacement of the generational model with a pyramid of populations.
-This pyramid is constructed iteratively, with both the number of levels and the number of solutions stored
-at each level growing as search progresses. P3 uses a model based crossover method
-which learns a linkage tree from gene epistasis. Combined with a simple hill climber, P3's design contains
-many synergistic features.
-
-Across a large number of problems and problem sizes P3 required less evaluations to reach the global optimum
-than optimally tuned state-of-the-art competitors. On single instance problems P3's improvement was by a
-constant factor, while for the three randomly generated problem classes P3's improvement increased with problem size. This
-quality extends to intermediate points during evolution, with P3 generally reaching at least as high
-of fitness as the competitive techniques when using the same number of evaluations. While P3 does require
-modeling overhead, the expense of this overhead is approximately linear with respect to genome size. There
-is some evidence that even when compared on wall clock time, P3 performs on par with
-the best comparison techniques. All of these achievements are made without any problem specific parameter
-tuning, making P3 easier to apply to new domains than its two closets competitors in quality.
-
-P3's quality is due to a number of desirable traits. First, mixing local search with model based
-crossover lets search focus on properly mixing high quality solutions. Second, by adding diversity only
-as necessary P3 tends to use the minimal amount of random initialization, unlike other techniques which must
-overcompensate with larger population sizes on single instance problems and consider the worst instance
-when solving problem classes. Third, by heavily exploiting existing diversity before adding more P3 is able
-to reach high quality intermediate fitnesses quickly without prematurely converging. Fourth, the very
-nature of the pyramid's shape allows search to preserve a desirable proportion of diversity at
-each fitness level, similar to a generational model using a decreasing population size.
-
-There are a number of meaningful avenues for future P3 experimentation. Perhaps the most
-pressing for practitioner acceptance is to apply P3 to real world problems and compare
-its results with other black box or even problem specific heuristics. While parameter-less,
-P3 is currently limited to discrete, fixed length genomes evaluated using single objective
-fitness. These limitations can be relaxed with future work to make P3 more widely applicable.
-While asymptotically linear in problem size, P3's modeling techniques and local search methods are likely going
-to be prohibitively expensive for genome sizes in the hundreds of thousands or millions of genes,
-and the inability of the model to capture overlapping linkage may be hindering
-search efficiency. Overcoming these limitations by using a new modeling technique may allow
-the pyramid model even greater flexibility. Similarly, while P3 is able to overcome low
-order deception via linkage learning, the iterative improvement method by which crossovers
-are made may mislead search on landscapes with higher order deception.
-
-However, even without
-these improvements our results show P3 is highly efficient at finding
-global optima on black box problems without any problem specific tuning.
-
-\section{Acknowledgements}
-This material is based in part upon work supported by the National Science Foundation
-under Cooperative Agreement No. DBI-0939454. Any opinions, findings, and conclusions
-or recommendations expressed in this material are those of the author(s) and do not
-necessarily reflect the views of the National Science Foundation.
-
-\small
-
-\bibliographystyle{apalike}
-\bibliography{../main}
-
-\normalsize
-\appendix
-\section{hBOA Simplification}
-\label{sec-appendix}
-To measure the quality of a decision forest, hBOA applies Equation~\ref{eq-hboa}.
-To favor compact
-models, Equation~\ref{eq-hboa} is scaled by Equation~\ref{eq-hboa-stop}, which provides increased cost
-for more total leaves.
-This quality is used for two purposes: comparison of potential changes from the existing model
-and comparison of that change with the existing model. The basis for our simplication
-is to rearrange $p(B)BDe(B) < p(B')BDe(B')$ to be $\frac{p(B)}{p(B')}<\frac{BDe(B')}{BDe(B)}$.
-
-\begin{equation}
-  BDe(B) = \prod_{i=0}^{N}\prod_{l\in L_i} \frac{\Gamma(m'_i(l))}{\Gamma(m_i(l) + m'_i(l))}
-  \prod_{x_i}\frac{\Gamma(m_i(x_i, l) + m'_i(x_i,l))}{\Gamma(m'_i(x_i,l))}
-  \label{eq-hboa}
-\end{equation}
-
-\begin{equation}
-  p(B) = c2^{-0.5(\sum_i|L_i|)log_2\mu}
-  \label{eq-hboa-stop}
-\end{equation}
-
-The outermost product of Equation~\ref{eq-hboa} iterates over all trees in the forest. However, each split
-can modify only one of the trees and therefore the contribution of all others can be canceled. The middle
-product is across all leaves in the tree. Again since only one leaf can be changed, all other terms can
-be canceled. By convention hBOA uses uninformed Bayesian priors of $m'_i(l)= 2$ and $m'_i(x_i, l)=1$ for
-binary alphabets. As $\Gamma(a) = (a-1)!$ this means the top term in the middle product and the bottom
-term in the third product reduce to 1. The only remaining terms are then $m_i(l)$ and $m_i(x_i, l)$ which
-represent the number of solutions which reached leaf $l$ and the number of solutions which reached leaf $l$
-with a specific value for $x_i$, respectively.
-
-Equation~\ref{eq-hboa-stop} can also be simplified when doing comparisons. If model $B'$ has exactly one more
-leaf than model $B$ then the ratio $\frac{p(B)}{p(B')}$ simplifies to $2^{0.5 log_2\mu}$ regardless of
-total model size.
-
-The resulting simplifications create Equation~\ref{eq-hboa-final},
-where $B'$ is different from $B$ by exactly 1 split, such that $l$ was split to create $l'$ and $l''$.
-The best split is whichever maximizes its improvement over $B$, which is equal to the right side
-of the inequality. Note that these factorials can still be exceedingly large and
-therefore it is imperative that implementations avoid rounding errors and overflows.
-
-\end{document}

### 2015-04-15 Dissertation/chapter-internal.tex
Major update from journal article polishing

--- 2015-02-19 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
@@
-\documentclass[twoside]{article}
-\usepackage{ecj,palatino,epsfig,latexsym,natbib}
-\usepackage{url}
-\usepackage{verbatim}
-\usepackage{subcaption}
-\usepackage[noend]{algpseudocode}
+\chapter{Internal Workings}
 
-\newcommand{\includegraphicswide}[1]
-{\includegraphics[width=.9\textwidth,height=\textheight,keepaspectratio]{#1}}
-
-\newcommand{\includegraphicsfit}[1]
-{\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}
-
-\usepackage{array}
-\newcolumntype{L}{>{\centering\arraybackslash}m{1.9cm}}
-
-%% do not add any other page- or text-size instruction here
-
-\parskip=0.00in
-
-\begin{document}
-
-% 45-character limit on header
-\ecjHeader{x}{x}{xxx-xxx}{201X}{Parameter-less Population Pyramid}{B. W. Goldman and W. F. Punch}
-\title{\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid}
-
-\author{\name{\bf B. W. Goldman} \hfill \addr{brianwgoldman@acm.org}\\ 
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-\AND
-       \name{\bf W. F. Punch} \hfill \addr{punch@msu.edu}\\
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-}
-
-\maketitle
-
-\begin{abstract}
-%Should be about 200 words
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
-evolutionary optimization without requiring any user specified parameters.
-P3's primary innovation is to replace the generational model with a pyramid of
-multiple populations that are iteratively created and expanded. In combination
-with local search and advanced crossover P3 scales to problem difficulty, exploiting
-previously learned information before adding more diversity.
-
-Across seven problems, each tested using on average 18 problem sizes, P3 outperformed
-all five advanced comparison algorithms. This improvement includes requiring less evaluations
-to find the global optimum and better fitness when using
-the same number of evaluations. Using both algorithm analysis and comparison we find P3's
-effectiveness is due to its ability to properly maintain, add, and exploit diversity.
-
-Unlike the best comparison algorithms, P3 was able to achieve this quality without any
-problem specific tuning. Thus, unlike previous parameter-less methods, P3 does not
-sacrifice quality for applicability. Therefore we conclude that
-P3 is an efficient, general, parameter-less approach to black box
-optimization which is more effective than existing state-of-the-art techniques.
-
-\end{abstract}
-
-\begin{keywords}
-
-Genetic algorithms, 
-linkage learning,
-local search,
-parameter-less.
-
-\end{keywords}
-
-\section{Introduction}
-A primary purpose of evolutionary optimization is to efficiently find good solutions
-to challenging real world problems with minimal prior knowledge about the problem itself.
-This driving goal has created search algorithms which can escape user bias to create
-truly novel results, sometimes publishable or patentable in their own right~\citep{kannappan:2014:humies}.
-While it is not possible for any algorithm to do better than random search across all possible
-problems~\citep{Wolpert:1997:nfl}, effectiveness can be achieved by assuming the search
-landscape has structure and then biasing the algorithm toward exploiting that structure.
-
-In evolutionary optimization, and genetic algorithms~(GAs) in particular, search is often
-biased through parameters. This can be beneficial as it allows practitioners to inject their
-knowledge about the shape of the search landscape into the algorithm.
-However, the quality of solutions found, and the speed at which they are found, is strongly tied to setting these parameters
-correctly~\citep{goldberg:1991:gasize}. As such, either expert knowledge or exceedingly
-expensive parameter tuning~\citep{grefenstette:1986:optimalga} are required to leverage
-this feature to its fullest potential. Furthermore,  parameters such as population size, mutation rate, crossover
-rate, tournament size, etc. usually have no clear relationship to the problem being solved, meaning even
-domain experts may not understand how the parameters will interact with the problem or with each other.
-To further complicate matters, there is mounting evidence that parameter values should change
-during search~\citep{goldman:2011:dynamic-parameters,laporte:2014:adaptivepop}.
-
-There have been periodic efforts to reduce or remove the need for parameter tuning.
-\cite{rechenberg:1973:es} introduced self-adaptive parameters, in which parameter values
-were included in each solution's genome and themselves underwent evolution. This allowed search
-to optimize some of its own parameters, resulting in a reduced need for expert tuning.
-\cite{harik:1999:parameterlessga} were able to design an entirely parameter-less GA by
-leveraging schema theory and parallel populations. Unfortunately these methods were provably less efficient
-than directly setting the parameters to optimal values~\citep{pelikan:1999:worstparameter-less}.
-
-One area that has been very effective at reducing the number of algorithm parameters is
-model based search. \cite{pelikan:2006:hboa}'s Hierarchical Bayesian Optimization
-Algorithm~(hBOA) and \cite{thierens:2010:ltga}'s Linkage Tree Genetic Algorithm~(LTGA)
-both require only a single parameter: population size. \cite{posik:2011:parameterless}
-leveraged model building to create a fully parameter-less algorithm, but it is restricted to
-only order-k, fully decomposable, noiseless problems.
-
-Most recently \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid~(P3).
-This method uses a pyramid structure of populations to combine model based search with local search
-to achieve parameter-less optimization. Initial results suggest that unlike
-previous parameter-less methods, P3 is actually more efficient than current state-of-the-art
-parameterized search algorithms. In this work we shall: extend these results to cover more
-comparison algorithms; compare both efficiency in reaching the global optimum and intermediate
-fitnesses; analyze algorithm complexity; and provide more in depth analysis of P3 itself.
-
-\section{Comparison Optimizers}
-\label{sec-optimizers}
-
-In order to fully understand the effectiveness of P3, we compare it with five advanced algorithms which
-have related features to P3. The Random Restart Hill Climber defined by ~\cite{goldman:2014:p3} was chosen as an efficient form of repeated
-local search. As P3 combines this hill climber with crossover,
-comparing with local search alone shows the advantages of P3's overall approach.
-The $1+(\lambda, \lambda)$ algorithm~\citep{doerr:2013:lambdalambda} is the current best theory supported simple genetic algorithm
-and its method of crossover is in some sense a macro-mutation just as in P3. hBOA and Parameter-less
-hBOA are advanced model building search techniques which are very effective at learning complex
-problem structure, designed to achieve similar goals as P3's linkage learning but using very different
-methods. Finally LTGA represents the current state-of-the-art in black box search and is the origin
-of P3's linkage learning and crossover methods.
-
-Only hBOA and LTGA require any parameters, with each of these only requiring a population
-size. This makes knowing the optimal behavior of these algorithms much more tractable. All of the
-algorithms are also gene order independent, fitness scale invariant, and unbiased. This means,
-for any problem, the order in which problem variables appear in the genome can be changed
-without changing the behavior of the search. The fitness can also be manipulated in any fashion
-as long as the rank ordering of solutions is unchanged. These algorithms are also unaffected by the meaning
-assigned to each bit, such that inverting a predetermined random subset of genes before evaluation
-will not impact search efficiency.
-
-Our implementations of all of these algorithms as well as all of the population size information, raw results,
-and processing scripts
-are available from our website.\footnote{\url{https://github.com/brianwgoldman/FastEfficientP3}}
-\subsection{Random Restart Hill Climber}
-\label{sec-hill-climber}
-Perhaps the simplest black box search heuristic is stochastic local search, or hill climbing.
-This optimization technique focuses on improving a single solution until it reaches a local
-optimum. Here we use the first improvement hill climber defined by \cite{goldman:2014:p3}
-and given in Figure~\ref{fig-hc}. This algorithm works by flipping each bit in a random
-order, keeping modifications when fitness is improved, until single bit flips cannot result in
-further fitness improvements.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Hill-Climber}{}
-    \State $options \leftarrow [0 \dots N-1]$\label{fig-hc-options}
-    \State $tried \leftarrow \emptyset$
-    \While{$|tried| < |options|$}
-      \ForAll{$index \in shuffled(options)$}
-        \If{$index \notin tried$}\label{fig-hc-tried}
-          \State Flip bit $index$ in solution
-          \If{solution's fitness increased}
-            \State $tried \leftarrow \emptyset$
-          \Else
-            \State Revert change
-          \EndIf
-          \State $tried \leftarrow tried \cup \{index\}$
-        \EndIf
-      \EndFor
-    \EndWhile
-  \EndProcedure
-\end{algorithmic}
-  \caption{Hill climbing algorithm used to improve randomly generated solutions until no single
-           bit change results in a fitness improvement.}
-  \label{fig-hc}
-\end{figure}
-
-The hill climber requires an amortized cost of $O(1)$ operations per evaluation. In order to
-terminate, at least one evaluation must be performed for each of the $N$ bits in the solution.
-As such any operation that happens only once per search can be amortized over at least $N$
-evaluations, covering the initialization of $options$ on Line~\ref{fig-hc-options}.
-Line~\ref{fig-hc-tried}, which prevents wasted evaluations, can be called at most twice
-per evaluation: once to add $index$ into $tried$ and once to prevent $index$ from being unnecessarily
-evaluated again. The only way three or more calls could happen is if
-no fitness improvement was made for the entire previous iteration, which contradicts the loop invariant.
-
-Due to its nature, this hill climber cannot escape basins of attraction. Once a solution is reached
-such that none of the single bit neighbors are fitness improvements, search stops. Thus
-this algorithm requires a restart mechanism to solve multimodal problems. We have
-chosen here to na\"{i}vely restart search from a random solution whenever a local optima is found. This
-ensures that on all landscapes there is always a non-zero probability of search finding the global optimum.
-
-\subsection{$1+(\lambda, \lambda)$}
-\cite{doerr:2013:lambdalambda} presented the first genetic algorithm to provably show
-the advantages of performing crossover on One Max. This comparatively simple
-algorithm maintains only a single individual and a self-controlled parameter $\lambda$.
-
-Each iteration, the number of bits to flip is chosen from the binomial distribution $b\sim B(N, \frac{\lambda}{N})$,
-where $N$ is the number of bits in the genome.
-Next, $\lfloor\lambda\rfloor$ offspring are produced by flipping $b$ bits. The
-best mutant then produces $\lfloor\lambda\rfloor$ offspring via uniform crossover with the original parent, such that each gene comes from the
-mutant with probability $\frac{1}{\lambda}$. In the original algorithm the best
-offspring produced by crossover then replaces the original parent if its fitness is no worse.
-The $\lambda$ parameter, which is initialized to 1, is decreased if the offspring replaced
-its parent and increased otherwise.
-
-The original formulation was designed specifically for unimodal landscapes and as such were
-not directly suitable for multimodal problems. \cite{goldman:2014:p3} extended $1+(\lambda, \lambda)$
-to include random restarts. As search stagnates, the $\lambda$ parameter increases in value. Eventually
-this results in $\lambda \ge N$ causing mutation to always flip all bits of the individual.
-As this prevents any future improvement, whenever $\lambda \ge N$ search is restarted from a random solution with $\lambda$
-reset to 1.
-
-A few other efficiency modifications were also made. If there is a tie in crossover offspring fitness,
-whichever has a larger hamming distance from the parent is retained. This encourages drifting across plateaus.
-The ``mod'' control strategy proposed by \cite{doerr:2013:lambdalambda} was not used as it conflicted with
-the random restart strategy.
-If a crossover individual is identical to either of its parents, it is not evaluated.
-If mutation produces an offspring which is better than the best crossover offspring, it is used to compare
-against the original parent.
-
-\subsection{Hierarchical Bayesian Optimization Algorithm}
-
-\cite{pelikan:2006:hboa} used statistical principles in combination with a decision tree structure
-to create the Hierarchical Bayesian Optimization Algorithm (hBOA). This method creates a model of
-epistatic relationships between genes which is then used to stochastically generate new solutions.
-Each generation a binary tournament with replacement is used to select $\mu$ solutions from
-the population. These solutions are then used to build the model, which in turn is used to generate $\mu$ new
-solutions. The new solutions are then integrated into the population using restricted tournament
-replacement.
-
-Conceptually, the model built by hBOA is trying to infer rules of the form ``Given that this
-subset of genes are set to these values, how frequently is gene $x_i$ set to value v?'' This can
-be represented using a directed acyclic decision forest, with each tree in the forest representing one gene
-in the solution. In the decision tree $T_i$, which is used to set the value of gene $x_i$,
-each internal node represents previous decisions on how to set
-some other gene $x_j$, with the children of that node representing how the decision was made. The
-leaves of each tree give the probability that $x_i$ should be set to each possible value.
-
-The forest is constructed iteratively, with each tree initially containing a single leaf
-and with each leaf storing a pointer for each selected solution. Each iteration the algorithm considers
-all possible ways of splitting an existing leaf using another gene $x_j$, such that solutions in the
-leaf are moved to the newly created leaves based on their value for $x_j$. The general goal is to
-separate the solutions such that all solutions with $x_i = 0$ move to one leaf while solutions with
-$x_j = 1$ move to the other.
-
-This goal is formalized using model scoring from Bayesian statistics. In its raw form this
-almost always creates near infinitesimal results, calculating fractions which include factorials of $\mu$
-and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{sec-appendix}, we
-derived a simplified form shown in Equation~\ref{eq-hboa-final}.
-Here $l$ is a leaf in tree $i$, with $l'$ and $l''$ the results of splitting $l$. $m_i(l)$ is the number
-of solutions which reach $l$ and $m_i(x_i, l)$ is the number of solutions which reach $l$ with the given value
-for $x_i$.
-If no proposed split satisfies the inequality, iteration stops.
-If multiple splits do, whichever maximizes the right side is chosen.
-
-\begin{equation}
-  2^{0.5 log_2\mu} < \frac{(m_i(l) + 1)!}{m_i(0, l)!m_i(1,l)!} \cdot
-  \frac{m_i(0, l')!m_i(1,l')!m_i(0, l'')!m_i(1,l'')!}{(m_i(l') + 1)!(m_i(l'') + 1)!}
-  \label{eq-hboa-final}
-\end{equation}
-
-Initially there are $\Theta(N^2)$ possible ways to split existing leaves, as each of the $N$ single node
-trees can be split by any of the other $N-1$ genes. Each iteration a new edge is added to the decision
-forest, meaning some of the previously tested splits cannot be used. For instance, if $T_i$, which is used
-to decide the value of $x_i$, is split using the value of $x_j$, $T_j$ can no longer be split using $x_i$.
-As a split creates two new leaves, $O(N)$ new potential splits must also be tested. Equation~\ref{eq-hboa-final}
-parses all solutions which reach a leaf to count gene frequencies, requiring $\mu$ time.
-The number of total leaves created depends heavily on the problem and $\mu$.
-However, assuming no splits are accepted or that the cost of testing all future splits is less than
-the initial $\Theta(N^2)$, constructing the model requires $\Omega(\mu N^2)$ time. Each
-model is used to generate $\mu$ solutions, leading to a cost per evaluation of $\Omega(N^2)$.
-
-To generate a solution, the value of each gene $x_i$ is set using its corresponding decision tree $T_i$. Because
-the forest is directed acyclic, there must be an ordering of $T_i$ such that before $T_i$ is executed all
-$x_j$ it uses to make decisions have already been set. As such, previous decisions made by other trees
-are used to follow each $T_i$ until a leaf is reached. The value of $x_i$ is then set based on the
-probability that other solutions reached that leaf with each value of $x_i$.
-
-To perform replacement, hBOA uses restricted tournament replacement. After each new solution is generated
-and evaluated, a set of $w$ solutions are chosen at random from the population, where $w=\min\{N, \frac{\mu}{20}\}$.
-From this set the solution which is most genetically similar to the offspring is chosen. If the offspring
-is at least as fit as the chosen solution, it replaces the chosen solution in the population. Otherwise the
-offspring is discarded. This method is designed to preserve genetic diversity as only genetically similar
-solutions must compete on fitness.
-
-hBOA is designed to work with large population sizes, resulting in a large number of
-evaluations per generation. As hBOA utilizes explicit diversity maintenance, standard methods for determining
-convergence are not considered very accurate. Therefore the authors suggest that an hBOA run should be
-terminated after performing generations equal to $N$.
-
-Like other model based techniques, hBOA has very few parameters. There is no mutation or crossover,
-and modeling does not rely on any explicit parameters. Solution selection, generation, and replacement
-are all derived from the population size, which must be set by the user.
-
-\subsection{Parameter-less hBOA}
-Using the methods first introduced by \cite{harik:1999:parameterlessga} for the Parameter-less GA,
-\cite{pelikan:2004:parameterlesshboa} created Parameter-less hBOA which automatically scales its
-population size to fit the problem. This is done by maintaining a list of concurrent populations
-using exponentially scaled population sizes.
-
-A run of Parameter-less hBOA starts with a single population of size $\mu_0$, conventionally set
-to $\mu_0=10$. After two generations are performed, a new population of size $\mu_1 = 2\mu_0$ is created
-and performs a generation. Evolution then continues with the $\mu_0$ population performing two generations
-for each one performed by $\mu_1$. Each time population $\mu_i$ performs its second generation a new population
-$\mu_{i+1}=2\mu_i$ is created, which performs generations half as often as $\mu_i$. In this way an infinite number of
-parallel population can be simulated, with each population receiving the same number of total evaluations.
-
-In all other aspects each population is identical to an hBOA population using a fixed $\mu$. No search information
-is shared among populations, and each search is independently terminated. As such Parameter-less hBOA cannot
-perform better than hBOA using the optimal population size for a given instance, as it must also spend evaluations
-on populations of different sizes. This inefficiency is bounded by a log multiple of the total number of
-evaluations~\citep{pelikan:1999:worstparameter-less}.
-
-\subsection{Linkage Tree Genetic Algorithm}
-\cite{thierens:2010:ltga} introduced the Linkage Tree Genetic Algorithm (LTGA) which automatically
-detects and exploits problem epistasis by examining pairwise gene entropy. Due to its enhanced
-ability to preserve high fitness gene subsets, LTGA was able to outperform state-of-the-art
-GAs across many benchmarks. Since its introduction, many variants of LTGA have been
-proposed~\citep{thierens:2011:gomea,goldman:2012:ltga} so for clarity we have chosen the version
-presented by \cite{thierens:2013:ltgahiff} as our model.
-
-LTGA's effectiveness comes from its method of performing crossover. Instead of blindly
-mixing genes between parents, LTGA attempts to preserve important interrelationships
-between genes. Before performing any crossovers in a generation, LTGA first builds
-a set of hierarchical gene clusters which are then used to dictate how genes are mixed
-during crossover.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Cluster-Creation}{}
-    \State $unmerged \leftarrow \{\{0\}, \{1\}, \{2\}, \dots, \{N-1\}\}$
-    \State $useful \leftarrow unmerged$
-    \While{$|unmerged|>1$}
-      \State $C_i, C_j \leftarrow \min_{C_i,C_j \in unmerged} D(C_i, C_j)$
-      \State $unmerged \leftarrow unmerged - \{C_i, C_j\} + \{C_i \cup C_j\}$
-      \State $useful \leftarrow useful + \{C_i \cup C_j\}$
-      \If{$D(C_i, C_j) = 0$}
-        \State $useful \leftarrow useful - \{C_i, C_j\}$
-      \EndIf
-    \EndWhile
-    \State Order $useful$ based based on last merged first\label{fig-cluster-creation-ordering}
-    \State Remove largest cluster from $useful$
-
-    \Return $useful$
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm describing how LTGA creates clusters using Equation~\ref{eq-distance}
-           for a population. $unmerged$ and $useful$ are ordered sets of sets of gene loci.}
-  \label{fig-cluster-creation}
-\end{figure}
-
-Figure~\ref{fig-cluster-creation} provides the agglomerative method LTGA uses to create gene clusters.
-This algorithm creates a tree of sets using pairwise gene entropy, such that the leaves of the tree contain a single gene and
-internal nodes are the union of their children's sets. These sets are then used by crossover to specify epistatic
-relationships which should be preserved.
-The process begins by creating the set of sets $unmerged$ which tracks all top level clusters. Initially
-$unmerged$ contains single member sets for each gene. After each iteration the two sets with the minimum average pairwise
-distance (given in Equation~\ref{eq-distance}) are merged to create a single cluster. This process is repeated
-until only a single set remains in $unmerged$ which contains all of the genes in the genome.
-
-\begin{equation}
-  D(C_i,C_j) = \frac{1}{\left | C_i \right |\cdot \left |C_j \right|}\sum_{c_i \in C_i}\sum_{c_j \in C_j}
-  2 - \frac{H(c_i) + H(c_j)}{H(c_i \cup c_j)}
-  \label{eq-distance}
-\end{equation}
-\begin{equation}
-  H(c) = -\sum_{s\in S} p_c(s)\log(p_c(s))
-  \label{eq-entropy}
-\end{equation}
-
-Throughout this process $useful$ tracks the set of all gene clusters which should be preserved for use by crossover.
-This set begins with all genes in separate clusters, and each time a new cluster is created it is added to $useful$.
-However, not all clusters are necessarily worth keeping. For instance, in all versions of LTGA the cluster
-containing all genes is removed from $useful$ as preserving all genes during crossover can only create clones.
-\cite{thierens:2013:ltgahiff} extended this removal to include any unsupported subsets. If the pairwise distance
-between two clusters is 0, this means there are no individuals in the population which disrupt the relationships between the two
-clusters. Therefore when performing crossover there is no reason to believe a fitness improvement can be achieved
-by breaking the stored pattern. As such a cluster is only kept if its direct superset has a non-zero distance.
-As a final step Line~\ref{fig-cluster-creation-ordering} reorders $useful$ such that clusters appear in reversed
-order from which they were added to $useful$. Thus the most linked clusters and those containing single genes
-appear at the end of the returned list.
-
-\cite{thierens:2013:ltgahiff}'s version of LTGA does not use the entire population when determining pairwise entropy.
-Instead, binary tournament is used to select half of the population. This is done to ensure the model is built
-using only high quality solutions, even during the first generation.
-
-In order to efficiently perform clustering, a pairwise gene frequency table is constructed from the selected solutions.
-To calculate Equation~\ref{eq-distance}, Equation~\ref{eq-entropy} is called for each
-gene ($H(c_i)$) and pair of genes ($H(c_i \cup c_j)$). Extracting this information requires $O(\mu N^2)$
-time, where $\mu$ is the population size and $N$ is the genome size. The process of converting
-this pairwise frequency information into clusters can be achieved in $O(N^2)$ using the bookkeeping
-methods presented by \cite{gronau:2007:upgma}. This cost is performed only once per generation,
-and is then used to perform approximately $O(\mu N)$ crossover evaluations. As a result, the amortized cost of
-LTGA's model building is $O(N)$.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Cluster-Usage}{}
-    \ForAll{$C_i \in useful$}
-      %\ForAll{$d \in shuffled(P_i)$}
-        \State $d \leftarrow rand\_choice(P)$\label{fig-cluster-usage-donate}
-        \State Copy $d$'s gene values for $C_i$ into solution
-        \If{solution was changed}
-          \If{solution's fitness decreased}
-            \State Revert changes
-          \EndIf
-          %\State \textbf{break}
-        \EndIf
-      %\EndFor
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm describing how clusters are used to perform crossover.}
-  \label{fig-cluster-usage}
-\end{figure}
-
-Figure~\ref{fig-cluster-usage} describes how the identified clusters are used by crossover to preserve
-gene linkage while still exploring the search space. Unlike more traditional crossover methods, LTGA
-crosses each individual with the entire population. Also, to produce a single offspring, multiple evaluations
-of the fitness function are performed.
-
-Each generation, each individual in the population undergoes crossover. In a single crossover event, each
-cluster of genes $C_i$ in $useful$ is applied as a crossover mask. A random donor $d$
-is chosen from the entire population (not just the model selected population), and $d$'s gene's for $C_i$ are copied
-into the working solution. If a modification is made, an evaluation is then performed. If the crossover
-resulted in no worse fitness then the changes are kept. This allows for neutral drift across plateaus.
-The resulting solution, which must be at least as fit as its parent, is then
-copied into the next generation.
-
-In total each individual can cause up to $|useful|$ evaluations. If all clusters were kept, even those deemed
-unhelpful, and all donations were evaluated, even those which did not change any genes, then \Call{Cluster-Usage}{}
-would perform exactly $2N-2$ evaluations for each of the $\mu$ solutions in the population. This provides the amortizing evaluations
-required to make clustering only $O(N)$ operations per evaluation. However, by skipping some evaluations, its
-possible that clustering may be super-linear.
-
-LTGA has no explicit form of diversity control and has no method for introducing new genetic information once
-the population has converged. Therefore an LTGA run is considered converged once two consecutive populations
-contain the same unique solutions.
-
-By design, LTGA only has a single parameter: population size. LTGA uses no mutation, and crossover is defined
-in terms of the clustering algorithm. Selection between generations is fully elitist and embedded in the crossover,
-with selection of model building solutions fixed to a binary tournament. Neither \Call{Cluster-Creation}{} nor
-\Call{Cluster-Usage}{} rely on parameter values. LTGA does not provide any method for controlling or setting
-the population size, relying instead on a fixed user specified size.
-
-\section{Parameter-less Population Pyramid}
-\label{sec-p3}
-\cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid (P3) as a method for
-performing optimization which does not require the user to provide any parameters. This is
-achieved by combining efficient local search with the model building methods of LTGA using
-an iteratively constructed hierarchy of populations.
-
-The high level algorithm of P3 is presented in Figure~\ref{fig-p3}. Unlike more traditional
-GAs, P3 does not follow a generational model. Instead, it maintains an iteratively
-expanding pyramid of expanding populations. Each iteration, a new random solution is generated.
-This solution is brought to a local optimum using the hill climbing algorithm in Figure~\ref{fig-hc}. If that
-local optimum has not yet been added to the pyramid, the solution is added to the lowest
-population $P_0$.
-
-Next, the solution is iteratively improved by applying LTGA's crossover algorithm (Figure~\ref{fig-cluster-usage})
-with each population $P_i$ in the pyramid. If this process results in a strict fitness improvement and has
-created a solution not yet stored in the pyramid, it is added to the next highest pyramid level $P_{i+1}$.
-If $P_{i+1}$ does not yet exist, it is created. In this way populations in the pyramid expand over time,
-and the number of populations stored increases over time. Initially the pyramid contains no solutions
-or populations, meaning the user does not need to specify a population size.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Iterate-P3}{}
-    \State Create random solution
-    \State Apply hill climber
-    \If{solution $\notin hashset$}
-      \State Add solution to $P_0$ and rebuild model
-      \State Add solution to $hashset$
-    \EndIf
-
-    \ForAll{$P_i \in pyramid$}
-      \State Mix solution with $P_i$
-      \If{solution's fitness has improved}
-        \If{solution $\notin hashset$}
-          \State Add solution to $P_{i+1}$ and rebuild model
-          \State Add solution to $hashset$
-        \EndIf
-      \EndIf
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-  \caption{One iteration of P3 optimization. $pyramid$ is an
-           ordered set of populations and $hashset$ is a set
-           of all solutions in $pyramid$.}
-  \label{fig-p3}
-\end{figure}
-
-To accommodate P3's unique population structure, some of LTGA's clustering procedures were modified. In LTGA,
-clusters are identified at the start of each generation and are used
-to create all offspring in that generation. As P3 does not perform serial
-generations, P3 instead rebuilds the model each
-time a solution is added to a population. Furthermore, unlike
-our chosen variant of LTGA, all solutions in the population are used to generate the model, not just the
-winners of a binary tournament. We can do this because even the
-worst solutions in the pyramid are already high quality due to local search.
-Using local search in LTGA was examined by \cite{bosman:2011:lsbbo} and found
-to provide no significant improvement. A likely cause was that that study applied
-local search to every solution, not just the initial population, resulting in significant overhead.
-
-Beyond the changes in population structuring, P3 modifies
-LTGA's version of \Call{Cluster-Creation}{} and \Call{Cluster-Usage}{}.
-P3 changes Line~\ref{fig-cluster-creation-ordering} in Figure~\ref{fig-cluster-creation}
-from \emph{last merged first} ordering to \emph{smallest first} ordering.
-This method applies gene clusters during crossover based on how many genes
-are in each cluster,\footnote{Ties are broken randomly.} and not on how tightly linked those genes are.
-\cite{goldman:2012:ltga} found that this alternative was better at preserving
-diversity, and therefore required smaller populations.
-
-P3 also modified Line~\ref{fig-cluster-usage-donate} in Figure~\ref{fig-cluster-usage}.
-Instead of choosing a single genetic donor for each cluster, P3 iterates over the
-population in a random order until a solution in the population is found which
-has a least one gene different for that cluster of genes from the improving solution.
-This process increases the likelihood of an evaluation being performed for every cluster,
-and helps test rare gene patterns in the population.
-
-In LTGA the cost of rebuilding the model is $O(\mu N^2)$ as it must collect pairwise
-gene frequency information for all $\mu$ solutions in the population. P3 does not store
-a single population, and it does not have a fixed $\mu$ size for any population. However,
-each time a solution is added to the population, it requires $O(N^2)$ time to update
-the table of pairwise frequencies and another $O(N^2)$ time to rebuild the linkage model.
-The model is then used immediately to perform up to one evaluation for each of the up
-to $2N-2$ clusters. Just as in LTGA, if no evaluation shortcuts were made, P3
-has an amortized cost of $O(N)$ modeling cost per evaluation. While P3 does
-rebuild the model more frequently per solution in the population, it also performs
-a number of local search evaluations which are quite efficient, meaning theoretical
-comparisons of their speed are difficult to perform. As a final note, P3's repeated
-attempts to find a useful donation make it less likely than LTGA to skip evaluations,
-but has an added cost to find these donations. Repeated donations could require as much
-as $O(\mu)$ attempts per evaluation, but experimental evidence suggest that this
-operation actually saves more overhead than it costs by increasing the number of evaluations
-per model rebuild.
-
-Each of the pieces of the P3 algorithm were selected not just for their stand alone efficacy,
-but for the ways in which they interact. By using the hill climber to ensure each solution
-is at a local optima, the underlying pairwise relationships in the problem are exposed. As
-a result, detecting clusters for use by crossover is much more effective. The crossover operator
-is extremely elitist, as each gene donation must result in no fitness loss, and a solution must
-strictly improve to be added to the next level of the pyramid. This is balanced by continual
-integration of new random solutions. Furthermore, each random restart decreases the probability
-of spurious linkages caused by shared ancestry. This diversity is further preserved by applying
-gene clusters in smallest first order during crossover as this reduces the probability of genetic
-hitchhikers.
-
-Other algorithms have proposed using multiple concurrent populations.
-\cite{hornby:2006:alps} had a hierarchy of populations with solutions periodically advancing
-upward. This allows for continuous integration of diversity as the lowest population is reseeded
-with random solutions. However, this method resulted in increased parameterization as not only was
-a population size required, but also new parameters for how frequently generations advanced between levels and
-how many total levels to have. \cite{harik:1999:parameterlessga} used multiple independent
-populations of different sizes as a method for removing the population size parameter, but
-doing was provably less efficient than using an optimal population size as no information is shared
-between the populations.
-
-\section{Problem Descriptions}
-\label{sec-problems}
-
-\subsection{Single Instance Problems}
-Understanding how a stochastic search algorithm will behave on arbitrary and complex
-search landscapes can be exceedingly difficult. Therefore a common practice for
-algorithm understanding is to perform search on well defined, well understood
-landscapes. To be of interest these landscapes need to represent interesting
-and important aspects of real world problems.
-
-Perhaps the most common such landscape is the Deceptive Trap problem \citep{goldberg:1991:gasize}.
-In this landscape the genome is broken up into $k$ bit non-overlapping subproblems referred
-to as traps.
-Each subproblem is scored using Equation~\ref{eq-deceptive-trap}, where $t$ is the number of
-bits in the trap set to 1. The global optimum in each trap is a string of all 1s, while all
-other solutions lead to a local optima of all 0s. This problem tests an algorithm's
-ability to overcome $k$ sized deception and is commonly used to determine how effective crossover
-is at preserving building blocks. Any crossover event that mixes bits from different parents in
-the same trap will likely result in that trap being optimized to the local optima. For our experiments
-we chose $k=7$ to ensure highly deceptive traps.
-
-\begin{equation}
-   trap(t) = \left\{
-     \begin{array}{rl}
-       k-1-t, &  t<k\\
-       k,   &  t = k
-     \end{array}
-   \right.
-  \label{eq-deceptive-trap}
-\end{equation}
-
-\cite{goldman:2012:ltga} found that mixing local search with linkage learning rendered the Deceptive
-Trap problem trivial. This is because local search is able to optimize each trap to one of the two
-optima (all 1s or all 0s), and then linkage learning can have perfect knowledge of gene interactions.
-In order to make the problem more difficult, the authors proposed the Deceptive
-Step Trap problem, given in Equation~\ref{eq-deceptive-step-trap}. This function modifies the results
-of Equation~\ref{eq-deceptive-trap} to include plateaus of size $s$, introducing an exponential number
-of local optima in each trap.
-With $k=7$ and $s=2$, as used in our experiments, all traps with 0, 1, 3, 5, and 7 bits set are local optima.
-This means that half of all ways to set the trap are 1 bit local optima.
-More generally, the number of local optima grows at $\Theta(2^{k-1})$.
-As a result, the Deceptive Step Trap is much more challenging for linkage
-learning techniques, while still being highly deceptive.
-\begin{equation}
-   step\_trap(t) = \left \lfloor \frac{(k-s)\pmod{s} + trap(t)}{s} \right \rfloor
-  \label{eq-deceptive-step-trap}
-\end{equation}
-
-Another challenging aspect of landscapes can be higher order relationships. The Hierarchical If
-and only If (HIFF) problem~\citep{watson:1998:hiff} is designed to capture the difficulties of this class of problem.
-In HIFF the genome is broken up into a complete binary tree, such that each gene appears in exactly one
-leaf and each internal node is the subset of genes contained in its children. If all genes represented
-in a node of the tree are set to the same value, they score equal to the size of the set. In this way
-small subsets lead toward solutions to larger subsets. However, a node can score if all genes are either
-all 1s or all 0s, meaning that to solve higher order subproblems it is necessary to perform crossovers
-that preserve lower order solutions. This problem is a natural fit for LTGA as the linkage tree can
-perfectly duplicate the problem's true relationships \citep{thierens:2013:ltgahiff}.
-
-As a final class of well known problems, we have chosen to borrow the Rastrigin problem from real valued
-optimization. This problem's landscape, determined by Equation~\ref{eq-rast}, is highly multimodal caused by
-the oscillating cosine function. \cite{goldman:2014:p3} proposed the Discretized Rastrigin problem, such
-that each floating point $x_i$ in Equation~\ref{eq-rast} is encoded using a 10 bit gray code.
-\begin{equation}
-  10n + \sum_{i=1}^{n}\left [ x_i^2-10\cos (2\pi x_i) \right ] \forall x\in [-5.12,5.12]
-  \label{eq-rast}
-\end{equation}
-
-\subsection{Randomly Generated Problem Classes}
-While well defined landscapes can provide specific insights into how an algorithm works,
-their static nature can be misleading. Specifically, algorithm quality might
-be very fragile such that it is only effective at searching well behaved landscapes. A more realistic
-test of an algorithm's black box effectiveness is to work with randomly generated instances
-drawn from a problem class. When tested over a sufficiently large sample it is then possible
-to draw more general conclusions about the algorithms effectiveness. The challenge with these
-landscapes is determining the global optimum to gauge if an algorithm was successful.
-
-Perhaps the most common model for generating random rugged landscapes is the NK model. An
-NK Landscape determines the fitness of each gene based on epistatic relationships with $K$
-other genes in the genome. This fitness is specified using a randomly generated table of
-fitness values, were each possible combination of the $K+1$ genes is mapped to some floating
-point value $[0-1]$. In unrestricted NK landscapes the relationships between genes are also
-randomly chosen and as a result finding the global optimum is $NP$-Hard for $K>1$. However,
-if epistasis is set such that each gene depends on the $K$ directly following it
-in the genome, the solution can be found in polynomial time~\citep{wright:2000:solvingnk}.
-These Nearest Neighbor NK landscapes are therefore ideal for search algorithm testing.
-For all of our experiments we fixed $K=5$ to ensure highly rugged landscapes.
-
-\cite{saul:1994:spinglass} presents a combinatorial benchmark problem derived from physics:
-Ising Spin Glasses. A spin glass is defined by a weighted graph of interaction terms between vertices.
-Each gene assigns a value to each vertex, with the fitness calculated by Equation~\ref{eq-ising}.
-In this equation, $E$ is the set of all edges, $e_{ij}$ is the edge weight connecting vertex $i$ to vertex $j$, and
-$x_i$ and $x_j$ are the gene values for vertex $i$ and $j$. Optimal fitness is when this sum is minimized.
-\begin{equation}
-\sum_{e_{ij} \in E} x_ie_{ij}x_j
-  \label{eq-ising}
-\end{equation}
-Similar to NK Landscapes, the general class is $NP$-Hard to optimize, but the $2D\pm J$ subset of
-Ising Spin Glasses can be polynomially solved.\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}
-In this subset the graph is restricted to be a two dimensional torus, edge weights are randomly set
-to either -1 or 1, and vertex values must be -1 or 1.
-
-As our final class of randomly generated problems we chose the Maximum Satisfiability (MAX-SAT) problem.
-Related to the more common 3-SAT problem, a MAX-SAT instance is defined by a set of three term clauses.
-Each term is a randomly chosen variable, which may also be negated. A clause scores if an and only if
-at least one term in the clause evaluates to true. In order to make MAX-SAT instances with a known global
-optimum, \cite{goldman:2014:p3} proposed constructing clauses around a fixed solution. In this way the
-signs of the terms are set to ensure the target solution satisfies the clause. To ensure each
-problem is challenging we chose a clause to variable ratio of $4.27$~\citep{selman:1996:maxsat}.
-
-\section{Comparison Algorithm Parameter Tuning}
-\label{sec-tuning}
-While four of the six algorithms in our experiments do not require any user specified parameters,
-hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
-handicapped, we extensively tuned each using the bisection method \citep{sastry:2001:bisection}
-to determine the optimal population size for each problem size.
-Extended by \cite{goldman:2012:ltga}, this method iteratively
-doubles the population size until some success criteria is met and then performs bisection
-between the lowest successful and highest unsuccessful sizes. In this way the minimum population size
-which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
-of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
-above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
-do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
-As bisection can make infinitely large population sizes, any run which had not found the global
-optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-hboa}
-      \end{centering}
-      \caption{Tuned Population Sizes for hBOA}
-      \label{fig-pop-hboa}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-ltga}
-      \end{centering}
-      \caption{Tuned Population Sizes for LTGA}
-      \label{fig-pop-ltga}
-    \end{subfigure}
-  \end{centering}
-  \caption{Optimal population sizes found using bisection on each size
-           for each problem.}
-  \label{fig-pop-sizes}
-\end{figure}
-
-Figure~\ref{fig-pop-sizes} shows the results from performing bisection. In general hBOA
-required population sizes which were at least an order of magnitude larger than LTGA. Due to
-runtime and memory overhead, finding the optimal value for hBOA was much less tractable than LTGA
-for moderate to large problem sizes. LTGA's population size also grew significantly slower
-than hBOA's as problem size increased, especially on the two Trap problems and Rastrigin.
-Both algorithms were very ineffective on MAX-SAT, neither able to tune to problems sizes over 60 bits.
-This is likely due to the fact that some randomly generated MAX-SAT landscapes are very flat
-and highly deceptive~\citep{rana:1998:gamaxsat}.
-\section{Finding the Global Optimum}
-\label{sec-optimum}
-
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{evals-to-success}
-  \end{center}
-  \caption{Comparison of the median number of evaluations to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-evals-to-success}
-\end{figure}
-
-Figure~\ref{fig-evals-to-success} shows the median number of evaluations required by each of the six
-algorithms to find the global optimum for multiple sizes of each problem. Each data point in
-Figure~\ref{fig-evals-to-success} represents the median of 100 runs, where unsuccessful runs
-are treated as requiring more evaluations than any successful run. If the median run was not successful
-no point is shown.
-Medians are used as the data is not normally distributed, and because it allows for more meaningful comparison
-between techniques with different success rates.
-The maximum problem size used for each problem was set to be the largest size we could
-feasibly determine LTGA's optimal population size. For many larger problems results are not shown
-for hBOA due to the extreme computational cost required to optimally set the population size.
-
-\subsection{Quantitative Comparison}
-Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
-The largest problem size for any problem where P3 was not the most efficient has 49 bits, with P3 achieving the best results on all 92 larger
-configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
-Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
-and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
-on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
-The likelihood that P3 would achieve these pairwise results assuming its median result is actually worse is $p < 10^{-15}$
-according to the binomial test.
-Pairwise comparison of LTGA and P3 on the largest problem size using the Mann-Whitney~U test results in
-$p < 10^{-5}$ for all problems.
-
-\subsection{Local Search}
-The Random Restart Hill Climber and $1+(\lambda, \lambda)$ are both relatively effective on small problem
-sizes. This is especially true for the three randomly generated problem classes. These problems
-may contain relatively few local optima or just be exceptionally difficult for the model based algorithms.
-On Deceptive Trap and Deceptive Step Trap using 4 or less traps, $1+(\lambda, \lambda)$ performs significantly
-better than any other algorithm. We believe this is
-because $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
-flipping entire traps.
-This ability also leads $1+(\lambda, \lambda)$
-to outperform the Random Restart Hill Climber on all problems except Nearest Neighbor NK.
-
-On larger problem sizes, the ability for local search to reach the global optimum quickly diminishes.
-Only on MAX-SAT are these optimizers competitive
-at larger tested problem sizes. However, we believe this is because the largest tested MAX-SAT
-was an order of magnitude smaller than the largest size tested for most other problems. As the problem
-size increases the number of local optima increases exponentially, which explains why Random Restart Hill
-Climbing was unable to scale.
-For larger problems it also becomes increasingly unlikely for $1+(\lambda, \lambda)$ to make the right combination
-of changes required to reach the global optimum. This
-behavior causes high variance in success rate, as evident by the occasional successes on large Deceptive Trap problems.
-
-\subsection{Model Building}
-Only techniques which explicitly built models of gene epistasis were able to solve the largest problem
-instances. On single instance problems LTGA was more effective than hBOA, with hBOA outperforming
-LTGA on Nearest Neighbor NK and Ising Spin Glasses. This may be caused by the differences in modeling method:
-unlike the single instance problems, gene epistasis in the randomly generated
-problem classes cannot be be perfectly represented with a linkage tree.
-
-Considering how different hBOA and LTGA are in performing optimization, it is somewhat surprising how similar
-their results are on HIFF.
-However, both techniques rely on populations large enough to support the diversity required to reach the global optimum
-and to model epistasis. Both techniques also only rebuild models once per generation. As the subproblems of HIFF
-are nested, it is unlikely that either technique can accurately model higher order epistasis before solving
-lower order subproblems. Therefore both methods require one generation per subproblem order.
-
-\subsection{P3}
-Unlike the other model based methods, P3 generally outperforms
-both Random Restart Hill Climber and $1+(\lambda, \lambda)$ even on small problem sizes.
-Unlike the other local search methods, P3 outperforms LTGA and hBOA even on large problem sizes.
-This implies that P3 is gaining the benefits of each, leveraging local search to solve easy
-problems and model building to solve harder ones.
-
-Furthermore, the interaction between these two optimization tools explains some of the reason P3
-outperforms each method alone.
-On Deceptive Trap P3's use of hill climbing ensures all traps are immediately optimized, allowing for perfect
-linkage detection and high quality donation.
-On HIFF local search solves all pairwise subproblems, saving P3 a generation over LTGA and hBOA.
-In comparison P3 is only a slight improvement on Deceptive Step Trap, which is less amenable to local search.
-
-\section{Fitness Over Time}
-\label{sec-overtime}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{fitness-over-time}
-  \end{center}
-  \caption{Compares the median best fitness reached during search for each of the six optimization methods.}
-  \label{fig-fitness-over-time}
-\end{figure}
-
-For some applications, finding the global optimum is less important than finding good solutions quickly. Therefore
-we examine this behavior in Figure~\ref{fig-fitness-over-time}. At regular intervals during optimization
-Figure~\ref{fig-fitness-over-time} shows the median of the best fitnesses found at that point of search across
-100 runs. For each problem we show the largest problem size for which we were able to successfully gather results
-for all six algorithms, but the trends shown are representative of all larger problem sizes. The maximum reporting interval
-is set to include the slowest P3 run to reach the global optimum.
-
-% 181 total
-% 121 has P3 best
-%   9 hc
-%  50 lambdalambda
-%  20 hboa
-%  18 phboa
-%  27 LTGA
-
-\subsection{Quantitative Comparison}
-Of 181 sample points, P3 had the highest median fitness in 121. In pairwise competition, $1+(\lambda, \lambda)$
-was the most likely to outperform P3, doing so on 50 sample points. LTGA, hBOA, and Parameter-less hBOA
-were the next best, outperforming P3 on 27, 20, and 18 sample points, respectively. Random Restart Hill Climbing almost
-never outperformed P3, doing so only 9 times.
-The likelihood that P3 would achieve these pairwise results assuming its median result is actually worse is $p < 10^{-9}$
-according to the binomial test.
-
-\subsection{Local Search}
-Perhaps the most striking result is the quality of $1+(\lambda, \lambda)$. Until quite far into search this
-method performs better than both LTGA and hBOA. Given sufficient evaluations $1+(\lambda, \lambda)$ also outperforms
-Random Restart Hill Climbing on all 7 problems. For brief periods in the middle of search it performs the
-best of all techniques on: Deceptive Trap; Deceptive Step Trap; HIFF; Ising Spin Glass; and MAX-SAT problems.
-$1+(\lambda, \lambda)$'s ability to efficiently incorporate gene modifications of larger
-than one bit allows it to overcome the deception and plateaus in Deceptive Trap and Deceptive Step Trap, solve medium sized subproblems
-in HIFF, flip the signs on multiple adjacent bits in Ising Spin Glass, and cross plateaus in MAX-SAT. However,
-this method is slow in reaching the global optima in many of these problems which causes it to eventually
-be overtaken by the model building techniques.
-
-\subsection{Model Building}
-Both hBOA and LTGA are marked by periods of little improvement followed by rapid improvement.
-In hBOA this is taken to the extreme, with all fitness
-improvement made at the very end of search. In both cases this is caused by model building. Before the model
-is accurate little improvement is made. Once it is accurate, fitness improves dramatically.
-
-At 58\% of the recording intervals hBOA has the worst fitness of any solver. Most of the exceptions
-occur when hBOA is still evaluating its initial population, allowing this random search to temporarily surpass the
-local search methods. After $N$ evaluations, however, hBOA and LTGA both fall behind until their models begin
-to improve.
-Parameter-less hBOA reaches intermediate fitnesses faster than hBOA, doing so on 62\% of intervals, as its models begin
-to optimize earlier than hBOA.  However, this trend is reversed after a sufficient number of evaluations, most clearly
-on Deceptive Step Trap and Ising Spin Glasses, as hBOA's tuned population overtakes Parameter-less hBOA's parallel populations.
-
-On every problem LTGA has five distinct periods: fitness plateau, near instantaneous improvement, fitness plateau,
-and improvement to global optimum. The early period corresponds with
-initialization of the population, with the first fitness gain achieved immediately upon completing the first generation.
-When using an inaccurate model, LTGA's mixing strategy performs a sort of less effective local search.
-Subsequent generations then make only minor fitness improvements. Once the model
-becomes accurate and the probability of a crossover using high quality genetic material increases sufficiently, LTGA
-enters a second period of rapid improvement.
-
-\begin{comment}
-This behavior is easiest to understand on the two Trap problems. LTGA's first generation will push individual traps toward a
-local optima, but it requires more than a single generation to do so completely. Until that occurs, the model is likely behaving
-no better than random as there is little apparent linkage between bits. Once the model begins to identify individual traps, crossover
-can begin to increase the frequency of higher fitness trap settings. This process begins slowly as low fitness local optima are more
-likely to be chosen for donation than high fitness local optima due to their frequency in the population. The process becomes
-self-catalysing as increased frequency of the global optimum trap genes means increased likelihood of the global optimum being spread by crossover.
-\end{comment}
-
-\subsection{P3}
-The integration of hill climbing into P3 makes it strictly better than using hill climbing alone.
-Early in optimization P3 and the Random Restart Hill Climber have effectively identical quality. This is because
-P3 performs the same evaluations as the Hill Climber for the first two restarts. Once P3 begins
-performing crossover it immediately improves over the Hill Climber. In 95\% of intervals P3 had a fitness at
-least as high as Hill Climbing. As such P3 is better than a simple
-hill climber regardless of how long each technique is run and irrespective of how high quality the solution
-found has to be.
-
-Unlike the model based methods, which struggle until model accuracy improves, P3's iterative solution
-integration allows it to improve much more quickly. This behavior exists in most problems, but is easiest to understand on Deceptive Trap.
-On this problem, P3 immediately brings all traps to local optima, equaled only by the Random Restart Hill Climber in quality.
-In comparison LTGA must evaluate the entire population and perform multiple generations to reach similar quality.
-P3 is able to immediately integrate optimal versions of each
-trap into a single individual as they are found by local search, resulting in smoother fitness improvement than LTGA.
-
-\section{Computational Expenses}
-While it is common in evolutionary computation to assume the evaluation function will dominate algorithm
-complexity, in some domains this will not be true. Model based methods are especially likely to violate
-this norm. Therefore, in order to assess P3's quality in solving problems with efficient fitness functions,
-we provide data on both its algorithmic complexity and wall clock time.
-
-\subsection{Operation Counting}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{rebuilds}
-      \end{centering}
-      \caption{Rebuilds}
-      \label{fig-rebuilds}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{donations}
-      \end{centering}
-      \caption{Donations}
-      \label{fig-donations}
-    \end{subfigure}
-  \end{centering}
-  \caption{Estimated computation costs incurred by model rebuilding (Figure~\ref{fig-rebuilds}) and
-           repeated donations (Figure~\ref{fig-donations}) per evaluation as problem size increases.}
-  \label{fig-costs}
-\end{figure}
-
-When discussing the asymptotic complexity of P3 in Section~\ref{sec-p3}, two aspects eluded precise
-analysis: how expensive is model rebuilding and how many gene donations are made. Figure~\ref{fig-costs}
-provides some insight into how often these two aspects of the algorithm are utilized.
-
-Figure~\ref{fig-rebuilds} reports in an algorithmic sense how expensive model rebuilding is for search.
-In order to calculate this value we recorded how many times search rebuilt the model during each run.
-Figure~\ref{fig-rebuilds} shows the estimated ratio of model rebuilding cost ($N^2$ per rebuild) over evaluation cost ($N$ per evaluation).
-If the cost of model building
-scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically constant.
-For Nearest Neighbor NK, Ising Spin Glasses, and Rastrigin this is the case. For both Trap problems
-and HIFF there is slow growth in the ratio. The problem sizes used for MAX-SAT were not sufficient
-to accurately gauge the asymptotic behavior. Together this suggests that while the cost of building the model
-is almost linear per evaluation, it can grow slowly. However, even in the worst case (HIFF) this growth was
-no more than twice the algorithmic cost of an evaluation even using 2048 bits.
-
-When applying a crossover subset, P3 tries random donors from the population until one is found with at least
-one bit different from the improving solution. In theory this can result in up to $O(\mu)$ operations.
-Figure~\ref{fig-donations} examines the observed average number of donations per evaluation performed.
-Ising Spin Glass, HIFF, and Rastrigin all achieve effectively constant behavior here, implying repeated
-donation does not impact the asymptotic runtime of P3. Both Trap functions and Nearest Neighbor NK all
-increase in number of donations as problem size increases, potentially increasing algorithmic costs. An
-important note is that each donation may range in size from a single bit up to $N-1$. However,
-repeated donation attempts are far more likely to happen with smaller clusters.
-As such this may cause some super-linear growth
-in P3, but its unlikely to be very high.
-
-\subsection{Wall Clock Performance}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{seconds-to-success}
-  \end{center}
-  \caption{Comparison of the median number of seconds to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-seconds-to-success}
-\end{figure}
-
-To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}.
-Similar to Figure~\ref{fig-evals-to-success}, each point represents the median
-of 100 runs, with unsuccessful runs treated as slower than successful runs.
-\begin{comment}
-In total these runs
-took over 6 computing years to complete. We were therefore forced to use
-diverse hardware potentially running multiple processes in parallel.
-As such these timings should be considered very skeptically as a number
-of sources may contribute to biased error in the results. However, due to
-the size of the dataset and the general regularity in the results some general conclusions
-can still be made.
-\end{comment}
-These results were collected using 2.5GHz Intel Xeon E5-2670v2 processors.
-
-\subsubsection{Model Building}
-hBOA and Parameter-less hBOA perform much worse when using wall clock time as the unit of comparison than when using
-evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while, under reasonable
-assumptions, P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
-hBOA goes from being slightly more efficient than LTGA in terms of evaluations to three orders of magnitude worse in terms of seconds.
-As P3 and LTGA require a similar asymptotic complexity per
-evaluation as the Hill Climber and $1+(\lambda, \lambda)$, no similar change in ordering occurs.
-
-\subsubsection{P3}
-When LTGA is optimally tuned to a single instance problem with an efficient evaluation function
-it can find the global optimum faster than P3 in terms of wall clock time. However, on randomly
-generated problem classes P3's efficient use of evaluations is enough to overtake LTGA.
-
-On the four single instance problems LTGA not only finds the global optimum using less wall clock time,
-the factor speedup increases as problem length does. Na\"{i}vely this suggests LTGA is achieving
-a lower order of complexity. However, for these experiments LTGA is growing at sub-linear time per
-evaluation, which is not asymptotically stable due to (at minimum) the time required to perform an evaluation.
-We suspect that the true cause is that $N$ is small enough to be overshadowed by lower order polynomial terms.
-For example, LTGA requires $O(N/\mu)$ time per evaluation to rebuild the linkage
-model from the frequency table. As a result, for small $\mu$ model building, and not extracting pairwise
-frequency, can dominate runtime.
-
-When applied to randomly generated problem classes, the differences in P3 and LTGA's evaluation complexity
-dominates runtime complexity. Similar to with Figure~\ref{fig-evals-to-success}, the amount of speedup P3
-achieves over LTGA increases with problem size on Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT.
-
-Across both types of problems we find that P3's time per evaluation grows approximately linearly. As such,
-we conclude that P3 requires asymptotically similar amounts of time per evaluation as the other efficient techniques.
-
+%%%%%%%%%%%%% Population Sizing %%%%%%%%%%%%%
 \section{Population Sizing}
@@
 
+%%%%%%%%%%%%%% P3 Inner Workings %%%%%%%%%%%%%%%%%%%%%%%
+\section{Inner Workings Specific to P3}
+
 \begin{comment}
@@
 Thus the level size and crossover success rates increase after contracting around level four.
-
-\section{Conclusions and Future Work}
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing black box
-optimization. P3's primary innovation is the replacement of the generational model with a pyramid of populations.
-This pyramid is constructed iteratively, with both the number of levels and the number of solutions stored
-at each level growing as search progresses. P3 uses a model based crossover method
-which learns a linkage tree from gene epistasis. Combined with a simple hill climber, P3's design contains
-many synergistic features.
-
-Across a large number of problems and problem sizes P3 required less evaluations to reach the global optimum
-than optimally tuned state-of-the-art competitors. On single instance problems P3's improvement was by a
-constant factor, while for the three randomly generated problem classes P3's improvement increased with problem size. This
-quality extends to intermediate points during evolution, with P3 generally reaching at least as high
-of fitness as the competitive techniques when using the same number of evaluations. While P3 does require
-modeling overhead, the expense of this overhead is approximately linear with respect to genome size. There
-is some evidence that even when compared on wall clock time, P3 performs on par with
-the best comparison techniques. All of these achievements are made without any problem specific parameter
-tuning, making P3 easier to apply to new domains than its two closets competitors in quality.
-
-P3's quality is due to a number of desirable traits. First, mixing local search with model based
-crossover lets search focus on properly mixing high quality solutions. Second, by adding diversity only
-as necessary P3 tends to use the minimal amount of random initialization, unlike other techniques which must
-overcompensate with larger population sizes on single instance problems and consider the worst instance
-when solving problem classes. Third, by heavily exploiting existing diversity before adding more P3 is able
-to reach high quality intermediate fitnesses quickly without prematurely converging. Fourth, the very
-nature of the pyramid's shape allows search to preserve a desirable proportion of diversity at
-each fitness level, similar to a generational model using a decreasing population size.
-
-There are a number of meaningful avenues for future P3 experimentation. Perhaps the most
-pressing for practitioner acceptance is to apply P3 to real world problems and compare
-its results with other black box or even problem specific heuristics. While parameter-less,
-P3 is currently limited to discrete, fixed length genomes evaluated using single objective
-fitness. These limitations can be relaxed with future work to make P3 more widely applicable.
-While asymptotically linear in problem size, P3's modeling techniques and local search methods are likely going
-to be prohibitively expensive for genome sizes in the hundreds of thousands or millions of genes,
-and the inability of the model to capture overlapping linkage may be hindering
-search efficiency. Overcoming these limitations by using a new modeling technique may allow
-the pyramid model even greater flexibility. Similarly, while P3 is able to overcome low
-order deception via linkage learning, the iterative improvement method by which crossovers
-are made may mislead search on landscapes with higher order deception.
-
-However, even without
-these improvements our results show P3 is highly efficient at finding
-global optima on black box problems without any problem specific tuning.
-
-\section{Acknowledgements}
-This material is based in part upon work supported by the National Science Foundation
-under Cooperative Agreement No. DBI-0939454. Any opinions, findings, and conclusions
-or recommendations expressed in this material are those of the author(s) and do not
-necessarily reflect the views of the National Science Foundation.
-
-\small
-
-\bibliographystyle{apalike}
-\bibliography{../main}
-
-\normalsize
-\appendix
-\section{hBOA Simplification}
-\label{sec-appendix}
-To measure the quality of a decision forest, hBOA applies Equation~\ref{eq-hboa}.
-To favor compact
-models, Equation~\ref{eq-hboa} is scaled by Equation~\ref{eq-hboa-stop}, which provides increased cost
-for more total leaves.
-This quality is used for two purposes: comparison of potential changes from the existing model
-and comparison of that change with the existing model. The basis for our simplication
-is to rearrange $p(B)BDe(B) < p(B')BDe(B')$ to be $\frac{p(B)}{p(B')}<\frac{BDe(B')}{BDe(B)}$.
-
-\begin{equation}
-  BDe(B) = \prod_{i=0}^{N}\prod_{l\in L_i} \frac{\Gamma(m'_i(l))}{\Gamma(m_i(l) + m'_i(l))}
-  \prod_{x_i}\frac{\Gamma(m_i(x_i, l) + m'_i(x_i,l))}{\Gamma(m'_i(x_i,l))}
-  \label{eq-hboa}
-\end{equation}
-
-\begin{equation}
-  p(B) = c2^{-0.5(\sum_i|L_i|)log_2\mu}
-  \label{eq-hboa-stop}
-\end{equation}
-
-The outermost product of Equation~\ref{eq-hboa} iterates over all trees in the forest. However, each split
-can modify only one of the trees and therefore the contribution of all others can be canceled. The middle
-product is across all leaves in the tree. Again since only one leaf can be changed, all other terms can
-be canceled. By convention hBOA uses uninformed Bayesian priors of $m'_i(l)= 2$ and $m'_i(x_i, l)=1$ for
-binary alphabets. As $\Gamma(a) = (a-1)!$ this means the top term in the middle product and the bottom
-term in the third product reduce to 1. The only remaining terms are then $m_i(l)$ and $m_i(x_i, l)$ which
-represent the number of solutions which reached leaf $l$ and the number of solutions which reached leaf $l$
-with a specific value for $x_i$, respectively.
-
-Equation~\ref{eq-hboa-stop} can also be simplified when doing comparisons. If model $B'$ has exactly one more
-leaf than model $B$ then the ratio $\frac{p(B)}{p(B')}$ simplifies to $2^{0.5 log_2\mu}$ regardless of
-total model size.
-
-The resulting simplifications create Equation~\ref{eq-hboa-final},
-where $B'$ is different from $B$ by exactly 1 split, such that $l$ was split to create $l'$ and $l''$.
-The best split is whichever maximizes its improvement over $B$, which is equal to the right side
-of the inequality. Note that these factorials can still be exceedingly large and
-therefore it is imperative that implementations avoid rounding errors and overflows.
-
-\end{document}

### 2015-04-15 Dissertation/chapter-introduction.tex
Copied Comprehensive to Disseration
Major update from journal article polishing

--- 2014-10-15 Comprehensive/chapter-introduction.tex
@@
 \label{chap-introduction}
-%TODO Expand motivation and assume less reader knowledge
+%TODO Explain epistasis
 A primary purpose of evolutionary optimization is to efficiently find good solutions
@@
 domain experts may not understand how the parameters will interact with the problem or with each other.
+To further complicate matters, there is mounting evidence that parameter values should change
+during search~\cite{goldman:2011:dynamic-parameters,laporte:2014:adaptivepop}.
 
 There have been periodic efforts to reduce or remove the need for parameter tuning.
-\cite{Back:1992:selfadapt} introduced self-adaptive parameters, in which parameter values
+\cite{rechenberg:1973:es} introduced self-adaptive parameters, in which parameter values
 were included in each solution's genome and themselves underwent evolution. This allowed search
-to optimize some of its own parameters resulting in a reduced the need for expert tuning.
+to optimize some of its own parameters, resulting in a reduced need for expert tuning.
 \cite{harik:1999:parameterlessga} were able to design an entirely parameter-less GA by
@@
 comparison algorithms; compare both efficiency in reaching the global optimum and intermediate
-fitnesses; algorithm complexity analysis; and provide more in depth analysis of P3 itself.
+fitnesses; analyze algorithm complexity; and provide more in depth analysis of P3 itself.

### 2015-04-15 Dissertation/chapter-p3.tex
Major update from journal article polishing

--- 2015-02-23 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
@@
-\documentclass[twoside]{article}
-\usepackage{ecj,palatino,epsfig,latexsym,natbib}
-\usepackage{url}
-\usepackage{verbatim}
-\usepackage{subcaption}
-\usepackage[noend]{algpseudocode}
-
-\newcommand{\includegraphicswide}[1]
-{\includegraphics[width=.9\textwidth,height=\textheight,keepaspectratio]{#1}}
-
-\newcommand{\includegraphicsfit}[1]
-{\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}
-
-\usepackage{array}
-\newcolumntype{L}{>{\centering\arraybackslash}m{1.9cm}}
-
-%% do not add any other page- or text-size instruction here
-
-\parskip=0.00in
-
-\begin{document}
-
-% 45-character limit on header
-\ecjHeader{x}{x}{xxx-xxx}{201X}{Parameter-less Population Pyramid}{B. W. Goldman and W. F. Punch}
-\title{\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid}
-
-\author{\name{\bf B. W. Goldman} \hfill \addr{brianwgoldman@acm.org}\\ 
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-\AND
-       \name{\bf W. F. Punch} \hfill \addr{punch@msu.edu}\\
-        \addr{Department of Computer Science and Engineering, Michigan State University, 
-        East Lansing, 48823, United States}
-}
-
-\maketitle
-
-\begin{abstract}
-%Should be about 200 words
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
-evolutionary optimization without requiring any user specified parameters.
-P3's primary innovation is to replace the generational model with a pyramid of
-multiple populations that are iteratively created and expanded. In combination
-with local search and advanced crossover P3 scales to problem difficulty, exploiting
-previously learned information before adding more diversity.
-
-Across seven problems, each tested using on average 18 problem sizes, P3 outperformed
-all five advanced comparison algorithms. This improvement includes requiring less evaluations
-to find the global optimum and better fitness when using
-the same number of evaluations. Using both algorithm analysis and comparison we find P3's
-effectiveness is due to its ability to properly maintain, add, and exploit diversity.
-
-Unlike the best comparison algorithms, P3 was able to achieve this quality without any
-problem specific tuning. Thus, unlike previous parameter-less methods, P3 does not
-sacrifice quality for applicability. Therefore we conclude that
-P3 is an efficient, general, parameter-less approach to black box
-optimization which is more effective than existing state-of-the-art techniques.
-
-\end{abstract}
-
-\begin{keywords}
-
-Genetic algorithms, 
-linkage learning,
-local search,
-parameter-less.
-
-\end{keywords}
-
-\section{Introduction}
-A primary purpose of evolutionary optimization is to efficiently find good solutions
-to challenging real world problems with minimal prior knowledge about the problem itself.
-This driving goal has created search algorithms which can escape user bias to create
-truly novel results, sometimes publishable or patentable in their own right~\citep{kannappan:2014:humies}.
-While it is not possible for any algorithm to do better than random search across all possible
-problems~\citep{Wolpert:1997:nfl}, effectiveness can be achieved by assuming the search
-landscape has structure and then biasing the algorithm toward exploiting that structure.
-
-In evolutionary optimization, and genetic algorithms~(GAs) in particular, search is often
-biased through parameters. This can be beneficial as it allows practitioners to inject their
-knowledge about the shape of the search landscape into the algorithm.
-However, the quality of solutions found, and the speed at which they are found, is strongly tied to setting these parameters
-correctly~\citep{goldberg:1991:gasize}. As such, either expert knowledge or exceedingly
-expensive parameter tuning~\citep{grefenstette:1986:optimalga} are required to leverage
-this feature to its fullest potential. Furthermore,  parameters such as population size, mutation rate, crossover
-rate, tournament size, etc. usually have no clear relationship to the problem being solved, meaning even
-domain experts may not understand how the parameters will interact with the problem or with each other.
-To further complicate matters, there is mounting evidence that parameter values should change
-during search~\citep{goldman:2011:dynamic-parameters,laporte:2014:adaptivepop}.
-
-There have been periodic efforts to reduce or remove the need for parameter tuning.
-\cite{rechenberg:1973:es} introduced self-adaptive parameters, in which parameter values
-were included in each solution's genome and themselves underwent evolution. This allowed search
-to optimize some of its own parameters, resulting in a reduced need for expert tuning.
-\cite{harik:1999:parameterlessga} were able to design an entirely parameter-less GA by
-leveraging schema theory and parallel populations. Unfortunately these methods were provably less efficient
-than directly setting the parameters to optimal values~\citep{pelikan:1999:worstparameter-less}.
-
-One area that has been very effective at reducing the number of algorithm parameters is
-model based search. \cite{pelikan:2006:hboa}'s Hierarchical Bayesian Optimization
-Algorithm~(hBOA) and \cite{thierens:2010:ltga}'s Linkage Tree Genetic Algorithm~(LTGA)
-both require only a single parameter: population size. \cite{posik:2011:parameterless}
-leveraged model building to create a fully parameter-less algorithm, but it is restricted to
-only order-k, fully decomposable, noiseless problems.
-
-Most recently \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid~(P3).
-This method uses a pyramid structure of populations to combine model based search with local search
-to achieve parameter-less optimization. Initial results suggest that unlike
-previous parameter-less methods, P3 is actually more efficient than current state-of-the-art
-parameterized search algorithms. In this work we shall: extend these results to cover more
-comparison algorithms; compare both efficiency in reaching the global optimum and intermediate
-fitnesses; analyze algorithm complexity; and provide more in depth analysis of P3 itself.
-
-\section{Comparison Optimizers}
-\label{sec-optimizers}
-
-In order to fully understand the effectiveness of P3, we compare it with five advanced algorithms which
-have related features to P3. The Random Restart Hill Climber defined by ~\cite{goldman:2014:p3} was chosen as an efficient form of repeated
-local search. As P3 combines this hill climber with crossover,
-comparing with local search alone shows the advantages of P3's overall approach.
-The $1+(\lambda, \lambda)$ algorithm~\citep{doerr:2013:lambdalambda} is the current best theory supported simple genetic algorithm
-and its method of crossover is in some sense a macro-mutation just as in P3. hBOA and Parameter-less
-hBOA are advanced model building search techniques which are very effective at learning complex
-problem structure, designed to achieve similar goals as P3's linkage learning but using very different
-methods. Finally LTGA represents the current state-of-the-art in black box search and is the origin
-of P3's linkage learning and crossover methods.
-
-Only hBOA and LTGA require any parameters, with each of these only requiring a population
-size. This makes knowing the optimal behavior of these algorithms much more tractable. All of the
-algorithms are also gene order independent, fitness scale invariant, and unbiased. This means,
-for any problem, the order in which problem variables appear in the genome can be changed
-without changing the behavior of the search. The fitness can also be manipulated in any fashion
-as long as the rank ordering of solutions is unchanged. These algorithms are also unaffected by the meaning
-assigned to each bit, such that inverting a predetermined random subset of genes before evaluation
-will not impact search efficiency.
-
-Our implementations of all of these algorithms as well as all of the population size information, raw results,
-and processing scripts
-are available from our website.\footnote{\url{https://github.com/brianwgoldman/FastEfficientP3}}
-\subsection{Random Restart Hill Climber}
-\label{sec-hill-climber}
-Perhaps the simplest black box search heuristic is stochastic local search, or hill climbing.
-This optimization technique focuses on improving a single solution until it reaches a local
-optimum. Here we use the first improvement hill climber defined by \cite{goldman:2014:p3}
-and given in Figure~\ref{fig-hc}. This algorithm works by flipping each bit in a random
-order, keeping modifications when fitness is improved, until single bit flips cannot result in
-further fitness improvements.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Hill-Climber}{}
-    \State $options \leftarrow [0 \dots N-1]$\label{fig-hc-options}
-    \State $tried \leftarrow \emptyset$
-    \While{$|tried| < |options|$}
-      \ForAll{$index \in shuffled(options)$}
-        \If{$index \notin tried$}\label{fig-hc-tried}
-          \State Flip bit $index$ in solution
-          \If{solution's fitness increased}
-            \State $tried \leftarrow \emptyset$
-          \Else
-            \State Revert change
-          \EndIf
-          \State $tried \leftarrow tried \cup \{index\}$
-        \EndIf
-      \EndFor
-    \EndWhile
-  \EndProcedure
-\end{algorithmic}
-  \caption{Hill climbing algorithm used to improve randomly generated solutions until no single
-           bit change results in a fitness improvement.}
-  \label{fig-hc}
-\end{figure}
-
-The hill climber requires an amortized cost of $O(1)$ operations per evaluation. In order to
-terminate, at least one evaluation must be performed for each of the $N$ bits in the solution.
-As such any operation that happens only once per search can be amortized over at least $N$
-evaluations, covering the initialization of $options$ on Line~\ref{fig-hc-options}.
-Line~\ref{fig-hc-tried}, which prevents wasted evaluations, can be called at most twice
-per evaluation: once to add $index$ into $tried$ and once to prevent $index$ from being unnecessarily
-evaluated again. The only way three or more calls could happen is if
-no fitness improvement was made for the entire previous iteration, which contradicts the loop invariant.
-
-Due to its nature, this hill climber cannot escape basins of attraction. Once a solution is reached
-such that none of the single bit neighbors are fitness improvements, search stops. Thus
-this algorithm requires a restart mechanism to solve multimodal problems. We have
-chosen here to na\"{i}vely restart search from a random solution whenever a local optima is found. This
-ensures that on all landscapes there is always a non-zero probability of search finding the global optimum.
-
-\subsection{$1+(\lambda, \lambda)$}
-\cite{doerr:2013:lambdalambda} presented the first genetic algorithm to provably show
-the advantages of performing crossover on One Max. This comparatively simple
-algorithm maintains only a single individual and a self-controlled parameter $\lambda$.
-
-Each iteration, the number of bits to flip is chosen from the binomial distribution $b\sim B(N, \frac{\lambda}{N})$,
-where $N$ is the number of bits in the genome.
-Next, $\lfloor\lambda\rfloor$ offspring are produced by flipping $b$ bits. The
-best mutant then produces $\lfloor\lambda\rfloor$ offspring via uniform crossover with the original parent, such that each gene comes from the
-mutant with probability $\frac{1}{\lambda}$. In the original algorithm the best
-offspring produced by crossover then replaces the original parent if its fitness is no worse.
-The $\lambda$ parameter, which is initialized to 1, is decreased if the offspring replaced
-its parent and increased otherwise.
-
-The original formulation was designed specifically for unimodal landscapes and as such were
-not directly suitable for multimodal problems. \cite{goldman:2014:p3} extended $1+(\lambda, \lambda)$
-to include random restarts. As search stagnates, the $\lambda$ parameter increases in value. Eventually
-this results in $\lambda \ge N$ causing mutation to always flip all bits of the individual.
-As this prevents any future improvement, whenever $\lambda \ge N$ search is restarted from a random solution with $\lambda$
-reset to 1.
-
-A few other efficiency modifications were also made. If there is a tie in crossover offspring fitness,
-whichever has a larger hamming distance from the parent is retained. This encourages drifting across plateaus.
-The ``mod'' control strategy proposed by \cite{doerr:2013:lambdalambda} was not used as it conflicted with
-the random restart strategy.
-If a crossover individual is identical to either of its parents, it is not evaluated.
-If mutation produces an offspring which is better than the best crossover offspring, it is used to compare
-against the original parent.
-
-\subsection{Hierarchical Bayesian Optimization Algorithm}
-
-\cite{pelikan:2006:hboa} used statistical principles in combination with a decision tree structure
-to create the Hierarchical Bayesian Optimization Algorithm (hBOA). This method creates a model of
-epistatic relationships between genes which is then used to stochastically generate new solutions.
-Each generation a binary tournament with replacement is used to select $\mu$ solutions from
-the population. These solutions are then used to build the model, which in turn is used to generate $\mu$ new
-solutions. The new solutions are then integrated into the population using restricted tournament
-replacement.
-
-Conceptually, the model built by hBOA is trying to infer rules of the form ``Given that this
-subset of genes are set to these values, how frequently is gene $x_i$ set to value v?'' This can
-be represented using a directed acyclic decision forest, with each tree in the forest representing one gene
-in the solution. In the decision tree $T_i$, which is used to set the value of gene $x_i$,
-each internal node represents previous decisions on how to set
-some other gene $x_j$, with the children of that node representing how the decision was made. The
-leaves of each tree give the probability that $x_i$ should be set to each possible value.
-
-The forest is constructed iteratively, with each tree initially containing a single leaf
-and with each leaf storing a pointer for each selected solution. Each iteration the algorithm considers
-all possible ways of splitting an existing leaf using another gene $x_j$, such that solutions in the
-leaf are moved to the newly created leaves based on their value for $x_j$. The general goal is to
-separate the solutions such that all solutions with $x_i = 0$ move to one leaf while solutions with
-$x_j = 1$ move to the other.
-
-This goal is formalized using model scoring from Bayesian statistics. In its raw form this
-almost always creates near infinitesimal results, calculating fractions which include factorials of $\mu$
-and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{sec-appendix}, we
-derived a simplified form shown in Equation~\ref{eq-hboa-final}.
-Here $l$ is a leaf in tree $i$, with $l'$ and $l''$ the results of splitting $l$. $m_i(l)$ is the number
-of solutions which reach $l$ and $m_i(x_i, l)$ is the number of solutions which reach $l$ with the given value
-for $x_i$.
-If no proposed split satisfies the inequality, iteration stops.
-If multiple splits do, whichever maximizes the right side is chosen.
-
-\begin{equation}
-  2^{0.5 log_2\mu} < \frac{(m_i(l) + 1)!}{m_i(0, l)!m_i(1,l)!} \cdot
-  \frac{m_i(0, l')!m_i(1,l')!m_i(0, l'')!m_i(1,l'')!}{(m_i(l') + 1)!(m_i(l'') + 1)!}
-  \label{eq-hboa-final}
-\end{equation}
-
-Initially there are $\Theta(N^2)$ possible ways to split existing leaves, as each of the $N$ single node
-trees can be split by any of the other $N-1$ genes. Each iteration a new edge is added to the decision
-forest, meaning some of the previously tested splits cannot be used. For instance, if $T_i$, which is used
-to decide the value of $x_i$, is split using the value of $x_j$, $T_j$ can no longer be split using $x_i$.
-As a split creates two new leaves, $O(N)$ new potential splits must also be tested. Equation~\ref{eq-hboa-final}
-parses all solutions which reach a leaf to count gene frequencies, requiring $\mu$ time.
-The number of total leaves created depends heavily on the problem and $\mu$.
-However, assuming no splits are accepted or that the cost of testing all future splits is less than
-the initial $\Theta(N^2)$, constructing the model requires $\Omega(\mu N^2)$ time. Each
-model is used to generate $\mu$ solutions, leading to a cost per evaluation of $\Omega(N^2)$.
-
-To generate a solution, the value of each gene $x_i$ is set using its corresponding decision tree $T_i$. Because
-the forest is directed acyclic, there must be an ordering of $T_i$ such that before $T_i$ is executed all
-$x_j$ it uses to make decisions have already been set. As such, previous decisions made by other trees
-are used to follow each $T_i$ until a leaf is reached. The value of $x_i$ is then set based on the
-probability that other solutions reached that leaf with each value of $x_i$.
-
-To perform replacement, hBOA uses restricted tournament replacement. After each new solution is generated
-and evaluated, a set of $w$ solutions are chosen at random from the population, where $w=\min\{N, \frac{\mu}{20}\}$.
-From this set the solution which is most genetically similar to the offspring is chosen. If the offspring
-is at least as fit as the chosen solution, it replaces the chosen solution in the population. Otherwise the
-offspring is discarded. This method is designed to preserve genetic diversity as only genetically similar
-solutions must compete on fitness.
-
-hBOA is designed to work with large population sizes, resulting in a large number of
-evaluations per generation. As hBOA utilizes explicit diversity maintenance, standard methods for determining
-convergence are not considered very accurate. Therefore the authors suggest that an hBOA run should be
-terminated after performing generations equal to $N$.
-
-Like other model based techniques, hBOA has very few parameters. There is no mutation or crossover,
-and modeling does not rely on any explicit parameters. Solution selection, generation, and replacement
-are all derived from the population size, which must be set by the user.
-
-\subsection{Parameter-less hBOA}
-Using the methods first introduced by \cite{harik:1999:parameterlessga} for the Parameter-less GA,
-\cite{pelikan:2004:parameterlesshboa} created Parameter-less hBOA which automatically scales its
-population size to fit the problem. This is done by maintaining a list of concurrent populations
-using exponentially scaled population sizes.
-
-A run of Parameter-less hBOA starts with a single population of size $\mu_0$, conventionally set
-to $\mu_0=10$. After two generations are performed, a new population of size $\mu_1 = 2\mu_0$ is created
-and performs a generation. Evolution then continues with the $\mu_0$ population performing two generations
-for each one performed by $\mu_1$. Each time population $\mu_i$ performs its second generation a new population
-$\mu_{i+1}=2\mu_i$ is created, which performs generations half as often as $\mu_i$. In this way an infinite number of
-parallel population can be simulated, with each population receiving the same number of total evaluations.
-
-In all other aspects each population is identical to an hBOA population using a fixed $\mu$. No search information
-is shared among populations, and each search is independently terminated. As such Parameter-less hBOA cannot
-perform better than hBOA using the optimal population size for a given instance, as it must also spend evaluations
-on populations of different sizes. This inefficiency is bounded by a log multiple of the total number of
-evaluations~\citep{pelikan:1999:worstparameter-less}.
-
-\subsection{Linkage Tree Genetic Algorithm}
-\cite{thierens:2010:ltga} introduced the Linkage Tree Genetic Algorithm (LTGA) which automatically
-detects and exploits problem epistasis by examining pairwise gene entropy. Due to its enhanced
-ability to preserve high fitness gene subsets, LTGA was able to outperform state-of-the-art
-GAs across many benchmarks. Since its introduction, many variants of LTGA have been
-proposed~\citep{thierens:2011:gomea,goldman:2012:ltga} so for clarity we have chosen the version
-presented by \cite{thierens:2013:ltgahiff} as our model.
-
-LTGA's effectiveness comes from its method of performing crossover. Instead of blindly
-mixing genes between parents, LTGA attempts to preserve important interrelationships
-between genes. Before performing any crossovers in a generation, LTGA first builds
-a set of hierarchical gene clusters which are then used to dictate how genes are mixed
-during crossover.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Cluster-Creation}{}
-    \State $unmerged \leftarrow \{\{0\}, \{1\}, \{2\}, \dots, \{N-1\}\}$
-    \State $useful \leftarrow unmerged$
-    \While{$|unmerged|>1$}
-      \State $C_i, C_j \leftarrow \min_{C_i,C_j \in unmerged} D(C_i, C_j)$
-      \State $unmerged \leftarrow unmerged - \{C_i, C_j\} + \{C_i \cup C_j\}$
-      \State $useful \leftarrow useful + \{C_i \cup C_j\}$
-      \If{$D(C_i, C_j) = 0$}
-        \State $useful \leftarrow useful - \{C_i, C_j\}$
-      \EndIf
-    \EndWhile
-    \State Order $useful$ based based on last merged first\label{fig-cluster-creation-ordering}
-    \State Remove largest cluster from $useful$
-
-    \Return $useful$
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm describing how LTGA creates clusters using Equation~\ref{eq-distance}
-           for a population. $unmerged$ and $useful$ are ordered sets of sets of gene loci.}
-  \label{fig-cluster-creation}
-\end{figure}
-
-Figure~\ref{fig-cluster-creation} provides the agglomerative method LTGA uses to create gene clusters.
-This algorithm creates a tree of sets using pairwise gene entropy, such that the leaves of the tree contain a single gene and
-internal nodes are the union of their children's sets. These sets are then used by crossover to specify epistatic
-relationships which should be preserved.
-The process begins by creating the set of sets $unmerged$ which tracks all top level clusters. Initially
-$unmerged$ contains single member sets for each gene. After each iteration the two sets with the minimum average pairwise
-distance (given in Equation~\ref{eq-distance}) are merged to create a single cluster. This process is repeated
-until only a single set remains in $unmerged$ which contains all of the genes in the genome.
-
-\begin{equation}
-  D(C_i,C_j) = \frac{1}{\left | C_i \right |\cdot \left |C_j \right|}\sum_{c_i \in C_i}\sum_{c_j \in C_j}
-  2 - \frac{H(c_i) + H(c_j)}{H(c_i \cup c_j)}
-  \label{eq-distance}
-\end{equation}
-\begin{equation}
-  H(c) = -\sum_{s\in S} p_c(s)\log(p_c(s))
-  \label{eq-entropy}
-\end{equation}
-
-Throughout this process $useful$ tracks the set of all gene clusters which should be preserved for use by crossover.
-This set begins with all genes in separate clusters, and each time a new cluster is created it is added to $useful$.
-However, not all clusters are necessarily worth keeping. For instance, in all versions of LTGA the cluster
-containing all genes is removed from $useful$ as preserving all genes during crossover can only create clones.
-\cite{thierens:2013:ltgahiff} extended this removal to include any unsupported subsets. If the pairwise distance
-between two clusters is 0, this means there are no individuals in the population which disrupt the relationships between the two
-clusters. Therefore when performing crossover there is no reason to believe a fitness improvement can be achieved
-by breaking the stored pattern. As such a cluster is only kept if its direct superset has a non-zero distance.
-As a final step Line~\ref{fig-cluster-creation-ordering} reorders $useful$ such that clusters appear in reversed
-order from which they were added to $useful$. Thus the most linked clusters and those containing single genes
-appear at the end of the returned list.
-
-\cite{thierens:2013:ltgahiff}'s version of LTGA does not use the entire population when determining pairwise entropy.
-Instead, binary tournament is used to select half of the population. This is done to ensure the model is built
-using only high quality solutions, even during the first generation.
-
-In order to efficiently perform clustering, a pairwise gene frequency table is constructed from the selected solutions.
-To calculate Equation~\ref{eq-distance}, Equation~\ref{eq-entropy} is called for each
-gene ($H(c_i)$) and pair of genes ($H(c_i \cup c_j)$). Extracting this information requires $O(\mu N^2)$
-time, where $\mu$ is the population size and $N$ is the genome size. The process of converting
-this pairwise frequency information into clusters can be achieved in $O(N^2)$ using the bookkeeping
-methods presented by \cite{gronau:2007:upgma}. This cost is performed only once per generation,
-and is then used to perform approximately $O(\mu N)$ crossover evaluations. As a result, the amortized cost of
-LTGA's model building is $O(N)$.
-
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Cluster-Usage}{}
-    \ForAll{$C_i \in useful$}
-      %\ForAll{$d \in shuffled(P_i)$}
-        \State $d \leftarrow rand\_choice(P)$\label{fig-cluster-usage-donate}
-        \State Copy $d$'s gene values for $C_i$ into solution
-        \If{solution was changed}
-          \If{solution's fitness decreased}
-            \State Revert changes
-          \EndIf
-          %\State \textbf{break}
-        \EndIf
-      %\EndFor
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm describing how clusters are used to perform crossover.}
-  \label{fig-cluster-usage}
-\end{figure}
-
-Figure~\ref{fig-cluster-usage} describes how the identified clusters are used by crossover to preserve
-gene linkage while still exploring the search space. Unlike more traditional crossover methods, LTGA
-crosses each individual with the entire population. Also, to produce a single offspring, multiple evaluations
-of the fitness function are performed.
-
-Each generation, each individual in the population undergoes crossover. In a single crossover event, each
-cluster of genes $C_i$ in $useful$ is applied as a crossover mask. A random donor $d$
-is chosen from the entire population (not just the model selected population), and $d$'s gene's for $C_i$ are copied
-into the working solution. If a modification is made, an evaluation is then performed. If the crossover
-resulted in no worse fitness then the changes are kept. This allows for neutral drift across plateaus.
-The resulting solution, which must be at least as fit as its parent, is then
-copied into the next generation.
-
-In total each individual can cause up to $|useful|$ evaluations. If all clusters were kept, even those deemed
-unhelpful, and all donations were evaluated, even those which did not change any genes, then \Call{Cluster-Usage}{}
-would perform exactly $2N-2$ evaluations for each of the $\mu$ solutions in the population. This provides the amortizing evaluations
-required to make clustering only $O(N)$ operations per evaluation. However, by skipping some evaluations, its
-possible that clustering may be super-linear.
-
-LTGA has no explicit form of diversity control and has no method for introducing new genetic information once
-the population has converged. Therefore an LTGA run is considered converged once two consecutive populations
-contain the same unique solutions.
-
-By design, LTGA only has a single parameter: population size. LTGA uses no mutation, and crossover is defined
-in terms of the clustering algorithm. Selection between generations is fully elitist and embedded in the crossover,
-with selection of model building solutions fixed to a binary tournament. Neither \Call{Cluster-Creation}{} nor
-\Call{Cluster-Usage}{} rely on parameter values. LTGA does not provide any method for controlling or setting
-the population size, relying instead on a fixed user specified size.
-
-\section{Parameter-less Population Pyramid}
-\label{sec-p3}
+\chapter{Parameter-less Population Pyramid}
+\label{chap-p3}
 \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid (P3) as a method for
@@
 between the populations.
-
-\section{Problem Descriptions}
-\label{sec-problems}
-
-\subsection{Single Instance Problems}
-Understanding how a stochastic search algorithm will behave on arbitrary and complex
-search landscapes can be exceedingly difficult. Therefore a common practice for
-algorithm understanding is to perform search on well defined, well understood
-landscapes. To be of interest these landscapes need to represent interesting
-and important aspects of real world problems.
-
-Perhaps the most common such landscape is the Deceptive Trap problem \citep{goldberg:1991:gasize}.
-In this landscape the genome is broken up into $k$ bit non-overlapping subproblems referred
-to as traps.
-Each subproblem is scored using Equation~\ref{eq-deceptive-trap}, where $t$ is the number of
-bits in the trap set to 1. The global optimum in each trap is a string of all 1s, while all
-other solutions lead to a local optima of all 0s. This problem tests an algorithm's
-ability to overcome $k$ sized deception and is commonly used to determine how effective crossover
-is at preserving building blocks. Any crossover event that mixes bits from different parents in
-the same trap will likely result in that trap being optimized to the local optima. For our experiments
-we chose $k=7$ to ensure highly deceptive traps.
-
-\begin{equation}
-   trap(t) = \left\{
-     \begin{array}{rl}
-       k-1-t, &  t<k\\
-       k,   &  t = k
-     \end{array}
-   \right.
-  \label{eq-deceptive-trap}
-\end{equation}
-
-\cite{goldman:2012:ltga} found that mixing local search with linkage learning rendered the Deceptive
-Trap problem trivial. This is because local search is able to optimize each trap to one of the two
-optima (all 1s or all 0s), and then linkage learning can have perfect knowledge of gene interactions.
-In order to make the problem more difficult, the authors proposed the Deceptive
-Step Trap problem, given in Equation~\ref{eq-deceptive-step-trap}. This function modifies the results
-of Equation~\ref{eq-deceptive-trap} to include plateaus of size $s$, introducing an exponential number
-of local optima in each trap.
-With $k=7$ and $s=2$, as used in our experiments, all traps with 0, 1, 3, 5, and 7 bits set are local optima.
-This means that half of all ways to set the trap are 1 bit local optima.
-More generally, the number of local optima grows at $\Theta(2^{k-1})$.
-As a result, the Deceptive Step Trap is much more challenging for linkage
-learning techniques, while still being highly deceptive.
-\begin{equation}
-   step\_trap(t) = \left \lfloor \frac{(k-s)\pmod{s} + trap(t)}{s} \right \rfloor
-  \label{eq-deceptive-step-trap}
-\end{equation}
-
-Another challenging aspect of landscapes can be higher order relationships. The Hierarchical If
-and only If (HIFF) problem~\citep{watson:1998:hiff} is designed to capture the difficulties of this class of problem.
-In HIFF the genome is broken up into a complete binary tree, such that each gene appears in exactly one
-leaf and each internal node is the subset of genes contained in its children. If all genes represented
-in a node of the tree are set to the same value, they score equal to the size of the set. In this way
-small subsets lead toward solutions to larger subsets. However, a node can score if all genes are either
-all 1s or all 0s, meaning that to solve higher order subproblems it is necessary to perform crossovers
-that preserve lower order solutions. This problem is a natural fit for LTGA as the linkage tree can
-perfectly duplicate the problem's true relationships \citep{thierens:2013:ltgahiff}.
-
-As a final class of well known problems, we have chosen to borrow the Rastrigin problem from real valued
-optimization. This problem's landscape, determined by Equation~\ref{eq-rast}, is highly multimodal caused by
-the oscillating cosine function. \cite{goldman:2014:p3} proposed the Discretized Rastrigin problem, such
-that each floating point $x_i$ in Equation~\ref{eq-rast} is encoded using a 10 bit gray code.
-\begin{equation}
-  10n + \sum_{i=1}^{n}\left [ x_i^2-10\cos (2\pi x_i) \right ] \forall x\in [-5.12,5.12]
-  \label{eq-rast}
-\end{equation}
-
-\subsection{Randomly Generated Problem Classes}
-While well defined landscapes can provide specific insights into how an algorithm works,
-their static nature can be misleading. Specifically, algorithm quality might
-be very fragile such that it is only effective at searching well behaved landscapes. A more realistic
-test of an algorithm's black box effectiveness is to work with randomly generated instances
-drawn from a problem class. When tested over a sufficiently large sample it is then possible
-to draw more general conclusions about the algorithms effectiveness. The challenge with these
-landscapes is determining the global optimum to gauge if an algorithm was successful.
-
-Perhaps the most common model for generating random rugged landscapes is the NK model. An
-NK Landscape determines the fitness of each gene based on epistatic relationships with $K$
-other genes in the genome. This fitness is specified using a randomly generated table of
-fitness values, were each possible combination of the $K+1$ genes is mapped to some floating
-point value $[0-1]$. In unrestricted NK landscapes the relationships between genes are also
-randomly chosen and as a result finding the global optimum is $NP$-Hard for $K>1$. However,
-if epistasis is set such that each gene depends on the $K$ directly following it
-in the genome, the solution can be found in polynomial time~\citep{wright:2000:solvingnk}.
-These Nearest Neighbor NK landscapes are therefore ideal for search algorithm testing.
-For all of our experiments we fixed $K=5$ to ensure highly rugged landscapes.
-
-\cite{saul:1994:spinglass} presents a combinatorial benchmark problem derived from physics:
-Ising Spin Glasses. A spin glass is defined by a weighted graph of interaction terms between vertices.
-Each gene assigns a value to each vertex, with the fitness calculated by Equation~\ref{eq-ising}.
-In this equation, $E$ is the set of all edges, $e_{ij}$ is the edge weight connecting vertex $i$ to vertex $j$, and
-$x_i$ and $x_j$ are the gene values for vertex $i$ and $j$. Optimal fitness is when this sum is minimized.
-\begin{equation}
-\sum_{e_{ij} \in E} x_ie_{ij}x_j
-  \label{eq-ising}
-\end{equation}
-Similar to NK Landscapes, the general class is $NP$-Hard to optimize, but the $2D\pm J$ subset of
-Ising Spin Glasses can be polynomially solved.\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}
-In this subset the graph is restricted to be a two dimensional torus, edge weights are randomly set
-to either -1 or 1, and vertex values must be -1 or 1.
-
-As our final class of randomly generated problems we chose the Maximum Satisfiability (MAX-SAT) problem.
-Related to the more common 3-SAT problem, a MAX-SAT instance is defined by a set of three term clauses.
-Each term is a randomly chosen variable, which may also be negated. A clause scores if an and only if
-at least one term in the clause evaluates to true. In order to make MAX-SAT instances with a known global
-optimum, \cite{goldman:2014:p3} proposed constructing clauses around a fixed solution. In this way the
-signs of the terms are set to ensure the target solution satisfies the clause. To ensure each
-problem is challenging we chose a clause to variable ratio of $4.27$~\citep{selman:1996:maxsat}.
-
-\section{Comparison Algorithm Parameter Tuning}
-\label{sec-tuning}
-While four of the six algorithms in our experiments do not require any user specified parameters,
-hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
-handicapped, we extensively tuned each using the bisection method \citep{sastry:2001:bisection}
-to determine the optimal population size for each problem size.
-Extended by \cite{goldman:2012:ltga}, this method iteratively
-doubles the population size until some success criteria is met and then performs bisection
-between the lowest successful and highest unsuccessful sizes. In this way the minimum population size
-which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
-of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
-above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
-do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
-As bisection can make infinitely large population sizes, any run which had not found the global
-optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-hboa}
-      \end{centering}
-      \caption{Tuned Population Sizes for hBOA}
-      \label{fig-pop-hboa}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-ltga}
-      \end{centering}
-      \caption{Tuned Population Sizes for LTGA}
-      \label{fig-pop-ltga}
-    \end{subfigure}
-  \end{centering}
-  \caption{Optimal population sizes found using bisection on each size
-           for each problem.}
-  \label{fig-pop-sizes}
-\end{figure}
-
-Figure~\ref{fig-pop-sizes} shows the results from performing bisection. In general hBOA
-required population sizes which were at least an order of magnitude larger than LTGA. Due to
-runtime and memory overhead, finding the optimal value for hBOA was much less tractable than LTGA
-for moderate to large problem sizes. LTGA's population size also grew significantly slower
-than hBOA's as problem size increased, especially on the two Trap problems and Rastrigin.
-Both algorithms were very ineffective on MAX-SAT, neither able to tune to problems sizes over 60 bits.
-This is likely due to the fact that some randomly generated MAX-SAT landscapes are very flat
-and highly deceptive~\citep{rana:1998:gamaxsat}.
-\section{Finding the Global Optimum}
-\label{sec-optimum}
-
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{evals-to-success}
-  \end{center}
-  \caption{Comparison of the median number of evaluations to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-evals-to-success}
-\end{figure}
-
-Figure~\ref{fig-evals-to-success} shows the median number of evaluations required by each of the six
-algorithms to find the global optimum for multiple sizes of each problem. Each data point in
-Figure~\ref{fig-evals-to-success} represents the median of 100 runs, where unsuccessful runs
-are treated as requiring more evaluations than any successful run. If the median run was not successful
-no point is shown.
-Medians are used as the data is not normally distributed, and because it allows for more meaningful comparison
-between techniques with different success rates.
-The maximum problem size used for each problem was set to be the largest size we could
-feasibly determine LTGA's optimal population size. For many larger problems results are not shown
-for hBOA due to the extreme computational cost required to optimally set the population size.
-
-\subsection{Quantitative Comparison}
-Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
-The largest problem size for any problem where P3 was not the most efficient has 49 bits, with P3 achieving the best results on all 92 larger
-configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
-Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
-and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
-on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
-The likelihood that P3 would achieve these pairwise results assuming its median result is actually worse is $p < 10^{-15}$
-according to the binomial test.
-Pairwise comparison of LTGA and P3 on the largest problem size using the Mann-Whitney~U test results in
-$p < 10^{-5}$ for all problems.
-
-\subsection{Local Search}
-The Random Restart Hill Climber and $1+(\lambda, \lambda)$ are both relatively effective on small problem
-sizes. This is especially true for the three randomly generated problem classes. These problems
-may contain relatively few local optima or just be exceptionally difficult for the model based algorithms.
-On Deceptive Trap and Deceptive Step Trap using 4 or less traps, $1+(\lambda, \lambda)$ performs significantly
-better than any other algorithm. We believe this is
-because $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
-flipping entire traps.
-This ability also leads $1+(\lambda, \lambda)$
-to outperform the Random Restart Hill Climber on all problems except Nearest Neighbor NK.
-
-On larger problem sizes, the ability for local search to reach the global optimum quickly diminishes.
-Only on MAX-SAT are these optimizers competitive
-at larger tested problem sizes. However, we believe this is because the largest tested MAX-SAT
-was an order of magnitude smaller than the largest size tested for most other problems. As the problem
-size increases the number of local optima increases exponentially, which explains why Random Restart Hill
-Climbing was unable to scale.
-For larger problems it also becomes increasingly unlikely for $1+(\lambda, \lambda)$ to make the right combination
-of changes required to reach the global optimum. This
-behavior causes high variance in success rate, as evident by the occasional successes on large Deceptive Trap problems.
-
-\subsection{Model Building}
-Only techniques which explicitly built models of gene epistasis were able to solve the largest problem
-instances. On single instance problems LTGA was more effective than hBOA, with hBOA outperforming
-LTGA on Nearest Neighbor NK and Ising Spin Glasses. This may be caused by the differences in modeling method:
-unlike the single instance problems, gene epistasis in the randomly generated
-problem classes cannot be be perfectly represented with a linkage tree.
-
-Considering how different hBOA and LTGA are in performing optimization, it is somewhat surprising how similar
-their results are on HIFF.
-However, both techniques rely on populations large enough to support the diversity required to reach the global optimum
-and to model epistasis. Both techniques also only rebuild models once per generation. As the subproblems of HIFF
-are nested, it is unlikely that either technique can accurately model higher order epistasis before solving
-lower order subproblems. Therefore both methods require one generation per subproblem order.
-
-\subsection{P3}
-Unlike the other model based methods, P3 generally outperforms
-both Random Restart Hill Climber and $1+(\lambda, \lambda)$ even on small problem sizes.
-Unlike the other local search methods, P3 outperforms LTGA and hBOA even on large problem sizes.
-This implies that P3 is gaining the benefits of each, leveraging local search to solve easy
-problems and model building to solve harder ones.
-
-Furthermore, the interaction between these two optimization tools explains some of the reason P3
-outperforms each method alone.
-On Deceptive Trap P3's use of hill climbing ensures all traps are immediately optimized, allowing for perfect
-linkage detection and high quality donation.
-On HIFF local search solves all pairwise subproblems, saving P3 a generation over LTGA and hBOA.
-In comparison P3 is only a slight improvement on Deceptive Step Trap, which is less amenable to local search.
-
-\section{Fitness Over Time}
-\label{sec-overtime}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{fitness-over-time}
-  \end{center}
-  \caption{Compares the median best fitness reached during search for each of the six optimization methods.}
-  \label{fig-fitness-over-time}
-\end{figure}
-
-For some applications, finding the global optimum is less important than finding good solutions quickly. Therefore
-we examine this behavior in Figure~\ref{fig-fitness-over-time}. At regular intervals during optimization
-Figure~\ref{fig-fitness-over-time} shows the median of the best fitnesses found at that point of search across
-100 runs. For each problem we show the largest problem size for which we were able to successfully gather results
-for all six algorithms, but the trends shown are representative of all larger problem sizes. The maximum reporting interval
-is set to include the slowest P3 run to reach the global optimum.
-
-% 181 total
-% 121 has P3 best
-%   9 hc
-%  50 lambdalambda
-%  20 hboa
-%  18 phboa
-%  27 LTGA
-
-\subsection{Quantitative Comparison}
-Of 181 sample points, P3 had the highest median fitness in 121. In pairwise competition, $1+(\lambda, \lambda)$
-was the most likely to outperform P3, doing so on 50 sample points. LTGA, hBOA, and Parameter-less hBOA
-were the next best, outperforming P3 on 27, 20, and 18 sample points, respectively. Random Restart Hill Climbing almost
-never outperformed P3, doing so only 9 times.
-The likelihood that P3 would achieve these pairwise results assuming its median result is actually worse is $p < 10^{-9}$
-according to the binomial test.
-
-\subsection{Local Search}
-Perhaps the most striking result is the quality of $1+(\lambda, \lambda)$. Until quite far into search this
-method performs better than both LTGA and hBOA. Given sufficient evaluations $1+(\lambda, \lambda)$ also outperforms
-Random Restart Hill Climbing on all 7 problems. For brief periods in the middle of search it performs the
-best of all techniques on: Deceptive Trap; Deceptive Step Trap; HIFF; Ising Spin Glass; and MAX-SAT problems.
-$1+(\lambda, \lambda)$'s ability to efficiently incorporate gene modifications of larger
-than one bit allows it to overcome the deception and plateaus in Deceptive Trap and Deceptive Step Trap, solve medium sized subproblems
-in HIFF, flip the signs on multiple adjacent bits in Ising Spin Glass, and cross plateaus in MAX-SAT. However,
-this method is slow in reaching the global optima in many of these problems which causes it to eventually
-be overtaken by the model building techniques.
-
-\subsection{Model Building}
-Both hBOA and LTGA are marked by periods of little improvement followed by rapid improvement.
-In hBOA this is taken to the extreme, with all fitness
-improvement made at the very end of search. In both cases this is caused by model building. Before the model
-is accurate little improvement is made. Once it is accurate, fitness improves dramatically.
-
-At 58\% of the recording intervals hBOA has the worst fitness of any solver. Most of the exceptions
-occur when hBOA is still evaluating its initial population, allowing this random search to temporarily surpass the
-local search methods. After $N$ evaluations, however, hBOA and LTGA both fall behind until their models begin
-to improve.
-Parameter-less hBOA reaches intermediate fitnesses faster than hBOA, doing so on 62\% of intervals, as its models begin
-to optimize earlier than hBOA.  However, this trend is reversed after a sufficient number of evaluations, most clearly
-on Deceptive Step Trap and Ising Spin Glasses, as hBOA's tuned population overtakes Parameter-less hBOA's parallel populations.
-
-On every problem LTGA has five distinct periods: fitness plateau, near instantaneous improvement, fitness plateau,
-and improvement to global optimum. The early period corresponds with
-initialization of the population, with the first fitness gain achieved immediately upon completing the first generation.
-When using an inaccurate model, LTGA's mixing strategy performs a sort of less effective local search.
-Subsequent generations then make only minor fitness improvements. Once the model
-becomes accurate and the probability of a crossover using high quality genetic material increases sufficiently, LTGA
-enters a second period of rapid improvement.
-
-\begin{comment}
-This behavior is easiest to understand on the two Trap problems. LTGA's first generation will push individual traps toward a
-local optima, but it requires more than a single generation to do so completely. Until that occurs, the model is likely behaving
-no better than random as there is little apparent linkage between bits. Once the model begins to identify individual traps, crossover
-can begin to increase the frequency of higher fitness trap settings. This process begins slowly as low fitness local optima are more
-likely to be chosen for donation than high fitness local optima due to their frequency in the population. The process becomes
-self-catalysing as increased frequency of the global optimum trap genes means increased likelihood of the global optimum being spread by crossover.
-\end{comment}
-
-\subsection{P3}
-The integration of hill climbing into P3 makes it strictly better than using hill climbing alone.
-Early in optimization P3 and the Random Restart Hill Climber have effectively identical quality. This is because
-P3 performs the same evaluations as the Hill Climber for the first two restarts. Once P3 begins
-performing crossover it immediately improves over the Hill Climber. In 95\% of intervals P3 had a fitness at
-least as high as Hill Climbing. As such P3 is better than a simple
-hill climber regardless of how long each technique is run and irrespective of how high quality the solution
-found has to be.
-
-Unlike the model based methods, which struggle until model accuracy improves, P3's iterative solution
-integration allows it to improve much more quickly. This behavior exists in most problems, but is easiest to understand on Deceptive Trap.
-On this problem, P3 immediately brings all traps to local optima, equaled only by the Random Restart Hill Climber in quality.
-In comparison LTGA must evaluate the entire population and perform multiple generations to reach similar quality.
-P3 is able to immediately integrate optimal versions of each
-trap into a single individual as they are found by local search, resulting in smoother fitness improvement than LTGA.
-
-\section{Computational Expenses}
-While it is common in evolutionary computation to assume the evaluation function will dominate algorithm
-complexity, in some domains this will not be true. Model based methods are especially likely to violate
-this norm. Therefore, in order to assess P3's quality in solving problems with efficient fitness functions,
-we provide data on both its algorithmic complexity and wall clock time.
-
-\subsection{Operation Counting}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{rebuilds}
-      \end{centering}
-      \caption{Rebuilds}
-      \label{fig-rebuilds}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{donations}
-      \end{centering}
-      \caption{Donations}
-      \label{fig-donations}
-    \end{subfigure}
-  \end{centering}
-  \caption{Estimated computation costs incurred by model rebuilding (Figure~\ref{fig-rebuilds}) and
-           repeated donations (Figure~\ref{fig-donations}) per evaluation as problem size increases.}
-  \label{fig-costs}
-\end{figure}
-
-When discussing the asymptotic complexity of P3 in Section~\ref{sec-p3}, two aspects eluded precise
-analysis: how expensive is model rebuilding and how many gene donations are made. Figure~\ref{fig-costs}
-provides some insight into how often these two aspects of the algorithm are utilized.
-
-Figure~\ref{fig-rebuilds} reports in an algorithmic sense how expensive model rebuilding is for search.
-In order to calculate this value we recorded how many times search rebuilt the model during each run.
-Figure~\ref{fig-rebuilds} shows the estimated ratio of model rebuilding cost ($N^2$ per rebuild) over evaluation cost ($N$ per evaluation).
-If the cost of model building
-scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically constant.
-For Nearest Neighbor NK, Ising Spin Glasses, and Rastrigin this is the case. For both Trap problems
-and HIFF there is slow growth in the ratio. The problem sizes used for MAX-SAT were not sufficient
-to accurately gauge the asymptotic behavior. Together this suggests that while the cost of building the model
-is almost linear per evaluation, it can grow slowly. However, even in the worst case (HIFF) this growth was
-no more than twice the algorithmic cost of an evaluation even using 2048 bits.
-
-When applying a crossover subset, P3 tries random donors from the population until one is found with at least
-one bit different from the improving solution. In theory this can result in up to $O(\mu)$ operations.
-Figure~\ref{fig-donations} examines the observed average number of donations per evaluation performed.
-Ising Spin Glass, HIFF, and Rastrigin all achieve effectively constant behavior here, implying repeated
-donation does not impact the asymptotic runtime of P3. Both Trap functions and Nearest Neighbor NK all
-increase in number of donations as problem size increases, potentially increasing algorithmic costs. An
-important note is that each donation may range in size from a single bit up to $N-1$. However,
-repeated donation attempts are far more likely to happen with smaller clusters.
-As such this may cause some super-linear growth
-in P3, but its unlikely to be very high.
-
-\subsection{Wall Clock Performance}
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{seconds-to-success}
-  \end{center}
-  \caption{Comparison of the median number of seconds to reach the global optimum for
-           the six different optimization methods with respect
-           to problem size.  If the median run did not reach the global optimum no data element
-           is shown.  Results given on a log-log scale.}
-  \label{fig-seconds-to-success}
-\end{figure}
-
-To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}.
-Similar to Figure~\ref{fig-evals-to-success}, each point represents the median
-of 100 runs, with unsuccessful runs treated as slower than successful runs.
-\begin{comment}
-In total these runs
-took over 6 computing years to complete. We were therefore forced to use
-diverse hardware potentially running multiple processes in parallel.
-As such these timings should be considered very skeptically as a number
-of sources may contribute to biased error in the results. However, due to
-the size of the dataset and the general regularity in the results some general conclusions
-can still be made.
-\end{comment}
-These results were collected using 2.5GHz Intel Xeon E5-2670v2 processors.
-
-\subsubsection{Model Building}
-hBOA and Parameter-less hBOA perform much worse when using wall clock time as the unit of comparison than when using
-evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while, under reasonable
-assumptions, P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
-hBOA goes from being slightly more efficient than LTGA in terms of evaluations to three orders of magnitude worse in terms of seconds.
-As P3 and LTGA require a similar asymptotic complexity per
-evaluation as the Hill Climber and $1+(\lambda, \lambda)$, no similar change in ordering occurs.
-
-\subsubsection{P3}
-When LTGA is optimally tuned to a single instance problem with an efficient evaluation function
-it can find the global optimum faster than P3 in terms of wall clock time. However, on randomly
-generated problem classes P3's efficient use of evaluations is enough to overtake LTGA.
-
-On the four single instance problems LTGA not only finds the global optimum using less wall clock time,
-the factor speedup increases as problem length does. Na\"{i}vely this suggests LTGA is achieving
-a lower order of complexity. However, for these experiments LTGA is growing at sub-linear time per
-evaluation, which is not asymptotically stable due to (at minimum) the time required to perform an evaluation.
-We suspect that the true cause is that $N$ is small enough to be overshadowed by lower order polynomial terms.
-For example, LTGA requires $O(N/\mu)$ time per evaluation to rebuild the linkage
-model from the frequency table. As a result, for small $\mu$ model building, and not extracting pairwise
-frequency, can dominate runtime.
-
-When applied to randomly generated problem classes, the differences in P3 and LTGA's evaluation complexity
-dominates runtime complexity. Similar to with Figure~\ref{fig-evals-to-success}, the amount of speedup P3
-achieves over LTGA increases with problem size on Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT.
-
-Across both types of problems we find that P3's time per evaluation grows approximately linearly. As such,
-we conclude that P3 requires asymptotically similar amounts of time per evaluation as the other efficient techniques.
-
-\section{Population Sizing}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{pop-p3}
-      \end{centering}
-      \caption{As problem size increases}
-      \label{fig-pop-p3}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{solutions-stored}
-      \end{centering}
-      \caption{Largest problem size}
-      \label{fig-solutions-stored}
-    \end{subfigure}
-  \end{centering}
-  \caption{The total number of solutions stored by P3 when the global optimum is found.
-           In Figure~\ref{fig-solutions-stored} the red ``+'' indicates LTGA's tuned population size.}
-  \label{fig-p3-storage}
-\end{figure}
-
-A major advantage to P3 is that it does not require the user to set a population size parameter.
-Beyond making P3 easier to apply, this also conveys two additional advantages: diversity scaled
-to initialization and no need to sacrifice intermediate fitness for eventual optimality.
-
-Figure~\ref{fig-pop-p3} shows how the number of total solutions stored in the pyramid changes
-as problem size increases, similar to Figure~\ref{fig-pop-sizes} for hBOA and LTGA's tuned population sizes.
-As expected, the number of concurrently stored solutions increases as problem difficult increases, with the
-exact behavior dependent on the problem landscape. Figure~\ref{fig-solutions-stored} examines how the number
-of solutions stored is distributed on the largest problem sizes. Here we see that the behavior depends
-on the type of problem. On single instance problems P3's stored variance is relatively low, and generally
-higher than optimally tuned LTGA's population size. On randomly generated problem classes P3 has a much
-higher variance in stored solutions, but in general requires smaller sizes than LTGA.
-
-\subsection{Problem Instance versus Problem Class}
-Our procedure for tuning LTGA and hBOA outlined in Section~\ref{sec-tuning} involved finding
-the optimal population size for each class of problem. For real world black box optimization this is
-realistically the best either algorithm could hope for as tuning to a problem
-instance or population initialization involves repeatedly solving the problem being tuned.
-This limitation does not exist in parameter-less methods, which scale their diversity
-based on the problem instance without needing to solve that instance repeatedly.
-
-To achieve high success rates on randomly generated problem classes, LTGA and hBOA must use a population size
-large enough to solve the hardest instances in that class. Therefore these methods will have
-population sizes larger than necessary to solve the easiest instances in the class. Even
-on single instance problems, both methods will require population sizes large enough to
-ensure the worst random initialization is diverse enough to solve the problem, which may be
-much larger than the best random initialization.
-
-\begin{figure}
-  \begin{center}
-  \includegraphicsfit{evals-to-success-boxplot}
-  \end{center}
-  \caption{Distribution of evaluations required to reach the global optimum for
-           P3 and LTGA on the largest size of each problem.}
-  \label{fig-evals-to-success-boxplot}
-\end{figure}
-
-Figure~\ref{fig-evals-to-success-boxplot}
-highlights how this can effect the required number of evaluations to reach the global optimum, showing
-the distribution of results when solving the largest size of each problem. On each problem
-LTGA has a much smaller difference between its best and worst runs. This makes sense as LTGA uses the same
-population size regardless of instance and progresses search generationally. In contrast P3 has a much
-higher split, with many runs finishing very quickly.
-On all problems except Deceptive Step Trap, P3's upper quartile is lower than LTGA's lower quartile.
-Furthermore, on Deceptive Trap, HIFF, and Ising Spin Glasses, P3's worst run is better than LTGA's best run.
-For Nearest Neighbor NK, most of P3's runs finish much faster than the fastest LTGA runs. However, some
-of P3's outliers take approximately as long as LTGA's tuned performance.
-This supports the hypothesis that P3 is able to scale its diversity not just
-to the problem class, but to the problem instance or even problem initialization, something
-wholly infeasible for tuned population sizing to do.
-
-This tuning distinction is also apparent when comparing Parameter-less hBOA with hBOA in Figure~\ref{fig-evals-to-success}.
-While generally performing worse than hBOA, the difference between the two algorithms is smallest
-on randomly generated problem classes. On MAX-SAT, Parameter-less hBOA actually outperformed both hBOA and LTGA,
-likely due to its ability to scale diversity to the problem instance instead of the entire problem class.
-
-\subsection{Fast versus Optimal}
-\label{sec-fast-vs-optimal}
-
-In Section~\ref{sec-overtime} we examined intermediate fitness qualities of LTGA and hBOA when using
-population sizes tuned to reach the global optimum. As a result, both were exceptionally ineffective
-at quickly reaching high quality solutions. This is because unlike P3, these methods have an explicit
-trade off between optimal performance and intermediate performance caused by their population size parameter.
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{small-pop-dst}
-      \end{centering}
-      \caption{Deceptive Step Trap 203}
-      \label{fig-small-pop-dst}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{small-pop-nk}
-      \end{centering}
-      \caption{Nearest Neighbor NK 200}
-      \label{fig-small-pop-nk}
-    \end{subfigure}
-  \end{centering}
-  \caption{Comparison of how reducing LTGA's population size effects
-           the median best fitness reached during search.}
-  \label{fig-small-pop}
-\end{figure}
-
-Figure~\ref{fig-small-pop} examines the effect of population size on LTGA's intermediate fitness by
-reducing LTGA's population size to one tenth of the tuned value. The two problems shown are representative of the behavior of using
-a smaller population size on the other five problems. Reducing the population size caused LTGA to improve earlier but plateau
-at lower fitnesses. This caused LTGA's success rate to drop from 100 to 0 on Deceptive Step Trap and from 98 to 68 on Nearest
-Neighbor NK.
-Even when using a reduced population size for LTGA, P3 still achieved a fitness at
-least as high as LTGA at 80\% of intervals.
-The likelihood that P3 would achieve these pairwise results assuming its median result is actually worse is $p < 10^{-15}$
-according to the binomial test.
-
-\begin{comment}
-It is important to note that hBOA and LTGA were tuned specifically
-to find the global optimum and are not necessarily using the optimal population size for finding intermediate fitnesses. Consider
-that for both reducing the population size would almost certainly improve their fitness for evaluation steps less than their population
-size. To examine the effect of population size we tested LTGA using one tenth of the population size required for reliably finding the
-global optimum, with the results given in Figure~\ref{fig-small-pop}.
-Reducing the population size resulted in LTGA's initial fitness improvement
-occurring earlier, as the first generation is completed much more quickly. On Deceptive Step Trap this means that
-LTGA leaps ahead of P3 due to its ability to overcome the two bit fitness plateaus.
-%
-Across all tested problems using the smaller
-population size reduced the quality of LTGA's first fitness plateau, likely due to missing required diversity to reach higher quality.
-The second period of improvement also comes earlier and reaches a lower quality fitness. On Deceptive Step Trap 0
-runs reach the global optimum, down from 100 successful for the full population size. On Nearest Neighbor NK this drops the number
-of successful runs from 98 to 68. As a result we conclude that while reducing the population size of LTGA may improve its quality
-at early points during optimization, doing so reduces the robustness of the final solution found. This is in contrast to P3
-which balances high quality intermediate fitness without trading away eventual optimality.
-\end{comment}
-
-\section{Inner Workings}
-
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{cross}
-      \end{centering}
-      \caption{Crossover Proportion}
-      \label{fig-cross}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{cross-success}
-      \end{centering}
-      \caption{Crossover Success}
-      \label{fig-cross-success}
-    \end{subfigure}
-  \end{centering}
-  \caption{For each problem Figure~\ref{fig-cross} shows the proportion of P3 evaluations spend on crossovers
-           and Figure~\ref{fig-cross-success} shows the percentage of fitness improving crossover evaluations.}
-\end{figure}
-
-While analysis of optimization speed is useful from a practitioner standpoint, doing so provides very
-little insight into algorithm behavior. To better understand how P3 works in detail we present here
-a look at some internal features specific to P3.
-
-\subsection{Crossover}
-Figure~\ref{fig-cross} shows the proportion of evaluations P3 spends on crossover, as opposed to hill climbing,
-and Figure~\ref{fig-cross-success} shows what percentage of crossover evaluations resulted in a fitness improvement.
-Together these figures provide some insight into the role of crossover within P3. The behavior
-for each is clearly problem dependent and generally asymptoticly stable as problem size increases.
-
-When solving problems where epistasis can be effectively detected and represented by a linkage tree, P3 tends to spend
-less evaluations performing crossover and each crossover is more likely to be successful.
-Deceptive Trap and Rastrigin are the easiest problems to model epistasis, with local search quickly
-reducing pairwise entropy in each. These are also the problems where P3 uses the least evaluations
-on crossover and has the highest success rates for crossover. At the other extreme are Nearest Neighbor
-NK and Ising Spin Glasses, which both have overlapping linkage not representable by a linkage tree.
-These problems have the highest crossover usage and lowest crossover success of any problem except Deceptive
-Step Trap.  While Deceptive Step Trap's epistasis can be accurately modeled by a linkage tree, the
-exponential number of plateaus makes detecting gene linkage very challenging.
-
-\begin{comment}
-P3 performs optimization using a mixture of hill climbing and crossover. Exactly how much time the algorithm
-dedicates to each is not clear from the algorithm description alone. As such, Figure~\ref{fig-cross} shows
-that the relationship between the two is strongly problem dependent but relatively independent of problem size.
-In general, problems which can be more accurately represented by crossover
-required less crossover evaluations. Deceptive Trap and Rastrigin spent the least evaluations on crossover, using
-15\% and 20\% respectively on the largest problem sizes. Both of these problems are perfectly separable and have relatively
-few local optima in each subproblem. HIFF and Ising Spin Glasses use crossover for about 60\% of evaluations. The HIFF structure can
-be perfectly learned by the linkage tree, but the higher order relationships require multiple stages of crossover to optimize.
-Ising Spin Glasses have an overlapping structure, making them impossible to perfectly represent in a single linkage tree, but
-that overlap is relatively minor in comparison with Nearest Neighbor NK, which uses 81\% of evaluations on crossover. Deceptive Step Trap spends the highest
-percentage of evaluations on crossover, 92\%. Unlike all of the other problems, the value taken by an individual bit in Deceptive
-Step Trap has very little meaning. This is because any trap value is either a local optima or 1 bit flip away from being a local optima.
-This has a combined result of meaning it takes very few local search evaluations to find a local optima but very many crossover operations
-to proliferate the global optima. Epistatic relationships are also going to be very weak and even small population sizes are likely
-to mark all clusters as $useful$.
-
-Solutions are only added to the next highest pyramid level if at least one crossover donation resulted in a fitness
-improvement. Figure~\ref{fig-cross-success} examines the percentage of crossover donations that resulted in a fitness
-improvement. The ordering in Figure~\ref{fig-cross-success} is an almost perfect inversion of Figure~\ref{fig-cross},
-implying that the more successful crossover is at achieving fitness improving donations the lower the proportion of evaluations P3 will
-spend performing crossover. P3 achieves an impressive 30\% success rate for crossover donations on the largest Deceptive Trap
-problem sizes. This makes sense as crossover is able to immediately and perfectly identify entire traps such that all donations
-always contain only complete traps. Furthermore, by forcing crossover to attempt donations of genes different from the current solution,
-donations containing a single trap will have a very high success rate, always improving suboptimal traps which are far more frequently
-created by local search that optimal traps. Rastrigin and HIFF have relatively high success rates of 11\% and 7\%, respectively. Again
-this is likely due to accurate modeling and relatively low diversity of building block alternatives.
-Ising Spin Glasses and Deceptive Step Trap both have low success rates near 0.01\%. This follows as both are difficult to model with
-a linkage tree. In comparison to Figure~\ref{fig-cross-success} Deceptive Step Trap performs better than expected for crossover.
-However, the reason for this is that even exceedingly uninformed crossover can make improvements to Deceptive Step Trap as
-flipping two bits in the same trap will move between two local optima. Nearest Neighbor NK had the least successful crossovers,
-with only 0.007\% successful on the largest tested problems. In comparison to Ising Spin Glass, NK's overlap is far more extensive.
-As a result it makes sense that individual crossovers are very unlikely to be fitness improvements.
-\end{comment}
-
-When compared with LTGA, P3's crossover success rates are lower but similar in problem ordering.
-Counterintuitively, the use of hill climbing on the initial population reduces
-P3's crossover success, not because it reduces model quality or the donation pool, but because it is much more challenging to improve
-locally optimal solutions than randomly generated ones. LTGA's crossover benefits from application to unoptimized solutions, which
-makes its aggregate crossover success incomparable to P3's.
-
-Even when crossover success rates are quite low, such as Nearest Neighbor NK's 0.007\% success,
-the results from Section~\ref{sec-optimum} and Section~\ref{sec-overtime} show it's still critical to optimization. Without
-crossover, P3's performance would be identical to the Random Restart Hill Climber, which was unable to solve even moderately sized problems
-and quickly fell behind P3 in intermediate fitness quality. Therefore even infrequently successful crossover donations are
-critical to success. This does, however, suggest a potential avenue for future improvement by using more successful
-modeling and donation algorithms.
-
-\subsection{Pyramid}
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{level-size}
-      \end{centering}
-      \caption{Population Size}
-      \label{fig-level-size}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{level-success}
-      \end{centering}
-      \caption{Crossover Success}
-      \label{fig-level-success}
-    \end{subfigure}
-  \end{centering}
-  \caption{For each problem Figure~\ref{fig-level-size} shows the number of solutions stored in each level of the pyramid
-           and Figure~\ref{fig-level-success} shows the percentage of fitness improving crossover evaluations at each level.}
-\end{figure}
-
-Another feature unique to P3 is the shape and size of the population pyramid
-constructed for each problem. Figure~\ref{fig-level-size} shows the number of solutions stored at
-each level of the pyramid for the largest tested problem sizes. Each point is the
-median size across 100 runs. If a run did not store any solutions at a level it is treated as 0.
-No point is drawn if the median run had 0 solutions stored at that level. While pyramid size
-is effected by problem size, the overall shape is not. As such the behavior shown in Figure~\ref{fig-level-size}
-is representative of that for all other tested problem sizes.
-
-With the exception of the dip in Deceptive Step Trap, all of the pyramids show a monotonic reduction
-in size as the level increases. This is because a solution must be a strict fitness improvement
-over its previous version to be added to a higher level, which becomes less likely
-each time the solution improves. \cite{lobo:2011:dynamicpop} found theoretical evidence
-and \cite{goldman:2011:dynamic-parameters} found empirical evidence that the optimal
-population size decreases each generation of traditional evolutionary search.
-By decreasing in size, P3 implicitly stores more diversity in low levels and focuses
-search on high quality solutions at higher levels. In comparison,
-LTGA and hBOA suboptimally use a fixed population size at each generation.
-
-Figure~\ref{fig-level-success} examines how crossover success changes at different
-levels of the pyramid. At low levels, success gets progressively lower
-as solution quality increases faster than the model's ability to improve solutions.
-At higher levels modeling becomes more accurate and donations contain higher frequencies
-of high quality building blocks, resulting in increased crossover success. The
-highest level of most problems has a low crossover success rate, as solutions
-crossing with that level have already been improved by previous operations to
-the point where the only improvement would be to create the global optimum, which can
-only happen once.
-
-\subsubsection{Deceptive Step Trap}
-The number of solutions stored at the fourth level of Deceptive Step Trap is
-significantly lower than that of the third or fifth levels, breaking the decreasing
-trend of the other six problems. Figure~\ref{fig-level-success} has a similar
-aberration, with crossover success dropping to  0.0004\% before rebounding and
-following the more common trajectory. This behavior exists in all other problem
-sizes tested, with the dips occurring at exactly the same level.
-
-This behavior is rooted in the peculiar nature
-of this landscape.
-After local search, all traps in all solutions have a total
-number of 1 bits equal to either 0, 1, 3, 5, or 7 as these correspond to the local
-optima when using $k=7$ and $s=2$. Crossover is very likely to overcome the
-two bit plateaus, and as a result solutions in the second level generally do not
-contain the lowest fitness local optima (5 bits set) and the third
-level has very few traps set to the next worst local optima (3 bits set). As a result,
-solutions which reach the third level can only be improved by replacing the deceptive
-local optima (0 and 1 bits set) with the global optimum (7 bits set).
-The global optimum is very rare in the population, and with 8 ways to represent local optima
-linkage learning is inaccurate. Therefore it is very unlikely for crossover to be successful,
-meaning few solutions will be added to the fourth level. Solutions that do improve by
-definition must have a higher frequency of optimal trap settings, meaning level four's model
-will be more accurate and donations are more likely to contain optimal trap values.
-Thus the level size and crossover success rates increase after contracting around level four.
-
-\section{Conclusions and Future Work}
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing black box
-optimization. P3's primary innovation is the replacement of the generational model with a pyramid of populations.
-This pyramid is constructed iteratively, with both the number of levels and the number of solutions stored
-at each level growing as search progresses. P3 uses a model based crossover method
-which learns a linkage tree from gene epistasis. Combined with a simple hill climber, P3's design contains
-many synergistic features.
-
-Across a large number of problems and problem sizes P3 required less evaluations to reach the global optimum
-than optimally tuned state-of-the-art competitors. On single instance problems P3's improvement was by a
-constant factor, while for the three randomly generated problem classes P3's improvement increased with problem size. This
-quality extends to intermediate points during evolution, with P3 generally reaching at least as high
-of fitness as the competitive techniques when using the same number of evaluations. While P3 does require
-modeling overhead, the expense of this overhead is approximately linear with respect to genome size. There
-is some evidence that even when compared on wall clock time, P3 performs on par with
-the best comparison techniques. All of these achievements are made without any problem specific parameter
-tuning, making P3 easier to apply to new domains than its two closets competitors in quality.
-
-P3's quality is due to a number of desirable traits. First, mixing local search with model based
-crossover lets search focus on properly mixing high quality solutions. Second, by adding diversity only
-as necessary P3 tends to use the minimal amount of random initialization, unlike other techniques which must
-overcompensate with larger population sizes on single instance problems and consider the worst instance
-when solving problem classes. Third, by heavily exploiting existing diversity before adding more P3 is able
-to reach high quality intermediate fitnesses quickly without prematurely converging. Fourth, the very
-nature of the pyramid's shape allows search to preserve a desirable proportion of diversity at
-each fitness level, similar to a generational model using a decreasing population size.
-
-There are a number of meaningful avenues for future P3 experimentation. Perhaps the most
-pressing for practitioner acceptance is to apply P3 to real world problems and compare
-its results with other black box or even problem specific heuristics. While parameter-less,
-P3 is currently limited to discrete, fixed length genomes evaluated using single objective
-fitness. These limitations can be relaxed with future work to make P3 more widely applicable.
-While asymptotically linear in problem size, P3's modeling techniques and local search methods are likely going
-to be prohibitively expensive for genome sizes in the hundreds of thousands or millions of genes,
-and the inability of the model to capture overlapping linkage may be hindering
-search efficiency. Overcoming these limitations by using a new modeling technique may allow
-the pyramid model even greater flexibility. Similarly, while P3 is able to overcome low
-order deception via linkage learning, the iterative improvement method by which crossovers
-are made may mislead search on landscapes with higher order deception.
-
-However, even without
-these improvements our results show P3 is highly efficient at finding
-global optima on black box problems without any problem specific tuning.
-
-\section{Acknowledgements}
-This material is based in part upon work supported by the National Science Foundation
-under Cooperative Agreement No. DBI-0939454. Any opinions, findings, and conclusions
-or recommendations expressed in this material are those of the author(s) and do not
-necessarily reflect the views of the National Science Foundation.
-
-\small
-
-\bibliographystyle{apalike}
-\bibliography{../main}
-
-\normalsize
-\appendix
-\section{hBOA Simplification}
-\label{sec-appendix}
-To measure the quality of a decision forest, hBOA applies Equation~\ref{eq-hboa}.
-To favor compact
-models, Equation~\ref{eq-hboa} is scaled by Equation~\ref{eq-hboa-stop}, which provides increased cost
-for more total leaves.
-This quality is used for two purposes: comparison of potential changes from the existing model
-and comparison of that change with the existing model. The basis for our simplication
-is to rearrange $p(B)BDe(B) < p(B')BDe(B')$ to be $\frac{p(B)}{p(B')}<\frac{BDe(B')}{BDe(B)}$.
-
-\begin{equation}
-  BDe(B) = \prod_{i=0}^{N}\prod_{l\in L_i} \frac{\Gamma(m'_i(l))}{\Gamma(m_i(l) + m'_i(l))}
-  \prod_{x_i}\frac{\Gamma(m_i(x_i, l) + m'_i(x_i,l))}{\Gamma(m'_i(x_i,l))}
-  \label{eq-hboa}
-\end{equation}
-
-\begin{equation}
-  p(B) = c2^{-0.5(\sum_i|L_i|)log_2\mu}
-  \label{eq-hboa-stop}
-\end{equation}
-
-The outermost product of Equation~\ref{eq-hboa} iterates over all trees in the forest. However, each split
-can modify only one of the trees and therefore the contribution of all others can be canceled. The middle
-product is across all leaves in the tree. Again since only one leaf can be changed, all other terms can
-be canceled. By convention hBOA uses uninformed Bayesian priors of $m'_i(l)= 2$ and $m'_i(x_i, l)=1$ for
-binary alphabets. As $\Gamma(a) = (a-1)!$ this means the top term in the middle product and the bottom
-term in the third product reduce to 1. The only remaining terms are then $m_i(l)$ and $m_i(x_i, l)$ which
-represent the number of solutions which reached leaf $l$ and the number of solutions which reached leaf $l$
-with a specific value for $x_i$, respectively.
-
-Equation~\ref{eq-hboa-stop} can also be simplified when doing comparisons. If model $B'$ has exactly one more
-leaf than model $B$ then the ratio $\frac{p(B)}{p(B')}$ simplifies to $2^{0.5 log_2\mu}$ regardless of
-total model size.
-
-The resulting simplifications create Equation~\ref{eq-hboa-final},
-where $B'$ is different from $B$ by exactly 1 split, such that $l$ was split to create $l'$ and $l''$.
-The best split is whichever maximizes its improvement over $B$, which is equal to the right side
-of the inequality. Note that these factorials can still be exceedingly large and
-therefore it is imperative that implementations avoid rounding errors and overflows.
-
-\end{document}

### 2015-04-15 Dissertation/chapter-problems.tex
Copied Comprehensive to Disseration
Major update from journal article polishing

--- 2014-10-15 Comprehensive/chapter-problems.tex
@@
 \label{chap-problems}
-TODO Details here
 
@@
 Another challenging aspect of landscapes can be higher order relationships. The Hierarchical If
-and only If (HIFF) problem is designed to capture the difficulties of this class of problem.
+and only If (HIFF) problem~\cite{watson:1998:hiff} is designed to capture the difficulties of this class of problem.
 In HIFF the genome is broken up into a complete binary tree, such that each gene appears in exactly one
@@
 signs of the terms are set to ensure the target solution satisfies the clause. To ensure each
-problem is challenging we chose a clause to variable ratio of $4.27$.
+problem is challenging we chose a clause to variable ratio of $4.27$~\citep{selman:1996:maxsat}.

### 2015-04-15 Dissertation/chapter-comparison_algorithms.tex
Broken reference

@@
 almost always creates near infinitesimal results, calculating fractions which include factorials of $\mu$
-and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{sec-appendix}, we
+and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{chap-appendix-hboa}, we
 derived a simplified form shown in Equation~\ref{eq-hboa-final}.

### 2015-04-15 Dissertation/chapter-understanding.tex
New chapter stuff

\chapter{Using Gray-Box Methods to understand Black-Box Effectiveness}

\section{Big Valley}

\section{Pyramid Levels}

### 2015-05-12 Dissertation/chapter-benchmarking.tex
Updated hyphens and Big O stuff

@@
 When applying a crossover subset, P3 tries random donors from the population until one is found with at least
-one bit different from the improving solution. In theory this can result in up to $O(\mu)$ operations.
+one bit different from the improving solution. In theory this can result in up to \BigO{\mu} operations.
 Figure~\ref{fig-donations} examines the observed average number of donations per evaluation performed.
@@
 evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while, under reasonable
-assumptions, P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
+assumptions, P3 and LTGA require \BigO{N} time per evaluation. This penalty is most clear on Ising Spin Glass where
 hBOA goes from being slightly more efficient than LTGA in terms of evaluations to three orders of magnitude worse in terms of seconds.
@@
 We suspect that the true cause is that $N$ is small enough to be overshadowed by lower order polynomial terms.
-For example, LTGA requires $O(N/\mu)$ time per evaluation to rebuild the linkage
+For example, LTGA requires \BigO{N/\mu} time per evaluation to rebuild the linkage
 model from the frequency table. As a result, for small $\mu$ model building, and not extracting pairwise

### 2015-05-12 Dissertation/chapter-comparison_algorithms.tex
Updated hyphens and Big O stuff

@@
 
-The hill climber requires an amortized cost of $O(1)$ operations per evaluation. In order to
+The hill climber requires an amortized cost of \BigO{1} operations per evaluation. In order to
 terminate, at least one evaluation must be performed for each of the $N$ bits in the solution.
@@
 to decide the value of $x_i$, is split using the value of $x_j$, $T_j$ can no longer be split using $x_i$.
-As a split creates two new leaves, $O(N)$ new potential splits must also be tested. Equation~\ref{eq-hboa-final}
+As a split creates two new leaves, \BigO{N} new potential splits must also be tested. Equation~\ref{eq-hboa-final}
 parses all solutions which reach a leaf to count gene frequencies, requiring $\mu$ time.
@@
 To calculate Equation~\ref{eq-distance}, Equation~\ref{eq-entropy} is called for each
-gene ($H(c_i)$) and pair of genes ($H(c_i \cup c_j)$). Extracting this information requires $O(\mu N^2)$
+gene ($H(c_i)$) and pair of genes ($H(c_i \cup c_j)$). Extracting this information requires \BigO{\mu N^2}
 time, where $\mu$ is the population size and $N$ is the genome size. The process of converting
-this pairwise frequency information into clusters can be achieved in $O(N^2)$ using the bookkeeping
+this pairwise frequency information into clusters can be achieved in \BigO{N^2} using the bookkeeping
 methods presented by \cite{gronau:2007:upgma}. This cost is performed only once per generation,
-and is then used to perform approximately $O(\mu N)$ crossover evaluations. As a result, the amortized cost of
-LTGA's model building is $O(N)$.
+and is then used to perform approximately \BigO{\mu N} crossover evaluations. As a result, the amortized cost of
+LTGA's model building is \BigO{N}.
 
@@
 would perform exactly $2N-2$ evaluations for each of the $\mu$ solutions in the population. This provides the amortizing evaluations
-required to make clustering only $O(N)$ operations per evaluation. However, by skipping some evaluations, its
+required to make clustering only \BigO{N} operations per evaluation. However, by skipping some evaluations, its
 possible that clustering may be super-linear.

### 2015-05-12 Dissertation/chapter-graybox.tex
Updated hyphens and Big O stuff

--- 2014-10-27 Comprehensive/chapter-graybox.tex
@@
-\chapter{New Problem Domain: Gray Box Optimization}
+\chapter{New Problem Domain: gray-box Optimization}
 \label{chap-graybox}
-All of the work presented so far has focused on the black box optimization
+All of the work presented so far has focused on the black-box optimization
 domain. These problems are characterized by a complete lack of available
 problem specific information. All the algorithm has to perform search
-is a ``black box'' which measures the quality of arbitrary proposed
+is a ``black-box'' which measures the quality of arbitrary proposed
 solutions. Optimizers which are successful in this domain are very general
@@
 
-In between those two domains there is another: gray box optimization. In this
+In between those two domains there is another: gray-box optimization. In this
 domain some features general to multiple problem classes are exploited, beyond
@@
 
-In addition to the black box evaluation function, we shall
+In addition to the black-box evaluation function, we shall
 add in two more exploitable features accessible to search algorithms:
@@
 Furthermore, each time a fitness improving move is made all previously tested moves must be tested again. As a result
-improving a random solution to be a local optimum using this algorithm requires between $\Omega(N^2)$ and $O(IN^2)$, where $I$ is the number
+improving a random solution to be a local optimum using this algorithm requires between $\Omega(N^2)$ and \BigO{IN^2}, where $I$ is the number
 of improving moves. The lower bound is achieved when the while loop in Figure~\ref{fig-hc} runs a constant number of times.
@@
 
-Local search in the gray box domain can be significantly more efficient~\cite{whitley:2013:greedy}. This is due
-to two consequences of the domain: the fitness effect of making a move can be calculated in $O(1)$ time, not $O(N)$,
-and the number of moves which must be reevaluated per improvement is $O(1)$, not $O(N)$. This results in local
-search requiring $O(N+I)$ time. If $I$ is within a constant factor of $N$, this means generating random local
+Local search in the gray-box domain can be significantly more efficient~\cite{whitley:2013:greedy}. This is due
+to two consequences of the domain: the fitness effect of making a move can be calculated in \BigO{1} time, not \BigO{N},
+and the number of moves which must be reevaluated per improvement is \BigO{1}, not \BigO{N}. This results in local
+search requiring \BigO{N+I} time. If $I$ is within a constant factor of $N$, this means generating random local
 optima is at most a constant factor slower than generating random solutions in the search space.
@@
 subfunction before and after making each of $k$ moves which overlap that subfunction. Combined this
-results in less than $cN2b(k)$ operations, which is $O(N)$ assuming $c$ and $b(k)$ do not grow with $N$.
+results in less than $cN2b(k)$ operations, which is \BigO{N} assuming $c$ and $b(k)$ do not grow with $N$.
 Also, this procedure calls $f_s$ only $k+1$ times more than is required to find the fitness of the solution itself.
@@
 it cannot be a fitness improvement and is therefore removed from $open$.  In total this requires less than
-$ck4b(k)$ time, which is $O(1)$ assuming $c$, $k$, and $b(k)$ do not scale with $N$.
+$ck4b(k)$ time, which is \BigO{1} assuming $c$, $k$, and $b(k)$ do not scale with $N$.
 
@@
 is no more than $N + Ick$. As a result the total cost of improving a random solution until it reaches
-a global optimum is only $O(N+I)$.
+a global optimum is only \BigO{N+I}.
 
@@
 in $delta$ can cause moves to change bins. Assuming the total number of bins does not increase with $N$, both steps
-can be done in $O(1)$ time. However, at least for MAX-SAT, there is evidence that when used in combination with subsequent
+can be done in \BigO{1} time. However, at least for MAX-SAT, there is evidence that when used in combination with subsequent
 search heuristics the less greedy first improvement algorithm was more effective.
 
-\subsection{Preliminary Empirical Comparison}
+\section{Efficient Hamming Ball Search}
+Another consequence of the gray-box domain is that increasing the neighborhood of local search becomes
+more tractable~\cite{chicano:2014:ball}.  Instead of improving solutions until no single bit flip
+is a fitness improvement, this can be expanded to no flip of $r$ or less bits can improve the solution.
+
+In a black-box setting, verifying that no $r$ bit flip can improve a solution requires testing all
+$n \choose r$ neighbors. This quickly becomes intractable as $r$ increases. However in the gray
+box domain not all combinations need to be tested. Consider that if two variables do not participate
+in the same subfunction, the relationship between their effects is by definition additive. As such
+there is no way for flipping both together to be a fitness improvement without flipping one of them
+being a fitness improvement. Therefore it is not necessary to try all possible $r$ sized subsets
+of the solution.
+
 \begin{figure}[t]

### 2015-05-12 Dissertation/chapter-p3.tex
Updated hyphens and Big O stuff

@@
 
-In LTGA the cost of rebuilding the model is $O(\mu N^2)$ as it must collect pairwise
+In LTGA the cost of rebuilding the model is \BigO{\mu N^2} as it must collect pairwise
 gene frequency information for all $\mu$ solutions in the population. P3 does not store
 a single population, and it does not have a fixed $\mu$ size for any population. However,
-each time a solution is added to the population, it requires $O(N^2)$ time to update
-the table of pairwise frequencies and another $O(N^2)$ time to rebuild the linkage model.
+each time a solution is added to the population, it requires \BigO{N^2} time to update
+the table of pairwise frequencies and another \BigO{N^2} time to rebuild the linkage model.
 The model is then used immediately to perform up to one evaluation for each of the up
 to $2N-2$ clusters. Just as in LTGA, if no evaluation shortcuts were made, P3
-has an amortized cost of $O(N)$ modeling cost per evaluation. While P3 does
+has an amortized cost of \BigO{N} modeling cost per evaluation. While P3 does
 rebuild the model more frequently per solution in the population, it also performs
@@
 but has an added cost to find these donations. Repeated donations could require as much
-as $O(\mu)$ attempts per evaluation, but experimental evidence suggest that this
+as \BigO{\mu} attempts per evaluation, but experimental evidence suggest that this
 operation actually saves more overhead than it costs by increasing the number of evaluations

### 2015-05-12 Dissertation/chapter-conclusions.tex
Updates from Gray-Box GECCO paper

@@
 global optima on black box problems without any problem specific tuning.
+
+In line with previous work, we have found that HBHC cannot effectively find global optima
+on problems with even moderate epistasis. In general it also obtains almost no improvement
+in fitness after only a few restarts. We designed TUX as a simplistic way of choosing
+restart locations based on previously found local optima. Unlike HBHC alone, TUX
+was able to continue improving given more time, finding global optima on problems three
+times as large.
+
+TUX's effectiveness is likely due to the HBHC acting as a super repair operator for uniform crossover.
+Given a sufficiently large $r$ the HBHC can return sections of the crossover offspring to either
+parents' original version of a given subfunction. The HBHC is elitist meaning there is a bias toward
+returning to the better of the two parent's versions. Furthermore, by being so disruptive,
+uniform crossover potentially allows for the HBHC to also find unrelated improvements.
+
+While TUX improves over plain HBHC, Gray-Box P3 is required to perform truly successful
+global optimization. Gray-Box P3 replaces na\"{i}ve local search with the HBHC and utilizes
+known non-linear relationships instead of statistical linkage learning. On NKq-Landscapes this drastically
+improves search effectiveness. A major source of this improvement is likely how difficult it is
+for Black-Box P3 to learn linkage relationships on these landscapes. Furthermore, Gray-Box P3
+can perform partial reevaluation and efficient hill climbing during the mixing phase.
+
+Gray-Box P3's success is less dramatic on Ising Spin Glasses.
+While it still outperformed all competitors, Black-Box P3 may actually scale better to larger
+problems. One explanation for this deviation is that Ising Spin Glasses require more
+exploration of equal fitness plateaus. For instance in Figure~\ref{fig-solver-over} and Figure~\ref{fig-radius-over}
+there is a significant pause in improvement when Gray-Box P3 reaches the second best fitness in the landscape.
+Nothing in its design suggests that Gray-Box P3 should be more effective at neutral drift. Another potential
+issue is that on these landscapes the importance of each non-linear relationship may be detectably unequal.
+As a result Black-Box linkage learning may better cluster variables which having meaningful
+impact on fitness while Gray-Box assumes all are equally important. A useful direction for future work
+would be to explore methods of performing efficient learning on top of the known variable interactions.
+
+Somewhat surprising is the difference in behavior between the polynomially solvable problems
+and Unrestricted NKq. While Black-Box P3 performed very well in the former, it was the least
+successful in the latter. The optimal radius for Gray-Box P3 also shifted from 1 to 2.
+A potential explanation is that in both Ising Spin Glass
+and Nearest Neighbor NKq the number of unique variables $r$
+or fewer steps away from a given variable is significantly lower than the worst case. This explains why on Unrestricted NKq
+even moderately high $r$ values hit our memory limit.
+For Black-Box P3 this may also be causing increased difficulty in linkage learning as variables
+become indirectly dependent on much larger sets. Furthermore, Black-Box
+P3 may be benefiting from an increased rate of duplicate dependencies on Nearest Neighbor NKq
+not present in Unrestricted NKq.
+
+While the inclusion of HBHC into Gray-Box P3 introduces a parameter, it requires
+trivial configuration. In the worst case there may be a handful of $r$ values to test.
+Furthermore, our evidence suggests setting $r=1$ is quite powerful, with higher values
+likely to be only a time shift in quality. This is in contrast to $r$'s role in HBHC,
+where low $r$ values are never expected to reach the same quality has higher $r$ values.
+Therefore we conclude that Gray-Box P3 maintains the out-of-the-box quality of Black-Box
+P3, while drastically improving efficiency for this new domain of problems.

### 2015-05-12 Dissertation/chapter-gray-box-benchmarking.tex
Updates from Gray-Box GECCO paper

--- 2015-04-02 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
@@
-\documentclass{sig-alternate}
-%\linespread{2.5}
-\usepackage{url}
-\usepackage{verbatim}
-\usepackage[noend]{algpseudocode}
-\newcommand{\includegraphicsfit}[1]
-{\includegraphics[width=.9\columnwidth,height=.4\textheight,keepaspectratio]{#1}}
-
-\newcommand{\includegraphicswide}[1]
-{\includegraphics[width=.9\textwidth,height=\textheight,keepaspectratio]{#1}}
-
-\newcommand{\BigO}[1]{$\mathcal{O}{(#1)}$}
-
-\usepackage{bm}
-
-\newfont{\mycrnotice}{ptmr8t at 7pt}
-\newfont{\myconfname}{ptmri8t at 7pt}
-\let\crnotice\mycrnotice%
-\let\confname\myconfname%
-
-\setcopyright{acmcopyright}
-\conferenceinfo{GECCO '15,}{July 11 - 16, 2015, Madrid, Spain}
-\isbn{978-1-4503-3472-3/15/07}\acmPrice{\$15.00}
-
-\doi{http://dx.doi.org/10.1145/2739480.2754775}
-\clubpenalty=10000
-\widowpenalty = 10000
-
-\begin{document}
-\title{Gray-Box Optimization using the Parameter-less Population Pyramid}
-%\subtitle{[Genetic Algorithms Track]}
-
-\numberofauthors{2} %
-%\begin{comment}
-\author{
-% 1st. author
-\alignauthor
-Brian W. Goldman\\
-       \affaddr{BEACON Center for the Study of Evolution in Action}\\
-       \affaddr{Michigan State University, U.S.A.}\\
-       \email{brianwgoldman@acm.org}
-% 2nd. author
-\alignauthor
-William F. Punch\\
-       \affaddr{BEACON Center for the Study of Evolution in Action}\\
-       \affaddr{Michigan State University, U.S.A.}\\
-       \email{punch@msu.edu}
-}
-%\end{comment}
-\begin{comment}
-\author{
-% 1st. author
-\alignauthor
-Anonymous\\
-       \affaddr{Group}\\
-       \affaddr{Group}\\
-       \affaddr{Organization}\\
-       \email{email@site.com}
-% 2nd. author
-\alignauthor
-Anonymous\\
-       \affaddr{Group}\\
-       \affaddr{Group}\\
-       \affaddr{Organization}\\
-       \email{email@site.com}
-}
-%\end{comment}
-
-\maketitle
-\begin{abstract}
-%Maximum 200 Words
-Unlike black-box optimization problems, gray-box optimization problems have known,
-limited, non-linear relationships between variables. Though more restrictive, gray-box problems include many
-real-world applications in network security, computational biology, VLSI design, and statistical physics.
-Leveraging these restrictions, the Hamming-Ball Hill Climber
-(HBHC) can efficiently find high quality local optima. We show how
-1) a simple memetic algorithm in conjunction with HBHC can find \emph{global}
-optima for some gray-box problems and 2) a gray-box version of the Parameter-less
-Population Pyramid (P3), utilizing both the HBHC and the known
-information about variable relationships, outperforms all of the
-examined algorithms. While HBHC's inclusion into P3 adds a parameter,
-we show experimentally it can be fixed to 1 without adversely effecting search.
-
-We provide experimental evidence on NKq-Landscapes and Ising Spin Glasses
-that Gray-Box P3 is effective at finding the global optima even for problems with
-thousands of variables. This capability is complemented by its efficiency, with
-running time and memory usage decreased by up to a linear factor from Black-Box P3.
-On NKq this results in a 375x speedup for problems with at least 1,000 variables.
-%While local search introduces a parameter, we experimentally show this parameter
-%can likely be fixed to 1 without significant loss in search quality.
-\end{abstract}
-
-% A category with the (minimum) three required fields
-%\category{Computing Methodologies}{Artificial Intelligence}{Search Methodologies}
-\category{I.2.8}{Artificial Intelligence}{Problem Solving, Control Methods, and Search}
-%A category including the fourth, optional field follows...
-%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]
-
-\terms{Algorithms, Performance, Experimentation}
-
-\keywords{Local Search; Global Search; Parameter-less}
-
-\newpage
-
-\section{Introduction}
-
-Recent work~\cite{whitley:2013:greedy,chicano:2014:ball} has shown that
-providing local search with more problem specific information than just an evaluation function can
-greatly improve search power and efficiency. These improvements
-require information easily derived for a broad range of interesting NP-Hard
-problems.
-
-Here we extend those benefits to global optimization on binary problems by integrating
-this local search with additional heuristics~\cite{chen:2011:memetic}.
-We introduce a novel algorithm designed to combine local
-and global search as simply as possible. To test the true power of this
-process we also extend the Parameter-less Population Pyramid (P3)~\cite{goldman:2014:p3}, the
-current state-of-the-art in black-box optimization~\cite{goldman:2015:fastp3}. This ``Gray-Box P3''
-utilizes both efficient local search and the additional problem information
-to drastically improve performance and reduce memory usage.
-
-\section{Gray-Box Domain}
-\label{sec-gray-box}
-In black-box optimization, the only information available is a specification
-for what constitutes a valid solution and a function which returns the quality
-of a given solution. While these very simple requirements allow almost any
-optimization task to fit into this domain, many problems have additional knowledge
-which could be used by search algorithms. At the other end of
-the spectrum is white-box optimization, in which everything about the problem
-is known. Search techniques in this domain specialize at the cost of generality.
-Gray-box optimization exists between these two extremes, exposing additional
-information which is available for a variety of problems to be exploited by search.
-
-Here we shall consider the gray-box optimization domain with the following
-characteristics~\cite{chicano:2014:ball}: The function which evaluates the quality of a solution is determined by the summation
-of subfunctions. Each subfunction uses at most $k$ variables in the given solution,
-such that $k$ remains constant as the number of problem variables $N$ increases.
-Each subfunction is independently evaluable and the variables each subfunction uses are known.
-
-While more restrictive than black-box optimization, this domain is still very general.
-MAX-SAT, and by extension all NP-hard problems, can be represented
-as a summation of subfunctions which evaluate each clause~\cite{whitley:2013:greedy}.
-Previous approximation studies of Vertex Cover~\cite{oliveto:2009:vertexcover},
-Set Cover~\cite{yu:2010:setcover}, MAX-CUT~\cite{festa:2002:maxcut}, and
-Ising Spin Glasses~\cite{pelikan:2003:hboaising} all work with problems which fit
-in the gray-box domain. These relate back to problems of real world interest such
-as network security, computational biology, VLSI design, and statistical physics.
-
-In order to obtain efficiency bounds for optimization algorithms, the following
-assertions are also included~\cite{chicano:2014:ball}: The total number of subfunctions grows no more than
-linearly with $N$, which holds true for randomized MAX-SAT, mesh spin glasses, and NK Landscapes.
-Furthermore, each variable participates in at most $c$ subfunctions, which holds true
-for mesh spin glasses and Nearest Neighbor NK. If violated, these assertions do not
-necessarily prevent the gray-box optimization algorithms from solving the problem, but may result
-in slower than predicted performance.
-%TODO Are these necessary?
-\begin{comment}
-We also include the trivial requirements that all problem variables must participate
-in at least one subfunction and that the problem is not separable. Any problem which
-violates these requirements can be restated either using fewer variables or as multiple separate problems.
-\end{comment}
-
-\section{Hamming-Ball Hill Climbing}
-\label{sec-hamming}
-The first algorithm to rigorously exploit the features of the gray-box domain
-described in Section~\ref{sec-gray-box} was \cite{whitley:2013:greedy}. In this study
-a hill climber was developed that found the approximate next-best move in \BigO{1} time given
-\BigO{N} setup.
-This is achieved by first determining the fitness effect of making each move, and then updating
-only effected moves each time a change to the solution is made. Due to the requirements of the domain,
-at most \BigO{ck} moves can be effected, with each requiring at most \BigO{1} subfunction calls
-to update their fitness effects. Combined with move binning, this technique is able to perform both
-first improvement and next-best hill climbing in \BigO{I+N}, where $I$ is the number of improving
-moves required to reach a local optimum. In comparison, the black-box approach for these searches can require
-\BigO{IN^2}.
-
-By again leveraging the properties of this domain, the Hamming-Ball Hill Climber (HBHC)~\cite{chicano:2014:ball} extends this
-method to find solutions that cannot be improved by flipping $r$
-or fewer bits called $r$-bit local optima.
-While na\"ively this would require testing \BigO{N^r} moves, in the gray-box domain only
-\BigO{(3ck)^rN} must be checked. This is due to the bounded relationships between variables.
-If $i$ and $j$ do not share a subfunction, the effect of flipping both variables
-must be equal to the sum of flipping both separately. Therefore, only variables that share
-a subfunction must be flipped together when checking $r=2$. The problem of determining
-which variables must be flipped together is equivalent to finding the connected induced
-subgraphs of $G$ with $r$ or fewer vertices, such that each vertex in $G$ is a variable
-and an edge exists between two vertices if and only if those variables share a subfunction.
-In \cite{chicano:2014:ball} an upper bound is derived for how many moves this could create,
-but no algorithm was given to efficiently discover these moves. This process was not included
-in their timing calculations as it only has to be performed once per problem instance.
-
-Under the assertion that $r \ll N$ and that $r$ does not increase with $N$, the
-Hamming-Ball Hill Climber requires only \BigO{I+N} time to find $r$-bit local optima starting
-from a random solution. As $I$ is likely in \BigO{N} this means finding random $r$-bit local optima
-is no more than a constant amount slower than generating random solutions.
-
-While these techniques make it possible to choose the best move just as efficiently
-as choosing any random improving move, the findings of~\cite{whitley:2013:greedy} suggest that,
-especially when paired with subsequent search heuristics, using random improving moves is more effective.
-In \cite{chicano:2014:ball} this decision was amended slightly to include the requirement that smaller
-distance moves are chosen before larger distance moves.
-
-\subsection{Novel r-order Subgraphs Algorithm}
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{ConnectedInducedSubgraphs}{}
-    \State $closed \leftarrow \emptyset$
-    \State $found \leftarrow [~]$
-    \ForAll{$v \in V$}
-      \State $closed \leftarrow closed \cup \{v\}$
-      \State $found \leftarrow found + $\Call{CISG}{$v$, $\emptyset$, $closed$, $\emptyset$}
-    \EndFor
-    \State \Return $found$
-  \EndProcedure
-  \Procedure{CISG}{$v$, $subset$, $closed$, $open$}
-    \State $subset' \leftarrow subset \cup \{v\}$
-    \State $found \leftarrow [subset']$
-    \If{$|subset'| \geq r$}
-      \Return $found$
-    \EndIf
-    \State $closed\_here \leftarrow \emptyset$
-    \State $open' \leftarrow open \cup adjacent(v)$
-    \ForAll{$v' \in open'$ such that $v' \notin closed$}
-        \State $closed\_here \leftarrow closed\_here \cup \{v'\}$
-        \State $closed \leftarrow closed \cup \{v'\}$
-        \State $recurse \leftarrow $\Call{CISG}{$v'$, $subset'$, $closed$, $open'$}
-        \State $found \leftarrow found + recurse$
-    \EndFor
-    \State $closed \leftarrow closed - closed\_here$
-    \State \Return $found$
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm to recursively find all connected induced subgraphs of size $r$ or fewer.}
-  \label{fig-connected-subgraphs}
-\end{figure}
-
-In order to determine which moves in the hamming-ball must be evaluated,
-we developed \Call{ConnectedInducedSubgraphs}{} given in Figure~\ref{fig-connected-subgraphs}.
-\Call{CISG}{} is a recursive helper function which finds all subgraphs
-which contain a given $subset$ and a given vertex $v$, while excluding
-any other vertices added to $closed$. To find all subgraphs, \Call{CISG}{}
-is called once for each vertex in the graph, such that $closed$ contains
-all previously searched vertices, and $subset=open=\emptyset$. In the initial
-call all desired subgraphs which contain $v$ are found, which is why $v$ remains
-in $closed$ to prevent duplicate subgraphs from being returned.
-
-At each recursive level \Call{CISG}{} expands $open$ to include any vertices
-adjacent to $v$ in the graph. By construction this means that $open'$ contains
-all possible ways of adding a single vertex to the current $subset'$. As each $v'$
-is tested it is temporarily added to $closed$ to prevent recursive calls
-from adding it again to $subset'$.
-
-When applied to the sparse graphs inherent in the gray-box domain, this algorithm
-requires \BigO{r!(ck)^rN} time, which reduces to \BigO{N}. The time spent
-in each call is dominated by the loop over $open'$. In the worst case, $open'$
-increases in size by the full adjacency of $v$, which is bounded by $ck$.
-This creates a worse case complexity for a single top level call of
-$\prod_{i}^{r} ick = r!(ck)^r$. This must be called once for each
-of the $N$ variables resulting in \BigO{r!(ck)^rN}.
-
-\section{Tournament Uniform\\ Crossover: TUX}
-\label{sec-tux}
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Iterate-TUX}{}
-    \State Create random $solution$
-    \State Hamming-Ball Hill Climb $solution$
-    \ForAll{$T_i \in T$}
-      \If{$T_i$ is empty}
-        \State $T_i \leftarrow solution$
-        \State \Return
-      \EndIf
-      \State Cross $solution$ with $T_i$ to create $2^{i+1}$ offspring
-      \State Hamming-Ball Hill Climb each offspring
-      \State $solution \leftarrow$ best of offspring, $solution$, and $T_i$
-      \State $T_i \leftarrow$ empty
-    \EndFor
-    \State Add $solution$ to end of $T$
-  \EndProcedure
-\end{algorithmic}
-  \caption{One iteration of TUX optimization. $T$ is an
-           ordered list of solutions, each position of which could be empty,
-           awaiting a crossover partner.}
-  \label{fig-TUX}
-\end{figure}
-
-Hamming-Ball Hill Climbing is not sufficient to efficiently find the global optima
-on problems with even moderate epistasis~\cite{chicano:2014:ball}. This is because,
-like all random restart hill climbers, it relies on random initialization to fall
-inside the global optima's basin of attraction.
-
-To remedy this limitation, we set out to develop a minimally complex memetic
-algorithm to help increase this probability. Figure~\ref{fig-TUX} presents
-the Tournament Uniform Crossover (TUX) algorithm, which combines simplistic
-selection with equal probability uniform crossover, the most basic unbiased crossover, to generate starting
-solutions likely to be in the global optima's basin of attraction.
-
-Conceptually TUX iteratively builds a structure similar to a single elimination
-bracket for solutions. Each ``match'' in the tournament takes in two candidate solutions,
-produces offspring via uniform crossover, applies hill climbing to each offspring, with
-the ``winner'' being the best of all those solutions. The tournament ``bracket'' is constructed
-iteratively, storing an initially empty list of solutions $T$, such that $|T|$ is equal to the height
-of the tournament. Iterative construction is possible because when a sub-bracket is complete only a single solution
-emerges and solutions only need to be stored until their partner is found.
-TUX is fully elitist but does not prematurely converge. This is because search continuously
-integrates new randomly generated solutions through other parts of the bracket. Whenever the
-top of the current bracket is reached, TUX doubles the size of the virtual tournament.
-
-When crossing solutions at $T_i$, TUX produces $2^{i+1}$ offspring. This relationship
-ensures that in total all levels of the tournament, including random initialization,
-perform the same number of hill climbing steps. It also shifts the focus of search toward
-areas expected to be of higher fitness. The expectation is also that it becomes progressively
-harder to improve solutions the higher up the tournament you advance, so more attempts
-are necessary to create new useful solutions.
-
-The primary advantages of TUX are that it does not introduce any new parameters (though it still
-requires an $r$ for the hill climber) and is relatively simple to implement. Even so
-it allows for learning from previous local optima and as Section~\ref{sec-experiments}
-will show it is quite effective at optimization.
-
-\section{Parameter-less Population\\ Pyramid: P3}
-\subsection{Origins}
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Iterate-P3}{}
-    \State Create random $solution$
-    \State Apply hill climber
-    \If{$solution \notin hashset$}
-      \State Add $solution$ to $P_0$
-      \State Add $solution$ to $hashset$
-    \EndIf
-
-    \ForAll{$P_i \in pyramid$}
-      \State Mix $solution$ with $P_i$
-      \If{$solution$'s fitness has improved}
-        \If{$solution \notin hashset$}
-          \State Add $solution$ to $P_{i+1}$
-          \State Add $solution$ to $hashset$
-        \EndIf
-      \EndIf
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-  \caption{One iteration of P3 optimization. $pyramid$ is an
-           ordered set of populations and $hashset$ is a set
-           of all solutions in $pyramid$.}
-  \label{fig-p3}
-\end{figure}
-
-The Parameter-less Population Pyramid (P3) is a recently introduced
-black-box optimization method which requires no problem specific configuration~\cite{goldman:2014:p3,goldman:2015:fastp3}.
-Unlike previous parameter-less methods it was shown to be more efficient than
-competing state-of-the-art algorithms across a number of problem classes.
-
-P3 combines hill climbing and model building using a novel, iteratively constructed, pyramid
-of populations. Figure~\ref{fig-p3} provides the high level description of how the algorithm
-works. To exploit information from multiple solutions, P3 performs global optimal mixing~\cite{thierens:2011:gomea}.
-Each time a solution is added to a level of the pyramid $P_i$, linkage learning is performed to
-create a collection of variable clusters. These clusters, collectively referred to as a linkage tree,
-are learned from the pairwise variable entropy of $P_i$. When a solution is mixed with $P_i$
-each cluster is used to donate variable values from a random donor in $P_i$ into the solution.
-After each donation the modified solution is evaluated,
-with the change being reverted if solution quality decreased. In this way P3 performs
-multiple evaluations per mixing event. This also means that unlike TUX, which uses information from only two
-solutions to perform crossover, P3 uses information from an entire population.
-Due to this model building, P3 requires \BigO{N} amortized time per evaluation and \BigO{N^2}
-memory usage.
-
-\subsection{Gray-Box Specialization}
-\begin{comment}
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{SubfunctionTree}{}
-    \State $clusters \leftarrow [\{0\}, \{1\}, \{2\}, \dots, \{N-1\}]$
-    \State $cluster\_number \leftarrow [0 .. N-1]$
-    \ForAll{$subfunction \in shuffled(subfunctions)$}
-      \State $to\_merge \leftarrow \emptyset$
-      \ForAll{$b \in subfunction$}
-        \State $to\_merge \leftarrow to\_merge \cup \{cluster\_number[b]\}$
-      \EndFor
-      \If{$|to\_merge| > 1$}
-        \State $new\_cluster \leftarrow \emptyset$
-        \ForAll{$i \in to\_merge$}
-          \State $new\_cluster \leftarrow new\_cluster \cup clusters[i]$
-        \EndFor
-        \ForAll{$b \in new\_cluster$}
-          \State $cluster\_number[b] \leftarrow |clusters|$
-        \EndFor
-        \State $clusters \leftarrow clusters + new\_cluster$
-      \EndIf
-    \EndFor
-    \State Remove first $N~clusters$
-    \State Remove any cluster containing all variables
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm used to convert a list of subfunctions into linkage tree clusters.}
-  \label{fig-sfx-tree}
-\end{figure}
-\end{comment}
-
-P3 represents a natural method for integrating the HBHC into
-a global optimization algorithm as P3 already utilizes local search.
-While HBHC's inclusion adds a parameter, Section~\ref{sec-radius} and
-Section~\ref{sec-over-time} provide evidence that $r$ can be fixed to 1, preserving
-the parameter-less nature of P3.
-Beyond using HBHC, there
-are also a number of ways in which P3 can be made more efficient by leveraging the
-additional information available in the gray-box domain.
-
-First and foremost, linkage learning is no longer necessary. By definition
-the direct non-linear relationships between variables are known. As a result
-P3 no longer needs to store the pairwise frequency information which leads
-the black-box version to require \BigO{N^2} memory. Instead,
-we have developed a method for creating a linkage tree which learns clusters
-from the same dependency graph defined in Section~\ref{sec-hamming}. The goal is for each
-cluster to be a connected induced subgraph, with the size of clusters
-mirroring those produced by the agglomerative linkage learning process normally
-used with P3. To form a single cluster, a random graph search is performed from
-a random starting vertex until a desired number of unique vertices have been explored.
-Cluster sizes are set recursively. For each cluster of size $l>1$
-a cluster of size $a$ and a cluster of size $l-a$ are also created, with $a$ chosen uniformly
-from the range $[1..l-1]$. This recursive process begins by initializing $l=N$ and splitting
-to form the first two clusters.
-
-This linking algorithm has a number of useful properties. First, it creates exactly
-$2N-2$ clusters, distributed in size similarly to the black-box clustering algorithm.
-Performing a random graph search to find $l$ unique vertices requires \BigO{lck} time,
-meaning cluster creation is optimally efficient.
-The cluster splitting process has identical properties as random pivot quicksort,
-meaning the sum of cluster sizes is \BigO{N\log N} in the average case.
-This efficiency allows new clusters to be created before every mixing event,
-unlike Black-Box P3 where they are created only when new solutions are added to the population.
-For simplicity the clusters are shuffled after they are created, not sorted on size like in
-Black-Box P3.
-
-Beyond efficiency, there are good reasons to believe these clusters will be useful
-for search. The closer two variables are in the dependency graph, the more likely
-they are to appear in the same cluster. All variables on average are expected to appear
-in at least one cluster, but variables which are central in the graph will appear in more
-clusters than those on the periphery. The more paths of a given
-length between two variables, the higher the probability of them being in the same cluster.
-Unlike Black-Box P3, this linkage tree does not require clusters to be nested, allowing
-more diversity in the types of clusters appearing in a single tree.
-
-In effect the clusters are sampling moves which the HBHC
-would make if $r \ge l$. For any solution in the search space which is not globally optimal, there
-must exist some move which will improve its quality. However, this move may
-be arbitrarily large and it is intractable to test all possible moves of even moderate size.
-By sampling from all possible large moves, we can maintain tractability while gaining
-a potentially non-zero probability of improvement.
-As clusters are used to donate values between solutions, these
-moves are always in the direction of previously found high quality solutions.
-This assumes that the density of high quality solutions is higher than average between good solutions.
-
-\begin{comment}
-Figure~\ref{fig-sfx-tree} provides an efficient method for creating a linkage
-tree similar to that used in Black-Box P3 given subfunction information.
-Each variable starts in its own cluster. Each subfunction in a random order
-is used to determine how to link existing clusters. If a subfunction overlaps
-multiple top level clusters, all overlapped clusters are merged to create a new
-top level cluster. This process creates a linkage tree similar to that learned
-for P3, with the difference that a node in the tree can have up to $k$ children.
-
-\Call{SubfunctionTree}{} requires \BigO{|cluster|} time to construct each $cluster$
-in the linkage tree. In the gray-box domain the total number of subfunctions is required
-to be $\Theta(N)$ in order to ensure the problem is not separable. As all variables must
-participate in a subfunction, $\Theta(N)$ clusters must be created. Therefore
-the only cost which cannot be amortized over the clusters to \BigO{1} is the creation
-of each new $cluster$, leading to a time of \BigO{|cluster|} per $cluster$.
-
-This method of constructing clusters has a number of desirable properties beyond its
-efficiency. As all non-linear relationships are denoted by subfunctions, mixing
-complete subfunctions between individuals has a high likelihood of preserving quality.
-Each cluster contains at least one entire subfunction, and two
-clusters must differ by at least one entire subfunction. The probability two
-variables appear in the same cluster increases with the number of subfunctions
-they share. The more paths that exist between two variables in the subfunction
-connectivity graph, the more likely those variables will appear in the same cluster.
-\end{comment}
-
-Another efficiency gain possible due to the gray-box domain is that each time a
-donation is made, only the effected part of the solution's fitness needs to be
-recalculated. As a result the number of subfunction evaluations required
-to determine the change in fitness is only \BigO{l}. This also
-allows for Gray-Box P3 to efficiently reapply hill climbing after each
-donation as only effected moves need to be rechecked. As a result, a donation
-and its resulting modifications from hill climbing are kept only if the
-new local optima is at least as fit as the solution before the donation occurred.
-All combined a single donation plus returning to a local optimum requires \BigO{l + I} time,
-while just the donation in Black-Box P3 requires \BigO{N}.
-
-\section{Experiments}
-\label{sec-experiments}
+\chapter{Benchmarking Gray-Box P3}
 To compare these gray-box optimization techniques we have chosen
@@
 
-\subsection{The Effect of Radius}
+\section{The Effect of Radius}
 \label{sec-radius}
@@
   \centering
-  \includegraphicswide{fitness}
+  \includegraphicsfit{fitness}
   \caption{Comparison of how radius effects solution quality at termination. For NKq-Landscapes $N=6,000$ and $K=4$ and
@@
   \centering
-  \includegraphicsfit{p3-seconds}
+  \includegraphicshalf{p3-seconds}
   \caption{Time required for Gray-Box P3 to reach the global optimum of Nearest Neighbor NKq instances with $N=6,000$.}
@@
 
-\subsection{Fitness Over Time}
+\section{Fitness Over Time}
 \label{sec-over-time}
@@
   \centering
-  \includegraphicswide{solver-over}
+  \includegraphicsfit{solver-over}
   \caption{Comparison of solution quality during optimization on a log-log scale for different algorithms. For NKq-Landscapes $N=6,000$ and $K=4$ and
@@
   \centering
-  \includegraphicswide{radius-over}
+  \includegraphicsfit{radius-over}
   \caption{Comparison of Gray-Box P3's solution quality during optimization on a log-log scale for different $r$ values.
@@
 
-\subsection{Scalability}
+\section{Scalability}
 Perhaps the most critical test of an optimization algorithm's quality is
@@
   \centering
-  \includegraphicsfit{length-nn}
+  \includegraphicshalf{length-nn}
   \caption{Comparison of how each algorithm's time required to reach the best fitness found scales with problem size
@@
   \centering
-  \includegraphicsfit{length-is}
+  \includegraphicshalf{length-is}
   \caption{Comparison of how each algorithm's time required to find the best fitness found scales with problem size
@@
   \centering
-  \includegraphicsfit{length-un}
+  \includegraphicshalf{length-un}
   \caption{Relative qualities of each method as problem size increases on Unrestricted NKq
@@
 \end{figure}
-
-\section{Discussion and Conclusions}
-In line with previous work, we have found that HBHC cannot effectively find global optima
-on problems with even moderate epistasis. In general it also obtains almost no improvement
-in fitness after only a few restarts. We designed TUX as a simplistic way of choosing
-restart locations based on previously found local optima. Unlike HBHC alone, TUX
-was able to continue improving given more time, finding global optima on problems three
-times as large.
-
-TUX's effectiveness is likely due to the HBHC acting as a super repair operator for uniform crossover.
-Given a sufficiently large $r$ the HBHC can return sections of the crossover offspring to either
-parents' original version of a given subfunction. The HBHC is elitist meaning there is a bias toward
-returning to the better of the two parent's versions. Furthermore, by being so disruptive,
-uniform crossover potentially allows for the HBHC to also find unrelated improvements.
-
-While TUX improves over plain HBHC, Gray-Box P3 is required to perform truly successful
-global optimization. Gray-Box P3 replaces na\"{i}ve local search with the HBHC and utilizes
-known non-linear relationships instead of statistical linkage learning. On NKq-Landscapes this drastically
-improves search effectiveness. A major source of this improvement is likely how difficult it is
-for Black-Box P3 to learn linkage relationships on these landscapes. Furthermore, Gray-Box P3
-can perform partial reevaluation and efficient hill climbing during the mixing phase.
-
-Gray-Box P3's success is less dramatic on Ising Spin Glasses.
-While it still outperformed all competitors, Black-Box P3 may actually scale better to larger
-problems. One explanation for this deviation is that Ising Spin Glasses require more
-exploration of equal fitness plateaus. For instance in Figure~\ref{fig-solver-over} and Figure~\ref{fig-radius-over}
-there is a significant pause in improvement when Gray-Box P3 reaches the second best fitness in the landscape.
-Nothing in its design suggests that Gray-Box P3 should be more effective at neutral drift. Another potential
-issue is that on these landscapes the importance of each non-linear relationship may be detectably unequal.
-As a result Black-Box linkage learning may better cluster variables which having meaningful
-impact on fitness while Gray-Box assumes all are equally important. A useful direction for future work
-would be to explore methods of performing efficient learning on top of the known variable interactions.
-
-Somewhat surprising is the difference in behavior between the polynomially solvable problems
-and Unrestricted NKq. While Black-Box P3 performed very well in the former, it was the least
-successful in the latter. The optimal radius for Gray-Box P3 also shifted from 1 to 2.
-A potential explanation is that in both Ising Spin Glass
-and Nearest Neighbor NKq the number of unique variables $r$
-or fewer steps away from a given variable is significantly lower than the worst case. This explains why on Unrestricted NKq
-even moderately high $r$ values hit our memory limit.
-For Black-Box P3 this may also be causing increased difficulty in linkage learning as variables
-become indirectly dependent on much larger sets. Furthermore, Black-Box
-P3 may be benefiting from an increased rate of duplicate dependencies on Nearest Neighbor NKq
-not present in Unrestricted NKq.
-
-While the inclusion of HBHC into Gray-Box P3 introduces a parameter, it requires
-trivial configuration. In the worst case there may be a handful of $r$ values to test.
-Furthermore, our evidence suggests setting $r=1$ is quite powerful, with higher values
-likely to be only a time shift in quality. This is in contrast to $r$'s role in HBHC,
-where low $r$ values are never expected to reach the same quality has higher $r$ values.
-Therefore we conclude that Gray-Box P3 maintains the out-of-the-box quality of Black-Box
-P3, while drastically improving efficiency for this new domain of problems.
-
-%
-% The following two commands are all you need in the
-% initial runs of your .tex file to
-% produce the bibliography for the citations in your paper.
-\bibliographystyle{abbrv}
-\bibliography{../main}  % sigproc.bib is the name of the Bibliography in this case
-% You must have a proper ".bib" file
-%  and remember to run:
-% latex bibtex latex latex
-% to resolve all references
-%
-% ACM needs 'a single self-contained file'!
-%
-%APPENDICES are optional
-\balancecolumns
-% That's all folks!
-\end{document}

### 2015-05-12 Dissertation/chapter-gray-box-p3.tex
Updates from Gray-Box GECCO paper

--- 2015-04-02 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
@@
-\documentclass{sig-alternate}
-%\linespread{2.5}
-\usepackage{url}
-\usepackage{verbatim}
-\usepackage[noend]{algpseudocode}
-\newcommand{\includegraphicsfit}[1]
-{\includegraphics[width=.9\columnwidth,height=.4\textheight,keepaspectratio]{#1}}
-
-\newcommand{\includegraphicswide}[1]
-{\includegraphics[width=.9\textwidth,height=\textheight,keepaspectratio]{#1}}
-
-\newcommand{\BigO}[1]{$\mathcal{O}{(#1)}$}
-
-\usepackage{bm}
-
-\newfont{\mycrnotice}{ptmr8t at 7pt}
-\newfont{\myconfname}{ptmri8t at 7pt}
-\let\crnotice\mycrnotice%
-\let\confname\myconfname%
-
-\setcopyright{acmcopyright}
-\conferenceinfo{GECCO '15,}{July 11 - 16, 2015, Madrid, Spain}
-\isbn{978-1-4503-3472-3/15/07}\acmPrice{\$15.00}
-
-\doi{http://dx.doi.org/10.1145/2739480.2754775}
-\clubpenalty=10000
-\widowpenalty = 10000
-
-\begin{document}
-\title{Gray-Box Optimization using the Parameter-less Population Pyramid}
-%\subtitle{[Genetic Algorithms Track]}
-
-\numberofauthors{2} %
-%\begin{comment}
-\author{
-% 1st. author
-\alignauthor
-Brian W. Goldman\\
-       \affaddr{BEACON Center for the Study of Evolution in Action}\\
-       \affaddr{Michigan State University, U.S.A.}\\
-       \email{brianwgoldman@acm.org}
-% 2nd. author
-\alignauthor
-William F. Punch\\
-       \affaddr{BEACON Center for the Study of Evolution in Action}\\
-       \affaddr{Michigan State University, U.S.A.}\\
-       \email{punch@msu.edu}
-}
-%\end{comment}
-\begin{comment}
-\author{
-% 1st. author
-\alignauthor
-Anonymous\\
-       \affaddr{Group}\\
-       \affaddr{Group}\\
-       \affaddr{Organization}\\
-       \email{email@site.com}
-% 2nd. author
-\alignauthor
-Anonymous\\
-       \affaddr{Group}\\
-       \affaddr{Group}\\
-       \affaddr{Organization}\\
-       \email{email@site.com}
-}
-%\end{comment}
-
-\maketitle
-\begin{abstract}
-%Maximum 200 Words
-Unlike black-box optimization problems, gray-box optimization problems have known,
-limited, non-linear relationships between variables. Though more restrictive, gray-box problems include many
-real-world applications in network security, computational biology, VLSI design, and statistical physics.
-Leveraging these restrictions, the Hamming-Ball Hill Climber
-(HBHC) can efficiently find high quality local optima. We show how
-1) a simple memetic algorithm in conjunction with HBHC can find \emph{global}
-optima for some gray-box problems and 2) a gray-box version of the Parameter-less
-Population Pyramid (P3), utilizing both the HBHC and the known
-information about variable relationships, outperforms all of the
-examined algorithms. While HBHC's inclusion into P3 adds a parameter,
-we show experimentally it can be fixed to 1 without adversely effecting search.
-
-We provide experimental evidence on NKq-Landscapes and Ising Spin Glasses
-that Gray-Box P3 is effective at finding the global optima even for problems with
-thousands of variables. This capability is complemented by its efficiency, with
-running time and memory usage decreased by up to a linear factor from Black-Box P3.
-On NKq this results in a 375x speedup for problems with at least 1,000 variables.
-%While local search introduces a parameter, we experimentally show this parameter
-%can likely be fixed to 1 without significant loss in search quality.
-\end{abstract}
-
-% A category with the (minimum) three required fields
-%\category{Computing Methodologies}{Artificial Intelligence}{Search Methodologies}
-\category{I.2.8}{Artificial Intelligence}{Problem Solving, Control Methods, and Search}
-%A category including the fourth, optional field follows...
-%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]
-
-\terms{Algorithms, Performance, Experimentation}
-
-\keywords{Local Search; Global Search; Parameter-less}
-
-\newpage
-
-\section{Introduction}
-
-Recent work~\cite{whitley:2013:greedy,chicano:2014:ball} has shown that
-providing local search with more problem specific information than just an evaluation function can
-greatly improve search power and efficiency. These improvements
-require information easily derived for a broad range of interesting NP-Hard
-problems.
-
-Here we extend those benefits to global optimization on binary problems by integrating
-this local search with additional heuristics~\cite{chen:2011:memetic}.
-We introduce a novel algorithm designed to combine local
-and global search as simply as possible. To test the true power of this
-process we also extend the Parameter-less Population Pyramid (P3)~\cite{goldman:2014:p3}, the
-current state-of-the-art in black-box optimization~\cite{goldman:2015:fastp3}. This ``Gray-Box P3''
-utilizes both efficient local search and the additional problem information
-to drastically improve performance and reduce memory usage.
-
-\section{Gray-Box Domain}
-\label{sec-gray-box}
-In black-box optimization, the only information available is a specification
-for what constitutes a valid solution and a function which returns the quality
-of a given solution. While these very simple requirements allow almost any
-optimization task to fit into this domain, many problems have additional knowledge
-which could be used by search algorithms. At the other end of
-the spectrum is white-box optimization, in which everything about the problem
-is known. Search techniques in this domain specialize at the cost of generality.
-Gray-box optimization exists between these two extremes, exposing additional
-information which is available for a variety of problems to be exploited by search.
-
-Here we shall consider the gray-box optimization domain with the following
-characteristics~\cite{chicano:2014:ball}: The function which evaluates the quality of a solution is determined by the summation
-of subfunctions. Each subfunction uses at most $k$ variables in the given solution,
-such that $k$ remains constant as the number of problem variables $N$ increases.
-Each subfunction is independently evaluable and the variables each subfunction uses are known.
-
-While more restrictive than black-box optimization, this domain is still very general.
-MAX-SAT, and by extension all NP-hard problems, can be represented
-as a summation of subfunctions which evaluate each clause~\cite{whitley:2013:greedy}.
-Previous approximation studies of Vertex Cover~\cite{oliveto:2009:vertexcover},
-Set Cover~\cite{yu:2010:setcover}, MAX-CUT~\cite{festa:2002:maxcut}, and
-Ising Spin Glasses~\cite{pelikan:2003:hboaising} all work with problems which fit
-in the gray-box domain. These relate back to problems of real world interest such
-as network security, computational biology, VLSI design, and statistical physics.
-
-In order to obtain efficiency bounds for optimization algorithms, the following
-assertions are also included~\cite{chicano:2014:ball}: The total number of subfunctions grows no more than
-linearly with $N$, which holds true for randomized MAX-SAT, mesh spin glasses, and NK Landscapes.
-Furthermore, each variable participates in at most $c$ subfunctions, which holds true
-for mesh spin glasses and Nearest Neighbor NK. If violated, these assertions do not
-necessarily prevent the gray-box optimization algorithms from solving the problem, but may result
-in slower than predicted performance.
-%TODO Are these necessary?
-\begin{comment}
-We also include the trivial requirements that all problem variables must participate
-in at least one subfunction and that the problem is not separable. Any problem which
-violates these requirements can be restated either using fewer variables or as multiple separate problems.
-\end{comment}
-
-\section{Hamming-Ball Hill Climbing}
-\label{sec-hamming}
-The first algorithm to rigorously exploit the features of the gray-box domain
-described in Section~\ref{sec-gray-box} was \cite{whitley:2013:greedy}. In this study
-a hill climber was developed that found the approximate next-best move in \BigO{1} time given
-\BigO{N} setup.
-This is achieved by first determining the fitness effect of making each move, and then updating
-only effected moves each time a change to the solution is made. Due to the requirements of the domain,
-at most \BigO{ck} moves can be effected, with each requiring at most \BigO{1} subfunction calls
-to update their fitness effects. Combined with move binning, this technique is able to perform both
-first improvement and next-best hill climbing in \BigO{I+N}, where $I$ is the number of improving
-moves required to reach a local optimum. In comparison, the black-box approach for these searches can require
-\BigO{IN^2}.
-
-By again leveraging the properties of this domain, the Hamming-Ball Hill Climber (HBHC)~\cite{chicano:2014:ball} extends this
-method to find solutions that cannot be improved by flipping $r$
-or fewer bits called $r$-bit local optima.
-While na\"ively this would require testing \BigO{N^r} moves, in the gray-box domain only
-\BigO{(3ck)^rN} must be checked. This is due to the bounded relationships between variables.
-If $i$ and $j$ do not share a subfunction, the effect of flipping both variables
-must be equal to the sum of flipping both separately. Therefore, only variables that share
-a subfunction must be flipped together when checking $r=2$. The problem of determining
-which variables must be flipped together is equivalent to finding the connected induced
-subgraphs of $G$ with $r$ or fewer vertices, such that each vertex in $G$ is a variable
-and an edge exists between two vertices if and only if those variables share a subfunction.
-In \cite{chicano:2014:ball} an upper bound is derived for how many moves this could create,
-but no algorithm was given to efficiently discover these moves. This process was not included
-in their timing calculations as it only has to be performed once per problem instance.
-
-Under the assertion that $r \ll N$ and that $r$ does not increase with $N$, the
-Hamming-Ball Hill Climber requires only \BigO{I+N} time to find $r$-bit local optima starting
-from a random solution. As $I$ is likely in \BigO{N} this means finding random $r$-bit local optima
-is no more than a constant amount slower than generating random solutions.
-
-While these techniques make it possible to choose the best move just as efficiently
-as choosing any random improving move, the findings of~\cite{whitley:2013:greedy} suggest that,
-especially when paired with subsequent search heuristics, using random improving moves is more effective.
-In \cite{chicano:2014:ball} this decision was amended slightly to include the requirement that smaller
-distance moves are chosen before larger distance moves.
-
-\subsection{Novel r-order Subgraphs Algorithm}
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{ConnectedInducedSubgraphs}{}
-    \State $closed \leftarrow \emptyset$
-    \State $found \leftarrow [~]$
-    \ForAll{$v \in V$}
-      \State $closed \leftarrow closed \cup \{v\}$
-      \State $found \leftarrow found + $\Call{CISG}{$v$, $\emptyset$, $closed$, $\emptyset$}
-    \EndFor
-    \State \Return $found$
-  \EndProcedure
-  \Procedure{CISG}{$v$, $subset$, $closed$, $open$}
-    \State $subset' \leftarrow subset \cup \{v\}$
-    \State $found \leftarrow [subset']$
-    \If{$|subset'| \geq r$}
-      \Return $found$
-    \EndIf
-    \State $closed\_here \leftarrow \emptyset$
-    \State $open' \leftarrow open \cup adjacent(v)$
-    \ForAll{$v' \in open'$ such that $v' \notin closed$}
-        \State $closed\_here \leftarrow closed\_here \cup \{v'\}$
-        \State $closed \leftarrow closed \cup \{v'\}$
-        \State $recurse \leftarrow $\Call{CISG}{$v'$, $subset'$, $closed$, $open'$}
-        \State $found \leftarrow found + recurse$
-    \EndFor
-    \State $closed \leftarrow closed - closed\_here$
-    \State \Return $found$
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm to recursively find all connected induced subgraphs of size $r$ or fewer.}
-  \label{fig-connected-subgraphs}
-\end{figure}
-
-In order to determine which moves in the hamming-ball must be evaluated,
-we developed \Call{ConnectedInducedSubgraphs}{} given in Figure~\ref{fig-connected-subgraphs}.
-\Call{CISG}{} is a recursive helper function which finds all subgraphs
-which contain a given $subset$ and a given vertex $v$, while excluding
-any other vertices added to $closed$. To find all subgraphs, \Call{CISG}{}
-is called once for each vertex in the graph, such that $closed$ contains
-all previously searched vertices, and $subset=open=\emptyset$. In the initial
-call all desired subgraphs which contain $v$ are found, which is why $v$ remains
-in $closed$ to prevent duplicate subgraphs from being returned.
-
-At each recursive level \Call{CISG}{} expands $open$ to include any vertices
-adjacent to $v$ in the graph. By construction this means that $open'$ contains
-all possible ways of adding a single vertex to the current $subset'$. As each $v'$
-is tested it is temporarily added to $closed$ to prevent recursive calls
-from adding it again to $subset'$.
-
-When applied to the sparse graphs inherent in the gray-box domain, this algorithm
-requires \BigO{r!(ck)^rN} time, which reduces to \BigO{N}. The time spent
-in each call is dominated by the loop over $open'$. In the worst case, $open'$
-increases in size by the full adjacency of $v$, which is bounded by $ck$.
-This creates a worse case complexity for a single top level call of
-$\prod_{i}^{r} ick = r!(ck)^r$. This must be called once for each
-of the $N$ variables resulting in \BigO{r!(ck)^rN}.
-
-\section{Tournament Uniform\\ Crossover: TUX}
-\label{sec-tux}
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Iterate-TUX}{}
-    \State Create random $solution$
-    \State Hamming-Ball Hill Climb $solution$
-    \ForAll{$T_i \in T$}
-      \If{$T_i$ is empty}
-        \State $T_i \leftarrow solution$
-        \State \Return
-      \EndIf
-      \State Cross $solution$ with $T_i$ to create $2^{i+1}$ offspring
-      \State Hamming-Ball Hill Climb each offspring
-      \State $solution \leftarrow$ best of offspring, $solution$, and $T_i$
-      \State $T_i \leftarrow$ empty
-    \EndFor
-    \State Add $solution$ to end of $T$
-  \EndProcedure
-\end{algorithmic}
-  \caption{One iteration of TUX optimization. $T$ is an
-           ordered list of solutions, each position of which could be empty,
-           awaiting a crossover partner.}
-  \label{fig-TUX}
-\end{figure}
-
-Hamming-Ball Hill Climbing is not sufficient to efficiently find the global optima
-on problems with even moderate epistasis~\cite{chicano:2014:ball}. This is because,
-like all random restart hill climbers, it relies on random initialization to fall
-inside the global optima's basin of attraction.
-
-To remedy this limitation, we set out to develop a minimally complex memetic
-algorithm to help increase this probability. Figure~\ref{fig-TUX} presents
-the Tournament Uniform Crossover (TUX) algorithm, which combines simplistic
-selection with equal probability uniform crossover, the most basic unbiased crossover, to generate starting
-solutions likely to be in the global optima's basin of attraction.
-
-Conceptually TUX iteratively builds a structure similar to a single elimination
-bracket for solutions. Each ``match'' in the tournament takes in two candidate solutions,
-produces offspring via uniform crossover, applies hill climbing to each offspring, with
-the ``winner'' being the best of all those solutions. The tournament ``bracket'' is constructed
-iteratively, storing an initially empty list of solutions $T$, such that $|T|$ is equal to the height
-of the tournament. Iterative construction is possible because when a sub-bracket is complete only a single solution
-emerges and solutions only need to be stored until their partner is found.
-TUX is fully elitist but does not prematurely converge. This is because search continuously
-integrates new randomly generated solutions through other parts of the bracket. Whenever the
-top of the current bracket is reached, TUX doubles the size of the virtual tournament.
-
-When crossing solutions at $T_i$, TUX produces $2^{i+1}$ offspring. This relationship
-ensures that in total all levels of the tournament, including random initialization,
-perform the same number of hill climbing steps. It also shifts the focus of search toward
-areas expected to be of higher fitness. The expectation is also that it becomes progressively
-harder to improve solutions the higher up the tournament you advance, so more attempts
-are necessary to create new useful solutions.
-
-The primary advantages of TUX are that it does not introduce any new parameters (though it still
-requires an $r$ for the hill climber) and is relatively simple to implement. Even so
-it allows for learning from previous local optima and as Section~\ref{sec-experiments}
-will show it is quite effective at optimization.
-
-\section{Parameter-less Population\\ Pyramid: P3}
-\subsection{Origins}
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{Iterate-P3}{}
-    \State Create random $solution$
-    \State Apply hill climber
-    \If{$solution \notin hashset$}
-      \State Add $solution$ to $P_0$
-      \State Add $solution$ to $hashset$
-    \EndIf
-
-    \ForAll{$P_i \in pyramid$}
-      \State Mix $solution$ with $P_i$
-      \If{$solution$'s fitness has improved}
-        \If{$solution \notin hashset$}
-          \State Add $solution$ to $P_{i+1}$
-          \State Add $solution$ to $hashset$
-        \EndIf
-      \EndIf
-    \EndFor
-  \EndProcedure
-\end{algorithmic}
-  \caption{One iteration of P3 optimization. $pyramid$ is an
-           ordered set of populations and $hashset$ is a set
-           of all solutions in $pyramid$.}
-  \label{fig-p3}
-\end{figure}
-
-The Parameter-less Population Pyramid (P3) is a recently introduced
-black-box optimization method which requires no problem specific configuration~\cite{goldman:2014:p3,goldman:2015:fastp3}.
-Unlike previous parameter-less methods it was shown to be more efficient than
-competing state-of-the-art algorithms across a number of problem classes.
-
-P3 combines hill climbing and model building using a novel, iteratively constructed, pyramid
-of populations. Figure~\ref{fig-p3} provides the high level description of how the algorithm
-works. To exploit information from multiple solutions, P3 performs global optimal mixing~\cite{thierens:2011:gomea}.
-Each time a solution is added to a level of the pyramid $P_i$, linkage learning is performed to
-create a collection of variable clusters. These clusters, collectively referred to as a linkage tree,
-are learned from the pairwise variable entropy of $P_i$. When a solution is mixed with $P_i$
-each cluster is used to donate variable values from a random donor in $P_i$ into the solution.
-After each donation the modified solution is evaluated,
-with the change being reverted if solution quality decreased. In this way P3 performs
-multiple evaluations per mixing event. This also means that unlike TUX, which uses information from only two
-solutions to perform crossover, P3 uses information from an entire population.
-Due to this model building, P3 requires \BigO{N} amortized time per evaluation and \BigO{N^2}
-memory usage.
-
-\subsection{Gray-Box Specialization}
-\begin{comment}
-\begin{figure}
-  \begin{algorithmic}[1]
-  \Procedure{SubfunctionTree}{}
-    \State $clusters \leftarrow [\{0\}, \{1\}, \{2\}, \dots, \{N-1\}]$
-    \State $cluster\_number \leftarrow [0 .. N-1]$
-    \ForAll{$subfunction \in shuffled(subfunctions)$}
-      \State $to\_merge \leftarrow \emptyset$
-      \ForAll{$b \in subfunction$}
-        \State $to\_merge \leftarrow to\_merge \cup \{cluster\_number[b]\}$
-      \EndFor
-      \If{$|to\_merge| > 1$}
-        \State $new\_cluster \leftarrow \emptyset$
-        \ForAll{$i \in to\_merge$}
-          \State $new\_cluster \leftarrow new\_cluster \cup clusters[i]$
-        \EndFor
-        \ForAll{$b \in new\_cluster$}
-          \State $cluster\_number[b] \leftarrow |clusters|$
-        \EndFor
-        \State $clusters \leftarrow clusters + new\_cluster$
-      \EndIf
-    \EndFor
-    \State Remove first $N~clusters$
-    \State Remove any cluster containing all variables
-  \EndProcedure
-\end{algorithmic}
-  \caption{Algorithm used to convert a list of subfunctions into linkage tree clusters.}
-  \label{fig-sfx-tree}
-\end{figure}
-\end{comment}
-
+\chapter{Gray-Box P3}
 P3 represents a natural method for integrating the HBHC into
@@
 
-\begin{comment}
-Figure~\ref{fig-sfx-tree} provides an efficient method for creating a linkage
-tree similar to that used in Black-Box P3 given subfunction information.
-Each variable starts in its own cluster. Each subfunction in a random order
-is used to determine how to link existing clusters. If a subfunction overlaps
-multiple top level clusters, all overlapped clusters are merged to create a new
-top level cluster. This process creates a linkage tree similar to that learned
-for P3, with the difference that a node in the tree can have up to $k$ children.
-
-\Call{SubfunctionTree}{} requires \BigO{|cluster|} time to construct each $cluster$
-in the linkage tree. In the gray-box domain the total number of subfunctions is required
-to be $\Theta(N)$ in order to ensure the problem is not separable. As all variables must
-participate in a subfunction, $\Theta(N)$ clusters must be created. Therefore
-the only cost which cannot be amortized over the clusters to \BigO{1} is the creation
-of each new $cluster$, leading to a time of \BigO{|cluster|} per $cluster$.
-
-This method of constructing clusters has a number of desirable properties beyond its
-efficiency. As all non-linear relationships are denoted by subfunctions, mixing
-complete subfunctions between individuals has a high likelihood of preserving quality.
-Each cluster contains at least one entire subfunction, and two
-clusters must differ by at least one entire subfunction. The probability two
-variables appear in the same cluster increases with the number of subfunctions
-they share. The more paths that exist between two variables in the subfunction
-connectivity graph, the more likely those variables will appear in the same cluster.
-\end{comment}
-
 Another efficiency gain possible due to the gray-box domain is that each time a
@@
 while just the donation in Black-Box P3 requires \BigO{N}.
-
-\section{Experiments}
-\label{sec-experiments}
-To compare these gray-box optimization techniques we have chosen
-NKq-Landscapes~\cite{chicano:2014:ball} and Ising Spin Glasses~\cite{saul:1994:spinglass}. NKq-Landscapes create
-a collection of randomly generated problem instances given a
-higher level problem class description. Each instance is described
-by a series of $N$ subfunctions, each corresponding to a variable
-in the solution. This subfunction uses its variable and $K$ other
-variables in the solution to calculate a fitness value. Fitness values
-are represented as a randomly generated lookup table, such that table
-entries are integers in the range $[0..q-1]$. As each subfunction reads
-$K+1$ variables, the table's size and $q$ are set to $2^{K+1}$. The quality
-of a solution is equal to the sum of the values returned by these subfunctions.
-
-In this work we shall consider two methods for choosing the $K$ variables
-each subfunction depends on: Nearest Neighbor NKq and Unrestricted NKq.
-In Nearest Neighbor NKq each variable depends on the $K$ variables which sequentially follow
-it in the solution, with dependencies wrapping around the end of the solution.
-Landscapes of this form can be solved in polynomial
-time~\cite{wright:2000:solvingnk}, allowing comparisons of how quickly each optimization
-algorithm can find the global optima. Nearest Neighbor NKq also ensures that $c=k=K+1$
-and that both $c$ and $k$ do not increase as $N$ increases, meaning the efficiency
-conclusions made in Section~\ref{sec-hamming} are applicable.
-
-Unrestricted NKq landscapes draw the $K$ dependencies at random without replacement. For $K > 1$ it is
-NP-Hard to find the global optimum of these landscapes. This also means that while
-$k$ remains fixed, the maximum number of subfunctions a variable appears in ($c$) can increase
-as $N$ increases. As a result some of the efficiency claims in Section~\ref{sec-hamming}
-may not be applicable.
-
-Ising Spin Glasses are a type of MAX-CUT problem relevant in statistical physics. Each spin glass
-encodes spins (vertices) and their relationships (edges) with the goal of assigning each spin a direction
-which minimizes relationship energy. Just as with NKq-Landscapes, Ising Spin Glasses as a whole
-are NP-Hard, but the $2D\pm J$ subset is polynomially solvable~\cite{saul:1994:spinglass}\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}. In this
-subset the graph is a 2D toroidal grid, with edge weights of $\pm 1$. In gray-box terms,
-problems of this subset have $k=2$ and $c=4$ regardless of $N$.
-%While Ising Spin Glasses are traditionally
-%a minimization problem, here we present them as a maximization problem by negating the normal fitness.
-
-For each problem class tested, we generated 50 instances.
-Extreme problem sizes were chosen to stress each algorithm.
-Each method was run once on each instance, and limited to 3 hours of computation and 4 GB of memory.
-Runs were performed on 2.5GHz Intel Xeon E5-2670v2 processors using the C++11 code available
-from our website.\footnote{\url{https://github.com/brianwgoldman/GrayBoxOptimization}}
-Each time the run achieved a new best fitness we recorded the current amount of processing time used.
-Timing includes the discovery of subgraphs to allow for comparison between different radius values.
-When reporting the ``best'' fitness for an instance we mean the best fitness found by any method before
-the time limit is reached. On all Nearest Neighbor NKq instances the ``best'' fitness
-is also the global optimum, verified using dynamic programming. For Ising Spin Glass the ``best'' fitness
-was the global optimum 44 out of 50 times.
-
-All figures report the median, upper and lower quartiles for either percentage error or seconds to reach the best. A run's percentage error is
-equal to the difference between its fitness and the best, divided by the best. When reporting seconds
-to reach the best fitness, any run which did not find the best fitness is treated as slower than any run that did.
-If the median run was unsuccessful, no data point is drawn.
-
-\subsection{The Effect of Radius}
-\label{sec-radius}
-\begin{figure*}
-  \centering
-  \includegraphicswide{fitness}
-  \caption{Comparison of how radius effects solution quality at termination. For NKq-Landscapes $N=6,000$ and $K=4$ and
-  for Ising Spin Glasses $N=6,084$. Range of radius values limited by memory constraints.}
-  \label{fig-fitness}
-\end{figure*}
-
-The only algorithm parameter in the HBHC, TUX, and Gray-Box P3
-is the radius of the hamming-ball. Therefore our first experiments are designed to determine
-the effect of this parameter on solution quality.
-
-Figure~\ref{fig-fitness} shows the effect on final solution fitness as $r$ increases. As expected
-from~\cite{chicano:2014:ball} the HBHC obtains higher quality as $r$ increases, with the magnitude
-of the improvement decreasing. TUX has a similar relationship and
-outperforms HBHC on all three problems for all $r$ values. Regardless of $r$, Gray-Box P3 outperforms both, with almost
-all $r$ values reaching the same best fitness. For Nearest Neighbor NKq, Gray-Box P3 finds the global
-optimum in every run for $r < 4$, with only a single unsuccessful run at $r=4$.
-
-\begin{figure}
-  \centering
-  \includegraphicsfit{p3-seconds}
-  \caption{Time required for Gray-Box P3 to reach the global optimum of Nearest Neighbor NKq instances with $N=6,000$.}
-  \label{fig-p3-seconds}
-\end{figure}
-
-To further examine the effect of $r$ on Gray-Box-P3, Figure~\ref{fig-p3-seconds} shows the number of
-seconds required to reach the global optimum. Setting $r=1$ was the most efficient configuration for all $K > 1$,
-supporting the trend that Gray-Box P3 works best with small $r$ values. With $K=1$, the landscape is smooth enough
-that with a sufficiently high $r$ the HBHC is able to find the global optimum.
-
-\subsection{Fitness Over Time}
-\label{sec-over-time}
-\begin{figure*}
-  \centering
-  \includegraphicswide{solver-over}
-  \caption{Comparison of solution quality during optimization on a log-log scale for different algorithms. For NKq-Landscapes $N=6,000$ and $K=4$ and
-  for Ising Spin Glasses $N=6,084$. Each algorithm uses its best found $r$ value.}
-  \label{fig-solver-over}
-\end{figure*}
-
-Reaching high quality solutions quickly can sometimes be more important than reaching the global optimum
-eventually. Figure~\ref{fig-solver-over} shows how solution quality progresses during
-search for each algorithm. HBHC and TUX have significant early delays caused by their high $r$ values. Larger
-$r$'s require a large amount of initial partial evaluation before performing hill climbing. After one full iteration
-HBHC effectively stalls, with TUX continuing to improve. Both are eclipsed by Gray-Box P3,
-which quickly descends to the global optimum, outperforming HBHC and TUX at every time point.
-
-Figure~\ref{fig-radius-over} further illustrates the effect of $r$ on Gray-Box P3. On Nearest Neighbor NKq
-and Ising Spin Glasses, increasing $r$ does not change the shape of the curve. Instead the quality reached is simply time shifted,
-such that given more time higher $r$ values will reach the same quality. As a result, for these problems we conclude
-that higher $r$ values simply add more expense for no overall gain.
-On Unrestricted NKq this relationship is less certain, with $r=1$ potentially having a different, and worse, shape than $r>1$.
-However, due to memory and time restrictions it is difficult to know if this trend continues.
-
-\begin{figure*}
-  \centering
-  \includegraphicswide{radius-over}
-  \caption{Comparison of Gray-Box P3's solution quality during optimization on a log-log scale for different $r$ values.
-  For NKq-Landscapes $N=6,000$ and $K=4$ and for Ising Spin Glasses $N=6,084$.}
-  \label{fig-radius-over}
-\end{figure*}
-
-\subsection{Scalability}
-Perhaps the most critical test of an optimization algorithm's quality is
-how it scales as problem difficulty increases. To test this behavior,
-we ran all three algorithms using the optimal $r$ values determined experimentally in Section~\ref{sec-radius},
-varying $N$ from 200 to 10,000 for NKq and 196 to 6,084 for Ising Spin Glass.
-In these plots we also include the black-box version of P3 to show the efficiency
-gains available for using gray-box information.
-
-Figure~\ref{fig-length-nn} and Figure~\ref{fig-length-is} show how long each
-algorithm required to reach the best overall quality found on Nearest Neighbor NKq
-and Ising Spin Glasses, respectively. For Nearest Neighbor NKq the best found
-is the global optimum for all runs of all lengths, while for Ising Spin Glasses
-the best quality found by any method was worse than the global optimum in 7 runs of
-$N=4,096$ and 21 runs of $N=6,084$. The median run of the HBHC was unable to reach the best fitness
-for any problems tested using more than 200 bits. TUX performed somewhat better, reaching
-the best fitness more than half of the time on problem sizes up to $N=800$ and $N=625$ for Nearest Neighbor NKq and Ising
-Spin Glass, respectively. Black-Box P3, which does not utilize partial reevaluation or the HBHC,
-was able to consistently reach the best fitness until it it hit the memory limit on $N=2,000$ for
-NKq and $N=2,916$ for Ising Spin Glass. This limitation is due to Black-Box P3's
-\BigO{N^2} memory requirements.
-
-\begin{figure}
-  \centering
-  \includegraphicsfit{length-nn}
-  \caption{Comparison of how each algorithm's time required to reach the best fitness found scales with problem size
-  on Nearest Neighbor NKq with $K=4$. With $N=1000$ Gray-Box P3 is 375x faster than Black-Box P3.}
-  \label{fig-length-nn}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphicsfit{length-is}
-  \caption{Comparison of how each algorithm's time required to find the best fitness found scales with problem size
-  on Ising Spin Glass. With $N=2025$ Gray-Box P3 is 4.6x faster than Black-Box P3.}
-  \label{fig-length-is}
-\end{figure}
-
-For all sizes of both problems, Gray-Box P3 was the fastest to reach the best fitness. On
-Nearest Neighbor NKq the improvement is very significant, with no alternative finishing within
-two orders of magnitude. On its largest successful problem size, the mean time to completion for
-Black-Box P3 was 375 times slower than Gray-Box P3. Using the Mann-Whitney test to compare their run times
-results in $p < 10^{-16}$. This is especially impressive considering
-previous work has shown Black-Box P3 is faster to reach the global optimum than other leading black-box
-methods~\cite{goldman:2015:fastp3}. Applying regression, we estimate that Black-Box P3's time to global optimum
-on Nearest Neighbor NKq is \BigO{N^{2.75}} while Gray-Box P3's is \BigO{N^{1.98}}.
-
-The results on Ising Spin Glass are similar, with a less extreme difference between
-Black-Box P3 and Gray-Box P3. In general Gray-Box is the fastest technique to
-find the global optimum by an order of magnitude, with Black-Box P3's mean run finishing
-4.6 times slower than Gray-Box on $N=2,025$. Using the Mann-Whitney test to compare their run times
-results in $p < 10^{-14}$. The regression line suggests that while
-Gray-Box P3 scales at \BigO{N^{3.35}}, Black-Box P3 scales at \BigO{N^{3.05}}.
-
-As Unrestricted NKq does not have a known global optimum, and the different algorithms
-rarely found the same best fitness, Figure~\ref{fig-length-un} compares the error
-for each technique at termination. Gray-Box P3 in general finds the best fitness,
-with TUX occasionally performing better. When $N>2,000$, Gray-Box P3 finds better quality
-solutions that all other methods for every instance.  Using the Mann-Whitney test to compare the fitness
-of Gray-Box P3 and TUX when $N=10,000$ results in $p < 10^{-16}$. HBHC and Black-Box P3 only reach similar
-qualities as TUX and Gray-Box P3 when $N=200$, doing so in 4 and 7 runs, respectively. As the problem
-size increase TUX begins to fall behind Gray-Box P3, with the HBHC stabilizing at about 2.5\% worse than
-the best found. Here Black-Box P3 performs worse than the other techniques,
-falling further behind as the problem size increases.
-
-\begin{figure}
-  \centering
-  \includegraphicsfit{length-un}
-  \caption{Relative qualities of each method as problem size increases on Unrestricted NKq
-  with $K=4$.}
-  \label{fig-length-un}
-\end{figure}
-
-\section{Discussion and Conclusions}
-In line with previous work, we have found that HBHC cannot effectively find global optima
-on problems with even moderate epistasis. In general it also obtains almost no improvement
-in fitness after only a few restarts. We designed TUX as a simplistic way of choosing
-restart locations based on previously found local optima. Unlike HBHC alone, TUX
-was able to continue improving given more time, finding global optima on problems three
-times as large.
-
-TUX's effectiveness is likely due to the HBHC acting as a super repair operator for uniform crossover.
-Given a sufficiently large $r$ the HBHC can return sections of the crossover offspring to either
-parents' original version of a given subfunction. The HBHC is elitist meaning there is a bias toward
-returning to the better of the two parent's versions. Furthermore, by being so disruptive,
-uniform crossover potentially allows for the HBHC to also find unrelated improvements.
-
-While TUX improves over plain HBHC, Gray-Box P3 is required to perform truly successful
-global optimization. Gray-Box P3 replaces na\"{i}ve local search with the HBHC and utilizes
-known non-linear relationships instead of statistical linkage learning. On NKq-Landscapes this drastically
-improves search effectiveness. A major source of this improvement is likely how difficult it is
-for Black-Box P3 to learn linkage relationships on these landscapes. Furthermore, Gray-Box P3
-can perform partial reevaluation and efficient hill climbing during the mixing phase.
-
-Gray-Box P3's success is less dramatic on Ising Spin Glasses.
-While it still outperformed all competitors, Black-Box P3 may actually scale better to larger
-problems. One explanation for this deviation is that Ising Spin Glasses require more
-exploration of equal fitness plateaus. For instance in Figure~\ref{fig-solver-over} and Figure~\ref{fig-radius-over}
-there is a significant pause in improvement when Gray-Box P3 reaches the second best fitness in the landscape.
-Nothing in its design suggests that Gray-Box P3 should be more effective at neutral drift. Another potential
-issue is that on these landscapes the importance of each non-linear relationship may be detectably unequal.
-As a result Black-Box linkage learning may better cluster variables which having meaningful
-impact on fitness while Gray-Box assumes all are equally important. A useful direction for future work
-would be to explore methods of performing efficient learning on top of the known variable interactions.
-
-Somewhat surprising is the difference in behavior between the polynomially solvable problems
-and Unrestricted NKq. While Black-Box P3 performed very well in the former, it was the least
-successful in the latter. The optimal radius for Gray-Box P3 also shifted from 1 to 2.
-A potential explanation is that in both Ising Spin Glass
-and Nearest Neighbor NKq the number of unique variables $r$
-or fewer steps away from a given variable is significantly lower than the worst case. This explains why on Unrestricted NKq
-even moderately high $r$ values hit our memory limit.
-For Black-Box P3 this may also be causing increased difficulty in linkage learning as variables
-become indirectly dependent on much larger sets. Furthermore, Black-Box
-P3 may be benefiting from an increased rate of duplicate dependencies on Nearest Neighbor NKq
-not present in Unrestricted NKq.
-
-While the inclusion of HBHC into Gray-Box P3 introduces a parameter, it requires
-trivial configuration. In the worst case there may be a handful of $r$ values to test.
-Furthermore, our evidence suggests setting $r=1$ is quite powerful, with higher values
-likely to be only a time shift in quality. This is in contrast to $r$'s role in HBHC,
-where low $r$ values are never expected to reach the same quality has higher $r$ values.
-Therefore we conclude that Gray-Box P3 maintains the out-of-the-box quality of Black-Box
-P3, while drastically improving efficiency for this new domain of problems.
-
-%
-% The following two commands are all you need in the
-% initial runs of your .tex file to
-% produce the bibliography for the citations in your paper.
-\bibliographystyle{abbrv}
-\bibliography{../main}  % sigproc.bib is the name of the Bibliography in this case
-% You must have a proper ".bib" file
-%  and remember to run:
-% latex bibtex latex latex
-% to resolve all references
-%
-% ACM needs 'a single self-contained file'!
-%
-%APPENDICES are optional
-\balancecolumns
-% That's all folks!
-\end{document}

### 2015-05-12 Dissertation/chapter-graybox.tex
Updates from Gray-Box GECCO paper

@@
-\chapter{New Problem Domain: gray-box Optimization}
+\chapter{New Problem Domain: Gray-Box Optimization}
 \label{chap-graybox}
@@
 
-\begin{figure}[t]
-  \begin{centering}
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{hc-time}
-      \end{centering}
-      \caption{Seconds per local optima.}
-      \label{fig-hc-time}
-    \end{subfigure}%
-    \begin{subfigure}{.5\textwidth}
-      \begin{centering}
-        \includegraphicsfit{hc-radius}
-      \end{centering}
-      \caption{Moves in Hamming Ball}
-      \label{fig-hc-radius}
-    \end{subfigure}
-  \end{centering}
-  \caption{Comparison of performing hill climbing
-           using the black box and gray box domains on Nearest Neighbor NK with $K=5$.
-           Figure~\ref{fig-hc-time} shows the time required to bring a random
-           solution to a local optima, averaged over 900 restarts across 30 instances.
-           Figure~\ref{fig-hc-radius} shows the number of moves which must be
-           tested by hill climbing when $N=500$ as $r$ increases.}
-  \label{fig-hc-compare}
-\end{figure}
-
 Consider the 3-Equal problem.  If there are at least two loci between $x_i$ and $x_j$ they do
@@
   \begin{algorithmic}[1]
-  \Procedure{ConnectedSubgraphs}{$v$, $subset$, $closed$, $open$}
-    \State $new\_subset \leftarrow subset \cup \{v\}$
-    \State $found \leftarrow [new\_subset]$
-    \If{$|new\_subset| \geq r$}
+  \Procedure{ConnectedInducedSubgraphs}{}
+    \State $closed \leftarrow \emptyset$
+    \State $found \leftarrow [~]$
+    \ForAll{$v \in V$}
+      \State $closed \leftarrow closed \cup \{v\}$
+      \State $found \leftarrow found + $\Call{CISG}{$v$, $\emptyset$, $closed$, $\emptyset$}
+    \EndFor
+    \State \Return $found$
+  \EndProcedure
+  \Procedure{CISG}{$v$, $subset$, $closed$, $open$}
+    \State $subset' \leftarrow subset \cup \{v\}$
+    \State $found \leftarrow [subset']$
+    \If{$|subset'| \geq r$}
       \Return $found$
@@
     \State $open' \leftarrow open \cup adjacent(v)$
-    \ForAll{$v' \in open'$}
-      \If{$v' \notin closed$}
+    \ForAll{$v' \in open'$ such that $v' \notin closed$}
         \State $closed\_here \leftarrow closed\_here \cup \{v'\}$
         \State $closed \leftarrow closed \cup \{v'\}$
-        \State $recurse \leftarrow $\Call{ConnectedSubgraphs}{$v'$, $new\_subset$, $closed$, $open'$}
+        \State $recurse \leftarrow $\Call{CISG}{$v'$, $subset'$, $closed$, $open'$}
         \State $found \leftarrow found + recurse$
-      \EndIf
     \EndFor
@@
 \end{algorithmic}
-  \caption{Algorithm to recursively find all connected induced subgraphs of size $r$ or less which
-           contain $subset \cup \{v\}$ and no additional vertices in $closed$.}
+  \caption{Algorithm to recursively find all connected induced subgraphs of size $r$ or fewer.}
   \label{fig-connected-subgraphs}
@@
 
-The problem of finding which subsets must be checked is therefore equivalent to
-finding all induced subgraphs of $r$ or less vertices which are connected. For sparse
-graphs this can be efficiently calculated using \Call{ConnectedSubgraphs}{} outlined
-in Figure~\ref{fig-connected-subgraphs}. To our knowledge this is a novel algorithm
-as previous work did not specify how connected subgraphs were found.
+In order to determine which moves in the hamming-ball must be evaluated,
+we developed \Call{ConnectedInducedSubgraphs}{} given in Figure~\ref{fig-connected-subgraphs}.
+\Call{CISG}{} is a recursive helper function which finds all subgraphs
+which contain a given $subset$ and a given vertex $v$, while excluding
+any other vertices added to $closed$. To find all subgraphs, \Call{CISG}{}
+is called once for each vertex in the graph, such that $closed$ contains
+all previously searched vertices, and $subset=open=\emptyset$. In the initial
+call all desired subgraphs which contain $v$ are found, which is why $v$ remains
+in $closed$ to prevent duplicate subgraphs from being returned.
 
-$subset$ and $open$ are both empty for the first top level call of \Call{ConnectedSubgraphs}{}, and
-$closed$ contains only $v$.
-The result of that call is to find all induced connected subgraphs which contain $v$.
-Subsequent top level calls expand $closed$ to also include their $v$.
-Each recursive level explores how $new\_subset$ can be expanded by including all $v'$ which are adjacent
-to at least one vertex in $new\_subset$. Once all subsets including $new\_subset \cup v'$ have been found,
-$v'$ is closed to prevent duplicates. However, when recursion ends $closed$ is returned to its previous state.
-The maximum recursive depth for this algorithm is $r$ as each level increases $|subset|$ by 1. Each level
-makes at most $|open'|$ recursive calls, with $|open'| \leq |open| + c(k-1)$. Finally there must be a top level
-call with each of the $N$ variables. This creates a loose bound of $O(r!(c(k-1))^rN)$, which is $O(N)$ under the
-assumption that $r$ and $c$ do not increase with $N$. The bound is loose because some portion of recursions
-are prevented by $closed$ and because overlapping adjacency will reduce the rate at which $|open'|$ increases.
+At each recursive level \Call{CISG}{} expands $open$ to include any vertices
+adjacent to $v$ in the graph. By construction this means that $open'$ contains
+all possible ways of adding a single vertex to the current $subset'$. As each $v'$
+is tested it is temporarily added to $closed$ to prevent recursive calls
+from adding it again to $subset'$.
 
-As this algorithm finds all connected subsets in $O(N)$ time for a fixed $r$, the number of moves
-which must be tested to determine if a solution is an $r$-bit local optima is $O(N)$. This means that
-while on Nearest Neighbor NK with $N=8000$, $K=5$, and $r=3$ the black box method would require 85 billion
-checks, the gray box method requires only 248,000. Even when allowing connections to be completely random,
-which results in $c$ increasing with $N$, the gray box method still only requires approximately 375 million checks,
-two and a half orders of magnitude less than assuming a black box.
+When applied to the sparse graphs inherent in the gray-box domain, this algorithm
+requires \BigO{r!(ck)^rN} time, which reduces to \BigO{N}. The time spent
+in each call is dominated by the loop over $open'$. In the worst case, $open'$
+increases in size by the full adjacency of $v$, which is bounded by $ck$.
+This creates a worse case complexity for a single top level call of
+$\prod_{i}^{r} ick = r!(ck)^r$. This must be called once for each
+of the $N$ variables resulting in \BigO{r!(ck)^rN}.
+
+As this algorithm finds all connected subsets in \BigO{N} time for a fixed $r$, the number of moves
+which must be tested to determine if a solution is an $r$-bit local optima is \BigO{N}. This means that
+while on Nearest Neighbor NK with $N=8000$, $K=5$, and $r=3$ the black-box method would require 85 billion
+checks, the gray-box method requires only 248,000. Even when allowing connections to be completely random,
+which results in $c$ increasing with $N$, the gray-box method still only requires approximately 375 million checks,
+two and a half orders of magnitude less than assuming a black-box.
 
@@
 the number of $delta$ values which must be updated each move increases. Still, with $c$, $r$, and $k$ constant
-with respect to $N$, it is possible to find $r$-bit hamming ball local optima in $O(N+I)$ time.
+with respect to $N$, it is possible to find $r$-bit hamming ball local optima in \BigO{N+I} time.
 
-\subsection{Preliminary Empirical Comparison}
-Figure~\ref{fig-hc-radius} shows how the number of moves increases as the radius of the hamming ball increases.
-Even on this relatively small problem ($N=500$) the black box method quickly becomes intractable. The gray box
-method is able to use less moves to find 4-bit local optima than the black box needs to find 2-bit local optima,
-and when solving 6-bit local optima the black box method requires 653,635 times as many moves as gray box. However,
-gray box still increases exponentially as $r$ increases, limiting how large $r$ can grow before becoming intractable.
-This relationship depends heavily on $c$ and how variables are distributed among subfunctions.
+\section{Tournament Uniform Crossover: TUX}
+\label{sec-tux}
+\begin{figure}
+  \begin{algorithmic}[1]
+  \Procedure{Iterate-TUX}{}
+    \State Create random $solution$
+    \State Hamming-Ball Hill Climb $solution$
+    \ForAll{$T_i \in T$}
+      \If{$T_i$ is empty}
+        \State $T_i \leftarrow solution$
+        \State \Return
+      \EndIf
+      \State Cross $solution$ with $T_i$ to create $2^{i+1}$ offspring
+      \State Hamming-Ball Hill Climb each offspring
+      \State $solution \leftarrow$ best of offspring, $solution$, and $T_i$
+      \State $T_i \leftarrow$ empty
+    \EndFor
+    \State Add $solution$ to end of $T$
+  \EndProcedure
+\end{algorithmic}
+  \caption{One iteration of TUX optimization. $T$ is an
+           ordered list of solutions, each position of which could be empty,
+           awaiting a crossover partner.}
+  \label{fig-TUX}
+\end{figure}
 
-\begin{comment}
-\section{Memetic Algorithms}
-~\cite{chen:2011:memetic}
+Hamming-Ball Hill Climbing is not sufficient to efficiently find the global optima
+on problems with even moderate epistasis~\cite{chicano:2014:ball}. This is because,
+like all random restart hill climbers, it relies on random initialization to fall
+inside the global optima's basin of attraction.
 
-Local search, even $r$-bit hamming ball search, is not sufficient to robustly find global optimums on
-highly multi-modal problems. This limitation was shown in Figure~\ref{fig-evals-to-success} with hill climbing
-unable to reach the global optima on almost any large size problem.
-\end{comment}
+To remedy this limitation, we set out to develop a minimally complex memetic
+algorithm to help increase this probability. Figure~\ref{fig-TUX} presents
+the Tournament Uniform Crossover (TUX) algorithm, which combines simplistic
+selection with equal probability uniform crossover, the most basic unbiased crossover, to generate starting
+solutions likely to be in the global optima's basin of attraction.
+
+Conceptually TUX iteratively builds a structure similar to a single elimination
+bracket for solutions. Each ``match'' in the tournament takes in two candidate solutions,
+produces offspring via uniform crossover, applies hill climbing to each offspring, with
+the ``winner'' being the best of all those solutions. The tournament ``bracket'' is constructed
+iteratively, storing an initially empty list of solutions $T$, such that $|T|$ is equal to the height
+of the tournament. Iterative construction is possible because when a sub-bracket is complete only a single solution
+emerges and solutions only need to be stored until their partner is found.
+TUX is fully elitist but does not prematurely converge. This is because search continuously
+integrates new randomly generated solutions through other parts of the bracket. Whenever the
+top of the current bracket is reached, TUX doubles the size of the virtual tournament.
+
+When crossing solutions at $T_i$, TUX produces $2^{i+1}$ offspring. This relationship
+ensures that in total all levels of the tournament, including random initialization,
+perform the same number of hill climbing steps. It also shifts the focus of search toward
+areas expected to be of higher fitness. The expectation is also that it becomes progressively
+harder to improve solutions the higher up the tournament you advance, so more attempts
+are necessary to create new useful solutions.
+
+The primary advantages of TUX are that it does not introduce any new parameters (though it still
+requires an $r$ for the hill climber) and is relatively simple to implement. Even so
+it allows for learning from previous local optima and as Section~\ref{sec-experiments}
+will show it is quite effective at optimization.

### 2015-05-12 Dissertation/chapter-understanding.tex
New section on big valley

@@
-\chapter{Using Gray-Box Methods to understand Black-Box Effectiveness}
+\chapter{Problem Understanding using Gray-Box Methods}
+The ruggedness and high dimensionality of most interesting landscapes makes them very challenging
+to visualize or otherwise understand. However, doing so can be very helpful in
+understanding the difficulty of a problem, and how best to design algorithms to deal
+with those difficulties. Conversely, understanding the characteristics which favor
+existing algorithms can help future researchers choose the algorithm which is
+most likely to perform well on their problem. To further this end, we shall
+explore the landscapes used in previous sections in an attempt to understand
+what about a landscape makes it a good problem for P3 to solve.
 
 \section{Big Valley}
+ In order to understand the global structure of a landscape
+\cite{boese:1994:bigvalley} examined the relationship between local optima.
+This process involved generating thousands of random solutions and then applying
+local search to each. To convert this information into something human interpretable,
+the authors created two dimensional plots: distance from nearest global optima and
+fitness difference from global optima.
+The authors found that for a number of interesting problems a ``Big Valley'' of
+local optima exist, such that the higher a solution's fitness the closer it is
+in representation space to the global optimum. As a result, this suggests that
+by focusing search around known high quality solutions, the likelihood of finding even
+higher quality solutions is increased. This relationship is an underlying assumption
+of all evolutionary computation based search methods, including P3.
+
+We have set out to extend this method of landscape visualization by considering not
+just a sample of local optima, but all local and global optima in the landscape.
+When considered as a black-box, each solution must be enumerated and all of its
+neighbors must be checked for improvements. This enumeration is prohibitively expensive for
+even trivial landscapes, requiring $\Omega(N2^N)$ operations.
+However, by leveraging the gray-box domain, some efficiency improvements can be made.
+
+\subsection{Quickly Finding All Local Optima}
+
+Consider that for gray-box problems, we can determine all improving moves from a given
+solution in \BigO{N} time and that this list can be updated in \BigO{1} time when flipping
+a single bit. As a direct result, checking if each new solution is locally optimal requires
+only \BigO{1} time and the overall enumeration process is reduced to \BigO{2^N}.
+
+Due to the limited non-linearity of the gray-box domain, its possible to exclude large
+parts of the search space without potentially missing any local optima. Consider a fitness
+improving bit flip. By definition the only way this flip can cease to be a fitness improvement
+is if one of the bits it shares a subfunction with is flipped. Therefore, all solutions which
+contain the same values for all of those bits as the current solution cannot be locally optimal.
+
+This knowledge can be effectively exploited by intelligently skipping parts of the enumeration.
+Before starting search we can determine the least significant bit in the solution which effects
+each possible fitness improving move. This is the bit which must be flipped in order to effect
+the quality of that move. As a result, when determine what solution to test next, all steps
+of the enumeration which do not result in the least significant bit of a fitness improve being
+flipped can be skipped. Furthermore, if there are multiple fitness improving moves, all must
+have at least one dependent bit flipped, which is the same as saying the fitness improving move
+with the most-significant least-significant bit must have its least significant bit flipped.
+
+Finding which move's least significant bit needs to be flipped can also be done efficiently via binning.
+Before starting search, each move can be binned based on its least significant bit. Initially
+search must check from the most significant bin down to find the first fitness improving move.
+This is the position from which enumeration must continue. If no move is fitness improving, a local
+optima has been found, and enumeration can continue with the least significant bit in the solution.
+Enumeration progresses by adding a 1 to the found position, with carry over as needed.
+If the carry ever exceeds the most significant bit in the solution, the entire space has been checked and iteration ends.
+
+When performing subsequent checks, not all moves needs to be retested for improvement. Instead, the highest
+bin which must be tested is the location of the most significant bit effected by the previous iteration. This is because
+the previous iteration has verified all higher moves are not fitness improving, and no action performed during
+that iteration can change that. Furthermore, no 1s can exist lower than that position, so bits
+below the least significant bit of the fitness improving move can be ignored.
+
+As a final efficiency, the order in which bits of the solution are considered can be remapped. When a flip is fitness
+improving, the amount of search space which is skipped depends on how significant its least significant bit is. Therefore,
+by rearranging the order to make its least significant bit more significant, more search space can be skipped. We perform
+this remapping in a greedy fashion, such that the flip with the least unmapped dependencies has all of its remaining dependencies
+mapped to the most significant remaining positions.
+
+All together, this can result in significant efficiency improvements for some landscapes. For instance, when applied to
+the OneMax problem (each bit scores one if set to 1, zero otherwise) this method finds all local optima in \BigO{N} time.
+Deceptive Trap, regardless of how the bits are arranged, requires \BigO{2^k2^{N/k}} time to find all $2^{N/k}$ local optima.
+As an example, this means a 60 bit
+Deceptive Trap problem with trap size of 4 requires 524,288 operations when using gray-box, but 1,152,921,504,606,846,976
+with black-box.
+
+As a final note, these methods extend to finding only $r$-bit local optima for all gray-box problems. However,
+the cost trade-off of doing so is somewhat unclear. By searching only for $r$-bit local optima, it is more likely
+that a move will exist which causes part of the landscape to be skipped. Yet the increase in total number of moves
+requires more updating information each time the solution is changed. The exact relationship between the two
+depends on the specific problem.
+
+\subsection{Looking at Problems}
+In comparison to the randomly generated problem instances of Nearest Neighbor NKQ and Ising Spin Glass,
+Deceptive Trap and Deceptive Step Trap have very comprehensible landscapes. Therefore, these problems represent
+a good place to begin understanding what ``big valley'' plots are showing. Figure~\ref{fig-valley-trap} shows
+how the local optima in each of these problems are distributed.
+
+\begin{figure*}
+  \centering
+  \includegraphicsfit{valley-trap}
+  \caption{Location and quality of local optima in comparison to the global optima for Deceptive Step Trap ($r=1$ and $r=2$)
+           and Deceptive Trap with $N=30$ and $k=5$.}
+  \label{fig-valley-trap}
+\end{figure*}
+
+The Deceptive Trap problem, shown in the rightmost panel, has very few local optima, each of which actually
+provide very little deception in terms of ``big valley'' properties. Given any two local optima, the one
+closer to the global optimum in fitness is also always closer to the global optimum in representation space.
+This makes sense as the worst local optimum contains all 0s, and each fitness improvement beyond that involves
+converting an entire trap to 1s, with the global optimum being all 1s. As a result, we should expect P3's method
+of elitist mixing to produce solutions which are progressively more and more similar to the global optimum until
+it is finally found.
+
+Deceptive Step Trap's inclusion of fitness plateaus creates an enormous number of local optima. In total there are 24
+million local optima on this problem, representing 2.2\% of the entire search space. In comparison Deceptive Trap of the
+same size has only 64 local optima. These additional local optima create a somewhat deceptive landscape, as selecting
+on fitness between two local optima can result in an increased genetic distance to the global optimum. In the middle
+panel of Figure~\ref{fig-valley-trap} we show how examining only 2-bit local optima causes Deceptive Step Trap using
+a step size of 2 to effectively revert to Deceptive Trap. This is because the hill climber can overcome all of the plateaus,
+leaving each trap either at the global optimum of all 1s or the local optimal value of all 0s or exactly one 1. Selection then
+becomes a near perfect predictor of distance to the global optimum.
+
+\begin{figure*}
+  \centering
+  \includegraphicsfit{valley-nn}
+  \caption{Location and quality of local optima in comparison to the global optima for a representative Nearest Neighbor NKQ problem
+           with $N=60$ and $k=2$.}
+  \label{fig-valley-nn}
+\end{figure*}
+
+Figure~\ref{fig-valley-nn} shows how the local optima are distributed for Nearest Neighbor NKQ. Unlike the trap problems,
+Nearest Neighbor NKQ exhibits a very traditional big valley shape. Increasing the radius of the local optima significantly
+reduces the total number of local optima, with those optima generally more similar to the global optima in both representation
+space and quality. Together this suggests that Nearest Neighbor NKQ is a good candidate for selection based methods like P3. Also,
+due to the general regularity and frequency of local optima, it may not be necessary to increase the radius in order to quickly find
+the global optimum.
+
+\begin{figure*}
+  \centering
+  \includegraphicsfit{valley-un}
+  \caption{Location and quality of local optima in comparison to the global optima for a representative Unrestricted NKQ problem
+           with $N=60$ and $k=2$.}
+  \label{fig-valley-un}
+\end{figure*}
+
+While visually somewhat similar, Unrestricted NKQ shown in Figure~\ref{fig-valley-nn} suggests selection may be misleading.
+Consider the bulge in the bottom right of $r=1$. If search finds points in that area, small modifications and elitist selection
+will not lead toward the global optimum. At higher radius values this issue becomes more apparent, with optima flattening out away
+from the global optima in representation space. For many of these points, the only way to improve would require flipping over 15 bits
+correctly.
+
+\begin{figure*}
+  \centering
+  \includegraphicsfit{valley-is}
+  \caption{Location and quality of local optima in comparison to the global optima for a representative Ising Spin Glass problem
+           with $N=36$.}
+  \label{fig-valley-is}
+\end{figure*}
+
+The Ising Spin Glass instance shown in Figure~\ref{fig-valley-nn} provide some insight into the stalled behavior of Gray-Box
+P3 when solving that problem. In this landscape there are many local optima which have the second best fitness, but are very
+different from the nearest global optima. From a search perspective this means selection can only get you within a certain distance
+of a global optima, and then it becomes unhelpful. This issue is not improved by increasing the local optima radius, as even with
+$r=4$ there are hundreds of local optima with the second best fitness in the landscape which are over 10 bits different (of 36) from
+the nearest global optima.
+
+\begin{figure*}
+  \centering
+  \includegraphicsfit{valley-sat}
+  \caption{Location and quality of local optima in comparison to the global optima for a representative MAX-SAT problem
+           with $N=36$.}
+  \label{fig-valley-sat}
+\end{figure*}
+
+Similar to Unrestricted NKQ, the MAXSAT instance shown in Figure~\ref{fig-valley-sat} suggests selection can be very misleading.
+There is almost a negative correlation between fitness difference from the global optimum and representation distance from the
+global optimum. As before, this problem is not solved by increasing the radius of the local optima.
+This may explain why methods like LTGA and hBOA, which extensively rely on selection,
+scale so poorly when tested on this problem. P3's comparative success may therefore be the result of the random restart hill climber
+eventually finding a solution in the right area of the search space, with crossover finding the global optimum mostly by chance.
 
 \section{Pyramid Levels}
+One method for understanding how P3 performs search is to examine the types of solutions being stored at each level
+of the pyramid. By comparing each solution with the nearest best found solution, we can create plots similar to the ``big valley''
+ones in the previous section, even for problems too large and complex to find all possible global optima. 
+This also provides a look at how the different levels of the pyramid focus search on different areas of the landscape.
+
+\begin{figure*}
+  \centering
+  \includegraphicsfit{level-dst}
+  \caption{Distribution of local optima stored at each level of Gray-Box P3 in relation to the global optimum on
+           the Deceptive Step Trap problem $N=6000$ and traps of size 5.}
+  \label{fig-level-dst}
+\end{figure*}
+
+Gray-Box P3's progression on Deceptive Step Trap, shown in Figure~\ref{fig-level-dst}, follows directly from our expectations.
+The local optima found using only hill climbing and stored in level 0 are quite poor and very far from the global optima.
+The first application of crossover dramatically improves their quality. However, this improvement is likely just overcoming
+the fitness plateaus, as solutions are actually being moved away from the global optimum. Subsequent levels store solutions
+which are generally closer to the global optimum in both fitness and representation space. This change means that at higher
+levels the frequency of traps being set to the global optima increases, resulting in better model building and better donation quality.
+
+\begin{figure*}
+  \centering
+  \includegraphicsfit{level-nn}
+  \caption{Distribution of local optima stored at each level of Gray-Box P3 in relation to the best found by the run on
+           a Nearest Neighbor NKQ problem $N=6000$ and $K=4$.}
+  \label{fig-level-nn}
+\end{figure*}
+
+Nearest Neighbor NKQ, shown in Figure~\ref{fig-level-nn}, provides a landscape almost as free of higher order deception
+as the Deceptive Step Trap problem.  Shown with a logarithmic y-axis, stored solutions exhibit a near perfect relationship
+between fitness and representational distance. Again, using only hill climbing creates local optima which are significantly
+worse and further away than even
+after a single application of crossover. While increasing the radius of the hill climber improves the quality of those initial
+solutions, even $r=3$ is only able to equal a single crossover application. As before the higher level a solution is stored,
+generally the closer it is to the global optima in both fitness and representational distance.
+
+\begin{figure*}
+  \centering
+  \includegraphicsfit{level-un}
+  \caption{Distribution of local optima stored at each level of Gray-Box P3 in relation to the best found by the run on
+           an Unrestricted NKQ problem $N=6000$ and $K=4$.}
+  \label{fig-level-un}
+\end{figure*}
+
+Figure~\ref{fig-level-un} gives insight into how different Unrestricted NKQ is from Nearest Neighbor NKQ. In this landscape
+Gray-Box P3, regardless of radius, appears to explore deceptive local optima which offer fitness improvements without moving
+closer to the global optima in representation space. In this problem there is a gap of over 1000 bits between the deceptive
+local optima and those likely to lead to the best found solution. An interesting behavior on this problem is that the branch
+between the two groups always seems to begin with a single solution in level 1. This suggests that the solutions found near
+the best found may all be similar due to sharing a common ancestor.
+
+\begin{figure*}
+  \centering
+  \includegraphicsfit{level-is}
+  \caption{Distribution of local optima stored at each level of Gray-Box P3 in relation to the best found by the run on an Ising Spin Glass
+           $N=6084$.}
+  \label{fig-level-is}
+\end{figure*}
+
+Gray-Box P3's progression on Ising Spin Glass, shown in Figure~\ref{fig-level-is}, is much more well behaved than on Unrestricted
+NKQ. As has been typical, each application of crossover results in improved fitness, with the largest gains between the first few levels.
+When approaching higher qualities these improvements also begin to translate into increased representational similarity to the global optimum.
+However, as discussed previously for this problem, we again see evidence for a large number of very diverse solutions with the second best
+fitness. These solutions can be over 1000 bits different from the eventual best found solution. Increasing the radius of the hill climber
+does not seem to significantly overcome this issue.
+
+\begin{figure*}
+  \centering
+  \includegraphicsfit{level-sat}
+  \caption{Distribution of local optima stored at each level of Gray-Box P3 in relation to the best found by the run on
+           a MAX-SAT problem $N=6000$.}
+  \label{fig-level-sat}
+\end{figure*}
+
+Figure~\ref{fig-level-sat} shows that Gray-Box P3 acts more similarly on MAX-SAT to Unrestricted NKQ than any of the other problems. Again there
+appears to be deceptive local optima between 500 and 1000 bits away from the nearest improving solution. Again a single solution appears
+to be the ancestor of all of the solutions near the best found. This suggests that similar to Unrestricted NKQ, Gray-Box P3 will have a
+very hard time finding the global optimum on MAX-SAT.
+
+In general it appears that the problems which have polynomial time solutions (Deceptive Step Trap, Nearest Neighbor NKQ, 2D Ising Spin Glass)
+share a similar behavior of selection leading to the global optimum. However, the NP-Hard problems (Unrestricted NKQ, MAX-SAT) seem to contain
+very large amounts of higher order deception, making them challenging for selection based methods. This relationship warrants further investigation
+into other NP-Hard problems to determine if the repeated selection found in P3 and other evolutionary computation methods are able to overcome
+this deception. One potential avenue for improvement to P3 may also be a method for dealing with this issue. For instance, it may be beneficial to
+explicitly partition solutions when this deceptive behavior is observed.

### 2015-05-22 Dissertation/chapter-understanding.tex
Fixed the text for new trap figures

@@
 
-\begin{figure*}
-  \centering
-  \includegraphicsfit{valley-trap}
-  \caption{Location and quality of local optima in comparison to the global optima for Deceptive Step Trap ($r=1$ and $r=2$)
-           and Deceptive Trap with $N=30$ and $k=5$.}
+\begin{figure}[t]
+  \begin{centering}
+    \begin{subfigure}{.44\textwidth}
+      \begin{centering}
+        \includegraphicsfit{valley-dt}
+      \end{centering}
+      \caption{Deceptive Trap}
+      \label{fig-valley-dt}
+    \end{subfigure}%
+    \begin{subfigure}{.56\textwidth}
+      \begin{centering}
+        \includegraphicsfit{valley-dst}
+      \end{centering}
+      \caption{Deceptive Step Trap}
+      \label{fig-valley-dst}
+    \end{subfigure}
+  \end{centering}
+  \caption{Location and quality of local optima in comparison to the global optima with $N=30$ and $k=5$.}
   \label{fig-valley-trap}
-\end{figure*}
+\end{figure}
 
-The Deceptive Trap problem, shown in the rightmost panel, has very few local optima, each of which actually
+The Deceptive Trap problem has very few local optima, each of which actually
 provide very little deception in terms of ``big valley'' properties. Given any two local optima, the one

### 2015-05-25 Dissertation/chapter-understanding.tex
Another pass of the "Understanding" chapter

@@
-\chapter{Problem Understanding using Gray-Box Methods}
+\chapter{Understanding When P3 Excels}
 The ruggedness and high dimensionality of most interesting landscapes makes them very challenging
@@
 explore the landscapes used in previous sections in an attempt to understand
-what about a landscape makes it a good problem for P3 to solve.
+what makes a landscape easy for P3 to solve.
 
 \section{Big Valley}
- In order to understand the global structure of a landscape
-\cite{boese:1994:bigvalley} examined the relationship between local optima.
-This process involved generating thousands of random solutions and then applying
+One method for understanding the global structure of a landscape
+is to examine the relationship between local optima~\cite{boese:1994:bigvalley}.
+In its original form, this process involved generating thousands of random solutions and then applying
 local search to each. To convert this information into something human interpretable,
@@
 Due to the limited non-linearity of the gray-box domain, its possible to exclude large
-parts of the search space without potentially missing any local optima. Consider a fitness
-improving bit flip. By definition the only way this flip can cease to be a fitness improvement
-is if one of the bits it shares a subfunction with is flipped. Therefore, all solutions which
-contain the same values for all of those bits as the current solution cannot be locally optimal.
+parts of the search space without potentially missing any local optima. Consider a bit flip $X_i$.
+For gray-box problems there are only $ck$ other bit values which are needed to determine if flipping
+$X_i$ is a fitness improvement. If a given set of values for $X_i$ and those $ck$ bits results in
+$X_i$ being a fitness improvement, then any solution which contains all of those values cannot be a local optima.
 
@@
 Deceptive Step Trap's inclusion of fitness plateaus creates an enormous number of local optima. In total there are 24
-million local optima on this problem, representing 2.2\% of the entire search space. In comparison Deceptive Trap of the
+million local optima on this problem, representing 2.2\% of the entire search space. In comparison, Deceptive Trap using the
 same size has only 64 local optima. These additional local optima create a somewhat deceptive landscape, as selecting
-on fitness between two local optima can result in an increased genetic distance to the global optimum. In the middle
-panel of Figure~\ref{fig-valley-trap} we show how examining only 2-bit local optima causes Deceptive Step Trap using
+on fitness between two local optima can result in an increased genetic distance to the global optimum.
+Figure~\ref{fig-valley-dst} with $r=2$ shows how examining only 2-bit local optima causes Deceptive Step Trap using
 a step size of 2 to effectively revert to Deceptive Trap. This is because the hill climber can overcome all of the plateaus,
@@
 While visually somewhat similar, Unrestricted NKQ shown in Figure~\ref{fig-valley-nn} suggests selection may be misleading.
-Consider the bulge in the bottom right of $r=1$. If search finds points in that area, small modifications and elitist selection
-will not lead toward the global optimum. At higher radius values this issue becomes more apparent, with optima flattening out away
+Consider the shape of the bottom of $r=1$. At a distance of about 15 bits there is a slight hump, such that high quality solutions
+are more frequent both closer and further from the global optimum than at the same distance.
+If search finds points in the area over 15 bits away from the global optimum, small modifications and elitist selection
+are more likely to lead away from the global optimum that toward it.
+At higher radius values this issue becomes more apparent, with optima flattening out away
 from the global optima in representation space. For many of these points, the only way to improve would require flipping over 15 bits
@@
 scale so poorly when tested on this problem. P3's comparative success may therefore be the result of the random restart hill climber
-eventually finding a solution in the right area of the search space, with crossover finding the global optimum mostly by chance.
+eventually finding a solution in the correct area of the search space, with crossover finding the global optimum mostly by chance.
 
@@
 local optima and those likely to lead to the best found solution. An interesting behavior on this problem is that the branch
-between the two groups always seems to begin with a single solution in level 1. This suggests that the solutions found near
+between the two groups always seems to begin with a single solution in level 1. This suggests that the stored solutions near
 the best found may all be similar due to sharing a common ancestor.

### 2015-05-26 Dissertation/chapter-graybox.tex
Minor fixes

@@
 partial reevaluation (score a single trap), the same as Rastrigin (separable). All NK problems
-where $K \ll N $ have a knowable epistasis table and the fitness contribution of each is subfunction
+where $K \ll N $ have a knowable epistasis table and the fitness contribution of each subfunction is
 calculable without evaluating the entire string. Similarly, MAX-SAT's clause list specifies epistasis,
@@
 the universe is epistatically linked with subsets that contain that element (add to uncovered size)
-and with each subset also contributing directly to fitness (add to set size). Used in
+and each subset also contributes directly to fitness (add to set size). Used in
 computational biology~\cite{painsky:2014:setcover}.

### 2015-05-26 Dissertation/chapter-introduction.tex
Minor fixes

@@
 to optimize some of its own parameters, resulting in a reduced need for expert tuning.
-\cite{harik:1999:parameterlessga} were able to design an entirely parameter-less GA by
+\cite{harik:1999:parameterlessga} was able to design an entirely parameter-less GA by
 leveraging schema theory and parallel populations. Unfortunately these methods were provably less efficient
@@
 
-Most recently \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid~(P3).
+Most recently we introduced the Parameter-less Population Pyramid~(P3)~\cite{goldman:2014:p3}.
 This method uses a pyramid structure of populations to combine model based search with local search

### 2015-05-26 Dissertation/abstract.tex
Minor formatting fixes
More hypens for black-box and gray-box

--- 2014-10-28 Comprehensive/abstract.tex
@@
 ABSTRACT\\ %
-Analyzing the Parameter-less Population Pyramid \\ %
+\MyTitle\\ %
 By \\ %
@@
 sacrifice quality for applicability. Therefore we conclude that
-P3 is an efficient, general, parameter-less approach to black box
+P3 is an efficient, general, parameter-less approach to black-box
 optimization which is more effective than existing state-of-the-art techniques.
 
-We propose to extend this research to include the ``gray box'' problem domain, which includes many
+We propose to extend this research to include the ``gray-box'' problem domain, which includes many
 interesting real world problems. We plan to develop a variant of P3 which exploits
@@
 global optimization, both through simple crossover and as part of P3.
-Finally, we plan compare this P3 variant with state of the art black box, gray box, and
+Finally, we plan compare this P3 variant with state of the art black-box, gray-box, and
 white box methods for solving challenging real world problems.

### 2015-05-26 Dissertation/chapter-comparison_algorithms.tex
More hypens for black-box and gray-box

@@
 problem structure, designed to achieve similar goals as P3's linkage learning but using very different
-methods. Finally LTGA represents the current state-of-the-art in black box search and is the origin
+methods. Finally LTGA represents the current state-of-the-art in black-box search and is the origin
 of P3's linkage learning and crossover methods.
@@
 \label{sec-hill-climber}
-Perhaps the simplest black box search heuristic is stochastic local search, or hill climbing.
+Perhaps the simplest black-box search heuristic is stochastic local search, or hill climbing.
 This optimization technique focuses on improving a single solution until it reaches a local

### 2015-05-26 Dissertation/chapter-conclusions.tex
More hypens for black-box and gray-box

@@
 \chapter{Conclusions and Future Work}
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing black box
+The Parameter-less Population Pyramid (P3) is a recently introduced method for performing black-box
 optimization. P3's primary innovation is the replacement of the generational model with a pyramid of populations.
@@
 pressing for practitioner acceptance is to apply P3 to real world problems and compare
-its results with other black box or even problem specific heuristics. While parameter-less,
+its results with other black-box or even problem specific heuristics. While parameter-less,
 P3 is currently limited to discrete, fixed length genomes evaluated using single objective
@@
 these improvements our results show P3 is highly efficient at finding
-global optima on black box problems without any problem specific tuning.
+global optima on black-box problems without any problem specific tuning.
 

### 2015-05-26 Dissertation/chapter-internal.tex
More hypens for black-box and gray-box

@@
 Our procedure for tuning LTGA and hBOA outlined in Section~\ref{sec-tuning} involved finding
-the optimal population size for each class of problem. For real world black box optimization this is
+the optimal population size for each class of problem. For real world black-box optimization this is
 realistically the best either algorithm could hope for as tuning to a problem

### 2015-05-26 Dissertation/chapter-proposed.tex
More hypens for black-box and gray-box

--- 2014-10-28 Comprehensive/chapter-proposed.tex
@@
 The Parameter-less Population Pyramid (P3) was shown to be very effective in Chapter~\ref{chap-benchmarking}
-when applied to black box optimization. Some of this effectiveness in due to P3's integration of
+when applied to black-box optimization. Some of this effectiveness in due to P3's integration of
 efficient local search and its ability to automatically detect and exploit variable epistasis.
-In Chapter~\ref{chap-graybox} we presented the gray box domain, which contains many interesting
+In Chapter~\ref{chap-graybox} we presented the gray-box domain, which contains many interesting
 real world problems, and the improvements that can be made to local search in this domain.
@@
 \item What improvements can be made to P3's crossover operator when linkage is known?
-\item How effective is a gray box P3 algorithm in comparison with other gray box techniques?
+\item How effective is a gray-box P3 algorithm in comparison with other gray-box techniques?
 \end{itemize}
@@
 On Nearest Neighbor NK $N=200$, $K=5$, $r=4$, SUT was successful 47 of 50 times when limited to 24 minutes of computation.
-Its median time to success was 1.47 minutes, which is on par with P3's black box median time of 1.27 minutes.
+Its median time to success was 1.47 minutes, which is on par with P3's black-box median time of 1.27 minutes.
 There is a good chance that this similarity will change on larger problems. Its possible that SUT will not
@@
 \section{Developing Efficient Linkage Aware Crossover}
-The reason for recording variable entropy in P3 is to detect unknown variable linage in the black box domain. However,
-the gray box domain makes high level variable linkage fully known. Therefore by leveraging this information it should
-be possible to more efficiently construct crossovers which are even more accurate than the inferred black box ones.
+The reason for recording variable entropy in P3 is to detect unknown variable linage in the black-box domain. However,
+the gray-box domain makes high level variable linkage fully known. Therefore by leveraging this information it should
+be possible to more efficiently construct crossovers which are even more accurate than the inferred black-box ones.
 
@@
 Assuming that the problem is not separable, this algorithm will create
-$\Omega(N)$ clusters. This is a safe assumption in the gray box domain as it is more
+$\Omega(N)$ clusters. This is a safe assumption in the gray-box domain as it is more
 efficient to solve each component independently.
@@
 \subsection{Preliminary Empirical Comparison}
-We tested P3 using the gray box \Call{SubfunctionTree}{} in place of the black box \Call{Cluster-Creation}{}
-and using the gray box hamming ball hill climber instead of black box hill climbing. As an added efficiency
+We tested P3 using the gray-box \Call{SubfunctionTree}{} in place of the black-box \Call{Cluster-Creation}{}
+and using the gray-box hamming ball hill climber instead of black-box hill climbing. As an added efficiency
 instead of recalculating all $delta$ values and assuming all moves are open each time a crossover donation is made,
@@
 $N=200$, $K=5$, $r=4$ this found the global optimum in all 50 runs, with a median time to success of 0.23 minutes. This makes it the
-most successful gray box tested so far, and 6 times faster than applying black box P3. Also, unlike the other
+most successful gray-box tested so far, and 6 times faster than applying black-box P3. Also, unlike the other
 methods decreasing $r$ actually makes it more efficient, achieving a median time to success of 0.013 minutes (0.78 seconds) with $r=1$,
-97 times faster than black box P3.
+97 times faster than black-box P3.
 %1.30066, 0.668016, 20.0872, 24.015, 6.84026, 0.620081, 1.43271, 10.9687, 0.625069, 24.0347, 2.84006, 1.57277, 2.81627, 7.82749, 0.63878, 1.52631, 1.31218, 1.41851, 0.650093, 2.79497, 3.01983, 0.28769, 24.0449, 0.292457, 0.316901, 2.957, 1.35011, 0.636912, 2.8645, 0.578051, 1.30444, 0.274158, 3.16518, 0.141998, 1.34107, 13.8565, 0.579429, 0.345084, 0.650234, 13.4685, 3.04961, 0.670062, 1.53499, 19.3596, 2.76979, 1.50853, 1.3763, 1.4152, 12.3199, 2.13644
@@
 
-Current research on these gray box methods have been very preliminary: a single size of a single problem class using
+Current research on these gray-box methods have been very preliminary: a single size of a single problem class using
 a single $r$, $k$, and $c$. To fully understand how to effectively perform search in this new domain, much more extensive testing
@@
 
-Just as when testing P3 in the black box domain, an important step will be to identify a representative set
-of gray box problems. Planned test problems are NK landscapes, Ising Spin Glasses, and MAX-SAT. Beyond testing the
+Just as when testing P3 in the black-box domain, an important step will be to identify a representative set
+of gray-box problems. Planned test problems are NK landscapes, Ising Spin Glasses, and MAX-SAT. Beyond testing the
 problems with known global optimum, we plan to try Unrestricted NK and Ising Spin Glasses with different linkage models.
-This is critical as the behavior of many gray box algorithms depend on the nature of the epistasis graph. In these domains
+This is critical as the behavior of many gray-box algorithms depend on the nature of the epistasis graph. In these domains
 the focus will shift from finding the global optimum (which is unknown) to finding the best solution in a fixed amount of time.
@@
 This will help solidify the conclusions about the applicability of these methods outside of the optimization community.
-Similarly, we plan to compare the results of our gray box methods with algorithms specifically designed for each problem
+Similarly, we plan to compare the results of our gray-box methods with algorithms specifically designed for each problem
 domain. This provides an interesting contention between the time required to develop problem specific solvers and the
-quality of using gray box methods.
+quality of using gray-box methods.
 
@@
 Another area of focus will be in how $N$ effects the relative effectiveness of the methods. One of the goals of moving
-to a gray box domain is how it impacts Big-$O$ complexity. As a result we plan to empirically verify the scalability
-of each algorithm, as well as the wall clock time to find the global optimum. This is necessary as unlike the black box
-domain where evaluation is considered arbitrarily expensive, evaluation in the gray box domain is perfectly efficient.
+to a gray-box domain is how it impacts Big-$O$ complexity. As a result we plan to empirically verify the scalability
+of each algorithm, as well as the wall clock time to find the global optimum. This is necessary as unlike the black-box
+domain where evaluation is considered arbitrarily expensive, evaluation in the gray-box domain is perfectly efficient.
 
@@
 \item Does preserving subfunctions intact during crossover improve search efficiency?
-\item How does black box P3 compare with gray box P3?
-\item How do the more generic gray box methods compare with problem specific methods for solving real world problems?
+\item How does black-box P3 compare with gray-box P3?
+\item How do the more generic gray-box methods compare with problem specific methods for solving real world problems?
 \end{itemize}

### 2015-05-26 Dissertation/chapter-conclusions.tex
Fixed up conclusion

--- 2015-04-15 Dissertation/chapter-conclusions.tex
@@
 \chapter{Conclusions and Future Work}
-The Parameter-less Population Pyramid (P3) is a recently introduced method for performing black box
+% P3 Journal
+The Parameter-less Population Pyramid (P3) is a recently introduced method for performing black-box
 optimization. P3's primary innovation is the replacement of the generational model with a pyramid of populations.
@@
 
+When leveraging features of the new domain, Gray-Box P3 can obtain substantial improvement
+over Black-Box P3. This comes from the integration of the Hamming-Ball Hill Climber (HBHC)
+and a novel method for deriving linkage clusters from the known problem epistasis,
+both of which result in a factor $N$ speedup over their black-box counterparts.
+Furthermore, Gray-Box P3
+can perform partial reevaluation and efficient hill climbing during the mixing phase
+without significant overhead. In specializing, Gray-Box P3 is able to find
+global optima orders of magnitude faster than Black-Box P3.
+
 There are a number of meaningful avenues for future P3 experimentation. Perhaps the most
 pressing for practitioner acceptance is to apply P3 to real world problems and compare
-its results with other black box or even problem specific heuristics. While parameter-less,
+its results with other black-box or even problem specific heuristics. While parameter-less,
 P3 is currently limited to discrete, fixed length genomes evaluated using single objective
 fitness. These limitations can be relaxed with future work to make P3 more widely applicable.
-While asymptotically linear in problem size, P3's modeling techniques and local search methods are likely going
+While asymptotically linear in problem size, Black-Box P3's modeling techniques and local search methods are likely going
 to be prohibitively expensive for genome sizes in the hundreds of thousands or millions of genes,
@@
 order deception via linkage learning, the iterative improvement method by which crossovers
-are made may mislead search on landscapes with higher order deception.
+are made may mislead search on landscapes with higher order deception. This appears
+to be especially important on NP-Hard problems, where high quality solutions may be
+very distant from the global optimum in representation space.
+
+While empirically effective, we believe its possible to derive theoretical foundations
+for Gray-Box P3. Specifically, this may be possible by considering the nature of crossover
+in terms similar to those use for finding $r$-bit local optima with HBHC. Consider
+that for any two parents the non-linear effects of crossover are constrained by
+where they containing different bit values. As a result, subsets of their differences
+can become linearly separable, allowing for each to be considered independently of all others.
+Furthermore, these conclusions about effective crossover in gray-box can likely be translated
+into a black-box setting to help explain the effectiveness of Black-Box P3.
 
@@
 these improvements our results show P3 is highly efficient at finding
-global optima on black box problems without any problem specific tuning.
+global optima on both black-box and gray-box problems without any problem specific tuning.

### 2015-05-26 Dissertation/chapter-gray-box-benchmarking.tex
Fixed up conclusion

@@
 \end{figure}
+
+\section{Discussion}
+In line with previous work, we have found that HBHC cannot effectively find global optima
+on problems with even moderate epistasis. In general it also obtains almost no improvement
+in fitness after only a few restarts. We designed TUX as a simplistic way of choosing
+restart locations based on previously found local optima. Unlike HBHC alone, TUX
+was able to continue improving given more time, finding global optima on problems three
+times as large.
+
+TUX's effectiveness is likely due to the HBHC acting as a super repair operator for uniform crossover.
+Given a sufficiently large $r$ the HBHC can return sections of the crossover offspring to either
+parents' original version of a given subfunction. The HBHC is elitist meaning there is a bias toward
+returning to the better of the two parent's versions. Furthermore, by being so disruptive,
+uniform crossover potentially allows for the HBHC to also find unrelated improvements.
+
+While TUX improves over plain HBHC, Gray-Box P3 is required to perform truly successful
+global optimization. Gray-Box P3 replaces na\"{i}ve local search with the HBHC and utilizes
+known non-linear relationships instead of statistical linkage learning. On NKq-Landscapes this drastically
+improves search effectiveness. A major source of this improvement is likely how difficult it is
+for Black-Box P3 to learn linkage relationships on these landscapes. Furthermore, Gray-Box P3
+can perform partial reevaluation and efficient hill climbing during the mixing phase.
+
+Gray-Box P3's success is less dramatic on Ising Spin Glasses.
+While it still outperformed all competitors, Black-Box P3 may actually scale better to larger
+problems. One explanation for this deviation is that Ising Spin Glasses require more
+exploration of equal fitness plateaus. For instance in Figure~\ref{fig-solver-over} and Figure~\ref{fig-radius-over}
+there is a significant pause in improvement when Gray-Box P3 reaches the second best fitness in the landscape.
+Nothing in its design suggests that Gray-Box P3 should be more effective at neutral drift. Another potential
+issue is that on these landscapes the importance of each non-linear relationship may be detectably unequal.
+As a result Black-Box linkage learning may better cluster variables which having meaningful
+impact on fitness while Gray-Box assumes all are equally important. A useful direction for future work
+would be to explore methods of performing efficient learning on top of the known variable interactions.
+
+Somewhat surprising is the difference in behavior between the polynomially solvable problems
+and Unrestricted NKq. While Black-Box P3 performed very well in the former, it was the least
+successful in the latter. The optimal radius for Gray-Box P3 also shifted from 1 to 2.
+A potential explanation is that in both Ising Spin Glass
+and Nearest Neighbor NKq the number of unique variables $r$
+or fewer steps away from a given variable is significantly lower than the worst case. This explains why on Unrestricted NKq
+even moderately high $r$ values hit our memory limit.
+For Black-Box P3 this may also be causing increased difficulty in linkage learning as variables
+become indirectly dependent on much larger sets. Furthermore, Black-Box
+P3 may be benefiting from an increased rate of duplicate dependencies on Nearest Neighbor NKq
+not present in Unrestricted NKq.
+
+While the inclusion of HBHC into Gray-Box P3 introduces a parameter, it requires
+trivial configuration. In the worst case there may be a handful of $r$ values to test.
+Furthermore, our evidence suggests setting $r=1$ is quite powerful, with higher values
+likely to be only a time shift in quality. This is in contrast to $r$'s role in HBHC,
+where low $r$ values are never expected to reach the same quality has higher $r$ values.
+Therefore we conclude that Gray-Box P3 maintains the out-of-the-box quality of Black-Box
+P3, while drastically improving efficiency for this new domain of problems.

### 2015-05-26 Dissertation/chapter-graybox.tex
Fixed up conclusion

@@
 \label{sec-whitley-ls}
-Section~\ref{sec-hill-climber} presented an efficient method for performing local search for the black
-box domain. Each potential bit flip, referred to here as a move, requires the entire solution to be reevaluated, which takes $\Omega(N)$ time.
+Section~\ref{sec-hill-climber} presented an efficient method for performing local search for the black-box
+domain. Each potential bit flip, referred to here as a move, requires the entire solution to be reevaluated, which takes $\Omega(N)$ time.
 Furthermore, each time a fitness improving move is made all previously tested moves must be tested again. As a result
@@
 In a black-box setting, verifying that no $r$ bit flip can improve a solution requires testing all
-$n \choose r$ neighbors. This quickly becomes intractable as $r$ increases. However in the gray
-box domain not all combinations need to be tested. Consider that if two variables do not participate
+$n \choose r$ neighbors. This quickly becomes intractable as $r$ increases. However in the gray-box
+domain not all combinations need to be tested. Consider that if two variables do not participate
 in the same subfunction, the relationship between their effects is by definition additive. As such
@@
 if the corresponding variables participate in at least one subfunction together. Restated,
-there is only an edge if the two variables have a non-linear relationship. The maximum
+there is only an edge if the two variables have a direct non-linear relationship. The maximum
 degree of a vertex in this graph is $c(k-1)$, making it sparse for sufficiently large $N$.
@@
 efficiently find $r$-bit local optima~\cite{chicano:2014:ball}. The only change is that instead of having a move
-and $delta$ for each bit, there must be a move and delta for each connected induced subgraph of $r$ or less bits.
+and $delta$ for each bit, there must be a move and $delta$ for each connected induced subgraph of $r$ or less bits.
 The efficiency analysis is unchanged with the exception that the constant increases exponentially with $r$ as

### 2015-05-26 Dissertation/abstract.tex
Updated abstract

@@
 
-We propose to extend this research to include the ``gray-box'' problem domain, which includes many
-interesting real world problems. We plan to develop a variant of P3 which exploits
-features of this domain, such as efficient local search and known variable linkage.
-Furthermore, we want to examine how ``hamming ball search'' can be integrated into
-global optimization, both through simple crossover and as part of P3.
-Finally, we plan compare this P3 variant with state of the art black-box, gray-box, and
-white box methods for solving challenging real world problems.
+P3 can be further improved by specializing to the gray-box domain, which includes
+partial solution evaluation and known, limited, non-linear relationships between variables.
+This Gray-Box P3 leverages the Hamming-Ball Hill Climber, an exceptionally efficient
+form of local search, as well as a novel method for performing crossover using the
+known variable interactions. In doing so Gray-Box P3 is able to find the global
+optimum to very large problems in seconds, improving over Black-Box P3 by up to
+two orders of magnitude.
 \end{doublespace}

### 2015-05-26 Dissertation/chapter-gray-box-benchmarking.tex
Fixed some label problems

@@
 \chapter{Benchmarking Gray-Box P3}
+\label{chapter-gray-box-benchmarking}
 To compare these gray-box optimization techniques we have chosen

### 2015-05-26 Dissertation/chapter-graybox.tex
Fixed some label problems

@@
 \section{Efficient Hamming Ball Search}
+\label{sec-hamming}
 Another consequence of the gray-box domain is that increasing the neighborhood of local search becomes
@@
 requires an $r$ for the hill climber) and is relatively simple to implement. Even so
-it allows for learning from previous local optima and as Section~\ref{sec-experiments}
+it allows for learning from previous local optima and as Section~\ref{chapter-gray-box-benchmarking}
 will show it is quite effective at optimization.

### 2015-06-01 Dissertation/abstract.tex
Formatting from Jared
Removed "very" and cleaned up some related text

@@
 Brian W. Goldman\\ %
-\ \\
 \end{centering}
@@
 known variable interactions. In doing so Gray-Box P3 is able to find the global
-optimum to very large problems in seconds, improving over Black-Box P3 by up to
+optimum of large problems in seconds, improving over Black-Box P3 by up to
 two orders of magnitude.

### 2015-06-01 Dissertation/chapter-benchmarking.tex
Removed "very" and cleaned up some related text

@@
 of 100 runs, with unsuccessful runs treated as slower than successful runs.
-\begin{comment}
-In total these runs
-took over 6 computing years to complete. We were therefore forced to use
-diverse hardware potentially running multiple processes in parallel.
-As such these timings should be considered very skeptically as a number
-of sources may contribute to biased error in the results. However, due to
-the size of the dataset and the general regularity in the results some general conclusions
-can still be made.
-\end{comment}
 These results were collected using 2.5GHz Intel Xeon E5-2670v2 processors.

### 2015-06-01 Dissertation/chapter-comparison_algorithms.tex
Removed "very" and cleaned up some related text

@@
 and its method of crossover is in some sense a macro-mutation just as in P3. hBOA and Parameter-less
-hBOA are advanced model building search techniques which are very effective at learning complex
+hBOA are advanced model building search techniques which are effective at learning complex
 problem structure, designed to achieve similar goals as P3's linkage learning but using very different
@@
 
-Like other model based techniques, hBOA has very few parameters. There is no mutation or crossover,
+Like other model based techniques, hBOA has few parameters. There is no mutation or crossover,
 and modeling does not rely on any explicit parameters. Solution selection, generation, and replacement
@@
 than hBOA's as problem size increased, especially on the two Trap problems and Rastrigin.
-Both algorithms were very ineffective on MAX-SAT, neither able to tune to problems sizes over 60 bits.
-This is likely due to the fact that some randomly generated MAX-SAT landscapes are very flat
+Both algorithms were ineffective on MAX-SAT, with neither able to tune to problems sizes over 60 bits.
+This is likely due to the fact that some randomly generated MAX-SAT landscapes are quite flat
 and highly deceptive~\citep{rana:1998:gamaxsat}.

### 2015-06-01 Dissertation/chapter-conclusions.tex
Removed "very" and cleaned up some related text

@@
 when solving problem classes. Third, by heavily exploiting existing diversity before adding more P3 is able
-to reach high quality intermediate fitnesses quickly without prematurely converging. Fourth, the very
+to reach high quality intermediate fitnesses quickly without prematurely converging. Fourth, the
 nature of the pyramid's shape allows search to preserve a desirable proportion of diversity at

### 2015-06-01 Dissertation/chapter-gray-box-benchmarking.tex
Removed "very" and cleaned up some related text

@@
 For all sizes of both problems, Gray-Box P3 was the fastest to reach the best fitness. On
-Nearest Neighbor NKq the improvement is very significant, with no alternative finishing within
+Nearest Neighbor NKq the improvement is substantial, with no alternative finishing within
 two orders of magnitude. On its largest successful problem size, the mean time to completion for
@@
 Somewhat surprising is the difference in behavior between the polynomially solvable problems
-and Unrestricted NKq. While Black-Box P3 performed very well in the former, it was the least
+and Unrestricted NKq. While Black-Box P3 performed well in the former, it was the least
 successful in the latter. The optimal radius for Gray-Box P3 also shifted from 1 to 2.

### 2015-06-01 Dissertation/chapter-graybox.tex
Removed "very" and cleaned up some related text

@@
 is a ``black-box'' which measures the quality of arbitrary proposed
-solutions. Optimizers which are successful in this domain are very general
+solutions. Optimizers which are successful in this domain generalize well
 as there are minimal requirements to apply them to a new problem.
@@
 To understand the types of problems in this domain, lets first examine
-a very simple artificial problem we'll call 3-Equal. The quality
+a simple artificial problem we'll call 3-Equal. The quality
 of a solution is determined by how often three consecutive bits are set

### 2015-06-01 Dissertation/chapter-internal.tex
Removed "very" and cleaned up some related text

@@
 
-While analysis of optimization speed is useful from a practitioner standpoint, doing so provides very
+While analysis of optimization speed is useful from a practitioner standpoint, doing so provides
 little insight into algorithm behavior. To better understand how P3 works in detail we present here
@@
 number of 1 bits equal to either 0, 1, 3, 5, or 7 as these correspond to the local
-optima when using $k=7$ and $s=2$. Crossover is very likely to overcome the
+optima when using $k=7$ and $s=2$. Crossover easily overcomes the
 two bit plateaus, and as a result solutions in the second level generally do not
 contain the lowest fitness local optima (5 bits set) and the third
-level has very few traps set to the next worst local optima (3 bits set). As a result,
+level has few traps set to the next worst local optima (3 bits set). As a result,
 solutions which reach the third level can only be improved by replacing the deceptive
 local optima (0 and 1 bits set) with the global optimum (7 bits set).
-The global optimum is very rare in the population, and with 8 ways to represent local optima
-linkage learning is inaccurate. Therefore it is very unlikely for crossover to be successful,
+The global optimum is rare in the population, and with 8 ways to represent local optima
+linkage learning is inaccurate. Therefore it is unlikely for crossover to be successful,
 meaning few solutions will be added to the fourth level. Solutions that do improve by

### 2015-06-01 Dissertation/chapter-problems.tex
More hypens for black-box and gray-box
Removed "very" and cleaned up some related text

@@
 their static nature can be misleading. Specifically, algorithm quality might
-be very fragile such that it is only effective at searching well behaved landscapes. A more realistic
-test of an algorithm's black box effectiveness is to work with randomly generated instances
+be so fragile that it is only effective at searching well behaved landscapes. A more realistic
+test of an algorithm's black-box effectiveness is to work with randomly generated instances
 drawn from a problem class. When tested over a sufficiently large sample it is then possible

### 2015-06-01 Dissertation/chapter-proposed.tex
Removed "very" and cleaned up some related text

@@
 \chapter{Proposed Research}
-The Parameter-less Population Pyramid (P3) was shown to be very effective in Chapter~\ref{chap-benchmarking}
+The Parameter-less Population Pyramid (P3) was shown to be effective in Chapter~\ref{chap-benchmarking}
 when applied to black-box optimization. Some of this effectiveness in due to P3's integration of
@@
 
-Current research on these gray-box methods have been very preliminary: a single size of a single problem class using
+Current research on these gray-box methods have been preliminary: a single size of a single problem class using
 a single $r$, $k$, and $c$. To fully understand how to effectively perform search in this new domain, much more extensive testing

### 2015-06-01 Dissertation/chapter-understanding.tex
Removed "very" and cleaned up some related text

@@
 \chapter{Understanding When P3 Excels}
-The ruggedness and high dimensionality of most interesting landscapes makes them very challenging
-to visualize or otherwise understand. However, doing so can be very helpful in
+The ruggedness and high dimensionality of most interesting landscapes makes them challenging
+to visualize or otherwise understand. However, doing so can be helpful in
 understanding the difficulty of a problem, and how best to design algorithms to deal
@@
 In comparison to the randomly generated problem instances of Nearest Neighbor NKQ and Ising Spin Glass,
-Deceptive Trap and Deceptive Step Trap have very comprehensible landscapes. Therefore, these problems represent
+Deceptive Trap and Deceptive Step Trap have comprehensible landscapes. Therefore, these problems represent
 a good place to begin understanding what ``big valley'' plots are showing. Figure~\ref{fig-valley-trap} shows
@@
 
-The Deceptive Trap problem has very few local optima, each of which actually
-provide very little deception in terms of ``big valley'' properties. Given any two local optima, the one
+The Deceptive Trap problem has few local optima, each of which actually
+provide little deception in terms of ``big valley'' properties. Given any two local optima, the one
 closer to the global optimum in fitness is also always closer to the global optimum in representation space.
@@
 Figure~\ref{fig-valley-nn} shows how the local optima are distributed for Nearest Neighbor NKQ. Unlike the trap problems,
-Nearest Neighbor NKQ exhibits a very traditional big valley shape. Increasing the radius of the local optima significantly
+Nearest Neighbor NKQ exhibits a traditional big valley shape. Increasing the radius of the local optima significantly
 reduces the total number of local optima, with those optima generally more similar to the global optima in both representation
@@
 
-Similar to Unrestricted NKQ, the MAXSAT instance shown in Figure~\ref{fig-valley-sat} suggests selection can be very misleading.
+Similar to Unrestricted NKQ, the MAXSAT instance shown in Figure~\ref{fig-valley-sat} suggests selection can be misleading.
 There is almost a negative correlation between fitness difference from the global optimum and representation distance from the
@@
 Gray-Box P3's progression on Deceptive Step Trap, shown in Figure~\ref{fig-level-dst}, follows directly from our expectations.
-The local optima found using only hill climbing and stored in level 0 are quite poor and very far from the global optima.
+The local optima found using only hill climbing and stored in level 0 are quite poor and far from the global optima.
 The first application of crossover dramatically improves their quality. However, this improvement is likely just overcoming
@@
 When approaching higher qualities these improvements also begin to translate into increased representational similarity to the global optimum.
-However, as discussed previously for this problem, we again see evidence for a large number of very diverse solutions with the second best
+However, as discussed previously for this problem, we again see evidence for a large number of diverse solutions with the second best
 fitness. These solutions can be over 1000 bits different from the eventual best found solution. Increasing the radius of the hill climber
@@
 to be the ancestor of all of the solutions near the best found. This suggests that similar to Unrestricted NKQ, Gray-Box P3 will have a
-very hard time finding the global optimum on MAX-SAT.
+hard time finding the global optimum on MAX-SAT.
 
@@
 share a similar behavior of selection leading to the global optimum. However, the NP-Hard problems (Unrestricted NKQ, MAX-SAT) seem to contain
-very large amounts of higher order deception, making them challenging for selection based methods. This relationship warrants further investigation
+large amounts of higher order deception, making them challenging for selection based methods. This relationship warrants further investigation
 into other NP-Hard problems to determine if the repeated selection found in P3 and other evolutionary computation methods are able to overcome

### 2015-06-01 Dissertation/abstract.tex
which -> that

@@
 P3 is an efficient, general, parameter-less approach to black-box
-optimization which is more effective than existing state-of-the-art techniques.
+optimization that is more effective than existing state-of-the-art techniques.
 

### 2015-06-01 Dissertation/appendix-hboa.tex
Major update from journal article polishing
which -> that

--- 2014-10-15 Comprehensive/appendix-hboa.tex
@@
 \chapter{hBOA Simplification}
-\label{chap-appendix}
+\label{chap-appendix-hboa}
 To measure the quality of a decision forest, hBOA applies Equation~\ref{eq-hboa}.
@@
 term in the third product reduce to 1. The only remaining terms are then $m_i(l)$ and $m_i(x_i, l)$ which
-represent the number of solutions which reached leaf $l$ and the number of solutions which reached leaf $l$
+represent the number of solutions that reached leaf $l$ and the number of solutions that reached leaf $l$
 with a specific value for $x_i$, respectively.

### 2015-06-01 Dissertation/chapter-comparison_algorithms.tex
which -> that

@@
 
-In order to fully understand the effectiveness of P3, we compare it with five advanced algorithms which
+In order to fully understand the effectiveness of P3, we compare it with five advanced algorithms that
 have related features to P3. The Random Restart Hill Climber defined by ~\cite{goldman:2014:p3} was chosen as an efficient form of repeated
@@
 and its method of crossover is in some sense a macro-mutation just as in P3. hBOA and Parameter-less
-hBOA are advanced model building search techniques which are effective at learning complex
+hBOA are advanced model building search techniques that are effective at learning complex
 problem structure, designed to achieve similar goals as P3's linkage learning but using very different
@@
 If a crossover individual is identical to either of its parents, it is not evaluated.
-If mutation produces an offspring which is better than the best crossover offspring, it is used to compare
+If mutation produces an offspring that is better than the best crossover offspring, it is used to compare
 against the original parent.
@@
 This goal is formalized using model scoring from Bayesian statistics. In its raw form this
-almost always creates near infinitesimal results, calculating fractions which include factorials of $\mu$
+almost always creates near infinitesimal results, calculating fractions that include factorials of $\mu$
 and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{chap-appendix-hboa}, we
@@
 Here $l$ is a leaf in tree $i$, with $l'$ and $l''$ the results of splitting $l$. $m_i(l)$ is the number
-of solutions which reach $l$ and $m_i(x_i, l)$ is the number of solutions which reach $l$ with the given value
+of solutions that reach $l$ and $m_i(x_i, l)$ is the number of solutions that reach $l$ with the given value
 for $x_i$.
@@
 As a split creates two new leaves, \BigO{N} new potential splits must also be tested. Equation~\ref{eq-hboa-final}
-parses all solutions which reach a leaf to count gene frequencies, requiring $\mu$ time.
+parses all solutions that reach a leaf to count gene frequencies, requiring $\mu$ time.
 The number of total leaves created depends heavily on the problem and $\mu$.
@@
 between genes. Before performing any crossovers in a generation, LTGA first builds
-a set of hierarchical gene clusters which are then used to dictate how genes are mixed
+a set of hierarchical gene clusters that are then used to dictate how genes are mixed
 during crossover.
@@
 internal nodes are the union of their children's sets. These sets are then used by crossover to specify epistatic
-relationships which should be preserved.
-The process begins by creating the set of sets $unmerged$ which tracks all top level clusters. Initially
+relationships that should be preserved.
+The process begins by creating the set of sets $unmerged$ that tracks all top level clusters. Initially
 $unmerged$ contains single member sets for each gene. After each iteration the two sets with the minimum average pairwise
@@
 
-Throughout this process $useful$ tracks the set of all gene clusters which should be preserved for use by crossover.
+Throughout this process $useful$ tracks the set of all gene clusters that should be preserved for use by crossover.
 This set begins with all genes in separate clusters, and each time a new cluster is created it is added to $useful$.
@@
 \cite{thierens:2013:ltgahiff} extended this removal to include any unsupported subsets. If the pairwise distance
-between two clusters is 0, this means there are no individuals in the population which disrupt the relationships between the two
+between two clusters is 0, this means there are no individuals in the population that disrupt the relationships between the two
 clusters. Therefore when performing crossover there is no reason to believe a fitness improvement can be achieved
@@
 In total each individual can cause up to $|useful|$ evaluations. If all clusters were kept, even those deemed
-unhelpful, and all donations were evaluated, even those which did not change any genes, then \Call{Cluster-Usage}{}
+unhelpful, and all donations were evaluated, even those that did not change any genes, then \Call{Cluster-Usage}{}
 would perform exactly $2N-2$ evaluations for each of the $\mu$ solutions in the population. This provides the amortizing evaluations
@@
 between the lowest successful and highest unsuccessful sizes. In this way the minimum population size
-which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
+that meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
 of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
@@
 do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
-As bisection can make infinitely large population sizes, any run which had not found the global
+As bisection can make infinitely large population sizes, any run that had not found the global
 optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.
@@
 Figure~\ref{fig-pop-sizes} shows the results from performing bisection. In general hBOA
-required population sizes which were at least an order of magnitude larger than LTGA. Due to
+required population sizes that were at least an order of magnitude larger than LTGA. Due to
 runtime and memory overhead, finding the optimal value for hBOA was much less tractable than LTGA

### 2015-06-01 Dissertation/chapter-conclusions.tex
which -> that

@@
 at each level growing as search progresses. P3 uses a model based crossover method
-which learns a linkage tree from gene epistasis. Combined with a simple hill climber, P3's design contains
+that learns a linkage tree from gene epistasis. Combined with a simple hill climber, P3's design contains
 many synergistic features.
@@
 crossover lets search focus on properly mixing high quality solutions. Second, by adding diversity only
-as necessary P3 tends to use the minimal amount of random initialization, unlike other techniques which must
+as necessary P3 tends to use the minimal amount of random initialization, unlike other techniques that must
 overcompensate with larger population sizes on single instance problems and consider the worst instance

### 2015-06-01 Dissertation/chapter-gray-box-benchmarking.tex
which -> that

@@
 each subfunction depends on: Nearest Neighbor NKq and Unrestricted NKq.
-In Nearest Neighbor NKq each variable depends on the $K$ variables which sequentially follow
+In Nearest Neighbor NKq each variable depends on the $K$ variables that sequentially follow
 it in the solution, with dependencies wrapping around the end of the solution.
@@
 encodes spins (vertices) and their relationships (edges) with the goal of assigning each spin a direction
-which minimizes relationship energy. Just as with NKq-Landscapes, Ising Spin Glasses as a whole
+that minimizes relationship energy. Just as with NKq-Landscapes, Ising Spin Glasses as a whole
 are NP-Hard, but the $2D\pm J$ subset is polynomially solvable~\cite{saul:1994:spinglass}\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}. In this
@@
 equal to the difference between its fitness and the best, divided by the best. When reporting seconds
-to reach the best fitness, any run which did not find the best fitness is treated as slower than any run that did.
+to reach the best fitness, any run that did not find the best fitness is treated as slower than any run that did.
 If the median run was unsuccessful, no data point is drawn.
@@
 issue is that on these landscapes the importance of each non-linear relationship may be detectably unequal.
-As a result Black-Box linkage learning may better cluster variables which having meaningful
+As a result, Black-Box linkage learning may better cluster variables that have meaningful
 impact on fitness while Gray-Box assumes all are equally important. A useful direction for future work

### 2015-06-01 Dissertation/chapter-gray-box-p3.tex
which -> that

@@
 the direct non-linear relationships between variables are known. As a result
-P3 no longer needs to store the pairwise frequency information which leads
+P3 no longer needs to store the pairwise frequency information that leads
 the black-box version to require \BigO{N^2} memory. Instead,
-we have developed a method for creating a linkage tree which learns clusters
+we have developed a method for creating a linkage tree that learns clusters
 from the same dependency graph defined in Section~\ref{sec-hamming}. The goal is for each
@@
 they are to appear in the same cluster. All variables on average are expected to appear
-in at least one cluster, but variables which are central in the graph will appear in more
+in at least one cluster, but variables that are central in the graph will appear in more
 clusters than those on the periphery. The more paths of a given
@@
 
-In effect the clusters are sampling moves which the HBHC
-would make if $r \ge l$. For any solution in the search space which is not globally optimal, there
-must exist some move which will improve its quality. However, this move may
+In effect the clusters are sampling moves that the HBHC
+would make if $r \ge l$. For any solution in the search space that is not globally optimal, there
+must exist some move that will improve its quality. However, this move may
 be arbitrarily large and it is intractable to test all possible moves of even moderate size.

### 2015-06-01 Dissertation/chapter-graybox.tex
which -> that

@@
 is a ``black-box'' which measures the quality of arbitrary proposed
-solutions. Optimizers which are successful in this domain generalize well
+solutions. Optimizers that are successful in this domain generalize well
 as there are minimal requirements to apply them to a new problem.
@@
 is dominated by problem specific search
-heuristics~\cite{een:2004:satsolver,sherrington:1975:isingsolver,wright:2000:solvingnk} which
+heuristics~\cite{een:2004:satsolver,sherrington:1975:isingsolver,wright:2000:solvingnk} that
 leverage all aspects of the problem to achieve efficiency. This makes such
@@
 domain some features general to multiple problem classes are exploited, beyond
-just an evaluation function. The goal is to create optimization methods which
+just an evaluation function. The goal is to create optimization methods that
 benefits from these features without those methods becoming specialized to
@@
 
-\item \textbf{Max-Cut}: Find the set of vertices which maximize the number of edges
+\item \textbf{Max-Cut}: Find the set of vertices that maximize the number of edges
 incident on exactly one vertex in the set. Epistasis and fitness are calculated
@@
 to two consequences of the domain: the fitness effect of making a move can be calculated in \BigO{1} time, not \BigO{N},
-and the number of moves which must be reevaluated per improvement is \BigO{1}, not \BigO{N}. This results in local
+and the number of moves that must be reevaluated per improvement is \BigO{1}, not \BigO{N}. This results in local
 search requiring \BigO{N+I} time. If $I$ is within a constant factor of $N$, this means generating random local
@@
 For each of the $\frac{cN}{k}$ subfunctions \Call{InitializeDelta}{} must determine the fitness of that
-subfunction before and after making each of $k$ moves which overlap that subfunction. Combined this
+subfunction before and after making each of $k$ moves that overlap that subfunction. Combined this
 results in less than $cN2b(k)$ operations, which is \BigO{N} assuming $c$ and $b(k)$ do not grow with $N$.
@@
 of the solution after making the move does not require any calls to $f_s$ as $delta[m]$ already stores the
-change in fitness. This process requires updating all the $delta$ values for the $k$ moves which interact with each of the $c$ subfunctions
-effected by $m$. This update replaces outdated information which used the original solution ($just\_m' - pre\_both$)
+change in fitness. This process requires updating all of the $delta$ values for the $k$ moves that interact with each of the $c$ subfunctions
+effected by $m$. This update replaces outdated information that used the original solution ($just\_m' - pre\_both$)
 with how much the move improves over the new solution ($post\_both - just\_m$). As $delta[m']$ has updated, $m'$ could
@@
 
-To determine which of the $n \choose r$ flips which must be checked, consider a graph where
+To determine which of the $n \choose r$ flips that must be checked, consider a graph where
 each vertex is a variable in the solution. An edge exists between two vertices if and only
@@
 we developed \Call{ConnectedInducedSubgraphs}{} given in Figure~\ref{fig-connected-subgraphs}.
-\Call{CISG}{} is a recursive helper function which finds all subgraphs
-which contain a given $subset$ and a given vertex $v$, while excluding
+\Call{CISG}{} is a recursive helper function that finds all subgraphs
+that contain a given $subset$ and a given vertex $v$, while excluding
 any other vertices added to $closed$. To find all subgraphs, \Call{CISG}{}
@@
 all previously searched vertices, and $subset=open=\emptyset$. In the initial
-call all desired subgraphs which contain $v$ are found, which is why $v$ remains
+call all desired subgraphs that contain $v$ are found, which is why $v$ remains
 in $closed$ to prevent duplicate subgraphs from being returned.
@@
 As this algorithm finds all connected subsets in \BigO{N} time for a fixed $r$, the number of moves
-which must be tested to determine if a solution is an $r$-bit local optima is \BigO{N}. This means that
+that must be tested to determine if a solution is an $r$-bit local optima is \BigO{N}. This means that
 while on Nearest Neighbor NK with $N=8000$, $K=5$, and $r=3$ the black-box method would require 85 billion
@@
 The efficiency analysis is unchanged with the exception that the constant increases exponentially with $r$ as
-the number of $delta$ values which must be updated each move increases. Still, with $c$, $r$, and $k$ constant
+the number of $delta$ values that must be updated each move increases. Still, with $c$, $r$, and $k$ constant
 with respect to $N$, it is possible to find $r$-bit hamming ball local optima in \BigO{N+I} time.

### 2015-06-01 Dissertation/chapter-internal.tex
which -> that

@@
 level has few traps set to the next worst local optima (3 bits set). As a result,
-solutions which reach the third level can only be improved by replacing the deceptive
+solutions that reach the third level can only be improved by replacing the deceptive
 local optima (0 and 1 bits set) with the global optimum (7 bits set).

### 2015-06-01 Dissertation/chapter-introduction.tex
Removed "very" and cleaned up some related text
which -> that

@@
 to challenging real world problems with minimal prior knowledge about the problem itself.
-This driving goal has created search algorithms which can escape user bias to create
+This driving goal has created search algorithms that can escape user bias to create
 truly novel results, sometimes publishable or patentable in their own right~\cite{kannappan:2014:humies}.
@@
 
-One area that has been very effective at reducing the number of algorithm parameters is
+One area that has been effective at reducing the number of algorithm parameters is
 model based search. \cite{pelikan:2006:hboa}'s Hierarchical Bayesian Optimization

### 2015-06-01 Dissertation/chapter-p3.tex
which -> that

@@
 \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid (P3) as a method for
-performing optimization which does not require the user to provide any parameters. This is
+performing optimization that does not require the user to provide any parameters. This is
 achieved by combining efficient local search with the model building methods of LTGA using
@@
 Instead of choosing a single genetic donor for each cluster, P3 iterates over the
-population in a random order until a solution in the population is found which
+population in a random order until a solution in the population is found that
 has a least one gene different for that cluster of genes from the improving solution.
@@
 rebuild the model more frequently per solution in the population, it also performs
-a number of local search evaluations which are quite efficient, meaning theoretical
+a number of local search evaluations that are quite efficient, meaning theoretical
 comparisons of their speed are difficult to perform. As a final note, P3's repeated

### 2015-06-01 Dissertation/chapter-proposed.tex
which -> that

@@
 Therefore as a first step in exploring this new domain we propose to explore simple memetic algorithms~\cite{chen:2011:memetic}
-which make use of this hill climber. The goal is to determine if combination with methods like uniform crossover
+that make use of this hill climber. The goal is to determine if combination with methods like uniform crossover
 can result in improved search quality.
@@
 the gray-box domain makes high level variable linkage fully known. Therefore by leveraging this information it should
-be possible to more efficiently construct crossovers which are even more accurate than the inferred black-box ones.
+be possible to more efficiently construct crossovers that are even more accurate than the inferred black-box ones.
 
@@
 static linkage model for crossover or multiple random approximations? It may be possible to compute for
-a given problem the set of all useful crossovers which best balance exploitation with exploration. As
+a given problem the set of all useful crossovers that best balance exploitation with exploration. As
 this only needs to be done once per problem, the algorithm for creating the set can be relatively time consuming.
@@
 crossover attempts to cross entire traps on Deceptive Trap, as mixing values within a trap does not lead to the global optimum.
-Under this principle it seems reasonable to assert that crossovers which ensure variables which share a non-linear relationship
+Under this principle it seems reasonable to assert that crossovers that ensure variables which share a non-linear relationship
 are crossed as a group are better than those that do not. The task then is to decide how to handle subfunction overlap to allow
@@
 
-Beyond the randomly generated instances, we plan to seek out real world problems which can be expressed in this domain.
+Beyond the randomly generated instances, we plan to seek out real world problems that can be expressed in this domain.
 This will help solidify the conclusions about the applicability of these methods outside of the optimization community.
@@
 As the hamming ball hill climber has a user specified parameter $r$, we will need to explore this parameter space.
-Specifically we will want to ensure all comparisons use an appropriate value for $r$ which does not scale with
+Specifically we will want to ensure all comparisons use an appropriate value for $r$ that does not scale with
 $N$. This process will likely involve brute force testing all small $r$ values for moderately large problems in each class
-and then choosing the one which performs best. That $r$ value will then be used for all problem sizes for that problem class.
+and then choosing the one that performs best. That $r$ value will then be used for all problem sizes for that problem class.
 

### 2015-06-01 Dissertation/chapter-understanding.tex
which -> that

@@
 understanding the difficulty of a problem, and how best to design algorithms to deal
-with those difficulties. Conversely, understanding the characteristics which favor
-existing algorithms can help future researchers choose the algorithm which is
+with those difficulties. Conversely, understanding the characteristics that favor
+existing algorithms can help future researchers choose which algorithm is
 most likely to perform well on their problem. To further this end, we shall
@@
 parts of the search space without potentially missing any local optima. Consider a bit flip $X_i$.
-For gray-box problems there are only $ck$ other bit values which are needed to determine if flipping
+For gray-box problems there are only $ck$ other bit values that are needed to determine if flipping
 $X_i$ is a fitness improvement. If a given set of values for $X_i$ and those $ck$ bits results in
-$X_i$ being a fitness improvement, then any solution which contains all of those values cannot be a local optima.
+$X_i$ being a fitness improvement, then any solution that contains all of those values cannot be a local optima.
 
 This knowledge can be effectively exploited by intelligently skipping parts of the enumeration.
-Before starting search we can determine the least significant bit in the solution which effects
-each possible fitness improving move. This is the bit which must be flipped in order to effect
-the quality of that move. As a result, when determine what solution to test next, all steps
-of the enumeration which do not result in the least significant bit of a fitness improve being
+Before starting search we can determine the least significant bit in the solution that effects
+each possible fitness improving move. This is the bit that must be flipped in order to effect
+the quality of that move. As a result, when determining what solution to test next, all steps
+of the enumeration that do not result in the least significant bit of a fitness improvement being
 flipped can be skipped. Furthermore, if there are multiple fitness improving moves, all must
@@
 When performing subsequent checks, not all moves needs to be retested for improvement. Instead, the highest
-bin which must be tested is the location of the most significant bit effected by the previous iteration. This is because
+bin that must be tested is the location of the most significant bit effected by the previous iteration. This is because
 the previous iteration has verified all higher moves are not fitness improving, and no action performed during
@@
 As a final efficiency, the order in which bits of the solution are considered can be remapped. When a flip is fitness
-improving, the amount of search space which is skipped depends on how significant its least significant bit is. Therefore,
+improving, the amount of search space that is skipped depends on how significant its least significant bit is. Therefore,
 by rearranging the order to make its least significant bit more significant, more search space can be skipped. We perform
@@
 the cost trade-off of doing so is somewhat unclear. By searching only for $r$-bit local optima, it is more likely
-that a move will exist which causes part of the landscape to be skipped. Yet the increase in total number of moves
+that a move will exist that causes part of the landscape to be skipped. Yet the increase in total number of moves
 requires more updating information each time the solution is changed. The exact relationship between the two
@@
 converting an entire trap to 1s, with the global optimum being all 1s. As a result, we should expect P3's method
-of elitist mixing to produce solutions which are progressively more and more similar to the global optimum until
+of elitist mixing to produce solutions that are progressively more and more similar to the global optimum until
 it is finally found.
@@
 The Ising Spin Glass instance shown in Figure~\ref{fig-valley-nn} provide some insight into the stalled behavior of Gray-Box
-P3 when solving that problem. In this landscape there are many local optima which have the second best fitness, but are very
+P3 when solving that problem. In this landscape there are many local optima that have the second best fitness, but are very
 different from the nearest global optima. From a search perspective this means selection can only get you within a certain distance
 of a global optima, and then it becomes unhelpful. This issue is not improved by increasing the local optima radius, as even with
-$r=4$ there are hundreds of local optima with the second best fitness in the landscape which are over 10 bits different (of 36) from
+$r=4$ there are hundreds of local optima with the second best fitness that are over 10 bits different (of 36) from
 the nearest global optima.
@@
 the fitness plateaus, as solutions are actually being moved away from the global optimum. Subsequent levels store solutions
-which are generally closer to the global optimum in both fitness and representation space. This change means that at higher
+that are generally closer to the global optimum in both fitness and representation space. This change means that at higher
 levels the frequency of traps being set to the global optima increases, resulting in better model building and better donation quality.
@@
 as the Deceptive Step Trap problem.  Shown with a logarithmic y-axis, stored solutions exhibit a near perfect relationship
-between fitness and representational distance. Again, using only hill climbing creates local optima which are significantly
+between fitness and representational distance. Again, using only hill climbing creates local optima that are significantly
 worse and further away than even
@@
 Figure~\ref{fig-level-un} gives insight into how different Unrestricted NKQ is from Nearest Neighbor NKQ. In this landscape
-Gray-Box P3, regardless of radius, appears to explore deceptive local optima which offer fitness improvements without moving
+Gray-Box P3, regardless of radius, appears to explore deceptive local optima that offer fitness improvements without moving
 closer to the global optima in representation space. In this problem there is a gap of over 1000 bits between the deceptive
@@
 
-In general it appears that the problems which have polynomial time solutions (Deceptive Step Trap, Nearest Neighbor NKQ, 2D Ising Spin Glass)
+In general it appears that the problems that have polynomial time solutions (Deceptive Step Trap, Nearest Neighbor NKQ, 2D Ising Spin Glass)
 share a similar behavior of selection leading to the global optimum. However, the NP-Hard problems (Unrestricted NKQ, MAX-SAT) seem to contain

### 2015-06-01 Dissertation/chapter-gray-box-benchmarking.tex
Removed "shall", tried to fix effect/affect

@@
 
-In this work we shall consider two methods for choosing the $K$ variables
+In this work we consider two methods for choosing the $K$ variables
 each subfunction depends on: Nearest Neighbor NKq and Unrestricted NKq.
@@
   \includegraphicsfit{fitness}
-  \caption{Comparison of how radius effects solution quality at termination. For NKq-Landscapes $N=6,000$ and $K=4$ and
+  \caption{Comparison of how radius affects solution quality at termination. For NKq-Landscapes $N=6,000$ and $K=4$ and
   for Ising Spin Glasses $N=6,084$. Range of radius values limited by memory constraints.}

### 2015-06-01 Dissertation/chapter-graybox.tex
Removed "shall", tried to fix effect/affect

@@
 
-In addition to the black-box evaluation function, we shall
+In addition to the black-box evaluation function, we
 add in two more exploitable features accessible to search algorithms:

### 2015-06-01 Dissertation/chapter-internal.tex
Removed "shall", tried to fix effect/affect

@@
   \end{centering}
-  \caption{Comparison of how reducing LTGA's population size effects
+  \caption{Comparison of how reducing LTGA's population size affects
            the median best fitness reached during search.}
@@
 No point is drawn if the median run had 0 solutions stored at that level. While pyramid size
-is effected by problem size, the overall shape is not. As such the behavior shown in Figure~\ref{fig-level-size}
+is affected by problem size, the overall shape is not. As such the behavior shown in Figure~\ref{fig-level-size}
 is representative of that for all other tested problem sizes.

### 2015-06-01 Dissertation/chapter-graybox.tex
fitness improving -> fitness-improving

@@
 domain. Each potential bit flip, referred to here as a move, requires the entire solution to be reevaluated, which takes $\Omega(N)$ time.
-Furthermore, each time a fitness improving move is made all previously tested moves must be tested again. As a result
+Furthermore, each time a fitness-improving move is made all previously tested moves must be tested again. As a result
 improving a random solution to be a local optimum using this algorithm requires between $\Omega(N^2)$ and \BigO{IN^2}, where $I$ is the number
@@
 
-When performing hill climbing, only moves $m$ such that $delta[m] > 0$ are fitness improving moves. Initially
+When performing hill climbing, only moves $m$ such that $delta[m] > 0$ are fitness-improving moves. Initially
 all moves are considered $open$, meaning that they could potentially be fitness improvements. Each iteration
-a random move $m$ is chosen from $open$, and $delta[m]$ is checked. If $m$ is a fitness improving move,
+a random move $m$ is chosen from $open$, and $delta[m]$ is checked. If $m$ is a fitness-improving move,
 \Call{MakeMove}{$m$}, shown in Figure~\ref{fig-make-move}, is called.
@@
 with how much the move improves over the new solution ($post\_both - just\_m$). As $delta[m']$ has updated, $m'$ could
-potentially become a fitness improving move and is therefore added into $open$. As $m$ was just flipped,
+potentially become a fitness-improving move and is therefore added into $open$. As $m$ was just flipped,
 it cannot be a fitness improvement and is therefore removed from $open$.  In total this requires less than

### 2015-06-01 Dissertation/chapter-internal.tex
fitness improving -> fitness-improving

@@
   \caption{For each problem Figure~\ref{fig-cross} shows the proportion of P3 evaluations spend on crossovers
-           and Figure~\ref{fig-cross-success} shows the percentage of fitness improving crossover evaluations.}
+           and Figure~\ref{fig-cross-success} shows the percentage of fitness-improving crossover evaluations.}
 \end{figure}
@@
   \caption{For each problem Figure~\ref{fig-level-size} shows the number of solutions stored in each level of the pyramid
-           and Figure~\ref{fig-level-success} shows the percentage of fitness improving crossover evaluations at each level.}
+           and Figure~\ref{fig-level-success} shows the percentage of fitness-improving crossover evaluations at each level.}
 \end{figure}

### 2015-06-01 Dissertation/chapter-understanding.tex
Removed "shall", tried to fix effect/affect
Rework of "Understanding" from writing group review

@@
 The ruggedness and high dimensionality of most interesting landscapes makes them challenging
-to visualize or otherwise understand. However, doing so can be helpful in
-understanding the difficulty of a problem, and how best to design algorithms to deal
-with those difficulties. Conversely, understanding the characteristics that favor
-existing algorithms can help future researchers choose which algorithm is
-most likely to perform well on their problem. To further this end, we shall
-explore the landscapes used in previous sections in an attempt to understand
-what makes a landscape easy for P3 to solve.
+to visualize or otherwise analyze. However, doing so can be helpful in
+quantifying the difficulty of a problem, and how to best design algorithms to deal
+with those difficulties. Similarly, knowing which characteristics favor
+each existing algorithm can help researchers choose the algorithm
+most likely to perform well on their problem. To further this end, we
+explore the landscapes used in previous sections to understand
+what makes a landscape suited for P3 optimization.
 
 \section{Big Valley}
-One method for understanding the global structure of a landscape
+One method for visualizing the global structure of a landscape
 is to examine the relationship between local optima~\cite{boese:1994:bigvalley}.
-In its original form, this process involved generating thousands of random solutions and then applying
-local search to each. To convert this information into something human interpretable,
-the authors created two dimensional plots: distance from nearest global optima and
-fitness difference from global optima.
-The authors found that for a number of interesting problems a ``Big Valley'' of
-local optima exist, such that the higher a solution's fitness the closer it is
-in representation space to the global optimum. As a result, this suggests that
-by focusing search around known high quality solutions, the likelihood of finding even
-higher quality solutions is increased. This relationship is an underlying assumption
-of all evolutionary computation based search methods, including P3.
+In its original form, this process involves generating thousands of random solutions and then applying
+local search to each. This information is then displayed in two-dimensional plots:
+distance from the nearest global optimum and fitness difference from the global optima.
+For a number of interesting problems, a ``Big Valley'' of
+local optima exist, such that the higher a solution's fitness is, the closer it is
+in representation space to the global optimum. This result suggests that
+focusing search around known high-quality solutions increases the likelihood of finding even
+higher-quality solutions. This relationship is an underlying assumption
+of all evolutionary-based search methods, including P3.
 
@@
 just a sample of local optima, but all local and global optima in the landscape.
-When considered as a black-box, each solution must be enumerated and all of its
-neighbors must be checked for improvements. This enumeration is prohibitively expensive for
+When considered as a black-box, finding all local optima requires each solution to be enumerated and all of its
+neighbors to be checked for fitness improvements. This enumeration is prohibitively expensive for
 even trivial landscapes, requiring $\Omega(N2^N)$ operations.
@@
 
-Consider that for gray-box problems, we can determine all improving moves from a given
+Consider that for gray-box problems, we can determine all fitness-improving moves from a given
 solution in \BigO{N} time and that this list can be updated in \BigO{1} time when flipping
-a single bit. As a direct result, checking if each new solution is locally optimal requires
+a single bit. Therefore, checking if each new solution is locally optimal requires
 only \BigO{1} time and the overall enumeration process is reduced to \BigO{2^N}.
 
-Due to the limited non-linearity of the gray-box domain, its possible to exclude large
+Due to the limited non-linearity of the gray-box domain, it is possible to exclude large
 parts of the search space without potentially missing any local optima. Consider a bit flip $X_i$.
 For gray-box problems there are only $ck$ other bit values that are needed to determine if flipping
-$X_i$ is a fitness improvement. If a given set of values for $X_i$ and those $ck$ bits results in
-$X_i$ being a fitness improvement, then any solution that contains all of those values cannot be a local optima.
+$X_i$ is a fitness improvement. If flipping $X_i$ in a given solution is a fitness improvement
+then any solution that contains the same values of $X_i$ and its $ck$ dependencies cannot be a local optima.
 
-This knowledge can be effectively exploited by intelligently skipping parts of the enumeration.
-Before starting search we can determine the least significant bit in the solution that effects
-each possible fitness improving move. This is the bit that must be flipped in order to effect
+This knowledge can be effectively exploited to skip parts of the enumeration.
+For each fitness-improving move, we can determine the least significant bit it depends on.
+This is the minimum bit that must be flipped in order to change
 the quality of that move. As a result, when determining what solution to test next, all steps
-of the enumeration that do not result in the least significant bit of a fitness improvement being
-flipped can be skipped. Furthermore, if there are multiple fitness improving moves, all must
-have at least one dependent bit flipped, which is the same as saying the fitness improving move
+of the enumeration that do not flip the least significant bit of a fitness-improving move
+can be skipped. Furthermore, if there are multiple fitness-improving moves, all must
+have at least one dependent bit flipped. Restated, the fitness-improving move
 with the most-significant least-significant bit must have its least significant bit flipped.
 
-Finding which move's least significant bit needs to be flipped can also be done efficiently via binning.
+Finding which move's least significant bit needs to be flipped can be done efficiently via binning.
 Before starting search, each move can be binned based on its least significant bit. Initially
-search must check from the most significant bin down to find the first fitness improving move.
-This is the position from which enumeration must continue. If no move is fitness improving, a local
-optima has been found, and enumeration can continue with the least significant bit in the solution.
-Enumeration progresses by adding a 1 to the found position, with carry over as needed.
-If the carry ever exceeds the most significant bit in the solution, the entire space has been checked and iteration ends.
+search examines bins from the most significant to the least significant to find the first fitness-improving move.
+This is the position from which enumeration must continue. If no move is fitness-improving, a local
+optimum has been found, and enumeration can continue with the least significant bit in the solution.
+Enumeration progresses by adding a 1 to the found position, with carry as needed.
+If the carry exceeds the most significant bit in the solution, the entire space has been checked and iteration ends.
 
-When performing subsequent checks, not all moves needs to be retested for improvement. Instead, the highest
-bin that must be tested is the location of the most significant bit effected by the previous iteration. This is because
-the previous iteration has verified all higher moves are not fitness improving, and no action performed during
-that iteration can change that. Furthermore, no 1s can exist lower than that position, so bits
-below the least significant bit of the fitness improving move can be ignored.
+When performing subsequent checks, not all moves need to be retested for improvement. Instead, the most significant
+bin that must be tested is the location of the most significant bit flipped by the previous iteration. This
+simplification is possible because
+the previous iteration has verified that all moves in more significant bins are not fitness-improving, and no action performed during
+that iteration can make them fitness-improving. Furthermore, no 1s exist in less significant positions, meaning
+iteration can continue immediately from the found position.
 
-As a final efficiency, the order in which bits of the solution are considered can be remapped. When a flip is fitness
-improving, the amount of search space that is skipped depends on how significant its least significant bit is. Therefore,
+As a final efficiency, the order in which bits of the solution are considered can be remapped. When a move is fitness-improving,
+the amount of search space that is skipped depends on how significant its least significant bit is. Therefore,
 by rearranging the order to make its least significant bit more significant, more search space can be skipped. We perform
-this remapping in a greedy fashion, such that the flip with the least unmapped dependencies has all of its remaining dependencies
+this remapping in a greedy fashion, such that the move with the least-unmapped dependencies has all of its remaining dependencies
 mapped to the most significant remaining positions.
 
-All together, this can result in significant efficiency improvements for some landscapes. For instance, when applied to
+All together, these optimization result in substantial efficiency improvements for some landscapes. For example, when applied to
 the OneMax problem (each bit scores one if set to 1, zero otherwise) this method finds all local optima in \BigO{N} time.
 Deceptive Trap, regardless of how the bits are arranged, requires \BigO{2^k2^{N/k}} time to find all $2^{N/k}$ local optima.
-As an example, this means a 60 bit
+This means a 60 bit
 Deceptive Trap problem with trap size of 4 requires 524,288 operations when using gray-box, but 1,152,921,504,606,846,976
@@
 As a final note, these methods extend to finding only $r$-bit local optima for all gray-box problems. However,
-the cost trade-off of doing so is somewhat unclear. By searching only for $r$-bit local optima, it is more likely
-that a move will exist that causes part of the landscape to be skipped. Yet the increase in total number of moves
-requires more updating information each time the solution is changed. The exact relationship between the two
+the cost trade-off of doing so is unclear. By searching for only $r$-bit local optima, it is more likely
+that a fitness-improving move will exist, allowing more of the landscape to be skipped.
+Yet the increase in total number of moves
+means each time a bit is flipped more computation must be performed. Therefore, the runtime effect of increasing $r$
 depends on the specific problem.

### 2015-06-05 Dissertation/abstract.tex
punch edits

@@
 multiple populations that are iteratively created and expanded. In combination
-with local search and advanced crossover P3 scales to problem difficulty, exploiting
+with local search and advanced crossover, %wfp comma 
+P3 scales to problem difficulty, exploiting
 previously learned information before adding more diversity.
@@
 Across seven problems, each tested using on average 18 problem sizes, P3 outperformed
-all five advanced comparison algorithms. This improvement includes requiring less evaluations
+all five advanced comparison algorithms. This improvement includes
+requiring 
+%wfp change less 
+fewer evaluations
 and less wall clock seconds to find the global optimum and better fitness when using
-the same number of evaluations. Using both algorithm analysis and comparison we find P3's
+the same number of evaluations. Using both algorithm analysis and
+comparison we 
+% wfp change find 
+show P3's
 effectiveness is due to its ability to properly maintain, add, and exploit diversity.
 
-Unlike the best comparison algorithms, P3 was able to achieve this quality without any
-problem specific tuning. Thus, unlike previous parameter-less methods, P3 does not
+Unlike the best comparison algorithms, P3 was able to achieve this
+quality 
+% wfp emphasis
+\textit{without any
+problem specific tuning}. Thus, unlike previous parameter-less methods, P3 does not
 sacrifice quality for applicability. Therefore we conclude that
@@
 
-P3 can be further improved by specializing to the gray-box domain, which includes
-partial solution evaluation and known, limited, non-linear relationships between variables.
-This Gray-Box P3 leverages the Hamming-Ball Hill Climber, an exceptionally efficient
+% wfp
+Furthermore, P3 can be specialized for gray-box problems, problems with known
+linkage such as those with known, limited, non-linear relationships
+between variables.
+
+% wfp
+% P3 can be further improved by specializing to the gray-box domain, which includes
+% partial solution evaluation and known, limited, non-linear
+% relationships between variables.
+
+% wfp This 
+Gray-Box P3 leverages the Hamming-Ball Hill Climber, an exceptionally efficient
 form of local search, as well as a novel method for performing crossover using the

### 2015-06-05 Dissertation/chapter-benchmarking.tex
which -> that
punch edits

@@
 between techniques with different success rates.
-The maximum problem size used for each problem was set to be the largest size we could
-feasibly determine LTGA's optimal population size. For many larger problems results are not shown
-for hBOA due to the extreme computational cost required to optimally set the population size.
+% wfp
+For LTGA, the
+% The 
+maximum problem size used for each problem was set to be the largest, %wfp 
+% wfp
+optimal problem
+size we could
+feasibly determine.
+% wfp LTGA's optimal population size. 
+%wfp
+For HBOA,
+many larger problems results are not shown due to the extreme
+computational cost required to optimally 
+% wfp set 
+determine the population size.
 
@@
 \subsection{Model Building}
-Only techniques which explicitly built models of gene epistasis were able to solve the largest problem
+Only techniques that explicitly built models of gene epistasis were able to solve the largest problem
 instances. On single instance problems LTGA was more effective than hBOA, with hBOA outperforming
@@
 we examine this behavior in Figure~\ref{fig-fitness-over-time}. At regular intervals during optimization
-Figure~\ref{fig-fitness-over-time} shows the median of the best fitnesses found at that point of search across
-100 runs. For each problem we show the largest problem size for which we were able to successfully gather results
+Figure~\ref{fig-fitness-over-time} shows the median of the best
+fitnesses found at that
+%wfp
+time 
+point of search across
+100 runs. 
+%wfp For each problem we 
+Figure~\ref{fig-fitness-over-time} shows the largest problem size for which we were able to successfully gather results
 for all six algorithms, but the trends shown are representative of all larger problem sizes. The maximum reporting interval

### 2015-06-05 Dissertation/chapter-comparison_algorithms.tex
punch edits

@@
 \label{sec-hill-climber}
-Perhaps the simplest black-box search heuristic is stochastic local search, or hill climbing.
+Perhaps the simplest black-box search heuristic is stochastic local
+search, 
+% wfp or
+also know as %wfp
+hill climbing.
 This optimization technique focuses on improving a single solution until it reaches a local
@@
 \cite{doerr:2013:lambdalambda} presented the first genetic algorithm to provably show
-the advantages of performing crossover on One Max. This comparatively simple
+the advantages of performing crossover on 
+the problem known as % wfp
+One Max. 
+% wfp could use a reference for OneMax
+This comparatively simple
 algorithm maintains only a single individual and a self-controlled parameter $\lambda$.
@@
 some other gene $x_j$, with the children of that node representing how the decision was made. The
-leaves of each tree give the probability that $x_i$ should be set to each possible value.
+leaves of each tree give the probability that $x_i$ should be set to 
+% wfp each possible value.
+one of the possible gene values.
 
@@
 To generate a solution, the value of each gene $x_i$ is set using its corresponding decision tree $T_i$. Because
-the forest is directed acyclic, there must be an ordering of $T_i$ such that before $T_i$ is executed all
+the forest is directed acyclic, there must be an ordering of $T_i$
+such that, %wfp comma
+before $T_i$ is executed, %wfp comma
+all
 $x_j$ it uses to make decisions have already been set. As such, previous decisions made by other trees
@@
 offspring is discarded. This method is designed to preserve genetic diversity as only genetically similar
-solutions must compete on fitness.
+solutions % wfp must 
+compete on fitness.
 
@@
 between two clusters is 0, this means there are no individuals in the population that disrupt the relationships between the two
-clusters. Therefore when performing crossover there is no reason to believe a fitness improvement can be achieved
+clusters. Therefore, %wfp comma
+when performing crossover, %wfp comma
+there is no reason to believe a fitness improvement can be achieved
 by breaking the stored pattern. As such a cluster is only kept if its direct superset has a non-zero distance.
-As a final step Line~\ref{fig-cluster-creation-ordering} reorders $useful$ such that clusters appear in reversed
+As a final step, %wfp comma
+Line~\ref{fig-cluster-creation-ordering} reorders $useful$ such that clusters appear in reversed
 order from which they were added to $useful$. Thus the most linked clusters and those containing single genes
@@
 
-Each generation, each individual in the population undergoes crossover. In a single crossover event, each
+During each % wfp
+generation, 
+%wfp each 
+every individual in the population undergoes crossover. In a single crossover event, each
 cluster of genes $C_i$ in $useful$ is applied as a crossover mask. A random donor $d$
@@
 into the working solution. If a modification is made, an evaluation is then performed. If the crossover
-resulted in no worse fitness then the changes are kept. This allows for neutral drift across plateaus.
+resulted in no worse fitness then the changes are kept, %wfp comma
+allowing for neutral drift across plateaus.
 The resulting solution, which must be at least as fit as its parent, is then
@@
 LTGA has no explicit form of diversity control and has no method for introducing new genetic information once
-the population has converged. Therefore an LTGA run is considered converged once two consecutive populations
-contain the same unique solutions.
+the population has converged. Therefore an LTGA run is considered
+converged
+%wfp once 
+when two consecutive populations contain the same unique solutions.
 
@@
 doubles the population size until some success criteria is met and then performs bisection
-between the lowest successful and highest unsuccessful sizes. In this way the minimum population size
+between the 
+% wfp lowest 
+smallest successful and 
+% wfp highest 
+largest unsuccessful sizes. In this way the minimum population size
 that meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
@@
 
-Figure~\ref{fig-pop-sizes} shows the results from performing bisection. In general hBOA
+Figure~\ref{fig-pop-sizes} shows the results from performing bisection
+on a number of problems to be used as comparison benchmarks in
+Chapter~\ref{chap-problems}. % wfp
+In general hBOA
 required population sizes that were at least an order of magnitude larger than LTGA. Due to
-runtime and memory overhead, finding the optimal value for hBOA was much less tractable than LTGA
-for moderate to large problem sizes. LTGA's population size also grew significantly slower
+runtime and memory overhead, finding the optimal value for hBOA was
+much less tractable than 
+for %wfp
+LTGA
+% wfp for 
+on moderate to large problem sizes. LTGA's population size also grew significantly slower
 than hBOA's as problem size increased, especially on the two Trap problems and Rastrigin.
@@
 While not currently treated as a parameter, we also performed preliminary tests of
-integrating hill climbing into LTGA and hBOA. To match P3, we applied first improvement
+integrating hill climbing into LTGA and hBOA
+as this is the procedure used in P3. %wfp
+To match P3, we applied first improvement
 hill climbing to each algorithm's initial population.
@@
 required to reach the global optimum. As such, all further experiments
-use the unmodified published versions.
+use the unmodified, %wfp comma
+published versions.

### 2015-06-05 Dissertation/chapter-gray-box-benchmarking.tex
punch edits

@@
 $k$ remains fixed, the maximum number of subfunctions a variable appears in ($c$) can increase
-as $N$ increases. As a result some of the efficiency claims in Section~\ref{sec-hamming}
+as $N$ increases. As a result, %wfp comma
+some of the efficiency claims in Section~\ref{sec-hamming}
 may not be applicable.
@@
 Figure~\ref{fig-fitness} shows the effect on final solution fitness as $r$ increases. As expected
-from~\cite{chicano:2014:ball} the HBHC obtains higher quality as $r$ increases, with the magnitude
+from~\cite{chicano:2014:ball}, %wfp comma
+the HBHC obtains higher quality as $r$ increases, with the magnitude
 of the improvement decreasing. TUX has a similar relationship and
@@
 Reaching high quality solutions quickly can sometimes be more important than reaching the global optimum
-eventually. Figure~\ref{fig-solver-over} shows how solution quality progresses during
-search for each algorithm. HBHC and TUX have significant early delays caused by their high $r$ values. Larger
+eventually. Figure~\ref{fig-solver-over} shows how solution quality
+progresses 
+% wfp during search 
+for each algorithm. HBHC and TUX have significant early delays caused by their high $r$ values. Larger
 $r$'s require a large amount of initial partial evaluation before performing hill climbing. After one full iteration
@@
 
-Figure~\ref{fig-radius-over} further illustrates the effect of $r$ on Gray-Box P3. On Nearest Neighbor NKq
-and Ising Spin Glasses, increasing $r$ does not change the shape of the curve. Instead the quality reached is simply time shifted,
+Figure~\ref{fig-radius-over} further illustrates the effect of $r$ on
+Gray-Box P3. On 
+%wfp
+both
+Nearest Neighbor NKq
+and Ising Spin Glasses, increasing $r$ does not change the shape of
+the curve. Instead, %wfp comma
+the quality reached is simply time shifted,
 such that given more time higher $r$ values will reach the same quality. As a result, for these problems we conclude
@@
 
+%wfp comment. In both these figures it is hard to pick out
+%HBHC. Perhaps a comment in the caption would help?
+
 \begin{figure}
@@
 The results on Ising Spin Glass are similar, with a less extreme difference between
-Black-Box P3 and Gray-Box P3. In general Gray-Box is the fastest technique to
+Black-Box P3 and Gray-Box P3. In general, %wfp comma
+Gray-Box is the fastest technique to
 find the global optimum by an order of magnitude, with Black-Box P3's mean run finishing
@@
 with TUX occasionally performing better. When $N>2,000$, Gray-Box P3 finds better quality
-solutions that all other methods for every instance.  Using the Mann-Whitney test to compare the fitness
+solutions 
+%wfp that 
+%wfp
+than
+all other methods for every instance.  Using the Mann-Whitney test to compare the fitness
 of Gray-Box P3 and TUX when $N=10,000$ results in $p < 10^{-16}$. HBHC and Black-Box P3 only reach similar
@@
 In line with previous work, we have found that HBHC cannot effectively find global optima
-on problems with even moderate epistasis. In general it also obtains almost no improvement
+on problems with even moderate epistasis. In general, %wfp comma
+it also obtains almost no improvement
 in fitness after only a few restarts. We designed TUX as a simplistic way of choosing
@@
 problems. One explanation for this deviation is that Ising Spin Glasses require more
-exploration of equal fitness plateaus. For instance in Figure~\ref{fig-solver-over} and Figure~\ref{fig-radius-over}
+exploration of equal fitness plateaus. For instance, %wfp comma
+in Figure~\ref{fig-solver-over} and Figure~\ref{fig-radius-over}
 there is a significant pause in improvement when Gray-Box P3 reaches the second best fitness in the landscape.
@@
 A potential explanation is that in both Ising Spin Glass
-and Nearest Neighbor NKq the number of unique variables $r$
-or fewer steps away from a given variable is significantly lower than the worst case. This explains why on Unrestricted NKq
+and Nearest Neighbor NKq
+
+%wfp comment, following doesn't make sense to me.
+the number of unique variables $r$
+or fewer steps away from a given variable is significantly lower than
+the worst case.
+
+This explains why on Unrestricted NKq
 even moderately high $r$ values hit our memory limit.

### 2015-06-05 Dissertation/chapter-gray-box-p3.tex
punch edits

@@
 \chapter{Gray-Box P3}
+%wfp comment. First use of acronym HBHC. introduce it.
 P3 represents a natural method for integrating the HBHC into
@@
 First and foremost, linkage learning is no longer necessary. By definition
-the direct non-linear relationships between variables are known. As a result
-P3 no longer needs to store the pairwise frequency information that leads
-the black-box version to require \BigO{N^2} memory. Instead,
+the direct non-linear relationships between variables are known. As a
+result, %wfp comma
+%wfp 
+Black-box 
+P3 no longer needs to store the pairwise frequency information 
+% wfp that leadsthe black-box version to require 
+%wfp
+requiring \BigO{N^2} memory. Instead,
 we have developed a method for creating a linkage tree that learns clusters
@@
 meaning cluster creation is optimally efficient.
-The cluster splitting process has identical properties as random pivot quicksort,
+The cluster splitting process has identical properties 
+% wfp as 
+%wfp
+to 
+random pivot quicksort,
 meaning the sum of cluster sizes is \BigO{N\log N} in the average case.
@@
 
-Another efficiency gain possible due to the gray-box domain is that each time a
-donation is made, only the effected part of the solution's fitness needs to be
-recalculated. As a result the number of subfunction evaluations required
+Another efficiency gain 
+% wfp possible due to the gray-box domain 
+is that each time a donation is made, only the effected part of the solution's fitness needs to be
+recalculated. As a result, %wfp comma
+the number of subfunction evaluations required
 to determine the change in fitness is only \BigO{l}. This also

### 2015-06-05 Dissertation/chapter-graybox.tex
punch edits

@@
 domain. These problems are characterized by a complete lack of available
-problem specific information. All the algorithm has to perform search
-is a ``black-box'' which measures the quality of arbitrary proposed
-solutions. Optimizers that are successful in this domain generalize well
+problem specific information. 
+%wfp
+A black-box algorithm can only propose a solution and measure the
+quality of that solution, using only that information to inform its search.
+% wfp All the algorithm has to perform search
+% is a ``black-box'' which measures the quality of arbitrary proposed
+% solutions. 
+Optimizers that are successful in this domain generalize well
 as there are minimal requirements to apply them to a new problem.
@@
 algorithms very specific, such that outside of their target domain they
-either cannot be applied or have no expectation of quality. As such,
+either cannot be applied or 
+%wfp
+their application has no guarantee of search quality.
+% wfp have no expectation of quality. 
+As such,
 each must also be designed by hand for each new problem class, requiring
@@
 
-In between those two domains there is another: gray-box optimization. In this
+In between those two domains there is another
+%wfp
+domain
+: gray-box optimization. In this
 domain some features general to multiple problem classes are exploited, beyond
@@
 of a solution is determined by how often three consecutive bits are set
-to the same value. As a result the maximum fitness is $N$ (all zero or all one) and the minimum fitness
+to the same value. As a result, %wfp comma
+the maximum fitness is $N$ (all zero or all one) and the minimum fitness
 is 0.
@@
 
+% wfp comment. Somewhere you should define subfunction. It could be
+% the elements of an NK but you really mean it more broadly than
+% that. For example, you refer to "clause" on MAX-SAT, subfunction on
+% NK, but then only continue to talk about subfunction.
+
 The overall quality of a solution must be equal to the sum of applying all subfunctions to that
@@
 The mapping of variables to subfunctions (epistasis) must also be known, and
-the maximum number of variables participating in each subfunction is $k$.
+the maximum number of variables participating in each subfunction 
+%wfp 
+known,
+%wfp 
+% is $k$.
+%wfp 
+represented by the variable $k$.
 To achieve maximum efficiency $k$ should be constant in regards to problem size. The cost
@@
 As the mapping of variables to subfunctions is known, it is also possible to calculate $c$, which
-is the maximum number of subfunctions any variable participates in. For best efficiency $c$ should
-be constant in regards to problem size. For example for all Nearest Neighbor NK landscapes all variables
+is the maximum number of subfunctions any variable participates in. 
+%wfp
+Algorithms can most efficiently utilize $c$ when it is constant with
+respect to
+%wfp For best efficiency $c$ should
+% be constant in regards to 
+problem size. For example for all Nearest Neighbor NK landscapes all variables
 participate in exactly $c=k=K+1$ subfunctions. Randomly generated MAX-SAT instances have no guarantee
-that $c$ is constant, but in general the expected number of subfunctions in which a variable appears is set to 12.81.
+that $c$ is constant, but in general the expected number of
+subfunctions in which a variable appears is set to 12.81. %wfp where does this come from?
 Assuming that all subfunctions use $k$ variables and each variable appears in $c$ subfunctions
-provides a bound on the total number of subfunctions of $\frac{cN}{k}$.
+provides a bound on the total number of subfunctions %wfp of
+$\frac{cN}{k}$.
 
@@
 Section~\ref{sec-hill-climber} presented an efficient method for performing local search for the black-box
-domain. Each potential bit flip, referred to here as a move, requires the entire solution to be reevaluated, which takes $\Omega(N)$ time.
-Furthermore, each time a fitness-improving move is made all previously tested moves must be tested again. As a result
+domain. Each potential bit flip, referred to here as a 
+%wfp emphasis
+\textit{move}, requires the entire solution to be reevaluated
+%wfp , which takes
+%wfp 
+taking  
+$\Omega(N)$ time.
+Furthermore, each time a fitness-improving move is made all previously
+tested moves must be tested again. As a result, %wfp comma
 improving a random solution to be a local optimum using this algorithm requires between $\Omega(N^2)$ and \BigO{IN^2}, where $I$ is the number
 of improving moves. The lower bound is achieved when the while loop in Figure~\ref{fig-hc} runs a constant number of times.
-For instance on One Max, Deceptive Trap, Deceptive Step Trap, and HIFF all possible single bit improvements are found during
-the first time through the loop. The worst case is when each loop is expected to only make a single improving move, which causes
+For instance, %wfp comman
+on One Max, Deceptive Trap, Deceptive Step Trap, and HIFF all possible single bit improvements are found during
+the first 
+% wfp time
+%wfp 
+pass
+through the loop. The worst case is when each loop is expected to only make a single improving move, which causes
 all $N$ moves to be reevaluated once per improvement, such as on the Leading Ones problem.
@@
 
-To achieve this performance, this local search technique begins by determining the change in fitness
-caused by making each potential move $m$, denoted as $delta[m]$. Figure~\ref{fig-initialize-delta}
-calculates the $delta$ for each $m$ starting at a given solution, as well as the fitness of the solution.
-Here, $f_s$ evaluates
-the subfunction $s$ on the given solution, and $solution \oplus m$ is the result of making move $m$ on $solution$.
-For each of the $\frac{cN}{k}$ subfunctions \Call{InitializeDelta}{} must determine the fitness of that
-subfunction before and after making each of $k$ moves that overlap that subfunction. Combined this
-results in less than $cN2b(k)$ operations, which is \BigO{N} assuming $c$ and $b(k)$ do not grow with $N$.
-Also, this procedure calls $f_s$ only $k+1$ times more than is required to find the fitness of the solution itself.
+To achieve this performance, this local search technique begins by
+determining the change in fitness caused by making each potential move
+$m$, denoted as $delta[m]$. Figure~\ref{fig-initialize-delta}
+calculates the $delta$ for each $m$ starting at a given solution, as
+well as the fitness of the solution.  Here, $f_s$ evaluates the
+subfunction $s$ on the given solution, and $solution \oplus m$ is the
+result of making move $m$ on $solution$.  For each of the
+$\frac{cN}{k}$ subfunctions \Call{InitializeDelta}{} must determine
+the fitness of that subfunction before and after making each of $k$
+moves that overlap that subfunction. Combined, %wfp comma
+this results in less than $cN2b(k)$ operations, which is \BigO{N}
+assuming $c$ and $b(k)$ do not grow with $N$.  Also, this procedure
+calls $f_s$ only $k+1$ times more than is required to find the fitness
+of the solution itself.
 
-When performing hill climbing, only moves $m$ such that $delta[m] > 0$ are fitness-improving moves. Initially
-all moves are considered $open$, meaning that they could potentially be fitness improvements. Each iteration
-a random move $m$ is chosen from $open$, and $delta[m]$ is checked. If $m$ is a fitness-improving move,
+When performing hill climbing, only moves $m$ such that $delta[m] > 0$
+are fitness-improving moves. Initially all moves are considered
+$open$, meaning that they could potentially be fitness
+improvements. 
+% wfp Each 
+% wfp
+During each iteration a random move $m$ is chosen from $open$,
+and $delta[m]$ is checked. If $m$ is a fitness-improving move,
 \Call{MakeMove}{$m$}, shown in Figure~\ref{fig-make-move}, is called.
@@
 can be done in \BigO{1} time. However, at least for MAX-SAT, there is evidence that when used in combination with subsequent
-search heuristics the less greedy first improvement algorithm was more effective.
+search heuristics the less greedy, %wfp commad
+first improvement algorithm was more effective.
 
@@
 In a black-box setting, verifying that no $r$ bit flip can improve a solution requires testing all
-$n \choose r$ neighbors. This quickly becomes intractable as $r$ increases. However in the gray-box
+$n \choose r$ neighbors. This quickly becomes intractable as $r$
+increases. However, %wfp commad
+in the gray-box
 domain not all combinations need to be tested. Consider that if two variables do not participate
-in the same subfunction, the relationship between their effects is by definition additive. As such
+in the same subfunction, the relationship between their effects is, % wfp comma
+by definition,  % wfp comma
+additive. As such
 there is no way for flipping both together to be a fitness improvement without flipping one of them
@@
 
-Consider the 3-Equal problem.  If there are at least two loci between $x_i$ and $x_j$ they do
+Consider the 3-Equal problem.  If there are at least two loci between
+$x_i$ and $x_j$ 
+%wfp 
+then 
+they do
 not share a common subfunction. Therefore, the change in fitness resulting from flipping
-both is equal to $delta[x_i] + delta[x_j]$. As such if $delta[x_i] \leq 0$ and $delta[x_j] \leq 0$
+both is equal to $delta[x_i] + delta[x_j]$. 
+%wfp As such
+If $delta[x_i] \leq 0$ and $delta[x_j] \leq 0$
 then $delta[x_i,x_j] \leq 0$. Now consider a solution such that the first half is set to 0
@@
 if the corresponding variables participate in at least one subfunction together. Restated,
-there is only an edge if the two variables have a direct non-linear relationship. The maximum
+there is only an edge if the two variables have a direct, %wfp comma
+non-linear relationship. The maximum
 degree of a vertex in this graph is $c(k-1)$, making it sparse for sufficiently large $N$.
@@
 
+%wfp comment. Hamming ball, first use here. Should be defined (or
+%somehow described, referenced, something.
 In order to determine which moves in the hamming-ball must be evaluated,
@@
 iteratively, storing an initially empty list of solutions $T$, such that $|T|$ is equal to the height
-of the tournament. Iterative construction is possible because when a sub-bracket is complete only a single solution
+of the tournament. Iterative construction is possible because, %wfp comma
+when a sub-bracket is complete, %wfp comma
+only a single solution
 emerges and solutions only need to be stored until their partner is found.
@@
 requires an $r$ for the hill climber) and is relatively simple to implement. Even so
-it allows for learning from previous local optima and as Section~\ref{chapter-gray-box-benchmarking}
-will show it is quite effective at optimization.
+it allows for learning from previous local optima and, %wfp comma
+as Section~\ref{chapter-gray-box-benchmarking} will show, %wfp comma
+it is quite effective at optimization.

### 2015-06-05 Dissertation/chapter-internal.tex
punch edits

@@
 
-To achieve high success rates on randomly generated problem classes, LTGA and hBOA must use a population size
-large enough to solve the hardest instances in that class. Therefore these methods will have
-population sizes larger than necessary to solve the easiest instances in the class. Even
+To achieve high success rates on randomly generated problem classes,
+LTGA and hBOA must use a population 
+%wfp
+which was optimally sized
+% large enough 
+to solve the hardest instances in that class. Therefore these methods will have
+population sizes larger than necessary to solve 
+% wfp the easiest 
+easier instances in the class. Even
 on single instance problems, both methods will require population sizes large enough to
@@
 highlights how this can effect the required number of evaluations to reach the global optimum, showing
-the distribution of results when solving the largest size of each problem. On each problem
+the distribution of results when solving the largest size of each
+problem. On each problem
+% wfp
+except Ising Spin Glass and MAX-SAT, 
 LTGA has a much smaller difference between its best and worst runs. This makes sense as LTGA uses the same
-population size regardless of instance and progresses search generationally. In contrast P3 has a much
+population size regardless of instance and 
+% wfp
+because its search progresses 
+%wfp search 
+generationally. In contrast P3 has a much
 higher split, with many runs finishing very quickly.
@@
 less evaluations performing crossover and each crossover is more likely to be successful.
-Deceptive Trap and Rastrigin are the easiest problems to model epistasis, with local search quickly
+Deceptive Trap and Rastrigin are the easiest problems to
+% wfp
+capture
+% wfp  model 
+epistasis, with local search quickly
 reducing pairwise entropy in each. These are also the problems where P3 uses the least evaluations
@@
 
-When compared with LTGA, P3's crossover success rates are lower but similar in problem ordering.
+When compared with LTGA, P3's crossover success rates are lower but 
+% wfp
+show the same problem ordering, from least to most effective.
+% wfp similar in problem ordering.
 Counterintuitively, the use of hill climbing on the initial population reduces
@@
 Even when crossover success rates are quite low, such as Nearest Neighbor NK's 0.007\% success,
-the results from Section~\ref{sec-optimum} and Section~\ref{sec-overtime} show it's still critical to optimization. Without
+the results from Section~\ref{sec-optimum} and
+Section~\ref{sec-overtime} show 
+% wfp
+how critical this small percentage is to optimization.
+% wfp it's still critical to optimization. 
+Without
 crossover, P3's performance would be identical to the Random Restart Hill Climber, which was unable to solve even moderately sized problems
@@
 By decreasing in size, P3 implicitly stores more diversity in low levels and focuses
-search on high quality solutions at higher levels. In comparison,
+search 
+%wfp on 
+%wfp
+for
+high quality solutions at higher levels. In comparison,
 LTGA and hBOA suboptimally use a fixed population size at each generation.

### 2015-06-05 Dissertation/chapter-introduction.tex
punch edits

@@
 However, the quality of solutions found, and the speed at which they are found, is strongly tied to setting these parameters
-correctly~\cite{goldberg:1991:gasize}. As such, either expert knowledge or exceedingly
+correctly~\cite{goldberg:1991:gasize}. As such, either expert
+knowledge or 
+% wfp exceedingly
 expensive parameter tuning~\cite{grefenstette:1986:optimalga} are required to leverage
@@
 \cite{rechenberg:1973:es} introduced self-adaptive parameters, in which parameter values
-were included in each solution's genome and themselves underwent evolution. This allowed search
+were included in each solution's genome and themselves underwent
+evolution. This allowed 
+% wfp search
+the search process itself 
 to optimize some of its own parameters, resulting in a reduced need for expert tuning.
@@
 This method uses a pyramid structure of populations to combine model based search with local search
-to achieve parameter-less optimization. Initial results suggest that unlike
-previous parameter-less methods, P3 is actually more efficient than current state-of-the-art
+to achieve parameter-less optimization. Initial results suggest that,
+% wfp comma
+unlike
+previous parameter-less methods, P3 is 
+% wfp actually 
+more efficient than current state-of-the-art
 parameterized search algorithms. In this work we shall: extend these results to cover more

### 2015-06-05 Dissertation/chapter-p3.tex
punch edits

@@
 with each population $P_i$ in the pyramid. If this process results in a strict fitness improvement and has
-created a solution not yet stored in the pyramid, it is added to the next highest pyramid level $P_{i+1}$.
+created a solution not yet stored in the pyramid, 
+% wfp it 
+that new solution is added to the next highest pyramid level $P_{i+1}$.
 If $P_{i+1}$ does not yet exist, it is created. In this way populations in the pyramid expand over time,
@@
 winners of a binary tournament. We can do this because even the
-worst solutions in the pyramid are already high quality due to local search.
+worst solutions in the pyramid are already high quality due to 
+%wfp
+the previous application of
+local search.
 Using local search in LTGA was examined by \cite{bosman:2011:lsbbo} and found
@@
 strictly improve to be added to the next level of the pyramid. This is balanced by continual
-integration of new random solutions. Furthermore, each random restart decreases the probability
+integration of new 
+%wfp 
+% random solutions.
+randomly generated, then locally optimized via hill climbing,
+solutions.
+Furthermore, each random restart decreases the probability
 of spurious linkages caused by shared ancestry. This diversity is further preserved by applying

### 2015-06-05 Dissertation/chapter-problems.tex
punch edits

@@
 
-Perhaps the most common such landscape is the Deceptive Trap problem \citep{goldberg:1991:gasize}.
+% wfp Perhaps the most common 
+One %wfp 
+such landscape is the Deceptive Trap problem \citep{goldberg:1991:gasize}.
 In this landscape the genome is broken up into $k$ bit non-overlapping subproblems referred
-to as traps.
+to as \textit{traps}. %wfp emphasis
 Each subproblem is scored using Equation~\ref{eq-deceptive-trap}, where $t$ is the number of
@@
 Trap problem trivial. This is because local search is able to optimize each trap to one of the two
-optima (all 1s or all 0s), and then linkage learning can have perfect knowledge of gene interactions.
+optima (all 1s or all 0s), 
+% wfp and then 
+providing linkage learning % wfp can have 
+with perfect knowledge of gene interactions.
 In order to make the problem more difficult, the authors proposed the Deceptive
@@
 These Nearest Neighbor NK landscapes are therefore ideal for search algorithm testing.
-For all of our experiments we fixed $K=5$ to ensure highly rugged landscapes.
+For all of our experiments
+%wfp
+using Nearest Neighbor NK
+we fixed $K=5$ to ensure highly rugged landscapes.
 
@@
 As our final class of randomly generated problems we chose the Maximum Satisfiability (MAX-SAT) problem.
-Related to the more common 3-SAT problem, a MAX-SAT instance is defined by a set of three term clauses.
+Related to the more common 3-SAT problem, a MAX-SAT instance is
+defined by a set of 
+three--term %wfp dash
+clauses.
 Each term is a randomly chosen variable, which may also be negated. A clause scores if an and only if

### 2015-06-05 Dissertation/chapter-understanding.tex
punch edits

@@
 quantifying the difficulty of a problem, and how to best design algorithms to deal
-with those difficulties. Similarly, knowing which characteristics favor
-each existing algorithm can help researchers choose the algorithm
+with those difficulties. 
+%wfp Similarly, knowing which characteristics favor
+%wfp
+Conversely, knowing which problem characteristics favor
+% wfp each existing 
+%wfp 
+a particular
+algorithm can help researchers choose the algorithm
 most likely to perform well on their problem. To further this end, we
@@
 local search to each. This information is then displayed in two-dimensional plots:
-distance from the nearest global optimum and fitness difference from the global optima.
+distance from the nearest
+%wfp global , did you mean global here or is local, below, correct?
+%wfp
+local
+optimum and fitness difference from the global optima.
 For a number of interesting problems, a ``Big Valley'' of
@@
 parts of the search space without potentially missing any local optima. Consider a bit flip $X_i$.
-For gray-box problems there are only $ck$ other bit values that are needed to determine if flipping
+For gray-box problems there are only $ck$ other bit values that 
+%wfp
+need to be checked to determine if
+% wfp are needed to determine if flipping
 $X_i$ is a fitness improvement. If flipping $X_i$ in a given solution is a fitness improvement
-then any solution that contains the same values of $X_i$ and its $ck$ dependencies cannot be a local optima.
+then any solution that contains the 
+%wfp same 
+%wfp
+original
+values of $X_i$ and its $ck$ dependencies cannot be a local optima.
 
 This knowledge can be effectively exploited to skip parts of the enumeration.
-For each fitness-improving move, we can determine the least significant bit it depends on.
+For each fitness-improving move, we can determine the least
+significant bit it depends on.
+%wfp comment. What does "minimum" mean here. Is it literally the least
+%significant bit of a subfunction?
 This is the minimum bit that must be flipped in order to change
@@
 of the pyramid. By comparing each solution with the nearest best found solution, we can create plots similar to the ``big valley''
-ones in the previous section, even for problems too large and complex to find all possible global optima. 
+%wfp ones 
+%wfp
+plots 
+in the previous section, even for problems too large and complex to find all possible global optima. 
 This also provides a look at how the different levels of the pyramid focus search on different areas of the landscape.

### 2015-06-09 Dissertation/abstract.tex
Accepted many of Bill's edits, made a few suggestions

@@
 multiple populations that are iteratively created and expanded. In combination
-with local search and advanced crossover, %wfp comma 
+with local search and advanced crossover,
 P3 scales to problem difficulty, exploiting
@@
 Across seven problems, each tested using on average 18 problem sizes, P3 outperformed
-all five advanced comparison algorithms. This improvement includes
-requiring 
-%wfp change less 
-fewer evaluations
+all five advanced comparison algorithms. This improvement includes requiring fewer evaluations
 and less wall clock seconds to find the global optimum and better fitness when using
-the same number of evaluations. Using both algorithm analysis and
-comparison we 
-% wfp change find 
-show P3's
+the same number of evaluations. Using both algorithm analysis and comparison we show P3's
 effectiveness is due to its ability to properly maintain, add, and exploit diversity.
 
-Unlike the best comparison algorithms, P3 was able to achieve this
-quality 
-% wfp emphasis
+Unlike the best comparison algorithms, P3 was able to achieve this quality
 \textit{without any
@@
 
-% wfp
-Furthermore, P3 can be specialized for gray-box problems, problems with known
-linkage such as those with known, limited, non-linear relationships
+% bwg: Slightly different from your version
+Furthermore, P3 can be specialized for gray-box problems, which have
+known, limited, non-linear relationships
 between variables.
-
-% wfp
-% P3 can be further improved by specializing to the gray-box domain, which includes
-% partial solution evaluation and known, limited, non-linear
-% relationships between variables.
-
-% wfp This 
 Gray-Box P3 leverages the Hamming-Ball Hill Climber, an exceptionally efficient

### 2015-06-09 Dissertation/chapter-benchmarking.tex
Accepted many of Bill's edits, made a few suggestions

@@
 between techniques with different success rates.
+% bwg: This section is talking about the largest problem size for ANY method (P3 and LTGA),
+% which is lost in your propsed edit.
+%The maximum problem size used for each problem was set to be the largest size we could
+%feasibly determine LTGA's optimal population size.
 % wfp
@@
 % wfp LTGA's optimal population size. 
-%wfp
 For HBOA,
-many larger problems results are not shown due to the extreme
+% bwg: Slighly reworded this.
+results on many large problems are not shown due to the extreme
 computational cost required to optimally 
-% wfp set 
 determine the population size.
@@
 Figure~\ref{fig-fitness-over-time} shows the median of the best
-fitnesses found at that
-%wfp
-time 
-point of search across
-100 runs. 
-%wfp For each problem we 
+fitnesses found at that time point of search across 100 runs.
+% bwg: Using "Figure~\ref{fig-fitness-over-time} shows" here seems repetative. Perhaps instead have "This figure displays"?
 Figure~\ref{fig-fitness-over-time} shows the largest problem size for which we were able to successfully gather results

### 2015-06-09 Dissertation/chapter-comparison_algorithms.tex
Accepted many of Bill's edits, made a few suggestions

--- 2015-06-01 Dissertation/chapter-comparison_algorithms.tex
@@
 \label{sec-hill-climber}
-Perhaps the simplest black-box search heuristic is stochastic local search, or hill climbing.
+Perhaps the simplest black-box search heuristic is stochastic local
+search, also know as hill climbing.
 This optimization technique focuses on improving a single solution until it reaches a local
@@
 \cite{doerr:2013:lambdalambda} presented the first genetic algorithm to provably show
-the advantages of performing crossover on One Max. This comparatively simple
+the advantages of performing crossover on
+% bwg: Instead of "the problem known as" how about "the One Max" problem?
+the problem known as % wfp
+One Max. 
+% wfp could use a reference for OneMax
+% bwg: I've looked, and One Max seems to be so common that no one cites a seminal paper for it.
+This comparatively simple
 algorithm maintains only a single individual and a self-controlled parameter $\lambda$.
@@
 some other gene $x_j$, with the children of that node representing how the decision was made. The
-leaves of each tree give the probability that $x_i$ should be set to each possible value.
+leaves of each tree give the probability that $x_i$ should be set to 
+one of the possible gene values.
 
@@
 To generate a solution, the value of each gene $x_i$ is set using its corresponding decision tree $T_i$. Because
-the forest is directed acyclic, there must be an ordering of $T_i$ such that before $T_i$ is executed all
+the forest is directed acyclic, there must be an ordering of $T_i$
+such that,
+before $T_i$ is executed,
+all
 $x_j$ it uses to make decisions have already been set. As such, previous decisions made by other trees
@@
 offspring is discarded. This method is designed to preserve genetic diversity as only genetically similar
-solutions must compete on fitness.
+solutions compete on fitness.
 
@@
 between two clusters is 0, this means there are no individuals in the population that disrupt the relationships between the two
-clusters. Therefore when performing crossover there is no reason to believe a fitness improvement can be achieved
+clusters. Therefore,
+when performing crossover,
+there is no reason to believe a fitness improvement can be achieved
 by breaking the stored pattern. As such a cluster is only kept if its direct superset has a non-zero distance.
-As a final step Line~\ref{fig-cluster-creation-ordering} reorders $useful$ such that clusters appear in reversed
+As a final step,
+Line~\ref{fig-cluster-creation-ordering} reorders $useful$ such that clusters appear in reversed
 order from which they were added to $useful$. Thus the most linked clusters and those containing single genes
@@
 
-Each generation, each individual in the population undergoes crossover. In a single crossover event, each
+During each
+generation,
+every individual in the population undergoes crossover. In a single crossover event, each
 cluster of genes $C_i$ in $useful$ is applied as a crossover mask. A random donor $d$
@@
 into the working solution. If a modification is made, an evaluation is then performed. If the crossover
-resulted in no worse fitness then the changes are kept. This allows for neutral drift across plateaus.
+resulted in no worse fitness then the changes are kept,
+% bwg: allowing -> which allows
+which allows for neutral drift across plateaus.
 The resulting solution, which must be at least as fit as its parent, is then
@@
 LTGA has no explicit form of diversity control and has no method for introducing new genetic information once
-the population has converged. Therefore an LTGA run is considered converged once two consecutive populations
-contain the same unique solutions.
+the population has converged. Therefore an LTGA run is considered
+converged
+when two consecutive populations contain the same unique solutions.
 
@@
 doubles the population size until some success criteria is met and then performs bisection
-between the lowest successful and highest unsuccessful sizes. In this way the minimum population size
+between the 
+smallest successful and the % bwg: added "the"
+largest unsuccessful sizes. In this way the minimum population size
 that meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
@@
 
-Figure~\ref{fig-pop-sizes} shows the results from performing bisection. In general hBOA
+% bwg: reworded slightly, changed chapter reference
+Figure~\ref{fig-pop-sizes} shows the results from performing bisection
+on all problems to be used as comparison benchmarks in
+Chapter~\ref{chap-benchmarking}.
+In general hBOA
 required population sizes that were at least an order of magnitude larger than LTGA. Due to
-runtime and memory overhead, finding the optimal value for hBOA was much less tractable than LTGA
-for moderate to large problem sizes. LTGA's population size also grew significantly slower
+runtime and memory overhead, finding the optimal value for hBOA was
+much less tractable than 
+for
+LTGA
+on moderate to large problem sizes. LTGA's population size also grew significantly slower
 than hBOA's as problem size increased, especially on the two Trap problems and Rastrigin.
@@
 While not currently treated as a parameter, we also performed preliminary tests of
-integrating hill climbing into LTGA and hBOA. To match P3, we applied first improvement
+integrating hill climbing into LTGA and hBOA
+as this is the procedure used in P3.
+To match P3, we applied first improvement
 hill climbing to each algorithm's initial population.
@@
 required to reach the global optimum. As such, all further experiments
-use the unmodified published versions.
+use the unmodified,
+published versions.

### 2015-06-09 Dissertation/chapter-gray-box-benchmarking.tex
Accepted many of Bill's edits, made a few suggestions

--- 2015-06-01 Dissertation/chapter-gray-box-benchmarking.tex
@@
 $k$ remains fixed, the maximum number of subfunctions a variable appears in ($c$) can increase
-as $N$ increases. As a result some of the efficiency claims in Section~\ref{sec-hamming}
+as $N$ increases. As a result,
+some of the efficiency claims in Section~\ref{sec-hamming}
 may not be applicable.
@@
 Figure~\ref{fig-fitness} shows the effect on final solution fitness as $r$ increases. As expected
-from~\cite{chicano:2014:ball} the HBHC obtains higher quality as $r$ increases, with the magnitude
+from~\cite{chicano:2014:ball},
+the HBHC obtains higher quality as $r$ increases, with the magnitude
 of the improvement decreasing. TUX has a similar relationship and
@@
 Reaching high quality solutions quickly can sometimes be more important than reaching the global optimum
-eventually. Figure~\ref{fig-solver-over} shows how solution quality progresses during
-search for each algorithm. HBHC and TUX have significant early delays caused by their high $r$ values. Larger
+eventually. Figure~\ref{fig-solver-over} shows how solution quality
+progresses 
+for each algorithm. HBHC and TUX have significant early delays caused by their high $r$ values. Larger
 $r$'s require a large amount of initial partial evaluation before performing hill climbing. After one full iteration
@@
 
-Figure~\ref{fig-radius-over} further illustrates the effect of $r$ on Gray-Box P3. On Nearest Neighbor NKq
-and Ising Spin Glasses, increasing $r$ does not change the shape of the curve. Instead the quality reached is simply time shifted,
+Figure~\ref{fig-radius-over} further illustrates the effect of $r$ on
+Gray-Box P3. On 
+both
+Nearest Neighbor NKq
+and Ising Spin Glasses, increasing $r$ does not change the shape of
+the curve. Instead,
+the quality reached is simply time shifted,
 such that given more time higher $r$ values will reach the same quality. As a result, for these problems we conclude
@@
 
+% bwg: Added HBHC comment to captions
+
 \begin{figure}
@@
   \caption{Comparison of how each algorithm's time required to reach the best fitness found scales with problem size
-  on Nearest Neighbor NKq with $K=4$. With $N=1000$ Gray-Box P3 is 375x faster than Black-Box P3.}
+  on Nearest Neighbor NKq with $K=4$. With $N=1000$ Gray-Box P3 is 375x faster than Black-Box P3.
+  HBHC was only successful on the smallest problem size.}
   \label{fig-length-nn}
@@
   \caption{Comparison of how each algorithm's time required to find the best fitness found scales with problem size
-  on Ising Spin Glass. With $N=2025$ Gray-Box P3 is 4.6x faster than Black-Box P3.}
+  on Ising Spin Glass. With $N=2025$ Gray-Box P3 is 4.6x faster than Black-Box P3.
+  HBHC was only successful on the smallest problem size.}
   \label{fig-length-is}
@@
 The results on Ising Spin Glass are similar, with a less extreme difference between
-Black-Box P3 and Gray-Box P3. In general Gray-Box is the fastest technique to
+Black-Box P3 and Gray-Box P3. In general,
+Gray-Box is the fastest technique to
 find the global optimum by an order of magnitude, with Black-Box P3's mean run finishing
@@
 with TUX occasionally performing better. When $N>2,000$, Gray-Box P3 finds better quality
-solutions that all other methods for every instance.  Using the Mann-Whitney test to compare the fitness
+solutions than all other methods for every instance.  Using the Mann-Whitney test to compare the fitness
 of Gray-Box P3 and TUX when $N=10,000$ results in $p < 10^{-16}$. HBHC and Black-Box P3 only reach similar
@@
 In line with previous work, we have found that HBHC cannot effectively find global optima
-on problems with even moderate epistasis. In general it also obtains almost no improvement
+on problems with even moderate epistasis. In general, it also obtains almost no improvement
 in fitness after only a few restarts. We designed TUX as a simplistic way of choosing
@@
 problems. One explanation for this deviation is that Ising Spin Glasses require more
-exploration of equal fitness plateaus. For instance in Figure~\ref{fig-solver-over} and Figure~\ref{fig-radius-over}
+exploration of equal fitness plateaus. For instance,
+in Figure~\ref{fig-solver-over} and Figure~\ref{fig-radius-over}
 there is a significant pause in improvement when Gray-Box P3 reaches the second best fitness in the landscape.
@@
 A potential explanation is that in both Ising Spin Glass
-and Nearest Neighbor NKq the number of unique variables $r$
-or fewer steps away from a given variable is significantly lower than the worst case. This explains why on Unrestricted NKq
+and Nearest Neighbor NKq 
+% bwg: Attempted to clarify this section
+the number of unique variables within a given radius
+from a given variable is significantly lower than
+the worst case.
+%
+This explains why on Unrestricted NKq
 even moderately high $r$ values hit our memory limit.

### 2015-06-09 Dissertation/chapter-gray-box-p3.tex
Accepted many of Bill's edits, made a few suggestions

--- 2015-06-01 Dissertation/chapter-gray-box-p3.tex
@@
 First and foremost, linkage learning is no longer necessary. By definition
-the direct non-linear relationships between variables are known. As a result
-P3 no longer needs to store the pairwise frequency information that leads
-the black-box version to require \BigO{N^2} memory. Instead,
+the direct non-linear relationships between variables are known. As a result,
+% bwg: Your suggestion for this section changed the meaning. I worked to clarify it
+Gray-Box P3 does not need to store the pairwise frequency information,
+reducing its required memory from \BigO{N^2} to \BigO{N}. Instead,
+%
 we have developed a method for creating a linkage tree that learns clusters
@@
 meaning cluster creation is optimally efficient.
-The cluster splitting process has identical properties as random pivot quicksort,
+The cluster splitting process has identical properties to random pivot quicksort,
 meaning the sum of cluster sizes is \BigO{N\log N} in the average case.
@@
 
-Another efficiency gain possible due to the gray-box domain is that each time a
+Another efficiency gain is that each time a
 donation is made, only the effected part of the solution's fitness needs to be
-recalculated. As a result the number of subfunction evaluations required
+recalculated. As a result, the number of subfunction evaluations required
 to determine the change in fitness is only \BigO{l}. This also

### 2015-06-09 Dissertation/chapter-graybox.tex
Accepted many of Bill's edits, made a few suggestions

@@
 problem specific information. 
-%wfp
 A black-box algorithm can only propose a solution and measure the
 quality of that solution, using only that information to inform its search.
-% wfp All the algorithm has to perform search
-% is a ``black-box'' which measures the quality of arbitrary proposed
-% solutions. 
 Optimizers that are successful in this domain generalize well
@@
 either cannot be applied or
-%wfp
 their application has no guarantee of search quality.
-% wfp have no expectation of quality. 
 As such,
@@
 
-In between those two domains there is another
-%wfp
-domain
-: gray-box optimization. In this
+In between those two domains there is another domain: gray-box optimization. In this
 domain some features general to multiple problem classes are exploited, beyond
@@
 of a solution is determined by how often three consecutive bits are set
-to the same value. As a result, %wfp comma
+to the same value. As a result,
 the maximum fitness is $N$ (all zero or all one) and the minimum fitness
@@
 
-% wfp comment. Somewhere you should define subfunction. It could be
-% the elements of an NK but you really mean it more broadly than
-% that. For example, you refer to "clause" on MAX-SAT, subfunction on
-% NK, but then only continue to talk about subfunction.
-
 The overall quality of a solution must be equal to the sum of applying all subfunctions to that
-solution. As a consequence each subfunction must be independently evaluable.
+solution,
+% bwg: added definition of subfunction
+where a subfunction can be any mapping from a subset of problem variables
+to a measure of quality.
+%
+As a consequence each subfunction must be independently evaluable.
 The mapping of variables to subfunctions (epistasis) must also be known, and
 the maximum number of variables participating in each subfunction 
-%wfp 
-known,
-%wfp 
-% is $k$.
-%wfp 
-represented by the variable $k$.
+% bwg: slightly different than your change
+is represented by the variable $k$.
+%
 To achieve maximum efficiency $k$ should be constant in regards to problem size. The cost
@@
 is the maximum number of subfunctions any variable participates in. 
-%wfp
-Algorithms can most efficiently utilize $c$ when it is constant with
+% bwg: Tried to clarify that algorithms don't "use" c, but work best when it is a constant
+Algorithms are most efficient when $c$ is constant with
 respect to
-%wfp For best efficiency $c$ should
-% be constant in regards to 
 problem size. For example for all Nearest Neighbor NK landscapes all variables
 participate in exactly $c=k=K+1$ subfunctions. Randomly generated MAX-SAT instances have no guarantee
-that $c$ is constant, but in general the expected number of
-subfunctions in which a variable appears is set to 12.81. %wfp where does this come from?
+% bwg: Filled in where the number comes from.
+that $c$ is constant, but the expected number of
+subfunctions in which a variable appears is equal to the clause to variable ratio times the clause size, or 12.81.
 Assuming that all subfunctions use $k$ variables and each variable appears in $c$ subfunctions
-provides a bound on the total number of subfunctions %wfp of
+provides a bound on the total number of subfunctions
 $\frac{cN}{k}$.
@@
 domain. Each potential bit flip, referred to here as a 
-%wfp emphasis
 \textit{move}, requires the entire solution to be reevaluated
-%wfp , which takes
-%wfp 
 taking  
@@
 Furthermore, each time a fitness-improving move is made all previously
-tested moves must be tested again. As a result, %wfp comma
+tested moves must be tested again. As a result,
 improving a random solution to be a local optimum using this algorithm requires between $\Omega(N^2)$ and \BigO{IN^2}, where $I$ is the number
 of improving moves. The lower bound is achieved when the while loop in Figure~\ref{fig-hc} runs a constant number of times.
-For instance, %wfp comman
+For instance,
 on One Max, Deceptive Trap, Deceptive Step Trap, and HIFF all possible single bit improvements are found during
-the first 
-% wfp time
-%wfp 
-pass
+the first pass
 through the loop. The worst case is when each loop is expected to only make a single improving move, which causes
@@
 the fitness of that subfunction before and after making each of $k$
-moves that overlap that subfunction. Combined, %wfp comma
+moves that overlap that subfunction. Combined,
 this results in less than $cN2b(k)$ operations, which is \BigO{N}
@@
 improvements. 
-% wfp Each 
-% wfp
 During each iteration a random move $m$ is chosen from $open$,
@@
 can be done in \BigO{1} time. However, at least for MAX-SAT, there is evidence that when used in combination with subsequent
-search heuristics the less greedy, %wfp commad
+search heuristics the less greedy,
 first improvement algorithm was more effective.
@@
 \label{sec-hamming}
-Another consequence of the gray-box domain is that increasing the neighborhood of local search becomes
-more tractable~\cite{chicano:2014:ball}.  Instead of improving solutions until no single bit flip
-is a fitness improvement, this can be expanded to no flip of $r$ or less bits can improve the solution.
+Another consequence of the gray-box domain is that increasing the hamming-ball radius of local search becomes
+more tractable~\cite{chicano:2014:ball}.
+% Added sentence here describing what a hamming-ball is.
+A hamming-ball is the collection of all solutions
+within a given hamming distance, or radius, from a given solution.
+Instead of improving solutions until no single bit flip
+% bwg: Introduced HBHC acronym here, reworked sentence a little
+is a fitness improvement, the Hamming-Ball Hill Climber (HBHC) finds solutions which
+cannot be improved by flipping $r$ or fewer bits.
 
@@
 $n \choose r$ neighbors. This quickly becomes intractable as $r$
-increases. However, %wfp commad
+increases. However,
 in the gray-box
 domain not all combinations need to be tested. Consider that if two variables do not participate
-in the same subfunction, the relationship between their effects is, % wfp comma
-by definition,  % wfp comma
+in the same subfunction, the relationship between their effects is,
+by definition,
 additive. As such
@@
 $x_i$ and $x_j$ 
-%wfp 
 then
@@
 both is equal to $delta[x_i] + delta[x_j]$. 
-%wfp As such
 If $delta[x_i] \leq 0$ and $delta[x_j] \leq 0$
@@
 if the corresponding variables participate in at least one subfunction together. Restated,
-there is only an edge if the two variables have a direct, %wfp comma
+there is only an edge if the two variables have a direct,
 non-linear relationship. The maximum
@@
 
-%wfp comment. Hamming ball, first use here. Should be defined (or
-%somehow described, referenced, something.
 In order to determine which moves in the hamming-ball must be evaluated,
@@
 iteratively, storing an initially empty list of solutions $T$, such that $|T|$ is equal to the height
-of the tournament. Iterative construction is possible because, %wfp comma
-when a sub-bracket is complete, %wfp comma
+of the tournament. Iterative construction is possible because,
+when a sub-bracket is complete,
 only a single solution
@@
 requires an $r$ for the hill climber) and is relatively simple to implement. Even so
-it allows for learning from previous local optima and, %wfp comma
-as Section~\ref{chapter-gray-box-benchmarking} will show, %wfp comma
+it allows for learning from previous local optima and,
+as Section~\ref{chapter-gray-box-benchmarking} will show,
 it is quite effective at optimization.

### 2015-06-09 Dissertation/chapter-internal.tex
Accepted many of Bill's edits, made a few suggestions

@@
 LTGA and hBOA must use a population 
-%wfp
-which was optimally sized
+% bwg: Went back to "large" as the key here is that hard instances cause bigger populations.
+which is large enough
+% which was optimally sized
 % large enough 
@@
 population sizes larger than necessary to solve 
-% wfp the easiest 
 easier instances in the class. Even
@@
 highlights how this can effect the required number of evaluations to reach the global optimum, showing
-the distribution of results when solving the largest size of each
-problem. On each problem
-% wfp
+the distribution of results when solving the largest size of each problem. On each problem
 except Ising Spin Glass and MAX-SAT, 
@@
 population size regardless of instance and 
-% wfp
 because its search progresses 
-%wfp search 
 generationally. In contrast P3 has a much
@@
 Deceptive Trap and Rastrigin are the easiest problems to
-% wfp
 capture
-% wfp  model 
 epistasis, with local search quickly
@@
 
-When compared with LTGA, P3's crossover success rates are lower but 
-% wfp
-show the same problem ordering, from least to most effective.
-% wfp similar in problem ordering.
+% bwg: Reworked this sentence
+P3's crossover success rates are lower than LTGA, but
+both produce the same ordering of the problems when using crossover success
+such that Nearest Neighbor NK is the least successful and Deceptive Trap is the most
+successful problem.
 Counterintuitively, the use of hill climbing on the initial population reduces
@@
 Section~\ref{sec-overtime} show 
-% wfp
 how critical this small percentage is to optimization.
-% wfp it's still critical to optimization. 
 Without
@@
 search 
-%wfp on 
-%wfp
-for
+% bwg: Search is focusing search in the area around high quality solutions, not "for" high quality solutions.
+around
 high quality solutions at higher levels. In comparison,

### 2015-06-09 Dissertation/chapter-introduction.tex
Accepted many of Bill's edits, made a few suggestions

--- 2015-06-01 Dissertation/chapter-introduction.tex
@@
 \label{chap-introduction}
-%TODO Explain epistasis
 A primary purpose of evolutionary optimization is to efficiently find good solutions
@@
 However, the quality of solutions found, and the speed at which they are found, is strongly tied to setting these parameters
-correctly~\cite{goldberg:1991:gasize}. As such, either expert knowledge or exceedingly
+correctly~\cite{goldberg:1991:gasize}. As such, either expert
+knowledge or 
 expensive parameter tuning~\cite{grefenstette:1986:optimalga} are required to leverage
@@
 \cite{rechenberg:1973:es} introduced self-adaptive parameters, in which parameter values
-were included in each solution's genome and themselves underwent evolution. This allowed search
+were included in each solution's genome and themselves underwent
+evolution. This allowed 
+the search process itself 
 to optimize some of its own parameters, resulting in a reduced need for expert tuning.
@@
 This method uses a pyramid structure of populations to combine model based search with local search
-to achieve parameter-less optimization. Initial results suggest that unlike
-previous parameter-less methods, P3 is actually more efficient than current state-of-the-art
+to achieve parameter-less optimization. Initial results suggest that,
+unlike
+previous parameter-less methods, P3 is 
+more efficient than current state-of-the-art
 parameterized search algorithms. In this work we shall: extend these results to cover more

### 2015-06-09 Dissertation/chapter-p3.tex
Accepted many of Bill's edits, made a few suggestions

--- 2015-06-01 Dissertation/chapter-p3.tex
@@
 with each population $P_i$ in the pyramid. If this process results in a strict fitness improvement and has
-created a solution not yet stored in the pyramid, it is added to the next highest pyramid level $P_{i+1}$.
+created a solution not yet stored in the pyramid, 
+that new solution is added to the next highest pyramid level $P_{i+1}$.
 If $P_{i+1}$ does not yet exist, it is created. In this way populations in the pyramid expand over time,
@@
 winners of a binary tournament. We can do this because even the
-worst solutions in the pyramid are already high quality due to local search.
+worst solutions in the pyramid are already high quality due to 
+the previous application of
+local search.
 Using local search in LTGA was examined by \cite{bosman:2011:lsbbo} and found
@@
 strictly improve to be added to the next level of the pyramid. This is balanced by continual
-integration of new random solutions. Furthermore, each random restart decreases the probability
+integration of new 
+% bwg: Removed "via hill climbing" as this sentence is already quite compound.
+randomly generated, then locally optimized,
+solutions.
+Furthermore, each random restart decreases the probability
 of spurious linkages caused by shared ancestry. This diversity is further preserved by applying

### 2015-06-09 Dissertation/chapter-problems.tex
Accepted many of Bill's edits, made a few suggestions

--- 2015-06-01 Dissertation/chapter-problems.tex
@@
 
-Perhaps the most common such landscape is the Deceptive Trap problem \citep{goldberg:1991:gasize}.
+One
+such landscape is the Deceptive Trap problem \citep{goldberg:1991:gasize}.
 In this landscape the genome is broken up into $k$ bit non-overlapping subproblems referred
-to as traps.
+to as \textit{traps}.
 Each subproblem is scored using Equation~\ref{eq-deceptive-trap}, where $t$ is the number of
@@
 Trap problem trivial. This is because local search is able to optimize each trap to one of the two
-optima (all 1s or all 0s), and then linkage learning can have perfect knowledge of gene interactions.
+optima (all 1s or all 0s), 
+providing linkage learning
+with perfect knowledge of gene interactions.
 In order to make the problem more difficult, the authors proposed the Deceptive
@@
 These Nearest Neighbor NK landscapes are therefore ideal for search algorithm testing.
-For all of our experiments we fixed $K=5$ to ensure highly rugged landscapes.
+For all of our experiments
+using Nearest Neighbor NK
+we fixed $K=5$ to ensure highly rugged landscapes.
 
@@
 As our final class of randomly generated problems we chose the Maximum Satisfiability (MAX-SAT) problem.
-Related to the more common 3-SAT problem, a MAX-SAT instance is defined by a set of three term clauses.
+Related to the more common 3-SAT problem, a MAX-SAT instance is
+defined by a set of 
+% bwg: Why -- (two hypens) and not - (one hypen)?
+three--term
+clauses.
 Each term is a randomly chosen variable, which may also be negated. A clause scores if an and only if

### 2015-06-09 Dissertation/chapter-understanding.tex
Accepted many of Bill's edits, made a few suggestions

@@
 with those difficulties. 
-%wfp Similarly, knowing which characteristics favor
-%wfp
-Conversely, knowing which problem characteristics favor
-% wfp each existing 
-%wfp 
+% bwg: This isn't really in opposition to the previous statement, so I think "similarly" is better than "conversely"
+Similarly, knowing which characteristics favor
+%Conversely, knowing which problem characteristics favor
 a particular
@@
 %wfp global , did you mean global here or is local, below, correct?
-%wfp
-local
+%local
+% bwg: The plot is taking a local optima, finding the representation distance
+% to the nearest global optimum (axis one) and the fitness different from the best possible fitness (axis two).
+global
 optimum and fitness difference from the global optima.
@@
 For gray-box problems there are only $ck$ other bit values that 
-%wfp
 need to be checked to determine if
-% wfp are needed to determine if flipping
-$X_i$ is a fitness improvement. If flipping $X_i$ in a given solution is a fitness improvement
+% bwg: put "flipping" back in as X_i itself is just a bit, not a flip.
+flipping $X_i$ is a fitness improvement. If flipping $X_i$ in a given solution is a fitness improvement
 then any solution that contains the
-%wfp same 
-%wfp
 original
@@
 of the pyramid. By comparing each solution with the nearest best found solution, we can create plots similar to the ``big valley''
-%wfp ones 
-%wfp
 plots 

### 2015-06-15 Dissertation/chapter-understanding.tex
Proposed enumeration algorithm
Added enumeration diagram
Xi is a move, not a flip
Better enumeration figure
Integrated algorithm and figure into "Understanding" chapter text

@@
 
+\begin{figure*}
+  \centering
+  \includegraphicsfit{Enumerate}
+  \caption{Example change of enumeration ordering. The gray loci represent all dependencies
+           for some move $m_i$. By reordering, $m_i$'s lowest $index$ dependency improves from 2 to 4.}
+  \label{fig-enumerate}
+\end{figure*}
+
 Due to the limited non-linearity of the gray-box domain, it is possible to exclude large
-parts of the search space without potentially missing any local optima. Consider a bit flip $X_i$.
-For gray-box problems there are only $ck$ other bit values that 
-need to be checked to determine if
-% bwg: put "flipping" back in as X_i itself is just a bit, not a flip.
-flipping $X_i$ is a fitness improvement. If flipping $X_i$ in a given solution is a fitness improvement
-then any solution that contains the
-original
-values of $X_i$ and its $ck$ dependencies cannot be a local optima.
+parts of the search space without missing any local optima.
+Consider the representation presented in the top of Figure~\ref{fig-enumerate}. In a black-box
+domain, enumeration would progress as a binary counter, treating $index$ zero (symbol $A$ in the genome) as
+the least significant bit. This ensures that before changing $index$ $i$, all possible settings of $index$
+0 through $i-1$ have been tested. The gray-box domain makes it possible to skip combinations
+which cannot be local optima. In this example there exists a move $m_i$ which is a fitness improvement
+when enumeration starts (all variables set to 0). Due to the known relationships between variables
+we know that the quality of $m_i$ only depends on variables $C$, $E$, $F$, and $H$.
+Therefore, until one of those four variables are modified, the solution cannot be a local optima.
+As a result, solutions 00000000, 10000000, 01000000, and 11000000 cannot be local optima and can
+therefore be skipped during enumeration. More generally, if at any point during enumeration
+there exists a fitness-improving move, no local optima can exist until at least one
+dependency of that move is modified.
 
-This knowledge can be effectively exploited to skip parts of the enumeration.
-For each fitness-improving move, we can determine the least
-significant bit it depends on.
-%wfp comment. What does "minimum" mean here. Is it literally the least
-%significant bit of a subfunction?
-This is the minimum bit that must be flipped in order to change
-the quality of that move. As a result, when determining what solution to test next, all steps
-of the enumeration that do not flip the least significant bit of a fitness-improving move
-can be skipped. Furthermore, if there are multiple fitness-improving moves, all must
-have at least one dependent bit flipped. Restated, the fitness-improving move
-with the most-significant least-significant bit must have its least significant bit flipped.
+\begin{figure}
+  \begin{algorithmic}[1]
+  \Procedure{AllLocalOptima}{}
+    \State $solution \leftarrow \{0\}^N$
+    \State $found \leftarrow [~]$
+    \State $index \leftarrow N-1$
+    \While{$index < N$}
+      \State $index \leftarrow $\Call{FindNextIndex}{$index$}
+      \If{$index = -1$}
+        \State $found \leftarrow found + [solution]$
+        \State $index \leftarrow 0$
+      \EndIf
+      \While{$index < N$ \textbf{and} $solution[index] = 1$}\label{fig-enumeration-algorithm-counter}
+        \State \Call{MakeMove}{$index$}
+        \State $index \leftarrow index + 1$
+      \EndWhile
+      \If{$index < N$}
+        \State \Call{MakeMove}{$index$}\label{fig-enumeration-algorithm-insert}
+      \EndIf
+    \EndWhile
+    \State \Return $found$
+  \EndProcedure
 
-Finding which move's least significant bit needs to be flipped can be done efficiently via binning.
-Before starting search, each move can be binned based on its least significant bit. Initially
-search examines bins from the most significant to the least significant to find the first fitness-improving move.
-This is the position from which enumeration must continue. If no move is fitness-improving, a local
-optimum has been found, and enumeration can continue with the least significant bit in the solution.
-Enumeration progresses by adding a 1 to the found position, with carry as needed.
-If the carry exceeds the most significant bit in the solution, the entire space has been checked and iteration ends.
+  \Procedure{FindNextIndex}{$index$}
+    \While{$index \ge 0$}
+      \ForAll{$m \in move\_bin[index]$}
+        \If{$delta[m] \ge 0$}
+          \Return $index$
+        \EndIf
+      \EndFor
+      \State $index \leftarrow index-1$
+    \EndWhile
+    \State \Return $-1$
+  \EndProcedure
+  \end{algorithmic}
+  \caption{Algorithm to find all local optima of a given gray-box problem.
+           \textproc{MakeMove}, described in Figure~\ref{fig-make-move}, flips bit $index$
+           and updates the fitness effect $deta$ of making all moves dependent on $index$.
+           $move\_bin$ stores moves based on their lowest $index$ dependency.}
+  \label{fig-enumerate-algorithm}
+\end{figure}
 
-When performing subsequent checks, not all moves need to be retested for improvement. Instead, the most significant
-bin that must be tested is the location of the most significant bit flipped by the previous iteration. This
+This knowledge can be effectively exploited to skip parts of the enumeration,
+as shown in Figure~\ref{fig-enumerate-algorithm}.
+Before starting, each move is put into a table $move\_bin$
+based on that move's dependency with the minimum $index$ value. This is the first $index$ which
+can be modified by enumeration to change the fitness effect of making that move.
+In order to determine how much of the enumeration can be skipped, we must find
+the highest $index$ in $move\_bin$ which contains a fitness-improving move,
+as done by \Call{FindNextIndex}{}. If no move is fitness-improving,
+then a local optima has been found.
+
+The \Call{AllLocalOptima}{} algorithm in Figure~\ref{fig-enumerate-algorithm} at a high level
+works by repeatedly calling \Call{FindNextIndex}{} and adding 1 to the resulting $index$ position's value.
+Initially \Call{AllLocalOptima}{} uses \Call{FindNextIndex}{} to check all moves (initializes $index$
+to $N-1$). If at any point \Call{FindNextIndex}{} returns -1 then no move is fitness-improving and
+the current solution is added to the list of local optima $found$. \Call{AllLocalOptima}{} then
+adds a 1 to the $index$ returned by \Call{FindNextIndex}{} using the loop on Line~\ref{fig-enumeration-algorithm-counter}
+to perform carry operations and Line~\ref{fig-enumeration-algorithm-insert} to create the new 1 value.
+Iteration stops when the carry exceeds the solution length.
+
+When performing subsequent checks, not all moves need to be retested for improvement. Instead, the highest $index$
+bin that must be tested is the highest $index$ flipped by the previous iteration. This
 simplification is possible because
-the previous iteration has verified that all moves in more significant bins are not fitness-improving, and no action performed during
-that iteration can make them fitness-improving. Furthermore, no 1s exist in less significant positions, meaning
-iteration can continue immediately from the found position.
+the previous iteration has verified that all moves in higher $index$ bins are not fitness-improving, and no action performed during
+that iteration can make them fitness-improving. Furthermore, no 1s can exist in lower $index$ positions, meaning
+iteration can continue immediately from the found $index$.
 
-As a final efficiency, the order in which bits of the solution are considered can be remapped. When a move is fitness-improving,
-the amount of search space that is skipped depends on how significant its least significant bit is. Therefore,
-by rearranging the order to make its least significant bit more significant, more search space can be skipped. We perform
+As a final efficiency, the order in which variables in the solution are $index$ed can be remapped. When a move is fitness-improving,
+the amount of search space that is skipped depends on how high its lowest $index$ dependency is. Therefore,
+by rearranging the order to make its lowest $index$ dependency higher, more search space can be skipped. We perform
 this remapping in a greedy fashion, such that the move with the least-unmapped dependencies has all of its remaining dependencies
-mapped to the most significant remaining positions.
+mapped to the most significant remaining positions. Figure~\ref{fig-enumerate} shows how changing the $index$ order of variables
+improves $m_i$'s lowest $index$ dependency from 2 to 4. Now whenever $m_i$ is a fitness improvement, \Call{FindNextIndex}{}
+skips 4 times as much search space.
 

### 2015-06-17 Dissertation/abstract.tex
Forgot to update abstract to remove "wall clock time"

@@
 all five advanced comparison algorithms. This improvement includes requiring fewer evaluations
-and less wall clock seconds to find the global optimum and better fitness when using
+to find the global optimum and better fitness when using
 the same number of evaluations. Using both algorithm analysis and comparison we show P3's

### 2015-06-17 Dissertation/chapter-understanding.tex
Improved Understanding from Bill's suggested edits

@@
 
-Consider that for gray-box problems, we can determine all fitness-improving moves from a given
-solution in \BigO{N} time and that this list can be updated in \BigO{1} time when flipping
+For gray-box problems, we can determine the set of all fitness-improving moves from a given
+solution in \BigO{N} time and this set can be updated in \BigO{1} time when flipping
 a single bit. Therefore, checking if each new solution is locally optimal requires
-only \BigO{1} time and the overall enumeration process is reduced to \BigO{2^N}.
+\BigO{1} time and the overall enumeration process requires \BigO{2^N}.
 
@@
 0 through $i-1$ have been tested. The gray-box domain makes it possible to skip combinations
-which cannot be local optima. In this example there exists a move $m_i$ which is a fitness improvement
-when enumeration starts (all variables set to 0). Due to the known relationships between variables
+which cannot be local optima. In Figure~\ref{fig-enumerate} there exists a move $m_i$ which is a fitness improvement
+when enumeration starts (all variables set to 0). Due to the known relationships between variables,
 we know that the quality of $m_i$ only depends on variables $C$, $E$, $F$, and $H$.
@@
 Before starting, each move is put into a table $move\_bin$
-based on that move's dependency with the minimum $index$ value. This is the first $index$ which
+based on that move's lowest $index$ dependency. This is the first $index$ which
 can be modified by enumeration to change the fitness effect of making that move.
@@
 
-The \Call{AllLocalOptima}{} algorithm in Figure~\ref{fig-enumerate-algorithm} at a high level
+The \Call{AllLocalOptima}{} algorithm in Figure~\ref{fig-enumerate-algorithm}
 works by repeatedly calling \Call{FindNextIndex}{} and adding 1 to the resulting $index$ position's value.

### 2015-06-24 Dissertation/abstract.tex
Grammar fixes from Erik

@@
 The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
-evolutionary optimization without requiring any user specified parameters.
+evolutionary optimization without requiring any user-specified parameters.
 P3's primary innovation is to replace the generational model with a pyramid of
@@
 \textit{without any
-problem specific tuning}. Thus, unlike previous parameter-less methods, P3 does not
+problem-specific tuning}. Thus, unlike previous parameter-less methods, P3 does not
 sacrifice quality for applicability. Therefore we conclude that

### 2015-06-24 Dissertation/chapter-comparison_algorithms.tex
Grammar fixes from Erik

@@
 comparing with local search alone shows the advantages of P3's overall approach.
-The $1+(\lambda, \lambda)$ algorithm~\citep{doerr:2013:lambdalambda} is the current best theory supported simple genetic algorithm
+The $(1+(\lambda, \lambda))$ algorithm~\citep{doerr:2013:lambdalambda} is the current best theory supported simple genetic algorithm
 and its method of crossover is in some sense a macro-mutation just as in P3. hBOA and Parameter-less
@@
 This optimization technique focuses on improving a single solution until it reaches a local
-optimum. Here we use the first improvement hill climber defined by \cite{goldman:2014:p3}
+optimum. Here we use the first-improvement hill climber defined by \cite{goldman:2014:p3}
 and given in Figure~\ref{fig-hc}. This algorithm works by flipping each bit in a random
@@
 %%%%%%%%%%%% Lambda Lambda %%%%%%%%%%%%%%%%
-\section{$1+(\lambda, \lambda)$}
+\section{$(1+(\lambda, \lambda))$}
 \cite{doerr:2013:lambdalambda} presented the first genetic algorithm to provably show
@@
 The original formulation was designed specifically for unimodal landscapes and as such were
-not directly suitable for multimodal problems. \cite{goldman:2014:p3} extended $1+(\lambda, \lambda)$
+not directly suitable for multimodal problems. \cite{goldman:2014:p3} extended $(1+(\lambda, \lambda))$
 to include random restarts. As search stagnates, the $\lambda$ parameter increases in value. Eventually
@@
 relationships that should be preserved.
-The process begins by creating the set of sets $unmerged$ that tracks all top level clusters. Initially
+The process begins by creating the set of sets $unmerged$ that tracks all top-level clusters. Initially
 $unmerged$ contains single member sets for each gene. After each iteration the two sets with the minimum average pairwise
@@
 Instead, binary tournament is used to select half of the population. This is done to ensure the model is built
-using only high quality solutions, even during the first generation.
+using only high-quality solutions, even during the first generation.
 
@@
 would perform exactly $2N-2$ evaluations for each of the $\mu$ solutions in the population. This provides the amortizing evaluations
-required to make clustering only \BigO{N} operations per evaluation. However, by skipping some evaluations, its
+required to make clustering only \BigO{N} operations per evaluation. However, by skipping some evaluations, it is
 possible that clustering may be super-linear.
@@
 \Call{Cluster-Usage}{} rely on parameter values. LTGA does not provide any method for controlling or setting
-the population size, relying instead on a fixed user specified size.
+the population size, relying instead on a fixed user-specified size.
 
@@
 
-While four of the six algorithms in our experiments do not require any user specified parameters,
+While four of the six algorithms in our experiments do not require any user-specified parameters,
 hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
@@
 as this is the procedure used in P3.
-To match P3, we applied first improvement
+To match P3, we applied first-improvement
 hill climbing to each algorithm's initial population.

### 2015-06-24 Dissertation/chapter-conclusions.tex
Grammar fixes from Erik

@@
 
-Across a large number of problems and problem sizes P3 required less evaluations to reach the global optimum
-than optimally tuned state-of-the-art competitors. On single instance problems P3's improvement was by a
+Across a large number of problems and problem sizes P3 required fewer evaluations to reach the global optimum
+than optimally tuned state-of-the-art competitors. On single-instance problems P3's improvement was by a
 constant factor, while for the three randomly generated problem classes P3's improvement increased with problem size. This
 quality extends to intermediate points during evolution, with P3 generally reaching at least as high
-of fitness as the competitive techniques when using the same number of evaluations. While P3 does require
+a fitness as the competitive techniques when using the same number of evaluations. While P3 does require
 modeling overhead, the expense of this overhead is approximately linear with respect to genome size. There
 is some evidence that even when compared on wall clock time, P3 performs on par with
-the best comparison techniques. All of these achievements are made without any problem specific parameter
-tuning, making P3 easier to apply to new domains than its two closets competitors in quality.
+the best comparison techniques. All of these achievements are made without any problem-specific parameter
+tuning, making P3 easier to apply to new domains than its two closest competitors in quality.
 
 P3's quality is due to a number of desirable traits. First, mixing local search with model based
-crossover lets search focus on properly mixing high quality solutions. Second, by adding diversity only
+crossover lets search focus on properly mixing high-quality solutions. Second, by adding diversity only
 as necessary P3 tends to use the minimal amount of random initialization, unlike other techniques that must
-overcompensate with larger population sizes on single instance problems and consider the worst instance
+overcompensate with larger population sizes on single-instance problems and consider the worst instance
 when solving problem classes. Third, by heavily exploiting existing diversity before adding more P3 is able
-to reach high quality intermediate fitnesses quickly without prematurely converging. Fourth, the
+to reach high-quality intermediate fitnesses quickly without prematurely converging. Fourth, the
 nature of the pyramid's shape allows search to preserve a desirable proportion of diversity at
@@
 There are a number of meaningful avenues for future P3 experimentation. Perhaps the most
-pressing for practitioner acceptance is to apply P3 to real world problems and compare
-its results with other black-box or even problem specific heuristics. While parameter-less,
-P3 is currently limited to discrete, fixed length genomes evaluated using single objective
+pressing for practitioner acceptance is to apply P3 to real-world problems and compare
+its results with other black-box or even problem-specific heuristics. While parameter-less,
+P3 is currently limited to discrete, fixed-length genomes evaluated using single-objective
 fitness. These limitations can be relaxed with future work to make P3 more widely applicable.
@@
 search efficiency. Overcoming these limitations by using a new modeling technique may allow
-the pyramid model even greater flexibility. Similarly, while P3 is able to overcome low
-order deception via linkage learning, the iterative improvement method by which crossovers
-are made may mislead search on landscapes with higher order deception. This appears
-to be especially important on NP-Hard problems, where high quality solutions may be
+the pyramid model even greater flexibility. Similarly, while P3 is able to overcome low-order
+deception via linkage learning, the iterative improvement method by which crossovers
+are made may mislead search on landscapes with higher-order deception. This appears
+to be especially important on NP-Hard problems, where high-quality solutions may be
 very distant from the global optimum in representation space.
 
-While empirically effective, we believe its possible to derive theoretical foundations
+While empirically effective, we believe it is possible to derive theoretical foundations
 for Gray-Box P3. Specifically, this may be possible by considering the nature of crossover
-in terms similar to those use for finding $r$-bit local optima with HBHC. Consider
+in terms similar to those used for finding $r$-bit local optima with HBHC. Consider
 that for any two parents the non-linear effects of crossover are constrained by
-where they containing different bit values. As a result, subsets of their differences
+where they contain different bit values. As a result, subsets of their differences
 can become linearly separable, allowing for each to be considered independently of all others.
@@
 these improvements our results show P3 is highly efficient at finding
-global optima on both black-box and gray-box problems without any problem specific tuning.
+global optima on both black-box and gray-box problems without any problem-specific tuning.

### 2015-06-24 Dissertation/chapter-gray-box-p3.tex
Grammar fixes from Erik

@@
 As clusters are used to donate values between solutions, these
-moves are always in the direction of previously found high quality solutions.
-This assumes that the density of high quality solutions is higher than average between good solutions.
+moves are always in the direction of previously found high-quality solutions.
+This assumes that the density of high-quality solutions is higher than average between good solutions.
 
 Another efficiency gain is that each time a
-donation is made, only the effected part of the solution's fitness needs to be
+donation is made, only the affected part of the solution's fitness needs to be
 recalculated. As a result, the number of subfunction evaluations required
@@
 allows for Gray-Box P3 to efficiently reapply hill climbing after each
-donation as only effected moves need to be rechecked. As a result, a donation
+donation as only affected moves need to be rechecked. As a result, a donation
 and its resulting modifications from hill climbing are kept only if the
-new local optima is at least as fit as the solution before the donation occurred.
-All combined a single donation plus returning to a local optimum requires \BigO{l + I} time,
+new local optimum is at least as fit as the solution before the donation occurred.
+All combined, a single donation plus returning to a local optimum requires \BigO{l + I} time,
 while just the donation in Black-Box P3 requires \BigO{N}.

### 2015-06-24 Dissertation/chapter-problems.tex
Grammar fixes from Erik

@@
 landscapes. To be of interest these landscapes need to represent interesting
-and important aspects of real world problems.
+and important aspects of real-world problems.
 
@@
 
-Another challenging aspect of landscapes can be higher order relationships. The Hierarchical If
+Another challenging aspect of landscapes can be higher-order relationships. The Hierarchical If
 and only If (HIFF) problem~\cite{watson:1998:hiff} is designed to capture the difficulties of this class of problem.
@@
 small subsets lead toward solutions to larger subsets. However, a node can score if all genes are either
-all 1s or all 0s, meaning that to solve higher order subproblems it is necessary to perform crossovers
+all 1s or all 0s, meaning that to solve higher-order subproblems it is necessary to perform crossovers
 that preserve lower order solutions. This problem is a natural fit for LTGA as the linkage tree can
@@
 drawn from a problem class. When tested over a sufficiently large sample it is then possible
-to draw more general conclusions about the algorithms effectiveness. The challenge with these
+to draw more general conclusions about an algorithm's effectiveness. The challenge with these
 landscapes is determining the global optimum to gauge if an algorithm was successful.
@@
 Ising Spin Glasses can be polynomially solved.\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}
-In this subset the graph is restricted to be a two dimensional torus, edge weights are randomly set
+In this subset the graph is restricted to be a two-dimensional torus, edge weights are randomly set
 to either -1 or 1, and vertex values must be -1 or 1.
@@
 signs of the terms are set to ensure the target solution satisfies the clause. To ensure each
-problem is challenging we chose a clause to variable ratio of $4.27$~\citep{selman:1996:maxsat}.
+problem is challenging we chose a clause-to-variable ratio of $4.27$~\citep{selman:1996:maxsat}.

### 2015-06-24 Dissertation/chapter-understanding.tex
Missing s
Grammar fixes from Erik

@@
 For a number of interesting problems, a ``Big Valley'' of
-local optima exist, such that the higher a solution's fitness is, the closer it is
+local optima exists, such that the higher a solution's fitness is, the closer it is
 in representation space to the global optimum. This result suggests that
@@
 we know that the quality of $m_i$ only depends on variables $C$, $E$, $F$, and $H$.
-Therefore, until one of those four variables are modified, the solution cannot be a local optima.
+Therefore, until one of those four variables are modified, the solution cannot be a local optimum.
 As a result, solutions 00000000, 10000000, 01000000, and 11000000 cannot be local optima and can
@@
            \textproc{MakeMove}, described in Figure~\ref{fig-make-move}, flips bit $index$
-           and updates the fitness effect $deta$ of making all moves dependent on $index$.
+           and updates the fitness effect $delta$ of making all moves dependent on $index$.
            $move\_bin$ stores moves based on their lowest $index$ dependency.}
@@
 as done by \Call{FindNextIndex}{}. If no move is fitness-improving,
-then a local optima has been found.
+then a local optimum has been found.
 
@@
 
-All together, these optimization result in substantial efficiency improvements for some landscapes. For example, when applied to
+All together, these optimizations result in substantial efficiency improvements for some landscapes. For example, when applied to
 the OneMax problem (each bit scores one if set to 1, zero otherwise) this method finds all local optima in \BigO{N} time.
 Deceptive Trap, regardless of how the bits are arranged, requires \BigO{2^k2^{N/k}} time to find all $2^{N/k}$ local optima.
-This means a 60 bit
+This means a 60-bit
 Deceptive Trap problem with trap size of 4 requires 524,288 operations when using gray-box, but 1,152,921,504,606,846,976
@@
 \subsection{Looking at Problems}
-In comparison to the randomly generated problem instances of Nearest Neighbor NKQ and Ising Spin Glass,
+In comparison to the randomly generated problem instances of Nearest Neighbor NKq and Ising Spin Glass,
 Deceptive Trap and Deceptive Step Trap have comprehensible landscapes. Therefore, these problems represent
@@
   \end{centering}
-  \caption{Location and quality of local optima in comparison to the global optima with $N=30$ and $k=5$.}
+  \caption{Location and quality of local optima in comparison to the global optimum with $N=30$ and $k=5$.}
   \label{fig-valley-trap}
@@
 The Deceptive Trap problem has few local optima, each of which actually
-provide little deception in terms of ``big valley'' properties. Given any two local optima, the one
+provides little deception in terms of ``big valley'' properties. Given any two local optima, the one
 closer to the global optimum in fitness is also always closer to the global optimum in representation space.
@@
 leaving each trap either at the global optimum of all 1s or the local optimal value of all 0s or exactly one 1. Selection then
-becomes a near perfect predictor of distance to the global optimum.
+becomes a nearly perfect predictor of distance to the global optimum.
 
@@
   \includegraphicsfit{valley-nn}
-  \caption{Location and quality of local optima in comparison to the global optima for a representative Nearest Neighbor NKQ problem
+  \caption{Location and quality of local optima in comparison to the global optima for a representative Nearest Neighbor NKq problem
            with $N=60$ and $k=2$.}
@@
 
-Figure~\ref{fig-valley-nn} shows how the local optima are distributed for Nearest Neighbor NKQ. Unlike the trap problems,
-Nearest Neighbor NKQ exhibits a traditional big valley shape. Increasing the radius of the local optima significantly
+Figure~\ref{fig-valley-nn} shows how the local optima are distributed for Nearest Neighbor NKq. Unlike the trap problems,
+Nearest Neighbor NKq exhibits a traditional big valley shape. Increasing the radius of the local optima significantly
 reduces the total number of local optima, with those optima generally more similar to the global optima in both representation
-space and quality. Together this suggests that Nearest Neighbor NKQ is a good candidate for selection based methods like P3. Also,
+space and quality. Together this suggests that Nearest Neighbor NKq is a good candidate for selection-based methods like P3. Also,
 due to the general regularity and frequency of local optima, it may not be necessary to increase the radius in order to quickly find
@@
   \includegraphicsfit{valley-un}
-  \caption{Location and quality of local optima in comparison to the global optima for a representative Unrestricted NKQ problem
+  \caption{Location and quality of local optima in comparison to the global optima for a representative Unrestricted NKq problem
            with $N=60$ and $k=2$.}
@@
 
-While visually somewhat similar, Unrestricted NKQ shown in Figure~\ref{fig-valley-nn} suggests selection may be misleading.
-Consider the shape of the bottom of $r=1$. At a distance of about 15 bits there is a slight hump, such that high quality solutions
+While visually somewhat similar, Unrestricted NKq shown in Figure~\ref{fig-valley-nn} suggests selection may be misleading.
+Consider the shape of the bottom of $r=1$. At a distance of about 15 bits there is a slight hump, such that high-quality solutions
 are more frequent both closer and further from the global optimum than at the same distance.
 If search finds points in the area over 15 bits away from the global optimum, small modifications and elitist selection
-are more likely to lead away from the global optimum that toward it.
+are more likely to lead away from the global optimum than toward it.
 At higher radius values this issue becomes more apparent, with optima flattening out away
@@
 
-Similar to Unrestricted NKQ, the MAXSAT instance shown in Figure~\ref{fig-valley-sat} suggests selection can be misleading.
+Similar to Unrestricted NKq, the MAXSAT instance shown in Figure~\ref{fig-valley-sat} suggests selection can be misleading.
 There is almost a negative correlation between fitness difference from the global optimum and representation distance from the
@@
 One method for understanding how P3 performs search is to examine the types of solutions being stored at each level
-of the pyramid. By comparing each solution with the nearest best found solution, we can create plots similar to the ``big valley''
+of the pyramid. By comparing each solution with the nearest best-found solution, we can create plots similar to the ``big valley''
 plots 
@@
 Gray-Box P3's progression on Deceptive Step Trap, shown in Figure~\ref{fig-level-dst}, follows directly from our expectations.
-The local optima found using only hill climbing and stored in level 0 are quite poor and far from the global optima.
+The local optima found using only hill climbing and stored in level 0 are quite poor and far from the global optimum.
 The first application of crossover dramatically improves their quality. However, this improvement is likely just overcoming
@@
 that are generally closer to the global optimum in both fitness and representation space. This change means that at higher
-levels the frequency of traps being set to the global optima increases, resulting in better model building and better donation quality.
+levels the frequency of traps being set to their values in the global optimum increases, resulting in better model building and better donation quality.
 
@@
   \caption{Distribution of local optima stored at each level of Gray-Box P3 in relation to the best found by the run on
-           a Nearest Neighbor NKQ problem $N=6000$ and $K=4$.}
+           a Nearest Neighbor NKq problem $N=6000$ and $K=4$.}
   \label{fig-level-nn}
@@
 
-Nearest Neighbor NKQ, shown in Figure~\ref{fig-level-nn}, provides a landscape almost as free of higher order deception
-as the Deceptive Step Trap problem.  Shown with a logarithmic y-axis, stored solutions exhibit a near perfect relationship
+Nearest Neighbor NKq, shown in Figure~\ref{fig-level-nn}, provides a landscape almost as free of higher-order deception
+as the Deceptive Step Trap problem.  Shown with a logarithmic y-axis, stored solutions exhibit a near-perfect relationship
 between fitness and representational distance. Again, using only hill climbing creates local optima that are significantly
@@
 solutions, even $r=3$ is only able to equal a single crossover application. As before the higher level a solution is stored,
-generally the closer it is to the global optima in both fitness and representational distance.
+generally the closer it is to a global optima in both fitness and representational distance.
 
@@
   \caption{Distribution of local optima stored at each level of Gray-Box P3 in relation to the best found by the run on
-           an Unrestricted NKQ problem $N=6000$ and $K=4$.}
+           an Unrestricted NKq problem $N=6000$ and $K=4$.}
   \label{fig-level-un}
@@
 
-Figure~\ref{fig-level-un} gives insight into how different Unrestricted NKQ is from Nearest Neighbor NKQ. In this landscape
+Figure~\ref{fig-level-un} gives insight into how different Unrestricted NKq is from Nearest Neighbor NKq. In this landscape
 Gray-Box P3, regardless of radius, appears to explore deceptive local optima that offer fitness improvements without moving
-closer to the global optima in representation space. In this problem there is a gap of over 1000 bits between the deceptive
-local optima and those likely to lead to the best found solution. An interesting behavior on this problem is that the branch
+closer to a global optima in representation space. In this problem there is a gap of over 1000 bits between the deceptive
+local optima and those likely to lead to the best-found solution. An interesting behavior on this problem is that the branch
 between the two groups always seems to begin with a single solution in level 1. This suggests that the stored solutions near
@@
 Gray-Box P3's progression on Ising Spin Glass, shown in Figure~\ref{fig-level-is}, is much more well behaved than on Unrestricted
-NKQ. As has been typical, each application of crossover results in improved fitness, with the largest gains between the first few levels.
+NKq. As has been typical, each application of crossover results in improved fitness, with the largest gains between the first few levels.
 When approaching higher qualities these improvements also begin to translate into increased representational similarity to the global optimum.
 However, as discussed previously for this problem, we again see evidence for a large number of diverse solutions with the second best
-fitness. These solutions can be over 1000 bits different from the eventual best found solution. Increasing the radius of the hill climber
+fitness. These solutions can be over 1000 bits different from the eventual best-found solution. Increasing the radius of the hill climber
 does not seem to significantly overcome this issue.
@@
 
-Figure~\ref{fig-level-sat} shows that Gray-Box P3 acts more similarly on MAX-SAT to Unrestricted NKQ than any of the other problems. Again there
-appears to be deceptive local optima between 500 and 1000 bits away from the nearest improving solution. Again a single solution appears
-to be the ancestor of all of the solutions near the best found. This suggests that similar to Unrestricted NKQ, Gray-Box P3 will have a
+Figure~\ref{fig-level-sat} shows that Gray-Box P3 acts more similarly on MAX-SAT to Unrestricted NKq than on any of the other problems. Again there
+appear to be deceptive local optima between 500 and 1000 bits away from the nearest improving solution. Again a single solution appears
+to be the ancestor of all of the solutions near the best found. This suggests that
+similar to its performance on Unrestricted NKq, Gray-Box P3 will have a
 hard time finding the global optimum on MAX-SAT.
 
-In general it appears that the problems that have polynomial time solutions (Deceptive Step Trap, Nearest Neighbor NKQ, 2D Ising Spin Glass)
-share a similar behavior of selection leading to the global optimum. However, the NP-Hard problems (Unrestricted NKQ, MAX-SAT) seem to contain
-large amounts of higher order deception, making them challenging for selection based methods. This relationship warrants further investigation
+In general it appears that the problems that have polynomial-time solutions (Deceptive Step Trap, Nearest Neighbor NKq, 2D Ising Spin Glass)
+share a similar behavior of selection leading to the global optimum. However, the NP-Hard problems (Unrestricted NKq, MAX-SAT) seem to contain
+large amounts of higher-order deception, making them challenging for selection-based methods. This relationship warrants further investigation
 into other NP-Hard problems to determine if the repeated selection found in P3 and other evolutionary computation methods are able to overcome

### 2015-06-24 Dissertation/chapter-benchmarking.tex

@@
 Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
-and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
+and the smallest Ising Spin Glass. $(1+(\lambda, \lambda))$ has the most success outperforming P3, doing so
 on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
@@
 \subsection{Local Search}
-The Random Restart Hill Climber and $1+(\lambda, \lambda)$ are both relatively effective on small problem
+The Random Restart Hill Climber and $(1+(\lambda, \lambda))$ are both relatively effective on small problem
 sizes. This is especially true for the three randomly generated problem classes. These problems
-may contain relatively few local optima or just be exceptionally difficult for the model based algorithms.
-On Deceptive Trap and Deceptive Step Trap using 4 or less traps, $1+(\lambda, \lambda)$ performs significantly
+may contain relatively few local optima or just be exceptionally difficult for the model-based algorithms.
+On Deceptive Trap and Deceptive Step Trap using 4 or fewer traps, $(1+(\lambda, \lambda))$ performs significantly
 better than any other algorithm. We believe this is
-because $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
+because $(1+(\lambda, \lambda))$ is able to overcome deception by probabilistically
 flipping entire traps.
-This ability also leads $1+(\lambda, \lambda)$
+This ability also leads $(1+(\lambda, \lambda))$
 to outperform the Random Restart Hill Climber on all problems except Nearest Neighbor NK.
@@
 Climbing was unable to scale.
-For larger problems it also becomes increasingly unlikely for $1+(\lambda, \lambda)$ to make the right combination
+For larger problems it also becomes increasingly unlikely for $(1+(\lambda, \lambda))$ to make the right combination
 of changes required to reach the global optimum. This
@@
 Only techniques that explicitly built models of gene epistasis were able to solve the largest problem
-instances. On single instance problems LTGA was more effective than hBOA, with hBOA outperforming
+instances. On single-instance problems LTGA was more effective than hBOA, with hBOA outperforming
 LTGA on Nearest Neighbor NK and Ising Spin Glasses. This may be caused by the differences in modeling method:
-unlike the single instance problems, gene epistasis in the randomly generated
+unlike the single-instance problems, gene epistasis in the randomly generated
 problem classes cannot be be perfectly represented with a linkage tree.
@@
 and to model epistasis. Both techniques also only rebuild models once per generation. As the subproblems of HIFF
-are nested, it is unlikely that either technique can accurately model higher order epistasis before solving
+are nested, it is unlikely that either technique can accurately model higher-order epistasis before solving
 lower order subproblems. Therefore both methods require one generation per subproblem order.
@@
 \subsection{P3}
-Unlike the other model based methods, P3 generally outperforms
-both Random Restart Hill Climber and $1+(\lambda, \lambda)$ even on small problem sizes.
+Unlike the other model-based methods, P3 generally outperforms
+both Random Restart Hill Climber and $(1+(\lambda, \lambda))$ even on small problem sizes.
 Unlike the other local search methods, P3 outperforms LTGA and hBOA even on large problem sizes.
@@
 On Deceptive Trap P3's use of hill climbing ensures all traps are immediately optimized, allowing for perfect
-linkage detection and high quality donation.
+linkage detection and high-quality donation.
 On HIFF local search solves all pairwise subproblems, saving P3 a generation over LTGA and hBOA.
@@
 \subsection{Quantitative Comparison}
-Of 181 sample points, P3 had the highest median fitness in 121. In pairwise competition, $1+(\lambda, \lambda)$
+Of 181 sample points, P3 had the highest median fitness in 121. In pairwise competition, $(1+(\lambda, \lambda))$
 was the most likely to outperform P3, doing so on 50 sample points. LTGA, hBOA, and Parameter-less hBOA
@@
 \subsection{Local Search}
-Perhaps the most striking result is the quality of $1+(\lambda, \lambda)$. Until quite far into search this
-method performs better than both LTGA and hBOA. Given sufficient evaluations $1+(\lambda, \lambda)$ also outperforms
+Perhaps the most striking result is the quality of $(1+(\lambda, \lambda))$. Until quite far into search this
+method performs better than both LTGA and hBOA. Given sufficient evaluations $(1+(\lambda, \lambda))$ also outperforms
 Random Restart Hill Climbing on all 7 problems. For brief periods in the middle of search it performs the
 best of all techniques on: Deceptive Trap; Deceptive Step Trap; HIFF; Ising Spin Glass; and MAX-SAT problems.
-$1+(\lambda, \lambda)$'s ability to efficiently incorporate gene modifications of larger
-than one bit allows it to overcome the deception and plateaus in Deceptive Trap and Deceptive Step Trap, solve medium sized subproblems
+$(1+(\lambda, \lambda))$'s ability to efficiently incorporate gene modifications of larger
+than one bit allows it to overcome the deception and plateaus in Deceptive Trap and Deceptive Step Trap, solve medium-sized subproblems
 in HIFF, flip the signs on multiple adjacent bits in Ising Spin Glass, and cross plateaus in MAX-SAT. However,
@@
 Subsequent generations then make only minor fitness improvements. Once the model
-becomes accurate and the probability of a crossover using high quality genetic material increases sufficiently, LTGA
+becomes accurate and the probability of a crossover using high-quality genetic material increases sufficiently, LTGA
 enters a second period of rapid improvement.
@@
 
-Unlike the model based methods, which struggle until model accuracy improves, P3's iterative solution
+Unlike the model-based methods, which struggle until model accuracy improves, P3's iterative solution
 integration allows it to improve much more quickly. This behavior exists in most problems, but is easiest to understand on Deceptive Trap.
@@
 While it is common in evolutionary computation to assume the evaluation function will dominate algorithm
-complexity, in some domains this will not be true. Model based methods are especially likely to violate
+complexity, in some domains this will not be true. Model-based methods are especially likely to violate
 this norm. Therefore, in order to assess P3's quality in solving problems with efficient fitness functions,
@@
 As such this may cause some super-linear growth
-in P3, but its unlikely to be very high.
+in P3, but it is unlikely to be very high.
 
@@
 As P3 and LTGA require a similar asymptotic complexity per
-evaluation as the Hill Climber and $1+(\lambda, \lambda)$, no similar change in ordering occurs.
+evaluation as the Hill Climber and $(1+(\lambda, \lambda))$, no similar change in ordering occurs.
 
 \subsubsection{P3}
-When LTGA is optimally tuned to a single instance problem with an efficient evaluation function
+When LTGA is optimally tuned to a single-instance problem with an efficient evaluation function
 it can find the global optimum faster than P3 in terms of wall clock time. However, on randomly
@@
 
-On the four single instance problems LTGA not only finds the global optimum using less wall clock time,
+On the four single-instance problems LTGA not only finds the global optimum using less wall clock time,
 the factor speedup increases as problem length does. Na\"{i}vely this suggests LTGA is achieving

### 2015-06-24 Dissertation/chapter-gray-box-benchmarking.tex

@@
 a collection of randomly generated problem instances given a
-higher level problem class description. Each instance is described
+higher-level problem class description. Each instance is described
 by a series of $N$ subfunctions, each corresponding to a variable
@@
 time~\cite{wright:2000:solvingnk}, allowing comparisons of how quickly each optimization
-algorithm can find the global optima. Nearest Neighbor NKq also ensures that $c=k=K+1$
+algorithm can find a global optima. Nearest Neighbor NKq also ensures that $c=k=K+1$
 and that both $c$ and $k$ do not increase as $N$ increases, meaning the efficiency
@@
 The only algorithm parameter in the HBHC, TUX, and Gray-Box P3
-is the radius of the hamming-ball. Therefore our first experiments are designed to determine
+is the radius of the hamming ball. Therefore our first experiments are designed to determine
 the effect of this parameter on solution quality.
@@
   \caption{Comparison of solution quality during optimization on a log-log scale for different algorithms. For NKq-Landscapes $N=6,000$ and $K=4$ and
-  for Ising Spin Glasses $N=6,084$. Each algorithm uses its best found $r$ value.}
+  for Ising Spin Glasses $N=6,084$. Each algorithm uses its best-found $r$ value.}
   \label{fig-solver-over}
@@
 
-Reaching high quality solutions quickly can sometimes be more important than reaching the global optimum
+Reaching high-quality solutions quickly can sometimes be more important than reaching the global optimum
 eventually. Figure~\ref{fig-solver-over} shows how solution quality
@@
 Given a sufficiently large $r$ the HBHC can return sections of the crossover offspring to either
-parents' original version of a given subfunction. The HBHC is elitist meaning there is a bias toward
-returning to the better of the two parent's versions. Furthermore, by being so disruptive,
+parent's original version of a given subfunction. The HBHC is elitist meaning there is a bias toward
+returning to the better of the two parents' versions. Furthermore, by being so disruptive,
 uniform crossover potentially allows for the HBHC to also find unrelated improvements.
@@
 in Figure~\ref{fig-solver-over} and Figure~\ref{fig-radius-over}
-there is a significant pause in improvement when Gray-Box P3 reaches the second best fitness in the landscape.
+there is a significant pause in improvement when Gray-Box P3 reaches the second-best fitness in the landscape.
 Nothing in its design suggests that Gray-Box P3 should be more effective at neutral drift. Another potential
@@
 from a given variable is significantly lower than
-the worst case.
+in the worst case.
 %
@@
 likely to be only a time shift in quality. This is in contrast to $r$'s role in HBHC,
-where low $r$ values are never expected to reach the same quality has higher $r$ values.
+where low $r$ values are never expected to reach the same quality as higher $r$ values.
 Therefore we conclude that Gray-Box P3 maintains the out-of-the-box quality of Black-Box

### 2015-06-24 Dissertation/chapter-graybox.tex

@@
 domain. These problems are characterized by a complete lack of available
-problem specific information. 
+problem-specific information.
 A black-box algorithm can only propose a solution and measure the
@@
 
-However, for many real world applications there is potentially more information
-available. The other end of the spectrum is white box optimization, in which
+However, for many real-world applications there is potentially more information
+available. The other end of the spectrum is white-box optimization, in which
 the algorithm knows the exact problem class it is trying to solve. This domain
-is dominated by problem specific search
+is dominated by problem-specific search
 heuristics~\cite{een:2004:satsolver,sherrington:1975:isingsolver,wright:2000:solvingnk} that
@@
 just an evaluation function. The goal is to create optimization methods that
-benefits from these features without those methods becoming specialized to
+benefit from these features without those methods becoming specialized to
 a small set of problem classes.
@@
 \section{Problems in this domain}
-To understand the types of problems in this domain, lets first examine
+To understand the types of problems in this domain, let us first examine
 a simple artificial problem we'll call 3-Equal. The quality
@@
 
-Every problem discussed in Chapter~\ref{chap-problems} except HIFF fit into this domain. Deceptive
+Every problem discussed in Chapter~\ref{chap-problems} except HIFF fits into this domain. Deceptive
 Trap and Deceptive Step Trap both can have known epistasis (which bits are in which traps) and
@@
 
-Beyond MAX-SAT and Ising Spin Glasses, which are interesting real world problems unto themselves,
-many $NP$-Hard real world problems can be expressed in these two requirements. This is especially true
+Beyond MAX-SAT and Ising Spin Glasses, which are interesting real-world problems unto themselves,
+many $NP$-Hard real-world problems can be expressed in these two requirements. This is especially true
 of graph problems.
@@
 that $c$ is constant, but the expected number of
-subfunctions in which a variable appears is equal to the clause to variable ratio times the clause size, or 12.81.
+subfunctions in which a variable appears is equal to the clause-to-variable ratio times the clause size, or 12.81.
 Assuming that all subfunctions use $k$ variables and each variable appears in $c$ subfunctions
@@
       \State $fitness \leftarrow fitness + pre\_move$
-      \ForAll{$m \in effected\_moves(s)$}
+      \ForAll{$m \in affected\_moves(s)$}
         \State $post\_move \leftarrow f_s(solution \oplus m)$
@@
     \State $fitness \leftarrow fitness + delta[m]$
-    \ForAll{$s \in effected\_subfunctions(m)$}
+    \ForAll{$s \in affected\_subfunctions(m)$}
       \State $pre\_both \leftarrow f_s(solution)$
       \State $just\_m \leftarrow f_s(solution \oplus m)$
-      \ForAll{$m' \in effected\_moves(s)$}
+      \ForAll{$m' \in affected\_moves(s)$}
         \State $just\_m' \leftarrow f_s(solution \oplus m')$
@@
 change in fitness. This process requires updating all of the $delta$ values for the $k$ moves that interact with each of the $c$ subfunctions
-effected by $m$. This update replaces outdated information that used the original solution ($just\_m' - pre\_both$)
+affected by $m$. This update replaces outdated information that used the original solution ($just\_m' - pre\_both$)
 with how much the move improves over the new solution ($post\_both - just\_m$). As $delta[m']$ has updated, $m'$ could
@@
 search heuristics the less greedy,
-first improvement algorithm was more effective.
+first-improvement algorithm was more effective.
 
@@
 % Added sentence here describing what a hamming-ball is.
-A hamming-ball is the collection of all solutions
+A hamming ball is the collection of all solutions
 within a given hamming distance, or radius, from a given solution.
@@
 
-In a black-box setting, verifying that no $r$ bit flip can improve a solution requires testing all
+In a black-box setting, verifying that no $r$-bit flip can improve a solution requires testing all
 $n \choose r$ neighbors. This quickly becomes intractable as $r$
@@
 there is no way for flipping both together to be a fitness improvement without flipping one of them
-being a fitness improvement. Therefore it is not necessary to try all possible $r$ sized subsets
+being a fitness improvement. Therefore it is not necessary to try all possible $r$-sized subsets
 of the solution.
@@
 
-In order to determine which moves in the hamming-ball must be evaluated,
+In order to determine which moves in the hamming ball must be evaluated,
 we developed \Call{ConnectedInducedSubgraphs}{} given in Figure~\ref{fig-connected-subgraphs}.
@@
 increases in size by the full adjacency of $v$, which is bounded by $ck$.
-This creates a worse case complexity for a single top level call of
+This creates a worse-case complexity for a single top-level call of
 $\prod_{i}^{r} ick = r!(ck)^r$. This must be called once for each
@@
 As this algorithm finds all connected subsets in \BigO{N} time for a fixed $r$, the number of moves
-that must be tested to determine if a solution is an $r$-bit local optima is \BigO{N}. This means that
+that must be tested to determine if a solution is an $r$-bit local optimum is \BigO{N}. This means that
 while on Nearest Neighbor NK with $N=8000$, $K=5$, and $r=3$ the black-box method would require 85 billion
@@
 which results in $c$ increasing with $N$, the gray-box method still only requires approximately 375 million checks,
-two and a half orders of magnitude less than assuming a black-box.
+two-and-a-half orders of magnitude less than assuming a black-box.
 
@@
 efficiently find $r$-bit local optima~\cite{chicano:2014:ball}. The only change is that instead of having a move
-and $delta$ for each bit, there must be a move and $delta$ for each connected induced subgraph of $r$ or less bits.
+and $delta$ for each bit, there must be a move and $delta$ for each connected induced subgraph of $r$ or fewer bits.
 The efficiency analysis is unchanged with the exception that the constant increases exponentially with $r$ as
@@
 
-Hamming-Ball Hill Climbing is not sufficient to efficiently find the global optima
+Hamming-Ball Hill Climbing is not sufficient to efficiently find the global optimum
 on problems with even moderate epistasis~\cite{chicano:2014:ball}. This is because,
 like all random restart hill climbers, it relies on random initialization to fall
-inside the global optima's basin of attraction.
+inside the global optimum's basin of attraction.
 
@@
 
-Conceptually TUX iteratively builds a structure similar to a single elimination
+Conceptually TUX iteratively builds a structure similar to a single-elimination
 bracket for solutions. Each ``match'' in the tournament takes in two candidate solutions,

### 2015-06-24 Dissertation/chapter-internal.tex

@@
 of solutions stored is distributed on the largest problem sizes. Here we see that the behavior depends
-on the type of problem. On single instance problems P3's stored variance is relatively low, and generally
+on the type of problem. On single-instance problems P3's stored variance is relatively low, and generally
 higher than optimally tuned LTGA's population size. On randomly generated problem classes P3 has a much
@@
 Our procedure for tuning LTGA and hBOA outlined in Section~\ref{sec-tuning} involved finding
-the optimal population size for each class of problem. For real world black-box optimization this is
+the optimal population size for each class of problem. For real-world black-box optimization this is
 realistically the best either algorithm could hope for as tuning to a problem
@@
 easier instances in the class. Even
-on single instance problems, both methods will require population sizes large enough to
+on single-instance problems, both methods will require population sizes large enough to
 ensure the worst random initialization is diverse enough to solve the problem, which may be
@@
 population sizes tuned to reach the global optimum. As a result, both were exceptionally ineffective
-at quickly reaching high quality solutions. This is because unlike P3, these methods have an explicit
+at quickly reaching high-quality solutions. This is because unlike P3, these methods have an explicit
 trade off between optimal performance and intermediate performance caused by their population size parameter.
@@
 at early points during optimization, doing so reduces the robustness of the final solution found. This is in contrast to P3
-which balances high quality intermediate fitness without trading away eventual optimality.
+which balances high-quality intermediate fitness without trading away eventual optimality.
 \end{comment}
@@
 Together these figures provide some insight into the role of crossover within P3. The behavior
-for each is clearly problem dependent and generally asymptoticly stable as problem size increases.
+for each is clearly problem dependent and generally asymptotically stable as problem size increases.
 
 When solving problems where epistasis can be effectively detected and represented by a linkage tree, P3 tends to spend
-less evaluations performing crossover and each crossover is more likely to be successful.
+fewer evaluations performing crossover and each crossover is more likely to be successful.
 Deceptive Trap and Rastrigin are the easiest problems to
@@
 % bwg: Reworked this sentence
-P3's crossover success rates are lower than LTGA, but
+P3's crossover success rates are lower than LTGA's, but
 both produce the same ordering of the problems when using crossover success
@@
 Without
-crossover, P3's performance would be identical to the Random Restart Hill Climber, which was unable to solve even moderately sized problems
+crossover, P3's performance would be identical to the Random Restart Hill Climber, which was unable to solve even moderately-sized problems
 and quickly fell behind P3 in intermediate fitness quality. Therefore even infrequently successful crossover donations are
@@
 search 
-% bwg: Search is focusing search in the area around high quality solutions, not "for" high quality solutions.
+% bwg: Search is focusing search in the area around high-quality solutions, not "for" high-quality solutions.
 around
-high quality solutions at higher levels. In comparison,
+high-quality solutions at higher levels. In comparison,
 LTGA and hBOA suboptimally use a fixed population size at each generation.
@@
 At higher levels modeling becomes more accurate and donations contain higher frequencies
-of high quality building blocks, resulting in increased crossover success. The
+of high-quality building blocks, resulting in increased crossover success. The
 highest level of most problems has a low crossover success rate, as solutions
@@
 optima when using $k=7$ and $s=2$. Crossover easily overcomes the
-two bit plateaus, and as a result solutions in the second level generally do not
+two-bit plateaus, and as a result solutions in the second level generally do not
 contain the lowest fitness local optima (5 bits set) and the third

### 2015-06-24 Dissertation/chapter-introduction.tex

@@
 A primary purpose of evolutionary optimization is to efficiently find good solutions
-to challenging real world problems with minimal prior knowledge about the problem itself.
+to challenging real-world problems with minimal prior knowledge about the problem itself.
 This driving goal has created search algorithms that can escape user bias to create

### 2015-06-24 Dissertation/chapter-proposed.tex
Removed "shall", tried to fix effect/affect

@@
 In Chapter~\ref{chap-graybox} we presented the gray-box domain, which contains many interesting
-real world problems, and the improvements that can be made to local search in this domain.
+real-world problems, and the improvements that can be made to local search in this domain.
 This presents the following questions:
@@
 
-We shall tentatively refer to the first proposed algorithm of this type as the Simple Uniform crossover Tournament (SUT)
+We tentatively refer to the first proposed algorithm of this type as the Simple Uniform crossover Tournament (SUT)
 algorithm. Conceptually, SUT is meant to be the simplest way to combine uniform crossover with hamming ball hill climbing.
@@
 Initially all variables start in their own cluster. Each subfunction is then used in a random order such that
-if that subfunction overlaps multiple top level clusters it merges them to create a new cluster.
+if that subfunction overlaps multiple top-level clusters it merges them to create a new cluster.
 
@@
 instead of recalculating all $delta$ values and assuming all moves are open each time a crossover donation is made,
-only $delta$s and moves effected by the donation are updated. This means that hill climbing requires $O(|cluster|+I)$
+only $delta$s and moves affected by the donation are updated. This means that hill climbing requires $O(|cluster|+I)$
 time.
@@
 
-Beyond the randomly generated instances, we plan to seek out real world problems that can be expressed in this domain.
+Beyond the randomly generated instances, we plan to seek out real-world problems that can be expressed in this domain.
 This will help solidify the conclusions about the applicability of these methods outside of the optimization community.
 Similarly, we plan to compare the results of our gray-box methods with algorithms specifically designed for each problem
-domain. This provides an interesting contention between the time required to develop problem specific solvers and the
+domain. This provides an interesting contention between the time required to develop problem-specific solvers and the
 quality of using gray-box methods.
 
-As the hamming ball hill climber has a user specified parameter $r$, we will need to explore this parameter space.
+As the hamming ball hill climber has a user-specified parameter $r$, we will need to explore this parameter space.
 Specifically we will want to ensure all comparisons use an appropriate value for $r$ that does not scale with
@@
 \item How does black-box P3 compare with gray-box P3?
-\item How do the more generic gray-box methods compare with problem specific methods for solving real world problems?
+\item How do the more generic gray-box methods compare with problem-specific methods for solving real-world problems?
 \end{itemize}

### 2015-07-13 CV/cv.tex
Fix to citation in CV
Updated CV
Quick patch for CV

@@
     \textbf{Michigan State University}, East Lansing, Michigan \vspace{2mm}\\\vspace{1mm}%
-    \textsl{Doctor of Philosophy in Computer Science \& Engineering} \hfill \textbf{ August 2012 -- Present}\vspace{-3mm}\\\vspace{-1mm}%
+    \textsl{Postdoctoral Research Assistant in Computer Science \& Engineering} \hfill \textbf{ July 2015 -- present}
+    \textsl{Doctor of Philosophy in Computer Science \& Engineering} \hfill \textbf{ August 2012 -- July 2015}\vspace{-3mm}\\\vspace{-1mm}%
     \begin{list2}
@@
         \item GPA: 4.0/4.0
-        \item ABD, Expected Graduation: June 2015
     \end{list2}\vspace{-1.5mm}
@@
     ``Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid,''
-    \textbf{Invited Paper}, \textsl{Evolutionary Computation Journal} (Impact Factor 3.733), Accepted With Revisions
+    \textbf{Invited Paper}, \textsl{Evolutionary Computation Journal} (Impact Factor 3.733), In Press
 
@@
     D. R. White, J. McDermott, M. Castelli, L. Manzoni, B. W. Goldman,
-    G. Kronberger, W. Jaskowski, U.-M. O’Reilly, and S. Luke,
+    G. Kronberger, W. Jaskowski, U.-M. O'Reilly, and S. Luke,
     ``Better GP Benchmarks: Community Survey Results and Proposals,''

### 2015-08-12 Hyperplane Enumeration/main.tex
Stared work on the enumeration paper

--- 2013-05-07 Reducing Wasted Evaluations in CGP/main.tex
@@
 \usepackage{graphicx}
-\usepackage{verbatim}
+
 \usepackage{url}
-\usepackage{hhline}
-\newcommand{\includegraphicsfit}[1]
-{\includegraphics[width=.9\textwidth,height=\textheight,keepaspectratio]{#1}}
-%\urldef{\mailsa}\path|{alfred.hofmann, ursula.barth, ingrid.haas, frank.holzwarth,|
-%\urldef{\mailsb}\path|anna.kramer, leonie.kunz, christine.reiss, nicole.sator,|
-%\urldef{\mailsc}\path|erika.siebert-cole, peter.strasser, lncs}@springer.com|    
 \newcommand{\keywords}[1]{\par\addvspace\baselineskip
 \noindent\keywordname\enspace\ignorespaces#1}
+
+\newcommand{\BigO}[1]{$\mathcal{O}{(#1)}$}
 
@@
 % first the title is needed
-\title{Reducing Wasted Evaluations in Cartesian Genetic Programming}
+\title{Hyperplane Elimination for Quickly Enumerating MK Landscapes}
 
@@
 %
-\author{Brian W.~Goldman\and William F. Punch}
-%\author{Anonymous Author\and Anonymous Author}
+%\author{Brian W.~Goldman\and William F. Punch}
+\author{Anonymous Author\and Anonymous Author}
+
 %
@@
 % unless you accept that it will be published
-\institute{BEACON Center for the Study of Evolution in Action,\\
-Michigan State University, U.S.A.\\
-brianwgoldman@acm.org, punch@msu.edu}
-%\institute{Department\\ Organization\\ Email}
+%\institute{BEACON Center for the Study of Evolution in Action,\\
+%Michigan State University, U.S.A.\\
+%brianwgoldman@acm.org, punch@msu.edu}
+\institute{Department\\ Organization\\ Email}
 
@@
 \begin{abstract}
-Cartesian Genetic Programming~(CGP) is a form of Genetic Programming~(GP)
-where a large proportion of the genome is identifiably unused
-by the phenotype.  This can lead mutation to create offspring that are 
-genotypically different but phenotypically identical, and therefore do
-not need to be evaluated.  We investigate theoretically and empirically
-the effects of avoiding these otherwise wasted evaluations, and provide
-evidence that doing so reduces the median number of evaluations to solve
-four benchmark problems, as well as reducing CGP's sensitivity to the mutation
-rate.  The similarity of results across the problem set in combination with the
-theoretical conclusions supports the general need for avoiding these unnecessary
-evaluations.
+TODO
 
@@
 %\emph{abstract} environment.
-
-\keywords{cartesian genetic programming, mutation}
+\keywords{Landscape Understanding, Gray-Box}
 \end{abstract}
@@
 \section{Introduction}
-%TODO Consider change to "best results"
-In Genetic Programming~(GP) the most common metric for measuring algorithm complexity
-is the number of fitness evaluations required to solve
-a black box problem.  This metric assumes that the cost of evaluating an
-individual will dominate search times on real world problems.  Under this
-metric, search algorithms that best exploit evaluation information and avoid
-unnecessary evaluations will receive the best results.  Cartesian Genetic
-Programming~(CGP), a branch of GP, has a particular structure that allows
-for the detection of some phenotypically identical genotypes without
-requiring fitness evaluation.  No previous work has investigated if this
-property can be exploited to meaningfully improve CGP efficiency.  After discussing the
-potential theoretical gains of avoiding these evaluations, we propose and
-investigate three novel techniques designed to improve CGP's
-search.
+TODO Motivate using network stuff.
 
-\section{Cartesian Genetic Programming}
-\label{sec:cgp}
-Cartesian Genetic Programming is a variant of genetic programming in which
-individuals encode directed acyclic graphs (DAG).
-An in-depth explanation of CGP is provided in~\cite{miller:2011:chapter2}, but for
-our purposes the important features are:
-\begin{itemize}
-  \item DAGs are represented by a collection of nodes and genes specifying output locations.
-  \item Nodes contain genes describing both what function they perform
-  and how they connect to other nodes.
-  \item Offspring are created using mutation.
-  \item Offspring replace parents if they are of the same or greater fitness.
-\end{itemize}
-CGP's encoding method does not require a node
-to connect to an output location, which is required for it to
-affect the phenotype of the individual.
-These unconnected nodes and their genes are commonly called ``inactive''
-and the connected nodes and their genes are called ``active.''
-As the topology of the DAG is constantly evolving, the number and location of
-active nodes can change from parent to child.
-There have been numerous studies on the usefulness of inactive nodes~\cite{vassilev:2000:neutrality,yu:2001:neutrality}
-including an argument that it is desirable to have up
-to $95\%$ of the genome inactive~\cite{miller:2006:redundancy}.
+\section{MK Landscapes and Gray-Box Optimization}
+TODO Explain all problems here
 
-CGP's canonical variation operator is point mutation.  Unfortunately the details
-of this operator are often ambiguously stated.  For instance, in some papers this
-operator chooses a set number of genes at random to be mutated~\cite{miller:2011:chapter2},
-while in other papers each gene can be mutated with a certain probability,
-allowing any number of genes to be mutated at once~\cite{harding:2012:mtcgp}.  Also, mutation in CGP can either
-force a mutated gene's value to change~\cite{harding:2012:mtcgp,miller:2011:chapter2},
-or randomly reset it to any valid value including the genes previous value~\cite{miller:2006:redundancy}.
-As much of our analysis hinges on
-the behavior of mutation, we specify mutation as follows:
-when performing mutation, all genes use the mutation rate as the
-probability that they are changed to a different value.
-%Only if a gene has exactly one valid value\footnote{Genes can have just a single valid
-%value in the case of the first node's connection genes in single input problem
-%or if there is only a single valid function.}
-%will mutation leave the gene unchanged.
+TODO Explain increased radius size~\cite{chicano:2014:ball}
 
-The most common form of CGP uses a $1+\lambda$ strategy, where all offspring are
-mutants of the parent.  The most fit offspring replaces the parent only if its
-fitness is no worse than the parent.  Ties between offspring are broken by random selection
-and ties between the offspring and the parent are awarded to the offspring:  in this manner
-inactive genes are allowed to drift, which has been shown to significantly improve
-performance~\cite{yu:2001:neutrality}.  We will refer to this standard form of
-mutation and replacement as \emph{Normal}.
+\section{Gray-Box Enumeration}
+TODO Explain how gray-box properties and gray coding can be used to
+enumerate in \BigO{2^N} time.
 
-\section{Wasted Evaluations}
-Preventing search from repeatedly evaluating the same search point can improve
-any search that is limited by the number of evaluations it can perform.
-In an evolutionary search employing indirect phenotype encoding, as CGP does,
-the same search point can be represented by multiple different genotypes
-making duplicate detection more difficult.  While it does not catch all
-identical phenotypes, if two CGP individuals are actively identical (contain
-the same active genes) then their phenotypes must be identical.
-Miller~\cite{miller:2011:chapter2} theorized
-that this may lead to wasted evaluations as any number of actively identical
-individuals can be evaluated even though they will receive identical fitnesses.  He did not,
-however, provide any analysis for how frequently these unnecessary evaluations may occur.
+\section{Hyperplane Elimination}
 
-\begin{figure}
-  \centering
-  \includegraphicsfit{Probability}
-  \caption{Probability of no active gene being mutated for different mutation rates} 
-  \label{fig:probability}
-\end{figure}
-\subsection{Detection}
-Before evaluating a CGP individual, it is common practice to determine which
-nodes are active allowing inactive nodes to be ignored during
-evaluation, leading to a reduction in runtime~\cite{vasicek:2012:efficient}.
-By storing the set of active genes as part of the individual, it is trivial
-to determine when active genes are mutated.  More generally, it takes $O(N)$ additional time to determine
-if two randomly chosen individuals contain different active genes, where $N$ is
-the number of active nodes, assuming at least one of the individuals has been
-evaluated.
+\section{Improved Enumeration Ordering}
 
-\begin{comment}
-To exploit this knowledge to its fullest, all newly created individuals could
-be compared against all previously evaluated individuals to determine if they
-need to be evaluated.  While this can be optimized by storing a hash table of
-actively unique individuals, the likelihood that two unrelated individuals are
-actively identical may be too low to warrant the memory overhead required by being
-this thorough.
-\end{comment}
+\section{Timing Comparisons}
 
-%A reasonable heuristic to reduce this overhead is to only check if
-A reasonable heuristic to avoid duplicate evaluations is to determine if
-a parent's active genes are identical to its offspring.  This will catch the
-majority of cases, as the probability of two individuals being actively identical
-decreases as the number of mutation applications
-between them increases.  This also has the advantage that
-extra individuals need not be stored in memory.  Due
-to these advantages, we only compare an offspring against its parent
-to detect unnecessary evaluations.
+\section{Examining Local Optima}
 
-\subsection{Frequency of Offspring Actively Identical to their Parent}
-\label{sec:freq}
-Using the definition of point mutation given in Section~\ref{sec:cgp},
-it is possible to quantify the probability that an offspring is actively
-identical to its parent.
-\begin{equation}
-  \left (1 - mutation\_rate  \right )^{active\_genes}
-\label{eq:waste}
-\end{equation}
-\begin{comment}
-First, the number of active genes in each individual
-is given by Equation~\ref{eq:active}.  This shows how the number of active genes
-can change greatly with the number of active nodes and the arity used for each node.
-\begin{equation}
-  active\_genes = active\_nodes\cdot\left(arity + 1 \right) + output\_length
-  \label{eq:active}
-\end{equation}
-\end{comment}
-Equation~\ref{eq:waste} shows the probability that none of the active genes are
-mutated at the given mutation rate.  To help display the effects of this relationship,
-Figure~\ref{fig:probability} illustrates the probability that an offspring is actively
-identical to its parent for various mutation rates and number of active genes.
-Figure~\ref{fig:probability} shows that as the mutation rate and the number of active genes decreases,
-the probability of an offspring being actively identical to a parent greatly increases.
-While the other versions of mutation discussed in Section~\ref{sec:cgp} will change the details
-of this equation, the overall relationship between the variables will be the same.
+\section{Conclusions}
 
-\begin{comment}
-Using the other versions of mutation in Section~\ref{sec:cgp} does not change the
-overall relationship of the number of active genes and the mutation rate to the
-frequency of wasted evaluations.  If mutated genes are not forced to change to a
-new value the probability of actively identical offspring will increase.  While this is intuitively
-straightforward, determining the exact amount of increase is complex as the probability
-that a randomly reset gene returns to its current value depends on which gene is being mutated.
-If mutation is changed to have a set percentage of the genome mutated, the relationship
-changes to use the percentage of the genome that is active, but it will still follow the general
-shape of Figure~\ref{fig:probability}.
-\end{comment}
+\subsubsection*{Acknowledgments.} TODO Acknowledge BEACON.
 
-\section{Methods to Avoid Wasting Evaluations}
-\subsection{Skip}
-The most straightforward method to use existing evaluation results is to
-set the offspring's fitness to the parent's
-if they are actively identical. This method will be referred to as \emph{Skip}.
-As \emph{Skip} makes no modification to evolutionary mechanisms, any problem
-solvable by \emph{Normal} will be solvable by \emph{Skip}, and vice versa. The
-primary difference is only that \emph{Normal} will always use at least as many
-evaluations as \emph{Skip}, with the potential to use significantly more.  As
-a result, \emph{Skip} may solve problems \emph{Normal} cannot if given a
-maximum number of evaluations.
+\bibliographystyle{splncs03}
+\bibliography{../main}
 
-A significant theoretic advantage to this technique is that it lowers CGP's
-sensitivity to the mutation rate.  In \emph{Normal}, as the mutation rate
-decreases the potential for wasted evaluations
-increases creating an evaluation penalty.  As the mutation rate increases the expected
-number of modified active genes increases making small
-changes in phenotype less probable. From this, \emph{Normal} will likely
-be most effective when the mutation rate
-is high enough to expect at least one active gene to be modified by
-mutation, but low enough to allow the small phenotypic moves required
-to reach optimal values.
-Unlike \emph{Normal}, \emph{Skip} can use mutation rates expected to produce
-frequent offspring that are not actively different from their parents without
-penalty.  As such \emph{Skip} can use very fine grained mutation allowing it to potentially
-hill climb much more effectively.
-
-\subsection{Accumulating Mutation}
-As was stated in Section~\ref{sec:cgp}, some of CGP's success stems
-from allowing genetic drift of inactive genes~\cite{yu:2001:neutrality}.
-%Inactive gene modification can build structures which can potentially be made active
-%by later mutations.
-This raises the following question: will evolution in CGP improve
-if inactive genes are allowed to change at an increased rate?
-
-To test this possibility we propose a method of mutation called \emph{Accumulate}.
-The primary concept behind this technique is to allow inactive genes to drift
-multiple generations worth of distance in a single actual generation of CGP.
-For our example we will refer to one of the offspring produced by the parent as $F_0$.  If $F_0$
-is actively identical to its parent, $F_0$ produces a mutant offspring $F_1$ in the same generation.
-This chain continues with $F_i$ mutating to produce $F_{i+1}$ until an individual
-is produced that has one or more active genes different from the original parent.
-This actively different individual is called $F_n$.  $F_n$ is then evaluated.
-If $F_n$ is no worse than the original parent, it replaces $F_0$ and CGP continues
-evolution and selection as normal.  If $F_n$ is worse than the original parent
-$F_{n-1}$ replaces $F_0$.  Thus accumulate's offspring can have zero to multiple
-active genes changed.
-
-By chaining mutations each $F_i$ will probabilistically have more changes to
-inactive genes than $F_{i-1}$ when compared with the original parent.  By using
-$F_{n-1}$ or $F_n$ instead of $F_0$, \emph{Accumulate} can drift faster than
-\emph{Normal} and \emph{Skip}.  This can be advantageous on problems that benefit
-from high exploration of the neutral space, with the potential drawback that
-previously active useful structures are less likely to survive to be reintegrated.
-
-\subsection{Single Active Mutation}
-%An alternative method for avoiding unnecessary evaluations is to
-%modify mutation to ensure offspring
-%are actively different from their parents. 
-\emph{Skip}
-and \emph{Accumulate} have the potential, similar to \emph{Normal}, to generate large numbers of individuals
-with no active differences, which can be computationally expensive even if they
-are not evaluated. To avoid
-this potential overhead, we propose a method to ensure a \emph{Single} active
-gene is mutated every time an offspring is generated.
-%
-Creating an offspring using \emph{Single} is an iterative process.  Until an
-active gene has been mutated, select a gene at random in the individual and
-mutate its value.  This has the following properties:
-\begin{itemize}
-\item Exactly one active gene is mutated for all offspring.
-\item Zero or more inactive genes can be mutated.
-\item A gene that is active will be mutated more frequently than one that is not.
-\item No mutation rate is required.
-\end{itemize}
-Many of these properties make \emph{Single} distinct from the other described mutation
-methods.  By forcing only a single active gene to change, \emph{Single} may have
-better results on problems where incremental improvements are possible with single
-gene changes and worse results on problems where larger changes are necessary.
-This is in contrast to all of the other methods that have the potential to change
-multiple active genes.
-Limiting mutation to a single active gene change does not, however, prevent \emph{Single}
-from making large phenotypic changes, which are still possible by connecting
-in inactive nodes.
-
-\emph{Single} has the property that the effective mutation rate is always
-$\frac{1}{a}$ for active genes and $\frac{1}{a + 1}$ for
-inactive genes, resulting in $\frac{n - a}{a + 1} + 1$ expected mutations, where
-$a$ is the number of active genes and $n$ is the total number of genes.  This
-means that the number of inactive genes does not affect the probability of an
-inactive gene being mutated, just the expected number of inactive genes that are
-mutated.  Furthermore, as the number of active genes approaches zero, the expected
-number of mutations increases dramatically, even though it is still limited to
-changing a single active gene.  As the evolved number of active genes is likely to be
-highly problem dependent, the behavior of \emph{Single} will likely change significantly
-from problem to problem, with some problems having very little changes to inactive
-genes, and others changing a great deal.  It is also possible that this will create
-a selective pressure on the number of active genes which may only become apparent
-if multiple parents are used.
-
-In comparison with \emph{Skip}, \emph{Single} is even easier to apply to new problems
-as it completely removes the need to set the mutation rate, but does so at the
-potential expense of lower maximum effectiveness.  In comparison with \emph{Accumulate},
-\emph{Single} mutates inactive genes less often with respect to active genes, making it
-better at preserving useful inactive structures but worse at exploring the inactive
-space.
-
-\section{Experimental Setup}
-In order to empirically test the suggested benefits of \emph{Skip},
-\emph{Accumulate}, and \emph{Single} in comparison with \emph{Normal}, we evaluated their performance on
-four benchmarks: Parity, Multiplier, Binary Decode and Binary Encode.
-Both Parity and Multiplier have been used extensively to test
-CGP~\cite{walker:2008:cgpmodules,miller:2006:redundancy}.
-The 3-bit even parity is included as it is the most commonly used benchmark for CGP and to
-help provide some level of comparability with previous work.  In order to try a
-more challenging binary problem, we chose to do the 3-bit multiplier, which has
-6 inputs and 6 outputs.
-
-Binary Decode is effectively a demultiplexer, where $N$
-input lines specify which of the $2^N$ output lines should have a value of one, with
-all others set to zero.  For example, if $101$ is the input to a 3-bit Binary Decode
-problem, the output is $00100000$.  Binary Encode is the reverse of Binary Decode.
-Given $2^N$ input lines, where exactly one line has a bit set to one and all others
-are zero, the binary value index of the one bit is output using $N$ lines.
-An example of this mapping is that $00100000$
-should return the output of $101$.  The reason we use Binary Decode and Binary Encode
-as benchmarks is their asymmetric number of inputs to outputs and their relatively
-small number of test cases.  For example, on the 16 to 4 Binary Encode and the 4 to 16 Binary
-Decode (the instances we use) only 16 test cases are needed.
-
-All problems used the operators \{and, or, nand, nor\} and determine fitness as the average
-percentage of correct bits across all possible inputs.  While this is a common function set used
-by CGP for these problems,~\cite{walker:2008:cgpmodules} used a different set of operators
-for Multiplier.  Preliminary testing shows our operator set makes Multiplier harder than theirs,
-but, as always, conclusions drawn by comparing experiments using different operator sets should
-be minimal.
-
-While a few of these benchmarks may be overly simple, with parity especially
-being considered for obsolescence~\cite{mcdermott:2012:benchmarks}, they are used here
-for the specific reason of their simplicity.  The primary
-difference between \emph{Normal} and \emph{Skip}
-is not that one can solve harder problems than the other, rather it is that \emph{Skip} is thought
-to solve problems in less evaluations than \emph{Normal}.  Being able to compare each method's
-median number of evaluations until success makes for the most logical metric in this case,
-so these problems are useful in that they can be solved often.  While much of the
-previous work on CGP uses Koza's \emph{computational effort} to analyze experimental
-results,~\cite{christensen:2002:comp_effort} explains how that can introduce
-significant underestimates, especially when using small population sizes.
-
-In order to compare with the effective standard CGP, we use the $1+4$
-evolutionary strategy.  We use a
-genotype size of 3000 nodes, similar to those found optimal in~\cite{miller:2006:redundancy}.
-Each run was allowed up to
-10 million evaluations before being terminated, and 50 runs of each experiment
-were performed.  To help focus the differences between each mutation method,
-all starting individuals on the same problem were identical for the same run number,
-meaning that the 50 different individuals \emph{Normal} started with were identical
-to the 50 starting individuals for \emph{Skip}, \emph{Accumulate}, and \emph{Single}.
-For statistical purposes this means all results should be compared as paired tests.
-As the number of evaluations to success is likely not normally distributed, we choose
-the Wilcoxon signed-rank test to compare each mutation method with \emph{Normal}.
-As an attempt to cover a wide range of mutation rates, the 16 mutation rates listed
-in Figure~\ref{fig:probability} were used for each method.  These cover from 0.0001
-to 0.8 on a log scale, and should encompass all previous levels of mutation used
-with CGP.
-
-\begin{comment}
-We had intended to include Koza-1 and Pagie-1 from~\cite{mcdermott:2012:benchmarks}
-as sample regression problems, but we had difficulty achieving acceptable success
-rates to allow for comparisons here.  Future work may be needed to understand why
-these problems were significantly harder when using our configuration settings
-than the four binary problems. 
-\end{comment}
-
-\section{Results}
-To summarize the results of testing each mutation rate, Figures~\ref{fig:parity},
-\ref{fig:multiply}, \ref{fig:encode}, and \ref{fig:decode} show the median number of evaluations
-required over 50 runs to solve the Parity, Multiply, Binary Encode, and Binary Decode problems.  Note
-that both axes are presented on a log scale.  In the event that the median run of a configuration
-did not find the optimal value before termination, the value is not displayed in the graph.  This can be
-clearly observed in Figure~\ref{fig:multiply} for mutation rates greater than 0.01.  As \emph{Single}
-does not use a mutation rate, its data is assumed constant for all mutation rates.
-%
-Table~\ref{table:results} shows fine details
-about the best configurations found for each mutation method.  This includes the mutation rate
-with the lowest median evaluations until success and the median evaluations it required, the median absolute
-deviation from that median, the single tailed P-value received from comparing with the best \emph{Normal}
-configuration using the Wilcoxon signed-rank test, and the median number of active genes in the solution.
-Note that the number of active genes is a different measure from the number of active nodes, which has
-been reported in other work, and is used here for compatibility with Equation~\ref{eq:waste}.
-%
-A complete package of our results as well as the source code used in testing
-is available from our website.\footnote{\url{https://github.com/brianwgoldman/ReducingWastedEvaluationsCGP}}
-
-\begin{figure}
-  \centering
-  \includegraphicsfit{parity}
-  \caption{Median evaluations to success for each method on the Parity problem} 
-  \label{fig:parity}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphicsfit{multiply}
-  \caption{Median evaluations to success for each method on the Multiply problem} 
-  \label{fig:multiply}
-\end{figure}
-
-\begin{comment}
-\begin{figure}
-  \centering
-  \includegraphicsfit{quartic}
-  \caption{Mean Evaluations to Success for each method on the Quartic problem} 
-  \label{fig:quartic}
-\end{figure}
-\end{comment}
-
-\begin{figure}
-  \centering
-  \includegraphicsfit{encode}
-  \caption{Median evaluations to success for each method on the Binary Encode problem} 
-  \label{fig:encode}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphicsfit{decode}
-  \caption{Median evaluations to success for each method on the Binary Decode problem} 
-  \label{fig:decode}
-\end{figure}
-
-\section{Discussion}
-Even though the Parity, Multiplier, Binary Encode, and Binary Decode problems have
-very different numbers of inputs and outputs, different input to output ratios, 
-different evolved numbers of active genes (see Table~\ref{table:results}), and
-different levels of difficulty, Figures~\ref{fig:parity}, \ref{fig:multiply},
-\ref{fig:encode}, and \ref{fig:decode} present surprisingly similar results.  
-\begin{comment}
-On all four
-problems \emph{Normal} has a fairly clear quadratic shape which at its lowest point
-approaches or sometimes crosses the line of \emph{Single}.  As the mutation rate increases
-beyond the optimal value for \emph{Normal} all techniques except \emph{Single} share
-very similar if not identical progressions.  Below \emph{Normal}'s optimal mutation
-rate, \emph{Normal} gets significantly worse while \emph{Skip} and \emph{Accumulate}
-diverge from \emph{Normal} by maintaining similar qualities as the optimal value.
-At exceptionally low mutation rates \emph{Normal} takes over an order of magnitude
-more evaluations to find success than all of the other techniques.
-
-The shape of these features is as expected.
-\end{comment}
-  The steep increases in median evaluations
-for \emph{Normal} corresponds with the prediction that overly high mutation rates will
-significantly increase the number of evaluations until success, 
-while low mutation rates will create a significant portion of
-wasted evaluations.  On Multiply, Binary Encode, and Binary Decode, overly high mutation rates
-prevent \emph{Normal} from reaching the optimum.
-Parity is the only problem CGP could solve using exceptionally high mutation rates, with even
-a slight improvement at the very top of the tested range.  This likely happens as the Parity problem
-we are using is so simple that a CGP using effectively random search is able to solve it.
-\emph{Skip} and \emph{Accumulate}'s tendency to mirror \emph{Normal}'s effectiveness as the
-mutation rate increases is because the probability of creating actively identical offspring
-approaches zero.  As the divergence point is always near
-\emph{Normal}'s optimum mutation rate, it can be inferred that \emph{Normal}'s significant decrease in
-quality when using lower mutation rates is likely caused by wasted evaluations.
-As \emph{Skip} also decreases
-somewhat in quality, some of the impaired fitness can also likely be attributed
-to improbability of making large enough genotypic changes to escape local optima
-with small mutation rates.
-
-All four figures suggest that \emph{Skip} and \emph{Accumulate}
-have a lower sensitivity to the mutation rate than \emph{Normal}.
-As the mutation rate goes below the optimal value,
-\emph{Skip} and \emph{Accumulate} encounter a relatively minor increase in their median number of evaluations
-when compared with \emph{Normal}. As confirmation of Equation~\ref{eq:waste}
-we recorded the predicted number of wasted evaluations and found
-the aggregate predicted waste was nearly identical to the differences between \emph{Normal} and \emph{Skip}.
-
-\emph{Single}, the only method without a mutation rate, is clearly not as generally effective as
-the other methods where mutation rate can be optimized.  However, when the optimal mutation rate is
-not known, \emph{Single} does show significant advantages.  When compared against the entire
-space of tested mutation rates, \emph{Single} almost always requires
-the least number of evaluations to succeed.  Below the optimum mutation rates for each, \emph{Skip}
-and \emph{Accumulate} are reasonable approximations of \emph{Single}, but come with the disadvantage
-that they are more computationally expensive.
-%
-As the number of evaluations required to solve a problem increases, \emph{Single}
-appears to improve in performance relative to \emph{Normal}.  Parity required the
-least number of evaluations to be solved, followed by Binary Encode and Binary Decode,
-with Multiply requiring by far the most.
-This ordering mirror's \emph{Single}'s relative scaling with \emph{Normal}, 
-as it goes from a lower peak performance than \emph{Normal}
-on Parity and Binary Encode to surpassing it on Binary Decode and Multiply.
-
-\begin{table}
-	\centering
-	\caption{Comparisons of Best Mutation Rate for Each Mutation Method on Each Benchmark.
-	  All results are from 100\% successful configurations.}
-	\begin{tabular}{r|c|c|c|c|c|c|c|c|c|c|c|}
-	  \cline{3-6}
-		 \multicolumn{2}{r|}{}    & \textbf{\emph{Normal}} & \textbf{\emph{Skip}} & \textbf{\emph{Accumulate}} & \textbf{\emph{Single}} \\ \cline{2-6} 
-		                    & Mutation Rate & 0.02 & 0.02 & 0.01 & N/A \\ \cline{2-6}
-	       \textbf{3-bit} & Median Evaluations & 1,018 & 959 & 928 & 1358 \\ \cline{2-6}
-        \textbf{Parity} & Median Absolute Deviation & 614 & 564 & 416 & 733 \\ \cline{2-6}
-	                      & P value  & N/A & 0.0000 & 0.4875 & 0.0790 \\ \cline{2-6}
-	                      & Median Final Active Genes & 250 & 250 & 262 & 223 \\ \hhline{~=====}
-
-		                    & Mutation Rate & 0.004 & 0.002 & 0.0008 & N/A \\ \cline{2-6}
-	       \textbf{3-bit} & Median Evaluations & 503,424 & 398,354 & 421,859 & 340,165 \\ \cline{2-6}
-	  \textbf{Multiplier} & Median Absolute Deviation & 191,037 & 184,979 & 165,218 & 151,063 \\ \cline{2-6}
-	                      & P value  & N/A & 0.0212 & 0.3632 & 0.0065 \\ \cline{2-6}
-	                      & Median Final Active Genes & 627 & 629 & 684 & 620 \\ \hhline{~=====}
-
-                        & Mutation Rate & 0.004 & 0.004 & 0.008 & N/A \\ \cline{2-6}
-	     \textbf{16 to 4} & Median Evaluations & 24,057 & 19,732 & 21,563 & 26,031 \\ \cline{2-6}
-	      \textbf{Encode} & Median Absolute Deviation & 8,829 & 7,425 & 7,449 & 10,170 \\ \cline{2-6}
-	                      & P value  & N/A & 0.0000 & 0.0864 & 0.3578 \\ \cline{2-6}
-	                      & Median Final Active Genes & 472 & 472 & 499 & 510 \\ \hhline{~=====}
-
-	                      & Mutation Rate & 0.004 & 0.002 & 0.004 & N/A \\ \cline{2-6} \cline{2-6}
-       \textbf{4 to 16} & Median Evaluations & 73,351 & 70,023 & 65,347 & 69,564 \\ \cline{2-6}
-        \textbf{Decode} & Median Absolute Deviation & 17,889 & 17,036 & 18,411 & 21,131 \\ \cline{2-6}
-	                      & P value  & N/A & 0.415 & 0.3276 & 0.2398 \\ \cline{2-6}
-	                      & Median Final Active Genes & 1,117 & 1,027 & 1,050 & 1,090 \\ \cline{2-6}
-
-  \end{tabular}
-  \label{table:results}
-\end{table}
-
-Using Table~\ref{table:results}, it is clear that \emph{Normal} is never the best option.
-In all cases \emph{Skip} and \emph{Accumulate} can obtain better results,
-with even \emph{Single} doing better on Binary Decode and Multiply.  Furthermore, \emph{Skip} always
-had a lower median absolute deviation than \emph{Normal}, with \emph{Accumulate} lower
-on all problems except Binary Decode.  The statistical analysis is less clear as the deviations
-are so large, but \emph{Skip}'s improvements are definitely significant on all problems except
-Binary Decode, with \emph{Single} statistically better than \emph{Normal} on Multiply.
-As support that our configuration
-for \emph{Normal} is valid to compare against, our peak quality is approximately 6 times
-better than results published in~\cite{walker:2008:cgpmodules} for their standard CGP and ECGP
-on Parity.
-%TODO Discuss length stuff
-
-While the true complexity measure for a GP implementation is how many evaluations
-it requires to solve a problem, we also performed preliminary timing tests to ensure
-our new techniques were less computationally complex than the evaluations they avoid.
-When using the best found mutation rate on Multiply, \emph{Normal} required the most time (4597 minutes)
-to complete all 50 runs, with \emph{Skip} using 9\% less time, \emph{Accumulate} 16\% less,
-and \emph{Single} 29\% less.
-
-\section{Conclusions and Future Work}
-CGP's current mutation operator has the potential to create offspring which can
-be identified as phenotypically identical to their parents without invoking
-the evaluation function.  Evaluating these offspring creates waste, as we can
-assign their fitness to be identical to their parent.  Through careful definition
-of how the mutation operator works, we have shown how the probability of having
-wasted evaluations is dependent on the mutation rate and the number of active genes.
-We predicted and then provided empirical evidence that avoiding this waste can
-reduce how many evaluations CGP requires to solve problems as well as reduce
-CGP's sensitivity to the mutation rate parameter.
-
-We proposed \emph{Skip} and \emph{Accumulate} as two methods for what to do when
-actively identical offspring are detected, and \emph{Single} as a method to ensure
-they are never created.  On four fairly diverse benchmark problems we showed that
-\emph{Skip} and \emph{Accumulate} do at least as well as \emph{Normal} using a
-variety of mutation rates and are frequently vastly more efficient than \emph{Normal}.
-As there was little discernible difference in their overall quality, for now we suggest \emph{Skip}
-be used as a replacement to \emph{Normal}, as it is less complex than \emph{Accumulate},
-although future experimentation may suggest otherwise.  In situations where the mutation
-rate cannot be optimized, we suggest \emph{Single}, as while it did not achieve
-the best peak performance on all problems, it has the least overhead with the widest applicability.
-Furthermore, \emph{Single} achieved the best results on the hardest test problem, suggesting more
-extensive testing of this method on difficult problems should be performed.
-
-Even though the results appear to be problem independent, \emph{Skip}, \emph{Accumulate},
-and \emph{Single} should be tested on a wider range of problem classes to make sure.
-More advanced techniques for handling or avoiding wasted evaluations may also improve
-results, such as a modified version of \emph{Single} to allow multiple active genes
-to be mutated at once.  Nevertheless, we feel that the theoretic and empirical evidence
-sufficiently supports a need to avoid wasting evaluations on actively identical
-individuals.
-
-\bibliographystyle{splncs}
-\bibliography{../main}
 \end{document}

### 2015-08-26 Hyperplane Enumeration/main.tex
Added a bunch more, fixed line ending problems

@@
 % first the title is needed
-\title{Hyperplane Elimination for Quickly Enumerating MK Landscapes}
+\title{Hyperplane Elimination for Quickly Enumerating Local Optima}
 
@@
 %\emph{abstract} environment.
-\keywords{Landscape Understanding, Gray-Box}
+\keywords{Landscape Understanding, Gray-Box, MK Landscapes}
 \end{abstract}
@@
 \section{Introduction}
-TODO Motivate using network stuff.
+The ruggedness and high dimensionality of most interesting landscapes makes them challenging
+to visualize or otherwise analyze. However, doing so can be helpful in
+quantifying the difficulty of a problem, and how to best design algorithms to deal
+with those difficulties. 
+Similarly, knowing which characteristics favor
+a particular
+algorithm can help researchers choose the algorithm
+most likely to perform well on their problem.
 
-\section{MK Landscapes and Gray-Box Optimization}
+One avenue for performing this problem understanding is to
+examine the frequency and distribution of local optima~\cite{boese:1994:bigvalley},
+and how operators allow search to transition between these
+optima~\cite{tomassini:2008:nknetworks,verel:2011:nknetworks,ochoa:2015:crossovernetworks}.
+However, finding all of the local optima is prohibitively time consuming
+for even small problems, with many studies limited to 18-bit
+problems~\cite{tomassini:2008:nknetworks,verel:2011:nknetworks}.
+By leveraging recent advancements in Gray-Box optimization which allow for
+constant time local search~\cite{chicano:2014:ball}, this limit was raised
+to 30-bit problems~\cite{ochoa:2015:crossovernetworks}.
+
+Here we will introduce a method for finding all local optima of problems
+with up to 80 bits. The cornerstone of this method is the identification
+of hyperplanes which cannot contain any local optima, allowing large
+portions of the search space to be skipped during enumeration.
+
+\section{Mk Landscapes and Gray-Box Optimization}
+An Mk Landscape is any function which can be expressed in the following form:
+\begin{equation}
+  f(x) = \sum_{i=1}^{M} f_i(mask(x, s_i))
+  \label{eq-mk}
+\end{equation}
+In this equation there are $M=$\BigO{N} subfunctions $f_i$, which act
+on some subset $s_i$ of variables in $x$. Each subfunction performs the mapping
+$f_i : \mathbb{B}^{k_i}\rightarrow \mathbb{R}$ where $k_i \leq k$ and $k$ is
+a constant with respect to $N$. The $mask$ function returns the values of $x$ for
+the variables in $s_i$.
+
+A critical aspect of this problem class is that it can be optimized using Gray-Box
+techniques, which can allow for highly efficient local~\cite{whitley:2013:greedy}
+and global~\cite{goldman:2015:GBO,tintos:2015:partitioncross} search. These algorithms
+exploit the known and limited non-linearity between variables expressed by all $s_i$,
+as well as perform partial solution evaluation using all $f_i$.
+
+This formulation can represent many problems of real-world interest, as well as
+the most commonly used combinatorial benchmark problems. In this work we shall
+examine 5 problems in particular: Concatenated Traps, Adjacent and Random NKq Landscapes,
+Ising Spin Glasses, and MAX-kSAT.
+
+The Concatenated Trap problem~\cite{deb:1992:trap} is a composition of $k$-order deceptive
+separable subfunctions. In Mk Landscape terms, $M=N/k$ such that $\forall_i |s_i| = k$ and
+$\forall_{i \neq j} s_i \cap s_j = \emptyset$. Each $f_i$ applies an identical function based
+on the number of variables set to 1:
+\begin{equation}
+   trap(t) = \left\{
+     \begin{array}{rl}
+       k-1-t, &  t<k\\
+       k,   &  t = k
+     \end{array}
+   \right.
+  \label{eq-trap}
+\end{equation}
+While this problem is still in common use~\cite{hsu:2015:dsmgaII,inoue:2015:adaptivep3},
+most advanced methods can solve it trivially~\cite{goldman:2012:ltga} and when expressed
+as an Mk Landscape can be solved exactly in \BigO{N} time~\cite{whitley:2015:mk}. However,
+we include it here as its simple structure allows for straight forward algorithm analysis.
+
 TODO Explain all problems here
@@
 TODO Explain how gray-box properties and gray coding can be used to
-enumerate in \BigO{2^N} time.
+enumerate in \BigO{2^N} time.~\cite{ochoa:2015:crossovernetworks}
 
@@
 
+% Proof sketch
+Moves can have their dependencies fixed in some permutation $O$ of all possible moves,
+such that $O_0$ is the first move to have its dependencies fixed.
+$D_O(i, m)$ is a function which returns how many unfixed dependencies move $m$ has after all
+$i$ moves in $O$ before $i$ have been fixed. A greedy solution to this problem works such
+that $O_i$ is set to be the move which minimizes $D_O(i, m)$.
+
+All non-greedy solutions must have some $i$ such that $D_O(i, O_i) > D_O(i, O_{i+1})$.
+This is because $D_O(i, m)$ can only decrease as $i$ increases, meaning that if some $m$
+has a lower value at $i$ than $O_i$, some index between $i$ and when $m$ appears in $O$
+this property must hold. Finally, all non-greedy solutions must have some $\hat{i}$
+which is the maximum $i$ value for which $D_O(i, O_i) > D_O(i, O_{i+1})$ is true.
+
+Consider the optimal ordering $O^\star$ which is not greedy.
+These criteria mean that $D_{O^\star}(\hat{i}, O_{\hat{i}}) > D_{O^\star}(\hat{i}, O_{\hat{i}+1})$.
+Swapping $O_{\hat{i}}$ and $O_{\hat{i}+1}$ cannot change the minimum dependency of any other moves.
+Performing the swap will improve the minimum dependency of $O_{\hat{i}+1}$ more than it hurts
+$O_{\hat{i}}$. Therefore, performing this swap will make $O^\star$ even better, implying it was
+not optimal before. Therefore there cannot exist an optimal ordering which is not greedy.
+
+\section{Complexity Classes for Simple Landscapes}
+
 \section{Timing Comparisons}
+\subsection{Increasing problem size}
+Consider including boxplots of largest size all are successful.
+
+\subsection{The Effect of Radius}
 

### 2015-09-01 Hyperplane Enumeration/main.tex
More hyperplane updates

@@
 A critical aspect of this problem class is that it can be optimized using Gray-Box
-techniques, which can allow for highly efficient local~\cite{whitley:2013:greedy}
+techniques, which can allow for highly efficient local~\cite{whitley:2012:constant}
 and global~\cite{goldman:2015:GBO,tintos:2015:partitioncross} search. These algorithms
@@
 
-TODO Explain all problems here
+NKq landscapes specify a class of randomly generated problem instances using 3 paramters:
+(1) the number of problem variables $N$ (2) the amount of variable epistasis $K$ where $K=k-1$
+and (3) the number of unique fitness values $q$. From these parameters a landscape is generated
+by creating $M=N$ subfunctions $f_i$, where $f_i$ uses variable $i$ and $K$ others to look up
+a fitness value in the range $[0..q-1]$ from a randomly generated table. This structure
+makes NKq landscapes a natural fit for Mk landscapes (sum of bounded subfunctions)
+and are the most studied problem class for Gray-Box
+optimization~\cite{whitley:2012:constant,goldman:2015:GBO,tintos:2015:partitioncross,ochoa:2015:crossovernetworks,whitley:2015:mk}.
+Here we choose $k=3$ with $q=2^{K+1}=2^{k}=8$.
+
+We use two common variants of NKq in our experiments which specify how the $K$ additional
+variables in each subfunction are chosen: Adjacent and Random. In Adjacent NKq, $f_i$ depends
+on variables $[i..(i+k)\%N]$. In Random NKq, $f_i$ depends on variables $i$ and a random
+set of $K$ unique variables which does not include $i$. While Random NKq is NP-Hard,
+the structure of Adjacent NKq allows for a polynomial time dynamic programming solution~\cite{wright:2000:solvingnk}.
+Adjacent NKq is therefore often used to understand a search algorithm's ability to reach the known global
+optimum.
+
+Ising Spin Glasses are a type of MAX-CUT problem derived from statistical physics.
+Each atom in the glass (vertex) can be assigned a spin, with the goal being to
+find the set of assignments which minimize the energy between nearby atoms (edges).
+Similar to Adjacent NKq, the $2D\pm J$ subset of Ising Spin Glasses can be polynomially
+solved~\cite{saul:1994:spinglass}\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}.
+In this subset, the graph is defined as a square two-dimensional torus with each edge
+weight chosen from ${-1, 1}$. Each vertex is assigned a spin from ${-1, 1}$ with
+the energy in the glass equal to
+\begin{equation}
+\sum_{e_{ij} \in E} x_ie_{ij}x_j
+  \label{eq-ising}
+\end{equation}
+where $e_{ij}$ is the edge connecting vertex $i$ to vertex $j$. In Mk landscape terms
+this type of spin glass has $M=2N$ and $k=2$.
+
+Our final examined problem is randomly generated maximum satisfiability or MAX-kSAT.
+This version of the canonical NP-Complete boolean satisfiability problem is formulated
+as the maximization of $M=4.27N$ clauses, each containing exactly $k=3$ unique literals.
+A clause is satisfied if any of its literals match how a solution's variables are set.
+
+\section{Gray-Box Enumeration}
+When considered as a black box, the process of finding all local optima
+in a landscape requires \BigO{N2^N} time. This is because evaluating
+a solution requires $N$ time, and all $2^N$ solutions in the landscape
+must be evaluated once. Finding the local
+optima then requires each solution to compare its fitness with each
+of its neighbors, which when properly cached takes \BigO{N} per solution.
+Extending this method to look for solutions which cannot be improved by
+flipping $r$ or less bits requires \BigO{N^r2^N} time. However,
+by exploiting Gray-Box optimization methods previous work~\cite{ochoa:2015:crossovernetworks}
+was able to find all $r$-bit local optima of Mk Landscapes in \BigO{2^N} for any small,
+fixed value of $r$.
+
+The first major result in Gray-Box optimization was the proof that the list
+of fitness improving moves during local search can be kept updated in \BigO{1}
+time~\cite{whitley:2012:constant}. Each time a modification is made to a solution
+only modifications which share a non-linear relationship with the changes can
+go from non-improving to improving and vice versa. In Gray-Box optimization
+the non-linearity between variables is known. Furthermore the number of
+non-linear relationships is bounded above by \BigO{Mk} as each of the $M$
+subfunctions can cause a non-linear relationship between each of its $k$
+variables. As $M=$\BigO{N} and $k$ is constant, the amortized number of affected
+flips per variable is \BigO{1}.
+
+This limited and known non-linearity can be used to allow search to find only $r$-bit
+local optima. Consider a variable interaction graph, such that each variable maps
+to a vertex in the graph and each edge represents a non-linear relationship between
+variables. TODO
 
 TODO Explain increased radius size~\cite{chicano:2014:ball}
-
-\section{Gray-Box Enumeration}
-TODO Explain how gray-box properties and gray coding can be used to
-enumerate in \BigO{2^N} time.~\cite{ochoa:2015:crossovernetworks}
 
@@
 
+\subsection{OneMax}
+In OneMax, each bin contains exactly 1 move. All variables in the current solution
+which disagree with the global optimum will be fitness improving moves.
+Initially $index$ is set to $N-1$ and is decreased by at most 1 in any step.
+Before finding the global optimum no carry operations can occur, meaning the
+cost of each iteration is equal to the amount by which $index$ is decreased.
+Therefore, Algorithm~\ref{alg-enumerate} requires \BigO{N} time to reach
+the global optimum.
+
+After finding the global optimum of OneMax, $index$ is set to 0. One iteration
+is then spent adding a 1 to $index$ 0, which in the worst case requires \BigO{N}
+carry operations. For all future iterations $index$ is a fitness improving move
+and is currently set to 1. In these iterations at least 1 carry operation must
+occur, and $index$ cannot be decreased. Iteration ends when $index$ exceeds $N$
+which requires at most $N$ carry operations. Therefore, Algorithm~\ref{alg-enumerate}
+requires \BigO{N} time to reach termination after finding the global optimum. Combined
+with initialization and the time to find the global optimum, the total complexity is \BigO{N}.
+
+\subsection{Concatenated Traps}
+TODO
+
 \section{Timing Comparisons}
@@
 \section{Examining Local Optima}
+
+TODO Consider discussing neutral networks
 

### 2015-09-02 Hyperplane Enumeration/main.tex
Added separable problems theory section

@@
 Extending this method to look for solutions which cannot be improved by
-flipping $r$ or less bits requires \BigO{N^r2^N} time. However,
+flipping $r$ or fewer bits requires \BigO{N^r2^N} time. However,
 by exploiting Gray-Box optimization methods previous work~\cite{ochoa:2015:crossovernetworks}
@@
 
-\subsection{Concatenated Traps}
-TODO
+\subsection{$k$-Bound Separable Problem}
+Consider any problem which is composed of non-overlapping
+subfunctions, each using using $k$ or fewer bits. Regardless
+of how they are initially ordered, reordering will ensure that
+variables of a function are always consecutively placed in enumeration.
+Consider a numbering of $f_i$ such that if $i < j$ then all of $f_i$'s
+variables appear before $f_j$'s in enumeration ordering.
+
+While any $f_i$ contains a fitness improving move, Algorithm~\ref{alg-enumerate}
+will find the one with the highest $i$ and enumerate it until it no longer
+contains an improving move. In total this enumeration requires less than $2^k$
+steps. Each $f_i$ is considered sequentially, meaning the first local optimum
+will be found in $2^kN/k$ time.
+
+Once the first local optimum is found, Algorithm~\ref{alg-enumerate} proceeds
+by finding all local optimum in larger and larger hyperplanes. Initially $f_0$
+is enumerated to find all local optima in the hyperplane with all $f_i, i>0$
+fixed. This process ends when a carry operation modifies a bit of $f_1$. At
+this point $f_1$ is enumerated, such that each time $f_1$ contains no improving moves
+$f_0$ is enumerated again. This finds all local optima in the hyperplane with $f_i, i>1$
+fixed. We represent the time required to find all local optima in the hyperplane
+with $f_j, j>i$ as $T(i)$. $T(0)=2^k$ as all values of $f_0$ must be enumerated.
+$T(i) = |l_i|*T(i-1)+2^k$ where $|l|$ is the number of ways $f_i$ can be set such
+that it contains no fitness improving moves. Each $l_i$ exposes a new hyperplane
+which can contain local optima, which causes the recursive call to $T(i-1)$.
+Assuming for simplicity that all subfunctions use exactly $k$ bits,
+the time required to find all local optima is $T(N/k-1)$.
+Assuming that $\forall_i |l_i|=c$, $T(N/k-1)=\sum_0^{N/k-1}2^kc^i<2^kc^{N/k}$.
+The number of local optima in the landscape is $c^{N/k}$ meaning that Algorithm~\ref{alg-enumerate}
+is within a constant factor of $2^k$ of optimal.
+
+Concatenated Traps is a commonly used $k$-bound separable problem
+with $c=2$.  Algorithm~\ref{alg-enumerate} requires $2^k2^{N/k}$ time
+to find all $2^{N/k}$ local optima in this problem.
 

### 2015-09-03 Hyperplane Enumeration/main.tex
Last push before writing group

@@
 \usepackage{graphicx}
+\usepackage{algorithm}
+\usepackage[noend]{algpseudocode}
+\renewcommand{\algorithmicrequire}{\textbf{Input:}}
+\renewcommand{\algorithmicensure}{\textbf{Output:}}
 
@@
 \noindent\keywordname\enspace\ignorespaces#1}
+
+\newcommand{\includegraphicsfit}[1]
+{\includegraphics[width=\columnwidth,height=\textheight,keepaspectratio]{#1}}
 
@@
 %\emph{abstract} environment.
-\keywords{Landscape Understanding, Gray-Box, MK Landscapes}
+\keywords{Landscape Understanding, Gray-Box, Mk Landscapes}
 \end{abstract}
@@
 constant time local search~\cite{chicano:2014:ball}, this limit was raised
-to 30-bit problems~\cite{ochoa:2015:crossovernetworks}.
+to 30-bit problems~\cite{ochoa:2015:crossovernetworks}. While sampling
+methods can approximate some of these statistics for much larger
+landscapes~\cite{iclanzan:2014:somnetworks}, they do not provide the complete
+picture necessary for many metrics~\cite{ochoa:2015:crossovernetworks}.
 
@@
 
-\section{Mk Landscapes and Gray-Box Optimization}
+\section{Mk Landscapes as a tool for Problem Generalization}
 An Mk Landscape is any function which can be expressed in the following form:
@@
 \end{equation}
-In this equation there are $M=$\BigO{N} subfunctions $f_i$, which act
-on some subset $s_i$ of variables in $x$. Each subfunction performs the mapping
-$f_i : \mathbb{B}^{k_i}\rightarrow \mathbb{R}$ where $k_i \leq k$ and $k$ is
-a constant with respect to $N$. The $mask$ function returns the values of $x$ for
-the variables in $s_i$.
+In this equation $f_i$ is a subfunction such that $f_i : \mathbb{B}^{|s_i|}\rightarrow \mathbb{R}$.
+The $mask$ function
+returns the values in $x$ associated with each variable in the set $s_i$.
+Each $|s_i| \leq k$, where $k$ is a constant with respect to $N=|x|$,
+the length of $x$. Finally, $M$ is constrained to be \BigO{N}.
 
 A critical aspect of this problem class is that it can be optimized using Gray-Box
-techniques, which can allow for highly efficient local~\cite{whitley:2012:constant}
+techniques, which can allow for highly efficient local~\cite{whitley:2012:constant,chicano:2014:ball}
 and global~\cite{goldman:2015:GBO,tintos:2015:partitioncross} search. These algorithms
@@
 most advanced methods can solve it trivially~\cite{goldman:2012:ltga} and when expressed
-as an Mk Landscape can be solved exactly in \BigO{N} time~\cite{whitley:2015:mk}. However,
-we include it here as its simple structure allows for straight forward algorithm analysis.
+as an Mk Landscape can be solved exactly in \BigO{N} time~\cite{whitley:2015:mk}.
+We therefore include it only because its structure allows for straight forward algorithm analysis.
+Here we set $k=5$.
 
@@
 and are the most studied problem class for Gray-Box
-optimization~\cite{whitley:2012:constant,goldman:2015:GBO,tintos:2015:partitioncross,ochoa:2015:crossovernetworks,whitley:2015:mk}.
+optimization~\cite{whitley:2012:constant,chicano:2014:ball,goldman:2015:GBO,tintos:2015:partitioncross,ochoa:2015:crossovernetworks,whitley:2015:mk}.
 Here we choose $k=3$ with $q=2^{K+1}=2^{k}=8$.
@@
 variables in each subfunction are chosen: Adjacent and Random. In Adjacent NKq, $f_i$ depends
-on variables $[i..(i+k)\%N]$. In Random NKq, $f_i$ depends on variables $i$ and a random
+on variables $[i..(i+k) \bmod N]$. In Random NKq, $f_i$ depends on variables $i$ and a random
 set of $K$ unique variables which does not include $i$. While Random NKq is NP-Hard,
@@
 In this subset, the graph is defined as a square two-dimensional torus with each edge
-weight chosen from ${-1, 1}$. Each vertex is assigned a spin from ${-1, 1}$ with
+weight chosen from $\{-1, 1\}$. Each vertex is assigned a spin from $\{-1, 1\}$ with
 the energy in the glass equal to
@@
 \end{equation}
-where $e_{ij}$ is the edge connecting vertex $i$ to vertex $j$. In Mk landscape terms
+where $e_{ij}$ is the weight of the edge vertex $i$ to vertex $j$. In Mk landscape terms
 this type of spin glass has $M=2N$ and $k=2$.
@@
 
-\section{Gray-Box Enumeration}
+\section{Gray-Box Enumeration of Mk Landscapes}
 When considered as a black box, the process of finding all local optima
@@
 flipping $r$ or fewer bits requires \BigO{N^r2^N} time. However,
-by exploiting Gray-Box optimization methods previous work~\cite{ochoa:2015:crossovernetworks}
+by exploiting Gray-Box optimization methods, previous work~\cite{ochoa:2015:crossovernetworks}
 was able to find all $r$-bit local optima of Mk Landscapes in \BigO{2^N} for any small,
@@
 The first major result in Gray-Box optimization was the proof that the list
-of fitness improving moves during local search can be kept updated in \BigO{1}
-time~\cite{whitley:2012:constant}. Each time a modification is made to a solution
-only modifications which share a non-linear relationship with the changes can
-go from non-improving to improving and vice versa. In Gray-Box optimization
-the non-linearity between variables is known. Furthermore the number of
-non-linear relationships is bounded above by \BigO{Mk} as each of the $M$
-subfunctions can cause a non-linear relationship between each of its $k$
-variables. As $M=$\BigO{N} and $k$ is constant, the amortized number of affected
-flips per variable is \BigO{1}.
+of fitness-improving moves can be kept update to date during local search in \BigO{1}
+time per bit flip~\cite{whitley:2012:constant}. In an Mk Landscape variables
+can only have a non-linear relationship if they appear together in at least one $s_i$.
+The fitness effect of flipping a bit can therefore only change when it has a non-linear
+relationship with the bit modified by local search. By definition the number of non-linear
+relationships in an Mk Landscape is linear with $N$, meaning on average the amortized
+number of updates is \BigO{1}.
 
-This limited and known non-linearity can be used to allow search to find only $r$-bit
-local optima. Consider a variable interaction graph, such that each variable maps
-to a vertex in the graph and each edge represents a non-linear relationship between
-variables. TODO
+This result was extended by~\cite{chicano:2014:ball} to include local search for local
+optima which cannot be improved using $r$ or fewer bits with no increase in asymptotic costs.
+Consider two variables $i$ and $j$ which do not appear in the same $s_i$.
+By definition the fitness effect of flipping both $i$ and $j$ is equal to the sum
+of flipping each independently. As a result if neither individual flip is fitness improving,
+flipping both together cannot be fitness improving. By examining the graph of non-linear
+interactions between variables this principle can be extended to any collection of $r$
+variables, with the number of useful collections growing at \BigO{N} when $r$ is a small constant.
+As a result the time to update the list of improving moves after a modification is still \BigO{1}.
 
-TODO Explain increased radius size~\cite{chicano:2014:ball}
+These advances can be applied directly to the task of finding all local optima in a
+landscape~\cite{ochoa:2015:crossovernetworks}.
+Instead of requiring \BigO{N} time to evaluate each solution and to check if it
+is a local optima, only \BigO{1} time is needed. Consider an enumeration ordering of the landscape
+which uses gray-codes, meaning that each transition between solutions requires exactly 1 bit flip.
+In Gray-Box optimization updating fitness and the list of improving moves after a single bit flip
+takes \BigO{1} time. That property holds even when looking for $r$-bit fitness-improving moves.
+Therefore Gray-Box enumeration is able to find all local optima in \BigO{2^N} time.
 
 \section{Hyperplane Elimination}
+\begin{figure*}
+  \centering
+  \includegraphicsfit{Enumerate}
+  \caption{Example change of enumeration ordering. The gray variables represent all dependencies
+           for some move $m_i$. By reordering, $m_i$'s lowest $index$ dependency improves from 2 to 4.}
+  \label{fig-enumerate}
+\end{figure*}
+
+Due to the limited non-linearity of the Gray-Box domain, it is possible to exclude large
+parts of the search space without missing any local optima.
+Consider the representation presented in the top of Figure~\ref{fig-enumerate}. In a Black-Box
+domain, enumeration would progress as a binary counter, treating $index$ zero (symbol $A$ in the solution) as
+the least significant bit. This ensures that before changing $index$ $i$, all possible settings of $index$
+0 through $i-1$ have been tested. This corresponds to examining the hyperplane with the lowest $i-1$ positions
+variable and all other positions fixed. The Gray-Box domain makes it possible to skip hyperplanes
+which cannot be local optima. In Figure~\ref{fig-enumerate} there exists a move $m_i$ which is a fitness improvement
+when enumeration starts (all variables set to 0). Due to the known relationships between variables,
+we know that the quality of $m_i$ only depends on variables $C$, $E$, $F$, and $H$.
+Therefore, until one of those four variables are modified, the solution cannot be a local optimum.
+As a result, the hyperplane **0*00*0 cannot contain a local optima and can
+therefore be eliminated from consideration during enumeration.
+More generally, if at any point during enumeration
+there exists a fitness-improving move, no local optima can exist until at least one
+dependency of that move is modified. Any hyperplane which has all of a fitness-improving
+move's dependencies fixed cannot contain any local optima.
+
+\begin{algorithm}
+  \caption{Find all local optima using Hyperplane Elimination.}
+  \label{alg-enumerate}
+  \begin{algorithmic}[1]
+    \Require $solution \leftarrow \{0\}^N$
+    \Require $move\_bin$ (List of sets of improving moves in $solution$)
+    \Require \textproc{MakeMove} (Function which flips $index$ in $solution$ and updates $move\_bin$)
+    \Ensure $found$ (list of all local optima)
+    \State $found \leftarrow [~]$
+    \State $index \leftarrow N-1$
+    \While{$index < N$}
+      \While{$index > 0$ \textbf{and} $move\_bin[index] = \emptyset$}\label{alg-enumerate-bincheck}
+        \State $index \leftarrow index-1$
+      \EndWhile
+      \If{$index = -1$}
+        \State $found \leftarrow found + [solution]$
+        \State $index \leftarrow 0$
+      \EndIf
+      \While{$index < N$ \textbf{and} $solution[index] = 1$}\label{alg-enumerate-counter}
+        \State \Call{MakeMove}{$index$}
+        \State $index \leftarrow index + 1$
+      \EndWhile
+      \If{$index < N$}
+        \State \Call{MakeMove}{$index$}\label{alg-enumerate-insert}
+      \EndIf
+    \EndWhile
+  \end{algorithmic}
+\end{algorithm}
+
+This knowledge can be exploited to skip parts of the enumeration,
+as shown in Algorithm~\ref{alg-enumerate}.
+Before starting, each fitness-improving move is put into a table $move\_bin$
+based on that move's lowest $index$ dependency. This is the first $index$ which
+can be modified by enumeration to change the fitness effect of making that move.
+In order to determine how much of the enumeration can be skipped, we must find
+the highest $index$ in $move\_bin$ which contains a fitness-improving move,
+as done by Line~\ref{alg-enumerate-bincheck}. If no move is fitness-improving,
+then a local optimum has been found.
+
+Algorithm~\ref{alg-enumerate} works by finding the highest $index$ in $move\_bin$
+which is not empty and then adding a 1 to that $index$ in $solution$. Initially
+all bins could contain a fitness-improving move, so $index$ starts at $N-1$.
+If at any point all bins are empty then the solution is added to the list of
+local optima $found$. Algorithm~\ref{alg-enumerate} then
+adds a 1 to $index$ using the loop on Line~\ref{alg-enumerate-counter}
+to perform carry operations and Line~\ref{alg-enumerate-insert} to create the new 1 value.
+Iteration stops when the carry exceeds the solution length.
+
+When performing subsequent iterations, not all bins need to be checked. Instead, the highest $index$
+bin that must be tested is the highest $index$ flipped by the previous iteration. This
+simplification is possible because
+the previous iteration has verified that all moves in higher $index$ bins are not fitness-improving, and no action performed during
+that iteration can make them fitness-improving.
+Furthermore, $index$ is always the location of the least significant 1 in $solution$, meaning
+iteration can continue immediately from the found $index$. This is true by construction
+as initially $solution$ contains all 0s. 1s are always inserted at the value of $index$
+with $index$ only increasing due to carry operations which clear all 1s lower than $index$.
+
+This hyperplane skipping extends to finding only $r$-bit local optima for all Mk landscapes.
+By increasing the number of potential fitness-improving moves, the number of hyperplanes which
+can be eliminated increases. However, the cost trade-off of increasing $r$ is unclear.
+While more solutions can be skipped, the time required to update $bin\_moves$ each iteration
+also increases. Therefore, the runtime effect of increasing $r$
+depends on the specific problem.
 
 \section{Improved Enumeration Ordering}
+As a final efficiency, the order in which variables in $solution$ are $index$ed can be reordered.
+When a move is fitness-improving,
+the amount of search space that is skipped depends on how high its lowest $index$ dependency is. Therefore,
+by rearranging the order to make its lowest $index$ dependency higher, more search space can be skipped. 
+Figure~\ref{fig-enumerate} shows how changing the $index$ order of variables
+improves $m_i$'s lowest $index$ dependency from 2 to 4. Consider that before reordering $m_i$ allows the hyperplane
+**0*00*0 to be skipped, while after reordering that becomes ****0000.
+Now whenever $m_i$ is a fitness improvement, Line~\ref{alg-enumerate-bincheck} in Algorithm~\ref{alg-enumerate}
+skips 4 times as much search space.
 
-% Proof sketch
+We perform this reordering in a greedy fashion, such that the move with the
+least-unmapped dependencies has all of its remaining dependencies mapped to the
+most significant remaining positions. With proper bookkeeping this algorithm requires
+\BigO{N} time. Assuming that all moves are equally likely to be fitness improving
+during enumeration, we argue below that the optimal bit ordering for skipping solutions is a greedy
+ordering.
+
+First we shall show that all optimal orderings of bits are optimal orderings of packed moves.
+Consider that lowering a bit in enumeration order can only decrease the number of skipped
+solutions if doing so decreases the minimum dependency of a move. However, doing so may
+allow the minimum dependency of some unrelated move to increase. Therefore when ordering
+bits all that matters is the order in which the last bit each move is assigned.
+
+% Proof
 Moves can have their dependencies fixed in some permutation $O$ of all possible moves,
@@
 $D_O(i, m)$ is a function which returns how many unfixed dependencies move $m$ has after all
-$i$ moves in $O$ before $i$ have been fixed. A greedy solution to this problem works such
+moves in $O$ before $i$ have been fixed. A greedy solution to this problem is one such
 that $O_i$ is set to be the move which minimizes $D_O(i, m)$.
@@
 This is because $D_O(i, m)$ can only decrease as $i$ increases, meaning that if some $m$
-has a lower value at $i$ than $O_i$, some index between $i$ and when $m$ appears in $O$
-this property must hold. Finally, all non-greedy solutions must have some $\hat{i}$
+has a lower value at $i$ than $O_i$, this property must hold for
+some index between $i$ and when $m$ appears in $O$.
+Finally, all non-greedy solutions must have some $\hat{i}$
 which is the maximum $i$ value for which $D_O(i, O_i) > D_O(i, O_{i+1})$ is true.
@@
 Performing the swap will improve the minimum dependency of $O_{\hat{i}+1}$ more than it hurts
-$O_{\hat{i}}$. Therefore, performing this swap will make $O^\star$ even better, implying it was
-not optimal before. Therefore there cannot exist an optimal ordering which is not greedy.
+$O_{\hat{i}}$. Therefore, performing this swap will make $O^\star$ skip even more solutions,
+contradicting the assertion that it was optimal. Therefore there cannot exist an optimal ordering which is not greedy.
 
@@
 In OneMax, each bin contains exactly 1 move. All variables in the current solution
-which disagree with the global optimum will be fitness improving moves.
+which disagree with the global optimum will be fitness-improving moves.
 Initially $index$ is set to $N-1$ and is decreased by at most 1 in any step.
@@
 is then spent adding a 1 to $index$ 0, which in the worst case requires \BigO{N}
-carry operations. For all future iterations $index$ is a fitness improving move
+carry operations. For all future iterations $index$ is a fitness-improving move
 and is currently set to 1. In these iterations at least 1 carry operation must
@@
 
-While any $f_i$ contains a fitness improving move, Algorithm~\ref{alg-enumerate}
+While any $f_i$ contains a fitness-improving move, Algorithm~\ref{alg-enumerate}
 will find the one with the highest $i$ and enumerate it until it no longer
@@
 $T(i) = |l_i|*T(i-1)+2^k$ where $|l|$ is the number of ways $f_i$ can be set such
-that it contains no fitness improving moves. Each $l_i$ exposes a new hyperplane
+that it contains no fitness-improving moves. Each $l_i$ exposes a new hyperplane
 which can contain local optima, which causes the recursive call to $T(i-1)$.
@@
 is within a constant factor of $2^k$ of optimal.
+
+When looking for only $r$-bit local optima of $k$-bound separable problems,
+Algorithm~\ref{alg-enumerate} will continue to be within $2^k$ of optimal.
+For $r < k$ the additional moves reduce $c$ without increasing the cost
+by more than a constant.
+When $r \geq k$ there can only be one local optima. Furthermore, as
+only non-linearly related subsets of variables must be tested to check
+for fitness improvements, no subsets can be added which are larger than $k$.
+Therefore even if $r=N-1$, the number of subsets only grows at \BigO{N}.
+As $c=1$ in this case, the cost of finding the single global optimum
+only requires the \BigO{2^kN/k} time to find the first local optimum, which is
+optimal.
 

### 2015-09-04 Hyperplane Enumeration/main.tex
Major revision after writing group.

@@
 \section{Introduction}
-The ruggedness and high dimensionality of most interesting landscapes makes them challenging
-to visualize or otherwise analyze. However, doing so can be helpful in
-quantifying the difficulty of a problem, and how to best design algorithms to deal
+The ruggedness and high dimensionality of most problem landscapes makes them challenging
+analyze and understand. However, doing so can be helpful in
+quantifying the difficulty of a problem, and how to design algorithms to deal
 with those difficulties. 
@@
 
-One avenue for performing this problem understanding is to
+A common way of analyzing landscapes is to
 examine the frequency and distribution of local optima~\cite{boese:1994:bigvalley},
@@
 However, finding all of the local optima is prohibitively time consuming
-for even small problems, with many studies limited to 18-bit
+even for small problems, with many studies limited to 18-bit
 problems~\cite{tomassini:2008:nknetworks,verel:2011:nknetworks}.
-By leveraging recent advancements in Gray-Box optimization which allow for
+By leveraging recent advancements in Gray-Box optimization that allow for
 constant time local search~\cite{chicano:2014:ball}, this limit was raised
@@
 methods can approximate some of these statistics for much larger
-landscapes~\cite{iclanzan:2014:somnetworks}, they do not provide the complete
-picture necessary for many metrics~\cite{ochoa:2015:crossovernetworks}.
+landscapes~\cite{iclanzan:2014:somnetworks}, they do not provide enough
+detail necessary for some metrics~\cite{ochoa:2015:crossovernetworks}.
 
-Here we will introduce a method for finding all local optima of problems
+Here we introduce a method for finding all local optima of problems
 with up to 80 bits. The cornerstone of this method is the identification
-of hyperplanes which cannot contain any local optima, allowing large
+of hyperplanes that cannot contain any local optima, allowing large
 portions of the search space to be skipped during enumeration.
+The techniques used apply to the generalized problem class of Mk Landscapes,
+which contains many real world and benchmark combinatorial problems.
 
 \section{Mk Landscapes as a tool for Problem Generalization}
-An Mk Landscape is any function which can be expressed in the following form:
+An Mk Landscape~\cite{whitley:2015:mk} is any function
+who's value is equal to the sum of a set of subfunctions.
+Each subfunction uses a small subset of variables from the original function's input.
+Combined with limitations on the number of subfunctions and the size of each subfunction's
+subset, Mk Landscapes allow for highly efficient local~\cite{whitley:2012:constant,chicano:2014:ball}
+and global~\cite{goldman:2015:GBO,tintos:2015:partitioncross} search.
+
+Formally, an Mk Landscape is any function $f : \mathbb{B}^{N}\rightarrow \mathbb{R}$
+that can be expressed in the following form:
 \begin{equation}
@@
 In this equation $f_i$ is a subfunction such that $f_i : \mathbb{B}^{|s_i|}\rightarrow \mathbb{R}$.
+Each $s_i$ is a set of variables in $x$, such that $|s_i| \leq k$.
 The $mask$ function
-returns the values in $x$ associated with each variable in the set $s_i$.
-Each $|s_i| \leq k$, where $k$ is a constant with respect to $N=|x|$,
-the length of $x$. Finally, $M$ is constrained to be \BigO{N}.
-
-A critical aspect of this problem class is that it can be optimized using Gray-Box
-techniques, which can allow for highly efficient local~\cite{whitley:2012:constant,chicano:2014:ball}
-and global~\cite{goldman:2015:GBO,tintos:2015:partitioncross} search. These algorithms
-exploit the known and limited non-linearity between variables expressed by all $s_i$,
-as well as perform partial solution evaluation using all $f_i$.
+returns the values in $x$ associated with each variable in $s_i$.
+The total number of subfunctions $M$ is constrained to grow at \BigO{N}, and $k$
+is constant with repect to $N$.
 
 This formulation can represent many problems of real-world interest, as well as
-the most commonly used combinatorial benchmark problems. In this work we shall
+the most commonly used combinatorial benchmark problems. In this work we
 examine 5 problems in particular: Concatenated Traps, Adjacent and Random NKq Landscapes,
@@
 most advanced methods can solve it trivially~\cite{goldman:2012:ltga} and when expressed
-as an Mk Landscape can be solved exactly in \BigO{N} time~\cite{whitley:2015:mk}.
+as an Mk Landscape it can be solved exactly in \BigO{N} time~\cite{whitley:2015:mk}.
 We therefore include it only because its structure allows for straight forward algorithm analysis.
@@
 and (3) the number of unique fitness values $q$. From these parameters a landscape is generated
-by creating $M=N$ subfunctions $f_i$, where $f_i$ uses variable $i$ and $K$ others to look up
+by creating $M=N$ subfunctions $f_i$, where $f_i$ uses variable $x_i$ and $K$ others to look up
 a fitness value in the range $[0..q-1]$ from a randomly generated table. This structure
@@
 variables in each subfunction are chosen: Adjacent and Random. In Adjacent NKq, $f_i$ depends
-on variables $[i..(i+k) \bmod N]$. In Random NKq, $f_i$ depends on variables $i$ and a random
-set of $K$ unique variables which does not include $i$. While Random NKq is NP-Hard,
+on variable indices $[i..(i+k) \bmod N]$. In Random NKq, $f_i$ depends on $x_i$ and a random
+set of $K$ unique variables that does not include $x_i$. While Random NKq is NP-Hard,
 the structure of Adjacent NKq allows for a polynomial time dynamic programming solution~\cite{wright:2000:solvingnk}.
@@
 Each atom in the glass (vertex) can be assigned a spin, with the goal being to
-find the set of assignments which minimize the energy between nearby atoms (edges).
+find the set of assignments that minimize the energy between nearby atoms (edges).
 Similar to Adjacent NKq, the $2D\pm J$ subset of Ising Spin Glasses can be polynomially
@@
 of its neighbors, which when properly cached takes \BigO{N} per solution.
-Extending this method to look for solutions which cannot be improved by
+Extending this method to look for solutions that cannot be improved by
 flipping $r$ or fewer bits requires \BigO{N^r2^N} time. However,
@@
 The first major result in Gray-Box optimization was the proof that the list
-of fitness-improving moves can be kept update to date during local search in \BigO{1}
+of fitness-improving moves can be updated after each flip in local search in \BigO{1}
 time per bit flip~\cite{whitley:2012:constant}. In an Mk Landscape variables
@@
 This result was extended by~\cite{chicano:2014:ball} to include local search for local
-optima which cannot be improved using $r$ or fewer bits with no increase in asymptotic costs.
-Consider two variables $i$ and $j$ which do not appear in the same $s_i$.
-By definition the fitness effect of flipping both $i$ and $j$ is equal to the sum
+optima that cannot be improved using $r$ or fewer bits with no increase in asymptotic costs.
+Consider two variables $x_i$ and $x_j$ that do not appear in the same $s_i$.
+By definition the fitness effect of flipping both $x_i$ and $x_j$ is equal to the sum
 of flipping each independently. As a result if neither individual flip is fitness improving,
@@
 is a local optima, only \BigO{1} time is needed. Consider an enumeration ordering of the landscape
-which uses gray-codes, meaning that each transition between solutions requires exactly 1 bit flip.
+that uses gray-codes, meaning that each transition between solutions requires exactly 1 bit flip.
 In Gray-Box optimization updating fitness and the list of improving moves after a single bit flip
@@
 variable and all other positions fixed. The Gray-Box domain makes it possible to skip hyperplanes
-which cannot be local optima. In Figure~\ref{fig-enumerate} there exists a move $m_i$ which is a fitness improvement
+that cannot be local optima. In Figure~\ref{fig-enumerate} there exists a move $m_i$ that is a fitness improvement
 when enumeration starts (all variables set to 0). Due to the known relationships between variables,
@@
 there exists a fitness-improving move, no local optima can exist until at least one
-dependency of that move is modified. Any hyperplane which has all of a fitness-improving
+dependency of that move is modified. Any hyperplane that has all of a fitness-improving
 move's dependencies fixed cannot contain any local optima.
@@
     \Require $move\_bin$ (List of sets of improving moves in $solution$)
-    \Require \textproc{MakeMove} (Function which flips $index$ in $solution$ and updates $move\_bin$)
+    \Require \textproc{MakeMove} (Function that flips $index$ in $solution$ and updates $move\_bin$)
     \Ensure $found$ (list of all local optima)
@@
 Before starting, each fitness-improving move is put into a table $move\_bin$
-based on that move's lowest $index$ dependency. This is the first $index$ which
+based on that move's lowest $index$ dependency. This is the first $index$ that
 can be modified by enumeration to change the fitness effect of making that move.
 In order to determine how much of the enumeration can be skipped, we must find
-the highest $index$ in $move\_bin$ which contains a fitness-improving move,
+the highest $index$ in $move\_bin$ that contains a fitness-improving move,
 as done by Line~\ref{alg-enumerate-bincheck}. If no move is fitness-improving,
@@
 Algorithm~\ref{alg-enumerate} works by finding the highest $index$ in $move\_bin$
-which is not empty and then adding a 1 to that $index$ in $solution$. Initially
+that is not empty and then adding a 1 to that $index$ in $solution$. Initially
 all bins could contain a fitness-improving move, so $index$ starts at $N-1$.
@@
 simplification is possible because
-the previous iteration has verified that all moves in higher $index$ bins are not fitness-improving, and no action performed during
-that iteration can make them fitness-improving.
+the previous iteration has verified that all moves in higher $index$ bins are not fitness improving, and no action performed during
+that iteration can make them fitness improving.
 Furthermore, $index$ is always the location of the least significant 1 in $solution$, meaning
@@
 This hyperplane skipping extends to finding only $r$-bit local optima for all Mk landscapes.
-By increasing the number of potential fitness-improving moves, the number of hyperplanes which
+By increasing the number of potential fitness-improving moves, the number of hyperplanes that
 can be eliminated increases. However, the cost trade-off of increasing $r$ is unclear.
-While more solutions can be skipped, the time required to update $bin\_moves$ each iteration
+While more solutions can be skipped, the time required to update $move\_bin$ each iteration
 also increases. Therefore, the runtime effect of increasing $r$
@@
 improves $m_i$'s lowest $index$ dependency from 2 to 4. Consider that before reordering $m_i$ allows the hyperplane
-**0*00*0 to be skipped, while after reordering that becomes ****0000.
+**000000 to be skipped by Algorithm~\ref{alg-enumerate}, while after reordering that becomes ****0000.
 Now whenever $m_i$ is a fitness improvement, Line~\ref{alg-enumerate-bincheck} in Algorithm~\ref{alg-enumerate}
-skips 4 times as much search space.
+skips 4 times as many solutions.
 
@@
 
-First we shall show that all optimal orderings of bits are optimal orderings of packed moves.
+First we show that all optimal orderings of bits are optimal orderings of packed moves.
 Consider that lowering a bit in enumeration order can only decrease the number of skipped
@@
 allow the minimum dependency of some unrelated move to increase. Therefore when ordering
-bits all that matters is the order in which the last bit each move is assigned.
+bits, all that matters is the order in which the last bit of each move is assigned.
 
@@
 such that $O_0$ is the first move to have its dependencies fixed.
-$D_O(i, m)$ is a function which returns how many unfixed dependencies move $m$ has after all
+$D_O(i, m)$ is a function that returns how many unfixed dependencies move $m$ has after all
 moves in $O$ before $i$ have been fixed. A greedy solution to this problem is one such
-that $O_i$ is set to be the move which minimizes $D_O(i, m)$.
+that $O_i$ is set to be the move that minimizes $D_O(i, m)$.
 
@@
 Finally, all non-greedy solutions must have some $\hat{i}$
-which is the maximum $i$ value for which $D_O(i, O_i) > D_O(i, O_{i+1})$ is true.
+that is the maximum $i$ value for which $D_O(i, O_i) > D_O(i, O_{i+1})$ is true.
 
-Consider the optimal ordering $O^\star$ which is not greedy.
+Consider an optimal ordering $O^\star$ that is not greedy.
 These criteria mean that $D_{O^\star}(\hat{i}, O_{\hat{i}}) > D_{O^\star}(\hat{i}, O_{\hat{i}+1})$.
 Swapping $O_{\hat{i}}$ and $O_{\hat{i}+1}$ cannot change the minimum dependency of any other moves.
-Performing the swap will improve the minimum dependency of $O_{\hat{i}+1}$ more than it hurts
-$O_{\hat{i}}$. Therefore, performing this swap will make $O^\star$ skip even more solutions,
-contradicting the assertion that it was optimal. Therefore there cannot exist an optimal ordering which is not greedy.
+Performing the swap improves the minimum dependency of $O_{\hat{i}+1}$ more than it hurts
+$O_{\hat{i}}$. Therefore, performing this swap makes $O^\star$ skip even more solutions,
+contradicting the assertion that it was optimal. Therefore there cannot exist an optimal ordering that is not greedy.
 
 \section{Complexity Classes for Simple Landscapes}
+Understanding the runtime complexity of Algorithm~\ref{alg-enumerate} requires
+knowledge of how often each move is fitness improving. For most landscapes
+this is intractable to do theoretically. However, for some restricted problem
+types its possible to rigorously determine Algorithm~\ref{alg-enumerate}'s complexity class.
 
-\subsection{OneMax}
-In OneMax, each bin contains exactly 1 move. All variables in the current solution
-which disagree with the global optimum will be fitness-improving moves.
-Initially $index$ is set to $N-1$ and is decreased by at most 1 in any step.
+\subsection{Linear Functions}
+A linear function is any $f : \mathbb{B}^{N}\rightarrow \mathbb{R}$ which contains
+no non-linear terms between variables. These functions are of the form:
+\begin{equation}
+  f(x) = \sum_{i=0}^{N-1} w_ix_i
+  \label{eq-linear}
+\end{equation}
+In MK Landscape terms, $M=N$ and $k=1$. The most well known linear function is OneMax,
+in which $w_i=1$ for all $i$.
+
+When Algorithm~\ref{alg-enumerate} is applied to a linear
+function, each index of $move\_bin$ can contain at most 1
+move, and a move is fitness improving whenever it disagrees with the global optimum.
+Initially $index$ is set to $N-1$ and is decreased by at least 1 in any step after the first.
 Before finding the global optimum no carry operations can occur, meaning the
 cost of each iteration is equal to the amount by which $index$ is decreased.
+Each step will flip a bit which disagreed with the global optimum.
 Therefore, Algorithm~\ref{alg-enumerate} requires \BigO{N} time to reach
@@
 
-After finding the global optimum of OneMax, $index$ is set to 0. One iteration
+After finding the global optimum, $index$ is set to 0. One iteration
 is then spent adding a 1 to $index$ 0, which in the worst case requires \BigO{N}
@@
 requires \BigO{N} time to reach termination after finding the global optimum. Combined
-with initialization and the time to find the global optimum, the total complexity is \BigO{N}.
+with initialization and the time to find the global optimum, the total complexity is \BigO{N},
+which is optimal.
 
 \subsection{$k$-Bound Separable Problem}
-Consider any problem which is composed of non-overlapping
+Consider any problem that is composed of non-overlapping
 subfunctions, each using using $k$ or fewer bits. Regardless
-of how they are initially ordered, reordering will ensure that
+of how they are initially ordered, reordering ensures that
 variables of a function are always consecutively placed in enumeration.
@@
 While any $f_i$ contains a fitness-improving move, Algorithm~\ref{alg-enumerate}
-will find the one with the highest $i$ and enumerate it until it no longer
+finds the one with the highest $i$ and enumerates its variables until it no longer
 contains an improving move. In total this enumeration requires less than $2^k$
 steps. Each $f_i$ is considered sequentially, meaning the first local optimum
-will be found in $2^kN/k$ time.
+is found in \BigO{2^kN/k} time.
 
@@
 fixed. We represent the time required to find all local optima in the hyperplane
-with $f_j, j>i$ as $T(i)$. $T(0)=2^k$ as all values of $f_0$ must be enumerated.
-$T(i) = |l_i|*T(i-1)+2^k$ where $|l|$ is the number of ways $f_i$ can be set such
+with $f_j, j>i$ fixed as $T(i)$. $T(0)=2^k$ as all values of $f_0$ must be enumerated.
+$T(i) = |l_i|*T(i-1)+2^k$ where $|l_i|$ is the number of ways $f_i$ can be set such
 that it contains no fitness-improving moves. Each $l_i$ exposes a new hyperplane
-which can contain local optima, which causes the recursive call to $T(i-1)$.
+that can contain local optima, which causes the recursive call to $T(i-1)$.
 Assuming for simplicity that all subfunctions use exactly $k$ bits,
@@
 The number of local optima in the landscape is $c^{N/k}$ meaning that Algorithm~\ref{alg-enumerate}
-is within a constant factor of $2^k$ of optimal.
+is within a constant factor ($2^k$) of optimal.
 
 When looking for only $r$-bit local optima of $k$-bound separable problems,
-Algorithm~\ref{alg-enumerate} will continue to be within $2^k$ of optimal.
+Algorithm~\ref{alg-enumerate} continues to be within a constant of optimal.
 For $r < k$ the additional moves reduce $c$ without increasing the cost
 by more than a constant.
-When $r \geq k$ there can only be one local optima. Furthermore, as
-only non-linearly related subsets of variables must be tested to check
-for fitness improvements, no subsets can be added which are larger than $k$.
+When $r \geq k$ there can only be one local optimum: the global optimum. Furthermore, as
+only non-linearly related subsets of variables must be checked
+for fitness improvements, no subsets can be added that are larger than $k$.
 Therefore even if $r=N-1$, the number of subsets only grows at \BigO{N}.
@@
 with $c=2$.  Algorithm~\ref{alg-enumerate} requires $2^k2^{N/k}$ time
-to find all $2^{N/k}$ local optima in this problem.
+to find all $2^{N/k}$ local optima in this problem. Note also that
+in the degenerate case of $k=1$, a $k$-bound separable problem
+is a linear function. As such, when $k=1$ and $c=1$,
+\BigO{2^kN/k} = \BigO{N}.
 

### 2015-09-09 Hyperplane Enumeration/main.tex
Added figures to text, added table, changed some results section titles

@@
 
-\section{Timing Comparisons}
-\subsection{Increasing problem size}
-Consider including boxplots of largest size all are successful.
+\section{Finding Local Optima}
+TODO Concatenated Traps using $N>40$ has slope of 2.
+\begin{figure*}
+  \centering
+  \includegraphicsfit{length-method}
+  \caption{Comparison of how each method scales with problem size on log-linear scales. Each point is the mean
+           runtime over 30 instances and each line the linear model. All confidence intervals too tight to see.}
+  \label{fig-length-method}
+\end{figure*}
 
-\subsection{The Effect of Radius}
+\begin{table}
+	\centering
+	\caption{Value of $m$ when fitting a model to $y = 2^c2^{mN}$ where $y$ is number of seconds required to
+	         find all local optima of a problem using $N$ bits. Optimal refers to the same model where
+	         $y$ is the number of local optima.}
+	\begin{tabular}{|r|c|c|c|c|}
+	  \hline
+	    & \textbf{Gray-Box} & \textbf{Hyper} & \textbf{Hyper-Reorder} & \textbf{Optimal} \\ \hline
+	  Concatenated Traps & 0.9809 & 0.5765 & 0.1601 & 0.2000 \\ \hline
+    Adjacent NKq & 0.9922 & 0.5972 & 0.3411 & 0.3603 \\ \hline
+    Random NKq & 0.9920 & 0.6202 & 0.3984 & 0.3531 \\ \hline
+    Ising Spin Glass & 0.9821 & 0.6576 & 0.5909 & 0.6015 \\ \hline
+    MAX-kSAT & 0.9949 & 0.8893 & 0.7735 & 0.5393 \\ \hline
+  \end{tabular}
+  \label{table-results}
+\end{table}
+
+\begin{figure*}
+  \centering
+  \includegraphicsfit{boxplot-method}
+  \caption{Comparison of completion time variance for the largest size of each problem
+           where all three methods were fully successful.}
+  \label{fig-boxplot-method}
+\end{figure*}
+
+\section{Finding $r$-Bit Local Optima}
 

### 2015-09-09 Hyperplane Enumeration/main.tex
Changed up sections, more figures

@@
 
-\section{Finding Local Optima}
+\section{Experiments}
+TODO Explain experimental setup details here.
+
+\subsection{Finding Local Optima}
 TODO Concatenated Traps using $N>40$ has slope of 2.
@@
 
-\section{Finding $r$-Bit Local Optima}
+\subsection{Finding $r$-Bit Local Optima}
 
-\section{Examining Local Optima}
+\begin{figure*}
+  \centering
+  \includegraphicsfit{slope-method}
+  \caption{Estimated slope with 95\% confidence intervals for each method to find $r$-bit local optima. Slope is $m$ in the model $y = 2^c2^{mN}$
+           where $y$ is number of seconds required to
+	         find all local optima of a problem using $N$ bits.}
+  \label{fig-boxplot-method}
+\end{figure*}
+
+\subsection{Examining Local Optima}
+
+\begin{figure*}
+  \centering
+  \includegraphicsfit{length-radius}
+  \caption{Number of $r$-bit local optima as $N$ increases for each problem.}
+  \label{fig-boxplot-method}
+\end{figure*}
 

### 2015-09-10 Hyperplane Enumeration/main.tex
Added text to the experiments section

@@
 \section{Mk Landscapes as a tool for Problem Generalization}
+\label{sec-mk}
 An Mk Landscape~\cite{whitley:2015:mk} is any function
@@
 \section{Improved Enumeration Ordering}
+\label{sec-reorder}
 As a final efficiency, the order in which variables in $solution$ are $index$ed can be reordered.
@@
 \section{Experiments}
-TODO Explain experimental setup details here.
+We compare three methods for finding all local optima: Gray-Box, Hyper, and Hyper-Reorder.
+The Gray-Box method refers to previous work~\cite{ochoa:2015:crossovernetworks} which does
+not perform hyperplane elimination, with Hyper and Hyper-Reorder being hyperplane elimination
+without and with the enumeration order improvements discussed in Section~\ref{sec-reorder}.
+
+For each of the five types of Mk Landscape outlined in Section~\ref{sec-mk}, we tested all
+algorithms for all $N$ in $[15..100]$ and $r$ in $[1..4]$. For each problem size we generated
+30 problem instances, with each $r$ of each algorithm run once per instance. Variables were randomly
+ordered in the genome for problems with natural structure. For example, variables in the same $s_i$
+of an Adjacent NKq instance are not likely to be consecutive in the genome. However, $s_i$ still overlaps
+$s_{i+1}$ in all but 1 variable.
+
+Each algorithm was given a maximum of four hours on a 2.5GHz Intel Xeon E5-2670v2 processor to find
+all $r$-bit local optima. Any run which failed was rerun once to prevent cluster issues from introducing
+noise. For the same reason any run which was over ten times slower than the next slowest run of the
+same configuration with a different seed was rerun. This resulted in 33 out of 36,000 runs being rerun.
+If the algorithm failed to complete in the alloted time for an instance,
+we declared it unsuccessful for that combination of $r$ and $N$ of that problem. All results
+we report are for successful combinations. For timing purposes local optima were counted but not
+recorded.
 
 \subsection{Finding Local Optima}
-TODO Concatenated Traps using $N>40$ has slope of 2.
+Our first experiment is to determine the empirical complexity of each method with respect to problem size.
+Of the 681 configurations where at least one method was successful, Hyper-Reorder had the lowest
+mean time to find all local optima all but 2 times. In both cases Hyper-Reorder was within a tenth of a second
+of the best method.
+Figure~\ref{fig-boxplot-method} shows the variance in runtime for the largest problem size all three methods
+were successful. In all cases it is clear that Hyper-Reorder outperforms the other methods.
+
+\begin{figure*}
+  \centering
+  \includegraphicsfit{boxplot-method}
+  \caption{Comparison of completion time variance for the largest size of each problem
+           where all three methods were successful at finding all 1-bit local optima.}
+  \label{fig-boxplot-method}
+\end{figure*}
+
+More critically, Figure~\ref{fig-length-method} shows how each method scales with problem size.
+Not only is Hyper-Reorder out performing the alternatives on all sizes, the rate at which it
+scales to problem complexity is better. Table~\ref{table-scaling} shows the model fit for each
+method on each problem. This value is the slope of each line shown in Figure~\ref{fig-length-method}.
+As expected Gray-Box scales at almost exactly $2^N$. Using Hyperplane elimination without
+reordering reduces this complexity, with reordering making further improvements. In the final
+column we should the growth rate of the number of local optima. It is impossible for any method
+which enumerates all local optima to grow slower than this rate. On Concatenated Traps, Adjacent
+NKq, and Ising Spin Glass, Hyper-Reorder has an estimated slope better than optimal. This is
+due to lower order terms dominating runtime for small problem sizes, as can be see in
+Figure~\ref{fig-length-method}. For instance, fitting
+the curve only to problems where $N>60$ makes Hyper-Reorder have $m=0.199$ on Concatenated Traps and
+$m=0.358$ on Adjacent NKq.
+
+With the exception of MAX-kSAT, Hyper-Reorder's time required to find all local optima
+appears to be growing at the same rate as the total number of local optima. We suspect
+some of this deficiency on MAX-kSAT is due to its significantly higher number of non-linear
+interactions. Concatenated Traps, Adjacent NKq, and Ising Spin Glass all have exactly 4
+non-linear relationships per variable. Random NKq and MAX-kSAT both increase with problem
+size, with Random NKq ranging between 5 and 6. With $N=15$, MAX-kSAT has on average nearly
+12 non-linear relationships per variable. By $N=30$ that grows to just over 17. Having each
+variable depend on over half the genome significantly limits the size of hyperplanes which
+can be eliminated.
+
 \begin{figure*}
@@
   \includegraphicsfit{length-method}
-  \caption{Comparison of how each method scales with problem size on log-linear scales. Each point is the mean
+  \caption{Comparison of how each method scales with problem size when finding 1-bit local optima
+           on log-linear scales. Each point is the mean
            runtime over 30 instances and each line the linear model. All confidence intervals too tight to see.}
@@
 	\caption{Value of $m$ when fitting a model to $y = 2^c2^{mN}$ where $y$ is number of seconds required to
-	         find all local optima of a problem using $N$ bits. Optimal refers to the same model where
+	         find all 1-bit local optima of a problem using $N$ bits. Optimal refers to the same model where
 	         $y$ is the number of local optima.}
@@
 	    & \textbf{Gray-Box} & \textbf{Hyper} & \textbf{Hyper-Reorder} & \textbf{Optimal} \\ \hline
-	  Concatenated Traps & 0.9809 & 0.5765 & 0.1601 & 0.2000 \\ \hline
-    Adjacent NKq & 0.9922 & 0.5972 & 0.3411 & 0.3603 \\ \hline
-    Random NKq & 0.9920 & 0.6202 & 0.3984 & 0.3531 \\ \hline
-    Ising Spin Glass & 0.9821 & 0.6576 & 0.5909 & 0.6015 \\ \hline
-    MAX-kSAT & 0.9949 & 0.8893 & 0.7735 & 0.5393 \\ \hline
+    Concatenated Traps & 0.9809 & 0.5831 & 0.1620 & 0.2000 \\ \hline
+    Adjacent NKq & 0.9922 & 0.5972 & 0.3419 & 0.3603 \\ \hline
+    Random NKq & 0.9920 & 0.6229 & 0.3989 & 0.3531 \\ \hline
+    Ising Spin Glass & 0.9821 & 0.6722 & 0.5976 & 0.6015 \\ \hline
+    MAX-kSAT & 0.9949 & 0.8922 & 0.7737 & 0.5393 \\ \hline
   \end{tabular}
-  \label{table-results}
+  \label{table-scaling}
 \end{table}
-
-\begin{figure*}
-  \centering
-  \includegraphicsfit{boxplot-method}
-  \caption{Comparison of completion time variance for the largest size of each problem
-           where all three methods were fully successful.}
-  \label{fig-boxplot-method}
-\end{figure*}
 
@@
 	         find all local optima of a problem using $N$ bits.}
-  \label{fig-boxplot-method}
+  \label{fig-slope-method}
 \end{figure*}
@@
   \caption{Number of $r$-bit local optima as $N$ increases for each problem.}
-  \label{fig-boxplot-method}
+  \label{fig-length-radius}
 \end{figure*}

### 2015-09-14 Hyperplane Enumeration/main.tex
More details at the extreme sizes, first pass of abstract

@@
 \begin{abstract}
-TODO
+A common tool for understanding combinatorial problem landscapes
+is to examine the properties of its local optima. Unfortunately,
+exhaustive methods for finding local optima are exceptionally time
+consuming and are therefore limited to very small problem sizes.
+We propose a method for exploiting generalized problem structure
+to significantly speed up this process by skipping hyperplanes
+which cannot contain local optima. In most of our experiments
+this hyperplane elimination allows enumeration time to scale with
+the total number of local optima instead of the total landscape size.
+We prove optimality for linear and $k$-bounded problems, and provide
+empirical evidence of optimality on NKq Landscapes and Ising Spin Glasses.
+Furthermore, we are able to extend this method to finding only solutions
+which cannot be improving by flipping $r$ or fewer bits. In some cases
+increasing $r$ actually reduces total runtime by drastically reducing the
+number of local optima. While previous methods were limited to
+problems with at most 18 or 30 bits, hyperplane elimination can enumerate the same problems with
+75 bits. When using higher $r$ values this bound can be pushed to 200 bit problems
+on average enumerated in under 6 minutes.
 
@@
 
-Here we introduce a method for finding all local optima of problems
-with up to 80 bits. The cornerstone of this method is the identification
+Here we introduce a method for finding all local optima for the same problems using
+up to 75 bits. The cornerstone of this method is the identification
 of hyperplanes that cannot contain any local optima, allowing large
@@
 Of the 681 configurations where at least one method was successful, Hyper-Reorder had the lowest
-mean time to find all local optima all but 2 times. In both cases Hyper-Reorder was within a tenth of a second
+mean time to find all local optima all but 2 times. In both of those cases Hyper-Reorder was within a tenth of a second
 of the best method.
@@
 reordering reduces this complexity, with reordering making further improvements. In the final
-column we should the growth rate of the number of local optima. It is impossible for any method
+column we show the growth rate of the number of local optima. It is impossible for any method
 which enumerates all local optima to grow slower than this rate. On Concatenated Traps, Adjacent
@@
 the curve only to problems where $N>60$ makes Hyper-Reorder have $m=0.199$ on Concatenated Traps and
-$m=0.358$ on Adjacent NKq.
+$m=0.358$ on Adjacent NKq. For Concatenated Traps this matches our theoretical predictions that
+Hyper-Reorder scales at $2^k2^{N/k}$ as $\frac{1}{k}=0.2$.
 
-With the exception of MAX-kSAT, Hyper-Reorder's time required to find all local optima
-appears to be growing at the same rate as the total number of local optima. We suspect
-some of this deficiency on MAX-kSAT is due to its significantly higher number of non-linear
-interactions. Concatenated Traps, Adjacent NKq, and Ising Spin Glass all have exactly 4
-non-linear relationships per variable. Random NKq and MAX-kSAT both increase with problem
-size, with Random NKq ranging between 5 and 6. With $N=15$, MAX-kSAT has on average nearly
-12 non-linear relationships per variable. By $N=30$ that grows to just over 17. Having each
-variable depend on over half the genome significantly limits the size of hyperplanes which
-can be eliminated.
+This reduction in growth complexity translates into a significant increase in the size of problems
+which can be enumerated. For instance, using the previous best enumeration methods~\cite{ochoa:2015:crossovernetworks}
+the largest Adjacent NKq problem size which can be enumerated in 4 hours is 34 bits. Given the same time
+Hyper-Reorder can enumerate problems with 75 bits. Using the regression model, we estimate solving
+a 75 bit problem using Gray-Box would require 230 million years. Similarly, Hyper-Reorder's largest
+Random NKq size of 69 bits would require Gray-Box 3 million years. These larger sizes make for
+much better approximations of landscapes interesting to search methods.
 
@@
 
+With the exception of MAX-kSAT, Hyper-Reorder's time required to find all local optima
+appears to be growing at the same rate as the total number of local optima. We suspect
+some of this deficiency on MAX-kSAT is due to its significantly higher number of non-linear
+interactions. Concatenated Traps, Adjacent NKq, and Ising Spin Glass all have exactly 4
+non-linear relationships per variable. For small problem sizes
+Random NKq and MAX-kSAT both increase with problem
+size. Random NKq's average ranges between 5 and 6 on problems we tested.
+With $N=15$, MAX-kSAT has on average nearly
+12 non-linear relationships per variable. By $N=30$ that grows to just over 17. Having each
+variable depend on over half the genome significantly limits the size of hyperplanes which
+can be eliminated. While the number of non-linear relationships per variable
+is asymptotically constant, this growth on small problems is likely affecting runtime.
+
 \subsection{Finding $r$-Bit Local Optima}
+A major advantage of Gray-Box enumeration is that it can find $r$-bit local optima
+without exponentially increasing runtime complexity. Figure~\ref{fig-slope-method}
+shows the estimated growth rate for each method as $r$ is increased. On Concatenated Traps,
+Adjacent NKq, Ising Spin Glass, and to a lesser extent Random NKq, none of our enumeration
+methods see an increase in growth complexity with increasing $r$. In all cases using Hyper-Reorder
+has a lower growth rate than the other two methods using the same $r$. On every problem
+except MAX-kSAT, Hyper-Reorder's slowest $r$ is faster than the fastest $r$
+of any other configuration.
+
+Perhaps most striking in Figure~\ref{fig-slope-method} is that on Adjacent NKq, Random NKq,
+and Ising Spin Glass, increasing $r$ \emph{reduces} Hyper-Reorder's growth complexity.
+
+TODO Mention bigger problem sizes.
 

### 2015-09-15 Hyperplane Enumeration/main.tex
Incorporated edits from Bill

@@
 \section{Introduction}
-The ruggedness and high dimensionality of most problem landscapes makes them challenging
+The ruggedness and high dimensionality of most problem landscapes makes them challenging to
 analyze and understand. However, doing so can be helpful in
-quantifying the difficulty of a problem, and how to design algorithms to deal
+quantifying the difficulty of a problem. Furthermore,
+this information can be useful to design search algorithms that specifically deal
 with those difficulties. 
-Similarly, knowing which characteristics favor
+Similarly, knowing problem characteristics that favor
 a particular
@@
 A common way of analyzing landscapes is to
-examine the frequency and distribution of local optima~\cite{boese:1994:bigvalley},
-and how operators allow search to transition between these
+examine both the frequency and distribution of local optima~\cite{boese:1994:bigvalley},
+as well as how different search operators transition between these
 optima~\cite{tomassini:2008:nknetworks,verel:2011:nknetworks,ochoa:2015:crossovernetworks}.
-However, finding all of the local optima is prohibitively time consuming
-even for small problems, with many studies limited to 18-bit
-problems~\cite{tomassini:2008:nknetworks,verel:2011:nknetworks}.
+However, finding all of the local optima of a complex problem is prohibitively time consuming
+even for small problems. Many studies are limited to 18-bit
+problems~\cite{tomassini:2008:nknetworks,verel:2011:nknetworks} due to this time constraint.
 By leveraging recent advancements in Gray-Box optimization that allow for
@@
 portions of the search space to be skipped during enumeration.
-The techniques used apply to the generalized problem class of Mk Landscapes,
+The techniques developed here can be applied to the generalized problem class of Mk Landscapes,
 which contains many real world and benchmark combinatorial problems.
@@
 Combined with limitations on the number of subfunctions and the size of each subfunction's
-subset, Mk Landscapes allow for highly efficient local~\cite{whitley:2012:constant,chicano:2014:ball}
-and global~\cite{goldman:2015:GBO,tintos:2015:partitioncross} search.
+subset, Mk Landscapes can be efficiently searched for both local~\cite{whitley:2012:constant,chicano:2014:ball}
+and global~\cite{goldman:2015:GBO,tintos:2015:partitioncross} optima.
 
@@
 
-The Concatenated Trap problem~\cite{deb:1992:trap} is a composition of $k$-order deceptive
+The Concatenated Trap problem~\cite{deb:1992:trap} is a composition of $k$-order deceptive,
 separable subfunctions. In Mk Landscape terms, $M=N/k$ such that $\forall_i |s_i| = k$ and
-$\forall_{i \neq j} s_i \cap s_j = \emptyset$. Each $f_i$ applies an identical function based
+$\forall_{i \neq j} s_i \cap s_j = \emptyset$. Each $f_i$ applies an identical subfunction based
 on the number of variables set to 1:
@@
 Similar to Adjacent NKq, the $2D\pm J$ subset of Ising Spin Glasses can be polynomially
-solved~\cite{saul:1994:spinglass}\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}.
-In this subset, the graph is defined as a square two-dimensional torus with each edge
+solved~\cite{saul:1994:spinglass}.
+In this subset, the graph is defined as a square two-dimensional toroidal grid with each edge
 weight chosen from $\{-1, 1\}$. Each vertex is assigned a spin from $\{-1, 1\}$ with
@@
 must be evaluated once. Finding the local
-optima then requires each solution to compare its fitness with each
+optima then requires comparing each solution's fitness with each
 of its neighbors, which when properly cached takes \BigO{N} per solution.
@@
 The first major result in Gray-Box optimization was the proof that the list
-of fitness-improving moves can be updated after each flip in local search in \BigO{1}
-time per bit flip~\cite{whitley:2012:constant}. In an Mk Landscape variables
+of fitness-improving moves from a solution can be updated after a bit flip in \BigO{1}
+time~\cite{whitley:2012:constant}. In an Mk Landscape, variables
 can only have a non-linear relationship if they appear together in at least one $s_i$.
-The fitness effect of flipping a bit can therefore only change when it has a non-linear
-relationship with the bit modified by local search. By definition the number of non-linear
+Therefore, the fitness effect of flipping $x_j$ can only change when flipping $x_i$ if
+there is a non-linear relationship between $x_i$ and $x_j$.
+By definition the total number of non-linear
 relationships in an Mk Landscape is linear with $N$, meaning on average the amortized
-number of updates is \BigO{1}.
+number of non-linear relationships per variable is \BigO{1}.
 
@@
 By definition the fitness effect of flipping both $x_i$ and $x_j$ is equal to the sum
-of flipping each independently. As a result if neither individual flip is fitness improving,
+of flipping each independently. As a result, if neither individual flip is fitness improving,
 flipping both together cannot be fitness improving. By examining the graph of non-linear
-interactions between variables this principle can be extended to any collection of $r$
+interactions between variables, this principle can be extended to any collection of $r$
 variables, with the number of useful collections growing at \BigO{N} when $r$ is a small constant.
-As a result the time to update the list of improving moves after a modification is still \BigO{1}.
+As a result, the time to update the list of improving moves after up to $r$ bit-flips is still \BigO{1}.
 
@@
   \caption{Example change of enumeration ordering. The gray variables represent all dependencies
-           for some move $m_i$. By reordering, $m_i$'s lowest $index$ dependency improves from 2 to 4.}
+           for move $m_i$ which flips variable $F$. By reordering, $m_i$'s lowest $index$ dependency improves from 2 to 4.}
   \label{fig-enumerate}
@@
 domain, enumeration would progress as a binary counter, treating $index$ zero (symbol $A$ in the solution) as
-the least significant bit. This ensures that before changing $index$ $i$, all possible settings of $index$
-0 through $i-1$ have been tested. This corresponds to examining the hyperplane with the lowest $i-1$ positions
-variable and all other positions fixed. The Gray-Box domain makes it possible to skip hyperplanes
-that cannot be local optima. In Figure~\ref{fig-enumerate} there exists a move $m_i$ that is a fitness improvement
+the least significant bit. This enumeration ordering ensures that before changing $index$ $i$, all possible settings of $index$
+0 through $i-1$ have been tested. This corresponds to examining the hyperplane where the lowest $i-1$ positions
+can vary and all other positions remain fixed. The Gray-Box domain makes it possible to skip hyperplanes
+that cannot contain local optima. In Figure~\ref{fig-enumerate}, move $m_i$ which flips variable $F$ is a fitness improvement
 when enumeration starts (all variables set to 0). Due to the known relationships between variables,
@@
 based on that move's lowest $index$ dependency. This is the first $index$ that
-can be modified by enumeration to change the fitness effect of making that move.
+can be modified by enumeration that can potentially change the fitness effect of making that move.
 In order to determine how much of the enumeration can be skipped, we must find
@@
 simplification is possible because
-the previous iteration has verified that all moves in higher $index$ bins are not fitness improving, and no action performed during
+the previous iterations have verified that all moves in higher $index$ bins are not fitness improving, and no action performed during
 that iteration can make them fitness improving.
 Furthermore, $index$ is always the location of the least significant 1 in $solution$, meaning
-iteration can continue immediately from the found $index$. This is true by construction
-as initially $solution$ contains all 0s. 1s are always inserted at the value of $index$
-with $index$ only increasing due to carry operations which clear all 1s lower than $index$.
+iteration can continue immediately from the found $index$. This is true by construction.
+Initially $solution$ contains all 0s. Each time a 1 is inserted its position is equal to $index$ and
+$index$ is only increased by carry operations which reset a position to 0 before increasing $index$.
 
@@
 improves $m_i$'s lowest $index$ dependency from 2 to 4. Consider that before reordering $m_i$ allows the hyperplane
-**000000 to be skipped by Algorithm~\ref{alg-enumerate}, while after reordering that becomes ****0000.
-Now whenever $m_i$ is a fitness improvement, Line~\ref{alg-enumerate-bincheck} in Algorithm~\ref{alg-enumerate}
+**000000 to be skipped by Algorithm~\ref{alg-enumerate}, while after reordering the hyperplane that can be skipped is ****0000.
+Now, whenever $m_i$ is a fitness improvement, Line~\ref{alg-enumerate-bincheck} in Algorithm~\ref{alg-enumerate}
 skips 4 times as many solutions.
@@
 least-unmapped dependencies has all of its remaining dependencies mapped to the
-most significant remaining positions. With proper bookkeeping this algorithm requires
+most significant remaining positions. With proper bookkeeping, this algorithm requires
 \BigO{N} time. Assuming that all moves are equally likely to be fitness improving
@@
 Swapping $O_{\hat{i}}$ and $O_{\hat{i}+1}$ cannot change the minimum dependency of any other moves.
-Performing the swap improves the minimum dependency of $O_{\hat{i}+1}$ more than it hurts
+Performing the swap improves the minimum dependency of $O_{\hat{i}+1}$ more than it worsens
 $O_{\hat{i}}$. Therefore, performing this swap makes $O^\star$ skip even more solutions,
@@
 \subsection{$k$-Bound Separable Problem}
-Consider any problem that is composed of non-overlapping
-subfunctions, each using using $k$ or fewer bits. Regardless
+A $k$-bound separable problem any problem that is composed of non-overlapping
+subfunctions, each using using $k$ or fewer bits. Formally this means
+$\forall_i |s_i| \leq k$ and $\forall_{i \neq j} s_i \cap s_j = \emptyset$.
+Regardless
 of how they are initially ordered, reordering ensures that
-variables of a function are always consecutively placed in enumeration.
-Consider a numbering of $f_i$ such that if $i < j$ then all of $f_i$'s
-variables appear before $f_j$'s in enumeration ordering.
+all variables which appear in the same $s_i$ are consecutively placed in enumeration.
+Consider a numbering of $f_i$ such that if $i < j$ then all of $s_i$'s
+variables appear before $s_j$'s in enumeration ordering.
 
-While any $f_i$ contains a fitness-improving move, Algorithm~\ref{alg-enumerate}
-finds the one with the highest $i$ and enumerates its variables until it no longer
-contains an improving move. In total this enumeration requires less than $2^k$
-steps. Each $f_i$ is considered sequentially, meaning the first local optimum
-is found in \BigO{2^kN/k} time.
+Algorithm~\ref{alg-enumerate} finds the $f_i$ with the highest $i$ which contains a
+fitness-improving move and enumerates its variables until it no longer
+contains an improving move. In total, enumerating $f_i$ requires less than $2^k$
+steps. Each $f_i$ is considered sequentially until none contains an improving move,
+meaning the first local optimum is found in \BigO{2^kN/k} time.
 
@@
 in the degenerate case of $k=1$, a $k$-bound separable problem
-is a linear function. As such, when $k=1$ and $c=1$,
+is a linear function. As such, when $k=1$, $c=1$ and
 \BigO{2^kN/k} = \BigO{N}.
@@
 \subsection{Finding Local Optima}
+\label{sec-1-bit-optima}
 Our first experiment is to determine the empirical complexity of each method with respect to problem size.
@@
 and Ising Spin Glass, increasing $r$ \emph{reduces} Hyper-Reorder's growth complexity.
-
-TODO Mention bigger problem sizes.
+Increasing $r$ creates more moves which can be fitness improving. As a result there
+are more hyperplanes which can potentially be detected and eliminated.
+On Concatenated Traps no flip using $1 < r < k$ bits can be fitness improving
+without a 1-bit flip being fitness improving, meaning no additional hyperplanes can be skipped.
+For MAX-kSAT, increasing $r$ likely compounds the high dependency problems discussed in
+Section~\ref{sec-1-bit-optima}.
 

### 2015-09-15 Hyperplane Enumeration/main.tex
Fixed largest Adjacent NKq size

@@
 problems with at most 18 or 30 bits, hyperplane elimination can enumerate the same problems with
-75 bits. When using higher $r$ values this bound can be pushed to 200 bit problems
+77 bits. When using higher $r$ values this bound can be pushed to 200 bit problems
 on average enumerated in under 6 minutes.
@@
 Here we introduce a method for finding all local optima for the same problems using
-up to 75 bits. The cornerstone of this method is the identification
+up to 77 bits. The cornerstone of this method is the identification
 of hyperplanes that cannot contain any local optima, allowing large
@@
 the largest Adjacent NKq problem size which can be enumerated in 4 hours is 34 bits. Given the same time
-Hyper-Reorder can enumerate problems with 75 bits. Using the regression model, we estimate solving
-a 75 bit problem using Gray-Box would require 230 million years. Similarly, Hyper-Reorder's largest
+Hyper-Reorder can enumerate problems with 77 bits. Using the regression model, we estimate solving
+a 77 bit problem using Gray-Box would require 900 million years. Similarly, Hyper-Reorder's largest
 Random NKq size of 69 bits would require Gray-Box 3 million years. These larger sizes make for

### 2015-09-15 Hyperplane Enumeration/main.tex
Filled out more r-bit local optima stuff

@@
 
-\subsection{Examining Local Optima}
+Increasing $r$ also results is a substantial reduction in the
+total number of local optima, as shown in Figure~\ref{fig-length-radius}.
+For every problem except Concatenated Traps, increasing $r$ not only reduces the
+number of local optima, it decreases the growth rate with respect to $N$. As there
+are fewer local optima to enumerate, Algorithm~\ref{alg-enumerate} requires less
+time to enumerate them all. In Concatenated Traps each trap requires exactly $k$ flips in
+order to move between local optima. Therefore, while $r < k$ the number of local optima
+cannot change and once $r \ge k$ there is exactly 1 local optima.
+
+Previous work~\cite{ochoa:2015:crossovernetworks} noted that counter-intuitively Adjacent
+NKq has more local optima than Random NKq for $r=1$ and $N=20$, even though the former is generally
+easier to solve. However, many of these optima are ``weak'' in the sense that they are not also
+2-bit local optima. Here we extend those conclusions for up to $N=59$ and up to $r=4$. Adjacent
+NKq tend to have more 1-bit local optima than Random NKq, most of which disappear as $r$
+increases. Furthermore, the growth rate of local optima between the two landscapes diverges
+as $r$ increases.
+
+MAX-kSAT and Ising Spin Glasses generally appear to contain far more local optima than
+any of the other problems we enumerated. With just $N=49$, Ising Spin Glasses average
+nearly 800 million local optima. For comparison, Adjacent NKq with $N=49$ averages fewer than 300 thousand,
+and with $N=78$ Adjacent NKq still only has fewer than 500 million.
+As with NKq, the polynomially solvable Ising Spin
+Glass problem had more 1-bit local optima than the NP-Hard MAX-kSAT. Unlike with NKq, this
+relationship generally does not change as $r$ increases. Even for $r=4$ the general
+number and growth rate of local optima appears to stay the same for both problems.
 
@@
   \includegraphicsfit{length-radius}
-  \caption{Number of $r$-bit local optima as $N$ increases for each problem.}
+  \caption{Number of $r$-bit local optima for each size of each problem. Note that
+           the number of local optima in Concatenated Traps does not change while $r<k$.}
   \label{fig-length-radius}
@@
 
-TODO Consider discussing neutral networks
+In our original experiment we tested problems up to size $N=100$, which was sufficient to
+cause all three methods using $r=1$ to fail all problems except for Hyper-Reorder on Concatenated
+Traps. However, using higher $r$ values, Hyper-Reorder is able to significantly extend its range.
+On Random NKq, Hyper-Reorder was successful at finding 2-bit local optima for $N=73$. On Adjacent
+NKq using all $r$ tested except 1 were successful on $N=100$. To push Algorithm~\ref{alg-enumerate}
+to the extreme we tested Adjacent NKq with $N=200$ and $r=4$, with all 30 runs successful. On
+average these runs required 5.6 minutes to complete. However, there was significant variance
+in both the runtime and number of local optima. 18 of 30 runs completed in under 1 minute,
+while the slowest run took 1.5 hours. 7 had less than one thousand 4-bit local optima, while 5
+had over one hundred thousand.
 
 \section{Conclusions}
+TODO Consider discussing neutral networks
 

### 2015-09-15 Hyperplane Enumeration/main.tex
Edit pass of the second half

@@
 
+\subsection{All Optimal Orderings are Greedy}
+
 First we show that all optimal orderings of bits are optimal orderings of packed moves.
@@
 
-\subsection{$k$-Bound Separable Problem}
-A $k$-bound separable problem any problem that is composed of non-overlapping
+\subsection{$k$-Bound Separable Problems}
+A $k$-bound separable problem is any problem that is composed of non-overlapping
 subfunctions, each using using $k$ or fewer bits. Formally this means
 $\forall_i |s_i| \leq k$ and $\forall_{i \neq j} s_i \cap s_j = \emptyset$.
-Regardless
-of how they are initially ordered, reordering ensures that
-all variables which appear in the same $s_i$ are consecutively placed in enumeration.
-Consider a numbering of $f_i$ such that if $i < j$ then all of $s_i$'s
+While there is no restriction on how variables are ordered in the problem,
+performing reordering will ensure that
+all variables which appear in the same $s_i$ are consecutively placed during enumeration.
+As such it is possible to create a numbering of $f_i$ such that if $i < j$ then all of $s_i$'s
 variables appear before $s_j$'s in enumeration ordering.
@@
 fitness-improving move and enumerates its variables until it no longer
-contains an improving move. In total, enumerating $f_i$ requires less than $2^k$
+contains an improving move. In total, enumerating $f_i$ requires $2^k$
 steps. Each $f_i$ is considered sequentially until none contains an improving move,
@@
 $T(i) = |l_i|*T(i-1)+2^k$ where $|l_i|$ is the number of ways $f_i$ can be set such
-that it contains no fitness-improving moves. Each $l_i$ exposes a new hyperplane
+that it contains no fitness-improving moves.
+Each setting in $l_i$ exposes a new hyperplane
 that can contain local optima, which causes the recursive call to $T(i-1)$.
+As all $|s_i| \leq k$ and $k$ is a
+constant, $\forall_i |l_i| \leq c \leq 2^k$ meaning that $c$ is a constant.
 Assuming for simplicity that all subfunctions use exactly $k$ bits,
-the time required to find all local optima is $T(N/k-1)$.
-Assuming that $\forall_i |l_i|=c$, $T(N/k-1)=\sum_0^{N/k-1}2^kc^i<2^kc^{N/k}$.
+the time required to find all local optima is
+$T(N/k-1)\leq \sum_0^{N/k-1}2^kc^i<2^kc^{N/k}$.
 The number of local optima in the landscape is $c^{N/k}$ meaning that Algorithm~\ref{alg-enumerate}
@@
 \BigO{2^kN/k} = \BigO{N}.
+
+\subsubsection{Importance of Reordering}
+The order in which variables are enumerated for $k$-bound separable problems
+has a significant impact on Algorithm~\ref{alg-enumerate}'s complexity.
+Instead of using the normal reordering procedure, consider an ordering
+such that if $i < j$, all variables in $s_i$ appear before the second variable
+in $s_j$ and after the first variable in $s_j$.
+
+Each time the second variable in $s_i$ is flipped, all of $s_{i-1}$ must be enumerated.
+This is because only carry operations can result in $index$ being set to the second
+variable in $s_i$. Each time $f_i$ contains a fitness improving move, $index$ is set
+to the first variable in $s_i$, which comes before all variables in $s_{i-1}$.
+As a result the time to solve this ordering becomes $T(i) = 2^{k-1}*T(i-1)+2^k$.
+Therefore, if $c < 2^{k-1}$ this ordering performs much worse than optimal.
 
@@
 	         find all 1-bit local optima of a problem using $N$ bits. Optimal refers to the same model where
-	         $y$ is the number of local optima.}
+	         $y$ is the number of local optima. All $R^2$ values over 0.95 except Hyper on Concatenated Traps
+	         which has 0.898.}
 	\begin{tabular}{|r|c|c|c|c|}
@@
 non-linear relationships per variable. For small problem sizes
+the number of non-linear relationships per variable in
 Random NKq and MAX-kSAT both increase with problem
@@
 A major advantage of Gray-Box enumeration is that it can find $r$-bit local optima
-without exponentially increasing runtime complexity. Figure~\ref{fig-slope-method}
-shows the estimated growth rate for each method as $r$ is increased. On Concatenated Traps,
+with only a constant increase in runtime. Figure~\ref{fig-slope-method}
+shows the estimated growth rate for each method as $r$ increases. On Concatenated Traps,
 Adjacent NKq, Ising Spin Glass, and to a lesser extent Random NKq, none of our enumeration
@@
 easier to solve. However, many of these optima are ``weak'' in the sense that they are not also
-2-bit local optima. Here we extend those conclusions for up to $N=59$ and up to $r=4$. Adjacent
-NKq tend to have more 1-bit local optima than Random NKq, most of which disappear as $r$
+2-bit local optima. Here we extend those observations for up to $N=59$ and up to $r=4$. Adjacent
+NKq tends to have more 1-bit local optima than Random NKq, most of which disappear as $r$
 increases. Furthermore, the growth rate of local optima between the two landscapes diverges
@@
 nearly 800 million local optima. For comparison, Adjacent NKq with $N=49$ averages fewer than 300 thousand,
-and with $N=78$ Adjacent NKq still only has fewer than 500 million.
+and with $N=77$ Adjacent NKq still only has fewer than 500 million.
 As with NKq, the polynomially solvable Ising Spin
 Glass problem had more 1-bit local optima than the NP-Hard MAX-kSAT. Unlike with NKq, this
-relationship generally does not change as $r$ increases. Even for $r=4$ the general
-number and growth rate of local optima appears to stay the same for both problems.
+relationship generally does not change as $r$ increases. However, both do still
+see large decreases in the number and growth rate of local optima. With $N=49$
+the number of 4-bit local optima Ising Spin Glass is on average under 70 thousand,
+a reduction of four orders of magnitude.
 
@@
 
-In our original experiment we tested problems up to size $N=100$, which was sufficient to
+In our original experiments we tested problems up to size $N=100$, which was sufficient to
 cause all three methods using $r=1$ to fail all problems except for Hyper-Reorder on Concatenated
@@
 average these runs required 5.6 minutes to complete. However, there was significant variance
-in both the runtime and number of local optima. 18 of 30 runs completed in under 1 minute,
-while the slowest run took 1.5 hours. 7 had less than one thousand 4-bit local optima, while 5
-had over one hundred thousand.
+in both the runtime and number of local optima. Eighteen runs completed in under 1 minute,
+while the slowest run took 1.5 hours. Three had fewer than 50 4-bit local optima, while five
+had over 100,000.
 

### 2015-09-17 Hyperplane Enumeration/main.tex
Updated to proof form

@@
 \newcommand{\BigO}[1]{$\mathcal{O}{(#1)}$}
+
+% Fancy proofs
+\let\proof\relax
+\let\endproof\relax
+\usepackage{amsthm}
 
@@
 solved~\cite{saul:1994:spinglass}.
-In this subset, the graph is defined as a square two-dimensional toroidal grid with each edge
-weight chosen from $\{-1, 1\}$. Each vertex is assigned a spin from $\{-1, 1\}$ with
+In this subset, the graph is defined as a square two-dimensional grid with
+periodic boundaries such that each edge
+weight is chosen from $\{-1, 1\}$. Each vertex is assigned a spin from $\{-1, 1\}$ with
 the energy in the glass equal to
@@
 \subsection{All Optimal Orderings are Greedy}
+The quality of an enumeration ordering is determined by how many solutions
+in the space can be skipped by Algorithm~\ref{alg-enumerate} using that ordering.
+Each time the loop on Line~\ref{alg-enumerate-bincheck} stops with $index>0$,
+$2^{index}$ solutions have been skipped. For practical purposes we will
+assume that all moves are equally likely to be fitness improving.
 
-First we show that all optimal orderings of bits are optimal orderings of packed moves.
-Consider that lowering a bit in enumeration order can only decrease the number of skipped
-solutions if doing so decreases the minimum dependency of a move. However, doing so may
-allow the minimum dependency of some unrelated move to increase. Therefore when ordering
-bits, all that matters is the order in which the last bit of each move is assigned.
+\begin{lemma}
+There exists an ordering of moves such that assigning all dependencies of
+each move sequentially to the highest remaining positions in the enumeration ordering
+will result in an optimal enumeration ordering.
+\end{lemma}
 
-% Proof
-Moves can have their dependencies fixed in some permutation $O$ of all possible moves,
-such that $O_0$ is the first move to have its dependencies fixed.
-$D_O(i, m)$ is a function that returns how many unfixed dependencies move $m$ has after all
-moves in $O$ before $i$ have been fixed. A greedy solution to this problem is one such
+\begin{proof}
+The number of solutions skipped each time a move is fitness improving depends
+on where it is placed in $move\_bin$. This position is determined by its
+minimum dependency. Therefore, the placement of variables which are not
+the minimum dependency of at least one move cannot affect the number of
+skipped solutions. Consider an ordering of moves which is determined by
+the minimum dependencies of each move in some enumeration ordering. This
+ordering of moves can then be used to create an enumeration ordering which
+skips the same number of solutions as the original enumeration ordering.
+\end{proof}
+
+Moves can have their dependencies assigned in some permutation $O$ of all possible moves,
+such that $O_0$ is the first move to have its dependencies assigned.
+$D_O(i, m)$ is a function that returns how many unassigned dependencies move $m$ has after all
+moves in $O$ before $i$ have been assigned. A greedy solution to this problem is one such
 that $O_i$ is set to be the move that minimizes $D_O(i, m)$.
 
+\begin{theorem}
+All optimal enumeration orderings are greedy orderings of moves.
+\end{theorem}
+
+\begin{proof}
 All non-greedy solutions must have some $i$ such that $D_O(i, O_i) > D_O(i, O_{i+1})$.
@@
 
-Consider an optimal ordering $O^\star$ that is not greedy.
-These criteria mean that $D_{O^\star}(\hat{i}, O_{\hat{i}}) > D_{O^\star}(\hat{i}, O_{\hat{i}+1})$.
-Swapping $O_{\hat{i}}$ and $O_{\hat{i}+1}$ cannot change the minimum dependency of any other moves.
-Performing the swap improves the minimum dependency of $O_{\hat{i}+1}$ more than it worsens
-$O_{\hat{i}}$. Therefore, performing this swap makes $O^\star$ skip even more solutions,
-contradicting the assertion that it was optimal. Therefore there cannot exist an optimal ordering that is not greedy.
+Consider the optimal ordering $O^\star$ that is not greedy and that has the minimum value for $\hat{i}$.
+Swapping $O_{\hat{i}}^\star$ and $O_{\hat{i}+1}^\star$ cannot change the minimum dependency of any other moves.
+Performing the swap causes $O_{\hat{i}}^\star$ to have the minimum dependency $O_{\hat{i}+1}^\star$ had before
+the swap, while $O_{\hat{i}+1}^\star$'s minimum dependency is raised by at least
+$D_{O^\star}(\hat{i}, O_{\hat{i}}^\star) - D_{O^\star}(\hat{i}, O_{\hat{i}+1}^\star)$. This value
+cannot be negative meaning that we have now constructed a solution which either skips more
+solutions than $O^\star$ or has a lower $\hat{i}$ than $O^\star$, contradicting our assertions.
+Therefore there cannot exist an optimal ordering that is not greedy.
+\end{proof}
+
+\begin{theorem}
+Not all greedy orderings are optimal enumeration orderings.
+\end{theorem}
+
+\begin{proof}
+Consider a problem with move dependencies $m_0 = \{A, B\}$, $m_1 = \{C, D\}$, and $m_2 = \{C, D, E\}$.
+There are two greedy orderings, $[m_0, m_1, m_2]$ and $[s_1, s_2, s_0]$. The former
+skips $2^3+2^1$ solutions, while the latter skips $2^3+2^2$. This is because $m_1$ and
+$m_2$ overlap. As a result, even though both a greedy, only the second is optimal.
+\end{proof}
+
+We have not been able to determine a method for finding the optimal move ordering efficiently.
+Therefore we will use the greedy method as it is probabilistically optimal.
 
@@
 
+\begin{theorem}
+Algorithm~\ref{alg-enumerate} finds all local optima of any linear function
+in \BigO{N} time.
+\end{theorem}
+
+\begin{proof}
 When Algorithm~\ref{alg-enumerate} is applied to a linear
@@
 Initially $index$ is set to $N-1$ and is decreased by at least 1 in any step after the first.
+Each step will flip a bit which disagreed with the global optimum.
 Before finding the global optimum no carry operations can occur, meaning the
 cost of each iteration is equal to the amount by which $index$ is decreased.
-Each step will flip a bit which disagreed with the global optimum.
 Therefore, Algorithm~\ref{alg-enumerate} requires \BigO{N} time to reach
@@
 is then spent adding a 1 to $index$ 0, which in the worst case requires \BigO{N}
-carry operations. For all future iterations $index$ is a fitness-improving move
-and is currently set to 1. In these iterations at least 1 carry operation must
+carry operations. For all future iterations $index$ is the position of the highest fitness-improving move
+which is currently set to 1. In these iterations at least 1 carry operation must
 occur, and $index$ cannot be decreased. Iteration ends when $index$ exceeds $N$
@@
 which is optimal.
+\end{proof}
 
@@
 $\forall_i |s_i| \leq k$ and $\forall_{i \neq j} s_i \cap s_j = \emptyset$.
+
+\begin{theorem}
+Algorithm~\ref{alg-enumerate} finds all \BigO{c^{M}} local optima of any $k$-bound separable function
+in \BigO{2^kc^{M}} time.
+\end{theorem}
+
+\begin{proof}
 While there is no restriction on how variables are ordered in the problem,
@@
 steps. Each $f_i$ is considered sequentially until none contains an improving move,
-meaning the first local optimum is found in \BigO{2^kN/k} time.
+meaning the first local optimum is found in \BigO{2^kM} time.
 
 Once the first local optimum is found, Algorithm~\ref{alg-enumerate} proceeds
-by finding all local optimum in larger and larger hyperplanes. Initially $f_0$
+by finding all local optima in larger and larger hyperplanes. Initially $f_0$
 is enumerated to find all local optima in the hyperplane with all $f_i, i>0$
@@
 constant, $\forall_i |l_i| \leq c \leq 2^k$ meaning that $c$ is a constant.
-Assuming for simplicity that all subfunctions use exactly $k$ bits,
-the time required to find all local optima is
-$T(N/k-1)\leq \sum_0^{N/k-1}2^kc^i<2^kc^{N/k}$.
-The number of local optima in the landscape is $c^{N/k}$ meaning that Algorithm~\ref{alg-enumerate}
+The time required to find all local optima is therefore
+$T(M-1)\leq \sum_0^{M-1}2^kc^i<2^kc^{M}$.
+The number of local optima in the landscape is \BigO{c^{M}} meaning that Algorithm~\ref{alg-enumerate}
 is within a constant factor ($2^k$) of optimal.
+\end{proof}
 
@@
 As $c=1$ in this case, the cost of finding the single global optimum
-only requires the \BigO{2^kN/k} time to find the first local optimum, which is
+only requires the \BigO{2^kM} time to find the first local optimum, which is
 optimal.
@@
 Concatenated Traps is a commonly used $k$-bound separable problem
-with $c=2$.  Algorithm~\ref{alg-enumerate} requires $2^k2^{N/k}$ time
+with $c=2$ and $M=N/k$.  Algorithm~\ref{alg-enumerate} requires $2^k2^{N/k}$ time
 to find all $2^{N/k}$ local optima in this problem. Note also that
@@
 is a linear function. As such, when $k=1$, $c=1$ and
-\BigO{2^kN/k} = \BigO{N}.
+\BigO{2^kM}~=~\BigO{N}.
 

### 2015-09-17 Hyperplane Enumeration/main.tex
Changed from figure* to figure
Conclusion and code URL

@@
 \section{Hyperplane Elimination}
-\begin{figure*}
+\begin{figure}
   \centering
@@
   \label{fig-enumerate}
-\end{figure*}
+\end{figure}
 
@@
 we report are for successful combinations. For timing purposes local optima were counted but not
-recorded.
+recorded. All of our code, data, and statistical analysis can be downloaded from our
+website.~\footnote{\url{https://name.of.site/removed/to/be/anonymous/}}
 
@@
 
-\begin{figure*}
+\begin{figure}
   \centering
@@
   \label{fig-boxplot-method}
-\end{figure*}
+\end{figure}
 
@@
   \label{fig-length-method}
-\end{figure*}
+\end{figure}
 
@@
   \label{fig-slope-method}
-\end{figure*}
+\end{figure}
 
@@
   \label{fig-length-radius}
-\end{figure*}
+\end{figure}
 
@@
 
-\section{Conclusions}
-TODO Consider discussing neutral networks
+\section{Conclusions and Future Work}
+By expressing a function as an Mk Landscapes, we have developed a method to find all
+local optima of that function very efficiently. The structure imposed by
+an Mk Landscape allows for the identification of hyperplanes in the search space
+which cannot contain any local optima. Combined with a greedy method for
+changing the order of how the space is enumerated, we were able to prove
+this method's runtime complexity scales with the number of local optima, unlike
+previous methods which scale with the number of possible solutions.
+When applied to NKq Landscapes and Ising Spin Glasses this allowed for a dramatic
+increase in the size of problems which can be enumerated. We believe that
+these larger problem sizes will be more representative of the types
+of search spaces where optimization is performed.
 
-\subsubsection*{Acknowledgments.} TODO Acknowledge BEACON.
+We show how this hyperplane elimination can be extended to find only solutions
+which cannot be improved by flipping $r$ or fewer bits. For some problems
+this counterintuitively reduces search times, and on Adjacent NKq allowed
+us to find all 4-bit local optima of a 200 bit problem.
+
+The obvious next step is to use the results of hyperplane elimination
+in conjunction with landscape analysis
+techniques~\cite{tomassini:2008:nknetworks,verel:2011:nknetworks,ochoa:2015:crossovernetworks}.
+We are especially interested to see how landscape properties change for larger $N$
+and $r$. As to the algorithm itself, we suspect there may be room for improvement
+in enumeration ordering. While we are able to prove that all optimal orderings
+are greedy, the reverse is not true due to how ties are broken. Therefore more
+gains may be possible by improving this ordering. Furthermore, we assumed
+that all moves are equally likely to be fitness improving. It is likely possible
+to determine the true probability of a move being fitness improving, resulting
+in potentially better ordering.
+
+\subsubsection*{Acknowledgments.}
+This material is based in part upon work supported by the National Science Foundation
+under Cooperative Agreement No. DBI-0939454. Any opinions, findings, and conclusions
+or recommendations expressed in this material are those of the author(s) and do not
+necessarily reflect the views of the National Science Foundation.
 

### 2015-09-18 Hyperplane Enumeration/main.tex
Final edit before sending to Bill

@@
 Furthermore, we are able to extend this method to finding only solutions
-which cannot be improving by flipping $r$ or fewer bits. In some cases
+which cannot be improved by flipping $r$ or fewer bits. In some cases
 increasing $r$ actually reduces total runtime by drastically reducing the
@@
 quantifying the difficulty of a problem. Furthermore,
-this information can be useful to design search algorithms that specifically deal
+this information can be used to design search algorithms that specifically deal
 with those difficulties. 
@@
 However, finding all of the local optima of a complex problem is prohibitively time consuming
-even for small problems. Many studies are limited to 18-bit
+even for small problems. Many studies have been limited to 18-bit
 problems~\cite{tomassini:2008:nknetworks,verel:2011:nknetworks} due to this time constraint.
@@
 The total number of subfunctions $M$ is constrained to grow at \BigO{N}, and $k$
-is constant with repect to $N$.
+is constant with respect to $N$.
 
@@
 NKq landscapes specify a class of randomly generated problem instances using 3 paramters:
-(1) the number of problem variables $N$ (2) the amount of variable epistasis $K$ where $K=k-1$
+(1) the number of problem variables $N$ (2) the amount of variable epistasis $K$ where $k=K+1$
 and (3) the number of unique fitness values $q$. From these parameters a landscape is generated
@@
 \end{equation}
-where $e_{ij}$ is the weight of the edge vertex $i$ to vertex $j$. In Mk landscape terms
+where $e_{ij}$ is the weight of the edge connecting vertex $i$ to vertex $j$. In Mk landscape terms
 this type of spin glass has $M=2N$ and $k=2$.
@@
 \section{Gray-Box Enumeration of Mk Landscapes}
+\label{sec-gray-box}
 When considered as a black box, the process of finding all local optima
-in a landscape requires \BigO{N2^N} time. This is because evaluating
-a solution requires $N$ time, and all $2^N$ solutions in the landscape
-must be evaluated once. Finding the local
-optima then requires comparing each solution's fitness with each
-of its neighbors, which when properly cached takes \BigO{N} per solution.
+in a landscape requires $\Omega(N2^N)$ time. This is because each
+of the $2^N$ solutions must be compared with each of its $N$ neighbors.
 Extending this method to look for solutions that cannot be improved by
-flipping $r$ or fewer bits requires \BigO{N^r2^N} time. However,
+flipping $r$ or fewer bits requires $\Omega(N^r2^N)$ time. However,
 by exploiting Gray-Box optimization methods, previous work~\cite{ochoa:2015:crossovernetworks}
@@
 of fitness-improving moves from a solution can be updated after a bit flip in \BigO{1}
-time~\cite{whitley:2012:constant}. In an Mk Landscape, variables
+time~\cite{whitley:2012:constant}.
+The fitness effect of flipping $x_j$ changes after flipping $x_i$ only if
+there is a non-linear relationship between $x_i$ and $x_j$.
+In an Mk Landscape, variables
 can only have a non-linear relationship if they appear together in at least one $s_i$.
-Therefore, the fitness effect of flipping $x_j$ can only change when flipping $x_i$ if
-there is a non-linear relationship between $x_i$ and $x_j$.
-By definition the total number of non-linear
+Therefore by definition the total number of non-linear
 relationships in an Mk Landscape is linear with $N$, meaning on average the amortized
@@
 
-This hyperplane skipping extends to finding only $r$-bit local optima for all Mk landscapes.
-By increasing the number of potential fitness-improving moves, the number of hyperplanes that
-can be eliminated increases. However, the cost trade-off of increasing $r$ is unclear.
-While more solutions can be skipped, the time required to update $move\_bin$ each iteration
-also increases. Therefore, the runtime effect of increasing $r$
-depends on the specific problem.
+Extending Algorithm~\ref{alg-enumerate} to search for $r$-bit optima only requires adding the
+necessary $r$-bit moves to $move\_bin$. As discussed in Section~\ref{sec-gray-box}, we know the number of these moves
+is \BigO{N} an can be kept updated in \BigO{1} time per flip. Therefore, for small $r$, the cost
+of finding $r$-bit local optima is no more than a constant slower than finding 1-bit local optima.
+In fact, we will provide evidence in Section~\ref{sec-r-bit} that increasing $r$ can actually
+decrease runtime.
 
@@
 The quality of an enumeration ordering is determined by how many solutions
-in the space can be skipped by Algorithm~\ref{alg-enumerate} using that ordering.
+Algorithm~\ref{alg-enumerate} can skip when using that ordering.
 Each time the loop on Line~\ref{alg-enumerate-bincheck} stops with $index>0$,
@@
 \begin{lemma}
-There exists an ordering of moves such that assigning all dependencies of
-each move sequentially to the highest remaining positions in the enumeration ordering
-will result in an optimal enumeration ordering.
+All optimal enumeration orderings are orderings of moves, such that
+all remaining dependencies of each move are sequentially assigned the
+highest remaining position in the enumeration ordering.
 \end{lemma}
@@
 \begin{proof}
-The number of solutions skipped each time a move is fitness improving depends
-on where it is placed in $move\_bin$. This position is determined by its
-minimum dependency. Therefore, the placement of variables which are not
-the minimum dependency of at least one move cannot affect the number of
-skipped solutions. Consider an ordering of moves which is determined by
-the minimum dependencies of each move in some enumeration ordering. This
-ordering of moves can then be used to create an enumeration ordering which
-skips the same number of solutions as the original enumeration ordering.
+Any enumeration ordering which cannot be described by an ordering of moves
+must contain the following property. There must exist a position $j$
+such that $x_j$ is not a dependency of any move with
+a minimum dependency of $i$ or higher where $i$ is also the minimum dependency of a move $m$.
+In this situation swapping $x_i$ and $x_j$ in enumeration ordering will raise $m$'s
+minimum dependency by at least one, but it will not lower the minimum dependency of any
+move. Therefore any ordering which contains this property cannot be optimal.
 \end{proof}
@@
 Consider a problem with move dependencies $m_0 = \{A, B\}$, $m_1 = \{C, D\}$, and $m_2 = \{C, D, E\}$.
-There are two greedy orderings, $[m_0, m_1, m_2]$ and $[s_1, s_2, s_0]$. The former
+There are two greedy orderings, $[m_0, m_1, m_2]$ and $[m_1, m_2, m_0]$. The former
 skips $2^3+2^1$ solutions, while the latter skips $2^3+2^2$. This is because $m_1$ and
@@
 
-Each time the second variable in $s_i$ is flipped, all of $s_{i-1}$ must be enumerated.
+In order to flip the second variable in $s_i$, all of $s_{i-1}$ must be enumerated.
 This is because only carry operations can result in $index$ being set to the second
@@
 recorded. All of our code, data, and statistical analysis can be downloaded from our
-website.~\footnote{\url{https://name.of.site/removed/to/be/anonymous/}}
+website.\footnote{\url{https://name.of.site/removed/to/be/anonymous/}}
 
@@
 method on each problem. This value is the slope of each line shown in Figure~\ref{fig-length-method}.
-As expected Gray-Box scales at almost exactly $2^N$. Using Hyperplane elimination without
+As expected Gray-Box scales at almost exactly $2^N$. Using hyperplane elimination without
 reordering reduces this complexity, with reordering making further improvements. In the final
@@
 Hyper-Reorder can enumerate problems with 77 bits. Using the regression model, we estimate solving
-a 77 bit problem using Gray-Box would require 900 million years. Similarly, Hyper-Reorder's largest
-Random NKq size of 69 bits would require Gray-Box 3 million years. These larger sizes make for
+a 77 bit problem using Gray-Box would require 900~million years. Similarly, Hyper-Reorder's largest
+Random NKq size of 69 bits would require Gray-Box 3~million years. These larger sizes make for
 much better approximations of landscapes interesting to search methods.
@@
 \subsection{Finding $r$-Bit Local Optima}
+\label{sec-r-bit}
 A major advantage of Gray-Box enumeration is that it can find $r$-bit local optima
@@
 
-Increasing $r$ also results is a substantial reduction in the
+Increasing $r$ also results in a substantial reduction in the
 total number of local optima, as shown in Figure~\ref{fig-length-radius}.
@@
 which cannot contain any local optima. Combined with a greedy method for
-changing the order of how the space is enumerated, we were able to prove
+changing the order of how the space is enumerated, we were able to show
 this method's runtime complexity scales with the number of local optima, unlike
@@
 this counterintuitively reduces search times, and on Adjacent NKq allowed
-us to find all 4-bit local optima of a 200 bit problem.
+us to find all 4-bit local optima of 200 bit problems.
 

### 2015-09-22 Hyperplane Enumeration/main.tex
Updated abstract from writing group

@@
 \begin{abstract}
-A common tool for understanding combinatorial problem landscapes
-is to examine the properties of its local optima. Unfortunately,
-exhaustive methods for finding local optima are exceptionally time
-consuming and are therefore limited to very small problem sizes.
+Examining the properties of local optima is a common method for understanding
+combinatorial-problem landscapes.
+Unfortunately,
+exhaustive methods for finding local optima are limited to very small problem sizes.
 We propose a method for exploiting generalized problem structure
-to significantly speed up this process by skipping hyperplanes
-which cannot contain local optima. In most of our experiments
+to skip hyperplanes
+that cannot contain local optima. In most of our experiments,
 this hyperplane elimination allows enumeration time to scale with
 the total number of local optima instead of the total landscape size.
-We prove optimality for linear and $k$-bounded problems, and provide
+We prove optimality for linear and $k$-bounded problems, and we provide
 empirical evidence of optimality on NKq Landscapes and Ising Spin Glasses.
-Furthermore, we are able to extend this method to finding only solutions
-which cannot be improved by flipping $r$ or fewer bits. In some cases
-increasing $r$ actually reduces total runtime by drastically reducing the
+Furthermore, we are able to refine this method to find only solutions
+that cannot be improved by flipping $r$ or fewer bits. Counterintuitively,
+increasing $r$ can reduce total runtime by drastically reducing the
 number of local optima. While previous methods were limited to
-problems with at most 18 or 30 bits, hyperplane elimination can enumerate the same problems with
-77 bits. When using higher $r$ values this bound can be pushed to 200 bit problems
-on average enumerated in under 6 minutes.
+landscapes with at most $2^{30}$ binary strings, hyperplane elimination can enumerate the same problems with
+$2^{77}$ binary strings, and find all 4-bit local optima to problems with $2^{200}$ binary strings.
+We believe this increase in size will help further our understanding of complex real-world landscapes.
 

### 2015-09-24 Hyperplane Enumeration/main.tex
Incorporated Bill's edits

@@
 exhaustive methods for finding local optima are limited to very small problem sizes.
-We propose a method for exploiting generalized problem structure
+We propose a method for exploiting problem structure
 to skip hyperplanes
@@
 this hyperplane elimination allows enumeration time to scale with
-the total number of local optima instead of the total landscape size.
+the total number of local optima instead of with the total landscape size.
 We prove optimality for linear and $k$-bounded problems, and we provide
@@
 Each subfunction uses a small subset of variables from the original function's input.
-Combined with limitations on the number of subfunctions and the size of each subfunction's
+Combined with limits on the number of subfunctions and the size of each subfunction's
 subset, Mk Landscapes can be efficiently searched for both local~\cite{whitley:2012:constant,chicano:2014:ball}
@@
      \begin{array}{rl}
-       k-1-t, &  t<k\\
-       k,   &  t = k
+       k-1-t,~ &  t<k\\
+       k,~   &  t = k
      \end{array}
@@
 
-Our final examined problem is randomly generated maximum satisfiability or MAX-kSAT.
+Our final problem is randomly generated maximum satisfiability or MAX-kSAT.
 This version of the canonical NP-Complete boolean satisfiability problem is formulated
@@
 Instead of requiring \BigO{N} time to evaluate each solution and to check if it
-is a local optima, only \BigO{1} time is needed. Consider an enumeration ordering of the landscape
+is a local optima, only \BigO{1} time is needed. Consider an enumeration of the landscape
 that uses gray-codes, meaning that each transition between solutions requires exactly 1 bit flip.
@@
   \includegraphicsfit{Enumerate}
-  \caption{Example change of enumeration ordering. The gray variables represent all dependencies
+  \caption{Example change of ordering. The gray variables represent all dependencies
            for move $m_i$ which flips variable $F$. By reordering, $m_i$'s lowest $index$ dependency improves from 2 to 4.}
@@
 domain, enumeration would progress as a binary counter, treating $index$ zero (symbol $A$ in the solution) as
-the least significant bit. This enumeration ordering ensures that before changing $index$ $i$, all possible settings of $index$
+the least significant bit. This ordering ensures that before changing $index$ $i$, all possible settings of $index$
 0 through $i-1$ have been tested. This corresponds to examining the hyperplane where the lowest $i-1$ positions
@@
 as shown in Algorithm~\ref{alg-enumerate}.
-Before starting, each fitness-improving move is put into a table $move\_bin$
+Initially all variables are set to 0 and
+each fitness-improving move is put into a table $move\_bin$
 based on that move's lowest $index$ dependency. This is the first $index$ that
@@
 that iteration can make them fitness improving.
-Furthermore, $index$ is always the location of the least significant 1 in $solution$, meaning
+Furthermore, $index$ is always the location of the least significant 1 bit in $solution$, meaning
 iteration can continue immediately from the found $index$. This is true by construction.
@@
 
-\section{Improved Enumeration Ordering}
+\section{Reordering Variables}
 \label{sec-reorder}
@@
 \subsection{All Optimal Orderings are Greedy}
-The quality of an enumeration ordering is determined by how many solutions
+The quality of an ordering is determined by how many solutions
 Algorithm~\ref{alg-enumerate} can skip when using that ordering.
@@
 \begin{lemma}
-All optimal enumeration orderings are orderings of moves, such that
+All optimal orderings of variables are orderings of moves, such that
 all remaining dependencies of each move are sequentially assigned the
@@
 \begin{proof}
-Any enumeration ordering which cannot be described by an ordering of moves
+Any ordering of variables which cannot be described by an ordering of moves
 must contain the following property. There must exist a position $j$
@@
 a minimum dependency of $i$ or higher where $i$ is also the minimum dependency of a move $m$.
-In this situation swapping $x_i$ and $x_j$ in enumeration ordering will raise $m$'s
+In this situation swapping the order of $x_i$ and $x_j$ will raise $m$'s
 minimum dependency by at least one, but it will not lower the minimum dependency of any
@@
 \begin{theorem}
-All optimal enumeration orderings are greedy orderings of moves.
+All optimal orderings of variables are greedy orderings of moves.
 \end{theorem}
@@
 \begin{theorem}
-Not all greedy orderings are optimal enumeration orderings.
+Not all greedy orderings are optimal orderings.
 \end{theorem}
@@
 
-We have not been able to determine a method for finding the optimal move ordering efficiently.
-Therefore we will use the greedy method as it is probabilistically optimal.
+We believe finding the optimal ordering of moves is NP-Hard and therefore rely on
+the greedy method as it is probabilistically optimal.
 
@@
 performing reordering will ensure that
-all variables which appear in the same $s_i$ are consecutively placed during enumeration.
+all variables which appear in the same $s_i$ are consecutive.
 As such it is possible to create a numbering of $f_i$ such that if $i < j$ then all of $s_i$'s
@@
 by finding all local optima in larger and larger hyperplanes. Initially $f_0$
-is enumerated to find all local optima in the hyperplane with all $f_i, i>0$
+is enumerated to find all local optima in the hyperplane where all $f_i, i>0$ remain
 fixed. This process ends when a carry operation modifies a bit of $f_1$. At
 this point $f_1$ is enumerated, such that each time $f_1$ contains no improving moves
-$f_0$ is enumerated again. This finds all local optima in the hyperplane with $f_i, i>1$
+$f_0$ is enumerated again. This finds all local optima in the hyperplane where $f_i, i>1$ remain
 fixed. We represent the time required to find all local optima in the hyperplane
@@
 \subsubsection{Importance of Reordering}
-The order in which variables are enumerated for $k$-bound separable problems
+The order of variables for $k$-bound separable problems
 has a significant impact on Algorithm~\ref{alg-enumerate}'s complexity.
@@
 
-Each algorithm was given a maximum of four hours on a 2.5GHz Intel Xeon E5-2670v2 processor to find
-all $r$-bit local optima. Any run which failed was rerun once to prevent cluster issues from introducing
+Each algorithm was given a maximum of four hours to find all $r$-bit local optima.
+Runs were distributed across a cluster using 2.5GHz Intel Xeon E5-2670v2 processors.
+Any run which failed was rerun once to prevent cluster issues from introducing
 noise. For the same reason any run which was over ten times slower than the next slowest run of the
@@
 Our first experiment is to determine the empirical complexity of each method with respect to problem size.
-Of the 681 configurations where at least one method was successful, Hyper-Reorder had the lowest
-mean time to find all local optima all but 2 times. In both of those cases Hyper-Reorder was within a tenth of a second
+We tested 932 task configurations (problem, $N$, $r$), with 681 completed by at least one method.
+Of those, Hyper-Reorder had the lowest
+mean time to find all local optima in all but 2. In both of those cases Hyper-Reorder was within a tenth of a second
 of the best method.
-Figure~\ref{fig-boxplot-method} shows the variance in runtime for the largest problem size all three methods
+Figure~\ref{fig-boxplot-method} shows the variance in runtime for the largest problem size where all three methods
 were successful. In all cases it is clear that Hyper-Reorder outperforms the other methods.
@@
 More critically, Figure~\ref{fig-length-method} shows how each method scales with problem size.
-Not only is Hyper-Reorder out performing the alternatives on all sizes, the rate at which it
+Not only does Hyper-Reorder outperform the alternatives on all sizes, the rate at which it
 scales to problem complexity is better. Table~\ref{table-scaling} shows the model fit for each
 method on each problem. This value is the slope of each line shown in Figure~\ref{fig-length-method}.
-As expected Gray-Box scales at almost exactly $2^N$. Using hyperplane elimination without
-reordering reduces this complexity, with reordering making further improvements. In the final
+As expected, Gray-Box scales at almost exactly $2^N$. Using hyperplane elimination without
+reordering reduces this complexity, and reordering makes further improvements. In the final
 column we show the growth rate of the number of local optima. It is impossible for any method
-which enumerates all local optima to grow slower than this rate. On Concatenated Traps, Adjacent
-NKq, and Ising Spin Glass, Hyper-Reorder has an estimated slope better than optimal. This is
+which enumerates all local optima to grow slower than this rate.
+However, Hyper-Reorder has an estimated slope better than optimal on Concatenated Traps, Adjacent
+NKq, and Ising Spin Glass.
+This is
 due to lower order terms dominating runtime for small problem sizes, as can be see in
@@
 This reduction in growth complexity translates into a significant increase in the size of problems
-which can be enumerated. For instance, using the previous best enumeration methods~\cite{ochoa:2015:crossovernetworks}
-the largest Adjacent NKq problem size which can be enumerated in 4 hours is 34 bits. Given the same time
+which can be enumerated. For instance, using the previous best enumeration methods~\cite{ochoa:2015:crossovernetworks},
+the largest Adjacent NKq problem size which can be enumerated in 4 hours is 34 bits. Given the same amount of time,
 Hyper-Reorder can enumerate problems with 77 bits. Using the regression model, we estimate solving
 a 77 bit problem using Gray-Box would require 900~million years. Similarly, Hyper-Reorder's largest
-Random NKq size of 69 bits would require Gray-Box 3~million years. These larger sizes make for
-much better approximations of landscapes interesting to search methods.
+Random NKq size of 69 bits would require Gray-Box 3~million years. We believe these larger problems
+are more likely to share characteristics with those where search algorithms are actually applied.
 
@@
 are more hyperplanes which can potentially be detected and eliminated.
-On Concatenated Traps no flip using $1 < r < k$ bits can be fitness improving
+On Concatenated Traps, no flip using $1 < r < k$ bits can be fitness improving
 without a 1-bit flip being fitness improving, meaning no additional hyperplanes can be skipped.
@@
 
-Previous work~\cite{ochoa:2015:crossovernetworks} noted that counter-intuitively Adjacent
+Previous work~\cite{ochoa:2015:crossovernetworks} noted that, counter-intuitively, Adjacent
 NKq has more local optima than Random NKq for $r=1$ and $N=20$, even though the former is generally
@@
 cause all three methods using $r=1$ to fail all problems except for Hyper-Reorder on Concatenated
-Traps. However, using higher $r$ values, Hyper-Reorder is able to significantly extend its range.
+Traps. For Gray-Box and Hyper, increasing $r$ only decreased the largest size they could
+enumerate.
+However, using higher $r$ values, Hyper-Reorder is able to significantly extend its range.
 On Random NKq, Hyper-Reorder was successful at finding 2-bit local optima for $N=73$. On Adjacent
-NKq using all $r$ tested except 1 were successful on $N=100$. To push Algorithm~\ref{alg-enumerate}
-to the extreme we tested Adjacent NKq with $N=200$ and $r=4$, with all 30 runs successful. On
+NKq for all $r$ tested except 1, Hyper-Reorder was successful on $N=100$. To push Algorithm~\ref{alg-enumerate}
+to the extreme, we tested Adjacent NKq with $N=200$ and $r=4$, with all 30 runs successful. On
 average these runs required 5.6 minutes to complete. However, there was significant variance
@@
 an Mk Landscape allows for the identification of hyperplanes in the search space
-which cannot contain any local optima. Combined with a greedy method for
-changing the order of how the space is enumerated, we were able to show
+which cannot contain any local optima and can therefore be eliminated from enumeration.
+Combined with a greedy method to
+change the ordering of variables, we were able to show
 this method's runtime complexity scales with the number of local optima, unlike
 previous methods which scale with the number of possible solutions.
-When applied to NKq Landscapes and Ising Spin Glasses this allowed for a dramatic
+When applied to NKq Landscapes and Ising Spin Glasses, this allowed for a dramatic
 increase in the size of problems which can be enumerated. We believe that
@@
 
-We show how this hyperplane elimination can be extended to find only solutions
+We also showed how this hyperplane elimination can be refined to find only solutions
 which cannot be improved by flipping $r$ or fewer bits. For some problems

### 2015-09-24 Hyperplane Enumeration/main.tex
Now under the page limit

@@
 to skip hyperplanes
-that cannot contain local optima. In most of our experiments,
-this hyperplane elimination allows enumeration time to scale with
-the total number of local optima instead of with the total landscape size.
-We prove optimality for linear and $k$-bounded problems, and we provide
+that cannot contain local optima, allowing runtime to scale with
+the number of local optima instead of with the landscape size.
+We prove optimality for linear and $k$-bounded separable problems, and we provide
 empirical evidence of optimality on NKq Landscapes and Ising Spin Glasses.
-Furthermore, we are able to refine this method to find only solutions
-that cannot be improved by flipping $r$ or fewer bits. Counterintuitively,
-increasing $r$ can reduce total runtime by drastically reducing the
-number of local optima. While previous methods were limited to
-landscapes with at most $2^{30}$ binary strings, hyperplane elimination can enumerate the same problems with
+We further refine this method to find only solutions
+that cannot be improved by flipping $r$ or fewer bits, which counterintuitively
+can reduce total runtime. While previous methods were limited to
+landscapes with at most $2^{34}$ binary strings, hyperplane elimination can enumerate the same problems with
 $2^{77}$ binary strings, and find all 4-bit local optima to problems with $2^{200}$ binary strings.
@@
 the structure of Adjacent NKq allows for a polynomial time dynamic programming solution~\cite{wright:2000:solvingnk}.
-Adjacent NKq is therefore often used to understand a search algorithm's ability to reach the known global
-optimum.
+Adjacent NKq is therefore often used to approximate a search algorithm's ability to
+solve Random NKq.
 
@@
 time~\cite{whitley:2012:constant}.
-The fitness effect of flipping $x_j$ changes after flipping $x_i$ only if
+The fitness effect of flipping $x_j$ only changes after flipping $x_i$ if
 there is a non-linear relationship between $x_i$ and $x_j$.
@@
 
-This result was extended by~\cite{chicano:2014:ball} to include local search for local
-optima that cannot be improved using $r$ or fewer bits with no increase in asymptotic costs.
+This result was extended, with no increase in asymptotic costs,
+to include search for
+solutions that cannot be improved by flipping $r$ or fewer bits~\cite{chicano:2014:ball}.
 Consider two variables $x_i$ and $x_j$ that do not appear in the same $s_i$.
@@
   \centering
-  \includegraphicsfit{Enumerate}
+  \includegraphics[width=.75\columnwidth,height=\textheight,keepaspectratio]{Enumerate}
+  %\includegraphicsfit{Enumerate}
   \caption{Example change of ordering. The gray variables represent all dependencies
@@
 Consider the representation presented in the top of Figure~\ref{fig-enumerate}. In a Black-Box
-domain, enumeration would progress as a binary counter, treating $index$ zero (symbol $A$ in the solution) as
+domain, enumeration would progress as a binary counter, treating $index$ 0 (symbol $A$ in the solution) as
 the least significant bit. This ordering ensures that before changing $index$ $i$, all possible settings of $index$
@@
     \Require $solution \leftarrow \{0\}^N$
-    \Require $move\_bin$ (List of sets of improving moves in $solution$)
+    \Require $move\_bin$ (How many improving moves have each minimum dependency)
     \Require \textproc{MakeMove} (Function that flips $index$ in $solution$ and updates $move\_bin$)
@@
     \While{$index < N$}
-      \While{$index > 0$ \textbf{and} $move\_bin[index] = \emptyset$}\label{alg-enumerate-bincheck}
+      \While{$index > 0$ \textbf{and} $move\_bin[index] = 0$}\label{alg-enumerate-bincheck}
         \State $index \leftarrow index-1$
@@
 **000000 to be skipped by Algorithm~\ref{alg-enumerate}, while after reordering the hyperplane that can be skipped is ****0000.
-Now, whenever $m_i$ is a fitness improvement, Line~\ref{alg-enumerate-bincheck} in Algorithm~\ref{alg-enumerate}
-skips 4 times as many solutions.
+Now, whenever $m_i$ is a fitness improvement, 4 times as many solutions are skipped.
 
@@
 most significant remaining positions. With proper bookkeeping, this algorithm requires
-\BigO{N} time. Assuming that all moves are equally likely to be fitness improving
-during enumeration, we argue below that the optimal bit ordering for skipping solutions is a greedy
-ordering.
+\BigO{N} time. We argue below that the optimal bit ordering for skipping solutions
+is a greedy ordering.
 
-\subsection{All Optimal Orderings are Greedy}
 The quality of an ordering is determined by how many solutions
@@
 When Algorithm~\ref{alg-enumerate} is applied to a linear
-function, each index of $move\_bin$ can contain at most 1
-move, and a move is fitness improving whenever it disagrees with the global optimum.
+function, an index $i$ of $move\_bin$ is non-zero if and only if
+$x_i$ disagrees with the global optimum.
 Initially $index$ is set to $N-1$ and is decreased by at least 1 in any step after the first.
@@
 not perform hyperplane elimination, with Hyper and Hyper-Reorder being hyperplane elimination
-without and with the enumeration order improvements discussed in Section~\ref{sec-reorder}.
+without and with the order improvements discussed in Section~\ref{sec-reorder}.
 
@@
 \label{sec-1-bit-optima}
-Our first experiment is to determine the empirical complexity of each method with respect to problem size.
+
+\begin{figure}
+  \centering
+  \includegraphicsfit{boxplot-method}
+  \caption{Comparison of completion time variance for the largest size of each problem
+           where all three methods were successful at finding all 1-bit local optima.}
+  \label{fig-boxplot-method}
+\end{figure}
+
+%Our first experiment is to determine the empirical complexity of each method with respect to problem size.
 We tested 932 task configurations (problem, $N$, $r$), with 681 completed by at least one method.
@@
   \centering
-  \includegraphicsfit{boxplot-method}
-  \caption{Comparison of completion time variance for the largest size of each problem
-           where all three methods were successful at finding all 1-bit local optima.}
-  \label{fig-boxplot-method}
+  \includegraphicsfit{length-method}
+  \caption{Comparison of how each method scales with problem size when finding 1-bit local optima
+           on log-linear scales. Each point is the mean
+           runtime over 30 instances and each line the linear model. All confidence intervals too tight to see.}
+  \label{fig-length-method}
 \end{figure}
@@
 
-This reduction in growth complexity translates into a significant increase in the size of problems
-which can be enumerated. For instance, using the previous best enumeration methods~\cite{ochoa:2015:crossovernetworks},
-the largest Adjacent NKq problem size which can be enumerated in 4 hours is 34 bits. Given the same amount of time,
-Hyper-Reorder can enumerate problems with 77 bits. Using the regression model, we estimate solving
-a 77 bit problem using Gray-Box would require 900~million years. Similarly, Hyper-Reorder's largest
-Random NKq size of 69 bits would require Gray-Box 3~million years. We believe these larger problems
-are more likely to share characteristics with those where search algorithms are actually applied.
-
-\begin{figure}
-  \centering
-  \includegraphicsfit{length-method}
-  \caption{Comparison of how each method scales with problem size when finding 1-bit local optima
-           on log-linear scales. Each point is the mean
-           runtime over 30 instances and each line the linear model. All confidence intervals too tight to see.}
-  \label{fig-length-method}
-\end{figure}
-
 \begin{table}
@@
 \end{table}
+
+This reduction in growth complexity translates into a significant increase in the size of problems
+which can be enumerated. For instance, using the previous best enumeration methods~\cite{ochoa:2015:crossovernetworks},
+the largest Adjacent NKq problem size which can be enumerated in 4 hours is 34 bits. Given the same amount of time,
+Hyper-Reorder can enumerate problems with 77 bits. Using the regression model, we estimate solving
+a 77 bit problem using Gray-Box would require 900~million years. Similarly, Hyper-Reorder's largest
+Random NKq size of 69 bits would require Gray-Box 3~million years. We believe these larger problems
+are more likely to share characteristics with those where search algorithms are actually applied.
 
@@
 Perhaps most striking in Figure~\ref{fig-slope-method} is that on Adjacent NKq, Random NKq,
-and Ising Spin Glass, increasing $r$ \emph{reduces} Hyper-Reorder's growth complexity.
+and Ising Spin Glass, increasing $r$ \textbf{reduces} Hyper-Reorder's growth complexity.
 Increasing $r$ creates more moves which can be fitness improving. As a result there
@@
 
+\begin{figure}
+  \centering
+  \includegraphicsfit{length-radius}
+  \caption{Number of $r$-bit local optima for each size of each problem. Note that
+           the number of local optima in Concatenated Traps does not change while $r<k$.}
+  \label{fig-length-radius}
+\end{figure}
+
 Previous work~\cite{ochoa:2015:crossovernetworks} noted that, counter-intuitively, Adjacent
@@
 see large decreases in the number and growth rate of local optima. With $N=49$
-the number of 4-bit local optima Ising Spin Glass is on average under 70 thousand,
+the number of 4-bit local optima in an Ising Spin Glass is on average under 70 thousand,
 a reduction of four orders of magnitude.
-
-\begin{figure}
-  \centering
-  \includegraphicsfit{length-radius}
-  \caption{Number of $r$-bit local optima for each size of each problem. Note that
-           the number of local optima in Concatenated Traps does not change while $r<k$.}
-  \label{fig-length-radius}
-\end{figure}
 

### 2015-10-06 Hyperplane Enumeration/main.tex
Minor fixes at Bill's suggestion
Integrated Mike's changes

@@
 empirical evidence of optimality on NKq Landscapes and Ising Spin Glasses.
-We further refine this method to find only solutions
+We further refine this method to find solutions
 that cannot be improved by flipping $r$ or fewer bits, which counterintuitively
@@
 $2^{77}$ binary strings, and find all 4-bit local optima to problems with $2^{200}$ binary strings.
-We believe this increase in size will help further our understanding of complex real-world landscapes.
+%We believe this increase in size will help further our understanding of complex real-world landscapes.
 
@@
 quantifying the difficulty of a problem. Furthermore,
-this information can be used to design search algorithms that specifically deal
+this understanding can be used to design search algorithms that specifically deal
 with those difficulties. 
@@
 optima~\cite{tomassini:2008:nknetworks,verel:2011:nknetworks,ochoa:2015:crossovernetworks}.
-However, finding all of the local optima of a complex problem is prohibitively time consuming
+However, finding all of the local optima of a complex problem is prohibitively time-consuming
 even for small problems. Many studies have been limited to 18-bit
@@
 to 30-bit problems~\cite{ochoa:2015:crossovernetworks}. While sampling
-methods can approximate some of these statistics for much larger
+methods can approximate the number and distribution of local optima for much larger
 landscapes~\cite{iclanzan:2014:somnetworks}, they do not provide enough
@@
 An Mk Landscape~\cite{whitley:2015:mk} is any function
-who's value is equal to the sum of a set of subfunctions.
+whose value is equal to the sum of a set of subfunctions.
 Each subfunction uses a small subset of variables from the original function's input.
@@
 We therefore include it only because its structure allows for straight forward algorithm analysis.
-Here we set $k=5$.
+Here we set $k=5$ to ensure sufficient deceptiveness.
 
-NKq landscapes specify a class of randomly generated problem instances using 3 paramters:
-(1) the number of problem variables $N$ (2) the amount of variable epistasis $K$ where $k=K+1$
+NKq Landscapes specify a class of randomly generated problem instances using 3 paramters:
+(1) the number of problem variables $N$, (2) the amount of variable epistasis $K$ where $k=K+1$,
 and (3) the number of unique fitness values $q$. From these parameters a landscape is generated
 by creating $M=N$ subfunctions $f_i$, where $f_i$ uses variable $x_i$ and $K$ others to look up
-a fitness value in the range $[0..q-1]$ from a randomly generated table. This structure
-makes NKq landscapes a natural fit for Mk landscapes (sum of bounded subfunctions)
+a fitness value in the range $[0..q-1]$ from a randomly generated table.
+As NKq Landscapes are structured as a sum of bounded subfunctions, they are a natural fit
+for Mk Landscapes
 and are the most studied problem class for Gray-Box
 optimization~\cite{whitley:2012:constant,chicano:2014:ball,goldman:2015:GBO,tintos:2015:partitioncross,ochoa:2015:crossovernetworks,whitley:2015:mk}.
-Here we choose $k=3$ with $q=2^{K+1}=2^{k}=8$.
+Here we choose $k=3$ with $q=2^{K+1}=2^{k}=8$ in line with previous research.
 
@@
 \end{equation}
-where $e_{ij}$ is the weight of the edge connecting vertex $i$ to vertex $j$. In Mk landscape terms
+where $e_{ij}$ is the weight of the edge connecting vertex $i$ to vertex $j$. In Mk Landscape terms
 this type of spin glass has $M=2N$ and $k=2$.
@@
 This version of the canonical NP-Complete boolean satisfiability problem is formulated
-as the maximization of $M=4.27N$ clauses, each containing exactly $k=3$ unique literals.
+as the maximization of $M=4.27N$ clauses, each containing exactly $k=3$ unique literals~\cite{selman:1996:maxsat}.
 A clause is satisfied if any of its literals match how a solution's variables are set.
@@
 
-This result was extended, with no increase in asymptotic costs,
+This result has been extended, with no increase in asymptotic costs,
 to include search for
@@
 By definition the fitness effect of flipping both $x_i$ and $x_j$ is equal to the sum
-of flipping each independently. As a result, if neither individual flip is fitness improving,
-flipping both together cannot be fitness improving. By examining the graph of non-linear
+of flipping each independently. As a result, if neither individual flip is fitness-improving,
+flipping both together cannot be fitness-improving. By examining the graph of non-linear
 interactions between variables, this principle can be extended to any collection of $r$
@@
 we know that the quality of $m_i$ only depends on variables $C$, $E$, $F$, and $H$.
-Therefore, until one of those four variables are modified, the solution cannot be a local optimum.
+Therefore, until one of those four variables is modified, the solution cannot be a local optimum.
 As a result, the hyperplane **0*00*0 cannot contain a local optima and can
@@
     \Require $solution \leftarrow \{0\}^N$
-    \Require $move\_bin$ (How many improving moves have each minimum dependency)
+    \Require $move\_bin$ (Count of improving moves at each minimum dependency)
     \Require \textproc{MakeMove} (Function that flips $index$ in $solution$ and updates $move\_bin$)
@@
 all bins could contain a fitness-improving move, so $index$ starts at $N-1$.
-If at any point all bins are empty then the solution is added to the list of
+If at any point all bins are empty, then the solution is added to the list of
 local optima $found$. Algorithm~\ref{alg-enumerate} then
-adds a 1 to $index$ using the loop on Line~\ref{alg-enumerate-counter}
+adds a 1 to $index$, using the loop on Line~\ref{alg-enumerate-counter}
 to perform carry operations and Line~\ref{alg-enumerate-insert} to create the new 1 value.
@@
 simplification is possible because
-the previous iterations have verified that all moves in higher $index$ bins are not fitness improving, and no action performed during
-that iteration can make them fitness improving.
-Furthermore, $index$ is always the location of the least significant 1 bit in $solution$, meaning
+the previous iterations have verified that all moves in higher $index$ bins are not fitness-improving, and no action performed during
+that iteration can make them fitness-improving.
+Furthermore, $index$ is always the location of the lowest index 1 bit in $solution$, meaning
 iteration can continue immediately from the found $index$. This is true by construction.
-Initially $solution$ contains all 0s. Each time a 1 is inserted its position is equal to $index$ and
+Initially $solution$ contains all 0s. Each time a 1 is inserted, its position is equal to $index$ and
 $index$ is only increased by carry operations which reset a position to 0 before increasing $index$.
@@
 \label{sec-reorder}
-As a final efficiency, the order in which variables in $solution$ are $index$ed can be reordered.
+As a final efficiency gain, the order in which variables in $solution$ are indexed can be reordered.
 When a move is fitness-improving,
@@
 Figure~\ref{fig-enumerate} shows how changing the $index$ order of variables
-improves $m_i$'s lowest $index$ dependency from 2 to 4. Consider that before reordering $m_i$ allows the hyperplane
+improves $m_i$'s lowest $index$ dependency from two to four. Consider that before reordering $m_i$ allows the hyperplane
 **000000 to be skipped by Algorithm~\ref{alg-enumerate}, while after reordering the hyperplane that can be skipped is ****0000.
-Now, whenever $m_i$ is a fitness improvement, 4 times as many solutions are skipped.
+Now, whenever $m_i$ is a fitness improvement, four times as many solutions are skipped.
 
@@
 least-unmapped dependencies has all of its remaining dependencies mapped to the
-most significant remaining positions. With proper bookkeeping, this algorithm requires
+highest remaining indices. With proper bookkeeping, this algorithm requires
 \BigO{N} time. We argue below that the optimal bit ordering for skipping solutions
@@
 $2^{index}$ solutions have been skipped. For practical purposes we will
-assume that all moves are equally likely to be fitness improving.
+assume that all moves are equally likely to be fitness-improving.
 
@@
 \begin{proof}
-Any ordering of variables which cannot be described by an ordering of moves
+Any ordering of variables that cannot be described by an ordering of moves
 must contain the following property. There must exist a position $j$
@@
 a minimum dependency of $i$ or higher where $i$ is also the minimum dependency of a move $m$.
-In this situation swapping the order of $x_i$ and $x_j$ will raise $m$'s
+In this situation, swapping the order of $x_i$ and $x_j$ will raise $m$'s
 minimum dependency by at least one, but it will not lower the minimum dependency of any
-move. Therefore any ordering which contains this property cannot be optimal.
+move. Therefore, any ordering which contains this property cannot be optimal.
 \end{proof}
 
-Moves can have their dependencies assigned in some permutation $O$ of all possible moves,
-such that $O_0$ is the first move to have its dependencies assigned.
-$D_O(i, m)$ is a function that returns how many unassigned dependencies move $m$ has after all
-moves in $O$ before $i$ have been assigned. A greedy solution to this problem is one such
-that $O_i$ is set to be the move that minimizes $D_O(i, m)$.
+Moves can have their dependencies assigned in some permutation $P$ of all possible moves,
+such that $P_0$ is the first move to have its dependencies assigned.
+$D_P(i, m)$ is a function that returns how many unassigned dependencies move $m$ has after all
+moves in $P$ before $i$ have been assigned. A greedy solution to this problem is one such
+that $P_i$ is set to be the move that minimizes $D_P(i, m)$.
 
@@
 \begin{proof}
-All non-greedy solutions must have some $i$ such that $D_O(i, O_i) > D_O(i, O_{i+1})$.
-This is because $D_O(i, m)$ can only decrease as $i$ increases, meaning that if some $m$
-has a lower value at $i$ than $O_i$, this property must hold for
-some index between $i$ and when $m$ appears in $O$.
+All non-greedy solutions must have some $i$ such that $D_P(i, P_i) > D_P(i, P_{i+1})$.
+This is because $D_P(i, m)$ can only decrease as $i$ increases, meaning that if some $m$
+has a lower value at $i$ than $P_i$, this property must hold for
+some index between $i$ and when $m$ appears in $P$.
 Finally, all non-greedy solutions must have some $\hat{i}$
-that is the maximum $i$ value for which $D_O(i, O_i) > D_O(i, O_{i+1})$ is true.
+that is the maximum $i$ value for which $D_P(i, P_i) > D_P(i, P_{i+1})$ is true.
 
-Consider the optimal ordering $O^\star$ that is not greedy and that has the minimum value for $\hat{i}$.
-Swapping $O_{\hat{i}}^\star$ and $O_{\hat{i}+1}^\star$ cannot change the minimum dependency of any other moves.
-Performing the swap causes $O_{\hat{i}}^\star$ to have the minimum dependency $O_{\hat{i}+1}^\star$ had before
-the swap, while $O_{\hat{i}+1}^\star$'s minimum dependency is raised by at least
-$D_{O^\star}(\hat{i}, O_{\hat{i}}^\star) - D_{O^\star}(\hat{i}, O_{\hat{i}+1}^\star)$. This value
-cannot be negative meaning that we have now constructed a solution which either skips more
-solutions than $O^\star$ or has a lower $\hat{i}$ than $O^\star$, contradicting our assertions.
-Therefore there cannot exist an optimal ordering that is not greedy.
+Consider the optimal ordering $P^\star$ that is not greedy and that has the minimum value for $\hat{i}$.
+Swapping $P_{\hat{i}}^\star$ and $P_{\hat{i}+1}^\star$ cannot change the minimum dependency of any other moves.
+Performing the swap causes $P_{\hat{i}}^\star$ to have the minimum dependency $P_{\hat{i}+1}^\star$ had before
+the swap, while $P_{\hat{i}+1}^\star$'s minimum dependency is raised by at least
+$D_{P^\star}(\hat{i}, P_{\hat{i}}^\star) - D_{P^\star}(\hat{i}, P_{\hat{i}+1}^\star)$. This value
+cannot be negative, meaning that we have now constructed a solution which either skips more
+solutions than $P^\star$ or has a lower $\hat{i}$ than $P^\star$, contradicting our assertions.
+Therefore, there cannot exist an optimal ordering that is not greedy.
 \end{proof}
@@
 skips $2^3+2^1$ solutions, while the latter skips $2^3+2^2$. This is because $m_1$ and
-$m_2$ overlap. As a result, even though both a greedy, only the second is optimal.
+$m_2$ overlap. As a result, even though both are greedy, only the second is optimal.
 \end{proof}
 
-We believe finding the optimal ordering of moves is NP-Hard and therefore rely on
+We suspect finding the optimal ordering of moves is NP-Hard and therefore rely on
 the greedy method as it is probabilistically optimal.
@@
 Understanding the runtime complexity of Algorithm~\ref{alg-enumerate} requires
-knowledge of how often each move is fitness improving. For most landscapes
+knowledge of how often each move is fitness-improving. For most landscapes,
 this is intractable to do theoretically. However, for some restricted problem
-types its possible to rigorously determine Algorithm~\ref{alg-enumerate}'s complexity class.
+types it is possible to rigorously determine Algorithm~\ref{alg-enumerate}'s complexity class.
 
@@
 $x_i$ disagrees with the global optimum.
-Initially $index$ is set to $N-1$ and is decreased by at least 1 in any step after the first.
-Each step will flip a bit which disagreed with the global optimum.
-Before finding the global optimum no carry operations can occur, meaning the
+Initially $index$ is set to $N-1$, and is decreased by at least one in any step after the first.
+Each step will flip a bit which disagrees with the global optimum.
+No carry operations can occur before finding the global optimum, meaning the
 cost of each iteration is equal to the amount by which $index$ is decreased.
@@
 is then spent adding a 1 to $index$ 0, which in the worst case requires \BigO{N}
-carry operations. For all future iterations $index$ is the position of the highest fitness-improving move
+carry operations. For all future iterations, $index$ is the position of the highest fitness-improving move
 which is currently set to 1. In these iterations at least 1 carry operation must
-occur, and $index$ cannot be decreased. Iteration ends when $index$ exceeds $N$
+occur, and $index$ cannot be decreased. Iteration ends when $index$ exceeds $N$,
 which requires at most $N$ carry operations. Therefore, Algorithm~\ref{alg-enumerate}
@@
 all variables which appear in the same $s_i$ are consecutive.
-As such it is possible to create a numbering of $f_i$ such that if $i < j$ then all of $s_i$'s
+As such, it is possible to create a numbering of $f_i$ such that if $i < j$ then all of $s_i$'s
 variables appear before $s_j$'s in enumeration ordering.
 
-Algorithm~\ref{alg-enumerate} finds the $f_i$ with the highest $i$ which contains a
+Algorithm~\ref{alg-enumerate} finds the $f_i$ with the highest $i$ that contains a
 fitness-improving move and enumerates its variables until it no longer
@@
 Once the first local optimum is found, Algorithm~\ref{alg-enumerate} proceeds
-by finding all local optima in larger and larger hyperplanes. Initially $f_0$
+by finding all local optima in larger and larger hyperplanes. Initially, $f_0$
 is enumerated to find all local optima in the hyperplane where all $f_i, i>0$ remain
 fixed. This process ends when a carry operation modifies a bit of $f_1$. At
-this point $f_1$ is enumerated, such that each time $f_1$ contains no improving moves
+this point, $f_1$ is enumerated such that each time $f_1$ contains no improving moves
 $f_0$ is enumerated again. This finds all local optima in the hyperplane where $f_i, i>1$ remain
@@
 $T(M-1)\leq \sum_0^{M-1}2^kc^i<2^kc^{M}$.
-The number of local optima in the landscape is \BigO{c^{M}} meaning that Algorithm~\ref{alg-enumerate}
+The number of local optima in the landscape is \BigO{c^{M}}, meaning that Algorithm~\ref{alg-enumerate}
 is within a constant factor ($2^k$) of optimal.
@@
 Algorithm~\ref{alg-enumerate} continues to be within a constant of optimal.
-For $r < k$ the additional moves reduce $c$ without increasing the cost
+For $r < k$, the additional moves reduce $c$ without increasing the cost
 by more than a constant.
@@
 for fitness improvements, no subsets can be added that are larger than $k$.
-Therefore even if $r=N-1$, the number of subsets only grows at \BigO{N}.
-As $c=1$ in this case, the cost of finding the single global optimum
+Therefore, even if $r=N-1$, the number of subsets only grows at \BigO{N}.
+Because $c=1$ in this case, the cost of finding the single global optimum
 only requires the \BigO{2^kM} time to find the first local optimum, which is
@@
 This is because only carry operations can result in $index$ being set to the second
-variable in $s_i$. Each time $f_i$ contains a fitness improving move, $index$ is set
+variable in $s_i$. Each time $f_i$ contains a fitness-improving move, $index$ is set
 to the first variable in $s_i$, which comes before all variables in $s_{i-1}$.
-As a result the time to solve this ordering becomes $T(i) = 2^{k-1}*T(i-1)+2^k$.
+As a result, the time to solve this ordering becomes $T(i) = 2^{k-1}*T(i-1)+2^k$.
 Therefore, if $c < 2^{k-1}$ this ordering performs much worse than optimal.
@@
 We compare three methods for finding all local optima: Gray-Box, Hyper, and Hyper-Reorder.
-The Gray-Box method refers to previous work~\cite{ochoa:2015:crossovernetworks} which does
-not perform hyperplane elimination, with Hyper and Hyper-Reorder being hyperplane elimination
-without and with the order improvements discussed in Section~\ref{sec-reorder}.
+The Gray-Box method refers to previous work~\cite{ochoa:2015:crossovernetworks} that does
+not perform hyperplane elimination. Hyper is Algorithm~\ref{alg-enumerate} without the order
+improvements discussed in Section~\ref{sec-reorder}. Hyper-Reorder is Algorithm~\ref{alg-enumerate} with
+reordering.
 
@@
 Any run which failed was rerun once to prevent cluster issues from introducing
-noise. For the same reason any run which was over ten times slower than the next slowest run of the
+noise. For the same reason, any run which was over ten times slower than the next slowest run of the
 same configuration with a different seed was rerun. This resulted in 33 out of 36,000 runs being rerun.
@@
 we declared it unsuccessful for that combination of $r$ and $N$ of that problem. All results
-we report are for successful combinations. For timing purposes local optima were counted but not
+we report are for successful combinations. For timing purposes, local optima were counted but not
 recorded. All of our code, data, and statistical analysis can be downloaded from our
@@
 Of those, Hyper-Reorder had the lowest
-mean time to find all local optima in all but 2. In both of those cases Hyper-Reorder was within a tenth of a second
+mean time to find all local optima in all but 2. In both of those cases, Hyper-Reorder was within a tenth of a second
 of the best method.
@@
 column we show the growth rate of the number of local optima. It is impossible for any method
-which enumerates all local optima to grow slower than this rate.
+that enumerates all local optima to grow slower than this rate.
 However, Hyper-Reorder has an estimated slope better than optimal on Concatenated Traps, Adjacent
@@
 
-This reduction in growth complexity translates into a significant increase in the size of problems
+This reduction in growth complexity translates into a substantial increase in the size of problems
 which can be enumerated. For instance, using the previous best enumeration methods~\cite{ochoa:2015:crossovernetworks},
-the largest Adjacent NKq problem size which can be enumerated in 4 hours is 34 bits. Given the same amount of time,
+the largest Adjacent NKq problem size that can be enumerated in 4 hours is 34 bits. Given the same amount of time,
 Hyper-Reorder can enumerate problems with 77 bits. Using the regression model, we estimate solving
@@
 interactions. Concatenated Traps, Adjacent NKq, and Ising Spin Glass all have exactly 4
-non-linear relationships per variable. For small problem sizes
+non-linear relationships per variable. For small problem sizes,
 the number of non-linear relationships per variable in
@@
 size. Random NKq's average ranges between 5 and 6 on problems we tested.
-With $N=15$, MAX-kSAT has on average nearly
-12 non-linear relationships per variable. By $N=30$ that grows to just over 17. Having each
-variable depend on over half the genome significantly limits the size of hyperplanes which
+With $N=15$, MAX-kSAT has, on average, nearly
+12 non-linear relationships per variable. By $N=30$, this grows to just over 17. Having each
+variable depend on over half the genome significantly limits the size of hyperplanes that
 can be eliminated. While the number of non-linear relationships per variable
@@
 Adjacent NKq, Ising Spin Glass, and to a lesser extent Random NKq, none of our enumeration
-methods see an increase in growth complexity with increasing $r$. In all cases using Hyper-Reorder
+methods have an increase in growth complexity with increasing $r$. In all cases, using Hyper-Reorder
 has a lower growth rate than the other two methods using the same $r$. On every problem
@@
 and Ising Spin Glass, increasing $r$ \textbf{reduces} Hyper-Reorder's growth complexity.
-Increasing $r$ creates more moves which can be fitness improving. As a result there
-are more hyperplanes which can potentially be detected and eliminated.
-On Concatenated Traps, no flip using $1 < r < k$ bits can be fitness improving
-without a 1-bit flip being fitness improving, meaning no additional hyperplanes can be skipped.
+Increasing $r$ creates more moves which can be fitness-improving. As a result there
+are more hyperplanes that can potentially be detected and eliminated.
+On Concatenated Traps, no flip using $1 < r < k$ bits can be fitness-improving
+without a 1-bit flip being fitness-improving, meaning no additional hyperplanes can be skipped.
 For MAX-kSAT, increasing $r$ likely compounds the high dependency problems discussed in
@@
 are fewer local optima to enumerate, Algorithm~\ref{alg-enumerate} requires less
-time to enumerate them all. In Concatenated Traps each trap requires exactly $k$ flips in
+time to enumerate them all. In Concatenated Traps, each trap requires exactly $k$ flips in
 order to move between local optima. Therefore, while $r < k$ the number of local optima
-cannot change and once $r \ge k$ there is exactly 1 local optima.
+cannot change, and once $r \ge k$ there is exactly one local optima.
 
@@
 nearly 800 million local optima. For comparison, Adjacent NKq with $N=49$ averages fewer than 300 thousand,
-and with $N=77$ Adjacent NKq still only has fewer than 500 million.
-As with NKq, the polynomially solvable Ising Spin
-Glass problem had more 1-bit local optima than the NP-Hard MAX-kSAT. Unlike with NKq, this
+and with $N=77$ Adjacent NKq still has fewer than 500 million.
+As with NKq, the polynomially-solvable Ising Spin
+Glass problem has more 1-bit local optima than the NP-Hard MAX-kSAT. Unlike with NKq, this
 relationship generally does not change as $r$ increases. However, both do still
@@
 to the extreme, we tested Adjacent NKq with $N=200$ and $r=4$, with all 30 runs successful. On
-average these runs required 5.6 minutes to complete. However, there was significant variance
-in both the runtime and number of local optima. Eighteen runs completed in under 1 minute,
+average, these runs required 5.6 minutes to complete. However, there was substantial variance
+in both the runtime and number of local optima. Eighteen runs completed in under one minute,
 while the slowest run took 1.5 hours. Three had fewer than 50 4-bit local optima, while five
@@
 \section{Conclusions and Future Work}
-By expressing a function as an Mk Landscapes, we have developed a method to find all
+By expressing a function as an Mk Landscape, we have developed a method to find all
 local optima of that function very efficiently. The structure imposed by
 an Mk Landscape allows for the identification of hyperplanes in the search space
-which cannot contain any local optima and can therefore be eliminated from enumeration.
+that cannot contain any local optima and can therefore be eliminated from enumeration.
 Combined with a greedy method to
@@
 this method's runtime complexity scales with the number of local optima, unlike
-previous methods which scale with the number of possible solutions.
+previous methods that scale with the number of possible solutions.
 When applied to NKq Landscapes and Ising Spin Glasses, this allowed for a dramatic
@@
 
-We also showed how this hyperplane elimination can be refined to find only solutions
-which cannot be improved by flipping $r$ or fewer bits. For some problems
+We further refined this hyperplane elimination to find only solutions
+that cannot be improved by flipping $r$ or fewer bits. For some problems
 this counterintuitively reduces search times, and on Adjacent NKq allowed
@@
 in enumeration ordering. While we are able to prove that all optimal orderings
-are greedy, the reverse is not true due to how ties are broken. Therefore more
+are greedy, the reverse is not true due to how ties are broken. Therefore, more
 gains may be possible by improving this ordering. Furthermore, we assumed
-that all moves are equally likely to be fitness improving. It is likely possible
-to determine the true probability of a move being fitness improving, resulting
+that all moves are equally likely to be fitness-improving. It is likely possible
+to determine the true probability of a move being fitness-improving, resulting
 in potentially better ordering.

### 2015-10-20 CV/Brian Goldman CV.tex
Added new CV

\documentclass[a4paper,11pt]{article}
\usepackage{cv}
\usepackage{verbatim}
\name{Brian W. Goldman}
\info{Address: & Department of Computer Science \& Engineering, Michigan State University, East Lansing, Michigan.\\
      Phone: & 314-313-1281\\
      Email: & brianwgoldman@acm.org\\
      Website: & \url{http://brianwgoldman.github.io/}}

\bibliography{../main}

\addtocategory{journals}{goldman:2015:cgpanalysis,goldman:2015:fastp3,white:2013:bgpb,lehmann:2013:mimics}
%At MSU
\addtocategory{papers}{goldman:2014:p3,goldman:2015:GBO}
%CGP
\addtocategory{papers}{goldman:2013:ordering,goldman:2013:cgpwaste}
%With Tauritz
\addtocategory{papers}{goldman:2012:ltga,goldman:2011:scx,kamrath:2013:supportivescx,goldman:2012:supportive}

\addtocategory{posters}{goldman:2011:dynamic-parameters}

\addtocategory{curpapers}{goldman:2016:hyperplane,goldman:2016:p3hiff,whitley:2016:mkl}
%\ref{sumpapers} A list of these appears on pages \pageref{papersstart}--\pageref{papersend}.

\begin{document}
\maketitle

\section{Education}
\begin{itemize}
\item 2015 Doctor of Philosophy in Computer Science \& Engineering, Michigan State University
\begin{itemize}
  \item Dissertation Advisor:  Dr. William F. Punch, GPA: 4.0/4.0
  \item Dissertation: Out of the Box Optimization using the Parameter-less Population Pyramid
\end{itemize}
\item 2012 Master of Science in Computer Science, Missouri University of Science \& Technology
\begin{itemize}
  \item Thesis Advisor:  Dr. Daniel R. Tauritz, GPA: 4.0/4.0
  \item Thesis: Robust Evolutionary Algorithms
\end{itemize}
\item 2010 Bachelor of Science in Computer Science, Missouri University of Science \& Technology
\begin{itemize}
  \item Minor in Mathematics, GPA: 3.84/4.0 in major, 3.81/4.0 overall
\end{itemize}
\end{itemize}
\section{Academic Positions}
\begin{itemize}
\item Research Assistant Professor, Colorado State University. Starting Spring 2016.
\item Postdoc, Hintze Lab, Michigan State University. Fall 2015.
\item Advanced Computing Solutions Program, Los Alamos National Laboratories. Summer 2012.
\item Center for Cyber Defenders, Sandia National Laboratories. Summer 2011.
\end{itemize}

\section{Awards}
\begin{itemize}
\item     Nominated Best Paper - Genetic Algorithms Track for GECCO 2015 ``Gray Box Optimization using the Parameter-less Population Pyramid''
\item     Best Paper - Genetic Algorithms Track for GECCO 2014 ``Parameter-less Population Pyramid''
\item     Received travel and proposal funding through BEACON NSF STC 2013--2015.
\item     Voted 2012 Leader of the Year, Missouri S\&T Computer Science Department
\item     Google AI challenge, ranked 25$^{th}$ out of 4619 world wide, 6$^{th}$ in USA, 2010
\item     Missouri S\&T Human versus Computer chess tournament, two first place AIs, 2009 and 2011
\end{itemize}

\section{Activities}
\begin{itemize}
\item     Co-chair, 2016 Genetic Programing Theory and Practice Conference
\item     Co-chair, GECCO 2015 Combinatorial Black Box Optimization Competition
\item     Peer Reviewer, IEEE Transactions on Evolutionary Computation Journal 2014--present
\item     Peer Reviewer, Genetic Programming and Evolvable Machines 2014--present
\item     Founder and organizer, Computational Evolution Discussion Group October 2013--2015
\item     Program Committee Member, GECCO 2012--present Genetic Algorithms Track
\item     Chair of ACM SIG for AI competition game development (SIG-Game), 2011--2012
\end{itemize}

\section{Teaching Experience}
\begin{itemize}
\item Evolutionary Computation, Fall 2014 and Fall 2015. Guest Lecturer for graduate level course.
\item Introduction to Programming II, Summer 2014.
      Lead instructor for C++11 course. In charge of lecturing, creating course content, and assessments
      for 37 students. Managed two graduate assistants who taught lab sections and graded assignments.
      Received 3.81/4 student evaluation.
\item Introduction to Programming II, Fall 2013 -- Spring 2014. Primary instructor for lab section.
\item Object Oriented Numerical Methods, Spring 2012. Teaching assistant for upper level intensive C++ course.
\item Introduction to C++ Lab, Spring 2011 -- Fall 2011. Primary instructor for lab section.
\item Parallel Computing, Spring 2011. Teaching assistant for upper level course.
\item Evolutionary Computation, Fall 2010. Teaching assistant for upper level course.
\end{itemize}

\begin{publications}
%\printbib{books}
\printbib{journals}
\printbib{papers}
\printbib{posters}
\printbib{curpapers}
%\printbib{chapters}

%\printbib{techreports}
%\printbib{bookreviews}
%\printbib{editorials}
\end{publications}

\end{document}

### 2015-10-21 Hyperplane Enumeration/main.tex
Changed "MakeMove" to "FlipBit"
Minor tweaks

@@
 Unfortunately,
-exhaustive methods for finding local optima are limited to very small problem sizes.
+exhaustive algorithms for finding local optima are limited to very small problem sizes.
 We propose a method for exploiting problem structure
@@
 (1) the number of problem variables $N$, (2) the amount of variable epistasis $K$ where $k=K+1$,
-and (3) the number of unique fitness values $q$. From these parameters a landscape is generated
+and (3) the number of unique subfunction values $q$. From these parameters a landscape is generated
 by creating $M=N$ subfunctions $f_i$, where $f_i$ uses variable $x_i$ and $K$ others to look up
@@
 optimization~\cite{whitley:2012:constant,chicano:2014:ball,goldman:2015:GBO,tintos:2015:partitioncross,ochoa:2015:crossovernetworks,whitley:2015:mk}.
-Here we choose $k=3$ with $q=2^{K+1}=2^{k}=8$ in line with previous research.
+We set $k=3$ and $q=2^{K+1}=2^{k}=8$ in line with previous work.
 
@@
 set of $K$ unique variables that does not include $x_i$. While Random NKq is NP-Hard,
-the structure of Adjacent NKq allows for a polynomial time dynamic programming solution~\cite{wright:2000:solvingnk}.
-Adjacent NKq is therefore often used to approximate a search algorithm's ability to
-solve Random NKq.
+the structure of Adjacent NKq allows for a polynomial time solution~\cite{wright:2000:solvingnk}.
+Adjacent NKq is therefore often used to test a search algorithm's ability to
+find the known global optimum.
 
@@
 can vary and all other positions remain fixed. The Gray-Box domain makes it possible to skip hyperplanes
-that cannot contain local optima. In Figure~\ref{fig-enumerate}, move $m_i$ which flips variable $F$ is a fitness improvement
+that cannot contain local optima. In Figure~\ref{fig-enumerate}, move $m_i$, which flips variable $F$, is a fitness improvement
 when enumeration starts (all variables set to 0). Due to the known relationships between variables,
@@
     \Require $move\_bin$ (Count of improving moves at each minimum dependency)
-    \Require \textproc{MakeMove} (Function that flips $index$ in $solution$ and updates $move\_bin$)
+    \Require \textproc{FlipBit} (Function that flips $index$ in $solution$ and updates $move\_bin$)
     \Ensure $found$ (list of all local optima)
@@
       \While{$index < N$ \textbf{and} $solution[index] = 1$}\label{alg-enumerate-counter}
-        \State \Call{MakeMove}{$index$}
+        \State \Call{FlipBit}{$index$}
         \State $index \leftarrow index + 1$
@@
       \If{$index < N$}
-        \State \Call{MakeMove}{$index$}\label{alg-enumerate-insert}
+        \State \Call{FlipBit}{$index$}\label{alg-enumerate-insert}
       \EndIf
@@
 \label{sec-reorder}
-As a final efficiency gain, the order in which variables in $solution$ are indexed can be reordered.
+Changing the order in which variables are indexed can provide further efficiency gains.
 When a move is fitness-improving,
@@
 Any ordering of variables that cannot be described by an ordering of moves
-must contain the following property. There must exist a position $j$
-such that $x_j$ is not a dependency of any move with
-a minimum dependency of $i$ or higher where $i$ is also the minimum dependency of a move $m$.
+must contain positions $i$ and $j$ with the following properties: (1) $i<j$
+(2) $i$ is the minimum dependency of a move $m$ (3) no move with minimum dependency
+of $i$ or higher depends on $x_j$.
 In this situation, swapping the order of $x_i$ and $x_j$ will raise $m$'s
@@
 \end{equation}
-In MK Landscape terms, $M=N$ and $k=1$. The most well known linear function is OneMax,
+In Mk Landscape terms, $M=N$ and $k=1$. The most well known linear function is OneMax,
 in which $w_i=1$ for all $i$.
@@
 Algorithm~\ref{alg-enumerate} finds all local optima of any linear function
-in \BigO{N} time.
+where $\forall_i w_i \neq 0$ in \BigO{N} time.
 \end{theorem}
@@
 $x_i$ disagrees with the global optimum.
-Initially $index$ is set to $N-1$, and is decreased by at least one in any step after the first.
+Initially $index$ is set to $N-1$, and is decreased by at least one in every step after the first.
 Each step will flip a bit which disagrees with the global optimum.
@@
 with $f_j, j>i$ fixed as $T(i)$. $T(0)=2^k$ as all values of $f_0$ must be enumerated.
-$T(i) = |l_i|*T(i-1)+2^k$ where $|l_i|$ is the number of ways $f_i$ can be set such
+$T(i) = |l_i|*T(i-1)+2^k$ where $l_i$ is the list of ways $f_i$ can be set such
 that it contains no fitness-improving moves.

### 2015-10-27 CV/Brian Goldman CV.tex
CBBOC is continuing
Updated CV from writing group feedback

@@
 \begin{itemize}
-  \item Dissertation Advisor:  Dr. William F. Punch, GPA: 4.0/4.0
+  \item Dissertation Advisor:  Dr. William F. Punch%, GPA: 4.0/4.0
   \item Dissertation: Out of the Box Optimization using the Parameter-less Population Pyramid
@@
 \begin{itemize}
-  \item Thesis Advisor:  Dr. Daniel R. Tauritz, GPA: 4.0/4.0
+  \item Thesis Advisor:  Dr. Daniel R. Tauritz%, GPA: 4.0/4.0
   \item Thesis: Robust Evolutionary Algorithms
@@
 \item 2010 Bachelor of Science in Computer Science, Missouri University of Science \& Technology
-\begin{itemize}
-  \item Minor in Mathematics, GPA: 3.84/4.0 in major, 3.81/4.0 overall
-\end{itemize}
+%\begin{itemize}
+%  \item Minor in Mathematics, GPA: 3.84/4.0 in major, 3.81/4.0 overall
+%\end{itemize}
 \end{itemize}
@@
 
+\begin{publications}
+%\printbib{books}
+\printbib{journals}
+\printbib{papers}
+\printbib{posters}
+\printbib{curpapers}
+%\printbib{chapters}
+%\printbib{techreports}
+%\printbib{bookreviews}
+%\printbib{editorials}
+\end{publications}
+
+\section{Grants}
+\begin{itemize}
+\item BEACON NSF STC 2015 funded proposal \$105,148
+\item BEACON NSF STC 2013 funded proposal \$62,284.
+\end{itemize}
+
 \section{Awards}
 \begin{itemize}
-\item     Nominated Best Paper - Genetic Algorithms Track for GECCO 2015 ``Gray Box Optimization using the Parameter-less Population Pyramid''
+\item     Runner-up, Best Paper - Genetic Algorithms Track for GECCO 2015 ``Gray Box Optimization using the Parameter-less Population Pyramid''
 \item     Best Paper - Genetic Algorithms Track for GECCO 2014 ``Parameter-less Population Pyramid''
-\item     Received travel and proposal funding through BEACON NSF STC 2013--2015.
 \item     Voted 2012 Leader of the Year, Missouri S\&T Computer Science Department
@@
 
+\section{Invited talks}
+\begin{itemize}
+\item Departmental Seminar, University of Memphis, April 2015
+\item BEACON NSF STC Congress, Michigan State University, August 2014 and August 2015
+\item Evolutionary Computation, Michigan State University, September 2014 and October 2015
+\item Heuristic and Evolutionary Algorithms Laboratory, University of Applied Sciences Upper Austria, December 2014
+\item BEACON Day, North Carolina A\&T, October 2013
+\end{itemize}
+
 \section{Activities}
@@
 \item     Co-chair, 2016 Genetic Programing Theory and Practice Conference
-\item     Co-chair, GECCO 2015 Combinatorial Black Box Optimization Competition
+\item     Co-chair, GECCO 2015--present Combinatorial Black Box Optimization Competition
 \item     Peer Reviewer, IEEE Transactions on Evolutionary Computation Journal 2014--present
 \item     Peer Reviewer, Genetic Programming and Evolvable Machines 2014--present
+\item     Program Committee Member, GECCO 2012--present Genetic Algorithms Track
 \item     Founder and organizer, Computational Evolution Discussion Group October 2013--2015
-\item     Program Committee Member, GECCO 2012--present Genetic Algorithms Track
 \item     Chair of ACM SIG for AI competition game development (SIG-Game), 2011--2012
@@
 \begin{itemize}
-\item Evolutionary Computation, Fall 2014 and Fall 2015. Guest Lecturer for graduate level course.
 \item Introduction to Programming II, Summer 2014.
@@
 
-\begin{publications}
-%\printbib{books}
-\printbib{journals}
-\printbib{papers}
-\printbib{posters}
-\printbib{curpapers}
-%\printbib{chapters}
-
-%\printbib{techreports}
-%\printbib{bookreviews}
-%\printbib{editorials}
-\end{publications}
-
 \end{document}

### 2015-10-27 Hyperplane Enumeration/main.tex
Minor edits from Bill

@@
 can only have a non-linear relationship if they appear together in at least one $s_i$.
-Therefore by definition the total number of non-linear
+By definition the total number of non-linear
 relationships in an Mk Landscape is linear with $N$, meaning on average the amortized
@@
 This reduction in growth complexity translates into a substantial increase in the size of problems
-which can be enumerated. For instance, using the previous best enumeration methods~\cite{ochoa:2015:crossovernetworks},
+which can be enumerated in reasonable time. For instance, using the previous best enumeration methods~\cite{ochoa:2015:crossovernetworks},
 the largest Adjacent NKq problem size that can be enumerated in 4 hours is 34 bits. Given the same amount of time,

### 2015-11-10 ResearchStatement/main.tex
First commit of my Research Statement

\documentclass[a4paper, 11pt]{article}

\usepackage{fancyhdr}
\usepackage[margin=1in]{geometry}
\usepackage{titlesec}

\titleformat{\section}
{\normalfont\Large\bfseries}{}{0pt}{}
\titleformat{\subsection}
{\normalfont\itshape}{}{0pt}{}

%\linespread{2.5}
\begin{document}
\thispagestyle{empty}

\pagestyle{fancy}
\lhead{Brian W. Goldman}
\rhead{Research Statement}

\begin{center}
{\LARGE \bf RESEARCH STATEMENT}\\
\vspace*{0.1cm}
{\normalsize Brian W. Goldman (brianwgoldman@acm.org)}
\end{center}

\noindent
For many problems in many disciplines, it is easier to determine the quality of a solution
than to find the best possible solution. For example, this is true of NP-Hard problems,
finite element analysis, chemical design, and physics simulations.
While coming from very different fields, these problems share
three critical characteristics:
\begin{itemize}
\item There is a defined set of possible solutions.
\item We can measure the quality of a proposed solution.
\item There is no tractable method for finding the best solution.
\end{itemize}
I design algorithms for optimizing problems with these characteristics. Instead of
attempting to create specialized heuristics for every possible problem,
my goal is to discover and utilize general principles which can
then be applied to a large set of problems without further expert knowledge.

\section{Algorithm Design}
Many generic optimization algorithms require problem specific
configuration. These algorithms are therefore of limited generality
as expert knowledge is required to properly configure them.

The primary focus of my dissertation was the creation of the
Parameter-less Population Pyramid (P3), an optimization algorithm
that requires no problem specific configuration~\cite{goldman:2014:p3,goldman:2015:fastp3}.
P3 is comprised of three parts: Stochastic local search, entropy based model building, and
population layering. The guiding principle behind its design is to exploit all previous
knowledge the algorithm has about the problem before resorting to further stochastic exploration.
Each iteration, P3 creates a new high quality solution using local search and a model of
problem structure. The model is built by examining the mutual information between variables
in previously found high quality solutions~\cite{goldman:2012:ltga}.
By storing multiple populations of solutions, P3 is able to progressively guide search
to better and better areas without becoming stuck or requiring large initial random search.
P3 is able to find the global optimum in less time than other techniques, as well as
maintain a better ``current best'' solution as optimization progresses across a diverse set of problems.
This quality is on top
of the fact that the comparison algorithms used unrealistically ideal configurations while
P3 requires no configuration.

~\cite{goldman:2015:GBO}

~\cite{goldman:2016:hyperplane}

~\cite{whitley:2016:mkl}
TODO Exact polynomial time

TODO SAT compression

\section{Interdisciplinary Applications}
In order to apply these generic optimization methods, I have formed a number of
interdisciplinary collaborations. I first began a collaboration with a mechanical engineer
at NASA. He was in the process of designing a lightweight pallet for carrying sensitive
equipment to other planetary bodies. Using finite element analysis he could measure
the quality of a given pallet, but he had no direct method for discovering the ``best'' pallet.
P3 was able to find a solution which used less mass than the one found by the industry standard
tool, and did so using 40\% of the time.

TODO Statistical Physics somewhere

Working with a chemist, we set out to
design better OLED molecules, which underly modern electronics displays.
Given a molecule, its possible to calculate its phosphorescence when excited. However,
there is no direct method for finding molecules which maximize this property.
By applying P3 and other generic optimization methods we found some interesting
molecules for future analysis.

Determining the placement of turbines in a wind farm is another candidate for
generic optimization. Aerospace engineers have developed models which can calculate
how the wake of each turbine can effect the electrical output of the entire field.
While turbines are commonly arranged in grids, we found using optimization methods
that fields with less regular layouts can be significantly more efficient.

TODO Neuro-evolution

\section{Future Work}
TODO Chemistry (Proteins / Batteries)

TODO Multi-objectiveness

TODO Generalized search space compression

%\vspace{0.5cm}
%\newpage

\small

\bibliographystyle{abbrv}
\bibliography{../main}

\end{document}

### 2015-11-10 ResearchStatement/main.tex
First full draft

@@
 I design algorithms for optimizing problems with these characteristics. Instead of
-attempting to create specialized heuristics for every possible problem,
+attempting to create specialized heuristics for every problem,
 my goal is to discover and utilize general principles which can
@@
 
-~\cite{goldman:2015:GBO}
-
-~\cite{goldman:2016:hyperplane}
-
-~\cite{whitley:2016:mkl}
-TODO Exact polynomial time
-
-TODO SAT compression
+By making slightly more information about the problem available to optimization, I was
+able to develop an even more efficient version of P3~\cite{goldman:2015:GBO}. Consider
+that many problems can be expressed as the sum of functions which use only a subset
+of problem variables. For example, Boolean Satisfiability is often expressed as the sum
+of clauses. The Maximum Cut of a graph is equal to the sum of checking each edge.
+Using this additional information, I was able to reduce P3's runtime by two
+orders of magnitude. I was also able to construct an efficient method
+for finding all local optima in an arbitrary problem~\cite{goldman:2016:hyperplane},
+and with a collaborator create a generic method which solves many common
+problems in the optimization literature in $O(N)$ time~\cite{whitley:2016:mkl}.
 
@@
 
-TODO Statistical Physics somewhere
+%TODO Statistical Physics somewhere
 
@@
 
-TODO Neuro-evolution
+A challenging task in machine learning is the training of large scale neural networks.
+I have recently been in collaboration with an evolutionary biologist
+who studies natural learning to develop better methods for training artificial brains.
 
 \section{Future Work}
-TODO Chemistry (Proteins / Batteries)
+In recently submitted work~\cite{whitley:2016:mkl}, we determined that many real-world
+instances of common NP-Hard problems contained a surprising amount of regularity.
+I believe this regularity can be exploited to significantly reduce the space
+which needs to be searched for high quality solutions. To that end, I plan to develop
+a problem independent technique for search space compression.
 
-TODO Multi-objectiveness
+A current limitation of P3 is the focus on single objective optimization. For many
+problems, there are many competing objectives which should be optimized. For example,
+strength, weight, and cost in an engineering task. Optimization gains significant
+benefits from treating this problem as having three separate goals instead of attempting
+to combine them into a single objective. I believe that integrating P3 with the concepts
+of multi-objective optimization will result in a powerful tool for real-world optimization.
 
-TODO Generalized search space compression
+Finally, and most critically, I plan to continue exploring interdisciplinary collaborations.
+As a tool for problem solving, optimization algorithms are only as useful as the problems they
+can solve. Some specific targets for future collaborations include better battery design,
+structural engineering tasks, and land use management.
 

### 2015-11-10 TeachingStatement/main.tex
Previous version of my Teaching Statement

\documentclass[a4paper, 11pt]{article}

\usepackage{fancyhdr}
\usepackage[margin=1in]{geometry}
\usepackage{titlesec}

\titleformat{\section}
{\normalfont\Large\bfseries}{}{0pt}{}
\titleformat{\subsection}
{\normalfont\itshape}{}{0pt}{}

%\linespread{2.5}
\begin{document}
\thispagestyle{empty}

\pagestyle{fancy}
\lhead{Brian W. Goldman}
\rhead{Teaching Statement}

\begin{center}
{\LARGE \bf TEACHING STATEMENT}\\
\vspace*{0.1cm}
{\normalsize Brian W. Goldman (brianwgoldman@acm.org)}
\end{center}

\noindent
A critical skill in both teaching and research is the ability to convey
complex topics in meaningful terms to an uninformed audience. In both
contexts this requires an understanding of what a listener or reader
already knows, what they will have difficulty with, and how best to
connect the two. This is especially important when trying to overcome
the cultural boundaries which turn away underrepresented groups from
the field of computer science.

\section{Moderate Class Sizes}
Due to my previous success as a Lab Instructor, I was offered a
position as Lead Instructor for the second level undergraduate
programming course. This class is designed to introduce the students,
who learn Python in the first semester, to C++11. They learn
everything from basic syntax and STL tools to templates and how
to building their own data structures with dynamic memory.
I was given full autonomy to create course content, including
lectures, assignments, and assessments. To help with grading
assignments and lab work created by the 37 students, the
department assignment me two Lab Instructors.
I am proud to report that I received an average
of 3.81 out of 4 in post course student assessment.

In order to create an inclusive environment, I worked to
show my concern for each student’s understanding. While
lecturing I frequently paused and prompted for questions,
and did my best to gauge whether students were understanding
the material. This would often result in me changing the pace
of the lecture to meet the students’ comprehension, either
skipping slides or moving to the chalkboard to give further
examples. For those students who did not want to ask questions
in class, I utilized Piazza, an online tool where students
can post their questions anonymously. These questions could be
visible to everyone in the class, allowing students to answer
each other's’ questions, or only visible only to me.

Recent studies have shown that women are more likely to stay
in computer science when courses are application oriented
as opposed to abstract problem solving. As such, when creating
assignments I worked hard to give them concrete,
real world significance. When learning about the ``map'' data
structure and the random number generator, I had the students
estimate the centrality of websites using random walks.
When learning the basics of object oriented programming,
I had them create a stock market simulator using real data.
For the first assignment, in which the students know so little
syntax that it is hard to do anything meaningful, I had the
students calculate the height of US debt when stacking different denominations of currency.

In classes and labs I focused on collaborative experiential learning.
Each class, to break up the lecture, I would give students
relevant programming tasks to complete in small groups.
These worksheets would prompt them to write code to solve a
problem or to explain the behavior of a block of code. The solution
required them to use the material I had just covered in lecture as
well as topics from the previous days. By allowing them to collaborate,
students were able to help each other understand the topics better.
The lab assignments had a similar design, such that students worked
in pairs to solve problems requiring similar skills as upcoming
homework assignments. In this way students saw material in five
contexts: Reading, Lecture, Worksheet, Lab, and Assignment.

I have also had the opportunity to give a handful of lectures in graduate level courses.
These invited talks focused on my research and how it related
to the general optimization principles students were learning.
As before I found the key to successfully teaching these students
was understanding how best to connect the new information to
what they already knew. I covered my research in terms they
were familiar with, contrasting them with algorithms they had
previously learned. As a result I received very positive
feedback from both the course instructor and the students.

\section{Large Class sizes}
While my primary experiences have been with moderately sized
classes (25-50 students), I have taught a few lectures to much
larger classes (120+). Lectures of that size make it much more
difficult to interactively gauge student comprehension, making
alternative teaching tools all the more important. The small
group worksheet approach became even more critical as, unlike
the lecture, it is unaffected by class size. It also provided
me an opportunity to work with small groups of students to
answer specific questions or to clarify ideas, something impossible
during large scale lecture. As a teaching assistant in these
courses I saw the effectiveness of video tutorials and online
classroom management systems which allow students to work together to answer each other’s questions.

Effective large scale education requires new approaches,
and computer science has some especially promising new
technologies. In the future I plan to provide students
unit tests and source code analysis tools, allowing them
to receive immediate and automated feedback. This could
be integrated with version control and continuous build systems,
simultaneously allowing these techniques to be used with even
the most introductory students and teaching them important
skills necessary for success in industry. Once set up these
methods can scale to accommodate any class size and support
distance and asynchronous learning.

\section{Outreach}
During my masters I was the chair of an ACM SIG set up to organize
the MegaMinerAI programming competition each semester.
In this competition small teams of students are given
24 hours to write programs which act as the player in a computer game.
These programs compete with those written by other students,
with the results visualized. The competition focuses on making
it as easy as possible to write useful AIs, and making the game
as visually interpretable as possible. These goals help create
a community building challenge which is open to creativity and
translates student’s work into visual improvement.
The competition saw a significant growth in participation under
my direction, with nearly 100 competitors from a department of
only 400 students. For my work with the organization the department
awarded me “Leader of the Year.”

\section{Conclusions}
The best way to communicate information is on the receiver's
terms, regardless of if the context is teaching or research.
As such developing better methods for reaching both specific
individuals and broad audiences is essential to my career goals.
\end{document}

### 2015-11-10 TeachingStatement/main.tex
Tried to rework last paragraph

@@
 
-\section{Conclusions}
+\section{Future Goals}
+While much of my teaching experience has been with introductory
+programming courses, I am also interested in teaching higher level
+and theory based courses. Artificial intelligence and machine learning
+are common courses which coincide strongly with my research interests.
+Algorithms, discrete math, and graph theory are topics that I believe
+present exciting challgenges for keeping students engaged.
+From previous work I have also acquired significant knowledge of
+introductory and high performance C++ and would enjoy teaching classes
+on those topics.
+
 The best way to communicate information is on the receiver's
@@
 As such developing better methods for reaching both specific
-individuals and broad audiences is essential to my career goals.
+individuals and broad audiences is essential to my career goals,
+and is something I plan to continue improving.
 \end{document}

### 2015-11-11 ResearchStatement/main.tex
Minor edits from Bill

@@
 design better OLED molecules, which underly modern electronics displays.
-Given a molecule, its possible to calculate its phosphorescence when excited. However,
+Given a molecule, it is possible to calculate its phosphorescence when excited. However,
 there is no direct method for finding molecules which maximize this property.
@@
 I have recently been in collaboration with an evolutionary biologist
-who studies natural learning to develop better methods for training artificial brains.
+who studies natural learning to develop better methods for training neural networks.
 

### 2015-11-14 TeachingStatement/main.tex
Fixes from Charles' suggestions

@@
 \usepackage{titlesec}
+\usepackage{verbatim}
 
@@
 
+\usepackage{enumitem}
+\setlist[itemize,enumerate]{noitemsep,nolistsep}
 %\linespread{2.5}
@@
 
-\section{Moderate Class Sizes}
+\section{Quotations from Student Evaluations}
+\begin{itemize}
+\item ``Brian is a great TA. I consistently attended the labs because I was always learning and benefiting.''
+\item ``Brian was always very helpful and also found a way to make you figure out what you were doing on your own.''
+\item ``Often explained things that were confusing in an easy way to understand.''
+\end{itemize}
+
+\section{Teaching Classes of Moderate Size}
 Due to my previous success as a Lab Instructor, I was offered a
@@
 In order to create an inclusive environment, I worked to
-show my concern for each student’s understanding. While
+show my concern for each student's understanding. While
 lecturing I frequently paused and prompted for questions,
@@
 the material. This would often result in me changing the pace
-of the lecture to meet the students’ comprehension, either
+of the lecture to meet the students' comprehension, either
 skipping slides or moving to the chalkboard to give further
@@
 visible to everyone in the class, allowing students to answer
-each other's’ questions, or only visible only to me.
+each others' questions, or only visible only to me.
 
@@
 
-\section{Large Class sizes}
+\section{Teaching Classes of Large Size}
 While my primary experiences have been with moderately sized
@@
 courses I saw the effectiveness of video tutorials and online
-classroom management systems which allow students to work together to answer each other’s questions.
+classroom management systems which allow students to work together to answer each other's questions.
 
@@
 
+\begin{comment}
 \section{Outreach}
@@
 a community building challenge which is open to creativity and
-translates student’s work into visual improvement.
+translates student's work into visual improvement.
 The competition saw a significant growth in participation under
@@
 awarded me “Leader of the Year.”
+\end{comment}
 

### 2015-11-19 CV/Brian Goldman CV.tex
Fixed spacing
Minor space saving
Split 'peer reviewed' from 'other' publications
Some forgotten edits to my application materials

@@
 \usepackage{verbatim}
+\setlist[itemize,enumerate]{noitemsep,nolistsep}
 \name{Brian W. Goldman}
@@
 \end{itemize}
-\item 2012 Master of Science in Computer Science, Missouri University of Science \& Technology
+\item 2012 Master of Science in Computer Science, Missouri University of Science and Technology
 \begin{itemize}
@@
 \end{itemize}
-\item 2010 Bachelor of Science in Computer Science, Missouri University of Science \& Technology
+\item 2010 Bachelor of Science in Computer Science, Missouri University of Science and Technology
 %\begin{itemize}
@@
 
-\begin{publications}
+\begin{peer-publications}
 %\printbib{books}
@@
 \printbib{papers}
-\printbib{posters}
-\printbib{curpapers}
 %\printbib{chapters}
@@
 %\printbib{editorials}
-\end{publications}
+\end{peer-publications}
+
+\begin{other-publications}
+\printbib{posters}
+\printbib{curpapers}
+\end{other-publications}
 
@@
 \item BEACON NSF STC 2015 funded proposal \$105,148
-\item BEACON NSF STC 2013 funded proposal \$62,284.
+\item BEACON NSF STC 2014 funded proposal \$62,284.
 \end{itemize}
@@
 \begin{itemize}
-\item Departmental Seminar, University of Memphis, April 2015
+\item Job Talk, University of Memphis, April 2015
 \item BEACON NSF STC Congress, Michigan State University, August 2014 and August 2015
@@
 \item Introduction to Programming II, Summer 2014.
-      Lead instructor for C++11 course. In charge of lecturing, creating course content, and assessments
+      Lead instructor for a C++11 course. In charge of lecturing, creating course content, and assessments
       for 37 students. Managed two graduate assistants who taught lab sections and graded assignments.
@@
 \item Introduction to Programming II, Fall 2013 -- Spring 2014. Primary instructor for lab section.
-\item Object Oriented Numerical Methods, Spring 2012. Teaching assistant for upper level intensive C++ course.
+\item Object Oriented Numerical Methods, Spring 2012. Teaching assistant for senior level course.
 \item Introduction to C++ Lab, Spring 2011 -- Fall 2011. Primary instructor for lab section.
-\item Parallel Computing, Spring 2011. Teaching assistant for upper level course.
-\item Evolutionary Computation, Fall 2010. Teaching assistant for upper level course.
+\item Parallel Computing, Spring 2011. Teaching assistant for senior level course.
+\item Evolutionary Computation, Fall 2010. Teaching assistant for senior level course.
 \end{itemize}

### 2015-11-19 ResearchStatement/main.tex
Fixes from Charles' suggestions
Some forgotten edits to my application materials

@@
 For many problems in many disciplines, it is easier to determine the quality of a solution
-than to find the best possible solution. For example, this is true of NP-Hard problems,
+than to find the best possible solution. For example, this is true of most NP-Hard problems,
 finite element analysis, chemical design, and physics simulations.
@@
 P3 is able to find the global optimum in less time than other techniques, as well as
-maintain a better ``current best'' solution as optimization progresses across a diverse set of problems.
+maintain a better ``current best'' solution as optimization progresses, across a diverse set of problems.
 This quality is on top
@@
 generic optimization. Aerospace engineers have developed models which can calculate
-how the wake of each turbine can effect the electrical output of the entire field.
+how the wake of each turbine effects the electrical output of the entire field.
 While turbines are commonly arranged in grids, we found using optimization methods
-that fields with less regular layouts can be significantly more efficient.
+that fields with less regular layouts can be substantially more efficient.
 
@@
 I have recently been in collaboration with an evolutionary biologist
-who studies natural learning to develop better methods for training neural networks.
+who studies natural learning to develop better optimization methods for training neural networks.
 
@@
 instances of common NP-Hard problems contained a surprising amount of regularity.
-I believe this regularity can be exploited to significantly reduce the space
+I believe this regularity can be exploited to reduce the space
 which needs to be searched for high quality solutions. To that end, I plan to develop
@@
 problems, there are many competing objectives which should be optimized. For example,
-strength, weight, and cost in an engineering task. Optimization gains significant
+strength, weight, and cost in an engineering task. Optimization gains substantial
 benefits from treating this problem as having three separate goals instead of attempting

### 2015-11-19 TeachingStatement/main.tex
Some forgotten edits to my application materials

@@
 
-\section{Teaching Classes of Moderate Size}
+\section{Teaching Medium-Sized Classes}
 Due to my previous success as a Lab Instructor, I was offered a
@@
 visible to everyone in the class, allowing students to answer
-each others' questions, or only visible only to me.
+each other's questions, or only visible only to me.
 
@@
 
-\section{Teaching Classes of Large Size}
+\section{Teaching Large Classes}
 While my primary experiences have been with moderately sized
@@
 Algorithms, discrete math, and graph theory are topics that I believe
-present exciting challgenges for keeping students engaged.
+present exciting challenges for keeping students engaged.
 From previous work I have also acquired significant knowledge of

### 2016-01-08 Hyperplane Enumeration/main.tex
Fixed >= problem
First pass of reviewer changes

@@
 %
-%\author{Brian W.~Goldman\and William F. Punch}
-\author{Anonymous Author\and Anonymous Author}
+\author{Brian W.~Goldman\and William F. Punch}
+%\author{Anonymous Author\and Anonymous Author}
 
@@
 the number of local optima instead of with the landscape size.
-We prove optimality for linear and $k$-bounded separable problems, and we provide
+We prove optimality for linear problems and the Concatenated Traps, and we provide
 empirical evidence of optimality on NKq Landscapes and Ising Spin Glasses.
@@
     \While{$index < N$}
-      \While{$index > 0$ \textbf{and} $move\_bin[index] = 0$}\label{alg-enumerate-bincheck}
+      \While{$index \geq 0$ \textbf{and} $move\_bin[index] = 0$}\label{alg-enumerate-bincheck}
         \State $index \leftarrow index-1$
@@
 necessary $r$-bit moves to $move\_bin$. As discussed in Section~\ref{sec-gray-box}, we know the number of these moves
-is \BigO{N} an can be kept updated in \BigO{1} time per flip. Therefore, for small $r$, the cost
+is \BigO{N} and can be kept updated in \BigO{1} time per flip. Therefore, for small $r$, the cost
 of finding $r$-bit local optima is no more than a constant slower than finding 1-bit local optima.
@@
 
+%TODO Reviewer 1: Explain how data structures allow this to be O(N).
 We perform this reordering in a greedy fashion, such that the move with the
 least-unmapped dependencies has all of its remaining dependencies mapped to the
-highest remaining indices. With proper bookkeeping, this algorithm requires
-\BigO{N} time. We argue below that the optimal bit ordering for skipping solutions
+highest remaining indices.
+%New
+Before iteration, each move is binned based on its number of dependencies, and mappings
+are created to convert a move to its bin location and a variable to the
+moves that depend on that variable. Each of these requires \BigO{N} time
+as you must consider for each move (\BigO{N}) all variables (\BigO{1})
+in all subfunctions (\BigO{1}) it overlaps.
+The algorithm then iteratively queries the move bins in ascending order
+(requiring at most $kM/N=$\BigO{1} steps) until the first non-empty bin
+is found. All dependencies of that move are assigned positions, and
+all moves that depend on that variable are updated in the move bin
+(requiring at most \BigO{1} amortized updates). Each iteration finishes
+a move, meaning after $M$ iterations the process is complete. Therefore
+reordering takes \BigO{N} time.
+%Old
+%With proper bookkeeping, this algorithm requires
+%\BigO{N} time.
+We argue below that the optimal bit ordering for skipping solutions
 is a greedy ordering.
@@
 
+%TODO Reviewer 2: O(N) isn't O(1).
 \begin{theorem}
@@
 with initialization and the time to find the global optimum, the total complexity is \BigO{N},
-which is optimal.
+%new
+which is the lowest possible bound for this problem.
 \end{proof}
@@
 
+%TODO Reviewer 2: State blocks are contiguous.
 \begin{theorem}
-Algorithm~\ref{alg-enumerate} finds all \BigO{c^{M}} local optima of any $k$-bound separable function
-in \BigO{2^kc^{M}} time.
+\label{theorem-k-bound-general}
+Algorithm~\ref{alg-enumerate} with reordering finds all $\prod_{i=0}^{M-1} |l_i|$ local optima
+in at most $2^k\sum_{j=0}^{M-1} \prod_{i=M-j}^{M-1} |l_i|$ + \BigO{M} time.
 \end{theorem}
@@
 performing reordering will ensure that
-all variables which appear in the same $s_i$ are consecutive.
+all variables that appear in the same $s_i$ are consecutive.
 As such, it is possible to create a numbering of $f_i$ such that if $i < j$ then all of $s_i$'s
@@
 fitness-improving move and enumerates its variables until it no longer
-contains an improving move. In total, enumerating $f_i$ requires $2^k$
+contains an improving move. In total, enumerating $f_i$ requires at most $2^k$
 steps. Each $f_i$ is considered sequentially until none contains an improving move,
@@
 fixed. We represent the time required to find all local optima in the hyperplane
-with $f_j, j>i$ fixed as $T(i)$. $T(0)=2^k$ as all values of $f_0$ must be enumerated.
-$T(i) = |l_i|*T(i-1)+2^k$ where $l_i$ is the list of ways $f_i$ can be set such
-that it contains no fitness-improving moves.
+with $f_j, j>i$ fixed as $T(i)$. $T(0)=2^k$ as all values of $f_0$ must be enumerated, and
+$T(i) = |l_i|*T(i-1)+2^k$.
 Each setting in $l_i$ exposes a new hyperplane
 that can contain local optima, which causes the recursive call to $T(i-1)$.
-As all $|s_i| \leq k$ and $k$ is a
-constant, $\forall_i |l_i| \leq c \leq 2^k$ meaning that $c$ is a constant.
+%As all $|s_i| \leq k$ and $k$ is a
+%constant, $\forall_i |l_i| \leq c \leq 2^k$ meaning that $c$ is a constant.
 The time required to find all local optima is therefore
-$T(M-1)\leq \sum_0^{M-1}2^kc^i<2^kc^{M}$.
-The number of local optima in the landscape is \BigO{c^{M}}, meaning that Algorithm~\ref{alg-enumerate}
-is within a constant factor ($2^k$) of optimal.
+$T(M-1) = 2^k\sum_{j=0}^{M-1} \prod_{i=M-j}^{M-1} |l_i|$.
 \end{proof}
@@
 optimal.
+
+\begin{theorem}
+If $\forall_i |l_i| = c$ and $c>1$ Algorithm~\ref{alg-enumerate} with reordering
+finds all $c^M$ local optima in at most $2^kc^M + $ \BigO{M} time.
+\end{theorem}
+
+\begin{proof}
+From Theorem~\ref{theorem-k-bound-general}, the number of local optima can
+be expressed as $\prod_{i=0}^{M-1} |l_i| = \prod_{i=0}^{M-1} c = c^M$. Similarly
+$2^k\sum_{j=0}^{M-1} \prod_{i=M-j}^{M-1} |l_i| = 2^k\sum_{j=0}^{M-1} c^{j-1} < 2^kc^M$.
+\end{proof}
 
@@
 
+%TODO Reviewer 1: State programming language and statistical tools
 Each algorithm was given a maximum of four hours to find all $r$-bit local optima.
@@
 website.\footnote{\url{https://name.of.site/removed/to/be/anonymous/}}
+%TODO Make this the actual web page
 

### 2016-01-08 Hyperplane Enumeration/main.tex
A few more

@@
 the number of local optima instead of with the landscape size.
-We prove optimality for linear problems and the Concatenated Traps, and we provide
+We prove optimality for linear problems and Concatenated Traps, and we provide
 empirical evidence of optimality on NKq Landscapes and Ising Spin Glasses.
@@
 $\forall_i |s_i| \leq k$ and $\forall_{i \neq j} s_i \cap s_j = \emptyset$.
+Let $l_i$ be the set of ways $f_i$ can be set
+such that it contains no fitness improving moves.
 
@@
 $f_0$ is enumerated again. This finds all local optima in the hyperplane where $f_i, i>1$ remain
-fixed. We represent the time required to find all local optima in the hyperplane
-with $f_j, j>i$ fixed as $T(i)$. $T(0)=2^k$ as all values of $f_0$ must be enumerated, and
+fixed. $T(i)$ represents the time required to find all local optima in the hyperplane
+with $f_j, j>i$ fixed. $T(0)=2^k$ as all values of $f_0$ must be enumerated, and
 $T(i) = |l_i|*T(i-1)+2^k$.
@@
 recorded. All of our code, data, and statistical analysis can be downloaded from our
-website.\footnote{\url{https://name.of.site/removed/to/be/anonymous/}}
+website.\footnote{\url{https://github.com/brianwgoldman/Enumerate-Local-Optima/releases}}
 %TODO Make this the actual web page

### 2016-01-08 Hyperplane Enumeration/main.tex
Finished my edit, back down to 16 pages

@@
 % unless you accept that it will be published
-%\institute{BEACON Center for the Study of Evolution in Action,\\
-%Michigan State University, U.S.A.\\
-%brianwgoldman@acm.org, punch@msu.edu}
-\institute{Department\\ Organization\\ Email}
+\institute{BEACON Center for the Study of Evolution in Action,\\
+Michigan State University, U.S.A.\\
+brianwgoldman@acm.org, punch@msu.edu}
+%\institute{Department\\ Organization\\ Email}
 
@@
 the number of local optima instead of with the landscape size.
-We prove optimality for linear problems and Concatenated Traps, and we provide
+We prove optimality for linear functions and Concatenated Traps, and we provide
 empirical evidence of optimality on NKq Landscapes and Ising Spin Glasses.
@@
 landscapes with at most $2^{34}$ binary strings, hyperplane elimination can enumerate the same problems with
-$2^{77}$ binary strings, and find all 4-bit local optima to problems with $2^{200}$ binary strings.
+$2^{77}$ binary strings, and find all 4-bit local optima of problems with $2^{200}$ binary strings.
 %We believe this increase in size will help further our understanding of complex real-world landscapes.
@@
 
-We use two common variants of NKq in our experiments which specify how the $K$ additional
+We use two common variants of NKq in our experiments that specify how the $K$ additional
 variables in each subfunction are chosen: Adjacent and Random. In Adjacent NKq, $f_i$ depends
@@
 the structure of Adjacent NKq allows for a polynomial time solution~\cite{wright:2000:solvingnk}.
-Adjacent NKq is therefore often used to test a search algorithm's ability to
-find the known global optimum.
 
@@
 by exploiting Gray-Box optimization methods, previous work~\cite{ochoa:2015:crossovernetworks}
-was able to find all $r$-bit local optima of Mk Landscapes in \BigO{2^N} for any small,
-fixed value of $r$.
+was able to find all $r$-bit local optima of Mk Landscapes in \BigO{2^N} for small $r$.
 
@@
   %\includegraphicsfit{Enumerate}
-  \caption{Example change of ordering. The gray variables represent all dependencies
-           for move $m_i$ which flips variable $F$. By reordering, $m_i$'s lowest $index$ dependency improves from 2 to 4.}
+  \caption{Example reordering. The gray variables are all dependencies
+           for move $m_i$ which flips variable $F$.
+           Reordering improves $m_i$'s lowest $index$ dependency from 2 to 4.
+           }
   \label{fig-enumerate}
@@
 the least significant bit. This ordering ensures that before changing $index$ $i$, all possible settings of $index$
-0 through $i-1$ have been tested. This corresponds to examining the hyperplane where the lowest $i-1$ positions
+0 through $i-1$ have been tested. This corresponds to examining the hyperplane where the lowest $i$ positions
 can vary and all other positions remain fixed. The Gray-Box domain makes it possible to skip hyperplanes
@@
 there exists a fitness-improving move, no local optima can exist until at least one
-dependency of that move is modified. Any hyperplane that has all of a fitness-improving
+dependency of that move is modified.
+%TODO Consider cutting this sentence
+Any hyperplane that has all of a fitness-improving
 move's dependencies fixed cannot contain any local optima.
@@
 can be modified by enumeration that can potentially change the fitness effect of making that move.
-In order to determine how much of the enumeration can be skipped, we must find
-the highest $index$ in $move\_bin$ that contains a fitness-improving move,
-as done by Line~\ref{alg-enumerate-bincheck}. If no move is fitness-improving,
-then a local optimum has been found.
-
 Algorithm~\ref{alg-enumerate} works by finding the highest $index$ in $move\_bin$
-that is not empty and then adding a 1 to that $index$ in $solution$. Initially
+that is not empty (Line~\ref{alg-enumerate-bincheck}) and then adding a 1 to that $index$ in $solution$. Initially
 all bins could contain a fitness-improving move, so $index$ starts at $N-1$.
@@
 %New
-Before iteration, each move is binned based on its number of dependencies, and mappings
+Before iteration, each move is binned based on its number of dependencies, and dictionaries
 are created to convert a move to its bin location and a variable to the
 moves that depend on that variable. Each of these requires \BigO{N} time
-as you must consider for each move (\BigO{N}) all variables (\BigO{1})
+as for each move (\BigO{N}) you must consider all variables (\BigO{N})
 in all subfunctions (\BigO{1}) it overlaps.
 The algorithm then iteratively queries the move bins in ascending order
-(requiring at most $kM/N=$\BigO{1} steps) until the first non-empty bin
+(requiring at most $kM/N=$\BigO{1} amortized steps) until the first non-empty bin
 is found. All dependencies of that move are assigned positions, and
@@
 (requiring at most \BigO{1} amortized updates). Each iteration finishes
-a move, meaning after $M$ iterations the process is complete. Therefore
-reordering takes \BigO{N} time.
+a move, meaning after \BigO{N} iterations the process is complete.
 %Old
 %With proper bookkeeping, this algorithm requires
-%\BigO{N} time.
-We argue below that the optimal bit ordering for skipping solutions
-is a greedy ordering.
+%\BigO{N} time. We argue below that the optimal bit ordering for skipping solutions
+%is a greedy ordering.
 
@@
 $D_{P^\star}(\hat{i}, P_{\hat{i}}^\star) - D_{P^\star}(\hat{i}, P_{\hat{i}+1}^\star)$. This value
-cannot be negative, meaning that we have now constructed a solution which either skips more
+cannot be negative, meaning that we have now constructed a solution that either skips more
 solutions than $P^\star$ or has a lower $\hat{i}$ than $P^\star$, contradicting our assertions.
-Therefore, there cannot exist an optimal ordering that is not greedy.
 \end{proof}
@@
 skips $2^3+2^1$ solutions, while the latter skips $2^3+2^2$. This is because $m_1$ and
-$m_2$ overlap. As a result, even though both are greedy, only the second is optimal.
+$m_2$ overlap. Even though both are greedy, only the second is optimal.
 \end{proof}
@@
 Initially $index$ is set to $N-1$, and is decreased by at least one in every step after the first.
-Each step will flip a bit which disagrees with the global optimum.
+Each step will flip a bit that disagrees with the global optimum.
 No carry operations can occur before finding the global optimum, meaning the
@@
 Algorithm~\ref{alg-enumerate} with reordering finds all $\prod_{i=0}^{M-1} |l_i|$ local optima
-in at most $2^k\sum_{j=0}^{M-1} \prod_{i=M-j}^{M-1} |l_i|$ + \BigO{M} time.
+in a $k$-bound problem
+in at most $2^k\sum_{j=0}^{M-1} \prod_{i=M-j}^{M-1} |l_i| + $ \BigO{M} time.
 \end{theorem}
@@
 When looking for only $r$-bit local optima of $k$-bound separable problems,
-Algorithm~\ref{alg-enumerate} continues to be within a constant of optimal.
-For $r < k$, the additional moves reduce $c$ without increasing the cost
+Algorithm~\ref{alg-enumerate}'s complexity only increases by a constant.
+For $r < k$, the additional moves may reduce $|l_i|$ without increasing the cost
 by more than a constant.
@@
 Therefore, even if $r=N-1$, the number of subsets only grows at \BigO{N}.
-Because $c=1$ in this case, the cost of finding the single global optimum
-only requires the \BigO{2^kM} time to find the first local optimum, which is
+Because all $|l_i|=1$ in this case, the cost of finding the single global optimum
+only requires the \BigO{M} time, which is
 optimal.
@@
 \begin{theorem}
-If $\forall_i |l_i| = c$ and $c>1$ Algorithm~\ref{alg-enumerate} with reordering
+If $\forall_i |l_i| = c$ and $c>1$ in a $k$-bound problem, Algorithm~\ref{alg-enumerate} with reordering
 finds all $c^M$ local optima in at most $2^kc^M + $ \BigO{M} time.
@@
 in the degenerate case of $k=1$, a $k$-bound separable problem
-is a linear function. As such, when $k=1$, $c=1$ and
-\BigO{2^kM}~=~\BigO{N}.
+is a linear function. As such, when $k=1$ and $c=1$,
+$2^kc^M + $ \BigO{M}~=~\BigO{N}.
 
@@
 As a result, the time to solve this ordering becomes $T(i) = 2^{k-1}*T(i-1)+2^k$.
-Therefore, if $c < 2^{k-1}$ this ordering performs much worse than optimal.
+Therefore, if $l_i < 2^{k-1}$ this ordering performs much worse than optimal.
 

### 2016-11-08 ResearchStatement/main.tex
Draft of new research statement

@@
 \usepackage{titlesec}
+\usepackage{url}
 
@@
 \noindent
-For many problems in many disciplines, it is easier to determine the quality of a solution
-than to find the best possible solution. For example, this is true of most NP-Hard problems,
-finite element analysis, chemical design, and physics simulations.
-While coming from very different fields, these problems share
-three critical characteristics:
-\begin{itemize}
-\item There is a defined set of possible solutions.
-\item We can measure the quality of a proposed solution.
-\item There is no tractable method for finding the best solution.
-\end{itemize}
-I design algorithms for optimizing problems with these characteristics. Instead of
-attempting to create specialized heuristics for every problem,
-my goal is to discover and utilize general principles which can
-then be applied to a large set of problems without further expert knowledge.
+Machine learning (ML) has made great advances in the last few years, and is gaining wide
+acceptance as a way to automate expertise at low cost. Yet most of ML's practical application has
+been by billion dollar companies to help them sell more products or reduce spending on labor.
+I believe machine learning has an untapped potential to help the poorest and most disadvantaged members
+of society by improving the effectiveness of underfunded social programs. However, effectively utilizing
+ML in these areas will require overcoming challenges not seen in most industrial uses: Sparsity of data,
+malicious manipulation of features, and a need to understand what unintentional biases exist in the model.
 
-\section{Algorithm Design}
-Many generic optimization algorithms require problem specific
-configuration. These algorithms are therefore of limited generality
-as expert knowledge is required to properly configure them.
+%NOTES:
+%Small grants to prevent homelessness -> reduce overhead http://science.sciencemag.org/content/353/6300/694
+%Social Workers overworked -> automatically filter cases
+%Automated Legal Help https://www.engadget.com/2016/06/29/ai-laywer-shoots-down-160000-parking-tickets/
+%Policing -> risk assessment http://fivethirtyeight.com/features/prison-reform-risk-assessment/
+
+\section{Previous Work}
+Many ML and Artificial intelligence (AI) algorithms require expert configuration
+to achieve high quality performance. This can make it very difficult for
+the benefits of these tools to spread beyond academics and well funded
+research departments.
 
@@
 Parameter-less Population Pyramid (P3), an optimization algorithm
-that requires no problem specific configuration~\cite{goldman:2014:p3,goldman:2015:fastp3}.
-P3 is comprised of three parts: Stochastic local search, entropy based model building, and
-population layering. The guiding principle behind its design is to exploit all previous
+that requires no problem specific configuration~\cite{goldman:2014:p3,goldman:2015:fastp3,goldman:2016:p3hiff}.
+The guiding principle behind its design is to exploit all previous
 knowledge the algorithm has about the problem before resorting to further stochastic exploration.
@@
 
-By making slightly more information about the problem available to optimization, I was
-able to develop an even more efficient version of P3~\cite{goldman:2015:GBO}. Consider
-that many problems can be expressed as the sum of functions which use only a subset
-of problem variables. For example, Boolean Satisfiability is often expressed as the sum
-of clauses. The Maximum Cut of a graph is equal to the sum of checking each edge.
-Using this additional information, I was able to reduce P3's runtime by two
-orders of magnitude. I was also able to construct an efficient method
-for finding all local optima in an arbitrary problem~\cite{goldman:2016:hyperplane},
-and with a collaborator create a generic method which solves many common
-problems in the optimization literature in $O(N)$ time~\cite{whitley:2016:mkl}.
+By exploiting additional information on a subset of problems, I was able to further
+improve P3's efficiency by two orders of magnitude~\cite{goldman:2015:GBO}.
+Working with a collaborator, we were able to develop a generic method using the same principles
+that solves many of common problems in the optimization literature in $O(N)$ time~\cite{whitley:2016:mkl}.
 
-\section{Interdisciplinary Applications}
-In order to apply these generic optimization methods, I have formed a number of
-interdisciplinary collaborations. I first began a collaboration with a mechanical engineer
-at NASA. He was in the process of designing a lightweight pallet for carrying sensitive
-equipment to other planetary bodies. Using finite element analysis he could measure
-the quality of a given pallet, but he had no direct method for discovering the ``best'' pallet.
-P3 was able to find a solution which used less mass than the one found by the industry standard
-tool, and did so using 40\% of the time.
+\section{Transitioning Focus}
+Throughout my Ph.D. I focused on abstract theoretical work. While enjoyable, I began to feel my time would be
+better spent focusing instead on how I could do more than just satisfy my intellectual curiosity.
+My postdoc at Colorado State University helped me make this transition, as the topic was treatments for
+Post Traumatic Stress Disorder (PTSD).
 
-%TODO Statistical Physics somewhere
+PTSD is the result of extreme stress causing the body to develop a new
+normal, in which the body's hormones are in a constant state of stress response. The illness is resistant
+to treatment as natural homeostatic processes reinforce this state. One way to create a lasting improvement
+is to disrupt the complex interaction of hormones and return them to a healthy state. To understand how to
+do this, our collaborators needed to understand the what steady states are possible. By modeling this problem as a search for local optima,
+I was able to apply my previous work for understanding optimization landscapes~\cite{goldman:2016:hyperplane}.
+Unlike naive enumeration, which would have required decades of computing, my technique was able to find
+all steady states in their most complete model in seconds.
 
-Working with a chemist, we set out to
-design better OLED molecules, which underly modern electronics displays.
-Given a molecule, it is possible to calculate its phosphorescence when excited. However,
-there is no direct method for finding molecules which maximize this property.
-By applying P3 and other generic optimization methods we found some interesting
-molecules for future analysis.
-
-Determining the placement of turbines in a wind farm is another candidate for
-generic optimization. Aerospace engineers have developed models which can calculate
-how the wake of each turbine effects the electrical output of the entire field.
-While turbines are commonly arranged in grids, we found using optimization methods
-that fields with less regular layouts can be substantially more efficient.
-
-A challenging task in machine learning is the training of large scale neural networks.
-I have recently been in collaboration with an evolutionary biologist
-who studies natural learning to develop better optimization methods for training neural networks.
+Recently my interests have centered on deep neural networks (DNNs). As a side project I
+investigated new ways of training the output layer of decapitated DNNs, and I am currently in the
+process of drafting a publication on my findings. At Google I recently completed the internal course
+for ML, and plan to bring a new ML feature into production next year.
 
 \section{Future Work}
-In recently submitted work~\cite{whitley:2016:mkl}, we determined that many real-world
-instances of common NP-Hard problems contained a surprising amount of regularity.
-I believe this regularity can be exploited to reduce the space
-which needs to be searched for high quality solutions. To that end, I plan to develop
-a problem independent technique for search space compression.
+A recent study in Science~\cite{evans:2016:homelessness} found that
+giving a one time payout of about \$1,000 to someone on the brink of homelessness
+can keep them out of shelters for two years. Furthermore, this saves an estimated
+\$20,000 in welfare, policing, and other expenses. Unfortunately the cost
+of ensuring only the needy receive benefits creates almost \$10,00 in overhead.
+I believe that the proper application of ML can help improve the effectiveness
+of this program by improving fund allocation and automatically filtering applications
+to reduce operation expenses, resulting in more money going to those who need it.
 
-A current limitation of P3 is the focus on single objective optimization. For many
-problems, there are many competing objectives which should be optimized. For example,
-strength, weight, and cost in an engineering task. Optimization gains substantial
-benefits from treating this problem as having three separate goals instead of attempting
-to combine them into a single objective. I believe that integrating P3 with the concepts
-of multi-objective optimization will result in a powerful tool for real-world optimization.
+Social workers are overworked and underpaid. When
+surveyed\footnote{\url{https://www.theguardian.com/society/2010/oct/06/overworked-social-workers-children-risk}},
+``95\% admitted that the increased pressure meant children's health and safety was at risk because social workers
+were having to make key decisions based on insufficient information.'' ML has repeatedly shown
+its ability to detect patterns and make decisions at or above human expert quality. With careful
+training, I believe ML could be used to help social workers make better decisions and rescue children
+from abusive homes.
 
-Finally, and most critically, I plan to continue exploring interdisciplinary collaborations.
-As a tool for problem solving, optimization algorithms are only as useful as the problems they
-can solve. Some specific targets for future collaborations include better battery design,
-structural engineering tasks, and land use management.
+Unlike traditional product applications, using ML for these kind of social action requires that
+we have deep understanding of how the model works. We have to know it is making
+decisions for the right reasons. This poses many challenging research questions
+as currently many ML systems are treated as black boxes. It will also be necessary to
+learn from comparatively small training sets, which will likely result in new advancements
+in how to repurpose existing models from other tasks. However, if we can rise to the challenge
+I believe ML can help those who need it most.
 

### 2016-11-12 ResearchStatement/main.tex
Updates to research statement after Darrell and Bill

@@
 \section{Previous Work}
-Many ML and Artificial intelligence (AI) algorithms require expert configuration
+Many ML and Artificial intelligence algorithms require expert configuration
 to achieve high quality performance. This can make it very difficult for
@@
 This quality is on top
-of the fact that the comparison algorithms used unrealistically ideal configurations while
+of the fact that previous algorithms used unrealistically ideal configurations while
 P3 requires no configuration.
@@
 Throughout my Ph.D. I focused on abstract theoretical work. While enjoyable, I began to feel my time would be
-better spent focusing instead on how I could do more than just satisfy my intellectual curiosity.
+better spent focusing on how I could do more than just satisfy my intellectual curiosity.
 My postdoc at Colorado State University helped me make this transition, as the topic was treatments for
@@
 is to disrupt the complex interaction of hormones and return them to a healthy state. To understand how to
-do this, our collaborators needed to understand the what steady states are possible. By modeling this problem as a search for local optima,
+do this, our collaborators needed to understand what steady states are possible. By modeling this problem as a search for local optima,
 I was able to apply my previous work for understanding optimization landscapes~\cite{goldman:2016:hyperplane}.
 Unlike naive enumeration, which would have required decades of computing, my technique was able to find
-all steady states in their most complete model in seconds.
+all steady states in our collaborators' most complete model in seconds.
 
@@
 process of drafting a publication on my findings. At Google I recently completed the internal course
-for ML, and plan to bring a new ML feature into production next year.
+for ML, and plan to bring a new ML service into production next year.
 
@@
 \$20,000 in welfare, policing, and other expenses. Unfortunately the cost
-of ensuring only the needy receive benefits creates almost \$10,00 in overhead.
+of ensuring only the needy receive benefits creates almost \$10,000 in overhead.
 I believe that the proper application of ML can help improve the effectiveness
@@
 
-Unlike traditional product applications, using ML for these kind of social action requires that
-we have deep understanding of how the model works. We have to know it is making
-decisions for the right reasons. This poses many challenging research questions
-as currently many ML systems are treated as black boxes. It will also be necessary to
-learn from comparatively small training sets, which will likely result in new advancements
+Unlike traditional product applications, using ML for social action
+requires that we know the model is making decisions for the right reasons.
+Much of my work in optimization has been in understanding how complex systems work~\cite{goldman:2015:cgpanalysis,goldman:2016:p3hiff},
+and ML will require such analysis as currently much of its effectiveness is poorly understood.
+Social action ML will also require learning
+from comparatively small training sets, which will likely result in new advancements
 in how to repurpose existing models from other tasks. However, if we can rise to the challenge

### 2016-11-17 CV/Brian Goldman CV.tex
Job at CSU is a postdoc for now
Updated CV to say that I've started the CSU Job
Updated CV

@@
 \name{Brian W. Goldman}
-\info{Address: & Department of Computer Science \& Engineering, Michigan State University, East Lansing, Michigan.\\
+\info{Address: & Google, Building 2600, Boulder, Colorado.\\
       Phone: & 314-313-1281\\
@@
 
-\addtocategory{journals}{goldman:2015:cgpanalysis,goldman:2015:fastp3,white:2013:bgpb,lehmann:2013:mimics}
+\addtocategory{journals}{whitley:2016:mkl,goldman:2015:cgpanalysis,goldman:2015:fastp3,white:2013:bgpb,lehmann:2013:mimics}
 %At MSU
-\addtocategory{papers}{goldman:2014:p3,goldman:2015:GBO}
+\addtocategory{papers}{goldman:2014:p3,goldman:2015:GBO,goldman:2016:hyperplane,goldman:2016:p3hiff}
 %CGP
@@
 
-\addtocategory{posters}{goldman:2011:dynamic-parameters}
+\addtocategory{posters}{goldman:2011:dynamic-parameters,chen:2016:sat}
 
-\addtocategory{curpapers}{goldman:2016:hyperplane,goldman:2016:p3hiff,whitley:2016:mkl}
+%\addtocategory{curpapers}{,}
 %\ref{sumpapers} A list of these appears on pages \pageref{papersstart}--\pageref{papersend}.
@@
 \end{itemize}
-\section{Academic Positions}
+\section{Employment}
 \begin{itemize}
-\item Research Assistant Professor, Colorado State University. Starting Spring 2016.
+\item Google Software Engineer. Fall 2016.
+\item Postdoc, Whitley Lab, Colorado State University. Spring 2016.
 \item Postdoc, Hintze Lab, Michigan State University. Fall 2015.
@@
 \begin{itemize}
+\item     Nominated, Best Paper - European Conference on Evolutionary Computation in Combinatorial Optimisation 2016 ``Hyperplane Elimination for Quickly Enumerating Local Optima''
 \item     Runner-up, Best Paper - Genetic Algorithms Track for GECCO 2015 ``Gray Box Optimization using the Parameter-less Population Pyramid''

### 2016-11-17 ResearchStatement/main.tex
Minor tweaks

@@
 Much of my work in optimization has been in understanding how complex systems work~\cite{goldman:2015:cgpanalysis,goldman:2016:p3hiff},
-and ML will require such analysis as currently much of its effectiveness is poorly understood.
+and ML will require such analysis to better understand its effectiveness.
 Social action ML will also require learning

### 2016-11-17 TeachingStatement/main.tex
Minor tweaks

@@
 everything from basic syntax and STL tools to templates and how
-to building their own data structures with dynamic memory.
+to build their own data structures with dynamic memory.
 I was given full autonomy to create course content, including
@@
 introductory and high performance C++ and would enjoy teaching classes
-on those topics.
+on those topics. The time I have spent working for Google has also given
+me new insight into the topics students will need in order to succeed
+in industry.
 

### 2016-11-20 ResearchStatement/main.tex
Rewrite of research statement

@@
 Machine learning (ML) has made great advances in the last few years, and is gaining wide
-acceptance as a way to automate expertise at low cost. Yet most of ML's practical application has
-been by billion dollar companies to help them sell more products or reduce spending on labor.
-I believe machine learning has an untapped potential to help the poorest and most disadvantaged members
-of society by improving the effectiveness of underfunded social programs. However, effectively utilizing
-ML in these areas will require overcoming challenges not seen in most industrial uses: Sparsity of data,
-malicious manipulation of features, and a need to understand what unintentional biases exist in the model.
+acceptance as a way to automate expertise at low cost. Yet most of ML's great successes
+have required enormous amounts of training data, and have produced models that
+defy human understanding. I believe that overcoming these limitations will lead to the
+next big expansion in ML use, enabling applications ranging from fighting poverty and
+protecting children from abuse to intrusion detection and mapping the human brain.
 
@@
 \section{Previous Work}
-Many ML and Artificial intelligence algorithms require expert configuration
-to achieve high quality performance. This can make it very difficult for
-the benefits of these tools to spread beyond academics and well funded
-research departments.
+The quality of an ML or Artificial Intelligence algorithm is often measured
+not by its theoretical rigor but by its empirical effectiveness. This leads
+to two problems: its often hard to know why one algorithm is better than
+another, and many design decisions are often left up to the user. My previous
+work has focused on overcoming these two issues with the goal of making
+more powerful algorithms that require less expert involvement to apply.
+
+Cartesian Genetic Programming (CGP) is a method for evolving digital circuits
+and other functions that has had continual use for almost two decades. Part of its
+success is that unlike other methods it tends to produce terse results. Through
+careful analysis, I was able to determine the reason for this desirable outcome
+was actually an unintended consequence of a seemingly unrelated design
+decision~\cite{goldman:2013:ordering,goldman:2015:cgpanalysis}. From this finding
+I was able to tweak that decision to improve performance, maintain
+terse solutions, and remove a parameter from user configuration.
 
@@
 that requires no problem specific configuration~\cite{goldman:2014:p3,goldman:2015:fastp3,goldman:2016:p3hiff}.
-The guiding principle behind its design is to exploit all previous
-knowledge the algorithm has about the problem before resorting to further stochastic exploration.
-Each iteration, P3 creates a new high quality solution using local search and a model of
-problem structure. The model is built by examining the mutual information between variables
-in previously found high quality solutions~\cite{goldman:2012:ltga}.
-By storing multiple populations of solutions, P3 is able to progressively guide search
-to better and better areas without becoming stuck or requiring large initial random search.
-P3 is able to find the global optimum in less time than other techniques, as well as
+Its design was inspired by my previous analysis of model building optimization algorithms~\cite{goldman:2012:ltga}.
+P3 combines model building, local search, and novel population structuring to minimize the amount
+of stochastic exploration.
+P3 is able to find the global optimum in less time than previous techniques, as well as
 maintain a better ``current best'' solution as optimization progresses, across a diverse set of problems.
 This quality is on top
-of the fact that previous algorithms used unrealistically ideal configurations while
-P3 requires no configuration.
-
+of the fact that P3 requires no configuration.
 By exploiting additional information on a subset of problems, I was able to further
@@
 Working with a collaborator, we were able to develop a generic method using the same principles
-that solves many of common problems in the optimization literature in $O(N)$ time~\cite{whitley:2016:mkl}.
+that solves many common problems in the optimization literature in $O(N)$ time~\cite{whitley:2016:mkl}.
 
 \section{Transitioning Focus}
-Throughout my Ph.D. I focused on abstract theoretical work. While enjoyable, I began to feel my time would be
-better spent focusing on how I could do more than just satisfy my intellectual curiosity.
-My postdoc at Colorado State University helped me make this transition, as the topic was treatments for
-Post Traumatic Stress Disorder (PTSD).
+During my postdoc at Colorado State University, my research interests move from abstract analysis
+and design of algorithms to how that pursuit could lead to practical societal benefits.
+The project was to investigate new methods for treating Post Traumatic Stress Disorder (PTSD).
 
@@
 to treatment as natural homeostatic processes reinforce this state. One way to create a lasting improvement
-is to disrupt the complex interaction of hormones and return them to a healthy state. To understand how to
-do this, our collaborators needed to understand what steady states are possible. By modeling this problem as a search for local optima,
-I was able to apply my previous work for understanding optimization landscapes~\cite{goldman:2016:hyperplane}.
-Unlike naive enumeration, which would have required decades of computing, my technique was able to find
-all steady states in our collaborators' most complete model in seconds.
+is to disrupt the complex interaction of hormones and return them to a healthy state. Our collaborators
+had developed a model for these interactions, but it was too complex for naive methods to determine what
+steady states were possible. In my previous work I developed a tool for analyzing optimization
+landscapes~\cite{goldman:2016:hyperplane}. By reformulating their model, my tool was able to complete
+the task in seconds.
 
-Recently my interests have centered on deep neural networks (DNNs). As a side project I
-investigated new ways of training the output layer of decapitated DNNs, and I am currently in the
+Recently my interests have centered on deep convolutional neural networks. As a side project I
+investigated new ways performing transfer learning, in which an existing highly trained network is
+repurposed to a new task with little training data. I am currently in the
 process of drafting a publication on my findings. At Google I recently completed the internal course
@@
 \section{Future Work}
+Any application that requires a large number of expert decisions be made at low
+cost can potentially benefit from ML. Unfortunately, our inability to understand
+why ML makes the decisions it does can prevent acceptance of its use. This is
+especially true when it comes to using ML to help the poorest and most disadvantaged
+members of our society.
+
 A recent study in Science~\cite{evans:2016:homelessness} found that
 giving a one time payout of about \$1,000 to someone on the brink of homelessness
-can keep them out of shelters for two years. Furthermore, this saves an estimated
-\$20,000 in welfare, policing, and other expenses. Unfortunately the cost
+can keep them out of shelters for two years and save the state \$20,000. Unfortunately the cost
 of ensuring only the needy receive benefits creates almost \$10,000 in overhead.
-I believe that the proper application of ML can help improve the effectiveness
-of this program by improving fund allocation and automatically filtering applications
-to reduce operation expenses, resulting in more money going to those who need it.
+I believe this is a case where ML could help reduce costs and allow more money to go to
+those who need it, as long as we can make certain the model is making decisions for the right reasons.
+Similarly, I believe ML can help overworked social workers who are
+``having to make key decisions based on insufficient
+information''\footnote{\url{https://www.theguardian.com/society/2010/oct/06/overworked-social-workers-children-risk}}
+about children's safety.
+Leveraging my previous experience in understanding how complex systems achieve high performance
+will be critical to me in making these applications a reality.
 
-Social workers are overworked and underpaid. When
-surveyed\footnote{\url{https://www.theguardian.com/society/2010/oct/06/overworked-social-workers-children-risk}},
-``95\% admitted that the increased pressure meant children's health and safety was at risk because social workers
-were having to make key decisions based on insufficient information.'' ML has repeatedly shown
-its ability to detect patterns and make decisions at or above human expert quality. With careful
-training, I believe ML could be used to help social workers make better decisions and rescue children
-from abusive homes.
-
-Unlike traditional product applications, using ML for social action
-requires that we know the model is making decisions for the right reasons.
-Much of my work in optimization has been in understanding how complex systems work~\cite{goldman:2015:cgpanalysis,goldman:2016:p3hiff},
-and ML will require such analysis to better understand its effectiveness.
-Social action ML will also require learning
-from comparatively small training sets, which will likely result in new advancements
-in how to repurpose existing models from other tasks. However, if we can rise to the challenge
-I believe ML can help those who need it most.
+Beyond the need to understand the model, these applications are going to require training
+with far smaller datasets than normal; it is expensive to gather data about giving money
+to the homeless, and dangerous to experiment on child protective services. This
+task is perfect driving new advancements in transfer learning. Breakthroughs
+in this area will also benefit other domains that have very limited training samples,
+such as intrusion detection from security. Finally, the combination of model understanding
+and small training sizes could reach such seemingly unrelated applications as neuroscience
+by allowing researchers to build models mapping brain activity from FMRIs (very expensive per sample)
+to behavior, and then understanding how that model is making its predictions.
+If we can rise to these challenges, I believe ML can greatly expand its reach
+and help those who need it most.
 

### 2016-11-22 ResearchStatement/main.tex
Back on just 2 pages
Language fixes

@@
 Cartesian Genetic Programming (CGP) is a method for evolving digital circuits
-and other functions that has had continual use for almost two decades. Part of its
-success is that unlike other methods it tends to produce terse results. Through
+and other functions that has had continual use for almost two decades. Part of CGP's
+success is that it tends to produce terse designs. Through
 careful analysis, I was able to determine the reason for this desirable outcome
@@
 improve P3's efficiency by two orders of magnitude~\cite{goldman:2015:GBO}.
-Working with a collaborator, we were able to develop a generic method using the same principles
-that solves many common problems in the optimization literature in $O(N)$ time~\cite{whitley:2016:mkl}.
+%Working with a collaborator, we were able to develop a generic method using the same principles
+%that solves many common problems in the optimization literature in $O(N)$ time~\cite{whitley:2016:mkl}.
 
@@
 \section{Future Work}
-Any application that requires a large number of expert decisions be made at low
+Any application that requires a large number of expert decisions to be made at low
 cost can potentially benefit from ML. Unfortunately, our inability to understand
-why ML makes the decisions it does can prevent acceptance of its use. This is
+why a learned model makes the decisions it does can prevent acceptance of its use. This is
 especially true when it comes to using ML to help the poorest and most disadvantaged
